### 2014-07-13 paper.tex
The first checkin

%------------------------------------------------------------%
\documentclass{llncs}
\pagestyle{plain}
\usepackage{longtable} 
%------------------------------------------------------------%
%- common stuff starts here....
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
%\usepackage{color}
%\usepackage{xcolor}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
%\usepackage{pst-all}
\usepackage{bcprules}
%\usepackage{subfigure}
%\usepackage{algorithmic}
%\usepackage{anysize}     
\input{fh_defs}
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}

\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}

\begin{document}
%  \title{Heap  Reference  Analysis  for Functional  Programs}
\title{Liveness-Based Garbage Collection}

% make a proper TOC despite llncs
\setcounter{tocdepth}{2}
\makeatletter
\renewcommand*\l@author[2]{}
\renewcommand*\l@title[2]{}
\makeatletter

  \author{Rahul Asati\inst{1} \and Amitabha Sanyal\inst{1} \and Amey Karkare\inst{2}
    \and   Alan Mycroft\inst{3}}
  \institute{IIT Bombay, Mumbai 400076, India\\
    \email{$\{$rahulasati,as$\}$@cse.iitb.ac.in}, \and 
             IIT Kanpur, Kanpur 208016, India\\
    \email{karkare@cse.iitk.ac.in}, \and 
             Computer Laboratory, University of Cambridge, CB3 0FD, UK\\
    \email{alan.mycroft@cl.cam.ac.uk}}
\maketitle

\begin{abstract}
Current garbage collectors  leave much heap-allocated data uncollected
because they preserve data {\em reachable} from a root set.
% However, it is only necessary to preserve {\em live} data---a subset of
% reachable data.
However, only  {\em live} data---a subset of  reachable data---need be
preserved.

Using   a    first-order   functional   language    we   formulate   a
context-sensitive liveness  analysis for structured data  and prove it
correct.   We  then use  a  0-CFA-like  conservative approximation  to
annotate  each  allocation  and  function-call program  point  with  a
finite-state  automaton---which   the  garbage-collector  inspects  to
curtail reachability  during marking. As  a result, fewer  objects are
marked  (albeit  with a  more  expensive  marker)  and then  preserved
(e.g.\ by a copy phase).
% and secondly because of the consequent
%reduced number of collections during a computation.

Experiments  confirm the  expected performance  benefits---increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections, a decrease  in the memory size required  to run programs,
and  reduced  overall  garbage  collection  time  for  a  majority  of
programs.
\end{abstract}

%------------------------------------------------------------%

%% \begin{figure}
%% \tableofcontents       %%%%%%% BEWARE: find the \setcounter{page}{2}
%%                        %%%%%%% and comment it out when deleting this TOC.
%% \vspace{5cm}

%% \section*{Page Budget}
%% CC'2014 and ESOP'2014 have a limit of 20 pages.
%% A draft page budget is:
%% \begin{itemize}
%% \item 1. abstract/introduction (3 pages)
%% {\color{red}Currently 2.5 pages.}
%% \item 2. core language syntax and semantics (2 pages).
%% {\color{red}Currently 2 pages.}
%% \item 3. liveness analysis and correctness (3 pages)
%% {\color{red}Currently 3.8 pages.}
%% \item 4. practical liveness computation (6 pages)
%% {\color{red}Currently 6.2 pages.}
%% \\ I think this is where should do all 3 of: the context-insensitive analysis
%% (one which has a summary demand for each function), then the
%% solving like differential equations, then using Mohri-Nederhof to
%% get a NFA for each (call/cons) program point.
%% \item 5. experimental framework and results (4 pages)
%% \item 6. (optional) monotonicity of GC (0.5 page)---our technique can never
%% cause more collections.
%% \item 7. conclusions/acknowledgements/bibliography (1.5 page)
%% \end{itemize}
%% \end{figure}

\section{Introduction}
\label{sec:intro}

Most modern  programming languages support dynamic  allocation of heap
data.  Static  analysis of heap data  is much harder  than analysis of
static   and   stack   data.    Garbage   collectors,   for   example,
conservatively  approximate  the liveness  of  heap  objects by  their
reachability  from a  set of  memory  locations called  the {\em  root
  set\/}.  Consequently, many objects  that are reachable but not live
remain  uncollected, causing  a  larger-than-necessary memory  demand.
This is confirmed  by empirical studies on Haskell~\cite{rojemo96lag},
Scheme~\cite{karkare06effectiveness}                                and
Java~\cite{shaham02estimating} programs.

Here we consider a first-order  pure functional language and propose a
liveness  analysis  which  annotates  various program  points  with  a
description  of variables and  fields whose  object references  may be
dereferenced  in the future.   The garbage  collector then  only marks
objects pointed by live references and leaves other, merely reachable,
objects  to  be  reclaimed.    (Although  not  strictly  necessary,  a
collector  would  normally {\em  nullify}  dead  variables and  fields
rather than leaving dangling  references.)  Since there are fewer live
objects   than   reachable   objects,   more  memory   is   reclaimed.
Additionally, since  the collector traverses a smaller  portion of the
heap, the time spent for each collection is also smaller.  The work is
presented in  the context of a  stop-the-world non-incremental garbage
collector (mark-and-sweep,  compacting or  copying) for which  we also
show a  monotonicity result: that  our technique can never  cause more
garbage  collections  to  occur   in  spite  of  changing  the  rather
unpredictable  execution  points   at  which  collections  occur.   We
anticipate that our technique  is applicable to more modern collectors
(generational,  concurrent, parallel),  but leave  such  extensions to
future work.

We first define a {\em fully context-sensitive} (in the sense that its
results  are unaffected  by function  inlining) liveness  analysis and
prove it  correct.  However, fully context-sensitive  methods often do
not scale,  and this analysis would  also require us  to determine, at
run-time,  the internal  liveness of  a  function body  at each  call.
Hence,   similarly   to   the   0-CFA   approach,   we   determine   a
context-independent {\em summary} of  liveness for each function which
safely   approximates   the   context-dependence   of   all   possible
calls~\cite{chatterjee99relevant,cherem07fast,lee05static}.       (Note
that  an intraprocedural context-insensitive  method which  assumes no
information  about function  callers would  be too  imprecise  for our
needs.)   In essence  our approach  sets up  interprocedural data-flow
equations for the liveness summaries  of functions and shows how these
can be solved symbolically  as context-free grammars (CFGs)\@.  We can
then determine  a CFG  for each program  point; these are  then safely
approximated with  finite-state automata  which are encoded  as tables
for each program point.  For  garbage collection purposes only {\em GC
  points}  (program points  representing  calls, to  $\CONS$  or to  a
user-function) need to be stored.
% which may call eventually call $\CONS$, need to be stored.

We  previously proposed  an intraprocedural  method for  heap liveness
analysis   for   a   Java-like   language~\cite{khedker07heap}   which
statically inserted  statements nullifying dead  references to improve
garbage collection; by  contrast nullification here occurs dynamically
(which can work better with  aliasing) when the garbage collector acts
on  liveness  annotations  to  avoid traversing  dead  references.   A
workshop    paper~\cite{karkare07liveness}    outlined    the    basic
0-CFA-style-summary  interprocedural  approach  to  functional-program
liveness  analysis.   The  current  paper adds  the  context-sensitive
analysis and better formalization along with experimental results.
%   to  substantiate our ideas.

\kern -3pt   %% AM: a very small tweak to get back to 2 pages for this
             %% section.
%% AM: rewrite the next paragraph to save 3-4 words and remove this \kern.

%------------------------------------------------------------%
\setcounter{page}{2}
\subsubsection{Motivating Example}
\label{sec:motiv}

Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
$\pi$ of an expression $e$  denotes a program point.  During execution
of  the program, it  represents the  instant of  time just  before the
evaluation  of  $e$.  
%%While we discuss liveness, it  will also be important to consider an
%%instance of time  just after the execution of  an expression, though
%%we shall not be required to mark it with a label.
% memory as  a (possibly  disconnected) directed
% acyclic   graph.\footnote{Since  the   language   under  consideration
%   (Section~\ref{sec:defs}) does not  have any imperative features, the
%   heap cannot  have cycles.}
% Here all data  is assumed to  be boxed, i.e.\
% stored in heap and accessed through references.
We  view the  heap  as a graph.  
Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
containing values.  Edges in the graph are {\em references} and
emanate from {\em variables} or fields.
Variable and field values may also be atomic values ($\NIL$, integers etc.)
While it is convenient to box these in diagrams, our presented
analysis treats them as non-heap values.

Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
The edges shown  by thick arrows are those  which are made live by the
program. In addition,  assuming that
the value of any reachable part of the program result may be explored
or printed, the edges marked by thick dashed arrows are also live.
A cell  is  marked and preserved  during  garbage collection,  only  if it  is
reachable from the  root set through a path of  live edges.  All other
cells can be  reclaimed.  Thus if  a garbage collection takes
% place   at    the   time   represented   by   the    heap   shown   in
at $\pi$ with the    heap   shown   in
Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
will be marked and preserved.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

\begin{figure}[t!]
%%%%{\color{Myblue}
\begin{picture}(100,130)(0,-90)
%  \begin{center}
    \begin{tabular}{cc}
\begin{boxedminipage}{.5\textwidth}
      {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
	       \listb
	  \UNL{1}      (\CONS\ (\CAR\  \lista)
%          \UNL{2}          (\CAR\  \lista)
	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
          \listb))))
          \UNL{0}
	  \UNL{0} \hspace*{-.49\TAL} (\LET\ z  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
	  \UNL{0} (\LET\ y  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
	  \UNL{1}   (\LET\ w $\leftarrow$ (\append\ y z)\ \IN
	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ w)))))
	\end{uprogram}
      }} 
\end{boxedminipage}
      &
      \raisebox{-25mm}{\scalebox{.75}{
	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
      \psset{unit=1mm}
      \psset{linewidth=.3mm}
      \begin{pspicture}(0,-5)(70,60)
	%\psframe(0,0)(73,60)
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	%      fillcolor=white]{$\times$}}}
        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
		  fillcolor=white]{$\times$}}}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
	\ncline[nodesepB=-.5]{*->}{c2}{d}
	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
        %here
        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}

%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
%	      fillcolor=white]{$\times$}}} %and here
	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
%	      fillcolor=white]{$\times$}}} %and here

	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      \end{pspicture}}} \\
      (a) Example program.&
      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
	(b) Memory graph at $\pi$. Thick edges denote \\
        {\white (b) }live links. Traversal stops at edges marked\\
        {\white (b) }$\times$  during garbage collection.
      \end{tabular}}
    \end{tabular}
\end{picture}

\kern -3ex

\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
%  \end{center}
%%%%}\color{Myblue} end
\end{figure}

\subsubsection{Organization of the paper}
Section~\ref{sec:defs} gives
the syntax and semantics of
the language used to illustrate our analysis along with basic concepts
and    notations.
% Section~\ref{sec:operational}   describes    the
% operational semantics  of the  language.
Liveness analysis is described in
Section~\ref{sec:liveness} followed by a sketch of a correctness
proof relative to a non-standard semantics.
Section~\ref{sec:computing} shows how to encode liveness as
finite-state automata.
Section~\ref{sec:experiments}   reports   experimental   results and
Section~\ref{sec:lgc-always-better} proves that a liveness based
collector can never do more garbage collections than a reachability
based collector. 
% and Section~\ref{sec:conclusion} concludes the paper.

\section{The target language---syntax and semantics}
\label{sec:defs}
We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
functions ($\CONS$, $\PRIM$ etc.).
The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
it has eager semantics and restricts programs to be in 
Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
where all actual parameters to functions are variables.
This restriction does not affect expressibility (and indeed we feel free
to ignore it in examples when inessential), but simplifies the analysis formulation.
Additionally, as in the three-address instruction form familiar from compiler texts,
it forces each temporary to be named and function calls to be serialized
(necessary to get an unambiguous definition of liveness).
We further require that each variable in a program is distinct, so that no scope shadowing
occurs---this simplifies proofs of soundness.
In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
or return ($\RETURN$).
The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.

The body of the program is
the  expression denoted  by  $e_\mainpgm$; for analysis purposes
it is convenient
to regard $e_\mainpgm$ as part of a function definition
$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
% As  mentioned earlier,
We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.

% \paragraph{Inlining.}
In spite of the ANF restrictions it is still possible
to inline non-recursive functions (a fact we use to prove the safety
of liveness analysis).  A user-function call
$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
to a function defined
(after renaming its formals and locals to be disjoint from existing variables)
by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
is replaced by a sequence of $\LET$'s of the form
$z_i\leftarrow (\ID\,\,y_i)$ followed by
the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.

\begin{figure}[t!]
\footnotesize
\begin{eqnarray*}
   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
    \hspace{7em} \mbox{\em --- program}\\
    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
    e) 
    \hspace{1.7em} \mbox{\em --- function definition} \\
e \in \mathit{Expr} & ::= &
\left\{\begin{array}{ll@{\hspace{4em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
       (\RETURN\,\, x) && \mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{Stmt} & ::= &
\left\{\begin{array}{lr@{\hspace{1em}}l}
       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
            & \mbox{\em --- function application} 
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

\subsubsection{Semantics}
We now give an operational
semantics for  our language.  Later, a refinement of
the  operational semantics,  which we  call minefield  semantics, will
serve to prove liveness analysis correct.
We give a small-step semantics because, unlike big-step semantics,
correctness for non-terminating programs does not need special treatment.
We start with the domains used by the semantics:
\[
\begin{array}{rlcl@{\hspace{2em}}l}
% \begin{eqnarray*}
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
\mbox{-- Environment} \\ 
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
  Val}) & \mbox{-- Heap}
% \end{eqnarray*}
\end{array}
\]
Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
A value is either a number, the empty list $\NIL$, or a location  $\ell$.
Our liveness analysis does not track numeric values, and thus is neutral
as to whether these are boxed or represented as immediates.
An environment is a finite mapping from variables to values, and a heap a finite mapping
from locations to pairs of values.
% Given  a program,  an  environment is  a
% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
% values in $\mathit{Val}$.    The heap is a mapping
% from  locations  to pairs of  values.
Finally, $\stk$ is a stack
(using $\bullet$ for push and $[\,]$ for empty stack)
of frames of unfinished function calls.  A
frame is a triple $(e,x,\rho)$ representing the call site
$
(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
$
being evaluated in environment $\rho$.
Frames can also be viewed as continuations, in this view
the ({\sc ord-return}) 
 rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.

The semantics 
of statements $s$ are given by the judgement form
$\rho, \heap, s \rightsquigarrow \heap', v$
and those for expressions $e$ by the form
$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
program terminates successfully with result value $\rho(x)$
on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$

Notation:  we write $\rho[x \mapsto v]$ for the environment which is
as $\rho$ but has value $v$ at $x$.
We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.

\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
a contained $s$).  Some of these we eliminate syntactically, e.g.\
ensuring all variables and functions are defined and are called with
the correct number of parameters.
Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
All but the first can be eliminated with a static type system but,
treating our program as dynamically typed, we regard all these
as stuck states.

% A program terminates successfully if it reaches the halt state
% in Figure~\ref{fig:lang-opsem-small}.
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

\begin{figure}[t!]  
% (Random text to ensure proper line-spacing (problem with infrule macro?).)

\footnotesize
\begin{minipage}{0.335\textwidth}
\infrule[ord-const]
        {}
        {\rho, \heap, \pp{\pi}{k}
          \rightsquigarrow \heap, k}
\end{minipage}\begin{minipage}{0.655\textwidth}
\infrule[ord-cons]
      {
\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
}
      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
        \ell}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \heap, v_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-cdr]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
          \rightsquigarrow \heap, v_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-id]
      {
}
      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
        \rightsquigarrow \heap, \rho(x)}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-prim]
      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
        \rightsquigarrow \heap, \rho(x) + \rho(y)}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-null-false]
\infrule[]
        {\rho(x)\not= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 0}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-null-true]
\infrule[ord-null]
        {\rho(x)= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 1}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-if-true]{
\infrule[]{
  \rho(x) \in \mathbb{N}\setminus\{0\}
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-if-false]{
\infrule[ord-if]{
  \rho(x) = 0
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-nonfn]{
%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
  \andalso
  \mbox{$s$ is not
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-fncall]{
  \mbox{$s$ is 
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
\andalso
\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
[\myvec{z} \mapsto \rho(\myvec{y})],
  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-return]{
}{
  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
  \pp{\pi}{(\RETURN\ x)}
  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
\end{minipage}

% \normalsize
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

  \caption{The small-step operational semantics}
  \label{fig:lang-opsem-small}
\figrule
\normalsize
\end{figure}

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point

\section{Liveness}\label{sec:liveness}

In classical liveness analysis a variable is either `live' (its value
may be used in future computation) or `dead' (definitely not used).
Semantically, a variable is dead at a given program point
if arbitrary changes to its value have no effect on the computation.
Later we will use $\bot$ to represent a value which `explodes' when
it is used in a computation; dead variables can safely have their
value replaced with $\bot$.
For heap-allocated data  we need a richer model of liveness
in that both variables and fields of $\CONS$ cells may be dead or live.
Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
liveness of the structure reachable from a variable is a set of
{\em access paths}
which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
grammar notation.
Thus the liveness of $x$ being
$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
only refer to the second and third members of $x$ considered as a list.
Semantically, access paths are prefix-closed, as accessing a field
requires accessing all the paths from the variable to the field,
and hence the above liveness is properly written 
$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
The classical notions of a scalar variable being live or dead correspond to
$\{\epsilon\}$ and $\{\}$.

The overall liveness (also written {\em liveness environment} for emphasis)
at a program point is conceptually a mapping from
variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
instead of the map
$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
Analogously to classical liveness, the liveness at program point $\pi$
in $\pi:e$ is the liveness just before executing $e$.

A complementary notion to liveness is {\em demand}.  The demand for
expression $e$ is again an access path---that subset of of $\{\acar,\acdr\}^\ast$ which the
context of $e$ may explore of $e$'s result.  So, for example given a
demand $\sigma$ and the expression $\pi: (\RETURN\ x)$, the liveness at
$\pi$ is exactly $x.\sigma$.
The classical analogy of this is in {\em strong liveness}, where
an assignment node $n: x:=y+z$ causes $y$ and $z$ to be live on entry
to $n$ if (and only if) $x$ is live at exit of $n$---the liveness
of $x$ at exit from $n$ becomes the demand on $y+z$.
Note that, for an operation like division which
may raise an exception, the assignment $n: x:=y/z$ makes $y$ and $z$ live
regardless of the liveness of $x$.

We use $\sigma$ to range over demands, $\alpha$ to range over access paths
and $\Lv$ to range over liveness environments. The notation
$\sigma_1\sigma_2$ denotes the set $\lbrace \alpha_1\alpha_2 \mid
\alpha_1 \in \sigma_1, \alpha_2 \in \sigma_2\rbrace$. Often we shall
abuse notation to juxtapose an edge label and a set of access paths:
$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
Finally, we use $\Lfonly$ to range over {\em demand transformers}.
These transform demands on a function call
into demands for its formal parameters:
given a user function $f$, defined by
$(\DEFINE\ (f\ x_1\ \ldots\ x_n)\ \ e_f)$ and called with demand $\sigma$,
then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.

Note that liveness refers to variables and fields, and not to $\CONS$ cells
(i.e.\ to edges in the memory graph, not to locations themselves).
Hence liveness of $\{x.\epsilon,x.\acar\}$ means that future computation
may refer to the value $\ell$ of variable $x$, and also to the $\CAR$ field
of location $\ell$.
In the absence of other pointers to heap location $\ell$, we are
certain that the $\CDR$ field of $\ell$ will not be referenced
and may hence be corrupted arbitrarily.
Note therefore, that while  $\ell$ cannot be garbage collected, any
location $\ell'$ stored in the $\CDR$ field of $\ell$ would be garbage
(again provided there are no other aliases to  $\ell$).

%% \comment{
%%   \begin{enumerate}
%% \item Set the framework for liveness. Basic notions: access
%%   paths and expressions.
%% \item An informal working definition of liveness. Minefield will
%%   provide the formal definition.
%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
%%   rules. One of the changes from  the earlier version is the lack of a
%%   post-liveness in  the judgements. This anticipates  the intended use
%%   of  the analysis  for garbage  collection and has  to be  carefully
%%   explained.
%%   \end{enumerate}
%% }

%% We now  set up  the framework for  liveness analysis.  Given  a memory
%% graph, a link in the graph  can be described by starting with a member
%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
%% \CDR\ dereferences to reach the link.  For notational convenience, the
%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
%% dereferences    is    represented   by    a    string   $\alpha    \in
%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
%% in the  memory graph that starts  at $x$ and  traverses adjacent links
%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
%% We call the string $x.\alpha$ an {\em access expression\/}, the string
%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
%% graph   by  the   access  expression   an  {\em   access   path}.   In
%% Figure~\ref{fig:mot-example},        the       access       expression
%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
%% containing the value  $4$.  Most often the memory  graph is clear from
%% the context,  and we shall use  access expressions to  refer to access
%% paths. An  access path  used to refer  to a  link in the  memory graph
%% refers    to     the    last     link    in    the     path.     Thus,
%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
%% set of access paths, then  $x.\sigma$ will represent the set of access
%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
%% \sigma\rbrace$.

%% \subsection{Liveness Analysis}
%% \label{sec:exp-liveness}
%% 
%% Consider the memory graph at a program point $\pi$ during an execution
%% of a program. A  link in the memory graph is said  to be {\em live} at
%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
%% execution  starting from  the  evaluation of  the expression  labelled
%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
%% be live at  $\pi$, if, during {\em any} execution  of the program, the
%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
%% In the first, the path  itself represents the sequence of dereferences
%% to  access  the last  edge,  as,  for  example, while  evaluating  the
%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
%%   implicitly  live} if it is live in the second sense.  Clearly, static
%% determination   of  implicit  liveness   requires  an   analysis  that
%% determines  potential sharing  during  any execution  of the  program.
%% However, the intended application  of our liveness analysis is garbage
%% collection  during which  we have  access  to the  entire heap.   This
%% obviates the need for a  separate sharing analysis and we restrict our
%% attention to explicit liveness only.
%% 
%% 
%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
%% %% \bcdr\ represent application of  the constructor \CONS.  The following
%% %% example motivates the need for these symbols.
%% 
%% %% \begin{example}\label{exmp:motiv-bar}
%% %%   Consider the expression at program point $\pi_1$ in
%% %%   \begin{center}
%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
%% %% 			    $\pi_3\!:${$\cdots$})}}
%% %%   \end{center}
%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
%% %%   \begin{itemize}
%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
%% %%     live at $\pi_1$.
%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
%% %%   \end{itemize}
%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
%% %% \end{example}
%% 
%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
%% %% resulting access patterns may not  be directly useful in that they may
%% %% not describe a  path in the memory graph.  We  define a {\em canonical
%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
%% %% to obtain a canonical access  pattern representing the same edge. This
%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
%% %% cannot  be  used for  nullification  during  garbage collection.   The
%% %% liveness of a such a path is irrelevant, and the path is not considered
%% %% any further.
%% 
%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
%% %% element in $\sigma_2$, i.e.
%% 
%% %% \begin{eqnarray*}
%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
%% %% \end{eqnarray*}

\subsection{Liveness Analysis}
First recall the classical formulation of liveness (as sets of simple variables)
on three-address instructions,
$
\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I)
     \setminus \mathit{def\/}(I)
     \cup   \mathit{ref\/}(I)
$,
and then note that {\em strong liveness} needs, when $I$ is the instruction $z:=x+y$,
that $\mathit{ref}(I)$ be refined to
$\{x,y\}$ if $z \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.

Our liveness analysis formulated in Figure~\ref{fig:live-judge}
is analogous.
% has three analogous aspects.
% \begin{itemize}
% \item
Firstly, the function $\mathit{ref}$, when given a statement $s$,
returns the incremental liveness {\em generated} by $s$.
Because we generalize {\em strong} liveness, $\mathit{ref}$ needs a second parameter,
specifying the demand $\sigma$ on the result of $s$,
to determine which access paths of its free variables are made live.
The cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
% The cases for
% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
argument (even if its result is never used).
This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
(note $\sigma$ may be $\{\}$).
In the opposite  sense, the  demand  of $\acar\sigma$  on
(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
that \CONS\ does not, by itself, dereference its arguments.
Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
$\mathit{ref}$
expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
Constants generate no liveness.

% \item
The function $\mathcal{L}$ now gives the (total) liveness of an expression $e$.
The cases $\RETURN$ and $\SIF$ are straightforward, but note the liveness $x.\epsilon$ generated
by the latter.
The case $(\LET\ z\leftarrow s\ \IN\ e')$
resembles a three-address instruction:
the liveness of $e$ is given by taking the liveness, $\Lv$, of $e'$,
killing any liveness of $z$ and adding any incremental liveness from $s$.
The main subtlety is how the liveness of $z$ in $\Lv$ is converted to a demand $\Lv(z)$ to be
placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.

% \item
%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
Finally, the judgement form $\mathit{Prog} \len \Lfonly$ is used to determine $\Lfonly$\@.
Analogously to classical liveness being computed as
a solution of dataflow equations, we require, via inference rule ({\sc live-define}),
$\Lfonly$ to satisfy the fixed-point property that:
 when we assume $\Lfonly$ to be the family of demand transformers for the program
 then the calculated liveness of each function body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees
 with the assumed  $\Lfonly$.
As usual, there are often multiple solutions to $\Lfonly$; all are safe
(see Section~\ref{sec:correctness})
but we prefer the least one as giving the least liveness subject to safety---and
hence greatest amount of garbage collected.
% \end{itemize}

We make three observations: firstly the rule ({\sc live-define}) has a least solution
as $\mathcal{L}(\cdot)$ is monotonic in $\sigma$; secondly that
({\sc live-define})
resembles the rule for type inference of mutually recursive function definitions,
and thirdly the asymmetry of demand and liveness (compared to post- and pre-liveness
classically) is due to the functional formulation here.

Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
(representing a fully context-sensitive analysis)
can be safely approximated, for each function, by a {\em procedure summary}
(unifying the contexts in the style of 0-CFA)\@.
The summary consists of a pair of a single demand and, for this demand,
the corresponding tuple of demands the function makes on its arguments.

%%% save two lines to bring up section 3.2.

% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
% enables us to do a context-dependent analysis of a call without having
% to enter the body of the called function.

%% We present  our liveness  analysis using judgement forms $\len$
%% for both expressions and programs.
%% These are given formally in Figure~\ref{fig:live-judge}, but
%% first we an intuitive understanding of the issues involved.
%% Consider the expression:
%% 
%% \begin{center}
%% \begin{uprogram}
%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
%%                         \px) \IN\  
%%            % \UNL{0}
%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
%%                         \pw ) \IN \ldots) 
%% \end{uprogram}
%% \end{center}
%% 
%% 
%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
%% \pw.\epsilon\}$.
%% Informally, this is so since \pw\ and \pz\ are used in
%% (\PRIM~\pz~\pw ).
%% 
%% \begin{enumerate}
%% \item The liveness  at $\pi$ is obtained by  removing the access paths
%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
%%     similar  to liveness  analysis of  imperative programs,  where the
%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
%%     liveness generated and  killed in $s$}. 
%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
%%   as a {\em  demand}.  The demand on an expression consists of  access
%%   patterns  representing all  possible future  uses of  its  value.  A
%%   demand  of $\epsilon$  means that  only the  reference resulting
%% %  from the evaluation  of
%%   from  $(\CAR\,\,\px)$ will be used.
%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
%% \end{enumerate}
%% 
%% In general,  an access path is  live before an  expression because the
%% link represented by it is used (i) while evaluating the expression, or
%% (ii) while using the result of the expression, or (iii) simply because
%% the link  is live at  a later program  point.  When the  expression in
%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
%% the access  path \pw.   Note that \px.\acar\  is live in  this example
%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
%%

% \paragraph{Judgements formally.}
% Let  $\Lv$  denote a  liveness  environment (a function from
% variables to access paths, abused using grammar notation) and
% $\sigma$ be a demand.
% The judgement form $e, \sigma \len \Lv$,
% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
% environment  before $e$  is $\Lv$ and
% is thus analogous to the classical notion of {\em transfer function}.

%    It is  interesting to  note that,
% unlike traditional liveness analysis,  our judgements do not require a
% liveness after the expression. This  is because of the coming together
% of  two reasons.   First, it  will be  clear while  discussing garbage
% collection,  that it is  enough for  our purposes  to assume  that the
% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
% restriction of our  programs to ANF ensures that  the post-liveness of
% each  expression  $e$  inside   the  function  body  continues  to  be
% $\emptyset$. Thus the post-liveness can be dropped altogether.

%==============================================================
\begin{figure}[t]  
% \footnotesize
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
          &=& \{x.\sigma\} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=& \{x.\epsilon,y.\epsilon\} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \\
\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \mathcal{L}(e_1,\sigma,\Lfonly) \cup
        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
        \{x.\epsilon\} \\
\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
        \Lv
           \setminus \{x.*\}
           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
\end{eqnarray*}

\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}

\normalsize
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%==============================================================

% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
% liveness generated  before the right-hand-side of  a \LET\ definition.
% Constants    do   not   generate    any   liveness.     The   operator
% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
% that \CONS\ does not, by itself, dereference its arguments.
% 
% 
% Unlike primitives, the mapping from the demand on a user-function call
% to the  demands on its arguments  has to be derived  from the function
% body itself---as a fixed point for a recursive function.
% The function  \Lfonly\ captures this mapping for
% all functions  defined in the  program.  For a  function $\mathit{f}$,
% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
% a    call     to    the     $i$th    argument.     The     rule
% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
% confirmed by the body of all the functions, then the guess is correct.
% {\color{blue}Note the similarity of this rule to type-inference
% for recursive program schemes.}

% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
% liveness generated before $s$.

\subsection{Minefield semantics and correctness}
\label{sec:minefield}
\label{sec:correctness}

This section gives a modified semantics
%(a reference semantics)
which checks liveness annotations
at run time, and `explodes' when these are found to be inconsistent with execution behavior,
but otherwise behaves as the standard semantics.
We show that such explosions never occur and hence run-time checks
can be elided.
%% We first assume some demand transformers $\Lfonly$;
%% our proof will rely on this being a solution (i.e.\ satisfies $\mathit{Prog} \len \Lfonly$)
%% of liveness analysis, but we do rely in this yet.
We first assume an arbitrary  demand transformers $\Lfonly$ which satisfies
$\mathit{Prog} \len \Lfonly$.
We then enrich the abstract machine state $\rho, \stk, \heap, e$ to
$\rho, \stk, \heap, e, \sigma, \Sigma$.
Here $\sigma$ is the demand to the currently active function,
thus the liveness $\Lv$ at $e$ is $\mathcal{L}(e,\sigma,\Lfonly)$,\footnote{
This a simple liveness propagation using $\mathcal{L}(\cdots)$
and $\mathit{ref}(\cdots)$ as $\Lfonly$ is assumed given.}
and $\Sigma$ is a stack of demands---one for each function frame
pushed in $\stk$.

Second, we augment $\mathit{Val}$ with a value $\bot$.
To model strong liveness $\bot$ may be copied freely,
including into a $\CONS$ cell,
% but reduces the state to $\bang$
but explodes
when used computationally (in a primitive operation other than a copy).
Additionally we define
$\mathit{GC}(\Lv,\Sigma): (\rho, \heap, \stk) \mapsto (\rho', \heap', \stk')$
which determines {\em live-reachability}\footnote{Reachability curtailed by liveness information.}
%%%%% \marginpar{Say more about live-reachability?}
using $\rho$ and the $\rho$'s in $\stk$ as the root set
and following links in $\heap$ {\em only as far as allowed by $\Lv$ and $\Sigma$}.
% The effect of $\mathit{GC}(\Lv,\Sigma)$ is to replace
Hence $\mathit{GC}(\Lv,\Sigma)$ replaces
live-unreachable values---in $\rho$, in $\heap$ and the $\rho$'s in \stk---with $\bot$.
For example, if $x.\epsilon \not\in \Lv$ then $\rho'(x) = \bot$.
%, regardless of $\rho(x)$.
Only $\mathit{GC}(\cdots)$ introduces $\bot$.

Third, we update the semantics in four ways: ({\em i\/}) we arrange that all
($\rightarrow$) transitions on expressions $e$ first use $\mathit{GC}(\cdots)$
to update the state and then continue as before;
and ({\em ii\/}) whenever the value $\bot$ is used computationally in a reduction
($\rightsquigarrow$), we enter a distinguished stuck state $\bang$.
For example, supposing $\rho(x)=\bot$
then
$\rho, \heap, s \rightsquigarrow \bang$ if $s$ is
$(\CAR\,\, x)$, $(\CDR\,\, x)$ or $(\PRIM\,\, x\ y)$,
but not if $s$ is $(\ID~x)$, $(\CONS~x~y)$ or $(f~x~y)$.
Finally ({\em iii\/}) we make a similar change to the ($\rightarrow$) reduction for ({\sc ord-if})
and ({\em iv\/}) augment the ({\sc ord-let}) rule for primitives to propagate
$\bang$ from ($\rightsquigarrow$) to ($\rightarrow$).
%%%%%% {\color{blue}The full minefield semantics is given as an appendix?}

% Suppose liveness for the current $e$ under reduction is $\Lv$; then:
% For every variable $x$ for which $x.\epsilon \not\in \Lv$ then replace
% $\rho(x)$ with $\bot$.
% Also, for every $\ell \in \mathit{Loc}$, replace the $\CAR$ field
% of $\heap(\ell)$ with $\bot$ when
% which there is no path $x.\alpha\acar$ in $\Lv$ such that following
% $\alpha$-links in $\heap$ from $\rho(x)$ reaches $\ell$.
% Correspondingly, replace the $\CDR$ field of $\ell$ if there
% is no path $x.\alpha\acdr$ in $\Lv$.
% \item
% 
% {
% \footnotesize
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-car]
%         {\rho(x) = \bot}
%         {\rho, \heap, {(\CAR\,\, x)}
%           \rightsquigarrow \bang}
% \end{minipage}
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-cdr]
%         {\rho(x) = \bot}
%         {\rho, \heap, {(\CDR\,\, x)}
%           \rightsquigarrow \bang}
% \end{minipage}\\[1ex]
% 
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-prim]
%       {\rho(x) = \bot \vee \rho(y) = \bot}
%       {\rho, \heap, {(\PRIM\,\, x\ y)}
%         \rightsquigarrow \bang}
% \end{minipage}
% \begin{minipage}{0.445\textwidth}
% \infrule[mine-null]
%         {\rho(x)= \bot
%         }{\rho, \heap, {(\NULLQ\,\, x)}
%           \rightsquigarrow \bang}
% \end{minipage}
% }
%

The resulting minefield semantics behaves identically
(identical heap, identical steps, including possible non-termination)
to the standard semantics, except for one possibility---the
minefield semantics going $\bang$ while the standard semantics
continues (either to a halt state, to a stuck state, or reduces forever).

 We now prove a result that relates
liveness analysis to the semantics.
\begin{proposition}
Given program $P$ with $P \len \Lfonly$, then in the minefield
 semantics $P \rightarrow \bang$ can never occur
 (cf.\ `well-typed programs do not go wrong').
\end{proposition}
\textit{Proof outline:}
Space does not permit a full proof, but we give the two main steps.
We proceed by contradiction and assume there is a program $P$ for which
$P \len \Lfonly$ can enter state $\bang$.
The first step is to construct a program $P'$ with identical behavior, but with
no user-function definitions, using inlining.
This is possible because a program which goes $\bang$ does so after a finite number
of reductions, and hence even recursive functions have only had a finite number of
invocations.
We hence repeatedly inline-expand user-function calls in $P$
until we obtain a program $P'$ which behaves identically\footnote{
Modulo replacement of
({\sc ord-call}) and
({\sc ord-return}) steps with
({\sc ord-id}) steps.
}
to $P$ in the standard semantics, but makes no user-function calls.
Any remaining, non-executed, calls can 
%be left or
replaced with a new primitive with the same
demand-to-liveness transfer function---thus making
$P'$ a simple expression $e'$.
%%%%%
Not only do the program points in reducing $e'$ correspond
one-one to states during evaluation of $P$,
but also the liveness associated with a point
in $e'$ is identical to the liveness at the corresponding state
$(\rho, \stk, \heap, e, \sigma, \Sigma)$ of $P$ (concatenating the liveness
$\mathcal{L}(e,\sigma,\Lfonly)$ at $e$
with the liveness, obtained from $\Sigma$, of the call sites in $\stk$ and
after renaming the variables correspondingly to the inlining which produced $e'$).
This assertion relies on the analysis being fully context-sensitive, and noting that
while the change of scope caused by inlining changes variable visibility
between $P$ and $e'$ it does not change the liveness---as local-to-a-function
$\LET$-variables whose scope has been prolonged due to inlining
are dead in the prolonged scope.

The second step of the proof is to show that $e'$ cannot go $\bang$.
We proceed by induction.
Correctness of the $\SIF$ and final $\RETURN$ forms are immediate;
the $(\LET~z\leftarrow s~\IN~e)$ form requires showing
the inference rules ensure that any value referenced via $z$ in $e$
was already live-reachable (via another variable and path) in any
enclosing expression (so that $\mathit{GC}(\cdots)$ could not re-write it to $\bot$).

% \newpage

\section{Computing liveness and its encoding as a table}\label{sec:computing}

Section~\ref{sec:liveness} gave a context-sensitive liveness analysis
and proved it correct with reference to a {\em minefield} semantics.
% Practical computation was not considered.
For practical use we need to solve the liveness equations {\em finitely and
symbolically}.  As expressed mathematically, and given a fixed program,
three things are potentially unbounded:
({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
({\em iii\/}) the number of such access paths.

We commonly first solve ({\em i\/}) by reducing the number of distinct
calling contexts
(e.g.\ to a single unified context in 0-CFA style).
However, it turns out that we can solve the equations for $\Lfonly$ symbolically
without this reduction, so here we defer this to Section~\ref{sec:bodylivenessbodies}.
(It also allows easier extension to  dynamically determined liveness---future work.)
%
We address ({\em ii\/}) and ({\em iii\/}) by 
% attempting to re-interpret
re-interpreting
the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
rather than a mutually recursive set of equations on sets of access paths.

This requires two ideas.  Firstly we have to control the use of functions---they
tend to be infinitary and do not occur naturally in CFGs; in particular our
$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
to a tuple of access paths.  The former is achieved by using a separate
meta-variable (later non-terminal)
$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
expressible finitely (it is a linear form).

Secondly, there are also two technical issues in re-interpreting
Figure~\ref{fig:live-judge} as a grammar.
One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
disappears after variables are separated ($\Lanv{i}{x}$ above). The other
is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
not a CFG\@.
Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.

Finally,  Section~\ref{sec:NFA-approx}
uses a construction due to
Mohri and Nederhof~\cite{mohri00regular} to over-approximate
context-free grammars with regular grammars;
these are more appropriate for run-time use.
%
Hence the overall `big picture' view is that each GC point
is annotated with a table encoding the DFA for that program point.
Such program points forming the call string causing the garbage collection
(and ending in $\CONS$)
are used by the garbage collector to curtail its reachability-marking phase
to this over-approximated liveness.

\subsubsection{Modifying the \CONS\ rule.}

The    $\mathit{ref}$    rule for \CONS,    shown    in
Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
\acar\  and  \acdr\  from  access paths in  $\sigma$.
  Mathematically  this  is fine  but causes problems  when
  solving  the liveness equations  {\em symbolically}  since such
  decomposition cannot be expressed as a context-free grammar.  
To  handle this,  we introduce  two  new symbols  \bcar\  and
\bcdr\ with the properties:
\[
 \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
\hspace{3em} \mbox{and} \hspace{3em}
 \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
\]
We can now rewrite the \CONS\ rule as:
\begin{eqnarray*}
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
\end{eqnarray*}

We  call the liveness equations with  this modification
  \Lfunonly$'$.
The  defintions of \bcar\  and  \bcdr\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{eqnarray*}
  \sigma_1\bcar\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
& \qquad\mbox{and}\qquad &
\sigma_1\bcdr\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
\end{eqnarray*}
%% \noindent \begin{minipage}{0.445\textwidth}
%% \begin{eqnarray*}
%%     \sigma_1\bcar\sigma_2 & \hookrightarrow &
%%     \left\lbrace\begin{array}{ll}
%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acar\sigma_2' \\
%%     \emptyset & \mbox{if } \sigma_2 = \acdr\sigma_2' \,\,\mbox{ or }
%%     \end{array}\right.\\
%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
%% %\label{eqn:eup-red1}
%% \end{eqnarray*}
%% \end{minipage}
%% \hspace*{.65cm}\begin{minipage}{0.445\textwidth}
%% \begin{eqnarray*}
%%     \sigma_1\bcdr\sigma_2 & \hookrightarrow &
%%     \left\lbrace\begin{array}{ll}
%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acdr\sigma_2' \\
%%     \emptyset & \mbox{if } \sigma_2 = \acar\sigma_2' \,\,\mbox{ or }
%%     \end{array}\right.\\
%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
%% \end{eqnarray*}
%% \end{minipage}\\

The reflexive transitive closure of $\hookrightarrow$ will be denoted
as $\stackrel{*}{\hookrightarrow}$. The following proposition relates \Lfunonly\   and \Lfunonly$'$:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]  
\begin{boxedminipage}{\textwidth}
\footnotesize
\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} (\DEFINE\ (\append~\lista~\listb)
	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \xtest\ $\leftarrow $\ (\NULLQ~\lista) \IN
	  \UNL{3}      $\pi_2\!\!:\, $(\SIF\ \xtest~$\pi_3\!\!:$(\RETURN\ \listb)  
          \UNL{4} \hspace*{.05cm}     $\pi_4\!\!:\, $(\LET\ \xtl\  $\leftarrow$\   (\CDR\ \lista) \IN
	  \UNL{5}  \hspace*{.05cm}    $\pi_5\!\!:\, $(\LET\ \xrec\  $\leftarrow$  (\append\ \ \xtl\ \  \listb) \IN
	  \UNL{6} \hspace*{.05cm}    $\pi_6\!\!:\, $(\LET\ \xhd\  $\leftarrow$  (\CAR\ \lista)  \IN
	  \UNL{7}  \hspace*{.05cm}   $\pi_7\!\!:\,
          $(\LET\ \xans\  $\leftarrow$ (\CONS\ \ \xhd\ \ \xrec)  \IN\
          $\pi_8\!\!:\, $(\RETURN\ \xans))))))))
	\end{uprogram}}

\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pz\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}              (\LET\ \py\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_9\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\append\ \py\ \pz) \IN
          \UNL{4}   \hspace*{.78cm}  $\pi_{10}\!\!:\, $  (\LET\ \pa\  $\leftarrow$  (\CDR\ \pw) \IN
	  \UNL{5}   \hspace*{.78cm}  $\pi_{11}\!\!:\, $
          (\LET\ \pb\  $\leftarrow$ (\CAR\  \pa) \IN~$\pi_{12}\!\!:\,$ (\RETURN\ \pb)))))))
\end{uprogram}}
\end{boxedminipage}
  \caption{An example program.}
\label{fig:examp-pgm}
\figrule
\normalsize
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{proposition}
Assume that  a liveness computation  based on  \Lfunonly\   gives  the liveness of
the variable $x$ at a program point $\pi_i$ as $\sigma$ (symbolically, $\Lanv{i}{x}=
\sigma$).  Further,  let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is
used instead of \Lfunonly.   Then  $\sigma' \stackrel{*}{\hookrightarrow}
\sigma$.
\end{proposition}

To see  why the proposition is  true, consider an analysis  based on
\Lfunonly$'$ in which   $\sigma$ appears in
the  context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.  Let $\alpha  \in \sigma$.  The
symbol  \bcar(\bcdr)  merely  marks  a  place in  $\alpha$  where  the
original  \CONS\  rule  would  have erased  an  immediately  following
\acar(\acdr),  or, in  absence of  such a  symbol, would  have dropped
$\alpha$ itself.  Since the application  of any rule in \Lfunonly$'$ merely
adds  symbols at  the beginning  of  $\alpha$, the  markers and  other
symbols in $\alpha$ are propagated to other dependent parts of program
in their same relative positions.  Consequently, the erasure carried out at the end of the analysis with
$\stackrel{*}{\hookrightarrow}$ gives the same result as obtained
through \Lfunonly.

\subsection{Generating equations for the demand transformer \Lfonly}
\label{sec:Lfonlysymbolically}

%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
We  shall  consider the  program  in  Figure~\ref{fig:examp-pgm} as  a
running example;
this is the  ANF version of the program in Figure~\ref{fig:mot-example}.

%%AM no longer used: In what follows,  we shall overload the semantics of
%% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
%mean the generation
%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.

To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
of $x_i$.  For our example program which has  a
single function \append, this generates the following equations:
% $\Lfonly_{\append}$  is defined by equations:
  \begin{eqnarray*}
    \Lf{\append}{1}{\sigma}
    &=& \epsilonset \cup \acar\bcar\sigma
    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
    \Lf{\append}{2}{\sigma}
    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
  \end{eqnarray*}
  In  general, the  equations  for \Lfonly\  are recursive  since
  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
    is  similar to  solving the  differential equation  
    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
    that    the    solution     has    the    form    $\mathit{y}
    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
    equation  yields  a  quadratic  equation  in  $r$,  and  each
    solution of $r$ gives rise  to a solution of the differential
    equation  (in our  setup we  can effectively  pick  the least
    solution rather than needing linear combinations).  }
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
\end{eqnarray}
where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
why \Lfonly\ has this form  is as follows: Recall that
\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
situation in which  the demand $\sigma$ does not  reach the $i$th
argument.   This happens,  for  instance, when  argument is  used
only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
transformation of the demand $\sigma$ before it reaches the $i$th
argument. This  transformation prefixes strings in 
$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.

Substituting  the  guessed   form  in  the  equations  describing
\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
containing $\sigma$ and those  without $\sigma$ on both sides of the
resulting equation, we
get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
  \append}$, we get:
  \begin{align*}
    \Uf{\append}{1} &= \epsilonset
    \cup \acdr.\Uf{\append}{1}  
    &\quad 
    \Uf{\append}{2} &= \Uf{\append}{2} \\
    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
    \cup \acdr\Df{\append}{1}\bcdr &\quad 
    \Df{\append}{2} &= \epsilonset \cup
    \Df{\append}{2}\bcdr 
  \end{align*}
Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.

\subsection{Generating liveness equations \Lv\  for function bodies}
\label{sec:bodylivenessbodies}

\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}

We now calculate a 0-CFA-style summary liveness for each GC point of a program.
There are two parts to this.
First, for each function $f$,  we determine a {\em summary demand} $\sigma_f$
over-approximating any demand $\sigma$ passed to $f$.
%$\Lf{f}{i}{\sigma}$.
So, consider a function $g$ containing a call site $\pi$ to $f$, say
$\pi\!\!:\!(\LET~x \leftarrow (f\,y_1\,\ldots\,y_n)~\IN~e)$.
Under the assumption that the demand on $g$ is $\sigma_g$, the liveness at $e$
is $\Lv = \mathcal{L}(e,\sigma_g,\Lfonly)$, and the $\LET$ case of
Figure~\ref{fig:live-judge} tells us this call site contributes $\Lv(x)$ to the
demand $\sigma_f$ placed on  $f$.
Write $\deltacall{f}{\pi}{g}$ for
this demand (contributed by call site $\pi$ in $g$).

Thus, supposing the $k$ call sites to function $f$ are
$\pi^1$ (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
%%% \marginpar{and writing $\sigma_f^i$ for the above $\deltacall{f}{\pi^i}{g^i}$?}
then the over-approximation requirement on $\sigma_f$
is achieved by taking $\sigma_f = \deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
% is achieved by taking $\sigma_f = \sigma_f^1 \cup \cdots \cup \sigma_f^k$.

The expression $e_{\mainpgm}$ is a special case; we assume it may be called externally
with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$ (denoted
$\sigma_{\!all}$).
This is because any  part  of its  value  may be  used  by  the environment---for
printing  the  result,  for  instance.

% In what follows,  we shall overload the semantics of
% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
% 
% 
% We assume an overall demand of $\{\acar,\acdr\}^*$ (denoted
% $\sigma_{\!all}$) on the main expression $e_\mainpgm$.  This is because 
% any  part  of its  value  may be  used  by  the environment---for
% printing  the  result,  for  instance.

%% To  generate  equations for  the  body $e_{\mathit{f}}$ of  a
%% function,  we evaluate \Lfun{e_{\mathit{f}}}{\sigma_{\mathit{f}}}{\Lfonly}.     For ${\sigma_{\mathit{f}}}$,  we have
%% to consider all calling  contexts during execution.  We therefore
%% do a 0-CFA approximation by  defining
%% % $\sigma_\mathit{f}$        as $\bigcup_\calls{f}{k}\!\!\sigma_\calls{f}{k}$,
%% $\sigma_\mathit{f}$        as $\sigma_\calls{f}{1} \cup \cdots \cup \sigma_\calls{f}{n}$,
%% where $\calls{f}{k}$,
%% $\mathit{k} = 1 \ldots n$ represent all the calls to $\mathit{f}$
%% in  the  program,   and  $\sigma_\calls{f}{k}$  are  the  demands
%% generated at the calls  by the \rn{let-live} rule while analyzing
%% the bodies of the caller.

For the running example, $\append$ has calls from $\mainpgm$ at $\pi_9$
and a recursive call at $\pi_5$.
So $\sigma_{\append} =
     \deltacall{\append}{\pi_9}{\mainpgm}  \cup \deltacall{\append}{\pi_5}{\append}$.
Filling in  the values gives:
%  where the former is the demand generated at the
%  call to \append\  in $e_\mainpgm$ the latter
%  is  the demand  at the  recursive call. 
%% Propagation of
%%  $\sigma_{\!all}$ in $e_\mainpgm$ gives    $\sigma_\calls{\append}{1}$   as   $\{\epsilon,   \acdr\}   \cup
%%  \acdr\acar\sigma_{\!all}$  and propagation of
%%  $\sigma_{\mathit{\append}}$ in $e_\append$ gives   $\sigma_\calls{\append}{2}$   as
%%  $\bcdr\sigma_{\mathit{\append}}$.  
%% Filling in  the values of $\sigma_\calls{\append}{1}$ and
%% $\sigma_\calls{\append}{2}$, we get:
\begin{align*}
\sigma_{\append}    &=
\left( \{\epsilon,  \acdr\} \cup  \acdr\acar\sigma_{\!all}  \right) ~\cup~
\bcdr\sigma_{\mathit{\append}}
\end{align*}

Second, for each function $f$ (possibly $\mainpgm$) we need the liveness at
each contained GC point
$\pi\!\!:\!(\LET~x \leftarrow s~\IN~e)$,
where $s$ is a call either to $\CONS$ or to a user function $g$.
Given $\sigma_f$ calculated above, this
is simply $\mathcal{L}(\pi,\sigma_f,\Lfonly)$.
%%% \marginpar{AM: we've always said this (safe), but actual parameters
%%% live before the call and dead after needn't be included!}

For the running example, containing GC points $\pi_5, \pi_7$ in $\append$ and
$\pi_9$ in $e_\mainpgm$, this gives
(recall Equation~(\ref{eq:LF:DI}) above states $\Lf{f}{i}{\sigma} = \Uf{f}{i} \cup \Df{f}{i}.\sigma$):
\begin{align*}
\Lanv{5}{\xtl} &= \Lf{\append}{1}{\bcdr\sigma_{\append}}  \quad &
                                \Lanv{7}{\xhd}\,\,\, &=  \bcar\sigma_{\append}     \\ 
\Lanv{5}{\listb} &= \Lf{\append}{2}{\bcdr\sigma_{\append}} &
                                \Lanv{7}{\xrec} &=  \bcdr\sigma_{\append} \\
\Lanv{5}{\lista} &= \epsilonset \cup \acar\bcar\sigma_{\append} 
\\[1ex]
  \Lanv{9}{\py}  &= \Lf{\append}{1}{\{\epsilon,
    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}} \\
  \Lanv{9}{\pz}  &= \Lf{\append}{2}{\{\epsilon,
    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}}
\end{align*}

\noindent 
In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
% \marginpar{Consistency: summary vs cumulative?}
\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
an equation defining
  liveness at each GC point of $e_{\mathit f}$.
% {\color{red}
%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
% }
\end{enumerate}

\subsection{Solving liveness equations---the grammar interpretation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
% being a grammar with multiple start symbols---one for
% for each nonterminal \var{\Lanv{i}{x}}.
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
 \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
  \acdr\var{\Uf{\append}{1}}\\         
  \var{\Df{\append}{1}}
  &\rightarrow&               \acar\bcar               \mid
  \acdr\var{\Df{\append}{1}}\bcdr\\
\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
\var{$\sigma_{\mathit{\!all}}$}
\end{eqnarray*}
Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.   
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is  undecidable.   We
circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
  strongly  regular}  CFGs which  have   easy
translations to non-deterministic finite state automata (NFA). The
NFAs are then simplified on the
basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
membership of forward access paths. The resulting NFAs are   finally
converted to DFAs for use during garbage collection.
%------------------------------------------------------------%
\subsubsection{Approximating CFGs using NFAs.}
\label{sec:NFA-approx}

We use the   
algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
approximate a CFG  to a   {\em strongly regular\/} grammar.  The
transformation has  the  property that  if $\Lv$  is a
non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
  L}_{G'}(\Lv)$.    This  is   required  for the
approximation to be safe with respect to liveness.

We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
transformation. We  pick the only production that  is affected by
the transformation---the production for \Df{\append}{1}.  The
production for \Uf{\append}{1}, while  recursive, is already in
strongly  regular  form  and   is  therefore  unaffected  by  the
transformation.
 \begin{eqnarray*}
  \var{\Df{\append}{1}}
  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
  \acdr\var{\Df{\append}{1}} \\ 
\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
\mid \epsilon 
\end{eqnarray*}
The  languages
generated for  \var{\Df{\append}{1}} in  the original grammar  and the
new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.

%------------------------------------------------------------%
\subsubsection{Transforming NFAs to Accept Forward Paths:}
%\label{sec:NFA-elim}
The strongly regular CFGs obtained after the Mohri-Nederhof
transformation are
  first converted  into  NFAs.  The        algorithm         described        in
Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
Thus  \nt\  can  be  used  to  check  
membership of forward paths.

\begin{figure}[t]
\begin{boxedminipage}{\textwidth}
\begin{center}
  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
%
  \raggedright{\bf Output:} NFA  \nt\ with  underlying
  alphabet $\lbrace \acar, \acdr \rbrace$ such that
  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
  \\
%

  \raggedright{\bf Steps:}
 \begin{algorithmic}
   \STATE $i \leftarrow 0$ 
   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
   without $\epsilon$-moves \cite{hopcraft90toc}

   \REPEAT
   \STATE $\nt'_{i+1} \leftarrow \nt_i$
   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
     edge from $q'$ with label $\bcar$ and
     outgoing edge to $q''$ with label $\acar$} 
   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
   label $\epsilon$.
   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
   \ENDFOR

   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
     edge from $q'$ with label $\bcdr$ and outgoing edge to
     $q''$ with label $\acdr$} 
   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
     label $\epsilon$.
   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
   \ENDFOR

   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
   $\nt'_{i+1}$ without $\epsilon$-moves

   \STATE $i \leftarrow i+1$
   \UNTIL ($\nt_{i} = \nt_{i-1}$)

   \STATE $\nt \leftarrow \nt_i$
%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
 \end{algorithmic}
\end{center}
\end{boxedminipage}
 \caption{Algorithm for transforming an NFA to  accept forward
   paths only.}\label{algo:simplify-NFA} \figrule
\end{figure}
%------------------------------------------------------------%
The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
process is  continued till  a fixed point  is reached.   When the
fixed point  is reached, the resulting NFA  contains all possible
reductions corresponding  to all the  paths in the  original NFA.
The proofs  of the termination  and correctness of  the algorithm
are given in our earlier paper~\cite{karkare07liveness}.

\input{liveness-dfa.tex}

We illustrate the algorithm in Figure~\ref{algo:simplify-NFA} by
constructing the automaton for \var{\Lanv{9}{y}}.
Figure~\ref{fig:automata-construction} (c) shows the automaton for
\var{\Lanv{9}{y}} constructed by composing the automata for
\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
get (d). We add an $\epsilon$ edge from $q_2$ to $q_3$ bypassing the
\bcdr\acdr\ pair from $q_2$ to $q_2$ and then to $q_3$. This is shown
in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
$q_2$.  Removing this $\epsilon$ edge gives the automaton shown in
(g).  Restricting this automaton to forward edges only, we get the
final automaton shown in (h). This automaton recognizes $\acdr^{*}\!\mid\acdr^{*}\acar \sigma_{\mathit \!all}$, showing that the entire
list \py, including its elements, is live at $\pi_9$. Also note that
the language accepted by the final automaton satisfies the
prefix-closed property.

\section{Prototype and evaluation}\label{sec:experiments}
To demonstrate the effectiveness of liveness-based garbage collection,
we  have  built a  prototype  consisting  of  an interpreter  for  our
language,  a  liveness  analyzer  and  a copying  collector  that  can
optionally use the results of liveness analysis for marking instead of
reachability. When the collector uses liveness for marking, we call it
a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
  reachability-based  collector}  (RGC).   The  copying  collector  is
neither incremental nor generational.  As a consequence, any cell that
becomes unreachable or  dead is assuredly collected in  the next round
of garbage collection.

While exploring the activation record of a function, LGC uses the DFAs
at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
is invoked  at the program point   $\pi$,  it would  use  the
liveness  DFAs at  $\pi$ itself for exploring the root set of the
current activation record. For any other activation record in the
stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
the DFAs at the point of call to  ${\mathit g}$ would be used.

Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
point  pair ($x$,  $\pi$).  We  assume that  there is  a function
$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
$\dfa{\pi}{x}$.    
Considering   a   DFA   as   a   table,
$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
\acar\      or     \acdr.       We      shall     also      write
$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
indicating  whether there  is a  transition from  $q$ on
$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
pointer to $x$.
The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
$x$  in the  other semi-space  and returns  the new  address.  It
hides  details such as  returning the  forwarding pointer  if the
value  of $x$  is  already copied,  and  creating the  forwarding
pointer otherwise.

The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
in the heap over time for RGC and LGC---here time is measured in terms
of the number of cons cells allocated. In addition, they also show the
number of  reachable cells and the  number of cells  that are actually
live\footnote{This   is  statically   approximated  by   our  liveness
  analysis}.  Since  the programs have  different memory requirements,
we have tuned the size of heap for each program to ensure a reasonable
number of  collections. An invocation  of RGC decreases the  number of
cells  in heap  till  it touches  the  curve of  reachable cells.   An
invocation of LGC decreases the number  of heap cells to no lower than
the curve of live cells.

To construct the  reachable and live curves, we  record for every cell
its creation time (\CreateTime), its last use time (\UseTime), and the
earliest time  when the  cell becomes unreachable  and can  be garbage
collected (\CollTime).  For accurate recording of  \CollTime, we force
frequent invocations  of a reachability based collector  in a separate
run.     A    cell   is    live    at    time   \mbox{\exectime}    if
\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.

The benchmark programs are drawn from the {\tt no-fib} suite and other
sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
  fibheap} increases  initially, comes  down before the  heap becomes gets
full and then  continues to remains low.  This results in  a single garbage
collection in both RGC and LGC.  The graphs also show the precision of
our liveness analysis.  For  all programs  except  {\tt nperms} and
{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
manages to collect a good portion of the data that are not live.

\subsection{Results}
\begin{figure}[t] 
\centering
\begin{boxedminipage}{.75\textwidth}
{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
  then} {\bf skip}\\ 
\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
\noindent \hspace*{.35cm}{\bf return} $y$
\end{boxedminipage}
\caption{Function       for       copying       a       root       set
  variable.}\label{algo:LGC-main-loop} \figrule

\end{figure}

\begin{figure}[p]
\begin{tabular}{@{}c@{}c@{}}
\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
&
{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
\\
\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
 &
{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
\\
\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
&
{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
\\
\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
&
{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
\end{tabular}
 \caption{Memory  usage of  programs. The blue and the red curves
  indicate the number of cons cells in the active semi-space for RGC and
  LGC respectively. The black curve represents the number of reachable
  cells and the grey curve represents the number of cells that are
  actually live (of which liveness analysis does  a static
  approximation). x-axis is the time measured in number of cons-cells
  allocated. y-axis is the number of cons-cells.}
\label{fig:memory-usage} \figrule
\end{figure}
The  increased effectiveness  of LGC  over RGC  is also  shown  in the
tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
provides  statistics  regarding the  analysis  itself.  The number  of
states and the analysis times  are within tolerable limits. Precision
of analysis refers to the percentage of dead cells that is collected by
LGC, averaged over all invocations.
The second table shows garbage  collection statistics for RGC and LGC.
LGC  collects larger garbage  per invocation,  drags cells  for lesser
time and requires a smaller heap size ({\em MinHeap}) for program to run
in comparison with  RGC.

There are a couple of  issues of concern.  The garbage collection time
is larger in the case of LGC for some programs. The reason is that the
cost of consulting the liveness DFA may outweigh the combined benefits
of  fewer   garbage  collections   and  fewer  markings   per  garbage
collection.   The  other issue  is  illustrated  by  the program  {\tt
  lambda}    (memory   usage  shown   separately    in
Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
in Figure~\ref{fig:experimental-results}, the  number of
touched cells\footnote{These  are the  cells which are  visited during
  marking phase, often  more than once due to  sharing.} in this
example is much higher for
LGC.  This  increase is due to  excessive sharing among  heap nodes in
this program.  Note  that a node re-visited because  of sharing is not
explored  any   further  during  a  RGC   collection.   However,  this
curtailment cannot happen in  LGC because of the possibility that the
node,  re-visited in a  different liveness  state, may  mark a  set of
cells different from the earlier visit.
\begin{figure}[t]
\footnotesize

\centering

\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline
{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
  nperm} &  {\sf fibheap} & {\sf knightstour} &
{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
\hline
\hline
{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
\hline
\end{tabular}}

\centerline{(a)}

\medskip

\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
\hline
    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\

                            &   \multicolumn{2}{c|}{cells per GC}  
                            &   \multicolumn{2}{c|}{cells per GC} 
                            &   \multicolumn{2}{c|}{\#GCs} 
                            &   \multicolumn{2}{c|}{MinHeap} 
                            &   \multicolumn{2}{c|}{Avg. Drag} 
                            &   \multicolumn{2}{c|}{(sec)} \\
\cline{2-13}
{Program}    &
RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
\hline
\hline
    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
%    {\sf  nboyer} & &  & &   & &  & & &  \\
%    {\sf   circsim?} & &  & &   & & & & &   \\
 \hline
\end{tabular}}

\centerline{(b)}
\vskip -2mm
\caption{Experimental results comparing RGC and LGC. Table (a) gives
  data related to liveness analysis, and (b) gives garbage collection
  data. The average drag time is measured in allocated cons cells.}
\label{fig:experimental-results}
\normalsize
\end{figure}

\section{Collecting more garbage can never slow things down}
\label{sec:lgc-always-better}
Since garbage collection is  effectively asynchronous to the allocator
thread, one might  worry as to how {\em  robust} our measurements are.
For example,  while LGC would,  in general, collect more  garbage than
RGC in the same heap state,  is it possible that on some programs, LGC
might do  a larger number  of collections? We  prove a result  to show
that   this  cannot   happen.    This  result   applies  to   classical
mark-and-sweep and copying garbage  collectors and, we hypothesize, to
generational collectors too.

\begin{lemma}
For the  same mutator,  a liveness-based collector  can never  do more
garbage collections than a reachability-based collector.
\end{lemma}

\begin{proof}Assume, as before,  that time is measured in terms of the number of
  cons cells allocated. Let us run  two copies of the mutator with the
  two    different   collectors    but    with   synchronous    memory
  allocations.   The   garbage    collections,   however,   happen
  asynchronously.

  To prove the lemma, it is enough to show the truth of the following 
statement: After
 every  LGC invocation,  the count of LGC invocations   is  no greater
than  RGC invocations.   The  base case  holds  since the  first
invocations of both GCs happen at the same time.  Assume the statement
to be true after $n$ invocations of LGC.  Since LGC copies a subset of
reachable cells, its heap would contain no more cells than RGC heap at
the end  of the $n$th invocation.   Thus either RGC  is invoked next before
LGC, or LGC and RGC are both invoked next at the same time.  In either
case, the statement holds after $n+1$ invocations of LGC.
\end{proof}

%% \begin{figure}[h]
%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
%% \caption{Memory  usage of  programs. The  solid  lines represents
%%   number of  live cells and the  dashed lines show  the number of
%%   reachable cells.  The red and blue  lines represent invocations
%%   of RGC and  LGC---their heights indicate  the amount of memory
%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
%%       paper and are being used as placeholder}}}
%% \label{fig:memory-usage} \figrule
%% \end{figure}
%%\clearpage

\section{Related Work} Previous attempts to  increase  the space efficiency
of functional programs by additional  reclamation of memory fall in two
broad categories. In the first,  the program itself is instrumented to
manage  reclamation and reallocation  without the  aid of  the garbage
collector.    Such    attempts   include:   sharing    analysis   based
reallocation~\cite{jones89compile},                       deforestation
techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
methods   that  enable   the   garbage  collector   to  collect   more
garbage~\cite{inoue88analysis,lee05static}   by
explicitly  nullifying  pointers  that  are not  live.   However,  the
nullification,  done   at  compile  time,   requires  sharing  (alias)
analysis.  Our  method, in contrast, does not  require sharing because
of the availability of the heap  itself at runtime.  To the best of our
knowledge, this is the first  attempt at liveness-based marking of the
heap during garbage collection.

\begin{figure}[t]
\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
 \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
\label{fig:memory-usage-lambda} \figrule
\end{figure}

\section{Conclusions}
\label{sec:conclusion}
We have defined a notion of liveness on structured data;
this generalizes classical liveness and strong liveness.
We started with a general fully context-sensitive analysis which
we proved correct with respect to a minefield semantics
(this models the effect of garbage collection between every
evaluation step).

To avoid scalability issues (and to avoid performing part of
the liveness computation at run time)
we defined an 0-CFA version of this liveness analysis in which
demands for function $f$ at all calling contexts are conflated into 
a single demand $\sigma_f$.
This enabled us to treat the liveness equations symbolically obtaining
context-free grammars for liveness at each GC point (calls to user
functions and to $\CONS$).
These were then converted to DFAs for run-time consultation by the
garbage collector. Experiments confirm the precision of the analysis.

To obtain performance figures we compared a reachability-based garbage
collector with a liveness-based  collector.  This showed a decrease in
the  number  of  GCs,   more  garbage  collected  per  invocation.   A
significant benefit of LGC is  that programs can run in smaller memory
when compared to  RGC. This is potentially useful  in situations where
memory  is  limited---as with  embedded  systems.  For  a majority  of
programs, the garbage collection times were reduced.

One issue we  highlighted was that while fewer  nodes were marked (and
more garbage collected), sometimes  cons cells could be visited and
traversed  multiple times  with different  sets of  liveness  paths to
explore.  Further work includes  improvements to the classical copying
collector to reduce the cost of this.

\subsection*{Acknowledgements}
The authors  would like  to thank Hemanshu  Vadehra for  a preliminary
implementation of the prototype system. Amey Karkare was supported for
this work by the DST/SERC fast track scheme for young scientists.

\bibliography{fun_hra}{}
\bibliographystyle{splncs}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
\section{Appendix stuff}

\paragraph{Safety of using the modified \CONS\  rule:}
Recall that while the correctness of the analysis was proved with
respect to  the $\len$ rules  in Figure~\ref{fig:live-judge}, the
equations   for  liveness   are  based   on  a   modification  of
$\mathit{ref}$    in    the    case    of   \CONS\  as  given    by
(\ref{eqn:mod-cons}).  We now show that  if the \bcar\,s and \bcdr\,s
in live access  paths resulting  from this  modification are
eliminated through  applications of $\hookrightarrow$,  we obtain
the same results as given by the original rules.

\comment{Proposal: We prove a result which goes as follows:

\begin{quote}
Lemma: Let  $\Lanv{i}{x}= \sigma$ for an a  program point $\pi_i$
and a variable  $x$, when the judgement is  based on the original
$\len$  rules.   Further, let  $\Lanv{i}{x}=  \sigma'$, when  the
judgement is based on $\len$ with the modified \CONS\ rule.  Then
$\sigma' \stackrel{*}{\hookrightarrow} \sigma$
\end{quote}

Here are some  ideas regarding how this can  be proved.  Consider
an arbitrary equation
$$ \Lanv{j}{y} = \ldots \Lanv{i}{x} \ldots$$ generated by the
analysis.

Now assume that $\Lanv{i}{x}= \sigma$.   We want to find out what
happens to the access patterns in $\sigma$ in the access patterns
of \Lanv{j}{y}.   To answer this,  we put \# markers  around each
$\alpha \in \sigma$  and redo the equation using  an extension of
$\len$ to take  care of \#-tagged patterns.  We  can then observe
occurrences of \#-tagged patterns  in the liveness information at
$\pi_j$.

Here is  the $\len$ extension for  the original \CONS\  clause to take
care of the \#-tagged patterns:
$$\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          = \{x.\#\alpha\# \mid \#\acar\alpha\# \in \sigma\} \cup
\{y.\#\alpha\# \mid \#\acdr\alpha\# \in \sigma\} $$
And here  is  the $\len$ extension for  the modified \CONS\  clause.
$$ \mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= \{x.\#\bcar\alpha\# \mid \#\alpha\# \in \sigma\} \cup
\{y.\#\bcdr\alpha\# \mid \#\alpha\# \in \sigma\} $$
All  other rules  remain unchanged.  Tracking a  pattern  through this
extension is sound because erasing the \#-tags would give us the same
liveness as the original rules.

Let us also extend $\hookrightarrow$ to \#-tagged strings as:
$$\sigma_1  {\hookrightarrow} \sigma_2\,\, \mathit{iff}\,\, \sigma_2 =
\{\alpha_1 \#\alpha\# \alpha_2 \mid  \alpha_1 \#\alpha'\# \alpha2 \in
  \sigma_1\,\, and \,\, \{\alpha'\} \stackrel{*}{\hookrightarrow} \{\alpha\}\}$$

Now we have an  observation represented by the following diagram:

\centerline{\epsfig{file=diag1.eps, height=4cm}}

We need  to make this  observation general enough.   Therefore we
should prove  the observation even if \Lanv{i}{x}  = $\sigma$ had
patterns with \bcar\ and \bcdr.   We should be able to prove this
by structural induction on the rules.

How does  the observation prove  the original lemma?   Suppose we
replayed  judgements  with  ($\len$  +  \#) and  with  ($\len$  +
modified \CONS +  \#).  Suppose that $\pi_i$ is  the last program
point till which the results  are the same (because no \CONS\ was
encountered).  Now  consider a  $\sigma \in \Lanv{i}{x}$  for any
$x$.  Put \#-tags around each string in $\sigma$. Now we have the
diagram:

\centerline{\epsfig{file=diag2.eps, height=7cm}}

This shows that modulo erasure of \#-tags, using the original $\len$
judgement and  $\len$ judgement with modified \CONS\ rule followed by
$\hookrightarrow$ give the same liveness.

And I would like  to conclude that it is all right  to do our analysis
with $\len$ + modified \CONS\ and then simplify using $\hookrightarrow$.

}

{\color{blue}\marginpar{From sec 3.1, liveness judgements}
\paragraph{Semantic considerations.}
The incremental liveness for generic primitive $\PRIM$ is given by
$\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly) = 
           \{x.\epsilon,y.\epsilon\}$.
Earlier we suggested that classical strong liveness for addition,
but not division, could use the stronger
$ \{x.\epsilon,y.\epsilon \mid \epsilon \in \sigma\}$.
However this is invalid: garbage collection
could reuse dead, but reachable, $\CONS$ cell fields (of $v$ below)
thereby producing a stuck state in a dynamically typed language.
Hence we treat $+$ as like division not like addition---suppose
% $(FOO)$ is a user function which may cause garbage collection, and consider:
calling $\mbox{\footnotesize (FOO)}$ garbage collects and reallocates the cell pointed
to by `v' in:

\vspace{-15pt}

\begin{quote}
\footnotesize
\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1}  (\LET\ v\ $\leftarrow $\ (\CONS~2~3) \IN
	  \UNL{2}  (\LET\ w\ $\leftarrow $\ (FOO) \IN
	  \UNL{3}  (\LET\ x\ $\leftarrow $\ (\PRIM\ (\CAR\ v) 4) \IN
	  \UNL{4}  (\RETURN\ w))))
	\end{uprogram}}
\end{quote}
In principle, {\em dead code removal} of
$(\PRIM\ (\CAR\ {\rm v})\ 4)$
would render the stronger version valid, but still requires static typing
to avoid the risk of {\em removing} a stuck state.
We restrict ourselves to liveness annotation rather than code transformation.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\footnotesize
Old version of the liveness rules...

\infrule[ret-live]
        {}
        {(\RETURN~x),\sigma,\Lfonly \len  x.\sigma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\infrule[if-live]
        { e_1, \sigma,\Lfonly \len \Lv_1
          \andalso e_2, \sigma,\Lfonly \len  \Lv_2}
        {(\SIF~x~e_1~e_2),\sigma,\Lfonly \len
               \Lv_1\cup\Lv_2\cup \{x.\epsilon\}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\infrule[let-live]
        {e, \sigma,\Lfonly  \len  \Lv}
        {(\LET~x \leftarrow  s~\IN~e),
          \sigma,\Lfonly \len   \Lv\setminus \{x.*\} \cup
          \mathit{ref\/}(s,\Lv(x),\Lfonly)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\infrule[live-define]
        {e_f, \sigma, \Lfonly  \len
%            \bigcup \{z_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
           \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \fen \Lfonly
\\ \mbox{where }
     (\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ \ e_f) \mbox{ is a member of }
        d_1 \ldots d_k}
\figrule
  \caption{Liveness Judgement Rules}\label{fig:live-judge-old}
\normalsize
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \newgray{lightgray}{.75}
%\newcommand{\dem}[1]{\red{\scriptsize{$#1$}}}
\newcommand{\dem}[1]{\psframebox[linestyle=none,fillstyle=none,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
%\newcommand{\liv}[1]{\blue{\scriptsize{$#1$}}}
\newcommand{\liv}[1]{\psframebox[linestyle=none,fillstyle=solid,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[h!]
\psset{unit=1mm,nodesep=1mm,labelsep=0.5mm,arrows=->,linewidth=.1,framesep=0.2mm}
\begin{center}
  \begin{pspicture}(0,0)(120,50)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(12,5)
 \footnotesize
   \putnode{let1}{origin}{15}{50}{$\pi_1$:\LET}

    \putnode{test1}{let1}{-12}{-7}{\xtest}
    \putnode{null}{let1}{0}{-7}{$\pi_2$:\NULLQ}
    \putnode{lista1}{null}{0}{-7}{$\pi_3$:\lista}
    \putnode{if}{let1}{18}{-7}{$\pi_4$:\SIF}

    \putnode{test2}{if}{-8}{-7}{$\pi_5$:\xtest}
    \putnode{listb1}{if}{3}{-7}{$\pi_6$:\listb}
    \putnode{let2}{if}{14}{-7}{$\pi_7$:\LET}

    \putnode{tl1}{let2}{-22}{-7}{\xtl}
    \putnode{cdr}{let2}{-5}{-7}{$\pi_8$:\CDR}
    \putnode{lista2}{cdr}{0}{-7}{$\pi_9$:\lista}
    \putnode{let3}{let2}{15}{-7}{$\pi_{10}$:\LET}

\putnode{rec1}{let3}{-13}{-7}{\xrec}
\putnode{append}{let3}{0}{-7}{$\pi_{11}$:\append}
\putnode{tl2}{append}{-10}{-7}{$\pi_{12}$:\xtl}
\putnode{listb2}{append}{5}{-7}{$\pi_{13}$:\listb}

\putnode{let4}{let3}{17}{-7}{$\pi_{14}$:\LET}
\putnode{hd1}{let4}{-5}{-7}{\xhd}
\putnode{car}{let4}{4}{-7}{$\pi_{15}$:\CAR}
\putnode{lista3}{car}{0}{-7}{$\pi_{16}$:\lista}

\putnode{let5}{let4}{16}{-7}{$\pi_{17}$:\LET}

\putnode{ans1}{let5}{-4}{-7}{\xans}
\putnode{cons}{let5}{6}{-7}{$\pi_{18}$:\CONS}
\putnode{hd2}{cons}{-10}{-7}{$\pi_{19}$:\xhd}
\putnode{rec2}{cons}{10}{-7}{$\pi_{20}$:\xrec}
\putnode{ans2}{let5}{19}{-7}{$\pi_{21}$:\xans}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5

%% \putnode{x}{let1}{0}{3}{\dem{\sigma}}
%% \putnode{x}{if}{0}{3}{\dem{\sigma}}
%% \putnode{x}{let2}{0}{3}{\dem{\sigma}}
%% \putnode{x}{let2}{-7}{0}{\liv{\Lv_4}}
%% \putnode{x}{let3}{0}{3}{\dem{\sigma}}
%% \putnode{x}{let4}{0}{3}{\dem{\sigma}}
%% \putnode{x}{ans2}{0}{3}{\dem{\sigma}}
%% \putnode{x}{ans2}{-8}{0}{\liv{\Lv_1}}
%% %\putnode{x}{ans2}{9}{0}{\liv{\Lv_0}}
%% \putnode{x}{plus}{0}{3}{\dem{\sigma}}
%% \putnode{x}{rec2}{1}{4}{\dem{\epsilonset}}
%% \putnode{x}{plus}{-7}{0}{\liv{\Lv_6}}
%% %\putnode{x}{let4}{-7}{0}{\liv{\{\xrec\}}}
%% \putnode{x}{len}{3}{3}{\dem{\epsilonset}}
%% \putnode{x}{len}{-7}{0}{\liv{\Lv_3}}
%% \putnode{x}{tl2}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
%% \putnode{x}{tl2}{-6}{0}{\liv{\Lv_3}}
%% %\putnode{x}{let3}{-10}{0}{\liv{\xtl.\Lf{\xlen}{1}(\epsilonset)}}
%% \putnode{x}{cdr}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
%% \putnode{x}{cdr}{-7}{0}{\liv{\Lv_4}}
%% \putnode{x}{l2}{0}{4}{\dem{\epsilonset \cup \acdr\Lf{\xlen}{1}(\epsilonset)}}
%% \putnode{x}{l2}{-5}{0}{\liv{\Lv_4}}
%% \putnode{x}{zero}{0}{3}{\dem{\sigma}}
%% \putnode{x}{zero}{-3}{0}{\liv{\Lv_0}}
%% \putnode{x}{test2}{0}{3}{\dem{\epsilonset}}
%% \putnode{x}{test2}{-7}{0}{\liv{\Lv_5}}
%% \putnode{x}{if}{-6}{0}{\liv{\Lv_3}}
%% \putnode{x}{null}{0}{3}{\dem{\epsilonset}}
%% \putnode{x}{null}{-9}{0}{\liv{\Lv_4}}
%% \putnode{x}{l1}{0}{3}{\dem{\epsilonset}}
%% \putnode{x}{vone2}{0}{3}{\dem{\epsilonset}}
%% \putnode{x}{one}{0}{3}{\dem{\epsilonset}}
%% \putnode{x}{let5}{0}{3}{\dem{\sigma}}
%% \putnode{x}{l1}{-5}{0}{\liv{\Lv_4}}
%% %\putnode{x}{let1}{-17}{-1}{\liv{\{\xl\} \cup \xl.\acdr\Lf{\xlen}{1}(\epsilonset)}}
%% \putnode{x}{let1}{8}{0}{\liv{\Lv_0}}
%% \putnode{x}{let1}{-7}{0}{\liv{\Lv_4}}
%% %\putnode{x}{if}{7}{0}{\liv{\Lv_0}}
%% %\putnode{x}{let2}{8}{0}{\liv{\Lv_0}}
%% %\putnode{x}{let3}{8}{0}{\liv{\Lv_0}}
%% \putnode{x}{let3}{-7}{0}{\liv{\Lv_3}}
%% %\putnode{x}{let4}{9}{0}{\liv{\Lv_0}}
%% \putnode{x}{let4}{-7}{0}{\liv{\Lv_2}}
%% \putnode{x}{vone2}{-8}{0}{\liv{\Lv_6}}
%% \putnode{x}{rec2}{-8}{0}{\liv{\Lv_6}}
%% \putnode{x}{let5}{-8}{0}{\liv{\Lv_6}}
%% \putnode{x}{one}{-6}{0}{\liv{\Lv_2}}
\ncline{-}{let1}{test1}
\ncline{-}{let1}{null}
\ncline{-}{let1}{if}
\ncline{-}{null}{lista1}

\ncline{-}{if}{test2}
\ncline{-}{if}{listb1}
\ncline{-}{if}{let2}
\ncline{-}{let2}{tl1}
\ncline{-}{let2}{cdr}
\ncline{-}{let2}{let3}
\ncline{-}{cdr}{lista2}

\ncline{-}{let3}{rec1}
\ncline{-}{let3}{append}
\ncline{-}{let3}{let4}

\ncline{-}{append}{tl2}
\ncline{-}{append}{listb2}

\ncline{-}{let4}{let5}
\ncline{-}{let4}{hd1}
\ncline{-}{let4}{car}
\ncline{-}{car}{lista3}
%% \ncline{-}{len4}{tl2}
%% \ncline{-}{let4}{vone1}
%% \ncline{-}{let4}{one}
%% \ncline{-}{let4}{let5}
\ncline{-}{let5}{ans1}
\ncline{-}{let5}{cons}
\ncline{-}{cons}{hd2}
\ncline{-}{cons}{rec2}
%% \ncline{-}{let5}{plus}
\ncline{-}{let5}{ans2}
\end{pspicture}
 \end{center}

  \caption{Example to illustrate liveness computation}\label{fig:live-examp}
\figrule
\normalsize
\end{figure}

\newpage

\begin{figure}[t]
  \begin{tabularx}{320pt}{|| c | X | X | X ||}
\hline
Pgm  Pt & Demand  & Pre & Post \\
\hline 
$\pi_1$ & $\sigma$ & \{\lista\} $\cup$ \listb.$\sigma$ $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & $\phi$ \\
\hline
%% $\pi_2$ & $\epsilonset$ & \{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
%% \hline
$\pi_3$ & $\epsilonset$ &\{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
\hline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% $\pi_4$ & $\sigma$ &\{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
%% \hline
%% $\pi_5$ & \epsilonset & \{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  \\
%% \hline
%% $\pi_6$ & $\sigma$ & \listb.$\sigma$ & $\phi$ \\
%% \hline
%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% $\pi_7$ & $\sigma$  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$\\
%% \hline
%% $\pi_8$ & \Lf{\append}{1}(\{\bcdr\}$\sigma$) & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
%% \hline
$\pi_9$ & \epsilonset $\cup$ \{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
\hline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% $\pi_{10}$ &$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
%% \hline
$\pi_{11}$ & \{\bcdr\}$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \lista $\cup$ \lista.\{\acar\bcar\} \\
\hline
$\pi_{12}$ &\Lf{\append}{1}(\{\bcdr\}$\sigma$)   &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
\hline
$\pi_{13}$ & \Lf{\append}{2}(\{\bcdr\}$\sigma$)  &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
\hline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
$\pi_{14}$ & $\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$ \\
\hline
%% $\pi_{15}$ & \{\bcar\}$\sigma$  & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   &  \xrec.\{\bcdr\}$\sigma$\\
%% \hline
%% $\pi_{16}$ & \epsilonset $\cup$ \{\acar\bcar\}$\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$  & \xrec.\{\bcdr\}$\sigma$ \\
%% \hline
$\pi_{17}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   & $\phi$  \\
\hline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
$\pi_{18}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$  \\
\hline
$\pi_{19}$ &  \{\bcar\}$\sigma$        & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
\hline
$\pi_{20}$ &   \{\bcdr\}$\sigma$       & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
\hline
$\pi_{21}$ & $\sigma$ & \xans.$\sigma$ & $\phi$  \\
\hline
\end{tabularx}
  \caption{Liveness annotations for the body of \append}\label{fig:xe_append}
\figrule
\normalsize
\end{figure}

\end{document}

### 2014-07-13 paper/paper.tex
organization added

--- 2014-07-13 paper.tex
@@
 \begin{abstract}
-Current garbage collectors  leave much heap-allocated data uncollected
-because they preserve data {\em reachable} from a root set.
-% However, it is only necessary to preserve {\em live} data---a subset of
-% reachable data.
-However, only  {\em live} data---a subset of  reachable data---need be
-preserved.
-
-Using   a    first-order   functional   language    we   formulate   a
-context-sensitive liveness  analysis for structured data  and prove it
-correct.   We  then use  a  0-CFA-like  conservative approximation  to
-annotate  each  allocation  and  function-call program  point  with  a
-finite-state  automaton---which   the  garbage-collector  inspects  to
-curtail reachability  during marking. As  a result, fewer  objects are
-marked  (albeit  with a  more  expensive  marker)  and then  preserved
-(e.g.\ by a copy phase).
-% and secondly because of the consequent
-%reduced number of collections during a computation.
-
-Experiments  confirm the  expected performance  benefits---increase in
-garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections, a decrease  in the memory size required  to run programs,
-and  reduced  overall  garbage  collection  time  for  a  majority  of
-programs.
+\input{sections/abstract}
 \end{abstract}
@@
 
-\section{Introduction}
-\label{sec:intro}
+\input{sections/intro}
 
-Most modern  programming languages support dynamic  allocation of heap
-data.  Static  analysis of heap data  is much harder  than analysis of
-static   and   stack   data.    Garbage   collectors,   for   example,
-conservatively  approximate  the liveness  of  heap  objects by  their
-reachability  from a  set of  memory  locations called  the {\em  root
-  set\/}.  Consequently, many objects  that are reachable but not live
-remain  uncollected, causing  a  larger-than-necessary memory  demand.
-This is confirmed  by empirical studies on Haskell~\cite{rojemo96lag},
-Scheme~\cite{karkare06effectiveness}                                and
-Java~\cite{shaham02estimating} programs.
-
-Here we consider a first-order  pure functional language and propose a
-liveness  analysis  which  annotates  various program  points  with  a
-description  of variables and  fields whose  object references  may be
-dereferenced  in the future.   The garbage  collector then  only marks
-objects pointed by live references and leaves other, merely reachable,
-objects  to  be  reclaimed.    (Although  not  strictly  necessary,  a
-collector  would  normally {\em  nullify}  dead  variables and  fields
-rather than leaving dangling  references.)  Since there are fewer live
-objects   than   reachable   objects,   more  memory   is   reclaimed.
-Additionally, since  the collector traverses a smaller  portion of the
-heap, the time spent for each collection is also smaller.  The work is
-presented in  the context of a  stop-the-world non-incremental garbage
-collector (mark-and-sweep,  compacting or  copying) for which  we also
-show a  monotonicity result: that  our technique can never  cause more
-garbage  collections  to  occur   in  spite  of  changing  the  rather
-unpredictable  execution  points   at  which  collections  occur.   We
-anticipate that our technique  is applicable to more modern collectors
-(generational,  concurrent, parallel),  but leave  such  extensions to
-future work.
-
-We first define a {\em fully context-sensitive} (in the sense that its
-results  are unaffected  by function  inlining) liveness  analysis and
-prove it  correct.  However, fully context-sensitive  methods often do
-not scale,  and this analysis would  also require us  to determine, at
-run-time,  the internal  liveness of  a  function body  at each  call.
-Hence,   similarly   to   the   0-CFA   approach,   we   determine   a
-context-independent {\em summary} of  liveness for each function which
-safely   approximates   the   context-dependence   of   all   possible
-calls~\cite{chatterjee99relevant,cherem07fast,lee05static}.       (Note
-that  an intraprocedural context-insensitive  method which  assumes no
-information  about function  callers would  be too  imprecise  for our
-needs.)   In essence  our approach  sets up  interprocedural data-flow
-equations for the liveness summaries  of functions and shows how these
-can be solved symbolically  as context-free grammars (CFGs)\@.  We can
-then determine  a CFG  for each program  point; these are  then safely
-approximated with  finite-state automata  which are encoded  as tables
-for each program point.  For  garbage collection purposes only {\em GC
-  points}  (program points  representing  calls, to  $\CONS$  or to  a
-user-function) need to be stored.
-% which may call eventually call $\CONS$, need to be stored.
-
-We  previously proposed  an intraprocedural  method for  heap liveness
-analysis   for   a   Java-like   language~\cite{khedker07heap}   which
-statically inserted  statements nullifying dead  references to improve
-garbage collection; by  contrast nullification here occurs dynamically
-(which can work better with  aliasing) when the garbage collector acts
-on  liveness  annotations  to  avoid traversing  dead  references.   A
-workshop    paper~\cite{karkare07liveness}    outlined    the    basic
-0-CFA-style-summary  interprocedural  approach  to  functional-program
-liveness  analysis.   The  current  paper adds  the  context-sensitive
-analysis and better formalization along with experimental results.
-%   to  substantiate our ideas.
-
-\kern -3pt   %% AM: a very small tweak to get back to 2 pages for this
-             %% section.
-%% AM: rewrite the next paragraph to save 3-4 words and remove this \kern.
-
-%------------------------------------------------------------%
-\setcounter{page}{2}
-\subsubsection{Motivating Example}
-\label{sec:motiv}
-
-Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
-$\pi$ of an expression $e$  denotes a program point.  During execution
-of  the program, it  represents the  instant of  time just  before the
-evaluation  of  $e$.  
-%%While we discuss liveness, it  will also be important to consider an
-%%instance of time  just after the execution of  an expression, though
-%%we shall not be required to mark it with a label.
-% memory as  a (possibly  disconnected) directed
-% acyclic   graph.\footnote{Since  the   language   under  consideration
-%   (Section~\ref{sec:defs}) does not  have any imperative features, the
-%   heap cannot  have cycles.}
-% Here all data  is assumed to  be boxed, i.e.\
-% stored in heap and accessed through references.
-We  view the  heap  as a graph.  
-Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
-containing values.  Edges in the graph are {\em references} and
-emanate from {\em variables} or fields.
-Variable and field values may also be atomic values ($\NIL$, integers etc.)
-While it is convenient to box these in diagrams, our presented
-analysis treats them as non-heap values.
-
-Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
-The edges shown  by thick arrows are those  which are made live by the
-program. In addition,  assuming that
-the value of any reachable part of the program result may be explored
-or printed, the edges marked by thick dashed arrows are also live.
-A cell  is  marked and preserved  during  garbage collection,  only  if it  is
-reachable from the  root set through a path of  live edges.  All other
-cells can be  reclaimed.  Thus if  a garbage collection takes
-% place   at    the   time   represented   by   the    heap   shown   in
-at $\pi$ with the    heap   shown   in
-Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
-will be marked and preserved.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% MOTIVATING EXAMPLE
-\newcommand{\nilfigure}
-{\scalebox{0.75}{
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,0)(1,1)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
-\putnode{start}{origin}{0}{0}{}
-\putnode{stop}{origin}{10}{10}{}
-\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
-\end{pspicture}
-}}
-
-\begin{figure}[t!]
-%%%%{\color{Myblue}
-\begin{picture}(100,130)(0,-90)
-%  \begin{center}
-    \begin{tabular}{cc}
-\begin{boxedminipage}{.5\textwidth}
-      {\sf
-	\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
-	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
-	       \listb
-	  \UNL{1}      (\CONS\ (\CAR\  \lista)
-%          \UNL{2}          (\CAR\  \lista)
-	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
-          \listb))))
-          \UNL{0}
-	  \UNL{0} \hspace*{-.49\TAL} (\LET\ z  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
-	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
-	  \UNL{0} (\LET\ y  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
-	  \UNL{1}   (\LET\ w $\leftarrow$ (\append\ y z)\ \IN
-	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ w)))))
-	\end{uprogram}
-      }} 
-\end{boxedminipage}
-      &
-      \raisebox{-25mm}{\scalebox{.75}{
-	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
-      \psset{unit=1mm}
-      \psset{linewidth=.3mm}
-      \begin{pspicture}(0,-5)(70,60)
-	%\psframe(0,0)(73,60)
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
-	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
-%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
-	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	%      fillcolor=white]{$\times$}}}
-        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
-	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
-	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-		  fillcolor=white]{$\times$}}}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
-	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
-	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
-	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
-	\ncline[nodesepB=-.5]{*->}{c2}{d}
-	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
-	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
-	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
-	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
-	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
-	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
-%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
-        %here
-        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
-        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-
-%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
-%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
-	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
-	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
-%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      \end{pspicture}}} \\
-      (a) Example program.&
-      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
-	(b) Memory graph at $\pi$. Thick edges denote \\
-        {\white (b) }live links. Traversal stops at edges marked\\
-        {\white (b) }$\times$  during garbage collection.
-      \end{tabular}}
-    \end{tabular}
-\end{picture}
-
-\kern -3ex
-
-\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
-%  \end{center}
-%%%%}\color{Myblue} end
-\end{figure}
-
-\subsubsection{Organization of the paper}
-Section~\ref{sec:defs} gives
-the syntax and semantics of
-the language used to illustrate our analysis along with basic concepts
-and    notations.
-% Section~\ref{sec:operational}   describes    the
-% operational semantics  of the  language.
-Liveness analysis is described in
-Section~\ref{sec:liveness} followed by a sketch of a correctness
-proof relative to a non-standard semantics.
-Section~\ref{sec:computing} shows how to encode liveness as
-finite-state automata.
-Section~\ref{sec:experiments}   reports   experimental   results and
-Section~\ref{sec:lgc-always-better} proves that a liveness based
-collector can never do more garbage collections than a reachability
-based collector. 
-% and Section~\ref{sec:conclusion} concludes the paper.
-
-\section{The target language---syntax and semantics}
-\label{sec:defs}
-We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
-functions ($\CONS$, $\PRIM$ etc.).
-The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
-it has eager semantics and restricts programs to be in 
-Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
-where all actual parameters to functions are variables.
-This restriction does not affect expressibility (and indeed we feel free
-to ignore it in examples when inessential), but simplifies the analysis formulation.
-Additionally, as in the three-address instruction form familiar from compiler texts,
-it forces each temporary to be named and function calls to be serialized
-(necessary to get an unambiguous definition of liveness).
-We further require that each variable in a program is distinct, so that no scope shadowing
-occurs---this simplifies proofs of soundness.
-In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
-or return ($\RETURN$).
-The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.
-
-The body of the program is
-the  expression denoted  by  $e_\mainpgm$; for analysis purposes
-it is convenient
-to regard $e_\mainpgm$ as part of a function definition
-$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
-% As  mentioned earlier,
-We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.
-
-% \paragraph{Inlining.}
-In spite of the ANF restrictions it is still possible
-to inline non-recursive functions (a fact we use to prove the safety
-of liveness analysis).  A user-function call
-$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
-to a function defined
-(after renaming its formals and locals to be disjoint from existing variables)
-by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
-is replaced by a sequence of $\LET$'s of the form
-$z_i\leftarrow (\ID\,\,y_i)$ followed by
-the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
-$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.
-
-\begin{figure}[t!]
-\footnotesize
-\begin{eqnarray*}
-   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
-    \hspace{7em} \mbox{\em --- program}\\
-    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
-    e) 
-    \hspace{1.7em} \mbox{\em --- function definition} \\
-e \in \mathit{Expr} & ::= &
-\left\{\begin{array}{ll@{\hspace{4em}}l}
-       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
-       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
-       (\RETURN\,\, x) && \mbox{\em --- return from function}
-    \end{array}\right. \\
-s \in \mathit{Stmt} & ::= &
-\left\{\begin{array}{lr@{\hspace{1em}}l}
-       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
-       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
-       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
-       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
-%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
-       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
-%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
-       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
-            & \mbox{\em --- function application} 
-    \end{array}\right.
-\end{eqnarray*}
-  \caption{The syntax of our language}\label{fig:lang-syntax}
-\figrule
-\normalsize
-\end{figure}
-
-\subsubsection{Semantics}
-We now give an operational
-semantics for  our language.  Later, a refinement of
-the  operational semantics,  which we  call minefield  semantics, will
-serve to prove liveness analysis correct.
-We give a small-step semantics because, unlike big-step semantics,
-correctness for non-terminating programs does not need special treatment.
-We start with the domains used by the semantics:
-\[
-\begin{array}{rlcl@{\hspace{2em}}l}
-% \begin{eqnarray*}
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
-\mbox{-- Environment} \\ 
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
-  Val}) & \mbox{-- Heap}
-% \end{eqnarray*}
-\end{array}
-\]
-Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
-A value is either a number, the empty list $\NIL$, or a location  $\ell$.
-Our liveness analysis does not track numeric values, and thus is neutral
-as to whether these are boxed or represented as immediates.
-An environment is a finite mapping from variables to values, and a heap a finite mapping
-from locations to pairs of values.
-% Given  a program,  an  environment is  a
-% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
-% values in $\mathit{Val}$.    The heap is a mapping
-% from  locations  to pairs of  values.
-Finally, $\stk$ is a stack
-(using $\bullet$ for push and $[\,]$ for empty stack)
-of frames of unfinished function calls.  A
-frame is a triple $(e,x,\rho)$ representing the call site
-$
-(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
-$
-being evaluated in environment $\rho$.
-Frames can also be viewed as continuations, in this view
-the ({\sc ord-return}) 
- rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
-
-The semantics 
-of statements $s$ are given by the judgement form
-$\rho, \heap, s \rightsquigarrow \heap', v$
-and those for expressions $e$ by the form
-$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
-The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
-program terminates successfully with result value $\rho(x)$
-on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$
-
-Notation:  we write $\rho[x \mapsto v]$ for the environment which is
-as $\rho$ but has value $v$ at $x$.
-We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
-has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
-$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
-$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
-
-\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
-with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
-a contained $s$).  Some of these we eliminate syntactically, e.g.\
-ensuring all variables and functions are defined and are called with
-the correct number of parameters.
-Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
-All but the first can be eliminated with a static type system but,
-treating our program as dynamically typed, we regard all these
-as stuck states.
-
-% A program terminates successfully if it reaches the halt state
-% in Figure~\ref{fig:lang-opsem-small}.
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-\begin{figure}[t!]  
-% (Random text to ensure proper line-spacing (problem with infrule macro?).)
-
-\footnotesize
-\begin{minipage}{0.335\textwidth}
-\infrule[ord-const]
-        {}
-        {\rho, \heap, \pp{\pi}{k}
-          \rightsquigarrow \heap, k}
-\end{minipage}\begin{minipage}{0.655\textwidth}
-\infrule[ord-cons]
-      {
-\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
-}
-      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
-        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
-        \ell}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-car]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
-          \rightsquigarrow \heap, v_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-cdr]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
-          \rightsquigarrow \heap, v_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-id]
-      {
-}
-      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
-        \rightsquigarrow \heap, \rho(x)}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-prim]
-      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
-      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
-        \rightsquigarrow \heap, \rho(x) + \rho(y)}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-null-false]
-\infrule[]
-        {\rho(x)\not= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 0}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-null-true]
-\infrule[ord-null]
-        {\rho(x)= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 1}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-if-true]{
-\infrule[]{
-  \rho(x) \in \mathbb{N}\setminus\{0\}
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-if-false]{
-\infrule[ord-if]{
-  \rho(x) = 0
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-nonfn]{
-%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
-  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
-  \andalso
-  \mbox{$s$ is not
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-fncall]{
-  \mbox{$s$ is 
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-\andalso
-\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
-[\myvec{z} \mapsto \rho(\myvec{y})],
-  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
-}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-return]{
-}{
-  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
-  \pp{\pi}{(\RETURN\ x)}
-  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
-\end{minipage}
-
-% \normalsize
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-  \caption{The small-step operational semantics}
-  \label{fig:lang-opsem-small}
-\figrule
-\normalsize
-\end{figure}
-
-%==============================================================
-\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-
-\section{Liveness}\label{sec:liveness}
-
-In classical liveness analysis a variable is either `live' (its value
-may be used in future computation) or `dead' (definitely not used).
-Semantically, a variable is dead at a given program point
-if arbitrary changes to its value have no effect on the computation.
-Later we will use $\bot$ to represent a value which `explodes' when
-it is used in a computation; dead variables can safely have their
-value replaced with $\bot$.
-For heap-allocated data  we need a richer model of liveness
-in that both variables and fields of $\CONS$ cells may be dead or live.
-Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
-liveness of the structure reachable from a variable is a set of
-{\em access paths}
-which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
-grammar notation.
-Thus the liveness of $x$ being
-$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
-only refer to the second and third members of $x$ considered as a list.
-Semantically, access paths are prefix-closed, as accessing a field
-requires accessing all the paths from the variable to the field,
-and hence the above liveness is properly written 
-$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
-The classical notions of a scalar variable being live or dead correspond to
-$\{\epsilon\}$ and $\{\}$.
-
-The overall liveness (also written {\em liveness environment} for emphasis)
-at a program point is conceptually a mapping from
-variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
-notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
-instead of the map
-$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
-Analogously to classical liveness, the liveness at program point $\pi$
-in $\pi:e$ is the liveness just before executing $e$.
-
-A complementary notion to liveness is {\em demand}.  The demand for
-expression $e$ is again an access path---that subset of of $\{\acar,\acdr\}^\ast$ which the
-context of $e$ may explore of $e$'s result.  So, for example given a
-demand $\sigma$ and the expression $\pi: (\RETURN\ x)$, the liveness at
-$\pi$ is exactly $x.\sigma$.
-The classical analogy of this is in {\em strong liveness}, where
-an assignment node $n: x:=y+z$ causes $y$ and $z$ to be live on entry
-to $n$ if (and only if) $x$ is live at exit of $n$---the liveness
-of $x$ at exit from $n$ becomes the demand on $y+z$.
-Note that, for an operation like division which
-may raise an exception, the assignment $n: x:=y/z$ makes $y$ and $z$ live
-regardless of the liveness of $x$.
-
-We use $\sigma$ to range over demands, $\alpha$ to range over access paths
-and $\Lv$ to range over liveness environments. The notation
-$\sigma_1\sigma_2$ denotes the set $\lbrace \alpha_1\alpha_2 \mid
-\alpha_1 \in \sigma_1, \alpha_2 \in \sigma_2\rbrace$. Often we shall
-abuse notation to juxtapose an edge label and a set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
-Finally, we use $\Lfonly$ to range over {\em demand transformers}.
-These transform demands on a function call
-into demands for its formal parameters:
-given a user function $f$, defined by
-$(\DEFINE\ (f\ x_1\ \ldots\ x_n)\ \ e_f)$ and called with demand $\sigma$,
-then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
-
-Note that liveness refers to variables and fields, and not to $\CONS$ cells
-(i.e.\ to edges in the memory graph, not to locations themselves).
-Hence liveness of $\{x.\epsilon,x.\acar\}$ means that future computation
-may refer to the value $\ell$ of variable $x$, and also to the $\CAR$ field
-of location $\ell$.
-In the absence of other pointers to heap location $\ell$, we are
-certain that the $\CDR$ field of $\ell$ will not be referenced
-and may hence be corrupted arbitrarily.
-Note therefore, that while  $\ell$ cannot be garbage collected, any
-location $\ell'$ stored in the $\CDR$ field of $\ell$ would be garbage
-(again provided there are no other aliases to  $\ell$).
-
-%% \comment{
-%%   \begin{enumerate}
-%% \item Set the framework for liveness. Basic notions: access
-%%   paths and expressions.
-%% \item An informal working definition of liveness. Minefield will
-%%   provide the formal definition.
-%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
-%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
-%%   rules. One of the changes from  the earlier version is the lack of a
-%%   post-liveness in  the judgements. This anticipates  the intended use
-%%   of  the analysis  for garbage  collection and has  to be  carefully
-%%   explained.
-%%   \end{enumerate}
-%% }
-
-%% We now  set up  the framework for  liveness analysis.  Given  a memory
-%% graph, a link in the graph  can be described by starting with a member
-%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
-%% \CDR\ dereferences to reach the link.  For notational convenience, the
-%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
-%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
-%% dereferences    is    represented   by    a    string   $\alpha    \in
-%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
-%% in the  memory graph that starts  at $x$ and  traverses adjacent links
-%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
-%% We call the string $x.\alpha$ an {\em access expression\/}, the string
-%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
-%% graph   by  the   access  expression   an  {\em   access   path}.   In
-%% Figure~\ref{fig:mot-example},        the       access       expression
-%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
-%% containing the value  $4$.  Most often the memory  graph is clear from
-%% the context,  and we shall use  access expressions to  refer to access
-%% paths. An  access path  used to refer  to a  link in the  memory graph
-%% refers    to     the    last     link    in    the     path.     Thus,
-%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
-%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
-%% set of access paths, then  $x.\sigma$ will represent the set of access
-%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
-%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
-%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
-%% \sigma\rbrace$.
-
-%% \subsection{Liveness Analysis}
-%% \label{sec:exp-liveness}
-%% 
-%% Consider the memory graph at a program point $\pi$ during an execution
-%% of a program. A  link in the memory graph is said  to be {\em live} at
-%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
-%% execution  starting from  the  evaluation of  the expression  labelled
-%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
-%% be live at  $\pi$, if, during {\em any} execution  of the program, the
-%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
-%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
-%% In the first, the path  itself represents the sequence of dereferences
-%% to  access  the last  edge,  as,  for  example, while  evaluating  the
-%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
-%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
-%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
-%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
-%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
-%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
-%%   implicitly  live} if it is live in the second sense.  Clearly, static
-%% determination   of  implicit  liveness   requires  an   analysis  that
-%% determines  potential sharing  during  any execution  of the  program.
-%% However, the intended application  of our liveness analysis is garbage
-%% collection  during which  we have  access  to the  entire heap.   This
-%% obviates the need for a  separate sharing analysis and we restrict our
-%% attention to explicit liveness only.
-%% 
-%% 
-%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
-%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
-%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
-%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
-%% %% \bcdr\ represent application of  the constructor \CONS.  The following
-%% %% example motivates the need for these symbols.
-%% 
-%% %% \begin{example}\label{exmp:motiv-bar}
-%% %%   Consider the expression at program point $\pi_1$ in
-%% %%   \begin{center}
-%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
-%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
-%% %% 			    $\pi_3\!:${$\cdots$})}}
-%% %%   \end{center}
-%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
-%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
-%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
-%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
-%% %%   \begin{itemize}
-%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
-%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
-%% %%     live at $\pi_1$.
-%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
-%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
-%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
-%% %%   \end{itemize}
-%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
-%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
-%% %% \end{example}
-%% 
-%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
-%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
-%% %% resulting access patterns may not  be directly useful in that they may
-%% %% not describe a  path in the memory graph.  We  define a {\em canonical
-%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
-%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
-%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
-%% %% to obtain a canonical access  pattern representing the same edge. This
-%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
-%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
-%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
-%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
-%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
-%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
-%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
-%% %% cannot  be  used for  nullification  during  garbage collection.   The
-%% %% liveness of a such a path is irrelevant, and the path is not considered
-%% %% any further.
-%% 
-%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
-%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
-%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
-%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
-%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
-%% %% element in $\sigma_2$, i.e.
-%% 
-%% %% \begin{eqnarray*}
-%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
-%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
-%% %% \end{eqnarray*}
-
-\subsection{Liveness Analysis}
-First recall the classical formulation of liveness (as sets of simple variables)
-on three-address instructions,
-$
-\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I)
-     \setminus \mathit{def\/}(I)
-     \cup   \mathit{ref\/}(I)
-$,
-and then note that {\em strong liveness} needs, when $I$ is the instruction $z:=x+y$,
-that $\mathit{ref}(I)$ be refined to
-$\{x,y\}$ if $z \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
-
-Our liveness analysis formulated in Figure~\ref{fig:live-judge}
-is analogous.
-% has three analogous aspects.
-% \begin{itemize}
-% \item
-Firstly, the function $\mathit{ref}$, when given a statement $s$,
-returns the incremental liveness {\em generated} by $s$.
-Because we generalize {\em strong} liveness, $\mathit{ref}$ needs a second parameter,
-specifying the demand $\sigma$ on the result of $s$,
-to determine which access paths of its free variables are made live.
-The cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
-% The cases for
-% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
-% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
-% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
-A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
-argument (even if its result is never used).
-This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
-(note $\sigma$ may be $\{\}$).
-In the opposite  sense, the  demand  of $\acar\sigma$  on
-(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-that \CONS\ does not, by itself, dereference its arguments.
-Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
-$\mathit{ref}$
-expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
-Constants generate no liveness.
-
-% \item
-The function $\mathcal{L}$ now gives the (total) liveness of an expression $e$.
-The cases $\RETURN$ and $\SIF$ are straightforward, but note the liveness $x.\epsilon$ generated
-by the latter.
-The case $(\LET\ z\leftarrow s\ \IN\ e')$
-resembles a three-address instruction:
-the liveness of $e$ is given by taking the liveness, $\Lv$, of $e'$,
-killing any liveness of $z$ and adding any incremental liveness from $s$.
-The main subtlety is how the liveness of $z$ in $\Lv$ is converted to a demand $\Lv(z)$ to be
-placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-% \item
-%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-Finally, the judgement form $\mathit{Prog} \len \Lfonly$ is used to determine $\Lfonly$\@.
-Analogously to classical liveness being computed as
-a solution of dataflow equations, we require, via inference rule ({\sc live-define}),
-$\Lfonly$ to satisfy the fixed-point property that:
- when we assume $\Lfonly$ to be the family of demand transformers for the program
- then the calculated liveness of each function body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees
- with the assumed  $\Lfonly$.
-As usual, there are often multiple solutions to $\Lfonly$; all are safe
-(see Section~\ref{sec:correctness})
-but we prefer the least one as giving the least liveness subject to safety---and
-hence greatest amount of garbage collected.
-% \end{itemize}
-
-We make three observations: firstly the rule ({\sc live-define}) has a least solution
-as $\mathcal{L}(\cdot)$ is monotonic in $\sigma$; secondly that
-({\sc live-define})
-resembles the rule for type inference of mutually recursive function definitions,
-and thirdly the asymmetry of demand and liveness (compared to post- and pre-liveness
-classically) is due to the functional formulation here.
-
-Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
-(representing a fully context-sensitive analysis)
-can be safely approximated, for each function, by a {\em procedure summary}
-(unifying the contexts in the style of 0-CFA)\@.
-The summary consists of a pair of a single demand and, for this demand,
-the corresponding tuple of demands the function makes on its arguments.
-
-%%% save two lines to bring up section 3.2.
-
-% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
-% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
-% enables us to do a context-dependent analysis of a call without having
-% to enter the body of the called function.
-
-%% We present  our liveness  analysis using judgement forms $\len$
-%% for both expressions and programs.
-%% These are given formally in Figure~\ref{fig:live-judge}, but
-%% first we an intuitive understanding of the issues involved.
-%% Consider the expression:
-%% 
-%% \begin{center}
-%% \begin{uprogram}
-%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
-%%                         \px) \IN\  
-%%            % \UNL{0}
-%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
-%%                         \pw ) \IN \ldots) 
-%% \end{uprogram}
-%% \end{center}
-%% 
-%% 
-%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
-%% \pw.\epsilon\}$.
-%% Informally, this is so since \pw\ and \pz\ are used in
-%% (\PRIM~\pz~\pw ).
-%% 
-%% \begin{enumerate}
-%% \item The liveness  at $\pi$ is obtained by  removing the access paths
-%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
-%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
-%%     similar  to liveness  analysis of  imperative programs,  where the
-%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
-%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
-%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
-%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
-%%     liveness generated and  killed in $s$}. 
-%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
-%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
-%%   as a {\em  demand}.  The demand on an expression consists of  access
-%%   patterns  representing all  possible future  uses of  its  value.  A
-%%   demand  of $\epsilon$  means that  only the  reference resulting
-%% %  from the evaluation  of
-%%   from  $(\CAR\,\,\px)$ will be used.
-%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
-%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
-%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
-%% \end{enumerate}
-%% 
-%% In general,  an access path is  live before an  expression because the
-%% link represented by it is used (i) while evaluating the expression, or
-%% (ii) while using the result of the expression, or (iii) simply because
-%% the link  is live at  a later program  point.  When the  expression in
-%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
-%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
-%% the access  path \pw.   Note that \px.\acar\  is live in  this example
-%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
-%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
-%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
-%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
-%%
-
-% \paragraph{Judgements formally.}
-% Let  $\Lv$  denote a  liveness  environment (a function from
-% variables to access paths, abused using grammar notation) and
-% $\sigma$ be a demand.
-% The judgement form $e, \sigma \len \Lv$,
-% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
-% environment  before $e$  is $\Lv$ and
-% is thus analogous to the classical notion of {\em transfer function}.
-
-%    It is  interesting to  note that,
-% unlike traditional liveness analysis,  our judgements do not require a
-% liveness after the expression. This  is because of the coming together
-% of  two reasons.   First, it  will be  clear while  discussing garbage
-% collection,  that it is  enough for  our purposes  to assume  that the
-% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
-% restriction of our  programs to ANF ensures that  the post-liveness of
-% each  expression  $e$  inside   the  function  body  continues  to  be
-% $\emptyset$. Thus the post-liveness can be dropped altogether.
-
-%==============================================================
-\begin{figure}[t]  
-% \footnotesize
-\begin{eqnarray*}
-\mathit{ref\/}(\kappa,\sigma,\Lfonly)
-          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
-          &=& \{x.\sigma\} \\
-\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
-          &=& \{x.\epsilon,y.\epsilon\} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \\
-\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
-        \mathcal{L}(e_1,\sigma,\Lfonly) \cup
-        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
-        \{x.\epsilon\} \\
-\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
-        \Lv
-           \setminus \{x.*\}
-           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
-\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
-\end{eqnarray*}
-
-\begin{minipage}{0.85\textwidth}
-\infrule[live-define]
-        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
-           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \len \Lfonly
-\\ \makebox[0mm]{where
-     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
-\end{minipage}
-
-\normalsize
-  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%==============================================================
-
-% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
-% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
-% liveness generated  before the right-hand-side of  a \LET\ definition.
-% Constants    do   not   generate    any   liveness.     The   operator
-% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
-% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
-% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
-% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
-% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-% that \CONS\ does not, by itself, dereference its arguments.
-% 
-% 
-% Unlike primitives, the mapping from the demand on a user-function call
-% to the  demands on its arguments  has to be derived  from the function
-% body itself---as a fixed point for a recursive function.
-% The function  \Lfonly\ captures this mapping for
-% all functions  defined in the  program.  For a  function $\mathit{f}$,
-% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
-% a    call     to    the     $i$th    argument.     The     rule
-% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
-% confirmed by the body of all the functions, then the guess is correct.
-% {\color{blue}Note the similarity of this rule to type-inference
-% for recursive program schemes.}
-
-% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
-% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
-% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
-% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
-% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
-% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
-% liveness generated before $s$.
-
-\subsection{Minefield semantics and correctness}
-\label{sec:minefield}
-\label{sec:correctness}
-
-This section gives a modified semantics
-%(a reference semantics)
-which checks liveness annotations
-at run time, and `explodes' when these are found to be inconsistent with execution behavior,
-but otherwise behaves as the standard semantics.
-We show that such explosions never occur and hence run-time checks
-can be elided.
-%% We first assume some demand transformers $\Lfonly$;
-%% our proof will rely on this being a solution (i.e.\ satisfies $\mathit{Prog} \len \Lfonly$)
-%% of liveness analysis, but we do rely in this yet.
-We first assume an arbitrary  demand transformers $\Lfonly$ which satisfies
-$\mathit{Prog} \len \Lfonly$.
-We then enrich the abstract machine state $\rho, \stk, \heap, e$ to
-$\rho, \stk, \heap, e, \sigma, \Sigma$.
-Here $\sigma$ is the demand to the currently active function,
-thus the liveness $\Lv$ at $e$ is $\mathcal{L}(e,\sigma,\Lfonly)$,\footnote{
-This a simple liveness propagation using $\mathcal{L}(\cdots)$
-and $\mathit{ref}(\cdots)$ as $\Lfonly$ is assumed given.}
-and $\Sigma$ is a stack of demands---one for each function frame
-pushed in $\stk$.
-
-Second, we augment $\mathit{Val}$ with a value $\bot$.
-To model strong liveness $\bot$ may be copied freely,
-including into a $\CONS$ cell,
-% but reduces the state to $\bang$
-but explodes
-when used computationally (in a primitive operation other than a copy).
-Additionally we define
-$\mathit{GC}(\Lv,\Sigma): (\rho, \heap, \stk) \mapsto (\rho', \heap', \stk')$
-which determines {\em live-reachability}\footnote{Reachability curtailed by liveness information.}
-%%%%% \marginpar{Say more about live-reachability?}
-using $\rho$ and the $\rho$'s in $\stk$ as the root set
-and following links in $\heap$ {\em only as far as allowed by $\Lv$ and $\Sigma$}.
-% The effect of $\mathit{GC}(\Lv,\Sigma)$ is to replace
-Hence $\mathit{GC}(\Lv,\Sigma)$ replaces
-live-unreachable values---in $\rho$, in $\heap$ and the $\rho$'s in \stk---with $\bot$.
-For example, if $x.\epsilon \not\in \Lv$ then $\rho'(x) = \bot$.
-%, regardless of $\rho(x)$.
-Only $\mathit{GC}(\cdots)$ introduces $\bot$.
-
-Third, we update the semantics in four ways: ({\em i\/}) we arrange that all
-($\rightarrow$) transitions on expressions $e$ first use $\mathit{GC}(\cdots)$
-to update the state and then continue as before;
-and ({\em ii\/}) whenever the value $\bot$ is used computationally in a reduction
-($\rightsquigarrow$), we enter a distinguished stuck state $\bang$.
-For example, supposing $\rho(x)=\bot$
-then
-$\rho, \heap, s \rightsquigarrow \bang$ if $s$ is
-$(\CAR\,\, x)$, $(\CDR\,\, x)$ or $(\PRIM\,\, x\ y)$,
-but not if $s$ is $(\ID~x)$, $(\CONS~x~y)$ or $(f~x~y)$.
-Finally ({\em iii\/}) we make a similar change to the ($\rightarrow$) reduction for ({\sc ord-if})
-and ({\em iv\/}) augment the ({\sc ord-let}) rule for primitives to propagate
-$\bang$ from ($\rightsquigarrow$) to ($\rightarrow$).
-%%%%%% {\color{blue}The full minefield semantics is given as an appendix?}
-
-% Suppose liveness for the current $e$ under reduction is $\Lv$; then:
-% For every variable $x$ for which $x.\epsilon \not\in \Lv$ then replace
-% $\rho(x)$ with $\bot$.
-% Also, for every $\ell \in \mathit{Loc}$, replace the $\CAR$ field
-% of $\heap(\ell)$ with $\bot$ when
-% which there is no path $x.\alpha\acar$ in $\Lv$ such that following
-% $\alpha$-links in $\heap$ from $\rho(x)$ reaches $\ell$.
-% Correspondingly, replace the $\CDR$ field of $\ell$ if there
-% is no path $x.\alpha\acdr$ in $\Lv$.
-% \item
-% 
-% {
-% \footnotesize
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-car]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CAR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-cdr]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CDR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}\\[1ex]
-% 
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-prim]
-%       {\rho(x) = \bot \vee \rho(y) = \bot}
-%       {\rho, \heap, {(\PRIM\,\, x\ y)}
-%         \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-null]
-%         {\rho(x)= \bot
-%         }{\rho, \heap, {(\NULLQ\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% }
-%
-
-The resulting minefield semantics behaves identically
-(identical heap, identical steps, including possible non-termination)
-to the standard semantics, except for one possibility---the
-minefield semantics going $\bang$ while the standard semantics
-continues (either to a halt state, to a stuck state, or reduces forever).
-
- We now prove a result that relates
-liveness analysis to the semantics.
-\begin{proposition}
-Given program $P$ with $P \len \Lfonly$, then in the minefield
- semantics $P \rightarrow \bang$ can never occur
- (cf.\ `well-typed programs do not go wrong').
-\end{proposition}
-\textit{Proof outline:}
-Space does not permit a full proof, but we give the two main steps.
-We proceed by contradiction and assume there is a program $P$ for which
-$P \len \Lfonly$ can enter state $\bang$.
-The first step is to construct a program $P'$ with identical behavior, but with
-no user-function definitions, using inlining.
-This is possible because a program which goes $\bang$ does so after a finite number
-of reductions, and hence even recursive functions have only had a finite number of
-invocations.
-We hence repeatedly inline-expand user-function calls in $P$
-until we obtain a program $P'$ which behaves identically\footnote{
-Modulo replacement of
-({\sc ord-call}) and
-({\sc ord-return}) steps with
-({\sc ord-id}) steps.
-}
-to $P$ in the standard semantics, but makes no user-function calls.
-Any remaining, non-executed, calls can 
-%be left or
-replaced with a new primitive with the same
-demand-to-liveness transfer function---thus making
-$P'$ a simple expression $e'$.
-%%%%%
-Not only do the program points in reducing $e'$ correspond
-one-one to states during evaluation of $P$,
-but also the liveness associated with a point
-in $e'$ is identical to the liveness at the corresponding state
-$(\rho, \stk, \heap, e, \sigma, \Sigma)$ of $P$ (concatenating the liveness
-$\mathcal{L}(e,\sigma,\Lfonly)$ at $e$
-with the liveness, obtained from $\Sigma$, of the call sites in $\stk$ and
-after renaming the variables correspondingly to the inlining which produced $e'$).
-This assertion relies on the analysis being fully context-sensitive, and noting that
-while the change of scope caused by inlining changes variable visibility
-between $P$ and $e'$ it does not change the liveness---as local-to-a-function
-$\LET$-variables whose scope has been prolonged due to inlining
-are dead in the prolonged scope.
-
-The second step of the proof is to show that $e'$ cannot go $\bang$.
-We proceed by induction.
-Correctness of the $\SIF$ and final $\RETURN$ forms are immediate;
-the $(\LET~z\leftarrow s~\IN~e)$ form requires showing
-the inference rules ensure that any value referenced via $z$ in $e$
-was already live-reachable (via another variable and path) in any
-enclosing expression (so that $\mathit{GC}(\cdots)$ could not re-write it to $\bot$).
-
-% \newpage
-
-\section{Computing liveness and its encoding as a table}\label{sec:computing}
-
-Section~\ref{sec:liveness} gave a context-sensitive liveness analysis
-and proved it correct with reference to a {\em minefield} semantics.
-% Practical computation was not considered.
-For practical use we need to solve the liveness equations {\em finitely and
-symbolically}.  As expressed mathematically, and given a fixed program,
-three things are potentially unbounded:
-({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
-({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
-({\em iii\/}) the number of such access paths.
-
-We commonly first solve ({\em i\/}) by reducing the number of distinct
-calling contexts
-(e.g.\ to a single unified context in 0-CFA style).
-However, it turns out that we can solve the equations for $\Lfonly$ symbolically
-without this reduction, so here we defer this to Section~\ref{sec:bodylivenessbodies}.
-(It also allows easier extension to  dynamically determined liveness---future work.)
-%
-We address ({\em ii\/}) and ({\em iii\/}) by 
-% attempting to re-interpret
-re-interpreting
-the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
-rather than a mutually recursive set of equations on sets of access paths.
-
-This requires two ideas.  Firstly we have to control the use of functions---they
-tend to be infinitary and do not occur naturally in CFGs; in particular our
-$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
-to a tuple of access paths.  The former is achieved by using a separate
-meta-variable (later non-terminal)
-$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
-($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
-Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
-expressible finitely (it is a linear form).
-
-Secondly, there are also two technical issues in re-interpreting
-Figure~\ref{fig:live-judge} as a grammar.
-One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
-disappears after variables are separated ($\Lanv{i}{x}$ above). The other
-is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
-not a CFG\@.
-Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.
-
-Finally,  Section~\ref{sec:NFA-approx}
-uses a construction due to
-Mohri and Nederhof~\cite{mohri00regular} to over-approximate
-context-free grammars with regular grammars;
-these are more appropriate for run-time use.
-%
-Hence the overall `big picture' view is that each GC point
-is annotated with a table encoding the DFA for that program point.
-Such program points forming the call string causing the garbage collection
-(and ending in $\CONS$)
-are used by the garbage collector to curtail its reachability-marking phase
-to this over-approximated liveness.
-
-\subsubsection{Modifying the \CONS\ rule.}
-
-The    $\mathit{ref}$    rule for \CONS,    shown    in
-Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
-\acar\  and  \acdr\  from  access paths in  $\sigma$.
-  Mathematically  this  is fine  but causes problems  when
-  solving  the liveness equations  {\em symbolically}  since such
-  decomposition cannot be expressed as a context-free grammar.  
-To  handle this,  we introduce  two  new symbols  \bcar\  and
-\bcdr\ with the properties:
-\[
- \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
-\hspace{3em} \mbox{and} \hspace{3em}
- \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
-\]
-We can now rewrite the \CONS\ rule as:
-\begin{eqnarray*}
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
-\end{eqnarray*}
-
-We  call the liveness equations with  this modification
-  \Lfunonly$'$.
-The  defintions of \bcar\  and  \bcdr\  induce   the  following
-relation $\hookrightarrow$ over sets of access paths:
-\begin{eqnarray*}
-  \sigma_1\bcar\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
-& \qquad\mbox{and}\qquad &
-\sigma_1\bcdr\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
-\end{eqnarray*}
-%% \noindent \begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcar\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acar\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acdr\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% %\label{eqn:eup-red1}
-%% \end{eqnarray*}
-%% \end{minipage}
-%% \hspace*{.65cm}\begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcdr\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acdr\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acar\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% \end{eqnarray*}
-%% \end{minipage}\\
-
-The reflexive transitive closure of $\hookrightarrow$ will be denoted
-as $\stackrel{*}{\hookrightarrow}$. The following proposition relates \Lfunonly\   and \Lfunonly$'$:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]  
-\begin{boxedminipage}{\textwidth}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\append~\lista~\listb)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \xtest\ $\leftarrow $\ (\NULLQ~\lista) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\SIF\ \xtest~$\pi_3\!\!:$(\RETURN\ \listb)  
-          \UNL{4} \hspace*{.05cm}     $\pi_4\!\!:\, $(\LET\ \xtl\  $\leftarrow$\   (\CDR\ \lista) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_5\!\!:\, $(\LET\ \xrec\  $\leftarrow$  (\append\ \ \xtl\ \  \listb) \IN
-	  \UNL{6} \hspace*{.05cm}    $\pi_6\!\!:\, $(\LET\ \xhd\  $\leftarrow$  (\CAR\ \lista)  \IN
-	  \UNL{7}  \hspace*{.05cm}   $\pi_7\!\!:\,
-          $(\LET\ \xans\  $\leftarrow$ (\CONS\ \ \xhd\ \ \xrec)  \IN\
-          $\pi_8\!\!:\, $(\RETURN\ \xans))))))))
-	\end{uprogram}}
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pz\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}              (\LET\ \py\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_9\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\append\ \py\ \pz) \IN
-          \UNL{4}   \hspace*{.78cm}  $\pi_{10}\!\!:\, $  (\LET\ \pa\  $\leftarrow$  (\CDR\ \pw) \IN
-	  \UNL{5}   \hspace*{.78cm}  $\pi_{11}\!\!:\, $
-          (\LET\ \pb\  $\leftarrow$ (\CAR\  \pa) \IN~$\pi_{12}\!\!:\,$ (\RETURN\ \pb)))))))
-\end{uprogram}}
-\end{boxedminipage}
-  \caption{An example program.}
-\label{fig:examp-pgm}
-\figrule
-\normalsize
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{proposition}
-Assume that  a liveness computation  based on  \Lfunonly\   gives  the liveness of
-the variable $x$ at a program point $\pi_i$ as $\sigma$ (symbolically, $\Lanv{i}{x}=
-\sigma$).  Further,  let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is
-used instead of \Lfunonly.   Then  $\sigma' \stackrel{*}{\hookrightarrow}
-\sigma$.
-\end{proposition}
-
-To see  why the proposition is  true, consider an analysis  based on
-\Lfunonly$'$ in which   $\sigma$ appears in
-the  context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.  Let $\alpha  \in \sigma$.  The
-symbol  \bcar(\bcdr)  merely  marks  a  place in  $\alpha$  where  the
-original  \CONS\  rule  would  have erased  an  immediately  following
-\acar(\acdr),  or, in  absence of  such a  symbol, would  have dropped
-$\alpha$ itself.  Since the application  of any rule in \Lfunonly$'$ merely
-adds  symbols at  the beginning  of  $\alpha$, the  markers and  other
-symbols in $\alpha$ are propagated to other dependent parts of program
-in their same relative positions.  Consequently, the erasure carried out at the end of the analysis with
-$\stackrel{*}{\hookrightarrow}$ gives the same result as obtained
-through \Lfunonly.
-
-\subsection{Generating equations for the demand transformer \Lfonly}
-\label{sec:Lfonlysymbolically}
-
-%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
-We  shall  consider the  program  in  Figure~\ref{fig:examp-pgm} as  a
-running example;
-this is the  ANF version of the program in Figure~\ref{fig:mot-example}.
-
-%%AM no longer used: In what follows,  we shall overload the semantics of
-%% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-%mean the generation
-%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
-%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.
-
-To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
-$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
-{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
-$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
-formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
-of $x_i$.  For our example program which has  a
-single function \append, this generates the following equations:
-% $\Lfonly_{\append}$  is defined by equations:
-  \begin{eqnarray*}
-    \Lf{\append}{1}{\sigma}
-    &=& \epsilonset \cup \acar\bcar\sigma
-    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
-    \Lf{\append}{2}{\sigma}
-    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
-  \end{eqnarray*}
-  In  general, the  equations  for \Lfonly\  are recursive  since
-  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
-  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
-    is  similar to  solving the  differential equation  
-    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
-    that    the    solution     has    the    form    $\mathit{y}
-    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
-    equation  yields  a  quadratic  equation  in  $r$,  and  each
-    solution of $r$ gives rise  to a solution of the differential
-    equation  (in our  setup we  can effectively  pick  the least
-    solution rather than needing linear combinations).  }
-\begin{eqnarray}
-\label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
-\end{eqnarray}
-where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
-alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
-why \Lfonly\ has this form  is as follows: Recall that
-\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
-the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
-situation in which  the demand $\sigma$ does not  reach the $i$th
-argument.   This happens,  for  instance, when  argument is  used
-only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
-transformation of the demand $\sigma$ before it reaches the $i$th
-argument. This  transformation prefixes strings in 
-$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.
-
-Substituting  the  guessed   form  in  the  equations  describing
-\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
-containing $\sigma$ and those  without $\sigma$ on both sides of the
-resulting equation, we
-get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
-% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
-$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
-solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
-  \append}$, we get:
-  \begin{align*}
-    \Uf{\append}{1} &= \epsilonset
-    \cup \acdr.\Uf{\append}{1}  
-    &\quad 
-    \Uf{\append}{2} &= \Uf{\append}{2} \\
-    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
-    \cup \acdr\Df{\append}{1}\bcdr &\quad 
-    \Df{\append}{2} &= \epsilonset \cup
-    \Df{\append}{2}\bcdr 
-  \end{align*}
-Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.
-
-\subsection{Generating liveness equations \Lv\  for function bodies}
-\label{sec:bodylivenessbodies}
-
-\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
-
-We now calculate a 0-CFA-style summary liveness for each GC point of a program.
-There are two parts to this.
-First, for each function $f$,  we determine a {\em summary demand} $\sigma_f$
-over-approximating any demand $\sigma$ passed to $f$.
-%$\Lf{f}{i}{\sigma}$.
-So, consider a function $g$ containing a call site $\pi$ to $f$, say
-$\pi\!\!:\!(\LET~x \leftarrow (f\,y_1\,\ldots\,y_n)~\IN~e)$.
-Under the assumption that the demand on $g$ is $\sigma_g$, the liveness at $e$
-is $\Lv = \mathcal{L}(e,\sigma_g,\Lfonly)$, and the $\LET$ case of
-Figure~\ref{fig:live-judge} tells us this call site contributes $\Lv(x)$ to the
-demand $\sigma_f$ placed on  $f$.
-Write $\deltacall{f}{\pi}{g}$ for
-this demand (contributed by call site $\pi$ in $g$).
-
-Thus, supposing the $k$ call sites to function $f$ are
-$\pi^1$ (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
-%%% \marginpar{and writing $\sigma_f^i$ for the above $\deltacall{f}{\pi^i}{g^i}$?}
-then the over-approximation requirement on $\sigma_f$
-is achieved by taking $\sigma_f = \deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
-% is achieved by taking $\sigma_f = \sigma_f^1 \cup \cdots \cup \sigma_f^k$.
-
-The expression $e_{\mainpgm}$ is a special case; we assume it may be called externally
-with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$ (denoted
-$\sigma_{\!all}$).
-This is because any  part  of its  value  may be  used  by  the environment---for
-printing  the  result,  for  instance.
-
-% In what follows,  we shall overload the semantics of
-% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-% 
-% 
-% We assume an overall demand of $\{\acar,\acdr\}^*$ (denoted
-% $\sigma_{\!all}$) on the main expression $e_\mainpgm$.  This is because 
-% any  part  of its  value  may be  used  by  the environment---for
-% printing  the  result,  for  instance.
-
-%% To  generate  equations for  the  body $e_{\mathit{f}}$ of  a
-%% function,  we evaluate \Lfun{e_{\mathit{f}}}{\sigma_{\mathit{f}}}{\Lfonly}.     For ${\sigma_{\mathit{f}}}$,  we have
-%% to consider all calling  contexts during execution.  We therefore
-%% do a 0-CFA approximation by  defining
-%% % $\sigma_\mathit{f}$        as $\bigcup_\calls{f}{k}\!\!\sigma_\calls{f}{k}$,
-%% $\sigma_\mathit{f}$        as $\sigma_\calls{f}{1} \cup \cdots \cup \sigma_\calls{f}{n}$,
-%% where $\calls{f}{k}$,
-%% $\mathit{k} = 1 \ldots n$ represent all the calls to $\mathit{f}$
-%% in  the  program,   and  $\sigma_\calls{f}{k}$  are  the  demands
-%% generated at the calls  by the \rn{let-live} rule while analyzing
-%% the bodies of the caller.
-
-For the running example, $\append$ has calls from $\mainpgm$ at $\pi_9$
-and a recursive call at $\pi_5$.
-So $\sigma_{\append} =
-     \deltacall{\append}{\pi_9}{\mainpgm}  \cup \deltacall{\append}{\pi_5}{\append}$.
-Filling in  the values gives:
-%  where the former is the demand generated at the
-%  call to \append\  in $e_\mainpgm$ the latter
-%  is  the demand  at the  recursive call. 
-%% Propagation of
-%%  $\sigma_{\!all}$ in $e_\mainpgm$ gives    $\sigma_\calls{\append}{1}$   as   $\{\epsilon,   \acdr\}   \cup
-%%  \acdr\acar\sigma_{\!all}$  and propagation of
-%%  $\sigma_{\mathit{\append}}$ in $e_\append$ gives   $\sigma_\calls{\append}{2}$   as
-%%  $\bcdr\sigma_{\mathit{\append}}$.  
-%% Filling in  the values of $\sigma_\calls{\append}{1}$ and
-%% $\sigma_\calls{\append}{2}$, we get:
-\begin{align*}
-\sigma_{\append}    &=
-\left( \{\epsilon,  \acdr\} \cup  \acdr\acar\sigma_{\!all}  \right) ~\cup~
-\bcdr\sigma_{\mathit{\append}}
-\end{align*}
-
-Second, for each function $f$ (possibly $\mainpgm$) we need the liveness at
-each contained GC point
-$\pi\!\!:\!(\LET~x \leftarrow s~\IN~e)$,
-where $s$ is a call either to $\CONS$ or to a user function $g$.
-Given $\sigma_f$ calculated above, this
-is simply $\mathcal{L}(\pi,\sigma_f,\Lfonly)$.
-%%% \marginpar{AM: we've always said this (safe), but actual parameters
-%%% live before the call and dead after needn't be included!}
-
-For the running example, containing GC points $\pi_5, \pi_7$ in $\append$ and
-$\pi_9$ in $e_\mainpgm$, this gives
-(recall Equation~(\ref{eq:LF:DI}) above states $\Lf{f}{i}{\sigma} = \Uf{f}{i} \cup \Df{f}{i}.\sigma$):
-\begin{align*}
-\Lanv{5}{\xtl} &= \Lf{\append}{1}{\bcdr\sigma_{\append}}  \quad &
-                                \Lanv{7}{\xhd}\,\,\, &=  \bcar\sigma_{\append}     \\ 
-\Lanv{5}{\listb} &= \Lf{\append}{2}{\bcdr\sigma_{\append}} &
-                                \Lanv{7}{\xrec} &=  \bcdr\sigma_{\append} \\
-\Lanv{5}{\lista} &= \epsilonset \cup \acar\bcar\sigma_{\append} 
-\\[1ex]
-  \Lanv{9}{\py}  &= \Lf{\append}{1}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}} \\
-  \Lanv{9}{\pz}  &= \Lf{\append}{2}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}}
-\end{align*}
-
-\noindent 
-In summary, the equations generated during liveness analysis are:
-\begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
-\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
-% {\color{red}
-%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
-%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
-% }
-\end{enumerate}
-
-\subsection{Solving liveness equations---the grammar interpretation}
-The  equations above can now be re-interpreted as a
-context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
-a variable  $X$ occurring on  the LHS of the  equations generated
-from the analysis.  We can  think of the resulting productions as
-% being a grammar with multiple start symbols---one for
-% for each nonterminal \var{\Lanv{i}{x}}.
-being associated with several  grammars, one for each non-terminal
-\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
-productions:
- \begin{eqnarray*}
-\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
-  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
- \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
-  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
-  \acdr\var{\Uf{\append}{1}}\\         
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar               \mid
-  \acdr\var{\Df{\append}{1}}\bcdr\\
-\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
-\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
-\var{$\sigma_{\mathit{\!all}}$}
-\end{eqnarray*}
-Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
-denoted $\mathscr{L}(\var{\Lanv{i}{x}})
-$,  is the desired  solution of
-\Lanv{i}{x}.   
-However, recall from  our earlier  discussion that
-the  decision problem that  we are  interested in  during garbage
-collection is:
-\begin{quote}
-Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
-\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
-\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
-of forward paths only. Then does $\alpha \in \sigma$?
-\end{quote}
-We could convert  the  rules  defining  $\hookrightarrow$  into
-productions and  add them to the grammar. However, this results
-in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is  undecidable.   We
-circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
-  strongly  regular}  CFGs which  have   easy
-translations to non-deterministic finite state automata (NFA). The
-NFAs are then simplified on the
-basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
-membership of forward access paths. The resulting NFAs are   finally
-converted to DFAs for use during garbage collection.
-%------------------------------------------------------------%
-\subsubsection{Approximating CFGs using NFAs.}
-\label{sec:NFA-approx}
-
-We use the   
-algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
-approximate a CFG  to a   {\em strongly regular\/} grammar.  The
-transformation has  the  property that  if $\Lv$  is a
-non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
-Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
-  L}_{G'}(\Lv)$.    This  is   required  for the
-approximation to be safe with respect to liveness.
-
-We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
-transformation. We  pick the only production that  is affected by
-the transformation---the production for \Df{\append}{1}.  The
-production for \Uf{\append}{1}, while  recursive, is already in
-strongly  regular  form  and   is  therefore  unaffected  by  the
-transformation.
- \begin{eqnarray*}
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
-  \acdr\var{\Df{\append}{1}} \\ 
-\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
-\mid \epsilon 
-\end{eqnarray*}
-The  languages
-generated for  \var{\Df{\append}{1}} in  the original grammar  and the
-new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
-\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
-
-%------------------------------------------------------------%
-\subsubsection{Transforming NFAs to Accept Forward Paths:}
-%\label{sec:NFA-elim}
-The strongly regular CFGs obtained after the Mohri-Nederhof
-transformation are
-  first converted  into  NFAs.  The        algorithm         described        in
-Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
-\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
-Thus  \nt\  can  be  used  to  check  
-membership of forward paths.
-
-\begin{figure}[t]
-\begin{boxedminipage}{\textwidth}
-\begin{center}
-  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
-  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
-%
-  \raggedright{\bf Output:} NFA  \nt\ with  underlying
-  alphabet $\lbrace \acar, \acdr \rbrace$ such that
-  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
-  \\
-%
-
-  \raggedright{\bf Steps:}
- \begin{algorithmic}
-   \STATE $i \leftarrow 0$ 
-   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-   without $\epsilon$-moves \cite{hopcraft90toc}
-
-   \REPEAT
-   \STATE $\nt'_{i+1} \leftarrow \nt_i$
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcar$ and
-     outgoing edge to $q''$ with label $\acar$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-   label $\epsilon$.
-   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-   \ENDFOR
-
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcdr$ and outgoing edge to
-     $q''$ with label $\acdr$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-     label $\epsilon$.
-   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-   \ENDFOR
-
-   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-   $\nt'_{i+1}$ without $\epsilon$-moves
-
-   \STATE $i \leftarrow i+1$
-   \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-   \STATE $\nt \leftarrow \nt_i$
-%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
- \end{algorithmic}
-\end{center}
-\end{boxedminipage}
- \caption{Algorithm for transforming an NFA to  accept forward
-   paths only.}\label{algo:simplify-NFA} \figrule
-\end{figure}
-%------------------------------------------------------------%
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
-pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
-process is  continued till  a fixed point  is reached.   When the
-fixed point  is reached, the resulting NFA  contains all possible
-reductions corresponding  to all the  paths in the  original NFA.
-The proofs  of the termination  and correctness of  the algorithm
-are given in our earlier paper~\cite{karkare07liveness}.
-
-\input{liveness-dfa.tex}
-
-We illustrate the algorithm in Figure~\ref{algo:simplify-NFA} by
-constructing the automaton for \var{\Lanv{9}{y}}.
-Figure~\ref{fig:automata-construction} (c) shows the automaton for
-\var{\Lanv{9}{y}} constructed by composing the automata for
-\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
-\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
-get (d). We add an $\epsilon$ edge from $q_2$ to $q_3$ bypassing the
-\bcdr\acdr\ pair from $q_2$ to $q_2$ and then to $q_3$. This is shown
-in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
-is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
-$q_2$.  Removing this $\epsilon$ edge gives the automaton shown in
-(g).  Restricting this automaton to forward edges only, we get the
-final automaton shown in (h). This automaton recognizes $\acdr^{*}\!\mid\acdr^{*}\acar \sigma_{\mathit \!all}$, showing that the entire
-list \py, including its elements, is live at $\pi_9$. Also note that
-the language accepted by the final automaton satisfies the
-prefix-closed property.
-
-\section{Prototype and evaluation}\label{sec:experiments}
-To demonstrate the effectiveness of liveness-based garbage collection,
-we  have  built a  prototype  consisting  of  an interpreter  for  our
-language,  a  liveness  analyzer  and  a copying  collector  that  can
-optionally use the results of liveness analysis for marking instead of
-reachability. When the collector uses liveness for marking, we call it
-a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-  reachability-based  collector}  (RGC).   The  copying  collector  is
-neither incremental nor generational.  As a consequence, any cell that
-becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.
-
-While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
-is invoked  at the program point   $\pi$,  it would  use  the
-liveness  DFAs at  $\pi$ itself for exploring the root set of the
-current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.
-
-Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-$\dfa{\pi}{x}$.    
-Considering   a   DFA   as   a   table,
-$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-\acar\      or     \acdr.       We      shall     also      write
-$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-indicating  whether there  is a  transition from  $q$ on
-$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-pointer to $x$.
-The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-$x$  in the  other semi-space  and returns  the new  address.  It
-hides  details such as  returning the  forwarding pointer  if the
-value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.
-
-The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-in the heap over time for RGC and LGC---here time is measured in terms
-of the number of cons cells allocated. In addition, they also show the
-number of  reachable cells and the  number of cells  that are actually
-live\footnote{This   is  statically   approximated  by   our  liveness
-  analysis}.  Since  the programs have  different memory requirements,
-we have tuned the size of heap for each program to ensure a reasonable
-number of  collections. An invocation  of RGC decreases the  number of
-cells  in heap  till  it touches  the  curve of  reachable cells.   An
-invocation of LGC decreases the number  of heap cells to no lower than
-the curve of live cells.
-
-To construct the  reachable and live curves, we  record for every cell
-its creation time (\CreateTime), its last use time (\UseTime), and the
-earliest time  when the  cell becomes unreachable  and can  be garbage
-collected (\CollTime).  For accurate recording of  \CollTime, we force
-frequent invocations  of a reachability based collector  in a separate
-run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
-  fibheap} increases  initially, comes  down before the  heap becomes gets
-full and then  continues to remains low.  This results in  a single garbage
-collection in both RGC and LGC.  The graphs also show the precision of
-our liveness analysis.  For  all programs  except  {\tt nperms} and
-{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.
-
-\subsection{Results}
-\begin{figure}[t] 
-\centering
-\begin{boxedminipage}{.75\textwidth}
-{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
-\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-  then} {\bf skip}\\ 
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
-\noindent \hspace*{.35cm}{\bf return} $y$
-\end{boxedminipage}
-\caption{Function       for       copying       a       root       set
-  variable.}\label{algo:LGC-main-loop} \figrule
-
-\end{figure}
-
-\begin{figure}[p]
-\begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
- &
-{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-\end{tabular}
- \caption{Memory  usage of  programs. The blue and the red curves
-  indicate the number of cons cells in the active semi-space for RGC and
-  LGC respectively. The black curve represents the number of reachable
-  cells and the grey curve represents the number of cells that are
-  actually live (of which liveness analysis does  a static
-  approximation). x-axis is the time measured in number of cons-cells
-  allocated. y-axis is the number of cons-cells.}
-\label{fig:memory-usage} \figrule
-\end{figure}
-The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-provides  statistics  regarding the  analysis  itself.  The number  of
-states and the analysis times  are within tolerable limits. Precision
-of analysis refers to the percentage of dead cells that is collected by
-LGC, averaged over all invocations.
-The second table shows garbage  collection statistics for RGC and LGC.
-LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to run
-in comparison with  RGC.
-
-There are a couple of  issues of concern.  The garbage collection time
-is larger in the case of LGC for some programs. The reason is that the
-cost of consulting the liveness DFA may outweigh the combined benefits
-of  fewer   garbage  collections   and  fewer  markings   per  garbage
-collection.   The  other issue  is  illustrated  by  the program  {\tt
-  lambda}    (memory   usage  shown   separately    in
-Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-in Figure~\ref{fig:experimental-results}, the  number of
-touched cells\footnote{These  are the  cells which are  visited during
-  marking phase, often  more than once due to  sharing.} in this
-example is much higher for
-LGC.  This  increase is due to  excessive sharing among  heap nodes in
-this program.  Note  that a node re-visited because  of sharing is not
-explored  any   further  during  a  RGC   collection.   However,  this
-curtailment cannot happen in  LGC because of the possibility that the
-node,  re-visited in a  different liveness  state, may  mark a  set of
-cells different from the earlier visit.
-\begin{figure}[t]
-\footnotesize
-
-\centering
-
-\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline
-{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-  nperm} &  {\sf fibheap} & {\sf knightstour} &
-{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-\hline
-\hline
-{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-\hline
-\end{tabular}}
-
-\centerline{(a)}
-
-\medskip
-
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
-  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-
-                            &   \multicolumn{2}{c|}{cells per GC}  
-                            &   \multicolumn{2}{c|}{cells per GC} 
-                            &   \multicolumn{2}{c|}{\#GCs} 
-                            &   \multicolumn{2}{c|}{MinHeap} 
-                            &   \multicolumn{2}{c|}{Avg. Drag} 
-                            &   \multicolumn{2}{c|}{(sec)} \\
-\cline{2-13}
-{Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
-%    {\sf  nboyer} & &  & &   & &  & & &  \\
-%    {\sf   circsim?} & &  & &   & & & & &   \\
- \hline
-\end{tabular}}
-
-\centerline{(b)}
-\vskip -2mm
-\caption{Experimental results comparing RGC and LGC. Table (a) gives
-  data related to liveness analysis, and (b) gives garbage collection
-  data. The average drag time is measured in allocated cons cells.}
-\label{fig:experimental-results}
-\normalsize
-\end{figure}
-
-\section{Collecting more garbage can never slow things down}
-\label{sec:lgc-always-better}
-Since garbage collection is  effectively asynchronous to the allocator
-thread, one might  worry as to how {\em  robust} our measurements are.
-For example,  while LGC would,  in general, collect more  garbage than
-RGC in the same heap state,  is it possible that on some programs, LGC
-might do  a larger number  of collections? We  prove a result  to show
-that   this  cannot   happen.    This  result   applies  to   classical
-mark-and-sweep and copying garbage  collectors and, we hypothesize, to
-generational collectors too.
-
-\begin{lemma}
-For the  same mutator,  a liveness-based collector  can never  do more
-garbage collections than a reachability-based collector.
-\end{lemma}
-
-\begin{proof}Assume, as before,  that time is measured in terms of the number of
-  cons cells allocated. Let us run  two copies of the mutator with the
-  two    different   collectors    but    with   synchronous    memory
-  allocations.   The   garbage    collections,   however,   happen
-  asynchronously.
-
-  To prove the lemma, it is enough to show the truth of the following 
-statement: After
- every  LGC invocation,  the count of LGC invocations   is  no greater
-than  RGC invocations.   The  base case  holds  since the  first
-invocations of both GCs happen at the same time.  Assume the statement
-to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-reachable cells, its heap would contain no more cells than RGC heap at
-the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-LGC, or LGC and RGC are both invoked next at the same time.  In either
-case, the statement holds after $n+1$ invocations of LGC.
-\end{proof}
-
-%% \begin{figure}[h]
-%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
-%% \caption{Memory  usage of  programs. The  solid  lines represents
-%%   number of  live cells and the  dashed lines show  the number of
-%%   reachable cells.  The red and blue  lines represent invocations
-%%   of RGC and  LGC---their heights indicate  the amount of memory
-%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
-%%       paper and are being used as placeholder}}}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-%%\clearpage
-
-\section{Related Work} Previous attempts to  increase  the space efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
-reallocation~\cite{jones89compile},                       deforestation
-techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
-methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
-
-\begin{figure}[t]
-\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
- \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
-\label{fig:memory-usage-lambda} \figrule
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-We have defined a notion of liveness on structured data;
-this generalizes classical liveness and strong liveness.
-We started with a general fully context-sensitive analysis which
-we proved correct with respect to a minefield semantics
-(this models the effect of garbage collection between every
-evaluation step).
-
-To avoid scalability issues (and to avoid performing part of
-the liveness computation at run time)
-we defined an 0-CFA version of this liveness analysis in which
-demands for function $f$ at all calling contexts are conflated into 
-a single demand $\sigma_f$.
-This enabled us to treat the liveness equations symbolically obtaining
-context-free grammars for liveness at each GC point (calls to user
-functions and to $\CONS$).
-These were then converted to DFAs for run-time consultation by the
-garbage collector. Experiments confirm the precision of the analysis.
-
-To obtain performance figures we compared a reachability-based garbage
-collector with a liveness-based  collector.  This showed a decrease in
-the  number  of  GCs,   more  garbage  collected  per  invocation.   A
-significant benefit of LGC is  that programs can run in smaller memory
-when compared to  RGC. This is potentially useful  in situations where
-memory  is  limited---as with  embedded  systems.  For  a majority  of
-programs, the garbage collection times were reduced.
-
-One issue we  highlighted was that while fewer  nodes were marked (and
-more garbage collected), sometimes  cons cells could be visited and
-traversed  multiple times  with different  sets of  liveness  paths to
-explore.  Further work includes  improvements to the classical copying
-collector to reduce the cost of this.
-
-\subsection*{Acknowledgements}
-The authors  would like  to thank Hemanshu  Vadehra for  a preliminary
-implementation of the prototype system. Amey Karkare was supported for
-this work by the DST/SERC fast track scheme for young scientists.
-
-\bibliography{fun_hra}{}
-\bibliographystyle{splncs}
+\input{sections/restOfOldPaper}
 
 \end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-\section{Appendix stuff}
-
-\paragraph{Safety of using the modified \CONS\  rule:}
-Recall that while the correctness of the analysis was proved with
-respect to  the $\len$ rules  in Figure~\ref{fig:live-judge}, the
-equations   for  liveness   are  based   on  a   modification  of
-$\mathit{ref}$    in    the    case    of   \CONS\  as  given    by
-(\ref{eqn:mod-cons}).  We now show that  if the \bcar\,s and \bcdr\,s
-in live access  paths resulting  from this  modification are
-eliminated through  applications of $\hookrightarrow$,  we obtain
-the same results as given by the original rules.
-
-\comment{Proposal: We prove a result which goes as follows:
-
-\begin{quote}
-Lemma: Let  $\Lanv{i}{x}= \sigma$ for an a  program point $\pi_i$
-and a variable  $x$, when the judgement is  based on the original
-$\len$  rules.   Further, let  $\Lanv{i}{x}=  \sigma'$, when  the
-judgement is based on $\len$ with the modified \CONS\ rule.  Then
-$\sigma' \stackrel{*}{\hookrightarrow} \sigma$
-\end{quote}
-
-Here are some  ideas regarding how this can  be proved.  Consider
-an arbitrary equation
-$$ \Lanv{j}{y} = \ldots \Lanv{i}{x} \ldots$$ generated by the
-analysis.
-
-Now assume that $\Lanv{i}{x}= \sigma$.   We want to find out what
-happens to the access patterns in $\sigma$ in the access patterns
-of \Lanv{j}{y}.   To answer this,  we put \# markers  around each
-$\alpha \in \sigma$  and redo the equation using  an extension of
-$\len$ to take  care of \#-tagged patterns.  We  can then observe
-occurrences of \#-tagged patterns  in the liveness information at
-$\pi_j$.
-
-Here is  the $\len$ extension for  the original \CONS\  clause to take
-care of the \#-tagged patterns:
-$$\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          = \{x.\#\alpha\# \mid \#\acar\alpha\# \in \sigma\} \cup
-\{y.\#\alpha\# \mid \#\acdr\alpha\# \in \sigma\} $$
-And here  is  the $\len$ extension for  the modified \CONS\  clause.
-$$ \mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-= \{x.\#\bcar\alpha\# \mid \#\alpha\# \in \sigma\} \cup
-\{y.\#\bcdr\alpha\# \mid \#\alpha\# \in \sigma\} $$
-All  other rules  remain unchanged.  Tracking a  pattern  through this
-extension is sound because erasing the \#-tags would give us the same
-liveness as the original rules.
-
-Let us also extend $\hookrightarrow$ to \#-tagged strings as:
-$$\sigma_1  {\hookrightarrow} \sigma_2\,\, \mathit{iff}\,\, \sigma_2 =
-\{\alpha_1 \#\alpha\# \alpha_2 \mid  \alpha_1 \#\alpha'\# \alpha2 \in
-  \sigma_1\,\, and \,\, \{\alpha'\} \stackrel{*}{\hookrightarrow} \{\alpha\}\}$$
-
-Now we have an  observation represented by the following diagram:
-
-\centerline{\epsfig{file=diag1.eps, height=4cm}}
-
-We need  to make this  observation general enough.   Therefore we
-should prove  the observation even if \Lanv{i}{x}  = $\sigma$ had
-patterns with \bcar\ and \bcdr.   We should be able to prove this
-by structural induction on the rules.
-
-How does  the observation prove  the original lemma?   Suppose we
-replayed  judgements  with  ($\len$  +  \#) and  with  ($\len$  +
-modified \CONS +  \#).  Suppose that $\pi_i$ is  the last program
-point till which the results  are the same (because no \CONS\ was
-encountered).  Now  consider a  $\sigma \in \Lanv{i}{x}$  for any
-$x$.  Put \#-tags around each string in $\sigma$. Now we have the
-diagram:
-
-\centerline{\epsfig{file=diag2.eps, height=7cm}}
-
-This shows that modulo erasure of \#-tags, using the original $\len$
-judgement and  $\len$ judgement with modified \CONS\ rule followed by
-$\hookrightarrow$ give the same liveness.
-
-And I would like  to conclude that it is all right  to do our analysis
-with $\len$ + modified \CONS\ and then simplify using $\hookrightarrow$.
-
-}
-
-{\color{blue}\marginpar{From sec 3.1, liveness judgements}
-\paragraph{Semantic considerations.}
-The incremental liveness for generic primitive $\PRIM$ is given by
-$\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly) = 
-           \{x.\epsilon,y.\epsilon\}$.
-Earlier we suggested that classical strong liveness for addition,
-but not division, could use the stronger
-$ \{x.\epsilon,y.\epsilon \mid \epsilon \in \sigma\}$.
-However this is invalid: garbage collection
-could reuse dead, but reachable, $\CONS$ cell fields (of $v$ below)
-thereby producing a stuck state in a dynamically typed language.
-Hence we treat $+$ as like division not like addition---suppose
-% $(FOO)$ is a user function which may cause garbage collection, and consider:
-calling $\mbox{\footnotesize (FOO)}$ garbage collects and reallocates the cell pointed
-to by `v' in:
-
-\vspace{-15pt}
-
-\begin{quote}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1}  (\LET\ v\ $\leftarrow $\ (\CONS~2~3) \IN
-	  \UNL{2}  (\LET\ w\ $\leftarrow $\ (FOO) \IN
-	  \UNL{3}  (\LET\ x\ $\leftarrow $\ (\PRIM\ (\CAR\ v) 4) \IN
-	  \UNL{4}  (\RETURN\ w))))
-	\end{uprogram}}
-\end{quote}
-In principle, {\em dead code removal} of
-$(\PRIM\ (\CAR\ {\rm v})\ 4)$
-would render the stronger version valid, but still requires static typing
-to avoid the risk of {\em removing} a stuck state.
-We restrict ourselves to liveness annotation rather than code transformation.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]
-\footnotesize
-Old version of the liveness rules...
-
-\infrule[ret-live]
-        {}
-        {(\RETURN~x),\sigma,\Lfonly \len  x.\sigma}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[if-live]
-        { e_1, \sigma,\Lfonly \len \Lv_1
-          \andalso e_2, \sigma,\Lfonly \len  \Lv_2}
-        {(\SIF~x~e_1~e_2),\sigma,\Lfonly \len
-               \Lv_1\cup\Lv_2\cup \{x.\epsilon\}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[let-live]
-        {e, \sigma,\Lfonly  \len  \Lv}
-        {(\LET~x \leftarrow  s~\IN~e),
-          \sigma,\Lfonly \len   \Lv\setminus \{x.*\} \cup
-          \mathit{ref\/}(s,\Lv(x),\Lfonly)}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\infrule[live-define]
-        {e_f, \sigma, \Lfonly  \len
-%            \bigcup \{z_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-           \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \fen \Lfonly
-\\ \mbox{where }
-     (\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ \ e_f) \mbox{ is a member of }
-        d_1 \ldots d_k}
-\figrule
-  \caption{Liveness Judgement Rules}\label{fig:live-judge-old}
-\normalsize
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \newgray{lightgray}{.75}
-%\newcommand{\dem}[1]{\red{\scriptsize{$#1$}}}
-\newcommand{\dem}[1]{\psframebox[linestyle=none,fillstyle=none,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-%\newcommand{\liv}[1]{\blue{\scriptsize{$#1$}}}
-\newcommand{\liv}[1]{\psframebox[linestyle=none,fillstyle=solid,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-\psset{unit=1mm,nodesep=1mm,labelsep=0.5mm,arrows=->,linewidth=.1,framesep=0.2mm}
-\begin{center}
-  \begin{pspicture}(0,0)(120,50)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(12,5)
- \footnotesize
-   \putnode{let1}{origin}{15}{50}{$\pi_1$:\LET}
-
-    \putnode{test1}{let1}{-12}{-7}{\xtest}
-    \putnode{null}{let1}{0}{-7}{$\pi_2$:\NULLQ}
-    \putnode{lista1}{null}{0}{-7}{$\pi_3$:\lista}
-    \putnode{if}{let1}{18}{-7}{$\pi_4$:\SIF}
-
-    \putnode{test2}{if}{-8}{-7}{$\pi_5$:\xtest}
-    \putnode{listb1}{if}{3}{-7}{$\pi_6$:\listb}
-    \putnode{let2}{if}{14}{-7}{$\pi_7$:\LET}
-
-    \putnode{tl1}{let2}{-22}{-7}{\xtl}
-    \putnode{cdr}{let2}{-5}{-7}{$\pi_8$:\CDR}
-    \putnode{lista2}{cdr}{0}{-7}{$\pi_9$:\lista}
-    \putnode{let3}{let2}{15}{-7}{$\pi_{10}$:\LET}
-
-\putnode{rec1}{let3}{-13}{-7}{\xrec}
-\putnode{append}{let3}{0}{-7}{$\pi_{11}$:\append}
-\putnode{tl2}{append}{-10}{-7}{$\pi_{12}$:\xtl}
-\putnode{listb2}{append}{5}{-7}{$\pi_{13}$:\listb}
-
-\putnode{let4}{let3}{17}{-7}{$\pi_{14}$:\LET}
-\putnode{hd1}{let4}{-5}{-7}{\xhd}
-\putnode{car}{let4}{4}{-7}{$\pi_{15}$:\CAR}
-\putnode{lista3}{car}{0}{-7}{$\pi_{16}$:\lista}
-
-\putnode{let5}{let4}{16}{-7}{$\pi_{17}$:\LET}
-
-\putnode{ans1}{let5}{-4}{-7}{\xans}
-\putnode{cons}{let5}{6}{-7}{$\pi_{18}$:\CONS}
-\putnode{hd2}{cons}{-10}{-7}{$\pi_{19}$:\xhd}
-\putnode{rec2}{cons}{10}{-7}{$\pi_{20}$:\xrec}
-\putnode{ans2}{let5}{19}{-7}{$\pi_{21}$:\xans}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-
-%% \putnode{x}{let1}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{if}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{let3}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let4}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{-8}{0}{\liv{\Lv_1}}
-%% %\putnode{x}{ans2}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{plus}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{rec2}{1}{4}{\dem{\epsilonset}}
-%% \putnode{x}{plus}{-7}{0}{\liv{\Lv_6}}
-%% %\putnode{x}{let4}{-7}{0}{\liv{\{\xrec\}}}
-%% \putnode{x}{len}{3}{3}{\dem{\epsilonset}}
-%% \putnode{x}{len}{-7}{0}{\liv{\Lv_3}}
-%% \putnode{x}{tl2}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{tl2}{-6}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let3}{-10}{0}{\liv{\xtl.\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l2}{0}{4}{\dem{\epsilonset \cup \acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{l2}{-5}{0}{\liv{\Lv_4}}
-%% \putnode{x}{zero}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{zero}{-3}{0}{\liv{\Lv_0}}
-%% \putnode{x}{test2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{test2}{-7}{0}{\liv{\Lv_5}}
-%% \putnode{x}{if}{-6}{0}{\liv{\Lv_3}}
-%% \putnode{x}{null}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{null}{-9}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l1}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{vone2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{one}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{let5}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{l1}{-5}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{let1}{-17}{-1}{\liv{\{\xl\} \cup \xl.\acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{let1}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let1}{-7}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{if}{7}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let2}{8}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let3}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let3}{-7}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let4}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let4}{-7}{0}{\liv{\Lv_2}}
-%% \putnode{x}{vone2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{rec2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{let5}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{one}{-6}{0}{\liv{\Lv_2}}
-\ncline{-}{let1}{test1}
-\ncline{-}{let1}{null}
-\ncline{-}{let1}{if}
-\ncline{-}{null}{lista1}
-
-\ncline{-}{if}{test2}
-\ncline{-}{if}{listb1}
-\ncline{-}{if}{let2}
-\ncline{-}{let2}{tl1}
-\ncline{-}{let2}{cdr}
-\ncline{-}{let2}{let3}
-\ncline{-}{cdr}{lista2}
-
-\ncline{-}{let3}{rec1}
-\ncline{-}{let3}{append}
-\ncline{-}{let3}{let4}
-
-\ncline{-}{append}{tl2}
-\ncline{-}{append}{listb2}
-
-\ncline{-}{let4}{let5}
-\ncline{-}{let4}{hd1}
-\ncline{-}{let4}{car}
-\ncline{-}{car}{lista3}
-%% \ncline{-}{len4}{tl2}
-%% \ncline{-}{let4}{vone1}
-%% \ncline{-}{let4}{one}
-%% \ncline{-}{let4}{let5}
-\ncline{-}{let5}{ans1}
-\ncline{-}{let5}{cons}
-\ncline{-}{cons}{hd2}
-\ncline{-}{cons}{rec2}
-%% \ncline{-}{let5}{plus}
-\ncline{-}{let5}{ans2}
-\end{pspicture}
- \end{center}
-
-  \caption{Example to illustrate liveness computation}\label{fig:live-examp}
-\figrule
-\normalsize
-\end{figure}
-
-\newpage
-
-\begin{figure}[t]
-  \begin{tabularx}{320pt}{|| c | X | X | X ||}
-\hline
-Pgm  Pt & Demand  & Pre & Post \\
-\hline 
-$\pi_1$ & $\sigma$ & \{\lista\} $\cup$ \listb.$\sigma$ $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & $\phi$ \\
-\hline
-%% $\pi_2$ & $\epsilonset$ & \{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_3$ & $\epsilonset$ &\{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_4$ & $\sigma$ &\{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-%% $\pi_5$ & \epsilonset & \{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  \\
-%% \hline
-%% $\pi_6$ & $\sigma$ & \listb.$\sigma$ & $\phi$ \\
-%% \hline
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_7$ & $\sigma$  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$\\
-%% \hline
-%% $\pi_8$ & \Lf{\append}{1}(\{\bcdr\}$\sigma$) & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_9$ & \epsilonset $\cup$ \{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_{10}$ &$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-$\pi_{11}$ & \{\bcdr\}$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \lista $\cup$ \lista.\{\acar\bcar\} \\
-\hline
-$\pi_{12}$ &\Lf{\append}{1}(\{\bcdr\}$\sigma$)   &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-$\pi_{13}$ & \Lf{\append}{2}(\{\bcdr\}$\sigma$)  &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{14}$ & $\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$ \\
-\hline
-%% $\pi_{15}$ & \{\bcar\}$\sigma$  & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   &  \xrec.\{\bcdr\}$\sigma$\\
-%% \hline
-%% $\pi_{16}$ & \epsilonset $\cup$ \{\acar\bcar\}$\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$  & \xrec.\{\bcdr\}$\sigma$ \\
-%% \hline
-$\pi_{17}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   & $\phi$  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{18}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$  \\
-\hline
-$\pi_{19}$ &  \{\bcar\}$\sigma$        & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{20}$ &   \{\bcdr\}$\sigma$       & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{21}$ & $\sigma$ & \xans.$\sigma$ & $\phi$  \\
-\hline
-\end{tabularx}
-  \caption{Liveness annotations for the body of \append}\label{fig:xe_append}
-\figrule
-\normalsize
-\end{figure}
-
-\end{document}

### 2014-07-13 paper/sections/abstract.tex
organization added

--- 2014-07-13 paper.tex
@@
-%------------------------------------------------------------%
-\documentclass{llncs}
-\pagestyle{plain}
-\usepackage{longtable} 
-%------------------------------------------------------------%
-%- common stuff starts here....
-\usepackage{etex}
-\usepackage{float}
-\usepackage{mathrsfs}
-\usepackage{booktabs}
-\usepackage{boxedminipage}
-\usepackage[T1]{fontenc}
-%\usepackage[pdf]{pstricks}
-\usepackage{epsfig}
-\usepackage{multirow}
-\usepackage{url}
-\usepackage[normalem]{ulem}
-\usepackage{schemepgm}
-\usepackage{graphicx}
-\usepackage{graphpap}
-\usepackage{tabularx}
-%\usepackage{color}
-%\usepackage{xcolor}
-\usepackage{algorithmic}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{pstricks}
-\usepackage{pst-text}
-\usepackage{pst-node}
-\usepackage{pst-tree}
-\usepackage{pst-rel-points}
-%\usepackage{pst-all}
-\usepackage{bcprules}
-%\usepackage{subfigure}
-%\usepackage{algorithmic}
-%\usepackage{anysize}     
-\input{fh_defs}
-%\newcommand{\comment}[1]{}
-\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
-\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
-    #1 \medskip \hrule \medskip}}}
-\newcommand{\scmin} {\mbox{\sf\em in}}
-\newcommand{\scmnull}{\mbox{\sf\em null?}}
-\newcommand{\scmpair}{\mbox{\sf\em pair?}}
-\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
-\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
-
-\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
-\newcommand{\stk}{\mbox{S}}       % stack
-\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
-\newcommand{\bang}{\mbox{\sc bang}}
-
-\begin{document}
-%  \title{Heap  Reference  Analysis  for Functional  Programs}
-\title{Liveness-Based Garbage Collection}
-
-% make a proper TOC despite llncs
-\setcounter{tocdepth}{2}
-\makeatletter
-\renewcommand*\l@author[2]{}
-\renewcommand*\l@title[2]{}
-\makeatletter
-
-  \author{Rahul Asati\inst{1} \and Amitabha Sanyal\inst{1} \and Amey Karkare\inst{2}
-    \and   Alan Mycroft\inst{3}}
-  \institute{IIT Bombay, Mumbai 400076, India\\
-    \email{$\{$rahulasati,as$\}$@cse.iitb.ac.in}, \and 
-             IIT Kanpur, Kanpur 208016, India\\
-    \email{karkare@cse.iitk.ac.in}, \and 
-             Computer Laboratory, University of Cambridge, CB3 0FD, UK\\
-    \email{alan.mycroft@cl.cam.ac.uk}}
-\maketitle
-
-\begin{abstract}
 Current garbage collectors  leave much heap-allocated data uncollected
@@
 programs.
-\end{abstract}
-
-%------------------------------------------------------------%
-
-%% \begin{figure}
-%% \tableofcontents       %%%%%%% BEWARE: find the \setcounter{page}{2}
-%%                        %%%%%%% and comment it out when deleting this TOC.
-%% \vspace{5cm}
-
-%% \section*{Page Budget}
-%% CC'2014 and ESOP'2014 have a limit of 20 pages.
-%% A draft page budget is:
-%% \begin{itemize}
-%% \item 1. abstract/introduction (3 pages)
-%% {\color{red}Currently 2.5 pages.}
-%% \item 2. core language syntax and semantics (2 pages).
-%% {\color{red}Currently 2 pages.}
-%% \item 3. liveness analysis and correctness (3 pages)
-%% {\color{red}Currently 3.8 pages.}
-%% \item 4. practical liveness computation (6 pages)
-%% {\color{red}Currently 6.2 pages.}
-%% \\ I think this is where should do all 3 of: the context-insensitive analysis
-%% (one which has a summary demand for each function), then the
-%% solving like differential equations, then using Mohri-Nederhof to
-%% get a NFA for each (call/cons) program point.
-%% \item 5. experimental framework and results (4 pages)
-%% \item 6. (optional) monotonicity of GC (0.5 page)---our technique can never
-%% cause more collections.
-%% \item 7. conclusions/acknowledgements/bibliography (1.5 page)
-%% \end{itemize}
-%% \end{figure}
-
-\section{Introduction}
-\label{sec:intro}
-
-Most modern  programming languages support dynamic  allocation of heap
-data.  Static  analysis of heap data  is much harder  than analysis of
-static   and   stack   data.    Garbage   collectors,   for   example,
-conservatively  approximate  the liveness  of  heap  objects by  their
-reachability  from a  set of  memory  locations called  the {\em  root
-  set\/}.  Consequently, many objects  that are reachable but not live
-remain  uncollected, causing  a  larger-than-necessary memory  demand.
-This is confirmed  by empirical studies on Haskell~\cite{rojemo96lag},
-Scheme~\cite{karkare06effectiveness}                                and
-Java~\cite{shaham02estimating} programs.
-
-Here we consider a first-order  pure functional language and propose a
-liveness  analysis  which  annotates  various program  points  with  a
-description  of variables and  fields whose  object references  may be
-dereferenced  in the future.   The garbage  collector then  only marks
-objects pointed by live references and leaves other, merely reachable,
-objects  to  be  reclaimed.    (Although  not  strictly  necessary,  a
-collector  would  normally {\em  nullify}  dead  variables and  fields
-rather than leaving dangling  references.)  Since there are fewer live
-objects   than   reachable   objects,   more  memory   is   reclaimed.
-Additionally, since  the collector traverses a smaller  portion of the
-heap, the time spent for each collection is also smaller.  The work is
-presented in  the context of a  stop-the-world non-incremental garbage
-collector (mark-and-sweep,  compacting or  copying) for which  we also
-show a  monotonicity result: that  our technique can never  cause more
-garbage  collections  to  occur   in  spite  of  changing  the  rather
-unpredictable  execution  points   at  which  collections  occur.   We
-anticipate that our technique  is applicable to more modern collectors
-(generational,  concurrent, parallel),  but leave  such  extensions to
-future work.
-
-We first define a {\em fully context-sensitive} (in the sense that its
-results  are unaffected  by function  inlining) liveness  analysis and
-prove it  correct.  However, fully context-sensitive  methods often do
-not scale,  and this analysis would  also require us  to determine, at
-run-time,  the internal  liveness of  a  function body  at each  call.
-Hence,   similarly   to   the   0-CFA   approach,   we   determine   a
-context-independent {\em summary} of  liveness for each function which
-safely   approximates   the   context-dependence   of   all   possible
-calls~\cite{chatterjee99relevant,cherem07fast,lee05static}.       (Note
-that  an intraprocedural context-insensitive  method which  assumes no
-information  about function  callers would  be too  imprecise  for our
-needs.)   In essence  our approach  sets up  interprocedural data-flow
-equations for the liveness summaries  of functions and shows how these
-can be solved symbolically  as context-free grammars (CFGs)\@.  We can
-then determine  a CFG  for each program  point; these are  then safely
-approximated with  finite-state automata  which are encoded  as tables
-for each program point.  For  garbage collection purposes only {\em GC
-  points}  (program points  representing  calls, to  $\CONS$  or to  a
-user-function) need to be stored.
-% which may call eventually call $\CONS$, need to be stored.
-
-We  previously proposed  an intraprocedural  method for  heap liveness
-analysis   for   a   Java-like   language~\cite{khedker07heap}   which
-statically inserted  statements nullifying dead  references to improve
-garbage collection; by  contrast nullification here occurs dynamically
-(which can work better with  aliasing) when the garbage collector acts
-on  liveness  annotations  to  avoid traversing  dead  references.   A
-workshop    paper~\cite{karkare07liveness}    outlined    the    basic
-0-CFA-style-summary  interprocedural  approach  to  functional-program
-liveness  analysis.   The  current  paper adds  the  context-sensitive
-analysis and better formalization along with experimental results.
-%   to  substantiate our ideas.
-
-\kern -3pt   %% AM: a very small tweak to get back to 2 pages for this
-             %% section.
-%% AM: rewrite the next paragraph to save 3-4 words and remove this \kern.
-
-%------------------------------------------------------------%
-\setcounter{page}{2}
-\subsubsection{Motivating Example}
-\label{sec:motiv}
-
-Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
-$\pi$ of an expression $e$  denotes a program point.  During execution
-of  the program, it  represents the  instant of  time just  before the
-evaluation  of  $e$.  
-%%While we discuss liveness, it  will also be important to consider an
-%%instance of time  just after the execution of  an expression, though
-%%we shall not be required to mark it with a label.
-% memory as  a (possibly  disconnected) directed
-% acyclic   graph.\footnote{Since  the   language   under  consideration
-%   (Section~\ref{sec:defs}) does not  have any imperative features, the
-%   heap cannot  have cycles.}
-% Here all data  is assumed to  be boxed, i.e.\
-% stored in heap and accessed through references.
-We  view the  heap  as a graph.  
-Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
-containing values.  Edges in the graph are {\em references} and
-emanate from {\em variables} or fields.
-Variable and field values may also be atomic values ($\NIL$, integers etc.)
-While it is convenient to box these in diagrams, our presented
-analysis treats them as non-heap values.
-
-Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
-The edges shown  by thick arrows are those  which are made live by the
-program. In addition,  assuming that
-the value of any reachable part of the program result may be explored
-or printed, the edges marked by thick dashed arrows are also live.
-A cell  is  marked and preserved  during  garbage collection,  only  if it  is
-reachable from the  root set through a path of  live edges.  All other
-cells can be  reclaimed.  Thus if  a garbage collection takes
-% place   at    the   time   represented   by   the    heap   shown   in
-at $\pi$ with the    heap   shown   in
-Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
-will be marked and preserved.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% MOTIVATING EXAMPLE
-\newcommand{\nilfigure}
-{\scalebox{0.75}{
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,0)(1,1)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
-\putnode{start}{origin}{0}{0}{}
-\putnode{stop}{origin}{10}{10}{}
-\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
-\end{pspicture}
-}}
-
-\begin{figure}[t!]
-%%%%{\color{Myblue}
-\begin{picture}(100,130)(0,-90)
-%  \begin{center}
-    \begin{tabular}{cc}
-\begin{boxedminipage}{.5\textwidth}
-      {\sf
-	\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
-	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
-	       \listb
-	  \UNL{1}      (\CONS\ (\CAR\  \lista)
-%          \UNL{2}          (\CAR\  \lista)
-	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
-          \listb))))
-          \UNL{0}
-	  \UNL{0} \hspace*{-.49\TAL} (\LET\ z  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
-	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
-	  \UNL{0} (\LET\ y  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
-	  \UNL{1}   (\LET\ w $\leftarrow$ (\append\ y z)\ \IN
-	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ w)))))
-	\end{uprogram}
-      }} 
-\end{boxedminipage}
-      &
-      \raisebox{-25mm}{\scalebox{.75}{
-	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
-      \psset{unit=1mm}
-      \psset{linewidth=.3mm}
-      \begin{pspicture}(0,-5)(70,60)
-	%\psframe(0,0)(73,60)
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
-	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
-%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
-	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	%      fillcolor=white]{$\times$}}}
-        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
-	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
-	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-		  fillcolor=white]{$\times$}}}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
-	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
-	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
-	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
-	\ncline[nodesepB=-.5]{*->}{c2}{d}
-	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
-	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
-	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
-	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
-	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
-	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
-%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
-        %here
-        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
-        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-
-%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
-%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
-	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
-	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
-%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      \end{pspicture}}} \\
-      (a) Example program.&
-      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
-	(b) Memory graph at $\pi$. Thick edges denote \\
-        {\white (b) }live links. Traversal stops at edges marked\\
-        {\white (b) }$\times$  during garbage collection.
-      \end{tabular}}
-    \end{tabular}
-\end{picture}
-
-\kern -3ex
-
-\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
-%  \end{center}
-%%%%}\color{Myblue} end
-\end{figure}
-
-\subsubsection{Organization of the paper}
-Section~\ref{sec:defs} gives
-the syntax and semantics of
-the language used to illustrate our analysis along with basic concepts
-and    notations.
-% Section~\ref{sec:operational}   describes    the
-% operational semantics  of the  language.
-Liveness analysis is described in
-Section~\ref{sec:liveness} followed by a sketch of a correctness
-proof relative to a non-standard semantics.
-Section~\ref{sec:computing} shows how to encode liveness as
-finite-state automata.
-Section~\ref{sec:experiments}   reports   experimental   results and
-Section~\ref{sec:lgc-always-better} proves that a liveness based
-collector can never do more garbage collections than a reachability
-based collector. 
-% and Section~\ref{sec:conclusion} concludes the paper.
-
-\section{The target language---syntax and semantics}
-\label{sec:defs}
-We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
-functions ($\CONS$, $\PRIM$ etc.).
-The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
-it has eager semantics and restricts programs to be in 
-Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
-where all actual parameters to functions are variables.
-This restriction does not affect expressibility (and indeed we feel free
-to ignore it in examples when inessential), but simplifies the analysis formulation.
-Additionally, as in the three-address instruction form familiar from compiler texts,
-it forces each temporary to be named and function calls to be serialized
-(necessary to get an unambiguous definition of liveness).
-We further require that each variable in a program is distinct, so that no scope shadowing
-occurs---this simplifies proofs of soundness.
-In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
-or return ($\RETURN$).
-The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.
-
-The body of the program is
-the  expression denoted  by  $e_\mainpgm$; for analysis purposes
-it is convenient
-to regard $e_\mainpgm$ as part of a function definition
-$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
-% As  mentioned earlier,
-We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.
-
-% \paragraph{Inlining.}
-In spite of the ANF restrictions it is still possible
-to inline non-recursive functions (a fact we use to prove the safety
-of liveness analysis).  A user-function call
-$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
-to a function defined
-(after renaming its formals and locals to be disjoint from existing variables)
-by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
-is replaced by a sequence of $\LET$'s of the form
-$z_i\leftarrow (\ID\,\,y_i)$ followed by
-the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
-$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.
-
-\begin{figure}[t!]
-\footnotesize
-\begin{eqnarray*}
-   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
-    \hspace{7em} \mbox{\em --- program}\\
-    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
-    e) 
-    \hspace{1.7em} \mbox{\em --- function definition} \\
-e \in \mathit{Expr} & ::= &
-\left\{\begin{array}{ll@{\hspace{4em}}l}
-       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
-       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
-       (\RETURN\,\, x) && \mbox{\em --- return from function}
-    \end{array}\right. \\
-s \in \mathit{Stmt} & ::= &
-\left\{\begin{array}{lr@{\hspace{1em}}l}
-       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
-       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
-       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
-       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
-%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
-       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
-%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
-       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
-            & \mbox{\em --- function application} 
-    \end{array}\right.
-\end{eqnarray*}
-  \caption{The syntax of our language}\label{fig:lang-syntax}
-\figrule
-\normalsize
-\end{figure}
-
-\subsubsection{Semantics}
-We now give an operational
-semantics for  our language.  Later, a refinement of
-the  operational semantics,  which we  call minefield  semantics, will
-serve to prove liveness analysis correct.
-We give a small-step semantics because, unlike big-step semantics,
-correctness for non-terminating programs does not need special treatment.
-We start with the domains used by the semantics:
-\[
-\begin{array}{rlcl@{\hspace{2em}}l}
-% \begin{eqnarray*}
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
-\mbox{-- Environment} \\ 
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
-  Val}) & \mbox{-- Heap}
-% \end{eqnarray*}
-\end{array}
-\]
-Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
-A value is either a number, the empty list $\NIL$, or a location  $\ell$.
-Our liveness analysis does not track numeric values, and thus is neutral
-as to whether these are boxed or represented as immediates.
-An environment is a finite mapping from variables to values, and a heap a finite mapping
-from locations to pairs of values.
-% Given  a program,  an  environment is  a
-% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
-% values in $\mathit{Val}$.    The heap is a mapping
-% from  locations  to pairs of  values.
-Finally, $\stk$ is a stack
-(using $\bullet$ for push and $[\,]$ for empty stack)
-of frames of unfinished function calls.  A
-frame is a triple $(e,x,\rho)$ representing the call site
-$
-(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
-$
-being evaluated in environment $\rho$.
-Frames can also be viewed as continuations, in this view
-the ({\sc ord-return}) 
- rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
-
-The semantics 
-of statements $s$ are given by the judgement form
-$\rho, \heap, s \rightsquigarrow \heap', v$
-and those for expressions $e$ by the form
-$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
-The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
-program terminates successfully with result value $\rho(x)$
-on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$
-
-Notation:  we write $\rho[x \mapsto v]$ for the environment which is
-as $\rho$ but has value $v$ at $x$.
-We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
-has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
-$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
-$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
-
-\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
-with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
-a contained $s$).  Some of these we eliminate syntactically, e.g.\
-ensuring all variables and functions are defined and are called with
-the correct number of parameters.
-Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
-All but the first can be eliminated with a static type system but,
-treating our program as dynamically typed, we regard all these
-as stuck states.
-
-% A program terminates successfully if it reaches the halt state
-% in Figure~\ref{fig:lang-opsem-small}.
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-\begin{figure}[t!]  
-% (Random text to ensure proper line-spacing (problem with infrule macro?).)
-
-\footnotesize
-\begin{minipage}{0.335\textwidth}
-\infrule[ord-const]
-        {}
-        {\rho, \heap, \pp{\pi}{k}
-          \rightsquigarrow \heap, k}
-\end{minipage}\begin{minipage}{0.655\textwidth}
-\infrule[ord-cons]
-      {
-\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
-}
-      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
-        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
-        \ell}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-car]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
-          \rightsquigarrow \heap, v_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-cdr]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
-          \rightsquigarrow \heap, v_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-id]
-      {
-}
-      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
-        \rightsquigarrow \heap, \rho(x)}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-prim]
-      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
-      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
-        \rightsquigarrow \heap, \rho(x) + \rho(y)}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-null-false]
-\infrule[]
-        {\rho(x)\not= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 0}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-null-true]
-\infrule[ord-null]
-        {\rho(x)= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 1}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-if-true]{
-\infrule[]{
-  \rho(x) \in \mathbb{N}\setminus\{0\}
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-if-false]{
-\infrule[ord-if]{
-  \rho(x) = 0
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-nonfn]{
-%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
-  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
-  \andalso
-  \mbox{$s$ is not
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-fncall]{
-  \mbox{$s$ is 
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-\andalso
-\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
-[\myvec{z} \mapsto \rho(\myvec{y})],
-  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
-}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-return]{
-}{
-  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
-  \pp{\pi}{(\RETURN\ x)}
-  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
-\end{minipage}
-
-% \normalsize
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-  \caption{The small-step operational semantics}
-  \label{fig:lang-opsem-small}
-\figrule
-\normalsize
-\end{figure}
-
-%==============================================================
-\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-
-\section{Liveness}\label{sec:liveness}
-
-In classical liveness analysis a variable is either `live' (its value
-may be used in future computation) or `dead' (definitely not used).
-Semantically, a variable is dead at a given program point
-if arbitrary changes to its value have no effect on the computation.
-Later we will use $\bot$ to represent a value which `explodes' when
-it is used in a computation; dead variables can safely have their
-value replaced with $\bot$.
-For heap-allocated data  we need a richer model of liveness
-in that both variables and fields of $\CONS$ cells may be dead or live.
-Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
-liveness of the structure reachable from a variable is a set of
-{\em access paths}
-which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
-grammar notation.
-Thus the liveness of $x$ being
-$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
-only refer to the second and third members of $x$ considered as a list.
-Semantically, access paths are prefix-closed, as accessing a field
-requires accessing all the paths from the variable to the field,
-and hence the above liveness is properly written 
-$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
-The classical notions of a scalar variable being live or dead correspond to
-$\{\epsilon\}$ and $\{\}$.
-
-The overall liveness (also written {\em liveness environment} for emphasis)
-at a program point is conceptually a mapping from
-variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
-notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
-instead of the map
-$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
-Analogously to classical liveness, the liveness at program point $\pi$
-in $\pi:e$ is the liveness just before executing $e$.
-
-A complementary notion to liveness is {\em demand}.  The demand for
-expression $e$ is again an access path---that subset of of $\{\acar,\acdr\}^\ast$ which the
-context of $e$ may explore of $e$'s result.  So, for example given a
-demand $\sigma$ and the expression $\pi: (\RETURN\ x)$, the liveness at
-$\pi$ is exactly $x.\sigma$.
-The classical analogy of this is in {\em strong liveness}, where
-an assignment node $n: x:=y+z$ causes $y$ and $z$ to be live on entry
-to $n$ if (and only if) $x$ is live at exit of $n$---the liveness
-of $x$ at exit from $n$ becomes the demand on $y+z$.
-Note that, for an operation like division which
-may raise an exception, the assignment $n: x:=y/z$ makes $y$ and $z$ live
-regardless of the liveness of $x$.
-
-We use $\sigma$ to range over demands, $\alpha$ to range over access paths
-and $\Lv$ to range over liveness environments. The notation
-$\sigma_1\sigma_2$ denotes the set $\lbrace \alpha_1\alpha_2 \mid
-\alpha_1 \in \sigma_1, \alpha_2 \in \sigma_2\rbrace$. Often we shall
-abuse notation to juxtapose an edge label and a set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
-Finally, we use $\Lfonly$ to range over {\em demand transformers}.
-These transform demands on a function call
-into demands for its formal parameters:
-given a user function $f$, defined by
-$(\DEFINE\ (f\ x_1\ \ldots\ x_n)\ \ e_f)$ and called with demand $\sigma$,
-then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
-
-Note that liveness refers to variables and fields, and not to $\CONS$ cells
-(i.e.\ to edges in the memory graph, not to locations themselves).
-Hence liveness of $\{x.\epsilon,x.\acar\}$ means that future computation
-may refer to the value $\ell$ of variable $x$, and also to the $\CAR$ field
-of location $\ell$.
-In the absence of other pointers to heap location $\ell$, we are
-certain that the $\CDR$ field of $\ell$ will not be referenced
-and may hence be corrupted arbitrarily.
-Note therefore, that while  $\ell$ cannot be garbage collected, any
-location $\ell'$ stored in the $\CDR$ field of $\ell$ would be garbage
-(again provided there are no other aliases to  $\ell$).
-
-%% \comment{
-%%   \begin{enumerate}
-%% \item Set the framework for liveness. Basic notions: access
-%%   paths and expressions.
-%% \item An informal working definition of liveness. Minefield will
-%%   provide the formal definition.
-%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
-%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
-%%   rules. One of the changes from  the earlier version is the lack of a
-%%   post-liveness in  the judgements. This anticipates  the intended use
-%%   of  the analysis  for garbage  collection and has  to be  carefully
-%%   explained.
-%%   \end{enumerate}
-%% }
-
-%% We now  set up  the framework for  liveness analysis.  Given  a memory
-%% graph, a link in the graph  can be described by starting with a member
-%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
-%% \CDR\ dereferences to reach the link.  For notational convenience, the
-%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
-%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
-%% dereferences    is    represented   by    a    string   $\alpha    \in
-%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
-%% in the  memory graph that starts  at $x$ and  traverses adjacent links
-%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
-%% We call the string $x.\alpha$ an {\em access expression\/}, the string
-%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
-%% graph   by  the   access  expression   an  {\em   access   path}.   In
-%% Figure~\ref{fig:mot-example},        the       access       expression
-%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
-%% containing the value  $4$.  Most often the memory  graph is clear from
-%% the context,  and we shall use  access expressions to  refer to access
-%% paths. An  access path  used to refer  to a  link in the  memory graph
-%% refers    to     the    last     link    in    the     path.     Thus,
-%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
-%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
-%% set of access paths, then  $x.\sigma$ will represent the set of access
-%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
-%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
-%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
-%% \sigma\rbrace$.
-
-%% \subsection{Liveness Analysis}
-%% \label{sec:exp-liveness}
-%% 
-%% Consider the memory graph at a program point $\pi$ during an execution
-%% of a program. A  link in the memory graph is said  to be {\em live} at
-%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
-%% execution  starting from  the  evaluation of  the expression  labelled
-%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
-%% be live at  $\pi$, if, during {\em any} execution  of the program, the
-%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
-%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
-%% In the first, the path  itself represents the sequence of dereferences
-%% to  access  the last  edge,  as,  for  example, while  evaluating  the
-%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
-%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
-%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
-%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
-%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
-%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
-%%   implicitly  live} if it is live in the second sense.  Clearly, static
-%% determination   of  implicit  liveness   requires  an   analysis  that
-%% determines  potential sharing  during  any execution  of the  program.
-%% However, the intended application  of our liveness analysis is garbage
-%% collection  during which  we have  access  to the  entire heap.   This
-%% obviates the need for a  separate sharing analysis and we restrict our
-%% attention to explicit liveness only.
-%% 
-%% 
-%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
-%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
-%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
-%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
-%% %% \bcdr\ represent application of  the constructor \CONS.  The following
-%% %% example motivates the need for these symbols.
-%% 
-%% %% \begin{example}\label{exmp:motiv-bar}
-%% %%   Consider the expression at program point $\pi_1$ in
-%% %%   \begin{center}
-%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
-%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
-%% %% 			    $\pi_3\!:${$\cdots$})}}
-%% %%   \end{center}
-%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
-%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
-%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
-%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
-%% %%   \begin{itemize}
-%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
-%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
-%% %%     live at $\pi_1$.
-%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
-%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
-%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
-%% %%   \end{itemize}
-%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
-%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
-%% %% \end{example}
-%% 
-%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
-%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
-%% %% resulting access patterns may not  be directly useful in that they may
-%% %% not describe a  path in the memory graph.  We  define a {\em canonical
-%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
-%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
-%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
-%% %% to obtain a canonical access  pattern representing the same edge. This
-%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
-%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
-%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
-%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
-%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
-%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
-%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
-%% %% cannot  be  used for  nullification  during  garbage collection.   The
-%% %% liveness of a such a path is irrelevant, and the path is not considered
-%% %% any further.
-%% 
-%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
-%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
-%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
-%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
-%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
-%% %% element in $\sigma_2$, i.e.
-%% 
-%% %% \begin{eqnarray*}
-%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
-%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
-%% %% \end{eqnarray*}
-
-\subsection{Liveness Analysis}
-First recall the classical formulation of liveness (as sets of simple variables)
-on three-address instructions,
-$
-\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I)
-     \setminus \mathit{def\/}(I)
-     \cup   \mathit{ref\/}(I)
-$,
-and then note that {\em strong liveness} needs, when $I$ is the instruction $z:=x+y$,
-that $\mathit{ref}(I)$ be refined to
-$\{x,y\}$ if $z \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
-
-Our liveness analysis formulated in Figure~\ref{fig:live-judge}
-is analogous.
-% has three analogous aspects.
-% \begin{itemize}
-% \item
-Firstly, the function $\mathit{ref}$, when given a statement $s$,
-returns the incremental liveness {\em generated} by $s$.
-Because we generalize {\em strong} liveness, $\mathit{ref}$ needs a second parameter,
-specifying the demand $\sigma$ on the result of $s$,
-to determine which access paths of its free variables are made live.
-The cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
-% The cases for
-% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
-% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
-% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
-A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
-argument (even if its result is never used).
-This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
-(note $\sigma$ may be $\{\}$).
-In the opposite  sense, the  demand  of $\acar\sigma$  on
-(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-that \CONS\ does not, by itself, dereference its arguments.
-Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
-$\mathit{ref}$
-expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
-Constants generate no liveness.
-
-% \item
-The function $\mathcal{L}$ now gives the (total) liveness of an expression $e$.
-The cases $\RETURN$ and $\SIF$ are straightforward, but note the liveness $x.\epsilon$ generated
-by the latter.
-The case $(\LET\ z\leftarrow s\ \IN\ e')$
-resembles a three-address instruction:
-the liveness of $e$ is given by taking the liveness, $\Lv$, of $e'$,
-killing any liveness of $z$ and adding any incremental liveness from $s$.
-The main subtlety is how the liveness of $z$ in $\Lv$ is converted to a demand $\Lv(z)$ to be
-placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-% \item
-%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-Finally, the judgement form $\mathit{Prog} \len \Lfonly$ is used to determine $\Lfonly$\@.
-Analogously to classical liveness being computed as
-a solution of dataflow equations, we require, via inference rule ({\sc live-define}),
-$\Lfonly$ to satisfy the fixed-point property that:
- when we assume $\Lfonly$ to be the family of demand transformers for the program
- then the calculated liveness of each function body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees
- with the assumed  $\Lfonly$.
-As usual, there are often multiple solutions to $\Lfonly$; all are safe
-(see Section~\ref{sec:correctness})
-but we prefer the least one as giving the least liveness subject to safety---and
-hence greatest amount of garbage collected.
-% \end{itemize}
-
-We make three observations: firstly the rule ({\sc live-define}) has a least solution
-as $\mathcal{L}(\cdot)$ is monotonic in $\sigma$; secondly that
-({\sc live-define})
-resembles the rule for type inference of mutually recursive function definitions,
-and thirdly the asymmetry of demand and liveness (compared to post- and pre-liveness
-classically) is due to the functional formulation here.
-
-Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
-(representing a fully context-sensitive analysis)
-can be safely approximated, for each function, by a {\em procedure summary}
-(unifying the contexts in the style of 0-CFA)\@.
-The summary consists of a pair of a single demand and, for this demand,
-the corresponding tuple of demands the function makes on its arguments.
-
-%%% save two lines to bring up section 3.2.
-
-% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
-% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
-% enables us to do a context-dependent analysis of a call without having
-% to enter the body of the called function.
-
-%% We present  our liveness  analysis using judgement forms $\len$
-%% for both expressions and programs.
-%% These are given formally in Figure~\ref{fig:live-judge}, but
-%% first we an intuitive understanding of the issues involved.
-%% Consider the expression:
-%% 
-%% \begin{center}
-%% \begin{uprogram}
-%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
-%%                         \px) \IN\  
-%%            % \UNL{0}
-%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
-%%                         \pw ) \IN \ldots) 
-%% \end{uprogram}
-%% \end{center}
-%% 
-%% 
-%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
-%% \pw.\epsilon\}$.
-%% Informally, this is so since \pw\ and \pz\ are used in
-%% (\PRIM~\pz~\pw ).
-%% 
-%% \begin{enumerate}
-%% \item The liveness  at $\pi$ is obtained by  removing the access paths
-%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
-%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
-%%     similar  to liveness  analysis of  imperative programs,  where the
-%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
-%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
-%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
-%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
-%%     liveness generated and  killed in $s$}. 
-%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
-%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
-%%   as a {\em  demand}.  The demand on an expression consists of  access
-%%   patterns  representing all  possible future  uses of  its  value.  A
-%%   demand  of $\epsilon$  means that  only the  reference resulting
-%% %  from the evaluation  of
-%%   from  $(\CAR\,\,\px)$ will be used.
-%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
-%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
-%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
-%% \end{enumerate}
-%% 
-%% In general,  an access path is  live before an  expression because the
-%% link represented by it is used (i) while evaluating the expression, or
-%% (ii) while using the result of the expression, or (iii) simply because
-%% the link  is live at  a later program  point.  When the  expression in
-%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
-%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
-%% the access  path \pw.   Note that \px.\acar\  is live in  this example
-%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
-%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
-%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
-%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
-%%
-
-% \paragraph{Judgements formally.}
-% Let  $\Lv$  denote a  liveness  environment (a function from
-% variables to access paths, abused using grammar notation) and
-% $\sigma$ be a demand.
-% The judgement form $e, \sigma \len \Lv$,
-% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
-% environment  before $e$  is $\Lv$ and
-% is thus analogous to the classical notion of {\em transfer function}.
-
-%    It is  interesting to  note that,
-% unlike traditional liveness analysis,  our judgements do not require a
-% liveness after the expression. This  is because of the coming together
-% of  two reasons.   First, it  will be  clear while  discussing garbage
-% collection,  that it is  enough for  our purposes  to assume  that the
-% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
-% restriction of our  programs to ANF ensures that  the post-liveness of
-% each  expression  $e$  inside   the  function  body  continues  to  be
-% $\emptyset$. Thus the post-liveness can be dropped altogether.
-
-%==============================================================
-\begin{figure}[t]  
-% \footnotesize
-\begin{eqnarray*}
-\mathit{ref\/}(\kappa,\sigma,\Lfonly)
-          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
-          &=& \{x.\sigma\} \\
-\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
-          &=& \{x.\epsilon,y.\epsilon\} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \\
-\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
-        \mathcal{L}(e_1,\sigma,\Lfonly) \cup
-        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
-        \{x.\epsilon\} \\
-\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
-        \Lv
-           \setminus \{x.*\}
-           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
-\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
-\end{eqnarray*}
-
-\begin{minipage}{0.85\textwidth}
-\infrule[live-define]
-        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
-           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \len \Lfonly
-\\ \makebox[0mm]{where
-     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
-\end{minipage}
-
-\normalsize
-  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%==============================================================
-
-% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
-% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
-% liveness generated  before the right-hand-side of  a \LET\ definition.
-% Constants    do   not   generate    any   liveness.     The   operator
-% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
-% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
-% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
-% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
-% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-% that \CONS\ does not, by itself, dereference its arguments.
-% 
-% 
-% Unlike primitives, the mapping from the demand on a user-function call
-% to the  demands on its arguments  has to be derived  from the function
-% body itself---as a fixed point for a recursive function.
-% The function  \Lfonly\ captures this mapping for
-% all functions  defined in the  program.  For a  function $\mathit{f}$,
-% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
-% a    call     to    the     $i$th    argument.     The     rule
-% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
-% confirmed by the body of all the functions, then the guess is correct.
-% {\color{blue}Note the similarity of this rule to type-inference
-% for recursive program schemes.}
-
-% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
-% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
-% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
-% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
-% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
-% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
-% liveness generated before $s$.
-
-\subsection{Minefield semantics and correctness}
-\label{sec:minefield}
-\label{sec:correctness}
-
-This section gives a modified semantics
-%(a reference semantics)
-which checks liveness annotations
-at run time, and `explodes' when these are found to be inconsistent with execution behavior,
-but otherwise behaves as the standard semantics.
-We show that such explosions never occur and hence run-time checks
-can be elided.
-%% We first assume some demand transformers $\Lfonly$;
-%% our proof will rely on this being a solution (i.e.\ satisfies $\mathit{Prog} \len \Lfonly$)
-%% of liveness analysis, but we do rely in this yet.
-We first assume an arbitrary  demand transformers $\Lfonly$ which satisfies
-$\mathit{Prog} \len \Lfonly$.
-We then enrich the abstract machine state $\rho, \stk, \heap, e$ to
-$\rho, \stk, \heap, e, \sigma, \Sigma$.
-Here $\sigma$ is the demand to the currently active function,
-thus the liveness $\Lv$ at $e$ is $\mathcal{L}(e,\sigma,\Lfonly)$,\footnote{
-This a simple liveness propagation using $\mathcal{L}(\cdots)$
-and $\mathit{ref}(\cdots)$ as $\Lfonly$ is assumed given.}
-and $\Sigma$ is a stack of demands---one for each function frame
-pushed in $\stk$.
-
-Second, we augment $\mathit{Val}$ with a value $\bot$.
-To model strong liveness $\bot$ may be copied freely,
-including into a $\CONS$ cell,
-% but reduces the state to $\bang$
-but explodes
-when used computationally (in a primitive operation other than a copy).
-Additionally we define
-$\mathit{GC}(\Lv,\Sigma): (\rho, \heap, \stk) \mapsto (\rho', \heap', \stk')$
-which determines {\em live-reachability}\footnote{Reachability curtailed by liveness information.}
-%%%%% \marginpar{Say more about live-reachability?}
-using $\rho$ and the $\rho$'s in $\stk$ as the root set
-and following links in $\heap$ {\em only as far as allowed by $\Lv$ and $\Sigma$}.
-% The effect of $\mathit{GC}(\Lv,\Sigma)$ is to replace
-Hence $\mathit{GC}(\Lv,\Sigma)$ replaces
-live-unreachable values---in $\rho$, in $\heap$ and the $\rho$'s in \stk---with $\bot$.
-For example, if $x.\epsilon \not\in \Lv$ then $\rho'(x) = \bot$.
-%, regardless of $\rho(x)$.
-Only $\mathit{GC}(\cdots)$ introduces $\bot$.
-
-Third, we update the semantics in four ways: ({\em i\/}) we arrange that all
-($\rightarrow$) transitions on expressions $e$ first use $\mathit{GC}(\cdots)$
-to update the state and then continue as before;
-and ({\em ii\/}) whenever the value $\bot$ is used computationally in a reduction
-($\rightsquigarrow$), we enter a distinguished stuck state $\bang$.
-For example, supposing $\rho(x)=\bot$
-then
-$\rho, \heap, s \rightsquigarrow \bang$ if $s$ is
-$(\CAR\,\, x)$, $(\CDR\,\, x)$ or $(\PRIM\,\, x\ y)$,
-but not if $s$ is $(\ID~x)$, $(\CONS~x~y)$ or $(f~x~y)$.
-Finally ({\em iii\/}) we make a similar change to the ($\rightarrow$) reduction for ({\sc ord-if})
-and ({\em iv\/}) augment the ({\sc ord-let}) rule for primitives to propagate
-$\bang$ from ($\rightsquigarrow$) to ($\rightarrow$).
-%%%%%% {\color{blue}The full minefield semantics is given as an appendix?}
-
-% Suppose liveness for the current $e$ under reduction is $\Lv$; then:
-% For every variable $x$ for which $x.\epsilon \not\in \Lv$ then replace
-% $\rho(x)$ with $\bot$.
-% Also, for every $\ell \in \mathit{Loc}$, replace the $\CAR$ field
-% of $\heap(\ell)$ with $\bot$ when
-% which there is no path $x.\alpha\acar$ in $\Lv$ such that following
-% $\alpha$-links in $\heap$ from $\rho(x)$ reaches $\ell$.
-% Correspondingly, replace the $\CDR$ field of $\ell$ if there
-% is no path $x.\alpha\acdr$ in $\Lv$.
-% \item
-% 
-% {
-% \footnotesize
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-car]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CAR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-cdr]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CDR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}\\[1ex]
-% 
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-prim]
-%       {\rho(x) = \bot \vee \rho(y) = \bot}
-%       {\rho, \heap, {(\PRIM\,\, x\ y)}
-%         \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-null]
-%         {\rho(x)= \bot
-%         }{\rho, \heap, {(\NULLQ\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% }
-%
-
-The resulting minefield semantics behaves identically
-(identical heap, identical steps, including possible non-termination)
-to the standard semantics, except for one possibility---the
-minefield semantics going $\bang$ while the standard semantics
-continues (either to a halt state, to a stuck state, or reduces forever).
-
- We now prove a result that relates
-liveness analysis to the semantics.
-\begin{proposition}
-Given program $P$ with $P \len \Lfonly$, then in the minefield
- semantics $P \rightarrow \bang$ can never occur
- (cf.\ `well-typed programs do not go wrong').
-\end{proposition}
-\textit{Proof outline:}
-Space does not permit a full proof, but we give the two main steps.
-We proceed by contradiction and assume there is a program $P$ for which
-$P \len \Lfonly$ can enter state $\bang$.
-The first step is to construct a program $P'$ with identical behavior, but with
-no user-function definitions, using inlining.
-This is possible because a program which goes $\bang$ does so after a finite number
-of reductions, and hence even recursive functions have only had a finite number of
-invocations.
-We hence repeatedly inline-expand user-function calls in $P$
-until we obtain a program $P'$ which behaves identically\footnote{
-Modulo replacement of
-({\sc ord-call}) and
-({\sc ord-return}) steps with
-({\sc ord-id}) steps.
-}
-to $P$ in the standard semantics, but makes no user-function calls.
-Any remaining, non-executed, calls can 
-%be left or
-replaced with a new primitive with the same
-demand-to-liveness transfer function---thus making
-$P'$ a simple expression $e'$.
-%%%%%
-Not only do the program points in reducing $e'$ correspond
-one-one to states during evaluation of $P$,
-but also the liveness associated with a point
-in $e'$ is identical to the liveness at the corresponding state
-$(\rho, \stk, \heap, e, \sigma, \Sigma)$ of $P$ (concatenating the liveness
-$\mathcal{L}(e,\sigma,\Lfonly)$ at $e$
-with the liveness, obtained from $\Sigma$, of the call sites in $\stk$ and
-after renaming the variables correspondingly to the inlining which produced $e'$).
-This assertion relies on the analysis being fully context-sensitive, and noting that
-while the change of scope caused by inlining changes variable visibility
-between $P$ and $e'$ it does not change the liveness---as local-to-a-function
-$\LET$-variables whose scope has been prolonged due to inlining
-are dead in the prolonged scope.
-
-The second step of the proof is to show that $e'$ cannot go $\bang$.
-We proceed by induction.
-Correctness of the $\SIF$ and final $\RETURN$ forms are immediate;
-the $(\LET~z\leftarrow s~\IN~e)$ form requires showing
-the inference rules ensure that any value referenced via $z$ in $e$
-was already live-reachable (via another variable and path) in any
-enclosing expression (so that $\mathit{GC}(\cdots)$ could not re-write it to $\bot$).
-
-% \newpage
-
-\section{Computing liveness and its encoding as a table}\label{sec:computing}
-
-Section~\ref{sec:liveness} gave a context-sensitive liveness analysis
-and proved it correct with reference to a {\em minefield} semantics.
-% Practical computation was not considered.
-For practical use we need to solve the liveness equations {\em finitely and
-symbolically}.  As expressed mathematically, and given a fixed program,
-three things are potentially unbounded:
-({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
-({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
-({\em iii\/}) the number of such access paths.
-
-We commonly first solve ({\em i\/}) by reducing the number of distinct
-calling contexts
-(e.g.\ to a single unified context in 0-CFA style).
-However, it turns out that we can solve the equations for $\Lfonly$ symbolically
-without this reduction, so here we defer this to Section~\ref{sec:bodylivenessbodies}.
-(It also allows easier extension to  dynamically determined liveness---future work.)
-%
-We address ({\em ii\/}) and ({\em iii\/}) by 
-% attempting to re-interpret
-re-interpreting
-the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
-rather than a mutually recursive set of equations on sets of access paths.
-
-This requires two ideas.  Firstly we have to control the use of functions---they
-tend to be infinitary and do not occur naturally in CFGs; in particular our
-$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
-to a tuple of access paths.  The former is achieved by using a separate
-meta-variable (later non-terminal)
-$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
-($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
-Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
-expressible finitely (it is a linear form).
-
-Secondly, there are also two technical issues in re-interpreting
-Figure~\ref{fig:live-judge} as a grammar.
-One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
-disappears after variables are separated ($\Lanv{i}{x}$ above). The other
-is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
-not a CFG\@.
-Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.
-
-Finally,  Section~\ref{sec:NFA-approx}
-uses a construction due to
-Mohri and Nederhof~\cite{mohri00regular} to over-approximate
-context-free grammars with regular grammars;
-these are more appropriate for run-time use.
-%
-Hence the overall `big picture' view is that each GC point
-is annotated with a table encoding the DFA for that program point.
-Such program points forming the call string causing the garbage collection
-(and ending in $\CONS$)
-are used by the garbage collector to curtail its reachability-marking phase
-to this over-approximated liveness.
-
-\subsubsection{Modifying the \CONS\ rule.}
-
-The    $\mathit{ref}$    rule for \CONS,    shown    in
-Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
-\acar\  and  \acdr\  from  access paths in  $\sigma$.
-  Mathematically  this  is fine  but causes problems  when
-  solving  the liveness equations  {\em symbolically}  since such
-  decomposition cannot be expressed as a context-free grammar.  
-To  handle this,  we introduce  two  new symbols  \bcar\  and
-\bcdr\ with the properties:
-\[
- \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
-\hspace{3em} \mbox{and} \hspace{3em}
- \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
-\]
-We can now rewrite the \CONS\ rule as:
-\begin{eqnarray*}
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
-\end{eqnarray*}
-
-We  call the liveness equations with  this modification
-  \Lfunonly$'$.
-The  defintions of \bcar\  and  \bcdr\  induce   the  following
-relation $\hookrightarrow$ over sets of access paths:
-\begin{eqnarray*}
-  \sigma_1\bcar\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
-& \qquad\mbox{and}\qquad &
-\sigma_1\bcdr\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
-\end{eqnarray*}
-%% \noindent \begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcar\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acar\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acdr\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% %\label{eqn:eup-red1}
-%% \end{eqnarray*}
-%% \end{minipage}
-%% \hspace*{.65cm}\begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcdr\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acdr\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acar\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% \end{eqnarray*}
-%% \end{minipage}\\
-
-The reflexive transitive closure of $\hookrightarrow$ will be denoted
-as $\stackrel{*}{\hookrightarrow}$. The following proposition relates \Lfunonly\   and \Lfunonly$'$:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]  
-\begin{boxedminipage}{\textwidth}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\append~\lista~\listb)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \xtest\ $\leftarrow $\ (\NULLQ~\lista) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\SIF\ \xtest~$\pi_3\!\!:$(\RETURN\ \listb)  
-          \UNL{4} \hspace*{.05cm}     $\pi_4\!\!:\, $(\LET\ \xtl\  $\leftarrow$\   (\CDR\ \lista) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_5\!\!:\, $(\LET\ \xrec\  $\leftarrow$  (\append\ \ \xtl\ \  \listb) \IN
-	  \UNL{6} \hspace*{.05cm}    $\pi_6\!\!:\, $(\LET\ \xhd\  $\leftarrow$  (\CAR\ \lista)  \IN
-	  \UNL{7}  \hspace*{.05cm}   $\pi_7\!\!:\,
-          $(\LET\ \xans\  $\leftarrow$ (\CONS\ \ \xhd\ \ \xrec)  \IN\
-          $\pi_8\!\!:\, $(\RETURN\ \xans))))))))
-	\end{uprogram}}
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pz\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}              (\LET\ \py\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_9\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\append\ \py\ \pz) \IN
-          \UNL{4}   \hspace*{.78cm}  $\pi_{10}\!\!:\, $  (\LET\ \pa\  $\leftarrow$  (\CDR\ \pw) \IN
-	  \UNL{5}   \hspace*{.78cm}  $\pi_{11}\!\!:\, $
-          (\LET\ \pb\  $\leftarrow$ (\CAR\  \pa) \IN~$\pi_{12}\!\!:\,$ (\RETURN\ \pb)))))))
-\end{uprogram}}
-\end{boxedminipage}
-  \caption{An example program.}
-\label{fig:examp-pgm}
-\figrule
-\normalsize
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{proposition}
-Assume that  a liveness computation  based on  \Lfunonly\   gives  the liveness of
-the variable $x$ at a program point $\pi_i$ as $\sigma$ (symbolically, $\Lanv{i}{x}=
-\sigma$).  Further,  let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is
-used instead of \Lfunonly.   Then  $\sigma' \stackrel{*}{\hookrightarrow}
-\sigma$.
-\end{proposition}
-
-To see  why the proposition is  true, consider an analysis  based on
-\Lfunonly$'$ in which   $\sigma$ appears in
-the  context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.  Let $\alpha  \in \sigma$.  The
-symbol  \bcar(\bcdr)  merely  marks  a  place in  $\alpha$  where  the
-original  \CONS\  rule  would  have erased  an  immediately  following
-\acar(\acdr),  or, in  absence of  such a  symbol, would  have dropped
-$\alpha$ itself.  Since the application  of any rule in \Lfunonly$'$ merely
-adds  symbols at  the beginning  of  $\alpha$, the  markers and  other
-symbols in $\alpha$ are propagated to other dependent parts of program
-in their same relative positions.  Consequently, the erasure carried out at the end of the analysis with
-$\stackrel{*}{\hookrightarrow}$ gives the same result as obtained
-through \Lfunonly.
-
-\subsection{Generating equations for the demand transformer \Lfonly}
-\label{sec:Lfonlysymbolically}
-
-%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
-We  shall  consider the  program  in  Figure~\ref{fig:examp-pgm} as  a
-running example;
-this is the  ANF version of the program in Figure~\ref{fig:mot-example}.
-
-%%AM no longer used: In what follows,  we shall overload the semantics of
-%% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-%mean the generation
-%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
-%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.
-
-To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
-$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
-{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
-$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
-formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
-of $x_i$.  For our example program which has  a
-single function \append, this generates the following equations:
-% $\Lfonly_{\append}$  is defined by equations:
-  \begin{eqnarray*}
-    \Lf{\append}{1}{\sigma}
-    &=& \epsilonset \cup \acar\bcar\sigma
-    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
-    \Lf{\append}{2}{\sigma}
-    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
-  \end{eqnarray*}
-  In  general, the  equations  for \Lfonly\  are recursive  since
-  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
-  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
-    is  similar to  solving the  differential equation  
-    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
-    that    the    solution     has    the    form    $\mathit{y}
-    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
-    equation  yields  a  quadratic  equation  in  $r$,  and  each
-    solution of $r$ gives rise  to a solution of the differential
-    equation  (in our  setup we  can effectively  pick  the least
-    solution rather than needing linear combinations).  }
-\begin{eqnarray}
-\label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
-\end{eqnarray}
-where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
-alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
-why \Lfonly\ has this form  is as follows: Recall that
-\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
-the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
-situation in which  the demand $\sigma$ does not  reach the $i$th
-argument.   This happens,  for  instance, when  argument is  used
-only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
-transformation of the demand $\sigma$ before it reaches the $i$th
-argument. This  transformation prefixes strings in 
-$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.
-
-Substituting  the  guessed   form  in  the  equations  describing
-\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
-containing $\sigma$ and those  without $\sigma$ on both sides of the
-resulting equation, we
-get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
-% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
-$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
-solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
-  \append}$, we get:
-  \begin{align*}
-    \Uf{\append}{1} &= \epsilonset
-    \cup \acdr.\Uf{\append}{1}  
-    &\quad 
-    \Uf{\append}{2} &= \Uf{\append}{2} \\
-    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
-    \cup \acdr\Df{\append}{1}\bcdr &\quad 
-    \Df{\append}{2} &= \epsilonset \cup
-    \Df{\append}{2}\bcdr 
-  \end{align*}
-Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.
-
-\subsection{Generating liveness equations \Lv\  for function bodies}
-\label{sec:bodylivenessbodies}
-
-\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
-
-We now calculate a 0-CFA-style summary liveness for each GC point of a program.
-There are two parts to this.
-First, for each function $f$,  we determine a {\em summary demand} $\sigma_f$
-over-approximating any demand $\sigma$ passed to $f$.
-%$\Lf{f}{i}{\sigma}$.
-So, consider a function $g$ containing a call site $\pi$ to $f$, say
-$\pi\!\!:\!(\LET~x \leftarrow (f\,y_1\,\ldots\,y_n)~\IN~e)$.
-Under the assumption that the demand on $g$ is $\sigma_g$, the liveness at $e$
-is $\Lv = \mathcal{L}(e,\sigma_g,\Lfonly)$, and the $\LET$ case of
-Figure~\ref{fig:live-judge} tells us this call site contributes $\Lv(x)$ to the
-demand $\sigma_f$ placed on  $f$.
-Write $\deltacall{f}{\pi}{g}$ for
-this demand (contributed by call site $\pi$ in $g$).
-
-Thus, supposing the $k$ call sites to function $f$ are
-$\pi^1$ (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
-%%% \marginpar{and writing $\sigma_f^i$ for the above $\deltacall{f}{\pi^i}{g^i}$?}
-then the over-approximation requirement on $\sigma_f$
-is achieved by taking $\sigma_f = \deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
-% is achieved by taking $\sigma_f = \sigma_f^1 \cup \cdots \cup \sigma_f^k$.
-
-The expression $e_{\mainpgm}$ is a special case; we assume it may be called externally
-with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$ (denoted
-$\sigma_{\!all}$).
-This is because any  part  of its  value  may be  used  by  the environment---for
-printing  the  result,  for  instance.
-
-% In what follows,  we shall overload the semantics of
-% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-% 
-% 
-% We assume an overall demand of $\{\acar,\acdr\}^*$ (denoted
-% $\sigma_{\!all}$) on the main expression $e_\mainpgm$.  This is because 
-% any  part  of its  value  may be  used  by  the environment---for
-% printing  the  result,  for  instance.
-
-%% To  generate  equations for  the  body $e_{\mathit{f}}$ of  a
-%% function,  we evaluate \Lfun{e_{\mathit{f}}}{\sigma_{\mathit{f}}}{\Lfonly}.     For ${\sigma_{\mathit{f}}}$,  we have
-%% to consider all calling  contexts during execution.  We therefore
-%% do a 0-CFA approximation by  defining
-%% % $\sigma_\mathit{f}$        as $\bigcup_\calls{f}{k}\!\!\sigma_\calls{f}{k}$,
-%% $\sigma_\mathit{f}$        as $\sigma_\calls{f}{1} \cup \cdots \cup \sigma_\calls{f}{n}$,
-%% where $\calls{f}{k}$,
-%% $\mathit{k} = 1 \ldots n$ represent all the calls to $\mathit{f}$
-%% in  the  program,   and  $\sigma_\calls{f}{k}$  are  the  demands
-%% generated at the calls  by the \rn{let-live} rule while analyzing
-%% the bodies of the caller.
-
-For the running example, $\append$ has calls from $\mainpgm$ at $\pi_9$
-and a recursive call at $\pi_5$.
-So $\sigma_{\append} =
-     \deltacall{\append}{\pi_9}{\mainpgm}  \cup \deltacall{\append}{\pi_5}{\append}$.
-Filling in  the values gives:
-%  where the former is the demand generated at the
-%  call to \append\  in $e_\mainpgm$ the latter
-%  is  the demand  at the  recursive call. 
-%% Propagation of
-%%  $\sigma_{\!all}$ in $e_\mainpgm$ gives    $\sigma_\calls{\append}{1}$   as   $\{\epsilon,   \acdr\}   \cup
-%%  \acdr\acar\sigma_{\!all}$  and propagation of
-%%  $\sigma_{\mathit{\append}}$ in $e_\append$ gives   $\sigma_\calls{\append}{2}$   as
-%%  $\bcdr\sigma_{\mathit{\append}}$.  
-%% Filling in  the values of $\sigma_\calls{\append}{1}$ and
-%% $\sigma_\calls{\append}{2}$, we get:
-\begin{align*}
-\sigma_{\append}    &=
-\left( \{\epsilon,  \acdr\} \cup  \acdr\acar\sigma_{\!all}  \right) ~\cup~
-\bcdr\sigma_{\mathit{\append}}
-\end{align*}
-
-Second, for each function $f$ (possibly $\mainpgm$) we need the liveness at
-each contained GC point
-$\pi\!\!:\!(\LET~x \leftarrow s~\IN~e)$,
-where $s$ is a call either to $\CONS$ or to a user function $g$.
-Given $\sigma_f$ calculated above, this
-is simply $\mathcal{L}(\pi,\sigma_f,\Lfonly)$.
-%%% \marginpar{AM: we've always said this (safe), but actual parameters
-%%% live before the call and dead after needn't be included!}
-
-For the running example, containing GC points $\pi_5, \pi_7$ in $\append$ and
-$\pi_9$ in $e_\mainpgm$, this gives
-(recall Equation~(\ref{eq:LF:DI}) above states $\Lf{f}{i}{\sigma} = \Uf{f}{i} \cup \Df{f}{i}.\sigma$):
-\begin{align*}
-\Lanv{5}{\xtl} &= \Lf{\append}{1}{\bcdr\sigma_{\append}}  \quad &
-                                \Lanv{7}{\xhd}\,\,\, &=  \bcar\sigma_{\append}     \\ 
-\Lanv{5}{\listb} &= \Lf{\append}{2}{\bcdr\sigma_{\append}} &
-                                \Lanv{7}{\xrec} &=  \bcdr\sigma_{\append} \\
-\Lanv{5}{\lista} &= \epsilonset \cup \acar\bcar\sigma_{\append} 
-\\[1ex]
-  \Lanv{9}{\py}  &= \Lf{\append}{1}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}} \\
-  \Lanv{9}{\pz}  &= \Lf{\append}{2}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}}
-\end{align*}
-
-\noindent 
-In summary, the equations generated during liveness analysis are:
-\begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
-\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
-% {\color{red}
-%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
-%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
-% }
-\end{enumerate}
-
-\subsection{Solving liveness equations---the grammar interpretation}
-The  equations above can now be re-interpreted as a
-context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
-a variable  $X$ occurring on  the LHS of the  equations generated
-from the analysis.  We can  think of the resulting productions as
-% being a grammar with multiple start symbols---one for
-% for each nonterminal \var{\Lanv{i}{x}}.
-being associated with several  grammars, one for each non-terminal
-\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
-productions:
- \begin{eqnarray*}
-\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
-  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
- \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
-  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
-  \acdr\var{\Uf{\append}{1}}\\         
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar               \mid
-  \acdr\var{\Df{\append}{1}}\bcdr\\
-\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
-\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
-\var{$\sigma_{\mathit{\!all}}$}
-\end{eqnarray*}
-Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
-denoted $\mathscr{L}(\var{\Lanv{i}{x}})
-$,  is the desired  solution of
-\Lanv{i}{x}.   
-However, recall from  our earlier  discussion that
-the  decision problem that  we are  interested in  during garbage
-collection is:
-\begin{quote}
-Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
-\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
-\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
-of forward paths only. Then does $\alpha \in \sigma$?
-\end{quote}
-We could convert  the  rules  defining  $\hookrightarrow$  into
-productions and  add them to the grammar. However, this results
-in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is  undecidable.   We
-circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
-  strongly  regular}  CFGs which  have   easy
-translations to non-deterministic finite state automata (NFA). The
-NFAs are then simplified on the
-basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
-membership of forward access paths. The resulting NFAs are   finally
-converted to DFAs for use during garbage collection.
-%------------------------------------------------------------%
-\subsubsection{Approximating CFGs using NFAs.}
-\label{sec:NFA-approx}
-
-We use the   
-algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
-approximate a CFG  to a   {\em strongly regular\/} grammar.  The
-transformation has  the  property that  if $\Lv$  is a
-non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
-Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
-  L}_{G'}(\Lv)$.    This  is   required  for the
-approximation to be safe with respect to liveness.
-
-We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
-transformation. We  pick the only production that  is affected by
-the transformation---the production for \Df{\append}{1}.  The
-production for \Uf{\append}{1}, while  recursive, is already in
-strongly  regular  form  and   is  therefore  unaffected  by  the
-transformation.
- \begin{eqnarray*}
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
-  \acdr\var{\Df{\append}{1}} \\ 
-\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
-\mid \epsilon 
-\end{eqnarray*}
-The  languages
-generated for  \var{\Df{\append}{1}} in  the original grammar  and the
-new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
-\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
-
-%------------------------------------------------------------%
-\subsubsection{Transforming NFAs to Accept Forward Paths:}
-%\label{sec:NFA-elim}
-The strongly regular CFGs obtained after the Mohri-Nederhof
-transformation are
-  first converted  into  NFAs.  The        algorithm         described        in
-Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
-\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
-Thus  \nt\  can  be  used  to  check  
-membership of forward paths.
-
-\begin{figure}[t]
-\begin{boxedminipage}{\textwidth}
-\begin{center}
-  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
-  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
-%
-  \raggedright{\bf Output:} NFA  \nt\ with  underlying
-  alphabet $\lbrace \acar, \acdr \rbrace$ such that
-  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
-  \\
-%
-
-  \raggedright{\bf Steps:}
- \begin{algorithmic}
-   \STATE $i \leftarrow 0$ 
-   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-   without $\epsilon$-moves \cite{hopcraft90toc}
-
-   \REPEAT
-   \STATE $\nt'_{i+1} \leftarrow \nt_i$
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcar$ and
-     outgoing edge to $q''$ with label $\acar$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-   label $\epsilon$.
-   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-   \ENDFOR
-
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcdr$ and outgoing edge to
-     $q''$ with label $\acdr$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-     label $\epsilon$.
-   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-   \ENDFOR
-
-   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-   $\nt'_{i+1}$ without $\epsilon$-moves
-
-   \STATE $i \leftarrow i+1$
-   \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-   \STATE $\nt \leftarrow \nt_i$
-%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
- \end{algorithmic}
-\end{center}
-\end{boxedminipage}
- \caption{Algorithm for transforming an NFA to  accept forward
-   paths only.}\label{algo:simplify-NFA} \figrule
-\end{figure}
-%------------------------------------------------------------%
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
-pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
-process is  continued till  a fixed point  is reached.   When the
-fixed point  is reached, the resulting NFA  contains all possible
-reductions corresponding  to all the  paths in the  original NFA.
-The proofs  of the termination  and correctness of  the algorithm
-are given in our earlier paper~\cite{karkare07liveness}.
-
-\input{liveness-dfa.tex}
-
-We illustrate the algorithm in Figure~\ref{algo:simplify-NFA} by
-constructing the automaton for \var{\Lanv{9}{y}}.
-Figure~\ref{fig:automata-construction} (c) shows the automaton for
-\var{\Lanv{9}{y}} constructed by composing the automata for
-\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
-\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
-get (d). We add an $\epsilon$ edge from $q_2$ to $q_3$ bypassing the
-\bcdr\acdr\ pair from $q_2$ to $q_2$ and then to $q_3$. This is shown
-in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
-is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
-$q_2$.  Removing this $\epsilon$ edge gives the automaton shown in
-(g).  Restricting this automaton to forward edges only, we get the
-final automaton shown in (h). This automaton recognizes $\acdr^{*}\!\mid\acdr^{*}\acar \sigma_{\mathit \!all}$, showing that the entire
-list \py, including its elements, is live at $\pi_9$. Also note that
-the language accepted by the final automaton satisfies the
-prefix-closed property.
-
-\section{Prototype and evaluation}\label{sec:experiments}
-To demonstrate the effectiveness of liveness-based garbage collection,
-we  have  built a  prototype  consisting  of  an interpreter  for  our
-language,  a  liveness  analyzer  and  a copying  collector  that  can
-optionally use the results of liveness analysis for marking instead of
-reachability. When the collector uses liveness for marking, we call it
-a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-  reachability-based  collector}  (RGC).   The  copying  collector  is
-neither incremental nor generational.  As a consequence, any cell that
-becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.
-
-While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
-is invoked  at the program point   $\pi$,  it would  use  the
-liveness  DFAs at  $\pi$ itself for exploring the root set of the
-current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.
-
-Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-$\dfa{\pi}{x}$.    
-Considering   a   DFA   as   a   table,
-$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-\acar\      or     \acdr.       We      shall     also      write
-$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-indicating  whether there  is a  transition from  $q$ on
-$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-pointer to $x$.
-The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-$x$  in the  other semi-space  and returns  the new  address.  It
-hides  details such as  returning the  forwarding pointer  if the
-value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.
-
-The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-in the heap over time for RGC and LGC---here time is measured in terms
-of the number of cons cells allocated. In addition, they also show the
-number of  reachable cells and the  number of cells  that are actually
-live\footnote{This   is  statically   approximated  by   our  liveness
-  analysis}.  Since  the programs have  different memory requirements,
-we have tuned the size of heap for each program to ensure a reasonable
-number of  collections. An invocation  of RGC decreases the  number of
-cells  in heap  till  it touches  the  curve of  reachable cells.   An
-invocation of LGC decreases the number  of heap cells to no lower than
-the curve of live cells.
-
-To construct the  reachable and live curves, we  record for every cell
-its creation time (\CreateTime), its last use time (\UseTime), and the
-earliest time  when the  cell becomes unreachable  and can  be garbage
-collected (\CollTime).  For accurate recording of  \CollTime, we force
-frequent invocations  of a reachability based collector  in a separate
-run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
-  fibheap} increases  initially, comes  down before the  heap becomes gets
-full and then  continues to remains low.  This results in  a single garbage
-collection in both RGC and LGC.  The graphs also show the precision of
-our liveness analysis.  For  all programs  except  {\tt nperms} and
-{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.
-
-\subsection{Results}
-\begin{figure}[t] 
-\centering
-\begin{boxedminipage}{.75\textwidth}
-{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
-\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-  then} {\bf skip}\\ 
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
-\noindent \hspace*{.35cm}{\bf return} $y$
-\end{boxedminipage}
-\caption{Function       for       copying       a       root       set
-  variable.}\label{algo:LGC-main-loop} \figrule
-
-\end{figure}
-
-\begin{figure}[p]
-\begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
- &
-{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-\end{tabular}
- \caption{Memory  usage of  programs. The blue and the red curves
-  indicate the number of cons cells in the active semi-space for RGC and
-  LGC respectively. The black curve represents the number of reachable
-  cells and the grey curve represents the number of cells that are
-  actually live (of which liveness analysis does  a static
-  approximation). x-axis is the time measured in number of cons-cells
-  allocated. y-axis is the number of cons-cells.}
-\label{fig:memory-usage} \figrule
-\end{figure}
-The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-provides  statistics  regarding the  analysis  itself.  The number  of
-states and the analysis times  are within tolerable limits. Precision
-of analysis refers to the percentage of dead cells that is collected by
-LGC, averaged over all invocations.
-The second table shows garbage  collection statistics for RGC and LGC.
-LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to run
-in comparison with  RGC.
-
-There are a couple of  issues of concern.  The garbage collection time
-is larger in the case of LGC for some programs. The reason is that the
-cost of consulting the liveness DFA may outweigh the combined benefits
-of  fewer   garbage  collections   and  fewer  markings   per  garbage
-collection.   The  other issue  is  illustrated  by  the program  {\tt
-  lambda}    (memory   usage  shown   separately    in
-Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-in Figure~\ref{fig:experimental-results}, the  number of
-touched cells\footnote{These  are the  cells which are  visited during
-  marking phase, often  more than once due to  sharing.} in this
-example is much higher for
-LGC.  This  increase is due to  excessive sharing among  heap nodes in
-this program.  Note  that a node re-visited because  of sharing is not
-explored  any   further  during  a  RGC   collection.   However,  this
-curtailment cannot happen in  LGC because of the possibility that the
-node,  re-visited in a  different liveness  state, may  mark a  set of
-cells different from the earlier visit.
-\begin{figure}[t]
-\footnotesize
-
-\centering
-
-\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline
-{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-  nperm} &  {\sf fibheap} & {\sf knightstour} &
-{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-\hline
-\hline
-{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-\hline
-\end{tabular}}
-
-\centerline{(a)}
-
-\medskip
-
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
-  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-
-                            &   \multicolumn{2}{c|}{cells per GC}  
-                            &   \multicolumn{2}{c|}{cells per GC} 
-                            &   \multicolumn{2}{c|}{\#GCs} 
-                            &   \multicolumn{2}{c|}{MinHeap} 
-                            &   \multicolumn{2}{c|}{Avg. Drag} 
-                            &   \multicolumn{2}{c|}{(sec)} \\
-\cline{2-13}
-{Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
-%    {\sf  nboyer} & &  & &   & &  & & &  \\
-%    {\sf   circsim?} & &  & &   & & & & &   \\
- \hline
-\end{tabular}}
-
-\centerline{(b)}
-\vskip -2mm
-\caption{Experimental results comparing RGC and LGC. Table (a) gives
-  data related to liveness analysis, and (b) gives garbage collection
-  data. The average drag time is measured in allocated cons cells.}
-\label{fig:experimental-results}
-\normalsize
-\end{figure}
-
-\section{Collecting more garbage can never slow things down}
-\label{sec:lgc-always-better}
-Since garbage collection is  effectively asynchronous to the allocator
-thread, one might  worry as to how {\em  robust} our measurements are.
-For example,  while LGC would,  in general, collect more  garbage than
-RGC in the same heap state,  is it possible that on some programs, LGC
-might do  a larger number  of collections? We  prove a result  to show
-that   this  cannot   happen.    This  result   applies  to   classical
-mark-and-sweep and copying garbage  collectors and, we hypothesize, to
-generational collectors too.
-
-\begin{lemma}
-For the  same mutator,  a liveness-based collector  can never  do more
-garbage collections than a reachability-based collector.
-\end{lemma}
-
-\begin{proof}Assume, as before,  that time is measured in terms of the number of
-  cons cells allocated. Let us run  two copies of the mutator with the
-  two    different   collectors    but    with   synchronous    memory
-  allocations.   The   garbage    collections,   however,   happen
-  asynchronously.
-
-  To prove the lemma, it is enough to show the truth of the following 
-statement: After
- every  LGC invocation,  the count of LGC invocations   is  no greater
-than  RGC invocations.   The  base case  holds  since the  first
-invocations of both GCs happen at the same time.  Assume the statement
-to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-reachable cells, its heap would contain no more cells than RGC heap at
-the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-LGC, or LGC and RGC are both invoked next at the same time.  In either
-case, the statement holds after $n+1$ invocations of LGC.
-\end{proof}
-
-%% \begin{figure}[h]
-%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
-%% \caption{Memory  usage of  programs. The  solid  lines represents
-%%   number of  live cells and the  dashed lines show  the number of
-%%   reachable cells.  The red and blue  lines represent invocations
-%%   of RGC and  LGC---their heights indicate  the amount of memory
-%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
-%%       paper and are being used as placeholder}}}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-%%\clearpage
-
-\section{Related Work} Previous attempts to  increase  the space efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
-reallocation~\cite{jones89compile},                       deforestation
-techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
-methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
-
-\begin{figure}[t]
-\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
- \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
-\label{fig:memory-usage-lambda} \figrule
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-We have defined a notion of liveness on structured data;
-this generalizes classical liveness and strong liveness.
-We started with a general fully context-sensitive analysis which
-we proved correct with respect to a minefield semantics
-(this models the effect of garbage collection between every
-evaluation step).
-
-To avoid scalability issues (and to avoid performing part of
-the liveness computation at run time)
-we defined an 0-CFA version of this liveness analysis in which
-demands for function $f$ at all calling contexts are conflated into 
-a single demand $\sigma_f$.
-This enabled us to treat the liveness equations symbolically obtaining
-context-free grammars for liveness at each GC point (calls to user
-functions and to $\CONS$).
-These were then converted to DFAs for run-time consultation by the
-garbage collector. Experiments confirm the precision of the analysis.
-
-To obtain performance figures we compared a reachability-based garbage
-collector with a liveness-based  collector.  This showed a decrease in
-the  number  of  GCs,   more  garbage  collected  per  invocation.   A
-significant benefit of LGC is  that programs can run in smaller memory
-when compared to  RGC. This is potentially useful  in situations where
-memory  is  limited---as with  embedded  systems.  For  a majority  of
-programs, the garbage collection times were reduced.
-
-One issue we  highlighted was that while fewer  nodes were marked (and
-more garbage collected), sometimes  cons cells could be visited and
-traversed  multiple times  with different  sets of  liveness  paths to
-explore.  Further work includes  improvements to the classical copying
-collector to reduce the cost of this.
-
-\subsection*{Acknowledgements}
-The authors  would like  to thank Hemanshu  Vadehra for  a preliminary
-implementation of the prototype system. Amey Karkare was supported for
-this work by the DST/SERC fast track scheme for young scientists.
-
-\bibliography{fun_hra}{}
-\bibliographystyle{splncs}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-\section{Appendix stuff}
-
-\paragraph{Safety of using the modified \CONS\  rule:}
-Recall that while the correctness of the analysis was proved with
-respect to  the $\len$ rules  in Figure~\ref{fig:live-judge}, the
-equations   for  liveness   are  based   on  a   modification  of
-$\mathit{ref}$    in    the    case    of   \CONS\  as  given    by
-(\ref{eqn:mod-cons}).  We now show that  if the \bcar\,s and \bcdr\,s
-in live access  paths resulting  from this  modification are
-eliminated through  applications of $\hookrightarrow$,  we obtain
-the same results as given by the original rules.
-
-\comment{Proposal: We prove a result which goes as follows:
-
-\begin{quote}
-Lemma: Let  $\Lanv{i}{x}= \sigma$ for an a  program point $\pi_i$
-and a variable  $x$, when the judgement is  based on the original
-$\len$  rules.   Further, let  $\Lanv{i}{x}=  \sigma'$, when  the
-judgement is based on $\len$ with the modified \CONS\ rule.  Then
-$\sigma' \stackrel{*}{\hookrightarrow} \sigma$
-\end{quote}
-
-Here are some  ideas regarding how this can  be proved.  Consider
-an arbitrary equation
-$$ \Lanv{j}{y} = \ldots \Lanv{i}{x} \ldots$$ generated by the
-analysis.
-
-Now assume that $\Lanv{i}{x}= \sigma$.   We want to find out what
-happens to the access patterns in $\sigma$ in the access patterns
-of \Lanv{j}{y}.   To answer this,  we put \# markers  around each
-$\alpha \in \sigma$  and redo the equation using  an extension of
-$\len$ to take  care of \#-tagged patterns.  We  can then observe
-occurrences of \#-tagged patterns  in the liveness information at
-$\pi_j$.
-
-Here is  the $\len$ extension for  the original \CONS\  clause to take
-care of the \#-tagged patterns:
-$$\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          = \{x.\#\alpha\# \mid \#\acar\alpha\# \in \sigma\} \cup
-\{y.\#\alpha\# \mid \#\acdr\alpha\# \in \sigma\} $$
-And here  is  the $\len$ extension for  the modified \CONS\  clause.
-$$ \mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-= \{x.\#\bcar\alpha\# \mid \#\alpha\# \in \sigma\} \cup
-\{y.\#\bcdr\alpha\# \mid \#\alpha\# \in \sigma\} $$
-All  other rules  remain unchanged.  Tracking a  pattern  through this
-extension is sound because erasing the \#-tags would give us the same
-liveness as the original rules.
-
-Let us also extend $\hookrightarrow$ to \#-tagged strings as:
-$$\sigma_1  {\hookrightarrow} \sigma_2\,\, \mathit{iff}\,\, \sigma_2 =
-\{\alpha_1 \#\alpha\# \alpha_2 \mid  \alpha_1 \#\alpha'\# \alpha2 \in
-  \sigma_1\,\, and \,\, \{\alpha'\} \stackrel{*}{\hookrightarrow} \{\alpha\}\}$$
-
-Now we have an  observation represented by the following diagram:
-
-\centerline{\epsfig{file=diag1.eps, height=4cm}}
-
-We need  to make this  observation general enough.   Therefore we
-should prove  the observation even if \Lanv{i}{x}  = $\sigma$ had
-patterns with \bcar\ and \bcdr.   We should be able to prove this
-by structural induction on the rules.
-
-How does  the observation prove  the original lemma?   Suppose we
-replayed  judgements  with  ($\len$  +  \#) and  with  ($\len$  +
-modified \CONS +  \#).  Suppose that $\pi_i$ is  the last program
-point till which the results  are the same (because no \CONS\ was
-encountered).  Now  consider a  $\sigma \in \Lanv{i}{x}$  for any
-$x$.  Put \#-tags around each string in $\sigma$. Now we have the
-diagram:
-
-\centerline{\epsfig{file=diag2.eps, height=7cm}}
-
-This shows that modulo erasure of \#-tags, using the original $\len$
-judgement and  $\len$ judgement with modified \CONS\ rule followed by
-$\hookrightarrow$ give the same liveness.
-
-And I would like  to conclude that it is all right  to do our analysis
-with $\len$ + modified \CONS\ and then simplify using $\hookrightarrow$.
-
-}
-
-{\color{blue}\marginpar{From sec 3.1, liveness judgements}
-\paragraph{Semantic considerations.}
-The incremental liveness for generic primitive $\PRIM$ is given by
-$\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly) = 
-           \{x.\epsilon,y.\epsilon\}$.
-Earlier we suggested that classical strong liveness for addition,
-but not division, could use the stronger
-$ \{x.\epsilon,y.\epsilon \mid \epsilon \in \sigma\}$.
-However this is invalid: garbage collection
-could reuse dead, but reachable, $\CONS$ cell fields (of $v$ below)
-thereby producing a stuck state in a dynamically typed language.
-Hence we treat $+$ as like division not like addition---suppose
-% $(FOO)$ is a user function which may cause garbage collection, and consider:
-calling $\mbox{\footnotesize (FOO)}$ garbage collects and reallocates the cell pointed
-to by `v' in:
-
-\vspace{-15pt}
-
-\begin{quote}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1}  (\LET\ v\ $\leftarrow $\ (\CONS~2~3) \IN
-	  \UNL{2}  (\LET\ w\ $\leftarrow $\ (FOO) \IN
-	  \UNL{3}  (\LET\ x\ $\leftarrow $\ (\PRIM\ (\CAR\ v) 4) \IN
-	  \UNL{4}  (\RETURN\ w))))
-	\end{uprogram}}
-\end{quote}
-In principle, {\em dead code removal} of
-$(\PRIM\ (\CAR\ {\rm v})\ 4)$
-would render the stronger version valid, but still requires static typing
-to avoid the risk of {\em removing} a stuck state.
-We restrict ourselves to liveness annotation rather than code transformation.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]
-\footnotesize
-Old version of the liveness rules...
-
-\infrule[ret-live]
-        {}
-        {(\RETURN~x),\sigma,\Lfonly \len  x.\sigma}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[if-live]
-        { e_1, \sigma,\Lfonly \len \Lv_1
-          \andalso e_2, \sigma,\Lfonly \len  \Lv_2}
-        {(\SIF~x~e_1~e_2),\sigma,\Lfonly \len
-               \Lv_1\cup\Lv_2\cup \{x.\epsilon\}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[let-live]
-        {e, \sigma,\Lfonly  \len  \Lv}
-        {(\LET~x \leftarrow  s~\IN~e),
-          \sigma,\Lfonly \len   \Lv\setminus \{x.*\} \cup
-          \mathit{ref\/}(s,\Lv(x),\Lfonly)}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\infrule[live-define]
-        {e_f, \sigma, \Lfonly  \len
-%            \bigcup \{z_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-           \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \fen \Lfonly
-\\ \mbox{where }
-     (\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ \ e_f) \mbox{ is a member of }
-        d_1 \ldots d_k}
-\figrule
-  \caption{Liveness Judgement Rules}\label{fig:live-judge-old}
-\normalsize
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \newgray{lightgray}{.75}
-%\newcommand{\dem}[1]{\red{\scriptsize{$#1$}}}
-\newcommand{\dem}[1]{\psframebox[linestyle=none,fillstyle=none,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-%\newcommand{\liv}[1]{\blue{\scriptsize{$#1$}}}
-\newcommand{\liv}[1]{\psframebox[linestyle=none,fillstyle=solid,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-\psset{unit=1mm,nodesep=1mm,labelsep=0.5mm,arrows=->,linewidth=.1,framesep=0.2mm}
-\begin{center}
-  \begin{pspicture}(0,0)(120,50)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(12,5)
- \footnotesize
-   \putnode{let1}{origin}{15}{50}{$\pi_1$:\LET}
-
-    \putnode{test1}{let1}{-12}{-7}{\xtest}
-    \putnode{null}{let1}{0}{-7}{$\pi_2$:\NULLQ}
-    \putnode{lista1}{null}{0}{-7}{$\pi_3$:\lista}
-    \putnode{if}{let1}{18}{-7}{$\pi_4$:\SIF}
-
-    \putnode{test2}{if}{-8}{-7}{$\pi_5$:\xtest}
-    \putnode{listb1}{if}{3}{-7}{$\pi_6$:\listb}
-    \putnode{let2}{if}{14}{-7}{$\pi_7$:\LET}
-
-    \putnode{tl1}{let2}{-22}{-7}{\xtl}
-    \putnode{cdr}{let2}{-5}{-7}{$\pi_8$:\CDR}
-    \putnode{lista2}{cdr}{0}{-7}{$\pi_9$:\lista}
-    \putnode{let3}{let2}{15}{-7}{$\pi_{10}$:\LET}
-
-\putnode{rec1}{let3}{-13}{-7}{\xrec}
-\putnode{append}{let3}{0}{-7}{$\pi_{11}$:\append}
-\putnode{tl2}{append}{-10}{-7}{$\pi_{12}$:\xtl}
-\putnode{listb2}{append}{5}{-7}{$\pi_{13}$:\listb}
-
-\putnode{let4}{let3}{17}{-7}{$\pi_{14}$:\LET}
-\putnode{hd1}{let4}{-5}{-7}{\xhd}
-\putnode{car}{let4}{4}{-7}{$\pi_{15}$:\CAR}
-\putnode{lista3}{car}{0}{-7}{$\pi_{16}$:\lista}
-
-\putnode{let5}{let4}{16}{-7}{$\pi_{17}$:\LET}
-
-\putnode{ans1}{let5}{-4}{-7}{\xans}
-\putnode{cons}{let5}{6}{-7}{$\pi_{18}$:\CONS}
-\putnode{hd2}{cons}{-10}{-7}{$\pi_{19}$:\xhd}
-\putnode{rec2}{cons}{10}{-7}{$\pi_{20}$:\xrec}
-\putnode{ans2}{let5}{19}{-7}{$\pi_{21}$:\xans}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-
-%% \putnode{x}{let1}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{if}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{let3}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let4}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{-8}{0}{\liv{\Lv_1}}
-%% %\putnode{x}{ans2}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{plus}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{rec2}{1}{4}{\dem{\epsilonset}}
-%% \putnode{x}{plus}{-7}{0}{\liv{\Lv_6}}
-%% %\putnode{x}{let4}{-7}{0}{\liv{\{\xrec\}}}
-%% \putnode{x}{len}{3}{3}{\dem{\epsilonset}}
-%% \putnode{x}{len}{-7}{0}{\liv{\Lv_3}}
-%% \putnode{x}{tl2}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{tl2}{-6}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let3}{-10}{0}{\liv{\xtl.\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l2}{0}{4}{\dem{\epsilonset \cup \acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{l2}{-5}{0}{\liv{\Lv_4}}
-%% \putnode{x}{zero}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{zero}{-3}{0}{\liv{\Lv_0}}
-%% \putnode{x}{test2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{test2}{-7}{0}{\liv{\Lv_5}}
-%% \putnode{x}{if}{-6}{0}{\liv{\Lv_3}}
-%% \putnode{x}{null}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{null}{-9}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l1}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{vone2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{one}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{let5}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{l1}{-5}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{let1}{-17}{-1}{\liv{\{\xl\} \cup \xl.\acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{let1}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let1}{-7}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{if}{7}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let2}{8}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let3}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let3}{-7}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let4}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let4}{-7}{0}{\liv{\Lv_2}}
-%% \putnode{x}{vone2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{rec2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{let5}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{one}{-6}{0}{\liv{\Lv_2}}
-\ncline{-}{let1}{test1}
-\ncline{-}{let1}{null}
-\ncline{-}{let1}{if}
-\ncline{-}{null}{lista1}
-
-\ncline{-}{if}{test2}
-\ncline{-}{if}{listb1}
-\ncline{-}{if}{let2}
-\ncline{-}{let2}{tl1}
-\ncline{-}{let2}{cdr}
-\ncline{-}{let2}{let3}
-\ncline{-}{cdr}{lista2}
-
-\ncline{-}{let3}{rec1}
-\ncline{-}{let3}{append}
-\ncline{-}{let3}{let4}
-
-\ncline{-}{append}{tl2}
-\ncline{-}{append}{listb2}
-
-\ncline{-}{let4}{let5}
-\ncline{-}{let4}{hd1}
-\ncline{-}{let4}{car}
-\ncline{-}{car}{lista3}
-%% \ncline{-}{len4}{tl2}
-%% \ncline{-}{let4}{vone1}
-%% \ncline{-}{let4}{one}
-%% \ncline{-}{let4}{let5}
-\ncline{-}{let5}{ans1}
-\ncline{-}{let5}{cons}
-\ncline{-}{cons}{hd2}
-\ncline{-}{cons}{rec2}
-%% \ncline{-}{let5}{plus}
-\ncline{-}{let5}{ans2}
-\end{pspicture}
- \end{center}
-
-  \caption{Example to illustrate liveness computation}\label{fig:live-examp}
-\figrule
-\normalsize
-\end{figure}
-
-\newpage
-
-\begin{figure}[t]
-  \begin{tabularx}{320pt}{|| c | X | X | X ||}
-\hline
-Pgm  Pt & Demand  & Pre & Post \\
-\hline 
-$\pi_1$ & $\sigma$ & \{\lista\} $\cup$ \listb.$\sigma$ $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & $\phi$ \\
-\hline
-%% $\pi_2$ & $\epsilonset$ & \{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_3$ & $\epsilonset$ &\{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_4$ & $\sigma$ &\{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-%% $\pi_5$ & \epsilonset & \{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  \\
-%% \hline
-%% $\pi_6$ & $\sigma$ & \listb.$\sigma$ & $\phi$ \\
-%% \hline
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_7$ & $\sigma$  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$\\
-%% \hline
-%% $\pi_8$ & \Lf{\append}{1}(\{\bcdr\}$\sigma$) & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_9$ & \epsilonset $\cup$ \{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_{10}$ &$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-$\pi_{11}$ & \{\bcdr\}$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \lista $\cup$ \lista.\{\acar\bcar\} \\
-\hline
-$\pi_{12}$ &\Lf{\append}{1}(\{\bcdr\}$\sigma$)   &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-$\pi_{13}$ & \Lf{\append}{2}(\{\bcdr\}$\sigma$)  &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{14}$ & $\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$ \\
-\hline
-%% $\pi_{15}$ & \{\bcar\}$\sigma$  & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   &  \xrec.\{\bcdr\}$\sigma$\\
-%% \hline
-%% $\pi_{16}$ & \epsilonset $\cup$ \{\acar\bcar\}$\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$  & \xrec.\{\bcdr\}$\sigma$ \\
-%% \hline
-$\pi_{17}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   & $\phi$  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{18}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$  \\
-\hline
-$\pi_{19}$ &  \{\bcar\}$\sigma$        & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{20}$ &   \{\bcdr\}$\sigma$       & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{21}$ & $\sigma$ & \xans.$\sigma$ & $\phi$  \\
-\hline
-\end{tabularx}
-  \caption{Liveness annotations for the body of \append}\label{fig:xe_append}
-\figrule
-\normalsize
-\end{figure}
-
-\end{document}

### 2014-07-13 paper/sections/intro.tex
organization added

--- 2014-07-13 paper.tex
@@
-%------------------------------------------------------------%
-\documentclass{llncs}
-\pagestyle{plain}
-\usepackage{longtable} 
-%------------------------------------------------------------%
-%- common stuff starts here....
-\usepackage{etex}
-\usepackage{float}
-\usepackage{mathrsfs}
-\usepackage{booktabs}
-\usepackage{boxedminipage}
-\usepackage[T1]{fontenc}
-%\usepackage[pdf]{pstricks}
-\usepackage{epsfig}
-\usepackage{multirow}
-\usepackage{url}
-\usepackage[normalem]{ulem}
-\usepackage{schemepgm}
-\usepackage{graphicx}
-\usepackage{graphpap}
-\usepackage{tabularx}
-%\usepackage{color}
-%\usepackage{xcolor}
-\usepackage{algorithmic}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{pstricks}
-\usepackage{pst-text}
-\usepackage{pst-node}
-\usepackage{pst-tree}
-\usepackage{pst-rel-points}
-%\usepackage{pst-all}
-\usepackage{bcprules}
-%\usepackage{subfigure}
-%\usepackage{algorithmic}
-%\usepackage{anysize}     
-\input{fh_defs}
-%\newcommand{\comment}[1]{}
-\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
-\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
-    #1 \medskip \hrule \medskip}}}
-\newcommand{\scmin} {\mbox{\sf\em in}}
-\newcommand{\scmnull}{\mbox{\sf\em null?}}
-\newcommand{\scmpair}{\mbox{\sf\em pair?}}
-\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
-\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
-
-\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
-\newcommand{\stk}{\mbox{S}}       % stack
-\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
-\newcommand{\bang}{\mbox{\sc bang}}
-
-\begin{document}
-%  \title{Heap  Reference  Analysis  for Functional  Programs}
-\title{Liveness-Based Garbage Collection}
-
-% make a proper TOC despite llncs
-\setcounter{tocdepth}{2}
-\makeatletter
-\renewcommand*\l@author[2]{}
-\renewcommand*\l@title[2]{}
-\makeatletter
-
-  \author{Rahul Asati\inst{1} \and Amitabha Sanyal\inst{1} \and Amey Karkare\inst{2}
-    \and   Alan Mycroft\inst{3}}
-  \institute{IIT Bombay, Mumbai 400076, India\\
-    \email{$\{$rahulasati,as$\}$@cse.iitb.ac.in}, \and 
-             IIT Kanpur, Kanpur 208016, India\\
-    \email{karkare@cse.iitk.ac.in}, \and 
-             Computer Laboratory, University of Cambridge, CB3 0FD, UK\\
-    \email{alan.mycroft@cl.cam.ac.uk}}
-\maketitle
-
-\begin{abstract}
-Current garbage collectors  leave much heap-allocated data uncollected
-because they preserve data {\em reachable} from a root set.
-% However, it is only necessary to preserve {\em live} data---a subset of
-% reachable data.
-However, only  {\em live} data---a subset of  reachable data---need be
-preserved.
-
-Using   a    first-order   functional   language    we   formulate   a
-context-sensitive liveness  analysis for structured data  and prove it
-correct.   We  then use  a  0-CFA-like  conservative approximation  to
-annotate  each  allocation  and  function-call program  point  with  a
-finite-state  automaton---which   the  garbage-collector  inspects  to
-curtail reachability  during marking. As  a result, fewer  objects are
-marked  (albeit  with a  more  expensive  marker)  and then  preserved
-(e.g.\ by a copy phase).
-% and secondly because of the consequent
-%reduced number of collections during a computation.
-
-Experiments  confirm the  expected performance  benefits---increase in
-garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections, a decrease  in the memory size required  to run programs,
-and  reduced  overall  garbage  collection  time  for  a  majority  of
-programs.
-\end{abstract}
-
-%------------------------------------------------------------%
-
-%% \begin{figure}
-%% \tableofcontents       %%%%%%% BEWARE: find the \setcounter{page}{2}
-%%                        %%%%%%% and comment it out when deleting this TOC.
-%% \vspace{5cm}
-
-%% \section*{Page Budget}
-%% CC'2014 and ESOP'2014 have a limit of 20 pages.
-%% A draft page budget is:
-%% \begin{itemize}
-%% \item 1. abstract/introduction (3 pages)
-%% {\color{red}Currently 2.5 pages.}
-%% \item 2. core language syntax and semantics (2 pages).
-%% {\color{red}Currently 2 pages.}
-%% \item 3. liveness analysis and correctness (3 pages)
-%% {\color{red}Currently 3.8 pages.}
-%% \item 4. practical liveness computation (6 pages)
-%% {\color{red}Currently 6.2 pages.}
-%% \\ I think this is where should do all 3 of: the context-insensitive analysis
-%% (one which has a summary demand for each function), then the
-%% solving like differential equations, then using Mohri-Nederhof to
-%% get a NFA for each (call/cons) program point.
-%% \item 5. experimental framework and results (4 pages)
-%% \item 6. (optional) monotonicity of GC (0.5 page)---our technique can never
-%% cause more collections.
-%% \item 7. conclusions/acknowledgements/bibliography (1.5 page)
-%% \end{itemize}
-%% \end{figure}
-
 \section{Introduction}
@@
 % and Section~\ref{sec:conclusion} concludes the paper.
-
-\section{The target language---syntax and semantics}
-\label{sec:defs}
-We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
-functions ($\CONS$, $\PRIM$ etc.).
-The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
-it has eager semantics and restricts programs to be in 
-Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
-where all actual parameters to functions are variables.
-This restriction does not affect expressibility (and indeed we feel free
-to ignore it in examples when inessential), but simplifies the analysis formulation.
-Additionally, as in the three-address instruction form familiar from compiler texts,
-it forces each temporary to be named and function calls to be serialized
-(necessary to get an unambiguous definition of liveness).
-We further require that each variable in a program is distinct, so that no scope shadowing
-occurs---this simplifies proofs of soundness.
-In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
-or return ($\RETURN$).
-The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.
-
-The body of the program is
-the  expression denoted  by  $e_\mainpgm$; for analysis purposes
-it is convenient
-to regard $e_\mainpgm$ as part of a function definition
-$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
-% As  mentioned earlier,
-We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.
-
-% \paragraph{Inlining.}
-In spite of the ANF restrictions it is still possible
-to inline non-recursive functions (a fact we use to prove the safety
-of liveness analysis).  A user-function call
-$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
-to a function defined
-(after renaming its formals and locals to be disjoint from existing variables)
-by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
-is replaced by a sequence of $\LET$'s of the form
-$z_i\leftarrow (\ID\,\,y_i)$ followed by
-the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
-$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.
-
-\begin{figure}[t!]
-\footnotesize
-\begin{eqnarray*}
-   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
-    \hspace{7em} \mbox{\em --- program}\\
-    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
-    e) 
-    \hspace{1.7em} \mbox{\em --- function definition} \\
-e \in \mathit{Expr} & ::= &
-\left\{\begin{array}{ll@{\hspace{4em}}l}
-       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
-       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
-       (\RETURN\,\, x) && \mbox{\em --- return from function}
-    \end{array}\right. \\
-s \in \mathit{Stmt} & ::= &
-\left\{\begin{array}{lr@{\hspace{1em}}l}
-       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
-       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
-       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
-       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
-%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
-       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
-%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
-       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
-            & \mbox{\em --- function application} 
-    \end{array}\right.
-\end{eqnarray*}
-  \caption{The syntax of our language}\label{fig:lang-syntax}
-\figrule
-\normalsize
-\end{figure}
-
-\subsubsection{Semantics}
-We now give an operational
-semantics for  our language.  Later, a refinement of
-the  operational semantics,  which we  call minefield  semantics, will
-serve to prove liveness analysis correct.
-We give a small-step semantics because, unlike big-step semantics,
-correctness for non-terminating programs does not need special treatment.
-We start with the domains used by the semantics:
-\[
-\begin{array}{rlcl@{\hspace{2em}}l}
-% \begin{eqnarray*}
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
-\mbox{-- Environment} \\ 
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
-  Val}) & \mbox{-- Heap}
-% \end{eqnarray*}
-\end{array}
-\]
-Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
-A value is either a number, the empty list $\NIL$, or a location  $\ell$.
-Our liveness analysis does not track numeric values, and thus is neutral
-as to whether these are boxed or represented as immediates.
-An environment is a finite mapping from variables to values, and a heap a finite mapping
-from locations to pairs of values.
-% Given  a program,  an  environment is  a
-% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
-% values in $\mathit{Val}$.    The heap is a mapping
-% from  locations  to pairs of  values.
-Finally, $\stk$ is a stack
-(using $\bullet$ for push and $[\,]$ for empty stack)
-of frames of unfinished function calls.  A
-frame is a triple $(e,x,\rho)$ representing the call site
-$
-(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
-$
-being evaluated in environment $\rho$.
-Frames can also be viewed as continuations, in this view
-the ({\sc ord-return}) 
- rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
-
-The semantics 
-of statements $s$ are given by the judgement form
-$\rho, \heap, s \rightsquigarrow \heap', v$
-and those for expressions $e$ by the form
-$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
-The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
-program terminates successfully with result value $\rho(x)$
-on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$
-
-Notation:  we write $\rho[x \mapsto v]$ for the environment which is
-as $\rho$ but has value $v$ at $x$.
-We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
-has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
-$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
-$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
-
-\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
-with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
-a contained $s$).  Some of these we eliminate syntactically, e.g.\
-ensuring all variables and functions are defined and are called with
-the correct number of parameters.
-Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
-All but the first can be eliminated with a static type system but,
-treating our program as dynamically typed, we regard all these
-as stuck states.
-
-% A program terminates successfully if it reaches the halt state
-% in Figure~\ref{fig:lang-opsem-small}.
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-\begin{figure}[t!]  
-% (Random text to ensure proper line-spacing (problem with infrule macro?).)
-
-\footnotesize
-\begin{minipage}{0.335\textwidth}
-\infrule[ord-const]
-        {}
-        {\rho, \heap, \pp{\pi}{k}
-          \rightsquigarrow \heap, k}
-\end{minipage}\begin{minipage}{0.655\textwidth}
-\infrule[ord-cons]
-      {
-\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
-}
-      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
-        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
-        \ell}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-car]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
-          \rightsquigarrow \heap, v_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-cdr]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
-          \rightsquigarrow \heap, v_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-id]
-      {
-}
-      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
-        \rightsquigarrow \heap, \rho(x)}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-prim]
-      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
-      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
-        \rightsquigarrow \heap, \rho(x) + \rho(y)}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-null-false]
-\infrule[]
-        {\rho(x)\not= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 0}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-null-true]
-\infrule[ord-null]
-        {\rho(x)= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 1}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-if-true]{
-\infrule[]{
-  \rho(x) \in \mathbb{N}\setminus\{0\}
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-if-false]{
-\infrule[ord-if]{
-  \rho(x) = 0
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-nonfn]{
-%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
-  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
-  \andalso
-  \mbox{$s$ is not
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-fncall]{
-  \mbox{$s$ is 
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-\andalso
-\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
-[\myvec{z} \mapsto \rho(\myvec{y})],
-  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
-}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-return]{
-}{
-  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
-  \pp{\pi}{(\RETURN\ x)}
-  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
-\end{minipage}
-
-% \normalsize
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-  \caption{The small-step operational semantics}
-  \label{fig:lang-opsem-small}
-\figrule
-\normalsize
-\end{figure}
-
-%==============================================================
-\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-
-\section{Liveness}\label{sec:liveness}
-
-In classical liveness analysis a variable is either `live' (its value
-may be used in future computation) or `dead' (definitely not used).
-Semantically, a variable is dead at a given program point
-if arbitrary changes to its value have no effect on the computation.
-Later we will use $\bot$ to represent a value which `explodes' when
-it is used in a computation; dead variables can safely have their
-value replaced with $\bot$.
-For heap-allocated data  we need a richer model of liveness
-in that both variables and fields of $\CONS$ cells may be dead or live.
-Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
-liveness of the structure reachable from a variable is a set of
-{\em access paths}
-which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
-grammar notation.
-Thus the liveness of $x$ being
-$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
-only refer to the second and third members of $x$ considered as a list.
-Semantically, access paths are prefix-closed, as accessing a field
-requires accessing all the paths from the variable to the field,
-and hence the above liveness is properly written 
-$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
-The classical notions of a scalar variable being live or dead correspond to
-$\{\epsilon\}$ and $\{\}$.
-
-The overall liveness (also written {\em liveness environment} for emphasis)
-at a program point is conceptually a mapping from
-variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
-notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
-instead of the map
-$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
-Analogously to classical liveness, the liveness at program point $\pi$
-in $\pi:e$ is the liveness just before executing $e$.
-
-A complementary notion to liveness is {\em demand}.  The demand for
-expression $e$ is again an access path---that subset of of $\{\acar,\acdr\}^\ast$ which the
-context of $e$ may explore of $e$'s result.  So, for example given a
-demand $\sigma$ and the expression $\pi: (\RETURN\ x)$, the liveness at
-$\pi$ is exactly $x.\sigma$.
-The classical analogy of this is in {\em strong liveness}, where
-an assignment node $n: x:=y+z$ causes $y$ and $z$ to be live on entry
-to $n$ if (and only if) $x$ is live at exit of $n$---the liveness
-of $x$ at exit from $n$ becomes the demand on $y+z$.
-Note that, for an operation like division which
-may raise an exception, the assignment $n: x:=y/z$ makes $y$ and $z$ live
-regardless of the liveness of $x$.
-
-We use $\sigma$ to range over demands, $\alpha$ to range over access paths
-and $\Lv$ to range over liveness environments. The notation
-$\sigma_1\sigma_2$ denotes the set $\lbrace \alpha_1\alpha_2 \mid
-\alpha_1 \in \sigma_1, \alpha_2 \in \sigma_2\rbrace$. Often we shall
-abuse notation to juxtapose an edge label and a set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
-Finally, we use $\Lfonly$ to range over {\em demand transformers}.
-These transform demands on a function call
-into demands for its formal parameters:
-given a user function $f$, defined by
-$(\DEFINE\ (f\ x_1\ \ldots\ x_n)\ \ e_f)$ and called with demand $\sigma$,
-then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
-
-Note that liveness refers to variables and fields, and not to $\CONS$ cells
-(i.e.\ to edges in the memory graph, not to locations themselves).
-Hence liveness of $\{x.\epsilon,x.\acar\}$ means that future computation
-may refer to the value $\ell$ of variable $x$, and also to the $\CAR$ field
-of location $\ell$.
-In the absence of other pointers to heap location $\ell$, we are
-certain that the $\CDR$ field of $\ell$ will not be referenced
-and may hence be corrupted arbitrarily.
-Note therefore, that while  $\ell$ cannot be garbage collected, any
-location $\ell'$ stored in the $\CDR$ field of $\ell$ would be garbage
-(again provided there are no other aliases to  $\ell$).
-
-%% \comment{
-%%   \begin{enumerate}
-%% \item Set the framework for liveness. Basic notions: access
-%%   paths and expressions.
-%% \item An informal working definition of liveness. Minefield will
-%%   provide the formal definition.
-%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
-%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
-%%   rules. One of the changes from  the earlier version is the lack of a
-%%   post-liveness in  the judgements. This anticipates  the intended use
-%%   of  the analysis  for garbage  collection and has  to be  carefully
-%%   explained.
-%%   \end{enumerate}
-%% }
-
-%% We now  set up  the framework for  liveness analysis.  Given  a memory
-%% graph, a link in the graph  can be described by starting with a member
-%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
-%% \CDR\ dereferences to reach the link.  For notational convenience, the
-%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
-%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
-%% dereferences    is    represented   by    a    string   $\alpha    \in
-%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
-%% in the  memory graph that starts  at $x$ and  traverses adjacent links
-%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
-%% We call the string $x.\alpha$ an {\em access expression\/}, the string
-%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
-%% graph   by  the   access  expression   an  {\em   access   path}.   In
-%% Figure~\ref{fig:mot-example},        the       access       expression
-%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
-%% containing the value  $4$.  Most often the memory  graph is clear from
-%% the context,  and we shall use  access expressions to  refer to access
-%% paths. An  access path  used to refer  to a  link in the  memory graph
-%% refers    to     the    last     link    in    the     path.     Thus,
-%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
-%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
-%% set of access paths, then  $x.\sigma$ will represent the set of access
-%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
-%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
-%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
-%% \sigma\rbrace$.
-
-%% \subsection{Liveness Analysis}
-%% \label{sec:exp-liveness}
-%% 
-%% Consider the memory graph at a program point $\pi$ during an execution
-%% of a program. A  link in the memory graph is said  to be {\em live} at
-%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
-%% execution  starting from  the  evaluation of  the expression  labelled
-%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
-%% be live at  $\pi$, if, during {\em any} execution  of the program, the
-%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
-%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
-%% In the first, the path  itself represents the sequence of dereferences
-%% to  access  the last  edge,  as,  for  example, while  evaluating  the
-%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
-%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
-%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
-%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
-%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
-%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
-%%   implicitly  live} if it is live in the second sense.  Clearly, static
-%% determination   of  implicit  liveness   requires  an   analysis  that
-%% determines  potential sharing  during  any execution  of the  program.
-%% However, the intended application  of our liveness analysis is garbage
-%% collection  during which  we have  access  to the  entire heap.   This
-%% obviates the need for a  separate sharing analysis and we restrict our
-%% attention to explicit liveness only.
-%% 
-%% 
-%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
-%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
-%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
-%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
-%% %% \bcdr\ represent application of  the constructor \CONS.  The following
-%% %% example motivates the need for these symbols.
-%% 
-%% %% \begin{example}\label{exmp:motiv-bar}
-%% %%   Consider the expression at program point $\pi_1$ in
-%% %%   \begin{center}
-%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
-%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
-%% %% 			    $\pi_3\!:${$\cdots$})}}
-%% %%   \end{center}
-%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
-%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
-%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
-%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
-%% %%   \begin{itemize}
-%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
-%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
-%% %%     live at $\pi_1$.
-%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
-%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
-%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
-%% %%   \end{itemize}
-%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
-%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
-%% %% \end{example}
-%% 
-%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
-%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
-%% %% resulting access patterns may not  be directly useful in that they may
-%% %% not describe a  path in the memory graph.  We  define a {\em canonical
-%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
-%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
-%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
-%% %% to obtain a canonical access  pattern representing the same edge. This
-%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
-%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
-%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
-%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
-%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
-%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
-%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
-%% %% cannot  be  used for  nullification  during  garbage collection.   The
-%% %% liveness of a such a path is irrelevant, and the path is not considered
-%% %% any further.
-%% 
-%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
-%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
-%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
-%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
-%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
-%% %% element in $\sigma_2$, i.e.
-%% 
-%% %% \begin{eqnarray*}
-%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
-%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
-%% %% \end{eqnarray*}
-
-\subsection{Liveness Analysis}
-First recall the classical formulation of liveness (as sets of simple variables)
-on three-address instructions,
-$
-\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I)
-     \setminus \mathit{def\/}(I)
-     \cup   \mathit{ref\/}(I)
-$,
-and then note that {\em strong liveness} needs, when $I$ is the instruction $z:=x+y$,
-that $\mathit{ref}(I)$ be refined to
-$\{x,y\}$ if $z \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
-
-Our liveness analysis formulated in Figure~\ref{fig:live-judge}
-is analogous.
-% has three analogous aspects.
-% \begin{itemize}
-% \item
-Firstly, the function $\mathit{ref}$, when given a statement $s$,
-returns the incremental liveness {\em generated} by $s$.
-Because we generalize {\em strong} liveness, $\mathit{ref}$ needs a second parameter,
-specifying the demand $\sigma$ on the result of $s$,
-to determine which access paths of its free variables are made live.
-The cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
-% The cases for
-% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
-% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
-% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
-A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
-argument (even if its result is never used).
-This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
-(note $\sigma$ may be $\{\}$).
-In the opposite  sense, the  demand  of $\acar\sigma$  on
-(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-that \CONS\ does not, by itself, dereference its arguments.
-Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
-$\mathit{ref}$
-expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
-Constants generate no liveness.
-
-% \item
-The function $\mathcal{L}$ now gives the (total) liveness of an expression $e$.
-The cases $\RETURN$ and $\SIF$ are straightforward, but note the liveness $x.\epsilon$ generated
-by the latter.
-The case $(\LET\ z\leftarrow s\ \IN\ e')$
-resembles a three-address instruction:
-the liveness of $e$ is given by taking the liveness, $\Lv$, of $e'$,
-killing any liveness of $z$ and adding any incremental liveness from $s$.
-The main subtlety is how the liveness of $z$ in $\Lv$ is converted to a demand $\Lv(z)$ to be
-placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-% \item
-%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-Finally, the judgement form $\mathit{Prog} \len \Lfonly$ is used to determine $\Lfonly$\@.
-Analogously to classical liveness being computed as
-a solution of dataflow equations, we require, via inference rule ({\sc live-define}),
-$\Lfonly$ to satisfy the fixed-point property that:
- when we assume $\Lfonly$ to be the family of demand transformers for the program
- then the calculated liveness of each function body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees
- with the assumed  $\Lfonly$.
-As usual, there are often multiple solutions to $\Lfonly$; all are safe
-(see Section~\ref{sec:correctness})
-but we prefer the least one as giving the least liveness subject to safety---and
-hence greatest amount of garbage collected.
-% \end{itemize}
-
-We make three observations: firstly the rule ({\sc live-define}) has a least solution
-as $\mathcal{L}(\cdot)$ is monotonic in $\sigma$; secondly that
-({\sc live-define})
-resembles the rule for type inference of mutually recursive function definitions,
-and thirdly the asymmetry of demand and liveness (compared to post- and pre-liveness
-classically) is due to the functional formulation here.
-
-Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
-(representing a fully context-sensitive analysis)
-can be safely approximated, for each function, by a {\em procedure summary}
-(unifying the contexts in the style of 0-CFA)\@.
-The summary consists of a pair of a single demand and, for this demand,
-the corresponding tuple of demands the function makes on its arguments.
-
-%%% save two lines to bring up section 3.2.
-
-% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
-% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
-% enables us to do a context-dependent analysis of a call without having
-% to enter the body of the called function.
-
-%% We present  our liveness  analysis using judgement forms $\len$
-%% for both expressions and programs.
-%% These are given formally in Figure~\ref{fig:live-judge}, but
-%% first we an intuitive understanding of the issues involved.
-%% Consider the expression:
-%% 
-%% \begin{center}
-%% \begin{uprogram}
-%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
-%%                         \px) \IN\  
-%%            % \UNL{0}
-%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
-%%                         \pw ) \IN \ldots) 
-%% \end{uprogram}
-%% \end{center}
-%% 
-%% 
-%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
-%% \pw.\epsilon\}$.
-%% Informally, this is so since \pw\ and \pz\ are used in
-%% (\PRIM~\pz~\pw ).
-%% 
-%% \begin{enumerate}
-%% \item The liveness  at $\pi$ is obtained by  removing the access paths
-%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
-%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
-%%     similar  to liveness  analysis of  imperative programs,  where the
-%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
-%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
-%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
-%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
-%%     liveness generated and  killed in $s$}. 
-%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
-%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
-%%   as a {\em  demand}.  The demand on an expression consists of  access
-%%   patterns  representing all  possible future  uses of  its  value.  A
-%%   demand  of $\epsilon$  means that  only the  reference resulting
-%% %  from the evaluation  of
-%%   from  $(\CAR\,\,\px)$ will be used.
-%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
-%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
-%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
-%% \end{enumerate}
-%% 
-%% In general,  an access path is  live before an  expression because the
-%% link represented by it is used (i) while evaluating the expression, or
-%% (ii) while using the result of the expression, or (iii) simply because
-%% the link  is live at  a later program  point.  When the  expression in
-%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
-%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
-%% the access  path \pw.   Note that \px.\acar\  is live in  this example
-%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
-%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
-%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
-%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
-%%
-
-% \paragraph{Judgements formally.}
-% Let  $\Lv$  denote a  liveness  environment (a function from
-% variables to access paths, abused using grammar notation) and
-% $\sigma$ be a demand.
-% The judgement form $e, \sigma \len \Lv$,
-% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
-% environment  before $e$  is $\Lv$ and
-% is thus analogous to the classical notion of {\em transfer function}.
-
-%    It is  interesting to  note that,
-% unlike traditional liveness analysis,  our judgements do not require a
-% liveness after the expression. This  is because of the coming together
-% of  two reasons.   First, it  will be  clear while  discussing garbage
-% collection,  that it is  enough for  our purposes  to assume  that the
-% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
-% restriction of our  programs to ANF ensures that  the post-liveness of
-% each  expression  $e$  inside   the  function  body  continues  to  be
-% $\emptyset$. Thus the post-liveness can be dropped altogether.
-
-%==============================================================
-\begin{figure}[t]  
-% \footnotesize
-\begin{eqnarray*}
-\mathit{ref\/}(\kappa,\sigma,\Lfonly)
-          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
-          &=& \{x.\sigma\} \\
-\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
-          &=& \{x.\epsilon,y.\epsilon\} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \\
-\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
-        \mathcal{L}(e_1,\sigma,\Lfonly) \cup
-        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
-        \{x.\epsilon\} \\
-\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
-        \Lv
-           \setminus \{x.*\}
-           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
-\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
-\end{eqnarray*}
-
-\begin{minipage}{0.85\textwidth}
-\infrule[live-define]
-        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
-           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \len \Lfonly
-\\ \makebox[0mm]{where
-     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
-\end{minipage}
-
-\normalsize
-  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%==============================================================
-
-% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
-% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
-% liveness generated  before the right-hand-side of  a \LET\ definition.
-% Constants    do   not   generate    any   liveness.     The   operator
-% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
-% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
-% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
-% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
-% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-% that \CONS\ does not, by itself, dereference its arguments.
-% 
-% 
-% Unlike primitives, the mapping from the demand on a user-function call
-% to the  demands on its arguments  has to be derived  from the function
-% body itself---as a fixed point for a recursive function.
-% The function  \Lfonly\ captures this mapping for
-% all functions  defined in the  program.  For a  function $\mathit{f}$,
-% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
-% a    call     to    the     $i$th    argument.     The     rule
-% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
-% confirmed by the body of all the functions, then the guess is correct.
-% {\color{blue}Note the similarity of this rule to type-inference
-% for recursive program schemes.}
-
-% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
-% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
-% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
-% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
-% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
-% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
-% liveness generated before $s$.
-
-\subsection{Minefield semantics and correctness}
-\label{sec:minefield}
-\label{sec:correctness}
-
-This section gives a modified semantics
-%(a reference semantics)
-which checks liveness annotations
-at run time, and `explodes' when these are found to be inconsistent with execution behavior,
-but otherwise behaves as the standard semantics.
-We show that such explosions never occur and hence run-time checks
-can be elided.
-%% We first assume some demand transformers $\Lfonly$;
-%% our proof will rely on this being a solution (i.e.\ satisfies $\mathit{Prog} \len \Lfonly$)
-%% of liveness analysis, but we do rely in this yet.
-We first assume an arbitrary  demand transformers $\Lfonly$ which satisfies
-$\mathit{Prog} \len \Lfonly$.
-We then enrich the abstract machine state $\rho, \stk, \heap, e$ to
-$\rho, \stk, \heap, e, \sigma, \Sigma$.
-Here $\sigma$ is the demand to the currently active function,
-thus the liveness $\Lv$ at $e$ is $\mathcal{L}(e,\sigma,\Lfonly)$,\footnote{
-This a simple liveness propagation using $\mathcal{L}(\cdots)$
-and $\mathit{ref}(\cdots)$ as $\Lfonly$ is assumed given.}
-and $\Sigma$ is a stack of demands---one for each function frame
-pushed in $\stk$.
-
-Second, we augment $\mathit{Val}$ with a value $\bot$.
-To model strong liveness $\bot$ may be copied freely,
-including into a $\CONS$ cell,
-% but reduces the state to $\bang$
-but explodes
-when used computationally (in a primitive operation other than a copy).
-Additionally we define
-$\mathit{GC}(\Lv,\Sigma): (\rho, \heap, \stk) \mapsto (\rho', \heap', \stk')$
-which determines {\em live-reachability}\footnote{Reachability curtailed by liveness information.}
-%%%%% \marginpar{Say more about live-reachability?}
-using $\rho$ and the $\rho$'s in $\stk$ as the root set
-and following links in $\heap$ {\em only as far as allowed by $\Lv$ and $\Sigma$}.
-% The effect of $\mathit{GC}(\Lv,\Sigma)$ is to replace
-Hence $\mathit{GC}(\Lv,\Sigma)$ replaces
-live-unreachable values---in $\rho$, in $\heap$ and the $\rho$'s in \stk---with $\bot$.
-For example, if $x.\epsilon \not\in \Lv$ then $\rho'(x) = \bot$.
-%, regardless of $\rho(x)$.
-Only $\mathit{GC}(\cdots)$ introduces $\bot$.
-
-Third, we update the semantics in four ways: ({\em i\/}) we arrange that all
-($\rightarrow$) transitions on expressions $e$ first use $\mathit{GC}(\cdots)$
-to update the state and then continue as before;
-and ({\em ii\/}) whenever the value $\bot$ is used computationally in a reduction
-($\rightsquigarrow$), we enter a distinguished stuck state $\bang$.
-For example, supposing $\rho(x)=\bot$
-then
-$\rho, \heap, s \rightsquigarrow \bang$ if $s$ is
-$(\CAR\,\, x)$, $(\CDR\,\, x)$ or $(\PRIM\,\, x\ y)$,
-but not if $s$ is $(\ID~x)$, $(\CONS~x~y)$ or $(f~x~y)$.
-Finally ({\em iii\/}) we make a similar change to the ($\rightarrow$) reduction for ({\sc ord-if})
-and ({\em iv\/}) augment the ({\sc ord-let}) rule for primitives to propagate
-$\bang$ from ($\rightsquigarrow$) to ($\rightarrow$).
-%%%%%% {\color{blue}The full minefield semantics is given as an appendix?}
-
-% Suppose liveness for the current $e$ under reduction is $\Lv$; then:
-% For every variable $x$ for which $x.\epsilon \not\in \Lv$ then replace
-% $\rho(x)$ with $\bot$.
-% Also, for every $\ell \in \mathit{Loc}$, replace the $\CAR$ field
-% of $\heap(\ell)$ with $\bot$ when
-% which there is no path $x.\alpha\acar$ in $\Lv$ such that following
-% $\alpha$-links in $\heap$ from $\rho(x)$ reaches $\ell$.
-% Correspondingly, replace the $\CDR$ field of $\ell$ if there
-% is no path $x.\alpha\acdr$ in $\Lv$.
-% \item
-% 
-% {
-% \footnotesize
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-car]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CAR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-cdr]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CDR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}\\[1ex]
-% 
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-prim]
-%       {\rho(x) = \bot \vee \rho(y) = \bot}
-%       {\rho, \heap, {(\PRIM\,\, x\ y)}
-%         \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-null]
-%         {\rho(x)= \bot
-%         }{\rho, \heap, {(\NULLQ\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% }
-%
-
-The resulting minefield semantics behaves identically
-(identical heap, identical steps, including possible non-termination)
-to the standard semantics, except for one possibility---the
-minefield semantics going $\bang$ while the standard semantics
-continues (either to a halt state, to a stuck state, or reduces forever).
-
- We now prove a result that relates
-liveness analysis to the semantics.
-\begin{proposition}
-Given program $P$ with $P \len \Lfonly$, then in the minefield
- semantics $P \rightarrow \bang$ can never occur
- (cf.\ `well-typed programs do not go wrong').
-\end{proposition}
-\textit{Proof outline:}
-Space does not permit a full proof, but we give the two main steps.
-We proceed by contradiction and assume there is a program $P$ for which
-$P \len \Lfonly$ can enter state $\bang$.
-The first step is to construct a program $P'$ with identical behavior, but with
-no user-function definitions, using inlining.
-This is possible because a program which goes $\bang$ does so after a finite number
-of reductions, and hence even recursive functions have only had a finite number of
-invocations.
-We hence repeatedly inline-expand user-function calls in $P$
-until we obtain a program $P'$ which behaves identically\footnote{
-Modulo replacement of
-({\sc ord-call}) and
-({\sc ord-return}) steps with
-({\sc ord-id}) steps.
-}
-to $P$ in the standard semantics, but makes no user-function calls.
-Any remaining, non-executed, calls can 
-%be left or
-replaced with a new primitive with the same
-demand-to-liveness transfer function---thus making
-$P'$ a simple expression $e'$.
-%%%%%
-Not only do the program points in reducing $e'$ correspond
-one-one to states during evaluation of $P$,
-but also the liveness associated with a point
-in $e'$ is identical to the liveness at the corresponding state
-$(\rho, \stk, \heap, e, \sigma, \Sigma)$ of $P$ (concatenating the liveness
-$\mathcal{L}(e,\sigma,\Lfonly)$ at $e$
-with the liveness, obtained from $\Sigma$, of the call sites in $\stk$ and
-after renaming the variables correspondingly to the inlining which produced $e'$).
-This assertion relies on the analysis being fully context-sensitive, and noting that
-while the change of scope caused by inlining changes variable visibility
-between $P$ and $e'$ it does not change the liveness---as local-to-a-function
-$\LET$-variables whose scope has been prolonged due to inlining
-are dead in the prolonged scope.
-
-The second step of the proof is to show that $e'$ cannot go $\bang$.
-We proceed by induction.
-Correctness of the $\SIF$ and final $\RETURN$ forms are immediate;
-the $(\LET~z\leftarrow s~\IN~e)$ form requires showing
-the inference rules ensure that any value referenced via $z$ in $e$
-was already live-reachable (via another variable and path) in any
-enclosing expression (so that $\mathit{GC}(\cdots)$ could not re-write it to $\bot$).
-
-% \newpage
-
-\section{Computing liveness and its encoding as a table}\label{sec:computing}
-
-Section~\ref{sec:liveness} gave a context-sensitive liveness analysis
-and proved it correct with reference to a {\em minefield} semantics.
-% Practical computation was not considered.
-For practical use we need to solve the liveness equations {\em finitely and
-symbolically}.  As expressed mathematically, and given a fixed program,
-three things are potentially unbounded:
-({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
-({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
-({\em iii\/}) the number of such access paths.
-
-We commonly first solve ({\em i\/}) by reducing the number of distinct
-calling contexts
-(e.g.\ to a single unified context in 0-CFA style).
-However, it turns out that we can solve the equations for $\Lfonly$ symbolically
-without this reduction, so here we defer this to Section~\ref{sec:bodylivenessbodies}.
-(It also allows easier extension to  dynamically determined liveness---future work.)
-%
-We address ({\em ii\/}) and ({\em iii\/}) by 
-% attempting to re-interpret
-re-interpreting
-the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
-rather than a mutually recursive set of equations on sets of access paths.
-
-This requires two ideas.  Firstly we have to control the use of functions---they
-tend to be infinitary and do not occur naturally in CFGs; in particular our
-$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
-to a tuple of access paths.  The former is achieved by using a separate
-meta-variable (later non-terminal)
-$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
-($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
-Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
-expressible finitely (it is a linear form).
-
-Secondly, there are also two technical issues in re-interpreting
-Figure~\ref{fig:live-judge} as a grammar.
-One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
-disappears after variables are separated ($\Lanv{i}{x}$ above). The other
-is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
-not a CFG\@.
-Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.
-
-Finally,  Section~\ref{sec:NFA-approx}
-uses a construction due to
-Mohri and Nederhof~\cite{mohri00regular} to over-approximate
-context-free grammars with regular grammars;
-these are more appropriate for run-time use.
-%
-Hence the overall `big picture' view is that each GC point
-is annotated with a table encoding the DFA for that program point.
-Such program points forming the call string causing the garbage collection
-(and ending in $\CONS$)
-are used by the garbage collector to curtail its reachability-marking phase
-to this over-approximated liveness.
-
-\subsubsection{Modifying the \CONS\ rule.}
-
-The    $\mathit{ref}$    rule for \CONS,    shown    in
-Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
-\acar\  and  \acdr\  from  access paths in  $\sigma$.
-  Mathematically  this  is fine  but causes problems  when
-  solving  the liveness equations  {\em symbolically}  since such
-  decomposition cannot be expressed as a context-free grammar.  
-To  handle this,  we introduce  two  new symbols  \bcar\  and
-\bcdr\ with the properties:
-\[
- \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
-\hspace{3em} \mbox{and} \hspace{3em}
- \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
-\]
-We can now rewrite the \CONS\ rule as:
-\begin{eqnarray*}
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
-\end{eqnarray*}
-
-We  call the liveness equations with  this modification
-  \Lfunonly$'$.
-The  defintions of \bcar\  and  \bcdr\  induce   the  following
-relation $\hookrightarrow$ over sets of access paths:
-\begin{eqnarray*}
-  \sigma_1\bcar\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
-& \qquad\mbox{and}\qquad &
-\sigma_1\bcdr\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
-\end{eqnarray*}
-%% \noindent \begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcar\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acar\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acdr\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% %\label{eqn:eup-red1}
-%% \end{eqnarray*}
-%% \end{minipage}
-%% \hspace*{.65cm}\begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcdr\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acdr\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acar\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% \end{eqnarray*}
-%% \end{minipage}\\
-
-The reflexive transitive closure of $\hookrightarrow$ will be denoted
-as $\stackrel{*}{\hookrightarrow}$. The following proposition relates \Lfunonly\   and \Lfunonly$'$:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]  
-\begin{boxedminipage}{\textwidth}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\append~\lista~\listb)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \xtest\ $\leftarrow $\ (\NULLQ~\lista) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\SIF\ \xtest~$\pi_3\!\!:$(\RETURN\ \listb)  
-          \UNL{4} \hspace*{.05cm}     $\pi_4\!\!:\, $(\LET\ \xtl\  $\leftarrow$\   (\CDR\ \lista) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_5\!\!:\, $(\LET\ \xrec\  $\leftarrow$  (\append\ \ \xtl\ \  \listb) \IN
-	  \UNL{6} \hspace*{.05cm}    $\pi_6\!\!:\, $(\LET\ \xhd\  $\leftarrow$  (\CAR\ \lista)  \IN
-	  \UNL{7}  \hspace*{.05cm}   $\pi_7\!\!:\,
-          $(\LET\ \xans\  $\leftarrow$ (\CONS\ \ \xhd\ \ \xrec)  \IN\
-          $\pi_8\!\!:\, $(\RETURN\ \xans))))))))
-	\end{uprogram}}
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pz\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}              (\LET\ \py\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_9\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\append\ \py\ \pz) \IN
-          \UNL{4}   \hspace*{.78cm}  $\pi_{10}\!\!:\, $  (\LET\ \pa\  $\leftarrow$  (\CDR\ \pw) \IN
-	  \UNL{5}   \hspace*{.78cm}  $\pi_{11}\!\!:\, $
-          (\LET\ \pb\  $\leftarrow$ (\CAR\  \pa) \IN~$\pi_{12}\!\!:\,$ (\RETURN\ \pb)))))))
-\end{uprogram}}
-\end{boxedminipage}
-  \caption{An example program.}
-\label{fig:examp-pgm}
-\figrule
-\normalsize
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{proposition}
-Assume that  a liveness computation  based on  \Lfunonly\   gives  the liveness of
-the variable $x$ at a program point $\pi_i$ as $\sigma$ (symbolically, $\Lanv{i}{x}=
-\sigma$).  Further,  let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is
-used instead of \Lfunonly.   Then  $\sigma' \stackrel{*}{\hookrightarrow}
-\sigma$.
-\end{proposition}
-
-To see  why the proposition is  true, consider an analysis  based on
-\Lfunonly$'$ in which   $\sigma$ appears in
-the  context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.  Let $\alpha  \in \sigma$.  The
-symbol  \bcar(\bcdr)  merely  marks  a  place in  $\alpha$  where  the
-original  \CONS\  rule  would  have erased  an  immediately  following
-\acar(\acdr),  or, in  absence of  such a  symbol, would  have dropped
-$\alpha$ itself.  Since the application  of any rule in \Lfunonly$'$ merely
-adds  symbols at  the beginning  of  $\alpha$, the  markers and  other
-symbols in $\alpha$ are propagated to other dependent parts of program
-in their same relative positions.  Consequently, the erasure carried out at the end of the analysis with
-$\stackrel{*}{\hookrightarrow}$ gives the same result as obtained
-through \Lfunonly.
-
-\subsection{Generating equations for the demand transformer \Lfonly}
-\label{sec:Lfonlysymbolically}
-
-%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
-We  shall  consider the  program  in  Figure~\ref{fig:examp-pgm} as  a
-running example;
-this is the  ANF version of the program in Figure~\ref{fig:mot-example}.
-
-%%AM no longer used: In what follows,  we shall overload the semantics of
-%% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-%mean the generation
-%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
-%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.
-
-To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
-$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
-{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
-$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
-formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
-of $x_i$.  For our example program which has  a
-single function \append, this generates the following equations:
-% $\Lfonly_{\append}$  is defined by equations:
-  \begin{eqnarray*}
-    \Lf{\append}{1}{\sigma}
-    &=& \epsilonset \cup \acar\bcar\sigma
-    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
-    \Lf{\append}{2}{\sigma}
-    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
-  \end{eqnarray*}
-  In  general, the  equations  for \Lfonly\  are recursive  since
-  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
-  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
-    is  similar to  solving the  differential equation  
-    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
-    that    the    solution     has    the    form    $\mathit{y}
-    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
-    equation  yields  a  quadratic  equation  in  $r$,  and  each
-    solution of $r$ gives rise  to a solution of the differential
-    equation  (in our  setup we  can effectively  pick  the least
-    solution rather than needing linear combinations).  }
-\begin{eqnarray}
-\label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
-\end{eqnarray}
-where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
-alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
-why \Lfonly\ has this form  is as follows: Recall that
-\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
-the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
-situation in which  the demand $\sigma$ does not  reach the $i$th
-argument.   This happens,  for  instance, when  argument is  used
-only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
-transformation of the demand $\sigma$ before it reaches the $i$th
-argument. This  transformation prefixes strings in 
-$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.
-
-Substituting  the  guessed   form  in  the  equations  describing
-\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
-containing $\sigma$ and those  without $\sigma$ on both sides of the
-resulting equation, we
-get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
-% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
-$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
-solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
-  \append}$, we get:
-  \begin{align*}
-    \Uf{\append}{1} &= \epsilonset
-    \cup \acdr.\Uf{\append}{1}  
-    &\quad 
-    \Uf{\append}{2} &= \Uf{\append}{2} \\
-    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
-    \cup \acdr\Df{\append}{1}\bcdr &\quad 
-    \Df{\append}{2} &= \epsilonset \cup
-    \Df{\append}{2}\bcdr 
-  \end{align*}
-Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.
-
-\subsection{Generating liveness equations \Lv\  for function bodies}
-\label{sec:bodylivenessbodies}
-
-\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
-
-We now calculate a 0-CFA-style summary liveness for each GC point of a program.
-There are two parts to this.
-First, for each function $f$,  we determine a {\em summary demand} $\sigma_f$
-over-approximating any demand $\sigma$ passed to $f$.
-%$\Lf{f}{i}{\sigma}$.
-So, consider a function $g$ containing a call site $\pi$ to $f$, say
-$\pi\!\!:\!(\LET~x \leftarrow (f\,y_1\,\ldots\,y_n)~\IN~e)$.
-Under the assumption that the demand on $g$ is $\sigma_g$, the liveness at $e$
-is $\Lv = \mathcal{L}(e,\sigma_g,\Lfonly)$, and the $\LET$ case of
-Figure~\ref{fig:live-judge} tells us this call site contributes $\Lv(x)$ to the
-demand $\sigma_f$ placed on  $f$.
-Write $\deltacall{f}{\pi}{g}$ for
-this demand (contributed by call site $\pi$ in $g$).
-
-Thus, supposing the $k$ call sites to function $f$ are
-$\pi^1$ (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
-%%% \marginpar{and writing $\sigma_f^i$ for the above $\deltacall{f}{\pi^i}{g^i}$?}
-then the over-approximation requirement on $\sigma_f$
-is achieved by taking $\sigma_f = \deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
-% is achieved by taking $\sigma_f = \sigma_f^1 \cup \cdots \cup \sigma_f^k$.
-
-The expression $e_{\mainpgm}$ is a special case; we assume it may be called externally
-with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$ (denoted
-$\sigma_{\!all}$).
-This is because any  part  of its  value  may be  used  by  the environment---for
-printing  the  result,  for  instance.
-
-% In what follows,  we shall overload the semantics of
-% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-% 
-% 
-% We assume an overall demand of $\{\acar,\acdr\}^*$ (denoted
-% $\sigma_{\!all}$) on the main expression $e_\mainpgm$.  This is because 
-% any  part  of its  value  may be  used  by  the environment---for
-% printing  the  result,  for  instance.
-
-%% To  generate  equations for  the  body $e_{\mathit{f}}$ of  a
-%% function,  we evaluate \Lfun{e_{\mathit{f}}}{\sigma_{\mathit{f}}}{\Lfonly}.     For ${\sigma_{\mathit{f}}}$,  we have
-%% to consider all calling  contexts during execution.  We therefore
-%% do a 0-CFA approximation by  defining
-%% % $\sigma_\mathit{f}$        as $\bigcup_\calls{f}{k}\!\!\sigma_\calls{f}{k}$,
-%% $\sigma_\mathit{f}$        as $\sigma_\calls{f}{1} \cup \cdots \cup \sigma_\calls{f}{n}$,
-%% where $\calls{f}{k}$,
-%% $\mathit{k} = 1 \ldots n$ represent all the calls to $\mathit{f}$
-%% in  the  program,   and  $\sigma_\calls{f}{k}$  are  the  demands
-%% generated at the calls  by the \rn{let-live} rule while analyzing
-%% the bodies of the caller.
-
-For the running example, $\append$ has calls from $\mainpgm$ at $\pi_9$
-and a recursive call at $\pi_5$.
-So $\sigma_{\append} =
-     \deltacall{\append}{\pi_9}{\mainpgm}  \cup \deltacall{\append}{\pi_5}{\append}$.
-Filling in  the values gives:
-%  where the former is the demand generated at the
-%  call to \append\  in $e_\mainpgm$ the latter
-%  is  the demand  at the  recursive call. 
-%% Propagation of
-%%  $\sigma_{\!all}$ in $e_\mainpgm$ gives    $\sigma_\calls{\append}{1}$   as   $\{\epsilon,   \acdr\}   \cup
-%%  \acdr\acar\sigma_{\!all}$  and propagation of
-%%  $\sigma_{\mathit{\append}}$ in $e_\append$ gives   $\sigma_\calls{\append}{2}$   as
-%%  $\bcdr\sigma_{\mathit{\append}}$.  
-%% Filling in  the values of $\sigma_\calls{\append}{1}$ and
-%% $\sigma_\calls{\append}{2}$, we get:
-\begin{align*}
-\sigma_{\append}    &=
-\left( \{\epsilon,  \acdr\} \cup  \acdr\acar\sigma_{\!all}  \right) ~\cup~
-\bcdr\sigma_{\mathit{\append}}
-\end{align*}
-
-Second, for each function $f$ (possibly $\mainpgm$) we need the liveness at
-each contained GC point
-$\pi\!\!:\!(\LET~x \leftarrow s~\IN~e)$,
-where $s$ is a call either to $\CONS$ or to a user function $g$.
-Given $\sigma_f$ calculated above, this
-is simply $\mathcal{L}(\pi,\sigma_f,\Lfonly)$.
-%%% \marginpar{AM: we've always said this (safe), but actual parameters
-%%% live before the call and dead after needn't be included!}
-
-For the running example, containing GC points $\pi_5, \pi_7$ in $\append$ and
-$\pi_9$ in $e_\mainpgm$, this gives
-(recall Equation~(\ref{eq:LF:DI}) above states $\Lf{f}{i}{\sigma} = \Uf{f}{i} \cup \Df{f}{i}.\sigma$):
-\begin{align*}
-\Lanv{5}{\xtl} &= \Lf{\append}{1}{\bcdr\sigma_{\append}}  \quad &
-                                \Lanv{7}{\xhd}\,\,\, &=  \bcar\sigma_{\append}     \\ 
-\Lanv{5}{\listb} &= \Lf{\append}{2}{\bcdr\sigma_{\append}} &
-                                \Lanv{7}{\xrec} &=  \bcdr\sigma_{\append} \\
-\Lanv{5}{\lista} &= \epsilonset \cup \acar\bcar\sigma_{\append} 
-\\[1ex]
-  \Lanv{9}{\py}  &= \Lf{\append}{1}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}} \\
-  \Lanv{9}{\pz}  &= \Lf{\append}{2}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}}
-\end{align*}
-
-\noindent 
-In summary, the equations generated during liveness analysis are:
-\begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
-\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
-% {\color{red}
-%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
-%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
-% }
-\end{enumerate}
-
-\subsection{Solving liveness equations---the grammar interpretation}
-The  equations above can now be re-interpreted as a
-context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
-a variable  $X$ occurring on  the LHS of the  equations generated
-from the analysis.  We can  think of the resulting productions as
-% being a grammar with multiple start symbols---one for
-% for each nonterminal \var{\Lanv{i}{x}}.
-being associated with several  grammars, one for each non-terminal
-\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
-productions:
- \begin{eqnarray*}
-\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
-  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
- \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
-  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
-  \acdr\var{\Uf{\append}{1}}\\         
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar               \mid
-  \acdr\var{\Df{\append}{1}}\bcdr\\
-\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
-\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
-\var{$\sigma_{\mathit{\!all}}$}
-\end{eqnarray*}
-Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
-denoted $\mathscr{L}(\var{\Lanv{i}{x}})
-$,  is the desired  solution of
-\Lanv{i}{x}.   
-However, recall from  our earlier  discussion that
-the  decision problem that  we are  interested in  during garbage
-collection is:
-\begin{quote}
-Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
-\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
-\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
-of forward paths only. Then does $\alpha \in \sigma$?
-\end{quote}
-We could convert  the  rules  defining  $\hookrightarrow$  into
-productions and  add them to the grammar. However, this results
-in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is  undecidable.   We
-circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
-  strongly  regular}  CFGs which  have   easy
-translations to non-deterministic finite state automata (NFA). The
-NFAs are then simplified on the
-basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
-membership of forward access paths. The resulting NFAs are   finally
-converted to DFAs for use during garbage collection.
-%------------------------------------------------------------%
-\subsubsection{Approximating CFGs using NFAs.}
-\label{sec:NFA-approx}
-
-We use the   
-algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
-approximate a CFG  to a   {\em strongly regular\/} grammar.  The
-transformation has  the  property that  if $\Lv$  is a
-non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
-Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
-  L}_{G'}(\Lv)$.    This  is   required  for the
-approximation to be safe with respect to liveness.
-
-We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
-transformation. We  pick the only production that  is affected by
-the transformation---the production for \Df{\append}{1}.  The
-production for \Uf{\append}{1}, while  recursive, is already in
-strongly  regular  form  and   is  therefore  unaffected  by  the
-transformation.
- \begin{eqnarray*}
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
-  \acdr\var{\Df{\append}{1}} \\ 
-\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
-\mid \epsilon 
-\end{eqnarray*}
-The  languages
-generated for  \var{\Df{\append}{1}} in  the original grammar  and the
-new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
-\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
-
-%------------------------------------------------------------%
-\subsubsection{Transforming NFAs to Accept Forward Paths:}
-%\label{sec:NFA-elim}
-The strongly regular CFGs obtained after the Mohri-Nederhof
-transformation are
-  first converted  into  NFAs.  The        algorithm         described        in
-Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
-\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
-Thus  \nt\  can  be  used  to  check  
-membership of forward paths.
-
-\begin{figure}[t]
-\begin{boxedminipage}{\textwidth}
-\begin{center}
-  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
-  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
-%
-  \raggedright{\bf Output:} NFA  \nt\ with  underlying
-  alphabet $\lbrace \acar, \acdr \rbrace$ such that
-  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
-  \\
-%
-
-  \raggedright{\bf Steps:}
- \begin{algorithmic}
-   \STATE $i \leftarrow 0$ 
-   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-   without $\epsilon$-moves \cite{hopcraft90toc}
-
-   \REPEAT
-   \STATE $\nt'_{i+1} \leftarrow \nt_i$
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcar$ and
-     outgoing edge to $q''$ with label $\acar$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-   label $\epsilon$.
-   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-   \ENDFOR
-
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcdr$ and outgoing edge to
-     $q''$ with label $\acdr$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-     label $\epsilon$.
-   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-   \ENDFOR
-
-   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-   $\nt'_{i+1}$ without $\epsilon$-moves
-
-   \STATE $i \leftarrow i+1$
-   \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-   \STATE $\nt \leftarrow \nt_i$
-%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
- \end{algorithmic}
-\end{center}
-\end{boxedminipage}
- \caption{Algorithm for transforming an NFA to  accept forward
-   paths only.}\label{algo:simplify-NFA} \figrule
-\end{figure}
-%------------------------------------------------------------%
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
-pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
-process is  continued till  a fixed point  is reached.   When the
-fixed point  is reached, the resulting NFA  contains all possible
-reductions corresponding  to all the  paths in the  original NFA.
-The proofs  of the termination  and correctness of  the algorithm
-are given in our earlier paper~\cite{karkare07liveness}.
-
-\input{liveness-dfa.tex}
-
-We illustrate the algorithm in Figure~\ref{algo:simplify-NFA} by
-constructing the automaton for \var{\Lanv{9}{y}}.
-Figure~\ref{fig:automata-construction} (c) shows the automaton for
-\var{\Lanv{9}{y}} constructed by composing the automata for
-\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
-\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
-get (d). We add an $\epsilon$ edge from $q_2$ to $q_3$ bypassing the
-\bcdr\acdr\ pair from $q_2$ to $q_2$ and then to $q_3$. This is shown
-in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
-is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
-$q_2$.  Removing this $\epsilon$ edge gives the automaton shown in
-(g).  Restricting this automaton to forward edges only, we get the
-final automaton shown in (h). This automaton recognizes $\acdr^{*}\!\mid\acdr^{*}\acar \sigma_{\mathit \!all}$, showing that the entire
-list \py, including its elements, is live at $\pi_9$. Also note that
-the language accepted by the final automaton satisfies the
-prefix-closed property.
-
-\section{Prototype and evaluation}\label{sec:experiments}
-To demonstrate the effectiveness of liveness-based garbage collection,
-we  have  built a  prototype  consisting  of  an interpreter  for  our
-language,  a  liveness  analyzer  and  a copying  collector  that  can
-optionally use the results of liveness analysis for marking instead of
-reachability. When the collector uses liveness for marking, we call it
-a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-  reachability-based  collector}  (RGC).   The  copying  collector  is
-neither incremental nor generational.  As a consequence, any cell that
-becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.
-
-While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
-is invoked  at the program point   $\pi$,  it would  use  the
-liveness  DFAs at  $\pi$ itself for exploring the root set of the
-current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.
-
-Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-$\dfa{\pi}{x}$.    
-Considering   a   DFA   as   a   table,
-$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-\acar\      or     \acdr.       We      shall     also      write
-$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-indicating  whether there  is a  transition from  $q$ on
-$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-pointer to $x$.
-The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-$x$  in the  other semi-space  and returns  the new  address.  It
-hides  details such as  returning the  forwarding pointer  if the
-value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.
-
-The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-in the heap over time for RGC and LGC---here time is measured in terms
-of the number of cons cells allocated. In addition, they also show the
-number of  reachable cells and the  number of cells  that are actually
-live\footnote{This   is  statically   approximated  by   our  liveness
-  analysis}.  Since  the programs have  different memory requirements,
-we have tuned the size of heap for each program to ensure a reasonable
-number of  collections. An invocation  of RGC decreases the  number of
-cells  in heap  till  it touches  the  curve of  reachable cells.   An
-invocation of LGC decreases the number  of heap cells to no lower than
-the curve of live cells.
-
-To construct the  reachable and live curves, we  record for every cell
-its creation time (\CreateTime), its last use time (\UseTime), and the
-earliest time  when the  cell becomes unreachable  and can  be garbage
-collected (\CollTime).  For accurate recording of  \CollTime, we force
-frequent invocations  of a reachability based collector  in a separate
-run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
-  fibheap} increases  initially, comes  down before the  heap becomes gets
-full and then  continues to remains low.  This results in  a single garbage
-collection in both RGC and LGC.  The graphs also show the precision of
-our liveness analysis.  For  all programs  except  {\tt nperms} and
-{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.
-
-\subsection{Results}
-\begin{figure}[t] 
-\centering
-\begin{boxedminipage}{.75\textwidth}
-{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
-\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-  then} {\bf skip}\\ 
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
-\noindent \hspace*{.35cm}{\bf return} $y$
-\end{boxedminipage}
-\caption{Function       for       copying       a       root       set
-  variable.}\label{algo:LGC-main-loop} \figrule
-
-\end{figure}
-
-\begin{figure}[p]
-\begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
- &
-{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-\end{tabular}
- \caption{Memory  usage of  programs. The blue and the red curves
-  indicate the number of cons cells in the active semi-space for RGC and
-  LGC respectively. The black curve represents the number of reachable
-  cells and the grey curve represents the number of cells that are
-  actually live (of which liveness analysis does  a static
-  approximation). x-axis is the time measured in number of cons-cells
-  allocated. y-axis is the number of cons-cells.}
-\label{fig:memory-usage} \figrule
-\end{figure}
-The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-provides  statistics  regarding the  analysis  itself.  The number  of
-states and the analysis times  are within tolerable limits. Precision
-of analysis refers to the percentage of dead cells that is collected by
-LGC, averaged over all invocations.
-The second table shows garbage  collection statistics for RGC and LGC.
-LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to run
-in comparison with  RGC.
-
-There are a couple of  issues of concern.  The garbage collection time
-is larger in the case of LGC for some programs. The reason is that the
-cost of consulting the liveness DFA may outweigh the combined benefits
-of  fewer   garbage  collections   and  fewer  markings   per  garbage
-collection.   The  other issue  is  illustrated  by  the program  {\tt
-  lambda}    (memory   usage  shown   separately    in
-Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-in Figure~\ref{fig:experimental-results}, the  number of
-touched cells\footnote{These  are the  cells which are  visited during
-  marking phase, often  more than once due to  sharing.} in this
-example is much higher for
-LGC.  This  increase is due to  excessive sharing among  heap nodes in
-this program.  Note  that a node re-visited because  of sharing is not
-explored  any   further  during  a  RGC   collection.   However,  this
-curtailment cannot happen in  LGC because of the possibility that the
-node,  re-visited in a  different liveness  state, may  mark a  set of
-cells different from the earlier visit.
-\begin{figure}[t]
-\footnotesize
-
-\centering
-
-\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline
-{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-  nperm} &  {\sf fibheap} & {\sf knightstour} &
-{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-\hline
-\hline
-{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-\hline
-\end{tabular}}
-
-\centerline{(a)}
-
-\medskip
-
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
-  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-
-                            &   \multicolumn{2}{c|}{cells per GC}  
-                            &   \multicolumn{2}{c|}{cells per GC} 
-                            &   \multicolumn{2}{c|}{\#GCs} 
-                            &   \multicolumn{2}{c|}{MinHeap} 
-                            &   \multicolumn{2}{c|}{Avg. Drag} 
-                            &   \multicolumn{2}{c|}{(sec)} \\
-\cline{2-13}
-{Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
-%    {\sf  nboyer} & &  & &   & &  & & &  \\
-%    {\sf   circsim?} & &  & &   & & & & &   \\
- \hline
-\end{tabular}}
-
-\centerline{(b)}
-\vskip -2mm
-\caption{Experimental results comparing RGC and LGC. Table (a) gives
-  data related to liveness analysis, and (b) gives garbage collection
-  data. The average drag time is measured in allocated cons cells.}
-\label{fig:experimental-results}
-\normalsize
-\end{figure}
-
-\section{Collecting more garbage can never slow things down}
-\label{sec:lgc-always-better}
-Since garbage collection is  effectively asynchronous to the allocator
-thread, one might  worry as to how {\em  robust} our measurements are.
-For example,  while LGC would,  in general, collect more  garbage than
-RGC in the same heap state,  is it possible that on some programs, LGC
-might do  a larger number  of collections? We  prove a result  to show
-that   this  cannot   happen.    This  result   applies  to   classical
-mark-and-sweep and copying garbage  collectors and, we hypothesize, to
-generational collectors too.
-
-\begin{lemma}
-For the  same mutator,  a liveness-based collector  can never  do more
-garbage collections than a reachability-based collector.
-\end{lemma}
-
-\begin{proof}Assume, as before,  that time is measured in terms of the number of
-  cons cells allocated. Let us run  two copies of the mutator with the
-  two    different   collectors    but    with   synchronous    memory
-  allocations.   The   garbage    collections,   however,   happen
-  asynchronously.
-
-  To prove the lemma, it is enough to show the truth of the following 
-statement: After
- every  LGC invocation,  the count of LGC invocations   is  no greater
-than  RGC invocations.   The  base case  holds  since the  first
-invocations of both GCs happen at the same time.  Assume the statement
-to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-reachable cells, its heap would contain no more cells than RGC heap at
-the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-LGC, or LGC and RGC are both invoked next at the same time.  In either
-case, the statement holds after $n+1$ invocations of LGC.
-\end{proof}
-
-%% \begin{figure}[h]
-%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
-%% \caption{Memory  usage of  programs. The  solid  lines represents
-%%   number of  live cells and the  dashed lines show  the number of
-%%   reachable cells.  The red and blue  lines represent invocations
-%%   of RGC and  LGC---their heights indicate  the amount of memory
-%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
-%%       paper and are being used as placeholder}}}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-%%\clearpage
-
-\section{Related Work} Previous attempts to  increase  the space efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
-reallocation~\cite{jones89compile},                       deforestation
-techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
-methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
-
-\begin{figure}[t]
-\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
- \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
-\label{fig:memory-usage-lambda} \figrule
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-We have defined a notion of liveness on structured data;
-this generalizes classical liveness and strong liveness.
-We started with a general fully context-sensitive analysis which
-we proved correct with respect to a minefield semantics
-(this models the effect of garbage collection between every
-evaluation step).
-
-To avoid scalability issues (and to avoid performing part of
-the liveness computation at run time)
-we defined an 0-CFA version of this liveness analysis in which
-demands for function $f$ at all calling contexts are conflated into 
-a single demand $\sigma_f$.
-This enabled us to treat the liveness equations symbolically obtaining
-context-free grammars for liveness at each GC point (calls to user
-functions and to $\CONS$).
-These were then converted to DFAs for run-time consultation by the
-garbage collector. Experiments confirm the precision of the analysis.
-
-To obtain performance figures we compared a reachability-based garbage
-collector with a liveness-based  collector.  This showed a decrease in
-the  number  of  GCs,   more  garbage  collected  per  invocation.   A
-significant benefit of LGC is  that programs can run in smaller memory
-when compared to  RGC. This is potentially useful  in situations where
-memory  is  limited---as with  embedded  systems.  For  a majority  of
-programs, the garbage collection times were reduced.
-
-One issue we  highlighted was that while fewer  nodes were marked (and
-more garbage collected), sometimes  cons cells could be visited and
-traversed  multiple times  with different  sets of  liveness  paths to
-explore.  Further work includes  improvements to the classical copying
-collector to reduce the cost of this.
-
-\subsection*{Acknowledgements}
-The authors  would like  to thank Hemanshu  Vadehra for  a preliminary
-implementation of the prototype system. Amey Karkare was supported for
-this work by the DST/SERC fast track scheme for young scientists.
-
-\bibliography{fun_hra}{}
-\bibliographystyle{splncs}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-\section{Appendix stuff}
-
-\paragraph{Safety of using the modified \CONS\  rule:}
-Recall that while the correctness of the analysis was proved with
-respect to  the $\len$ rules  in Figure~\ref{fig:live-judge}, the
-equations   for  liveness   are  based   on  a   modification  of
-$\mathit{ref}$    in    the    case    of   \CONS\  as  given    by
-(\ref{eqn:mod-cons}).  We now show that  if the \bcar\,s and \bcdr\,s
-in live access  paths resulting  from this  modification are
-eliminated through  applications of $\hookrightarrow$,  we obtain
-the same results as given by the original rules.
-
-\comment{Proposal: We prove a result which goes as follows:
-
-\begin{quote}
-Lemma: Let  $\Lanv{i}{x}= \sigma$ for an a  program point $\pi_i$
-and a variable  $x$, when the judgement is  based on the original
-$\len$  rules.   Further, let  $\Lanv{i}{x}=  \sigma'$, when  the
-judgement is based on $\len$ with the modified \CONS\ rule.  Then
-$\sigma' \stackrel{*}{\hookrightarrow} \sigma$
-\end{quote}
-
-Here are some  ideas regarding how this can  be proved.  Consider
-an arbitrary equation
-$$ \Lanv{j}{y} = \ldots \Lanv{i}{x} \ldots$$ generated by the
-analysis.
-
-Now assume that $\Lanv{i}{x}= \sigma$.   We want to find out what
-happens to the access patterns in $\sigma$ in the access patterns
-of \Lanv{j}{y}.   To answer this,  we put \# markers  around each
-$\alpha \in \sigma$  and redo the equation using  an extension of
-$\len$ to take  care of \#-tagged patterns.  We  can then observe
-occurrences of \#-tagged patterns  in the liveness information at
-$\pi_j$.
-
-Here is  the $\len$ extension for  the original \CONS\  clause to take
-care of the \#-tagged patterns:
-$$\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          = \{x.\#\alpha\# \mid \#\acar\alpha\# \in \sigma\} \cup
-\{y.\#\alpha\# \mid \#\acdr\alpha\# \in \sigma\} $$
-And here  is  the $\len$ extension for  the modified \CONS\  clause.
-$$ \mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-= \{x.\#\bcar\alpha\# \mid \#\alpha\# \in \sigma\} \cup
-\{y.\#\bcdr\alpha\# \mid \#\alpha\# \in \sigma\} $$
-All  other rules  remain unchanged.  Tracking a  pattern  through this
-extension is sound because erasing the \#-tags would give us the same
-liveness as the original rules.
-
-Let us also extend $\hookrightarrow$ to \#-tagged strings as:
-$$\sigma_1  {\hookrightarrow} \sigma_2\,\, \mathit{iff}\,\, \sigma_2 =
-\{\alpha_1 \#\alpha\# \alpha_2 \mid  \alpha_1 \#\alpha'\# \alpha2 \in
-  \sigma_1\,\, and \,\, \{\alpha'\} \stackrel{*}{\hookrightarrow} \{\alpha\}\}$$
-
-Now we have an  observation represented by the following diagram:
-
-\centerline{\epsfig{file=diag1.eps, height=4cm}}
-
-We need  to make this  observation general enough.   Therefore we
-should prove  the observation even if \Lanv{i}{x}  = $\sigma$ had
-patterns with \bcar\ and \bcdr.   We should be able to prove this
-by structural induction on the rules.
-
-How does  the observation prove  the original lemma?   Suppose we
-replayed  judgements  with  ($\len$  +  \#) and  with  ($\len$  +
-modified \CONS +  \#).  Suppose that $\pi_i$ is  the last program
-point till which the results  are the same (because no \CONS\ was
-encountered).  Now  consider a  $\sigma \in \Lanv{i}{x}$  for any
-$x$.  Put \#-tags around each string in $\sigma$. Now we have the
-diagram:
-
-\centerline{\epsfig{file=diag2.eps, height=7cm}}
-
-This shows that modulo erasure of \#-tags, using the original $\len$
-judgement and  $\len$ judgement with modified \CONS\ rule followed by
-$\hookrightarrow$ give the same liveness.
-
-And I would like  to conclude that it is all right  to do our analysis
-with $\len$ + modified \CONS\ and then simplify using $\hookrightarrow$.
-
-}
-
-{\color{blue}\marginpar{From sec 3.1, liveness judgements}
-\paragraph{Semantic considerations.}
-The incremental liveness for generic primitive $\PRIM$ is given by
-$\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly) = 
-           \{x.\epsilon,y.\epsilon\}$.
-Earlier we suggested that classical strong liveness for addition,
-but not division, could use the stronger
-$ \{x.\epsilon,y.\epsilon \mid \epsilon \in \sigma\}$.
-However this is invalid: garbage collection
-could reuse dead, but reachable, $\CONS$ cell fields (of $v$ below)
-thereby producing a stuck state in a dynamically typed language.
-Hence we treat $+$ as like division not like addition---suppose
-% $(FOO)$ is a user function which may cause garbage collection, and consider:
-calling $\mbox{\footnotesize (FOO)}$ garbage collects and reallocates the cell pointed
-to by `v' in:
-
-\vspace{-15pt}
-
-\begin{quote}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1}  (\LET\ v\ $\leftarrow $\ (\CONS~2~3) \IN
-	  \UNL{2}  (\LET\ w\ $\leftarrow $\ (FOO) \IN
-	  \UNL{3}  (\LET\ x\ $\leftarrow $\ (\PRIM\ (\CAR\ v) 4) \IN
-	  \UNL{4}  (\RETURN\ w))))
-	\end{uprogram}}
-\end{quote}
-In principle, {\em dead code removal} of
-$(\PRIM\ (\CAR\ {\rm v})\ 4)$
-would render the stronger version valid, but still requires static typing
-to avoid the risk of {\em removing} a stuck state.
-We restrict ourselves to liveness annotation rather than code transformation.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]
-\footnotesize
-Old version of the liveness rules...
-
-\infrule[ret-live]
-        {}
-        {(\RETURN~x),\sigma,\Lfonly \len  x.\sigma}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[if-live]
-        { e_1, \sigma,\Lfonly \len \Lv_1
-          \andalso e_2, \sigma,\Lfonly \len  \Lv_2}
-        {(\SIF~x~e_1~e_2),\sigma,\Lfonly \len
-               \Lv_1\cup\Lv_2\cup \{x.\epsilon\}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[let-live]
-        {e, \sigma,\Lfonly  \len  \Lv}
-        {(\LET~x \leftarrow  s~\IN~e),
-          \sigma,\Lfonly \len   \Lv\setminus \{x.*\} \cup
-          \mathit{ref\/}(s,\Lv(x),\Lfonly)}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\infrule[live-define]
-        {e_f, \sigma, \Lfonly  \len
-%            \bigcup \{z_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-           \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \fen \Lfonly
-\\ \mbox{where }
-     (\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ \ e_f) \mbox{ is a member of }
-        d_1 \ldots d_k}
-\figrule
-  \caption{Liveness Judgement Rules}\label{fig:live-judge-old}
-\normalsize
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \newgray{lightgray}{.75}
-%\newcommand{\dem}[1]{\red{\scriptsize{$#1$}}}
-\newcommand{\dem}[1]{\psframebox[linestyle=none,fillstyle=none,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-%\newcommand{\liv}[1]{\blue{\scriptsize{$#1$}}}
-\newcommand{\liv}[1]{\psframebox[linestyle=none,fillstyle=solid,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-\psset{unit=1mm,nodesep=1mm,labelsep=0.5mm,arrows=->,linewidth=.1,framesep=0.2mm}
-\begin{center}
-  \begin{pspicture}(0,0)(120,50)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(12,5)
- \footnotesize
-   \putnode{let1}{origin}{15}{50}{$\pi_1$:\LET}
-
-    \putnode{test1}{let1}{-12}{-7}{\xtest}
-    \putnode{null}{let1}{0}{-7}{$\pi_2$:\NULLQ}
-    \putnode{lista1}{null}{0}{-7}{$\pi_3$:\lista}
-    \putnode{if}{let1}{18}{-7}{$\pi_4$:\SIF}
-
-    \putnode{test2}{if}{-8}{-7}{$\pi_5$:\xtest}
-    \putnode{listb1}{if}{3}{-7}{$\pi_6$:\listb}
-    \putnode{let2}{if}{14}{-7}{$\pi_7$:\LET}
-
-    \putnode{tl1}{let2}{-22}{-7}{\xtl}
-    \putnode{cdr}{let2}{-5}{-7}{$\pi_8$:\CDR}
-    \putnode{lista2}{cdr}{0}{-7}{$\pi_9$:\lista}
-    \putnode{let3}{let2}{15}{-7}{$\pi_{10}$:\LET}
-
-\putnode{rec1}{let3}{-13}{-7}{\xrec}
-\putnode{append}{let3}{0}{-7}{$\pi_{11}$:\append}
-\putnode{tl2}{append}{-10}{-7}{$\pi_{12}$:\xtl}
-\putnode{listb2}{append}{5}{-7}{$\pi_{13}$:\listb}
-
-\putnode{let4}{let3}{17}{-7}{$\pi_{14}$:\LET}
-\putnode{hd1}{let4}{-5}{-7}{\xhd}
-\putnode{car}{let4}{4}{-7}{$\pi_{15}$:\CAR}
-\putnode{lista3}{car}{0}{-7}{$\pi_{16}$:\lista}
-
-\putnode{let5}{let4}{16}{-7}{$\pi_{17}$:\LET}
-
-\putnode{ans1}{let5}{-4}{-7}{\xans}
-\putnode{cons}{let5}{6}{-7}{$\pi_{18}$:\CONS}
-\putnode{hd2}{cons}{-10}{-7}{$\pi_{19}$:\xhd}
-\putnode{rec2}{cons}{10}{-7}{$\pi_{20}$:\xrec}
-\putnode{ans2}{let5}{19}{-7}{$\pi_{21}$:\xans}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-
-%% \putnode{x}{let1}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{if}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{let3}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let4}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{-8}{0}{\liv{\Lv_1}}
-%% %\putnode{x}{ans2}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{plus}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{rec2}{1}{4}{\dem{\epsilonset}}
-%% \putnode{x}{plus}{-7}{0}{\liv{\Lv_6}}
-%% %\putnode{x}{let4}{-7}{0}{\liv{\{\xrec\}}}
-%% \putnode{x}{len}{3}{3}{\dem{\epsilonset}}
-%% \putnode{x}{len}{-7}{0}{\liv{\Lv_3}}
-%% \putnode{x}{tl2}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{tl2}{-6}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let3}{-10}{0}{\liv{\xtl.\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l2}{0}{4}{\dem{\epsilonset \cup \acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{l2}{-5}{0}{\liv{\Lv_4}}
-%% \putnode{x}{zero}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{zero}{-3}{0}{\liv{\Lv_0}}
-%% \putnode{x}{test2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{test2}{-7}{0}{\liv{\Lv_5}}
-%% \putnode{x}{if}{-6}{0}{\liv{\Lv_3}}
-%% \putnode{x}{null}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{null}{-9}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l1}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{vone2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{one}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{let5}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{l1}{-5}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{let1}{-17}{-1}{\liv{\{\xl\} \cup \xl.\acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{let1}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let1}{-7}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{if}{7}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let2}{8}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let3}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let3}{-7}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let4}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let4}{-7}{0}{\liv{\Lv_2}}
-%% \putnode{x}{vone2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{rec2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{let5}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{one}{-6}{0}{\liv{\Lv_2}}
-\ncline{-}{let1}{test1}
-\ncline{-}{let1}{null}
-\ncline{-}{let1}{if}
-\ncline{-}{null}{lista1}
-
-\ncline{-}{if}{test2}
-\ncline{-}{if}{listb1}
-\ncline{-}{if}{let2}
-\ncline{-}{let2}{tl1}
-\ncline{-}{let2}{cdr}
-\ncline{-}{let2}{let3}
-\ncline{-}{cdr}{lista2}
-
-\ncline{-}{let3}{rec1}
-\ncline{-}{let3}{append}
-\ncline{-}{let3}{let4}
-
-\ncline{-}{append}{tl2}
-\ncline{-}{append}{listb2}
-
-\ncline{-}{let4}{let5}
-\ncline{-}{let4}{hd1}
-\ncline{-}{let4}{car}
-\ncline{-}{car}{lista3}
-%% \ncline{-}{len4}{tl2}
-%% \ncline{-}{let4}{vone1}
-%% \ncline{-}{let4}{one}
-%% \ncline{-}{let4}{let5}
-\ncline{-}{let5}{ans1}
-\ncline{-}{let5}{cons}
-\ncline{-}{cons}{hd2}
-\ncline{-}{cons}{rec2}
-%% \ncline{-}{let5}{plus}
-\ncline{-}{let5}{ans2}
-\end{pspicture}
- \end{center}
-
-  \caption{Example to illustrate liveness computation}\label{fig:live-examp}
-\figrule
-\normalsize
-\end{figure}
-
-\newpage
-
-\begin{figure}[t]
-  \begin{tabularx}{320pt}{|| c | X | X | X ||}
-\hline
-Pgm  Pt & Demand  & Pre & Post \\
-\hline 
-$\pi_1$ & $\sigma$ & \{\lista\} $\cup$ \listb.$\sigma$ $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & $\phi$ \\
-\hline
-%% $\pi_2$ & $\epsilonset$ & \{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_3$ & $\epsilonset$ &\{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_4$ & $\sigma$ &\{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-%% $\pi_5$ & \epsilonset & \{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  \\
-%% \hline
-%% $\pi_6$ & $\sigma$ & \listb.$\sigma$ & $\phi$ \\
-%% \hline
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_7$ & $\sigma$  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$\\
-%% \hline
-%% $\pi_8$ & \Lf{\append}{1}(\{\bcdr\}$\sigma$) & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_9$ & \epsilonset $\cup$ \{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_{10}$ &$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-$\pi_{11}$ & \{\bcdr\}$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \lista $\cup$ \lista.\{\acar\bcar\} \\
-\hline
-$\pi_{12}$ &\Lf{\append}{1}(\{\bcdr\}$\sigma$)   &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-$\pi_{13}$ & \Lf{\append}{2}(\{\bcdr\}$\sigma$)  &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{14}$ & $\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$ \\
-\hline
-%% $\pi_{15}$ & \{\bcar\}$\sigma$  & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   &  \xrec.\{\bcdr\}$\sigma$\\
-%% \hline
-%% $\pi_{16}$ & \epsilonset $\cup$ \{\acar\bcar\}$\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$  & \xrec.\{\bcdr\}$\sigma$ \\
-%% \hline
-$\pi_{17}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   & $\phi$  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{18}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$  \\
-\hline
-$\pi_{19}$ &  \{\bcar\}$\sigma$        & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{20}$ &   \{\bcdr\}$\sigma$       & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{21}$ & $\sigma$ & \xans.$\sigma$ & $\phi$  \\
-\hline
-\end{tabularx}
-  \caption{Liveness annotations for the body of \append}\label{fig:xe_append}
-\figrule
-\normalsize
-\end{figure}
-
-\end{document}

### 2014-07-13 paper/sections/restOfOldPaper.tex
organization added

--- 2014-07-13 paper.tex
@@
-%------------------------------------------------------------%
-\documentclass{llncs}
-\pagestyle{plain}
-\usepackage{longtable} 
-%------------------------------------------------------------%
-%- common stuff starts here....
-\usepackage{etex}
-\usepackage{float}
-\usepackage{mathrsfs}
-\usepackage{booktabs}
-\usepackage{boxedminipage}
-\usepackage[T1]{fontenc}
-%\usepackage[pdf]{pstricks}
-\usepackage{epsfig}
-\usepackage{multirow}
-\usepackage{url}
-\usepackage[normalem]{ulem}
-\usepackage{schemepgm}
-\usepackage{graphicx}
-\usepackage{graphpap}
-\usepackage{tabularx}
-%\usepackage{color}
-%\usepackage{xcolor}
-\usepackage{algorithmic}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{pstricks}
-\usepackage{pst-text}
-\usepackage{pst-node}
-\usepackage{pst-tree}
-\usepackage{pst-rel-points}
-%\usepackage{pst-all}
-\usepackage{bcprules}
-%\usepackage{subfigure}
-%\usepackage{algorithmic}
-%\usepackage{anysize}     
-\input{fh_defs}
-%\newcommand{\comment}[1]{}
-\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
-\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
-    #1 \medskip \hrule \medskip}}}
-\newcommand{\scmin} {\mbox{\sf\em in}}
-\newcommand{\scmnull}{\mbox{\sf\em null?}}
-\newcommand{\scmpair}{\mbox{\sf\em pair?}}
-\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
-\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
-
-\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
-\newcommand{\stk}{\mbox{S}}       % stack
-\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
-\newcommand{\bang}{\mbox{\sc bang}}
-
-\begin{document}
-%  \title{Heap  Reference  Analysis  for Functional  Programs}
-\title{Liveness-Based Garbage Collection}
-
-% make a proper TOC despite llncs
-\setcounter{tocdepth}{2}
-\makeatletter
-\renewcommand*\l@author[2]{}
-\renewcommand*\l@title[2]{}
-\makeatletter
-
-  \author{Rahul Asati\inst{1} \and Amitabha Sanyal\inst{1} \and Amey Karkare\inst{2}
-    \and   Alan Mycroft\inst{3}}
-  \institute{IIT Bombay, Mumbai 400076, India\\
-    \email{$\{$rahulasati,as$\}$@cse.iitb.ac.in}, \and 
-             IIT Kanpur, Kanpur 208016, India\\
-    \email{karkare@cse.iitk.ac.in}, \and 
-             Computer Laboratory, University of Cambridge, CB3 0FD, UK\\
-    \email{alan.mycroft@cl.cam.ac.uk}}
-\maketitle
-
-\begin{abstract}
-Current garbage collectors  leave much heap-allocated data uncollected
-because they preserve data {\em reachable} from a root set.
-% However, it is only necessary to preserve {\em live} data---a subset of
-% reachable data.
-However, only  {\em live} data---a subset of  reachable data---need be
-preserved.
-
-Using   a    first-order   functional   language    we   formulate   a
-context-sensitive liveness  analysis for structured data  and prove it
-correct.   We  then use  a  0-CFA-like  conservative approximation  to
-annotate  each  allocation  and  function-call program  point  with  a
-finite-state  automaton---which   the  garbage-collector  inspects  to
-curtail reachability  during marking. As  a result, fewer  objects are
-marked  (albeit  with a  more  expensive  marker)  and then  preserved
-(e.g.\ by a copy phase).
-% and secondly because of the consequent
-%reduced number of collections during a computation.
-
-Experiments  confirm the  expected performance  benefits---increase in
-garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections, a decrease  in the memory size required  to run programs,
-and  reduced  overall  garbage  collection  time  for  a  majority  of
-programs.
-\end{abstract}
-
-%------------------------------------------------------------%
-
-%% \begin{figure}
-%% \tableofcontents       %%%%%%% BEWARE: find the \setcounter{page}{2}
-%%                        %%%%%%% and comment it out when deleting this TOC.
-%% \vspace{5cm}
-
-%% \section*{Page Budget}
-%% CC'2014 and ESOP'2014 have a limit of 20 pages.
-%% A draft page budget is:
-%% \begin{itemize}
-%% \item 1. abstract/introduction (3 pages)
-%% {\color{red}Currently 2.5 pages.}
-%% \item 2. core language syntax and semantics (2 pages).
-%% {\color{red}Currently 2 pages.}
-%% \item 3. liveness analysis and correctness (3 pages)
-%% {\color{red}Currently 3.8 pages.}
-%% \item 4. practical liveness computation (6 pages)
-%% {\color{red}Currently 6.2 pages.}
-%% \\ I think this is where should do all 3 of: the context-insensitive analysis
-%% (one which has a summary demand for each function), then the
-%% solving like differential equations, then using Mohri-Nederhof to
-%% get a NFA for each (call/cons) program point.
-%% \item 5. experimental framework and results (4 pages)
-%% \item 6. (optional) monotonicity of GC (0.5 page)---our technique can never
-%% cause more collections.
-%% \item 7. conclusions/acknowledgements/bibliography (1.5 page)
-%% \end{itemize}
-%% \end{figure}
-
-\section{Introduction}
-\label{sec:intro}
-
-Most modern  programming languages support dynamic  allocation of heap
-data.  Static  analysis of heap data  is much harder  than analysis of
-static   and   stack   data.    Garbage   collectors,   for   example,
-conservatively  approximate  the liveness  of  heap  objects by  their
-reachability  from a  set of  memory  locations called  the {\em  root
-  set\/}.  Consequently, many objects  that are reachable but not live
-remain  uncollected, causing  a  larger-than-necessary memory  demand.
-This is confirmed  by empirical studies on Haskell~\cite{rojemo96lag},
-Scheme~\cite{karkare06effectiveness}                                and
-Java~\cite{shaham02estimating} programs.
-
-Here we consider a first-order  pure functional language and propose a
-liveness  analysis  which  annotates  various program  points  with  a
-description  of variables and  fields whose  object references  may be
-dereferenced  in the future.   The garbage  collector then  only marks
-objects pointed by live references and leaves other, merely reachable,
-objects  to  be  reclaimed.    (Although  not  strictly  necessary,  a
-collector  would  normally {\em  nullify}  dead  variables and  fields
-rather than leaving dangling  references.)  Since there are fewer live
-objects   than   reachable   objects,   more  memory   is   reclaimed.
-Additionally, since  the collector traverses a smaller  portion of the
-heap, the time spent for each collection is also smaller.  The work is
-presented in  the context of a  stop-the-world non-incremental garbage
-collector (mark-and-sweep,  compacting or  copying) for which  we also
-show a  monotonicity result: that  our technique can never  cause more
-garbage  collections  to  occur   in  spite  of  changing  the  rather
-unpredictable  execution  points   at  which  collections  occur.   We
-anticipate that our technique  is applicable to more modern collectors
-(generational,  concurrent, parallel),  but leave  such  extensions to
-future work.
-
-We first define a {\em fully context-sensitive} (in the sense that its
-results  are unaffected  by function  inlining) liveness  analysis and
-prove it  correct.  However, fully context-sensitive  methods often do
-not scale,  and this analysis would  also require us  to determine, at
-run-time,  the internal  liveness of  a  function body  at each  call.
-Hence,   similarly   to   the   0-CFA   approach,   we   determine   a
-context-independent {\em summary} of  liveness for each function which
-safely   approximates   the   context-dependence   of   all   possible
-calls~\cite{chatterjee99relevant,cherem07fast,lee05static}.       (Note
-that  an intraprocedural context-insensitive  method which  assumes no
-information  about function  callers would  be too  imprecise  for our
-needs.)   In essence  our approach  sets up  interprocedural data-flow
-equations for the liveness summaries  of functions and shows how these
-can be solved symbolically  as context-free grammars (CFGs)\@.  We can
-then determine  a CFG  for each program  point; these are  then safely
-approximated with  finite-state automata  which are encoded  as tables
-for each program point.  For  garbage collection purposes only {\em GC
-  points}  (program points  representing  calls, to  $\CONS$  or to  a
-user-function) need to be stored.
-% which may call eventually call $\CONS$, need to be stored.
-
-We  previously proposed  an intraprocedural  method for  heap liveness
-analysis   for   a   Java-like   language~\cite{khedker07heap}   which
-statically inserted  statements nullifying dead  references to improve
-garbage collection; by  contrast nullification here occurs dynamically
-(which can work better with  aliasing) when the garbage collector acts
-on  liveness  annotations  to  avoid traversing  dead  references.   A
-workshop    paper~\cite{karkare07liveness}    outlined    the    basic
-0-CFA-style-summary  interprocedural  approach  to  functional-program
-liveness  analysis.   The  current  paper adds  the  context-sensitive
-analysis and better formalization along with experimental results.
-%   to  substantiate our ideas.
-
-\kern -3pt   %% AM: a very small tweak to get back to 2 pages for this
-             %% section.
-%% AM: rewrite the next paragraph to save 3-4 words and remove this \kern.
-
-%------------------------------------------------------------%
-\setcounter{page}{2}
-\subsubsection{Motivating Example}
-\label{sec:motiv}
-
-Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
-$\pi$ of an expression $e$  denotes a program point.  During execution
-of  the program, it  represents the  instant of  time just  before the
-evaluation  of  $e$.  
-%%While we discuss liveness, it  will also be important to consider an
-%%instance of time  just after the execution of  an expression, though
-%%we shall not be required to mark it with a label.
-% memory as  a (possibly  disconnected) directed
-% acyclic   graph.\footnote{Since  the   language   under  consideration
-%   (Section~\ref{sec:defs}) does not  have any imperative features, the
-%   heap cannot  have cycles.}
-% Here all data  is assumed to  be boxed, i.e.\
-% stored in heap and accessed through references.
-We  view the  heap  as a graph.  
-Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
-containing values.  Edges in the graph are {\em references} and
-emanate from {\em variables} or fields.
-Variable and field values may also be atomic values ($\NIL$, integers etc.)
-While it is convenient to box these in diagrams, our presented
-analysis treats them as non-heap values.
-
-Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
-The edges shown  by thick arrows are those  which are made live by the
-program. In addition,  assuming that
-the value of any reachable part of the program result may be explored
-or printed, the edges marked by thick dashed arrows are also live.
-A cell  is  marked and preserved  during  garbage collection,  only  if it  is
-reachable from the  root set through a path of  live edges.  All other
-cells can be  reclaimed.  Thus if  a garbage collection takes
-% place   at    the   time   represented   by   the    heap   shown   in
-at $\pi$ with the    heap   shown   in
-Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
-will be marked and preserved.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% MOTIVATING EXAMPLE
-\newcommand{\nilfigure}
-{\scalebox{0.75}{
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,0)(1,1)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
-\putnode{start}{origin}{0}{0}{}
-\putnode{stop}{origin}{10}{10}{}
-\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
-\end{pspicture}
-}}
-
-\begin{figure}[t!]
-%%%%{\color{Myblue}
-\begin{picture}(100,130)(0,-90)
-%  \begin{center}
-    \begin{tabular}{cc}
-\begin{boxedminipage}{.5\textwidth}
-      {\sf
-	\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
-	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
-	       \listb
-	  \UNL{1}      (\CONS\ (\CAR\  \lista)
-%          \UNL{2}          (\CAR\  \lista)
-	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
-          \listb))))
-          \UNL{0}
-	  \UNL{0} \hspace*{-.49\TAL} (\LET\ z  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
-	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
-	  \UNL{0} (\LET\ y  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
-	  \UNL{1}   (\LET\ w $\leftarrow$ (\append\ y z)\ \IN
-	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ w)))))
-	\end{uprogram}
-      }} 
-\end{boxedminipage}
-      &
-      \raisebox{-25mm}{\scalebox{.75}{
-	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
-      \psset{unit=1mm}
-      \psset{linewidth=.3mm}
-      \begin{pspicture}(0,-5)(70,60)
-	%\psframe(0,0)(73,60)
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
-	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
-%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
-	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	%      fillcolor=white]{$\times$}}}
-        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
-	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
-	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-		  fillcolor=white]{$\times$}}}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
-	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
-	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
-	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
-	\ncline[nodesepB=-.5]{*->}{c2}{d}
-	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
-	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
-	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
-	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
-	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
-	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
-%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
-        %here
-        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
-        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-
-%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
-%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
-	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
-	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
-%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      \end{pspicture}}} \\
-      (a) Example program.&
-      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
-	(b) Memory graph at $\pi$. Thick edges denote \\
-        {\white (b) }live links. Traversal stops at edges marked\\
-        {\white (b) }$\times$  during garbage collection.
-      \end{tabular}}
-    \end{tabular}
-\end{picture}
-
-\kern -3ex
-
-\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
-%  \end{center}
-%%%%}\color{Myblue} end
-\end{figure}
-
-\subsubsection{Organization of the paper}
-Section~\ref{sec:defs} gives
-the syntax and semantics of
-the language used to illustrate our analysis along with basic concepts
-and    notations.
-% Section~\ref{sec:operational}   describes    the
-% operational semantics  of the  language.
-Liveness analysis is described in
-Section~\ref{sec:liveness} followed by a sketch of a correctness
-proof relative to a non-standard semantics.
-Section~\ref{sec:computing} shows how to encode liveness as
-finite-state automata.
-Section~\ref{sec:experiments}   reports   experimental   results and
-Section~\ref{sec:lgc-always-better} proves that a liveness based
-collector can never do more garbage collections than a reachability
-based collector. 
-% and Section~\ref{sec:conclusion} concludes the paper.
-
 \section{The target language---syntax and semantics}
@@
 \end{figure}
-
-\end{document}

### 2014-07-14 paper/sections/abstract.tex
Still playing with git; modified the abstract slightly

@@
-Current garbage collectors  leave much heap-allocated data uncollected
-because they preserve data {\em reachable} from a root set.
-% However, it is only necessary to preserve {\em live} data---a subset of
-% reachable data.
-However, only  {\em live} data---a subset of  reachable data---need be
-preserved.
+{\color{red}
+The memory  requirement of a lazy functional  programming language can
+be decreased  in two ways. One can  use the result a  of {\em liveness
+analysis}  to preserve only  {\em live}  data---a subset  of reachable
+data---during garbage collection. Apart  from evaluated data this will
+also  results  in  early  release  of  memory  hogging  closures.   In
+addition, one  can also  evaluate closures that  are guaranteed  to be
+evaluated in  the future.  The  information that is required  for this
+eager evaluation comes from {\em strictness analysis}.
+}
 
@@
 (e.g.\ by a copy phase).
-% and secondly because of the consequent
-%reduced number of collections during a computation.
 

### 2014-07-14 paper/paper.tex
Changed the abstract slightly
Changed the names of the authors

@@
 
-  \author{Rahul Asati\inst{1} \and Amitabha Sanyal\inst{1} \and Amey Karkare\inst{2}
-    \and   Alan Mycroft\inst{3}}
+  \author{K. Prasanna Kumar\inst{1} \and Amey Karkare\inst{2}
+    \and   Amitabha Sanyal\inst{1}}
   \institute{IIT Bombay, Mumbai 400076, India\\
-    \email{$\{$rahulasati,as$\}$@cse.iitb.ac.in}, \and 
+    \email{$\{$prasanna,as$\}$@cse.iitb.ac.in}, \and 
              IIT Kanpur, Kanpur 208016, India\\
-    \email{karkare@cse.iitk.ac.in}, \and 
-             Computer Laboratory, University of Cambridge, CB3 0FD, UK\\
-    \email{alan.mycroft@cl.cam.ac.uk}}
+    \email{karkare@cse.iitk.ac.in}}
 \maketitle
@@
 
-\input{sections/restOfOldPaper}
+%\input{sections/restOfOldPaper}
 

### 2014-07-14 paper/sections/abstract.tex
Playing with commit

@@
-{\color{red}
 The memory  requirement of a lazy functional  programming language can
-be decreased  in two ways. One can  use the result a  of {\em liveness
-analysis}  to preserve only  {\em live}  data---a subset  of reachable
-data---during garbage collection. Apart  from evaluated data this will
-also  results  in  early  release  of  memory  hogging  closures.   In
-addition, one  can also  evaluate closures that  are guaranteed  to be
-evaluated in  the future.  The  information that is required  for this
-eager evaluation comes from {\em strictness analysis}.
-}
+be decreased  in two ways. The  first is to  use the result of  a {\em
+liveness  analysis} during  garbage collection  to preserve  only {\em
+live}  data---a  subset  of  {\em  reachable data}.  This  results  in
+collection of greater amount of garbage and consequently fewer garbage
+collections. Apart from evaluated data this will also results in early
+release  of memory  hogging closures.   The second  way is  to eagerly
+evaluate closures that  are guaranteed to be evaluated  in the future.
+The information that is required  for this eager evaluation comes from
+{\em strictness analysis}.
 
-Using   a    first-order   functional   language    we   formulate   a
-context-sensitive liveness  analysis for structured data  and prove it
-correct.   We  then use  a  0-CFA-like  conservative approximation  to
-annotate  each  allocation  and  function-call program  point  with  a
-finite-state  automaton---which   the  garbage-collector  inspects  to
-curtail reachability  during marking. As  a result, fewer  objects are
-marked  (albeit  with a  more  expensive  marker)  and then  preserved
-(e.g.\ by a copy phase).
+For  a first-order  lazy functional  language, we  present  an uniform
+framework  for liveness  and strictness  ana   lysis. We  start by  for a
+first-order functional  for language we  formulate a context-sensitive
+liveness analysis for  structured data and prove it  correct.  We then
+use   a  0-CFA-like  conservative   approximation  to   annotate  each
+allocation  and  function-call   program  point  with  a  finite-state
+automaton---which   the    garbage-collector   inspects   to   curtail
+reachability  during marking. As  a result,  fewer objects  are marked
+(albeit with a  more expensive marker) and then  preserved (e.g.\ by a
+copy phase).
 

### 2014-07-14 paper/sections/abstract.tex
abstract first version

--- 2014-07-13 paper/sections/abstract.tex
@@
-Current garbage collectors  leave much heap-allocated data uncollected
-because they preserve data {\em reachable} from a root set.
-% However, it is only necessary to preserve {\em live} data---a subset of
-% reachable data.
-However, only  {\em live} data---a subset of  reachable data---need be
-preserved.
+We  consider  two  ways  to  reduce the  memory  requirement  of  lazy
+functional  programming  languages. The  first  is  to modify  garbage
+collectors  to  preserve  only  {\em  live} data---a  subset  of  {\em
+reachable data}.  This results in an increase in the garbage reclaimed
+and  consequently in  fewer  garbage collections.   The  second is  to
+eagerly evaluate closures  that are guaranteed to be  evaluated in the
+future.
 
-Using   a    first-order   functional   language    we   formulate   a
-context-sensitive liveness  analysis for structured data  and prove it
-correct.   We  then use  a  0-CFA-like  conservative approximation  to
-annotate  each  allocation  and  function-call program  point  with  a
-finite-state  automaton---which   the  garbage-collector  inspects  to
-curtail reachability  during marking. As  a result, fewer  objects are
-marked  (albeit  with a  more  expensive  marker)  and then  preserved
-(e.g.\ by a copy phase).
-% and secondly because of the consequent
-%reduced number of collections during a computation.
+The  main contribution  of  this  paper is  an  uniform framework  for
+obtaining the information for the optimizations mentioned above---{\em
+liveness}  and {\em strictness}  analysis. Using  a central  notion of
+demand,  we  formulate   a  context-sensitive  liveness  analysis  for
+structured  data and  prove  it  correct.  We  then present strictness
+as a dual of liveness. We use a  0-CFA-like
+conservative   approximation   to   annotate   each   allocation   and
+function-call program point  with a finite-state automaton---which the
+garbage-collector inspects to  curtail reachability during
+marking. The result of the strictness analysis, being necessarily
+finite,  can  be directly used to insert code to evaluate closures.
 
-Experiments  confirm the  expected performance  benefits---increase in
+%As a result,  fewer objects are marked (albeit  with a more expensive
+%marker) and then preserved (e.g.\ by a copy phase).
+
+Experiments  confirm  the  expected  performance benefits.  The  early
+evaluation   of   closures  decreases   the   memory   demands  of   a
+program. Further,  liveness-based collection results in  a increase in
 garbage  reclaimed  and  a   consequent  decrease  in  the  number  of

### 2014-07-23 paper/sections/intro.tex
intro new

@@
 
-Most modern  programming languages support dynamic  allocation of heap
-data.  Static  analysis of heap data  is much harder  than analysis of
-static   and   stack   data.    Garbage   collectors,   for   example,
-conservatively  approximate  the liveness  of  heap  objects by  their
-reachability  from a  set of  memory  locations called  the {\em  root
-  set\/}.  Consequently, many objects  that are reachable but not live
-remain  uncollected, causing  a  larger-than-necessary memory  demand.
-This is confirmed  by empirical studies on Haskell~\cite{rojemo96lag},
-Scheme~\cite{karkare06effectiveness}                                and
-Java~\cite{shaham02estimating} programs.
+Functional  programming  languages  depend  on  garbage  collector  to
+efficiently  reclaim  unused  memory   allocated  by  programs  either
+explicitly (for example using constructors) or implicitly (for example
+for      closures).       However,      empirical      studies      on
+Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
+Java~\cite{shaham02estimating}  programs   have  shown   that  garbage
+collectors leave uncollected a lot of memory objects that are not live
+(here {\em live}ness  means the object can potentially be  used by the
+program at  a later  stage).  Due  to lack of  a suitable  analysis to
+distinguish  reachable  objects  from live  objects.  current  garbage
+collectors conservatively approximate the  liveness of heap objects by
+their reachability from a set of memory locations called the {\em root
+  set\/} resulting in unnecessary memory usage.
 
-Here we consider a first-order  pure functional language and propose a
-liveness  analysis  which  annotates  various program  points  with  a
-description  of variables and  fields whose  object references  may be
-dereferenced  in the future.   The garbage  collector then  only marks
-objects pointed by live references and leaves other, merely reachable,
-objects  to  be  reclaimed.    (Although  not  strictly  necessary,  a
-collector  would  normally {\em  nullify}  dead  variables and  fields
-rather than leaving dangling  references.)  Since there are fewer live
-objects   than   reachable   objects,   more  memory   is   reclaimed.
-Additionally, since  the collector traverses a smaller  portion of the
-heap, the time spent for each collection is also smaller.  The work is
-presented in  the context of a  stop-the-world non-incremental garbage
-collector (mark-and-sweep,  compacting or  copying) for which  we also
-show a  monotonicity result: that  our technique can never  cause more
-garbage  collections  to  occur   in  spite  of  changing  the  rather
-unpredictable  execution  points   at  which  collections  occur.   We
-anticipate that our technique  is applicable to more modern collectors
-(generational,  concurrent, parallel),  but leave  such  extensions to
-future work.
+Languages having lazy  semantics for evaluation put  further demand on
+memory as they  require heap allocated closure to be  carried from the
+point of occurrence  of an expression to the point  of (first) real use
+of  that expression.  Strictness analysis~\cite{some-refs}  statically
+determines  the  patterns  of  future usage  of  an  expression,  thus
+allowing some  of the  closures to  be evaluated at  the point  of the
+occurrence of the corresponding expression.
 
-We first define a {\em fully context-sensitive} (in the sense that its
-results  are unaffected  by function  inlining) liveness  analysis and
-prove it  correct.  However, fully context-sensitive  methods often do
-not scale,  and this analysis would  also require us  to determine, at
-run-time,  the internal  liveness of  a  function body  at each  call.
-Hence,   similarly   to   the   0-CFA   approach,   we   determine   a
-context-independent {\em summary} of  liveness for each function which
-safely   approximates   the   context-dependence   of   all   possible
-calls~\cite{chatterjee99relevant,cherem07fast,lee05static}.       (Note
-that  an intraprocedural context-insensitive  method which  assumes no
-information  about function  callers would  be too  imprecise  for our
-needs.)   In essence  our approach  sets up  interprocedural data-flow
-equations for the liveness summaries  of functions and shows how these
-can be solved symbolically  as context-free grammars (CFGs)\@.  We can
-then determine  a CFG  for each program  point; these are  then safely
-approximated with  finite-state automata  which are encoded  as tables
-for each program point.  For  garbage collection purposes only {\em GC
-  points}  (program points  representing  calls, to  $\CONS$  or to  a
-user-function) need to be stored.
-% which may call eventually call $\CONS$, need to be stored.
+We propose the  use of liveness analysis of heap  cells and strictness
+analysis  of expressions  to  reduce the  memory  requirement of  lazy
+functional  programming languages.   We modify  garbage collectors  to
+preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
+This results in an increase  in the garbage reclaimed and consequently
+in fewer  garbage collections.  Further, we eagerly evaluate closures
+that are guaranteed to be evaluated in the future (strict closures).
 
-We  previously proposed  an intraprocedural  method for  heap liveness
-analysis   for   a   Java-like   language~\cite{khedker07heap}   which
-statically inserted  statements nullifying dead  references to improve
-garbage collection; by  contrast nullification here occurs dynamically
-(which can work better with  aliasing) when the garbage collector acts
-on  liveness  annotations  to  avoid traversing  dead  references.   A
-workshop    paper~\cite{karkare07liveness}    outlined    the    basic
-0-CFA-style-summary  interprocedural  approach  to  functional-program
-liveness  analysis.   The  current  paper adds  the  context-sensitive
-analysis and better formalization along with experimental results.
-%   to  substantiate our ideas.
+The main contribution  of this paper is an uniform  framework for {\em
+  liveness} and {\em strictness}  analysis to enable the optimizations
+mentioned  above. Using  a central  notion of  demand, we  formulate a
+context-sensitive liveness  analysis for structured data  and prove it
+correct.  We then  present strictness as a dual of  liveness. We use a
+0-CFA-like conservative approximation to  annotate each allocation and
+function-call program point with  a finite-state automaton---which the
+garbage-collector inspects to curtail reachability during marking. The
+result of  the strictness analysis,  being necessarily finite,  can be
+directly used to insert code to evaluate closures.
 
-\kern -3pt   %% AM: a very small tweak to get back to 2 pages for this
-             %% section.
-%% AM: rewrite the next paragraph to save 3-4 words and remove this \kern.
+We previously proposed an interprocedural liveness analysis to improve
+garbage collection first order eager subset of Scheme~\cite{cc-paper}.
+The current paper extends the work to handle lazy evaluation semantics
+for the same language.
+
+Experiments  confirm  the  expected  performance benefits.  The  early
+evaluation   of   closures  decreases   the   memory   demands  of   a
+program. Further,  liveness-based collection results in  a increase in
+garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
+collections, a decrease  in the memory size required  to run programs,
+and  reduced  overall  garbage  collection  time  for  a  majority  of
+programs.
 
 %------------------------------------------------------------%
-\setcounter{page}{2}
 \subsubsection{Motivating Example}

### 2014-08-30 paper/sections/syntax-and-semantics.tex
The syntax and op. semantics section

--- 2014-07-13 paper/sections/restOfOldPaper.tex
@@
 \label{sec:defs}
-We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
-functions ($\CONS$, $\PRIM$ etc.).
-The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
-it has eager semantics and restricts programs to be in 
-Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
-where all actual parameters to functions are variables.
-This restriction does not affect expressibility (and indeed we feel free
-to ignore it in examples when inessential), but simplifies the analysis formulation.
-Additionally, as in the three-address instruction form familiar from compiler texts,
-it forces each temporary to be named and function calls to be serialized
-(necessary to get an unambiguous definition of liveness).
-We further require that each variable in a program is distinct, so that no scope shadowing
-occurs---this simplifies proofs of soundness.
-In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
-or return ($\RETURN$).
-The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.
+We let $x$, $y$, $z$ range over variables, $f$ over user-functions and
+$p$ over  primitive functions ($\CONS$, $\PRIM$ etc.).   The syntax of
+our language  is shown  in Figure~\ref{fig:lang-syntax}; it  has eager
+semantics and  restricts programs to be in  Administrative Normal Form
+(ANF)~\cite{chakravarty03perspective}  where all actual  parameters to
+functions   are   variables.   This   restriction   does  not   affect
+expressibility (and indeed we feel  free to ignore it in examples when
+inessential). In addition, this  form also explicates the formation of
+closures.  We further  require  that  each variable  in  a program  is
+distinct, so  that no scope shadowing  occurs---this simplifies proofs
+of soundness.  In this  formulation expressions: either perform a test
+($\SIF$), make a computation step ($\LET$) or return ($\RETURN$).  The
+$\RETURN$ keyword is logically redundant, but we find it clarifies the
+semantics and analysis.
 
@@
 % As  mentioned earlier,
-We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.
+We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
+language syntax) with the program point just before expression $e$.
 
@@
 \begin{array}{rlcl@{\hspace{2em}}l}
-% \begin{eqnarray*}
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
-\mbox{-- Environment} \\ 
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
-  Val}) & \mbox{-- Heap}
-% \end{eqnarray*}
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Loc} & \mbox{-- Environment} \\ 
+v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Val \times
+  Val}& \mbox{-- Values}\\
+c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (\mathit{Val} +
+\mathit{Clo}) & \mbox{-- Heap}
 \end{array}
 \]
+%Note that in the eager case, if a cons had a numeric argument, this
+%fact was known when the cons was created. Thus you could have unboxed
+%numbers. This is not possible now. For instance, the argument of a
+%cons could be a closure which evaluates to a number.
+%Decision - numbers are boxed
 Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
@@
         {}
-        {\rho, \heap, \pp{\pi}{k}
-          \rightsquigarrow \heap, k}
+        {\rho, \stk, \heap,   \pp{\pi}{k}
+          \rightsquigarrow \rho, \stk, \heap, k}
 \end{minipage}\begin{minipage}{0.655\textwidth}
@@
       {
-\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
+%%\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
 }
-      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
-        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
+      {\rho, \ell \bullet \stk,\heap, \pp{\pi}{(\CONS\,\, x\ y)}
+        \rightsquigarrow \rho, \stk, \heap[\ell\mapsto(\rho(x),\rho(y))],
         \ell}
@@
 \infrule[ord-car]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
-          \rightsquigarrow \heap, v_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
+        {\heap(\rho(x)) = (e_1, e_2)}
+        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
+          \rightsquigarrow \rho, \stk, \heap, e_1}
+\end{minipage}
+\begin{minipage}{0.395\textwidth}
+\infrule[ord-car]
+        {\heap(\rho(x)) = e}
+        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
+          \rightsquigarrow \rho, (\CAR \bullet):\stk, \heap, v_1}
+\end{minipage}\\[0.5ex]
+
+\begin{minipage}{0.595\textwidth}
 \infrule[ord-cdr]
@@
 \renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-
-\section{Liveness}\label{sec:liveness}
-
-In classical liveness analysis a variable is either `live' (its value
-may be used in future computation) or `dead' (definitely not used).
-Semantically, a variable is dead at a given program point
-if arbitrary changes to its value have no effect on the computation.
-Later we will use $\bot$ to represent a value which `explodes' when
-it is used in a computation; dead variables can safely have their
-value replaced with $\bot$.
-For heap-allocated data  we need a richer model of liveness
-in that both variables and fields of $\CONS$ cells may be dead or live.
-Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
-liveness of the structure reachable from a variable is a set of
-{\em access paths}
-which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
-grammar notation.
-Thus the liveness of $x$ being
-$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
-only refer to the second and third members of $x$ considered as a list.
-Semantically, access paths are prefix-closed, as accessing a field
-requires accessing all the paths from the variable to the field,
-and hence the above liveness is properly written 
-$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
-The classical notions of a scalar variable being live or dead correspond to
-$\{\epsilon\}$ and $\{\}$.
-
-The overall liveness (also written {\em liveness environment} for emphasis)
-at a program point is conceptually a mapping from
-variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
-notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
-instead of the map
-$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
-Analogously to classical liveness, the liveness at program point $\pi$
-in $\pi:e$ is the liveness just before executing $e$.
-
-A complementary notion to liveness is {\em demand}.  The demand for
-expression $e$ is again an access path---that subset of of $\{\acar,\acdr\}^\ast$ which the
-context of $e$ may explore of $e$'s result.  So, for example given a
-demand $\sigma$ and the expression $\pi: (\RETURN\ x)$, the liveness at
-$\pi$ is exactly $x.\sigma$.
-The classical analogy of this is in {\em strong liveness}, where
-an assignment node $n: x:=y+z$ causes $y$ and $z$ to be live on entry
-to $n$ if (and only if) $x$ is live at exit of $n$---the liveness
-of $x$ at exit from $n$ becomes the demand on $y+z$.
-Note that, for an operation like division which
-may raise an exception, the assignment $n: x:=y/z$ makes $y$ and $z$ live
-regardless of the liveness of $x$.
-
-We use $\sigma$ to range over demands, $\alpha$ to range over access paths
-and $\Lv$ to range over liveness environments. The notation
-$\sigma_1\sigma_2$ denotes the set $\lbrace \alpha_1\alpha_2 \mid
-\alpha_1 \in \sigma_1, \alpha_2 \in \sigma_2\rbrace$. Often we shall
-abuse notation to juxtapose an edge label and a set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
-Finally, we use $\Lfonly$ to range over {\em demand transformers}.
-These transform demands on a function call
-into demands for its formal parameters:
-given a user function $f$, defined by
-$(\DEFINE\ (f\ x_1\ \ldots\ x_n)\ \ e_f)$ and called with demand $\sigma$,
-then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
-
-Note that liveness refers to variables and fields, and not to $\CONS$ cells
-(i.e.\ to edges in the memory graph, not to locations themselves).
-Hence liveness of $\{x.\epsilon,x.\acar\}$ means that future computation
-may refer to the value $\ell$ of variable $x$, and also to the $\CAR$ field
-of location $\ell$.
-In the absence of other pointers to heap location $\ell$, we are
-certain that the $\CDR$ field of $\ell$ will not be referenced
-and may hence be corrupted arbitrarily.
-Note therefore, that while  $\ell$ cannot be garbage collected, any
-location $\ell'$ stored in the $\CDR$ field of $\ell$ would be garbage
-(again provided there are no other aliases to  $\ell$).
-
-%% \comment{
-%%   \begin{enumerate}
-%% \item Set the framework for liveness. Basic notions: access
-%%   paths and expressions.
-%% \item An informal working definition of liveness. Minefield will
-%%   provide the formal definition.
-%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
-%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
-%%   rules. One of the changes from  the earlier version is the lack of a
-%%   post-liveness in  the judgements. This anticipates  the intended use
-%%   of  the analysis  for garbage  collection and has  to be  carefully
-%%   explained.
-%%   \end{enumerate}
-%% }
-
-%% We now  set up  the framework for  liveness analysis.  Given  a memory
-%% graph, a link in the graph  can be described by starting with a member
-%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
-%% \CDR\ dereferences to reach the link.  For notational convenience, the
-%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
-%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
-%% dereferences    is    represented   by    a    string   $\alpha    \in
-%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
-%% in the  memory graph that starts  at $x$ and  traverses adjacent links
-%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
-%% We call the string $x.\alpha$ an {\em access expression\/}, the string
-%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
-%% graph   by  the   access  expression   an  {\em   access   path}.   In
-%% Figure~\ref{fig:mot-example},        the       access       expression
-%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
-%% containing the value  $4$.  Most often the memory  graph is clear from
-%% the context,  and we shall use  access expressions to  refer to access
-%% paths. An  access path  used to refer  to a  link in the  memory graph
-%% refers    to     the    last     link    in    the     path.     Thus,
-%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
-%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
-%% set of access paths, then  $x.\sigma$ will represent the set of access
-%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
-%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
-%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
-%% \sigma\rbrace$.
-
-%% \subsection{Liveness Analysis}
-%% \label{sec:exp-liveness}
-%% 
-%% Consider the memory graph at a program point $\pi$ during an execution
-%% of a program. A  link in the memory graph is said  to be {\em live} at
-%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
-%% execution  starting from  the  evaluation of  the expression  labelled
-%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
-%% be live at  $\pi$, if, during {\em any} execution  of the program, the
-%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
-%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
-%% In the first, the path  itself represents the sequence of dereferences
-%% to  access  the last  edge,  as,  for  example, while  evaluating  the
-%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
-%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
-%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
-%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
-%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
-%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
-%%   implicitly  live} if it is live in the second sense.  Clearly, static
-%% determination   of  implicit  liveness   requires  an   analysis  that
-%% determines  potential sharing  during  any execution  of the  program.
-%% However, the intended application  of our liveness analysis is garbage
-%% collection  during which  we have  access  to the  entire heap.   This
-%% obviates the need for a  separate sharing analysis and we restrict our
-%% attention to explicit liveness only.
-%% 
-%% 
-%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
-%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
-%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
-%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
-%% %% \bcdr\ represent application of  the constructor \CONS.  The following
-%% %% example motivates the need for these symbols.
-%% 
-%% %% \begin{example}\label{exmp:motiv-bar}
-%% %%   Consider the expression at program point $\pi_1$ in
-%% %%   \begin{center}
-%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
-%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
-%% %% 			    $\pi_3\!:${$\cdots$})}}
-%% %%   \end{center}
-%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
-%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
-%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
-%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
-%% %%   \begin{itemize}
-%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
-%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
-%% %%     live at $\pi_1$.
-%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
-%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
-%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
-%% %%   \end{itemize}
-%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
-%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
-%% %% \end{example}
-%% 
-%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
-%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
-%% %% resulting access patterns may not  be directly useful in that they may
-%% %% not describe a  path in the memory graph.  We  define a {\em canonical
-%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
-%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
-%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
-%% %% to obtain a canonical access  pattern representing the same edge. This
-%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
-%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
-%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
-%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
-%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
-%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
-%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
-%% %% cannot  be  used for  nullification  during  garbage collection.   The
-%% %% liveness of a such a path is irrelevant, and the path is not considered
-%% %% any further.
-%% 
-%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
-%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
-%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
-%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
-%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
-%% %% element in $\sigma_2$, i.e.
-%% 
-%% %% \begin{eqnarray*}
-%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
-%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
-%% %% \end{eqnarray*}
-
-\subsection{Liveness Analysis}
-First recall the classical formulation of liveness (as sets of simple variables)
-on three-address instructions,
-$
-\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I)
-     \setminus \mathit{def\/}(I)
-     \cup   \mathit{ref\/}(I)
-$,
-and then note that {\em strong liveness} needs, when $I$ is the instruction $z:=x+y$,
-that $\mathit{ref}(I)$ be refined to
-$\{x,y\}$ if $z \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
-
-Our liveness analysis formulated in Figure~\ref{fig:live-judge}
-is analogous.
-% has three analogous aspects.
-% \begin{itemize}
-% \item
-Firstly, the function $\mathit{ref}$, when given a statement $s$,
-returns the incremental liveness {\em generated} by $s$.
-Because we generalize {\em strong} liveness, $\mathit{ref}$ needs a second parameter,
-specifying the demand $\sigma$ on the result of $s$,
-to determine which access paths of its free variables are made live.
-The cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
-% The cases for
-% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
-% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
-% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
-A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
-argument (even if its result is never used).
-This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
-(note $\sigma$ may be $\{\}$).
-In the opposite  sense, the  demand  of $\acar\sigma$  on
-(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-that \CONS\ does not, by itself, dereference its arguments.
-Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
-$\mathit{ref}$
-expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
-Constants generate no liveness.
-
-% \item
-The function $\mathcal{L}$ now gives the (total) liveness of an expression $e$.
-The cases $\RETURN$ and $\SIF$ are straightforward, but note the liveness $x.\epsilon$ generated
-by the latter.
-The case $(\LET\ z\leftarrow s\ \IN\ e')$
-resembles a three-address instruction:
-the liveness of $e$ is given by taking the liveness, $\Lv$, of $e'$,
-killing any liveness of $z$ and adding any incremental liveness from $s$.
-The main subtlety is how the liveness of $z$ in $\Lv$ is converted to a demand $\Lv(z)$ to be
-placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-% \item
-%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-Finally, the judgement form $\mathit{Prog} \len \Lfonly$ is used to determine $\Lfonly$\@.
-Analogously to classical liveness being computed as
-a solution of dataflow equations, we require, via inference rule ({\sc live-define}),
-$\Lfonly$ to satisfy the fixed-point property that:
- when we assume $\Lfonly$ to be the family of demand transformers for the program
- then the calculated liveness of each function body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees
- with the assumed  $\Lfonly$.
-As usual, there are often multiple solutions to $\Lfonly$; all are safe
-(see Section~\ref{sec:correctness})
-but we prefer the least one as giving the least liveness subject to safety---and
-hence greatest amount of garbage collected.
-% \end{itemize}
-
-We make three observations: firstly the rule ({\sc live-define}) has a least solution
-as $\mathcal{L}(\cdot)$ is monotonic in $\sigma$; secondly that
-({\sc live-define})
-resembles the rule for type inference of mutually recursive function definitions,
-and thirdly the asymmetry of demand and liveness (compared to post- and pre-liveness
-classically) is due to the functional formulation here.
-
-Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
-(representing a fully context-sensitive analysis)
-can be safely approximated, for each function, by a {\em procedure summary}
-(unifying the contexts in the style of 0-CFA)\@.
-The summary consists of a pair of a single demand and, for this demand,
-the corresponding tuple of demands the function makes on its arguments.
-
-%%% save two lines to bring up section 3.2.
-
-% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
-% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
-% enables us to do a context-dependent analysis of a call without having
-% to enter the body of the called function.
-
-%% We present  our liveness  analysis using judgement forms $\len$
-%% for both expressions and programs.
-%% These are given formally in Figure~\ref{fig:live-judge}, but
-%% first we an intuitive understanding of the issues involved.
-%% Consider the expression:
-%% 
-%% \begin{center}
-%% \begin{uprogram}
-%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
-%%                         \px) \IN\  
-%%            % \UNL{0}
-%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
-%%                         \pw ) \IN \ldots) 
-%% \end{uprogram}
-%% \end{center}
-%% 
-%% 
-%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
-%% \pw.\epsilon\}$.
-%% Informally, this is so since \pw\ and \pz\ are used in
-%% (\PRIM~\pz~\pw ).
-%% 
-%% \begin{enumerate}
-%% \item The liveness  at $\pi$ is obtained by  removing the access paths
-%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
-%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
-%%     similar  to liveness  analysis of  imperative programs,  where the
-%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
-%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
-%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
-%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
-%%     liveness generated and  killed in $s$}. 
-%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
-%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
-%%   as a {\em  demand}.  The demand on an expression consists of  access
-%%   patterns  representing all  possible future  uses of  its  value.  A
-%%   demand  of $\epsilon$  means that  only the  reference resulting
-%% %  from the evaluation  of
-%%   from  $(\CAR\,\,\px)$ will be used.
-%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
-%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
-%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
-%% \end{enumerate}
-%% 
-%% In general,  an access path is  live before an  expression because the
-%% link represented by it is used (i) while evaluating the expression, or
-%% (ii) while using the result of the expression, or (iii) simply because
-%% the link  is live at  a later program  point.  When the  expression in
-%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
-%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
-%% the access  path \pw.   Note that \px.\acar\  is live in  this example
-%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
-%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
-%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
-%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
-%%
-
-% \paragraph{Judgements formally.}
-% Let  $\Lv$  denote a  liveness  environment (a function from
-% variables to access paths, abused using grammar notation) and
-% $\sigma$ be a demand.
-% The judgement form $e, \sigma \len \Lv$,
-% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
-% environment  before $e$  is $\Lv$ and
-% is thus analogous to the classical notion of {\em transfer function}.
-
-%    It is  interesting to  note that,
-% unlike traditional liveness analysis,  our judgements do not require a
-% liveness after the expression. This  is because of the coming together
-% of  two reasons.   First, it  will be  clear while  discussing garbage
-% collection,  that it is  enough for  our purposes  to assume  that the
-% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
-% restriction of our  programs to ANF ensures that  the post-liveness of
-% each  expression  $e$  inside   the  function  body  continues  to  be
-% $\emptyset$. Thus the post-liveness can be dropped altogether.
-
-%==============================================================
-\begin{figure}[t]  
-% \footnotesize
-\begin{eqnarray*}
-\mathit{ref\/}(\kappa,\sigma,\Lfonly)
-          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
-          &=& \{x.\sigma\} \\
-\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
-          &=& \{x.\epsilon,y.\epsilon\} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \\
-\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
-        \mathcal{L}(e_1,\sigma,\Lfonly) \cup
-        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
-        \{x.\epsilon\} \\
-\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
-        \Lv
-           \setminus \{x.*\}
-           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
-\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
-\end{eqnarray*}
-
-\begin{minipage}{0.85\textwidth}
-\infrule[live-define]
-        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
-           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \len \Lfonly
-\\ \makebox[0mm]{where
-     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
-\end{minipage}
-
-\normalsize
-  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%==============================================================
-
-% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
-% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
-% liveness generated  before the right-hand-side of  a \LET\ definition.
-% Constants    do   not   generate    any   liveness.     The   operator
-% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
-% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
-% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
-% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
-% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-% that \CONS\ does not, by itself, dereference its arguments.
-% 
-% 
-% Unlike primitives, the mapping from the demand on a user-function call
-% to the  demands on its arguments  has to be derived  from the function
-% body itself---as a fixed point for a recursive function.
-% The function  \Lfonly\ captures this mapping for
-% all functions  defined in the  program.  For a  function $\mathit{f}$,
-% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
-% a    call     to    the     $i$th    argument.     The     rule
-% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
-% confirmed by the body of all the functions, then the guess is correct.
-% {\color{blue}Note the similarity of this rule to type-inference
-% for recursive program schemes.}
-
-% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
-% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
-% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
-% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
-% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
-% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
-% liveness generated before $s$.
-
-\subsection{Minefield semantics and correctness}
-\label{sec:minefield}
-\label{sec:correctness}
-
-This section gives a modified semantics
-%(a reference semantics)
-which checks liveness annotations
-at run time, and `explodes' when these are found to be inconsistent with execution behavior,
-but otherwise behaves as the standard semantics.
-We show that such explosions never occur and hence run-time checks
-can be elided.
-%% We first assume some demand transformers $\Lfonly$;
-%% our proof will rely on this being a solution (i.e.\ satisfies $\mathit{Prog} \len \Lfonly$)
-%% of liveness analysis, but we do rely in this yet.
-We first assume an arbitrary  demand transformers $\Lfonly$ which satisfies
-$\mathit{Prog} \len \Lfonly$.
-We then enrich the abstract machine state $\rho, \stk, \heap, e$ to
-$\rho, \stk, \heap, e, \sigma, \Sigma$.
-Here $\sigma$ is the demand to the currently active function,
-thus the liveness $\Lv$ at $e$ is $\mathcal{L}(e,\sigma,\Lfonly)$,\footnote{
-This a simple liveness propagation using $\mathcal{L}(\cdots)$
-and $\mathit{ref}(\cdots)$ as $\Lfonly$ is assumed given.}
-and $\Sigma$ is a stack of demands---one for each function frame
-pushed in $\stk$.
-
-Second, we augment $\mathit{Val}$ with a value $\bot$.
-To model strong liveness $\bot$ may be copied freely,
-including into a $\CONS$ cell,
-% but reduces the state to $\bang$
-but explodes
-when used computationally (in a primitive operation other than a copy).
-Additionally we define
-$\mathit{GC}(\Lv,\Sigma): (\rho, \heap, \stk) \mapsto (\rho', \heap', \stk')$
-which determines {\em live-reachability}\footnote{Reachability curtailed by liveness information.}
-%%%%% \marginpar{Say more about live-reachability?}
-using $\rho$ and the $\rho$'s in $\stk$ as the root set
-and following links in $\heap$ {\em only as far as allowed by $\Lv$ and $\Sigma$}.
-% The effect of $\mathit{GC}(\Lv,\Sigma)$ is to replace
-Hence $\mathit{GC}(\Lv,\Sigma)$ replaces
-live-unreachable values---in $\rho$, in $\heap$ and the $\rho$'s in \stk---with $\bot$.
-For example, if $x.\epsilon \not\in \Lv$ then $\rho'(x) = \bot$.
-%, regardless of $\rho(x)$.
-Only $\mathit{GC}(\cdots)$ introduces $\bot$.
-
-Third, we update the semantics in four ways: ({\em i\/}) we arrange that all
-($\rightarrow$) transitions on expressions $e$ first use $\mathit{GC}(\cdots)$
-to update the state and then continue as before;
-and ({\em ii\/}) whenever the value $\bot$ is used computationally in a reduction
-($\rightsquigarrow$), we enter a distinguished stuck state $\bang$.
-For example, supposing $\rho(x)=\bot$
-then
-$\rho, \heap, s \rightsquigarrow \bang$ if $s$ is
-$(\CAR\,\, x)$, $(\CDR\,\, x)$ or $(\PRIM\,\, x\ y)$,
-but not if $s$ is $(\ID~x)$, $(\CONS~x~y)$ or $(f~x~y)$.
-Finally ({\em iii\/}) we make a similar change to the ($\rightarrow$) reduction for ({\sc ord-if})
-and ({\em iv\/}) augment the ({\sc ord-let}) rule for primitives to propagate
-$\bang$ from ($\rightsquigarrow$) to ($\rightarrow$).
-%%%%%% {\color{blue}The full minefield semantics is given as an appendix?}
-
-% Suppose liveness for the current $e$ under reduction is $\Lv$; then:
-% For every variable $x$ for which $x.\epsilon \not\in \Lv$ then replace
-% $\rho(x)$ with $\bot$.
-% Also, for every $\ell \in \mathit{Loc}$, replace the $\CAR$ field
-% of $\heap(\ell)$ with $\bot$ when
-% which there is no path $x.\alpha\acar$ in $\Lv$ such that following
-% $\alpha$-links in $\heap$ from $\rho(x)$ reaches $\ell$.
-% Correspondingly, replace the $\CDR$ field of $\ell$ if there
-% is no path $x.\alpha\acdr$ in $\Lv$.
-% \item
-% 
-% {
-% \footnotesize
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-car]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CAR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-cdr]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CDR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}\\[1ex]
-% 
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-prim]
-%       {\rho(x) = \bot \vee \rho(y) = \bot}
-%       {\rho, \heap, {(\PRIM\,\, x\ y)}
-%         \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-null]
-%         {\rho(x)= \bot
-%         }{\rho, \heap, {(\NULLQ\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% }
-%
-
-The resulting minefield semantics behaves identically
-(identical heap, identical steps, including possible non-termination)
-to the standard semantics, except for one possibility---the
-minefield semantics going $\bang$ while the standard semantics
-continues (either to a halt state, to a stuck state, or reduces forever).
-
- We now prove a result that relates
-liveness analysis to the semantics.
-\begin{proposition}
-Given program $P$ with $P \len \Lfonly$, then in the minefield
- semantics $P \rightarrow \bang$ can never occur
- (cf.\ `well-typed programs do not go wrong').
-\end{proposition}
-\textit{Proof outline:}
-Space does not permit a full proof, but we give the two main steps.
-We proceed by contradiction and assume there is a program $P$ for which
-$P \len \Lfonly$ can enter state $\bang$.
-The first step is to construct a program $P'$ with identical behavior, but with
-no user-function definitions, using inlining.
-This is possible because a program which goes $\bang$ does so after a finite number
-of reductions, and hence even recursive functions have only had a finite number of
-invocations.
-We hence repeatedly inline-expand user-function calls in $P$
-until we obtain a program $P'$ which behaves identically\footnote{
-Modulo replacement of
-({\sc ord-call}) and
-({\sc ord-return}) steps with
-({\sc ord-id}) steps.
-}
-to $P$ in the standard semantics, but makes no user-function calls.
-Any remaining, non-executed, calls can 
-%be left or
-replaced with a new primitive with the same
-demand-to-liveness transfer function---thus making
-$P'$ a simple expression $e'$.
-%%%%%
-Not only do the program points in reducing $e'$ correspond
-one-one to states during evaluation of $P$,
-but also the liveness associated with a point
-in $e'$ is identical to the liveness at the corresponding state
-$(\rho, \stk, \heap, e, \sigma, \Sigma)$ of $P$ (concatenating the liveness
-$\mathcal{L}(e,\sigma,\Lfonly)$ at $e$
-with the liveness, obtained from $\Sigma$, of the call sites in $\stk$ and
-after renaming the variables correspondingly to the inlining which produced $e'$).
-This assertion relies on the analysis being fully context-sensitive, and noting that
-while the change of scope caused by inlining changes variable visibility
-between $P$ and $e'$ it does not change the liveness---as local-to-a-function
-$\LET$-variables whose scope has been prolonged due to inlining
-are dead in the prolonged scope.
-
-The second step of the proof is to show that $e'$ cannot go $\bang$.
-We proceed by induction.
-Correctness of the $\SIF$ and final $\RETURN$ forms are immediate;
-the $(\LET~z\leftarrow s~\IN~e)$ form requires showing
-the inference rules ensure that any value referenced via $z$ in $e$
-was already live-reachable (via another variable and path) in any
-enclosing expression (so that $\mathit{GC}(\cdots)$ could not re-write it to $\bot$).
-
-% \newpage
-
-\section{Computing liveness and its encoding as a table}\label{sec:computing}
-
-Section~\ref{sec:liveness} gave a context-sensitive liveness analysis
-and proved it correct with reference to a {\em minefield} semantics.
-% Practical computation was not considered.
-For practical use we need to solve the liveness equations {\em finitely and
-symbolically}.  As expressed mathematically, and given a fixed program,
-three things are potentially unbounded:
-({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
-({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
-({\em iii\/}) the number of such access paths.
-
-We commonly first solve ({\em i\/}) by reducing the number of distinct
-calling contexts
-(e.g.\ to a single unified context in 0-CFA style).
-However, it turns out that we can solve the equations for $\Lfonly$ symbolically
-without this reduction, so here we defer this to Section~\ref{sec:bodylivenessbodies}.
-(It also allows easier extension to  dynamically determined liveness---future work.)
-%
-We address ({\em ii\/}) and ({\em iii\/}) by 
-% attempting to re-interpret
-re-interpreting
-the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
-rather than a mutually recursive set of equations on sets of access paths.
-
-This requires two ideas.  Firstly we have to control the use of functions---they
-tend to be infinitary and do not occur naturally in CFGs; in particular our
-$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
-to a tuple of access paths.  The former is achieved by using a separate
-meta-variable (later non-terminal)
-$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
-($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
-Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
-expressible finitely (it is a linear form).
-
-Secondly, there are also two technical issues in re-interpreting
-Figure~\ref{fig:live-judge} as a grammar.
-One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
-disappears after variables are separated ($\Lanv{i}{x}$ above). The other
-is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
-not a CFG\@.
-Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.
-
-Finally,  Section~\ref{sec:NFA-approx}
-uses a construction due to
-Mohri and Nederhof~\cite{mohri00regular} to over-approximate
-context-free grammars with regular grammars;
-these are more appropriate for run-time use.
-%
-Hence the overall `big picture' view is that each GC point
-is annotated with a table encoding the DFA for that program point.
-Such program points forming the call string causing the garbage collection
-(and ending in $\CONS$)
-are used by the garbage collector to curtail its reachability-marking phase
-to this over-approximated liveness.
-
-\subsubsection{Modifying the \CONS\ rule.}
-
-The    $\mathit{ref}$    rule for \CONS,    shown    in
-Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
-\acar\  and  \acdr\  from  access paths in  $\sigma$.
-  Mathematically  this  is fine  but causes problems  when
-  solving  the liveness equations  {\em symbolically}  since such
-  decomposition cannot be expressed as a context-free grammar.  
-To  handle this,  we introduce  two  new symbols  \bcar\  and
-\bcdr\ with the properties:
-\[
- \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
-\hspace{3em} \mbox{and} \hspace{3em}
- \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
-\]
-We can now rewrite the \CONS\ rule as:
-\begin{eqnarray*}
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
-\end{eqnarray*}
-
-We  call the liveness equations with  this modification
-  \Lfunonly$'$.
-The  defintions of \bcar\  and  \bcdr\  induce   the  following
-relation $\hookrightarrow$ over sets of access paths:
-\begin{eqnarray*}
-  \sigma_1\bcar\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
-& \qquad\mbox{and}\qquad &
-\sigma_1\bcdr\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
-\end{eqnarray*}
-%% \noindent \begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcar\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acar\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acdr\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% %\label{eqn:eup-red1}
-%% \end{eqnarray*}
-%% \end{minipage}
-%% \hspace*{.65cm}\begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcdr\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acdr\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acar\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% \end{eqnarray*}
-%% \end{minipage}\\
-
-The reflexive transitive closure of $\hookrightarrow$ will be denoted
-as $\stackrel{*}{\hookrightarrow}$. The following proposition relates \Lfunonly\   and \Lfunonly$'$:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]  
-\begin{boxedminipage}{\textwidth}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\append~\lista~\listb)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \xtest\ $\leftarrow $\ (\NULLQ~\lista) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\SIF\ \xtest~$\pi_3\!\!:$(\RETURN\ \listb)  
-          \UNL{4} \hspace*{.05cm}     $\pi_4\!\!:\, $(\LET\ \xtl\  $\leftarrow$\   (\CDR\ \lista) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_5\!\!:\, $(\LET\ \xrec\  $\leftarrow$  (\append\ \ \xtl\ \  \listb) \IN
-	  \UNL{6} \hspace*{.05cm}    $\pi_6\!\!:\, $(\LET\ \xhd\  $\leftarrow$  (\CAR\ \lista)  \IN
-	  \UNL{7}  \hspace*{.05cm}   $\pi_7\!\!:\,
-          $(\LET\ \xans\  $\leftarrow$ (\CONS\ \ \xhd\ \ \xrec)  \IN\
-          $\pi_8\!\!:\, $(\RETURN\ \xans))))))))
-	\end{uprogram}}
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pz\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}              (\LET\ \py\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_9\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\append\ \py\ \pz) \IN
-          \UNL{4}   \hspace*{.78cm}  $\pi_{10}\!\!:\, $  (\LET\ \pa\  $\leftarrow$  (\CDR\ \pw) \IN
-	  \UNL{5}   \hspace*{.78cm}  $\pi_{11}\!\!:\, $
-          (\LET\ \pb\  $\leftarrow$ (\CAR\  \pa) \IN~$\pi_{12}\!\!:\,$ (\RETURN\ \pb)))))))
-\end{uprogram}}
-\end{boxedminipage}
-  \caption{An example program.}
-\label{fig:examp-pgm}
-\figrule
-\normalsize
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{proposition}
-Assume that  a liveness computation  based on  \Lfunonly\   gives  the liveness of
-the variable $x$ at a program point $\pi_i$ as $\sigma$ (symbolically, $\Lanv{i}{x}=
-\sigma$).  Further,  let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is
-used instead of \Lfunonly.   Then  $\sigma' \stackrel{*}{\hookrightarrow}
-\sigma$.
-\end{proposition}
-
-To see  why the proposition is  true, consider an analysis  based on
-\Lfunonly$'$ in which   $\sigma$ appears in
-the  context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.  Let $\alpha  \in \sigma$.  The
-symbol  \bcar(\bcdr)  merely  marks  a  place in  $\alpha$  where  the
-original  \CONS\  rule  would  have erased  an  immediately  following
-\acar(\acdr),  or, in  absence of  such a  symbol, would  have dropped
-$\alpha$ itself.  Since the application  of any rule in \Lfunonly$'$ merely
-adds  symbols at  the beginning  of  $\alpha$, the  markers and  other
-symbols in $\alpha$ are propagated to other dependent parts of program
-in their same relative positions.  Consequently, the erasure carried out at the end of the analysis with
-$\stackrel{*}{\hookrightarrow}$ gives the same result as obtained
-through \Lfunonly.
-
-\subsection{Generating equations for the demand transformer \Lfonly}
-\label{sec:Lfonlysymbolically}
-
-%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
-We  shall  consider the  program  in  Figure~\ref{fig:examp-pgm} as  a
-running example;
-this is the  ANF version of the program in Figure~\ref{fig:mot-example}.
-
-%%AM no longer used: In what follows,  we shall overload the semantics of
-%% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-%mean the generation
-%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
-%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.
-
-To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
-$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
-{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
-$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
-formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
-of $x_i$.  For our example program which has  a
-single function \append, this generates the following equations:
-% $\Lfonly_{\append}$  is defined by equations:
-  \begin{eqnarray*}
-    \Lf{\append}{1}{\sigma}
-    &=& \epsilonset \cup \acar\bcar\sigma
-    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
-    \Lf{\append}{2}{\sigma}
-    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
-  \end{eqnarray*}
-  In  general, the  equations  for \Lfonly\  are recursive  since
-  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
-  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
-    is  similar to  solving the  differential equation  
-    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
-    that    the    solution     has    the    form    $\mathit{y}
-    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
-    equation  yields  a  quadratic  equation  in  $r$,  and  each
-    solution of $r$ gives rise  to a solution of the differential
-    equation  (in our  setup we  can effectively  pick  the least
-    solution rather than needing linear combinations).  }
-\begin{eqnarray}
-\label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
-\end{eqnarray}
-where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
-alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
-why \Lfonly\ has this form  is as follows: Recall that
-\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
-the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
-situation in which  the demand $\sigma$ does not  reach the $i$th
-argument.   This happens,  for  instance, when  argument is  used
-only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
-transformation of the demand $\sigma$ before it reaches the $i$th
-argument. This  transformation prefixes strings in 
-$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.
-
-Substituting  the  guessed   form  in  the  equations  describing
-\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
-containing $\sigma$ and those  without $\sigma$ on both sides of the
-resulting equation, we
-get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
-% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
-$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
-solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
-  \append}$, we get:
-  \begin{align*}
-    \Uf{\append}{1} &= \epsilonset
-    \cup \acdr.\Uf{\append}{1}  
-    &\quad 
-    \Uf{\append}{2} &= \Uf{\append}{2} \\
-    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
-    \cup \acdr\Df{\append}{1}\bcdr &\quad 
-    \Df{\append}{2} &= \epsilonset \cup
-    \Df{\append}{2}\bcdr 
-  \end{align*}
-Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.
-
-\subsection{Generating liveness equations \Lv\  for function bodies}
-\label{sec:bodylivenessbodies}
-
-\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
-
-We now calculate a 0-CFA-style summary liveness for each GC point of a program.
-There are two parts to this.
-First, for each function $f$,  we determine a {\em summary demand} $\sigma_f$
-over-approximating any demand $\sigma$ passed to $f$.
-%$\Lf{f}{i}{\sigma}$.
-So, consider a function $g$ containing a call site $\pi$ to $f$, say
-$\pi\!\!:\!(\LET~x \leftarrow (f\,y_1\,\ldots\,y_n)~\IN~e)$.
-Under the assumption that the demand on $g$ is $\sigma_g$, the liveness at $e$
-is $\Lv = \mathcal{L}(e,\sigma_g,\Lfonly)$, and the $\LET$ case of
-Figure~\ref{fig:live-judge} tells us this call site contributes $\Lv(x)$ to the
-demand $\sigma_f$ placed on  $f$.
-Write $\deltacall{f}{\pi}{g}$ for
-this demand (contributed by call site $\pi$ in $g$).
-
-Thus, supposing the $k$ call sites to function $f$ are
-$\pi^1$ (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
-%%% \marginpar{and writing $\sigma_f^i$ for the above $\deltacall{f}{\pi^i}{g^i}$?}
-then the over-approximation requirement on $\sigma_f$
-is achieved by taking $\sigma_f = \deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
-% is achieved by taking $\sigma_f = \sigma_f^1 \cup \cdots \cup \sigma_f^k$.
-
-The expression $e_{\mainpgm}$ is a special case; we assume it may be called externally
-with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$ (denoted
-$\sigma_{\!all}$).
-This is because any  part  of its  value  may be  used  by  the environment---for
-printing  the  result,  for  instance.
-
-% In what follows,  we shall overload the semantics of
-% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-% 
-% 
-% We assume an overall demand of $\{\acar,\acdr\}^*$ (denoted
-% $\sigma_{\!all}$) on the main expression $e_\mainpgm$.  This is because 
-% any  part  of its  value  may be  used  by  the environment---for
-% printing  the  result,  for  instance.
-
-%% To  generate  equations for  the  body $e_{\mathit{f}}$ of  a
-%% function,  we evaluate \Lfun{e_{\mathit{f}}}{\sigma_{\mathit{f}}}{\Lfonly}.     For ${\sigma_{\mathit{f}}}$,  we have
-%% to consider all calling  contexts during execution.  We therefore
-%% do a 0-CFA approximation by  defining
-%% % $\sigma_\mathit{f}$        as $\bigcup_\calls{f}{k}\!\!\sigma_\calls{f}{k}$,
-%% $\sigma_\mathit{f}$        as $\sigma_\calls{f}{1} \cup \cdots \cup \sigma_\calls{f}{n}$,
-%% where $\calls{f}{k}$,
-%% $\mathit{k} = 1 \ldots n$ represent all the calls to $\mathit{f}$
-%% in  the  program,   and  $\sigma_\calls{f}{k}$  are  the  demands
-%% generated at the calls  by the \rn{let-live} rule while analyzing
-%% the bodies of the caller.
-
-For the running example, $\append$ has calls from $\mainpgm$ at $\pi_9$
-and a recursive call at $\pi_5$.
-So $\sigma_{\append} =
-     \deltacall{\append}{\pi_9}{\mainpgm}  \cup \deltacall{\append}{\pi_5}{\append}$.
-Filling in  the values gives:
-%  where the former is the demand generated at the
-%  call to \append\  in $e_\mainpgm$ the latter
-%  is  the demand  at the  recursive call. 
-%% Propagation of
-%%  $\sigma_{\!all}$ in $e_\mainpgm$ gives    $\sigma_\calls{\append}{1}$   as   $\{\epsilon,   \acdr\}   \cup
-%%  \acdr\acar\sigma_{\!all}$  and propagation of
-%%  $\sigma_{\mathit{\append}}$ in $e_\append$ gives   $\sigma_\calls{\append}{2}$   as
-%%  $\bcdr\sigma_{\mathit{\append}}$.  
-%% Filling in  the values of $\sigma_\calls{\append}{1}$ and
-%% $\sigma_\calls{\append}{2}$, we get:
-\begin{align*}
-\sigma_{\append}    &=
-\left( \{\epsilon,  \acdr\} \cup  \acdr\acar\sigma_{\!all}  \right) ~\cup~
-\bcdr\sigma_{\mathit{\append}}
-\end{align*}
-
-Second, for each function $f$ (possibly $\mainpgm$) we need the liveness at
-each contained GC point
-$\pi\!\!:\!(\LET~x \leftarrow s~\IN~e)$,
-where $s$ is a call either to $\CONS$ or to a user function $g$.
-Given $\sigma_f$ calculated above, this
-is simply $\mathcal{L}(\pi,\sigma_f,\Lfonly)$.
-%%% \marginpar{AM: we've always said this (safe), but actual parameters
-%%% live before the call and dead after needn't be included!}
-
-For the running example, containing GC points $\pi_5, \pi_7$ in $\append$ and
-$\pi_9$ in $e_\mainpgm$, this gives
-(recall Equation~(\ref{eq:LF:DI}) above states $\Lf{f}{i}{\sigma} = \Uf{f}{i} \cup \Df{f}{i}.\sigma$):
-\begin{align*}
-\Lanv{5}{\xtl} &= \Lf{\append}{1}{\bcdr\sigma_{\append}}  \quad &
-                                \Lanv{7}{\xhd}\,\,\, &=  \bcar\sigma_{\append}     \\ 
-\Lanv{5}{\listb} &= \Lf{\append}{2}{\bcdr\sigma_{\append}} &
-                                \Lanv{7}{\xrec} &=  \bcdr\sigma_{\append} \\
-\Lanv{5}{\lista} &= \epsilonset \cup \acar\bcar\sigma_{\append} 
-\\[1ex]
-  \Lanv{9}{\py}  &= \Lf{\append}{1}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}} \\
-  \Lanv{9}{\pz}  &= \Lf{\append}{2}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}}
-\end{align*}
-
-\noindent 
-In summary, the equations generated during liveness analysis are:
-\begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
-\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
-% {\color{red}
-%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
-%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
-% }
-\end{enumerate}
-
-\subsection{Solving liveness equations---the grammar interpretation}
-The  equations above can now be re-interpreted as a
-context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
-a variable  $X$ occurring on  the LHS of the  equations generated
-from the analysis.  We can  think of the resulting productions as
-% being a grammar with multiple start symbols---one for
-% for each nonterminal \var{\Lanv{i}{x}}.
-being associated with several  grammars, one for each non-terminal
-\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
-productions:
- \begin{eqnarray*}
-\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
-  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
- \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
-  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
-  \acdr\var{\Uf{\append}{1}}\\         
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar               \mid
-  \acdr\var{\Df{\append}{1}}\bcdr\\
-\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
-\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
-\var{$\sigma_{\mathit{\!all}}$}
-\end{eqnarray*}
-Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
-denoted $\mathscr{L}(\var{\Lanv{i}{x}})
-$,  is the desired  solution of
-\Lanv{i}{x}.   
-However, recall from  our earlier  discussion that
-the  decision problem that  we are  interested in  during garbage
-collection is:
-\begin{quote}
-Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
-\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
-\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
-of forward paths only. Then does $\alpha \in \sigma$?
-\end{quote}
-We could convert  the  rules  defining  $\hookrightarrow$  into
-productions and  add them to the grammar. However, this results
-in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is  undecidable.   We
-circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
-  strongly  regular}  CFGs which  have   easy
-translations to non-deterministic finite state automata (NFA). The
-NFAs are then simplified on the
-basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
-membership of forward access paths. The resulting NFAs are   finally
-converted to DFAs for use during garbage collection.
-%------------------------------------------------------------%
-\subsubsection{Approximating CFGs using NFAs.}
-\label{sec:NFA-approx}
-
-We use the   
-algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
-approximate a CFG  to a   {\em strongly regular\/} grammar.  The
-transformation has  the  property that  if $\Lv$  is a
-non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
-Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
-  L}_{G'}(\Lv)$.    This  is   required  for the
-approximation to be safe with respect to liveness.
-
-We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
-transformation. We  pick the only production that  is affected by
-the transformation---the production for \Df{\append}{1}.  The
-production for \Uf{\append}{1}, while  recursive, is already in
-strongly  regular  form  and   is  therefore  unaffected  by  the
-transformation.
- \begin{eqnarray*}
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
-  \acdr\var{\Df{\append}{1}} \\ 
-\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
-\mid \epsilon 
-\end{eqnarray*}
-The  languages
-generated for  \var{\Df{\append}{1}} in  the original grammar  and the
-new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
-\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
-
-%------------------------------------------------------------%
-\subsubsection{Transforming NFAs to Accept Forward Paths:}
-%\label{sec:NFA-elim}
-The strongly regular CFGs obtained after the Mohri-Nederhof
-transformation are
-  first converted  into  NFAs.  The        algorithm         described        in
-Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
-\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
-Thus  \nt\  can  be  used  to  check  
-membership of forward paths.
-
-\begin{figure}[t]
-\begin{boxedminipage}{\textwidth}
-\begin{center}
-  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
-  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
-%
-  \raggedright{\bf Output:} NFA  \nt\ with  underlying
-  alphabet $\lbrace \acar, \acdr \rbrace$ such that
-  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
-  \\
-%
-
-  \raggedright{\bf Steps:}
- \begin{algorithmic}
-   \STATE $i \leftarrow 0$ 
-   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-   without $\epsilon$-moves \cite{hopcraft90toc}
-
-   \REPEAT
-   \STATE $\nt'_{i+1} \leftarrow \nt_i$
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcar$ and
-     outgoing edge to $q''$ with label $\acar$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-   label $\epsilon$.
-   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-   \ENDFOR
-
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcdr$ and outgoing edge to
-     $q''$ with label $\acdr$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-     label $\epsilon$.
-   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-   \ENDFOR
-
-   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-   $\nt'_{i+1}$ without $\epsilon$-moves
-
-   \STATE $i \leftarrow i+1$
-   \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-   \STATE $\nt \leftarrow \nt_i$
-%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
- \end{algorithmic}
-\end{center}
-\end{boxedminipage}
- \caption{Algorithm for transforming an NFA to  accept forward
-   paths only.}\label{algo:simplify-NFA} \figrule
-\end{figure}
-%------------------------------------------------------------%
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
-pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
-process is  continued till  a fixed point  is reached.   When the
-fixed point  is reached, the resulting NFA  contains all possible
-reductions corresponding  to all the  paths in the  original NFA.
-The proofs  of the termination  and correctness of  the algorithm
-are given in our earlier paper~\cite{karkare07liveness}.
-
-\input{liveness-dfa.tex}
-
-We illustrate the algorithm in Figure~\ref{algo:simplify-NFA} by
-constructing the automaton for \var{\Lanv{9}{y}}.
-Figure~\ref{fig:automata-construction} (c) shows the automaton for
-\var{\Lanv{9}{y}} constructed by composing the automata for
-\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
-\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
-get (d). We add an $\epsilon$ edge from $q_2$ to $q_3$ bypassing the
-\bcdr\acdr\ pair from $q_2$ to $q_2$ and then to $q_3$. This is shown
-in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
-is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
-$q_2$.  Removing this $\epsilon$ edge gives the automaton shown in
-(g).  Restricting this automaton to forward edges only, we get the
-final automaton shown in (h). This automaton recognizes $\acdr^{*}\!\mid\acdr^{*}\acar \sigma_{\mathit \!all}$, showing that the entire
-list \py, including its elements, is live at $\pi_9$. Also note that
-the language accepted by the final automaton satisfies the
-prefix-closed property.
-
-\section{Prototype and evaluation}\label{sec:experiments}
-To demonstrate the effectiveness of liveness-based garbage collection,
-we  have  built a  prototype  consisting  of  an interpreter  for  our
-language,  a  liveness  analyzer  and  a copying  collector  that  can
-optionally use the results of liveness analysis for marking instead of
-reachability. When the collector uses liveness for marking, we call it
-a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-  reachability-based  collector}  (RGC).   The  copying  collector  is
-neither incremental nor generational.  As a consequence, any cell that
-becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.
-
-While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
-is invoked  at the program point   $\pi$,  it would  use  the
-liveness  DFAs at  $\pi$ itself for exploring the root set of the
-current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.
-
-Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-$\dfa{\pi}{x}$.    
-Considering   a   DFA   as   a   table,
-$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-\acar\      or     \acdr.       We      shall     also      write
-$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-indicating  whether there  is a  transition from  $q$ on
-$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-pointer to $x$.
-The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-$x$  in the  other semi-space  and returns  the new  address.  It
-hides  details such as  returning the  forwarding pointer  if the
-value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.
-
-The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-in the heap over time for RGC and LGC---here time is measured in terms
-of the number of cons cells allocated. In addition, they also show the
-number of  reachable cells and the  number of cells  that are actually
-live\footnote{This   is  statically   approximated  by   our  liveness
-  analysis}.  Since  the programs have  different memory requirements,
-we have tuned the size of heap for each program to ensure a reasonable
-number of  collections. An invocation  of RGC decreases the  number of
-cells  in heap  till  it touches  the  curve of  reachable cells.   An
-invocation of LGC decreases the number  of heap cells to no lower than
-the curve of live cells.
-
-To construct the  reachable and live curves, we  record for every cell
-its creation time (\CreateTime), its last use time (\UseTime), and the
-earliest time  when the  cell becomes unreachable  and can  be garbage
-collected (\CollTime).  For accurate recording of  \CollTime, we force
-frequent invocations  of a reachability based collector  in a separate
-run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
-  fibheap} increases  initially, comes  down before the  heap becomes gets
-full and then  continues to remains low.  This results in  a single garbage
-collection in both RGC and LGC.  The graphs also show the precision of
-our liveness analysis.  For  all programs  except  {\tt nperms} and
-{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.
-
-\subsection{Results}
-\begin{figure}[t] 
-\centering
-\begin{boxedminipage}{.75\textwidth}
-{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
-\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-  then} {\bf skip}\\ 
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
-\noindent \hspace*{.35cm}{\bf return} $y$
-\end{boxedminipage}
-\caption{Function       for       copying       a       root       set
-  variable.}\label{algo:LGC-main-loop} \figrule
-
-\end{figure}
-
-\begin{figure}[p]
-\begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
- &
-{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-\end{tabular}
- \caption{Memory  usage of  programs. The blue and the red curves
-  indicate the number of cons cells in the active semi-space for RGC and
-  LGC respectively. The black curve represents the number of reachable
-  cells and the grey curve represents the number of cells that are
-  actually live (of which liveness analysis does  a static
-  approximation). x-axis is the time measured in number of cons-cells
-  allocated. y-axis is the number of cons-cells.}
-\label{fig:memory-usage} \figrule
-\end{figure}
-The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-provides  statistics  regarding the  analysis  itself.  The number  of
-states and the analysis times  are within tolerable limits. Precision
-of analysis refers to the percentage of dead cells that is collected by
-LGC, averaged over all invocations.
-The second table shows garbage  collection statistics for RGC and LGC.
-LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to run
-in comparison with  RGC.
-
-There are a couple of  issues of concern.  The garbage collection time
-is larger in the case of LGC for some programs. The reason is that the
-cost of consulting the liveness DFA may outweigh the combined benefits
-of  fewer   garbage  collections   and  fewer  markings   per  garbage
-collection.   The  other issue  is  illustrated  by  the program  {\tt
-  lambda}    (memory   usage  shown   separately    in
-Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-in Figure~\ref{fig:experimental-results}, the  number of
-touched cells\footnote{These  are the  cells which are  visited during
-  marking phase, often  more than once due to  sharing.} in this
-example is much higher for
-LGC.  This  increase is due to  excessive sharing among  heap nodes in
-this program.  Note  that a node re-visited because  of sharing is not
-explored  any   further  during  a  RGC   collection.   However,  this
-curtailment cannot happen in  LGC because of the possibility that the
-node,  re-visited in a  different liveness  state, may  mark a  set of
-cells different from the earlier visit.
-\begin{figure}[t]
-\footnotesize
-
-\centering
-
-\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline
-{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-  nperm} &  {\sf fibheap} & {\sf knightstour} &
-{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-\hline
-\hline
-{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-\hline
-\end{tabular}}
-
-\centerline{(a)}
-
-\medskip
-
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
-  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-
-                            &   \multicolumn{2}{c|}{cells per GC}  
-                            &   \multicolumn{2}{c|}{cells per GC} 
-                            &   \multicolumn{2}{c|}{\#GCs} 
-                            &   \multicolumn{2}{c|}{MinHeap} 
-                            &   \multicolumn{2}{c|}{Avg. Drag} 
-                            &   \multicolumn{2}{c|}{(sec)} \\
-\cline{2-13}
-{Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
-%    {\sf  nboyer} & &  & &   & &  & & &  \\
-%    {\sf   circsim?} & &  & &   & & & & &   \\
- \hline
-\end{tabular}}
-
-\centerline{(b)}
-\vskip -2mm
-\caption{Experimental results comparing RGC and LGC. Table (a) gives
-  data related to liveness analysis, and (b) gives garbage collection
-  data. The average drag time is measured in allocated cons cells.}
-\label{fig:experimental-results}
-\normalsize
-\end{figure}
-
-\section{Collecting more garbage can never slow things down}
-\label{sec:lgc-always-better}
-Since garbage collection is  effectively asynchronous to the allocator
-thread, one might  worry as to how {\em  robust} our measurements are.
-For example,  while LGC would,  in general, collect more  garbage than
-RGC in the same heap state,  is it possible that on some programs, LGC
-might do  a larger number  of collections? We  prove a result  to show
-that   this  cannot   happen.    This  result   applies  to   classical
-mark-and-sweep and copying garbage  collectors and, we hypothesize, to
-generational collectors too.
-
-\begin{lemma}
-For the  same mutator,  a liveness-based collector  can never  do more
-garbage collections than a reachability-based collector.
-\end{lemma}
-
-\begin{proof}Assume, as before,  that time is measured in terms of the number of
-  cons cells allocated. Let us run  two copies of the mutator with the
-  two    different   collectors    but    with   synchronous    memory
-  allocations.   The   garbage    collections,   however,   happen
-  asynchronously.
-
-  To prove the lemma, it is enough to show the truth of the following 
-statement: After
- every  LGC invocation,  the count of LGC invocations   is  no greater
-than  RGC invocations.   The  base case  holds  since the  first
-invocations of both GCs happen at the same time.  Assume the statement
-to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-reachable cells, its heap would contain no more cells than RGC heap at
-the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-LGC, or LGC and RGC are both invoked next at the same time.  In either
-case, the statement holds after $n+1$ invocations of LGC.
-\end{proof}
-
-%% \begin{figure}[h]
-%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
-%% \caption{Memory  usage of  programs. The  solid  lines represents
-%%   number of  live cells and the  dashed lines show  the number of
-%%   reachable cells.  The red and blue  lines represent invocations
-%%   of RGC and  LGC---their heights indicate  the amount of memory
-%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
-%%       paper and are being used as placeholder}}}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-%%\clearpage
-
-\section{Related Work} Previous attempts to  increase  the space efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
-reallocation~\cite{jones89compile},                       deforestation
-techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
-methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
-
-\begin{figure}[t]
-\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
- \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
-\label{fig:memory-usage-lambda} \figrule
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-We have defined a notion of liveness on structured data;
-this generalizes classical liveness and strong liveness.
-We started with a general fully context-sensitive analysis which
-we proved correct with respect to a minefield semantics
-(this models the effect of garbage collection between every
-evaluation step).
-
-To avoid scalability issues (and to avoid performing part of
-the liveness computation at run time)
-we defined an 0-CFA version of this liveness analysis in which
-demands for function $f$ at all calling contexts are conflated into 
-a single demand $\sigma_f$.
-This enabled us to treat the liveness equations symbolically obtaining
-context-free grammars for liveness at each GC point (calls to user
-functions and to $\CONS$).
-These were then converted to DFAs for run-time consultation by the
-garbage collector. Experiments confirm the precision of the analysis.
-
-To obtain performance figures we compared a reachability-based garbage
-collector with a liveness-based  collector.  This showed a decrease in
-the  number  of  GCs,   more  garbage  collected  per  invocation.   A
-significant benefit of LGC is  that programs can run in smaller memory
-when compared to  RGC. This is potentially useful  in situations where
-memory  is  limited---as with  embedded  systems.  For  a majority  of
-programs, the garbage collection times were reduced.
-
-One issue we  highlighted was that while fewer  nodes were marked (and
-more garbage collected), sometimes  cons cells could be visited and
-traversed  multiple times  with different  sets of  liveness  paths to
-explore.  Further work includes  improvements to the classical copying
-collector to reduce the cost of this.
-
-\subsection*{Acknowledgements}
-The authors  would like  to thank Hemanshu  Vadehra for  a preliminary
-implementation of the prototype system. Amey Karkare was supported for
-this work by the DST/SERC fast track scheme for young scientists.
-
-\bibliography{fun_hra}{}
-\bibliographystyle{splncs}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-\section{Appendix stuff}
-
-\paragraph{Safety of using the modified \CONS\  rule:}
-Recall that while the correctness of the analysis was proved with
-respect to  the $\len$ rules  in Figure~\ref{fig:live-judge}, the
-equations   for  liveness   are  based   on  a   modification  of
-$\mathit{ref}$    in    the    case    of   \CONS\  as  given    by
-(\ref{eqn:mod-cons}).  We now show that  if the \bcar\,s and \bcdr\,s
-in live access  paths resulting  from this  modification are
-eliminated through  applications of $\hookrightarrow$,  we obtain
-the same results as given by the original rules.
-
-\comment{Proposal: We prove a result which goes as follows:
-
-\begin{quote}
-Lemma: Let  $\Lanv{i}{x}= \sigma$ for an a  program point $\pi_i$
-and a variable  $x$, when the judgement is  based on the original
-$\len$  rules.   Further, let  $\Lanv{i}{x}=  \sigma'$, when  the
-judgement is based on $\len$ with the modified \CONS\ rule.  Then
-$\sigma' \stackrel{*}{\hookrightarrow} \sigma$
-\end{quote}
-
-Here are some  ideas regarding how this can  be proved.  Consider
-an arbitrary equation
-$$ \Lanv{j}{y} = \ldots \Lanv{i}{x} \ldots$$ generated by the
-analysis.
-
-Now assume that $\Lanv{i}{x}= \sigma$.   We want to find out what
-happens to the access patterns in $\sigma$ in the access patterns
-of \Lanv{j}{y}.   To answer this,  we put \# markers  around each
-$\alpha \in \sigma$  and redo the equation using  an extension of
-$\len$ to take  care of \#-tagged patterns.  We  can then observe
-occurrences of \#-tagged patterns  in the liveness information at
-$\pi_j$.
-
-Here is  the $\len$ extension for  the original \CONS\  clause to take
-care of the \#-tagged patterns:
-$$\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          = \{x.\#\alpha\# \mid \#\acar\alpha\# \in \sigma\} \cup
-\{y.\#\alpha\# \mid \#\acdr\alpha\# \in \sigma\} $$
-And here  is  the $\len$ extension for  the modified \CONS\  clause.
-$$ \mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-= \{x.\#\bcar\alpha\# \mid \#\alpha\# \in \sigma\} \cup
-\{y.\#\bcdr\alpha\# \mid \#\alpha\# \in \sigma\} $$
-All  other rules  remain unchanged.  Tracking a  pattern  through this
-extension is sound because erasing the \#-tags would give us the same
-liveness as the original rules.
-
-Let us also extend $\hookrightarrow$ to \#-tagged strings as:
-$$\sigma_1  {\hookrightarrow} \sigma_2\,\, \mathit{iff}\,\, \sigma_2 =
-\{\alpha_1 \#\alpha\# \alpha_2 \mid  \alpha_1 \#\alpha'\# \alpha2 \in
-  \sigma_1\,\, and \,\, \{\alpha'\} \stackrel{*}{\hookrightarrow} \{\alpha\}\}$$
-
-Now we have an  observation represented by the following diagram:
-
-\centerline{\epsfig{file=diag1.eps, height=4cm}}
-
-We need  to make this  observation general enough.   Therefore we
-should prove  the observation even if \Lanv{i}{x}  = $\sigma$ had
-patterns with \bcar\ and \bcdr.   We should be able to prove this
-by structural induction on the rules.
-
-How does  the observation prove  the original lemma?   Suppose we
-replayed  judgements  with  ($\len$  +  \#) and  with  ($\len$  +
-modified \CONS +  \#).  Suppose that $\pi_i$ is  the last program
-point till which the results  are the same (because no \CONS\ was
-encountered).  Now  consider a  $\sigma \in \Lanv{i}{x}$  for any
-$x$.  Put \#-tags around each string in $\sigma$. Now we have the
-diagram:
-
-\centerline{\epsfig{file=diag2.eps, height=7cm}}
-
-This shows that modulo erasure of \#-tags, using the original $\len$
-judgement and  $\len$ judgement with modified \CONS\ rule followed by
-$\hookrightarrow$ give the same liveness.
-
-And I would like  to conclude that it is all right  to do our analysis
-with $\len$ + modified \CONS\ and then simplify using $\hookrightarrow$.
-
-}
-
-{\color{blue}\marginpar{From sec 3.1, liveness judgements}
-\paragraph{Semantic considerations.}
-The incremental liveness for generic primitive $\PRIM$ is given by
-$\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly) = 
-           \{x.\epsilon,y.\epsilon\}$.
-Earlier we suggested that classical strong liveness for addition,
-but not division, could use the stronger
-$ \{x.\epsilon,y.\epsilon \mid \epsilon \in \sigma\}$.
-However this is invalid: garbage collection
-could reuse dead, but reachable, $\CONS$ cell fields (of $v$ below)
-thereby producing a stuck state in a dynamically typed language.
-Hence we treat $+$ as like division not like addition---suppose
-% $(FOO)$ is a user function which may cause garbage collection, and consider:
-calling $\mbox{\footnotesize (FOO)}$ garbage collects and reallocates the cell pointed
-to by `v' in:
-
-\vspace{-15pt}
-
-\begin{quote}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1}  (\LET\ v\ $\leftarrow $\ (\CONS~2~3) \IN
-	  \UNL{2}  (\LET\ w\ $\leftarrow $\ (FOO) \IN
-	  \UNL{3}  (\LET\ x\ $\leftarrow $\ (\PRIM\ (\CAR\ v) 4) \IN
-	  \UNL{4}  (\RETURN\ w))))
-	\end{uprogram}}
-\end{quote}
-In principle, {\em dead code removal} of
-$(\PRIM\ (\CAR\ {\rm v})\ 4)$
-would render the stronger version valid, but still requires static typing
-to avoid the risk of {\em removing} a stuck state.
-We restrict ourselves to liveness annotation rather than code transformation.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]
-\footnotesize
-Old version of the liveness rules...
-
-\infrule[ret-live]
-        {}
-        {(\RETURN~x),\sigma,\Lfonly \len  x.\sigma}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[if-live]
-        { e_1, \sigma,\Lfonly \len \Lv_1
-          \andalso e_2, \sigma,\Lfonly \len  \Lv_2}
-        {(\SIF~x~e_1~e_2),\sigma,\Lfonly \len
-               \Lv_1\cup\Lv_2\cup \{x.\epsilon\}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[let-live]
-        {e, \sigma,\Lfonly  \len  \Lv}
-        {(\LET~x \leftarrow  s~\IN~e),
-          \sigma,\Lfonly \len   \Lv\setminus \{x.*\} \cup
-          \mathit{ref\/}(s,\Lv(x),\Lfonly)}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\infrule[live-define]
-        {e_f, \sigma, \Lfonly  \len
-%            \bigcup \{z_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-           \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \fen \Lfonly
-\\ \mbox{where }
-     (\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ \ e_f) \mbox{ is a member of }
-        d_1 \ldots d_k}
-\figrule
-  \caption{Liveness Judgement Rules}\label{fig:live-judge-old}
-\normalsize
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \newgray{lightgray}{.75}
-%\newcommand{\dem}[1]{\red{\scriptsize{$#1$}}}
-\newcommand{\dem}[1]{\psframebox[linestyle=none,fillstyle=none,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-%\newcommand{\liv}[1]{\blue{\scriptsize{$#1$}}}
-\newcommand{\liv}[1]{\psframebox[linestyle=none,fillstyle=solid,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-\psset{unit=1mm,nodesep=1mm,labelsep=0.5mm,arrows=->,linewidth=.1,framesep=0.2mm}
-\begin{center}
-  \begin{pspicture}(0,0)(120,50)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(12,5)
- \footnotesize
-   \putnode{let1}{origin}{15}{50}{$\pi_1$:\LET}
-
-    \putnode{test1}{let1}{-12}{-7}{\xtest}
-    \putnode{null}{let1}{0}{-7}{$\pi_2$:\NULLQ}
-    \putnode{lista1}{null}{0}{-7}{$\pi_3$:\lista}
-    \putnode{if}{let1}{18}{-7}{$\pi_4$:\SIF}
-
-    \putnode{test2}{if}{-8}{-7}{$\pi_5$:\xtest}
-    \putnode{listb1}{if}{3}{-7}{$\pi_6$:\listb}
-    \putnode{let2}{if}{14}{-7}{$\pi_7$:\LET}
-
-    \putnode{tl1}{let2}{-22}{-7}{\xtl}
-    \putnode{cdr}{let2}{-5}{-7}{$\pi_8$:\CDR}
-    \putnode{lista2}{cdr}{0}{-7}{$\pi_9$:\lista}
-    \putnode{let3}{let2}{15}{-7}{$\pi_{10}$:\LET}
-
-\putnode{rec1}{let3}{-13}{-7}{\xrec}
-\putnode{append}{let3}{0}{-7}{$\pi_{11}$:\append}
-\putnode{tl2}{append}{-10}{-7}{$\pi_{12}$:\xtl}
-\putnode{listb2}{append}{5}{-7}{$\pi_{13}$:\listb}
-
-\putnode{let4}{let3}{17}{-7}{$\pi_{14}$:\LET}
-\putnode{hd1}{let4}{-5}{-7}{\xhd}
-\putnode{car}{let4}{4}{-7}{$\pi_{15}$:\CAR}
-\putnode{lista3}{car}{0}{-7}{$\pi_{16}$:\lista}
-
-\putnode{let5}{let4}{16}{-7}{$\pi_{17}$:\LET}
-
-\putnode{ans1}{let5}{-4}{-7}{\xans}
-\putnode{cons}{let5}{6}{-7}{$\pi_{18}$:\CONS}
-\putnode{hd2}{cons}{-10}{-7}{$\pi_{19}$:\xhd}
-\putnode{rec2}{cons}{10}{-7}{$\pi_{20}$:\xrec}
-\putnode{ans2}{let5}{19}{-7}{$\pi_{21}$:\xans}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-
-%% \putnode{x}{let1}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{if}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{let3}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let4}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{-8}{0}{\liv{\Lv_1}}
-%% %\putnode{x}{ans2}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{plus}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{rec2}{1}{4}{\dem{\epsilonset}}
-%% \putnode{x}{plus}{-7}{0}{\liv{\Lv_6}}
-%% %\putnode{x}{let4}{-7}{0}{\liv{\{\xrec\}}}
-%% \putnode{x}{len}{3}{3}{\dem{\epsilonset}}
-%% \putnode{x}{len}{-7}{0}{\liv{\Lv_3}}
-%% \putnode{x}{tl2}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{tl2}{-6}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let3}{-10}{0}{\liv{\xtl.\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l2}{0}{4}{\dem{\epsilonset \cup \acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{l2}{-5}{0}{\liv{\Lv_4}}
-%% \putnode{x}{zero}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{zero}{-3}{0}{\liv{\Lv_0}}
-%% \putnode{x}{test2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{test2}{-7}{0}{\liv{\Lv_5}}
-%% \putnode{x}{if}{-6}{0}{\liv{\Lv_3}}
-%% \putnode{x}{null}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{null}{-9}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l1}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{vone2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{one}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{let5}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{l1}{-5}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{let1}{-17}{-1}{\liv{\{\xl\} \cup \xl.\acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{let1}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let1}{-7}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{if}{7}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let2}{8}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let3}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let3}{-7}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let4}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let4}{-7}{0}{\liv{\Lv_2}}
-%% \putnode{x}{vone2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{rec2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{let5}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{one}{-6}{0}{\liv{\Lv_2}}
-\ncline{-}{let1}{test1}
-\ncline{-}{let1}{null}
-\ncline{-}{let1}{if}
-\ncline{-}{null}{lista1}
-
-\ncline{-}{if}{test2}
-\ncline{-}{if}{listb1}
-\ncline{-}{if}{let2}
-\ncline{-}{let2}{tl1}
-\ncline{-}{let2}{cdr}
-\ncline{-}{let2}{let3}
-\ncline{-}{cdr}{lista2}
-
-\ncline{-}{let3}{rec1}
-\ncline{-}{let3}{append}
-\ncline{-}{let3}{let4}
-
-\ncline{-}{append}{tl2}
-\ncline{-}{append}{listb2}
-
-\ncline{-}{let4}{let5}
-\ncline{-}{let4}{hd1}
-\ncline{-}{let4}{car}
-\ncline{-}{car}{lista3}
-%% \ncline{-}{len4}{tl2}
-%% \ncline{-}{let4}{vone1}
-%% \ncline{-}{let4}{one}
-%% \ncline{-}{let4}{let5}
-\ncline{-}{let5}{ans1}
-\ncline{-}{let5}{cons}
-\ncline{-}{cons}{hd2}
-\ncline{-}{cons}{rec2}
-%% \ncline{-}{let5}{plus}
-\ncline{-}{let5}{ans2}
-\end{pspicture}
- \end{center}
-
-  \caption{Example to illustrate liveness computation}\label{fig:live-examp}
-\figrule
-\normalsize
-\end{figure}
-
-\newpage
-
-\begin{figure}[t]
-  \begin{tabularx}{320pt}{|| c | X | X | X ||}
-\hline
-Pgm  Pt & Demand  & Pre & Post \\
-\hline 
-$\pi_1$ & $\sigma$ & \{\lista\} $\cup$ \listb.$\sigma$ $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & $\phi$ \\
-\hline
-%% $\pi_2$ & $\epsilonset$ & \{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_3$ & $\epsilonset$ &\{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_4$ & $\sigma$ &\{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-%% $\pi_5$ & \epsilonset & \{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  \\
-%% \hline
-%% $\pi_6$ & $\sigma$ & \listb.$\sigma$ & $\phi$ \\
-%% \hline
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_7$ & $\sigma$  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$\\
-%% \hline
-%% $\pi_8$ & \Lf{\append}{1}(\{\bcdr\}$\sigma$) & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_9$ & \epsilonset $\cup$ \{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_{10}$ &$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-$\pi_{11}$ & \{\bcdr\}$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \lista $\cup$ \lista.\{\acar\bcar\} \\
-\hline
-$\pi_{12}$ &\Lf{\append}{1}(\{\bcdr\}$\sigma$)   &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-$\pi_{13}$ & \Lf{\append}{2}(\{\bcdr\}$\sigma$)  &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{14}$ & $\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$ \\
-\hline
-%% $\pi_{15}$ & \{\bcar\}$\sigma$  & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   &  \xrec.\{\bcdr\}$\sigma$\\
-%% \hline
-%% $\pi_{16}$ & \epsilonset $\cup$ \{\acar\bcar\}$\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$  & \xrec.\{\bcdr\}$\sigma$ \\
-%% \hline
-$\pi_{17}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   & $\phi$  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{18}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$  \\
-\hline
-$\pi_{19}$ &  \{\bcar\}$\sigma$        & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{20}$ &   \{\bcdr\}$\sigma$       & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{21}$ & $\sigma$ & \xans.$\sigma$ & $\phi$  \\
-\hline
-\end{tabularx}
-  \caption{Liveness annotations for the body of \append}\label{fig:xe_append}
-\figrule
-\normalsize
-\end{figure}

### 2014-09-10 lazy-scheme/src/SchemeAST.h
Adding implementation

@@ -1,0 +1,552 @@
+#ifndef __SCHEME_AST_H__
+#define __SCHEME_AST_H__ 1
+
+#include "Output.h"
+#include <vector>
+#include<map>
+#include<iostream>
+#include <boost/variant.hpp>
+#include "Demands.h"
+#include <functional>
+#include <cstring>
+#include<utility>
+#include<set>
+using namespace std;
+
+namespace Scheme {
+
+namespace AST {
+
+class ExprNode;
+
+typedef int state_index;
+typedef std::set<state_index> stateset;
+
+enum cell_type
+{
+	nilExprClosure,
+	constIntExprClosure,
+	constBoolExprClosure,
+	constStringExprClosure,
+	funcApplicationExprClosure,
+	consExprClosure,
+	unaryprimopExprClosure,
+	binaryprimopExprClosure,
+	defaultClosure
+};
+typedef unsigned long clock_tick;
+
+struct cons
+{
+  cell_type typecell;
+  union
+  {
+	  struct
+	  {
+		 cons *car, *cdr;
+	  }cell;
+	  int intVal;
+	  bool boolval;
+	  std::string* stringVal;
+	  struct
+	  {
+		  ExprNode* expr;
+		  cons* arg1;
+		  cons* arg2;
+	  }closure;
+  }val;
+  void *forward;
+  stateset *setofStates;
+  int depth;
+  bool inWHNF;
+#ifdef GC_ENABLE_STATS
+    /*----------------------------------------------------------------------
+     * Following fields are added by Amey Karkare to
+     * generate gc related statistics
+     */
+    /* size is used only if we support vectors */
+    clock_tick created;       /* creation time of cell */
+    clock_tick first_use;     /* first use time of cell */
+    clock_tick last_use;      /* last use time of cell */
+    char       is_reachable:1,  /* true if cell is reachable during current gc */
+               is_used:1;       /* true if cell is dereferenced for use */
+    /*----------------------------------------------------------------------*/
+#endif
+#ifdef ENABLE_SHARING_STATS
+    int visited;
+#endif
+
+
+};
+
+enum exprType
+{
+	nilExpr,
+	constIntExpr,
+	constBoolExpr,
+	constStringExpr,
+	varExpr,
+	letExpr,
+	funcDefine,
+	funcApplicationExpr,
+	consExpr,
[…]

### 2014-09-10 lazy-scheme/src/SchemeAST_Defs.cpp

@@ -1,0 +1,1066 @@
+//#include "SchemeAST.h"
+#include <list>
+#include <sstream>
+#include <algorithm>
+#include <assert.h>
+#include <string>
+#include <cstring>
+#include <ctime>
+#include <unordered_map>
+#include <boost/variant.hpp>
+#include "Simulator.h"
+#include "gc.h"
+using namespace std;
+using namespace Scheme::AST;
+
+
+extern deque<actRec> actRecStack;
+extern GCStatus gc_status;
+vector<double> gc_times;
+int conscount;
+extern int gcinvoke;
+int maxcount;
+extern double gctime;
+ProgramNode* pgm;
+unsigned long lbl_count = 0;
+map< string, vector<ExprNode*>> func_call_points;
+
+std::string getNextLabel()
+{
+	std::stringstream sstream;
+	//sstream <<  "p#" << lbl_count++;
+	sstream <<  lbl_count++;
+	return sstream.str();
+}
+
+cons* reduceParamToWHNF(cons* heap_cell)
+{
+	if (heap_cell->inWHNF)
+		return heap_cell;
+	else
+	{
+		cons* retval = heap_cell;
+		while(!retval->inWHNF)
+			retval = retval->val.closure.expr->evaluate(heap_cell);
+		heap_cell->typecell = retval->typecell;
+		heap_cell->val = retval->val;
+		heap_cell->inWHNF = retval->inWHNF;
+		return heap_cell;
+	}
+}
+
+Node::~Node() {}
+
+std::string Node::getLabel() const
+{
+	return label;
+}
+
+std::string IdExprNode::getLabel() const
+{
+	return label;
+}
+
+std::string ReturnExprNode::getLabel() const
+{
+	return this->pID->getLabel();
+}
+
+Node::Node(const std::string name) : type(defaultType), label(""), node_name(name) {}
+
+
+ExprNode::~ExprNode() {}
+
+ExprNode::ExprNode(const std::string name) : Node(name) {}
+
+
+IdExprNode::~IdExprNode() {
+	delete pID;
+}
+
+std::string IdExprNode::getIDStr() const {
+	return *pID;
+}
+
+IdExprNode::IdExprNode(std::string * id) : ExprNode("ID"), pID(id) {}
+
+IdExprNode * IdExprNode::clone() const
+{
[…]

### 2014-09-10 lazy-scheme/src/Simulator.cpp

@@ -1,0 +1,276 @@
+//============================================================================
+// Name        : Simulator.cpp
+// Author      : 
+// Version     :
+// Copyright   : Your copyright notice
+// Description : Hello World in C++, Ansi-style
+//============================================================================
+#include <iostream>
+#include <fstream>
+//#include "gc.h"
+//#include "Utils.hpp"
+#include "Simulator.h"
+#include <unordered_map>
+#include "Demands.h"
+#include <ctime>
+#include <sys/stat.h>
+
+using namespace std;
+using namespace Scheme::AST;
+using namespace Scheme::Demands;
+using namespace Scheme::output;
+
+double gctime=0;
+int gcinvoke=0;
+extern demand_grammar gLivenessData;
+
+
+
+unsigned int FileRead( istream & is, vector <char> & buff ) {
+    is.read( &buff[0], buff.size() );
+    return is.gcount();
+}
+
+unsigned int CountLines( const vector <char> & buff, int sz ) {
+    int newlines = 0;
+    const char * p = &buff[0];
+    for ( int i = 0; i < sz; i++ ) {
+    	if ( p[i] == '\n' ) {
+    		newlines++;
+    	}
+    }
+    return newlines;
+}
+
+
+struct StringListCompare
+{
+	bool operator () (const list<std::string> & p_lhs, const list<std::string> & p_rhs)
+	{
+		const size_t lhsLength = p_lhs.size() ;
+		const size_t rhsLength = p_rhs.size() ;
+
+		return (lhsLength < rhsLength) ; // compares with the length
+	}
+} StringListComparer;
+
+
+Simulator::Simulator(int gctype)
+{
+	pgm = NULL;
+	conscount = 0;
+	gc_invoke = 0;
+	gc_type = (GCStatus)gctype;
+	heap_size = 0;
+	cout << "Simulator created successfully" << endl;
+	//Do all global initialization here
+}
+
+void print_path (path prod_path, ostream& out)
+{
+
+	std::stringstream ss;
+	for_each(prod_path.begin(), prod_path.end(), [&ss] (const std::string& s) { ss << "." << s; });
+	if (ss.str().size() > 1)
+	{
+		std::string a = ss.str().substr(1); //Remove the first "." character
+		out << a;
+	}
+}
+
[…]

### 2014-09-10 lazy-scheme/src/gc.cpp

@@ -1,0 +1,1833 @@
+#include <malloc.h>
+#include <string.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <assert.h>
+#include <vector>
+#include <deque>
+#include <string>
+#include <iostream>
+#include <fstream>
+#include <unordered_map>
+#include <set>
+#include <utility>
+#include <algorithm>
+#include "gc.h"
+using namespace std;
+
+int numcopied = 0; //TODO:Delete after debugging
+
+
+extern GCStatus gc_status;
+
+//globals
+void *freept;
+void *scan;
+void *buffer_live;
+void *buffer_dead;
+void *boundary_live;
+void *boundary_dead;
+unsigned long memorysize;
+unsigned long newheapcells,oldheapcells;
+unsigned long copycells,heapslot,garbagecells;
+actRec *stack_start, *stack_top;
+GCStatus gc_status = gc_plain;
+int gccount = 1;
+
+
+
+//Use only push_front and pop_front to insert and remove elements from actRecStack
+//Use iterator to iterate over the elements of actRecStack during garbage collection
+deque<actRec> actRecStack = deque<actRec>(0);
+deque<actRec>::iterator actRecIter;
+vector<var_heap>::iterator heapRefIter;
+vector<var_stack>::iterator nonRefIter;
+stateMap statemap;
+state_index **state_transition_table; //Dynamically allocating a 2-D array might be error prone. See if vectors can be used instead.
+//call_next_data call_next; //Should we use a map?
+std::unordered_map<std::string, std::string> call_next;
+
+//ofstream gcout;
+
+
+void reachable_dfs();
+void clear_liveness_data();
+void print_heap();
+void print_buffer_data();
+void print_cell_data();
+int is_valid_address(void*);
+int lt_scan_free();
+int lt_scan_freept();
+void* getscan();
+void update_scan();
+//long diff_scan_free();
+state_index lookup_dfanode(string nodename);
+void* followpaths(void* loc, state_index index);
+state_index get_target_dfastate(state_index i1, state_index i2);
+void set_car(void* loc,  void* ref);
+void set_cdr(void* loc,  void* ref);
+void copy_children(void* cellptr, stateset* st);
[…]

### 2014-09-10 paper/sections/syntax-and-semantics.tex
Op sematics first version

@@
 %Decision - numbers are boxed
-Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
-A value is either a number, the empty list $\NIL$, or a location  $\ell$.
-Our liveness analysis does not track numeric values, and thus is neutral
-as to whether these are boxed or represented as immediates.
-An environment is a finite mapping from variables to values, and a heap a finite mapping
-from locations to pairs of values.
-% Given  a program,  an  environment is  a
-% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
-% values in $\mathit{Val}$.    The heap is a mapping
-% from  locations  to pairs of  values.
-Finally, $\stk$ is a stack
-(using $\bullet$ for push and $[\,]$ for empty stack)
-of frames of unfinished function calls.  A
-frame is a triple $(e,x,\rho)$ representing the call site
-$
-(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
-$
-being evaluated in environment $\rho$.
-Frames can also be viewed as continuations, in this view
-the ({\sc ord-return}) 
- rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
+Here $\mathit{Loc}$ is countable set  of locations in the heap.  Since
+all data objects are boxed, we  model an environment as a mapping from
+the set of variables of the program $\mathit{Var}$ to locations in the
+heap.  These either contain a value  in WHNF (a number, the empty list
+$\NIL$, or  a \CONS\ cell)  or a {\em  closure}.  A closure is  a pair
+$(s, \rho)$ in which $s$ is an unevaluated application with the values
+of its free variables supplied by $\rho$.
 
-The semantics 
-of statements $s$ are given by the judgement form
-$\rho, \heap, s \rightsquigarrow \heap', v$
-and those for expressions $e$ by the form
-$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
-The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
-program terminates successfully with result value $\rho(x)$
-on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$
+Finally, $\stk$  is a stack (using  $\bullet$ for push  and $[\,]$ for
+empty stack)  of frames  of unfinished function  calls.  A frame  is a
+triple  $(e,x,\rho)$   representing  the  call  site   $  (\LET\,\,  x
+\leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\,  e) $ being evaluated in
+environment $\rho$.   Frames can also  be viewed as  continuations, in
+this view  the ({\sc ord-return})  rule in the  small-step operational
+semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
 
-Notation:  we write $\rho[x \mapsto v]$ for the environment which is
-as $\rho$ but has value $v$ at $x$.
-We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
-has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
-$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
-$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
+The semantics of statements $s$ are given by the judgement form $\rho,
+\heap, s \rightsquigarrow \heap', v$  and those for expressions $e$ by
+the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
+The  start state is  $(\{\},[\,], \{\},  e_\mainpgm)$ and  the program
+terminates successfully  with result  value $\rho(x)$ on  reaching the
+halt state $(\rho,[\,], \heap, (\RETURN\ x))$
+
+Notation: we write $\rho[x \mapsto v]$ for the environment which is as
+$\rho$ but  has value $v$ at  $x$.  We also  write $[\myvec{x} \mapsto
+  \myvec{v}]$  which respectively  has  values $v_1,  \ldots, v_n$  at
+$x_1,  \ldots, x_n$  and write  $[\myvec{x}  \mapsto \rho(\myvec{y})]$
+when $v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
 
 \paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
-with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
-a contained $s$).  Some of these we eliminate syntactically, e.g.\
-ensuring all variables and functions are defined and are called with
-the correct number of parameters.
-Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
-All but the first can be eliminated with a static type system but,
-treating our program as dynamically typed, we regard all these
-as stuck states.
-
-% A program terminates successfully if it reaches the halt state
-% in Figure~\ref{fig:lang-opsem-small}.
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
+with  $\rightarrow$  (perhaps  because  $\rightsquigarrow$  could  not
+reduce a  contained $s$).  Some  of these we  eliminate syntactically,
+e.g.\ ensuring all variables and  functions are defined and are called
+with the correct number  of parameters.  Others include $(\CDR\ \NIL),
+(\CAR\ 3),  (+\ \NIL\ 4)$ and  $(\SIF\ \NIL\ e_1\ e_2)$.   All but the
+first can  be eliminated with a  static type system  but, treating our
+program as dynamically typed, we regard all these as stuck states.
 
 \begin{figure}[t!]
-% (Random text to ensure proper line-spacing (problem with infrule macro?).)
-
-\footnotesize
-\begin{minipage}{0.335\textwidth}
-\infrule[ord-const]
-        {}
-        {\rho, \stk, \heap,   \pp{\pi}{k}
-          \rightsquigarrow \rho, \stk, \heap, k}
-\end{minipage}\begin{minipage}{0.655\textwidth}
-\infrule[ord-cons]
-      {
-%%\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
-}
-      {\rho, \ell \bullet \stk,\heap, \pp{\pi}{(\CONS\,\, x\ y)}
-        \rightsquigarrow \rho, \stk, \heap[\ell\mapsto(\rho(x),\rho(y))],
-        \ell}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-car]
-        {\heap(\rho(x)) = (e_1, e_2)}
-        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
-          \rightsquigarrow \rho, \stk, \heap, e_1}
-\end{minipage}
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-car]
-        {\heap(\rho(x)) = e}
-        {\rho, \stk, \heap, \pp{\pi}{(\CAR\,\, x)}
-          \rightsquigarrow \rho, (\CAR \bullet):\stk, \heap, v_1}
-\end{minipage}\\[0.5ex]
-
-\begin{minipage}{0.595\textwidth}
-\infrule[ord-cdr]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
-          \rightsquigarrow \heap, v_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-id]
-      {
-}
-      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
-        \rightsquigarrow \heap, \rho(x)}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-prim]
-      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
-      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
-        \rightsquigarrow \heap, \rho(x) + \rho(y)}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-null-false]
-\infrule[]
-        {\rho(x)\not= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 0}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-null-true]
-\infrule[ord-null]
-        {\rho(x)= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 1}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-if-true]{
-\infrule[]{
-  \rho(x) \in \mathbb{N}\setminus\{0\}
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-if-false]{
-\infrule[ord-if]{
-  \rho(x) = 0
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-nonfn]{
-%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
-  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
-  \andalso
-  \mbox{$s$ is not
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-fncall]{
-  \mbox{$s$ is 
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-\andalso
-\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
-[\myvec{z} \mapsto \rho(\myvec{y})],
-  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
-}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-return]{
-}{
-  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
-  \pp{\pi}{(\RETURN\ x)}
-  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
-\end{minipage}
-
-% \normalsize
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-  \caption{The small-step operational semantics}
-  \label{fig:lang-opsem-small}
-\figrule
-\normalsize
+\begin{tabular}{|l|l|c|}
+\hline
+Condition & Transition & Rule name \\
+\hline
+          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
+  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
+\hline
+          & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
+$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (v_1, v_2) & \rho, (\rho', z, e)\!:\!S, H,
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v_1], e$      &  \sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (\CAR~x)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
+$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
+\hline
+\shortstack[l]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
+[\rho(\myvec{x})/\myvec{y}], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
+\hline
+$\ell$ is a new location& \shortstack[l]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+  \rightsquigarrow$ \\
+$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
+  \ell)\oplus\lfloor \rho\rfloor_{FV(S)-\{x\}})], e$} & \sc{let} \\ 
+\hline
+$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_1$ & \sc{if-true} \\ 
+\hline
+$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_2$ & \sc{if-false} \\ 
+\hline
+$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H,
+  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
+$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
+\hline
+\shortstack[l]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
+(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
+\hline
+$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H, (\RETURN~x)
+  \rightsquigarrow$ \\
+$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-clo} \\
+\hline
+\end{tabular}
+\caption{A small-step semantics for the language}\label{fig:lang-semantics}
 \end{figure}
-
 %==============================================================

### 2014-09-10 paper/sections/liveness.tex
Just started

--- 2014-07-13 paper/sections/restOfOldPaper.tex
@@
-\section{The target language---syntax and semantics}
-\label{sec:defs}
-We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
-functions ($\CONS$, $\PRIM$ etc.).
-The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
-it has eager semantics and restricts programs to be in 
-Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
-where all actual parameters to functions are variables.
-This restriction does not affect expressibility (and indeed we feel free
-to ignore it in examples when inessential), but simplifies the analysis formulation.
-Additionally, as in the three-address instruction form familiar from compiler texts,
-it forces each temporary to be named and function calls to be serialized
-(necessary to get an unambiguous definition of liveness).
-We further require that each variable in a program is distinct, so that no scope shadowing
-occurs---this simplifies proofs of soundness.
-In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
-or return ($\RETURN$).
-The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.
-
-The body of the program is
-the  expression denoted  by  $e_\mainpgm$; for analysis purposes
-it is convenient
-to regard $e_\mainpgm$ as part of a function definition
-$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
-% As  mentioned earlier,
-We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.
-
-% \paragraph{Inlining.}
-In spite of the ANF restrictions it is still possible
-to inline non-recursive functions (a fact we use to prove the safety
-of liveness analysis).  A user-function call
-$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
-to a function defined
-(after renaming its formals and locals to be disjoint from existing variables)
-by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
-is replaced by a sequence of $\LET$'s of the form
-$z_i\leftarrow (\ID\,\,y_i)$ followed by
-the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
-$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.
-
-\begin{figure}[t!]
-\footnotesize
-\begin{eqnarray*}
-   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
-    \hspace{7em} \mbox{\em --- program}\\
-    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
-    e) 
-    \hspace{1.7em} \mbox{\em --- function definition} \\
-e \in \mathit{Expr} & ::= &
-\left\{\begin{array}{ll@{\hspace{4em}}l}
-       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
-       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
-       (\RETURN\,\, x) && \mbox{\em --- return from function}
-    \end{array}\right. \\
-s \in \mathit{Stmt} & ::= &
-\left\{\begin{array}{lr@{\hspace{1em}}l}
-       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
-       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
-       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
-       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
-%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
-       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
-%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
-       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
-            & \mbox{\em --- function application} 
-    \end{array}\right.
-\end{eqnarray*}
-  \caption{The syntax of our language}\label{fig:lang-syntax}
-\figrule
-\normalsize
-\end{figure}
-
-\subsubsection{Semantics}
-We now give an operational
-semantics for  our language.  Later, a refinement of
-the  operational semantics,  which we  call minefield  semantics, will
-serve to prove liveness analysis correct.
-We give a small-step semantics because, unlike big-step semantics,
-correctness for non-terminating programs does not need special treatment.
-We start with the domains used by the semantics:
-\[
-\begin{array}{rlcl@{\hspace{2em}}l}
-% \begin{eqnarray*}
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
-\mbox{-- Environment} \\ 
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
-  Val}) & \mbox{-- Heap}
-% \end{eqnarray*}
-\end{array}
-\]
-Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
-A value is either a number, the empty list $\NIL$, or a location  $\ell$.
-Our liveness analysis does not track numeric values, and thus is neutral
-as to whether these are boxed or represented as immediates.
-An environment is a finite mapping from variables to values, and a heap a finite mapping
-from locations to pairs of values.
-% Given  a program,  an  environment is  a
-% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
-% values in $\mathit{Val}$.    The heap is a mapping
-% from  locations  to pairs of  values.
-Finally, $\stk$ is a stack
-(using $\bullet$ for push and $[\,]$ for empty stack)
-of frames of unfinished function calls.  A
-frame is a triple $(e,x,\rho)$ representing the call site
-$
-(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
-$
-being evaluated in environment $\rho$.
-Frames can also be viewed as continuations, in this view
-the ({\sc ord-return}) 
- rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
-
-The semantics 
-of statements $s$ are given by the judgement form
-$\rho, \heap, s \rightsquigarrow \heap', v$
-and those for expressions $e$ by the form
-$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
-The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
-program terminates successfully with result value $\rho(x)$
-on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$
-
-Notation:  we write $\rho[x \mapsto v]$ for the environment which is
-as $\rho$ but has value $v$ at $x$.
-We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
-has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
-$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
-$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
-
-\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
-with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
-a contained $s$).  Some of these we eliminate syntactically, e.g.\
-ensuring all variables and functions are defined and are called with
-the correct number of parameters.
-Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
-All but the first can be eliminated with a static type system but,
-treating our program as dynamically typed, we regard all these
-as stuck states.
-
-% A program terminates successfully if it reaches the halt state
-% in Figure~\ref{fig:lang-opsem-small}.
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-\begin{figure}[t!]  
-% (Random text to ensure proper line-spacing (problem with infrule macro?).)
-
-\footnotesize
-\begin{minipage}{0.335\textwidth}
-\infrule[ord-const]
-        {}
-        {\rho, \heap, \pp{\pi}{k}
-          \rightsquigarrow \heap, k}
-\end{minipage}\begin{minipage}{0.655\textwidth}
-\infrule[ord-cons]
-      {
-\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
-}
-      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
-        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
-        \ell}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-car]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
-          \rightsquigarrow \heap, v_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-cdr]
-        {\heap(\rho(x)) = (v_1, v_2)}
-        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
-          \rightsquigarrow \heap, v_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-\infrule[ord-id]
-      {
-}
-      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
-        \rightsquigarrow \heap, \rho(x)}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-\infrule[ord-prim]
-      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
-      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
-        \rightsquigarrow \heap, \rho(x) + \rho(y)}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-null-false]
-\infrule[]
-        {\rho(x)\not= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 0}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-null-true]
-\infrule[ord-null]
-        {\rho(x)= \NIL
-        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
-          \rightsquigarrow \heap, 1}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.395\textwidth}
-% \infrule[ord-if-true]{
-\infrule[]{
-  \rho(x) \in \mathbb{N}\setminus\{0\}
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_1}
-\end{minipage}\begin{minipage}{0.595\textwidth}
-% \infrule[ord-if-false]{
-\infrule[ord-if]{
-  \rho(x) = 0
-}{
-  \rho,\stk, \heap,
-  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
-  \longrightarrow \rho,\stk, \heap, e_2}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-nonfn]{
-%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
-  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
-  \andalso
-  \mbox{$s$ is not
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-let-fncall]{
-  \mbox{$s$ is 
-$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
-       }
-\andalso
-\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
-}{
-  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
-    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
-    \pp{\pi_2}{e})}
-  \longrightarrow
-% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
-[\myvec{z} \mapsto \rho(\myvec{y})],
-  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
-}
-\end{minipage} \\[0.5ex]
-
-\begin{minipage}{0.990\textwidth}
-\infrule[ord-return]{
-}{
-  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
-  \pp{\pi}{(\RETURN\ x)}
-  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
-\end{minipage}
-
-% \normalsize
-% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
-% is the halt state (with result value $\rho(x)$).
-
-  \caption{The small-step operational semantics}
-  \label{fig:lang-opsem-small}
-\figrule
-\normalsize
-\end{figure}
-
-%==============================================================
-\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-
 \section{Liveness}\label{sec:liveness}
@@
 may be  used in future  computation) or `dead' (definitely  not used).
-Semantically, a variable is dead at a given program point
-if arbitrary changes to its value have no effect on the computation.
-Later we will use $\bot$ to represent a value which `explodes' when
-it is used in a computation; dead variables can safely have their
-value replaced with $\bot$.
-For heap-allocated data  we need a richer model of liveness
-in that both variables and fields of $\CONS$ cells may be dead or live.
-Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
-liveness of the structure reachable from a variable is a set of
-{\em access paths}
-which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
-grammar notation.
-Thus the liveness of $x$ being
-$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
-only refer to the second and third members of $x$ considered as a list.
-Semantically, access paths are prefix-closed, as accessing a field
-requires accessing all the paths from the variable to the field,
-and hence the above liveness is properly written 
-$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
-The classical notions of a scalar variable being live or dead correspond to
-$\{\epsilon\}$ and $\{\}$.
+Semantically, a variable is dead at a given program point if arbitrary
+changes to its value have no effect on the computation.  Later we will
+use $\bot$ to represent a value  which `explodes' when it is used in a
+computation; dead variables can  safely have their value replaced with
+$\bot$.  For heap-allocated data we need a richer model of liveness in
+that both variables  and fields of $\CONS$ cells may  be dead or live.
+Using  $\acar$,  $\acdr$  to  represent access  using  $\CAR$,  $\CDR$
+respectively, liveness of the structure reachable from a variable is a
+set  of  {\em  access  paths}  which  we  represent  as  a  subset  of
+$\{\acar,\acdr\}^\ast$, and  use conventional grammar  notation.  Thus
+the liveness of $x$ being $\{\acdr\acar, \acdr\acdr\acar\}$ means that
+future computation can  only refer to the second  and third members of
+$x$   considered  as   a   list.   Semantically,   access  paths   are
+prefix-closed, as  accessing a field requires accessing  all the paths
+from  the variable  to  the field,  and  hence the  above liveness  is
+properly   written   $\{\epsilon,   \acdr,   \acdr\acar,   \acdr\acdr,
+\acdr\acdr\acar\}$.  The classical notions  of a scalar variable being
+live or dead correspond to $\{\epsilon\}$ and $\{\}$.
 
-The overall liveness (also written {\em liveness environment} for emphasis)
-at a program point is conceptually a mapping from
-variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
-notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
-instead of the map
-$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
-Analogously to classical liveness, the liveness at program point $\pi$
-in $\pi:e$ is the liveness just before executing $e$.
+The  overall liveness  (also  written {\em  liveness environment}  for
+emphasis) at a program point  is conceptually a mapping from variables
+to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse notation, for
+example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$ instead of the
+map           $[x          \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},
+  y\mapsto\{\epsilon\},  z\mapsto\{\}]$.    Analogously  to  classical
+liveness,  the liveness  at  program  point $\pi$  in  $\pi:e$ is  the
+liveness just before executing $e$.
 
@@
 
-We use $\sigma$ to range over demands, $\alpha$ to range over access paths
-and $\Lv$ to range over liveness environments. The notation
+We use $\sigma$  to range over demands, $\alpha$  to range over access
+paths  and $\Lv$  to range  over liveness  environments.  The notation
 $\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
@@
 Finally,  we use $\Lfonly$  to range  over {\em  demand transformers}.
-These transform demands on a function call
-into demands for its formal parameters:
-given a user function $f$, defined by
-$(\DEFINE\ (f\ x_1\ \ldots\ x_n)\ \ e_f)$ and called with demand $\sigma$,
-then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
+These transform demands on a function call into demands for its formal
+parameters:    given    a    user    function    $f$,    defined    by
+$(\DEFINE\  (f\ x_1\  \ldots\ x_n)\  \  e_f)$ and  called with  demand
+$\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
 
-Note that liveness refers to variables and fields, and not to $\CONS$ cells
-(i.e.\ to edges in the memory graph, not to locations themselves).
-Hence liveness of $\{x.\epsilon,x.\acar\}$ means that future computation
-may refer to the value $\ell$ of variable $x$, and also to the $\CAR$ field
-of location $\ell$.
-In the absence of other pointers to heap location $\ell$, we are
-certain that the $\CDR$ field of $\ell$ will not be referenced
-and may hence be corrupted arbitrarily.
-Note therefore, that while  $\ell$ cannot be garbage collected, any
-location $\ell'$ stored in the $\CDR$ field of $\ell$ would be garbage
-(again provided there are no other aliases to  $\ell$).
-
-%% \comment{
-%%   \begin{enumerate}
-%% \item Set the framework for liveness. Basic notions: access
-%%   paths and expressions.
-%% \item An informal working definition of liveness. Minefield will
-%%   provide the formal definition.
-%% \item Liveness judgements. The form  of the judgement ($e, \sigma \len
-%%   \Lv$)  and the  intuition  behind it.   Justification of  individual
-%%   rules. One of the changes from  the earlier version is the lack of a
-%%   post-liveness in  the judgements. This anticipates  the intended use
-%%   of  the analysis  for garbage  collection and has  to be  carefully
-%%   explained.
-%%   \end{enumerate}
-%% }
-
-%% We now  set up  the framework for  liveness analysis.  Given  a memory
-%% graph, a link in the graph  can be described by starting with a member
-%% of  the  root   set  and  going  through  a   sequence  of  \CAR\  and
-%% \CDR\ dereferences to reach the link.  For notational convenience, the
-%% links corresponding to  the \CAR\ and \CDR\ fields  of the \CONS\ cell
-%% are  denoted by  \acar\  and \acdr,  and  thus a  sequence of  several
-%% dereferences    is    represented   by    a    string   $\alpha    \in
-%% \lbrace\acar,\acdr\rbrace^*$.  The string  $x.\alpha$ describes a path
-%% in the  memory graph that starts  at $x$ and  traverses adjacent links
-%% specified by  $\alpha$.  $x.\epsilon$ is  written more simply  as $x$.
-%% We call the string $x.\alpha$ an {\em access expression\/}, the string
-%% $\alpha$ an {\em access pattern\/},  and the path traced in the memory
-%% graph   by  the   access  expression   an  {\em   access   path}.   In
-%% Figure~\ref{fig:mot-example},        the       access       expression
-%% \ww.\acdr\acar\acar\ represents the access  path from \ww\ to the node
-%% containing the value  $4$.  Most often the memory  graph is clear from
-%% the context,  and we shall use  access expressions to  refer to access
-%% paths. An  access path  used to refer  to a  link in the  memory graph
-%% refers    to     the    last     link    in    the     path.     Thus,
-%% \ww.\acdr\acar\acar\ denotes the link  incident on the node containing
-%% the value $4$.  As a  notational convenience, if $\sigma$ represents a
-%% set of access paths, then  $x.\sigma$ will represent the set of access
-%% paths $  \lbrace x.\alpha \mid \alpha  \in \sigma\rbrace$.  Similarly,
-%% if  $r$ is  the single  symbol \acar\  or \acdr,  then  $r\sigma$ will
-%% denote the  set of access patterns  $ \lbrace r\alpha  \mid \alpha \in
-%% \sigma\rbrace$.
-
-%% \subsection{Liveness Analysis}
-%% \label{sec:exp-liveness}
-%% 
-%% Consider the memory graph at a program point $\pi$ during an execution
-%% of a program. A  link in the memory graph is said  to be {\em live} at
-%% $\pi$,  if it  is used  (dereferenced or  tested) during  a  period of
-%% execution  starting from  the  evaluation of  the expression  labelled
-%% $\pi$, to the end of the program.  An access path $x.\alpha$ is said to
-%% be live at  $\pi$, if, during {\em any} execution  of the program, the
-%% link denoted by $x.\alpha$ in the memory graph is live at $\pi$.  Note
-%% that an access path such as  $\px.\acdr\acar$ can be live in two ways:
-%% In the first, the path  itself represents the sequence of dereferences
-%% to  access  the last  edge,  as,  for  example, while  evaluating  the
-%% expression  $(\LET\,\,   \pz  \leftarrow  (\CDR\,\,\,\px)\,\,  \IN\,\,
-%% (\CAR\,\,\pz))$.  In the second, the  path may be made live indirectly
-%% by  evaluating $(\LET\,\,  \pz \leftarrow  (\CAR\,\,\,\py)\,\, \IN\,\,
-%% (\CAR\,\,\pz))$  in an execution  context where  $(\CDR\,\,\,\px)$ and
-%% $(\CAR\,\,\,\py)$ share  the same node.   We call an access  path {\em
-%%   explicitly  live},  if  it is  live  in  the  first sense  and  {\em
-%%   implicitly  live} if it is live in the second sense.  Clearly, static
-%% determination   of  implicit  liveness   requires  an   analysis  that
-%% determines  potential sharing  during  any execution  of the  program.
-%% However, the intended application  of our liveness analysis is garbage
-%% collection  during which  we have  access  to the  entire heap.   This
-%% obviates the need for a  separate sharing analysis and we restrict our
-%% attention to explicit liveness only.
-%% 
-%% 
-%% %% We  extend  our  alphabet   of  access  patterns  to  include  symbols
-%% %% \bcar\ and \bcdr\   representing backward traversals over a  $\CAR\ $ or
-%% %% $\CDR\  $  edge.   If  traversals  over \acar\  and  \acdr\  represent
-%% %% dereferencing    or   selection,    traversals    over   \bcar\    and
-%% %% \bcdr\ represent application of  the constructor \CONS.  The following
-%% %% example motivates the need for these symbols.
-%% 
-%% %% \begin{example}\label{exmp:motiv-bar}
-%% %%   Consider the expression at program point $\pi_1$ in
-%% %%   \begin{center}
-%% %%     {\sf $\pi_1\!:${(\LET\ w $\leftarrow$
-%% %% 	                    $\pi_2\!:${(\CONS\ x y)} \scmin\ 
-%% %% 			    $\pi_3\!:${$\cdots$})}}
-%% %%   \end{center}
-%% %%   Assuming  that  the   set  of  live  access  paths   at  $\pi_3$  is
-%% %%   $\lbrace\ww.\alpha\rbrace$, we  would like to find  out which access
-%% %%   paths  rooted   at  \xx\   are  live  at   $\pi_1$.   Let
-%% %%   \xx.$\alpha'$ be live at $\pi_1$.  Then, the two possible cases are:
-%% %%   \begin{itemize}
-%% %%   \item If $\alpha = \acdr\beta$ or $\alpha = \epsilon$, then there is
-%% %%     no link  that is accessible through  a path rooted at  \xx\ and is
-%% %%     live at $\pi_1$.
-%% %%   \item  If  $\alpha  =  \acar\beta$  then  the  link  represented  by
-%% %%     $\ww.\alpha$  is also  represented  by $\xx.\beta\ $and  is live  at
-%% %%     $\pi_1$.  Thus, $\alpha' = \beta$.
-%% %%   \end{itemize}
-%% %%   This  relation  between  $\alpha$  and  $\alpha'$  is  expressed  by
-%% %%   $\alpha' = \bcar\alpha$.  \bcdr\ can be interpreted similarly. \qed
-%% %% \end{example}
-%% 
-%% %% While the inclusion of \bcar\ and \bcdr\ in the alphabet enables us to
-%% %% treat  applications  of  selectors  and  constructors  uniformly,  the
-%% %% resulting access patterns may not  be directly useful in that they may
-%% %% not describe a  path in the memory graph.  We  define a {\em canonical
-%% %%   access pattern\/}  as a pattern  that is restricted to  the alphabet
-%% %% $\lbrace\acar,  \acdr\rbrace$.   Given   an  access  pattern  $\alpha$
-%% %% containing \bcar\ and \bcdr, it  may be possible in certain cases
-%% %% to obtain a canonical access  pattern representing the same edge. This
-%% %% is  through reductions  based on  the identities  $(\CAR\  (\CONS\ \px
-%% %% \  \py))  =  \px$  and  $(\CDR\  (\CONS\  \px  \  \py))  =  \py$.   In
-%% %% Figure~~\ref{fig:mot-example},   the  path  $\pz.\bcdr\acdr\acar\acar$
-%% %% reduces  to $\pz.\acar\acar$  in this  sense.  If  $\alpha$  cannot be
-%% %% reduced  to path without  \bcar\ and  \bcdr\ by  these rules,  then we
-%% %% reduce it  to the special  canonical access pattern $\bot$.   Since an
-%% %% access  path with  $\bot$ cannot  be used  for traversing  a  heap, it
-%% %% cannot  be  used for  nullification  during  garbage collection.   The
-%% %% liveness of a such a path is irrelevant, and the path is not considered
-%% %% any further.
-%% 
-%% %% $\alpha  \rightk{k}  \alpha'$ denotes  the  reduction  of $\alpha$  to
-%% %% $\alpha'$  in $k$ steps,  and $\rightstar$  denotes the  reflexive and
-%% %% transitive closure of $\rightarrow$.  The concatenation ($\cdot$) of a
-%% %% set of access patterns $\sigma_1$  with $\sigma_2$ is defined as a set
-%% %% containing  concatenation  of each  element  in  $\sigma_1$ with  each
-%% %% element in $\sigma_2$, i.e.
-%% 
-%% %% \begin{eqnarray*}
-%% %%   \sigma_1\cdot\sigma_2 &\eqdef& \lbrace \alpha_1\alpha_2 \mid \alpha_1 \in
-%% %%   \sigma_1, \alpha_2 \in \sigma_2 \rbrace
-%% %% \end{eqnarray*}
+Note that liveness refers to  variables and fields, and not to $\CONS$
+cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
+themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
+future computation may refer to  the value $\ell$ of variable $x$, and
+also to the $\CAR$ field of  location $\ell$.  In the absence of other
+pointers to heap location $\ell$, we are certain that the $\CDR$ field
+of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
+arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
+collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
+would  be  garbage (again  provided  there  are  no other  aliases  to
+$\ell$).
 
@@
 % \item
-The function $\mathcal{L}$ now gives the (total) liveness of an expression $e$.
-The cases $\RETURN$ and $\SIF$ are straightforward, but note the liveness $x.\epsilon$ generated
-by the latter.
-The case $(\LET\ z\leftarrow s\ \IN\ e')$
-resembles a three-address instruction:
-the liveness of $e$ is given by taking the liveness, $\Lv$, of $e'$,
-killing any liveness of $z$ and adding any incremental liveness from $s$.
-The main subtlety is how the liveness of $z$ in $\Lv$ is converted to a demand $\Lv(z)$ to be
-placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
+The  function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
+expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
+but note the liveness $x.\epsilon$  generated by the latter.  The case
+$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
+instruction:  the liveness  of $e$  is given  by taking  the liveness,
+$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
+liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
+$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
+$\mathit{ref}(s,\Lv(z),\Lfonly)$.
 
@@
 %%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-Finally, the judgement form $\mathit{Prog} \len \Lfonly$ is used to determine $\Lfonly$\@.
-Analogously to classical liveness being computed as
-a solution of dataflow equations, we require, via inference rule ({\sc live-define}),
-$\Lfonly$ to satisfy the fixed-point property that:
- when we assume $\Lfonly$ to be the family of demand transformers for the program
- then the calculated liveness of each function body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees
- with the assumed  $\Lfonly$.
-As usual, there are often multiple solutions to $\Lfonly$; all are safe
-(see Section~\ref{sec:correctness})
-but we prefer the least one as giving the least liveness subject to safety---and
-hence greatest amount of garbage collected.
+Finally, the  judgement form $\mathit{Prog}  \len \Lfonly$ is  used to
+determine  $\Lfonly$\@.   Analogously   to  classical  liveness  being
+computed  as  a  solution  of  dataflow  equations,  we  require,  via
+inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
+fixed-point property that:  when we assume $\Lfonly$ to  be the family
+of demand transformers for the program then the calculated liveness of
+each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
+assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
+$\Lfonly$;  all are  safe (see  Section~\ref{sec:correctness})  but we
+prefer  the  least  one  as  giving  the  least  liveness  subject  to
+safety---and hence greatest amount of garbage collected.
 % \end{itemize}
 
-We make three observations: firstly the rule ({\sc live-define}) has a least solution
-as $\mathcal{L}(\cdot)$ is monotonic in $\sigma$; secondly that
-({\sc live-define})
-resembles the rule for type inference of mutually recursive function definitions,
-and thirdly the asymmetry of demand and liveness (compared to post- and pre-liveness
+We make three observations: firstly the rule ({\sc live-define}) has a
+least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
+secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
+inference of mutually recursive  function definitions, and thirdly the
+asymmetry of  demand and liveness (compared to  post- and pre-liveness
 classically) is due to the functional formulation here.
@@
 the corresponding tuple of demands the function makes on its arguments.
-
-%%% save two lines to bring up section 3.2.
-
-% The  next  subsection {\color{red}will  describe}  how  to  derive a  function  for
-% \Lfonly\  from this  rule.  \Lfonly\  is a  summary of  functions that
-% enables us to do a context-dependent analysis of a call without having
-% to enter the body of the called function.
-
-%% We present  our liveness  analysis using judgement forms $\len$
-%% for both expressions and programs.
-%% These are given formally in Figure~\ref{fig:live-judge}, but
-%% first we an intuitive understanding of the issues involved.
-%% Consider the expression:
-%% 
-%% \begin{center}
-%% \begin{uprogram}
-%% 	  \UFL\ \hspace*{-1\TAL} $\pi\!:$(\LET\, \pz\   $\leftarrow$ (\CAR\,\,
-%%                         \px) \IN\  
-%%            % \UNL{0}
-%%              $\pi'\!:\!$ (\LET\, \py\  $\leftarrow$ (\PRIM\,  \pz\,\,
-%%                         \pw ) \IN \ldots) 
-%% \end{uprogram}
-%% \end{center}
-%% 
-%% 
-%% Assume as a starting point that liveness at $\pi'$ is $\{\pz.\epsilon,
-%% \pw.\epsilon\}$.
-%% Informally, this is so since \pw\ and \pz\ are used in
-%% (\PRIM~\pz~\pw ).
-%% 
-%% \begin{enumerate}
-%% \item The liveness  at $\pi$ is obtained by  removing the access paths
-%%   starting  with \pz\  from the  liveness  at $\pi'$  and adding  the
-%%   liveness  generated ($\mathit{ref}()$) by $(\CAR~\px)$\footnote{This is
-%%     similar  to liveness  analysis of  imperative programs,  where the
-%%     liveness before a statement $s$ is given as $\mathit{live_{in}}(s)
-%%     =   \mathit{live_{out}}(s)    \setminus   \mathit{def\/}(s)   \cup
-%%     \mathit{ref\/}(s)$. Here  $\mathit{live_{out}}(s)$ is the liveness
-%%     after $s$ and  $\mathit{ref\/}(s)$ and $\mathit{def\/}(s)$ are the
-%%     liveness generated and  killed in $s$}. 
-%% \item  Since  \pz\ is  bound  to  the  value of  $(\CAR~\px)$,  the
-%%   liveness of \pz, namely $\epsilon$, is transferred to $(\CAR~\px)$
-%%   as a {\em  demand}.  The demand on an expression consists of  access
-%%   patterns  representing all  possible future  uses of  its  value.  A
-%%   demand  of $\epsilon$  means that  only the  reference resulting
-%% %  from the evaluation  of
-%%   from  $(\CAR\,\,\px)$ will be used.
-%% \item  Because of this  demand, the  liveness generated  by (\CAR\,\,
-%%   \px) are $\px$ and $\px.\acar$.  Thus liveness at $\pi$ is
-%%   is $\{\px.\epsilon, \px.\acar, \pw.\epsilon\}$.
-%% \end{enumerate}
-%% 
-%% In general,  an access path is  live before an  expression because the
-%% link represented by it is used (i) while evaluating the expression, or
-%% (ii) while using the result of the expression, or (iii) simply because
-%% the link  is live at  a later program  point.  When the  expression in
-%% consideration  is (\CAR~\px), (i)  is illustrated  by the  access path
-%% \px, (ii)  is illustrated by the  access path \px.\acar,  and (iii) by
-%% the access  path \pw.   Note that \px.\acar\  is live in  this example
-%% because  the   result  of  (\CAR~\px)   is  used  in   the  expression
-%% (\PRIM~\pz~\pw  ) through  the variable  \pz.  In  contrast, replacing
-%% (\PRIM~\pz~\pw ) by the expression (\CONS~\pz~\pw) would not have made
-%% (\CAR~\px) live, since \CONS\ would not have dereferenced \pz.
-%%
-
-% \paragraph{Judgements formally.}
-% Let  $\Lv$  denote a  liveness  environment (a function from
-% variables to access paths, abused using grammar notation) and
-% $\sigma$ be a demand.
-% The judgement form $e, \sigma \len \Lv$,
-% reads  as:  if  the demand  on  $e$  is  $\sigma$, then  the  liveness
-% environment  before $e$  is $\Lv$ and
-% is thus analogous to the classical notion of {\em transfer function}.
-
-%    It is  interesting to  note that,
-% unlike traditional liveness analysis,  our judgements do not require a
-% liveness after the expression. This  is because of the coming together
-% of  two reasons.   First, it  will be  clear while  discussing garbage
-% collection,  that it is  enough for  our purposes  to assume  that the
-% post-liveness  of  each function  body  is  $\emptyset$.  Second,  the
-% restriction of our  programs to ANF ensures that  the post-liveness of
-% each  expression  $e$  inside   the  function  body  continues  to  be
-% $\emptyset$. Thus the post-liveness can be dropped altogether.
 
@@
 \end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%==============================================================
-
-% The liveness judgement rules are shown in Figure~\ref{fig:live-judge}.
-% Given  a  demand $\sigma$,  the  function  $\mathit{ref\/}$ gives  the
-% liveness generated  before the right-hand-side of  a \LET\ definition.
-% Constants    do   not   generate    any   liveness.     The   operator
-% \PRIM\ dereferences its arguments  resulting in the liveness shown.  A
-% demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-% $\acar\sigma$  on  $x$.   In  addition, \CAR\  also  dereferences  its
-% argument.  This  generates the liveness  $\{x\} \cup  x.\acar\sigma$ before
-% (\CAR~$x$).   In a  opposite  sense, the  demand  of $\acar\sigma$  on
-% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-% that \CONS\ does not, by itself, dereference its arguments.
-% 
-% 
-% Unlike primitives, the mapping from the demand on a user-function call
-% to the  demands on its arguments  has to be derived  from the function
-% body itself---as a fixed point for a recursive function.
-% The function  \Lfonly\ captures this mapping for
-% all functions  defined in the  program.  For a  function $\mathit{f}$,
-% \Lf{f}{i}{\sigma} gives the transformation of the demand $\sigma$ from
-% a    call     to    the     $i$th    argument.     The     rule
-% (\rn{live-define})  says that  if  a  guess  of \Lfonly\  is
-% confirmed by the body of all the functions, then the guess is correct.
-% {\color{blue}Note the similarity of this rule to type-inference
-% for recursive program schemes.}
-
-% The   liveness   judgements    for   \RETURN\    and   \SIF\   are
-% straightforward. $x.\epsilon$  appears in the liveness  of \SIF\ since
-% $x$  is  tested as  a  conditional  variable.   Finally, as  mentioned
-% earlier,  the liveness  before  a let  expression $(\LET~x  \leftarrow
-% s~\IN~e)$ is  obtained by  first finding the  liveness before  the let
-% body  $e$,  removing the  liveness  of $x$  from  it,  and adding  the
-% liveness generated before $s$.
-
-\subsection{Minefield semantics and correctness}
-\label{sec:minefield}
-\label{sec:correctness}
-
-This section gives a modified semantics
-%(a reference semantics)
-which checks liveness annotations
-at run time, and `explodes' when these are found to be inconsistent with execution behavior,
-but otherwise behaves as the standard semantics.
-We show that such explosions never occur and hence run-time checks
-can be elided.
-%% We first assume some demand transformers $\Lfonly$;
-%% our proof will rely on this being a solution (i.e.\ satisfies $\mathit{Prog} \len \Lfonly$)
-%% of liveness analysis, but we do rely in this yet.
-We first assume an arbitrary  demand transformers $\Lfonly$ which satisfies
-$\mathit{Prog} \len \Lfonly$.
-We then enrich the abstract machine state $\rho, \stk, \heap, e$ to
-$\rho, \stk, \heap, e, \sigma, \Sigma$.
-Here $\sigma$ is the demand to the currently active function,
-thus the liveness $\Lv$ at $e$ is $\mathcal{L}(e,\sigma,\Lfonly)$,\footnote{
-This a simple liveness propagation using $\mathcal{L}(\cdots)$
-and $\mathit{ref}(\cdots)$ as $\Lfonly$ is assumed given.}
-and $\Sigma$ is a stack of demands---one for each function frame
-pushed in $\stk$.
-
-Second, we augment $\mathit{Val}$ with a value $\bot$.
-To model strong liveness $\bot$ may be copied freely,
-including into a $\CONS$ cell,
-% but reduces the state to $\bang$
-but explodes
-when used computationally (in a primitive operation other than a copy).
-Additionally we define
-$\mathit{GC}(\Lv,\Sigma): (\rho, \heap, \stk) \mapsto (\rho', \heap', \stk')$
-which determines {\em live-reachability}\footnote{Reachability curtailed by liveness information.}
-%%%%% \marginpar{Say more about live-reachability?}
-using $\rho$ and the $\rho$'s in $\stk$ as the root set
-and following links in $\heap$ {\em only as far as allowed by $\Lv$ and $\Sigma$}.
-% The effect of $\mathit{GC}(\Lv,\Sigma)$ is to replace
-Hence $\mathit{GC}(\Lv,\Sigma)$ replaces
-live-unreachable values---in $\rho$, in $\heap$ and the $\rho$'s in \stk---with $\bot$.
-For example, if $x.\epsilon \not\in \Lv$ then $\rho'(x) = \bot$.
-%, regardless of $\rho(x)$.
-Only $\mathit{GC}(\cdots)$ introduces $\bot$.
-
-Third, we update the semantics in four ways: ({\em i\/}) we arrange that all
-($\rightarrow$) transitions on expressions $e$ first use $\mathit{GC}(\cdots)$
-to update the state and then continue as before;
-and ({\em ii\/}) whenever the value $\bot$ is used computationally in a reduction
-($\rightsquigarrow$), we enter a distinguished stuck state $\bang$.
-For example, supposing $\rho(x)=\bot$
-then
-$\rho, \heap, s \rightsquigarrow \bang$ if $s$ is
-$(\CAR\,\, x)$, $(\CDR\,\, x)$ or $(\PRIM\,\, x\ y)$,
-but not if $s$ is $(\ID~x)$, $(\CONS~x~y)$ or $(f~x~y)$.
-Finally ({\em iii\/}) we make a similar change to the ($\rightarrow$) reduction for ({\sc ord-if})
-and ({\em iv\/}) augment the ({\sc ord-let}) rule for primitives to propagate
-$\bang$ from ($\rightsquigarrow$) to ($\rightarrow$).
-%%%%%% {\color{blue}The full minefield semantics is given as an appendix?}
-
-% Suppose liveness for the current $e$ under reduction is $\Lv$; then:
-% For every variable $x$ for which $x.\epsilon \not\in \Lv$ then replace
-% $\rho(x)$ with $\bot$.
-% Also, for every $\ell \in \mathit{Loc}$, replace the $\CAR$ field
-% of $\heap(\ell)$ with $\bot$ when
-% which there is no path $x.\alpha\acar$ in $\Lv$ such that following
-% $\alpha$-links in $\heap$ from $\rho(x)$ reaches $\ell$.
-% Correspondingly, replace the $\CDR$ field of $\ell$ if there
-% is no path $x.\alpha\acdr$ in $\Lv$.
-% \item
-% 
-% {
-% \footnotesize
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-car]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CAR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-cdr]
-%         {\rho(x) = \bot}
-%         {\rho, \heap, {(\CDR\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}\\[1ex]
-% 
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-prim]
-%       {\rho(x) = \bot \vee \rho(y) = \bot}
-%       {\rho, \heap, {(\PRIM\,\, x\ y)}
-%         \rightsquigarrow \bang}
-% \end{minipage}
-% \begin{minipage}{0.445\textwidth}
-% \infrule[mine-null]
-%         {\rho(x)= \bot
-%         }{\rho, \heap, {(\NULLQ\,\, x)}
-%           \rightsquigarrow \bang}
-% \end{minipage}
-% }
-%
-
-The resulting minefield semantics behaves identically
-(identical heap, identical steps, including possible non-termination)
-to the standard semantics, except for one possibility---the
-minefield semantics going $\bang$ while the standard semantics
-continues (either to a halt state, to a stuck state, or reduces forever).
-
- We now prove a result that relates
-liveness analysis to the semantics.
-\begin{proposition}
-Given program $P$ with $P \len \Lfonly$, then in the minefield
- semantics $P \rightarrow \bang$ can never occur
- (cf.\ `well-typed programs do not go wrong').
-\end{proposition}
-\textit{Proof outline:}
-Space does not permit a full proof, but we give the two main steps.
-We proceed by contradiction and assume there is a program $P$ for which
-$P \len \Lfonly$ can enter state $\bang$.
-The first step is to construct a program $P'$ with identical behavior, but with
-no user-function definitions, using inlining.
-This is possible because a program which goes $\bang$ does so after a finite number
-of reductions, and hence even recursive functions have only had a finite number of
-invocations.
-We hence repeatedly inline-expand user-function calls in $P$
-until we obtain a program $P'$ which behaves identically\footnote{
-Modulo replacement of
-({\sc ord-call}) and
-({\sc ord-return}) steps with
-({\sc ord-id}) steps.
-}
-to $P$ in the standard semantics, but makes no user-function calls.
-Any remaining, non-executed, calls can 
-%be left or
-replaced with a new primitive with the same
-demand-to-liveness transfer function---thus making
-$P'$ a simple expression $e'$.
-%%%%%
-Not only do the program points in reducing $e'$ correspond
-one-one to states during evaluation of $P$,
-but also the liveness associated with a point
-in $e'$ is identical to the liveness at the corresponding state
-$(\rho, \stk, \heap, e, \sigma, \Sigma)$ of $P$ (concatenating the liveness
-$\mathcal{L}(e,\sigma,\Lfonly)$ at $e$
-with the liveness, obtained from $\Sigma$, of the call sites in $\stk$ and
-after renaming the variables correspondingly to the inlining which produced $e'$).
-This assertion relies on the analysis being fully context-sensitive, and noting that
-while the change of scope caused by inlining changes variable visibility
-between $P$ and $e'$ it does not change the liveness---as local-to-a-function
-$\LET$-variables whose scope has been prolonged due to inlining
-are dead in the prolonged scope.
-
-The second step of the proof is to show that $e'$ cannot go $\bang$.
-We proceed by induction.
-Correctness of the $\SIF$ and final $\RETURN$ forms are immediate;
-the $(\LET~z\leftarrow s~\IN~e)$ form requires showing
-the inference rules ensure that any value referenced via $z$ in $e$
-was already live-reachable (via another variable and path) in any
-enclosing expression (so that $\mathit{GC}(\cdots)$ could not re-write it to $\bot$).
-
-% \newpage
-
-\section{Computing liveness and its encoding as a table}\label{sec:computing}
-
-Section~\ref{sec:liveness} gave a context-sensitive liveness analysis
-and proved it correct with reference to a {\em minefield} semantics.
-% Practical computation was not considered.
-For practical use we need to solve the liveness equations {\em finitely and
-symbolically}.  As expressed mathematically, and given a fixed program,
-three things are potentially unbounded:
-({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
-({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
-({\em iii\/}) the number of such access paths.
-
-We commonly first solve ({\em i\/}) by reducing the number of distinct
-calling contexts
-(e.g.\ to a single unified context in 0-CFA style).
-However, it turns out that we can solve the equations for $\Lfonly$ symbolically
-without this reduction, so here we defer this to Section~\ref{sec:bodylivenessbodies}.
-(It also allows easier extension to  dynamically determined liveness---future work.)
-%
-We address ({\em ii\/}) and ({\em iii\/}) by 
-% attempting to re-interpret
-re-interpreting
-the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
-rather than a mutually recursive set of equations on sets of access paths.
-
-This requires two ideas.  Firstly we have to control the use of functions---they
-tend to be infinitary and do not occur naturally in CFGs; in particular our
-$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
-to a tuple of access paths.  The former is achieved by using a separate
-meta-variable (later non-terminal)
-$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
-($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
-Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
-expressible finitely (it is a linear form).
-
-Secondly, there are also two technical issues in re-interpreting
-Figure~\ref{fig:live-judge} as a grammar.
-One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
-disappears after variables are separated ($\Lanv{i}{x}$ above). The other
-is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
-not a CFG\@.
-Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.
-
-Finally,  Section~\ref{sec:NFA-approx}
-uses a construction due to
-Mohri and Nederhof~\cite{mohri00regular} to over-approximate
-context-free grammars with regular grammars;
-these are more appropriate for run-time use.
-%
-Hence the overall `big picture' view is that each GC point
-is annotated with a table encoding the DFA for that program point.
-Such program points forming the call string causing the garbage collection
-(and ending in $\CONS$)
-are used by the garbage collector to curtail its reachability-marking phase
-to this over-approximated liveness.
-
-\subsubsection{Modifying the \CONS\ rule.}
-
-The    $\mathit{ref}$    rule for \CONS,    shown    in
-Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
-\acar\  and  \acdr\  from  access paths in  $\sigma$.
-  Mathematically  this  is fine  but causes problems  when
-  solving  the liveness equations  {\em symbolically}  since such
-  decomposition cannot be expressed as a context-free grammar.  
-To  handle this,  we introduce  two  new symbols  \bcar\  and
-\bcdr\ with the properties:
-\[
- \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
-\hspace{3em} \mbox{and} \hspace{3em}
- \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
-\]
-We can now rewrite the \CONS\ rule as:
-\begin{eqnarray*}
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
-\end{eqnarray*}
-
-We  call the liveness equations with  this modification
-  \Lfunonly$'$.
-The  defintions of \bcar\  and  \bcdr\  induce   the  following
-relation $\hookrightarrow$ over sets of access paths:
-\begin{eqnarray*}
-  \sigma_1\bcar\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
-& \qquad\mbox{and}\qquad &
-\sigma_1\bcdr\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
-\end{eqnarray*}
-%% \noindent \begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcar\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acar\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acdr\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% %\label{eqn:eup-red1}
-%% \end{eqnarray*}
-%% \end{minipage}
-%% \hspace*{.65cm}\begin{minipage}{0.445\textwidth}
-%% \begin{eqnarray*}
-%%     \sigma_1\bcdr\sigma_2 & \hookrightarrow &
-%%     \left\lbrace\begin{array}{ll}
-%%     \sigma_1\sigma_2' & \mbox{if } \sigma_2 = \acdr\sigma_2' \\
-%%     \emptyset & \mbox{if } \sigma_2 = \acar\sigma_2' \,\,\mbox{ or }
-%%     \end{array}\right.\\
-%% &&\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\sigma_2 = \epsilon
-%% \end{eqnarray*}
-%% \end{minipage}\\
-
-The reflexive transitive closure of $\hookrightarrow$ will be denoted
-as $\stackrel{*}{\hookrightarrow}$. The following proposition relates \Lfunonly\   and \Lfunonly$'$:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]  
-\begin{boxedminipage}{\textwidth}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\append~\lista~\listb)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \xtest\ $\leftarrow $\ (\NULLQ~\lista) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\SIF\ \xtest~$\pi_3\!\!:$(\RETURN\ \listb)  
-          \UNL{4} \hspace*{.05cm}     $\pi_4\!\!:\, $(\LET\ \xtl\  $\leftarrow$\   (\CDR\ \lista) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_5\!\!:\, $(\LET\ \xrec\  $\leftarrow$  (\append\ \ \xtl\ \  \listb) \IN
-	  \UNL{6} \hspace*{.05cm}    $\pi_6\!\!:\, $(\LET\ \xhd\  $\leftarrow$  (\CAR\ \lista)  \IN
-	  \UNL{7}  \hspace*{.05cm}   $\pi_7\!\!:\,
-          $(\LET\ \xans\  $\leftarrow$ (\CONS\ \ \xhd\ \ \xrec)  \IN\
-          $\pi_8\!\!:\, $(\RETURN\ \xans))))))))
-	\end{uprogram}}
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pz\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}              (\LET\ \py\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_9\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\append\ \py\ \pz) \IN
-          \UNL{4}   \hspace*{.78cm}  $\pi_{10}\!\!:\, $  (\LET\ \pa\  $\leftarrow$  (\CDR\ \pw) \IN
-	  \UNL{5}   \hspace*{.78cm}  $\pi_{11}\!\!:\, $
-          (\LET\ \pb\  $\leftarrow$ (\CAR\  \pa) \IN~$\pi_{12}\!\!:\,$ (\RETURN\ \pb)))))))
-\end{uprogram}}
-\end{boxedminipage}
-  \caption{An example program.}
-\label{fig:examp-pgm}
-\figrule
-\normalsize
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{proposition}
-Assume that  a liveness computation  based on  \Lfunonly\   gives  the liveness of
-the variable $x$ at a program point $\pi_i$ as $\sigma$ (symbolically, $\Lanv{i}{x}=
-\sigma$).  Further,  let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is
-used instead of \Lfunonly.   Then  $\sigma' \stackrel{*}{\hookrightarrow}
-\sigma$.
-\end{proposition}
-
-To see  why the proposition is  true, consider an analysis  based on
-\Lfunonly$'$ in which   $\sigma$ appears in
-the  context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.  Let $\alpha  \in \sigma$.  The
-symbol  \bcar(\bcdr)  merely  marks  a  place in  $\alpha$  where  the
-original  \CONS\  rule  would  have erased  an  immediately  following
-\acar(\acdr),  or, in  absence of  such a  symbol, would  have dropped
-$\alpha$ itself.  Since the application  of any rule in \Lfunonly$'$ merely
-adds  symbols at  the beginning  of  $\alpha$, the  markers and  other
-symbols in $\alpha$ are propagated to other dependent parts of program
-in their same relative positions.  Consequently, the erasure carried out at the end of the analysis with
-$\stackrel{*}{\hookrightarrow}$ gives the same result as obtained
-through \Lfunonly.
-
-\subsection{Generating equations for the demand transformer \Lfonly}
-\label{sec:Lfonlysymbolically}
-
-%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
-We  shall  consider the  program  in  Figure~\ref{fig:examp-pgm} as  a
-running example;
-this is the  ANF version of the program in Figure~\ref{fig:mot-example}.
-
-%%AM no longer used: In what follows,  we shall overload the semantics of
-%% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-%mean the generation
-%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
-%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.
-
-To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
-$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
-{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
-$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
-formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
-of $x_i$.  For our example program which has  a
-single function \append, this generates the following equations:
-% $\Lfonly_{\append}$  is defined by equations:
-  \begin{eqnarray*}
-    \Lf{\append}{1}{\sigma}
-    &=& \epsilonset \cup \acar\bcar\sigma
-    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
-    \Lf{\append}{2}{\sigma}
-    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
-  \end{eqnarray*}
-  In  general, the  equations  for \Lfonly\  are recursive  since
-  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
-  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
-    is  similar to  solving the  differential equation  
-    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
-    that    the    solution     has    the    form    $\mathit{y}
-    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
-    equation  yields  a  quadratic  equation  in  $r$,  and  each
-    solution of $r$ gives rise  to a solution of the differential
-    equation  (in our  setup we  can effectively  pick  the least
-    solution rather than needing linear combinations).  }
-\begin{eqnarray}
-\label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
-\end{eqnarray}
-where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
-alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
-why \Lfonly\ has this form  is as follows: Recall that
-\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
-the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
-situation in which  the demand $\sigma$ does not  reach the $i$th
-argument.   This happens,  for  instance, when  argument is  used
-only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
-transformation of the demand $\sigma$ before it reaches the $i$th
-argument. This  transformation prefixes strings in 
-$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.
-
-Substituting  the  guessed   form  in  the  equations  describing
-\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
-containing $\sigma$ and those  without $\sigma$ on both sides of the
-resulting equation, we
-get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
-% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
-$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
-solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
-  \append}$, we get:
-  \begin{align*}
-    \Uf{\append}{1} &= \epsilonset
-    \cup \acdr.\Uf{\append}{1}  
-    &\quad 
-    \Uf{\append}{2} &= \Uf{\append}{2} \\
-    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
-    \cup \acdr\Df{\append}{1}\bcdr &\quad 
-    \Df{\append}{2} &= \epsilonset \cup
-    \Df{\append}{2}\bcdr 
-  \end{align*}
-Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.
-
-\subsection{Generating liveness equations \Lv\  for function bodies}
-\label{sec:bodylivenessbodies}
-
-\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
-
-We now calculate a 0-CFA-style summary liveness for each GC point of a program.
-There are two parts to this.
-First, for each function $f$,  we determine a {\em summary demand} $\sigma_f$
-over-approximating any demand $\sigma$ passed to $f$.
-%$\Lf{f}{i}{\sigma}$.
-So, consider a function $g$ containing a call site $\pi$ to $f$, say
-$\pi\!\!:\!(\LET~x \leftarrow (f\,y_1\,\ldots\,y_n)~\IN~e)$.
-Under the assumption that the demand on $g$ is $\sigma_g$, the liveness at $e$
-is $\Lv = \mathcal{L}(e,\sigma_g,\Lfonly)$, and the $\LET$ case of
-Figure~\ref{fig:live-judge} tells us this call site contributes $\Lv(x)$ to the
-demand $\sigma_f$ placed on  $f$.
-Write $\deltacall{f}{\pi}{g}$ for
-this demand (contributed by call site $\pi$ in $g$).
-
-Thus, supposing the $k$ call sites to function $f$ are
-$\pi^1$ (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
-%%% \marginpar{and writing $\sigma_f^i$ for the above $\deltacall{f}{\pi^i}{g^i}$?}
-then the over-approximation requirement on $\sigma_f$
-is achieved by taking $\sigma_f = \deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
-% is achieved by taking $\sigma_f = \sigma_f^1 \cup \cdots \cup \sigma_f^k$.
-
-The expression $e_{\mainpgm}$ is a special case; we assume it may be called externally
-with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$ (denoted
-$\sigma_{\!all}$).
-This is because any  part  of its  value  may be  used  by  the environment---for
-printing  the  result,  for  instance.
-
-% In what follows,  we shall overload the semantics of
-% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
-% 
-% 
-% We assume an overall demand of $\{\acar,\acdr\}^*$ (denoted
-% $\sigma_{\!all}$) on the main expression $e_\mainpgm$.  This is because 
-% any  part  of its  value  may be  used  by  the environment---for
-% printing  the  result,  for  instance.
-
-%% To  generate  equations for  the  body $e_{\mathit{f}}$ of  a
-%% function,  we evaluate \Lfun{e_{\mathit{f}}}{\sigma_{\mathit{f}}}{\Lfonly}.     For ${\sigma_{\mathit{f}}}$,  we have
-%% to consider all calling  contexts during execution.  We therefore
-%% do a 0-CFA approximation by  defining
-%% % $\sigma_\mathit{f}$        as $\bigcup_\calls{f}{k}\!\!\sigma_\calls{f}{k}$,
-%% $\sigma_\mathit{f}$        as $\sigma_\calls{f}{1} \cup \cdots \cup \sigma_\calls{f}{n}$,
-%% where $\calls{f}{k}$,
-%% $\mathit{k} = 1 \ldots n$ represent all the calls to $\mathit{f}$
-%% in  the  program,   and  $\sigma_\calls{f}{k}$  are  the  demands
-%% generated at the calls  by the \rn{let-live} rule while analyzing
-%% the bodies of the caller.
-
-For the running example, $\append$ has calls from $\mainpgm$ at $\pi_9$
-and a recursive call at $\pi_5$.
-So $\sigma_{\append} =
-     \deltacall{\append}{\pi_9}{\mainpgm}  \cup \deltacall{\append}{\pi_5}{\append}$.
-Filling in  the values gives:
-%  where the former is the demand generated at the
-%  call to \append\  in $e_\mainpgm$ the latter
-%  is  the demand  at the  recursive call. 
-%% Propagation of
-%%  $\sigma_{\!all}$ in $e_\mainpgm$ gives    $\sigma_\calls{\append}{1}$   as   $\{\epsilon,   \acdr\}   \cup
-%%  \acdr\acar\sigma_{\!all}$  and propagation of
-%%  $\sigma_{\mathit{\append}}$ in $e_\append$ gives   $\sigma_\calls{\append}{2}$   as
-%%  $\bcdr\sigma_{\mathit{\append}}$.  
-%% Filling in  the values of $\sigma_\calls{\append}{1}$ and
-%% $\sigma_\calls{\append}{2}$, we get:
-\begin{align*}
-\sigma_{\append}    &=
-\left( \{\epsilon,  \acdr\} \cup  \acdr\acar\sigma_{\!all}  \right) ~\cup~
-\bcdr\sigma_{\mathit{\append}}
-\end{align*}
-
-Second, for each function $f$ (possibly $\mainpgm$) we need the liveness at
-each contained GC point
-$\pi\!\!:\!(\LET~x \leftarrow s~\IN~e)$,
-where $s$ is a call either to $\CONS$ or to a user function $g$.
-Given $\sigma_f$ calculated above, this
-is simply $\mathcal{L}(\pi,\sigma_f,\Lfonly)$.
-%%% \marginpar{AM: we've always said this (safe), but actual parameters
-%%% live before the call and dead after needn't be included!}
-
-For the running example, containing GC points $\pi_5, \pi_7$ in $\append$ and
-$\pi_9$ in $e_\mainpgm$, this gives
-(recall Equation~(\ref{eq:LF:DI}) above states $\Lf{f}{i}{\sigma} = \Uf{f}{i} \cup \Df{f}{i}.\sigma$):
-\begin{align*}
-\Lanv{5}{\xtl} &= \Lf{\append}{1}{\bcdr\sigma_{\append}}  \quad &
-                                \Lanv{7}{\xhd}\,\,\, &=  \bcar\sigma_{\append}     \\ 
-\Lanv{5}{\listb} &= \Lf{\append}{2}{\bcdr\sigma_{\append}} &
-                                \Lanv{7}{\xrec} &=  \bcdr\sigma_{\append} \\
-\Lanv{5}{\lista} &= \epsilonset \cup \acar\bcar\sigma_{\append} 
-\\[1ex]
-  \Lanv{9}{\py}  &= \Lf{\append}{1}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}} \\
-  \Lanv{9}{\pz}  &= \Lf{\append}{2}{\{\epsilon,
-    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}}
-\end{align*}
-
-\noindent 
-In summary, the equations generated during liveness analysis are:
-\begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
-\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
-% {\color{red}
-%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
-%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
-% }
-\end{enumerate}
-
-\subsection{Solving liveness equations---the grammar interpretation}
-The  equations above can now be re-interpreted as a
-context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
-a variable  $X$ occurring on  the LHS of the  equations generated
-from the analysis.  We can  think of the resulting productions as
-% being a grammar with multiple start symbols---one for
-% for each nonterminal \var{\Lanv{i}{x}}.
-being associated with several  grammars, one for each non-terminal
-\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
-productions:
- \begin{eqnarray*}
-\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
-  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
- \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
-  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
-  \acdr\var{\Uf{\append}{1}}\\         
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar               \mid
-  \acdr\var{\Df{\append}{1}}\bcdr\\
-\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
-\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
-\var{$\sigma_{\mathit{\!all}}$}
-\end{eqnarray*}
-Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
-denoted $\mathscr{L}(\var{\Lanv{i}{x}})
-$,  is the desired  solution of
-\Lanv{i}{x}.   
-However, recall from  our earlier  discussion that
-the  decision problem that  we are  interested in  during garbage
-collection is:
-\begin{quote}
-Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
-\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
-\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
-of forward paths only. Then does $\alpha \in \sigma$?
-\end{quote}
-We could convert  the  rules  defining  $\hookrightarrow$  into
-productions and  add them to the grammar. However, this results
-in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is  undecidable.   We
-circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
-  strongly  regular}  CFGs which  have   easy
-translations to non-deterministic finite state automata (NFA). The
-NFAs are then simplified on the
-basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
-membership of forward access paths. The resulting NFAs are   finally
-converted to DFAs for use during garbage collection.
-%------------------------------------------------------------%
-\subsubsection{Approximating CFGs using NFAs.}
-\label{sec:NFA-approx}
-
-We use the   
-algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
-approximate a CFG  to a   {\em strongly regular\/} grammar.  The
-transformation has  the  property that  if $\Lv$  is a
-non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
-Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
-  L}_{G'}(\Lv)$.    This  is   required  for the
-approximation to be safe with respect to liveness.
-
-We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
-transformation. We  pick the only production that  is affected by
-the transformation---the production for \Df{\append}{1}.  The
-production for \Uf{\append}{1}, while  recursive, is already in
-strongly  regular  form  and   is  therefore  unaffected  by  the
-transformation.
- \begin{eqnarray*}
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
-  \acdr\var{\Df{\append}{1}} \\ 
-\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
-\mid \epsilon 
-\end{eqnarray*}
-The  languages
-generated for  \var{\Df{\append}{1}} in  the original grammar  and the
-new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
-\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
-
-%------------------------------------------------------------%
-\subsubsection{Transforming NFAs to Accept Forward Paths:}
-%\label{sec:NFA-elim}
-The strongly regular CFGs obtained after the Mohri-Nederhof
-transformation are
-  first converted  into  NFAs.  The        algorithm         described        in
-Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
-\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
-Thus  \nt\  can  be  used  to  check  
-membership of forward paths.
-
-\begin{figure}[t]
-\begin{boxedminipage}{\textwidth}
-\begin{center}
-  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
-  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
-%
-  \raggedright{\bf Output:} NFA  \nt\ with  underlying
-  alphabet $\lbrace \acar, \acdr \rbrace$ such that
-  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
-  \\
-%
-
-  \raggedright{\bf Steps:}
- \begin{algorithmic}
-   \STATE $i \leftarrow 0$ 
-   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-   without $\epsilon$-moves \cite{hopcraft90toc}
-
-   \REPEAT
-   \STATE $\nt'_{i+1} \leftarrow \nt_i$
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcar$ and
-     outgoing edge to $q''$ with label $\acar$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-   label $\epsilon$.
-   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-   \ENDFOR
-
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcdr$ and outgoing edge to
-     $q''$ with label $\acdr$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-     label $\epsilon$.
-   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-   \ENDFOR
-
-   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-   $\nt'_{i+1}$ without $\epsilon$-moves
-
-   \STATE $i \leftarrow i+1$
-   \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-   \STATE $\nt \leftarrow \nt_i$
-%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
- \end{algorithmic}
-\end{center}
-\end{boxedminipage}
- \caption{Algorithm for transforming an NFA to  accept forward
-   paths only.}\label{algo:simplify-NFA} \figrule
-\end{figure}
-%------------------------------------------------------------%
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
-pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
-process is  continued till  a fixed point  is reached.   When the
-fixed point  is reached, the resulting NFA  contains all possible
-reductions corresponding  to all the  paths in the  original NFA.
-The proofs  of the termination  and correctness of  the algorithm
-are given in our earlier paper~\cite{karkare07liveness}.
-
-\input{liveness-dfa.tex}
-
-We illustrate the algorithm in Figure~\ref{algo:simplify-NFA} by
-constructing the automaton for \var{\Lanv{9}{y}}.
-Figure~\ref{fig:automata-construction} (c) shows the automaton for
-\var{\Lanv{9}{y}} constructed by composing the automata for
-\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
-\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
-get (d). We add an $\epsilon$ edge from $q_2$ to $q_3$ bypassing the
-\bcdr\acdr\ pair from $q_2$ to $q_2$ and then to $q_3$. This is shown
-in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
-is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
-$q_2$.  Removing this $\epsilon$ edge gives the automaton shown in
-(g).  Restricting this automaton to forward edges only, we get the
-final automaton shown in (h). This automaton recognizes $\acdr^{*}\!\mid\acdr^{*}\acar \sigma_{\mathit \!all}$, showing that the entire
-list \py, including its elements, is live at $\pi_9$. Also note that
-the language accepted by the final automaton satisfies the
-prefix-closed property.
-
-\section{Prototype and evaluation}\label{sec:experiments}
-To demonstrate the effectiveness of liveness-based garbage collection,
-we  have  built a  prototype  consisting  of  an interpreter  for  our
-language,  a  liveness  analyzer  and  a copying  collector  that  can
-optionally use the results of liveness analysis for marking instead of
-reachability. When the collector uses liveness for marking, we call it
-a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-  reachability-based  collector}  (RGC).   The  copying  collector  is
-neither incremental nor generational.  As a consequence, any cell that
-becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.
-
-While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
-is invoked  at the program point   $\pi$,  it would  use  the
-liveness  DFAs at  $\pi$ itself for exploring the root set of the
-current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.
-
-Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-$\dfa{\pi}{x}$.    
-Considering   a   DFA   as   a   table,
-$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-\acar\      or     \acdr.       We      shall     also      write
-$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-indicating  whether there  is a  transition from  $q$ on
-$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-pointer to $x$.
-The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-$x$  in the  other semi-space  and returns  the new  address.  It
-hides  details such as  returning the  forwarding pointer  if the
-value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.
-
-The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-in the heap over time for RGC and LGC---here time is measured in terms
-of the number of cons cells allocated. In addition, they also show the
-number of  reachable cells and the  number of cells  that are actually
-live\footnote{This   is  statically   approximated  by   our  liveness
-  analysis}.  Since  the programs have  different memory requirements,
-we have tuned the size of heap for each program to ensure a reasonable
-number of  collections. An invocation  of RGC decreases the  number of
-cells  in heap  till  it touches  the  curve of  reachable cells.   An
-invocation of LGC decreases the number  of heap cells to no lower than
-the curve of live cells.
-
-To construct the  reachable and live curves, we  record for every cell
-its creation time (\CreateTime), its last use time (\UseTime), and the
-earliest time  when the  cell becomes unreachable  and can  be garbage
-collected (\CollTime).  For accurate recording of  \CollTime, we force
-frequent invocations  of a reachability based collector  in a separate
-run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
-  fibheap} increases  initially, comes  down before the  heap becomes gets
-full and then  continues to remains low.  This results in  a single garbage
-collection in both RGC and LGC.  The graphs also show the precision of
-our liveness analysis.  For  all programs  except  {\tt nperms} and
-{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.
-
-\subsection{Results}
-\begin{figure}[t] 
-\centering
-\begin{boxedminipage}{.75\textwidth}
-{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
-\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-  then} {\bf skip}\\ 
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
-\noindent \hspace*{.35cm}{\bf return} $y$
-\end{boxedminipage}
-\caption{Function       for       copying       a       root       set
-  variable.}\label{algo:LGC-main-loop} \figrule
-
-\end{figure}
-
-\begin{figure}[p]
-\begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
- &
-{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-\end{tabular}
- \caption{Memory  usage of  programs. The blue and the red curves
-  indicate the number of cons cells in the active semi-space for RGC and
-  LGC respectively. The black curve represents the number of reachable
-  cells and the grey curve represents the number of cells that are
-  actually live (of which liveness analysis does  a static
-  approximation). x-axis is the time measured in number of cons-cells
-  allocated. y-axis is the number of cons-cells.}
-\label{fig:memory-usage} \figrule
-\end{figure}
-The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-provides  statistics  regarding the  analysis  itself.  The number  of
-states and the analysis times  are within tolerable limits. Precision
-of analysis refers to the percentage of dead cells that is collected by
-LGC, averaged over all invocations.
-The second table shows garbage  collection statistics for RGC and LGC.
-LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to run
-in comparison with  RGC.
-
-There are a couple of  issues of concern.  The garbage collection time
-is larger in the case of LGC for some programs. The reason is that the
-cost of consulting the liveness DFA may outweigh the combined benefits
-of  fewer   garbage  collections   and  fewer  markings   per  garbage
-collection.   The  other issue  is  illustrated  by  the program  {\tt
-  lambda}    (memory   usage  shown   separately    in
-Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-in Figure~\ref{fig:experimental-results}, the  number of
-touched cells\footnote{These  are the  cells which are  visited during
-  marking phase, often  more than once due to  sharing.} in this
-example is much higher for
-LGC.  This  increase is due to  excessive sharing among  heap nodes in
-this program.  Note  that a node re-visited because  of sharing is not
-explored  any   further  during  a  RGC   collection.   However,  this
-curtailment cannot happen in  LGC because of the possibility that the
-node,  re-visited in a  different liveness  state, may  mark a  set of
-cells different from the earlier visit.
-\begin{figure}[t]
-\footnotesize
-
-\centering
-
-\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline
-{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-  nperm} &  {\sf fibheap} & {\sf knightstour} &
-{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-\hline
-\hline
-{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-\hline
-\end{tabular}}
-
-\centerline{(a)}
-
-\medskip
-
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
-  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-
-                            &   \multicolumn{2}{c|}{cells per GC}  
-                            &   \multicolumn{2}{c|}{cells per GC} 
-                            &   \multicolumn{2}{c|}{\#GCs} 
-                            &   \multicolumn{2}{c|}{MinHeap} 
-                            &   \multicolumn{2}{c|}{Avg. Drag} 
-                            &   \multicolumn{2}{c|}{(sec)} \\
-\cline{2-13}
-{Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
-%    {\sf  nboyer} & &  & &   & &  & & &  \\
-%    {\sf   circsim?} & &  & &   & & & & &   \\
- \hline
-\end{tabular}}
-
-\centerline{(b)}
-\vskip -2mm
-\caption{Experimental results comparing RGC and LGC. Table (a) gives
-  data related to liveness analysis, and (b) gives garbage collection
-  data. The average drag time is measured in allocated cons cells.}
-\label{fig:experimental-results}
-\normalsize
-\end{figure}
-
-\section{Collecting more garbage can never slow things down}
-\label{sec:lgc-always-better}
-Since garbage collection is  effectively asynchronous to the allocator
-thread, one might  worry as to how {\em  robust} our measurements are.
-For example,  while LGC would,  in general, collect more  garbage than
-RGC in the same heap state,  is it possible that on some programs, LGC
-might do  a larger number  of collections? We  prove a result  to show
-that   this  cannot   happen.    This  result   applies  to   classical
-mark-and-sweep and copying garbage  collectors and, we hypothesize, to
-generational collectors too.
-
-\begin{lemma}
-For the  same mutator,  a liveness-based collector  can never  do more
-garbage collections than a reachability-based collector.
-\end{lemma}
-
-\begin{proof}Assume, as before,  that time is measured in terms of the number of
-  cons cells allocated. Let us run  two copies of the mutator with the
-  two    different   collectors    but    with   synchronous    memory
-  allocations.   The   garbage    collections,   however,   happen
-  asynchronously.
-
-  To prove the lemma, it is enough to show the truth of the following 
-statement: After
- every  LGC invocation,  the count of LGC invocations   is  no greater
-than  RGC invocations.   The  base case  holds  since the  first
-invocations of both GCs happen at the same time.  Assume the statement
-to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-reachable cells, its heap would contain no more cells than RGC heap at
-the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-LGC, or LGC and RGC are both invoked next at the same time.  In either
-case, the statement holds after $n+1$ invocations of LGC.
-\end{proof}
-
-%% \begin{figure}[h]
-%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
-%% \caption{Memory  usage of  programs. The  solid  lines represents
-%%   number of  live cells and the  dashed lines show  the number of
-%%   reachable cells.  The red and blue  lines represent invocations
-%%   of RGC and  LGC---their heights indicate  the amount of memory
-%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
-%%       paper and are being used as placeholder}}}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-%%\clearpage
-
-\section{Related Work} Previous attempts to  increase  the space efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
-reallocation~\cite{jones89compile},                       deforestation
-techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
-methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
-
-\begin{figure}[t]
-\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
- \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
-\label{fig:memory-usage-lambda} \figrule
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-We have defined a notion of liveness on structured data;
-this generalizes classical liveness and strong liveness.
-We started with a general fully context-sensitive analysis which
-we proved correct with respect to a minefield semantics
-(this models the effect of garbage collection between every
-evaluation step).
-
-To avoid scalability issues (and to avoid performing part of
-the liveness computation at run time)
-we defined an 0-CFA version of this liveness analysis in which
-demands for function $f$ at all calling contexts are conflated into 
-a single demand $\sigma_f$.
-This enabled us to treat the liveness equations symbolically obtaining
-context-free grammars for liveness at each GC point (calls to user
-functions and to $\CONS$).
-These were then converted to DFAs for run-time consultation by the
-garbage collector. Experiments confirm the precision of the analysis.
-
-To obtain performance figures we compared a reachability-based garbage
-collector with a liveness-based  collector.  This showed a decrease in
-the  number  of  GCs,   more  garbage  collected  per  invocation.   A
-significant benefit of LGC is  that programs can run in smaller memory
-when compared to  RGC. This is potentially useful  in situations where
-memory  is  limited---as with  embedded  systems.  For  a majority  of
-programs, the garbage collection times were reduced.
-
-One issue we  highlighted was that while fewer  nodes were marked (and
-more garbage collected), sometimes  cons cells could be visited and
-traversed  multiple times  with different  sets of  liveness  paths to
-explore.  Further work includes  improvements to the classical copying
-collector to reduce the cost of this.
-
-\subsection*{Acknowledgements}
-The authors  would like  to thank Hemanshu  Vadehra for  a preliminary
-implementation of the prototype system. Amey Karkare was supported for
-this work by the DST/SERC fast track scheme for young scientists.
-
-\bibliography{fun_hra}{}
-\bibliographystyle{splncs}
-
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-\section{Appendix stuff}
-
-\paragraph{Safety of using the modified \CONS\  rule:}
-Recall that while the correctness of the analysis was proved with
-respect to  the $\len$ rules  in Figure~\ref{fig:live-judge}, the
-equations   for  liveness   are  based   on  a   modification  of
-$\mathit{ref}$    in    the    case    of   \CONS\  as  given    by
-(\ref{eqn:mod-cons}).  We now show that  if the \bcar\,s and \bcdr\,s
-in live access  paths resulting  from this  modification are
-eliminated through  applications of $\hookrightarrow$,  we obtain
-the same results as given by the original rules.
-
-\comment{Proposal: We prove a result which goes as follows:
-
-\begin{quote}
-Lemma: Let  $\Lanv{i}{x}= \sigma$ for an a  program point $\pi_i$
-and a variable  $x$, when the judgement is  based on the original
-$\len$  rules.   Further, let  $\Lanv{i}{x}=  \sigma'$, when  the
-judgement is based on $\len$ with the modified \CONS\ rule.  Then
-$\sigma' \stackrel{*}{\hookrightarrow} \sigma$
-\end{quote}
-
-Here are some  ideas regarding how this can  be proved.  Consider
-an arbitrary equation
-$$ \Lanv{j}{y} = \ldots \Lanv{i}{x} \ldots$$ generated by the
-analysis.
-
-Now assume that $\Lanv{i}{x}= \sigma$.   We want to find out what
-happens to the access patterns in $\sigma$ in the access patterns
-of \Lanv{j}{y}.   To answer this,  we put \# markers  around each
-$\alpha \in \sigma$  and redo the equation using  an extension of
-$\len$ to take  care of \#-tagged patterns.  We  can then observe
-occurrences of \#-tagged patterns  in the liveness information at
-$\pi_j$.
-
-Here is  the $\len$ extension for  the original \CONS\  clause to take
-care of the \#-tagged patterns:
-$$\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          = \{x.\#\alpha\# \mid \#\acar\alpha\# \in \sigma\} \cup
-\{y.\#\alpha\# \mid \#\acdr\alpha\# \in \sigma\} $$
-And here  is  the $\len$ extension for  the modified \CONS\  clause.
-$$ \mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-= \{x.\#\bcar\alpha\# \mid \#\alpha\# \in \sigma\} \cup
-\{y.\#\bcdr\alpha\# \mid \#\alpha\# \in \sigma\} $$
-All  other rules  remain unchanged.  Tracking a  pattern  through this
-extension is sound because erasing the \#-tags would give us the same
-liveness as the original rules.
-
-Let us also extend $\hookrightarrow$ to \#-tagged strings as:
-$$\sigma_1  {\hookrightarrow} \sigma_2\,\, \mathit{iff}\,\, \sigma_2 =
-\{\alpha_1 \#\alpha\# \alpha_2 \mid  \alpha_1 \#\alpha'\# \alpha2 \in
-  \sigma_1\,\, and \,\, \{\alpha'\} \stackrel{*}{\hookrightarrow} \{\alpha\}\}$$
-
-Now we have an  observation represented by the following diagram:
-
-\centerline{\epsfig{file=diag1.eps, height=4cm}}
-
-We need  to make this  observation general enough.   Therefore we
-should prove  the observation even if \Lanv{i}{x}  = $\sigma$ had
-patterns with \bcar\ and \bcdr.   We should be able to prove this
-by structural induction on the rules.
-
-How does  the observation prove  the original lemma?   Suppose we
-replayed  judgements  with  ($\len$  +  \#) and  with  ($\len$  +
-modified \CONS +  \#).  Suppose that $\pi_i$ is  the last program
-point till which the results  are the same (because no \CONS\ was
-encountered).  Now  consider a  $\sigma \in \Lanv{i}{x}$  for any
-$x$.  Put \#-tags around each string in $\sigma$. Now we have the
-diagram:
-
-\centerline{\epsfig{file=diag2.eps, height=7cm}}
-
-This shows that modulo erasure of \#-tags, using the original $\len$
-judgement and  $\len$ judgement with modified \CONS\ rule followed by
-$\hookrightarrow$ give the same liveness.
-
-And I would like  to conclude that it is all right  to do our analysis
-with $\len$ + modified \CONS\ and then simplify using $\hookrightarrow$.
-
-}
-
-{\color{blue}\marginpar{From sec 3.1, liveness judgements}
-\paragraph{Semantic considerations.}
-The incremental liveness for generic primitive $\PRIM$ is given by
-$\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly) = 
-           \{x.\epsilon,y.\epsilon\}$.
-Earlier we suggested that classical strong liveness for addition,
-but not division, could use the stronger
-$ \{x.\epsilon,y.\epsilon \mid \epsilon \in \sigma\}$.
-However this is invalid: garbage collection
-could reuse dead, but reachable, $\CONS$ cell fields (of $v$ below)
-thereby producing a stuck state in a dynamically typed language.
-Hence we treat $+$ as like division not like addition---suppose
-% $(FOO)$ is a user function which may cause garbage collection, and consider:
-calling $\mbox{\footnotesize (FOO)}$ garbage collects and reallocates the cell pointed
-to by `v' in:
-
-\vspace{-15pt}
-
-\begin{quote}
-\footnotesize
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1}  (\LET\ v\ $\leftarrow $\ (\CONS~2~3) \IN
-	  \UNL{2}  (\LET\ w\ $\leftarrow $\ (FOO) \IN
-	  \UNL{3}  (\LET\ x\ $\leftarrow $\ (\PRIM\ (\CAR\ v) 4) \IN
-	  \UNL{4}  (\RETURN\ w))))
-	\end{uprogram}}
-\end{quote}
-In principle, {\em dead code removal} of
-$(\PRIM\ (\CAR\ {\rm v})\ 4)$
-would render the stronger version valid, but still requires static typing
-to avoid the risk of {\em removing} a stuck state.
-We restrict ourselves to liveness annotation rather than code transformation.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]
-\footnotesize
-Old version of the liveness rules...
-
-\infrule[ret-live]
-        {}
-        {(\RETURN~x),\sigma,\Lfonly \len  x.\sigma}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[if-live]
-        { e_1, \sigma,\Lfonly \len \Lv_1
-          \andalso e_2, \sigma,\Lfonly \len  \Lv_2}
-        {(\SIF~x~e_1~e_2),\sigma,\Lfonly \len
-               \Lv_1\cup\Lv_2\cup \{x.\epsilon\}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\infrule[let-live]
-        {e, \sigma,\Lfonly  \len  \Lv}
-        {(\LET~x \leftarrow  s~\IN~e),
-          \sigma,\Lfonly \len   \Lv\setminus \{x.*\} \cup
-          \mathit{ref\/}(s,\Lv(x),\Lfonly)}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\infrule[live-define]
-        {e_f, \sigma, \Lfonly  \len
-%            \bigcup \{z_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-           \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \fen \Lfonly
-\\ \mbox{where }
-     (\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ \ e_f) \mbox{ is a member of }
-        d_1 \ldots d_k}
-\figrule
-  \caption{Liveness Judgement Rules}\label{fig:live-judge-old}
-\normalsize
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \newgray{lightgray}{.75}
-%\newcommand{\dem}[1]{\red{\scriptsize{$#1$}}}
-\newcommand{\dem}[1]{\psframebox[linestyle=none,fillstyle=none,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-%\newcommand{\liv}[1]{\blue{\scriptsize{$#1$}}}
-\newcommand{\liv}[1]{\psframebox[linestyle=none,fillstyle=solid,fillcolor=lightgray]{\rule[-1pt]{0pt}{8pt}{\scriptsize $#1$}}\hspace*{1pt}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-\psset{unit=1mm,nodesep=1mm,labelsep=0.5mm,arrows=->,linewidth=.1,framesep=0.2mm}
-\begin{center}
-  \begin{pspicture}(0,0)(120,50)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(12,5)
- \footnotesize
-   \putnode{let1}{origin}{15}{50}{$\pi_1$:\LET}
-
-    \putnode{test1}{let1}{-12}{-7}{\xtest}
-    \putnode{null}{let1}{0}{-7}{$\pi_2$:\NULLQ}
-    \putnode{lista1}{null}{0}{-7}{$\pi_3$:\lista}
-    \putnode{if}{let1}{18}{-7}{$\pi_4$:\SIF}
-
-    \putnode{test2}{if}{-8}{-7}{$\pi_5$:\xtest}
-    \putnode{listb1}{if}{3}{-7}{$\pi_6$:\listb}
-    \putnode{let2}{if}{14}{-7}{$\pi_7$:\LET}
-
-    \putnode{tl1}{let2}{-22}{-7}{\xtl}
-    \putnode{cdr}{let2}{-5}{-7}{$\pi_8$:\CDR}
-    \putnode{lista2}{cdr}{0}{-7}{$\pi_9$:\lista}
-    \putnode{let3}{let2}{15}{-7}{$\pi_{10}$:\LET}
-
-\putnode{rec1}{let3}{-13}{-7}{\xrec}
-\putnode{append}{let3}{0}{-7}{$\pi_{11}$:\append}
-\putnode{tl2}{append}{-10}{-7}{$\pi_{12}$:\xtl}
-\putnode{listb2}{append}{5}{-7}{$\pi_{13}$:\listb}
-
-\putnode{let4}{let3}{17}{-7}{$\pi_{14}$:\LET}
-\putnode{hd1}{let4}{-5}{-7}{\xhd}
-\putnode{car}{let4}{4}{-7}{$\pi_{15}$:\CAR}
-\putnode{lista3}{car}{0}{-7}{$\pi_{16}$:\lista}
-
-\putnode{let5}{let4}{16}{-7}{$\pi_{17}$:\LET}
-
-\putnode{ans1}{let5}{-4}{-7}{\xans}
-\putnode{cons}{let5}{6}{-7}{$\pi_{18}$:\CONS}
-\putnode{hd2}{cons}{-10}{-7}{$\pi_{19}$:\xhd}
-\putnode{rec2}{cons}{10}{-7}{$\pi_{20}$:\xrec}
-\putnode{ans2}{let5}{19}{-7}{$\pi_{21}$:\xans}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-
-%% \putnode{x}{let1}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{if}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let2}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{let3}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{let4}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{ans2}{-8}{0}{\liv{\Lv_1}}
-%% %\putnode{x}{ans2}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{plus}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{rec2}{1}{4}{\dem{\epsilonset}}
-%% \putnode{x}{plus}{-7}{0}{\liv{\Lv_6}}
-%% %\putnode{x}{let4}{-7}{0}{\liv{\{\xrec\}}}
-%% \putnode{x}{len}{3}{3}{\dem{\epsilonset}}
-%% \putnode{x}{len}{-7}{0}{\liv{\Lv_3}}
-%% \putnode{x}{tl2}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{tl2}{-6}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let3}{-10}{0}{\liv{\xtl.\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{0}{4}{\dem{\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{cdr}{-7}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l2}{0}{4}{\dem{\epsilonset \cup \acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{l2}{-5}{0}{\liv{\Lv_4}}
-%% \putnode{x}{zero}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{zero}{-3}{0}{\liv{\Lv_0}}
-%% \putnode{x}{test2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{test2}{-7}{0}{\liv{\Lv_5}}
-%% \putnode{x}{if}{-6}{0}{\liv{\Lv_3}}
-%% \putnode{x}{null}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{null}{-9}{0}{\liv{\Lv_4}}
-%% \putnode{x}{l1}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{vone2}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{one}{0}{3}{\dem{\epsilonset}}
-%% \putnode{x}{let5}{0}{3}{\dem{\sigma}}
-%% \putnode{x}{l1}{-5}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{let1}{-17}{-1}{\liv{\{\xl\} \cup \xl.\acdr\Lf{\xlen}{1}(\epsilonset)}}
-%% \putnode{x}{let1}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let1}{-7}{0}{\liv{\Lv_4}}
-%% %\putnode{x}{if}{7}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let2}{8}{0}{\liv{\Lv_0}}
-%% %\putnode{x}{let3}{8}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let3}{-7}{0}{\liv{\Lv_3}}
-%% %\putnode{x}{let4}{9}{0}{\liv{\Lv_0}}
-%% \putnode{x}{let4}{-7}{0}{\liv{\Lv_2}}
-%% \putnode{x}{vone2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{rec2}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{let5}{-8}{0}{\liv{\Lv_6}}
-%% \putnode{x}{one}{-6}{0}{\liv{\Lv_2}}
-\ncline{-}{let1}{test1}
-\ncline{-}{let1}{null}
-\ncline{-}{let1}{if}
-\ncline{-}{null}{lista1}
-
-\ncline{-}{if}{test2}
-\ncline{-}{if}{listb1}
-\ncline{-}{if}{let2}
-\ncline{-}{let2}{tl1}
-\ncline{-}{let2}{cdr}
-\ncline{-}{let2}{let3}
-\ncline{-}{cdr}{lista2}
-
-\ncline{-}{let3}{rec1}
-\ncline{-}{let3}{append}
-\ncline{-}{let3}{let4}
-
-\ncline{-}{append}{tl2}
-\ncline{-}{append}{listb2}
-
-\ncline{-}{let4}{let5}
-\ncline{-}{let4}{hd1}
-\ncline{-}{let4}{car}
-\ncline{-}{car}{lista3}
-%% \ncline{-}{len4}{tl2}
-%% \ncline{-}{let4}{vone1}
-%% \ncline{-}{let4}{one}
-%% \ncline{-}{let4}{let5}
-\ncline{-}{let5}{ans1}
-\ncline{-}{let5}{cons}
-\ncline{-}{cons}{hd2}
-\ncline{-}{cons}{rec2}
-%% \ncline{-}{let5}{plus}
-\ncline{-}{let5}{ans2}
-\end{pspicture}
- \end{center}
-
-  \caption{Example to illustrate liveness computation}\label{fig:live-examp}
-\figrule
-\normalsize
-\end{figure}
-
-\newpage
-
-\begin{figure}[t]
-  \begin{tabularx}{320pt}{|| c | X | X | X ||}
-\hline
-Pgm  Pt & Demand  & Pre & Post \\
-\hline 
-$\pi_1$ & $\sigma$ & \{\lista\} $\cup$ \listb.$\sigma$ $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & $\phi$ \\
-\hline
-%% $\pi_2$ & $\epsilonset$ & \{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_3$ & $\epsilonset$ &\{\lista\} $\cup$ \listb.$\sigma$  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_4$ & $\sigma$ &\{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-%% $\pi_5$ & \epsilonset & \{\xtest\} $\cup$ \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \listb.$\sigma$ $\cup$ \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  \\
-%% \hline
-%% $\pi_6$ & $\sigma$ & \listb.$\sigma$ & $\phi$ \\
-%% \hline
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_7$ & $\sigma$  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$\\
-%% \hline
-%% $\pi_8$ & \Lf{\append}{1}(\{\bcdr\}$\sigma$) & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-%% \hline
-$\pi_9$ & \epsilonset $\cup$ \{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  & \{\lista\}  $\cup$ \lista.\{\acdr\}\Lf{\append}{1}(\{\bcdr\}$\sigma$)  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)   & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% $\pi_{10}$ &$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  & $\phi$ \\
-%% \hline
-$\pi_{11}$ & \{\bcdr\}$\sigma$ & \{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$) & \lista $\cup$ \lista.\{\acar\bcar\} \\
-\hline
-$\pi_{12}$ &\Lf{\append}{1}(\{\bcdr\}$\sigma$)   &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-$\pi_{13}$ & \Lf{\append}{2}(\{\bcdr\}$\sigma$)  &\{\lista\}  $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xtl.\Lf{\append}{1}(\{\bcdr\}$\sigma$) $\cup$  \listb.\Lf{\append}{2}(\{\bcdr\}$\sigma$)  &  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{14}$ & $\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$ \\
-\hline
-%% $\pi_{15}$ & \{\bcar\}$\sigma$  & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   &  \xrec.\{\bcdr\}$\sigma$\\
-%% \hline
-%% $\pi_{16}$ & \epsilonset $\cup$ \{\acar\bcar\}$\sigma$ & \{\lista\} $\cup$ \lista.\{\acar\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$  & \xrec.\{\bcdr\}$\sigma$ \\
-%% \hline
-$\pi_{17}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$   & $\phi$  \\
-\hline
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-$\pi_{18}$ & $\sigma$ & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ & $\phi$  \\
-\hline
-$\pi_{19}$ &  \{\bcar\}$\sigma$        & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{20}$ &   \{\bcdr\}$\sigma$       & \xhd.\{\bcar\}$\sigma$ $\cup$ \xrec.\{\bcdr\}$\sigma$ &   \\
-\hline
-$\pi_{21}$ & $\sigma$ & \xans.$\sigma$ & $\phi$  \\
-\hline
-\end{tabularx}
-  \caption{Liveness annotations for the body of \append}\label{fig:xe_append}
-\figrule
-\normalsize
-\end{figure}

### 2014-11-10 paper/sections/liveness.tex
pushing liveness
Updated demand transformation
Updated Liveness Rules

@@
 \section{Liveness}\label{sec:liveness}
+%\comment{
+To describe the issues that arise while formulating liveness of a lazy
+language, consider the example shown in Figure ???
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{verbatim}
+pi_0: let pi_1: x = (cons 1 x)
+      in let  pi_2: y = (cdr x)
+         in  let  pi_3: z = (car y)
+             in pi_4: if (z > 0) then 
+                       pi_5: let pi_6: w = (take 5 x)
+                             in let pi_7: u = (cdr  w)
+                                in l1:return u
+                      else l2:return y
+
+\end{verbatim}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The program is annotated with symbols standing for program points. For
+reasons that will  become clear later, we use  different symbols for
+the program points corresponding to return points (l instead of $\pi$)
+Similar  to classical  liveness  analysis for  imperative programs,  a
+demand  of  $\sigma$ on  the  expression  propagates  to a  demand  of
+$\sigma$  on the return  values $u$  and $y$.  This gives  liveness of
+$\lbrace u.\sigma \rbrace$ and $\{y.\sigma\}$ for the program points l1 and l2. We now
+show how the liveness at l1 is propagated backwards; l2 can be treated
+similarly.
+
+Since u is (cdr w), a liveness of $\sigma$ for u translates into a
+demand of $\sigma$ on $(cdr~w)$. This is turn translates into a liveness
+of $\{\epsilon,~1.\sigma\}$ of $w$. In the case of eager evaluation, we
+would have the livenesses of $\pi_7$ and l1 as $\{\epsilon,~1.\sigma\}$ and
+${u.\sigma}$. The liveness of $u$ would have been killed at $\pi_7$. The
+liveness of lazy evaluation differs in the following two respects:
+\begin{enumerate}
+\item The definition  $u = (cdr~w)$, does not  result in an evaluation
+  of $(cdr~w)$; rather a  closure is created.  The evaluation actually
+  takes place  while evaluating $(return~u)$. Thus it  is necessary to
+  record the liveness of $\{\epsilon,  1.\sigma\}$ for $w$ also at the
+  program point l1.
+
+  Regard the  expression as a  tree and consider  paths from
+  the  root  to   program  points  corresponding  to  return
+  expressions.   The  above discussion  shows  that there  a
+  single liveness  environment for each  such path. Further,
+  if a program  point happens to fall on  more than one such
+  path, then  the liveness environment of  the program point
+  is the variable-wise union of the liveness environments of
+  the individual paths.
+\item The second difference is that since our lets are lazy,
+  the let  definition can be  recursive as at  program point
+  $\pi_1$. In such a case, the liveness has to be formulated
+  as a fix-point computation.
+\end{enumerate}
+
+We therefore  formulate the per-program-point  liveness that
+we require for garbage collection in terms of two ....
+
+A  map $\cal{M}$  that gives  the common  liveness  for each
+return path.
+
+A  map that  gives  the  return paths  that  pass through  a
+program point.
+
+%}
 In classical liveness analysis a  variable is either `live' (its value
@@
 \mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
 \mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
-\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
-          &=& \{x.\sigma\} \\
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
 \mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
-          &=& \{x.\epsilon,y.\epsilon\} \\
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
 \mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
-          &=& \{x.\epsilon\} \\
-\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
 %          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
           &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
+               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
                \end{array}
@@
 \mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
+        \begin{array}{l l}
                     \mathcal{L}(e_1,\sigma,\Lfonly) \cup
         \mathcal{L}(e_2,\sigma,\Lfonly) \cup
-        \{x.\epsilon\} \\
-\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
+        \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \mathcal{L}(e_1,\sigma,\Lfonly) \cup
+        \mathcal{L}(e_2,\sigma,\Lfonly)  & \mbox{otherwise}
+                 \end{array} \\
+        \mathcal{L}(\LET~\myvec{x} \leftarrow \myvec{s}~\IN~e),\sigma,\Lfonly) &=&
         \Lv
-           \setminus \{x.*\}
-           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
-\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
+           \setminus \{\myvec{x}.*\}
+           \cup LE(L)
+\mbox{ where } \Lv = \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\myvec{x}}
 \end{eqnarray*}
@@
 \end{minipage}
-
+\begin{minipage}{0.85\textwidth}
+ \infrule[live-let-define]
+         {f~\stackrel{\text{\tiny def}}{=}~(\DEFINE~(f~ x)~(\SIF~*~x~s[(f~ x)/x],~\mbox{and}~\Ldonly_{x \leftarrow s}~\stackrel{\text{\tiny def}}{=}~\Lfonly_f}       
+         { x \leftarrow s \len \Ldonly_{x \leftarrow s}}
+\end{minipage}
 \normalsize

### 2014-11-10 paper/sections/liveness.tex
liveness version 0

@@
 \\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
+\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
 \mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
         \begin{array}{l l}
-                    \mathcal{L}(e_1,\sigma,\Lfonly) \cup
-        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
-        \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
-                    \mathcal{L}(e_1,\sigma,\Lfonly) \cup
-        \mathcal{L}(e_2,\sigma,\Lfonly)  & \mbox{otherwise}
+                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
+        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
+        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
+        \emptyset  & \mbox{otherwise}
                  \end{array} \\
-        \mathcal{L}(\LET~\myvec{x} \leftarrow \myvec{s}~\IN~e),\sigma,\Lfonly) &=&
-        \Lv
-           \setminus \{\myvec{x}.*\}
-           \cup LE(L)
-\mbox{ where } \Lv = \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\myvec{x}}
+ \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
+\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
+        \in \mathcal{M}\}
+\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
+\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
+ \makebox[0mm]{\hspace*{7cm}   
+ $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
+           x_2\, \ldots\, x_n)/x]))$,} \\
+ \makebox[0mm]{\hspace*{5.5cm} and 
+     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
 \end{eqnarray*}
@@
 \end{minipage}
+
+\bigskip
+\begin{center}
 \begin{minipage}{0.85\textwidth}
- \infrule[live-let-define]
-         {f~\stackrel{\text{\tiny def}}{=}~(\DEFINE~(f~ x)~(\SIF~*~x~s[(f~ x)/x],~\mbox{and}~\Ldonly_{x \leftarrow s}~\stackrel{\text{\tiny def}}{=}~\Lfonly_f}       
-         { x \leftarrow s \len \Ldonly_{x \leftarrow s}}
 \end{minipage}
+\end{center}
+%% \begin{minipage}{0.85\textwidth}
+%%  \infrule[live-let-define]
+%%          {f~\stackrel{\text{\tiny def}}{=}~(\DEFINE~(f~x_1
+%%            ~x_2 \ldots x_n)~(\SIF~*~x~s[(f~x_1~
+%%            x_2\ldots x_n)/x]),~\mbox{and}~\Ldonly_{x \leftarrow s}~\stackrel{\text{\tiny def}}{=}~\Lfonly_f}       
+%%          { x \leftarrow s \len \Ldonly_{x \leftarrow s}
+%% \\ \makebox[0mm]{where
+%%      $(\ x_1\ \ldots\ x_n)$ are the variables in $s$}}
+%% \end{minipage}
 \normalsize

### 2014-12-17 paper/sections/syntax-and-semantics.tex
Updated Liveness Rules
Nearing the end of syntax and semantics

@@
 We let $x$, $y$, $z$ range over variables, $f$ over user-functions and
-$p$ over  primitive functions ($\CONS$, $\PRIM$ etc.).   The syntax of
-our language  is shown  in Figure~\ref{fig:lang-syntax}; it  has eager
-semantics and  restricts programs to be in  Administrative Normal Form
+$p$    over    primitive    functions   ($\CONS$,    $\PRIM$    etc.).
+Figure~\ref{fig:lang-syntax} describes the  syntax of our language. It
+is a first  order language with lazy semantics  and restricts programs
+to        be        in        Administrative        Normal        Form
 (ANF)~\cite{chakravarty03perspective}  where all actual  parameters to
 functions   are   variables.   This   restriction   does  not   affect
-expressibility (and indeed we feel  free to ignore it in examples when
-inessential). In addition, this  form also explicates the formation of
-closures.  We further  require  that  each variable  in  a program  is
+expressibility  (and we ignore  it in  examples when  inessential). In
+addition, this  form also explicates the creation  of closures through
+$\LET$s which are lazy.  A consequence  of lazy $\LET$ is that $x$ can
+also occur  in $s$  in $\LET\,\, x  \leftarrow s\,\, \IN\,\,  e$. This
+enables  creation of  graph-like structures  in a  pure  language. The
+restriction of \LET\ to a single definition is for ease of exposition;
+generalization  to  multiple  definitions  does  not  pose  conceptual
+difficulties.  We  further restrict each  variable in a program  to be
 distinct, so  that no scope shadowing  occurs---this simplifies proofs
-of soundness.  In this  formulation expressions: either perform a test
-($\SIF$), make a computation step ($\LET$) or return ($\RETURN$).  The
-$\RETURN$ keyword is logically redundant, but we find it clarifies the
-semantics and analysis.
+of soundness.  The $\RETURN$  and \SIF\ expressions trigger evaluation
+of  closures.
 
-The body of the program is
-the  expression denoted  by  $e_\mainpgm$; for analysis purposes
-it is convenient
-to regard $e_\mainpgm$ as part of a function definition
-$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
-% As  mentioned earlier,
+The body of the program is the expression denoted by $e_\mainpgm$; for
+analysis purposes it is convenient to regard $e_\mainpgm$ as part of a
+function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
 We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
@@
 % \paragraph{Inlining.}
-In spite of the ANF restrictions it is still possible
-to inline non-recursive functions (a fact we use to prove the safety
-of liveness analysis).  A user-function call
-$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
-to a function defined
-(after renaming its formals and locals to be disjoint from existing variables)
-by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
-is replaced by a sequence of $\LET$'s of the form
-$z_i\leftarrow (\ID\,\,y_i)$ followed by
-the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
-$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.
+{\color{blue}In spite of the ANF  restrictions it is still possible to
+  inline non-recursive functions (a fact we use to prove the safety of
+  liveness  analysis).  A user-function  call $(\LET\,\,  x \leftarrow
+  (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$  to a function defined (after
+  renaming  its  formals  and  locals  to be  disjoint  from  existing
+  variables) by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by
+  a  sequence of  $\LET$'s  of the  form $z_i\leftarrow  (\ID\,\,y_i)$
+  followed by the body $e_f$ but with its $(\RETURN\,\,w)$ expressions
+  replaced by $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}
 
@@
     \end{array}\right. \\
-s \in \mathit{Stmt} & ::= &
+s \in \mathit{Application} & ::= &
 \left\{\begin{array}{lr@{\hspace{1em}}l}
@@
 \subsubsection{Semantics}
-We now give an operational
-semantics for  our language.  Later, a refinement of
-the  operational semantics,  which we  call minefield  semantics, will
-serve to prove liveness analysis correct.
-We give a small-step semantics because, unlike big-step semantics,
-correctness for non-terminating programs does not need special treatment.
+We now give  a small-step semantics for our  language.  We shall later
+augment this semantics so that it enables us to reason about liveness.
 We start with the domains used by the semantics:
@@
 \begin{array}{rlcl@{\hspace{2em}}l}
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Loc} & \mbox{-- Environment} \\ 
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Val \times
-  Val}& \mbox{-- Values}\\
-c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (\mathit{Val} +
-\mathit{Clo}) & \mbox{-- Heap}
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\ 
+v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times
+  Data}& \mbox{-- Values}\\
+c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{--
+  Closures}\\
+d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values and Closures} \\ 
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
 \end{array}
 \]
-%Note that in the eager case, if a cons had a numeric argument, this
-%fact was known when the cons was created. Thus you could have unboxed
-%numbers. This is not possible now. For instance, the argument of a
-%cons could be a closure which evaluates to a number.
-%Decision - numbers are boxed
-Here $\mathit{Loc}$ is countable set  of locations in the heap.  Since
+
+Here $\mathit{Loc}$ is a countable set  of locations in the heap.  Since
 all data objects are boxed, we  model an environment as a mapping from
@@
 heap.  These either contain a value  in WHNF (a number, the empty list
-$\NIL$, or  a \CONS\ cell)  or a {\em  closure}.  A closure is  a pair
-$(s, \rho)$ in which $s$ is an unevaluated application with the values
-of its free variables supplied by $\rho$.
+$\NIL$, or  a \CONS\ cell with possibly unevaluated constituents)
+or a {\em closure}.   A closure is a pair $(s, \rho)$  in which $s$ is
+an  unevaluated application  with  the values  of  its free  variables
+given by $\rho$.
 
-Finally, $\stk$  is a stack (using  $\bullet$ for push  and $[\,]$ for
-empty stack)  of frames  of unfinished function  calls.  A frame  is a
-triple  $(e,x,\rho)$   representing  the  call  site   $  (\LET\,\,  x
-\leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\,  e) $ being evaluated in
-environment $\rho$.   Frames can also  be viewed as  continuations, in
-this view  the ({\sc ord-return})  rule in the  small-step operational
-semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
-
-The semantics of statements $s$ are given by the judgement form $\rho,
-\heap, s \rightsquigarrow \heap', v$  and those for expressions $e$ by
-the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
-The  start state is  $(\{\},[\,], \{\},  e_\mainpgm)$ and  the program
-terminates successfully  with result  value $\rho(x)$ on  reaching the
-halt state $(\rho,[\,], \heap, (\RETURN\ x))$
+The  semantics of  expressions  (and applications)  are  given by  the
+judgement form $\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap',
+e'$.  Here \stk\ is a stack  a list of continuation frames viewed as a
+list.   Each continuation  is of  the form  $(\rho, x,  e)$  where the
+location of $x$ is updated when the expression that is currently being
+evaluated reaches WHNF, and $e$ is the next expression to be evaluated
+in  the  environment  $\rho$.    The  start  state  is  $(\{\},[(\{\},
+  \mainpgm,  \HALT)], \{\},  e_\mainpgm)$, and  the  program terminates
+successfully on  reaching the  halt state $(\rho,[\,],  \heap, \HALT)$
+with $\rho(\mainpgm)$ containing the  final result. Note that since we
+assume that our language is not statically typed, a program with other
+than syntactic errors (including those that could have been eliminated
+using a static type system) lead to {\em stuck states} i.e. states not
+covered by any rules in the small-step semantics.
 
@@
 
-\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
-with  $\rightarrow$  (perhaps  because  $\rightsquigarrow$  could  not
-reduce a  contained $s$).  Some  of these we  eliminate syntactically,
-e.g.\ ensuring all variables and  functions are defined and are called
-with the correct number  of parameters.  Others include $(\CDR\ \NIL),
-(\CAR\ 3),  (+\ \NIL\ 4)$ and  $(\SIF\ \NIL\ e_1\ e_2)$.   All but the
-first can  be eliminated with a  static type system  but, treating our
-program as dynamically typed, we regard all these as stuck states.
-
 \begin{figure}[t!]
-\begin{tabular}{|l|l|c|}
+\begin{tabular}{|c|c|c|}
 \hline
-Condition & Transition & Rule name \\
+\hline
+Premise & Transition & Rule name \\
+\hline
 \hline
@@
 \hline
-          & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
+          & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
 $ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
 \hline
-$H(\rho(x)) \mbox{ is } (v_1, v_2) & \rho, (\rho', z, e)\!:\!S, H,
-(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v_1], e$      &  \sc{car-whnf} \\
+$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto d_1], e$      &  \sc{car-whnf} \\
 \hline
-$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (\CAR~x)  \rightsquigarrow
+$H(\rho(x)) \mbox{ is } (e, \rho')$ & $\rho, S, H, (\CAR~x)  \rightsquigarrow
 \rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
@@
 $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
+ & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
 $\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
 \hline
-$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
+$H(\rho(x)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
 \rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
 \hline
-$H(\rho(y)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
+$H(\rho(y)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
 \rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
 \hline
-\shortstack[l]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
+\shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
 [\rho(\myvec{x})/\myvec{y}], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
 \hline
-$\ell$ is a new location& \shortstack[l]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+$\ell$ is a new location& \shortstack[c]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
   \rightsquigarrow$ \\
@@
 \hline
-$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H,
+$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H,
   (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
@@
 \hline
-\shortstack[l]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
+\shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
 (\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
 \hline
-$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H, (\RETURN~x)
+$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\RETURN~x)
   \rightsquigarrow$ \\

### 2014-12-17 paper/sections/syntax-and-semantics.tex
Done -- the first version of syntax and semantics

@@
 % \paragraph{Inlining.}
-{\color{blue}In spite of the ANF  restrictions it is still possible to
+{\color{Myblue}In spite of the ANF  restrictions it is still possible to
   inline non-recursive functions (a fact we use to prove the safety of
@@
 
-The  semantics of  expressions  (and applications)  are  given by  the
-judgement form $\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap',
-e'$.  Here \stk\ is a stack  a list of continuation frames viewed as a
-list.   Each continuation  is of  the form  $(\rho, x,  e)$  where the
-location of $x$ is updated when the expression that is currently being
-evaluated reaches WHNF, and $e$ is the next expression to be evaluated
-in  the  environment  $\rho$.    The  start  state  is  $(\{\},[(\{\},
-  \mainpgm,  \HALT)], \{\},  e_\mainpgm)$, and  the  program terminates
-successfully on  reaching the  halt state $(\rho,[\,],  \heap, \HALT)$
-with $\rho(\mainpgm)$ containing the  final result. Note that since we
+The semantics of expressions (and similarly applications) are given by
+the  judgement form $\rho,  \stk, \heap,  e \rightarrow  \rho', \stk',
+\heap', e'$.  Here \stk\ is a stack of continuation frames represented
+as  a list  in the  description of  the semantics.   Each continuation
+frame  is of the  form $(\rho,  x, e)$  where the  location of  $x$ is
+updated when the expression  that is currently being evaluated reaches
+WHNF,  and  $e$  is  the  next  expression  to  be  evaluated  in  the
+environment  $\rho$.   The  start  state is  $(\{\},[(\{\},  \mainpgm,
+  \HALT)], \{\}, e_\mainpgm)$, and the program terminates successfully
+on  reaching   the  halt   state  $(\rho,[\,],  \heap,   \HALT)$  with
+$\rho(\mainpgm)$  containing  the final  result.  Note  that since  we
 assume that our language is not statically typed, a program with other
 than syntactic errors (including those that could have been eliminated
-using a static type system) lead to {\em stuck states} i.e. states not
-covered by any rules in the small-step semantics.
+using a  static type system) lead  to {\em stuck  states} i.e.  states
+not covered by any rules in the small-step semantics.
 
-Notation: we write $\rho[x \mapsto v]$ for the environment which is as
-$\rho$ but  has value $v$ at  $x$.  We also  write $[\myvec{x} \mapsto
-  \myvec{v}]$  which respectively  has  values $v_1,  \ldots, v_n$  at
-$x_1,  \ldots, x_n$  and write  $[\myvec{x}  \mapsto \rho(\myvec{y})]$
-when $v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
+The  notation $[\myvec{x} \mapsto  \myvec{\ell}]$ represents
+an  environment that maps  variables $x_1,  \ldots, x_n$  to locations
+$\ell_1,  \ldots,  \ell_n$.   The  notation  $\heap[\ell  \mapsto  d]$
+indicates the  updation of  a heap \heap\  at $\ell$ with  $d$.  $\rho
+\oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
+and $\lfloor \rho \rfloor_X$  represents the environment restricted to
+the locations in $X$. Finally $FV(s)$ represents the free variables in
+the application $s$.
+
+As an example, consider the rules related to the application $(+~x~y)$
+involving a strict binary operator.  When both $x$ and $y$ are already
+in WHNF  ({\sc prim-whnf}), the  heap is updated  with the sum  of the
+arguments  and  the  continuation  gives  the next  expression  to  be
+evaluated.   If  either  of  the   arguments  is  not  in  WHNF  ({\sc
+  prim-1-clo} and {\sc prim-2-clo}), it  is sent for evaluation, and a
+continuation marks that the evaluation  of $(+~x~y)$ has to be resumed
+after the argument is evaluated.
 
@@
 \begin{tabular}{|c|c|c|}
-\hline
+
 \hline
@@
 \shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
-[\rho(\myvec{x})/\myvec{y}], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
+[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
 \hline
@@
 $[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
-  \ell)\oplus\lfloor \rho\rfloor_{FV(S)-\{x\}})], e$} & \sc{let} \\ 
+  \ell)\oplus\lfloor \rho\rfloor_{FV(S)\setminus\{x\}})], e$} & \sc{let} \\ 
 \hline

### 2015-01-19 lazy-scheme/src/Output.cpp
LGC working for simple cases

@@ -12,7 +12,8 @@ void Scheme::output::dumpGrammar(std::ostream & out,
                                  Scheme::Demands::regular_demand_grammar * rgram) {
     out << std::left;
     auto gram = rgram->first;
-    for(auto ipart = 0; ipart < rgram->second->size(); ++ipart) {
+    for(auto ipart = 0; ipart < rgram->second->size(); ++ipart)
+    {
         auto type = "NONE";
         if(rgram->second->at(ipart).first == Scheme::Demands::LEFT)         type = "LEFT";
         else if(rgram->second->at(ipart).first == Scheme::Demands::RIGHT)   type = "RIGHT";
@@ -24,14 +25,16 @@ void Scheme::output::dumpGrammar(std::ostream & out,
         for(auto & non_terminal : rgram->second->at(ipart).second) {
             out << std::setw(16) << non_terminal << arrow;
 
-            if(!gram->at(non_terminal).size()) {
+            if(!gram->at(non_terminal).size())
+            {
                 out << empty << "\n";
                 continue;
             }
 
             auto path_iter = gram->at(non_terminal).begin();
 
-            if(path_iter->size()) {
+            if(path_iter->size())
+            {
                 auto token_iter = path_iter->begin();
                 out << *token_iter;
                 while(++token_iter != path_iter->end())
@@ -41,7 +44,8 @@ void Scheme::output::dumpGrammar(std::ostream & out,
 
             while(++path_iter != gram->at(non_terminal).end()) {
                 out << vpipe;
-                if(path_iter->size()) {
+                if(path_iter->size())
+                {
                     auto token_iter = path_iter->begin();
                     out << *token_iter;
                     while(++token_iter != path_iter->end())
@@ -61,29 +65,34 @@ void Scheme::output::dumpGrammar(std::ostream & out, Scheme::Demands::demand_gra
     for(auto & non_terminal : *gram) {
         out << std::setw(16) << non_terminal.first << arrow;
 
-        if(!non_terminal.second.size()) {
+        if(!non_terminal.second.size())
+        {
             out << empty << std::endl;
             continue;
         }
 
         auto path_iter = non_terminal.second.begin();
 
-        if(path_iter->size()) {
+        if(path_iter->size())
+        {
             auto token_iter = path_iter->begin();
             out << *token_iter;
             while(++token_iter != path_iter->end())
                 out << cdot << *token_iter;
-        } else
+        }
+        else
             out << empty;
 
-        while(++path_iter != non_terminal.second.end()) {
[…]

### 2015-01-19 lazy-scheme/src/SchemeAST.h

@@ -197,6 +197,8 @@ public:
 	virtual std::string getLabel() const;
 	virtual cons* make_closure();
 
+	std::string idVarLabel;
+
 protected:
 	IdExprNode * pID;
 

### 2015-01-19 lazy-scheme/src/SchemeAST_Defs.cpp

@@ -32,6 +32,9 @@ extern unsigned int num_of_allocations;
 extern map<cons*, int> heap_map;
 extern map<int, string> root_var_map;
 
+
+std::string curr_return_addr;
+
 std::string getNextLabel()
 {
 	std::stringstream sstream;
@@ -160,6 +163,7 @@ void ReturnExprNode::doLabel(bool shouldAddLabel)
 {
 	//label = (shouldAddLabel) ? getNextLabel() : "";
 	pID->doLabel(shouldAddLabel);
+	idVarLabel = pID->getLabel();
 	label = pID->getLabel();
 }
 
@@ -185,12 +189,15 @@ cons* ReturnExprNode::evaluate(cons* heap_cell = NULL)
 
 		retval->inWHNF = true;
 		retval->typecell = temp->typecell;
-		retval->val = temp->val;\
+		retval->val = temp->val;
 
 		update_heap_refs.pop();
 
 	}
 	
+	cout << "Setting to current return address to " << this->pID->getLabel() << " or " << idVarLabel << endl;
+	curr_return_addr = this->getLabel();
+
 	return retval;
 }
 cons* ReturnExprNode::make_closure()
@@ -424,6 +431,7 @@ cons* IfExprNode::evaluate(cons* heap_cell = NULL)
 		IdExprNode* i = (IdExprNode*)this->pCond;
 		cons* cond_heap_ref = (cons*)lookup_addr(i->getIDStr().c_str());
 
+		cout << "Evaluating condition corrsponding to variable " << i->getIDStr() << endl;
 		update_heap_refs.push(cond_heap_ref);
 		//cout << "Pushing on to stack in if expr " << cond_heap_ref << endl;
 		cons* cond_resultValue = this->pCond->evaluate();
@@ -513,8 +521,11 @@ LetExprNode * LetExprNode::clone() const
 
 cons* LetExprNode::evaluate(cons* heap_cell = NULL)
 {
-//	cout << "Processing let variable " << this->pID->getIDStr() << " at label " << getLabel() <<  endl;
-	
+
+	cout << "Processing let variable " << this->pID->getIDStr() << " at label " << getLabel() <<  endl;
+	curr_return_addr = getLabel();
+
+
 	if ((gc_status != gc_disable && current_heap() < 1) ||
 			(getVarExpr()->isFunctionCallExpression() && (current_heap < (0 + ((FuncExprNode*)(getVarExpr()))->pListArgs->size()))) )
 	{
[…]

### 2015-01-19 lazy-scheme/src/SchemeAST_Demands.cpp

@@ -16,6 +16,8 @@ unordered_map<string, Node*> gfunc_prog_pts;
 bool in_function_define = false;
 
 
+
+
 void print_label_set(unordered_set<string> s)
 {
 	cout << "Printing label set " << endl;
@@ -38,6 +40,8 @@ unordered_map<string, expr_demand_grammars*> ReturnExprNode::transformDemand(con
 	//cout << "Created new label for return " << label << endl;
 	label_set.insert(label);
 	pID->label_set.insert(label);
+
+
 	return gLivenessMap;
 }
 
@@ -175,8 +179,8 @@ expr_demand_grammars * BinaryPrimExprNode::transformDemandRef(const rule & deman
     expr_demand_grammars * result =  merge(pArg1->transformDemandRef(arg_1_demand),
                                            pArg2->transformDemandRef(arg_2_demand));
     result->first->emplace(label, demand);
-    cout << "Printing result of BinaryPrimOp Demand transform " << endl;
-    Scheme::output::dumpGrammar(cout, result->second);
+//    cout << "Printing result of BinaryPrimOp Demand transform " << endl;
+//    Scheme::output::dumpGrammar(cout, result->second);
     
 
     return result;

### 2015-01-19 lazy-scheme/src/gc.cpp

@@ -109,6 +109,8 @@ void set_max_reachability();
 #endif
 
 
+cons* copy_deep(cons* node);
+
 /*-----Memory manager-------*/
 
 //!!!caution remove memorysize from this function
@@ -275,7 +277,7 @@ void make_environment(const char *functionname, std::string returnpoint)
 	std::cout << "Creating new environment for function " << functionname << " with return " << returnpoint << endl;
 	actRec newenv;
     actRecStack.front().return_point = returnpoint;
-    newenv.return_point=-1;
+    newenv.return_point = "-1";
     newenv.funcname=functionname;
     actRecStack.push_front(newenv);
     mmc=mmc+(sizeof(actRec));
@@ -284,6 +286,7 @@ void make_environment(const char *functionname, std::string returnpoint)
 void delete_environment()
 {
 	
+	cout << "Removing function " << actRecStack.front().funcname << endl;
 	actRecStack.pop_front();
     mmc=mmc-(sizeof(actRec));
     actRecStack.front().return_point=-1; //reset
@@ -722,7 +725,20 @@ void update_heap_ref_stack()
 		cons* heap_ref = update_heap_refs.top();
 		cout << ++i << " " << heap_ref << "  " << heap_ref->forward << endl;
 		//cout << heap_ref->typecell << endl;
-		assert(heap_ref->forward != NULL);
+
+
+		//assert(heap_ref->forward != NULL);
+
+		//forward pointer should not be NULL, so do a deep copy and update the pointer.
+		//This might be a hack, need to do something better. Maybe put the print method on stack and
+		//ensure that references that are being printed are fully copied.
+		if (heap_ref->forward == NULL)
+		{
+			cons* new_ref = copy_deep(heap_ref);
+			cout << "In heap ref stack copying " << heap_ref << " to " << new_ref << " or " << heap_ref->forward << endl;
+			//heap_ref = new_ref;
+		}
+
 		heap_ref = heap_ref->forward;
 		assert(is_valid_address(heap_ref));
 		temp.push(heap_ref);
@@ -737,13 +753,22 @@ void update_heap_ref_stack()
 	assert(temp.empty());
 	//Update print stack
 
-	cout << "Updating temp stack with size " << print_stack.size() << endl;
[…]

### 2015-01-19 paper/sections/syntax-and-semantics.tex
my changes

@@
+<<<<<<< HEAD
+
+\section{The target language---syntax and semantics}
+\label{sec:defs}
+We let $x$, $y$, $z$ range over variables, $f$ over user-functions and
+$p$ over  primitive functions ($\CONS$, $\PRIM$ etc.).   The syntax of
+our language  is shown  in Figure~\ref{fig:lang-syntax}; it  has eager
+semantics and  restricts programs to be in  Administrative Normal Form
+(ANF)~\cite{chakravarty03perspective}  where all actual  parameters to
+functions   are   variables.   This   restriction   does  not   affect
+expressibility (and indeed we feel  free to ignore it in examples when
+inessential). In addition, this  form also explicates the formation of
+closures.  We further  require  that  each variable  in  a program  is
+distinct, so  that no scope shadowing  occurs---this simplifies proofs
+of soundness.  In this  formulation expressions: either perform a test
+($\SIF$), make a computation step ($\LET$) or return ($\RETURN$).  The
+$\RETURN$ keyword is logically redundant, but we find it clarifies the
+semantics and analysis.
+
+The body of the program is
+the  expression denoted  by  $e_\mainpgm$; for analysis purposes
+it is convenient
+to regard $e_\mainpgm$ as part of a function definition
+$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
+% As  mentioned earlier,
+We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
+language syntax) with the program point just before expression $e$.
+
+% \paragraph{Inlining.}
+In spite of the ANF restrictions it is still possible
+to inline non-recursive functions (a fact we use to prove the safety
+of liveness analysis).  A user-function call
+$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
+to a function defined
+(after renaming its formals and locals to be disjoint from existing variables)
+by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
+is replaced by a sequence of $\LET$'s of the form
+$z_i\leftarrow (\ID\,\,y_i)$ followed by
+the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
+$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.
+
+\begin{figure}[t!]
+\footnotesize
+\begin{eqnarray*}
+   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
+    \hspace{7em} \mbox{\em --- program}\\
+    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
+    e) 
+    \hspace{1.7em} \mbox{\em --- function definition} \\
+e \in \mathit{Expr} & ::= &
+\left\{\begin{array}{ll@{\hspace{4em}}l}
+       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
+       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
+       (\RETURN\,\, x) && \mbox{\em --- return from function}
+    \end{array}\right. \\
+s \in \mathit{Stmt} & ::= &
+\left\{\begin{array}{lr@{\hspace{1em}}l}
+       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
+       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
+       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
+       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
+%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
+       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
+%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
+       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
+            & \mbox{\em --- function application} 
+    \end{array}\right.
+\end{eqnarray*}
+  \caption{The syntax of our language}\label{fig:lang-syntax}
+\figrule
+\normalsize
+\end{figure}
+
+\subsubsection{Semantics}
+We now give an operational
+semantics for  our language.  Later, a refinement of
+the  operational semantics,  which we  call minefield  semantics, will
+serve to prove liveness analysis correct.
+We give a small-step semantics because, unlike big-step semantics,
+correctness for non-terminating programs does not need special treatment.
+We start with the domains used by the semantics:
+\[
+\begin{array}{rlcl@{\hspace{2em}}l}
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Loc} & \mbox{-- Environment} \\ 
+v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Val \times
+  Val}& \mbox{-- Values}\\
+c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (\mathit{Val} +
+\mathit{Clo}) & \mbox{-- Heap}
+\end{array}
+\]
+%Note that in the eager case, if a cons had a numeric argument, this
+%fact was known when the cons was created. Thus you could have unboxed
+%numbers. This is not possible now. For instance, the argument of a
+%cons could be a closure which evaluates to a number.
+%Decision - numbers are boxed
+Here $\mathit{Loc}$ is countable set  of locations in the heap.  Since
+all data objects are boxed, we  model an environment as a mapping from
+the set of variables of the program $\mathit{Var}$ to locations in the
+heap.  These either contain a value  in WHNF (a number, the empty list
+$\NIL$, or  a \CONS\ cell)  or a {\em  closure}.  A closure is  a pair
+$(s, \rho)$ in which $s$ is an unevaluated application with the values
+of its free variables supplied by $\rho$.
+
+Finally, $\stk$  is a stack (using  $\bullet$ for push  and $[\,]$ for
+empty stack)  of frames  of unfinished function  calls.  A frame  is a
+triple  $(e,x,\rho)$   representing  the  call  site   $  (\LET\,\,  x
+\leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\,  e) $ being evaluated in
+environment $\rho$.   Frames can also  be viewed as  continuations, in
+this view  the ({\sc ord-return})  rule in the  small-step operational
+semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.
+
+The semantics of statements $s$ are given by the judgement form $\rho,
+\heap, s \rightsquigarrow \heap', v$  and those for expressions $e$ by
+the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
+The  start state is  $(\{\},[\,], \{\},  e_\mainpgm)$ and  the program
+terminates successfully  with result  value $\rho(x)$ on  reaching the
+halt state $(\rho,[\,], \heap, (\RETURN\ x))$
+
+Notation: we write $\rho[x \mapsto v]$ for the environment which is as
+$\rho$ but  has value $v$ at  $x$.  We also  write $[\myvec{x} \mapsto
+  \myvec{v}]$  which respectively  has  values $v_1,  \ldots, v_n$  at
+$x_1,  \ldots, x_n$  and write  $[\myvec{x}  \mapsto \rho(\myvec{y})]$
+when $v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.
+
+\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
+with  $\rightarrow$  (perhaps  because  $\rightsquigarrow$  could  not
+reduce a  contained $s$).  Some  of these we  eliminate syntactically,
+e.g.\ ensuring all variables and  functions are defined and are called
+with the correct number  of parameters.  Others include $(\CDR\ \NIL),
+(\CAR\ 3),  (+\ \NIL\ 4)$ and  $(\SIF\ \NIL\ e_1\ e_2)$.   All but the
+first can  be eliminated with a  static type system  but, treating our
+program as dynamically typed, we regard all these as stuck states.
+
+\begin{figure}[t!]
+\begin{tabular}{|l|l|c|}
+\hline
+Condition & Transition & Rule name \\
+\hline
+          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
+  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
+\hline
+          & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
+$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (v_1, v_2) & \rho, (\rho', z, e)\!:\!S, H,
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v_1], e$      &  \sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (\CAR~x)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & \shortstack[l]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
+$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } (e, \rho') & \rho, S, H, (+~x~y)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
+\hline
+\shortstack[l]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
+[\rho(\myvec{x})/\myvec{y}], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
+\hline
+$\ell$ is a new location& \shortstack[l]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+  \rightsquigarrow$ \\
+$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
+  \ell)\oplus\lfloor \rho\rfloor_{FV(S)-\{x\}})], e$} & \sc{let} \\ 
+\hline
+$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_1$ & \sc{if-true} \\ 
+\hline
+$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_2$ & \sc{if-false} \\ 
+\hline
+$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H,
+  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
+$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
+\hline
+\shortstack[l]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
+(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
+\hline
+$H(\rho(x)) = (e,\rho')$ & \shortstack[l]{$\rho, S, H, (\RETURN~x)
+  \rightsquigarrow$ \\
+$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-clo} \\
+\hline
+\end{tabular}
+\caption{A small-step semantics for the language}\label{fig:lang-semantics}
+\end{figure}
+%==============================================================
+\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
+
+=======
+
 \section{The target language---syntax and semantics}
@@
 \renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
+
+>>>>>>> 4bb03a5c3d7b9f8e48b056d77b2ef7801e7effa3

### 2015-03-13 paper/paper.tex
updates...
added liveness section
pushing liveness
initializing
Updated Liveness Rules
my changes
fh_defs.tex

@@
-%------------------------------------------------------------%
-\documentclass{llncs}
-\pagestyle{plain}
-\usepackage{longtable} 
-%------------------------------------------------------------%
-%- common stuff starts here....
+\documentclass{sig-alternate}
 \usepackage{etex}
@@
 \usepackage[T1]{fontenc}
-%\usepackage[pdf]{pstricks}
+\usepackage[pdf]{pstricks}
 \usepackage{epsfig}
@@
 \usepackage{tabularx}
-%\usepackage{color}
-%\usepackage{xcolor}
 \usepackage{algorithmic}
@@
 \usepackage{pst-rel-points}
-%\usepackage{pst-all}
 \usepackage{bcprules}
-%\usepackage{subfigure}
-%\usepackage{algorithmic}
-%\usepackage{anysize}     
 \input{fh_defs}
+\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
+      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
+      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
+      #1 \vfill}}
+      %Poor man's typography
+\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
+      %It would be better to specify these in font-relative measures, but it 
+      %probably doesn't scale anyway.
 %\newcommand{\comment}[1]{}
@@
 \newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
-
+%\nonstopmode
 \def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
@@
 \begin{document}
-%  \title{Heap  Reference  Analysis  for Functional  Programs}
-\title{Liveness-Based Garbage Collection}
+%
+% --- Author Metadata here ---
+\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
+%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
+% --- End of Author Metadata ---
 
-% make a proper TOC despite llncs
-\setcounter{tocdepth}{2}
-\makeatletter
-\renewcommand*\l@author[2]{}
-\renewcommand*\l@title[2]{}
-\makeatletter
+\title{Liveness-Based Garbage Collection for Lazy Languages}
 
-  \author{K. Prasanna Kumar\inst{1} \and Amey Karkare\inst{2}
-    \and   Amitabha Sanyal\inst{1}}
-  \institute{IIT Bombay, Mumbai 400076, India\\
-    \email{$\{$prasanna,as$\}$@cse.iitb.ac.in}, \and 
-             IIT Kanpur, Kanpur 208016, India\\
-    \email{karkare@cse.iitk.ac.in}}
+%
+% You need the command \numberofauthors to handle the 'placement
+% and alignment' of the authors beneath the title.
+%
+% For aesthetic reasons, we recommend 'three authors at a time'
+% i.e. three 'name/affiliation blocks' be placed beneath the title.
+%
+% NOTE: You are NOT restricted in how many 'rows' of
+% "name/affiliations" may appear. We just ask that you restrict
+% the number of 'columns' to three.
+%
+% Because of the available 'opening page real-estate'
+% we ask you to refrain from putting more than six authors
+% (two rows with three columns) beneath the article title.
+% More than six makes the first-page appear very cluttered indeed.
+%
+% Use the \alignauthor commands to handle the names
+% and affiliations for an 'aesthetic maximum' of six authors.
+% Add names, affiliations, addresses for
+% the seventh etc. author(s) as the argument for the
+% \additionalauthors command.
+% These 'additional authors' will be output/set for you
+% without further effort on your part as the last section in
+% the body of your article BEFORE References or any Appendices.
+
+\numberofauthors{3} %  in this sample file, there are a *total*
+% of EIGHT authors. SIX appear on the 'first-page' (for formatting
+% reasons) and the remaining two appear in the \additionalauthors section.
+%
+\author{
+% 1st. author
+\alignauthor K. Prasanna Kumar \\
+       \affaddr{IIT Bombay,}\\
+       \affaddr{Mumbai 400076, India}
+       \email{prasanna@cse.iitb.ac.in}
+% 2nd. author
+\alignauthor Amey Karkare\\
+       \affaddr{IIT Kanpur}\\
+       \affaddr{Kanpur 208016, India}
+       \email{karkare@cse.iitk.ac.in}
+% 3rd. author
+\alignauthor Amitabha Sanyal \\
+       \affaddr{IIT Bombay}\\
+       \affaddr{Mumbai 400076, India}\\
+       \email{as@cse.iitb.ac.in}
+}
 \maketitle
+\begin{abstract}
+We  consider  two  ways  to  reduce the  memory  requirement  of  lazy
+functional  programming  languages. The  first  is  to modify  garbage
+collectors  to  preserve  only  {\em  live} data---a  subset  of  {\em
+reachable data}.  This results in an increase in the garbage reclaimed
+and  consequently in  fewer  garbage collections.   The  second is  to
+eagerly evaluate closures  that are guaranteed to be  evaluated in the
+future.
 
-\begin{abstract}
-\input{sections/abstract}
+The  main contribution  of  this  paper is  an  uniform framework  for
+obtaining the information for the optimizations mentioned above---{\em
+liveness}  and {\em strictness}  analysis. Using  a central  notion of
+demand,  we  formulate   a  context-sensitive  liveness  analysis  for
+structured  data and  prove  it  correct.  We  then present strictness
+as a dual of liveness. We use a  0-CFA-like
+conservative   approximation   to   annotate   each   allocation   and
+function-call program point  with a finite-state automaton---which the
+garbage-collector inspects to  curtail reachability during
+marking. The result of the strictness analysis, being necessarily
+finite,  can  be directly used to insert code to evaluate closures.
+
+%As a result,  fewer objects are marked (albeit  with a more expensive
+%marker) and then preserved (e.g.\ by a copy phase).
+
+Experiments  confirm  the  expected  performance benefits.  The  early
+evaluation   of   closures  decreases   the   memory   demands  of   a
+program. Further,  liveness-based collection results in  a increase in
+garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
+collections, a decrease  in the memory size required  to run programs,
+and  reduced  overall  garbage  collection  time  for  a  majority  of
+programs.
 \end{abstract}
 
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+\category{D.2.8}{Software  Engineering}{Metrics}[complexity  measures,
+  performance measures]
+
+\terms{Theory}
+
+\keywords{ACM proceedings, \LaTeX, text tagging}
+
+By Friday, February 27 2015, 23:59 (UTC-11), submit a full paper of at
+most  12 pages (6  pages for  an Experience  Report), in  standard ACM
+conference format, including bibliography, figures, and appendices.
+
+\section{Introduction}
+\label{sec:intro}
+
+Functional  programming  languages  depend  on  garbage  collector  to
+efficiently  reclaim  unused  memory   allocated  by  programs  either
+explicitly (for example using constructors) or implicitly (for example
+for      closures).       However,      empirical      studies      on
+Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
+Java~\cite{shaham02estimating}  programs   have  shown   that  garbage
+collectors leave uncollected a lot of memory objects that are not live
+(here {\em live}ness  means the object can potentially be  used by the
+program at  a later  stage).  Due  to lack of  a suitable  analysis to
+distinguish  reachable  objects  from live  objects.  current  garbage
+collectors conservatively approximate the  liveness of heap objects by
+their reachability from a set of memory locations called the {\em root
+  set\/} resulting in unnecessary memory usage.
+
+Languages having lazy  semantics for evaluation put  further demand on
+memory as they  require heap allocated closure to be  carried from the
+point of occurrence  of an expression to the point  of (first) real use
+of  that expression.  Strictness analysis~\cite{some-refs}  statically
+determines  the  patterns  of  future usage  of  an  expression,  thus
+allowing some  of the  closures to  be evaluated at  the point  of the
+occurrence of the corresponding expression.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% MOTIVATING EXAMPLE
+\newcommand{\nilfigure}
+{\scalebox{0.75}{
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,0)(1,1)
+%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
+\putnode{start}{origin}{0}{0}{}
+\putnode{stop}{origin}{10}{10}{}
+\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
+\end{pspicture}
+}}
+
+\begin{figure*}[t!]
+%%%%{\color{Myblue}
+\begin{picture}(100,130)(0,-90)
+%  \begin{center}
+    \begin{tabular}{cc}
+\begin{boxedminipage}{.5\textwidth}
+      {\sf
+	\renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
+	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
+	       \listb
+	  \UNL{1}      (\CONS\ (\CAR\  \lista)
+%          \UNL{2}          (\CAR\  \lista)
+	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
+          \listb))))
+          \UNL{0}
+	  \UNL{0} \hspace*{-.49\TAL} (\LET\ \pz  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
+	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
+	  \UNL{0} (\LET\ \py  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
+	  \UNL{1}   (\LET\ \pw $\leftarrow$ (\append\ \py\  \pz)\ \IN
+	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ \pw)))))
+	\end{uprogram}
+      }} 
+\end{boxedminipage}
+      &
+      \raisebox{-25mm}{\scalebox{.75}{
+	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
+      \psset{unit=1mm}
+      \psset{linewidth=.3mm}
+      \begin{pspicture}(0,-5)(70,60)
+	%\psframe(0,0)(73,60)
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
+	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
+%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
+	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
+	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
+	%      fillcolor=white]{$\times$}}}
+        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
+%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
+	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
+	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
+	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
+	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
+		  fillcolor=white]{$\times$}}}
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
+	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
+	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
+	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
+	\ncline[nodesepB=-.5]{*->}{c2}{d}
+	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
+	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
+	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
+	      fillcolor=white]{$\times$}}}
+	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
+	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
+	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
+	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
+	      fillcolor=white]{$\times$}}}
+	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
+	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
+	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
+	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
+	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
+%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
+	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
+	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
+        %here
+        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
+        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
+
+%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
+%	      fillcolor=white]{$\times$}}} %and here
+	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
+	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
+%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
+%	      fillcolor=white]{$\times$}}} %and here
+
+	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
+	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
+	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
+	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
+	      fillcolor=white]{$\times$}}}
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
+%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
+	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
+	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
+%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+      \end{pspicture}}} \\
+      (a) Example program.&
+      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+	(b) Memory graph at $\pi$. Thick edges denote \\
+        {\white (b) }live links. Traversal stops at edges marked\\
+        {\white (b) }$\times$  during garbage collection.
+      \end{tabular}}
+    \end{tabular}
+\end{picture}
+
+\kern -3ex
+
+\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
+%  \end{center}
+%%%%}\color{Myblue} end
+\end{figure*}
+We propose the  use of liveness analysis of heap  cells and strictness
+analysis  of expressions  to  reduce the  memory  requirement of  lazy
+functional  programming languages.   We modify  garbage collectors  to
+preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
+This results in an increase  in the garbage reclaimed and consequently
+in fewer  garbage collections.  Further, we eagerly evaluate closures
+that are guaranteed to be evaluated in the future (strict closures).
+
+The main contribution  of this paper is an uniform  framework for {\em
+  liveness} and {\em strictness}  analysis to enable the optimizations
+mentioned  above. Using  a central  notion of  demand, we  formulate a
+context-sensitive liveness  analysis for structured data  and prove it
+correct.  We then  present strictness as a dual of  liveness. We use a
+0-CFA-like conservative approximation to  annotate each allocation and
+function-call program point with  a finite-state automaton---which the
+garbage-collector inspects to curtail reachability during marking. The
+result of  the strictness analysis,  being necessarily finite,  can be
+directly used to insert code to evaluate closures.
+
+We previously proposed an interprocedural liveness analysis to improve
+garbage collection first order eager subset of Scheme~\cite{cc-paper}.
+The current paper extends the work to handle lazy evaluation semantics
+for the same language.
+
+Experiments  confirm  the  expected  performance benefits.  The  early
+evaluation   of   closures  decreases   the   memory   demands  of   a
+program. Further,  liveness-based collection results in  a increase in
+garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
+collections, a decrease  in the memory size required  to run programs,
+and  reduced  overall  garbage  collection  time  for  a  majority  of
+programs.
+
 %------------------------------------------------------------%
+\subsubsection{Motivating Example}
+\label{sec:motiv}
 
-%% \begin{figure}
-%% \tableofcontents       %%%%%%% BEWARE: find the \setcounter{page}{2}
-%%                        %%%%%%% and comment it out when deleting this TOC.
-%% \vspace{5cm}
+Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
+$\pi$ of an expression $e$  denotes a program point.  During execution
+of  the program, it  represents the  instant of  time just  before the
+evaluation  of  $e$.  
+We  view the  heap  as a graph.  
+Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
+containing values.  Edges in the graph are {\em references} and
+emanate from {\em variables} or fields.
+Variable and field values may also be atomic values ($\NIL$, integers etc.)
+While it is convenient to box these in diagrams, our presented
+analysis treats them as non-heap values.
 
-%% \section*{Page Budget}
-%% CC'2014 and ESOP'2014 have a limit of 20 pages.
-%% A draft page budget is:
-%% \begin{itemize}
-%% \item 1. abstract/introduction (3 pages)
-%% {\color{red}Currently 2.5 pages.}
-%% \item 2. core language syntax and semantics (2 pages).
-%% {\color{red}Currently 2 pages.}
-%% \item 3. liveness analysis and correctness (3 pages)
-%% {\color{red}Currently 3.8 pages.}
-%% \item 4. practical liveness computation (6 pages)
-%% {\color{red}Currently 6.2 pages.}
-%% \\ I think this is where should do all 3 of: the context-insensitive analysis
-%% (one which has a summary demand for each function), then the
-%% solving like differential equations, then using Mohri-Nederhof to
-%% get a NFA for each (call/cons) program point.
-%% \item 5. experimental framework and results (4 pages)
-%% \item 6. (optional) monotonicity of GC (0.5 page)---our technique can never
-%% cause more collections.
-%% \item 7. conclusions/acknowledgements/bibliography (1.5 page)
-%% \end{itemize}
-%% \end{figure}
+Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
+The edges shown  by thick arrows are those  which are made live by the
+program. In addition,  assuming that
+the value of any reachable part of the program result may be explored
+or printed, the edges marked by thick dashed arrows are also live.
+A cell  is  marked and preserved  during  garbage collection,  only  if it  is
+reachable from the  root set through a path of  live edges.  All other
+cells can be  reclaimed.  Thus if  a garbage collection takes
+% place   at    the   time   represented   by   the    heap   shown   in
+at $\pi$ with the    heap   shown   in
+Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
+will be marked and preserved.
 
-\input{sections/intro}
+\subsubsection{Organization of the paper}
+Section~\ref{sec:defs} gives
+the syntax and semantics of
+the language used to illustrate our analysis along with basic concepts
+and    notations.
+% Section~\ref{sec:operational}   describes    the
+% operational semantics  of the  language.
+Liveness analysis is described in
+Section~\ref{sec:liveness} followed by a sketch of a correctness
+proof relative to a non-standard semantics.
+Section~\ref{sec:computing} shows how to encode liveness as
+finite-state automata.
+Section~\ref{sec:experiments}   reports   experimental   results and
+Section~\ref{sec:lgc-always-better} proves that a liveness based
+collector can never do more garbage collections than a reachability
+based collector. 
+% and Section~\ref{sec:conclusion} concludes the paper.
 
-%\input{sections/restOfOldPaper}
+\section{The target language---syntax and semantics}
+\label{sec:defs}
+We let $x$, $y$, $z$ range over variables, $f$ over user-functions and
+$p$    over    primitive    functions   ($\CONS$,    $\PRIM$    etc.).
+Figure~\ref{fig:lang-syntax} describes the  syntax of our language. It
+is a first  order language with lazy semantics  and restricts programs
+to        be        in        Administrative        Normal        Form
+(ANF)~\cite{chakravarty03perspective}  where all actual  parameters to
+functions   are   variables.   This   restriction   does  not   affect
+expressibility  (and we ignore  it in  examples when  inessential). In
+addition, this  form also explicates the creation  of closures through
+$\LET$s which are lazy.  A consequence  of lazy $\LET$ is that $x$ can
+also occur  in $s$  in $\LET\,\, x  \leftarrow s\,\, \IN\,\,  e$. This
+enables  creation of  graph-like structures  in a  pure  language. The
+restriction of \LET\ to a single definition is for ease of exposition;
+generalization  to  multiple  definitions  does  not  pose  conceptual
+difficulties.  We  further restrict each  variable in a program  to be
+distinct, so  that no scope shadowing  occurs---this simplifies proofs
+of soundness.  The $\RETURN$  and \SIF\ expressions trigger evaluation
+of  closures.
 
+The body of the program is the expression denoted by $e_\mainpgm$; for
+analysis purposes it is convenient to regard $e_\mainpgm$ as part of a
+function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
+We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
+language syntax) with the program point just before expression $e$.
+
+% \paragraph{Inlining.}
+{\color{Myblue}In spite of the ANF  restrictions it is still possible to
+  inline non-recursive functions (a fact we use to prove the safety of
+  liveness  analysis).  A user-function  call $(\LET\,\,  x \leftarrow
+  (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$  to a function defined (after
+  renaming  its  formals  and  locals  to be  disjoint  from  existing
+  variables) by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by
+  a  sequence of  $\LET$'s  of the  form $z_i\leftarrow  (\ID\,\,y_i)$
+  followed by the body $e_f$ but with its $(\RETURN\,\,w)$ expressions
+  replaced by $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}
+
+\begin{figure*}[]
+\footnotesize
+\begin{eqnarray*}
+   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
+    \hspace{7em} \mbox{\em --- program}\\
+    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
+    e) 
+    \hspace{1.7em} \mbox{\em --- function definition} \\
+e \in \mathit{Expr} & ::= &
+\left\{\begin{array}{ll@{\hspace{4em}}l}
+       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
+       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
+       (\RETURN\,\, x) && \mbox{\em --- return from function}
+    \end{array}\right. \\
+s \in \mathit{Application} & ::= &
+\left\{\begin{array}{lr@{\hspace{1em}}l}
+       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
+       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
+       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
+       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
+%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
+       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
+%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
+       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
+            & \mbox{\em --- function application} 
+    \end{array}\right.
+\end{eqnarray*}
+  \caption{The syntax of our language}\label{fig:lang-syntax}
+\figrule
+\normalsize
+\end{figure*}
+
+\subsubsection{Semantics}
+We now give  a small-step semantics for our  language.  We shall later
+augment this semantics so that it enables us to reason about liveness.
+We start with the domains used by the semantics:
+\[
+\begin{array}{rlcl@{\hspace{2em}}l}
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\ 
+v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times
+  Data}& \mbox{-- Values}\\
+c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{--
+  Closures}\\
+d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values/Closures} \\ 
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
+\end{array}
+\]
+
+Here $\mathit{Loc}$ is a countable set  of locations in the heap.  Since
+all data objects are boxed, we  model an environment as a mapping from
+the set of variables of the program $\mathit{Var}$ to locations in the
+heap.  These either contain a value  in WHNF (a number, the empty list
+$\NIL$, or  a \CONS\ cell with possibly unevaluated constituents)
+or a {\em closure}.   A closure is a pair $(s, \rho)$  in which $s$ is
+an  unevaluated application  with  the values  of  its free  variables
+given by $\rho$.
+
+The semantics of expressions (and applications) are given by
+the  judgment form $\rho,  \stk, \heap,  e \rightarrow  \rho', \stk',
+\heap', e'$.  Here \stk\ is a stack of continuation frames represented
+as  a list  in the  description of  the semantics.   Each continuation
+frame  is of the  form $(\rho,  x, e)$,  where the  location of  $x$ is
+updated when the currently evaluated expression reaches
+WHNF,  and  $e$  is  the  next  expression  to  be  evaluated  in  the
+environment  $\rho$.   The  start  state is  $(\{\},[(\{\},  \mainpgm,
+  \HALT)], \{\}, e_\mainpgm)$, and the program terminates successfully
+on  reaching   the  halt   state  $(\rho,[\,],  \heap,   \HALT)$  with
+$\rho(\mainpgm)$  containing  the final  result.  Note  that since  we
+assume that our language is not statically typed, a program with other
+than syntactic errors (including those that could have been eliminated
+using a  static type system) lead  to {\em stuck  states} i.e.  states
+not covered by any rules in the small-step semantics.
+
+The  notation $[\myvec{x} \mapsto  \myvec{\ell}]$ represents
+an  environment that maps  variables $x_1,  \ldots, x_n$  to locations
+$\ell_1,  \ldots,  \ell_n$.   The  notation  $\heap[\ell  \mapsto  d]$
+indicates the  updation of  a heap \heap\  at $\ell$ with  $d$.  $\rho
+\oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
+and $\lfloor \rho \rfloor_X$  represents the environment restricted to
+the locations in $X$. Finally $FV(s)$ represents the free variables in
+the application $s$.
+
+As an example, consider the rules related to the application $(+~x~y)$
+involving a strict binary operator.  When both $x$ and $y$ are already
+in WHNF  ({\sc prim-whnf}), the  heap is updated  with the sum  of the
+arguments  and  the  continuation  gives  the next  expression  to  be
+evaluated.   If  either  of  the   arguments  is  not  in  WHNF  ({\sc
+  prim-1-clo} and {\sc prim-2-clo}), it  is sent for evaluation, and a
+continuation marks that the evaluation  of $(+~x~y)$ has to be resumed
+after the argument is evaluated.
+
+\begin{figure*}[t!]
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Premise & Transition & Rule name \\
+\hline
+\hline
+          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
+  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
+\hline
+          & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
+$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto d_1], e$      &  \sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } (e, \rho')$ & $\rho, S, H, (\CAR~x)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
+$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
+\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
+\hline
+\shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
+[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
+\hline
+$\ell$ is a new location& \shortstack[c]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+  \rightsquigarrow$ \\
+$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
+  \ell)\oplus\lfloor \rho\rfloor_{FV(S)\setminus\{x\}})], e$} & \sc{let} \\ 
+\hline
+$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_1$ & \sc{if-true} \\ 
+\hline
+$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_2$ & \sc{if-false} \\ 
+\hline
+$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H,
+  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
+$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
+\hline
+\shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
+(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
+\hline
+$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\RETURN~x)
+  \rightsquigarrow$ \\
+$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-coo} \\
+\hline
+\end{tabular}
+\caption{A small-step semantics for the language}\label{fig:lang-semantics}
+\end{center}
+\end{figure*}
+
+%==============================================================
+\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
+
+\section{Liveness}\label{sec:liveness}
+
+A variable is {\em live} it  there is a possibility of its value being
+used  in future computation  and dead  if it  is definitely  not used.
+Heap-allocated data needs a richer model of liveness which talks about
+the liveness  of pointers including  fields of $\CONS$ cells.   We use
+$\acar$, $\acdr$ to represent  access using $\CAR$, $\CDR$ fields. The
+liveness of the structure reachable  from a variable is represented by
+a  set  of  {\em   access  paths}  i.e.   prefix-closed  strings  from
+$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
+$\{\epsilon,  \acdr, \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
+future computations can  only refer up to the  second and third members
+of $x$.  The classical notions of a scalar variable being live or dead
+correspond to $\{\epsilon\}$ and $\{\}$.
+
+A {\em liveness environment} is a mapping from variables to subsets of
+$\{\acar,\acdr\}^\ast$. But  we often express it as a  set, for
+example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$ instead of $[x
+  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
+  z\mapsto\{\}]$.   For  eager  languages,  we  associate  a  liveness
+environment  with each program  point; the  liveness at  program point
+$\pi$ in $\pi:e$ is the liveness just before executing $e$. Due to the
+inherent uncertainty  in determining the  point of evaluation  in lazy
+languages, we  associate liveness  environments with certain  kinds of
+paths in the program. All programs  points lying on the path share the
+same liveness environment.  We explain this point later.
+
+A complementary  notion to liveness  is {\em demand}.  The  demand for
+expression  $e$   is  again  an   access  path---that  subset   of  of
+$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
+result.  So,  for example given  a demand $\sigma$ and  the expression
+$\pi: (\RETURN\ x)$, the liveness at $\pi$ is exactly $x.\sigma$.  The
+classical  analogy of  this  is  in {\em  strong  liveness}, where  an
+assignment node $n: x:=y+z$ causes $y$  and $z$ to be live on entry to
+$n$ if (and only if) $x$ is  live at exit of $n$---the liveness of $x$
+at  exit from  $n$ becomes  the demand  on $y+z$.
+
+%% Note  that, for  an  operation  like division  which  may raise  an
+%% exception,  the  assignment $n:  x:=y/z$  makes  $y$  and $z$  live
+%% regardless of the liveness of $x$.
+
+We use $\sigma$  to range over demands, $\alpha$  to range over access
+paths  and $\Lv$  to range  over liveness  environments.  The notation
+$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
+\alpha_1 \in  \sigma_1, \alpha_2 \in \sigma_2\rbrace$.  Often we shall
+abuse notation to  juxtapose an edge label and a  set of access paths:
+$\acar\sigma$   is   a   shorthand  for   $\lbrace\acar\rbrace\sigma$.
+Finally,  we use $\Lfonly$  to range  over {\em  demand transformers}.
+These transform demands on a function call into demands for its formal
+parameters:    given    a    user    function    $f$,    defined    by
+$(\DEFINE\  (f\ x_1\  \ldots\ x_n)\  \  e_f)$ and  called with  demand
+$\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
+
+Note that liveness refers to  variables and fields, and not to $\CONS$
+cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
+themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
+future computation may refer to  the value $\ell$ of variable $x$, and
+also to the $\CAR$ field of  location $\ell$.  In the absence of other
+pointers to heap location $\ell$, we are certain that the $\CDR$ field
+of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
+arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
+collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
+would  be  garbage (again  provided  there  are  no other  aliases  to
+$\ell$).
+
+\bigskip
+\hrule 
+\bigskip
+
+\subsection{Liveness Analysis}
+
+To describe  the issues  in formulating liveness  of a  lazy language,
+consider the function definition shown::
+
+\renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} (\DEFINE\ (\pf~\pa)
+	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
+	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
+          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
+          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
+	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
+          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
+	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
+	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
+          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
+	\end{uprogram}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Consider a call to $\pf$  with a demand $\sigma$. For obvious reasons,
+this is also the demand on the return values $u$ and $y$, and thus the
+liveness  at  the  program  points  $\pi_7$ and  $\pi_8$  re  $\lbrace
+\pu.\sigma \rbrace$ and $\{\py.\sigma\}$.  The liveness at $\pi_7$ and
+$\pi_8$ are propagated backwards as in traditional liveness analysis.
+
+Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
+\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
+turn generates  a liveness  of $\{\pw,~\pw.1\sigma\}$ at  $\pi_6$. The
+liveness of  \pu\ is  also killed at  $\pi_6$ since  it is not  in the
+scope of  \pu. However, for simplicity  of exposition, we  do not kill
+the variable in our  formulation.  This does not introduce imprecision
+since  the liveness  of  \pu\ will  not  be consulted  by the  garbage
+collector at $\pi_6$.
+%==============================================================
+\begin{figure*}[t]  
+\begin{eqnarray*}
+\mathit{ref\/}(\kappa,\sigma,\Lfonly)
+          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
+\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
+          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
+\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
+%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
+          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
+               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
+               \end{array}
+%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
+\\[1ex]
+\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
+\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
+        \begin{array}{l l}
+                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
+        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
+        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
+        \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+ \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
+\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
+        \in \mathcal{M}\}
+\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
+\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
+ \makebox[0mm]{\hspace*{7cm}   
+ $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
+           x_2\, \ldots\, x_n)/x]))$,} \\
+ \makebox[0mm]{\hspace*{5.5cm} and 
+     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
+\end{eqnarray*}
+\begin{minipage}{0.85\textwidth}
+\infrule[live-define]
+        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
+           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
+              \mbox{ for each $f$ and $\sigma$}
+        }
+        { d_1 \ldots d_k \len \Lfonly
+\\ \makebox[0mm]{where
+     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
+\end{minipage}
+  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
+\end{figure*}
+%======================================================================================
+The liveness of lazy and eager evaluation differ in two respects:
+\begin{enumerate}
+\item  In an  eager  language, \pw\  would  not have  been live  after
+  $\pi_6$, the point of its last use. In a lazy language, however, the
+  definition  $\pu  \leftarrow  (\CDR~\pw)$,  does not  result  in  an
+  evaluation  of  $(\CDR~\pw)$; instead  a  closure  is created.   The
+  evaluation actually takes place while evaluating $(\RETURN~u)$. Thus
+  it is necessary to record the liveness of $\pw$ at the program point
+  $\pi_8$.
+
+\item The second difference is that  our \LET s can be recursive as in
+  program point $\pi_1$.  In such a case, the liveness of a let has to
+  be formulated as a fix-point computation.
+\end{enumerate}
+
+  To address the first issue, we  regard the body of the function as a
+  tree and consider paths from the root of this tree to program points
+  which     trigger     a     evaluation---\SIF\    conditions     and
+  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
+  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
+  \pi_3,  \pi_4, \pi_5$, $EP_1  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
+  \pi_7, \pi_8$, and $EP_2 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
+  analysis consists of two maps:  The map $\cal{M}$ maps a return path
+  to a liveness  environment that is common to  all the program points
+  in the path. And a second  map $\cal{P}$ maps a program point to the
+  set of  return paths on  which the point  lies.  If a  program point
+  happens  to fall  on  more than  one  such path,  then the  liveness
+  environment of the  program point is the variable-wise  union of the
+  liveness  environments  of the  individual  paths.   In the  example
+  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
+  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
+  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.
+
+  In  the  sequel we  shall  describe  how  to obtain  $\cal{M}$.  The
+  formulation  of $\cal{P}$  is easy  and we  do not  describe  it any
+  further.
+
+\bigskip
+\hrule 
+\bigskip
+
+First recall the classical formulation  of liveness (as sets of simple
+variables)        on        three-address       instructions,        $
+\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
+\mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
+  strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
+$\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
+\in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
+
+Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
+analogous.
+% has three analogous aspects.
+% \begin{itemize}
+% \item
+Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
+returns the  incremental liveness {\em generated} by  $s$.  Because we
+generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
+parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
+determine which access paths of its free variables are made live.  The
+cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
+% The cases for
+% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
+% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
+% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
+A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
+$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
+argument (even if its result is never used).
+This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
+(note $\sigma$ may be $\{\}$).
+In the opposite  sense, the  demand  of $\acar\sigma$  on
+(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
+that \CONS\ does not, by itself, dereference its arguments.
+Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
+$\mathit{ref}$
+expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
+Constants generate no liveness.
+
+The  function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
+expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
+but note the liveness $x.\epsilon$  generated by the latter.  The case
+$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
+instruction:  the liveness  of $e$  is given  by taking  the liveness,
+$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
+liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
+$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
+$\mathit{ref}(s,\Lv(z),\Lfonly)$.
+
+% \item
+%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
+Finally, the  judgement form $\mathit{Prog}  \len \Lfonly$ is  used to
+determine  $\Lfonly$\@.   Analogously   to  classical  liveness  being
+computed  as  a  solution  of  dataflow  equations,  we  require,  via
+inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
+fixed-point property that:  when we assume $\Lfonly$ to  be the family
+of demand transformers for the program then the calculated liveness of
+each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
+assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
+$\Lfonly$;  all are  safe (see  Section~\ref{sec:correctness})  but we
+prefer  the  least  one  as  giving  the  least  liveness  subject  to
+safety---and hence greatest amount of garbage collected.
+% \end{itemize}
+
+We make three observations: firstly the rule ({\sc live-define}) has a
+least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
+secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
+inference of mutually recursive  function definitions, and thirdly the
+asymmetry of  demand and liveness (compared to  post- and pre-liveness
+classically) is due to the functional formulation here.
+
+Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
+(representing a fully context-sensitive analysis)
+can be safely approximated, for each function, by a {\em procedure summary}
+(unifying the contexts in the style of 0-CFA)\@.
+The summary consists of a pair of a single demand and, for this demand,
+the corresponding tuple of demands the function makes on its
+arguments.
+
+%\comment{
+
+%}
+
+% \item
+
+\subsection{References}
+\bibliography{fun_hra}{}
+\bibliographystyle{abbrv}
+
+Generated by bibtex from your ~.bib file.  Run latex,
+then bibtex, then latex twice (to resolve references)
+to create the ~.bbl file.  Insert that ~.bbl file into
+the .tex source file and comment out
+the command \texttt{{\char'134}thebibliography}.
+% This next section command marks the start of
+% Appendix B, and does not continue the present hierarchy
+\section{More Help for the Hardy}
+The sig-alternate.cls file itself is chock-full of succinct
+and helpful comments.  If you consider yourself a moderately
+experienced to expert user of \LaTeX, you may find reading
+it useful but please remember not to change it.
+%\balancecolumns % GM June 2007
+% That's all folks!
 \end{document}

### 2015-03-13 paper/paper.tex
update paper.tex
Updated section on liveness analysis for lazy languages

@@
 \newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
-%\nonstopmode
 \def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
@@
 \begin{document}
-%
 % --- Author Metadata here ---
@@
 
-Consider a call to $\pf$  with a demand $\sigma$. For obvious reasons,
-this is also the demand on the return values $u$ and $y$, and thus the
-liveness  at  the  program  points  $\pi_7$ and  $\pi_8$  re  $\lbrace
+The concept of demand and liveness for lazy languages are almost similar to their 
+counterparts in the eager language. The important distinction comes from the fact 
+that in eager languages, variables can point only to data values (either atomic values or cons cells). 
+Hence any demand on the variable can easily be transferred onto the data being pointed. 
+In lazy languages, a variable can either point to a fully evaluated data value or it 
+could point to a closure which on evaluation generates the required data value. For example, 
+consider the statement {\tt let \py~ $\leftarrow $ (\CDR~\px)}. In case of an eager language 
+the variable \py~ points to  the \CDR~ part of \px~ which is guaranteed to be a cons cell. 
+Hence any demand on \py~ can be immediately transferred to the \CDR~ part of \px. In case of 
+a lazy language, \px~ might not be evaluated and  hence pointing to a closure. In our 
+example, it points to the closure corresponding to (\CONS~1~\px). It does not make any sense 
+to transfer the demand onto the \CDR~ part of \px~ as it still does not exist. In such a case 
+we treat the closure as a demand transformer and tranform the demand on \py~ to the arguments 
+of the closure pointed by \px. This procedure is similar to the process of transferring the demand 
+from the return value of a function to the arguments of the function call in an eager language. In case 
+of lazy languages this process cannot be done statically as we cannot statically determine whether a variable points to 
+a data value or a closure. Our method approximates the liveness of closure arguments at runtime 
+by tagging extra information with each closure. Since the arguments of the closures could themselves be closures in turn, 
+we keep recursively applying the same procedure until we hit a data value. The only case where we do not 
+follow this procedure is if the demand on a particular closure is null. In such a case we safely collect 
+the entire closure as garbage.
+
+We now describe the process of finding out the liveness 
+of each variable defined in the function $\pf$ assuming a symbolic demand $\sigma$ on it. 
+For obvious reasons, this is also the demand on the return values $u$ and $y$, and thus the
+liveness  at  the  program  points  $\pi_7$ and  $\pi_8$  are  $\lbrace
 \pu.\sigma \rbrace$ and $\{\py.\sigma\}$.  The liveness at $\pi_7$ and
@@
 \item  In an  eager  language, \pw\  would  not have  been live  after
-  $\pi_6$, the point of its last use. In a lazy language, however, the
+  $\pi_7$, the point of its last use. In a lazy language, however, the
   definition  $\pu  \leftarrow  (\CDR~\pw)$,  does not  result  in  an
@@
   program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
-  \pi_3,  \pi_4, \pi_5$, $EP_1  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
-  \pi_7, \pi_8$, and $EP_2 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
+  \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
+  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
   analysis consists of two maps:  The map $\cal{M}$ maps a return path

### 2015-03-13 lazy-scheme/src/gc.cpp

@@ -60,7 +60,6 @@ vector<var_heap>::iterator heapRefIter;
 vector<var_stack>::iterator nonRefIter;
 stateMap statemap;
 state_index **state_transition_table; //Dynamically allocating a 2-D array might be error prone. See if vectors can be used instead.
-//call_next_data call_next; //Should we use a map?
 std::unordered_map<std::string, std::string> call_next;
 stack<cons*> update_heap_refs;
 stack<cons*> print_stack;
@@ -70,15 +69,12 @@ map<int, string> root_var_map;
 
 #ifdef __DEBUG__GC
 ofstream gcout("gc_messages.txt", ios::out);
-#define P(x) x;
 #else
 ostream &gcout = null_stream;
-#define P(x) (void)0;
 #endif
 
-//void reachable_dfs();
+
 void clear_liveness_data();
-//void print_heap();
 void print_buffer_data();
 void print_cell_data();
 int is_valid_address(void*);
@@ -86,13 +82,11 @@ int lt_scan_free();
 int lt_scan_freept();
 void* getscan();
 void update_scan();
-//long diff_scan_free();
 state_index lookup_dfanode(string nodename);
-//void* followpaths(void* loc, state_index index);
 state_index get_target_dfastate(state_index i1, state_index i2);
 void set_car(void* loc,  void* ref);
 void set_cdr(void* loc,  void* ref);
-//void copy_children(void* cellptr, stateset* st, ostream&  out = null_stream);
+
 
 #ifdef GC_ENABLE_STATS
 /* declarations of GC statistics related functions */
@@ -140,6 +134,86 @@ void depthfirstpaths(Scheme::AST::cons* loc, Scheme::AST::state_index index);
 
 
 
+
+#define FIND_REACHABLE
+#ifdef FIND_REACHABLE
+void reach_dfs(cons* loc, vector<cons*>& v)
+{
+	if (NULL == loc)
+		return;
+
+	if (find(v.begin(), v.end(), loc) == v.end())
+		v.push_back(loc);
+	else
+		return;
+
+	if (loc->typecell == consExprClosure)
+	{
+
+		reach_dfs(loc->val.cell.car, v);
+
+		reach_dfs(loc->val.cell.cdr, v);
+	}
+	else
+	{
+		switch(loc->typecell)
+		{
+		case constIntExprClosure:
+		case constBoolExprClosure:
+		case constStringExprClosure:
+		case nilExprClosure:
+			break;
[…]

### 2015-03-13 lazy-scheme/src/gc.h

@@ -40,7 +40,7 @@ using namespace Scheme::AST;
  * Author : Amey Karkare
  * */
 //#define GC_ENABLE_STATS
-#if 0
+#if __DEBUG__GC
 #define DBG(stmt) stmt
 #else
 #define DBG(stmt) (void)0

### 2015-03-13 lazy-scheme/src/gc.cpp
optimizing lgc

@@ -229,8 +229,8 @@ void print_gc_move(cons* from, cons* to, ostream& out)
 		if (to)
 			to_index = (to - (cons*)buffer_live);
 
-		out << from_index << "-->" << to_index << "(" << to << ")" << endl;
-		out << "Type of cell " << from->typecell << endl;
+		DBG(out << from_index << "-->" << to_index << "(" << to << ")" << endl);
+		DBG(out << "Type of cell " << from->typecell << endl);
 	}
 	return;
 }
@@ -1575,9 +1575,9 @@ void liveness_gc()
 	  ostream &pre = null_stream;
 #endif
 	  DBG(pre << "Doing liveness based GC #" << gccount << " after " << num_of_allocations << " allocations"<<endl);
-#ifdef FIND_REACHABLE
-	  size_t num_reachable_cells = find_num_reachable_cells();
-#endif
+//#ifdef FIND_REACHABLE
+//	  size_t num_reachable_cells = find_num_reachable_cells();
+//#endif
 
 	  swap_buffer();
 
@@ -1623,10 +1623,10 @@ void liveness_gc()
   double gc_time = ((double(pend - pstart)/CLOCKS_PER_SEC));
 //  cerr << "GC Time for " << gccount << " = " << gc_time << endl;
   gctime += gc_time;
-#ifdef FIND_REACHABLE
-  cout << "Num of reachable cells = " << num_reachable_cells << endl;
-  cout << "Num of live cells = " << numcopied << endl;
-#endif
+//#ifdef FIND_REACHABLE
+//  cout << "Num of reachable cells = " << num_reachable_cells << endl;
+//  cout << "Num of live cells = " << numcopied << endl;
+//#endif
   return;
 }
 
@@ -1646,7 +1646,7 @@ cons* followpaths(cons* loc, state_index index, ostream& out)
 #endif
   //Can we add a check here too to find out if a cons cell has been copied using reachability?
   //That might improve the efficiency
-  if (loc->typecell == consExprClosure)
+  if (loc->typecell == consExprClosure && loc->copied_using_rgc == false)
   {
 	  state_index a0 = state_transition_table[index][0]; //get_target_dfastate(index, 0);
 	  if (a0 > 0)

### 2015-03-13 lazy-scheme/src/gc.cpp
Added code to measure the number of reachable cells in each GC

@@ -201,6 +201,7 @@ size_t find_num_reachable_cells()
 	while(!print_stack.empty())
 	{
 		cons* heap_ref = print_stack.top();
+		print_stack.pop();
 		temp.push(heap_ref);
 		reach_dfs(heap_ref, reachable_cells);
 	}
@@ -1575,9 +1576,10 @@ void liveness_gc()
 	  ostream &pre = null_stream;
 #endif
 	  DBG(pre << "Doing liveness based GC #" << gccount << " after " << num_of_allocations << " allocations"<<endl);
-//#ifdef FIND_REACHABLE
-//	  size_t num_reachable_cells = find_num_reachable_cells();
-//#endif
+#ifdef FIND_REACHABLE
+	  numcopied = 0;
+	  size_t num_reachable_cells = find_num_reachable_cells();
+#endif
 
 	  swap_buffer();
 
@@ -1623,10 +1625,10 @@ void liveness_gc()
   double gc_time = ((double(pend - pstart)/CLOCKS_PER_SEC));
 //  cerr << "GC Time for " << gccount << " = " << gc_time << endl;
   gctime += gc_time;
-//#ifdef FIND_REACHABLE
-//  cout << "Num of reachable cells = " << num_reachable_cells << endl;
-//  cout << "Num of live cells = " << numcopied << endl;
-//#endif
+#ifdef FIND_REACHABLE
+  cout << "Num of reachable cells = " << num_reachable_cells << endl;
+  cout << "Num of live cells = " << numcopied << endl;
+#endif
   return;
 }
 

### 2015-03-13 lazy-scheme/src/run_in_loop.py
added ratio of RGC and LGC

@@ -29,6 +29,7 @@ def collect_gc_results(filename, prog_dir_path, gc_type):
     heap_total_cmd = grep_cmd + " \"Heap total=[0-9]* \"  " + result_file_name  + int_grep
     heap_left_cmd = grep_cmd + " \"Heap left=[0-9]*\"  " +  result_file_name  + int_grep
     heap_used_cmd = grep_cmd + " \"Heap used=[0-9]*\"   " +  result_file_name  + int_grep
+A
     gc_invocations_cmd = grep_cmd + " \"GC Invocations=[0-9]*\"   " +  result_file_name  + int_grep
     gc_time_cmd = grep_cmd + " \"GC Time=[0-9]*(\.[0-9]*)?\"   " +  result_file_name  + float_grep
     prog_execution_time_cmd = grep_cmd + " \"Program Execution Time=[0-9]*(\.[0-9]*)?\"   " +  result_file_name  + float_grep
@@ -82,11 +83,17 @@ def main():
         f.write('\\begin{document}\n')
         f.write('\\begin{center}\n')
         f.write('\\begin{tabular}\n')
-        f.write('{| l | l | l | l | l |}\n')
+        f.write('{| l | r | r | r | r | r |}\n')
         f.write('\\hline\n')
-        f.write('Prog Name & Num RGC & Num LGC & Avg. RGC Time & Avg. LGC Time \\\\ \\hline\n')
+        f.write('Prog Name & Num RGC & Num LGC & Avg. RGC Time & Avg. LGC Time & RGC/LGC \\\\ \\hline\n')
         for g in gc_stat_list:
-            f.write('\\verb@' + g.prog_name + '@'  + '&' + str(g.reachability.gc_invocations/float(numTimes)) + '&' + str(g.liveness.gc_invocations/float(numTimes)) + '&' + str(g.reachability.gc_time/float(numTimes)) + '&' + str(g.liveness.gc_time/float(numTimes)) + '\\\\ \\hline\n') 
+            f.write('\\verb@' + g.prog_name + '@'  + 
+                    '&' + str(g.reachability.gc_invocations/float(numTimes)) +
+                    '&' + str(g.liveness.gc_invocations/float(numTimes)) +
+                    '&' + str(g.reachability.gc_time/float(numTimes)) +
+                    '&' + str(g.liveness.gc_time/float(numTimes)) +
+                    '&' + str(g.liveness.gc_time/g.reachability.gc_time) +
[…]

### 2015-03-14 paper/paper.tex
Updated section on liveness analysis for lazy languages
Updated section on liveness analysis

@@
 
-\subsection{Liveness Analysis}
+\subsection{Liveness Analysis for lazy languages}
 
@@
 
-The concept of demand and liveness for lazy languages are almost similar to their 
-counterparts in the eager language. The important distinction comes from the fact 
-that in eager languages, variables can point only to data values (either atomic values or cons cells). 
-Hence any demand on the variable can easily be transferred onto the data being pointed. 
-In lazy languages, a variable can either point to a fully evaluated data value or it 
+The main challenges in doing liveness analysis for lazy languages are, 
+1) a variable can either point to a fully evaluated data value or it 
 could point to a closure which on evaluation generates the required data value. For example, 
@@
 example, it points to the closure corresponding to (\CONS~1~\px). It does not make any sense 
-to transfer the demand onto the \CDR~ part of \px~ as it still does not exist. In such a case 
-we treat the closure as a demand transformer and tranform the demand on \py~ to the arguments 
-of the closure pointed by \px. This procedure is similar to the process of transferring the demand 
-from the return value of a function to the arguments of the function call in an eager language. In case 
-of lazy languages this process cannot be done statically as we cannot statically determine whether a variable points to 
-a data value or a closure. Our method approximates the liveness of closure arguments at runtime 
+to transfer the demand onto the \CDR~ part of \px~ as it still does not exist.
+, 2) it is not possible to statically determine when a closure gets evaluated.  Therefore 
+during GC we treat a closure as a demand transformer and tranform the demand on to the arguments 
+of the closure pointed. This is similar to the process of transferring the demand 
+from the return value of a function to the arguments of the function call in an eager language.
+
+In our example, during GC if we find that \px~ points to a closure instead of a cons cell we 
+find out the demand/liveness on the arguments of the closure, namely \py.
+Statically it cannot be determined what values a closure points to or when it gets evaluated,
+hence our method approximates the liveness of closure arguments at runtime 
 by tagging extra information with each closure. Since the arguments of the closures could themselves be closures in turn, 
-we keep recursively applying the same procedure until we hit a data value. The only case where we do not 
+we recursively apply the same procedure until we hit a data value. The only case where we do not 
 follow this procedure is if the demand on a particular closure is null. In such a case we safely collect 
@@
   further.
+  We observe that the ${ref}$ rules neither use ${\cal M}$ directly nor 
+  create new labels. We only need to consider the rules for {\LET}, {\SIF}, and
+  {\RETURN}.
+
+  A {\RETURN~\px}\  statement always forces an evaluation and hence a new 
+  label {\em l} needs to be created and added to $\cal{M}$. A new liveness
+  environment is associated with {\em l}. The demand $\sigma$ that is placed 
+  on this expression is now transferred to the variable \px. {\em l} will now
+  contain an entry for \px~ with liveness $\sigma$.
+
+  An \SIF~ statement will modify  $\cal{M}$ by doing a union of 
+  the maps from the {\em then} and the {\em else} part. Note that each part can contain 
+  more than one label. Since the variable corresponding to the condition may not
+  be part of either the {\em then} or the {\em else} part, a new label {\em l} is created 
+  and added to $\cal{M}$. The only liveness associated with {\em l} is the liveness 
+  of the condition variable.
+
+  The processing rule for {\LET} is the most complex among all the rules. A {\LET} does
+  not add any new label but might modify the liveness values of some variables or might add 
+  liveness values for the let bound variable. Since there might be more than one {\RETURN} 
+  statement following a {\LET}, there might be multiple {\em E-paths} reaching the {\LET}.
+  The demand on the let variable might be different along different {\em E-paths}. Therefore,
+  for each label {\em l} in $\cal{M}$, the liveness of the let-expression is calculated using 
+  the liveness environment $\Lv$ corresponding to {\em l}. Since the {\LET} is lazy, 
+  calculating its livness might require a fixed-point processing.
 

### 2015-03-15 paper/paper.tex
pulled parts written by prasanna

--- 2015-03-13 paper/paper.tex
@@
 Heap-allocated data needs a richer model of liveness which talks about
-the liveness  of pointers including  fields of $\CONS$ cells.   We use
-$\acar$, $\acdr$ to represent  access using $\CAR$, $\CDR$ fields. The
-liveness of the structure reachable  from a variable is represented by
-a  set  of  {\em   access  paths}  i.e.   prefix-closed  strings  from
-$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
-$\{\epsilon,  \acdr, \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
-future computations can  only refer up to the  second and third members
-of $x$.  The classical notions of a scalar variable being live or dead
-correspond to $\{\epsilon\}$ and $\{\}$.
+the  liveness of pointers  including fields  of $\CONS$  cells.  Using
+$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
+the  liveness  of the  structure  reachable  from  a variable  can  be
+represented by a set of {\em access paths} i.e.  prefix-closed strings
+from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
+liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
+\acdr\acdr\acar\}$, then future computations  can only refer up to the
+second and third  members of $x$. In this  notation, a scalar variable
+being live or dead are represented as $\{\epsilon\}$ and $\{\}$ respectively.
 
 A {\em liveness environment} is a mapping from variables to subsets of
-$\{\acar,\acdr\}^\ast$. But  we often express it as a  set, for
-example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$ instead of $[x
+$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
+writing  $\{x.\acar\acdr,   x.\acdr,  y.\epsilon\}$  instead   of  $[x
   \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
-  z\mapsto\{\}]$.   For  eager  languages,  we  associate  a  liveness
-environment  with each program  point; the  liveness at  program point
-$\pi$ in $\pi:e$ is the liveness just before executing $e$. Due to the
-inherent uncertainty  in determining the  point of evaluation  in lazy
-languages, we  associate liveness  environments with certain  kinds of
-paths in the program. All programs  points lying on the path share the
-same liveness environment.  We explain this point later.
-
-A complementary  notion to liveness  is {\em demand}.  The  demand for
+  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  each
+program  point $\pi$:  the liveness  at $\pi:e$  is the  liveness just
+before executing $e$.
+%%
+%% Due to the inherent uncertainty in
+%% determining the  point of evaluation  in lazy languages,  we associate
+%% liveness environments with certain kinds  of paths in the program. All
+%% programs points lying on the path share the same liveness environment.
+%% We explain this point later.
+%% 
+A  notion  related  to  liveness  is {\em  demand}.   The  demand on
 expression  $e$   is  again  an   access  path---that  subset   of  of

### 2015-03-15 paper/paper.tex
"Liveness analysis - clarified the difference in LA of eager and lazy"

@@
 To describe  the issues  in formulating liveness  of a  lazy language,
-consider the function definition shown::
+consider the function definition shown:
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\medskip
 
-The main challenges in doing liveness analysis for lazy languages are, 
-1) a variable can either point to a fully evaluated data value or it 
-could point to a closure which on evaluation generates the required data value. For example, 
-consider the statement {\tt let \py~ $\leftarrow $ (\CDR~\px)}. In case of an eager language 
-the variable \py~ points to  the \CDR~ part of \px~ which is guaranteed to be a cons cell. 
-Hence any demand on \py~ can be immediately transferred to the \CDR~ part of \px. In case of 
-a lazy language, \px~ might not be evaluated and  hence pointing to a closure. In our 
-example, it points to the closure corresponding to (\CONS~1~\px). It does not make any sense 
-to transfer the demand onto the \CDR~ part of \px~ as it still does not exist.
-, 2) it is not possible to statically determine when a closure gets evaluated.  Therefore 
-during GC we treat a closure as a demand transformer and tranform the demand on to the arguments 
-of the closure pointed. This is similar to the process of transferring the demand 
-from the return value of a function to the arguments of the function call in an eager language.
+Liveness analysis for lazy languages differ from eager languages in
+two respects:
+\begin{enumerate}
+\item Unlike eager languages, a variable may be  bound to an
+  expression that is not fully evaluated. i.e. a closure.
+\item The point where an expression gets evaluated cannot be statically
+  determined.
+\end{enumerate}
 
-In our example, during GC if we find that \px~ points to a closure instead of a cons cell we 
-find out the demand/liveness on the arguments of the closure, namely \py.
-Statically it cannot be determined what values a closure points to or when it gets evaluated,
-hence our method approximates the liveness of closure arguments at runtime 
-by tagging extra information with each closure. Since the arguments of the closures could themselves be closures in turn, 
-we recursively apply the same procedure until we hit a data value. The only case where we do not 
-follow this procedure is if the demand on a particular closure is null. In such a case we safely collect 
-the entire closure as garbage.
+To  deal  with the  first  issue, we  apply  the  results of  liveness
+analysis only  to fully evaluated  data.  If the marking  phase during
+garbage collection leads to a closure, and the closure is not dead, we
+mark the entire reachable memory  under the closure for collection.  A
+better scheme  would be  to garbage collect  the closure based  on the
+liveness  at the point  of its  creation.  This  is sound  because our
+language is statically scoped.  However, this is work in progress, and
+the method described in this paper switches over to reachability while
+collecting closures.
 
-We now describe the process of finding out the liveness 
-of each variable defined in the function $\pf$ assuming a symbolic demand $\sigma$ on it. 
-For obvious reasons, this is also the demand on the return values $u$ and $y$, and thus the
-liveness  at  the  program  points  $\pi_7$ and  $\pi_8$  are  $\lbrace
-\pu.\sigma \rbrace$ and $\{\py.\sigma\}$.  The liveness at $\pi_7$ and
-$\pi_8$ are propagated backwards as in traditional liveness analysis.
+%% However, we GC,  if we find a variable can either  point to a fully
+%% evaluated  data value  or  it could  point  to a  closure which  on
+%% evaluation  generates  the   required  data  value.   For  example,
+%% consider the statement {\tt  let \py~ $\leftarrow $ (\CDR~\px)}. In
+%% case of  an eager  language the variable  \py~ points to  the \CDR~
+%% part of  \px~ which  is guaranteed  to be a  cons cell.   Hence any
+%% demand on \py~ can be  immediately transferred to the \CDR~ part of
+%% \px.  In case  of a lazy language, \px~ might  not be evaluated and
+%% hence  pointing to  a closure.  In our  example, it  points  to the
+%% closure corresponding to (\CONS~1~\px).  It does not make any sense
+%% to transfer the demand onto the \CDR~ part of \px~ as it still does
+%% not exist.  , 2) it is  not possible to statically determine when a
+%% closure gets evaluated.  Therefore during  GC we treat a closure as
+%% a demand transformer and tranform the demand on to the arguments of
+%% the  closure   pointed.   This  is   similar  to  the   process  of
+%% transferring the demand from the  return value of a function to the
+%% arguments of the function call in an eager language.
+
+%% In our example, during GC if  we find that \px~ points to a closure
+%% instead  of a  cons cell  we find  out the  demand/liveness  on the
+%% arguments  of the  closure, namely  \py.  Statically  it  cannot be
+%% determined  what  values  a  closure  points to  or  when  it  gets
+%% evaluated, hence  our method  approximates the liveness  of closure
+%% arguments  at  runtime  by  tagging  extra  information  with  each
+%% closure.  Since  the arguments of the closures  could themselves be
+%% closures in turn, we recursively  apply the same procedure until we
+%% hit  a data  value.  The only  case  where we  do  not follow  this
+%% procedure is if the demand on a particular closure is null. In such
+%% a case we safely collect the entire closure as garbage.
+
+To see  our way of  handling the second  issue, we trace  the liveness
+analysis to  an extent. Assume  that the calls  to the $\pf$  place an
+overall demand  of $\sigma$ on the  body of \pf.  on  the now describe
+the process  of finding out the  liveness of each  variable defined in
+the function  $\pf$ assuming  a symbolic demand  $\sigma$ on  it.  For
+obvious reasons, this is also the  demand on the return values $u$ and
+$y$, and thus  the liveness at the program  points $\pi_7$ and $\pi_8$
+are $\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The liveness
+at  $\pi_7$ and  $\pi_8$ are  propagated backwards  as  in traditional
+liveness analysis.
 
@@
 the variable in our  formulation.  This does not introduce imprecision
-since  the liveness  of  \pu\ will  not  be consulted  by the  garbage
-collector at $\pi_6$.
+since  \pu\  is yet  to  come into  existence at  $\pi_6$ and  its
+liveness will not be consulted by the garbage collector at this point.
 %==============================================================
@@
 %======================================================================================
-The liveness of lazy and eager evaluation differ in two respects:
-\begin{enumerate}
-\item  In an  eager  language, \pw\  would  not have  been live  after
+In an  eager  language, \pw\  would  not have  been live  after
  $\pi_7$, the point of its last use. In a lazy language, however, the
@@
   $\pi_8$.
-
-\item The second difference is that  our \LET s can be recursive as in
-  program point $\pi_1$.  In such a case, the liveness of a let has to
-  be formulated as a fix-point computation.
-\end{enumerate}
-
-  To address the first issue, we  regard the body of the function as a
+%% The second  difference is that  our \LET s  can be recursive  as in
+%%   program point $\pi_1$.  In such a case, the liveness of a let has
+%%   to be formulated as a fix-point computation.
+  To address this issue, we  regard the body of the function as a
   tree and consider paths from the root of this tree to program points

### 2015-03-16 paper/paper.tex
Added implementation section

@@
 
+\section{Implementation}
+% GC assumptions
+By the time the program starts executing, the liveness automata are created and stored in a file. 
+This is re-used every time the program executes. 
+% GC points - mention reduction of gc points due to e-paths
+An important difference in GC of lazy language compared to an eager language is the point of 
+garbage collection. In case of eager languages only \CONS~ points and function calls are 
+the GC points. In a lazy language every \LET~ statement is a potential GC point. Hence we need to store 
+liveness automata corresponding to every \LET~ statement in the program. This leads to an explosion 
+in the number of automata. From our observation that along an {\em E-path} all program points have 
+the same liveness, we can reduce the number of automata state required by sharing the same automata 
+for all program points along an {\em E-path}. In the example program, we only need to create automata 
+for program points \ldots, since they are the only distinct {\em E-path}s in the program. 
+% Abstract description of the GC algorithm
+% Liveness for cons cells & reachability for closures
+During GC, if the heap cell being processed is a cons cell, we consult the liveness automaton 
+corresponding to it and copy only the live paths. In case the heap cell points to a closure, 
+we safely approximate the liveness to be {\{0+1\}\*} and copy everything under the closure.
+% Handling of mixture of LGC & RGC
+A consequence of copying everything under a closure is the fact that we can end up trying to copy 
+things which are not live and corrupt the heap. This can happen if at some $GC_k$ a cons cell is copied 
+using liveness. At a later stage this cons cell becomes part of a closure and hence during $GC_{k+n}$ the 
+algorithm tries to copy everything under the cons cell thus copying heap locations which might not be valid. 
+To avoid this behavior we either have to choose to do a full liveness based GC for closure or we should 
+ensure that any non-live references are not carried across GC's. We take the latter approach and ensure that 
+non-live references are set to null. After every GC, the live part of the buffer is scanned and any reference 
+which is not live is set to null. The correctness of this follows from the fact that any reference which is 
+decalred not live in a $GC_k$ will not be derefernced during later execution.
+% A reference copied using LGC could become part of a closure in a later GC
+% and copy wrong locations. Avoiding it by setting non-live references to null after every GC.
+% Proof for the correctness. 
+% Difference between eager and lazy GC
+Another difference that is present in GC of lazy lanaguage is the presence of a print stack. In lazy 
+languages the evaluation of closures is driven by the print function. Hence, during a GC there will be 
+heap references that will be present on the stack which also have to copied/collected. One solution is to 
+assume that all references present on the print stack are fully live and copy everything pointed by them 
+akin to reachability. This is safe but copies heap cells which have already been printed. Since the print 
+function is not part of the executing program its liveness automata are not considered. Therefore we 
+simulate the liveness of the print function by using a flag in the heap cell. The flag ensures that only the 
+required parts of a cons cell are copied. The behavior of the GC algorithm is as described below,
+\begin{verbatim}
+copy cons cell to new loc
+if flag == false
+  copy cdr part of the cell
+  update cdr part of new loc
+end if
+\end{verbatim}
+We provide an intuition about the correctness of this algorithm. Consider the following facts about the print stack 
+during a GC,
+\begin{enumerate}
+\item The top of the stack always contains a reference to a closure.
+\item All other references on the print stack point to atomic values or cons cells, it can never contain a
+reference to a closure. The components of the cons cell can contain reference to closures.
+\item If a reference on the print stack is a cons cell, either its car part or its cdr part will also be 
+present on the stack.
+\end{enumerate}
+%Needs to be formatted and re-written
+Sketch of the proof
+Since all closures are copied fully and atomic values are copied, only cons cells need to be handled. 
+Consider the following cases,
+\begin{enumerate}
+\item the car part of the cons cell is on the print stack. This means that the GC happened while 
+the car part was being processed. The cdr part is yet to be processed and the reference to the cdr part 
+is not on the print stack. Hence it is sufficient to copy only the cdr part of the cons cell. Any references 
+necessary for processing the car part will be copied due to references on the print stack.
+\item the cdr part of the cons cell is on the print stack. This means that GC happened while 
+the cdr part was being processed.  The car part has already been processed and printed, hence we do not 
+need to copy the car part. Additionally any references that are needed for printing the cdr part will 
+also be copied due to the references on the print stack and hence nothing needs to be copied in this case.
+\end{enumerate}
+% Doing GC for references on the print stack
+%  - using a flag to avoid copying non-live cells
+% Optimizing GC for closures 
+% - use a copied using rgc flag for closures to avoid multiple traversals
+One of the reasosn why reachability based GC is faster than liveness based GC is due to the fact that 
+in case we try to copy a cons cell which is already copied we do not need to traverse it any further while
+ doing RGC. The same is not true in the case of LGC, as the set of liveness paths may not be similar in 
+both the traversals. One optimization is to keep track of all the liveness automaton states that a particular 
+cons cell has been visited with and ensure that we do not traverse it with the same state more than once. 
+This optimization requires extra memory for each heap cell and potentially it can hold all the automata states. 
+Therefore it is not a feasible solution. In case of lazy languages since we use reachability to copy closures, 
+some cons cells might be copied using reachability. We can avoid repeated traversals of these cons cells without 
+tracking the states. This only requires the knowledge of whether the cons cell was copied using reachability. 
+Hence we use a flag to remember if the heap cell was copied using reachability at any time, if it is set we do 
+not traverse any further.
+
 \subsection{References}

### 2015-03-18 paper/paper.tex
committing

--- 2015-03-15 paper/paper.tex
@@
 $\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
-result.  So,  for example given  a demand $\sigma$ and  the expression
-$\pi: (\RETURN\ x)$, the liveness at $\pi$ is exactly $x.\sigma$.  The
-classical  analogy of  this  is  in {\em  strong  liveness}, where  an
-assignment node $n: x:=y+z$ causes $y$  and $z$ to be live on entry to
-$n$ if (and only if) $x$ is  live at exit of $n$---the liveness of $x$
-at exit from $n$ becomes the demand on $y+z$.
+result.   The  demand on  an  expression  generates  liveness for  its
+constituent  variables.  So  for  example, given  a (non-null)  demand
+$\sigma =  \lbrace \epsilon, \acar  \rbrace$ and the  expression $\pi:
+(\CDR\  x)$, the liveness  at $\pi$  is $\lbrace  x.\epsilon, x.\acdr,
+x.\acdr\acar \rbrace$.  This indicates  that to explore $\sigma$ parts
+of the  result of $(\CDR\  x)$, one has  to first dereference  $x$ and
+\CDR\  field of  the cons  cell referred  by $x$.  Note  that liveness
+refers to  variables and  fields, and not  to $\CONS$ cells  (i.e.\ to
+edges  in  the memory  graph,  not  to  locations themselves).   Hence
+liveness of $\{x.\epsilon,x.\acar\}$ means that future computation may
+refer to  the value  $\ell$ of  variable $x$, and  also to  the $\CAR$
+field of location $\ell$.
+
+%% The classical  analogy of this  is in
+%% {\em strong liveness},  which can be expressed in  our terminology as:
+%% $y$ and $z$  have a liveness of $\lbrace\epsilon\rbrace$  at the entry
+%% to $n: x:=y+z$,  c$y$ and $z$ if  (and only if) $x$ has  a liveness of
+%% $\lbrace\epsilon\rbrace$ at exit of  $n$---the liveness of $x$ at exit
+%% from $n$ becomes the demand on $y+z$.
 
@@
 
-Note that liveness refers to  variables and fields, and not to $\CONS$
-cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
-themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
-future computation may refer to  the value $\ell$ of variable $x$, and
-also to the $\CAR$ field of  location $\ell$.  In the absence of other
-pointers to heap location $\ell$, we are certain that the $\CDR$ field
-of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
-arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
-collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
-would  be  garbage (again  provided  there  are  no other  aliases  to
-$\ell$).
+%% Note that liveness refers to  variables and fields, and not to $\CONS$
+%% cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
+%% themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
+%% future computation may refer to  the value $\ell$ of variable $x$, and
+%% also to the $\CAR$ field of  location $\ell$.  In the absence of other
+%% pointers to heap location $\ell$, we are certain that the $\CDR$ field
+%% of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
+%% arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
+%% collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
+%% would  be  garbage (again  provided  there  are  no other  aliases  to
+%% $\ell$).
 
@@
 \subsection{Liveness Analysis for lazy languages}
-
-To describe  the issues  in formulating liveness  of a  lazy language,
-consider the function definition shown:
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\pf~\pa)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
-          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
-          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
-          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
-	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
-	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
-          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
-	\end{uprogram}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\medskip
 
@@
 
-To see  our way of  handling the second  issue, we trace  the liveness
-analysis to  an extent. Assume  that the calls  to the $\pf$  place an
-overall demand  of $\sigma$ on the  body of \pf.  on  the now describe
-the process  of finding out the  liveness of each  variable defined in
-the function  $\pf$ assuming  a symbolic demand  $\sigma$ on  it.  For
-obvious reasons, this is also the  demand on the return values $u$ and
-$y$, and thus  the liveness at the program  points $\pi_7$ and $\pi_8$
-are $\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The liveness
-at  $\pi_7$ and  $\pi_8$ are  propagated backwards  as  in traditional
-liveness analysis.
+To see  our way of  handling the second  issue, consider the liveness
+analysis of the following function:
+
+\renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} (\DEFINE\ (\pf~\pa)
+	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
+	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
+          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
+          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
+	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
+          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
+	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
+	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
+          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
+	\end{uprogram}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\medskip
+
+Assume that the calls to the $\pf$ place an overall demand of $\sigma$
+on the body  of \pf.  For obvious reasons, this is  also the demand on
+the return  values $u$ and $y$,  and thus the liveness  at the program
+points  $\pi_7$  and  $\pi_8$  are $\lbrace  \pu.\sigma  \rbrace$  and
+$\{\py.\sigma\}$.  The liveness at  $\pi_7$ and $\pi_8$ are propagated
+backwards as in traditional liveness analysis.
 
@@
 scope of  \pu. However, for simplicity  of exposition, we  do not kill
-the variable in our  formulation.  This does not introduce imprecision
-since  \pu\  is yet  to  come into  existence at  $\pi_6$ and  its
-liveness will not be consulted by the garbage collector at this point.
+  livenesses  of  variables  in  our formulation.   This  does  not
+introduce any imprecision since \pu\  is yet to come into existence at
+$\pi_6$  and  its  liveness  will  not be  consulted  by  the  garbage
+collector at this point.
 %==============================================================
@@
 %======================================================================================
+
 In an  eager  language, \pw\  would  not have  been live  after
@@
   evaluation  of  $(\CDR~\pw)$; instead  a  closure  is created.   The
-  evaluation actually takes place while evaluating $(\RETURN~u)$. Thus
+  evaluation actually takes place while evaluating $(\RETURN~\pu)$. Thus
   it is necessary to record the liveness of $\pw$ at the program point
-  $\pi_8$.
+  $\pi_8$, indeed at every program point  from $\pi_8$ up to the
+  beginning of the program.
 %% The second  difference is that  our \LET s  can be recursive  as in
@@
 %%   to be formulated as a fix-point computation.
-  To address this issue, we  regard the body of the function as a
+  We thus  regard the body of the function as a
   tree and consider paths from the root of this tree to program points
   which     trigger     a     evaluation---\SIF\    conditions     and
-  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
+  \RETURN\  expressions. We call  such paths  {\em e-paths}.   For the
   program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
   \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
-  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
-  analysis consists of two maps:  The map $\cal{M}$ maps a return path
-  to a liveness  environment that is common to  all the program points
-  in the path. And a second  map $\cal{P}$ maps a program point to the
-  set of  return paths on  which the point  lies.  If a  program point
+  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+
+The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
+path  to a  liveness environment  that is  common to  all  the program
+points in the path. And a second map $\cal{P}$ maps a program point to
+the set of  return paths on which the point lies.   If a program point
 happens  to  fall  on more  than  one  such  path, then  the  liveness
@@
 liveness  environments  of  the  individual  paths.   In  the  example
-  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
-  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
-  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.
-
-  In  the  sequel we  shall  describe  how  to obtain  $\cal{M}$.  The
-  formulation  of $\cal{P}$  is easy  and we  do not  describe  it any
-  further.
-  We observe that the ${ref}$ rules neither use ${\cal M}$ directly nor 
-  create new labels. We only need to consider the rules for {\LET}, {\SIF}, and
-  {\RETURN}.
+program, ${\cal M}(EP_1)=  \ldots$, ${\cal M}(EP_2)=\ldots$ and ${\cal
+  M}(EP_3)= \ldots$.   Since ${\cal P}(pi_4)$ is  $\lbrace EP_1, EP_2,
+EP_3 \rbrace$,  the liveness  of $pi_4$ is  \ldots.  In the  sequel we
+shall describe how to  obtain $\cal{M}$.  The formulation of $\cal{P}$
+is obvious  and we do  not describe it.   We observe that  the ${ref}$
+rules neither use  ${\cal M}$ directly nor create  new labels. We only
+need to consider the rules for {\LET}, {\SIF}, and {\RETURN}.
 

### 2015-03-19 paper/paper.tex
sigplan style
fkdk

@@
     #1 \medskip \hrule \medskip}}}
+\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
+    #1 \medskip \hrule \medskip}}}
+
 \newcommand{\scmin} {\mbox{\sf\em in}}
@@
 \subsection{Liveness Analysis for lazy languages}
+\comment{analysis for liveness has two components - transfering
+  demands on expressions to liveness of results
+and propagating liveness. first discuss demand --> liveness}
+
+%% First recall the classical formulation  of liveness (as sets of simple
+%% variables)        on        three-address       instructions,        $
+%% \mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
+%% \mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
+%%   strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
+%% $\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
+%% \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
+
+%%Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
+%%analogous.
+% has three analogous aspects.
+% \begin{itemize}
+% \item
+\alan{Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
+returns the  incremental liveness {\em generated} by  $s$.  Because we
+generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
+parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
+determine which access paths of its free variables are made live.  The
+cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
+% The cases for
+% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
+% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
+% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
+A  demand of  $\sigma$  on  (\CAR~$x$) is  transformed  to the  demand
+$\acar\sigma$  on $x$.   In  addition, \CAR\  always dereferences  its
+argument  (even if  its result  is  never used).   This generates  the
+liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$ (note  $\sigma$  may be
+$\{\}$).   In  the opposite  sense,  the  demand  of $\acar\sigma$  on
+(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
+that \CONS\ does not,  by itself, dereference its arguments.  Thirdly,
+for the case  of a user-function call, a  third parameter $\Lfonly$ to
+$\mathit{ref}$  expresses how  the demand  $\sigma$ on  the  result is
+transformed  into demands  on its  parameters.  Constants  generate no
+liveness.}
 
@@
 \begin{enumerate}
-\item Unlike eager languages, a variable may be  bound to an
-  expression that is not fully evaluated. i.e. a closure.
+\item Unlike eager languages, a variable may be bound to an expression
+  that is not fully evaluated. i.e. a closure.
 \item The point where an expression gets evaluated cannot be statically
@@
         \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
-        \emptyset  & \mbox{otherwise}
+        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                  \end{array} \\
@@
 The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-path  to a  liveness environment  that is  common to  all  the program
-points in the path. And a second map $\cal{P}$ maps a program point to
-the set of  return paths on which the point lies.   If a program point
-happens  to  fall  on more  than  one  such  path, then  the  liveness
-environment of  the program  point is the  variable-wise union  of the
-liveness  environments  of  the  individual  paths.   In  the  example
-program, ${\cal M}(EP_1)=  \ldots$, ${\cal M}(EP_2)=\ldots$ and ${\cal
-  M}(EP_3)= \ldots$.   Since ${\cal P}(pi_4)$ is  $\lbrace EP_1, EP_2,
-EP_3 \rbrace$,  the liveness  of $pi_4$ is  \ldots.  In the  sequel we
-shall describe how to  obtain $\cal{M}$.  The formulation of $\cal{P}$
-is obvious  and we do  not describe it.   We observe that  the ${ref}$
-rules neither use  ${\cal M}$ directly nor create  new labels. We only
-need to consider the rules for {\LET}, {\SIF}, and {\RETURN}.
-
-  A {\RETURN~\px}\  statement always forces an evaluation and hence a new 
-  label {\em l} needs to be created and added to $\cal{M}$. A new liveness
-  environment is associated with {\em l}. The demand $\sigma$ that is placed 
-  on this expression is now transferred to the variable \px. {\em l} will now
-  contain an entry for \px~ with liveness $\sigma$.
-
-  An \SIF~ statement will modify  $\cal{M}$ by doing a union of 
-  the maps from the {\em then} and the {\em else} part. Note that each part can contain 
-  more than one label. Since the variable corresponding to the condition may not
-  be part of either the {\em then} or the {\em else} part, a new label {\em l} is created 
-  and added to $\cal{M}$. The only liveness associated with {\em l} is the liveness 
-  of the condition variable.
-
-  The processing rule for {\LET} is the most complex among all the rules. A {\LET} does
-  not add any new label but might modify the liveness values of some variables or might add 
-  liveness values for the let bound variable. Since there might be more than one {\RETURN} 
-  statement following a {\LET}, there might be multiple {\em E-paths} reaching the {\LET}.
-  The demand on the let variable might be different along different {\em E-paths}. Therefore,
-  for each label {\em l} in $\cal{M}$, the liveness of the let-expression is calculated using 
-  the liveness environment $\Lv$ corresponding to {\em l}. Since the {\LET} is lazy, 
-  calculating its livness might require a fixed-point processing.
+path to a common liveness environment, and a second map $\cal{P}$ maps
+a program  point to the set of  return paths on which  the point lies.
+If a  program point happens to fall  on more than one  such path, then
+the  liveness environment of  the program  point is  the variable-wise
+union of  the liveness environments  of the individual paths.   In the
+example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
+${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
+EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In the sequel
+we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
+$\cal{P}$ is obvious  and we do not describe it.   We observe that the
+${ref}$  rules  neither  use   ${\cal  M}$  directly  nor  create  new
+labels. We  only need  to consider the  rules for {\LET},  {\SIF}, and
+{\RETURN}.
+
+  A  {\RETURN~\px}\ with a  demand $\sigma$  forces an  evaluation and
+  hence a  new ${\cal M}$ needs to  be created with {\em  l} mapped to
+  $\lbrace \px.\sigma \rbrace$.  An \SIF~ statement is an union of the
+  ${\cal M}$-maps  from the {\em then}  and the {\em  else} part. Note
+  that each part  can contain more than one  label. Since the variable
+  corresponding to  the condition may not  be part of  either the {\em
+    then} or the  {\em else} part, a new label {\em  l} is created and
+  added to $\cal{M}$. The only liveness associated with {\em l} is the
+  liveness of the condition variable. Also notice that the
+  consequences of laziness---the entire statement including the
+  condition is not evaluated if the demand on it is
+  $\emptyset$. This results in the empty map denoted by
+  $\emptyset_{\cal{M}}$.
+
+  The processing  {\LET} rule has an elegant formulation. Recollect
+  that our language permits lazy \LET s whose definitions can be
+  recursive   Thus, in the expression
+  $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in $s$.  We model a let statement as a recursive function which captures an
+  arbitrary  number
+  of unrollings. We pass to this function the demand arising out of
+  the liveness of $x$  at the
+  beginning of $e$, and let the $the  
+  that is the  most complex among  all the
+  rules. A  {\LET} does  not add  any new label  but might  modify the
+  liveness values of  some variables or might add  liveness values for
+  the let bound variable. Since there might be more than one {\RETURN}
+  statement following a {\LET},  there might be multiple {\em E-paths}
+  reaching  the {\LET}.   The  demand  on the  let  variable might  be
+  different along  different {\em E-paths}. Therefore,  for each label
+  {\em  l}  in  $\cal{M}$,  the  liveness  of  the  let-expression  is
+  calculated  using the  liveness environment  $\Lv$  corresponding to
+  {\em l}.  Since the  {\LET} is lazy,  calculating its  livness might
+  require a fixed-point processing.
 
@@
 
-First recall the classical formulation  of liveness (as sets of simple
-variables)        on        three-address       instructions,        $
-\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
-\mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
-  strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
-$\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
-\in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
-
-Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
-analogous.
-% has three analogous aspects.
-% \begin{itemize}
-% \item
-Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
-returns the  incremental liveness {\em generated} by  $s$.  Because we
-generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
-parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
-determine which access paths of its free variables are made live.  The
-cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
-% The cases for
-% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
-% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
-% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
-A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
-argument (even if its result is never used).
-This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
-(note $\sigma$ may be $\{\}$).
-In the opposite  sense, the  demand  of $\acar\sigma$  on
-(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-that \CONS\ does not, by itself, dereference its arguments.
-Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
-$\mathit{ref}$
-expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
-Constants generate no liveness.
+%% First recall the classical formulation  of liveness (as sets of simple
+%% variables)        on        three-address       instructions,        $
+%% \mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
+%% \mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
+%%   strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
+%% $\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
+%% \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
+
+%% Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
+%% analogous.
+%% % has three analogous aspects.
+%% % \begin{itemize}
+%% % \item
+%% Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
+%% returns the  incremental liveness {\em generated} by  $s$.  Because we
+%% generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
+%% parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
+%% determine which access paths of its free variables are made live.  The
+%% cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
+%% % The cases for
+%% % $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
+%% % is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
+%% % The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
+%% A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
+%% $\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
+%% argument (even if its result is never used).
+%% This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
+%% (note $\sigma$ may be $\{\}$).
+%% In the opposite  sense, the  demand  of $\acar\sigma$  on
+%% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
+%% that \CONS\ does not, by itself, dereference its arguments.
+%% Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
+%% $\mathit{ref}$
+%% expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
+%% Constants generate no liveness.
 

### 2015-03-20 paper/paper.tex
still on liveness

@@
-\documentclass{sig-alternate}
+\documentclass[10pt]{sigplanconf}
 \usepackage{etex}
@@
     #1 \medskip \hrule \medskip}}}
+\newcommand{\blankout}[1]{}
 
@@
 % the body of your article BEFORE References or any Appendices.
-
+\cmt{{
 \numberofauthors{3} %  in this sample file, there are a *total*
@@
 }
+}}
+\authorinfo{Double Blind Review}{}{}
 \maketitle
+
 \begin{abstract}
@@
 
+\pagebreak
+
 \section{Liveness}\label{sec:liveness}
@@
 \acdr\acdr\acar\}$, then future computations  can only refer up to the
-second and third  members of $x$. In this  notation, a scalar variable
-being live or dead are represented as $\{\epsilon\}$ and $\{\}$ respectively.
-
-A {\em liveness environment} is a mapping from variables to subsets of
-$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
-writing  $\{x.\acar\acdr,   x.\acdr,  y.\epsilon\}$  instead   of  $[x
+second  and  third  members  of  $x$. In  this  notation,  $x  \mapsto
+\{\epsilon\}$ represents access using  $x$ itself and $x \mapsto \{\}$
+indicates that $x$ is dead.  A {\em liveness environment} is a mapping
+from  variables  to   subsets  of  $\{\acar,\acdr\}^\ast$,  but  often
+expressed as  a set, for example by  writing $\{x.\acar\acdr, x.\acdr,
+y.\epsilon\}$               instead               of               $[x
   \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
-  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  each
-program  point $\pi$:  the liveness  at $\pi:e$  is the  liveness just
+  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  
+program  points $\pi$.  The liveness  at $\pi:e$  is the  liveness just
 before executing $e$.
-%%
-%% Due to the inherent uncertainty in
-%% determining the  point of evaluation  in lazy languages,  we associate
-%% liveness environments with certain kinds  of paths in the program. All
-%% programs points lying on the path share the same liveness environment.
-%% We explain this point later.
-%% 
-A  notion  related  to  liveness  is  {\em  demand}.   The  demand  on
-expression  $e$   is  again  an   access  path---that  subset   of  of
+
+A  notion related  to  liveness is  {\em  demand}.  The  demand on  an
+expression  $e$  is again  a  set  of  access paths---that  subset  of
 $\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
-result.   The  demand on  an  expression  generates  liveness for  its
-constituent  variables.  So  for  example, given  a (non-null)  demand
-$\sigma =  \lbrace \epsilon, \acar  \rbrace$ and the  expression $\pi:
-(\CDR\  x)$, the liveness  at $\pi$  is $\lbrace  x.\epsilon, x.\acdr,
-x.\acdr\acar \rbrace$.  This indicates  that to explore $\sigma$ parts
-of the  result of $(\CDR\  x)$, one has  to first dereference  $x$ and
-\CDR\  field of  the cons  cell referred  by $x$.  Note  that liveness
-refers to  variables and  fields, and not  to $\CONS$ cells  (i.e.\ to
-edges  in  the memory  graph,  not  to  locations themselves).   Hence
-liveness of $\{x.\epsilon,x.\acar\}$ means that future computation may
-refer to  the value  $\ell$ of  variable $x$, and  also to  the $\CAR$
-field of location $\ell$.
-
-%% The classical  analogy of this  is in
-%% {\em strong liveness},  which can be expressed in  our terminology as:
-%% $y$ and $z$  have a liveness of $\lbrace\epsilon\rbrace$  at the entry
-%% to $n: x:=y+z$,  c$y$ and $z$ if  (and only if) $x$ has  a liveness of
-%% $\lbrace\epsilon\rbrace$ at exit of  $n$---the liveness of $x$ at exit
-%% from $n$ becomes the demand on $y+z$.
-
-%% Note  that, for  an  operation  like division  which  may raise  an
-%% exception,  the  assignment $n:  x:=y/z$  makes  $y$  and $z$  live
-%% regardless of the liveness of $x$.
+result.   To see  the need  for demands,  consider the  let expression
+$\pi:(\LET~x\leftarrow (\CDR~y)~\IN~ \pi':\RETURN~x)$. Assume that the
+context of  this expression produces the liveness  $[x \mapsto \lbrace
+  \epsilon, \acar  \rbrace]$ at $\pi'$.   Due to the  \LET\ definition
+which  binds $x$  to $(\CDR~y)$,  the liveness  of $x$  at  $\pi'$ now
+becomes  the  demand on  $(\CDR~y)$.   This,  in  turn, generates  the
+liveness $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar \rbrace$ at $\pi$
+(the set of $y$-rooted accesses required to explore $\lbrace \epsilon,
+\acar  \rbrace$  paths  of  the  result  of  $(\CDR~y)$)\footnote{The
+  classical analogue  is strong liveness  analysis, where $y$  and $z$
+  have  a  liveness  of  $\lbrace\epsilon\rbrace$  at  the  entry  $n:
+  x:=y+z$,    if   and    only   if    $x$   has    a    liveness   of
+  $\lbrace\epsilon\rbrace$ at  exit of  $n$.  In our  terminology, the
+  liveness of $x$  at exit from $n$ becomes the  demand on $y+z$. This
+  generates the  liveness $\lbrace y.\epsilon,  z.\epsilon \rbrace$ at
+  the entry of $n$.}.
+%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
+%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
+%% $y$.
+Note that  liveness refers to  access using variables and  fields, and
+not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
+locations themselves).
 
@@
 $\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
-Finally,  we use $\Lfonly$  to range  over {\em  demand transformers}.
-These transform demands on a function call into demands for its formal
-parameters:    given    a    user    function    $f$,    defined    by
-$(\DEFINE\  (f\ x_1\  \ldots\ x_n)\  \  e_f)$ and  called with  demand
-$\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
-
-%% Note that liveness refers to  variables and fields, and not to $\CONS$
-%% cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
-%% themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
-%% future computation may refer to  the value $\ell$ of variable $x$, and
-%% also to the $\CAR$ field of  location $\ell$.  In the absence of other
-%% pointers to heap location $\ell$, we are certain that the $\CDR$ field
-%% of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
-%% arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
-%% collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
-%% would  be  garbage (again  provided  there  are  no other  aliases  to
-%% $\ell$).
-
-\bigskip
-\hrule 
-\bigskip
+%% Finally,  we use $\Lfonly$  to range  over {\em  demand transformers}.
+%% These transform demands on a function call into demands for its formal
+%% parameters:    given    a    user    function    $f$,    defined    by
+%% $(\DEFINE\  (f\ x_1\  \ldots\ x_n)\  \  e_f)$ and  called with  demand
+%% $\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
 
 \subsection{Liveness Analysis for lazy languages}
-\comment{analysis for liveness has two components - transfering
-  demands on expressions to liveness of results
-and propagating liveness. first discuss demand --> liveness}
-
-%% First recall the classical formulation  of liveness (as sets of simple
-%% variables)        on        three-address       instructions,        $
-%% \mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
-%% \mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
-%%   strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
-%% $\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
-%% \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
-
-%%Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
-%%analogous.
-% has three analogous aspects.
-% \begin{itemize}
-% \item
-\alan{Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
-returns the  incremental liveness {\em generated} by  $s$.  Because we
-generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
-parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
-determine which access paths of its free variables are made live.  The
-cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
-% The cases for
-% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
-% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
-% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
-A  demand of  $\sigma$  on  (\CAR~$x$) is  transformed  to the  demand
-$\acar\sigma$  on $x$.   In  addition, \CAR\  always dereferences  its
-argument  (even if  its result  is  never used).   This generates  the
-liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$ (note  $\sigma$  may be
-$\{\}$).   In  the opposite  sense,  the  demand  of $\acar\sigma$  on
-(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-that \CONS\ does not,  by itself, dereference its arguments.  Thirdly,
-for the case  of a user-function call, a  third parameter $\Lfonly$ to
-$\mathit{ref}$  expresses how  the demand  $\sigma$ on  the  result is
-transformed  into demands  on its  parameters.  Constants  generate no
-liveness.}
-
-Liveness analysis for lazy languages differ from eager languages in
-two respects:
-\begin{enumerate}
-\item Unlike eager languages, a variable may be bound to an expression
-  that is not fully evaluated. i.e. a closure.
-\item The point where an expression gets evaluated cannot be statically
-  determined.
-\end{enumerate}
-
-To  deal  with the  first  issue, we  apply  the  results of  liveness
-analysis only  to fully evaluated  data.  If the marking  phase during
-garbage collection leads to a closure, and the closure is not dead, we
-mark the entire reachable memory  under the closure for collection.  A
-better scheme  would be  to garbage collect  the closure based  on the
-liveness  at the point  of its  creation.  This  is sound  because our
-language is statically scoped.  However, this is work in progress, and
-the method described in this paper switches over to reachability while
-collecting closures.
-
-%% However, we GC,  if we find a variable can either  point to a fully
-%% evaluated  data value  or  it could  point  to a  closure which  on
-%% evaluation  generates  the   required  data  value.   For  example,
-%% consider the statement {\tt  let \py~ $\leftarrow $ (\CDR~\px)}. In
-%% case of  an eager  language the variable  \py~ points to  the \CDR~
-%% part of  \px~ which  is guaranteed  to be a  cons cell.   Hence any
-%% demand on \py~ can be  immediately transferred to the \CDR~ part of
-%% \px.  In case  of a lazy language, \px~ might  not be evaluated and
-%% hence  pointing to  a closure.  In our  example, it  points  to the
-%% closure corresponding to (\CONS~1~\px).  It does not make any sense
-%% to transfer the demand onto the \CDR~ part of \px~ as it still does
-%% not exist.  , 2) it is  not possible to statically determine when a
-%% closure gets evaluated.  Therefore during  GC we treat a closure as
-%% a demand transformer and tranform the demand on to the arguments of
-%% the  closure   pointed.   This  is   similar  to  the   process  of
-%% transferring the demand from the  return value of a function to the
-%% arguments of the function call in an eager language.
-
-%% In our example, during GC if  we find that \px~ points to a closure
-%% instead  of a  cons cell  we find  out the  demand/liveness  on the
-%% arguments  of the  closure, namely  \py.  Statically  it  cannot be
-%% determined  what  values  a  closure  points to  or  when  it  gets
-%% evaluated, hence  our method  approximates the liveness  of closure
-%% arguments  at  runtime  by  tagging  extra  information  with  each
-%% closure.  Since  the arguments of the closures  could themselves be
-%% closures in turn, we recursively  apply the same procedure until we
-%% hit  a data  value.  The only  case  where we  do  not follow  this
-%% procedure is if the demand on a particular closure is null. In such
-%% a case we safely collect the entire closure as garbage.
-
-To see  our way of  handling the second  issue, consider the liveness
-analysis of the following function:
+
+Figure~\ref{fig:live-judge}  describes  our  analysis.   Firstly,  the
+function $\mathit{ref}$,  when given an  application $s$ and  a demand
+$\sigma$,  returns the liveness  generated for  the free  variables of
+$s$.   In  each  case,  the  null demand  ($\emptyset$)  generates  no
+liveness.   Notice  that this  is  different  from  liveness of  eager
+languages, where, for instance,  (\CDR~$x$) would be evaluated even if
+the   demand   on   it   is  $\emptyset$   generating   the   liveness
+$\{x.\epsilon\}$.  The  cases (\PRIM~x~y)  and (\NULLQ~x) are  easy. A
+non-null  demand of  $\sigma$  on (\CDR~$x$),  is  transformed to  the
+liveness  $x.\epsilon,  x.\acdr\sigma$.   In  an opposite  sense,  the
+demand  of  $\acar\sigma$ on  (\CONS~$x$~$y$)  is  transformed to  the
+demand  $\sigma$  on $x$.   Note  that  \CONS\  does not,  by  itself,
+dereference its arguments. Constants generate no liveness.
+
+For  the  case  of a  user-function  call  we  use a  third  parameter
+$\Lfonly$  that  represents the  summaries  of  all  functions in  the
+program. $\Lfonly$ expresses how the  demand $\sigma$ on the result of
+a function call is transformed  into liveness on its parameters at the
+beginning of the call.  $\Lfonly$  is determined by the judgement form
+$\mathit{Prog}    \len   \Lfonly$    using   inference    rule   ({\sc
+  live-define}). This states the  fixpoint property to be satisfied by
+$\Lfonly$, namely, the demand transformation assumed for each function
+in  the  program should  be  the  same  as the  demand  transformation
+calculated from their bodies. The algorithmic formulation of this rule
+through  a  grammar will  be  described  in [???].   This yields  the  least
+solution  satisfying  the  fixpoint  property  and  thus  ensures  the
+greatest amount of garbage collected.
+
+%%   Analogously   to  classical  liveness  being
+%% computed  as  a  solution  of  dataflow  equations,  we  require,  via
+%% inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
+%% fixed-point property that:  when we assume $\Lfonly$ to  be the family
+%% of demand transformers for the program then the calculated liveness of
+%% each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
+%% assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
+%% $\Lfonly$;  all are  safe (see  Section~\ref{sec:correctness})  but we
+%% prefer  the  least  one  as  giving  the  least  liveness  subject  to
+%% safety---and hence greatest amount of garbage collected.
+
+We now describe the  propagation of liveness through expressions. This
+brings  us to  the  second way  in  which liveness  analysis for  lazy
+languages  differ from  eager languages---the point where  an
+expression gets evaluated cannot be statically determined.  To see how
+this  is handled,  consider  the liveness  analysis  of the  following
+function:
+%%%%%%%%%%%%%%%%%%%%%%
+
+% Push this to a later point
+
+%% Unlike eager languages, a variable may be bound to an expression
+%%   that is not fully evaluated. i.e. a closure.
+
+%% To  deal  with the  first  issue, we  apply  the  results of  liveness
+%% analysis only  to fully evaluated  data.  If the marking  phase during
+%% garbage collection leads to a closure, and the closure is not dead, we
+%% mark the entire reachable memory  under the closure for collection.  A
+%% better scheme  would be  to garbage collect  the closure based  on the
+%% liveness  at the point  of its  creation.  This  is sound  because our
+%% language is statically scoped.  However, this is work in progress, and
+%% the method described in this paper switches over to reachability while
+%% collecting closures.
+
+%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\pf~\pa)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
-          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
+	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
+	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
+	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
+          \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
           $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
+	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
           $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
-	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
-	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
-          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
+	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
+	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
+          \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
 	\end{uprogram}}
@@
 scope of \pu.   However, for simplicity of exposition,  we do not kill
-  livenesses  of  variables  in  our formulation.   This  does  not
-introduce any imprecision since \pu\  is yet to come into existence at
-$\pi_6$  and  its  liveness  will  not be  consulted  by  the  garbage
-collector at this point.
+livenesses of  variables in our formulation.  This  does not introduce
+any imprecision  since \pu\ is yet  to come into  existence at $\pi_6$
+and its  liveness will  not be consulted  by the garbage  collector at
+this point.
 %==============================================================
@@
 
-In an  eager  language, \pw\  would  not have  been live  after
- $\pi_7$, the point of its last use. In a lazy language, however, the
-  definition  $\pu  \leftarrow  (\CDR~\pw)$,  does not  result  in  an
-  evaluation  of  $(\CDR~\pw)$; instead  a  closure  is created.   The
-  evaluation actually takes place while evaluating $(\RETURN~\pu)$. Thus
-  it is necessary to record the liveness of $\pw$ at the program point
-  $\pi_8$, indeed at every program point  from $\pi_8$ up to the
-  beginning of the program.
+In an eager language, \pw\ would not have been live after $\pi_7$, the
+point of  its last  use. In a  lazy language, however,  the definition
+$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
+$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
+takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
+record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
+every program point from $\pi_8$ up to the beginning of the program.
 %% The second  difference is that  our \LET s  can be recursive  as in
@@
 %%   to be formulated as a fix-point computation.
-  We thus  regard the body of the function as a
-  tree and consider paths from the root of this tree to program points
-  which     trigger     a     evaluation---\SIF\    conditions     and
-  \RETURN\  expressions. We call  such paths  {\em e-paths}.   For the
-  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
-  \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
-  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+  We thus regard the body of the function as a tree and consider paths
+  from  the  root of  this  tree to  program  points  which trigger  a
+  evaluation---\SIF\  conditions and  \RETURN\  expressions.  We  call
+  such paths  {\em e-paths}.  For  the program shown, there  are three
+  such  paths $EP_1  = \pi_1,  \pi_2,  \pi_3, \pi_4,  \pi_5$, $EP_2  =
+  \pi_1, \pi_2, \pi_3, \pi_4, \pi_6, \pi_7, \pi_8$, and $EP_3 = \pi_1,
+  \pi_2, \pi_3, \pi_4, \pi_9$.
 
@@
   that  our language  permits lazy  \LET  s whose  definitions can  be
-  recursive   Thus, in the expression
-  $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in $s$.  We model a let statement as a recursive function which captures an
-  arbitrary  number
-  of unrollings. We pass to this function the demand arising out of
-  the liveness of $x$  at the
-  beginning of $e$, and let the $the  
-  that is the  most complex among  all the
-  rules. A  {\LET} does  not add  any new label  but might  modify the
-  liveness values of  some variables or might add  liveness values for
-  the let bound variable. Since there might be more than one {\RETURN}
-  statement following a {\LET},  there might be multiple {\em E-paths}
-  reaching  the {\LET}.   The  demand  on the  let  variable might  be
-  different along  different {\em E-paths}. Therefore,  for each label
-  {\em  l}  in  $\cal{M}$,  the  liveness  of  the  let-expression  is
-  calculated  using the  liveness environment  $\Lv$  corresponding to
-  {\em l}.  Since the  {\LET} is lazy,  calculating its  livness might
-  require a fixed-point processing.
-
-\bigskip
-\hrule 
-\bigskip
+  recursive Thus,  in the expression  $\LET~x \leftarrow~s~\IN~e$, $x$
+  can occur  in $s$.   We model the  definition $x \leftarrow~s$  as a
+  possibly recursive  function which  captures an arbitrary  number of
+  unrollings of the  definition.  We pass to this  function the demand
+  arising  out of the  liveness of  $x$ along  all paths  reaching the
+  beginning of $e$, and let  the \Lfonly abstraction derive the demand
+  on the  variables of $s$.
+
+%% that is the  most complex among  all the
+%%   rules.  A  {\LET} does not  add any new  label but might  modify the
+%%   liveness values of  some variables or might add  liveness values for
+%%   the  let  bound  variable.   Since  there might  be  more  than  one
+%%   {\RETURN} statement following a {\LET}, there might be multiple {\em
+%%     E-paths}  reaching the  {\LET}.  The  demand on  the  let variable
+%%   might be different along different {\em E-paths}.
 
@@
 
-The  function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
+\alan{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
 expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
@@
 $\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
-$\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-% \item
+$\mathit{ref}(s,\Lv(z),\Lfonly)$.}
+
 %%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-Finally, the  judgement form $\mathit{Prog}  \len \Lfonly$ is  used to
-determine  $\Lfonly$\@.   Analogously   to  classical  liveness  being
-computed  as  a  solution  of  dataflow  equations,  we  require,  via
-inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
-fixed-point property that:  when we assume $\Lfonly$ to  be the family
-of demand transformers for the program then the calculated liveness of
-each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
-assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
-$\Lfonly$;  all are  safe (see  Section~\ref{sec:correctness})  but we
-prefer  the  least  one  as  giving  the  least  liveness  subject  to
-safety---and hence greatest amount of garbage collected.
+
 % \end{itemize}
@@
 
-Section~\ref{sec:computing} shows how the demand transformers $\Lfonly$ for a program
-(representing a fully context-sensitive analysis)
-can be safely approximated, for each function, by a {\em procedure summary}
-(unifying the contexts in the style of 0-CFA)\@.
-The summary consists of a pair of a single demand and, for this demand,
-the corresponding tuple of demands the function makes on its
+Section~\ref{sec:computing}   shows   how   the  demand   transformers
+$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
+analysis) can  be safely  approximated, for each  function, by  a {\em
+  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
+The  summary consists  of a  pair  of a  single demand  and, for  this
+demand, the corresponding  tuple of demands the function  makes on its
 arguments.
 
-%\comment{
-
-%}
-
 % \item
+\pagebreak
 
@@
 % GC assumptions
-By the time the program starts executing, the liveness automata are created and stored in a file. 
-This is re-used every time the program executes. 
+By the  time the program  starts executing, the liveness  automata are
+created and stored in a file.   This is re-used every time the program
+executes.
 % GC points - mention reduction of gc points due to e-paths
-An important difference in GC of lazy language compared to an eager language is the point of 
-garbage collection. In case of eager languages only \CONS~ points and function calls are 
-the GC points. In a lazy language every \LET~ statement is a potential GC point. Hence we need to store 
-liveness automata corresponding to every \LET~ statement in the program. This leads to an explosion 
-in the number of automata. From our observation that along an {\em E-path} all program points have 
-the same liveness, we can reduce the number of automata state required by sharing the same automata 
-for all program points along an {\em E-path}. In the example program, we only need to create automata 
-for program points \ldots, since they are the only distinct {\em E-path}s in the program. 
+An important  difference in GC of  lazy language compared  to an eager
+language  is  the  point  of  garbage collection.  In  case  of  eager
+languages only \CONS~ points and  function calls are the GC points. In
+a lazy language  every \LET~ statement is a  potential GC point. Hence
+we  need  to store  liveness  automata  corresponding  to every  \LET~
+statement in the program. This leads  to an explosion in the number of
+automata. From our observation that  along an {\em E-path} all program
+points have  the same liveness, we  can reduce the  number of automata
+state required  by sharing  the same automata  for all  program points
+along an {\em E-path}. In the  example program, we only need to create
+automata for program  points \ldots, since they are  the only distinct
+{\em E-path}s in the program.
 % Abstract description of the GC algorithm
 % Liveness for cons cells & reachability for closures
-During GC, if the heap cell being processed is a cons cell, we consult the liveness automaton 
-corresponding to it and copy only the live paths. In case the heap cell points to a closure, 
-we safely approximate the liveness to be {\{0+1\}\*} and copy everything under the closure.
+During GC, if the heap cell being processed is a cons cell, we consult
+the  liveness automaton  corresponding to  it and  copy only  the live
+paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
+approximate the  liveness to be {\{0+1\}\*} and  copy everything under
+the closure.
 % Handling of mixture of LGC & RGC
-A consequence of copying everything under a closure is the fact that we can end up trying to copy 
-things which are not live and corrupt the heap. This can happen if at some $GC_k$ a cons cell is copied 
-using liveness. At a later stage this cons cell becomes part of a closure and hence during $GC_{k+n}$ the 
-algorithm tries to copy everything under the cons cell thus copying heap locations which might not be valid. 
-To avoid this behavior we either have to choose to do a full liveness based GC for closure or we should 
-ensure that any non-live references are not carried across GC's. We take the latter approach and ensure that 
-non-live references are set to null. After every GC, the live part of the buffer is scanned and any reference 
-which is not live is set to null. The correctness of this follows from the fact that any reference which is 
-decalred not live in a $GC_k$ will not be derefernced during later execution.
+A consequence of  copying everything under a closure  is the fact that
+we can end up trying to copy things which are not live and corrupt the
+heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
+liveness. At  a later stage this  cons cell becomes part  of a closure
+and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
+under the  cons cell  thus copying heap  locations which might  not be
+valid.  To avoid  this behavior we either have to choose  to do a full
+liveness based  GC for closure or  we should ensure  that any non-live
+references are  not carried across  GC's. We take the  latter approach
+and ensure that  non-live references are set to  null. After every GC,
+the live part of the buffer  is scanned and any reference which is not
+live is  set to null.  The correctness of  this follows from  the fact
+that any reference which is decalred  not live in a $GC_k$ will not be
+derefernced during later execution.
 % A reference copied using LGC could become part of a closure in a later GC
@@
 % Difference between eager and lazy GC
-Another difference that is present in GC of lazy lanaguage is the presence of a print stack. In lazy 
-languages the evaluation of closures is driven by the print function. Hence, during a GC there will be 
-heap references that will be present on the stack which also have to copied/collected. One solution is to 
-assume that all references present on the print stack are fully live and copy everything pointed by them 
-akin to reachability. This is safe but copies heap cells which have already been printed. Since the print 
-function is not part of the executing program its liveness automata are not considered. Therefore we 
-simulate the liveness of the print function by using a flag in the heap cell. The flag ensures that only the 
-required parts of a cons cell are copied. The behavior of the GC algorithm is as described below,
+Another  difference that is  present in  GC of  lazy lanaguage  is the
+presence  of  a print  stack.  In  lazy  languages the  evaluation  of
+closures is  driven by  the print function.  Hence, during a  GC there
+will be heap  references that will be present on  the stack which also
+have  to  copied/collected.  One   solution  is  to  assume  that  all
+references  present  on  the  print  stack are  fully  live  and  copy
+everything  pointed by  them akin  to reachability.  This is  safe but
+copies heap  cells which  have already been  printed. Since  the print
+function is  not part of  the executing program its  liveness automata
+are not  considered. Therefore we  simulate the liveness of  the print
+function by using a flag in  the heap cell. The flag ensures that only
+the required parts  of a cons cell are copied. The  behavior of the GC
+algorithm is as described below,
 \begin{verbatim}
@@
 \end{verbatim}
-We provide an intuition about the correctness of this algorithm. Consider the following facts about the print stack 
-during a GC,
+We   provide   an   intuition    about   the   correctness   of   this
+algorithm. Consider the following facts about the print stack during a
+GC,
 \begin{enumerate}
@@
 % That's all folks!
+
 \end{document}

### 2015-03-21 paper/paper.tex

--- 2015-03-18 paper/paper.tex
@@
 \usepackage{tabularx}
-\usepackage{algorithmic}
+\usepackage {algorithmic}
 \usepackage{amssymb}
@@
 \usepackage{bcprules}
+%\usepackage{program}
 \input{fh_defs}
@@
 }
+}}
+\authorinfo{Double Blind Review}{}{}
 \maketitle
@@
 $\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
-result.   The  demand on  an  expression  generates  liveness for  its
-constituent  variables.  So  for  example, given  a (non-null)  demand
-$\sigma =  \lbrace \epsilon, \acar  \rbrace$ and the  expression $\pi:
-(\CDR\  x)$, the liveness  at $\pi$  is $\lbrace  x.\epsilon, x.\acdr,
-x.\acdr\acar \rbrace$.  This indicates  that to explore $\sigma$ parts
-of the  result of $(\CDR\  x)$, one has  to first dereference  $x$ and
-\CDR\  field of  the cons  cell referred  by $x$.  Note  that liveness
-refers to  variables and  fields, and not  to $\CONS$ cells  (i.e.\ to
-edges  in  the memory  graph,  not  to  locations themselves).   Hence
-liveness of $\{x.\epsilon,x.\acar\}$ means that future computation may
-refer to  the value  $\ell$ of  variable $x$, and  also to  the $\CAR$
-field of location $\ell$.
-
-%% The classical  analogy of this  is in
-%% {\em strong liveness},  which can be expressed in  our terminology as:
-%% $y$ and $z$  have a liveness of $\lbrace\epsilon\rbrace$  at the entry
-%% to $n: x:=y+z$,  c$y$ and $z$ if  (and only if) $x$ has  a liveness of
-%% $\lbrace\epsilon\rbrace$ at exit of  $n$---the liveness of $x$ at exit
-%% from $n$ becomes the demand on $y+z$.
+result.  So,  for example given  a demand $\sigma$ and  the expression
+$\pi: (\RETURN\ x)$, the liveness at $\pi$ is exactly $x.\sigma$.  The
+classical  analogy of  this  is  in {\em  strong  liveness}, where  an
+assignment node $n: x:=y+z$ causes $y$  and $z$ to be live on entry to
+$n$ if (and only if) $x$ is  live at exit of $n$---the liveness of $x$
+at exit from $n$ becomes the demand on $y+z$.
 
@@
 
-%% Note that liveness refers to  variables and fields, and not to $\CONS$
-%% cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
-%% themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
-%% future computation may refer to  the value $\ell$ of variable $x$, and
-%% also to the $\CAR$ field of  location $\ell$.  In the absence of other
-%% pointers to heap location $\ell$, we are certain that the $\CDR$ field
-%% of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
-%% arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
-%% collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
-%% would  be  garbage (again  provided  there  are  no other  aliases  to
-%% $\ell$).
+Note that liveness refers to  variables and fields, and not to $\CONS$
+cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
+themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
+future computation may refer to  the value $\ell$ of variable $x$, and
+also to the $\CAR$ field of  location $\ell$.  In the absence of other
+pointers to heap location $\ell$, we are certain that the $\CDR$ field
+of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
+arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
+collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
+would  be  garbage (again  provided  there  are  no other  aliases  to
+$\ell$).
 
@@
 \subsection{Liveness Analysis for lazy languages}
+
+To describe  the issues  in formulating liveness  of a  lazy language,
+consider the function definition shown:
+
+\renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} (\DEFINE\ (\pf~\pa)
+	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
+	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
+          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
+          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
+	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
+          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
+	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
+	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
+          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
+	\end{uprogram}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\medskip
 
@@
 
-To see  our way of  handling the second  issue, consider the liveness
-analysis of the following function:
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} (\DEFINE\ (\pf~\pa)
-	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
-	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
-          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
-          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
-	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
-          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
-	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
-	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
-          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
-	\end{uprogram}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\medskip
-
-Assume that the calls to the $\pf$ place an overall demand of $\sigma$
-on the body  of \pf.  For obvious reasons, this is  also the demand on
-the return  values $u$ and $y$,  and thus the liveness  at the program
-points  $\pi_7$  and  $\pi_8$  are $\lbrace  \pu.\sigma  \rbrace$  and
-$\{\py.\sigma\}$.  The liveness at  $\pi_7$ and $\pi_8$ are propagated
-backwards as in traditional liveness analysis.
+To see  our way of  handling the second  issue, we trace  the liveness
+analysis to  an extent. Assume  that the calls  to the $\pf$  place an
+overall demand  of $\sigma$ on the  body of \pf.  on  the now describe
+the process  of finding out the  liveness of each  variable defined in
+the function  $\pf$ assuming  a symbolic demand  $\sigma$ on  it.  For
+obvious reasons, this is also the  demand on the return values $u$ and
+$y$, and thus  the liveness at the program  points $\pi_7$ and $\pi_8$
+are $\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The liveness
+at  $\pi_7$ and  $\pi_8$ are  propagated backwards  as  in traditional
+liveness analysis.
 
@@
 scope of  \pu. However, for simplicity  of exposition, we  do not kill
-  livenesses  of  variables  in  our formulation.   This  does  not
-introduce any imprecision since \pu\  is yet to come into existence at
-$\pi_6$  and  its  liveness  will  not be  consulted  by  the  garbage
-collector at this point.
+the variable in our  formulation.  This does not introduce imprecision
+since  \pu\  is yet  to  come into  existence at  $\pi_6$ and  its
+liveness will not be consulted by the garbage collector at this point.
 %==============================================================
@@
 %======================================================================================
-
 In an  eager  language, \pw\  would  not have  been live  after
@@
   evaluation  of  $(\CDR~\pw)$; instead  a  closure  is created.   The
-  evaluation actually takes place while evaluating $(\RETURN~\pu)$. Thus
+  evaluation actually takes place while evaluating $(\RETURN~u)$. Thus
   it is necessary to record the liveness of $\pw$ at the program point
-  $\pi_8$, indeed at every program point  from $\pi_8$ up to the
-  beginning of the program.
+  $\pi_8$.
 %% The second  difference is that  our \LET s  can be recursive  as in
@@
 %%   to be formulated as a fix-point computation.
-  We thus  regard the body of the function as a
+  To address this issue, we  regard the body of the function as a
   tree and consider paths from the root of this tree to program points
   which     trigger     a     evaluation---\SIF\    conditions     and
-  \RETURN\  expressions. We call  such paths  {\em e-paths}.   For the
+  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
   program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
   \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
-  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
-
-The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-path  to a  liveness environment  that is  common to  all  the program
-points in the path. And a second map $\cal{P}$ maps a program point to
-the set of  return paths on which the point lies.   If a program point
+  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
+  analysis consists of two maps:  The map $\cal{M}$ maps a return path
+  to a liveness  environment that is common to  all the program points
+  in the path. And a second  map $\cal{P}$ maps a program point to the
+  set of  return paths on  which the point  lies.  If a  program point
   happens  to fall  on  more than  one  such path,  then the  liveness
@@
   liveness  environments  of the  individual  paths.   In the  example
-program, ${\cal M}(EP_1)=  \ldots$, ${\cal M}(EP_2)=\ldots$ and ${\cal
-  M}(EP_3)= \ldots$.   Since ${\cal P}(pi_4)$ is  $\lbrace EP_1, EP_2,
-EP_3 \rbrace$,  the liveness  of $pi_4$ is  \ldots.  In the  sequel we
-shall describe how to  obtain $\cal{M}$.  The formulation of $\cal{P}$
-is obvious  and we do  not describe it.   We observe that  the ${ref}$
-rules neither use  ${\cal M}$ directly nor create  new labels. We only
-need to consider the rules for {\LET}, {\SIF}, and {\RETURN}.
+  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
+  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
+  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.
+
+  In  the  sequel we  shall  describe  how  to obtain  $\cal{M}$.  The
+  formulation  of $\cal{P}$  is easy  and we  do not  describe  it any
+  further.
+  We observe that the ${ref}$ rules neither use ${\cal M}$ directly nor 
+  create new labels. We only need to consider the rules for {\LET}, {\SIF}, and
+  {\RETURN}.
 
@@
 % Abstract description of the GC algorithm
+\begin{algorithmic}
+\STATE $new\_cell\gets$ dup\_heapcell(cell)
+\IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
+\STATE $new\_car\gets$ copy($cell.car$)
+\STATE $new\_cell.car\gets$ $new\_car$ 
+\STATE $new\_cdr\gets$ copy(cell.cdr)
+\STATE $new\_cell.cdr\gets$ $new\_cdr$
+\ELSE
+\IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
+\STATE $new\_cell\gets$ dup\_heapcell(cell)
+\STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
+\STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
+\ELSE
+\STATE $new\_cell\gets$ dup\_heapcell(cell)
+\ENDIF
+\ENDIF
+\end{algorithmic}
+
 % Liveness for cons cells & reachability for closures
@@
 required parts of a cons cell are copied. The behavior of the GC algorithm is as described below,
-\begin{verbatim}
-copy cons cell to new loc
-if flag == false
-  copy cdr part of the cell
-  update cdr part of new loc
-end if
-\end{verbatim}
+\begin{algorithmic}
+\STATE $new\_cell\gets$ dup\_heapcell(cell)
+\IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
+\STATE $new\_cdr\gets$ copy(cell.cdr)
+\STATE $new\_cell.cdr\gets$ $new\_cdr$
+\ENDIF
+\end{algorithmic}
 We provide an intuition about the correctness of this algorithm. Consider the following facts about the print stack 
@@
 
+\section{Results}
+%Add graphs for memort usage LGC vs RGC
+%Add graphs/table showing the GC times and other statistics
+\section{Conclusion and Future Work}
+%Mention lower peak memory usage. 
+%Mention faster GC than full LGC and more garbage collected than RGC.
+%Pitch it as sweet spot between full LGC and RGC.
+%Future work
+%Mention full LGC
+%Mention k-liveness, use length example to show 1-liveness is not sufficient.
+%Mention mixed mode GC, 
+% - doing RGC most of the times and switching to LGC in case RGC fails to collect any garbage
+% - doing LGC most of the times but regularly checking the reachability of the heap and 
+% switching to RGC if LGC almost equals RGC
+
 \subsection{References}

### 2015-03-21 paper/paper.tex
End of liveness

--- 2015-03-20 paper/paper.tex
@@
       %probably doesn't scale anyway.
-%\newcommand{\comment}[1]{}
+\newcommand{\mycomment}[1]{}
 \definecolor{Myblue}{rgb}{0.0,0.0,0.5}
@@
   (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
-$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
+$\rho'\oplus\rho, (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
 \hline
@@
 
-\pagebreak
-
 \section{Liveness}\label{sec:liveness}
 
-A variable is {\em live} it  there is a possibility of its value being
+A variable is {\em live} if  there is a possibility of its value being
 used  in future computation  and dead  if it  is definitely  not used.
 Heap-allocated data needs a richer model of liveness which talks about
-the  liveness of pointers  including fields  of $\CONS$  cells.  Using
+liveness  of  pointers  including  fields  of  $\CONS$  cells.   Using
 $\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
@@
 \acdr\acdr\acar\}$, then future computations  can only refer up to the
-second  and  third  members  of  $x$. In  this  notation,  $x  \mapsto
-\{\epsilon\}$ represents access using  $x$ itself and $x \mapsto \{\}$
-indicates that $x$ is dead.  A {\em liveness environment} is a mapping
-from  variables  to   subsets  of  $\{\acar,\acdr\}^\ast$,  but  often
-expressed as  a set, for example by  writing $\{x.\acar\acdr, x.\acdr,
+second and  third members of $x$.   A {\em liveness  environment} is a
+mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
+expressed as  a set, for example by  writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, 
 y.\epsilon\}$               instead               of               $[x
-  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
-  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  
-program  points $\pi$.  The liveness  at $\pi:e$  is the  liveness just
-before executing $e$.
+  \mapsto\{\epsilon,\acdr,\acdr\acdr\},     y\mapsto\{\epsilon\},
+  z\mapsto\{\}]$.   In  this   notation,   $x  \mapsto   \{\epsilon\}$
+represents access using $x$ itself and $x \mapsto \{\}$ indicates that
+$x$ is dead.  We associate  a liveness environment with program points
+$\pi$.  The liveness at $\pi:e$  is the liveness just before executing
+$e$.
 
@@
 abuse notation to  juxtapose an edge label and a  set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
+$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$. A more
+elaborate explanation of these notions appear in \cite{asati14lgc}
+
 %% Finally, we use $\Lfonly$  to range over {\em demand transformers}.
-%% These transform demands on a function call into demands for its formal
-%% parameters:    given    a    user    function    $f$,    defined    by
+%% These transform  demands on  a function call  into demands  for its
+%% formal  parameters:   given  a   user  function  $f$,   defined  by
 %% $(\DEFINE\ (f\  x_1\ \ldots\ x_n)\  \ e_f)$ and called  with demand
-%% $\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.
+%% $\sigma$,  then $\Lf{f}{i}{\sigma}$  is  the liveness  of $x_i$  at
+%% $e_f$.
 
@@
 
-Figure~\ref{fig:live-judge}  describes  our  analysis.   Firstly,  the
-function $\mathit{ref}$,  when given an  application $s$ and  a demand
-$\sigma$,  returns the liveness  generated for  the free  variables of
-$s$.   In  each  case,  the  null demand  ($\emptyset$)  generates  no
-liveness.   Notice  that this  is  different  from  liveness of  eager
-languages, where, for instance,  (\CDR~$x$) would be evaluated even if
-the   demand   on   it   is  $\emptyset$   generating   the   liveness
-$\{x.\epsilon\}$.  The  cases (\PRIM~x~y)  and (\NULLQ~x) are  easy. A
-non-null  demand of  $\sigma$  on (\CDR~$x$),  is  transformed to  the
-liveness  $x.\epsilon,  x.\acdr\sigma$.   In  an opposite  sense,  the
-demand  of  $\acar\sigma$ on  (\CONS~$x$~$y$)  is  transformed to  the
-demand  $\sigma$  on $x$.   Note  that  \CONS\  does not,  by  itself,
-dereference its arguments. Constants generate no liveness.
+Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
+parts. The function $\mathit{ref}$,  when given an application $s$ and
+a demand $\sigma$, returns  the incremental liveness generated for the
+free  variables   of  $s$  due  to  the   application.   The  function
+$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
+expressions.
+
+ In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
+ generates no  liveness.  Notice that this is  different from liveness
+ of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
+ evaluated even for the null  demand, and this would require an access
+ using  $x$.   A  non-null   demand  of  $\sigma$  on  (\CDR~$x$),  is
+ transformed  to  the  liveness  $\{x.\epsilon, x.\acdr\sigma\}$.   In  an
+ opposite  sense, the  demand of  $\acdr\sigma$ on  (\CONS~$y$~$z$) is
+ transformed to  the demand  $\sigma$ on $z$.   Note that  \CONS\ does
+ not, by itself, dereference  its arguments. The rules for (\PRIM~x~y)
+ and (\NULLQ~x) are on similar lines. Constants generate no liveness.
 
@@
 $\Lfonly$  that  represents the  summaries  of  all  functions in  the
-program. $\Lfonly$ expresses how the  demand $\sigma$ on the result of
-a function call is transformed  into liveness on its parameters at the
+program. $\Lfonly_f$ (the component of $\Lfonly$ for a specific
+function $f$),  expresses how the  demand $\sigma$ on the result of
+a call to $f$ is transformed  into the liveness of its parameters at the
 beginning of the call.  $\Lfonly$  is determined by the judgement form
-$\mathit{Prog}    \len   \Lfonly$    using   inference    rule   ({\sc
-  live-define}). This states the  fixpoint property to be satisfied by
+$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
+This  rule  describes the  fixed-point  property  to  be satisfied  by
 $\Lfonly$, namely, the demand transformation assumed for each function
 in  the  program should  be  the  same  as the  demand  transformation
-calculated from their bodies. The algorithmic formulation of this rule
-through  a  grammar will  be  described  in [???].   This yields  the  least
-solution  satisfying  the  fixpoint  property  and  thus  ensures  the
-greatest amount of garbage collected.
-
-%%   Analogously   to  classical  liveness  being
-%% computed  as  a  solution  of  dataflow  equations,  we  require,  via
-%% inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
-%% fixed-point property that:  when we assume $\Lfonly$ to  be the family
-%% of demand transformers for the program then the calculated liveness of
-%% each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
-%% assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
+calculated    from    their   bodies.    As    we    shall   see    in
+Section~\ref{sec:grammar-formulation},  the  least solution  satisfying
+the  rule  is obtained  by  converting  the  rule into  an  equivalent
+grammar.  The least solution also  ensures that the greatest amount of
+garbage is collected.
+
+%%   Analogously to classical liveness being computed as a solution of
+%% dataflow   equations,  we   require,  via   inference   rule  ({\sc
+%% live-define}), $\Lfonly$ to  satisfy the fixed-point property that:
+%% when we  assume $\Lfonly$ to  be the family of  demand transformers
+%% for the program then the  calculated liveness of each function body
+%% $\mathcal{L}(e_f,\sigma,\Lfonly)$    agrees   with    the   assumed
+%% $\Lfonly$.   As  usual,  there  are  often  multiple  solutions  to
 %% $\Lfonly$; all are  safe (see Section~\ref{sec:correctness}) but we
@@
 
-We now describe the  propagation of liveness through expressions. This
-brings  us to  the  second way  in  which liveness  analysis for  lazy
-languages  differ from  eager languages---the point where  an
-expression gets evaluated cannot be statically determined.  To see how
-this  is handled,  consider  the liveness  analysis  of the  following
-function:
+We next  describe the function $\mathcal{L}$  that propagates liveness
+across  expressions.   This brings  us  to  the  second way  in  which
+liveness analysis for lazy languages differ from eager languages---the
+point  where  an  expression   gets  evaluated  cannot  be  statically
+determined.  To  deal with this, we introduce  flow insensitivity into
+our  analysis.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 %% analysis only to fully evaluated data.  If the marking phase during
-%% garbage collection leads to a closure, and the closure is not dead, we
-%% mark the entire reachable memory  under the closure for collection.  A
-%% better scheme  would be  to garbage collect  the closure based  on the
-%% liveness  at the point  of its  creation.  This  is sound  because our
-%% language is statically scoped.  However, this is work in progress, and
-%% the method described in this paper switches over to reachability while
-%% collecting closures.
+%% garbage collection leads to a closure, and the closure is not dead,
+%% we  mark  the  entire   reachable  memory  under  the  closure  for
+%% collection.   A  better scheme  would  be  to  garbage collect  the
+%% closure based on  the liveness at the point  of its creation.  This
+%% is sound because our  language is statically scoped.  However, this
+%% is  work  in progress,  and  the  method  described in  this  paper
+%% switches over to reachability while collecting closures.
 
@@
 
-Assume that the calls to the $\pf$ place an overall demand of $\sigma$
-on the body  of \pf.  For obvious reasons, this is  also the demand on
-the return  values $u$ and $y$,  and thus the liveness  at the program
-points  $\pi_7$  and  $\pi_8$  are $\lbrace  \pu.\sigma  \rbrace$  and
-$\{\py.\sigma\}$.  The liveness at  $\pi_7$ and $\pi_8$ are propagated
-backwards as in traditional liveness analysis.
+Consider  an analysis of  the body  of $\pf$  with a  demand $\sigma$.
+Clearly, this is also the demand on the return values $u$ and $y$
+giving the liveness  at  the  program points  $\pi_7$  and $\pi_8$  as
+$\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The  liveness at
+$\pi_7$  and  $\pi_8$  are  propagated  backwards  as  in  traditional
+liveness analysis.
 
@@
 \pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
-turn generates  a liveness of $\{\pw,~\pw.1\sigma\}$  at $\pi_6$.  The
-liveness of  \pu\ is  also killed at  $\pi_6$ since  it is not  in the
-scope of \pu.   However, for simplicity of exposition,  we do not kill
-livenesses of  variables in our formulation.  This  does not introduce
-any imprecision  since \pu\ is yet  to come into  existence at $\pi_6$
-and its  liveness will  not be consulted  by the garbage  collector at
-this point.
+turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
+$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
+not in the scope of \pu.  However, this does not cause any imprecision
+in the way  of the GC marking more  cells as live, since \pu\  is yet to
+come into existence at $\pi_7$  and its liveness will not be consulted
+by the garbage collector at this point.
 %==============================================================
@@
 record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
-every program point from $\pi_8$ up to the beginning of the program.
-%% The second  difference is that  our \LET s  can be recursive  as in
-%%   program point $\pi_1$.  In such a case, the liveness of a let has
-%%   to be formulated as a fix-point computation.
-  We thus regard the body of the function as a tree and consider paths
-  from  the  root of  this  tree to  program  points  which trigger  a
-  evaluation---\SIF\  conditions and  \RETURN\  expressions.  We  call
-  such paths  {\em e-paths}.  For  the program shown, there  are three
-  such  paths $EP_1  = \pi_1,  \pi_2,  \pi_3, \pi_4,  \pi_5$, $EP_2  =
-  \pi_1, \pi_2, \pi_3, \pi_4, \pi_6, \pi_7, \pi_8$, and $EP_3 = \pi_1,
-  \pi_2, \pi_3, \pi_4, \pi_9$.
+every program  point from $\pi_8$ up  to the beginning  of the program
+(since  we do  not kill  liveness).  We  thus regard  the body  of the
+function as  a tree and consider paths  from the root of  this tree to
+program  points  which  trigger  a evaluation---\SIF\  conditions  and
+\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
+\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
+\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
 
@@
 path to a common liveness environment, and a second map $\cal{P}$ maps
-a program  point to the set of  return paths on which  the point lies.
+a program  point to the set of  e-paths on which  the point lies.
 If a  program  point happens to fall  on more than one  such path, then
@@
 ${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
-EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In the sequel
+EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
+our propagation of liveness is flow-insensitive but path-sensitive.
+In the sequel
 we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
-$\cal{P}$ is obvious  and we do not describe it.   We observe that the
-${ref}$  rules  neither  use   ${\cal  M}$  directly  nor  create  new
-labels. We  only need  to consider the  rules for {\LET},  {\SIF}, and
-{\RETURN}.
-
-  A  {\RETURN~\px}\ with a  demand $\sigma$  forces an  evaluation and
-  hence a  new ${\cal M}$ needs to  be created with {\em  l} mapped to
-  $\lbrace \px.\sigma \rbrace$.  An \SIF~ statement is an union of the
-  ${\cal M}$-maps  from the {\em then}  and the {\em  else} part. Note
-  that each part  can contain more than one  label. Since the variable
-  corresponding to  the condition may not  be part of  either the {\em
-    then} or the  {\em else} part, a new label {\em  l} is created and
-  added to $\cal{M}$. The only liveness associated with {\em l} is the
-  liveness of the condition variable. Also notice that the
-  consequences of laziness---the entire statement including the
-  condition is not evaluated if the demand on it is
-  $\emptyset$. This results in the empty map denoted by
-  $\emptyset_{\cal{M}}$.
-
-  The  processing {\LET}  rule has  an elegant  formulation. Recollect
-  that  our language  permits lazy  \LET  s whose  definitions can  be
-  recursive Thus,  in the expression  $\LET~x \leftarrow~s~\IN~e$, $x$
-  can occur  in $s$.   We model the  definition $x \leftarrow~s$  as a
-  possibly recursive  function which  captures an arbitrary  number of
-  unrollings of the  definition.  We pass to this  function the demand
-  arising  out of the  liveness of  $x$ along  all paths  reaching the
-  beginning of $e$, and let  the \Lfonly abstraction derive the demand
-  on the  variables of $s$.
-
-%% that is the  most complex among  all the
-%%   rules.  A  {\LET} does not  add any new  label but might  modify the
-%%   liveness values of  some variables or might add  liveness values for
-%%   the  let  bound  variable.   Since  there might  be  more  than  one
-%%   {\RETURN} statement following a {\LET}, there might be multiple {\em
-%%     E-paths}  reaching the  {\LET}.  The  demand on  the  let variable
-%%   might be different along different {\em E-paths}.
-
-%% First recall the classical formulation  of liveness (as sets of simple
-%% variables)        on        three-address       instructions,        $
-%% \mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
-%% \mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
-%%   strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
-%% $\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
-%% \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.
-
-%% Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
-%% analogous.
-%% % has three analogous aspects.
-%% % \begin{itemize}
-%% % \item
-%% Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
-%% returns the  incremental liveness {\em generated} by  $s$.  Because we
-%% generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
-%% parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
-%% determine which access paths of its free variables are made live.  The
-%% cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
-%% % The cases for
-%% % $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
-%% % is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
-%% % The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
-%% A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
-%% $\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
-%% argument (even if its result is never used).
-%% This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
-%% (note $\sigma$ may be $\{\}$).
-%% In the opposite  sense, the  demand  of $\acar\sigma$  on
-%% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
-%% that \CONS\ does not, by itself, dereference its arguments.
-%% Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
-%% $\mathit{ref}$
-%% expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
-%% Constants generate no liveness.
-
-\alan{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
+$\cal{P}$ is easy and we do not elaborate it any further.
+
+Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
+The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
+  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
+expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
+  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
+triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
+is created and added to  the union.  Also notice that the consequences
+of  laziness---the entire  expression including  the condition  is not
+evaluated if  the demand  on it is  $\emptyset$.  This results  in the
+empty map denoted by $\emptyset_{\cal{M}}$.
+
+ The  {\LET} rule  has  an elegant  formulation.   Recollect that  our
+ language permits lazy \LET~s whose definitions can be recursive Thus,
+ in the expression $\LET~x  \leftarrow~s~\IN~e$, $x$ can occur in $s$.
+ We  model the  definition $x  \leftarrow~s$ as  a  possibly recursive
+ function which captures an  arbitrary number of ``unrollings'' of the
+ definition.  We pass  to this function the demand  arising out of the
+ liveness   of  $x$   at   the   beginning  of   $e$,   and  let   the
+ \Lfonly\ abstraction derive the liveness on the variables of $s$.
+
+\mycomment{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
 expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
@@
 $\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
-$\mathit{ref}(s,\Lv(z),\Lfonly)$.}
+$\mathit{ref}(s,\Lv(z),\Lfonly)$.
 
@@
 classically) is due to the functional formulation here.
+}%mycomment
 
@@
 arguments.
-
-% \item
-\pagebreak
 
@@
 % Difference between eager and lazy GC
-Another  difference that is  present in  GC of  lazy lanaguage  is the
+
+Another  difference that is  present in  GC of  lazy language  is the
 presence  of  a print  stack.  In  lazy  languages the  evaluation  of
@@
 algorithm is as described below,
-\begin{verbatim}
-copy cons cell to new loc
-if flag == false
-  copy cdr part of the cell
-  update cdr part of new loc
-end if
-\end{verbatim}
+\begin{algorithmic}
+\STATE $new\_cell\gets$ dup\_heapcell(cell)
+\IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
+\STATE $new\_cdr\gets$ copy(cell.cdr)
+\STATE $new\_cell.cdr\gets$ $new\_cdr$
+\ENDIF
+\end{algorithmic}
 We   provide   an   intuition    about   the   correctness   of   this
@@
 GC,
+
 \begin{enumerate}
@@
 \begin{enumerate}
-\item the car part of the cons cell is on the print stack. This means that the GC happened while 
-the car part was being processed. The cdr part is yet to be processed and the reference to the cdr part 
-is not on the print stack. Hence it is sufficient to copy only the cdr part of the cons cell. Any references 
-necessary for processing the car part will be copied due to references on the print stack.
-\item the cdr part of the cons cell is on the print stack. This means that GC happened while 
-the cdr part was being processed.  The car part has already been processed and printed, hence we do not 
-need to copy the car part. Additionally any references that are needed for printing the cdr part will 
-also be copied due to the references on the print stack and hence nothing needs to be copied in this case.
+\item the car part of the cons  cell is on the print stack. This means
+  that the GC happened while the car part was being processed. The cdr
+  part is yet to be processed and the reference to the cdr part is not
+  on the print stack. Hence it is sufficient to copy only the cdr part
+  of the  cons cell. Any  references necessary for processing  the car
+  part will be copied due to references on the print stack.
+\item the cdr part of the cons  cell is on the print stack. This means
+  that GC  happened while the cdr  part was being  processed.  The car
+  part has already been processed and printed, hence we do not need to
+  copy the car  part. Additionally any references that  are needed for
+  printing the cdr  part will also be copied due  to the references on
+  the print stack and hence nothing needs to be copied in this case.
 \end{enumerate}
-% Doing GC for references on the print stack
-%  - using a flag to avoid copying non-live cells
-% Optimizing GC for closures 
-% - use a copied using rgc flag for closures to avoid multiple traversals
-One of the reasosn why reachability based GC is faster than liveness based GC is due to the fact that 
-in case we try to copy a cons cell which is already copied we do not need to traverse it any further while
- doing RGC. The same is not true in the case of LGC, as the set of liveness paths may not be similar in 
-both the traversals. One optimization is to keep track of all the liveness automaton states that a particular 
-cons cell has been visited with and ensure that we do not traverse it with the same state more than once. 
-This optimization requires extra memory for each heap cell and potentially it can hold all the automata states. 
-Therefore it is not a feasible solution. In case of lazy languages since we use reachability to copy closures, 
-some cons cells might be copied using reachability. We can avoid repeated traversals of these cons cells without 
-tracking the states. This only requires the knowledge of whether the cons cell was copied using reachability. 
-Hence we use a flag to remember if the heap cell was copied using reachability at any time, if it is set we do 
-not traverse any further.
+% Doing GC for  references on the print stack - using  a flag to avoid
+% copying non-live  cells Optimizing  GC for closures  - use  a copied
+% using rgc flag for closures to avoid multiple traversals
+One of the  reasosn why reachability based GC  is faster than liveness
+based GC is  due to the fact that  in case we try to copy  a cons cell
+which is  already copied  we do  not need to  traverse it  any further
+while doing RGC. The  same is not true in the case  of LGC, as the set
+of  liveness paths  may not  be similar  in both  the  traversals. One
+optimization is  to keep  track of all  the liveness  automaton states
+that a particular  cons cell has been visited with  and ensure that we
+do  not  traverse  it  with  the  same state  more  than  once.   This
+optimization requires extra memory  for each heap cell and potentially
+it can hold  all the automata states.  Therefore it  is not a feasible
+solution. In case of lazy  languages since we use reachability to copy
+closures, some cons  cells might be copied using  reachability. We can
+avoid  repeated traversals of  these cons  cells without  tracking the
+states. This only requires the  knowledge of whether the cons cell was
+copied using  reachability.  Hence  we use a  flag to remember  if the
+heap cell was  copied using reachability at any time, if  it is set we
+do not traverse any further.
+
+\section{Results}
+%Add graphs for memort usage LGC vs RGC
+%Add graphs/table showing the GC times and other statistics
+\section{Conclusion and Future Work}
+%Mention lower peak memory usage. 
+%Mention faster GC than full LGC and more garbage collected than RGC.
+%Pitch it as sweet spot between full LGC and RGC.
+%Future work
+%Mention full LGC
+%Mention k-liveness, use length example to show 1-liveness is not sufficient.
+%Mention mixed mode GC, 
+% - doing RGC most of the times and switching to LGC in case RGC fails to collect any garbage
+% - doing LGC most of the times but regularly checking the reachability of the heap and 
+% switching to RGC if LGC almost equals RGC
 

### 2015-03-21 paper/paper.tex
Added generating liveness equations section

--- 2015-03-21 paper/paper.tex
@@
 
+  Consider the example program in Figure~\ref{fig:mot-example2}.
+
 \bigskip
@@
 %\comment{
+
+\section{Generating Liveness Equations}
+\begin{figure*}[t!]
+\begin{picture}(100,130)(0,-90)
+  \begin{tabular}{cc}
+    \begin{boxedminipage}{.5\textwidth}
+      {\sf
+        \renewcommand{\arraystretch}{1}{
+          \begin{uprogram}
+            \UNL{1} (\DEFINE\ (\length~\xl)
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
+            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_2~(1~+~\py))$)))))
+        \end{uprogram}}
+        \renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \py) \IN~$\pi_{9}\!\!:\,$ (\RETURN\ \pw)))))))
+\end{uprogram}}
+        } 
+    \end{boxedminipage}
+    & 
+    \begin{boxedminipage}{.5\textwidth}
+      {\sf
+        \begin{tabular}{lc}
+        $\cal{M} = $ 
+            \begin{tabular}{lc}            
+              $[$
+              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
+              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
+              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
+              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $]$
+            \end{tabular} \\
+            $\cal{P} = [$
+              \begin{tabular}{lc}
+              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
+              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
+              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
+              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
+              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
+              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
+              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
+              $]$
+            \end{tabular}
+        \end{tabular}
+      }
+    \end{boxedminipage}
+    \\
+    (a) Example program.&
+    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+        (b) {\em e-path} to liveness environment map \\
+        (c) Prog pt to {\em e-paths } map
+    \end{tabular}}
+  \end{tabular}
+\end{picture}
+\kern -3ex
+
+\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
+\end{figure*}
+
+ For our example program which has  a
+single function \length, this generates the following equations:
+
+  \begin{eqnarray*}
+    \Lf{\length}{1}{\sigma}
+    &=& \epsilonset \cup \acdr\Lf{\length}{1}{\epsilon}
+   \end{eqnarray*}
+
+\subsection{Generating Liveness Equations for Function bodies}
+For the running example, $\length$ has calls from $\mainpgm$ at $\pi_8$
+and a recursive call at $\pi_6$.
+\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
+So $\sigma_{\length} =
+     \deltacall{\length}{\pi_8}{\mainpgm}  \cup \deltacall{\length}{\pi_6}{\length}$.
+Filling in  the values gives:
+
+\begin{align*}
+\sigma_{\length}    &=
+ \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
+\end{align*}
+
+For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $\pi_3, \pi_4, \pi_7$ in $\length$ and
+$\pi_9$ in $e_\mainpgm$, this gives
+
+\begin{align*}
+\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\epsilon} \cup \epsilonset \\
+\Lanv{7}{\pu} &= \Lf{\length}{1}{\epsilon} \\
+\Lanv{7}{\py} &= \epsilonset \\
+\Lanv{3}{\xl} &= \epsilonset \\
+\Lanv{3}{\px} &= \epsilonset 
+\\[1ex]
+  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
+  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
+\end{align*}
+
+\subsection{Solving liveness equations---the grammar interpretation}
+The  equations above can now be re-interpreted as a
+context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
+\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
+a variable  $X$ occurring on  the LHS of the  equations generated
+from the analysis.  We can  think of the resulting productions as
+being associated with several  grammars, one for each non-terminal
+\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
+grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
+productions:
+ \begin{eqnarray*}
+\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
+  {\length}{1}{\epsilon}} \mid \epsilon \\
+\end{eqnarray*}
 

### 2015-03-21 paper/paper.tex

--- 2015-03-21 paper/paper.tex
@@
 %======================================================================================
-
 In an eager language, \pw\ would not have been live after $\pi_7$, the
@@
 arguments.
+
+%\comment{
+
+\section{Generating Liveness Equations}
+\begin{figure*}[t!]
+\begin{picture}(100,130)(0,-90)
+  \begin{tabular}{cc}
+    \begin{boxedminipage}{.5\textwidth}
+      {\sf
+        \renewcommand{\arraystretch}{1}{
+          \begin{uprogram}
+            \UNL{1} (\DEFINE\ (\length~\xl)
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
+            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_2~(1~+~\py))$)))))
+        \end{uprogram}}
+        \renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \py) \IN~$\pi_{9}\!\!:\,$ (\RETURN\ \pw)))))))
+\end{uprogram}}
+        } 
+    \end{boxedminipage}
+    & 
+    \begin{boxedminipage}{.5\textwidth}
+      {\sf
+        \begin{tabular}{lc}
+        $\cal{M} = $ 
+            \begin{tabular}{lc}            
+              $[$
+              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
+              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
+              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
+              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $]$
+            \end{tabular} \\
+            $\cal{P} = [$
+              \begin{tabular}{lc}
+              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
+              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
+              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
+              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
+              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
+              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
+              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
+              $]$
+            \end{tabular}
+        \end{tabular}
+      }
+    \end{boxedminipage}
+    \\
+    (a) Example program.&
+    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+        (b) {\em e-path} to liveness environment map \\
+        (c) Prog pt to {\em e-paths } map
+    \end{tabular}}
+  \end{tabular}
+\end{picture}
+\kern -3ex
+
+\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
+\end{figure*}
+
+ For our example program which has  a
+single function \length, this generates the following equations:
+
+  \begin{eqnarray*}
+    \Lf{\length}{1}{\sigma}
+    &=& \epsilonset \cup \acdr\Lf{\length}{1}{\epsilon}
+   \end{eqnarray*}
+
+\subsection{Generating Liveness Equations for Function bodies}
+For the running example, $\length$ has calls from $\mainpgm$ at $\pi_8$
+and a recursive call at $\pi_6$.
+\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
+So $\sigma_{\length} =
+     \deltacall{\length}{\pi_8}{\mainpgm}  \cup \deltacall{\length}{\pi_6}{\length}$.
+Filling in  the values gives:
+
+\begin{align*}
+\sigma_{\length}    &=
+ \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
+\end{align*}
+
+For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $\pi_3, \pi_4, \pi_7$ in $\length$ and
+$\pi_9$ in $e_\mainpgm$, this gives
+
+\begin{align*}
+\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\epsilon} \cup \epsilonset \\
+\Lanv{7}{\pu} &= \Lf{\length}{1}{\epsilon} \\
+\Lanv{7}{\py} &= \epsilonset \\
+\Lanv{3}{\xl} &= \epsilonset \\
+\Lanv{3}{\px} &= \epsilonset 
+\\[1ex]
+  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
+  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
+\end{align*}
+
+\subsection{Solving liveness equations---the grammar interpretation}
+The  equations above can now be re-interpreted as a
+context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
+\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
+a variable  $X$ occurring on  the LHS of the  equations generated
+from the analysis.  We can  think of the resulting productions as
+being associated with several  grammars, one for each non-terminal
+\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
+grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
+productions:
+ \begin{eqnarray*}
+\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
+  {\length}{1}{\epsilon}} \mid \epsilon \\
+\end{eqnarray*}
+
+%}
+
+% \item
 

### 2015-03-22 lazy-scheme/src/SchemeAST_Defs.cpp
Fixed a bug handling 0-ary functions

@@ -551,7 +551,11 @@ cons* LetExprNode::evaluate()
 		if (!isFunctionCall )
 			num_cells_reqd = 1;
 		else if (isFunctionCall)
+		{
 			num_cells_reqd = ((FuncExprNode*)(getVarExpr()))->pListArgs->size();
+			//We need to handle 0-ary functions. We need at least one cons cell even for a 0-ary function call.
+			num_cells_reqd = (num_cells_reqd == 0) ? 1 : num_cells_reqd;
+		}
 //		cerr << "Num of cons cells required is " << num_cells_reqd<<endl;
 		if (check_space(num_cells_reqd * sizeof(cons)) == 0)
 		{

### 2015-03-22 lazy-scheme/src/gc.cpp

@@ -67,6 +67,8 @@ unsigned int num_of_allocations = 0;
 map<cons*, int> heap_map;
 map<int, string> root_var_map;
 
+
+
 #ifdef __DEBUG__GC
 ofstream gcout("gc_messages.txt", ios::out);
 #else
@@ -1044,10 +1046,10 @@ void printval(void *ref, ostream& out)
 		out<<")";
 	}
 	else if(cref->typecell == constIntExprClosure)
-	  {
-	    out << cref->val.intVal;
-	    out << endl;
-	  }
+	{
+		out << cref->val.intVal;
+		out << endl;
+	}
 	else if(cref->typecell == constBoolExprClosure)
 		out << cref->val.boolval;
 	else if(cref->typecell == constStringExprClosure)
@@ -1412,6 +1414,7 @@ void reachability_gc()
 	}
 	update_heap_ref_stack(pre, 0);
 #ifdef __DEBUG__GC
+	DBG(pre << "Number of cells copied " << ((cons*)freept - (cons*)buffer_live) << " = " << copycells << endl);
 	ofstream postgc("PostGC" + to_string(gccount) + ".txt", ios_base::out);
 	create_heap_bft(postgc);
 	postgc.close();
@@ -2451,7 +2454,9 @@ void print_heap_cell_list(vector<cons*> heap_cell_list, map<cons*, int> heap_map
 	out << "Starting GC " << endl;
 	for(auto elem : heap_cell_list)
 	{
-		assert(elem);
+		//assert(elem);
+		if(!elem)
+			continue;
 
 		if (index < num_root_vars)
 		{
@@ -2562,58 +2567,61 @@ void create_heap_bft(ostream& out)
 	{
 
 		cons* curr_cell = heap_cell_list[curr_index];
-
-		switch(curr_cell->typecell)
+		if (curr_cell)
 		{
-		case consExprClosure: {
-			cons* car_part = curr_cell->val.cell.car;
 
-			if (heap_map.find(car_part) == heap_map.end())
+			switch(curr_cell->typecell)
 			{
-				heap_map[car_part] = index;
-				++index;
-				heap_cell_list.push_back(car_part);
-			}
-			cons* cdr_part = curr_cell->val.cell.cdr;
-			if (heap_map.find(cdr_part) == heap_map.end())
-			{
-				heap_map[cdr_part] = index;
-				++index;
-				heap_cell_list.push_back(cdr_part);
-			}
-		}
-		break;
-		case unaryprimopExprClosure: {
[…]

### 2015-03-22 lazy-scheme/src/Simulator.cpp
no fixed (../benchmarks/programs/) path now

@@ -121,6 +121,8 @@ void write_grammar_to_text_file(demand_grammar *g, string filename)
 	return;
 }
 
+string outdir = "./output/";
+
 Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //This method or the constructor should take other parameters like gctype, heap size etc...
 {
 	//Do any per program initialization here.
@@ -138,8 +140,10 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 ///////////////////////////////////////////////////////////////////////
 	bool filesCached = true;
 	struct stat buffer;
-	bool state_map_file =  (stat (("../benchmarks/programs/" + pgmname + "/fsmdump-" + pgmname + "-state-map").c_str(), &buffer) == 0);
-	bool state_transition_file =  (stat (("../benchmarks/programs/" + pgmname + "/fsmdump-" + pgmname + "-state-transition-table").c_str(), &buffer) == 0);
+    mkdir(outdir.c_str(), 0755);
+    mkdir((outdir+pgmname).c_str(), 0755);
+	bool state_map_file =  (stat ((outdir + pgmname + "/fsmdump-" + pgmname + "-state-map").c_str(), &buffer) == 0);
+	bool state_transition_file =  (stat ((outdir + pgmname + "/fsmdump-" + pgmname + "-state-transition-table").c_str(), &buffer) == 0);
 	filesCached = state_map_file && state_transition_file;
 
 	if (gc_type == gc_live && !filesCached)
@@ -158,30 +162,30 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 
 
 		//cout << "program name " << pgmname << endl;
-		write_grammar_to_text_file(&gLivenessData, "../benchmarks/programs/" + pgmname + "/program-cfg.txt");
+		write_grammar_to_text_file(&gLivenessData, outdir + pgmname + "/program-cfg.txt");
 		//Simplify grammar
 		simplifyCFG(&gLivenessData);
-		write_grammar_to_text_file(&gLivenessData, "../benchmarks/programs/" + pgmname + "/simplified-program-cfg.txt");
+		write_grammar_to_text_file(&gLivenessData, outdir + pgmname + "/simplified-program-cfg.txt");
 		//Convert CFG to strongly regular grammar
[…]

### 2015-03-22 lazy-scheme/src/gc.cpp

@@ -35,6 +35,7 @@ std::ostream null_stream(&null_buffer);
 extern GCStatus gc_status;
 extern double gctime;
 extern int gcinvoke;
+extern string outdir;
 
 //globals
 void *freept;
@@ -1040,7 +1041,6 @@ cons* getCar(void* ref, const char fromGC)
 	   cout << "Error dereferencing a NULL pointer for car" << endl;
 	cons *conscell=(cons*)ref;
 #ifdef GC_ENABLE_STATS
-    assert(0);
     if (!fromGC)
         update_last_use(conscell);
 #endif
@@ -1059,7 +1059,6 @@ cons* getCdr(void* ref, const char fromGC)
 		   cout << "Error dereferencing a NULL pointer for cdr" << endl;
     cons *conscell=(cons*)ref;
 #ifdef GC_ENABLE_STATS
-    assert(0);
     if (!fromGC)
         update_last_use(conscell);
 #endif
@@ -1074,7 +1073,6 @@ cons* getCdr(void* ref, const char fromGC)
 int isEqualCons (cons* v1,cons* v2, const char fromGC)
 {
 #ifdef GC_ENABLE_STATS
-    assert(0);
     if (!fromGC)
     {
         update_last_use(v1);
@@ -2845,7 +2843,7 @@ void initialize (string program_name, state_index numkeys, GCStatus gc_type)
 			state_transition_table[i][1]=0;
 		}
 
-		string filepath = "../benchmarks/programs/" + program_name + "/fsmdump-" + program_name;
+		string filepath = outdir + program_name + "/fsmdump-" + program_name;
 		//cout << "Reading " << (filepath + "-state-map") << endl;
 		read_state_map_from_file(filepath + "-state-map");
 		//cout << "Reading " << (filepath + "-state-transition-table") << endl;
@@ -2863,7 +2861,7 @@ void initialize (string program_name, state_index numkeys, GCStatus gc_type)
 			state_transition_table[i][1]=0;
 		}
 
-		string filepath = "../benchmarks/programs/" + program_name + "/fsmdump-" + program_name;
+		string filepath = outdir + program_name + "/fsmdump-" + program_name;
 		//cout << "Reading " << (filepath + "-state-map") << endl;
 		read_state_map_from_file(filepath + "-state-map");
 		//cout << "Reading " << (filepath + "-state-transition-table") << endl;

### 2015-03-22 lazy-scheme/src/run_in_loop.py

@@ -5,7 +5,7 @@ import time
 import datetime
 from collections import namedtuple
 
-prog_size = {'nqueens':10000, 'nperm':40000, 'fibheap':500, 'treejoin':1616535, 'sudoku':4075, 'lcss':550, 'lambda':2790}
+prog_size = {'nqueens':10000, 'nperm':40000, 'fibheap':500, 'treejoin':1616535, 'sudoku':4075, 'lcss':550, 'lambda':27900}
 
 gc_info = namedtuple("gc_info", "heap_total heap_left heap_used gc_invocations gc_time exec_time")
 gc_stat = namedtuple("gc_stat", "prog_name reachability liveness")
@@ -29,7 +29,6 @@ def collect_gc_results(filename, prog_dir_path, gc_type):
     heap_total_cmd = grep_cmd + " \"Heap total=[0-9]* \"  " + result_file_name  + int_grep
     heap_left_cmd = grep_cmd + " \"Heap left=[0-9]*\"  " +  result_file_name  + int_grep
     heap_used_cmd = grep_cmd + " \"Heap used=[0-9]*\"   " +  result_file_name  + int_grep
-A
     gc_invocations_cmd = grep_cmd + " \"GC Invocations=[0-9]*\"   " +  result_file_name  + int_grep
     gc_time_cmd = grep_cmd + " \"GC Time=[0-9]*(\.[0-9]*)?\"   " +  result_file_name  + float_grep
     prog_execution_time_cmd = grep_cmd + " \"Program Execution Time=[0-9]*(\.[0-9]*)?\"   " +  result_file_name  + float_grep

### 2015-03-22 paper/paper.tex
Added unit tests and fixed run_in_loop.py

@@
 \RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
-program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
-\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
-\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
+$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
+ and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.
 
@@
 union of  the liveness environments  of the individual paths.   In the
-example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
-${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
-EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
+example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
+${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
+Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
+the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
 our propagation of liveness is flow-insensitive but path-sensitive.
@@
 \begin{figure*}[t!]
-\begin{picture}(100,130)(0,-90)
+%%\begin{picture}(100,130)(0,-90)
+\begin{picture}(100,130)(0, -50)
   \begin{tabular}{cc}
-    \begin{boxedminipage}{.5\textwidth}
+    \begin{boxedminipage}{.51\textwidth}
       {\sf
+        \small
         \renewcommand{\arraystretch}{1}{
@@
             \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
-            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
             \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
             \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_2~(1~+~\py))$)))))
+            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
         \end{uprogram}}
@@
 	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \py) \IN~$\pi_{9}\!\!:\,$ (\RETURN\ \pw)))))))
+	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
 \end{uprogram}}
@@
     & 
-    \begin{boxedminipage}{.5\textwidth}
+    \begin{boxedminipage}{.49\textwidth}
       {\sf
+        \small
         \begin{tabular}{lc}
-        $\cal{M} = $ 
+        $\cal{M_{\mbox{\tiny length}}} = $ 
             \begin{tabular}{lc}            
               $[$
-              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
+              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
               \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
-              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
-              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $ep_1 \mapsto \lbrace \rbrace , $\\
+              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
               $]$
             \end{tabular} \\
-            $\cal{P} = [$
+            $\cal{P_{\mbox{\tiny length}}} = [$
               \begin{tabular}{lc}
-              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
-              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
-              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
-              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
-              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
-              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
-              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
+                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
+                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
+                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
+                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
+                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
+                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
+                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
+                $]$
+              \end{tabular}\\
+              $\cal{M_{\mbox{\tiny main}}} = $ 
+              \begin{tabular}{lc}            
+                $[$
+                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $]$
+              \end{tabular} \\
+              $\cal{P_{\mbox{\tiny main}}} = [$
+                \begin{tabular}{lc}
+                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
+                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \ldots  $ \
                   $]$  
@@
 \kern -3ex
-
+\vspace{0.5in}
 \caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
@@
 \subsection{Generating Liveness Equations for Function bodies}
-For the running example, $\length$ has calls from $\mainpgm$ at $\pi_8$
+For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
 and a recursive call at $\pi_6$.
@@
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_8}{\mainpgm}  \cup \deltacall{\length}{\pi_6}{\length}$.
+     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
 Filling in  the values gives:
@@
 
-For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $\pi_3, \pi_4, \pi_7$ in $\length$ and
-$\pi_9$ in $e_\mainpgm$, this gives
+For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $ep_1, ep_2, ep_3$ in $\length$ and
+$ep_4$ in $e_\main$, this gives
 
 \begin{align*}
-\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\epsilon} \cup \epsilonset \\
-\Lanv{7}{\pu} &= \Lf{\length}{1}{\epsilon} \\
-\Lanv{7}{\py} &= \epsilonset \\
-\Lanv{3}{\xl} &= \epsilonset \\
-\Lanv{3}{\px} &= \epsilonset 
+\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \cup \bX\sigma_{\mathit{\length}} \\
+\Lanv{7}{\pu} &= \Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \\
+\Lanv{7}{\py} &= \bX\sigma_{\mathit{\length}} \\
+\Lanv{3}{\xl} &= \bX\sigma_{\mathit{\length}} \\
+\Lanv{3}{\px} &= \bX\sigma_{\mathit{\length}}
 \\[1ex]
@@
 % Abstract description of the GC algorithm
-\begin{algorithmic}
-\STATE $new\_cell\gets$ dup\_heapcell(cell)
-\IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
-\STATE $new\_car\gets$ copy($cell.car$)
-\STATE $new\_cell.car\gets$ $new\_car$ 
-\STATE $new\_cdr\gets$ copy(cell.cdr)
-\STATE $new\_cell.cdr\gets$ $new\_cdr$
-\ELSE
-\IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
-\STATE $new\_cell\gets$ dup\_heapcell(cell)
-\STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
-\STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
-\ELSE
-\STATE $new\_cell\gets$ dup\_heapcell(cell)
-\ENDIF
-\ENDIF
-\end{algorithmic}
+%% \begin{algorithmic}
+%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
+%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
+%% \STATE $new\_car\gets$ copy($cell.car$)
+%% \STATE $new\_cell.car\gets$ $new\_car$ 
+%% \STATE $new\_cdr\gets$ copy(cell.cdr)
+%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
+%% \ELSE
+%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
+%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
+%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
+%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
+%% \ELSE
+%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
+%% \ENDIF
+%% \ENDIF
+%% \end{algorithmic}
 
@@
 algorithm is as described below,
-\begin{algorithmic}
-\STATE $new\_cell\gets$ dup\_heapcell(cell)
-\IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
-\STATE $new\_cdr\gets$ copy(cell.cdr)
-\STATE $new\_cell.cdr\gets$ $new\_cdr$
-\ENDIF
-\end{algorithmic}
+%% \begin{algorithmic}
+%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
+%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
+%% \STATE $new\_cdr\gets$ copy(cell.cdr)
+%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
+%% \ENDIF
+%% \end{algorithmic}
 We   provide   an   intuition    about   the   correctness   of   this

### 2015-03-22 lazy-scheme/src/run_in_loop.py

@@ -29,7 +29,6 @@ def collect_gc_results(filename, prog_dir_path, gc_type):
     heap_total_cmd = grep_cmd + " \"Heap total=[0-9]* \"  " + result_file_name  + int_grep
     heap_left_cmd = grep_cmd + " \"Heap left=[0-9]*\"  " +  result_file_name  + int_grep
     heap_used_cmd = grep_cmd + " \"Heap used=[0-9]*\"   " +  result_file_name  + int_grep
-A
     gc_invocations_cmd = grep_cmd + " \"GC Invocations=[0-9]*\"   " +  result_file_name  + int_grep
     gc_time_cmd = grep_cmd + " \"GC Time=[0-9]*(\.[0-9]*)?\"   " +  result_file_name  + float_grep
     prog_execution_time_cmd = grep_cmd + " \"Program Execution Time=[0-9]*(\.[0-9]*)?\"   " +  result_file_name  + float_grep

### 2015-03-25 paper/paper.tex
Introducing 0bar 1 bar and 2bar

--- 2015-03-21 paper/paper.tex
@@
 
+%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
 \newcommand{\scmin} {\mbox{\sf\em in}}
@@
 \newcommand{\bang}{\mbox{\sc bang}}
-
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{proposition}[theorem]{Proposition} 
 \begin{document}
@@
 program.  $\Lfonly_f$  (the  component  of $\Lfonly$  for  a  specific
-function $f$),  expresses how the  demand $\sigma$ on the result of
-a call to $f$ is transformed  into the liveness of its parameters at the
+function $f$),  expresses how the demand  $\sigma$ on the  result of a
+call to $f$ is transformed into  the liveness of its parameters at the
 beginning of the call.  $\Lfonly$  is determined by the judgement form
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure*}[t!]
+\begin{picture}(100,130)(0,-90)
+  \begin{tabular}{cc}
+    \begin{boxedminipage}{.5\textwidth}
+      {\sf
+        \renewcommand{\arraystretch}{1}{
+          \begin{uprogram}
+            \UNL{1} $(\DEFINE\ (\length~\xl)$
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow \ (\NULLQ~\xl)~\IN$
+            \UNL{3} \hspace*{.05cm} $\psi_1\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
+            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\py\  $\leftarrow  (\CDR~\xl)$  \IN
+            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\,
+            $(\LET~\pz\  $\leftarrow  (\length~\py)~\IN$
+            \UNL{6} \hspace*{1.15cm}  $\pi_7\!\!:\,
+           $(\LET~\pw\  $\leftarrow  (\PRIM~1~\pz)~\IN$
+            \UNL{7} \hspace*{.05cm}\hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN_1~\pw)))))))$
+        \end{uprogram}}
+
+        \renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
+          $(\LET\ \pc\  $\leftarrow (\length\ \pb)~\IN$
+          \UNL{4}\hspace*{.05cm}\hspace*{.78cm} $\psi_{3}\!\!:\, (\RETURN\ \pc)))))))$
+\end{uprogram}}
+        } 
+    \end{boxedminipage}
+    & 
+    \begin{boxedminipage}{.5\textwidth}
+      {\sf
+        \begin{tabular}{lc}
+        $\cal{M} = $ 
+            \begin{tabular}{lc}            
+              $[$
+              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
+              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
+              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
+              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $]$
+            \end{tabular} \\
+            $\cal{P} = [$
+              \begin{tabular}{lc}
+              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
+              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
+              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
+              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
+              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
+              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
+              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
+              $]$
+            \end{tabular}
+        \end{tabular}
+      }
+    \end{boxedminipage}
+    \\
+    (a) Example program.&
+    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+        (b) {\em e-path} to liveness environment map \\
+        (c) Prog pt to {\em e-paths } map
+    \end{tabular}}
+  \end{tabular}
+\end{picture}
+\kern -3ex
+
+\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
+\end{figure*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
-	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
-	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
-          \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
-          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
-	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
-          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
-	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
-	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
-          \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
-	\end{uprogram}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\medskip
+%% %%%%%%%%%%%%%%%%%%%%%%%%
+
+%% \renewcommand{\arraystretch}{1}{
+%% 	  \begin{uprogram}
+%% 	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
+%% 	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
+%% 	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
+%%           \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
+%%           $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
+%% 	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
+%%           $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
+%% 	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
+%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
+%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
+%% 	\end{uprogram}}
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% \medskip
 
@@
 by the garbage collector at this point.
+
+In an eager language, \pw\ would not have been live after $\pi_7$, the
+point of  its last  use. In a  lazy language, however,  the definition
+$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
+$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
+takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
+record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
+every program  point from $\pi_8$ up  to the beginning  of the program
+(since  we do  not kill  liveness).  We  thus regard  the body  of the
+function as  a tree and consider paths  from the root of  this tree to
+program  points  which  trigger  a evaluation---\SIF\  conditions  and
+\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
+\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
+\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+
+The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
+path to a common liveness environment, and a second map $\cal{P}$ maps
+a program  point to the set of  e-paths on which  the point lies.
+If a  program  point happens to fall  on more than one  such path, then
+the  liveness environment of  the program  point is  the variable-wise
+union of  the liveness environments  of the individual paths.   In the
+example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
+${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
+EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
+our propagation of liveness is flow-insensitive but path-sensitive.
+In the sequel
+we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
+$\cal{P}$ is easy and we do not elaborate it any further.
+
+Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
+The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
+  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
+expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
+  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
+triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
+is created and added to  the union.  Also notice that the consequences
+of  laziness---the entire  expression including  the condition  is not
+evaluated if  the demand  on it is  $\emptyset$.  This results  in the
+empty map denoted by $\emptyset_{\cal{M}}$.
 %==============================================================
@@
 %======================================================================================
-In an eager language, \pw\ would not have been live after $\pi_7$, the
-point of  its last  use. In a  lazy language, however,  the definition
-$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
-$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
-takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
-record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
-every program  point from $\pi_8$ up  to the beginning  of the program
-(since  we do  not kill  liveness).  We  thus regard  the body  of the
-function as  a tree and consider paths  from the root of  this tree to
-program  points  which  trigger  a evaluation---\SIF\  conditions  and
-\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
-program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
-\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
-\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
-
-The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-path to a common liveness environment, and a second map $\cal{P}$ maps
-a program  point to the set of  e-paths on which  the point lies.
-If a  program  point happens to fall  on more than one  such path, then
-the  liveness environment of  the program  point is  the variable-wise
-union of  the liveness environments  of the individual paths.   In the
-example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
-${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
-EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
-our propagation of liveness is flow-insensitive but path-sensitive.
-In the sequel
-we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
-$\cal{P}$ is easy and we do not elaborate it any further.
-
-Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
-evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
-  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
-expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
-  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
-triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
-is created and added to  the union.  Also notice that the consequences
-of  laziness---the entire  expression including  the condition  is not
-evaluated if  the demand  on it is  $\emptyset$.  This results  in the
-empty map denoted by $\emptyset_{\cal{M}}$.
 
@@
 arguments.
+\section{Computing liveness and its encoding as a table}\label{sec:computing}
+
+Section~\ref{sec:liveness} gave a context-sensitive liveness
+analysis and {\color {red}  proved it correct with reference
+  to   a  {\em   minefield}   semantics}.   In   particular,
+\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
+describes  a  liveness  set  for  each program  point  in  a
+function in terms of a  given $\sigma$ and \Lfonly. We still
+have not mentioned what is the specific sigma that should be
+used for each function How is LF to be computed.
+
+It helps to cast the equations  in the form of a grammar. To
+do  this  we  first  have  to  cast  massage  the  equations
+slightly:
+
+\subsubsection{Modifying the liveness rules.}
+
+The      $\mathit{ref}$     rule      for     \CONS,      shown     in
+Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
+\acar\ and \acdr\ from  access paths in $\sigma$.  Mathematically this
+is fine but  causes problems when solving the  liveness equations {\em
+  symbolically}  since such  decomposition  cannot be  expressed as  a
+context-free  grammar.  Similarly,  the rules  for \CAR,  \CDR, \PRIM,
+\NULLQ,  and \SIF\  require  us  to return  $\emptyset$,  on the  null
+demand. While this is mathematically fine, for our grammar formulation
+we would like to treat these operations symbolically.
+
+To  handle  this,  we  first  introduce  three  new  symbols.
+First, \bcar\ and \bcdr\ are defined as:
+\begin{align*}
+ \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
+ \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
+\end{align*}
+
+and \cbar\ as:
+
+\begin{align*}
+ \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
+                       & \{\epsilon\} ~\mbox{otherwise}
+\end{align*}
+
+We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
+\begin{align*}
+&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
+= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
+\mbox{and} \\
+&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
+          =   x.\cbar\sigma \cup x.0\sigma
+\end{align*}
+and the \Lfunonly\ rule
+for \SIF\ as:
+\begin{align*}
+\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
+                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
+        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
+        &[\,l \mapsto  x.\cbar \sigma]
+\end{align*}
+
+The  defintions of \bcar\  and  \bcdr\  induce   the  following
+relation $\hookrightarrow$ over sets of access paths:
+\begin{align*}
+  \sigma_1\bcar\sigma_2  &\hookrightarrow 
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \bcar\sigma_2',\\
+\sigma_1\bcdr\sigma_2  &\hookrightarrow 
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
+  \bcdr\sigma_2',~\mbox{and}\\  
+\sigma_1\cbar \emptyset &\hookrightarrow \emptyset
+\end{align*}
+
+We  call the liveness equations with  this modification
+  \Lfunonly$'$.
+
+The reflexive  transitive closure of  $\hookrightarrow$ will
+be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
+following proposition relates  the original and the modified
+rules.
+
+\begin{proposition}
+Assume that a liveness computation based on the original set
+of rules gives the liveness of the variable $x$ at a program
+point  $\pi_i$   as  $\sigma$  (symbolically,  $\Lanv{i}{x}=
+\sigma$).   Further,  let  $\Lanv{i}{x}= \sigma'$  when  the
+modified rules are used instead of \Lfunonly.  Then $\sigma'
+\stackrel{*}{\hookrightarrow} \sigma$.
+\end{proposition}
+
+%% To see why the  proposition is true, consider an analysis
+%% based on  \Lfunonly$'$ in  which $\sigma$ appears  in the
+%% context      $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.
+%% Let $\alpha \in  \sigma$.  The symbol \bcar(\bcdr) merely
+%% marks a place in  $\alpha$ where the original \CONS\ rule
+%% would have erased  an immediately following \acar(\acdr),
+%% or,  in absence  of  such a  symbol,  would have  dropped
+%% $\alpha$ itself.   Since the  application of any  rule in
+%% \Lfunonly$'$  merely  adds symbols  at  the beginning  of
+%% $\alpha$, the  markers and other symbols  in $\alpha$ are
+%% propagated to  other dependent parts of  program in their
+%% same  relative   positions.   Consequently,  the  erasure
+%% carried   out   at  the   end   of   the  analysis   with
+%% $\stackrel{*}{\hookrightarrow}$ gives  the same result as
+%% obtained through \Lfunonly.
+
+\alan{For practical use we need to solve the liveness equations {\em finitely and
+symbolically}.  As expressed mathematically, and given a fixed program,
+three things are potentially unbounded:
+({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
+({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
+({\em iii\/}) the number of such access paths.
+
+We commonly first solve ({\em i\/}) by reducing the number of distinct
+calling contexts (e.g.\  to a single unified context  in 0-CFA style).
+However, it  turns out that we  can solve the  equations for $\Lfonly$
+symbolically  without  this  reduction,  so  here  we  defer  this  to
+Section~\ref{sec:bodylivenessbodies}.    (It    also   allows   easier
+extension to dynamically determined liveness---future work.)
+%
+We address ({\em ii\/}) and ({\em iii\/}) by 
+% attempting to re-interpret
+re-interpreting
+the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
+rather than a mutually recursive set of equations on sets of access paths.
+
+This requires two ideas.  Firstly we have to control the use of functions---they
+tend to be infinitary and do not occur naturally in CFGs; in particular our
+$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
+to a tuple of access paths.  The former is achieved by using a separate
+meta-variable (later non-terminal)
+$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
+($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
+Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
+expressible finitely (it is a linear form).
+
+Secondly, there are also two technical issues in re-interpreting
+Figure~\ref{fig:live-judge} as a grammar.
+One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
+disappears after variables are separated ($\Lanv{i}{x}$ above). The other
+is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
+not a CFG\@.
+Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.
+
+Finally,  Section~\ref{sec:NFA-approx}
+uses a construction due to
+Mohri and Nederhof~\cite{mohri00regular} to over-approximate
+context-free grammars with regular grammars;
+these are more appropriate for run-time use.
+%
+Hence the overall `big picture' view is that each GC point
+is annotated with a table encoding the DFA for that program point.
+Such program points forming the call string causing the garbage collection
+(and ending in $\CONS$)
+are used by the garbage collector to curtail its reachability-marking phase
+to this over-approximated liveness.
+
+\subsubsection{Modifying the \CONS\ rule.}
+
+The    $\mathit{ref}$    rule for \CONS,    shown    in
+Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
+\acar\  and  \acdr\  from  access paths in  $\sigma$.
+  Mathematically  this  is fine  but causes problems  when
+  solving  the liveness equations  {\em symbolically}  since such
+  decomposition cannot be expressed as a context-free grammar.  
+To  handle this,  we introduce  two  new symbols  \bcar\  and
+\bcdr\ with the properties:
+\[
+ \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
+\hspace{3em} \mbox{and} \hspace{3em}
+ \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
+\]
+We can now rewrite the \CONS\ rule as:
+\begin{eqnarray*}
+\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
+&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
+\end{eqnarray*}
+
+We  call the liveness equations with  this modification
+  \Lfunonly$'$.
+The  defintions of \bcar\  and  \bcdr\  induce   the  following
+relation $\hookrightarrow$ over sets of access paths:
+\begin{eqnarray*}
+  \sigma_1\bcar\sigma_2  \hookrightarrow 
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
+& \qquad\mbox{and}\qquad &
+\sigma_1\bcdr\sigma_2  \hookrightarrow 
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
+\end{eqnarray*}
+
+The reflexive  transitive closure of  $\hookrightarrow$ will
+be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
+following proposition relates \Lfunonly\ and \Lfunonly$'$:
+
+\begin{proposition}
+Assume that a liveness computation based on \Lfunonly\ gives
+the liveness of the variable  $x$ at a program point $\pi_i$
+as $\sigma$ (symbolically, $\Lanv{i}{x}= \sigma$).  Further,
+let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is used instead
+of  \Lfunonly.  Then  $\sigma' \stackrel{*}{\hookrightarrow}
+\sigma$.
+\end{proposition}
+
+To  see why the  proposition is  true, consider  an analysis
+based  on  \Lfunonly$'$ in  which  $\sigma$  appears in  the
+context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.   Let
+$\alpha \in \sigma$.  The symbol \bcar(\bcdr) merely marks a
+place in $\alpha$ where  the original \CONS\ rule would have
+erased an immediately following \acar(\acdr), or, in absence
+of such a symbol, would have dropped $\alpha$ itself.  Since
+the  application of  any  rule in  \Lfunonly$'$ merely  adds
+symbols at the beginning  of $\alpha$, the markers and other
+symbols in $\alpha$ are  propagated to other dependent parts
+of program in  their same relative positions.  Consequently,
+the  erasure carried  out at  the end  of the  analysis with
+$\stackrel{*}{\hookrightarrow}$  gives  the  same result  as
+obtained through \Lfunonly.
+
+\subsection{Generating equations for the demand transformer \Lfonly}
+\label{sec:Lfonlysymbolically}
+
+%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
+We shall consider  the program in Figure~\ref{fig:examp-pgm}
+as a running example; this is the ANF version of the program
+in Figure~\ref{fig:mot-example}.
+
+%%AM no longer used: In  what follows, we shall overload the semantics
+%% of   \Lfunonly\  so  that   \Lfun{e}{\sigma}{\Lfonly}  additionally
+%% records  the liveness  environment at  every sub-expression  of $e$
+%% during its computation.
+%mean the generation
+%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
+%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.
+
+To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
+$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
+{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
+$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
+formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
+of $x_i$.  For our example program which has  a
+single function \append, this generates the following equations:
+% $\Lfonly_{\append}$  is defined by equations:
+  \begin{eqnarray*}
+    \Lf{\append}{1}{\sigma}
+    &=& \epsilonset \cup \acar\bcar\sigma
+    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
+    \Lf{\append}{2}{\sigma}
+    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
+  \end{eqnarray*}
+  In  general, the  equations  for \Lfonly\  are recursive  since
+  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
+  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
+    is  similar to  solving the  differential equation  
+    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
+    that    the    solution     has    the    form    $\mathit{y}
+    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
+    equation  yields  a  quadratic  equation  in  $r$,  and  each
+    solution of $r$ gives rise  to a solution of the differential
+    equation  (in our  setup we  can effectively  pick  the least
+    solution rather than needing linear combinations).  }
+\begin{eqnarray}
+\label{eq:LF:DI}
+  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
+\end{eqnarray}
+where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
+alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
+why \Lfonly\ has this form  is as follows: Recall that
+\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
+the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
+situation in which  the demand $\sigma$ does not  reach the $i$th
+argument.   This happens,  for  instance, when  argument is  used
+only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
+transformation of the demand $\sigma$ before it reaches the $i$th
+argument. This  transformation prefixes strings in 
+$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.
+
+Substituting  the  guessed   form  in  the  equations  describing
+\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
+containing $\sigma$ and those  without $\sigma$ on both sides of the
+resulting equation, we
+get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
+% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
+$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
+solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
+  \append}$, we get:
+  \begin{align*}
+    \Uf{\append}{1} &= \epsilonset
+    \cup \acdr.\Uf{\append}{1}  
+    &\quad 
+    \Uf{\append}{2} &= \Uf{\append}{2} \\
+    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
+    \cup \acdr\Df{\append}{1}\bcdr &\quad 
+    \Df{\append}{2} &= \epsilonset \cup
+    \Df{\append}{2}\bcdr 
+  \end{align*}
+Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.
+
+\subsection{Generating liveness equations \Lv\  for function bodies}
+\label{sec:bodylivenessbodies}
+
+We now calculate a 0-CFA-style summary liveness for each GC point of a program.
+There are two parts to this.
+First, for each function $f$,  we determine a {\em summary demand} $\sigma_f$
+over-approximating any demand $\sigma$ passed to $f$.
+%$\Lf{f}{i}{\sigma}$.
+So, consider a function $g$ containing a call site $\pi$ to $f$, say
+$\pi\!\!:\!(\LET~x \leftarrow (f\,y_1\,\ldots\,y_n)~\IN~e)$.
+Under the assumption that the demand on $g$ is $\sigma_g$, the liveness at $e$
+is $\Lv = \mathcal{L}(e,\sigma_g,\Lfonly)$, and the $\LET$ case of
+Figure~\ref{fig:live-judge} tells us this call site contributes $\Lv(x)$ to the
+demand $\sigma_f$ placed on  $f$.
+Write $\deltacall{f}{\pi}{g}$ for
+this demand (contributed by call site $\pi$ in $g$).
+
+Thus, supposing the $k$ call sites to function $f$ are
+$\pi^1$ (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
+%%% \marginpar{and writing $\sigma_f^i$ for the above $\deltacall{f}{\pi^i}{g^i}$?}
+then the over-approximation requirement on $\sigma_f$
+is achieved by taking $\sigma_f = \deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
+% is achieved by taking $\sigma_f = \sigma_f^1 \cup \cdots \cup \sigma_f^k$.
+
+The expression $e_{\mainpgm}$ is a special case; we assume it may be called externally
+with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$ (denoted
+$\sigma_{\!all}$).
+This is because any  part  of its  value  may be  used  by  the environment---for
+printing  the  result,  for  instance.
+
+% In what follows,  we shall overload the semantics of
+% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
+% 
+% 
+% We assume an overall demand of $\{\acar,\acdr\}^*$ (denoted
+% $\sigma_{\!all}$) on the main expression $e_\mainpgm$.  This is because 
+% any  part  of its  value  may be  used  by  the environment---for
+% printing  the  result,  for  instance.
+
+%% To  generate  equations for  the  body $e_{\mathit{f}}$ of  a
+%% function,  we evaluate \Lfun{e_{\mathit{f}}}{\sigma_{\mathit{f}}}{\Lfonly}.     For ${\sigma_{\mathit{f}}}$,  we have
+%% to consider all calling  contexts during execution.  We therefore
+%% do a 0-CFA approximation by  defining
+%% % $\sigma_\mathit{f}$        as $\bigcup_\calls{f}{k}\!\!\sigma_\calls{f}{k}$,
+%% $\sigma_\mathit{f}$        as $\sigma_\calls{f}{1} \cup \cdots \cup \sigma_\calls{f}{n}$,
+%% where $\calls{f}{k}$,
+%% $\mathit{k} = 1 \ldots n$ represent all the calls to $\mathit{f}$
+%% in  the  program,   and  $\sigma_\calls{f}{k}$  are  the  demands
+%% generated at the calls  by the \rn{let-live} rule while analyzing
+%% the bodies of the caller.
+
+For the running example, $\append$ has calls from $\mainpgm$ at $\pi_9$
+and a recursive call at $\pi_5$.
+So $\sigma_{\append} =
+     \deltacall{\append}{\pi_9}{\mainpgm}  \cup \deltacall{\append}{\pi_5}{\append}$.
+Filling in  the values gives:
+%  where the former is the demand generated at the
+%  call to \append\  in $e_\mainpgm$ the latter
+%  is  the demand  at the  recursive call. 
+%% Propagation of
+%%  $\sigma_{\!all}$ in $e_\mainpgm$ gives    $\sigma_\calls{\append}{1}$   as   $\{\epsilon,   \acdr\}   \cup
+%%  \acdr\acar\sigma_{\!all}$  and propagation of
+%%  $\sigma_{\mathit{\append}}$ in $e_\append$ gives   $\sigma_\calls{\append}{2}$   as
+%%  $\bcdr\sigma_{\mathit{\append}}$.  
+%% Filling in  the values of $\sigma_\calls{\append}{1}$ and
+%% $\sigma_\calls{\append}{2}$, we get:
+\begin{align*}
+\sigma_{\append}    &=
+\left( \{\epsilon,  \acdr\} \cup  \acdr\acar\sigma_{\!all}  \right) ~\cup~
+\bcdr\sigma_{\mathit{\append}}
+\end{align*}
+
+Second, for each function $f$ (possibly $\mainpgm$) we need the liveness at
+each contained GC point
+$\pi\!\!:\!(\LET~x \leftarrow s~\IN~e)$,
+where $s$ is a call either to $\CONS$ or to a user function $g$.
+Given $\sigma_f$ calculated above, this
+is simply $\mathcal{L}(\pi,\sigma_f,\Lfonly)$.
+%%% \marginpar{AM: we've always said this (safe), but actual parameters
+%%% live before the call and dead after needn't be included!}
+
+For the running example, containing GC points $\pi_5, \pi_7$ in $\append$ and
+$\pi_9$ in $e_\mainpgm$, this gives
+(recall Equation~(\ref{eq:LF:DI}) above states $\Lf{f}{i}{\sigma} = \Uf{f}{i} \cup \Df{f}{i}.\sigma$):
+\begin{align*}
+\Lanv{5}{\xtl} &= \Lf{\append}{1}{\bcdr\sigma_{\append}}  \quad &
+                                \Lanv{7}{\xhd}\,\,\, &=  \bcar\sigma_{\append}     \\ 
+\Lanv{5}{\listb} &= \Lf{\append}{2}{\bcdr\sigma_{\append}} &
+                                \Lanv{7}{\xrec} &=  \bcdr\sigma_{\append} \\
+\Lanv{5}{\lista} &= \epsilonset \cup \acar\bcar\sigma_{\append} 
+\\[1ex]
+  \Lanv{9}{\py}  &= \Lf{\append}{1}{\{\epsilon,
+    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}} \\
+  \Lanv{9}{\pz}  &= \Lf{\append}{2}{\{\epsilon,
+    \acdr\} \cup \acdr\acar\sigma_{\mathit{\!all}}}
+\end{align*}
+
+\noindent 
+In summary, the equations generated during liveness analysis are:
+\begin{enumerate}
+\item   For    each   function   $\mathit{f}$,    equations   defining
+\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
+\item  For  each   function  $\mathit{f}$, an  equation  defining
+the summary demand  $\sigma_{\mathit f}$ on $e_f$.
+% \marginpar{Consistency: summary vs cumulative?}
+\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
+an equation defining
+  liveness at each GC point of $e_{\mathit f}$.
+% {\color{red}
+%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
+%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
+% }
+\end{enumerate}
+
+\subsection{Solving liveness equations---the grammar interpretation}
+The  equations above can now be re-interpreted as a
+context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
+\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
+a variable  $X$ occurring on  the LHS of the  equations generated
+from the analysis.  We can  think of the resulting productions as
+% being a grammar with multiple start symbols---one for
+% for each nonterminal \var{\Lanv{i}{x}}.
+being associated with several  grammars, one for each non-terminal
+\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
+grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
+productions:
+ \begin{eqnarray*}
+\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
+  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
+ \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
+  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
+  \acdr\var{\Uf{\append}{1}}\\         
+  \var{\Df{\append}{1}}
+  &\rightarrow&               \acar\bcar               \mid
+  \acdr\var{\Df{\append}{1}}\bcdr\\
+\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
+\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
+\var{$\sigma_{\mathit{\!all}}$}
+\end{eqnarray*}
+Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
+denoted $\mathscr{L}(\var{\Lanv{i}{x}})
+$,  is the desired  solution of
+\Lanv{i}{x}.   
+However, recall from  our earlier  discussion that
+the  decision problem that  we are  interested in  during garbage
+collection is:
+\begin{quote}
+Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
+\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
+\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
+of forward paths only. Then does $\alpha \in \sigma$?
+\end{quote}
+We could convert  the  rules  defining  $\hookrightarrow$  into
+productions and  add them to the grammar. However, this results
+in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
+membership  problem   for  such  grammars   is  undecidable.   We
+circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
+  strongly  regular}  CFGs which  have   easy
+translations to non-deterministic finite state automata (NFA). The
+NFAs are then simplified on the
+basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
+membership of forward access paths. The resulting NFAs are   finally
+converted to DFAs for use during garbage collection.
+%------------------------------------------------------------%
+\subsubsection{Approximating CFGs using NFAs.}
+\label{sec:NFA-approx}
+
+We use the   
+algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
+approximate a CFG  to a   {\em strongly regular\/} grammar.  The
+transformation has  the  property that  if $\Lv$  is a
+non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
+Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
+  L}_{G'}(\Lv)$.    This  is   required  for the
+approximation to be safe with respect to liveness.
+
+We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
+transformation. We  pick the only production that  is affected by
+the transformation---the production for \Df{\append}{1}.  The
+production for \Uf{\append}{1}, while  recursive, is already in
+strongly  regular  form  and   is  therefore  unaffected  by  the
+transformation.
+ \begin{eqnarray*}
+  \var{\Df{\append}{1}}
+  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
+  \acdr\var{\Df{\append}{1}} \\ 
+\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
+\mid \epsilon 
+\end{eqnarray*}
+The  languages
+generated for  \var{\Df{\append}{1}} in  the original grammar  and the
+new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
+\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
+
+%------------------------------------------------------------%
+\subsubsection{Transforming NFAs to Accept Forward Paths:}
+%\label{sec:NFA-elim}
+The strongly regular CFGs obtained after the Mohri-Nederhof
+transformation are
+  first converted  into  NFAs.  The        algorithm         described        in
+Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
+\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
+Thus  \nt\  can  be  used  to  check  
+membership of forward paths.
+
+\begin{figure*}[t]
+\begin{boxedminipage}{\textwidth}
+\begin{center}
+  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
+  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
+%
+  \raggedright{\bf Output:} NFA  \nt\ with  underlying
+  alphabet $\lbrace \acar, \acdr \rbrace$ such that
+  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
+  \\
+%
+
+  \raggedright{\bf Steps:}
+ \begin{algorithmic}
+   \STATE $i \leftarrow 0$ 
+   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
+   without $\epsilon$-moves \cite{hopcraft90toc}
+
+   \REPEAT
+   \STATE $\nt'_{i+1} \leftarrow \nt_i$
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcar$ and
+     outgoing edge to $q''$ with label $\acar$} 
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+   label $\epsilon$.
+   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
+   \ENDFOR
+
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcdr$ and outgoing edge to
+     $q''$ with label $\acdr$} 
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+     label $\epsilon$.
+   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
+   \ENDFOR
+
+   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
+   $\nt'_{i+1}$ without $\epsilon$-moves
+
+   \STATE $i \leftarrow i+1$
+   \UNTIL ($\nt_{i} = \nt_{i-1}$)
+
+   \STATE $\nt \leftarrow \nt_i$
+%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
+ \end{algorithmic}
+\end{center}
+\end{boxedminipage}
+ \caption{Algorithm for transforming an NFA to  accept forward
+   paths only.}\label{algo:simplify-NFA} \figrule
+\end{figure*}
+%------------------------------------------------------------%
+The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
+pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
+process is  continued till  a fixed point  is reached.   When the
+fixed point  is reached, the resulting NFA  contains all possible
+reductions corresponding  to all the  paths in the  original NFA.
+The proofs  of the termination  and correctness of  the algorithm
+are given in our earlier paper~\cite{karkare07liveness}.
+
+\input{liveness-dfa.tex}
+
+We illustrate the algorithm in Figure~\ref{algo:simplify-NFA} by
+constructing the automaton for \var{\Lanv{9}{y}}.
+Figure~\ref{fig:automata-construction} (c) shows the automaton for
+\var{\Lanv{9}{y}} constructed by composing the automata for
+\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
+\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
+get (d). We add an $\epsilon$ edge from $q_2$ to $q_3$ bypassing the
+\bcdr\acdr\ pair from $q_2$ to $q_2$ and then to $q_3$. This is shown
+in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
+is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
+$q_2$.  Removing this $\epsilon$ edge gives the automaton shown in
+(g).  Restricting this automaton to forward edges only, we get the
+final automaton shown in (h). This automaton recognizes $\acdr^{*}\!\mid\acdr^{*}\acar \sigma_{\mathit \!all}$, showing that the entire
+list \py, including its elements, is live at $\pi_9$. Also note that
+the language accepted by the final automaton satisfies the
+prefix-closed property.
 
 %\comment{
-
+}%%%%%%%%%%%%%%%%
 \section{Generating Liveness Equations}
-\begin{figure*}[t!]
-\begin{picture}(100,130)(0,-90)
-  \begin{tabular}{cc}
-    \begin{boxedminipage}{.5\textwidth}
-      {\sf
-        \renewcommand{\arraystretch}{1}{
-          \begin{uprogram}
-            \UNL{1} (\DEFINE\ (\length~\xl)
-            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
-            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
-            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_2~(1~+~\py))$)))))
-        \end{uprogram}}
-        \renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \py) \IN~$\pi_{9}\!\!:\,$ (\RETURN\ \pw)))))))
-\end{uprogram}}
-        } 
-    \end{boxedminipage}
-    & 
-    \begin{boxedminipage}{.5\textwidth}
-      {\sf
-        \begin{tabular}{lc}
-        $\cal{M} = $ 
-            \begin{tabular}{lc}            
-              $[$
-              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
-              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
-              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
-              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
-              $]$
-            \end{tabular} \\
-            $\cal{P} = [$
-              \begin{tabular}{lc}
-              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
-              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
-              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
-              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
-              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
-              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
-              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
-              $]$
-            \end{tabular}
-        \end{tabular}
-      }
-    \end{boxedminipage}
-    \\
-    (a) Example program.&
-    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
-        (b) {\em e-path} to liveness environment map \\
-        (c) Prog pt to {\em e-paths } map
-    \end{tabular}}
-  \end{tabular}
-\end{picture}
-\kern -3ex
-
-\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
-\end{figure*}
 
@@
 and a recursive call at $\pi_6$.
-\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
+%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
 So $\sigma_{\length} =

### 2015-03-27 paper/paper.tex

--- 2015-03-22 paper/paper.tex
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-
-% Push this to a later point
-
-%% Unlike eager languages, a variable may be bound to an expression
-%%   that is not fully evaluated. i.e. a closure.
-
-%% To  deal with the  first issue,  we apply  the results  of liveness
-%% analysis only to fully evaluated data.  If the marking phase during
-%% garbage collection leads to a closure, and the closure is not dead,
-%% we  mark  the  entire   reachable  memory  under  the  closure  for
-%% collection.   A  better scheme  would  be  to  garbage collect  the
-%% closure based on  the liveness at the point  of its creation.  This
-%% is sound because our  language is statically scoped.  However, this
-%% is  work  in progress,  and  the  method  described in  this  paper
-%% switches over to reachability while collecting closures.
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-
-\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
-	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
-	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
-          \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
-          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
-	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
-          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
-	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
-	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
-          \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
-	\end{uprogram}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\medskip
-
-Consider  an analysis of  the body  of $\pf$  with a  demand $\sigma$.
-Clearly, this is also the demand on the return values $u$ and $y$
-giving the liveness  at  the  program points  $\pi_7$  and $\pi_8$  as
-$\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The  liveness at
-$\pi_7$  and  $\pi_8$  are  propagated  backwards  as  in  traditional
-liveness analysis.
-
-Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
-\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
-turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
-$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
-not in the scope of \pu.  However, this does not cause any imprecision
-in the way  of the GC marking more  cells as live, since \pu\  is yet to
-come into existence at $\pi_7$  and its liveness will not be consulted
-by the garbage collector at this point.
-%==============================================================
-\begin{figure*}[t]  
-\begin{eqnarray*}
-\mathit{ref\/}(\kappa,\sigma,\Lfonly)
-          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
-\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
-\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
-        \begin{array}{l l}
-                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
-        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
-        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
-        \emptyset_{\cal{M}?}  & \mbox{otherwise}
-                 \end{array} \\
- \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
-\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
-        \in \mathcal{M}\}
-\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
-\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
- \makebox[0mm]{\hspace*{7cm}   
- $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
-           x_2\, \ldots\, x_n)/x]))$,} \\
- \makebox[0mm]{\hspace*{5.5cm} and 
-     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
-\end{eqnarray*}
-\begin{minipage}{0.85\textwidth}
-\infrule[live-define]
-        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
-           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \len \Lfonly
-\\ \makebox[0mm]{where
-     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
-\end{minipage}
-  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
-\end{figure*}
-%======================================================================================
-In an eager language, \pw\ would not have been live after $\pi_7$, the
-point of  its last  use. In a  lazy language, however,  the definition
-$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
-$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
-takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
-record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
-every program  point from $\pi_8$ up  to the beginning  of the program
-(since  we do  not kill  liveness).  We  thus regard  the body  of the
-function as  a tree and consider paths  from the root of  this tree to
-program  points  which  trigger  a evaluation---\SIF\  conditions  and
-\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
-program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
-$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
- and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.
-
-The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-path to a common liveness environment, and a second map $\cal{P}$ maps
-a program  point to the set of  e-paths on which  the point lies.
-If a  program  point happens to fall  on more than one  such path, then
-the  liveness environment of  the program  point is  the variable-wise
-union of  the liveness environments  of the individual paths.   In the
-example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
-${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
-Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
-the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
-our propagation of liveness is flow-insensitive but path-sensitive.
-In the sequel
-we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
-$\cal{P}$ is easy and we do not elaborate it any further.
-
-Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
-evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
-  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
-expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
-  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
-triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
-is created and added to  the union.  Also notice that the consequences
-of  laziness---the entire  expression including  the condition  is not
-evaluated if  the demand  on it is  $\emptyset$.  This results  in the
-empty map denoted by $\emptyset_{\cal{M}}$.
-
- The  {\LET} rule  has  an elegant  formulation.   Recollect that  our
- language permits lazy \LET~s whose definitions can be recursive Thus,
- in the expression $\LET~x  \leftarrow~s~\IN~e$, $x$ can occur in $s$.
- We  model the  definition $x  \leftarrow~s$ as  a  possibly recursive
- function which captures an  arbitrary number of ``unrollings'' of the
- definition.  We pass  to this function the demand  arising out of the
- liveness   of  $x$   at   the   beginning  of   $e$,   and  let   the
- \Lfonly\ abstraction derive the liveness on the variables of $s$.
-
-\mycomment{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
-expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
-but note the liveness $x.\epsilon$  generated by the latter.  The case
-$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
-instruction:  the liveness  of $e$  is given  by taking  the liveness,
-$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
-liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
-$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
-$\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-
-% \end{itemize}
-
-We make three observations: firstly the rule ({\sc live-define}) has a
-least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
-secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
-inference of mutually recursive  function definitions, and thirdly the
-asymmetry of  demand and liveness (compared to  post- and pre-liveness
-classically) is due to the functional formulation here.
-}%mycomment
-
-Section~\ref{sec:computing}   shows   how   the  demand   transformers
-$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
-analysis) can  be safely  approximated, for each  function, by  a {\em
-  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
-The  summary consists  of a  pair  of a  single demand  and, for  this
-demand, the corresponding  tuple of demands the function  makes on its
-arguments.
-
-%\comment{
-
-\section{Generating Liveness Equations}
 \begin{figure*}[t!]
@@
 \caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
+%% >>>>>>> 67d5a8ebe45703d9726af91b752e9d74152ae455
 \end{figure*}
 
- For our example program which has  a
-single function \length, this generates the following equations:
-
+% Push this to a later point
+
+%% Unlike eager languages, a variable may be bound to an expression
+%%   that is not fully evaluated. i.e. a closure.
+
+%% To  deal with the  first issue,  we apply  the results  of liveness
+%% analysis only to fully evaluated data.  If the marking phase during
+%% garbage collection leads to a closure, and the closure is not dead,
+%% we  mark  the  entire   reachable  memory  under  the  closure  for
+%% collection.   A  better scheme  would  be  to  garbage collect  the
+%% closure based on  the liveness at the point  of its creation.  This
+%% is sound because our  language is statically scoped.  However, this
+%% is  work  in progress,  and  the  method  described in  this  paper
+%% switches over to reachability while collecting closures.
+
+%% %%%%%%%%%%%%%%%%%%%%%%%%
+
+%% \renewcommand{\arraystretch}{1}{
+%% 	  \begin{uprogram}
+%% 	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
+%% 	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
+%% 	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
+%%           \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
+%%           $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
+%% 	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
+%%           $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
+%% 	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
+%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
+%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
+%% 	\end{uprogram}}
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% \medskip
+
+Consider  an analysis of  the body  of $\pf$  with a  demand $\sigma$.
+Clearly, this is also the demand on the return values $u$ and $y$
+giving the liveness  at  the  program points  $\pi_7$  and $\pi_8$  as
+$\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The  liveness at
+$\pi_7$  and  $\pi_8$  are  propagated  backwards  as  in  traditional
+liveness analysis.
+
+Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
+\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
+turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
+$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
+not in the scope of \pu.  However, this does not cause any imprecision
+in the way of the GC marking  more cells as live, since \pu\ is yet to
+come into existence at $\pi_7$  and its liveness will not be consulted
+by the garbage collector at this point.
+
+In an eager language, \pw\ would not have been live after $\pi_7$, the
+point of  its last  use. In a  lazy language, however,  the definition
+$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
+$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
+takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
+record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
+every program  point from $\pi_8$ up  to the beginning  of the program
+(since  we do  not kill  liveness).  We  thus regard  the body  of the
+function as  a tree and consider paths  from the root of  this tree to
+program  points  which  trigger  a evaluation---\SIF\  conditions  and
+\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
+\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
+\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+
+The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
+path to a common liveness environment, and a second map $\cal{P}$ maps
+a program  point to the set of  e-paths on which  the point lies.
+If a  program  point happens to fall  on more than one  such path, then
+the  liveness environment of  the program  point is  the variable-wise
+union of  the liveness environments  of the individual paths.   In the
+example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
+${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
+EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
+our propagation of liveness is flow-insensitive but path-sensitive.
+In the sequel
+we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
+$\cal{P}$ is easy and we do not elaborate it any further.
+
+Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
+The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
+  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
+expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
+  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
+triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
+is created and added to  the union.  Also notice that the consequences
+of  laziness---the entire  expression including  the condition  is not
+evaluated if  the demand  on it is  $\emptyset$.  This results  in the
+empty map denoted by $\emptyset_{\cal{M}}$.
+%==============================================================
+\begin{figure*}[t]  
+\begin{eqnarray*}
+\mathit{ref\/}(\kappa,\sigma,\Lfonly)
+          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
+\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
+          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
+\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
+%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
+          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
+               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
+               \end{array}
+%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
+\\[1ex]
+\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
+\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
+        \begin{array}{l l}
+                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
+        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
+        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
+        \emptyset_{\cal{M}?}  & \mbox{otherwise}
+                 \end{array} \\
+ \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
+\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
+        \in \mathcal{M}\}
+\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
+\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
+ \makebox[0mm]{\hspace*{7cm}   
+ $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
+           x_2\, \ldots\, x_n)/x]))$,} \\
+ \makebox[0mm]{\hspace*{5.5cm} and 
+     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
+\end{eqnarray*}
+\begin{minipage}{0.85\textwidth}
+\infrule[live-define]
+        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
+           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
+              \mbox{ for each $f$ and $\sigma$}
+        }
+        { d_1 \ldots d_k \len \Lfonly
+\\ \makebox[0mm]{where
+     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
+\end{minipage}
+  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
+\end{figure*}
+
+ The  {\LET} rule  has  an elegant  formulation.   Recollect that  our
+ language permits lazy \LET~s whose definitions can be recursive Thus,
+ in the expression $\LET~x  \leftarrow~s~\IN~e$, $x$ can occur in $s$.
+ We  model the  definition $x  \leftarrow~s$ as  a  possibly recursive
+ function which captures an  arbitrary number of ``unrollings'' of the
+ definition.  We pass  to this function the demand  arising out of the
+ liveness   of  $x$   at   the   beginning  of   $e$,   and  let   the
+ \Lfonly\ abstraction derive the liveness on the variables of $s$.
+
+\mycomment{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
+expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
+but note the liveness $x.\epsilon$  generated by the latter.  The case
+$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
+instruction:  the liveness  of $e$  is given  by taking  the liveness,
+$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
+liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
+$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
+$\mathit{ref}(s,\Lv(z),\Lfonly)$.
+
+%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
+
+% \end{itemize}
+
+We make three observations: firstly the rule ({\sc live-define}) has a
+least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
+secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
+inference of mutually recursive  function definitions, and thirdly the
+asymmetry of  demand and liveness (compared to  post- and pre-liveness
+classically) is due to the functional formulation here.
+}%mycomment
+
+Section~\ref{sec:computing}   shows   how   the  demand   transformers
+$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
+analysis) can  be safely  approximated, for each  function, by  a {\em
+  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
+The  summary consists  of a  pair  of a  single demand  and, for  this
+demand, the corresponding  tuple of demands the function  makes on its
+arguments.
+\section{Computing liveness and its encoding as a table}\label{sec:computing}
+
+Section~\ref{sec:liveness} gave a context-sensitive liveness
+analysis and {\color {red}  proved it correct with reference
+  to   a  {\em   minefield}   semantics}.   In   particular,
+\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
+describes  a  liveness  set  for  each program  point  in  a
+function in terms of a  given $\sigma$ and \Lfonly. We still
+have not mentioned what is the specific sigma that should be
+used for each function How is LF to be computed.
+
+It helps to cast the equations  in the form of a grammar. To
+do  this  we  first  have  to  cast  massage  the  equations
+slightly:
+
+\subsubsection{Modifying the liveness rules.}
+
+The      $\mathit{ref}$     rule      for     \CONS,      shown     in
+Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
+\acar\ and \acdr\  from the access paths in  $\sigma$.  Similarly, the
+rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
+$\emptyset$, if $\sigma$ itself $\emptyset$. This creates difficulties
+since  we  want  to  solve  the  equations  arising  out  of  liveness
+symbolically, and realizing these rules requires $\sigma$ to be known.
+
+The  solution  is  to   also  treat  the  operations  mentioned  above
+symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
+\bcdr\ are defined as:
+\begin{align*}
+ \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
+ \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
+\end{align*}
+
+and \cbar\ as:
+
+\begin{align*}
+ \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
+                       & \{\epsilon\} ~\mbox{otherwise}
+\end{align*}
+
+We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
+\begin{align*}
+&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
+= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
+\mbox{and} \\
+&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
+          =   x.\cbar\sigma \cup x.0\sigma
+\end{align*}
+and the \Lfunonly\ rule
+for \SIF\ as:
+\begin{align*}
+\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
+                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
+        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
+        &[\,l \mapsto  x.\cbar \sigma]
+\end{align*}
+The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
+similar to \CAR.
+
+The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
+relation $\hookrightarrow$ over sets of access paths:
+\begin{align*}
+  &\sigma_1\bcar\sigma_2  \hookrightarrow 
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \bcar\sigma_2',\\
+&\sigma_1\bcdr\sigma_2  \hookrightarrow 
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
+  \bcdr\sigma_2',~\mbox{and}\\  
+&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
+\end{align*}
+
+The reflexive  transitive closure of  $\hookrightarrow$ will
+be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
+following proposition relates  the original and the modified
+liveness rules.
+
+\begin{proposition}
+Assume that a liveness computation  based on the original set of rules
+gives the liveness  of the variable $x$ at a  program point $\pi_i$ as
+$\sigma$   (symbolically,   $\Lanv{i}{x}=   \sigma$).   Further,   let
+$\Lanv{i}{x}=  \sigma'$ when the  modified rules  are used  instead of
+\Lfunonly.  Then $\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
+\end{proposition}
+
+For  an explanation  of why  the  proposition holds  for the  modified
+\CONS\  rule,   we  refer   the  reader  to   \cite{asati14lgc}.   The
+proposition also holds for the the other rules for similar reasons.
+
+%% To see why the  proposition is true, consider an analysis
+%% based on  \Lfunonly$'$ in  which $\sigma$ appears  in the
+%% context      $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.
+%% Let $\alpha \in  \sigma$.  The symbol \bcar(\bcdr) merely
+%% marks a place in  $\alpha$ where the original \CONS\ rule
+%% would have erased  an immediately following \acar(\acdr),
+%% or,  in absence  of  such a  symbol,  would have  dropped
+%% $\alpha$ itself.   Since the  application of any  rule in
+%% \Lfunonly$'$  merely  adds symbols  at  the beginning  of
+%% $\alpha$, the  markers and other symbols  in $\alpha$ are
+%% propagated to  other dependent parts of  program in their
+%% same  relative   positions.   Consequently,  the  erasure
+%% carried   out   at  the   end   of   the  analysis   with
+%% $\stackrel{*}{\hookrightarrow}$ gives  the same result as
+%% obtained through \Lfunonly.
+
+\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
+
+\section{Generating Liveness Equations}
+The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
+example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
+how  to generate  the equations  describing the  demand transformation
+\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
+determine    \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   Of   particular
+interest  is the liveness  at the  program point  at the  beginning of
+beginning  of $e_{\mathit{f}}$  (assumed to  be $\pi_1$).  We generate
+equations  of the  form $\Lf{\mathit{f}}{i}{\sigma}  = \Lanv{1}{x_i}$,
+where  $x_i$   is  the  $i^{\mbox{\footnotesize   th}}$  parameter  of
+$\mathit{f}$. Doing this for  \length, we get:
+\begin{align*}
+ \Lanv{1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
+\end{align*}
+Which gives the following equation for \Lfone{\length}: 
   \begin{eqnarray*}
     \Lf{\length}{1}{\sigma}
-    &=& \epsilonset \cup \acdr\Lf{\length}{1}{\epsilon}
+    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
    \end{eqnarray*}
 
-\subsection{Generating Liveness Equations for Function bodies}
+In general, the equations for  \Lfonly\ are recursive since, as in the
+case  of \length,  \Lan{1}\ may,  in turn,  be expressed  in  terms of
+\Lfone{\mathit{f}}.
+
+As mentioned earlier, each of  the liveness rules modifies a demand by
+prefixing   with   symbols   in   the   above   set,   and   therefore
+$\Lf{f}{i}{\sigma}$ has the closed form:
+\begin{eqnarray}
+\label{eq:LF:DI}
+  \Lf{f}{i}{\sigma} &=& \Df{f}{i}.\sigma
+\end{eqnarray}
+where \Df{f}{i} are sets of  strings over the alphabet $\lbrace \acar,
+\acdr,\bcar, \bcdr, \cbar \rbrace$.   Substituting the guessed form in
+the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
+$\sigma$,  we  get the  following  equation  for  \Df{f}{i}: that  are
+independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
+for \Lfonly$_{\mathit f}$.  Applied to \Lfonly$_{\mathit \length}$, we
+get:
+  \begin{align*}
+    \Df{\length}{1} &= \lbrace\cbar\rbrace 
+    \cup \acdr\Df{\length}{1}
+  \end{align*}
+  Note that these equations can also be viewed as CFGs.
+
+\subsection{Generating liveness equations \Lv\  for function bodies}
+\label{sec:bodylivenessbodies}
+
+We now calculate a 0-CFA-style summary liveness for each GC point of a
+program.  There are two parts  to this.  First, for each function $f$,
+we determine a {\em  summary demand} $\sigma_f$ over-approximating any
+demand $\sigma$ passed to $f$.  So, consider a function $g$ containing
+a  call   site  $\pi$   to  $f$,  say   $\pi\!\!:\!(\LET~x  \leftarrow
+(f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.   Let   the  demand  on   $g$  be
+$\sigma_g$ and the  liveness of $x$ at $\pi_i$  be $\Lanv{i}{x}$. Then
+the $\LET$ case of Figure~\ref{fig:live-judge} tells us this call site
+contributes  $\Lanv{i}{x}$ to  the  demand $\sigma_f$  placed on  $f$.
+Write $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
+site $\pi$  in $g$).  Thus, supposing  the $k$ call  sites to function
+$f$  are  $\pi^1$ (in  function  $g^1$)  \ldots  $\pi^k$ (in  function
+$g^k$),  then  the  over-approximation  requirement on  $\sigma_f$  is
+achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
+\cup \deltacall{f}{\pi^k}{g^k}$.
+
+The expression $e_{\mainpgm}$  is a special case; we  assume it may be
+called externally with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$
+(denoted $\sigma_{\!all}$).  This is because any part of its value may
+be used by the environment---for printing the result, for instance.
+{\color {red} What about print?}
+
 For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
 and a recursive call at $\pi_6$.
-\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
+%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
+     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{length}{\pi_6}{\length}$.
 Filling in  the values gives:
@@
 \sigma_{\length}    &=
- \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
+ \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{length}}}
 \end{align*}
@@
 \begin{align*}
-\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \cup \bX\sigma_{\mathit{\length}} \\
-\Lanv{7}{\pu} &= \Lf{\length}{1}{\bX\sigma_{\mathit{\length}}} \\
-\Lanv{7}{\py} &= \bX\sigma_{\mathit{\length}} \\
-\Lanv{3}{\xl} &= \bX\sigma_{\mathit{\length}} \\
-\Lanv{3}{\px} &= \bX\sigma_{\mathit{\length}}
+\Lanv{7}{\xl} &= \acdr\Lf{length}{1}{\bX\sigma_{\mathit{length}}} \cup \bX\sigma_{\mathit{length}} \\
+\Lanv{7}{\pu} &= \Lf{length}{1}{\bX\sigma_{\mathit{\length}}} \\
+\Lanv{7}{\py} &= \bX\sigma_{\mathit{length}} \\
+\Lanv{3}{\xl} &= \bX\sigma_{\mathit{length}} \\
+\Lanv{3}{\px} &= \bX\sigma_{\mathit{length}}
 \\[1ex]
@@
 
-%}
-
-% \item
+In summary, the equations generated during liveness analysis are:
+\begin{enumerate}
+\item   For    each   function   $\mathit{f}$,    equations   defining
+\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
+\item  For  each   function  $\mathit{f}$, an  equation  defining
+the summary demand  $\sigma_{\mathit f}$ on $e_f$.
+% \marginpar{Consistency: summary vs cumulative?}
+\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
+an equation defining
+  liveness at each GC point of $e_{\mathit f}$.
+
+\subsection{Solving liveness equations---the grammar interpretation}
+The  equations above can now be re-interpreted as a
+context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
+\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
+a variable  $X$ occurring on  the LHS of the  equations generated
+from the analysis.  We can  think of the resulting productions as
+% being a grammar with multiple start symbols---one for
+% for each nonterminal \var{\Lanv{i}{x}}.
+being associated with several  grammars, one for each non-terminal
+\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
+grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
+productions:
+ \begin{eqnarray*}
+\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
+  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
+ \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
+  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
+  \acdr\var{\Uf{\append}{1}}\\         
+  \var{\Df{\append}{1}}
+  &\rightarrow&               \acar\bcar               \mid
+  \acdr\var{\Df{\append}{1}}\bcdr\\
+\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
+\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
+\var{$\sigma_{\mathit{\!all}}$}
+\end{eqnarray*}
+Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
+denoted $\mathscr{L}(\var{\Lanv{i}{x}})
+$,  is the desired  solution of
+\Lanv{i}{x}.   
+However, recall from  our earlier  discussion that
+the  decision problem that  we are  interested in  during garbage
+collection is:
+\begin{quote}
+Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
+\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
+\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
+of forward paths only. Then does $\alpha \in \sigma$?
+\end{quote}
+We could convert  the  rules  defining  $\hookrightarrow$  into
+productions and  add them to the grammar. However, this results
+in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
+membership  problem   for  such  grammars   is  undecidable.   We
+circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
+  strongly  regular}  CFGs which  have   easy
+translations to non-deterministic finite state automata (NFA). The
+NFAs are then simplified on the
+basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
+membership of forward access paths. The resulting NFAs are   finally
+converted to DFAs for use during garbage collection.
+%------------------------------------------------------------%
+\subsubsection{Approximating CFGs using NFAs.}
+\label{sec:NFA-approx}
+
+We use the   
+algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
+approximate a CFG  to a   {\em strongly regular\/} grammar.  The
+transformation has  the  property that  if $\Lv$  is a
+non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
+Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
+  L}_{G'}(\Lv)$.    This  is   required  for the
+approximation to be safe with respect to liveness.
+
+We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
+transformation. We  pick the only production that  is affected by
+the transformation---the production for \Df{\append}{1}.  The
+production for \Uf{\append}{1}, while  recursive, is already in
+strongly  regular  form  and   is  therefore  unaffected  by  the
+transformation.
+ \begin{eqnarray*}
+  \var{\Df{\append}{1}}
+  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
+  \acdr\var{\Df{\append}{1}} \\ 
+\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
+\mid \epsilon 
+\end{eqnarray*}
+The  languages
+generated for  \var{\Df{\append}{1}} in  the original grammar  and the
+new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
+\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
+
+%------------------------------------------------------------%
+\subsubsection{Transforming NFAs to Accept Forward Paths:}
+%\label{sec:NFA-elim}
+The strongly regular CFGs obtained after the Mohri-Nederhof
+transformation are
+  first converted  into  NFAs.  The        algorithm         described        in
+Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
+\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
+Thus  \nt\  can  be  used  to  check  
+membership of forward paths.
+
+\begin{figure*}[t]
+\begin{boxedminipage}{\textwidth}
+\begin{center}
+  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
+  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
+%
+  \raggedright{\bf Output:} NFA  \nt\ with  underlying
+  alphabet $\lbrace \acar, \acdr \rbrace$ such that
+  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
+  \\
+%
+
+  \raggedright{\bf Steps:}
+ \begin{algorithmic}
+   \STATE $i \leftarrow 0$ 
+   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
+   without $\epsilon$-moves \cite{hopcraft90toc}
+
+   \REPEAT
+   \STATE $\nt'_{i+1} \leftarrow \nt_i$
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcar$ and
+     outgoing edge to $q''$ with label $\acar$} 
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+   label $\epsilon$.
+   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
+   \ENDFOR
+
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcdr$ and outgoing edge to
+     $q''$ with label $\acdr$} 
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+     label $\epsilon$.
+   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
+   \ENDFOR
+
+   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
+   $\nt'_{i+1}$ without $\epsilon$-moves
+
+   \STATE $i \leftarrow i+1$
+   \UNTIL ($\nt_{i} = \nt_{i-1}$)
+
+   \STATE $\nt \leftarrow \nt_i$
+%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
+ \end{algorithmic}
+\end{center}
+\end{boxedminipage}
+ \caption{Algorithm for transforming an NFA to  accept forward
+   paths only.}\label{algo:simplify-NFA} \figrule
+\end{figure*}
+
+The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
+pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
+process is  continued till  a fixed point  is reached.   When the
+fixed point  is reached, the resulting NFA  contains all possible
+reductions corresponding  to all the  paths in the  original NFA.
+The proofs  of the termination  and correctness of  the algorithm
+are given in our earlier paper~\cite{karkare07liveness}.
+
+\input{liveness-dfa.tex}
+
+We  illustrate  the  algorithm  in  Figure~\ref{algo:simplify-NFA}  by
+constructing      the       automaton      for      \var{\Lanv{9}{y}}.
+Figure~\ref{fig:automata-construction}  (c)  shows  the automaton  for
+\var{\Lanv{9}{y}}   constructed   by   composing  the   automata   for
+\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
+\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
+get (d). We  add an $\epsilon$ edge from $q_2$  to $q_3$ bypassing the
+\bcdr\acdr\ pair from $q_2$ to $q_2$  and then to $q_3$. This is shown
+in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
+is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
+$q_2$.   Removing this $\epsilon$  edge gives  the automaton  shown in
+(g).  Restricting  this automaton  to forward edges  only, we  get the
+final   automaton   shown    in   (h).   This   automaton   recognizes
+$\acdr^{*}\!\mid\acdr^{*}\acar  \sigma_{\mathit \!all}$,  showing that
+the entire list \py, including  its elements, is live at $\pi_9$. Also
+note that the  language accepted by the final  automaton satisfies the
+prefix-closed property.
+
+\hrule
+\hrule
 
 \section{Implementation}
-% GC assumptions
 By the  time the program  starts executing, the liveness  automata are
 created and stored in a file.   This is re-used every time the program
-executes.
-% GC points - mention reduction of gc points due to e-paths
-An important  difference in GC of  lazy language compared  to an eager
-language  is  the  point  of  garbage collection.  In  case  of  eager
-languages only \CONS~ points and  function calls are the GC points. In
-a lazy language  every \LET~ statement is a  potential GC point. Hence
-we  need  to store  liveness  automata  corresponding  to every  \LET~
-statement in the program. This leads  to an explosion in the number of
-automata. From our observation that  along an {\em E-path} all program
-points have  the same liveness, we  can reduce the  number of automata
-state required  by sharing  the same automata  for all  program points
-along an {\em E-path}. In the  example program, we only need to create
-automata for program  points \ldots, since they are  the only distinct
-{\em E-path}s in the program.
+executes.  An important difference in  GC of lazy language compared to
+an  eager language is  the point  of garbage  collection.  In  case of
+eager  languages only  \CONS~ points  and  function calls  are the  GC
+points. In  a lazy  language every \LET~  statement is a  potential GC
+point. Hence we need to store liveness automata corresponding to every
+\LET~  statement in the  program. This  leads to  an explosion  in the
+number of  automata. From our  observation that along an  {\em E-path}
+all program points have the same liveness, we can reduce the number of
+automata state required  by sharing the same automata  for all program
+points along an {\em E-path}. In  the example program, we only need to
+create automata  for program  points \ldots, since  they are  the only
+distinct {\em E-path}s in the program.
 % Abstract description of the GC algorithm
@@
 
+\end{enumerate}
+
+% {\color{red}
+%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
+%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
+% }
+
+%% =======
+
+%------------------------------------------------------------%
+
 \end{document}

### 2015-03-27 paper/paper.tex
Updated Implementation Section

--- 2015-03-22 paper/paper.tex
@@
 	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
+	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  (\CAR~\pa) \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pa) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
 \end{uprogram}}
@@
                 $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pa.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                   $]$
@@
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
+     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{length}{\pi_6}{\length}$.
 Filling in  the values gives:
@@
 executes.
+\subsection{GC points and automata sharing}
 % GC points - mention reduction of gc points due to e-paths
-An important  difference in GC of  lazy language compared  to an eager
-language  is  the  point  of  garbage collection.  In  case  of  eager
-languages only \CONS~ points and  function calls are the GC points. In
-a lazy language  every \LET~ statement is a  potential GC point. Hence
-we  need  to store  liveness  automata  corresponding  to every  \LET~
-statement in the program. This leads  to an explosion in the number of
-automata. From our observation that  along an {\em E-path} all program
-points have  the same liveness, we  can reduce the  number of automata
-state required  by sharing  the same automata  for all  program points
-along an {\em E-path}. In the  example program, we only need to create
-automata for program  points \ldots, since they are  the only distinct
-{\em E-path}s in the program.
+We describe a simple outline of the GC algorthm in figure {\em add refernce to algorithm}.
+In a lazy language each \LET~ expression creates a closure, and hence allocates memory 
+on the heap. Therefore in contrast to an eager language, every \LET~ statement is a  potential GC point. 
+As a GC can be triggered at any of the \LET~ expressions, we  need  to store  liveness  automata  
+corresponding  to every  \LET~ statement in the program. 
+Even though this leads  to an explosion in the number of
+automata, we can optimize it using our observation that  along an {\em E-path} all program
+points have  the same liveness. We therefore create a single liveness environment corresponding to each 
+unique {\em E-path} sets and share the same automata  for all  program points
+having a common {\em E-path} set. In the  example program {\em refer to the example program}, we only need to create
+automata for the {\em E-path} sets \ldots, since they are  the only distinct
+{\em E-path} sets in the program.
 % Abstract description of the GC algorithm
@@
 
+\subsection{Liveness based GC algorithm}
 % Liveness for cons cells & reachability for closures
-During GC, if the heap cell being processed is a cons cell, we consult
-the  liveness automaton  corresponding to  it and  copy only  the live
-paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
-approximate the  liveness to be {\{0+1\}\*} and  copy everything under
-the closure.
-% Handling of mixture of LGC & RGC
-A consequence of  copying everything under a closure  is the fact that
-we can end up trying to copy things which are not live and corrupt the
-heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
-liveness. At  a later stage this  cons cell becomes part  of a closure
-and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
-under the  cons cell  thus copying heap  locations which might  not be
-valid.  To avoid  this behavior we either have to choose  to do a full
-liveness based  GC for closure or  we should ensure  that any non-live
-references are  not carried across  GC's. We take the  latter approach
-and ensure that  non-live references are set to  null. After every GC,
-the live part of the buffer  is scanned and any reference which is not
-live is  set to null.  The correctness of  this follows from  the fact
-that any reference which is decalred  not live in a $GC_k$ will not be
-derefernced during later execution.
+When ever the program runs out of memory, we scan the current buffer and we check if 
+each reference is live or not and using a depth first traversal copy its children recursively.
+The handling of a heap cell depends on the type of data it stores, as described in the algorithm{\em add reference to algorithm}.
+\begin{itemize}
+\item An atomic value is copied if it is live.
+\item A \CONS~ cell is copied if it is live. Its \CAR~ and \CDR~ part are copied recursively if they are live.
+\item A closure and everything reachable from it are copied if it is live otherwise nothing is copied. 
+\end{itemize}
+
+Observe that we are using liveness based GC for \CONS~ cells whereas we do a reachability based GC for closures. 
+Since we do not nullify references that are not live this can lead to copying of incorrect references. 
+Consider a \CONS~ cell {\em c} whose \CAR~ part is live but \CDR~ part is not live. During GC the pointer corresponding to the 
+\CDR~ part is neither copied and nor nullified. Post GC if this cell becomes part of a closure then during a subsequent GC we will do 
+a reachability based GC on {\em c} and hence try to chase the pointer corresponding to its \CDR~ part. 
+One way to avoid this problem is by setting references which are not live to null. Doing this during GC is expensive as it 
+requires an alias analysis. We avoid this overhead by waiting for the GC to end and then setting any references which point to 
+the dead buffer to null. From the definition of liveness any reference which is declared not live will not be referenced beyond that 
+execution point till the end of the execution. Hence once the GC is complete any pointer in the live buffer which points to a location 
+in the dead buffer can be nullified safely. Although this involves scanning the entire live buffer again, the overhead is not too bad as 
+the expected number of live heap cells in the live buffer (immediately after a GC) is usually much less compared to the size of the live buffer. Also, by 
+copying closures using reachability we can perform certain optimizations which will not be possible if we use only liveness  based GC.
+We describe this optimization in section{\em add reference}.
 % A reference copied using LGC could become part of a closure in a later GC
@@
 
+\subsection{Printing and GC on the print stack}
 Another  difference that is  present in  GC of  lazy language  is the

### 2015-03-27 paper/paper.tex

--- 2015-03-27 paper/paper.tex
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+
 % Push this to a later point
@@
 by the garbage collector at this point.
-
 In an eager language, \pw\ would not have been live after $\pi_7$, the
@@
 \end{figure*}
+%======================================================================================
+In an eager language, \pw\ would not have been live after $\pi_7$, the
+point of  its last  use. In a  lazy language, however,  the definition
+$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
+$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
+takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
+record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
+every program  point from $\pi_8$ up  to the beginning  of the program
+(since  we do  not kill  liveness).  We  thus regard  the body  of the
+function as  a tree and consider paths  from the root of  this tree to
+program  points  which  trigger  a evaluation---\SIF\  conditions  and
+\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
+$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
+ and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.
+
+The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
+path to a common liveness environment, and a second map $\cal{P}$ maps
+a program  point to the set of  e-paths on which  the point lies.
+If a  program  point happens to fall  on more than one  such path, then
+the  liveness environment of  the program  point is  the variable-wise
+union of  the liveness environments  of the individual paths.   In the
+example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
+${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
+Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
+the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
+our propagation of liveness is flow-insensitive but path-sensitive.
+In the sequel
+we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
+$\cal{P}$ is easy and we do not elaborate it any further.
+
+Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
+The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
+  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
+expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
+  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
+triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
+is created and added to  the union.  Also notice that the consequences
+of  laziness---the entire  expression including  the condition  is not
+evaluated if  the demand  on it is  $\emptyset$.  This results  in the
+empty map denoted by $\emptyset_{\cal{M}}$.
 
@@
 \section{Computing liveness and its encoding as a table}\label{sec:computing}
+
+%\comment{
+
+\section{Generating Liveness Equations}
+\begin{figure*}[t!]
+%%\begin{picture}(100,130)(0,-90)
+\begin{picture}(100,130)(0, -50)
+  \begin{tabular}{cc}
+    \begin{boxedminipage}{.51\textwidth}
+      {\sf
+        \small
+        \renewcommand{\arraystretch}{1}{
+          \begin{uprogram}
+            \UNL{1} (\DEFINE\ (\length~\xl)
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
+            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
+        \end{uprogram}}
+        \renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
+\end{uprogram}}
+        } 
+    \end{boxedminipage}
+    & 
+    \begin{boxedminipage}{.49\textwidth}
+      {\sf
+        \small
+        \begin{tabular}{lc}
+        $\cal{M_{\mbox{\tiny length}}} = $ 
+            \begin{tabular}{lc}            
+              $[$
+              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
+              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
+              $ep_1 \mapsto \lbrace \rbrace , $\\
+              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $]$
+            \end{tabular} \\
+            $\cal{P_{\mbox{\tiny length}}} = [$
+              \begin{tabular}{lc}
+                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
+                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
+                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
+                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
+                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
+                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
+                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
+                $]$
+              \end{tabular}\\
+              $\cal{M_{\mbox{\tiny main}}} = $ 
+              \begin{tabular}{lc}            
+                $[$
+                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $]$
+              \end{tabular} \\
+              $\cal{P_{\mbox{\tiny main}}} = [$
+                \begin{tabular}{lc}
+                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
+                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \ldots  $ \
+                  $]$  
+                \end{tabular}
+        \end{tabular}
+      }
+    \end{boxedminipage}
+    \\
+    (a) Example program.&
+    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+        (b) {\em e-path} to liveness environment map \\
+        (c) Prog pt to {\em e-paths } map
+    \end{tabular}}
+  \end{tabular}
+\end{picture}
+\kern -3ex
+\vspace{0.5in}
+\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
+\end{figure*}
+
+ For our example program which has  a
+single function \length, this generates the following equations:
 
@@
 {\color {red} What about print?}
-
 For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
@@
 \section{Implementation}
+% GC assumptions
 By the  time the program  starts executing, the liveness  automata are
@@
 % That's all folks!
-
 \end{enumerate}

### 2015-03-27 paper/sections/restOfOldPaper.tex
updates...

@@
 The reflexive transitive closure  of $\hookrightarrow$ will be denoted
-as $\stackrel{*}{\hookrightarrow}$. The following proposition relates \Lfunonly\   and \Lfunonly$'$:
+as $\stackrel{*}{\hookrightarrow}$.  The following proposition relates
+\Lfunonly\ and \Lfunonly$'$:
 
@@
 \begin{proposition}
-Assume that  a liveness computation  based on  \Lfunonly\   gives  the liveness of
-the variable $x$ at a program point $\pi_i$ as $\sigma$ (symbolically, $\Lanv{i}{x}=
-\sigma$).  Further,  let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is
-used instead of \Lfunonly.   Then  $\sigma' \stackrel{*}{\hookrightarrow}
-\sigma$.
+Assume  that a  liveness  computation based  on  \Lfunonly\ gives  the
+liveness of  the variable $x$ at  a program point  $\pi_i$ as $\sigma$
+(symbolically,  $\Lanv{i}{x}=  \sigma$).   Further, let  $\Lanv{i}{x}=
+\sigma'$  when  \Lfunonly$'$  is  used  instead  of  \Lfunonly.   Then
+$\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
 \end{proposition}
@@
 To  see why the  proposition is  true, consider  an analysis  based on
-\Lfunonly$'$ in which   $\sigma$ appears in
-the  context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.  Let $\alpha  \in \sigma$.  The
-symbol  \bcar(\bcdr)  merely  marks  a  place in  $\alpha$  where  the
-original  \CONS\  rule  would  have erased  an  immediately  following
-\acar(\acdr),  or, in  absence of  such a  symbol, would  have dropped
-$\alpha$ itself.  Since the application  of any rule in \Lfunonly$'$ merely
-adds  symbols at  the beginning  of  $\alpha$, the  markers and  other
-symbols in $\alpha$ are propagated to other dependent parts of program
-in their same relative positions.  Consequently, the erasure carried out at the end of the analysis with
+\Lfunonly$'$    in   which   $\sigma$    appears   in    the   context
+$\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.     Let    $\alpha   \in
+\sigma$.   The symbol \bcar(\bcdr)  merely marks  a place  in $\alpha$
+where  the  original \CONS\  rule  would  have  erased an  immediately
+following \acar(\acdr),  or, in absence  of such a symbol,  would have
+dropped  $\alpha$  itself.   Since  the  application of  any  rule  in
+\Lfunonly$'$  merely adds symbols  at the  beginning of  $\alpha$, the
+markers  and  other  symbols  in  $\alpha$  are  propagated  to  other
+dependent  parts   of  program  in  their   same  relative  positions.
+Consequently, the erasure carried out  at the end of the analysis with
 $\stackrel{*}{\hookrightarrow}$  gives  the  same result  as  obtained
@@
 
-%%AM no longer used: In what follows,  we shall overload the semantics of
-%% \Lfunonly\ so that  \Lfun{e}{\sigma}{\Lfonly} additionally records the liveness environment at every sub-expression of $e$ during its computation.  
+%%AM no longer used: In  what follows, we shall overload the semantics
+%% of   \Lfunonly\  so  that   \Lfun{e}{\sigma}{\Lfonly}  additionally
+%% records  the liveness  environment at  every sub-expression  of $e$
+%% during its computation.
 %mean the generation
@@
 this work by the DST/SERC fast track scheme for young scientists.
-
-\bibliography{fun_hra}{}
-\bibliographystyle{splncs}
 

### 2015-04-07 paper/paper.tex
Committing minor changes

@@
             \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
-            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
-            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_3\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\psi_1\!\!:\, (\RETURN~0)$
+            \UNL{4} \hspace*{1.15cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{1.15cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN~(1~+~\py))$)))))
         \end{uprogram}}
@@
 	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
+	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_5\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}   $\pi_6\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_{7}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
 \end{uprogram}}
@@
         \begin{tabular}{lc}
-        $\cal{M_{\mbox{\tiny length}}} = $ 
+        $\cal{M_{\mbox{\tiny \length}}} = $ 
             \begin{tabular}{lc}            
               $[$
-              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
+              $\psi_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
               \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
-              $ep_1 \mapsto \lbrace \rbrace , $\\
-              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $\psi_1 \mapsto \lbrace \rbrace , $\\
+              $\psi_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
               $]$
             \end{tabular} \\
-            $\cal{P_{\mbox{\tiny length}}} = [$
+            $\cal{P_{\mbox{\tiny \length}}} = [$
               \begin{tabular}{lc}
-                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
-                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
-                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
-                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
-                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
-                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
-                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
+                $\pi_{\mbox{1}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \
+                $ \pi_{\mbox{2}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \\
+                $ \pi_{\mbox{3}} \mapsto \lbrace \psi_3 \rbrace, $ \
+                $ \pi_{\mbox{4}} \mapsto \lbrace \psi_1 \rbrace, $ \
+                $ \pi_{\mbox{5}} \mapsto \lbrace \psi_2 \rbrace, $ \\
+                $ \pi_{\mbox{6}} \mapsto \lbrace \psi_2 \rbrace, $ \
+                $ \pi_{\mbox{7}} \mapsto \lbrace \psi_2 \rbrace $ \\
                 $]$
               \end{tabular}\\
-              $\cal{M_{\mbox{\tiny main}}} = $ 
+              $\cal{M_{\mbox{\tiny \main}}} = $ 
               \begin{tabular}{lc}            
                 $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $\psi_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                   $]$
               \end{tabular} \\
-              $\cal{P_{\mbox{\tiny main}}} = [$
+              $\cal{P_{\mbox{\tiny \main}}} = [$
                 \begin{tabular}{lc}
-                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
-                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $\pi_{\mbox{8}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \pi_{\mbox{9}} \mapsto \lbrace \psi_4 \rbrace, $ \\
+                  $ \pi_{\mbox{10}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \pi_{\mbox{11}} \mapsto \lbrace \psi_4 \rbrace, $ \
                   $ \ldots  $ \
@@
   \raggedright{\bf Steps:}
- \begin{algorithmic}
-   \STATE $i \leftarrow 0$ 
-   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-   without $\epsilon$-moves \cite{hopcraft90toc}
-
-   \REPEAT
-   \STATE $\nt'_{i+1} \leftarrow \nt_i$
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcar$ and
-     outgoing edge to $q''$ with label $\acar$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-   label $\epsilon$.
-   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-   \ENDFOR
-
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcdr$ and outgoing edge to
-     $q''$ with label $\acdr$} 
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-     label $\epsilon$.
-   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-   \ENDFOR
-
-   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-   $\nt'_{i+1}$ without $\epsilon$-moves
-
-   \STATE $i \leftarrow i+1$
-   \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-   \STATE $\nt \leftarrow \nt_i$
-%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
- \end{algorithmic}
+ %% \begin{algorithmic}
+%%    \COMMENT Fix the algorithm issue
+%%    %% \STATE $i \leftarrow 0$ 
+%%    %% \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
+%%    %% without $\epsilon$-moves \cite{hopcraft90toc}
+
+%%    %% \REPEAT
+%%    %% \STATE $\nt'_{i+1} \leftarrow \nt_i$
+%%    %% \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+%%    %%   edge from $q'$ with label $\bcar$ and
+%%    %%   outgoing edge to $q''$ with label $\acar$} 
+%%    %% \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+%%    %% label $\epsilon$.
+%%    %% \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
+%%    %% \ENDFOR
+
+%%    %% \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+%%    %%   edge from $q'$ with label $\bcdr$ and outgoing edge to
+%%    %%   $q''$ with label $\acdr$} 
+%%    %% \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+%%    %%   label $\epsilon$.
+%%    %% \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
+%%    %% \ENDFOR
+
+%%    %% \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
+%%    %% $\nt'_{i+1}$ without $\epsilon$-moves
+
+%%    %% \STATE $i \leftarrow i+1$
+%%    %% \UNTIL ($\nt_{i} = \nt_{i-1}$)
+
+%%    %% \STATE $\nt \leftarrow \nt_i$
+%% %   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
+%%  \end{algorithmic}
 \end{center}

### 2015-04-07 lazy-scheme/src/gc.cpp

@@ -1226,8 +1226,7 @@ void update_heap_ref_stack(ostream& out, int gc_type)
 		update_heap_refs.push(temp.top());
 		temp.pop();
 	}
-//cerr<<"Completed updated print & heap ref stacks"<<endl;
-//	assert(temp.empty());
+
 }
 
 //garbage collector functions

### 2015-04-26 paper/paper.tex
UPdated liveness analysis section

--- 2015-03-27 paper/paper.tex
@@
 \definecolor{Myblue}{rgb}{0.0,0.0,0.5}
-\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
-    #1 \medskip \hrule \medskip}}}
+\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
 \newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
@@
 A variable is {\em live} if  there is a possibility of its value being
-used  in future computation  and dead  if it  is definitely  not used.
+used in  future computations  and dead if  it is definitely  not used.
 Heap-allocated data needs a richer model of liveness which talks about
@@
 mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
-expressed as  a set, for example by  writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, 
-y.\epsilon\}$               instead               of               $[x
+expressed  as a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
+x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
   \mapsto\{\epsilon,\acdr,\acdr\acdr\},           y\mapsto\{\epsilon\},
@@
 $x$ is dead.  We associate  a liveness environment with program points
-$\pi$.  The liveness at $\pi:e$  is the liveness just before executing
+denoted by $\pi$ and $\psi$.  The liveness at $\pi:e$  is the liveness just before executing
 $e$.
@@
 the  rule  is obtained  by  converting  the  rule into  an  equivalent
-grammar.  The least solution also  ensures that the greatest amount of
+grammar (actually the language recognized by the grammar is the least solution).  The least solution also  ensures that the greatest amount of
 garbage is collected.
@@
 point  where  an  expression   gets  evaluated  cannot  be  statically
-determined.  To  deal with this, we introduce  flow insensitivity into
+determined.  To  deal with this, we introduce an analysis which is
+flow insensitive but path-sensitive\comment{Check the terminology} into
 our  analysis.
@@
             \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
-            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
             \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
             \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
+            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
+            $(\LET~\pz\ $\leftarrow$ (1~+~\py))))))) \IN
+            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$)))))
         \end{uprogram}}
@@
 	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
+          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
+          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
+          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
 \end{uprogram}}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-
 % Push this to a later point
@@
 
-Consider  an analysis of  the body  of $\pf$  with a  demand $\sigma$.
-Clearly, this is also the demand on the return values $u$ and $y$
-giving the liveness  at  the  program points  $\pi_7$  and $\pi_8$  as
-$\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The  liveness at
-$\pi_7$  and  $\pi_8$  are  propagated  backwards  as  in  traditional
+Consider  an analysis of  the body  of $\length$  with a  demand $\sigma$.
+Clearly, this is also the demand on the return value $z$ giving the
+liveness  at  the  program point  $\psi_3$ as $\{\pz.\sigma\}$.   
+The  liveness at $\psi_3$  is  propagated  backwards  as  in  traditional
 liveness analysis.
 
-Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
-\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
-turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
-$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
-not in the scope of \pu.  However, this does not cause any imprecision
-in the way of the GC marking  more cells as live, since \pu\ is yet to
-come into existence at $\pi_7$  and its liveness will not be consulted
+Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
+\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
+turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
+$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
+beyond the scope of the definition of \py.  However, this does not cause any imprecision
+in the way of the GC marking  more cells as live, since \py\ is yet to
+come into existence at $\pi_6$  and its liveness will not be consulted
 by the garbage collector at this point.
-In an eager language, \pw\ would not have been live after $\pi_7$, the
+
+In an eager language, \py\ would not have been live at $\psi_3$, the
 point of  its last  use. In a  lazy language, however,  the definition
-$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
-$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
-takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
-record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
-every program  point from $\pi_8$ up  to the beginning  of the program
+$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
+$(1+~\py)$; instead  a closure is created.   The evaluation actually
+takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
+record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
+every program  point from $\psi_3$ up  to the beginning  of the program
 (since  we do  not kill  liveness).  We  thus regard  the body  of the
 function as  a tree and consider paths  from the root of  this tree to
-program  points  which  trigger  a evaluation---\SIF\  conditions  and
-\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+program  points  which  trigger  a evaluation\footnote{Program points
+  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
+\RETURN\  expressions.  We  call such  paths {\em
+  e-paths}. \comment{Update the e-path labels}  For the
 program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
@@
 
-The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-path to a common liveness environment, and a second map $\cal{P}$ maps
+The analysis  consists of  two maps: The  map $\cal{M}$ maps  a e-path to a common liveness environment, and a second map $\cal{P}$ maps
 a program  point to the set of  e-paths on which  the point lies.
@@
 union of  the liveness environments  of the individual paths.   In the
-example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
-${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
+example program, ${\cal M}(\psi_1)= \ldots$, ${\cal M}(\psi_2)=\ldots$ and
+${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
 EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
@@
 Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
-evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
+The  expression $\psi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\psi  \mapsto
   \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
-expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
+expression $(\SIF~\psi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
   M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
-triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
-is created and added to  the union.  Also notice that the consequences
+triggers an evaluation of  $x$, the map $[\psi \mapsto \{x.\epsilon\}]$
+is created and added to  the union.  Also notice the consequences
 of  laziness---the entire  expression including  the condition  is not
@@
 \end{figure*}
-%======================================================================================
-In an eager language, \pw\ would not have been live after $\pi_7$, the
-point of  its last  use. In a  lazy language, however,  the definition
-$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
-$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
-takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
-record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
-every program  point from $\pi_8$ up  to the beginning  of the program
-(since  we do  not kill  liveness).  We  thus regard  the body  of the
-function as  a tree and consider paths  from the root of  this tree to
-program  points  which  trigger  a evaluation---\SIF\  conditions  and
-\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
-program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
-$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
- and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.
-
-The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-path to a common liveness environment, and a second map $\cal{P}$ maps
-a program  point to the set of  e-paths on which  the point lies.
-If a  program  point happens to fall  on more than one  such path, then
-the  liveness environment of  the program  point is  the variable-wise
-union of  the liveness environments  of the individual paths.   In the
-example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
-${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
-Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
-the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
-our propagation of liveness is flow-insensitive but path-sensitive.
-In the sequel
-we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
-$\cal{P}$ is easy and we do not elaborate it any further.
-
-Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
-evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
-  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
-expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
-  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
-triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
-is created and added to  the union.  Also notice that the consequences
-of  laziness---the entire  expression including  the condition  is not
-evaluated if  the demand  on it is  $\emptyset$.  This results  in the
-empty map denoted by $\emptyset_{\cal{M}}$.
-
- The  {\LET} rule  has  an elegant  formulation.   Recollect that  our
- language permits lazy \LET~s whose definitions can be recursive Thus,
+
+ The  liveness of {\LET} has  an elegant  formulation.   Recollect that  our
+ language permits lazy {\LET}s whose definitions can be recursive. Thus,
  in the expression $\LET~x  \leftarrow~s~\IN~e$, $x$ can occur in $s$.
@@
 
-%\comment{
-
-\section{Generating Liveness Equations}
-\begin{figure*}[t!]
-%%\begin{picture}(100,130)(0,-90)
-\begin{picture}(100,130)(0, -50)
-  \begin{tabular}{cc}
-    \begin{boxedminipage}{.51\textwidth}
-      {\sf
-        \small
-        \renewcommand{\arraystretch}{1}{
-          \begin{uprogram}
-            \UNL{1} (\DEFINE\ (\length~\xl)
-            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
-            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
-            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
-        \end{uprogram}}
-        \renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
-\end{uprogram}}
-        } 
-    \end{boxedminipage}
-    & 
-    \begin{boxedminipage}{.49\textwidth}
-      {\sf
-        \small
-        \begin{tabular}{lc}
-        $\cal{M_{\mbox{\tiny length}}} = $ 
-            \begin{tabular}{lc}            
-              $[$
-              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
-              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
-              $ep_1 \mapsto \lbrace \rbrace , $\\
-              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
-              $]$
-            \end{tabular} \\
-            $\cal{P_{\mbox{\tiny length}}} = [$
-              \begin{tabular}{lc}
-                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
-                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
-                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
-                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
-                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
-                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
-                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
-                $]$
-              \end{tabular}\\
-              $\cal{M_{\mbox{\tiny main}}} = $ 
-              \begin{tabular}{lc}            
-                $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
-                  $]$
-              \end{tabular} \\
-              $\cal{P_{\mbox{\tiny main}}} = [$
-                \begin{tabular}{lc}
-                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
-                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \ldots  $ \
-                  $]$  
-                \end{tabular}
-        \end{tabular}
-      }
-    \end{boxedminipage}
-    \\
-    (a) Example program.&
-    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
-        (b) {\em e-path} to liveness environment map \\
-        (c) Prog pt to {\em e-paths } map
-    \end{tabular}}
-  \end{tabular}
-\end{picture}
-\kern -3ex
-\vspace{0.5in}
-\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
-\end{figure*}
-
- For our example program which has  a
-single function \length, this generates the following equations:
-
 Section~\ref{sec:liveness} gave a context-sensitive liveness
@@
 rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
-$\emptyset$, if $\sigma$ itself $\emptyset$. This creates difficulties
+$\emptyset$, if $\sigma$ itself $\emptyset$.  To realize these rules
+$\sigma$ needs to be known.This creates difficulties
 since  we  want  to  solve  the  equations  arising  out  of  liveness
-symbolically, and realizing these rules requires $\sigma$ to be known.
+symbolically.
 
@@
 beginning of  $e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
-equations  of the  form $\Lf{\mathit{f}}{i}{\sigma}  = \Lanv{1}{x_i}$,
-where  $x_i$   is  the  $i^{\mbox{\footnotesize   th}}$  parameter  of
-$\mathit{f}$. Doing this for  \length, we get:
+equations     of     the     form    $\Lf{\mathit{f}}{i}{\sigma}     =
+\Lanv{\pi_1}{x_i}$, where  $x_i$ is the  $i^{\mbox{\footnotesize th}}$
+parameter of $\mathit{f}$. Doing this for \length, we get:
 \begin{align*}
- \Lanv{1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
+ \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
 \end{align*}
@@
 In general, the equations for  \Lfonly\ are recursive since, as in the
-case  of \length,  \Lan{1}\ may,  in turn,  be expressed  in  terms of
+case of \length,  \Lan{\pi_1}\ may, in turn, be  expressed in terms of
 \Lfone{\mathit{f}}.
@@
   \end{align*}
-  Note that these equations can also be viewed as CFGs.
+
+Note that this equation can also be viewed as CFGs.
 
@@
 {\color {red} What about print?}
+
 For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
@@
 \sigma_{\length}    &=
- \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{length}}}
+ \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
 \end{align*}
 
-For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $ep_1, ep_2, ep_3$ in $\length$ and
-$ep_4$ in $e_\main$, this gives
-
+For some of the program points in \length\  and \main,  this gives
+{\color {red} Prasanna, please fill this correctly}. You could use the
+set notation if the map notation spills over: 
 \begin{align*}
-\Lanv{7}{\xl} &= \acdr\Lf{length}{1}{\bX\sigma_{\mathit{length}}} \cup \bX\sigma_{\mathit{length}} \\
-\Lanv{7}{\pu} &= \Lf{length}{1}{\bX\sigma_{\mathit{\length}}} \\
-\Lanv{7}{\py} &= \bX\sigma_{\mathit{length}} \\
-\Lanv{3}{\xl} &= \bX\sigma_{\mathit{length}} \\
-\Lanv{3}{\px} &= \bX\sigma_{\mathit{length}}
-\\[1ex]
-  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
-  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
+\Lan{\pi_1} &= \Lan{\pi_2} = \ldots \\
+\Lan{\pi_3} &= \ldots \\
+\Lan{\pi_4} &= \ldots \\
+\Lan{\pi_5} &= \Lan{\pi_6} = \Lan{\pi_7} \ldots \\
+\Lan{\pi_8} &= \Lan{\pi_9} = \Lan{\pi_{10}} \ldots
 \end{align*}
-
 \subsection{Solving liveness equations---the grammar interpretation}
@@
 context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
+\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal for
 a variable  $X$ occurring on  the LHS of the  equations generated
@@
 \section{Implementation}
-% GC assumptions
 By the  time the program  starts executing, the liveness  automata are
@@
 % That's all folks!
+
 \end{enumerate}

### 2015-04-26 paper/paper.tex
Added comments for fixing liveness equation part

@@
   &\sigma_1\bcar\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \bcar\sigma_2',\\
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
 &\sigma_1\bcdr\sigma_2  \hookrightarrow 
   \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
-  \bcdr\sigma_2',~\mbox{and}\\  
+  \acdr\sigma_2',~\mbox{and}\\  
 &\sigma_1\cbar \emptyset \hookrightarrow \emptyset
@@
 interest  is the liveness  at the  program point  at the  beginning of
-beginning of  $e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
+$e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
 equations     of     the     form    $\Lf{\mathit{f}}{i}{\sigma}     =
@@
 
-As mentioned earlier, each of  the liveness rules modifies a demand by
-prefixing   with   symbols   in   the   above   set,   and   therefore
+As mentioned earlier\comment{Check if already mentioned}, each of  the liveness rules modifies a demand by
+prefixing   it with   symbols   in   the   above   set,   and   therefore
 $\Lf{f}{i}{\sigma}$ has the closed form:
@@
 the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
-$\sigma$,  we  get the  following  equation  for  \Df{f}{i}: that  are
+$\sigma$,  we  get an  equation  for  \Df{f}{i} that is
 independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
@@
 
-Note that this equation can also be viewed as CFGs.
+Note that this equation can also be viewed as a CFG with \{\acdr,
+\cbar\} as the terminal symbols and \Df{\length}{1} as the sole non-terminal.
 
@@
 achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
-\cup \deltacall{f}{\pi^k}{g^k}$.
+\cup \deltacall{f}{\pi^k}{g^k}$\comment{Rewrite, copied from earlier paper}.
 
@@
 \end{align*}
+
+In summary, the equations generated during liveness analysis are:
+\begin{enumerate}
+\item   For    each   function   $\mathit{f}$,    equations   defining
+\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
+\item  For  each   function  $\mathit{f}$, an  equation  defining
+the summary demand  $\sigma_{\mathit f}$ on $e_f$.
+% \marginpar{Consistency: summary vs cumulative?}
+\item For each function $\mathit{f}$ (including $\mainpgm$ for
+  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages?}
+an equation defining
+  liveness at each GC point of $e_{\mathit f}$.
+
 \subsection{Solving liveness equations---the grammar interpretation}
@@
 \end{eqnarray*}
-
-In summary, the equations generated during liveness analysis are:
-\begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
-\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
-
-\subsection{Solving liveness equations---the grammar interpretation}
-The  equations above can now be re-interpreted as a
-context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
-a variable  $X$ occurring on  the LHS of the  equations generated
-from the analysis.  We can  think of the resulting productions as
-% being a grammar with multiple start symbols---one for
-% for each nonterminal \var{\Lanv{i}{x}}.
-being associated with several  grammars, one for each non-terminal
-\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
-productions:
- \begin{eqnarray*}
-\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
-  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
- \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
-  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
-  \acdr\var{\Uf{\append}{1}}\\         
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar               \mid
-  \acdr\var{\Df{\append}{1}}\bcdr\\
-\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
-\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
-\var{$\sigma_{\mathit{\!all}}$}
-\end{eqnarray*}
+\comment{Pick up a richer equation or more than one liveness equations}
 Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
@@
 in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is  undecidable.   We
-circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
+membership  problem   for  such  grammars   is
+undecidable. \comment{Rewrite it to say that this is a CFG with 3
+  unrestricted rules and the membership for even this grammar is undecidable}
+\comment{State the undecidability of the decision problem above.}
+
+  We circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
   strongly  regular}  CFGs which  have   easy
@@
 converted to DFAs for use during garbage collection.
+\comment{Fix all comments and produce first draft}
 %------------------------------------------------------------%

### 2015-04-27 paper/paper.tex
To dos for Prasanna

@@
-\documentclass[10pt]{sigplanconf}
+\documentclass[9pt]{sigplanconf}
 \usepackage{etex}
@@
 \label{sec:NFA-approx}
-
+\comment{Change the language and the example}
 We use the   
@@
 Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
-\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
+\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow}
+\lang{\nt}$, where {\nt} accepts forward paths only.
+\comment{Rewrite the algorithm to handle \cbar and update the example automata}
 Thus  \nt\  can  be  used  to  check  
@@
 The proofs  of the termination  and correctness of  the algorithm
-are given in our earlier paper~\cite{karkare07liveness}.
+are given in an earlier paper~\cite{karkare07liveness}.
 
 \input{liveness-dfa.tex}
-
+\comment{change according to example}
 We  illustrate  the  algorithm  in  Figure~\ref{algo:simplify-NFA}  by
@@
 \LET~  statement in the  program. This  leads to  an explosion  in the
-number of  automata. From our  observation that along an  {\em E-path}
+number of  automata. \comment{Directly say that we share the automata}From our  observation that along an  {\em E-path}
 all program points have the same liveness, we can reduce the number of
@@
 % Liveness for cons cells & reachability for closures
+
+\comment{Mention the fact that liveness at a program point indicates
+  what references will be used by the program eventually. The exact
+  point where a reference will be used cannot be determined statically
+due to the language being lazy. }
 During GC, if the heap cell being processed is a cons cell, we consult
@@
 presence  of  a print  stack.  In  lazy  languages the  evaluation  of
-closures is  driven by the print  function.  Hence, during  a GC there
+closures is  driven by the print  function\comment{Refer to SPJ's book}.  Hence, during  a GC there
 will be heap  references that will be present on  the stack which also

### 2015-04-27 paper/paper.tex
Added Results and Conclusion section from previous paper

@@
 
-%% Finally, we use $\Lfonly$  to range over {\em demand transformers}.
-%% These transform  demands on  a function call  into demands  for its
-%% formal  parameters:   given  a   user  function  $f$,   defined  by
-%% $(\DEFINE\ (f\  x_1\ \ldots\ x_n)\  \ e_f)$ and called  with demand
-%% $\sigma$,  then $\Lf{f}{i}{\sigma}$  is  the liveness  of $x_i$  at
-%% $e_f$.
-
 \subsection{Liveness Analysis for lazy languages}
@@
 
-%%   Analogously to classical liveness being computed as a solution of
-%% dataflow   equations,  we   require,  via   inference   rule  ({\sc
-%% live-define}), $\Lfonly$ to  satisfy the fixed-point property that:
-%% when we  assume $\Lfonly$ to  be the family of  demand transformers
-%% for the program then the  calculated liveness of each function body
-%% $\mathcal{L}(e_f,\sigma,\Lfonly)$    agrees   with    the   assumed
-%% $\Lfonly$.   As  usual,  there  are  often  multiple  solutions  to
-%% $\Lfonly$; all are  safe (see Section~\ref{sec:correctness}) but we
-%% prefer  the least  one  as  giving the  least  liveness subject  to
-%% safety---and hence greatest amount of garbage collected.
-
 We next  describe the function $\mathcal{L}$  that propagates liveness
@@
 determined.  To  deal with this, we introduce an analysis which is
-flow insensitive but path-sensitive\comment{Check the terminology} into
+flow-insensitive but path-sensitive\comment{Check the terminology} into
 our  analysis.
@@
 \begin{figure*}[t!]
-%%\begin{picture}(100,130)(0,-90)
 \begin{picture}(100,130)(0, -50)
@@
         \begin{tabular}{lc}
-        $\cal{M_{\mbox{\tiny length}}} = $ 
+        $\cal{M_{\mbox{\tiny \length}}} = $ 
             \begin{tabular}{lc}            
@@
             \end{tabular} \\
-            $\cal{P_{\mbox{\tiny length}}} = [$
+            $\cal{P_{\mbox{\tiny \length}}} = [$
               \begin{tabular}{lc}
@@
               \end{tabular}\\
-              $\cal{M_{\mbox{\tiny main}}} = $ 
+              $\cal{M_{\mbox{\tiny \main}}} = $ 
               \begin{tabular}{lc}            
@@
               \end{tabular} \\
-              $\cal{P_{\mbox{\tiny main}}} = [$
+              $\cal{P_{\mbox{\tiny \main}}} = [$
                 \begin{tabular}{lc}
@@
 \caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
-%% >>>>>>> 67d5a8ebe45703d9726af91b752e9d74152ae455
 \end{figure*}
@@
 arguments.
+
 \section{Computing liveness and its encoding as a table}\label{sec:computing}
-
 Section~\ref{sec:liveness} gave a context-sensitive liveness
@@
 \section{Results}
+To demonstrate the effectiveness of liveness-based garbage collection,
+we  have  built a  prototype  consisting  of  an interpreter  for  our
+language,  a  liveness  analyzer  and  a copying  collector  that  can
+optionally use the results of liveness analysis for marking instead of
+reachability. When the collector uses liveness for marking, we call it
+a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
+  reachability-based  collector}  (RGC).   The  copying  collector  is
+neither incremental nor generational.  As a consequence, any cell that
+becomes unreachable or  dead is assuredly collected in  the next round
+of garbage collection.
+
+While exploring the activation record of a function, LGC uses the DFAs
+at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
+is invoked  at the program point   $\pi$,  it would  use  the
+liveness  DFAs at  $\pi$ itself for exploring the root set of the
+current activation record. For any other activation record in the
+stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
+the DFAs at the point of call to  ${\mathit g}$ would be used.
+
+Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
+point  pair ($x$,  $\pi$).  We  assume that  there is  a function
+$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
+$\dfa{\pi}{x}$.    
+Considering   a   DFA   as   a   table,
+$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
+state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
+\acar\      or     \acdr.       We      shall     also      write
+$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
+indicating  whether there  is a  transition from  $q$ on
+$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
+$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
+empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
+$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
+in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
+pointer to $x$.
+The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
+$x$  in the  other semi-space  and returns  the new  address.  It
+hides  details such as  returning the  forwarding pointer  if the
+value  of $x$  is  already copied,  and  creating the  forwarding
+pointer otherwise.
+
+The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
+in the heap over time for RGC and LGC---here time is measured in terms
+of the number of cons cells allocated. In addition, they also show the
+number of  reachable cells and the  number of cells  that are actually
+live\footnote{This   is  statically   approximated  by   our  liveness
+  analysis}.  Since  the programs have  different memory requirements,
+we have tuned the size of heap for each program to ensure a reasonable
+number of  collections. An invocation  of RGC decreases the  number of
+cells  in heap  till  it touches  the  curve of  reachable cells.   An
+invocation of LGC decreases the number  of heap cells to no lower than
+the curve of live cells.
+
+To construct the  reachable and live curves, we  record for every cell
+its creation time (\CreateTime), its last use time (\UseTime), and the
+earliest time  when the  cell becomes unreachable  and can  be garbage
+collected (\CollTime).  For accurate recording of  \CollTime, we force
+frequent invocations  of a reachability based collector  in a separate
+run.     A    cell   is    live    at    time   \mbox{\exectime}    if
+\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
+  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
+  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
+
+The benchmark programs are drawn from the {\tt no-fib} suite and other
+sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
+collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
+  fibheap} increases  initially, comes  down before the  heap becomes gets
+full and then  continues to remains low.  This results in  a single garbage
+collection in both RGC and LGC.  The graphs also show the precision of
+our liveness analysis.  For  all programs  except  {\tt nperms} and
+{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
+manages to collect a good portion of the data that are not live.
+
+\subsection{Results}
+\begin{figure}[t] 
+\centering
+\begin{boxedminipage}{.75\textwidth}
+{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
+\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
+\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
+  then} {\bf skip}\\ 
+\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
+\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
+\noindent \hspace*{.35cm}{\bf return} $y$
+\end{boxedminipage}
+\caption{Function       for       copying       a       root       set
+  variable.}\label{algo:LGC-main-loop} \figrule
+
+\end{figure}
+
+\begin{figure}[p]
+\begin{tabular}{@{}c@{}c@{}}
+\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
+&
+{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
+\\
+\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
+ &
+{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
+\\
+\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
+&
+{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
+\\
+\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
+&
+{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
+\end{tabular}
+ \caption{Memory  usage of  programs. The blue and the red curves
+  indicate the number of cons cells in the active semi-space for RGC and
+  LGC respectively. The black curve represents the number of reachable
+  cells and the grey curve represents the number of cells that are
+  actually live (of which liveness analysis does  a static
+  approximation). x-axis is the time measured in number of cons-cells
+  allocated. y-axis is the number of cons-cells.}
+\label{fig:memory-usage} \figrule
+\end{figure}
+The  increased effectiveness  of LGC  over RGC  is also  shown  in the
+tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
+provides  statistics  regarding the  analysis  itself.  The number  of
+states and the analysis times  are within tolerable limits. Precision
+of analysis refers to the percentage of dead cells that is collected by
+LGC, averaged over all invocations.
+The second table shows garbage  collection statistics for RGC and LGC.
+LGC  collects larger garbage  per invocation,  drags cells  for lesser
+time and requires a smaller heap size ({\em MinHeap}) for program to run
+in comparison with  RGC.
+
+There are a couple of  issues of concern.  The garbage collection time
+is larger in the case of LGC for some programs. The reason is that the
+cost of consulting the liveness DFA may outweigh the combined benefits
+of  fewer   garbage  collections   and  fewer  markings   per  garbage
+collection.   The  other issue  is  illustrated  by  the program  {\tt
+  lambda}    (memory   usage  shown   separately    in
+Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
+in Figure~\ref{fig:experimental-results}, the  number of
+touched cells\footnote{These  are the  cells which are  visited during
+  marking phase, often  more than once due to  sharing.} in this
+example is much higher for
+LGC.  This  increase is due to  excessive sharing among  heap nodes in
+this program.  Note  that a node re-visited because  of sharing is not
+explored  any   further  during  a  RGC   collection.   However,  this
+curtailment cannot happen in  LGC because of the possibility that the
+node,  re-visited in a  different liveness  state, may  mark a  set of
+cells different from the earlier visit.
+\begin{figure}[t]
+\footnotesize
+
+\centering
+
+\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+\hline
+{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
+  nperm} &  {\sf fibheap} & {\sf knightstour} &
+{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
+\hline
+\hline
+{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
+{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
+{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
+\hline
+\end{tabular}}
+
+\centerline{(a)}
+
+\medskip
+
+\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
+\hline
+    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
+  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
+\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
+
+                            &   \multicolumn{2}{c|}{cells per GC}  
+                            &   \multicolumn{2}{c|}{cells per GC} 
+                            &   \multicolumn{2}{c|}{\#GCs} 
+                            &   \multicolumn{2}{c|}{MinHeap} 
+                            &   \multicolumn{2}{c|}{Avg. Drag} 
+                            &   \multicolumn{2}{c|}{(sec)} \\
+\cline{2-13}
+{Program}    &
+RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
+\hline
+\hline
+    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
+    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
+    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
+     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
+    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
+    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
+    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
+    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
+    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
+%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
+%    {\sf  nboyer} & &  & &   & &  & & &  \\
+%    {\sf   circsim?} & &  & &   & & & & &   \\
+ \hline
+\end{tabular}}
+
+\centerline{(b)}
+\vskip -2mm
+\caption{Experimental results comparing RGC and LGC. Table (a) gives
+  data related to liveness analysis, and (b) gives garbage collection
+  data. The average drag time is measured in allocated cons cells.}
+\label{fig:experimental-results}
+\normalsize
+\end{figure}
+
+\section{Collecting more garbage can never slow things down}
+\label{sec:lgc-always-better}
+Since garbage collection is  effectively asynchronous to the allocator
+thread, one might  worry as to how {\em  robust} our measurements are.
+For example,  while LGC would,  in general, collect more  garbage than
+RGC in the same heap state,  is it possible that on some programs, LGC
+might do  a larger number  of collections? We  prove a result  to show
+that   this  cannot   happen.    This  result   applies  to   classical
+mark-and-sweep and copying garbage  collectors and, we hypothesize, to
+generational collectors too.
+
+\begin{lemma}
+For the  same mutator,  a liveness-based collector  can never  do more
+garbage collections than a reachability-based collector.
+\end{lemma}
+
+\begin{proof}Assume, as before,  that time is measured in terms of the number of
+  cons cells allocated. Let us run  two copies of the mutator with the
+  two    different   collectors    but    with   synchronous    memory
+  allocations.   The   garbage    collections,   however,   happen
+  asynchronously.
+
+  To prove the lemma, it is enough to show the truth of the following 
+statement: After
+ every  LGC invocation,  the count of LGC invocations   is  no greater
+than  RGC invocations.   The  base case  holds  since the  first
+invocations of both GCs happen at the same time.  Assume the statement
+to be true after $n$ invocations of LGC.  Since LGC copies a subset of
+reachable cells, its heap would contain no more cells than RGC heap at
+the end  of the $n$th invocation.   Thus either RGC  is invoked next before
+LGC, or LGC and RGC are both invoked next at the same time.  In either
+case, the statement holds after $n+1$ invocations of LGC.
+\end{proof}
+
+%% \begin{figure}[h]
+%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
+%% \caption{Memory  usage of  programs. The  solid  lines represents
+%%   number of  live cells and the  dashed lines show  the number of
+%%   reachable cells.  The red and blue  lines represent invocations
+%%   of RGC and  LGC---their heights indicate  the amount of memory
+%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
+%%       paper and are being used as placeholder}}}
+%% \label{fig:memory-usage} \figrule
+%% \end{figure}
+%%\clearpage
+
+\section{Related Work} Previous attempts to  increase  the space efficiency
+of functional programs by additional  reclamation of memory fall in two
+broad categories. In the first,  the program itself is instrumented to
+manage  reclamation and reallocation  without the  aid of  the garbage
+collector.    Such    attempts   include:   sharing    analysis   based
+reallocation~\cite{jones89compile},                       deforestation
+techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
+methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
+analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
+methods   that  enable   the   garbage  collector   to  collect   more
+garbage~\cite{inoue88analysis,lee05static}   by
+explicitly  nullifying  pointers  that  are not  live.   However,  the
+nullification,  done   at  compile  time,   requires  sharing  (alias)
+analysis.  Our  method, in contrast, does not  require sharing because
+of the availability of the heap  itself at runtime.  To the best of our
+knowledge, this is the first  attempt at liveness-based marking of the
+heap during garbage collection.
+
+\begin{figure}[t]
+\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
+ \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
+\label{fig:memory-usage-lambda} \figrule
+\end{figure}
+
+\section{Conclusions}
+\label{sec:conclusion}
+We have defined a notion of liveness on structured data;
+this generalizes classical liveness and strong liveness.
+We started with a general fully context-sensitive analysis which
+we proved correct with respect to a minefield semantics
+(this models the effect of garbage collection between every
+evaluation step).
+
+To avoid scalability issues (and to avoid performing part of
+the liveness computation at run time)
+we defined an 0-CFA version of this liveness analysis in which
+demands for function $f$ at all calling contexts are conflated into 
+a single demand $\sigma_f$.
+This enabled us to treat the liveness equations symbolically obtaining
+context-free grammars for liveness at each GC point (calls to user
+functions and to $\CONS$).
+These were then converted to DFAs for run-time consultation by the
+garbage collector. Experiments confirm the precision of the analysis.
+
+To obtain performance figures we compared a reachability-based garbage
+collector with a liveness-based  collector.  This showed a decrease in
+the  number   of  GCs,  more  garbage  collected   per  invocation.  A
+significant benefit of LGC is that programs can run in smaller memory when
+compared to RGC. This is potentially useful in situations where memory
+is limited---as with embedded systems. For a majority of programs, the
+garbage collection times were reduced.
+
+One issue we  highlighted was that while fewer  nodes were marked (and
+more garbage collected), sometimes  cons cells could be visited and
+traversed  multiple times  with different  sets of  liveness  paths to
+explore.  Further work includes  improvements to the classical copying
+collector to reduce the cost of this.
+
 %Add graphs for memort usage LGC vs RGC
 %Add graphs/table showing the GC times and other statistics
-\section{Conclusion and Future Work}
+
 %Mention lower peak memory usage. 

### 2015-04-28 paper/paper.tex

@@
 by the garbage collector at this point.
-
 In an eager language, \py\ would not have been live at $\psi_3$, the

### 2015-04-28 paper/paper.tex
Added latest graphs

@@
 \label{sec:lgc-always-better}
-Since garbage collection is  effectively asynchronous to the allocator
-thread, one might  worry as to how {\em  robust} our measurements are.
-For example,  while LGC would,  in general, collect more  garbage than
-RGC in the same heap state,  is it possible that on some programs, LGC
-might do  a larger number  of collections? We  prove a result  to show
-that   this  cannot   happen.    This  result   applies  to   classical
-mark-and-sweep and copying garbage  collectors and, we hypothesize, to
-generational collectors too.
-
-\begin{lemma}
-For the  same mutator,  a liveness-based collector  can never  do more
-garbage collections than a reachability-based collector.
-\end{lemma}
-
-\begin{proof}Assume, as before,  that time is measured in terms of the number of
-  cons cells allocated. Let us run  two copies of the mutator with the
-  two    different   collectors    but    with   synchronous    memory
-  allocations.   The   garbage    collections,   however,   happen
-  asynchronously.
-
-  To prove the lemma, it is enough to show the truth of the following 
-statement: After
- every  LGC invocation,  the count of LGC invocations   is  no greater
-than  RGC invocations.   The  base case  holds  since the  first
-invocations of both GCs happen at the same time.  Assume the statement
-to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-reachable cells, its heap would contain no more cells than RGC heap at
-the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-LGC, or LGC and RGC are both invoked next at the same time.  In either
-case, the statement holds after $n+1$ invocations of LGC.
-\end{proof}
+
+%% Since garbage collection is  effectively asynchronous to the allocator
+%% thread, one might  worry as to how {\em  robust} our measurements are.
+%% For example,  while LGC would,  in general, collect more  garbage than
+%% RGC in the same heap state,  is it possible that on some programs, LGC
+%% might do  a larger number  of collections? We  prove a result  to show
+%% that   this  cannot   happen.    This  result   applies  to   classical
+%% mark-and-sweep and copying garbage  collectors and, we hypothesize, to
+%% generational collectors too.
+
+%% \begin{lemma}
+%% For the  same mutator,  a liveness-based collector  can never  do more
+%% garbage collections than a reachability-based collector.
+%% \end{lemma}
+
+%% \begin{proof}Assume, as before,  that time is measured in terms of the number of
+%%   cons cells allocated. Let us run  two copies of the mutator with the
+%%   two    different   collectors    but    with   synchronous    memory
+%%   allocations.   The   garbage    collections,   however,   happen
+%%   asynchronously.
+
+%%   To prove the lemma, it is enough to show the truth of the following 
+%% statement: After
+%%  every  LGC invocation,  the count of LGC invocations   is  no greater
+%% than  RGC invocations.   The  base case  holds  since the  first
+%% invocations of both GCs happen at the same time.  Assume the statement
+%% to be true after $n$ invocations of LGC.  Since LGC copies a subset of
+%% reachable cells, its heap would contain no more cells than RGC heap at
+%% the end  of the $n$th invocation.   Thus either RGC  is invoked next before
+%% LGC, or LGC and RGC are both invoked next at the same time.  In either
+%% case, the statement holds after $n+1$ invocations of LGC.
+%% \end{proof}
 

### 2015-04-28 paper/paper.tex
Updated paper as per discussion

@@
 determined.  To  deal with this, we introduce an analysis which is
-flow-insensitive but path-sensitive\comment{Check the terminology} into
+flow-insensitive but path-sensitive\comment{Checked the terminology, path-sensitive means certain prooperties holding along a path only, http://en.wikipedia.org/wiki/Data-flow\_analysis} into
 our  analysis.
@@
             \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
-            $(\LET~\pz\ $\leftarrow$ (1~+~\py))))))) \IN
-            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$)))))
+            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
+            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
         \end{uprogram}}
@@
             \begin{tabular}{lc}            
-              $[$
-              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
-              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
-              $ep_1 \mapsto \lbrace \rbrace , $\\
-              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $[$ \\
+              $\psi_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,$ \\ 
+              $\psi_2 \mapsto \lbrace \rbrace , $\\
+              $\psi_3 \mapsto \lbrace \pz.\sigma_{\length}, \py.\cbar\sigma_{\length}, $\\
+              $\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace $  \\
               $]$
             \end{tabular} \\
-            $\cal{P_{\mbox{\tiny \length}}} = [$
+            $\cal{P_{\mbox{\tiny \length}}} = $
               \begin{tabular}{lc}
-                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
-                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
-                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
-                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
-                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
-                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
-                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
+                $[$ \\
+                $\pi_{\mbox{1}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \
+                $ \pi_{\mbox{2}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \\
+                $ \psi_{\mbox{1}} \mapsto \lbrace \psi_1 \rbrace, $ \
+                $ \psi_{\mbox{2}} \mapsto \lbrace \psi_2 \rbrace, $ \
+                $ \pi_{\mbox{5}} \mapsto \lbrace \psi_3 \rbrace, $ \\
+                $ \pi_{\mbox{6}} \mapsto \lbrace \psi_3 \rbrace, $ \
+                $ \pi_{\mbox{7}} \mapsto \lbrace \psi_3 \rbrace $ \\
+                $ \psi_{\mbox{3}} \mapsto \lbrace \psi_3 \rbrace $ \\
                 $]$
@@
                 $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $\psi_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                   $]$
               \end{tabular} \\
-              $\cal{P_{\mbox{\tiny \main}}} = [$
+              $\cal{P_{\mbox{\tiny \main}}} = $
                 \begin{tabular}{lc}
-                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
-                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $[$ \
+                  $\pi_{\mbox{8}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \pi_{\mbox{9}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \pi_{\mbox{10}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \psi_{\mbox{4}} \mapsto \lbrace \psi_4 \rbrace, $ \
                   $ \ldots  $ \
@@
   e-paths}. \comment{Update the e-path labels}  For the
-program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
-\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
-\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+function \length, there are three such paths $\psi_1 = \pi_1, \pi_2, \pi_3,
+\pi_4,  \pi_5$, $\psi_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
+\pi_8$, and $\psi_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
 
@@
 example program, ${\cal M}(\psi_1)= \ldots$, ${\cal M}(\psi_2)=\ldots$ and
-${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
-EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
+${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace \psi_1,
+\psi_2, \psi_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
 our propagation of liveness is flow-insensitive but path-sensitive.
@@
 \begin{align*}
- \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
+ \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
 \end{align*}
@@
     \Lf{\length}{1}{\sigma}
-    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
+    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
    \end{eqnarray*}
@@
 
-As mentioned earlier\comment{Check if already mentioned}, each of  the liveness rules modifies a demand by
+As mentioned earlier\comment{Check if already mentioned, Not mentioned earlier}, each of  the liveness rules modifies a demand by
 prefixing   it with   symbols   in   the   above   set,   and   therefore
@@
     \Df{\length}{1} &= \lbrace\cbar\rbrace 
-    \cup \acdr\Df{\length}{1}
+    \cup \acdr\Df{\length}{1}\cbar
   \end{align*}
@@
 {\color {red} What about print?}
-For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
+For the running example, $\length$ has calls from $\main$ at $\pi_{10}$
 and a recursive call at $\pi_6$.
@@
 \sigma_{\length}    &=
- \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
+ \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\cbar\sigma_{\mathit{\length}}}
 \end{align*}
@@
 \begin{align*}
-\Lan{\pi_1} &= \Lan{\pi_2} = \ldots \\
-\Lan{\pi_3} &= \ldots \\
-\Lan{\pi_4} &= \ldots \\
-\Lan{\pi_5} &= \Lan{\pi_6} = \Lan{\pi_7} \ldots \\
-\Lan{\pi_8} &= \Lan{\pi_9} = \Lan{\pi_{10}} \ldots
+\Lan{\pi_1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
+              \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
+\Lan{\pi_2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
+                  \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
+\Lan{\pi_5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
+\Lan{\pi_8} &= \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace 
 \end{align*}
@@
 \item For each function $\mathit{f}$ (including $\mainpgm$ for
-  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages?}
+  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages? No, we define it only in the implementation section as of now.}
 an equation defining
@@
 \var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
+grammar    for   \var{\Lanv{5}{\xl}}   comprises    the   following
 productions:
  \begin{eqnarray*}
-\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
-  {\length}{1}{\epsilon}} \mid \epsilon \\
+\var{\Lanv{5}{\xl}}  &\rightarrow&   \var{\acdr \Lf
+  {\length}{1}{\cbar\sigma_{\length}}} \mid \cbar\sigma_{\length} \\
 \end{eqnarray*}

### 2015-05-07 paper/paper.tex
Updated paper with subsection on handling 2bar

@@
 \usepackage{tabularx}
-\usepackage {algorithmic}
+\usepackage [plain,noend,noline,boxed]{algorithm2e}
 \usepackage{amssymb}
@@
 
+To avoid analysing the body of a function we assumed a symbolic demand $\sigma$
+and created a summary liveness for the function arguments and the variables 
+in the function. To get the actual liveness we have to substitute the actual 
+demands that are placed on the function. We calculate an over-approximation 
+of the actual demand, a {\em summary demand} and calculate the liveness at each GC point inside the 
+function. Even though we lose a bit of precision due to this, our 
+experiments show that this is acceptable \comment{Is this claim true?}.
+The {\em summary demand} is calculated by taking a union of the demand at 
+every call site of a function. A function call in our language can only be of the
+form $\pi\!\!:\!(\LET~x  \leftarrow (f\,y_1\,\ldots\,y_n)\ldots$. The demand on 
+$f$ due to this function call will be the liveness of $x$ at the program 
+point $\pi$. Denoting this demand as $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
+site $\pi$  in function $g$) we can write the {\em summary demand} as 
+$\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
+\cup \deltacall{f}{\pi^k}{g^k}$.
+
+\comment{
 We now calculate a 0-CFA-style summary liveness for each GC point of a
@@
 achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
-\cup \deltacall{f}{\pi^k}{g^k}$\comment{Rewrite, copied from earlier paper}.
+\cup \deltacall{f}{\pi^k}{g^k}$Rewrite, copied from earlier paper}.
 
@@
 
+\subsubsection{Removing the $\cbar$ edges}
+Once the NFA's are created, we need to remove the transitions on 
+$\cbar$. This edge is replaced by an $\epsilon$ transition and the 
+$\cbar$ edge is deleted. If the original grammar had a non-null 
+$\sigma$ then the NFA reaches a final state throught this path 
+otherwise it does not reach a final state.\comment{Need to rewrite this}
 %------------------------------------------------------------%
@@
 algorithm is as described below,
-%% \begin{algorithmic}
-%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
-%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
-%% \STATE $new\_cdr\gets$ copy(cell.cdr)
-%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
-%% \ENDIF
-%% \end{algorithmic}
+
+%% \begin{figure}
+%% \begin{algorithm}[H]
+%%  \KwData{this text}
+%%  \KwResult{how to write algorithm with \LaTeX2e }
+%%  initialization\;
+%%  \While{not at end of this document}{
+%%   read current\;
+%%   \eIf{understand}{
+%%    go to next section\;
+%%    current section becomes this one\;
+%%    }{
+%%    go back to the beginning of current section\;
+%%   }
+%%  }
+%%  \caption{How to write algorithms}
+%% \end{algorithm}
+%% \label{algo:LGC-Algo} \figrule
+%% \end{figure}
 We   provide   an   intuition    about   the   correctness   of   this
-algorithm. Consider the following facts about the print stack during a
+algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts about the print stack during a
 GC,

### 2015-05-07 paper/paper.tex
changed liveness

--- 2015-04-28 paper/paper.tex
@@
 $\Lfonly$  that  represents the  summaries  of  all  functions in  the
-program.  $\Lfonly_f$  (the  component  of $\Lfonly$  for  a  specific
-function $f$),  expresses how the demand  $\sigma$ on the  result of a
-call to $f$ is transformed into  the liveness of its parameters at the
-beginning of the call.  $\Lfonly$  is determined by the judgement form
-$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
-This  rule  describes the  fixed-point  property  to  be satisfied  by
-$\Lfonly$, namely, the demand transformation assumed for each function
-in  the  program should  be  the  same  as the  demand  transformation
-calculated    from   their    bodies.     As   we    shall   see    in
-Section~\ref{sec:grammar-formulation},  the least  solution satisfying
-the  rule  is obtained  by  converting  the  rule into  an  equivalent
-grammar (actually the language recognized by the grammar is the least solution).  The least solution also  ensures that the greatest amount of
-garbage is collected.
+program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
+specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
+result  of a  call to  $f$  is transformed  into the  liveness of  its
+parameters at the  beginning of the call.  $\Lfonly$  is determined by
+the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
+({\sc live-define}).  This rule  describes the fixed-point property to
+be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
+for each  function in  the program  should be the  same as  the demand
+transformation  calculated from  their  bodies.  As  we  shall see  in
+Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
+grammar and  the least  solution satisfying the  rule is  the language
+generated by this  grammar.  The least solution also  ensures the safe
+collection of the greatest amount of garbage.
 
@@
 point  where  an  expression   gets  evaluated  cannot  be  statically
-determined.  To  deal with this, we introduce an analysis which is
-flow-insensitive but path-sensitive\comment{Checked the terminology, path-sensitive means certain prooperties holding along a path only, http://en.wikipedia.org/wiki/Data-flow\_analysis} into
-our  analysis.
+determined.   %% To deal  with this,  we introduce  an analysis  which is
+%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
+%%   path-sensitive means certain prooperties  holding along a path only,
+%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
+To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
+demand $\sigma$.  Clearly, this is also the demand on the return value
+$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
+$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
+in traditional liveness analysis.
 
@@
             \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
-            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
+            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
+            \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{.34cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{.34cm} $\pi_5\!\!:\,
             $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
+            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
         \end{uprogram}}
@@
 	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
+	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_6\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}   $\pi_7\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
           $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
@@
     \begin{boxedminipage}{.49\textwidth}
-      {\sf
+      {
         \small
@@
             \begin{tabular}{lc}            
-              $[$ \\
-              $\psi_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,$ \\ 
-              $\psi_2 \mapsto \lbrace \rbrace , $\\
-              $\psi_3 \mapsto \lbrace \pz.\sigma_{\length}, \py.\cbar\sigma_{\length}, $\\
-              $\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace $  \\
-              $]$
+              $[$ 
+              $ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
+              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
+                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
+              $~~~~~~~~~~~~~~\;\; \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace ]$
             \end{tabular} \\
@@
               \begin{tabular}{lc}
-                $[$ \\
-                $\pi_{\mbox{1}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \
-                $ \pi_{\mbox{2}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \\
-                $ \psi_{\mbox{1}} \mapsto \lbrace \psi_1 \rbrace, $ \
-                $ \psi_{\mbox{2}} \mapsto \lbrace \psi_2 \rbrace, $ \
-                $ \pi_{\mbox{5}} \mapsto \lbrace \psi_3 \rbrace, $ \\
-                $ \pi_{\mbox{6}} \mapsto \lbrace \psi_3 \rbrace, $ \
-                $ \pi_{\mbox{7}} \mapsto \lbrace \psi_3 \rbrace $ \\
-                $ \psi_{\mbox{3}} \mapsto \lbrace \psi_3 \rbrace $ \\
-                $]$
+                $[~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
+                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
+                  \rbrace, $\\
+                $~~\psi_1 \mapsto \lbrace ep_1 \rbrace, $ \
+                $ \psi_2 \mapsto \lbrace ep_2 \rbrace, $ \
+                $ \pi_5 \mapsto \lbrace ep_3 \rbrace, $ \\
+                $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
+                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace\rbrack$
               \end{tabular}\\
@@
                 $[$
-                  $\psi_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                   $]$
@@
                 \begin{tabular}{lc}
-                  $[$ \
-                  $\pi_{\mbox{8}} \mapsto \lbrace \psi_4 \rbrace, $ \
-                  $ \pi_{\mbox{9}} \mapsto \lbrace \psi_4 \rbrace, $ \
-                  $ \pi_{\mbox{10}} \mapsto \lbrace \psi_4 \rbrace, $ \
-                  $ \psi_{\mbox{4}} \mapsto \lbrace \psi_4 \rbrace, $ \
-                  $ \ldots  $ \
-                  $]$  
+                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
+                  $\pi_8 \mapsto \lbrace ep_4 \rbrace, $ \\
+                  $~~~ep_4 \mapsto \lbrace ep_4 \rbrace,\ldots~\rbrack$  
                 \end{tabular}
@@
 %% \medskip
-
-Consider  an analysis of  the body  of $\length$  with a  demand $\sigma$.
-Clearly, this is also the demand on the return value $z$ giving the
-liveness  at  the  program point  $\psi_3$ as $\{\pz.\sigma\}$.   
-The  liveness at $\psi_3$  is  propagated  backwards  as  in  traditional
-liveness analysis.
 
@@
   e-paths}. \comment{Update the e-path labels}  For the
-function \length, there are three such paths $\psi_1 = \pi_1, \pi_2, \pi_3,
-\pi_4,  \pi_5$, $\psi_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
-\pi_8$, and $\psi_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
-
-The analysis  consists of  two maps: The  map $\cal{M}$ maps  a e-path to a common liveness environment, and a second map $\cal{P}$ maps
-a program  point to the set of  e-paths on which  the point lies.
-If a  program  point happens to fall  on more than one  such path, then
-the  liveness environment of  the program  point is  the variable-wise
-union of  the liveness environments  of the individual paths.   In the
-example program, ${\cal M}(\psi_1)= \ldots$, ${\cal M}(\psi_2)=\ldots$ and
-${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace \psi_1,
-\psi_2, \psi_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
-our propagation of liveness is flow-insensitive but path-sensitive.
-In the sequel
-we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
-$\cal{P}$ is easy and we do not elaborate it any further.
+function \length, there are three such paths $ep_1 = \pi_1, \pi_2, \psi_1$,
+$ep_2 = \pi_1,  \pi_2, \psi_2$ and  $ep_3  =  \pi_1, \pi_2,  \pi_3,
+\pi_4, \pi_5,  \psi_3$.
+
+The analysis consists of two maps:  The map $\cal{M}$ maps a e-path to
+a liveness environment arising out  of all uses of variables along the
+path, and  a second map $\cal{P}$ maps  a program point to  the set of
+e-paths sharing the point.  If a program point happens to fall on more
+than one such path, then the liveness environment of the program point
+is  the  variable-wise  union  of  the liveness  environments  of  the
+individual paths.  In the example program, ${\cal M}(\psi_1)= \ldots$,
+${\cal  M}(\psi_2)=\ldots$  and  ${\cal  M}(\psi_3)=  \ldots$.   Since
+${\cal  P}(pi_4)$ is  $\lbrace  \psi_1, \psi_2,  \psi_3 \rbrace$,  the
+liveness of $pi_4$ is \ldots.  In summary, our propagation of liveness
+is  flow-insensitive  but  path-sensitive.   In the  sequel  we  shall
+describe  how to obtain  $\cal{M}$.  The  formulation of  $\cal{P}$ is
+easy and we do not elaborate it any further.
 

### 2015-05-09 paper/paper.tex
commented unwanted input
Liveness

@@
 We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
-language syntax) with the program point just before expression $e$.
+language syntax) with the program point just before expression
+$e$.
 
@@
 represents access using $x$ itself and $x \mapsto \{\}$ indicates that
-$x$ is dead.  We associate  a liveness environment with program points
-denoted by $\pi$ and $\psi$.  The liveness at $\pi:e$  is the liveness just before executing
+$x$ is dead.  We associate  liveness environments with program points
+(denoted by $\pi$ and $\psi$).  The liveness at $\pi:e$  is the liveness just before executing
 $e$.
 
-A  notion related  to  liveness is  {\em  demand}.  The  demand on  an
-expression  $e$  is again  a  set  of  access paths---that  subset  of
-$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
-result.   To see  the need  for demands,  consider the  let expression
-$\pi:(\LET~x\leftarrow (\CDR~y)~\IN~ \pi':\RETURN~x)$. Assume that the
-context of  this expression produces the liveness  $[x \mapsto \lbrace
-  \epsilon, \acar  \rbrace]$ at $\pi'$.   Due to the  \LET\ definition
-which  binds $x$  to $(\CDR~y)$,  the liveness  of $x$  at  $\pi'$ now
-becomes  the  demand on  $(\CDR~y)$.   This,  in  turn, generates  the
-liveness $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar \rbrace$ at $\pi$
-(the set of $y$-rooted accesses required to explore $\lbrace \epsilon,
-\acar  \rbrace$  paths  of  the  result  of  $(\CDR~y)$)\footnote{The
-  classical analogue  is strong liveness  analysis, where $y$  and $z$
-  have  a  liveness  of  $\lbrace\epsilon\rbrace$  at  the  entry  $n:
-  x:=y+z$,    if   and    only   if    $x$   has    a    liveness   of
-  $\lbrace\epsilon\rbrace$ at  exit of  $n$.  In our  terminology, the
-  liveness of $x$  at exit from $n$ becomes the  demand on $y+z$. This
-  generates the  liveness $\lbrace y.\epsilon,  z.\epsilon \rbrace$ at
-  the entry of $n$.}.
+A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
+represents the  future use of variables, demands  represent the future
+use of the  values of expressions. The demand on  an expression $e$ is
+again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
+which the context of $e$ may explore of $e$'s result.  To see the need
+for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
+(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
+expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
+  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
+$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
+$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
+y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
+$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
+\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
+  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
+  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
+  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
+  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
+  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
+  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
 %% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
@@
 locations themselves).
-
-We use $\sigma$  to range over demands, $\alpha$  to range over access
-paths  and $\Lv$ to  range over  liveness environments.   The notation
-$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
-\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
-abuse notation to  juxtapose an edge label and a  set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$. A more
-elaborate explanation of these notions appear in \cite{asati14lgc}
-
-\subsection{Liveness Analysis for lazy languages}
-
-Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
-parts. The function $\mathit{ref}$,  when given an application $s$ and
-a demand $\sigma$, returns  the incremental liveness generated for the
-free  variables   of  $s$  due  to  the   application.   The  function
-$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
-expressions.
-
- In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
- generates no  liveness.  Notice that this is  different from liveness
- of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
- evaluated even for the null  demand, and this would require an access
- using  $x$.   A  non-null   demand  of  $\sigma$  on  (\CDR~$x$),  is
- transformed  to  the  liveness  $\{x.\epsilon, x.\acdr\sigma\}$.   In  an
- opposite  sense, the  demand of  $\acdr\sigma$ on  (\CONS~$y$~$z$) is
- transformed to  the demand  $\sigma$ on $z$.   Note that  \CONS\ does
- not, by itself,  dereference  its arguments. The rules for (\PRIM~x~y)
- and (\NULLQ~x) are on similar lines. Constants generate no liveness.
-
-For  the  case  of a  user-function  call  we  use a  third  parameter
-$\Lfonly$  that  represents the  summaries  of  all  functions in  the
-program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
-specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
-result  of a  call to  $f$  is transformed  into the  liveness of  its
-parameters at the  beginning of the call.  $\Lfonly$  is determined by
-the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
-({\sc live-define}).  This rule  describes the fixed-point property to
-be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
-for each  function in  the program  should be the  same as  the demand
-transformation  calculated from  their  bodies.  As  we  shall see  in
-Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar and  the least  solution satisfying the  rule is  the language
-generated by this  grammar.  The least solution also  ensures the safe
-collection of the greatest amount of garbage.
-
-We next  describe the function $\mathcal{L}$  that propagates liveness
-across  expressions.   This brings  us  to  the  second way  in  which
-liveness analysis for lazy languages differ from eager languages---the
-point  where  an  expression   gets  evaluated  cannot  be  statically
-determined.   %% To deal  with this,  we introduce  an analysis  which is
-%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
-%%   path-sensitive means certain prooperties  holding along a path only,
-%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
-To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
-demand $\sigma$.  Clearly, this is also the demand on the return value
-$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
-$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
-in traditional liveness analysis.
-
 \begin{figure*}[t!]
-\begin{picture}(100,130)(0, -50)
   \begin{tabular}{cc}
     \begin{boxedminipage}{.51\textwidth}
-      {\sf
         \small
@@
 \end{uprogram}}
-        } 
+        %%} 
     \end{boxedminipage}
+
     &
-    \begin{boxedminipage}{.49\textwidth}
-      {
+
+    \begin{boxedminipage}{.46\textwidth}
+
         \small
@@
             \begin{tabular}{lc}            
-              $[$ 
-              $ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
+              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
               $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                 \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
-              $~~~~~~~~~~~~~~\;\; \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace ]$
+              $~~~~~~~~~~~~~~\;\;
+                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
+                \rbrace~\rbrack$ \\
             \end{tabular} \\
@@
               \begin{tabular}{lc}
-                $[~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
+                $~[
+                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                 $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
@@
                 $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
-                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace\rbrack$
+                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace~\rbrack$
               \end{tabular}\\
@@
                 $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
                   $]$
@@
                   $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
-                  $\pi_8 \mapsto \lbrace ep_4 \rbrace, $ \\
-                  $~~~ep_4 \mapsto \lbrace ep_4 \rbrace,\ldots~\rbrack$  
+                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
                 \end{tabular}
         \end{tabular}
-      } 
+
     \end{boxedminipage} 
-    \\
-    (a) Example program.&
-    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
-        (b) {\em e-path} to liveness environment map \\
-        (c) Prog pt to {\em e-paths } map
-    \end{tabular}}
+    %% \end{ tabular}}
   \end{tabular}
-\end{picture}
+%\end{picture}
 \kern -3ex
 \vspace{0.5in}
-\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
+\caption{Example Program and its liveness maps.{\color{red} Fix the
+    formatting of the table on the right}}\label{fig:mot-example2}    
 \end{figure*}
+
+We use $\sigma$  to range over demands, $\alpha$  to range over access
+paths  and $\Lv$ to  range over  liveness environments.   The notation
+$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
+\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
+abuse notation to  juxtapose an edge label and a  set of access paths:
+$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$. A more
+elaborate explanation of these notions appear in \cite{asati14lgc}
+
+\subsection{Liveness Analysis for lazy languages}
+
+Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
+parts. The function $\mathit{ref}$,  when given an application $s$ and
+a demand $\sigma$, returns  the incremental liveness generated for the
+free  variables   of  $s$  due  to  the   application.   The  function
+$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
+expressions.
+
+ In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
+ generates no  liveness.  Notice that this is  different from liveness
+ of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
+ evaluated  even for  the null  demand. This  would require  an access
+ using $x$ resulting in  $x.\epsilon$ being live.  A non-null demand of
+ $\sigma$ on (\CDR~$x$), is transformed to the liveness $\{x.\epsilon,
+ x.\acdr\sigma\}$.  In an opposite  sense, the demand of $\acdr\sigma$
+ on  (\CONS~$y$~$z$) is  transformed to  the demand  $\sigma$  on $z$.
+ Note that \CONS\ does not,  by itself, dereference its arguments. The
+ rules for (\PRIM~x~y) and  (\NULLQ~x) are on similar lines. Constants
+ generate no liveness.
+
+For  the  case  of a  user-function  call  we  use a  third  parameter
+$\Lfonly$  that  represents the  summaries  of  all  functions in  the
+program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
+specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
+result  of a  call to  $f$  is transformed  into the  liveness of  its
+parameters at the  beginning of the call.  $\Lfonly$  is determined by
+the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
+({\sc live-define}).  This rule  describes the fixed-point property to
+be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
+for each  function in  the program  should be the  same as  the demand
+transformation  calculated from  their  bodies.  As  we  shall see  in
+Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
+grammar and  the least  solution satisfying the  rule is  the language
+generated by this  grammar.  The least solution also  ensures the safe
+collection of the greatest amount of garbage.
+
+We next  describe the function $\mathcal{L}$  that propagates liveness
+across  expressions.   This brings  us  to  the  second reason why 
+liveness analysis for lazy and eager languages are different---the
+point  where  an  expression   gets  evaluated  cannot  be  statically
+determined.   %% To deal  with this,  we introduce  an analysis  which is
+%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
+%%   path-sensitive means certain prooperties  holding along a path only,
+%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
+To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
+demand $\sigma$.  Clearly, this is also the demand on the return value
+$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
+$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
+in traditional liveness analysis.
 
@@
 
-Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
-\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
-turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
-$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
-beyond the scope of the definition of \py.  However, this does not cause any imprecision
-in the way of the GC marking  more cells as live, since \py\ is yet to
-come into existence at $\pi_6$  and its liveness will not be consulted
-by the garbage collector at this point.
-In an eager language, \py\ would not have been live at $\psi_3$, the
-point of  its last  use. In a  lazy language, however,  the definition
-$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
-$(1+~\py)$; instead  a closure is created.   The evaluation actually
-takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
-record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
-every program  point from $\psi_3$ up  to the beginning  of the program
-(since  we do  not kill  liveness).  We  thus regard  the body  of the
-function as  a tree and consider  paths  from the root of  this tree to
-program  points  which  trigger  a evaluation\footnote{Program points
-  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
-\RETURN\  expressions.  We  call such  paths {\em
-  e-paths}. \comment{Update the e-path labels}  For the
-function \length, there are three such paths $ep_1 = \pi_1, \pi_2, \psi_1$,
-$ep_2 = \pi_1,  \pi_2, \psi_2$ and  $ep_3  =  \pi_1, \pi_2,  \pi_3,
-\pi_4, \pi_5,  \psi_3$.
-
-The analysis consists of two maps:  The map $\cal{M}$ maps a e-path to
-a liveness environment arising out  of all uses of variables along the
-path, and  a second map $\cal{P}$ maps  a program point to  the set of
-e-paths sharing the point.  If a program point happens to fall on more
-than one such path, then the liveness environment of the program point
-is  the  variable-wise  union  of  the liveness  environments  of  the
-individual paths.  In the example program, ${\cal M}(\psi_1)= \ldots$,
-${\cal  M}(\psi_2)=\ldots$  and  ${\cal  M}(\psi_3)=  \ldots$.   Since
-${\cal  P}(pi_4)$ is  $\lbrace  \psi_1, \psi_2,  \psi_3 \rbrace$,  the
-liveness of $pi_4$ is \ldots.  In summary, our propagation of liveness
-is  flow-insensitive  but  path-sensitive.   In the  sequel  we  shall
-describe  how to obtain  $\cal{M}$.  The  formulation of  $\cal{P}$ is
-easy and we do not elaborate it any further.
+Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
+\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
+generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_5$.  We now make
+the following observations:
+\begin{enumerate}
+\item While the liveness of \py\  should be killed at $\pi_4$ since it
+  is  beyond the scope  of the  definition of  \py, we  do not  do so.
+  However, this  does not cause any  imprecision in the way  of the GC
+  marking more cells as live, since \py\ is yet to come into existence
+  at $\pi_4$,  and its liveness will  not be consulted  by the garbage
+  collector at  this point. 
+\item In an eager language, \py\ would not have been live at $\psi_3$,
+  a program point  beyond its last use.  In  a lazy language, however,
+  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
+  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
+  evaluation  actually takes  place while  evaluating $(\RETURN~\pz)$.
+  We therefore  record  the liveness of $\py$ at the program  
+  point $\psi_3$,  indeed at every  program point from $\psi_3$  up to
+  the beginning  of the program (since  we do not  kill liveness).
+\end{enumerate}
+
+We thus regard  the body of the function as a  tree and consider paths
+from  the  root  of this  tree  to  program  points which  trigger  an
+evaluation\footnote{Program  points which  trigger  evaluation  are
+  labelled with  $\psi$ {\color {red} ---slightly late in the day}}---\SIF\ conditions and  \RETURN\ expressions.
+We call such paths {\em e-paths}.  For the function \length, there are
+three such paths $ep_1 = \langle \pi_1, \pi_2, \psi_1\rangle$, $ep_2 =
+\langle\pi_1, \pi_2, \psi_2\rangle$ and  $ep_3 = \langle \pi_1, \pi_2,
+\pi_3, \pi_4, \pi_5, \psi_3\rangle$.
+
+Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
+analysis   comes   the    map   $\cal{M}_{\mathit   f}^{\sigma}$   (or
+$\cal{M}_{\mathit  f}$ when  the  demand $\sigma$  is  clear from  the
+context) maps  a e-path to a  liveness environment arising  out of all
+uses of variables  along the path. A second  map $\cal{P}_{\mathit f}$
+maps a  program point to the set  of e-paths sharing the  point.  If a
+program point  happens to fall  on more than  one such path,  then the
+liveness environment  of the program point is  the variable-wise union
+of the liveness environments of  the individual paths.  In the example
+program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
+ep_3\rbrace$,  the liveness  at $\pi_1$  is  ${\cal M_{\length}}(ep_1)
+\cup {\cal  M_{\length}}(ep_2)\cup {\cal M_{\length}}(ep_3)$.   In the
+sequel   we   shall  describe   how   to   obtain  $\cal{M}$   through
+$\mathcal{L}$.  The  formulation of  $\cal{P}$ is easy  and we  do not
+elaborate it any further.
+
+{\color{red}  \hrule \hrule \hrule  \medskip }
 
@@
 %======================================================================================
-In an eager language, \pw\ would not have been live after $\pi_7$, the
-point of  its last  use. In a  lazy language, however,  the definition
-$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
-$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
-takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
-record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
-every program  point from $\pi_8$ up  to the beginning  of the program
-(since  we do  not kill  liveness).  We  thus regard  the body  of the
-function as  a tree and consider paths  from the root of  this tree to
-program  points  which  trigger  a evaluation---\SIF\  conditions  and
-\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
-program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
-$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
- and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.
-
-The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-path to a common liveness environment, and a second map $\cal{P}$ maps
-a program  point to the set of  e-paths on which  the point lies.
-If a  program  point happens to fall  on more than one  such path, then
-the  liveness environment of  the program  point is  the variable-wise
-union of  the liveness environments  of the individual paths.   In the
-example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
-${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
-Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
-the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
-our propagation of liveness is flow-insensitive but path-sensitive.
-In the sequel
-we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
-$\cal{P}$ is easy and we do not elaborate it any further.
-
-Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
-evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
-  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
-expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
-  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
-triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
-is created and added to  the union.  Also notice that the consequences
-of  laziness---the entire  expression including  the condition  is not
-evaluated if  the demand  on it is  $\emptyset$.  This results  in the
-empty map denoted by $\emptyset_{\cal{M}}$.
+%% \mycomment{In an eager language, \pw\ would not have been live after $\pi_7$, the
+%% point of  its last  use. In a  lazy language, however,  the definition
+%% $\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
+%% $(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
+%% takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
+%% record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
+%% every program  point from $\pi_8$ up  to the beginning  of the program
+%% (since  we do  not kill  liveness).  We  thus regard  the body  of the
+%% function as  a tree and consider paths  from the root of  this tree to
+%% program  points  which  trigger  a evaluation---\SIF\  conditions  and
+%% \RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+%% program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
+%% $ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
+%%  and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.
+
+%% The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
+%% path to a common liveness environment, and a second map $\cal{P}$ maps
+%% a program  point to the set of  e-paths on which  the point lies.
+%% If a  program  point happens to fall  on more than one  such path, then
+%% the  liveness environment of  the program  point is  the variable-wise
+%% union of  the liveness environments  of the individual paths.   In the
+%% example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
+%% ${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
+%% Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
+%% the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
+%% our propagation of liveness is flow-insensitive but path-sensitive.
+%% In the sequel
+%% we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
+%% $\cal{P}$ is easy and we do not elaborate it any further.
+
+%% Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
+%% The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+%% evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
+%%   \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
+%% expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
+%%   M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
+%% triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
+%% is created and added to  the union.  Also notice that the consequences
+%% of  laziness---the entire  expression including  the condition  is not
+%% evaluated if  the demand  on it is  $\emptyset$.  This results  in the
+%% empty map denoted by $\emptyset_{\cal{M}}$.}
 

### 2015-05-09 paper/paper.tex

@@
 
-\input{liveness-dfa.tex}
+%%\input{liveness-dfa.tex}
 \comment{change according to example}

### 2015-05-11 paper/paper.tex
Updated implementation section

--- 2015-05-07 paper/paper.tex
@@
 $\Lfonly$  that  represents the  summaries  of  all  functions in  the
-program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
-specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
-result  of a  call to  $f$  is transformed  into the  liveness of  its
-parameters at the  beginning of the call.  $\Lfonly$  is determined by
-the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
-({\sc live-define}).  This rule  describes the fixed-point property to
-be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
-for each  function in  the program  should be the  same as  the demand
-transformation  calculated from  their  bodies.  As  we  shall see  in
-Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar and  the least  solution satisfying the  rule is  the language
-generated by this  grammar.  The least solution also  ensures the safe
-collection of the greatest amount of garbage.
+program.  $\Lfonly_f$  (the  component  of $\Lfonly$  for  a  specific
+function $f$),  expresses how the demand  $\sigma$ on the  result of a
+call to $f$ is transformed into  the liveness of its parameters at the
+beginning of the call.  $\Lfonly$  is determined by the judgement form
+$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
+This  rule  describes the  fixed-point  property  to  be satisfied  by
+$\Lfonly$, namely, the demand transformation assumed for each function
+in  the  program should  be  the  same  as the  demand  transformation
+calculated    from   their    bodies.     As   we    shall   see    in
+Section~\ref{sec:grammar-formulation},  the least  solution satisfying
+the  rule  is obtained  by  converting  the  rule into  an  equivalent
+grammar (actually the language recognized by the grammar is the least solution).  The least solution also  ensures that the greatest amount of
+garbage is collected.
 
@@
 point  where  an  expression   gets  evaluated  cannot  be  statically
-determined.   %% To deal  with this,  we introduce  an analysis  which is
-%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
-%%   path-sensitive means certain prooperties  holding along a path only,
-%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
-To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
-demand $\sigma$.  Clearly, this is also the demand on the return value
-$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
-$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
-in traditional liveness analysis.
+determined.  To  deal with this, we introduce an analysis which is
+flow-insensitive but path-sensitive\comment{Checked the terminology, path-sensitive means certain prooperties holding along a path only, http://en.wikipedia.org/wiki/Data-flow\_analysis} into
+our  analysis.
 
@@
             \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
-            \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{.34cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{.34cm} $\pi_5\!\!:\,
+            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
+            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
             $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
+            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
         \end{uprogram}}
@@
 	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_6\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}   $\pi_7\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
+	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
           $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
@@
     \begin{boxedminipage}{.49\textwidth}
-      {
+      {\sf
         \small
@@
             \begin{tabular}{lc}            
-              $[$ 
-              $ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
-              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
-                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
-              $~~~~~~~~~~~~~~\;\; \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace ]$
+              $[$ \\
+              $\psi_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,$ \\ 
+              $\psi_2 \mapsto \lbrace \rbrace , $\\
+              $\psi_3 \mapsto \lbrace \pz.\sigma_{\length}, \py.\cbar\sigma_{\length}, $\\
+              $\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace $  \\
+              $]$
             \end{tabular} \\
@@
               \begin{tabular}{lc}
-                $[~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
-                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
-                  \rbrace, $\\
-                $~~\psi_1 \mapsto \lbrace ep_1 \rbrace, $ \
-                $ \psi_2 \mapsto \lbrace ep_2 \rbrace, $ \
-                $ \pi_5 \mapsto \lbrace ep_3 \rbrace, $ \\
-                $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
-                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace\rbrack$
+                $[$ \\
+                $\pi_{\mbox{1}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \
+                $ \pi_{\mbox{2}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \\
+                $ \psi_{\mbox{1}} \mapsto \lbrace \psi_1 \rbrace, $ \
+                $ \psi_{\mbox{2}} \mapsto \lbrace \psi_2 \rbrace, $ \
+                $ \pi_{\mbox{5}} \mapsto \lbrace \psi_3 \rbrace, $ \\
+                $ \pi_{\mbox{6}} \mapsto \lbrace \psi_3 \rbrace, $ \
+                $ \pi_{\mbox{7}} \mapsto \lbrace \psi_3 \rbrace $ \\
+                $ \psi_{\mbox{3}} \mapsto \lbrace \psi_3 \rbrace $ \\
+                $]$
               \end{tabular}\\
@@
                 $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $\psi_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                   $]$
@@
                 \begin{tabular}{lc}
-                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
-                  $\pi_8 \mapsto \lbrace ep_4 \rbrace, $ \\
-                  $~~~ep_4 \mapsto \lbrace ep_4 \rbrace,\ldots~\rbrack$  
+                  $[$ \
+                  $\pi_{\mbox{8}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \pi_{\mbox{9}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \pi_{\mbox{10}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \psi_{\mbox{4}} \mapsto \lbrace \psi_4 \rbrace, $ \
+                  $ \ldots  $ \
+                  $]$  
                 \end{tabular}
@@
 %% \medskip
+
+Consider  an analysis of  the body  of $\length$  with a  demand $\sigma$.
+Clearly, this is also the demand on the return value $z$ giving the
+liveness  at  the  program point  $\psi_3$ as $\{\pz.\sigma\}$.   
+The  liveness at $\psi_3$  is  propagated  backwards  as  in  traditional
+liveness analysis.
 
@@
   e-paths}. \comment{Update the e-path labels}  For the
-function \length, there are three such paths $ep_1 = \pi_1, \pi_2, \psi_1$,
-$ep_2 = \pi_1,  \pi_2, \psi_2$ and  $ep_3  =  \pi_1, \pi_2,  \pi_3,
-\pi_4, \pi_5,  \psi_3$.
-
-The analysis consists of two maps:  The map $\cal{M}$ maps a e-path to
-a liveness environment arising out  of all uses of variables along the
-path, and  a second map $\cal{P}$ maps  a program point to  the set of
-e-paths sharing the point.  If a program point happens to fall on more
-than one such path, then the liveness environment of the program point
-is  the  variable-wise  union  of  the liveness  environments  of  the
-individual paths.  In the example program, ${\cal M}(\psi_1)= \ldots$,
-${\cal  M}(\psi_2)=\ldots$  and  ${\cal  M}(\psi_3)=  \ldots$.   Since
-${\cal  P}(pi_4)$ is  $\lbrace  \psi_1, \psi_2,  \psi_3 \rbrace$,  the
-liveness of $pi_4$ is \ldots.  In summary, our propagation of liveness
-is  flow-insensitive  but  path-sensitive.   In the  sequel  we  shall
-describe  how to obtain  $\cal{M}$.  The  formulation of  $\cal{P}$ is
-easy and we do not elaborate it any further.
+function \length, there are three such paths $\psi_1 = \pi_1, \pi_2, \pi_3,
+\pi_4,  \pi_5$, $\psi_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
+\pi_8$, and $\psi_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+
+The analysis  consists of  two maps: The  map $\cal{M}$ maps  a e-path to a common liveness environment, and a second map $\cal{P}$ maps
+a program  point to the set of  e-paths on which  the point lies.
+If a  program  point happens to fall  on more than one  such path, then
+the  liveness environment of  the program  point is  the variable-wise
+union of  the liveness environments  of the individual paths.   In the
+example program, ${\cal M}(\psi_1)= \ldots$, ${\cal M}(\psi_2)=\ldots$ and
+${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace \psi_1,
+\psi_2, \psi_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
+our propagation of liveness is flow-insensitive but path-sensitive.
+In the sequel
+we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
+$\cal{P}$ is easy and we do not elaborate it any further.
 
@@
 % GC assumptions
+\subsection{GC Points}
+Unlike eager languages where GC happens only at \CONS~ points or at 
+function calls, every \LET~ point is a potential GC point. The number of 
+automata that needs to be stored is reduced by sharing automata along all 
+the program points along an {\em E-path}.
+\comment{
 By the  time the program  starts executing, the liveness  automata are
@@
 \LET~  statement in the  program. This  leads to  an explosion  in the
-number of  automata. \comment{Directly say that we share the automata}From our  observation that along an  {\em E-path}
+number of  automata. From our  observation that along an  {\em E-path}
 all program points have the same liveness, we can reduce the number of
 automata state required  by sharing the same automata  for all program
-points along an {\em E-path}. In  the example program, we only need to
+points along an {\em E-path}.} In  the example program, we only need to
 create automata  for program  points \ldots, since  they are  the only

### 2015-05-13 paper/paper.tex
liveness

--- 2015-05-09 paper/paper.tex
@@
             \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
-            \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{.34cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{.34cm} $\pi_5\!\!:\,
+            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
+            $(\LET\ \pv\ $\leftarrow 0$ \IN
+            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\RETURN~\psi_2:\pv)$
+            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
             $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
+            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\RETURN~\psi_3:\pz)$)))))))
         \end{uprogram}}
@@
 	  \begin{uprogram}
-	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_6\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
-	  \UNL{2}   \hspace*{.78cm}   $\pi_7\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
+	  \UNL{1} $(\DEFINE\ (\main)$
+          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
+	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
           $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
-          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
+          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\RETURN~\psi_4:\pw)))))$
 \end{uprogram}}
@@
                   \rbrace, $\\
-                $~~\psi_1 \mapsto \lbrace ep_1 \rbrace, $ \
-                $ \psi_2 \mapsto \lbrace ep_2 \rbrace, $ \
-                $ \pi_5 \mapsto \lbrace ep_3 \rbrace, $ \\
-                $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
-                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace~\rbrack$
+$ ~~~\pi_3 \mapsto \lbrace ep_2
+                  \rbrace $, $ \pi_4 \mapsto \lbrace ep_2
+                  \rbrace$, $\pi_5 \mapsto \lbrace  ep_3 \rbrace$, \\
+                $~~~\pi_6 \mapsto \lbrace ep_3 \rbrace,  \pi_7 \mapsto
+                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3 \rbrace~\rbrack$
               \end{tabular}\\
@@
 We next  describe the function $\mathcal{L}$  that propagates liveness
-across  expressions.   This brings  us  to  the  second reason why 
-liveness analysis for lazy and eager languages are different---the
-point  where  an  expression   gets  evaluated  cannot  be  statically
-determined.   %% To deal  with this,  we introduce  an analysis  which is
-%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
-%%   path-sensitive means certain prooperties  holding along a path only,
-%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
-To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
-demand $\sigma$.  Clearly, this is also the demand on the return value
-$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
-$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
-in traditional liveness analysis.
+across expressions.  Note that,  unlike eager languages, the syntactic
+occurrence  of an expression  in a  program denotes  the point  of 
+creation of its closure, not necessarily its evaluation.
+%% To   deal  with   this,   we  introduce   an   analysis  which   is
+%%flow-insensitive but path-sensitive\comment{Checked the terminology,
+%%path-sensitive means certain properties  holding along a path only,
+%%http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
+To see the  consequences of this, consider an analysis  of the body of
+$\length$ with a demand $\sigma$.  Clearly, this is also the demand on
+the return value $z$ giving the liveness at the program point $\pi_8$
+as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated backwards
+as in traditional liveness analysis.
 
@@
 \pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
-generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_5$.  We now make
+generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
 the following observations:
 \begin{enumerate}
-\item While the liveness of \py\  should be killed at $\pi_4$ since it
+\item While the liveness of \py\  should be killed at $\pi_6$ since it
   is  beyond the scope  of the  definition of  \py, we  do not  do so.
   However, this  does not cause any  imprecision in the way  of the GC
-  marking more cells as live, since \py\ is yet to come into existence
-  at $\pi_4$,  and its liveness will  not be consulted  by the garbage
-  collector at  this point. 
-\item In an eager language, \py\ would not have been live at $\psi_3$,
+  marking more cells as live,  since, during execution, \py\ is yet to
+  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
+  consulted by the garbage collector at this point.
+\item In an eager language, \py\ would not have been live at $\pi_8$,
   a program point  beyond its last use.  In  a lazy language, however,
@@
   We therefore  record  the liveness of $\py$ at the program  
-  point $\psi_3$,  indeed at every  program point from $\psi_3$  up to
-  the beginning  of the program (since  we do not  kill liveness).
+  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
+  the beginning  of the program $\pi_1$ (since  we do not  kill liveness).
 \end{enumerate}
@@
 We thus regard  the body of the function as a  tree and consider paths
-from  the  root  of this  tree  to  program  points which  trigger  an
+from  the  root   of  this  tree  to  program   points  which  trigger
 evaluation\footnote{Program   points  which  trigger   evaluation  are
-  labelled with  $\psi$ {\color {red} ---slightly late in the day}}---\SIF\ conditions and  \RETURN\ expressions.
-We call such paths {\em e-paths}.  For the function \length, there are
-three such paths $ep_1 = \langle \pi_1, \pi_2, \psi_1\rangle$, $ep_2 =
-\langle\pi_1, \pi_2, \psi_2\rangle$ and  $ep_3 = \langle \pi_1, \pi_2,
-\pi_3, \pi_4, \pi_5, \psi_3\rangle$.
+  labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
+    day}}---\SIF\ conditions  and \RETURN\ expressions.   We call such
+paths {\em e-paths}.   For the function \length, there  are three such
+paths  $ep_1   =  \langle   \pi_1,  \pi_2, \psi_1\rangle$,   $ep_2  =
+\langle\pi_1, \pi_2, \pi_3, \pi_4, \psi_2\rangle$ and  $ep_3 = \langle \pi_1, \pi_2,
+\pi_5,  \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
 
 Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the  liveness
-analysis   comes   the    map   $\cal{M}_{\mathit   f}^{\sigma}$   (or
-$\cal{M}_{\mathit  f}$ when  the  demand $\sigma$  is  clear from  the
-context) maps  a e-path to a  liveness environment arising  out of all
-uses of variables  along the path. A second  map $\cal{P}_{\mathit f}$
-maps a  program point to the set  of e-paths sharing the  point.  If a
-program point  happens to fall  on more than  one such path,  then the
-liveness environment  of the program point is  the variable-wise union
-of the liveness environments of  the individual paths.  In the example
+analysis computes two maps:  the function $\mathcal{L}$ computes a map
+from a  e-path to a  liveness environment arising  out of all  uses of
+variables  along   the  path.  We  call   this  map  $\cal{M}_{\mathit
+  f}^{\sigma}$ (or  $\cal{M}_{\mathit f}$ when the  demand $\sigma$ is
+clear from  the context).  A  second map $\cal{P}_{\mathit f}$  maps a
+program point to  the set of e-paths sharing the  point.  If a program
+point happens  to fall on more  than one such path,  then the liveness
+environment of  the program  point is the  variable-wise union  of the
+liveness  environments  of  the  individual  paths.   In  the  example
 program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
@@
 Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $\psi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+The  expression $(\RETURN~\psi:x)$  with a  demand $\sigma$  forces an
 evaluation  of  $x$  and  hence  a new  map  ${\cal  M}=[\psi  \mapsto
@@
 triggers an evaluation of $x$, the map $[\psi \mapsto \{x.\epsilon\}]$
-is created and added to  the union.  Also notice the consequences
-of  laziness---the entire  expression including  the condition  is not
+is created  and added to the  union.  Also notice  a consequences of
+laziness---the  entire  expression  including  the  condition  is  not
 evaluated if  the demand  on it is  $\emptyset$.  This results  in the

### 2015-05-16 paper/paper.tex
algo working.

--- 2015-05-11 paper/paper.tex
@@
 %\usepackage{program}
+\usepackage{algorithmic}
+
 \input{fh_defs}
@@
 distinct, so  that no scope shadowing  occurs---this simplifies proofs
-of soundness.  The $\RETURN$  and \SIF\ expressions trigger evaluation
+of soundness.  The $\SRETURN$  and \SIF\ expressions trigger evaluation
 of  closures.
@@
   a  sequence of  $\LET$'s  of the  form $z_i\leftarrow  (\ID\,\,y_i)$
-  followed by the body $e_f$ but with its $(\RETURN\,\,w)$ expressions
+  followed by the body $e_f$ but with its $(\SRETURN\,\,w)$ expressions
   replaced by $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}
@@
        (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
-       (\RETURN\,\, x) && \mbox{\em --- return from function}
+       (\SRETURN\,\, x) && \mbox{\em --- return from function}
     \end{array}\right. \\
@@
 \shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
-(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
-\hline
-$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\RETURN~x)
+(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
+\hline
+$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\SRETURN~x)
   \rightsquigarrow$ \\
-$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-coo} \\
+$\rho'\oplus\rho,~ (\rho, x, (\SRETURN~x))\!:\!S, H,  e$} & \sc{return-coo} \\
 \hline
@@
 for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
-(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
+(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context  of  this
 expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
@@
             \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
+            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\SRETURN ~0)$
             \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
@@
             $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
+            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\SRETURN ~\pz)$))))))
         \end{uprogram}}
@@
           $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
-          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
+          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\SRETURN\ \pw)))))))
 \end{uprogram}}
@@
 %% 	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
-%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
-%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
+%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\SRETURN_1~\pu)$))
+%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\SRETURN_2~\py)$)))))
 %% 	\end{uprogram}}
@@
   evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
-  evaluation  actually takes  place while  evaluating $(\RETURN~\pz)$.
+  evaluation  actually takes  place while  evaluating $(\SRETURN~\pz)$.
   We therefore  record  the liveness of $\py$ at the program  
@@
 evaluation\footnote{Program  points which  trigger  evaluation  are
-  labelled with  $\psi$ {\color {red} ---slightly late in the day}}---\SIF\ conditions and  \RETURN\ expressions.
+  labelled with  $\psi$ {\color {red} ---slightly late in the day}}---\SIF\ conditions and  \SRETURN\ expressions.
 We call such paths {\em e-paths}.  For the function \length, there are
@@
 
-Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $\psi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
+The  expression $\psi:(\SRETURN~x)$  with  a demand  $\sigma$ forces  an
 evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\psi  \mapsto
@@
 \\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
+\mathcal{L}((\SRETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
 \mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
@@
 %% $(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
-%% takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
+%% takes place while evaluating  $(\SRETURN~\pu)$. Thus it is necessary to
 %% record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
@@
 %% program  points  which  trigger  a evaluation---\SIF\  conditions  and
-%% \RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+%% \SRETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
 %% program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
@@
 
-%% Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-%% The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
+%% Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
+%% The  expression $\pi:(\SRETURN~x)$  with  a demand  $\sigma$ forces  an
 %% evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
@@
 \mycomment{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
-expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
+expression $e$.   The cases $\SRETURN$ and  $\SIF$ are straightforward,
 but note the liveness $x.\epsilon$  generated by the latter.  The case
@@
 %------------------------------------------------------------%
+\input{lazy-liveness}
 

### 2015-05-18 paper/paper.tex
Intro

--- 2015-05-13 paper/paper.tex
@@
 efficiently  reclaim  unused   memory  allocated  by  programs  either
-explicitly (for example using constructors) or implicitly (for example
-for      closures).       However,      empirical      studies      on
+explicitly  (for  example using  constructors)  or implicitly  (during
+creation    of    closures).     However,   empirical    studies    on
 Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
@@
 program at  a later  stage).  Due  to lack of  a suitable  analysis to
-distinguish  reachable  objects  from live  objects.  current  garbage
+distinguish  reachable  objects  from  live objects,  current  garbage
 collectors conservatively approximate the  liveness of heap objects by
 their reachability from a set of memory locations called the {\em root
-  set\/} resulting in unnecessary memory usage.
-
-Languages having lazy  semantics for evaluation put  further demand on
-memory as they  require heap allocated closure to be  carried from the
-point of occurrence  of an expression to the point  of (first) real use
-of  that expression.  Strictness analysis~\cite{some-refs}  statically
-determines  the  patterns  of  future usage  of  an  expression,  thus
-allowing some  of the  closures to  be evaluated at  the point  of the
-occurrence of the corresponding expression.
+  set\/}. This results in unnecessary memory retention.
+
+Programs written in languages having lazy semantics for evaluation put
+an further demands on  memory as they require heap allocated
+closures to be  carried from the point of  occurrence of an expression
+to the point of (first) real use of that expression.
+%% This   may   be   mitigated  by   strictness
+%% analysis~\cite{some-refs} that  can statically determine  the patterns
+%% of future usage  of an expression, thus allowing  some of the closures
+%% to be  evaluated at the point  of the occurrence  of the corresponding
+%% expression.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 This results in an increase  in the garbage reclaimed and consequently
-in fewer  garbage collections.  Further, we eagerly evaluate closures
-that are guaranteed to be evaluated in the future (strict closures).
-
-The main contribution  of this paper is an uniform  framework for {\em
-  liveness} and {\em strictness}  analysis to enable the optimizations
-mentioned  above. Using  a central  notion of  demand, we  formulate a
-context-sensitive liveness  analysis for structured data  and prove it
-correct.  We then  present strictness as a dual of  liveness. We use a
-0-CFA-like conservative approximation to  annotate each allocation and
+in fewer  garbage collections.  
+Using  a central notion  of demand,  we formulate  a context-sensitive
+liveness  analysis  and  prove   it  correct.   We  use  a  0-CFA-like
+conservative   approximation   to   annotate   each   allocation   and
 function-call program point  with a finite-state automaton---which the
-garbage-collector inspects to curtail reachability during marking. The
-result of  the strictness analysis,  being necessarily finite,  can be
-directly used to insert code to evaluate closures.
-
-We previously proposed an interprocedural liveness analysis to improve
-garbage collection first order eager subset of Scheme~\cite{cc-paper}.
-The current paper extends the work to handle lazy evaluation semantics
-for the same language.
+garbage-collector inspects to curtail reachability during marking.
+
+While this idea has been shown to be effective~\cite{asati14lgc} for a
+first-order  eager  language,  a  straightforward  extension  of  this
+earlier  technique  is  not   possible  in  lazy  languages,  where  a
+heap-allocated   object  may  either   be  a   evaluated  data   or  a
+closure. Since liveness is brought about by evaluation of closures and
+given the  uncertainty in lazy  languages regarding the  program point
+where this  evaluation takes  place, this introduces  complications in
+the analysis itself.  Further,  since we want to extend liveness-based
+collection to closures, we  track closures  to their
+points of creation, from where we obtain their liveness. Since this is
+done  at  runtime,  this  requires considerable  modification  to  the
+earlier GC scheme.
+
+%% We  previously  proposed an  interprocedural  liveness analysis  to
+%% improve   garbage   collection   first   order  eager   subset   of
+%% Scheme~\cite{asati14lgc}.   The current paper  extends the  work to
+%% handle lazy evaluation semantics for the same language.
 
@@
 garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections, a decrease  in the memory size required  to run programs,
-and  reduced  overall  garbage  collection  time  for  a  majority  of
-programs.
+collections,  a  (some quantification)  decrease  in  the memory  size
+required to run programs,  and reduced overall garbage collection time
+for a majority of programs.
 
@@
 and    notations.
-% Section~\ref{sec:operational}   describes    the
-% operational semantics  of the  language.
-Liveness analysis is described in
-Section~\ref{sec:liveness} followed by a sketch of a correctness
-proof relative to a non-standard semantics.
-Section~\ref{sec:computing} shows how to encode liveness as
-finite-state automata.
-Section~\ref{sec:experiments}   reports   experimental   results and
-Section~\ref{sec:lgc-always-better} proves that a liveness based
-collector can never do more garbage collections than a reachability
-based collector. 
+% Section~\ref{sec:operational} describes the operational semantics of
+% the language.
+Liveness analysis is  described in Section~\ref{sec:liveness} followed
+by  a  sketch  of  a  correctness proof  relative  to  a  non-standard
+semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
+as   finite-state  automata.    Section~\ref{sec:experiments}  reports
+experimental  results  and Section~\ref{sec:lgc-always-better}  proves
+that a liveness based collector  can never do more garbage collections
+than a reachability based collector.
 % and Section~\ref{sec:conclusion} concludes the paper.
@@
 
-The body of the program is the expression denoted by $e_\mainpgm$; for
-analysis purposes it is convenient to regard $e_\mainpgm$ as part of a
-function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
-We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
-language syntax) with the program point just before expression
-$e$.
+The body  of a function ${\mathit  f}$ will be  denoted as $e_{\mathit
+  f}$. for  analysis purposes it is convenient  to regard $e_\mainpgm$
+as part of a function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$
+as in  C\@.  We  write $\pi\!:\!e$ to  associate the label  $\pi$ (not
+part  of the  language  syntax)  with the  program  point just  before
+expression $e$.
 
@@
 
-Here $\mathit{Loc}$ is a countable set  of locations in the heap.  Since
-all data objects are boxed, we  model an environment as a mapping from
-the set of variables of the program $\mathit{Var}$ to locations in the
-heap.  These either contain a value  in WHNF (a number, the empty list
-$\NIL$, or  a \CONS\ cell with possibly unevaluated constituents)
-or a {\em closure}.   A closure is a pair $(s, \rho)$  in which $s$ is
-an  unevaluated application  with  the values  of  its free  variables
-given by $\rho$.
+Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
+Since all data objects are boxed, we model an environment as a mapping
+from the set  of variables of the program  $\mathit{Var}$ to locations
+in the  heap.  These  either contain  a value in  WHNF (a  number, the
+empty  list  $\NIL$,  or  a  \CONS\  cell  with  possibly  unevaluated
+constituents) or a {\em closure}.  A  closure is a pair $(s, \rho)$ in
+which $s$  is an unevaluated application  with the values  of its free
+variables given by $\rho$.
 
@@
 $x$ is  dead.  We associate liveness environments  with program points
-(denoted by $\pi$ and $\psi$).  The liveness at $\pi:e$  is the liveness just before executing
-$e$.
+(denoted  by  $\pi$ and  $\psi$).   The  liveness  at $\pi:e$  is  the
+liveness just before executing $e$.
 
@@
   \begin{tabular}{cc}
-    \begin{boxedminipage}{.51\textwidth}
+    \begin{boxedminipage}{.45\textwidth}
         \small
@@
 
-    \begin{boxedminipage}{.46\textwidth}
+    \begin{boxedminipage}{.51\textwidth}
 
@@
         \begin{tabular}{lc}
-        $\cal{M_{\mbox{\tiny \length}}} = $ 
+        $\mathcal{L}(e_{\length},\sigma,\Lfonly) = $ 
             \begin{tabular}{lc}            
@@
               \end{tabular}\\
-              $\cal{M_{\mbox{\tiny \main}}} = $ 
+              $\mathcal{L}(e_{\main},\sigma,\Lfonly) = $ 
               \begin{tabular}{lc}            
@@
 creation of its closure, not necessarily its evaluation.
-%% To   deal  with   this,   we  introduce   an   analysis  which   is
-%%flow-insensitive but path-sensitive\comment{Checked the terminology,
-%%path-sensitive means certain properties  holding along a path only,
-%%http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
 To see the  consequences of this, consider an analysis  of the body of
@@
 as in traditional liveness analysis.
-
-%%%%%%%%%%%%%%%%%%%%%%
-
-% Push this to a later point
-
-%% Unlike eager languages, a variable may be bound to an expression
-%%   that is not fully evaluated. i.e. a closure.
-
-%% To  deal with the  first issue,  we apply  the results  of liveness
-%% analysis only to fully evaluated data.  If the marking phase during
-%% garbage collection leads to a closure, and the closure is not dead,
-%% we  mark  the  entire   reachable  memory  under  the  closure  for
-%% collection.   A  better scheme  would  be  to  garbage collect  the
-%% closure based on  the liveness at the point  of its creation.  This
-%% is sound because our  language is statically scoped.  However, this
-%% is  work  in progress,  and  the  method  described in  this  paper
-%% switches over to reachability while collecting closures.
-
-%% %%%%%%%%%%%%%%%%%%%%%%%%
-
-%% \renewcommand{\arraystretch}{1}{
-%% 	  \begin{uprogram}
-%% 	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
-%% 	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
-%% 	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
-%%           \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
-%%           $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
-%% 	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
-%%           $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
-%% 	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
-%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
-%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
-%% 	\end{uprogram}}
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% \medskip
 
@@
 paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
-\langle\pi_1, \pi_2, \pi_3, \pi_4, \psi_2\rangle$ and  $ep_3 = \langle \pi_1, \pi_2,
-\pi_5,  \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
+\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
+\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
 
 Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
-analysis computes two maps:  the function $\mathcal{L}$ computes a map
-from a  e-path to a  liveness environment arising  out of all  uses of
-variables  along   the  path.  We  call   this  map  $\cal{M}_{\mathit
-  f}^{\sigma}$ (or  $\cal{M}_{\mathit f}$ when the  demand $\sigma$ is
-clear from  the context).  A  second map $\cal{P}_{\mathit f}$  maps a
-program point to  the set of e-paths sharing the  point.  If a program
-point happens  to fall on more  than one such path,  then the liveness
-environment of  the program  point is the  variable-wise union  of the
-liveness  environments  of  the  individual  paths.   In  the  example
-program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
-ep_3\rbrace$,  the liveness  at $\pi_1$  is  ${\cal M_{\length}}(ep_1)
-\cup {\cal  M_{\length}}(ep_2)\cup {\cal M_{\length}}(ep_3)$.   In the
-sequel   we   shall  describe   how   to   obtain  $\cal{M}$   through
-$\mathcal{L}$.  The  formulation of  $\cal{P}$ is easy  and we  do not
-elaborate it any further.
-
-{\color{red}  \hrule \hrule \hrule  \medskip }
+analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
+computes a  map that takes a  e-path in ${\mathit f}$  and returns the
+liveness environment  arising out of  all uses of variables  along the
+e-path.  A  second map $\cal{P}_{\mathit  f}$ maps any program  point in
+$\mathit f$ to
+the set of  e-paths sharing the point.  If a  program point happens to
+fall on  more than one such  e-path, then the  liveness environment of
+the  program  point  is   the  variable-wise  union  of  the  liveness
+environments of the individual e-paths.  In the example program, since
+${\mathcal P}_{\length}(\pi_1) = \lbrace ep_1, ep_2, ep_3\rbrace$, the
+liveness at $\pi_1$ is ${\cal  M}(ep_1) \cup {\cal M}(ep_2) \cup {\cal
+  M}(ep_3)$, where ${\cal M} = \mathcal{L}(e_f, \sigma, \Lfonly)$.  In
+the  sequel we shall  only define  $\mathcal{L}$.  The  formulation of
+$\cal{P}$ is easy and we do not elaborate it any further.
 
@@
 The  expression $(\RETURN~\psi:x)$  with a  demand $\sigma$  forces an
-evaluation  of  $x$  and  hence  a new  map  ${\cal  M}=[\psi  \mapsto
-  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
-expression $(\SIF~\psi:x~e_1~e_2)$ is  a pointwise-union of the ${\cal
-  M}$-maps of $e_1$  and $e_2$. In addition, since  the condition also
-triggers an evaluation of $x$, the map $[\psi \mapsto \{x.\epsilon\}]$
-is created  and added to the  union.  Also notice  a consequences of
-laziness---the  entire  expression  including  the  condition  is  not
-evaluated if  the demand  on it is  $\emptyset$.  This results  in the
-empty map denoted by $\emptyset_{\cal{M}}$.
+evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
+  \{x.\sigma\}] $  needs to  be created.  The  map for  the expression
+$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
+$e_2$. In addition, since the condition also triggers an evaluation of
+$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
+added  to the  union.  Also  notice a  consequences  of laziness---the
+entire  expression including  the condition  is not  evaluated  if the
+demand on it is $\emptyset$.  This results in the empty map, which, by
+abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.
 %==============================================================
@@
 \\[1ex]
-\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
+\mathcal{L}((\RETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
+x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
+  new e-path terminating with $\psi$}\\
 \mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
@@
         \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
-        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
+        \lbrack\lbrace \mathit{ep} \mapsto  \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
         \emptyset_{\cal{M}?}  & \mbox{otherwise}
                  \end{array} \\
- \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
+&&\mbox{where $\mathit{ep}$ is a new e-path
+  terminating with $\psi$}\\
 \mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
-        \in \mathcal{M}\}
-\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
-\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
- \makebox[0mm]{\hspace*{7cm}   
+        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
+\mbox{where}~\mathcal{M} =
+\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in \mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
+&& \makebox[0mm]{\hspace*{7cm}   
  $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
            x_2\, \ldots\, x_n)/x]))$,} \\  
- \makebox[0mm]{\hspace*{5.5cm} and 
-     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
+ \makebox[0mm]{\hspace*{8cm} where
+     $x_1\, \ldots\, x_n$ are the variables in $s$}
 \end{eqnarray*}
@@
 \end{figure*}
-%======================================================================================
-%% \mycomment{In an eager language, \pw\ would not have been live after $\pi_7$, the
-%% point of  its last  use. In a  lazy language, however,  the definition
-%% $\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
-%% $(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
-%% takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
-%% record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
-%% every program  point from $\pi_8$ up  to the beginning  of the program
-%% (since  we do  not kill  liveness).  We  thus regard  the body  of the
-%% function as  a tree and consider paths  from the root of  this tree to
-%% program  points  which  trigger  a evaluation---\SIF\  conditions  and
-%% \RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
-%% program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
-%% $ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
-%%  and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.
-
-%% The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
-%% path to a common liveness environment, and a second map $\cal{P}$ maps
-%% a program  point to the set of  e-paths on which  the point lies.
-%% If a  program  point happens to fall  on more than one  such path, then
-%% the  liveness environment of  the program  point is  the variable-wise
-%% union of  the liveness environments  of the individual paths.   In the
-%% example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
-%% ${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
-%% Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
-%% the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
-%% our propagation of liveness is flow-insensitive but path-sensitive.
-%% In the sequel
-%% we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
-%% $\cal{P}$ is easy and we do not elaborate it any further.
-
-%% Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-%% The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
-%% evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
-%%   \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
-%% expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
-%%   M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
-%% triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
-%% is created and added to  the union.  Also notice that the consequences
-%% of  laziness---the entire  expression including  the condition  is not
-%% evaluated if  the demand  on it is  $\emptyset$.  This results  in the
-%% empty map denoted by $\emptyset_{\cal{M}}$.}
-
- The  liveness of {\LET} has  an elegant  formulation.   Recollect that  our
- language permits lazy {\LET}s whose definitions can be recursive. Thus,
- in the expression $\LET~x  \leftarrow~s~\IN~e$, $x$ can occur in $s$.
- We  model the  definition $x  \leftarrow~s$ as  a  possibly recursive
- function which captures an  arbitrary number of ``unrollings'' of the
- definition.  We pass  to this function the demand  arising out of the
- liveness   of  $x$   at   the   beginning  of   $e$,   and  let   the
- \Lfonly\ abstraction derive the liveness on the variables of $s$.
-
-\mycomment{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
-expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
-but note the liveness $x.\epsilon$  generated by the latter.  The case
-$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
-instruction:  the liveness  of $e$  is given  by taking  the liveness,
-$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
-liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
-$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
-$\mathit{ref}(s,\Lv(z),\Lfonly)$.
+
+ The liveness  of {\LET} has  an elegant formulation.   Recollect that
+ our language permits lazy {\LET}s whose definitions can be recursive,
+ i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
+ $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
+ recursive function (named $\mathit{f}$ in the rule) which captures an
+ arbitrary number of ``unrollings'' of the definition\footnote{The $*$
+   in the definition of ${\mathit f}$ represents  non-deterministic choice,
+   but can be treated as a constant for purposes of analysis.}.  We pass to the
+ demand transformer $\Lfonly_{\mathit f}$  of this function the demand
+ arising out  of the  liveness of  $x$ at the  beginning of  $e$. This
+ gives the liveness of the variables of $s$.
+
+\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
+  of  an   expression  $e$.   The  cases  $\RETURN$   and  $\SIF$  are
+  straightforward, but note the liveness $x.\epsilon$ generated by the
+  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
+  three-address instruction:  the liveness of  $e$ is given  by taking
+  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
+  any incremental  liveness from  $s$.  The main  subtlety is  how the
+  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
+  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
 
@@
 $\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
-analysis) can  be safely  approximated, for each  function, by  a {\em
-  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
-The  summary consists  of a  pair  of a  single demand  and, for  this
-demand, the corresponding  tuple of demands the function  makes on its
+analysis) can be safely approximated  by a {\em procedure summary} for
+each function  (unifying the contexts  in the style of  0-CFA)\@.  The
+summary consists  of a pair of  a single demand and,  for this demand,
+the  corresponding  tuple  of   demands  the  function  makes  on  its
 arguments.
 
+{\color{red} Proof of correctness of the liveness analysis...}
+
 \section{Computing liveness and its encoding as a table}\label{sec:computing}
-Section~\ref{sec:liveness} gave a context-sensitive liveness
-analysis and {\color {red}  proved it correct with reference
-  to   a  {\em   minefield}   semantics}.   In   particular,
-\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
-describes  a  liveness  set  for  each program  point  in  a
-function in terms of a  given $\sigma$ and \Lfonly. We still
-have not mentioned what is the specific sigma that should be
-used for each function How is LF to be computed.
-
-It helps to cast the equations  in the form of a grammar. To
-do  this  we  first  have  to  cast  massage  the  equations
-slightly:
+Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
+and  {\color  {red}  proved  it  correct  with  reference  to  a  {\em
+    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
+together described a liveness set for each program point in a function
+in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
+describe how to obtain demand transformers \Lfonly\ from the rule {\sc
+  live-define} and how to compute the specific demand $\sigma$ on each
+function.
+
+To  answer these  questions, it  is convenient  to cast  the equations
+arising out of the $\mathit{ref}$  and $\mathcal{L}$ as the rules of a
+grammar.   But first  we modify  the rules  themselves to  a different
+form.
 
@@
 $\emptyset$, if  $\sigma$ itself $\emptyset$.  To  realize these rules
-$\sigma$ needs to be known.This creates difficulties
-since  we  want  to  solve  the  equations  arising  out  of  liveness
-symbolically.
+$\sigma$ needs to be known. This creates difficulties since we want to
+solve the equations arising out of liveness symbolically.
 
@@
 \end{align*}
-
 and \cbar\ as:
-
 \begin{align*}
@@
         \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
-        &[\,l \mapsto  x.\cbar \sigma]
+        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
 \end{align*}
@@
 similar to \CAR. 
-
 The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
@@
 \CONS\  rule,   we  refer   the  reader  to   \cite{asati14lgc}.   The
-proposition also holds for the the other rules for similar reasons.
-
-%% To see why the  proposition is true, consider an analysis
-%% based on  \Lfunonly$'$ in  which $\sigma$ appears  in the
-%% context      $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.
-%% Let $\alpha \in  \sigma$.  The symbol \bcar(\bcdr) merely
-%% marks a place in  $\alpha$ where the original \CONS\ rule
-%% would have erased  an immediately following \acar(\acdr),
-%% or,  in absence  of  such a  symbol,  would have  dropped
-%% $\alpha$ itself.   Since the  application of any  rule in
-%% \Lfunonly$'$  merely  adds symbols  at  the beginning  of
-%% $\alpha$, the  markers and other symbols  in $\alpha$ are
-%% propagated to  other dependent parts of  program in their
-%% same  relative   positions.   Consequently,  the  erasure
-%% carried   out   at  the   end   of   the  analysis   with
-%% $\stackrel{*}{\hookrightarrow}$ gives  the same result as
-%% obtained through \Lfunonly.
+proposition also holds for other modified rules for similar reasons.
+
+{\color{red} \medskip \hrule \hrule \hrule  \medskip }
 
@@
 The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running 
-example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
-how  to generate  the equations  describing the  demand transformation
+example.   Given a  function $\mathit{f}$,  we now  describe  how to
+generate    equations    for     the    demand    transformation
 \Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
-determine   \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.    Of   particular
-interest  is the liveness  at the  program point  at the  beginning of
-$e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
-equations     of     the     form    $\Lf{\mathit{f}}{i}{\sigma}     =
-\Lanv{\pi_1}{x_i}$, where  $x_i$ is the  $i^{\mbox{\footnotesize th}}$
-parameter of $\mathit{f}$. Doing this for \length, we get:
-\begin{align*}
- \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
-\end{align*}
-Which gives the following equation for \Lfone{\length}: 
+determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
+consider $\Lanv{1}{x_i}$,  the  liveness of    the $i^{\mbox{\footnotesize  th}}$
+parameter $x_i$ at the  program point at the  beginning of
+$e_{\mathit{f}}$ (assumed  to be  $\pi_1$). By the rule {\sc
+  live-define}, this should be the
+same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
 \begin{eqnarray*}
-    \Lf{\length}{1}{\sigma}
-    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
+&& \Lanv{1}{\xl} = \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
 \end{eqnarray*}
+and  the only equation defining \Lfone{\length} is:  
+  \begin{eqnarray*}
+   && \Lf{\length}{1}{\sigma}
+    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
+   \end{eqnarray*}
 
 In general, the equations for  \Lfonly\ are recursive since, as in the
-case of \length,  \Lan{\pi_1}\ may, in turn, be  expressed in terms of
-\Lfone{\mathit{f}}.
-
-As mentioned earlier\comment{Check if already mentioned, Not mentioned earlier}, each of  the liveness rules modifies a demand by
-prefixing   it with   symbols   in   the   above   set,   and   therefore
-$\Lf{f}{i}{\sigma}$ has the closed form:
+case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
+\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
+solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
+each  of the liveness  rules modifies  a demand  by prefixing  it with
+symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
+\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
+closed form:
 \begin{eqnarray}
 \label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} &=& \Df{f}{i}.\sigma
+  \Lf{f}{i}{\sigma} = \Df{f}{i}.\sigma
 \end{eqnarray}
-where \Df{f}{i} are sets of  strings over the alphabet $\lbrace \acar,
-\acdr,\bcar, \bcdr, \cbar \rbrace$.   Substituting the guessed form in
+where \Df{f}{i} are sets of  strings over the alphabet mentioned above.   Substituting the guessed form in
 the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
-$\sigma$,  we  get an  equation  for  \Df{f}{i} that is
-independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
-for \Lfonly$_{\mathit f}$.  Applied to \Lfonly$_{\mathit \length}$, we
+$\sigma$,  we get  an equation  for \Df{f}{i}  that is  independent of
+$\sigma$.   Any   solution  for   \Df{f}{i}  yields  a   solution  for
+\Lfonly$_{\mathit  f}$.  Applied  to  \Lfonly$_{\mathit \length}$,  we
 get:
-  \begin{align*}
-    \Df{\length}{1} &= \lbrace\cbar\rbrace 
+  \begin{eqnarray*}
+  &&  \Df{\length}{1}.\sigma = \cbar.\sigma
+    \cup \acdr\Df{\length}{1}(\cbar\sigma), ~\mbox{or}\\
+  &&  \Df{\length}{1} = \lbrace\cbar\rbrace 
     \cup \acdr\Df{\length}{1}\cbar
-  \end{align*}
+  \end{eqnarray*}
 
@@
 
-To avoid analysing the body of a function we assumed a symbolic demand $\sigma$
-and created a summary liveness for the function arguments and the variables 
-in the function. To get the actual liveness we have to substitute the actual 
-demands that are placed on the function. We calculate an over-approximation 
-of the actual demand, a {\em summary demand} and calculate the liveness at each GC point inside the 
-function. Even though we lose a bit of precision due to this, our 
-experiments show that this is acceptable \comment{Is this claim true?}.
-The {\em summary demand} is calculated by taking a union of the demand at 
-every call site of a function. A function call in our language can only be of the
-form $\pi\!\!:\!(\LET~x  \leftarrow (f\,y_1\,\ldots\,y_n)\ldots$. The demand on 
-$f$ due to this function call will be the liveness of $x$ at the program 
-point $\pi$. Denoting this demand as $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
-site $\pi$  in function $g$) we can write the {\em summary demand} as 
-$\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
-\cup \deltacall{f}{\pi^k}{g^k}$.
-
-\comment{
-We now calculate a 0-CFA-style summary liveness for each GC point of a
-program.  There are two parts  to this.  First, for each function $f$,
-we determine a {\em  summary demand} $\sigma_f$ over-approximating any
-demand $\sigma$ passed to $f$.  So, consider a function $g$ containing
-a  call   site  $\pi$   to  $f$,  say   $\pi\!\!:\!(\LET~x  \leftarrow
-(f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.   Let   the  demand  on   $g$  be
-$\sigma_g$ and the  liveness of $x$ at $\pi_i$  be $\Lanv{i}{x}$. Then
-the $\LET$ case of Figure~\ref{fig:live-judge} tells us this call site
-contributes  $\Lanv{i}{x}$ to  the  demand $\sigma_f$  placed on  $f$.
-Write $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
-site $\pi$  in $g$).  Thus, supposing  the $k$ call  sites to function
-$f$  are  $\pi^1$ (in  function  $g^1$)  \ldots  $\pi^k$ (in  function
-$g^k$),  then  the  over-approximation  requirement on  $\sigma_f$  is
-achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
-\cup \deltacall{f}{\pi^k}{g^k}$Rewrite, copied from earlier paper}.
-
-The expression $e_{\mainpgm}$  is a special case; we  assume it may be
-called externally with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$
-(denoted $\sigma_{\!all}$).  This is because any part of its value may
-be used by the environment---for printing the result, for instance.
-{\color {red} What about print?}
-For the running example, $\length$ has calls from $\main$ at $\pi_{10}$
-and a recursive call at $\pi_6$.
+To avoid analysing the body of a function for each call, we calculated
+the liveness  for the arguments and  the variables in  a function with
+respect to a symbolic demand  $\sigma$.  To get the actual liveness we
+calculate an over-approximation of the  actual demands made by all the
+calls by a {\em summary demand} and calculate the liveness at each GC
+point inside the  function based on this summary  demand.  Even though
+we  lose precision  due to  this, our  experiments show  that  this is
+acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
+  summary demand}  is calculated by taking  a union of  the demands at
+every call site of a function.
+
+Consider a function $g$ containing a  call to $f$ at a site $\pi$, say
+$\pi\!\!:\!(\LET~x   \leftarrow   (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.
+Let the  demand on $g$  be $\sigma_g$ and,  based on this  demand, the
+liveness of  $x$ at $\pi_i$ be  $\Lanv{i}{x}$.  By the  $\LET$ rule of
+Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
+$\Lanv{i}{x}$ to  the demand $\sigma_f$.  Assuming that  there are $k$
+call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
+(in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
+$\deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
+The distinguished function  \mainpgm\ is a special case.  We assume it
+is called through a  printing mechanism with demand $\sigma_\mainpgm =
+\{\acar,\acdr\}^\ast$ (denoted $\sigma_{\!all}$).
+
+For  the  running  example,   $\length$  has  calls  from  $\main$  at
+$\pi_{10}$ and a recursive call at $\pi_6$.
 %\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{length}{\pi_6}{\length}$.
+     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
 Filling in  the values gives:
-
-\begin{align*}
-\sigma_{\length}    &=
- \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\cbar\sigma_{\mathit{\length}}}
-\end{align*}
-
+\begin{eqnarray*}
+\sigma_{\length}    &=&
+ \sigma_{\!all}  ~\cup~ \acdr\Lf{\length}{1}{\cbar\sigma_{\mathit{\length}}}
+\end{eqnarray*}
 For some of the program points in \length\  and \main,  this gives
@@
 \begin{align*}
-\Lan{\pi_1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
-              \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{\pi_2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
-                  \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{\pi_5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
-\Lan{\pi_8} &= \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace 
+\Lan{1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
+              \py.\cbar\sigma_{\length}, \\ &~~~~~~~\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
+\Lan{2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
+                  \py.\cbar\sigma_{\length}, \\ &\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
+\Lan{5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
+\Lan{8} &= \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace 
 \end{align*}

### 2015-05-18 paper/paper.tex

@@
 %% $y$.
+
 Note that  liveness refers to  access using variables and  fields, and
@@
             $(\LET\ \pv\ $\leftarrow 0$ \IN
-            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\RETURN~\psi_2:\pv)$
+            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\SRETURN~\psi_2:\pv)$
             \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
@@
             $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\RETURN~\psi_3:\pz)$)))))))
+            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\SRETURN~\psi_3:\pz)$)))))))
         \end{uprogram}}
@@
           $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
-          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\RETURN~\psi_4:\pw)))))$
+          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\SRETURN~\psi_4:\pw)))))$
 \end{uprogram}}
@@
   labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
-    day}}---\SIF\ conditions  and \RETURN\ expressions.   We call such
+    day}}---\SIF\ conditions  and \SRETURN\ expressions.   We call such
 paths {\em e-paths}.   For the function \length, there  are three such
@@
 
-Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
-The  expression $(\RETURN~\psi:x)$  with a  demand $\sigma$  forces an
+Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
+The  expression $(\SRETURN~\psi:x)$  with a  demand $\sigma$  forces an
 evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
@@
 abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.
+
 %==============================================================
@@
 \\[1ex]
-\mathcal{L}((\RETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
+\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
 x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
@@
 \mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
-  of  an   expression  $e$.   The  cases  $\RETURN$   and  $\SIF$  are
+  of  an   expression  $e$.   The  cases  $\SRETURN$   and  $\SIF$  are
   straightforward, but note the liveness $x.\epsilon$ generated by the
@@
   placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
-
-% \end{itemize}
 
@@
 %------------------------------------------------------------%
+%\input{lazy-liveness}
 

### 2015-06-02 paper/paper.tex
comitting

@@
 \usepackage[T1]{fontenc}
-\usepackage[pdf]{pstricks}
+%\usepackage[pdf]{pstricks}
 \usepackage{epsfig}
@@
 \newcommand{\mycomment}[1]{}
-\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
+\definecolor{Myblue}{rgb}{.2,0,1}
 \newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
@@
 % of EIGHT authors. SIX appear on the 'first-page' (for formatting
-% reasons) and the remaining two appear in the \additionalauthors section.
+% reasons) and the remaining two appear in the \additionalauthors
+section.
 %
@@
 \begin{abstract}
-We  consider  two  ways  to  reduce the  memory  requirement  of  lazy
-functional  programming  languages. The  first  is  to modify  garbage
-collectors  to  preserve  only  {\em  live} data---a  subset  of  {\em
-reachable data}.  This results in an increase in the garbage reclaimed
-and  consequently in  fewer  garbage collections.   The  second is  to
-eagerly evaluate closures  that are guaranteed to be  evaluated in the
-future.
-
-The  main contribution  of  this  paper is  an  uniform framework  for
-obtaining the information for the optimizations mentioned above---{\em
-liveness}  and {\em strictness}  analysis. Using  a central  notion of
-demand,  we  formulate   a  context-sensitive  liveness  analysis  for
-structured  data and  prove  it  correct.  We  then present strictness
-as a dual of liveness. We use a  0-CFA-like
-conservative   approximation   to   annotate   each   allocation   and
-function-call program point  with a finite-state automaton---which the
-garbage-collector inspects to  curtail reachability during
-marking. The result of the strictness analysis, being necessarily
-finite,  can  be directly used to insert code to evaluate closures.
-
-%As a result,  fewer objects are marked (albeit  with a more expensive
-%marker) and then preserved (e.g.\ by a copy phase).
-
-Experiments  confirm  the  expected  performance benefits.  The  early
-evaluation   of   closures  decreases   the   memory   demands  of   a
-program. Further,  liveness-based collection results in  a increase in
+We consider  the problem of reducing  the memory required  to run lazy
+first-order functional programs.  To  do this, we analyze programs for
+liveness of heap  allocated data, i.e. data that  can possibly be used
+in the future.  The accompanying garbage collector is  modified to use
+the result of the analysis to preserve only {\em live} data---a subset
+of {\em reachable data}---during  garbage collection.  This results in
+an  increase in  the garbage  reclaimed and  consequently  reduces the
+memory footprints of programs.   While this technique has already been
+shown to yield benefits  for eager first-order languages, the presence
+of  closures   at  run-time  poses  additional   challenges  for  lazy
+languages.   Closures require  changes  both in  the earlier  liveness
+analysis for heap allocated memory  and also in the garbage collection
+scheme required to reclaim memory that is not live.
+
+We  formulate  a context-sensitive  liveness  analysis  and using  the
+results annotate each potential allocation point in the program with a
+set     of    deterministic    finite-state     automata(DFA).     The
+garbage-collector  inspects  the DFA  to  curtail reachability  during
+marking.  As a result, fewer  objects are marked (with a possibly more
+expensive  marker)  and  preserved  by the  copy  phase.   Experiments
+confirm  that  liveness-based  collection  results in  a  increase  in
 garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections, a decrease  in the memory size required  to run programs,
-and  reduced  overall  garbage  collection  time  for  a  majority  of
-programs.
+collections. On  our benchmark programs, we obtain  a decrease between
+x\% and y\% in the memory size required to run programs, when compared
+to reachability-based garbage collection.
 \end{abstract}
@@
 
-Functional  programming  languages  depend  on  garbage  collector  to
-efficiently  reclaim  unused   memory  allocated  by  programs  either
-explicitly  (for  example using  constructors)  or implicitly  (during
-creation    of    closures).     However,   empirical    studies    on
-Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
-Java~\cite{shaham02estimating}   programs  have  shown   that  garbage
-collectors leave uncollected a lot of memory objects that are not live
-(here {\em live}ness  means the object can potentially  be used by the
-program at  a later  stage).  Due  to lack of  a suitable  analysis to
-distinguish  reachable  objects  from  live objects,  current  garbage
-collectors conservatively approximate the  liveness of heap objects by
-their reachability from a set of memory locations called the {\em root
-  set\/}. This results in unnecessary memory retention.
-
-Programs written in languages having lazy semantics for evaluation put
-an further demands on  memory as they require heap allocated
-closures to be  carried from the point of  occurrence of an expression
-to the point of (first) real use of that expression.
-%% This   may   be   mitigated  by   strictness
-%% analysis~\cite{some-refs} that  can statically determine  the patterns
-%% of future usage  of an expression, thus allowing  some of the closures
-%% to be  evaluated at the point  of the occurrence  of the corresponding
-%% expression.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Functional programs extensively use dynamically allocated memory.  The
+allocation  is either  explicit (for  example, using  constructors) or
+implicit  (during creation  of  closures).  Programs  written in  lazy
+functional languages put additional  demands on memory as they require
+closures to  be carried  from the  point of creation  to the  point of
+evaluation.
+
+While  the runtime  system  of most  functional  languages includes  a
+garbage collector to efficiently  reclaim memory, empirical studies on
+Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
+Haskell~\cite{rojemo96lag}
+%and Java~\cite{shaham02estimating}
+programs have shown that  garbage collectors leave uncollected a large
+number of  memory objects that are  reachable but not  live (here {\em
+  live} means the  object can potentially be used by  the program at a
+later stage).  This results in unnecessary memory retention.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%
 %%% MOTIVATING EXAMPLE
@@
 \begin{pspicture}(0,0)(1,1)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
+%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
+subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
 \putnode{start}{origin}{0}{0}{}
@@
     \begin{tabular}{cc}
-\begin{boxedminipage}{.5\textwidth}
+\begin{boxedminipage}{.42\textwidth}
       {\sf
@@
 	  \begin{uprogram}
-	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
-	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
+          \UFL\
+\UNL{0} (\DEFINE\ (\length\  \pl)
+	  \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
+          (\PRIM\ 1\ (\CAR\  \pl))))
+           \UNL{0}
+	   \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
+	  \UNL{1}  (\SIF~(\NULLQ \ \lista)
 	       \listb
-	  \UNL{1}      (\CONS\ (\CAR\  \lista)
-%          \UNL{2}          (\CAR\  \lista)
-	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
+	  \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
+(\CDR\  \lista)
           \listb))))
           \UNL{0}
-	  \UNL{0} \hspace*{-.49\TAL} (\LET\ \pz  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
-	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
-	  \UNL{0} (\LET\ \py  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
-	  \UNL{1}   (\LET\ \pw $\leftarrow$ (\append\ \py\  \pz)\ \IN
-	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ \pw)))))
+          \UNL{0} (\LET\ \px\
+          $\leftarrow$\ (\CONS\ $5$
+%          \UNL{1}\hspace*{1.5cm}
+          (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
+	  \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
+	  \UNL{2}
+          (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
+          \UNL{3} (\SIF~(\NULLQ~(\CAR~(\CDR~\pz)))~$0$~$\pi$:\
+(\length\ \pz))))))
 	\end{uprogram}
@@
 &
-      \raisebox{-25mm}{\scalebox{.75}{
-	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
-      \psset{unit=1mm}
-      \psset{linewidth=.3mm}
-      \begin{pspicture}(0,-5)(70,60)
-	%\psframe(0,0)(73,60)
+\begin{minipage}{.45\textwidth}
+
+\hspace*{2cm}{\epsfig{file=motiv-example.eps, height=4cm}}
+
+\end{minipage}
+
+\\
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%       &
+%%       \raisebox{-25mm}{\scalebox{.75}{
+%% 	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
+%%       \psset{unit=1mm}
+%%       \psset{linewidth=.3mm}
+%%       \begin{pspicture}(0,-5)(70,60)
+%% 	%\psframe(0,0)(73,60)
+%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
-	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
-%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
-	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	%      fillcolor=white]{$\times$}}}
-        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
-	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
-	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-		  fillcolor=white]{$\times$}}}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
-	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
-	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
-	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
-	\ncline[nodesepB=-.5]{*->}{c2}{d}
-	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
-	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
-	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
-	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
-	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
-	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
-	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
-	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
+%% 	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
+%% 	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
+%% %
+%%      \putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
+%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
+%%
+%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
+%%      fillstyle=solid,
+%% 	%      fillcolor=white]{$\times$}}}
+%%
+%%  \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox
+%%{.63}{\nilfigure}}}
+%% %	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
+%% 	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
+%%
+%% \putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
+%% 	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
+%%
+%% \aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
+%%fillstyle=solid,
+%% 		  fillcolor=white]{$\times$}}}
+%% 	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% 	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
+%% 	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
+%% 	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
+%% 	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
+%% 	\ncline[nodesepB=-.5]{*->}{c2}{d}
+%% 	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
+%% 	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
+%%
+%% \aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
+%% fillstyle=solid,
+%% 	      fillcolor=white]{$\times$}}}
+%% 	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
+%%
+%% \nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
+%% 	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
+%%
+%% \aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
+%% fillstyle=solid,
+%% 	      fillcolor=white]{$\times$}}}
+%%
+%% \putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
+%%
+%%\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
+%%
+%%\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
+%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
-	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
-	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
-%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
-        %here
-        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
-        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
-
-%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
-%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%	      fillcolor=white]{$\times$}}} %and here
-
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
-	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
-	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
-	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-	      fillcolor=white]{$\times$}}}
+%% 	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
+%% 	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
+%% 	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
+%% %
+%%\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
+%% 	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
+%%
+%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
+%%         %here
+%%
+%%\putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
+%%
+%%\putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{
+%%\scalebox{.63}{\nilfigure}}}
+
+%% %
+%%\aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,
+%%fillstyle=solid,
+%% %	      fillcolor=white]{$\times$}}} %and here
+%% 	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
+%%
+%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h}
+%% %
+%%\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,
+%%fillstyle=solid,
+%% %	      fillcolor=white]{$\times$}}} %and here
+
+%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
+%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
+%% 	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
+%%
+%%\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
+%% 	      fillcolor=white]{$\times$}}}
+%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
-%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
-	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
-%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      \end{pspicture}}} \\
-      (a) Example program.&
+%% 	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
+%% %
+%%\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
+%% 	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
+%%
+%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
+%% %	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
+%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        %%
+%%       \end{pspicture}}} \\
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+      (a) Example program.&\hspace*{2cm}
       \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
@@
 
-\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
-%  \end{center}
-%%%%}\color{Myblue} end
+\caption{Example Program and its Memory Graph.{\color{red}Amey will combine this with the figure for syntax}}\label{fig:mot-example}
 \end{figure*}
-We propose the  use of liveness analysis of heap  cells and strictness
-analysis  of expressions  to  reduce the  memory  requirement of  lazy
-functional  programming languages.   We modify  garbage collectors  to
-preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
-This results in an increase  in the garbage reclaimed and consequently
-in fewer  garbage collections.  
-Using  a central notion  of demand,  we formulate  a context-sensitive
-liveness  analysis  and  prove   it  correct.   We  use  a  0-CFA-like
-conservative   approximation   to   annotate   each   allocation   and
-function-call program point  with a finite-state automaton---which the
-garbage-collector inspects to curtail reachability during marking.
+
+In this paper,  we propose the use of liveness  analysis of heap cells
+for garbage collection in a lazy first-order functional language.  The
+central notion in our analysis  is a generalization of liveness called
+{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
+expression.   The analysis  has two  parts.  First,  each  function is
+summarized in a context-sensitive manner as a {\em demand transformer}
+that transforms a {\em symbolic} demand  on its body to demands on its
+arguments.  This summary  is then used to step  through function calls
+during analysis.   Second, the concrete  demand on a function  body is
+obtained through a 0-CFA-like conservative approximation that combines
+the  demands on  all the  calls to  the function.   The result  of the
+analysis  is the annotation  of each  program point  with finite-state
+automata capturing  the liveness of variables at  the point. Depending
+on the program point where  garbage collection has been triggered, the
+garbage-collector consults  a set of automata  to curtail reachability
+during marking.  This results in  an increase in the garbage reclaimed
+and consequently in fewer garbage collections.
 
 While this idea has been shown to be effective~\cite{asati14lgc} for a
-first-order  eager  language,  a  straightforward  extension  of  this
-earlier  technique  is  not   possible  in  lazy  languages,  where  a
-heap-allocated   object  may  either   be  a   evaluated  data   or  a
-closure. Since liveness is brought about by evaluation of closures and
-given the  uncertainty in lazy  languages regarding the  program point
-where this  evaluation takes  place, this introduces  complications in
-the analysis itself.  Further,  since we want to extend liveness-based
-collection to closures, we  track closures  to their
-points of creation, from where we obtain their liveness. Since this is
-done  at  runtime,  this  requires considerable  modification  to  the
-earlier GC scheme.
-
-%% We  previously  proposed an  interprocedural  liveness analysis  to
-%% improve   garbage   collection   first   order  eager   subset   of
-%% Scheme~\cite{asati14lgc}.   The current paper  extends the  work to
-%% handle lazy evaluation semantics for the same language.
-
-Experiments  confirm  the expected  performance  benefits.  The  early
-evaluation   of   closures  decreases   the   memory   demands  of   a
-program. Further,  liveness-based collection results in  a increase in
-garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections,  a  (some quantification)  decrease  in  the memory  size
-required to run programs,  and reduced overall garbage collection time
-for a majority of programs.
+first-order {\em  eager} language, a straightforward  extension of the
+earlier  technique  is  not  possible  for  lazy  languages,  where  a
+heap-allocated  object may either  be an  unevaluated expression  or a
+closure. Since  data is  made live by  evaluation of closures,  and in
+lazy languages  the place in  the program where this  evaluation takes
+place cannot  be statically determined,  laziness complicates liveness
+analysis itself.  In  addition, apart from evaluated data,  we want to
+extend  liveness-based collection also  to closures---closures with no
+future use should also  be collected.  This involves tracking closures
+to  their points of  creation at  runtime from  where we  obtain their
+liveness.  This  requires significant  modification to the  earlier GC
+scheme.
+
+{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
+single
+  generation  copying  collector   confirm  the  expected  performance
+  benefits.   Liveness-based  collection  results  in an  increase  in
+  garbage reclaimed.   As a  consequence, there is  a decrease  in the
+  number of  collections and a  (some quantification) decrease  in the
+  memory size required  to run programs. Besides there  is a reduction
+  in the overall execution time for some programs.}
 
 %------------------------------------------------------------%
-\subsubsection{Motivating Example}
+\subsection{Motivating Example}
 \label{sec:motiv}
 
-Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
-$\pi$ of an expression $e$  denotes a program point.  During execution
-of  the program, it  represents the  instant of  time just  before the
-evaluation  of  $e$.  
-We  view the  heap  as a graph.  
-Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
-containing values.  Edges in the graph are {\em references} and
-emanate from {\em variables} or fields.
-Variable and field values may also be atomic values ($\NIL$, integers etc.)
-While it is convenient to box these in diagrams, our presented
-analysis treats them as non-heap values.
-
-Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
-The edges shown  by thick arrows are those  which are made live by the
-program. In addition,  assuming that
-the value of any reachable part of the program result may be explored
-or printed, the edges marked by thick dashed arrows are also live.
-A cell  is  marked and preserved  during  garbage collection,  only  if it  is
+Figure~\ref{fig:mot-example} shows an example program and the state of
+the heap at the program point $\pi$, i.e. just before the  evaluation
+of $(\length\ \pz)$.  The heap is represented by a graph in which
+a node  either represents atomic values ($\NIL$,  integers etc.), or
+a \CONS\  cell  containing $\CAR$  and  $\CDR$  fields, or a chunk of
+memory representing an unevaluated
+value  also called a  {\em closure}  (represented by  hatched boxes).
+Edges in the graph are  {\em references} and emanate from variables or
+fields.  The figure shows the lists \px\ and \pz\ partially evaluated.
+The evaluation was triggered by the test (\NULLQ~(\CAR~(\CDR~\pz))) in
+the \SIF\ expression.
+
+The edges shown by thick arrows  are those which are live at $\pi$.  A
+cell is marked and preserved  during garbage collection only if it is
 reachable from the  root set through a path of  live edges.  All other
-cells can be  reclaimed.  Thus if  a garbage collection takes
-% place   at    the   time   represented   by   the    heap   shown   in
-at $\pi$ with the    heap   shown   in
-Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
-will be marked and preserved.
-
-\subsubsection{Organization of the paper}
-Section~\ref{sec:defs} gives
-the syntax and semantics of
-the language used to illustrate our analysis along with basic concepts
-and    notations.
+cells can be  reclaimed.  Thus if a garbage  collection takes place at
+$\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b), only the
+cells referenced by $\pz$,  $(\CDR~ \pz)$ and (the cells constituting)
+the closure  referenced by $(\CDR~(\CDR~\pz))$  need to be  marked and
+preserved.  In  contrast, a  reachability-based  collector would  have
+preserved all cells.
+
+{\color {Myblue}Rewrite at the end: The  specific contributions of the
+paper
+  are: The formulation of a  liveness analysis for a lazy language and
+  the proof of its correctness, the undecidability of the grammar that
+  results from the analysis and  its approximation by DFAs, the design
+  of a  garbage collector that can  use the DFAs to  retain live cells
+  that  also include  closures, and  empirical results  that  show the
+  effectiveness of liveness-based garbage collection.}
+
+\subsection{Organization of the paper}
+
+Section~\ref{sec:defs} introduces the syntax and semantics of the
+language used to illustrate our
+analysis.
 % Section~\ref{sec:operational} describes the operational semantics of
 % the language.
-Liveness analysis is  described in Section~\ref{sec:liveness} followed
-by  a  sketch  of  a  correctness proof  relative  to  a  non-standard
+Liveness analysis for this language is  described in
+Section~\ref{sec:liveness}. This section also has by  a  sketch  of  a
+correctness proof  relative  to  a  non-standard
 semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
@@
 \label{sec:defs}
-We let $x$, $y$, $z$ range over variables, $f$ over user-functions and
-$p$    over    primitive    functions   ($\CONS$,    $\PRIM$    etc.).
+%%We let $x$, $y$, $z$ range over variables, $f$ over user-functions
+and
+%% $p$    over    primitive    functions   ($\CONS$,    $\PRIM$ etc.).
 Figure~\ref{fig:lang-syntax} describes the  syntax of our language. It
-is a first  order language with lazy semantics  and restricts programs
+is a first order language with lazy semantics. Programs are restricted
 to        be        in        Administrative        Normal        Form
 (ANF)~\cite{chakravarty03perspective}  where all actual  parameters to
-functions   are   variables.   This   restriction   does  not   affect
-expressibility  (and we ignore  it in  examples when  inessential). In
-addition, this  form also explicates the creation  of closures through
-$\LET$s which are lazy.  A consequence  of lazy $\LET$ is that $x$ can
-also occur  in $s$  in $\LET\,\, x  \leftarrow s\,\, \IN\,\,  e$. This
-enables  creation of  graph-like structures  in a  pure  language. The
-restriction of \LET\ to a single definition is for ease of exposition;
-generalization  to  multiple  definitions  does  not  pose  conceptual
-difficulties.  We  further restrict each  variable in a program  to be
-distinct, so  that no scope shadowing  occurs---this simplifies proofs
-of soundness.  The $\SRETURN$  and \SIF\ expressions trigger evaluation
-of  closures.
-
-The body  of a function ${\mathit  f}$ will be  denoted as $e_{\mathit
-  f}$. for  analysis purposes it is convenient  to regard $e_\mainpgm$
-as part of a function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$
-as in  C\@.  We  write $\pi\!:\!e$ to  associate the label  $\pi$ (not
-part  of the  language  syntax)  with the  program  point just  before
-expression $e$.
+functions  are  variables.  While  this  restriction  does not  affect
+expressibility,  this form  has  the benefit  of  making explicit  the
+creation  of closures through  the $\LET$  construct.  $\LET$s  in our
+language  are lazy;  in the  expression $\LET\,\,  x  \leftarrow s\,\,
+\IN\,\, e$, $x$ may occur  in $s$. This enables creation of graph-like
+structures in  a pure language. The  restriction of \LET\  to a single
+definition  is  for ease  of  exposition---generalization to  multiple
+definitions  does  not add conceptual  difficulties.  We  further
+restrict each variable  in a program to be distinct,  so that no scope
+shadowing   occurs---this  simplifies   proofs   of  soundness.    The
+$\SRETURN$ and \SIF\ expressions trigger evaluation of closures.
+
+The body  of a function ${\mathit  f}$ is denoted  as $e_{\mathit f}$.
+We   shall   extend   this   convention   to   the   main   expression
+$e_\mainpgm$. We assume that each program has a distinguished function
+\mainpgm\      with      the      definition     $(\DEFINE\      ({\tt
+  \mainpgm})\  e_\mainpgm)$ and  the execution  of the  program starts
+with the call to \mainpgm.  We write $\pi\!:\!e$ to associate the label
+$\pi$ (not  part of the language  syntax) with the  program point just
+before expression $e$.
 
 % \paragraph{Inlining.}
-{\color{Myblue}In spite of the ANF  restrictions it is still possible to
-  inline non-recursive functions (a fact we use to prove the safety of
-  liveness  analysis).  A user-function  call $(\LET\,\,  x \leftarrow
-  (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$  to a function defined (after
-  renaming  its  formals  and  locals  to be  disjoint  from  existing
-  variables) by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by
-  a  sequence of  $\LET$'s  of the  form $z_i\leftarrow  (\ID\,\,y_i)$
-  followed by the body $e_f$ but with its $(\SRETURN\,\,w)$ expressions
-  replaced by $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}
+{ \color {Myblue} Relook after proof: In spite of the  ANF restrictions
+it
+  is still possible  to inline non-recursive functions (a  fact we use
+  to  prove the safety  of liveness  analysis).  A  user-function call
+  $(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n)  \,\, \IN\,\, e)$ to a
+  function  defined  (after renaming  its  formals  and  locals to  be
+  disjoint         from         existing         variables)         by
+  $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by a sequence of
+  $\LET$'s of  the form  $z_i\leftarrow (\ID\,\,y_i)$ followed  by the
+  body $e_f$  but with  its $(\SRETURN\,\,w)$ expressions  replaced by
+  $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}
 
@@
     \hspace{7em} \mbox{\em --- program}\\
-    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
+    \mathit{df} \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
     e) 
-    \hspace{1.7em} \mbox{\em --- function definition} \\
+    \hspace{2.2em} \ \ \ \ \mbox{\em --- function definition} \\
 e \in \mathit{Expr} & ::= &
 \left\{\begin{array}{ll@{\hspace{4em}}l}
-       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
-       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
-       (\SRETURN\,\, x) && \mbox{\em --- return from function}
+       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
+       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
+binding} \\
+       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
     \end{array}\right. \\
@@
        (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\
-       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
+       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester
+and generic arithmetic} \\
 %        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\
-       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
-%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
+       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)}
+\\
+%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic}
+\\
        \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)}
@@
 \subsubsection{Semantics}
-We now give  a small-step semantics for our  language.  We shall later
-augment this semantics so that it enables us to reason about liveness.
-We start with the domains used by the semantics:
+We now give  a small-step semantics for our  language.
+We first specify the domains used by the semantics:
 \[
-\begin{array}{rlcl@{\hspace{2em}}l}
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\ 
+\begin{array}{rlcl@{\hspace{0.5em}}l}
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
+\mbox{-- Environment} \\
 v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times
@@
   Closures}\\
-d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values/Closures} \\ 
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
+d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \&
+  Closures} \\
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{--
+Heap}
 \end{array}
@@
 empty  list  $\NIL$,  or  a  \CONS\  cell  with  possibly  unevaluated
-constituents) or a {\em closure}.  A  closure is a pair $(s, \rho)$ in
-which $s$  is an unevaluated application  with the values  of its free
-variables given by $\rho$.
-
-The semantics of expressions (and applications) are given by
-the  judgment form $\rho,  \stk, \heap,  e \rightarrow  \rho', \stk',
-\heap', e'$.  Here \stk\ is a stack of continuation frames represented
-as  a list  in the  description of  the semantics.   Each continuation
-frame  is of the  form $(\rho,  x, e)$,  where the  location of  $x$ is
-updated when the currently evaluated expression reaches
-WHNF,  and  $e$  is  the  next  expression  to  be  evaluated  in  the
-environment  $\rho$.   The  start  state is  $(\{\},[(\{\},  \mainpgm,
-  \HALT)], \{\}, e_\mainpgm)$, and the program terminates successfully
-on  reaching   the  halt   state  $(\rho,[\,],  \heap,   \HALT)$  with
-$\rho(\mainpgm)$  containing  the final  result.  Note  that since  we
-assume that our language is not statically typed, a program with other
-than syntactic errors (including those that could have been eliminated
-using a  static type system) lead  to {\em stuck  states} i.e.  states
-not covered by any rules in the small-step semantics.
-
-The  notation $[\myvec{x} \mapsto  \myvec{\ell}]$ represents
-an  environment that maps  variables $x_1,  \ldots, x_n$  to locations
-$\ell_1,  \ldots,  \ell_n$.   The  notation  $\heap[\ell  \mapsto  d]$
+constituents) or  a {\em  closure}.  A closure  is a pair  $\langle s,
+\rho\rangle$  in which  $s$  is an  unevaluated  application, and  the
+values of its free variables is given by $\rho$.
+
+The  semantics  of   expressions  (and  applications\footnote{In  most
+  contexts, we  shall use  the term 'expression' and the notation $e$  for both  expressions and
+  applications.}) are  given by transitions  of the form  $\rho, \stk,
+\heap, e \rightarrow \rho', \stk', \heap', e'$.  Here \stk\ is a stack
+of  continuation  frames.  Each  continuation  frame  is  of the  form
+$(\rho, x, e)$, signifying that the  location of $x$ has to be updated
+with the  value of the currently  evaluating expression, and $e$  is to be
+evaluated  next  in  the  environment  $\rho$.   The  start  state  is
+$([\;]_\rho,([\;]_\rho,  \ans,   (\print~\ans)):[\;]_{S}  ,  [\;]_{H},
+(\mainpgm)$.  In  this, the initial  environment and the  initial heap
+are  denoted  by  $[\;]_\rho$  and  $[\;]_H$, and  the  initial  stack
+consists  of  a  single  continuation   frame  in  which  \ans\  is  a
+distinguished  variable  that  would  be  updated with  the  value  of
+(\mainpgm).  In  addition, \print\ is a function  modelling a printing
+mechanism---a standard  runtime support assumption  for lazy languages
+that prints the  value of (\mainpgm).  Notice the  use of the operator
+$:$ for adding elements to the top of the stack.
+
+%% In absence of a static type-checking mechanism, programs with other
+%% than syntactic  errors lead to states  not covered by  any rules in
+%% the small-step semantics.
+The notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
+environment that maps variables $x_i$
+%$x_1,  \ldots, x_n$
+to locations $\ell_i$
+%$\ell_1,  \ldots,  \ell_n$,
+and  $\heap[\ell :=  d]$
 indicates the  updation of  a heap \heap\  at $\ell$ with  $d$.  $\rho
@@
 
-As an example, consider the rules related to the application $(+~x~y)$
-involving a strict binary operator.  When both $x$ and $y$ are already
-in WHNF  ({\sc prim-whnf}), the  heap is updated  with the sum  of the
-arguments  and  the  continuation  gives  the next  expression  to  be
-evaluated.   If  either  of  the   arguments  is  not  in  WHNF  ({\sc
-  prim-1-clo} and {\sc prim-2-clo}), it  is sent for evaluation, and a
-continuation marks that the evaluation  of $(+~x~y)$ has to be resumed
-after the argument is evaluated.
+The small-step semantics  is shown in Figure~\ref{fig:lang-semantics}.
+As  a sample,  consider  the  rules  for  $(+~x~y)$, an  application
+involving a binary operator strict in both arguments.   When
+both $x$  and $y$ are already  in WHNF ({\sc prim-whnf}),  the heap is
+updated with the  sum of the arguments and  the next expression to be
+evaluated is picked from the continuation on the stack. If either of the arguments is not in
+WHNF  ({\sc  prim-1-clo}  and   {\sc  prim-2-clo}),  it  is  sent  for
+evaluation, and a continuation  marks that the evaluation of $(+~x~y)$
+has to be resumed after the argument is evaluated.
 
@@
           & $\rho, (\rho', x, e)\!:\!S, H, \kappa
-  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
-\hline
-          & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
-$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
+  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
+\\
+\hline
+          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
+\rightsquigarrow
+$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
 \hline
 $H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
-(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto d_1], e$      &  \sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } (e, \rho')$ & $\rho, S, H, (\CAR~x)  \rightsquigarrow
-\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
+\sc{car-whnf} \\ 
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
+\rightsquigarrow
+{\color {red}\rho\oplus\rho'}, (\rho, x, (\CAR~x))\!:\!S, H, s$      &  \sc{car-clo}
+\\
 \hline
 $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
-$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
-\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
-\hline
-$H(\rho(y)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
-\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
-\hline
-\shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
-[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
-\hline
-$\ell$ is a new location& \shortstack[c]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
-  \rightsquigarrow$ \\
-$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
-  \ell)\oplus\lfloor \rho\rfloor_{FV(S)\setminus\{x\}})], e$} & \sc{let} \\ 
+ & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
+$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
+\sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
+\rightsquigarrow
+{\color {red}\rho\oplus\rho'}, (\rho, x, (+~x~y))\!:\!S, H, s$      &
+\sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
+\rightsquigarrow
+{\color {red}\rho\oplus\rho'}, (\rho, x, (+~x~y))\!:\!S, H, s$      &
+\sc{prim-2-clo} \\
+\hline
+{$\mathit{f}~\mbox{defined as}$
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
+(f~\myvec{x})  \rightsquigarrow
+[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
+\sc{funcall} \\
+\hline
+$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+  \rightsquigarrow$
+$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
+    \lfloor\rho\rfloor_{FV(s){\color {red}\setminus\{x\}}}  \oplus [x \mapsto
+  \ell]\rangle], e$} &
+\sc{let} \\
 \hline
@@
 \hline
-$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H,
-  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
-$\rho'\oplus\rho, (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
-\hline
-\shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
-(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
-\hline
-$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\SRETURN~x)
-  \rightsquigarrow$ \\
-$\rho'\oplus\rho,~ (\rho, x, (\SRETURN~x))\!:\!S, H,  e$} & \sc{return-coo} \\
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
+  (\SIF~x~e_1~e_2)   \rightsquigarrow
+{\color {red} \rho \oplus \rho'}, (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$} &
+\sc{if-clo} \\
+\hline
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
+e)\!:\!S, H,
+(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
+\sc{return-whnf}\\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
+  \rightsquigarrow$
+${\color {red}\rho\oplus\rho'},~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
+\sc{return-clo} \\
 \hline
 \end{tabular}
-\caption{A small-step semantics for the language}\label{fig:lang-semantics}
+\caption{A small-step semantics for the language. {\color {red} Can we
+    replace  $\rho   \oplus  \rho'$  by  $\rho'$?  Can   we  drop  the
+    $\setminus\{x\}$         in          the         {\sc         let}
+    rule?}\label{fig:lang-semantics}}
 \end{center}
@@
 Heap-allocated data needs a richer model of liveness which talks about
-liveness  of  pointers  including  fields  of  $\CONS$  cells.   Using
-$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
-the  liveness  of the  structure  reachable  from  a variable  can  be
-represented by a set of {\em access paths} i.e.  prefix-closed strings
-from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
-liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
-\acdr\acdr\acar\}$, then future computations  can only refer up to the
-second and  third members of $x$.   A {\em liveness  environment} is a
-mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
-expressed  as a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
-x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
-  \mapsto\{\epsilon,\acdr,\acdr\acdr\},           y\mapsto\{\epsilon\},
+liveness  of references.   Using  $\acar$, $\acdr$  to represent  access
+using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
+reachable from a  variable can be represented by a  set of {\em access
+  paths} i.e.  prefix-closed  strings from $\{\acar,\acdr\}^\ast$.  As
+an  example,  if $x$  is  a  list  with liveness  $\{\epsilon,  \acdr,
+\acdr\acar,  \acdr\acdr, \acdr\acdr\acar\}$, then  future computations
+can only  refer up  to the second  and third  members of $x$.   A {\em
+  liveness  environment} is  a mapping  from variables  to  subsets of
+$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
+writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$ instead of
+$[x     \mapsto\{\epsilon,\acdr,\acdr\acdr\},    y\mapsto\{\epsilon\},
   z\mapsto\{\}]$.    In  this   notation,  $x   \mapsto  \{\epsilon\}$
 represents access using $x$ itself and $x \mapsto \{\}$ indicates that
-$x$ is  dead.  We associate liveness environments  with program points
-(denoted  by  $\pi$ and  $\psi$).   The  liveness  at $\pi:e$  is  the
-liveness just before executing $e$.
+$x$ is dead.  We  associate liveness environments with program points.
+The liveness at $\pi:e$ is the liveness just before executing $e$.
 
 A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
-represents the  future use of variables, demands  represent the future
-use of the  values of expressions. The demand on  an expression $e$ is
+represents the future use of variables, a demand represents the future
+use of the value of an  expression. The demand on an expression $e$ is
 again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
@@
 $(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
-y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
-$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
-\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
+y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
+of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
+\rbrace$  paths of  the result  of  $(\CDR~y)$\footnote{The classical
   analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
@@
   of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
-  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
-  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
+  becomes the demand on $y+z$.   This, in turn, generates the liveness
+  $\lbrace y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
 %% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
-%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
+%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\
+%% field of the cons cell pointed by
 %% $y$.
@@
 locations themselves).
-\begin{figure*}[t!]
-  \begin{tabular}{cc}
-    \begin{boxedminipage}{.45\textwidth}
-        \small
-        \renewcommand{\arraystretch}{1}{
-          \begin{uprogram}
-            \UNL{1} (\DEFINE\ (\length~\xl)
-            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
-            $(\LET\ \pv\ $\leftarrow 0$ \IN
-            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\SRETURN~\psi_2:\pv)$
-            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
-            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\SRETURN~\psi_3:\pz)$)))))))
-        \end{uprogram}}
-        \renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $(\DEFINE\ (\main)$
-          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
-	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
-          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
-          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\SRETURN~\psi_4:\pw)))))$
-\end{uprogram}}
-        %%} 
-    \end{boxedminipage}
-
-    &
-
-    \begin{boxedminipage}{.51\textwidth}
-
-        \small
-        \begin{tabular}{lc}
-        $\mathcal{L}(e_{\length},\sigma,\Lfonly) = $ 
-            \begin{tabular}{lc}            
-              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
-              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
-                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
-              $~~~~~~~~~~~~~~\;\;
-                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
-                \rbrace~\rbrack$ \\
-            \end{tabular} \\
-            $\cal{P_{\mbox{\tiny \length}}} = $
-              \begin{tabular}{lc}
-                $~[
-                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
-                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
-                  \rbrace, $\\
-$ ~~~\pi_3 \mapsto \lbrace ep_2
-                  \rbrace $, $ \pi_4 \mapsto \lbrace ep_2
-                  \rbrace$, $\pi_5 \mapsto \lbrace  ep_3 \rbrace$, \\
-                $~~~\pi_6 \mapsto \lbrace ep_3 \rbrace,  \pi_7 \mapsto
-                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3 \rbrace~\rbrack$
-              \end{tabular}\\
-              $\mathcal{L}(e_{\main},\sigma,\Lfonly) = $ 
-              \begin{tabular}{lc}            
-                $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
-                  $]$
-              \end{tabular} \\
-              $\cal{P_{\mbox{\tiny \main}}} = $
-                \begin{tabular}{lc}
-                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
-                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
-                \end{tabular}
-        \end{tabular}
-
-    \end{boxedminipage} 
-    %% \end{ tabular}}
-  \end{tabular}
-%\end{picture}
-\kern -3ex
-\vspace{0.5in}
-\caption{Example Program and its liveness maps.{\color{red} Fix the
-    formatting of the table on the right}}\label{fig:mot-example2}    
-\end{figure*}
-
 We use $\sigma$  to range over demands, $\alpha$  to range over access
@@
 creation of its closure, not necessarily its evaluation.
-To see the  consequences of this, consider an analysis  of the body of
-$\length$ with a demand $\sigma$.  Clearly, this is also the demand on
-the return value $z$ giving the liveness at the program point $\pi_8$
-as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated backwards
-as in traditional liveness analysis.
+To see the  consequences of this, consider
+Figure~\cite{fig:mot-example2} which shows an analysis  of the body of
+a function $\length$ with a demand $\sigma$.  Clearly, this is also the
+demand on the return value $z$ giving the liveness at the program point
+$\pi_8$ as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated
+backwards as in traditional liveness analysis.
 
@@
   point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
-  the beginning  of the program $\pi_1$ (since  we do not  kill liveness).
+  the beginning  of the program $\pi_1$ (since  we do not  kill
+liveness).
 \end{enumerate}
@@
 evaluation\footnote{Program   points  which  trigger   evaluation  are
-  labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
-    day}}---\SIF\ conditions  and \SRETURN\ expressions.   We call such
+  labelled  with   $\psi$  }---\SIF\ conditions  and \SRETURN\
+expressions.   We call such
 paths {\em e-paths}.   For the function \length, there  are three such
@@
 \pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
+
+\begin{figure*}[t!]
+  \begin{tabular}{cc}
+    \begin{minipage}{.45\textwidth}
+        \small
+        \renewcommand{\arraystretch}{1}{
+          \begin{uprogram}
+            \UNL{1} (\DEFINE\ (\length~\xl)
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
+(\NULLQ~\xl) \IN
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
+$\psi_1\!\!:\,$ \px
+            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
+            $(\LET\ \pv\ $\leftarrow 0$ \IN
+            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
+(\SRETURN~\psi_2:\pv)$
+            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
+$\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
+$\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
+            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
+            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
+(\SRETURN~\psi_3:\pz)$)))))))
+        \end{uprogram}}
+        \renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} $(\DEFINE\ (\main)$
+          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots
+\IN
+	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\
+$\leftarrow$  \ldots \IN
+          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
+          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
+          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\,
+(\SRETURN~\psi_4:\pw)))))$
+\end{uprogram}}
+        %%}
+    \end{minipage}
+
+    &
+
+    \begin{minipage}{.51\textwidth}
+
+        \small
+\begin{eqnarray*}
+        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
+              [\;ep_1 \mapsto \lbrace \px.\sigma_{\length},
+\xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace,\\
+ &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma_{\length},
+                \py.\cbar\sigma_{\length},
+\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},\\
+&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
+\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
+                \rbrace~\rbrack \\
+            \cal{P_{\mbox{\tiny \length}}} &=&
+                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
+                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
+                  \rbrace, \\
+&&\;\;  \pi_3 \mapsto \lbrace ep_2
+                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
+                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
+ &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
+\pi_7 \mapsto
+                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
+\rbrace~\rbrack\\
+              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
+                [\;
+                  ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}},
+\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots\rbrace
+                  ]\\
+              \cal{P_{\mbox{\tiny \main}}} &=&
+                  \lbrack~\pi_6 \mapsto \lbrace ep_4 \rbrace,  \
+                   \pi_7 \mapsto \lbrace ep_4 \rbrace,  \
+                  \pi_8 \mapsto \lbrace ep_4 \rbrace~\rbrack
+\end{eqnarray*}
+    \end{minipage} \\
+(a)&
+(b)
+    %% \end{ tabular}}
+  \end{tabular}
+%\end{picture}
+\kern -3ex
+%\vspace{0.5in}
+\caption{(a) Example program and (b) its liveness
+maps.}\label{fig:mot-example2}
+\end{figure*}
 
@@
 e-path.  A second map $\cal{P}_{\mathit  f}$ maps any program point in
-$\mathit f$ to
-the set of  e-paths sharing the point.  If a  program point happens to
-fall on  more than one such  e-path, then the  liveness environment of
-the  program  point  is   the  variable-wise  union  of  the  liveness
-environments of the individual e-paths.  In the example program, since
-${\mathcal P}_{\length}(\pi_1) = \lbrace ep_1, ep_2, ep_3\rbrace$, the
-liveness at $\pi_1$ is ${\cal  M}(ep_1) \cup {\cal M}(ep_2) \cup {\cal
-  M}(ep_3)$, where ${\cal M} = \mathcal{L}(e_f, \sigma, \Lfonly)$.  In
-the  sequel we shall  only define  $\mathcal{L}$.  The  formulation of
-$\cal{P}$ is easy and we do not elaborate it any further.
+$\mathit f$  to the set  of e-paths sharing  the point.  If  a program
+point happens to fall on more  than one such e-path, then the liveness
+environment of  the program  point is the  variable-wise union  of the
+liveness  environments  of the  individual  e-paths.   In the  example
+program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
+ep_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(ep_1)  \cup {\cal
+  M}(ep_2) \cup  {\cal M}(ep_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
+\sigma, \Lfonly)$.  In the  sequel we shall only define $\mathcal{L}$.
+The formulation  of $\cal{P}$ is easy  and we do not  elaborate it any
+further.
 
@@
 evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
-  \{x.\sigma\}] $  needs to  be created.  The  map for  the expression
+  \{x.\sigma\}] $  needs to  be created, where the $ep$ is a fresh
+evaluation path terminating in $\psi$.  The  map for  the expression
 $(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
@@
 
-%==============================================================
-\begin{figure*}[t]  
-\begin{eqnarray*}
-\mathit{ref\/}(\kappa,\sigma,\Lfonly)
-          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
-\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
-\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
-x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
-  new e-path terminating with $\psi$}\\
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
-        \begin{array}{l l}
-                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
-        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
-        \lbrack\lbrace \mathit{ep} \mapsto  \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
-        \emptyset_{\cal{M}?}  & \mbox{otherwise}
-                 \end{array} \\
-&&\mbox{where $\mathit{ep}$ is a new e-path
-  terminating with $\psi$}\\
-\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
-\mbox{where}~\mathcal{M} =
-\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in \mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
-&& \makebox[0mm]{\hspace*{7cm}   
- $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
-           x_2\, \ldots\, x_n)/x]))$,} \\  
- \makebox[0mm]{\hspace*{8cm} where
-     $x_1\, \ldots\, x_n$ are the variables in $s$}
-\end{eqnarray*}
-\begin{minipage}{0.85\textwidth}
-\infrule[live-define]
-        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
-           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { d_1 \ldots d_k \len \Lfonly
-\\ \makebox[0mm]{where
-     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
-\end{minipage}
-  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
-\end{figure*}
-
  The liveness  of {\LET} has  an elegant formulation.   Recollect that
@@
  arbitrary number of ``unrollings'' of the definition\footnote{The $*$
-   in the definition of ${\mathit f}$ represents  non-deterministic choice,
-   but can be treated as a constant for purposes of analysis.}.  We pass to the
- demand transformer $\Lfonly_{\mathit f}$  of this function the demand
- arising out  of the  liveness of  $x$ at the  beginning of  $e$. This
- gives the liveness of the variables of $s$.
+   in  the definition  of ${\mathit  f}$  represents non-deterministic
+   choice,  but  can  be  treated   as  a  constant  for  purposes  of
+   analysis.}.   We pass to  the demand  transformer $\Lfonly_{\mathit
+   f}$ of this function the demand  arising out of the liveness of $x$
+ at the beginning of $e$. This  gives the liveness of the variables of
+ $s$.
 
@@
 
-\section{Computing liveness and its encoding as a table}\label{sec:computing}
+\section{Computing liveness and its encoding as a
+table}\label{sec:computing}
 Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
@@
 
+%==============================================================
+\begin{figure*}[t]
+\begin{eqnarray*}
+\mathit{ref\/}(\kappa,\sigma,\Lfonly)
+          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
+$\NIL$}\\
+\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
+          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha
+\mid \acdr\alpha \in \sigma\} \\
+\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
+\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne
+\emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
+          &=&    \begin{array}{l l}
+                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
+%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
+          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
+               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
+               \end{array}
+%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
+\\[1ex]
+\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep}
+\mapsto
+x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
+  new e-path terminating with $\psi$}\\
+\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
+        \begin{array}{l l}
+                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
+        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
+        \lbrack\lbrace \mathit{ep} \mapsto
+\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
+        \emptyset_{\cal{M}?}  & \mbox{otherwise}
+                 \end{array} \\
+&&\mbox{where $\mathit{ep}$ is a new e-path
+  terminating with $\psi$}\\
+\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
+\mbox{where}~\mathcal{M} =
+\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in
+\mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
+&& \makebox[0mm]{\hspace*{7cm}
+ $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
+           x_2\, \ldots\, x_n)/x]))$,} \\
+ \makebox[0mm]{\hspace*{8cm} where
+     $x_1\, \ldots\, x_n$ are the variables in $s$}
+\end{eqnarray*}
+\begin{minipage}{0.85\textwidth}
+\infrule[live-define]
+        {\mathcal{L}(e_f,\sigma,\Lfonly) =
+           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
+              \mbox{ for each $f$ and $\sigma$}
+        }
+        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
+\\ \makebox[0mm]{where
+     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
+\ldots \mathit{df_k}$}}
+\end{minipage}
+  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
+\end{figure*}
+
 \begin{proposition}
@@
 determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
-consider $\Lanv{1}{x_i}$,  the  liveness of    the $i^{\mbox{\footnotesize  th}}$
+consider $\Lanv{1}{x_i}$,  the  liveness of    the
+$i^{\mbox{\footnotesize  th}}$
 parameter $x_i$ at the  program point at the  beginning of
@@
 \end{eqnarray}
-where \Df{f}{i} are sets of  strings over the alphabet mentioned above.   Substituting the guessed form in
+where \Df{f}{i} are sets of  strings over the alphabet mentioned above.
+  Substituting the guessed form in
 the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
@@
 Note that this equation can also be viewed as a CFG with \{\acdr,
-\cbar\} as the terminal symbols and \Df{\length}{1} as the sole non-terminal.
+\cbar\} as the terminal symbols and \Df{\length}{1} as the sole
+non-terminal.
 
@@
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
+     \deltacall{\length}{\pi_{11}}{\main}  \cup
+\deltacall{\length}{\pi_6}{\length}$.
 Filling in  the values gives:
@@
 \sigma_{\length}    &=&
- \sigma_{\!all}  ~\cup~ \acdr\Lf{\length}{1}{\cbar\sigma_{\mathit{\length}}}
+ \sigma_{\!all}  ~\cup~
+\acdr\Lf{\length}{1}{\cbar\sigma_{\mathit{\length}}}
 \end{eqnarray*}
@@
 \begin{align*}
-\Lan{1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
-              \py.\cbar\sigma_{\length}, \\ &~~~~~~~\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
-                  \py.\cbar\sigma_{\length}, \\ &\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
-\Lan{8} &= \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace 
+\Lan{1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length},
+\pz.\sigma_{\length},\
+              \py.\cbar\sigma_{\length}, \\
+&~~~~~~~\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
+\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
+\Lan{2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length},
+\pz.\sigma_{\length},\
+                  \py.\cbar\sigma_{\length}, \\
+&\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
+\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
+\Lan{5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
+\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
+\Lan{8} &= \lbrace \pw.\sigma_{\mathit{\!all}},
+\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace
 \end{align*}
@@
 \item For each function $\mathit{f}$ (including $\mainpgm$ for
-  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages? No, we define it only in the implementation section as of now.}
+  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages?
+No, we define it only in the implementation section as of now.}
 an equation defining
@@
 
-\subsection{Solving liveness equations---the grammar interpretation}
+\subsection{Solving liveness equations---the grammar
+interpretation}\label{sec:grammar-formulation}
 The  equations above can now be re-interpreted as a
 context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal for
+\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal
+for
 a variable  $X$ occurring on  the LHS of the  equations generated
@@
 \comment{Pick up a richer equation or more than one liveness equations}
-Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
+Other equations can be converted  similarly.     The  language
+generated  by  \var{\Lanv{i}{x}},
 denoted $\mathscr{L}(\var{\Lanv{i}{x}})
@@
 Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
-\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
+\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let
+$\mathscr{L}(\var{\Lanv{i}{x}})
 \stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists
@@
 undecidable. \comment{Rewrite it to say that this is a CFG with 3
-  unrestricted rules and the membership for even this grammar is undecidable}
+  unrestricted rules and the membership for even this grammar is
+undecidable}
 \comment{State the undecidability of the decision problem above.}
 
-  We circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
+  We circumvent the problem by over-approximating the CFG generated by
+the analysis to  {\em
   strongly  regular}  CFGs which  have   easy
@@
 non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
-Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
+Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq
+{\mathscr
   L}_{G'}(\Lv)$.    This  is   required  for the
@@
 $\sigma$ then the NFA reaches a final state throught this path
-otherwise it does not reach a final state.\comment{Need to rewrite this}
+otherwise it does not reach a final state.\comment{Need to rewrite
+this}
 %------------------------------------------------------------%
@@
 transformation are
-  first converted  into  NFAs.  The        algorithm         described        in
+  first converted  into  NFAs.  The        algorithm         described
+      in
 Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
@@
 \lang{\nt}$, where {\nt} accepts forward paths only.
-\comment{Rewrite the algorithm to handle \cbar and update the example automata}
+\comment{Rewrite the algorithm to handle \cbar and update the example
+automata}
 Thus  \nt\  can  be  used  to  check
@@
 Unlike eager languages where GC happens only at \CONS~ points or at
-function calls, every \LET~ point is a potential GC point. The number of 
-automata that needs to be stored is reduced by sharing automata along all 
+function calls, every \LET~ point is a potential GC point. The number
+of
+automata that needs to be stored is reduced by sharing automata along
+all
 the program points along an {\em E-path}.
@@
 derefernced during later execution.
-% A reference copied using LGC could become part of a closure in a later GC
-% and copy wrong locations. Avoiding it by setting non-live references to null after every GC.
+% A reference copied using LGC could become part of a closure in a
+later GC
+% and copy wrong locations. Avoiding it by setting non-live references
+to null after every GC.
 % Proof for the correctness.
@@
 presence  of  a print  stack.  In  lazy  languages the  evaluation  of
-closures is  driven by the print  function\comment{Refer to SPJ's book}.  Hence, during  a GC there
+closures is  driven by the print  function\comment{Refer to SPJ's
+book}.  Hence, during  a GC there
 will be heap  references that will be present on  the stack which also
@@
 We   provide   an   intuition    about   the   correctness   of   this
-algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts about the print stack during a
+algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts
+about the print stack during a
 GC,
@@
 \item The top of the stack always contains a reference to a closure.
-\item All other references on the print stack point to atomic values or cons cells, it can never contain a
-reference to a closure. The components of the cons cell can contain reference to closures.
-\item If a reference on the print stack is a cons cell, either its car part or its cdr part will also be 
+\item All other references on the print stack point to atomic values or
+cons cells, it can never contain a
+reference to a closure. The components of the cons cell can contain
+reference to closures.
+\item If a reference on the print stack is a cons cell, either its car
+part or its cdr part will also be
 present on the stack.
@@
 Sketch of the proof
-Since all closures are copied fully and atomic values are copied, only cons cells need to be handled. 
+Since all closures are copied fully and atomic values are copied, only
+cons cells need to be handled.
 Consider the following cases,
@@
 While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
+at  the latest  program  point traversed  in the function's body.
+Thus, if  LGC
 is invoked  at the program point   $\pi$,  it would  use  the
@@
 current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
+stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
+call-chain,
 the DFAs at the point of call to  ${\mathit g}$ would be used.
@@
 The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
-  fibheap} increases  initially, comes  down before the  heap becomes gets
-full and then  continues to remains low.  This results in  a single garbage
+sources and have been  manually converted to  ANF.  All  graphs  except
+ {\tt  fibheap}  show  fewer  garbage
+collector  invocations for  LGC. The  number  of reachable  cells in
+{\tt
+  fibheap} increases  initially, comes  down before the  heap becomes
+gets
+full and then  continues to remains low.  This results in  a single
+garbage
 collection in both RGC and LGC.  The graphs also show the precision of
@@
   then} {\bf skip}\\
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
+\noindent \hspace*{.25cm}  {\bf else} {\bf if}
+$\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR =
+\mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\
+\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$
+{\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa},
+\mathsf{dfa}(q,1),x.\CDR)$\\
 \noindent \hspace*{.35cm}{\bf return} $y$
@@
 
-\begin{figure}[p]
-\begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
- &
-{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-\end{tabular}
- \caption{Memory  usage of  programs. The blue and the red curves
-  indicate the number of cons cells in the active semi-space for RGC and
-  LGC respectively. The black curve represents the number of reachable
-  cells and the grey curve represents the number of cells that are
-  actually live (of which liveness analysis does  a static
-  approximation). x-axis is the time measured in number of cons-cells
-  allocated. y-axis is the number of cons-cells.}
-\label{fig:memory-usage} \figrule
-\end{figure}
+%% \begin{figure}[p]
+%% \begin{tabular}{@{}c@{}c@{}}
+%% \hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
+%% &
+%% {\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
+%% \\
+%% \hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
+%%  &
+%% {\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
+%% \\
+%% \hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
+%% &
+%% {\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
+%% \\
+%% \hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
+%% &
+%% {\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
+%% \end{tabular}
+%%  \caption{Memory  usage of  programs. The blue and the red curves
+%%   indicate the number of cons cells in the active semi-space for RGC
+%% and
+%%   LGC respectively. The black curve represents the number of reachable
+%%   cells and the grey curve represents the number of cells that are
+%%   actually live (of which liveness analysis does  a static
+%%   approximation). x-axis is the time measured in number of cons-cells
+%%   allocated. y-axis is the number of cons-cells.}
+%% \label{fig:memory-usage} \figrule
+%% \end{figure}
 The  increased effectiveness  of LGC  over RGC  is also  shown  in the
@@
 LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to run
+time and requires a smaller heap size ({\em MinHeap}) for program to
+run
 in comparison with  RGC.
@@
 
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
+\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r |
+r | r | r | r | r |}
+\hline
+    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\#
+Touched}
   & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
@@
 {Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
+RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC &
+LGC \\
+\hline
+\hline
+    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5
+&.007 & .034 \\
+    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147
+&588 &.045 & .144 \\
+    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970
+ &4 &.086 & .075 \\
+     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507
+&171878 &76618 &1.406 & .9  \\
+    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558
+&78720 &0 &.006 & .014  \\
+    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225
+&307092 &206729 &82112 &464.902 & 14.124  \\
+    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150
+&212653 &1954 &.356 & .217 \\
+    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579
+&501093 &521826 &39465 &70.314 & 24.811 \\
+    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95
+&.093 &2.49  \\
+%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????
+\\
 %    {\sf  nboyer} & &  & &   & &  & & &  \\
@@
 
-%% Since garbage collection is  effectively asynchronous to the allocator
-%% thread, one might  worry as to how {\em  robust} our measurements are.
-%% For example,  while LGC would,  in general, collect more  garbage than
-%% RGC in the same heap state,  is it possible that on some programs, LGC
-%% might do  a larger number  of collections? We  prove a result  to show
-%% that   this  cannot   happen.    This  result   applies  to   classical
-%% mark-and-sweep and copying garbage  collectors and, we hypothesize, to
+%% Since garbage collection is  effectively asynchronous to the
+allocator
+%% thread, one might  worry as to how {\em  robust} our measurements
+are.
+%% For example,  while LGC would,  in general, collect more  garbage
+than
+%% RGC in the same heap state,  is it possible that on some programs,
+LGC
+%% might do  a larger number  of collections? We  prove a result  to
+show
+%% that   this  cannot   happen.    This  result   applies  to
+classical
+%% mark-and-sweep and copying garbage  collectors and, we hypothesize,
+to
 %% generational collectors too.
@@
 %% \begin{lemma}
-%% For the  same mutator,  a liveness-based collector  can never  do more
+%% For the  same mutator,  a liveness-based collector  can never  do
+more
 %% garbage collections than a reachability-based collector.
@@
 
-%% \begin{proof}Assume, as before,  that time is measured in terms of the number of
-%%   cons cells allocated. Let us run  two copies of the mutator with the
-%%   two    different   collectors    but    with   synchronous    memory
+%% \begin{proof}Assume, as before,  that time is measured in terms of
+the number of
+%%   cons cells allocated. Let us run  two copies of the mutator with
+the
+%%   two    different   collectors    but    with   synchronous
+memory
 %%   allocations.   The   garbage    collections,   however,   happen
@@
 
-%%   To prove the lemma, it is enough to show the truth of the following 
+%%   To prove the lemma, it is enough to show the truth of the
+following
 %% statement: After
-%%  every  LGC invocation,  the count of LGC invocations   is  no greater
+%%  every  LGC invocation,  the count of LGC invocations   is  no
+greater
 %% than  RGC invocations.   The  base case  holds  since the  first
-%% invocations of both GCs happen at the same time.  Assume the statement
-%% to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-%% reachable cells, its heap would contain no more cells than RGC heap at
-%% the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-%% LGC, or LGC and RGC are both invoked next at the same time.  In either
+%% invocations of both GCs happen at the same time.  Assume the
+statement
+%% to be true after $n$ invocations of LGC.  Since LGC copies a subset
+of
+%% reachable cells, its heap would contain no more cells than RGC heap
+at
+%% the end  of the $n$th invocation.   Thus either RGC  is invoked next
+before
+%% LGC, or LGC and RGC are both invoked next at the same time.  In
+either
 %% case, the statement holds after $n+1$ invocations of LGC.
@@
 
-\section{Related Work} Previous attempts to  increase  the space efficiency
+\section{Related Work} Previous attempts to  increase  the space
+efficiency
 of functional programs by additional  reclamation of memory fall in two
@@
 
-\begin{figure}[t]
-\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
- \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
-\label{fig:memory-usage-lambda} \figrule
-\end{figure}
+%% \begin{figure}[t]
+%% \centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
+%%  \caption{Memory  usage and garbage collection pattern of  {\tt
+%% lambda}.}
+%% \label{fig:memory-usage-lambda} \figrule
+%% \end{figure}
 
@@
 the  number   of  GCs,  more  garbage  collected   per  invocation.  A
-significant benefit of LGC is that programs can run in smaller memory when
+significant benefit of LGC is that programs can run in smaller memory
+when
 compared to RGC. This is potentially useful in situations where memory
@@
 %Mention full LGC
-%Mention k-liveness, use length example to show 1-liveness is not sufficient.
+%Mention k-liveness, use length example to show 1-liveness is not
+sufficient.
 %Mention mixed mode GC,
-% - doing RGC most of the times and switching to LGC in case RGC fails to collect any garbage
-% - doing LGC most of the times but regularly checking the reachability of the heap and 
+% - doing RGC most of the times and switching to LGC in case RGC fails
+to collect any garbage
+% - doing LGC most of the times but regularly checking the reachability
+of the heap and
 % switching to RGC if LGC almost equals RGC

### 2015-06-02 paper/paper.tex
added algo back
fix for compile error

--- 2015-05-18 paper/paper.tex
@@
-\documentclass[9pt]{sigplanconf}
+\documentclass[10pt]{sigplanconf}
 \usepackage{etex}
@@
 \usepackage{tabularx}
-\usepackage [plain,noend,noline,boxed]{algorithm2e}
+\usepackage {algorithmic}
 \usepackage{amssymb}
@@
 %\usepackage{program}
-\usepackage{algorithmic}
-
 \input{fh_defs}
@@
 \definecolor{Myblue}{rgb}{0.0,0.0,0.5}
-\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
+\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
+    #1 \medskip \hrule \medskip}}}
 \newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
@@
 efficiently  reclaim  unused  memory   allocated  by  programs  either
-explicitly  (for  example using  constructors)  or implicitly  (during
-creation    of    closures).     However,   empirical    studies    on
+explicitly (for example using constructors) or implicitly (for example
+for      closures).       However,      empirical      studies      on
 Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
@@
 program at  a later  stage).  Due  to lack of  a suitable  analysis to
-distinguish  reachable  objects  from  live objects,  current  garbage
+distinguish  reachable  objects  from live  objects.  current  garbage
 collectors conservatively approximate the  liveness of heap objects by
 their reachability from a set of memory locations called the {\em root
-  set\/}. This results in unnecessary memory retention.
-
-Programs written in languages having lazy semantics for evaluation put
-an further demands on  memory as they require heap allocated
-closures to be  carried from the point of  occurrence of an expression
-to the point of (first) real use of that expression.
-%% This   may   be   mitigated  by   strictness
-%% analysis~\cite{some-refs} that  can statically determine  the patterns
-%% of future usage  of an expression, thus allowing  some of the closures
-%% to be  evaluated at the point  of the occurrence  of the corresponding
-%% expression.
+  set\/} resulting in unnecessary memory usage.
+
+Languages having lazy  semantics for evaluation put  further demand on
+memory as they  require heap allocated closure to be  carried from the
+point of occurrence  of an expression to the point  of (first) real use
+of  that expression.  Strictness analysis~\cite{some-refs}  statically
+determines  the  patterns  of  future usage  of  an  expression,  thus
+allowing some  of the  closures to  be evaluated at  the point  of the
+occurrence of the corresponding expression.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 This results in an increase  in the garbage reclaimed and consequently
-in fewer  garbage collections.  
-Using  a central notion  of demand,  we formulate  a context-sensitive
-liveness  analysis  and  prove   it  correct.   We  use  a  0-CFA-like
-conservative   approximation   to   annotate   each   allocation   and
+in fewer  garbage collections.  Further, we eagerly evaluate closures
+that are guaranteed to be evaluated in the future (strict closures).
+
+The main contribution  of this paper is an uniform  framework for {\em
+  liveness} and {\em strictness}  analysis to enable the optimizations
+mentioned  above. Using  a central  notion of  demand, we  formulate a
+context-sensitive liveness  analysis for structured data  and prove it
+correct.  We then  present strictness as a dual of  liveness. We use a
+0-CFA-like conservative approximation to  annotate each allocation and
 function-call program point with  a finite-state automaton---which the
-garbage-collector inspects to curtail reachability during marking.
-
-While this idea has been shown to be effective~\cite{asati14lgc} for a
-first-order  eager  language,  a  straightforward  extension  of  this
-earlier  technique  is  not   possible  in  lazy  languages,  where  a
-heap-allocated   object  may  either   be  a   evaluated  data   or  a
-closure. Since liveness is brought about by evaluation of closures and
-given the  uncertainty in lazy  languages regarding the  program point
-where this  evaluation takes  place, this introduces  complications in
-the analysis itself.  Further,  since we want to extend liveness-based
-collection to closures, we  track closures  to their
-points of creation, from where we obtain their liveness. Since this is
-done  at  runtime,  this  requires considerable  modification  to  the
-earlier GC scheme.
-
-%% We  previously  proposed an  interprocedural  liveness analysis  to
-%% improve   garbage   collection   first   order  eager   subset   of
-%% Scheme~\cite{asati14lgc}.   The current paper  extends the  work to
-%% handle lazy evaluation semantics for the same language.
+garbage-collector inspects to curtail reachability during marking. The
+result of  the strictness analysis,  being necessarily finite,  can be
+directly used to insert code to evaluate closures.
+
+We previously proposed an interprocedural liveness analysis to improve
+garbage collection first order eager subset of Scheme~\cite{cc-paper}.
+The current paper extends the work to handle lazy evaluation semantics
+for the same language.
 
@@
 garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections,  a  (some quantification)  decrease  in  the memory  size
-required to run programs,  and reduced overall garbage collection time
-for a majority of programs.
+collections, a decrease  in the memory size required  to run programs,
+and  reduced  overall  garbage  collection  time  for  a  majority  of
+programs.
 
@@
 and    notations.
-% Section~\ref{sec:operational} describes the operational semantics of
-% the language.
-Liveness analysis is  described in Section~\ref{sec:liveness} followed
-by  a  sketch  of  a  correctness proof  relative  to  a  non-standard
-semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
-as   finite-state  automata.    Section~\ref{sec:experiments}  reports
-experimental  results  and Section~\ref{sec:lgc-always-better}  proves
-that a liveness based collector  can never do more garbage collections
-than a reachability based collector.
+% Section~\ref{sec:operational}   describes    the
+% operational semantics  of the  language.
+Liveness analysis is described in
+Section~\ref{sec:liveness} followed by a sketch of a correctness
+proof relative to a non-standard semantics.
+Section~\ref{sec:computing} shows how to encode liveness as
+finite-state automata.
+Section~\ref{sec:experiments}   reports   experimental   results and
+Section~\ref{sec:lgc-always-better} proves that a liveness based
+collector can never do more garbage collections than a reachability
+based collector. 
 % and Section~\ref{sec:conclusion} concludes the paper.
@@
 
-The body  of a function ${\mathit  f}$ will be  denoted as $e_{\mathit
-  f}$. for  analysis purposes it is convenient  to regard $e_\mainpgm$
-as part of a function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$
-as in  C\@.  We  write $\pi\!:\!e$ to  associate the label  $\pi$ (not
-part  of the  language  syntax)  with the  program  point just  before
-expression $e$.
+The body of the program is the expression denoted by $e_\mainpgm$; for
+analysis purposes it is convenient to regard $e_\mainpgm$ as part of a
+function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
+We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
+language syntax) with the program point just before expression $e$.
 
@@
 
-Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
-Since all data objects are boxed, we model an environment as a mapping
-from the set  of variables of the program  $\mathit{Var}$ to locations
-in the  heap.  These  either contain  a value in  WHNF (a  number, the
-empty  list  $\NIL$,  or  a  \CONS\  cell  with  possibly  unevaluated
-constituents) or a {\em closure}.  A  closure is a pair $(s, \rho)$ in
-which $s$  is an unevaluated application  with the values  of its free
-variables given by $\rho$.
+Here $\mathit{Loc}$ is a countable set  of locations in the heap.  Since
+all data objects are boxed, we  model an environment as a mapping from
+the set of variables of the program $\mathit{Var}$ to locations in the
+heap.  These either contain a value  in WHNF (a number, the empty list
+$\NIL$, or  a \CONS\ cell with possibly unevaluated constituents)
+or a {\em closure}.   A closure is a pair $(s, \rho)$  in which $s$ is
+an  unevaluated application  with  the values  of  its free  variables
+given by $\rho$.
 
@@
 A variable is {\em live} if  there is a possibility of its value being
-used in  future computations  and dead if  it is definitely  not used.
+used  in future computation  and dead  if it  is definitely  not used.
 Heap-allocated data needs a richer model of liveness which talks about
@@
 mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
-expressed  as a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
-x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
+expressed as  a set, for example by  writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, 
+y.\epsilon\}$               instead               of               $[x
   \mapsto\{\epsilon,\acdr,\acdr\acdr\},     y\mapsto\{\epsilon\},
@@
 represents access using $x$ itself and $x \mapsto \{\}$ indicates that
-$x$ is  dead.  We associate liveness environments  with program points
-(denoted  by  $\pi$ and  $\psi$).   The  liveness  at $\pi:e$  is  the
-liveness just before executing $e$.
-
-A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
-represents the  future use of variables, demands  represent the future
-use of the  values of expressions. The demand on  an expression $e$ is
-again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
-which the context of $e$ may explore of $e$'s result.  To see the need
-for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
-(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context  of  this
-expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
-  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
-$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
-$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
-y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
-$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
-\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
-  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
-  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
-  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
-  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
-  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
-  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
+$x$ is dead.  We associate  a liveness environment with program points
+$\pi$.  The liveness at $\pi:e$  is the liveness just before executing
+$e$.
+
+A  notion related  to  liveness is  {\em  demand}.  The  demand on  an
+expression  $e$  is again  a  set  of  access paths---that  subset  of
+$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
+result.   To see  the need  for demands,  consider the  let expression
+$\pi:(\LET~x\leftarrow (\CDR~y)~\IN~ \pi':\SRETURN~x)$. Assume that the
+context of  this expression produces the liveness  $[x \mapsto \lbrace
+  \epsilon, \acar  \rbrace]$ at $\pi'$.   Due to the  \LET\ definition
+which  binds $x$  to $(\CDR~y)$,  the liveness  of $x$  at  $\pi'$ now
+becomes  the  demand on  $(\CDR~y)$.   This,  in  turn, generates  the
+liveness $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar \rbrace$ at $\pi$
+(the set of $y$-rooted accesses required to explore $\lbrace \epsilon,
+\acar  \rbrace$  paths  of  the  result  of  $(\CDR~y)$)\footnote{The
+  classical analogue  is strong liveness  analysis, where $y$  and $z$
+  have  a  liveness  of  $\lbrace\epsilon\rbrace$  at  the  entry  $n:
+  x:=y+z$,    if   and    only   if    $x$   has    a    liveness   of
+  $\lbrace\epsilon\rbrace$ at  exit of  $n$.  In our  terminology, the
+  liveness of $x$  at exit from $n$ becomes the  demand on $y+z$. This
+  generates the  liveness $\lbrace y.\epsilon,  z.\epsilon \rbrace$ at
+  the entry of $n$.}.
 %% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
@@
 %% $y$.
-
 Note that  liveness refers to  access using variables and  fields, and
@@
 locations themselves).
-\begin{figure*}[t!]
-  \begin{tabular}{cc}
-    \begin{boxedminipage}{.45\textwidth}
-        \small
-        \renewcommand{\arraystretch}{1}{
-          \begin{uprogram}
-            \UNL{1} (\DEFINE\ (\length~\xl)
-            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
-            $(\LET\ \pv\ $\leftarrow 0$ \IN
-            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\SRETURN~\psi_2:\pv)$
-            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
-            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\SRETURN~\psi_3:\pz)$)))))))
-        \end{uprogram}}
-        \renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $(\DEFINE\ (\main)$
-          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
-	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
-          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
-          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
-          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\SRETURN~\psi_4:\pw)))))$
-\end{uprogram}}
-        %%} 
-    \end{boxedminipage}
-
-    &
-
-    \begin{boxedminipage}{.51\textwidth}
-
-        \small
-        \begin{tabular}{lc}
-        $\mathcal{L}(e_{\length},\sigma,\Lfonly) = $ 
-            \begin{tabular}{lc}            
-              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
-              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
-                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
-              $~~~~~~~~~~~~~~\;\;
-                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
-                \rbrace~\rbrack$ \\
-            \end{tabular} \\
-            $\cal{P_{\mbox{\tiny \length}}} = $
-              \begin{tabular}{lc}
-                $~[
-                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
-                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
-                  \rbrace, $\\
-$ ~~~\pi_3 \mapsto \lbrace ep_2
-                  \rbrace $, $ \pi_4 \mapsto \lbrace ep_2
-                  \rbrace$, $\pi_5 \mapsto \lbrace  ep_3 \rbrace$, \\
-                $~~~\pi_6 \mapsto \lbrace ep_3 \rbrace,  \pi_7 \mapsto
-                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3 \rbrace~\rbrack$
-              \end{tabular}\\
-              $\mathcal{L}(e_{\main},\sigma,\Lfonly) = $ 
-              \begin{tabular}{lc}            
-                $[$
-                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
-                  $]$
-              \end{tabular} \\
-              $\cal{P_{\mbox{\tiny \main}}} = $
-                \begin{tabular}{lc}
-                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
-                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
-                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
-                \end{tabular}
-        \end{tabular}
-
-    \end{boxedminipage} 
-    %% \end{ tabular}}
-  \end{tabular}
-%\end{picture}
-\kern -3ex
-\vspace{0.5in}
-\caption{Example Program and its liveness maps.{\color{red} Fix the
-    formatting of the table on the right}}\label{fig:mot-example2}    
-\end{figure*}
 
@@
 
+%% Finally, we use $\Lfonly$  to range over {\em demand transformers}.
+%% These transform  demands on  a function call  into demands  for its
+%% formal  parameters:   given  a   user  function  $f$,   defined  by
+%% $(\DEFINE\ (f\  x_1\ \ldots\ x_n)\  \ e_f)$ and called  with demand
+%% $\sigma$,  then $\Lf{f}{i}{\sigma}$  is  the liveness  of $x_i$  at
+%% $e_f$.
+
 \subsection{Liveness Analysis for lazy languages}
@@
  of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
- evaluated  even for  the null  demand. This  would require  an access
- using $x$ resulting in  $x.\epsilon$ being live.  A non-null demand of
- $\sigma$ on (\CDR~$x$), is transformed to the liveness $\{x.\epsilon,
- x.\acdr\sigma\}$.  In an opposite  sense, the demand of $\acdr\sigma$
- on  (\CONS~$y$~$z$) is  transformed to  the demand  $\sigma$  on $z$.
- Note that \CONS\ does not,  by itself, dereference its arguments. The
- rules for (\PRIM~x~y) and  (\NULLQ~x) are on similar lines. Constants
- generate no liveness.
+ evaluated even for the null  demand, and this would require an access
+ using  $x$.   A  non-null   demand  of  $\sigma$  on  (\CDR~$x$),  is
+ transformed  to  the  liveness  $\{x.\epsilon, x.\acdr\sigma\}$.   In  an
+ opposite  sense, the  demand of  $\acdr\sigma$ on  (\CONS~$y$~$z$) is
+ transformed to  the demand  $\sigma$ on $z$.   Note that  \CONS\ does
+ not, by itself, dereference  its arguments. The rules for (\PRIM~x~y)
+ and (\NULLQ~x) are on similar lines. Constants generate no liveness.
 
@@
 $\Lfonly$  that  represents the  summaries  of  all  functions in  the
-program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
-specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
-result  of a  call to  $f$  is transformed  into the  liveness of  its
-parameters at the  beginning of the call.  $\Lfonly$  is determined by
-the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
-({\sc live-define}).  This rule  describes the fixed-point property to
-be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
-for each  function in  the program  should be the  same as  the demand
-transformation  calculated from  their  bodies.  As  we  shall see  in
-Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar and  the least  solution satisfying the  rule is  the language
-generated by this  grammar.  The least solution also  ensures the safe
-collection of the greatest amount of garbage.
+program.  $\Lfonly_f$  (the  component  of $\Lfonly$  for  a  specific
+function $f$),  expresses how the demand  $\sigma$ on the  result of a
+call to $f$ is transformed into  the liveness of its parameters at the
+beginning of the call.  $\Lfonly$  is determined by the judgement form
+$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
+This  rule  describes the  fixed-point  property  to  be satisfied  by
+$\Lfonly$, namely, the demand transformation assumed for each function
+in  the  program should  be  the  same  as the  demand  transformation
+calculated    from   their    bodies.     As   we    shall   see    in
+Section~\ref{sec:grammar-formulation},  the least  solution satisfying
+the  rule  is obtained  by  converting  the  rule into  an  equivalent
+grammar.  The least solution also  ensures that the greatest amount of
+garbage is collected.
+
+%%   Analogously to classical liveness being computed as a solution of
+%% dataflow   equations,  we   require,  via   inference   rule  ({\sc
+%% live-define}), $\Lfonly$ to  satisfy the fixed-point property that:
+%% when we  assume $\Lfonly$ to  be the family of  demand transformers
+%% for the program then the  calculated liveness of each function body
+%% $\mathcal{L}(e_f,\sigma,\Lfonly)$    agrees   with    the   assumed
+%% $\Lfonly$.   As  usual,  there  are  often  multiple  solutions  to
+%% $\Lfonly$; all are  safe (see Section~\ref{sec:correctness}) but we
+%% prefer  the least  one  as  giving the  least  liveness subject  to
+%% safety---and hence greatest amount of garbage collected.
 
 We next  describe the function $\mathcal{L}$  that propagates liveness
-across expressions.  Note that,  unlike eager languages, the syntactic
-occurrence  of an expression  in a  program denotes  the point  of 
-creation of its closure, not necessarily its evaluation.
-To see the  consequences of this, consider an analysis  of the body of
-$\length$ with a demand $\sigma$.  Clearly, this is also the demand on
-the return value $z$ giving the liveness at the program point $\pi_8$
-as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated backwards
-as in traditional liveness analysis.
-
-Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
-\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
-generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
-the following observations:
-\begin{enumerate}
-\item While the liveness of \py\  should be killed at $\pi_6$ since it
-  is  beyond the scope  of the  definition of  \py, we  do not  do so.
-  However, this  does not cause any  imprecision in the way  of the GC
-  marking more cells as live,  since, during execution, \py\ is yet to
-  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
-  consulted by the garbage collector at this point.
-\item In an eager language, \py\ would not have been live at $\pi_8$,
-  a program point  beyond its last use.  In  a lazy language, however,
-  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
-  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
-  evaluation  actually takes  place while  evaluating $(\SRETURN~\pz)$.
-  We therefore  record  the liveness of $\py$ at the program  
-  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
-  the beginning  of the program $\pi_1$ (since  we do not  kill liveness).
-\end{enumerate}
-
-We thus regard  the body of the function as a  tree and consider paths
-from  the  root   of  this  tree  to  program   points  which  trigger
-evaluation\footnote{Program   points  which  trigger   evaluation  are
-  labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
-    day}}---\SIF\ conditions  and \SRETURN\ expressions.   We call such
-paths {\em e-paths}.   For the function \length, there  are three such
-paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
-\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
-\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
-
-Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
-analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
-computes a  map that takes a  e-path in ${\mathit f}$  and returns the
-liveness environment  arising out of  all uses of variables  along the
-e-path.  A  second map $\cal{P}_{\mathit  f}$ maps any program  point in
-$\mathit f$ to
-the set of  e-paths sharing the point.  If a  program point happens to
-fall on  more than one such  e-path, then the  liveness environment of
-the  program  point  is   the  variable-wise  union  of  the  liveness
-environments of the individual e-paths.  In the example program, since
-${\mathcal P}_{\length}(\pi_1) = \lbrace ep_1, ep_2, ep_3\rbrace$, the
-liveness at $\pi_1$ is ${\cal  M}(ep_1) \cup {\cal M}(ep_2) \cup {\cal
-  M}(ep_3)$, where ${\cal M} = \mathcal{L}(e_f, \sigma, \Lfonly)$.  In
-the  sequel we shall  only define  $\mathcal{L}$.  The  formulation of
+across  expressions.   This brings  us  to  the  second way  in  which
+liveness analysis for lazy languages differ from eager languages---the
+point  where  an  expression   gets  evaluated  cannot  be  statically
+determined.  To  deal with this, we introduce  flow insensitivity into
+our  analysis.
+
+\begin{figure*}[t!]
+%%\begin{picture}(100,130)(0,-90)
+\begin{picture}(100,130)(0, -50)
+  \begin{tabular}{cc}
+    \begin{boxedminipage}{.51\textwidth}
+      {\sf
+        \small
+        \renewcommand{\arraystretch}{1}{
+          \begin{uprogram}
+            \UNL{1} (\DEFINE\ (\length~\xl)
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
+            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\SRETURN_{ep_1}~0)$
+            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\SRETURN_{ep_2}~(1~+~\py))$)))))
+        \end{uprogram}}
+        \renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
+	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
+          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\SRETURN$_{ep_4}$\ \pw)))))))
+\end{uprogram}}
+        } 
+    \end{boxedminipage}
+    & 
+    \begin{boxedminipage}{.49\textwidth}
+      {\sf
+        \small
+        \begin{tabular}{lc}
+        $\cal{M_{\mbox{\tiny length}}} = $ 
+            \begin{tabular}{lc}            
+              $[$
+              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
+              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
+              $ep_1 \mapsto \lbrace \rbrace , $\\
+              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
+              $]$
+            \end{tabular} \\
+            $\cal{P_{\mbox{\tiny length}}} = [$
+              \begin{tabular}{lc}
+                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
+                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
+                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
+                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
+                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
+                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
+                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
+                $]$
+              \end{tabular}\\
+              $\cal{M_{\mbox{\tiny main}}} = $ 
+              \begin{tabular}{lc}            
+                $[$
+                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
+                  $]$
+              \end{tabular} \\
+              $\cal{P_{\mbox{\tiny main}}} = [$
+                \begin{tabular}{lc}
+                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
+                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
+                  $ \ldots  $ \
+                  $]$  
+                \end{tabular}
+        \end{tabular}
+      }
+    \end{boxedminipage}
+    \\
+    (a) Example program.&
+    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+        (b) {\em e-path} to liveness environment map \\
+        (c) Prog pt to {\em e-paths } map
+    \end{tabular}}
+  \end{tabular}
+\end{picture}
+\kern -3ex
+\vspace{0.5in}
+\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
+%% >>>>>>> 67d5a8ebe45703d9726af91b752e9d74152ae455
+\end{figure*}
+
+% Push this to a later point
+
+%% Unlike eager languages, a variable may be bound to an expression
+%%   that is not fully evaluated. i.e. a closure.
+
+%% To  deal with the  first issue,  we apply  the results  of liveness
+%% analysis only to fully evaluated data.  If the marking phase during
+%% garbage collection leads to a closure, and the closure is not dead,
+%% we  mark  the  entire   reachable  memory  under  the  closure  for
+%% collection.   A  better scheme  would  be  to  garbage collect  the
+%% closure based on  the liveness at the point  of its creation.  This
+%% is sound because our  language is statically scoped.  However, this
+%% is  work  in progress,  and  the  method  described in  this  paper
+%% switches over to reachability while collecting closures.
+
+%% %%%%%%%%%%%%%%%%%%%%%%%%
+
+%% \renewcommand{\arraystretch}{1}{
+%% 	  \begin{uprogram}
+%% 	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
+%% 	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
+%% 	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
+%%           \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
+%%           $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
+%% 	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
+%%           $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
+%% 	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
+%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\SRETURN_1~\pu)$))
+%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\SRETURN_2~\py)$)))))
+%% 	\end{uprogram}}
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% \medskip
+
+Consider  an analysis of  the body  of $\pf$  with a  demand $\sigma$.
+Clearly, this is also the demand on the return values $u$ and $y$
+giving the liveness  at  the  program points  $\pi_7$  and $\pi_8$  as
+$\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The  liveness at
+$\pi_7$  and  $\pi_8$  are  propagated  backwards  as  in  traditional
+liveness analysis.
+
+Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
+\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
+turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
+$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
+not in the scope of \pu.  However, this does not cause any imprecision
+in the way of the GC marking  more cells as live, since \pu\ is yet to
+come into existence at $\pi_7$  and its liveness will not be consulted
+by the garbage collector at this point.
+
+In an eager language, \pw\ would not have been live after $\pi_7$, the
+point of  its last  use. In a  lazy language, however,  the definition
+$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
+$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
+takes place while evaluating  $(\SRETURN~\pu)$. Thus it is necessary to
+record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
+every program  point from $\pi_8$ up  to the beginning  of the program
+(since  we do  not kill  liveness).  We  thus regard  the body  of the
+function as  a tree and consider paths  from the root of  this tree to
+program  points  which  trigger  a evaluation---\SIF\  conditions  and
+\SRETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
+program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
+\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
+\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.
+
+The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
+path to a common liveness environment, and a second map $\cal{P}$ maps
+a program  point to the set of  e-paths on which  the point lies.
+If a  program  point happens to fall  on more than one  such path, then
+the  liveness environment of  the program  point is  the variable-wise
+union of  the liveness environments  of the individual paths.   In the
+example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
+${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
+EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
+our propagation of liveness is flow-insensitive but path-sensitive.
+In the sequel
+we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
 $\cal{P}$ is easy and we do not elaborate it any further.
@@
 Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
-The  expression $(\SRETURN~\psi:x)$  with a  demand $\sigma$  forces an
-evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
-  \{x.\sigma\}] $  needs to  be created.  The  map for  the expression
-$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
-$e_2$. In addition, since the condition also triggers an evaluation of
-$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
-added  to the  union.  Also  notice a  consequences  of laziness---the
-entire  expression including  the condition  is not  evaluated  if the
-demand on it is $\emptyset$.  This results in the empty map, which, by
-abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.
-
+The  expression $\pi:(\SRETURN~x)$  with  a demand  $\sigma$ forces  an
+evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
+  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
+expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
+  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
+triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
+is created and added to  the union.  Also notice that the consequences
+of  laziness---the entire  expression including  the condition  is not
+evaluated if  the demand  on it is  $\emptyset$.  This results  in the
+empty map denoted by $\emptyset_{\cal{M}}$.
 %==============================================================
@@
 \\[1ex]
-\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep} \mapsto
-x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
-  new e-path terminating with $\psi$}\\
+\mathcal{L}((\SRETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
 \mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
@@
         \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
-        \lbrack\lbrace \mathit{ep} \mapsto  \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
+        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
         \emptyset_{\cal{M}?}  & \mbox{otherwise}
                  \end{array} \\
-&&\mbox{where $\mathit{ep}$ is a new e-path
-  terminating with $\psi$}\\
+ \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
 \mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
-\mbox{where}~\mathcal{M} =
-\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in \mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
-&& \makebox[0mm]{\hspace*{7cm}   
+        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
+        \in \mathcal{M}\}
+\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
+\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
+ \makebox[0mm]{\hspace*{7cm}   
  $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
            x_2\, \ldots\, x_n)/x]))$,} \\
- \makebox[0mm]{\hspace*{8cm} where
-     $x_1\, \ldots\, x_n$ are the variables in $s$}
+ \makebox[0mm]{\hspace*{5.5cm} and 
+     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
 \end{eqnarray*}
@@
 
- The liveness  of {\LET} has  an elegant formulation.   Recollect that
- our language permits lazy {\LET}s whose definitions can be recursive,
- i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
- $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
- recursive function (named $\mathit{f}$ in the rule) which captures an
- arbitrary number of ``unrollings'' of the definition\footnote{The $*$
-   in the definition of ${\mathit f}$ represents  non-deterministic choice,
-   but can be treated as a constant for purposes of analysis.}.  We pass to the
- demand transformer $\Lfonly_{\mathit f}$  of this function the demand
- arising out  of the  liveness of  $x$ at the  beginning of  $e$. This
- gives the liveness of the variables of $s$.
-
-\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
-  of  an   expression  $e$.   The  cases  $\SRETURN$   and  $\SIF$  are
-  straightforward, but note the liveness $x.\epsilon$ generated by the
-  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
-  three-address instruction:  the liveness of  $e$ is given  by taking
-  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
-  any incremental  liveness from  $s$.  The main  subtlety is  how the
-  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
-  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
+ The  {\LET} rule  has  an elegant  formulation.   Recollect that  our
+ language permits lazy \LET~s whose definitions can be recursive Thus,
+ in the expression $\LET~x  \leftarrow~s~\IN~e$, $x$ can occur in $s$.
+ We  model the  definition $x  \leftarrow~s$ as  a  possibly recursive
+ function which captures an  arbitrary number of ``unrollings'' of the
+ definition.  We pass  to this function the demand  arising out of the
+ liveness   of  $x$   at   the   beginning  of   $e$,   and  let   the
+ \Lfonly\ abstraction derive the liveness on the variables of $s$.
+
+\mycomment{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
+expression $e$.   The cases $\SRETURN$ and  $\SIF$ are straightforward,
+but note the liveness $x.\epsilon$  generated by the latter.  The case
+$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
+instruction:  the liveness  of $e$  is given  by taking  the liveness,
+$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
+liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
+$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
+$\mathit{ref}(s,\Lv(z),\Lfonly)$.
+
+%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
+
+% \end{itemize}
 
@@
 $\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
-analysis) can be safely approximated  by a {\em procedure summary} for
-each function  (unifying the contexts  in the style of  0-CFA)\@.  The
-summary consists  of a pair of  a single demand and,  for this demand,
-the  corresponding  tuple  of   demands  the  function  makes  on  its
+analysis) can  be safely  approximated, for each  function, by  a {\em
+  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
+The  summary consists  of a  pair  of a  single demand  and, for  this
+demand, the corresponding  tuple of demands the function  makes on its
 arguments.
-
-{\color{red} Proof of correctness of the liveness analysis...}
-
 \section{Computing liveness and its encoding as a table}\label{sec:computing}
-Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
-and  {\color  {red}  proved  it  correct  with  reference  to  a  {\em
-    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
-together described a liveness set for each program point in a function
-in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
-describe how to obtain demand transformers \Lfonly\ from the rule {\sc
-  live-define} and how to compute the specific demand $\sigma$ on each
-function.
-
-To  answer these  questions, it  is convenient  to cast  the equations
-arising out of the $\mathit{ref}$  and $\mathcal{L}$ as the rules of a
-grammar.   But first  we modify  the rules  themselves to  a different
-form.
+
+Section~\ref{sec:liveness} gave a context-sensitive liveness
+analysis and {\color {red}  proved it correct with reference
+  to   a  {\em   minefield}   semantics}.   In   particular,
+\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
+describes  a  liveness  set  for  each program  point  in  a
+function in terms of a  given $\sigma$ and \Lfonly. We still
+have not mentioned what is the specific sigma that should be
+used for each function How is LF to be computed.
+
+It helps to cast the equations  in the form of a grammar. To
+do  this  we  first  have  to  cast  massage  the  equations
+slightly:
 
@@
 rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
-$\emptyset$, if  $\sigma$ itself $\emptyset$.  To  realize these rules
-$\sigma$ needs to be known. This creates difficulties since we want to
-solve the equations arising out of liveness symbolically.
+$\emptyset$, if $\sigma$ itself $\emptyset$. This creates difficulties
+since  we  want  to  solve  the  equations  arising  out  of  liveness
+symbolically, and realizing these rules requires $\sigma$ to be known.
 
@@
 \end{align*}
+
 and \cbar\ as:
+
 \begin{align*}
@@
         \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
-        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
+        &[\,l \mapsto  x.\cbar \sigma]
 \end{align*}
@@
 similar to \CAR.
+
 The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
@@
   &\sigma_1\bcar\sigma_2  \hookrightarrow 
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \bcar\sigma_2',\\
 &\sigma_1\bcdr\sigma_2  \hookrightarrow 
   \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
-  \acdr\sigma_2',~\mbox{and}\\  
+  \bcdr\sigma_2',~\mbox{and}\\  
 &\sigma_1\cbar \emptyset \hookrightarrow \emptyset
@@
 \CONS\  rule,   we  refer   the  reader  to   \cite{asati14lgc}.   The
-proposition also holds for other modified rules for similar reasons.
-
-{\color{red} \medskip \hrule \hrule \hrule  \medskip }
+proposition also holds for the the other rules for similar reasons.
+
+%% To see why the  proposition is true, consider an analysis
+%% based on  \Lfunonly$'$ in  which $\sigma$ appears  in the
+%% context      $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.
+%% Let $\alpha \in  \sigma$.  The symbol \bcar(\bcdr) merely
+%% marks a place in  $\alpha$ where the original \CONS\ rule
+%% would have erased  an immediately following \acar(\acdr),
+%% or,  in absence  of  such a  symbol,  would have  dropped
+%% $\alpha$ itself.   Since the  application of any  rule in
+%% \Lfunonly$'$  merely  adds symbols  at  the beginning  of
+%% $\alpha$, the  markers and other symbols  in $\alpha$ are
+%% propagated to  other dependent parts of  program in their
+%% same  relative   positions.   Consequently,  the  erasure
+%% carried   out   at  the   end   of   the  analysis   with
+%% $\stackrel{*}{\hookrightarrow}$ gives  the same result as
+%% obtained through \Lfunonly.
 
@@
 The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
-example.   Given a  function $\mathit{f}$,  we now  describe  how to
-generate    equations    for     the    demand    transformation
+example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
+how  to generate  the equations  describing the  demand transformation
 \Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
-determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
-consider $\Lanv{1}{x_i}$,  the  liveness of    the $i^{\mbox{\footnotesize  th}}$
-parameter $x_i$ at the  program point at the  beginning of
-$e_{\mathit{f}}$ (assumed  to be  $\pi_1$). By the rule {\sc
-  live-define}, this should be the
-same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
+determine    \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   Of   particular
+interest  is the liveness  at the  program point  at the  beginning of
+beginning  of $e_{\mathit{f}}$  (assumed to  be $\pi_1$).  We generate
+equations  of the  form $\Lf{\mathit{f}}{i}{\sigma}  = \Lanv{1}{x_i}$,
+where  $x_i$   is  the  $i^{\mbox{\footnotesize   th}}$  parameter  of
+$\mathit{f}$. Doing this for  \length, we get:
+\begin{align*}
+ \Lanv{1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
+\end{align*}
+Which gives the following equation for \Lfone{\length}: 
   \begin{eqnarray*}
-&& \Lanv{1}{\xl} = \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
+    \Lf{\length}{1}{\sigma}
+    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
    \end{eqnarray*}
-and  the only equation defining \Lfone{\length} is:  
-  \begin{eqnarray*}
-   && \Lf{\length}{1}{\sigma}
-    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
-   \end{eqnarray*}
 
 In general, the equations for  \Lfonly\ are recursive since, as in the
-case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
-\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
-solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
-each  of the liveness  rules modifies  a demand  by prefixing  it with
-symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
-\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
-closed form:
+case  of \length,  \Lan{1}\ may,  in turn,  be expressed  in  terms of
+\Lfone{\mathit{f}}.
+
+As mentioned earlier, each of  the liveness rules modifies a demand by
+prefixing   with   symbols   in   the   above   set,   and   therefore
+$\Lf{f}{i}{\sigma}$ has the closed form:
 \begin{eqnarray}
 \label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} = \Df{f}{i}.\sigma
+  \Lf{f}{i}{\sigma} &=& \Df{f}{i}.\sigma
 \end{eqnarray}
-where \Df{f}{i} are sets of  strings over the alphabet mentioned above.   Substituting the guessed form in
+where \Df{f}{i} are sets of  strings over the alphabet $\lbrace \acar,
+\acdr,\bcar, \bcdr, \cbar \rbrace$.   Substituting the guessed form in
 the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
-$\sigma$,  we get  an equation  for \Df{f}{i}  that is  independent of
-$\sigma$.   Any   solution  for   \Df{f}{i}  yields  a   solution  for
-\Lfonly$_{\mathit  f}$.  Applied  to  \Lfonly$_{\mathit \length}$,  we
+$\sigma$,  we  get the  following  equation  for  \Df{f}{i}: that  are
+independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
+for \Lfonly$_{\mathit f}$.  Applied to \Lfonly$_{\mathit \length}$, we
 get:
-  \begin{eqnarray*}
-  &&  \Df{\length}{1}.\sigma = \cbar.\sigma
-    \cup \acdr\Df{\length}{1}(\cbar\sigma), ~\mbox{or}\\
-  &&  \Df{\length}{1} = \lbrace\cbar\rbrace 
-    \cup \acdr\Df{\length}{1}\cbar
-  \end{eqnarray*}
-
-Note that this equation can also be viewed as a CFG with \{\acdr,
-\cbar\} as the terminal symbols and \Df{\length}{1} as the sole non-terminal.
+  \begin{align*}
+    \Df{\length}{1} &= \lbrace\cbar\rbrace 
+    \cup \acdr\Df{\length}{1}
+  \end{align*}
+  Note that these equations can also be viewed as CFGs.
 
@@
 
-To avoid analysing the body of a function for each call, we calculated
-the liveness  for the arguments and  the variables in  a function with
-respect to a symbolic demand  $\sigma$.  To get the actual liveness we
-calculate an over-approximation of the  actual demands made by all the
-calls by a {\em summary demand} and calculate the liveness at each GC
-point inside the  function based on this summary  demand.  Even though
-we  lose precision  due to  this, our  experiments show  that  this is
-acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
-  summary demand}  is calculated by taking  a union of  the demands at
-every call site of a function.
-
-Consider a function $g$ containing a  call to $f$ at a site $\pi$, say
-$\pi\!\!:\!(\LET~x   \leftarrow   (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.
-Let the  demand on $g$  be $\sigma_g$ and,  based on this  demand, the
-liveness of  $x$ at $\pi_i$ be  $\Lanv{i}{x}$.  By the  $\LET$ rule of
-Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
-$\Lanv{i}{x}$ to  the demand $\sigma_f$.  Assuming that  there are $k$
-call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
-(in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
-$\deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
-The distinguished function  \mainpgm\ is a special case.  We assume it
-is called through a  printing mechanism with demand $\sigma_\mainpgm =
-\{\acar,\acdr\}^\ast$ (denoted $\sigma_{\!all}$).
-
-For  the  running  example,   $\length$  has  calls  from  $\main$  at
-$\pi_{10}$ and a recursive call at $\pi_6$.
+We now calculate a 0-CFA-style summary liveness for each GC point of a
+program.  There are two parts  to this.  First, for each function $f$,
+we determine a {\em  summary demand} $\sigma_f$ over-approximating any
+demand $\sigma$ passed to $f$.  So, consider a function $g$ containing
+a  call   site  $\pi$   to  $f$,  say   $\pi\!\!:\!(\LET~x  \leftarrow
+(f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.   Let   the  demand  on   $g$  be
+$\sigma_g$ and the  liveness of $x$ at $\pi_i$  be $\Lanv{i}{x}$. Then
+the $\LET$ case of Figure~\ref{fig:live-judge} tells us this call site
+contributes  $\Lanv{i}{x}$ to  the  demand $\sigma_f$  placed on  $f$.
+Write $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
+site $\pi$  in $g$).  Thus, supposing  the $k$ call  sites to function
+$f$  are  $\pi^1$ (in  function  $g^1$)  \ldots  $\pi^k$ (in  function
+$g^k$),  then  the  over-approximation  requirement on  $\sigma_f$  is
+achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
+\cup \deltacall{f}{\pi^k}{g^k}$.
+
+The expression $e_{\mainpgm}$  is a special case; we  assume it may be
+called externally with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$
+(denoted $\sigma_{\!all}$).  This is because any part of its value may
+be used by the environment---for printing the result, for instance.
+{\color {red} What about print?}
+
+For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
+and a recursive call at $\pi_6$.
 %\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
+     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{length}{\pi_6}{\length}$.
 Filling in  the values gives:
-\begin{eqnarray*}
-\sigma_{\length}    &=&
- \sigma_{\!all}  ~\cup~ \acdr\Lf{\length}{1}{\cbar\sigma_{\mathit{\length}}}
-\end{eqnarray*}
-For some of the program points in \length\  and \main,  this gives
-{\color {red} Prasanna, please fill this correctly}. You could use the
-set notation if the map notation spills over: 
+
 \begin{align*}
-\Lan{1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
-              \py.\cbar\sigma_{\length}, \\ &~~~~~~~\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
-                  \py.\cbar\sigma_{\length}, \\ &\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
-\Lan{8} &= \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace 
+\sigma_{\length}    &=
+ \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{length}}}
 \end{align*}
 
-In summary, the equations generated during liveness analysis are:
-\begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
-\item For each function $\mathit{f}$ (including $\mainpgm$ for
-  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages? No, we define it only in the implementation section as of now.}
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
+For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $ep_1, ep_2, ep_3$ in $\length$ and
+$ep_4$ in $e_\main$, this gives
+
+\begin{align*}
+\Lanv{7}{\xl} &= \acdr\Lf{length}{1}{\bX\sigma_{\mathit{length}}} \cup \bX\sigma_{\mathit{length}} \\
+\Lanv{7}{\pu} &= \Lf{length}{1}{\bX\sigma_{\mathit{\length}}} \\
+\Lanv{7}{\py} &= \bX\sigma_{\mathit{length}} \\
+\Lanv{3}{\xl} &= \bX\sigma_{\mathit{length}} \\
+\Lanv{3}{\px} &= \bX\sigma_{\mathit{length}}
+\\[1ex]
+  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
+  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
+\end{align*}
 
@@
 context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal for
+\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
 a variable  $X$ occurring on  the LHS of the  equations generated
@@
 \var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{5}{\xl}}   comprises    the   following
+grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
 productions:
  \begin{eqnarray*}
-\var{\Lanv{5}{\xl}}  &\rightarrow&   \var{\acdr \Lf
-  {\length}{1}{\cbar\sigma_{\length}}} \mid \cbar\sigma_{\length} \\
+\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
+  {\length}{1}{\epsilon}} \mid \epsilon \\
 \end{eqnarray*}
-\comment{Pick up a richer equation or more than one liveness equations}
+
+In summary, the equations generated during liveness analysis are:
+\begin{enumerate}
+\item   For    each   function   $\mathit{f}$,    equations   defining
+\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
+\item  For  each   function  $\mathit{f}$, an  equation  defining
+the summary demand  $\sigma_{\mathit f}$ on $e_f$.
+% \marginpar{Consistency: summary vs cumulative?}
+\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
+an equation defining
+  liveness at each GC point of $e_{\mathit f}$.
+
+\subsection{Solving liveness equations---the grammar interpretation}
+The  equations above can now be re-interpreted as a
+context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
+\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
+a variable  $X$ occurring on  the LHS of the  equations generated
+from the analysis.  We can  think of the resulting productions as
+% being a grammar with multiple start symbols---one for
+% for each nonterminal \var{\Lanv{i}{x}}.
+being associated with several  grammars, one for each non-terminal
+\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
+grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
+productions:
+ \begin{eqnarray*}
+\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
+  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
+ \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
+  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
+  \acdr\var{\Uf{\append}{1}}\\         
+  \var{\Df{\append}{1}}
+  &\rightarrow&               \acar\bcar               \mid
+  \acdr\var{\Df{\append}{1}}\bcdr\\
+\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
+\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
+\var{$\sigma_{\mathit{\!all}}$}
+\end{eqnarray*}
 Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
@@
 in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is
-undecidable. \comment{Rewrite it to say that this is a CFG with 3
-  unrestricted rules and the membership for even this grammar is undecidable}
-\comment{State the undecidability of the decision problem above.}
-
-  We circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
+membership  problem   for  such  grammars   is  undecidable.   We
+circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
   strongly  regular}  CFGs which  have   easy
@@
 converted to DFAs for use during garbage collection.
-\comment{Fix all comments and produce first draft}
 %------------------------------------------------------------%
@@
 \label{sec:NFA-approx}
-\comment{Change the language and the example}
+
 We use the   
@@
 
-\subsubsection{Removing the $\cbar$ edges}
-Once the NFA's are created, we need to remove the transitions on 
-$\cbar$. This edge is replaced by an $\epsilon$ transition and the 
-$\cbar$ edge is deleted. If the original grammar had a non-null 
-$\sigma$ then the NFA reaches a final state throught this path 
-otherwise it does not reach a final state.\comment{Need to rewrite this}
 %------------------------------------------------------------%
@@
 Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
-\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow}
-\lang{\nt}$, where {\nt} accepts forward paths only.
-\comment{Rewrite the algorithm to handle \cbar and update the example automata}
+\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
 Thus  \nt\  can  be  used  to  check  
@@
   \raggedright{\bf Steps:}
- %% \begin{algorithmic}
-%%    \COMMENT Fix the algorithm issue
-%%    %% \STATE $i \leftarrow 0$ 
-%%    %% \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-%%    %% without $\epsilon$-moves \cite{hopcraft90toc}
-
-%%    %% \REPEAT
-%%    %% \STATE $\nt'_{i+1} \leftarrow \nt_i$
-%%    %% \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-%%    %%   edge from $q'$ with label $\bcar$ and
-%%    %%   outgoing edge to $q''$ with label $\acar$} 
-%%    %% \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-%%    %% label $\epsilon$.
-%%    %% \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-%%    %% \ENDFOR
-
-%%    %% \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-%%    %%   edge from $q'$ with label $\bcdr$ and outgoing edge to
-%%    %%   $q''$ with label $\acdr$} 
-%%    %% \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-%%    %%   label $\epsilon$.
-%%    %% \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-%%    %% \ENDFOR
-
-%%    %% \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-%%    %% $\nt'_{i+1}$ without $\epsilon$-moves
-
-%%    %% \STATE $i \leftarrow i+1$
-%%    %% \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-%%    %% \STATE $\nt \leftarrow \nt_i$
-%% %   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
-%%  \end{algorithmic}
+ \begin{algorithmic}
+   \STATE $i \leftarrow 0$ 
+   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
+   without $\epsilon$-moves \cite{hopcraft90toc}
+
+   \REPEAT
+   \STATE $\nt'_{i+1} \leftarrow \nt_i$
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcar$ and
+     outgoing edge to $q''$ with label $\acar$} 
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+   label $\epsilon$.
+   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
+   \ENDFOR
+
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcdr$ and outgoing edge to
+     $q''$ with label $\acdr$} 
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+     label $\epsilon$.
+   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
+   \ENDFOR
+
+   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
+   $\nt'_{i+1}$ without $\epsilon$-moves
+
+   \STATE $i \leftarrow i+1$
+   \UNTIL ($\nt_{i} = \nt_{i-1}$)
+
+   \STATE $\nt \leftarrow \nt_i$
+%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
+ \end{algorithmic}
 \end{center}
@@
 The proofs  of the termination  and correctness of  the algorithm
-are given in an earlier paper~\cite{karkare07liveness}.
-
-%%\input{liveness-dfa.tex}
-\comment{change according to example}
+are given in our earlier paper~\cite{karkare07liveness}.
+
+\input{liveness-dfa.tex}
+
 We  illustrate  the  algorithm  in  Figure~\ref{algo:simplify-NFA}  by
@@
 \section{Implementation}
-% GC assumptions
-\subsection{GC Points}
-Unlike eager languages where GC happens only at \CONS~ points or at 
-function calls, every \LET~ point is a potential GC point. The number of 
-automata that needs to be stored is reduced by sharing automata along all 
-the program points along an {\em E-path}.
-\comment{
 By the  time the program  starts executing, the liveness  automata are
@@
 automata state required  by sharing the same automata  for all program
-points along an {\em E-path}.} In  the example program, we only need to
+points along an {\em E-path}. In  the example program, we only need to
 create automata  for program  points \ldots, since  they are  the only
@@
 % Liveness for cons cells & reachability for closures
-
-\comment{Mention the fact that liveness at a program point indicates
-  what references will be used by the program eventually. The exact
-  point where a reference will be used cannot be determined statically
-due to the language being lazy. }
 During GC, if the heap cell being processed is a cons cell, we consult
@@
 presence  of  a print  stack.  In  lazy  languages the  evaluation  of
-closures is  driven by the print  function\comment{Refer to SPJ's book}.  Hence, during  a GC there
+closures is  driven by the print  function.  Hence, during  a GC there
 will be heap  references that will be present on  the stack which also
@@
 algorithm is as described below,
-
-%% \begin{figure}
-%% \begin{algorithm}[H]
-%%  \KwData{this text}
-%%  \KwResult{how to write algorithm with \LaTeX2e }
-%%  initialization\;
-%%  \While{not at end of this document}{
-%%   read current\;
-%%   \eIf{understand}{
-%%    go to next section\;
-%%    current section becomes this one\;
-%%    }{
-%%    go back to the beginning of current section\;
-%%   }
-%%  }
-%%  \caption{How to write algorithms}
-%% \end{algorithm}
-%% \label{algo:LGC-Algo} \figrule
-%% \end{figure}
+%% \begin{algorithmic}
+%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
+%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
+%% \STATE $new\_cdr\gets$ copy(cell.cdr)
+%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
+%% \ENDIF
+%% \end{algorithmic}
 We   provide   an   intuition    about   the   correctness   of   this
-algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts about the print stack during a
+algorithm. Consider the following facts about the print stack during a
 GC,
@@
 \section{Results}
-To demonstrate the effectiveness of liveness-based garbage collection,
-we  have  built a  prototype  consisting  of  an interpreter  for  our
-language,  a  liveness  analyzer  and  a copying  collector  that  can
-optionally use the results of liveness analysis for marking instead of
-reachability. When the collector uses liveness for marking, we call it
-a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-  reachability-based  collector}  (RGC).   The  copying  collector  is
-neither incremental nor generational.  As a consequence, any cell that
-becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.
-
-While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
-is invoked  at the program point   $\pi$,  it would  use  the
-liveness  DFAs at  $\pi$ itself for exploring the root set of the
-current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.
-
-Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-$\dfa{\pi}{x}$.    
-Considering   a   DFA   as   a   table,
-$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-\acar\      or     \acdr.       We      shall     also      write
-$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-indicating  whether there  is a  transition from  $q$ on
-$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-pointer to $x$.
-The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-$x$  in the  other semi-space  and returns  the new  address.  It
-hides  details such as  returning the  forwarding pointer  if the
-value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.
-
-The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-in the heap over time for RGC and LGC---here time is measured in terms
-of the number of cons cells allocated. In addition, they also show the
-number of  reachable cells and the  number of cells  that are actually
-live\footnote{This   is  statically   approximated  by   our  liveness
-  analysis}.  Since  the programs have  different memory requirements,
-we have tuned the size of heap for each program to ensure a reasonable
-number of  collections. An invocation  of RGC decreases the  number of
-cells  in heap  till  it touches  the  curve of  reachable cells.   An
-invocation of LGC decreases the number  of heap cells to no lower than
-the curve of live cells.
-
-To construct the  reachable and live curves, we  record for every cell
-its creation time (\CreateTime), its last use time (\UseTime), and the
-earliest time  when the  cell becomes unreachable  and can  be garbage
-collected (\CollTime).  For accurate recording of  \CollTime, we force
-frequent invocations  of a reachability based collector  in a separate
-run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
-  fibheap} increases  initially, comes  down before the  heap becomes gets
-full and then  continues to remains low.  This results in  a single garbage
-collection in both RGC and LGC.  The graphs also show the precision of
-our liveness analysis.  For  all programs  except  {\tt nperms} and
-{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.
-
-\subsection{Results}
-\begin{figure}[t] 
-\centering
-\begin{boxedminipage}{.75\textwidth}
-{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
-\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-  then} {\bf skip}\\ 
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
-\noindent \hspace*{.35cm}{\bf return} $y$
-\end{boxedminipage}
-\caption{Function       for       copying       a       root       set
-  variable.}\label{algo:LGC-main-loop} \figrule
-
-\end{figure}
-
-\begin{figure}[p]
-\begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
- &
-{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-&
-{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-\end{tabular}
- \caption{Memory  usage of  programs. The blue and the red curves
-  indicate the number of cons cells in the active semi-space for RGC and
-  LGC respectively. The black curve represents the number of reachable
-  cells and the grey curve represents the number of cells that are
-  actually live (of which liveness analysis does  a static
-  approximation). x-axis is the time measured in number of cons-cells
-  allocated. y-axis is the number of cons-cells.}
-\label{fig:memory-usage} \figrule
-\end{figure}
-The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-provides  statistics  regarding the  analysis  itself.  The number  of
-states and the analysis times  are within tolerable limits. Precision
-of analysis refers to the percentage of dead cells that is collected by
-LGC, averaged over all invocations.
-The second table shows garbage  collection statistics for RGC and LGC.
-LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to run
-in comparison with  RGC.
-
-There are a couple of  issues of concern.  The garbage collection time
-is larger in the case of LGC for some programs. The reason is that the
-cost of consulting the liveness DFA may outweigh the combined benefits
-of  fewer   garbage  collections   and  fewer  markings   per  garbage
-collection.   The  other issue  is  illustrated  by  the program  {\tt
-  lambda}    (memory   usage  shown   separately    in
-Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-in Figure~\ref{fig:experimental-results}, the  number of
-touched cells\footnote{These  are the  cells which are  visited during
-  marking phase, often  more than once due to  sharing.} in this
-example is much higher for
-LGC.  This  increase is due to  excessive sharing among  heap nodes in
-this program.  Note  that a node re-visited because  of sharing is not
-explored  any   further  during  a  RGC   collection.   However,  this
-curtailment cannot happen in  LGC because of the possibility that the
-node,  re-visited in a  different liveness  state, may  mark a  set of
-cells different from the earlier visit.
-\begin{figure}[t]
-\footnotesize
-
-\centering
-
-\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline
-{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-  nperm} &  {\sf fibheap} & {\sf knightstour} &
-{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-\hline
-\hline
-{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-\hline
-\end{tabular}}
-
-\centerline{(a)}
-
-\medskip
-
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
-  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-
-                            &   \multicolumn{2}{c|}{cells per GC}  
-                            &   \multicolumn{2}{c|}{cells per GC} 
-                            &   \multicolumn{2}{c|}{\#GCs} 
-                            &   \multicolumn{2}{c|}{MinHeap} 
-                            &   \multicolumn{2}{c|}{Avg. Drag} 
-                            &   \multicolumn{2}{c|}{(sec)} \\
-\cline{2-13}
-{Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
-%    {\sf  nboyer} & &  & &   & &  & & &  \\
-%    {\sf   circsim?} & &  & &   & & & & &   \\
- \hline
-\end{tabular}}
-
-\centerline{(b)}
-\vskip -2mm
-\caption{Experimental results comparing RGC and LGC. Table (a) gives
-  data related to liveness analysis, and (b) gives garbage collection
-  data. The average drag time is measured in allocated cons cells.}
-\label{fig:experimental-results}
-\normalsize
-\end{figure}
-
-\section{Collecting more garbage can never slow things down}
-\label{sec:lgc-always-better}
-
-%% Since garbage collection is  effectively asynchronous to the allocator
-%% thread, one might  worry as to how {\em  robust} our measurements are.
-%% For example,  while LGC would,  in general, collect more  garbage than
-%% RGC in the same heap state,  is it possible that on some programs, LGC
-%% might do  a larger number  of collections? We  prove a result  to show
-%% that   this  cannot   happen.    This  result   applies  to   classical
-%% mark-and-sweep and copying garbage  collectors and, we hypothesize, to
-%% generational collectors too.
-
-%% \begin{lemma}
-%% For the  same mutator,  a liveness-based collector  can never  do more
-%% garbage collections than a reachability-based collector.
-%% \end{lemma}
-
-%% \begin{proof}Assume, as before,  that time is measured in terms of the number of
-%%   cons cells allocated. Let us run  two copies of the mutator with the
-%%   two    different   collectors    but    with   synchronous    memory
-%%   allocations.   The   garbage    collections,   however,   happen
-%%   asynchronously.
-
-%%   To prove the lemma, it is enough to show the truth of the following 
-%% statement: After
-%%  every  LGC invocation,  the count of LGC invocations   is  no greater
-%% than  RGC invocations.   The  base case  holds  since the  first
-%% invocations of both GCs happen at the same time.  Assume the statement
-%% to be true after $n$ invocations of LGC.  Since LGC copies a subset of
-%% reachable cells, its heap would contain no more cells than RGC heap at
-%% the end  of the $n$th invocation.   Thus either RGC  is invoked next before
-%% LGC, or LGC and RGC are both invoked next at the same time.  In either
-%% case, the statement holds after $n+1$ invocations of LGC.
-%% \end{proof}
-
-%% \begin{figure}[h]
-%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
-%% \caption{Memory  usage of  programs. The  solid  lines represents
-%%   number of  live cells and the  dashed lines show  the number of
-%%   reachable cells.  The red and blue  lines represent invocations
-%%   of RGC and  LGC---their heights indicate  the amount of memory
-%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
-%%       paper and are being used as placeholder}}}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-%%\clearpage
-
-\section{Related Work} Previous attempts to  increase  the space efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
-reallocation~\cite{jones89compile},                       deforestation
-techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
-methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
-
-\begin{figure}[t]
-\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
- \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
-\label{fig:memory-usage-lambda} \figrule
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusion}
-We have defined a notion of liveness on structured data;
-this generalizes classical liveness and strong liveness.
-We started with a general fully context-sensitive analysis which
-we proved correct with respect to a minefield semantics
-(this models the effect of garbage collection between every
-evaluation step).
-
-To avoid scalability issues (and to avoid performing part of
-the liveness computation at run time)
-we defined an 0-CFA version of this liveness analysis in which
-demands for function $f$ at all calling contexts are conflated into 
-a single demand $\sigma_f$.
-This enabled us to treat the liveness equations symbolically obtaining
-context-free grammars for liveness at each GC point (calls to user
-functions and to $\CONS$).
-These were then converted to DFAs for run-time consultation by the
-garbage collector. Experiments confirm the precision of the analysis.
-
-To obtain performance figures we compared a reachability-based garbage
-collector with a liveness-based  collector.  This showed a decrease in
-the  number   of  GCs,  more  garbage  collected   per  invocation.  A
-significant benefit of LGC is that programs can run in smaller memory when
-compared to RGC. This is potentially useful in situations where memory
-is limited---as with embedded systems. For a majority of programs, the
-garbage collection times were reduced.
-
-One issue we  highlighted was that while fewer  nodes were marked (and
-more garbage collected), sometimes  cons cells could be visited and
-traversed  multiple times  with different  sets of  liveness  paths to
-explore.  Further work includes  improvements to the classical copying
-collector to reduce the cost of this.
-
 %Add graphs for memort usage LGC vs RGC
 %Add graphs/table showing the GC times and other statistics
-
+\section{Conclusion and Future Work}
 %Mention lower peak memory usage. 
@@
 %------------------------------------------------------------%
-%\input{lazy-liveness}
 

### 2015-06-02 lazy-scheme/post_process/gen_stats_and_plotall.sh

@@ -44,14 +44,16 @@ declare -A limit
 #limit["sudoku"]=28940
 limit["sudoku"]=1728
 limit["fft"]=3300
-limit["lcss"]=52321
+#imit["lcss"]=52321
+limit["lcss"]=22243
 limit["gc_bench"]=131091
-limit["nperm"]=27429
-limit["fibheap"]=300000
+limit["nperm"]=27428
+limit["fibheap"]=38200
 limit["knightstour"]=508245
 limit["treejoin"]=1800000
-limit["nqueens"]=210000
-limit["lambda"]=20466
+limit["nqueens"]=25000
+#imit["lambda"]=20466
+limit["lambda"]=100000
 
 declare -A freq
 
@@ -76,7 +78,7 @@ xtics["nperm"]=145000; ytics["nperm"]=35000
 xtics["fibheap"]=70000; ytics["fibheap"]=30000
 xtics["knightstour"]=850000; ytics["knightstour"]=40000
 xtics["treejoin"]=130000; ytics["treejoin"]=110000
-xtics["nqueens"]=3000000; ytics["nqueens"]=210000
+xtics["nqueens"]=6000000; ytics["nqueens"]=105000
 xtics["lambda"]=90000; ytics["lambda"]=10000
 
 if [ "x$BMs" = "x" ]; then
@@ -85,8 +87,8 @@ fi
 
 for bm in $BMs
 do
-    #for opt in gc-plain gc-live
-    for opt in gc-plain gc-live gc-freq=${freq[$bm]}
+    for opt in gc-plain gc-live
+    #for opt in gc-plain gc-live gc-freq=${freq[$bm]}
     do 
      	echo "Processing $bm [$opt]"
      	$DIR/gen_gc_stats.sh $OPTARG \

### 2015-06-03 paper/paper.tex
new paper
Updated Implementation Section

--- 2015-06-02 paper/paper.tex
@@
 
-\hspace*{2cm}{\epsfig{file=motiv-example.eps, height=4cm}}
+%\hspace*{2cm}{\epsfig{file=motiv-example.eps, height=4cm}}
 
@@
 % GC assumptions
-\subsection{GC Points}
-Unlike eager languages where GC happens only at \CONS~ points or at
-function calls, every \LET~ point is a potential GC point. The number
-of
-automata that needs to be stored is reduced by sharing automata along
-all
-the program points along an {\em E-path}.
-\comment{
-By the  time the program  starts executing, the liveness  automata are
-created and stored in a file.   This is re-used every time the program
-executes.  An important difference in  GC of lazy language compared to
-an  eager language is  the point  of garbage  collection.  In  case of
-eager  languages only  \CONS~ points  and  function calls  are the  GC
-points. In  a lazy  language every \LET~  statement is a  potential GC
-point. Hence we need to store liveness automata corresponding to every
-\LET~  statement in the  program. This  leads to  an explosion  in the
-number of  automata. From our  observation that along an  {\em E-path}
-all program points have the same liveness, we can reduce the number of
-automata state required  by sharing the same automata  for all program
-points along an {\em E-path}.} In  the example program, we only need to
-create automata  for program  points \ldots, since  they are  the only
-distinct {\em E-path}s in the program.
+In this section we describe in detail the process of liveness based garbage collection. 
+During the execution of a program if it runs of out of memory, execution is stopped and the 
+garbage collector is invoked. At this point all the references on the stack are checked for 
+liveness and only live references are copied and the program continues execution. 
+The main complexity in our algorithm is the way closures are handled. The following sub-sections 
+show two diffrent ways of handling closures along with their pros and cons.
+\subsection{A simple LGC Algorithm}
 % Abstract description of the GC algorithm
-%% \begin{algorithmic}
-%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
-%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
-%% \STATE $new\_car\gets$ copy($cell.car$)
-%% \STATE $new\_cell.car\gets$ $new\_car$
-%% \STATE $new\_cdr\gets$ copy(cell.cdr)
-%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
-%% \ELSE
-%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
-%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
-%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)
-%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)
-%% \ELSE
-%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
-%% \ENDIF
-%% \ENDIF
-%% \end{algorithmic}
-
-% Liveness for cons cells & reachability for closures
-
-\comment{Mention the fact that liveness at a program point indicates
-  what references will be used by the program eventually. The exact
-  point where a reference will be used cannot be determined statically
-due to the language being lazy. }
-During GC, if the heap cell being processed is a cons cell, we consult
-the  liveness automaton  corresponding to  it and  copy only  the live
-paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
-approximate the  liveness to be {\{0+1\}\*} and  copy everything under
-the closure.
-% Handling of mixture of LGC & RGC
+\begin{verbatim}
+procedure copy (cell)
+ new_cell <-  dup_heapcell(cell)
+ IF{celltype is cons}
+     IF {cell.car is live}
+        new_car        <-  copy(cell.car)
+        new_cell.car   <-  new_car
+     ENDIF
+     IF {cell.cdr is live}
+        new_cdr        <-  copy(cell.cdr)
+        new_cell.cdr   <-  new_cdr
+     ENDIF
+ ELSE
+  IF {celltype is closure}
+    new_cell      <-  dup_heapcell(cell)
+    IF {cell.arg1 is live}
+       new_cell.arg1 <-  copy(cell.arg1)
+    ENDIF
+    IF {cell.arg2 is live}
+       new_cell.arg2 <-  copy(cell.arg2)
+    ENDIF
+ ENDIF
+ ENDIF
+ copy_references_on_print_stack()
+\end{verbatim}
+In the simple algorithm presented we consult liveness automata for copying both \CONS cells and closures. 
+Copying \CONS cells are simple and just involve consulting the liveness automaton corresponding to them 
+and recursively copying. In case of closures, it is more complicated as they might refer to data/closures 
+which might be created in a different environment(function). To overcome this we we carry some extra data 
+with each closure, we carry the program point at which it was evaluated and also the names of the free 
+variables contained in the closure. This gives us precise liveness information of the free variables of a 
+closure but involves run-time overhead of updating the program point correctly. To avoid this overhead we 
+go for a safer approximation by storing the creation point of the closure and using it to check liveness. 
+This is safe since the liveness at the creation point of the closure encompasses all possible uses of the 
+closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
+what is found at the creation point.
+
+\subsection{Modified GC Algorithm}
+Since the first algorithm requires a lot of bookkeeping and is not precise, we use a modified algorithm 
+which does not use liveness for copying closures. 
+\begin{verbatim}
+procedure copy (cell)
+ new_cell <-  dup_heapcell(cell)
+ IF{celltype is cons}
+     IF {cell.car is live}
+        new_car        <-  copy(cell.car)
+        new_cell.car   <-  new_car
+     ENDIF
+     IF {cell.cdr is live}
+        new_cdr        <-  copy(cell.cdr)
+        new_cell.cdr   <-  new_cdr
+     ENDIF
+ ELSE
+  IF {celltype is closure && cell is not copied}
+    new_cell      <-  dup_heapcell(cell)
+    new_cell.arg1 <-  copy(cell.arg1)
+    new_cell.arg2 <-  copy(cell.arg2)
+  ENDIF
+ ENDIF
+ copy_references_on_print_stack()
+\end{verbatim}
+
+In this algorithm we consult the liveness automata only if it is a cons cell, for closures we 
+assume everything under it to be live and copy everything that is reachable. Although this might 
+lead to copying more cells it is faster as there are no liveness checks. Additionally we can 
+optimize some of the processing by not traversing cells more than once. This is described in 
+sub-section[]. \comment{Once experimental results are obtained compare the two algorithms here}
+Since we mix both liveness based GC and reachability the garbage collector has to be more careful 
+and ensure references marked non-live are never copied using reachability. We discuss this issue 
+and its solution in sub-section[].
+
+\subsection{Optimization for RGC for closures}
+One of the  reasosn why reachability based GC  is faster than liveness
+based GC is  due to the fact that  in case we try to copy  a cons cell
+which is  already copied  we do  not need to  traverse it  any further.
+ The  same is not true in the case  of LGC, as the set
+of  liveness paths  may not  be similar  in both  the  traversals. One
+optimization is  to keep  track of all  the liveness  automaton states
+that a particular  cons cell has been visited with  and ensure that we
+do  not  traverse  it  with  the  same state  more  than  once.   This
+optimization requires extra memory  for each heap cell and potentially
+it can hold  all the automata states.  Therefore it  is not a feasible
+solution. In case of lazy  languages since we use reachability to copy
+closures, some cons  cells might be copied using  reachability. We can
+avoid  repeated traversals of  these cons  cells without  tracking the
+states. This only requires the  knowledge of whether the cons cell was
+copied using  reachability.  Hence  we use a  flag to remember  if the
+heap cell was  copied using reachability at any time, if  it is set we
+do not traverse any further.\comment{Maybe add a diagram to make it clearer}
+
+\subsection{Issues with mixing LGC and RGC}
 A consequence of  copying everything under a closure  is the fact that
@@
 valid.  To avoid  this behavior we either have to choose  to do a full
-liveness based  GC for closure or  we should ensure  that any non-live
+liveness based  GC for closures or  we should ensure  that any non-live
 references are  not carried across  GC's. We take the  latter approach
@@
 % Proof for the correctness.
-% Difference between eager and lazy GC
-
+\subsection{Garbage collection for references on print stack}
 Another  difference that is  present in  GC of  lazy language  is the
@@
 function by using a flag in  the heap cell. The flag ensures that only
-the required parts  of a cons cell are copied. The  behavior of the GC
-algorithm is as described below,
-
+the required parts  of a cons cell are copied.
 %% \begin{figure}
@@
 %% \end{figure}
-We   provide   an   intuition    about   the   correctness   of   this
-algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts
-about the print stack during a
-GC,
-
-\begin{enumerate}
-\item The top of the stack always contains a reference to a closure.
-\item All other references on the print stack point to atomic values or
-cons cells, it can never contain a
-reference to a closure. The components of the cons cell can contain
-reference to closures.
-\item If a reference on the print stack is a cons cell, either its car
-part or its cdr part will also be
-present on the stack.
-\end{enumerate}
-%Needs to be formatted and re-written
-Sketch of the proof
-Since all closures are copied fully and atomic values are copied, only
-cons cells need to be handled.
-Consider the following cases,
-\begin{enumerate}
-\item the car part of the cons  cell is on the print stack. This means
-  that the GC happened while the car part was being processed. The cdr
-  part is yet to be processed and the reference to the cdr part is not
-  on the print stack. Hence it is sufficient to copy only the cdr part
-  of the  cons cell. Any  references necessary for processing  the car
-  part will be copied due to references on the print stack.
-\item the cdr part of the cons  cell is on the print stack. This means
-  that GC  happened while the cdr  part was being  processed.  The car
-  part has already been processed and printed, hence we do not need to
-  copy the car  part. Additionally any references that  are needed for
-  printing the cdr  part will also be copied due  to the references on
-  the print stack and hence nothing needs to be copied in this case.
-\end{enumerate}
+%% We   provide   an   intuition    about   the   correctness   of   this
+%% algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts
+%% about the print stack during a
+%% GC,
+
+%% \begin{enumerate}
+%% \item The top of the stack always contains a reference to a closure.
+%% \item All other references on the print stack point to atomic values or
+%% cons cells, it can never contain a
+%% reference to a closure. The components of the cons cell can contain
+%% reference to closures.
+%% \item If a reference on the print stack is a cons cell, either its car
+%% part or its cdr part will also be
+%% present on the stack.
+%% \end{enumerate}
+%% %Needs to be formatted and re-written
+%% Sketch of the proof
+%% Since all closures are copied fully and atomic values are copied, only
+%% cons cells need to be handled.
+%% Consider the following cases,
+%% \begin{enumerate}
+%% \item the car part of the cons  cell is on the print stack. This means
+%%   that the GC happened while the car part was being processed. The cdr
+%%   part is yet to be processed and the reference to the cdr part is not
+%%   on the print stack. Hence it is sufficient to copy only the cdr part
+%%   of the  cons cell. Any  references necessary for processing  the car
+%%   part will be copied due to references on the print stack.
+%% \item the cdr part of the cons  cell is on the print stack. This means
+%%   that GC  happened while the cdr  part was being  processed.  The car
+%%   part has already been processed and printed, hence we do not need to
+%%   copy the car  part. Additionally any references that  are needed for
+%%   printing the cdr  part will also be copied due  to the references on
+%%   the print stack and hence nothing needs to be copied in this case.
+%% \end{enumerate}
+\hrule
+\comment{
+Unlike eager languages where GC happens only at \CONS~ points or at
+function calls, every \LET~ point is a potential GC point. The number
+of
+automata that needs to be stored is reduced by sharing automata along
+all
+the program points along an {\em E-path}.
+
+By the  time the program  starts executing, the liveness  automata are
+created and stored in a file.   This is re-used every time the program
+executes.  An important difference in  GC of lazy language compared to
+an  eager language is  the point  of garbage  collection.  In  case of
+eager  languages only  \CONS~ points  and  function calls  are the  GC
+points. In  a lazy  language every \LET~  statement is a  potential GC
+point. Hence we need to store liveness automata corresponding to every
+\LET~  statement in the  program. This  leads to  an explosion  in the
+number of  automata. From our  observation that along an  {\em E-path}
+all program points have the same liveness, we can reduce the number of
+automata state required  by sharing the same automata  for all program
+points along an {\em E-path}.} In  the example program, we only need to
+create automata  for program  points \ldots, since  they are  the only
+distinct {\em E-path}s in the program.
+
+% Liveness for cons cells & reachability for closures
+
+\comment{Mention the fact that liveness at a program point indicates
+  what references will be used by the program eventually. The exact
+  point where a reference will be used cannot be determined statically
+due to the language being lazy. }
+During GC, if the heap cell being processed is a cons cell, we consult
+the  liveness automaton  corresponding to  it and  copy only  the live
+paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
+approximate the  liveness to be {\{0+1\}\*} and  copy everything under
+the closure.
+% Handling of mixture of LGC & RGC
+
+% Difference between eager and lazy GC
+
 % Doing GC for  references on the print stack - using  a flag to avoid
@@
 % using rgc flag for closures to avoid multiple traversals
-One of the  reasosn why reachability based GC  is faster than liveness
-based GC is  due to the fact that  in case we try to copy  a cons cell
-which is  already copied  we do  not need to  traverse it  any further
-while doing RGC. The  same is not true in the case  of LGC, as the set
-of  liveness paths  may not  be similar  in both  the  traversals. One
-optimization is  to keep  track of all  the liveness  automaton states
-that a particular  cons cell has been visited with  and ensure that we
-do  not  traverse  it  with  the  same state  more  than  once.   This
-optimization requires extra memory  for each heap cell and potentially
-it can hold  all the automata states.  Therefore it  is not a feasible
-solution. In case of lazy  languages since we use reachability to copy
-closures, some cons  cells might be copied using  reachability. We can
-avoid  repeated traversals of  these cons  cells without  tracking the
-states. This only requires the  knowledge of whether the cons cell was
-copied using  reachability.  Hence  we use a  flag to remember  if the
-heap cell was  copied using reachability at any time, if  it is set we
-do not traverse any further.
 

### 2015-06-03 paper/paper.tex
liveness

--- 2015-06-02 paper/paper.tex
@@
 
-%
 % You need the command \numberofauthors to handle the 'placement
@@
 
-\subsubsection{Semantics}
+\subsection{Semantics}
 We now give  a small-step semantics for our  language.
@@
 of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
-\rbrace$  paths of  the result  of  $(\CDR~y)$\footnote{The classical
-  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
-  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
-  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
-  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
-  becomes the demand on $y+z$.   This, in turn, generates the liveness
-  $\lbrace y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
+\rbrace$  paths   of  the  result   of  $(\CDR~y)$\footnote{Reproduced
+  from~\cite{asati14lgc}: In  classical strong liveness  analysis, $y$
+  and $z$  are live at the  entry $n: x:=y+z$,  if and only if  $x$ is
+  live  at exit  of $n$.   In our  terminology, the  liveness $\lbrace
+  \epsilon\rbrace $ of $x$ at the  exit from $n$ becomes the demand on
+  $y+z$, and this, in turn generates the liveness $\lbrace y.\epsilon,
+  z.\epsilon \rbrace$ at the entry of $n$.}.
 %% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
@@
 Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
-parts. The function $\mathit{ref}$,  when given an application $s$ and
-a demand $\sigma$, returns  the incremental liveness generated for the
+parts. The function $\mathit{ref}$,  takes  an application $s$ and
+a demand $\sigma$, and returns  the incremental liveness generated for the
 free  variables   of  $s$  due  to  the   application.   The  function
@@
 
- In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
- generates no  liveness.  Notice that this is  different from liveness
- of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
- evaluated  even for  the null  demand. This  would require  an access
- using $x$ resulting in  $x.\epsilon$ being live.  A non-null demand of
- $\sigma$ on (\CDR~$x$), is transformed to the liveness $\{x.\epsilon,
- x.\acdr\sigma\}$.  In an opposite  sense, the demand of $\acdr\sigma$
- on  (\CONS~$y$~$z$) is  transformed to  the demand  $\sigma$  on $z$.
- Note that \CONS\ does not,  by itself, dereference its arguments. The
- rules for (\PRIM~x~y) and  (\NULLQ~x) are on similar lines. Constants
- generate no liveness.
+ Since  in a  lazy language,  an  expression is  not evaluated  unless
+ required, the null demand ($\emptyset$) does not generate liveness in
+ any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
+ non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed  to the
+ liveness $\{x.\epsilon, x.\acdr\sigma\}$.   In an opposite sense, the
+ demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
+ demand  $\sigma$ on  $z$.   Note  that \CONS\  does  not, by  itself,
+ dereference its  arguments. The rules for  (\PRIM~x~y) and (\NULLQ~x)
+ are similar. Constants do not generate any liveness.
 
@@
 program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
-specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
-result  of a  call to  $f$  is transformed  into the  liveness of  its
-parameters at the  beginning of the call.  $\Lfonly$  is determined by
-the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
-({\sc live-define}).  This rule  describes the fixed-point property to
-be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
-for each  function in  the program  should be the  same as  the demand
-transformation  calculated from  their  bodies.  As  we  shall see  in
+specific function $f$), expresses how the demand $\sigma$ on a call to
+$f$  is  transformed  into  the  liveness of  its  parameters  at  the
+beginning of the call.  $\Lfonly$  is determined by the judgement form
+$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
+This  rule  describes the  fixed-point  property  to  be satisfied  by
+$\Lfonly$, namely, the demand transformation assumed for each function
+in  the  program should  be  the  same  as the  demand  transformation
+calculated    from    its    body.      As    we    shall    see    in
 Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar and  the least  solution satisfying the  rule is  the language
-generated by this  grammar.  The least solution also  ensures the safe
-collection of the greatest amount of garbage.
+grammar  and the  the  language  generated by  this  grammar is  least
+solution satisfying  the rule. We  prefer the least solution  since it
+ensures the safe collection of the greatest amount of garbage.
 
@@
 To see the  consequences of this, consider
-Figure~\cite{fig:mot-example2} which shows an analysis  of the body of
+Figure~\ref{fig:mot-example2} which shows an analysis  of the body of
 a function $\length$ with a demand $\sigma$.  Clearly, this is also the
@@
 \pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
-
-\begin{figure*}[t!]
-  \begin{tabular}{cc}
-    \begin{minipage}{.45\textwidth}
-        \small
-        \renewcommand{\arraystretch}{1}{
-          \begin{uprogram}
-            \UNL{1} (\DEFINE\ (\length~\xl)
-            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
-(\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
-$\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
-            $(\LET\ \pv\ $\leftarrow 0$ \IN
-            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
-(\SRETURN~\psi_2:\pv)$
-            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
-$\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
-$\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
-            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
-(\SRETURN~\psi_3:\pz)$)))))))
-        \end{uprogram}}
-        \renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $(\DEFINE\ (\main)$
-          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots
-\IN
-	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\
-$\leftarrow$  \ldots \IN
-          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
-          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
-          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\,
-(\SRETURN~\psi_4:\pw)))))$
-\end{uprogram}}
-        %%}
-    \end{minipage}
-
-    &
-
-    \begin{minipage}{.51\textwidth}
-
-        \small
-\begin{eqnarray*}
-        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
-              [\;ep_1 \mapsto \lbrace \px.\sigma_{\length},
-\xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace,\\
- &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma_{\length},
-                \py.\cbar\sigma_{\length},
-\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},\\
-&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
-\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
-                \rbrace~\rbrack \\
-            \cal{P_{\mbox{\tiny \length}}} &=&
-                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
-                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
-                  \rbrace, \\
-&&\;\;  \pi_3 \mapsto \lbrace ep_2
-                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
-                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
- &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
-\pi_7 \mapsto
-                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
-\rbrace~\rbrack\\
-              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
-                [\;
-                  ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}},
-\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots\rbrace
-                  ]\\
-              \cal{P_{\mbox{\tiny \main}}} &=&
-                  \lbrack~\pi_6 \mapsto \lbrace ep_4 \rbrace,  \
-                   \pi_7 \mapsto \lbrace ep_4 \rbrace,  \
-                  \pi_8 \mapsto \lbrace ep_4 \rbrace~\rbrack
-\end{eqnarray*}
-    \end{minipage} \\
-(a)&
-(b)
-    %% \end{ tabular}}
-  \end{tabular}
-%\end{picture}
-\kern -3ex
-%\vspace{0.5in}
-\caption{(a) Example program and (b) its liveness
-maps.}\label{fig:mot-example2}
-\end{figure*}
-
-Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
-analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
-computes a  map that takes a  e-path in ${\mathit f}$  and returns the
-liveness environment  arising out of  all uses of variables  along the
-e-path.  A second map $\cal{P}_{\mathit  f}$ maps any program point in
-$\mathit f$  to the set  of e-paths sharing  the point.  If  a program
-point happens to fall on more  than one such e-path, then the liveness
-environment of  the program  point is the  variable-wise union  of the
-liveness  environments  of the  individual  e-paths.   In the  example
-program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
-ep_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(ep_1)  \cup {\cal
-  M}(ep_2) \cup  {\cal M}(ep_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
-\sigma, \Lfonly)$.  In the  sequel we shall only define $\mathcal{L}$.
-The formulation  of $\cal{P}$ is easy  and we do not  elaborate it any
-further.
-
-Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
-The  expression $(\SRETURN~\psi:x)$  with a  demand $\sigma$  forces an
-evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
-  \{x.\sigma\}] $  needs to  be created, where the $ep$ is a fresh
-evaluation path terminating in $\psi$.  The  map for  the expression
-$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
-$e_2$. In addition, since the condition also triggers an evaluation of
-$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
-added  to the  union.  Also  notice a  consequences  of laziness---the
-entire  expression including  the condition  is not  evaluated  if the
-demand on it is $\emptyset$.  This results in the empty map, which, by
-abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.
-
- The liveness  of {\LET} has  an elegant formulation.   Recollect that
- our language permits lazy {\LET}s whose definitions can be recursive,
- i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
- $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
- recursive function (named $\mathit{f}$ in the rule) which captures an
- arbitrary number of ``unrollings'' of the definition\footnote{The $*$
-   in  the definition  of ${\mathit  f}$  represents non-deterministic
-   choice,  but  can  be  treated   as  a  constant  for  purposes  of
-   analysis.}.   We pass to  the demand  transformer $\Lfonly_{\mathit
-   f}$ of this function the demand  arising out of the liveness of $x$
- at the beginning of $e$. This  gives the liveness of the variables of
- $s$.
-
-\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
-  of  an   expression  $e$.   The  cases  $\SRETURN$   and  $\SIF$  are
-  straightforward, but note the liveness $x.\epsilon$ generated by the
-  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
-  three-address instruction:  the liveness of  $e$ is given  by taking
-  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
-  any incremental  liveness from  $s$.  The main  subtlety is  how the
-  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
-  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-We make three observations: firstly the rule ({\sc live-define}) has a
-least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
-secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
-inference of mutually recursive  function definitions, and thirdly the
-asymmetry of  demand and liveness (compared to  post- and pre-liveness
-classically) is due to the functional formulation here.
-}%mycomment
-
-Section~\ref{sec:computing}   shows   how   the  demand   transformers
-$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
-analysis) can be safely approximated  by a {\em procedure summary} for
-each function  (unifying the contexts  in the style of  0-CFA)\@.  The
-summary consists  of a pair of  a single demand and,  for this demand,
-the  corresponding  tuple  of   demands  the  function  makes  on  its
-arguments.
-
-{\color{red} Proof of correctness of the liveness analysis...}
-
-\section{Computing liveness and its encoding as a
-table}\label{sec:computing}
-Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
-and  {\color  {red}  proved  it  correct  with  reference  to  a  {\em
-    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
-together described a liveness set for each program point in a function
-in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
-describe how to obtain demand transformers \Lfonly\ from the rule {\sc
-  live-define} and how to compute the specific demand $\sigma$ on each
-function.
-
-To  answer these  questions, it  is convenient  to cast  the equations
-arising out of the $\mathit{ref}$  and $\mathcal{L}$ as the rules of a
-grammar.   But first  we modify  the rules  themselves to  a different
-form.
-
-\subsubsection{Modifying the liveness rules.}
-
-The      $\mathit{ref}$     rule      for     \CONS,      shown     in
-Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
-\acar\ and \acdr\  from the access paths in  $\sigma$.  Similarly, the
-rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
-$\emptyset$, if  $\sigma$ itself $\emptyset$.  To  realize these rules
-$\sigma$ needs to be known. This creates difficulties since we want to
-solve the equations arising out of liveness symbolically.
-
-The  solution  is  to   also  treat  the  operations  mentioned  above
-symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
-\bcdr\ are defined as:
-\begin{align*}
- \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
- \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
-\end{align*}
-and \cbar\ as:
-\begin{align*}
- \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
-                       & \{\epsilon\} ~\mbox{otherwise}
-\end{align*}
-
-We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
-\begin{align*}
-&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
-\mbox{and} \\
-&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          =   x.\cbar\sigma \cup x.0\sigma
-\end{align*}
-and the \Lfunonly\ rule
-for \SIF\ as:
-\begin{align*}
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
-                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
-        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
-        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
-\end{align*}
-The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
-similar to \CAR.
-The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
-relation $\hookrightarrow$ over sets of access paths:
-\begin{align*}
-  &\sigma_1\bcar\sigma_2  \hookrightarrow
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
-&\sigma_1\bcdr\sigma_2  \hookrightarrow
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
-  \acdr\sigma_2',~\mbox{and}\\
-&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
-\end{align*}
-
-The reflexive  transitive closure of  $\hookrightarrow$ will
-be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
-following proposition relates  the original and the modified
-liveness rules.
 
@@
 
+Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
+analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
+computes a  map that takes a  e-path in ${\mathit f}$  and returns the
+liveness environment  arising out of  all uses of variables  along the
+e-path.  A second map $\cal{P}_{\mathit  f}$ maps any program point in
+$\mathit f$  to the set  of e-paths sharing  the point.  If  a program
+point happens to fall on more  than one such e-path, then the liveness
+environment of  the program  point is the  variable-wise union  of the
+liveness  environments  of the  individual  e-paths.   In the  example
+program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
+ep_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(ep_1)  \cup {\cal
+  M}(ep_2) \cup  {\cal M}(ep_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
+\sigma, \Lfonly)$.  In the  sequel we shall only define $\mathcal{L}$.
+The formulation  of $\cal{P}$ is easy  and we do not  elaborate it any
+further.
+
+Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\SRETURN}.
+The  expression $(\SRETURN~\psi:x)$  with a  demand $\sigma$  forces an
+evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
+  \{x.\sigma\}] $  needs to  be created, where the $ep$ is a fresh
+evaluation path terminating in $\psi$.  The  map for  the expression
+$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union  of the maps of $e_1$ and
+$e_2$. In addition, since the condition also triggers an evaluation of
+$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
+added  to the  union.  Also  notice a  consequences  of laziness---the
+entire  expression including  the condition  is not  evaluated  if the
+demand on it is $\emptyset$.  This results in the empty map, which, by
+abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.
+
+ The liveness  of {\LET} has  an elegant formulation.   Recollect that
+ our language permits lazy {\LET}s whose definitions can be recursive,
+ i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
+ $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
+ recursive function (named $\mathit{f}$ in the rule) which captures an
+ arbitrary number of ``unrollings'' of the definition\footnote{The $*$
+   in  the definition  of ${\mathit  f}$  represents non-deterministic
+   choice.  During   liveness  it  can  be  assumed   to  generate  no
+   liveness.}.   We pass to  the demand  transformer $\Lfonly_{\mathit
+   f}$ of this function the demand  arising out of the liveness of $x$
+ at the beginning of $e$. This  gives the liveness of the variables of
+ $s$.
+
+\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
+  of  an   expression  $e$.   The  cases  $\SRETURN$   and  $\SIF$  are
+  straightforward, but note the liveness $x.\epsilon$ generated by the
+  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
+  three-address instruction:  the liveness of  $e$ is given  by taking
+  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
+  any incremental  liveness from  $s$.  The main  subtlety is  how the
+  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
+  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
+
+We make three observations: firstly the rule ({\sc live-define}) has a
+least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
+secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
+inference of mutually recursive  function definitions, and thirdly the
+asymmetry of  demand and liveness (compared to  post- and pre-liveness
+classically) is due to the functional formulation here.
+}%mycomment
+
+Section~\ref{sec:computing}   shows   how   the  demand   transformers
+$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
+analysis) can be safely approximated by a {\em procedure summary}. for
+each function.   The summary  is in the  form of a  demand transformer
+that maps a demand on a call to the function to demands on each of its
+arguments.
+
+{\color{red} Proof of correctness of the liveness analysis...}
+
+\section{Computing liveness and its encoding as DFA}\label{sec:computing}
+Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
+and  {\color  {red}  proved  it  correct  with  reference  to  a  {\em
+    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
+together described a liveness set for each program point in a function
+in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
+describe how to obtain demand transformers \Lfonly\ from the rule {\sc
+  live-define} and how to compute the specific demand $\sigma$ on each
+function.
+
+To  answer these  questions, it  is convenient  to cast  the equations
+arising out of $\mathit{ref}$  and $\mathcal{L}$ as the rules of a
+grammar.   But first  we modify  the rules  themselves to  a different
+form.
+
+\subsection{Modifying the liveness rules.}
+
+The      $\mathit{ref}$     rule      for     \CONS,      shown     in
+Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
+\acar\ and \acdr\  from the access paths in  $\sigma$.  Similarly, the
+rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
+$\emptyset$, if  $\sigma$ itself $\emptyset$.  To  realize these rules
+$\sigma$ needs to be known. This creates difficulties since we want to
+solve the equations arising out of liveness symbolically.
+
+The  solution  is  to   also  treat  the  operations  mentioned  above
+symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
+\bcdr\ are defined as:
+\begin{align*}
+ \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
+ \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
+\end{align*}
+and \cbar\ as:
+\begin{align*}
+ \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
+                       & \{\epsilon\} ~\mbox{otherwise}
+\end{align*}
+
+We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
+\begin{align*}
+&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
+= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
+\mbox{and} \\
+&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
+          =   x.\cbar\sigma \cup x.0\sigma
+\end{align*}
+and the \Lfunonly\ rule
+for \SIF\ as:
+\begin{align*}
+\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
+                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
+        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
+        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
+\end{align*}
+The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
+similar to \CAR.
+The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
+relation $\hookrightarrow$ over sets of access paths:
+\begin{align*}
+  &\sigma_1\bcar\sigma_2  \hookrightarrow
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
+&\sigma_1\bcdr\sigma_2  \hookrightarrow
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
+  \acdr\sigma_2',~\mbox{and}\\
+&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
+\end{align*}
+
+The reflexive  transitive closure of  $\hookrightarrow$ will
+be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
+following proposition relates  the original and the modified
+liveness rules.
+
+\begin{figure*}[t!]
+  \begin{tabular}{cc}
+    \begin{minipage}{.45\textwidth}
+        \small
+        \renewcommand{\arraystretch}{1}{
+          \begin{uprogram}
+            \UNL{1} (\DEFINE\ (\length~\xl)
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
+(\NULLQ~\xl) \IN
+            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
+$\psi_1\!\!:\,$ \px
+            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
+            $(\LET\ \pv\ $\leftarrow 0$ \IN
+            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
+(\SRETURN~\psi_2:\pv)$
+            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
+$\leftarrow$  (\CDR~\xl)  \IN
+            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
+$\leftarrow$  (\length~\pu)  \IN
+            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
+            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
+            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
+(\SRETURN~\psi_3:\pz)$)))))))
+        \end{uprogram}}
+        \renewcommand{\arraystretch}{1}{
+	  \begin{uprogram}
+	  \UNL{1} $(\DEFINE\ (\main)$
+          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots
+\IN
+	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\
+$\leftarrow$  \ldots \IN
+          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
+          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
+          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\,
+(\SRETURN~\psi_4:\pw)))))$
+\end{uprogram}}
+        %%}
+    \end{minipage}
+
+    &
+
+    \begin{minipage}{.51\textwidth}
+
+        \small
+\begin{eqnarray*}
+        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
+              [\;ep_1 \mapsto \lbrace \px.\sigma_{\length},
+\xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace,\\
+ &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma_{\length},
+                \py.\cbar\sigma_{\length},
+\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},\\
+&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
+\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
+                \rbrace~\rbrack \\
+            \cal{P_{\mbox{\tiny \length}}} &=&
+                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
+                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
+                  \rbrace, \\
+&&\;\;  \pi_3 \mapsto \lbrace ep_2
+                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
+                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
+ &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
+\pi_7 \mapsto
+                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
+\rbrace~\rbrack\\
+              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
+                [\;
+                  ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}},
+\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots\rbrace
+                  ]\\
+              \cal{P_{\mbox{\tiny \main}}} &=&
+                  \lbrack~\pi_6 \mapsto \lbrace ep_4 \rbrace,  \
+                   \pi_7 \mapsto \lbrace ep_4 \rbrace,  \
+                  \pi_8 \mapsto \lbrace ep_4 \rbrace~\rbrack
+\end{eqnarray*}
+    \end{minipage} \\
+(a)&
+(b)
+    %% \end{ tabular}}
+  \end{tabular}
+%\end{picture}
+\kern -3ex
+%\vspace{0.5in}
+\caption{(a) Example program and (b) its liveness
+maps.}\label{fig:mot-example2}
+\end{figure*}
+%================================================================
+
 \begin{proposition}
@@
 
-{\color{red} \medskip \hrule \hrule \hrule  \medskip }
+%Thick red line 
+%{\color{red} \medskip \hrule \hrule \hrule  \medskip }
 
@@
 
-\section{Generating Liveness Equations}
+\subsection{Generating Liveness Equations}
 The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running

### 2015-06-03 paper/paper.tex
implemenatation

@@
 \newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
+\newcommand{\cred}[1]{{\color{red}{(#1)}}}
 \newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
@@
 
-%\hspace*{2cm}{\epsfig{file=motiv-example.eps, height=4cm}}
+\hspace*{1cm}{\epsfig{file=motiv-example.eps, height=4cm}}
 
@@
 % GC assumptions
-In this section we describe in detail the process of liveness based garbage collection. 
-During the execution of a program if it runs of out of memory, execution is stopped and the 
-garbage collector is invoked. At this point all the references on the stack are checked for 
-liveness and only live references are copied and the program continues execution. 
-The main complexity in our algorithm is the way closures are handled. The following sub-sections 
-show two diffrent ways of handling closures along with their pros and cons.
+\cred{Introduction}
+
+In this  section we describe in  detail the process  of liveness based
+garbage collection.  During  the execution of a program  if it runs of
+out  of memory,  execution is  stopped  and the  garbage collector  is
+invoked. At this point all the references on the stack are checked for
+liveness and only live references are copied and the program continues
+execution.  The main  complexity in our algorithm is  the way closures
+are  handled. The  following sub-sections  show two  different  ways of
+handling closures along with their pros and cons.
 \subsection{A simple LGC Algorithm}
+\cred{Introduce terminology}
 % Abstract description of the GC algorithm
@@
 procedure copy (cell)
+ IF {cell is live}
    new_cell <-  dup_heapcell(cell)
    IF{celltype is cons}
-     IF {cell.car is live}
         new_car        <-  copy(cell.car)
         new_cell.car   <-  new_car
-     ENDIF
-     IF {cell.cdr is live}
+
         new_cdr        <-  copy(cell.cdr)
@@
   IF {celltype is closure}
-    new_cell      <-  dup_heapcell(cell)
-    IF {cell.arg1 is live}
        new_cell.arg1 <-  copy(cell.arg1)
-    ENDIF
-    IF {cell.arg2 is live}
        new_cell.arg2 <-  copy(cell.arg2)
-    ENDIF
    ENDIF
@@
 and recursively copying. In case of closures, it is more complicated as they might refer to data/closures 
-which might be created in a different environment(function). To overcome this we we carry some extra data 
+which might be created in a different environment(function), all of
+which might not even exist on the program stack. To overcome this we we carry some extra data 
 with each closure, we carry the program point at which it was evaluated and also the names of the free 
@@
 closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
-what is found at the creation point.
+what is found at the creation point. \cred{repeated traversals if sharing}.
 
 \subsection{Modified GC Algorithm}
-Since the first algorithm requires a lot of bookkeeping and is not precise, we use a modified algorithm 
+Since the first algorithm requires a lot of bookkeeping and is not
+precise, we have a second version of the algorithm 
 which does not use liveness for copying closures. 
@@
 and ensure references marked non-live are never copied using reachability. We discuss this issue 
-and its solution in sub-section[].
+and its solution in sub-section[]\comment{hanging like a sore thumb}.
 
@@
  The  same is not true in the case  of LGC, as the set
-of  liveness paths  may not  be similar  in both  the  traversals. One
+of  liveness paths  may not  be similar  in both  the
+traversals. \cred{Push earlier: One
 optimization is  to keep  track of all  the liveness  automaton states
@@
 it can hold  all the automata states.  Therefore it  is not a feasible
-solution. In case of lazy  languages since we use reachability to copy
+solution.} In case of lazy  languages since we use reachability to copy
 closures, some cons  cells might be copied using  reachability. We can
@@
 function by using a flag in  the heap cell. The flag ensures that only
-the required parts  of a cons cell are copied.
+the required parts  of a cons cell are copied.\cred{Very sketchy: Make
+  another attempt to write}
+
 %% \begin{figure}

### 2015-06-04 paper/paper.tex
Partially fixed full liveness for closures and updated implementation section

@@
 \begin{verbatim}
+proceudre lgc()
+ FOREACH reference in root set
+ DO
+   IF reference is live
+     copy(cell)
+   ENDIF
+ DONE
+
 procedure copy (cell)
@@
 which might be created in a different environment(function), all of
-which might not even exist on the program stack. To overcome this we we carry some extra data 
-with each closure, we carry the program point at which it was evaluated and also the names of the free 
+which might not even exist on the program stack. To overcome this we need to carry extra information 
+with each closure, the program point at which it was evaluated and also the names of the free 
 variables contained in the closure. This gives us precise liveness information of the free variables of a 
@@
 closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
-what is found at the creation point. \cred{repeated traversals if sharing}.
+what is found at the creation point.
+
+The advantage of this method is that it collects all dead cells even if they are part of a closure and hence 
+will reduce the memory requirements and the number of garbage collections. The main drawbacks of this method 
+is the fact that the algorithm might do multiple traversals over the same structure. Assume that two references 
+point to the same list, now using reachability if it is garbage collected using one reference there is no need 
+to traverse using the other reference as everything reachable from the root node would have been copied. 
+Unfortunately, this is not the case with liveness based GC as the two references might make different sections 
+of the list as live and copy them. Liveness based algorithm will have to make one traversal per reference. 
+Experimentally we found that this was a major source of overhead in the liveness based algorithm. 
+This can be mitigated by maintaining a list of liveness automata states that a particular heap cell was 
+traversed with and ensuring that we do not repeat traversals with the same state. This requires that each 
+cell maintain a list of state which potentially might contain all the states. Hence this is not a feasible 
+solution.
+\cred{repeated traversals if sharing}.
 
@@
 procedure copy (cell)
+  IF{celltype is cons}
+    IF cell is live &&
+         cell is not copied using reachability
        new_cell <-  dup_heapcell(cell) 
- IF{celltype is cons}
-     IF {cell.car is live}
        new_car        <-  copy(cell.car)
        new_cell.car   <-  new_car
-     ENDIF
-     IF {cell.cdr is live}
        new_cdr        <-  copy(cell.cdr)
        new_cell.cdr   <-  new_cdr
+    ELSE
+       cell <- NULL
     ENDIF
@@
 assume everything under it to be live and copy everything that is reachable. Although this might 
-lead to copying more cells it is faster as there are no liveness checks. Additionally we can 
-optimize some of the processing by not traversing cells more than once. This is described in 
-sub-section[]. \comment{Once experimental results are obtained compare the two algorithms here}
+lead to copying more cells it is faster as there are no liveness checks. 
+\comment{Once experimental results are obtained compare the two algorithms here}
 Since we mix both liveness based GC and reachability the garbage collector has to be more careful 
 and ensure references marked non-live are never copied using reachability. We discuss this issue 
-and its solution in sub-section[]\comment{hanging like a sore thumb}.
-
-\subsection{Optimization for RGC for closures}
-One of the  reasosn why reachability based GC  is faster than liveness
-based GC is  due to the fact that  in case we try to copy  a cons cell
-which is  already copied  we do  not need to  traverse it  any further.
- The  same is not true in the case  of LGC, as the set
-of  liveness paths  may not  be similar  in both  the
-traversals. \cred{Push earlier: One
-optimization is  to keep  track of all  the liveness  automaton states
-that a particular  cons cell has been visited with  and ensure that we
-do  not  traverse  it  with  the  same state  more  than  once.   This
-optimization requires extra memory  for each heap cell and potentially
-it can hold  all the automata states.  Therefore it  is not a feasible
-solution.} In case of lazy  languages since we use reachability to copy
-closures, some cons  cells might be copied using  reachability. We can
-avoid  repeated traversals of  these cons  cells without  tracking the
-states. This only requires the  knowledge of whether the cons cell was
-copied using  reachability.  Hence  we use a  flag to remember  if the
-heap cell was  copied using reachability at any time, if  it is set we
-do not traverse any further.\comment{Maybe add a diagram to make it clearer}
-
-\subsection{Issues with mixing LGC and RGC}
+and its solution in sub-section[]\cred{hanging like a sore thumb}.
+
+\subsubsection{Optimization for RGC for closures}
+An advantage of using reachablity for closures is that it opens up 
+certain opportunities of optimization. As we saw in section[], multiple 
+traversals over the same structure was a major inefficiency in the full 
+liveness based algorithm and overcoming it required a lot of memory.
+Using reachability for closures allows us to overcome this partially 
+using limited amount of memory. Any heap cell which is part of a closure 
+is copied using reachaibilty and hence by maintaining a single flag in each 
+cell to indicate whether it was copied using reachability prevents 
+traversing the same cell multiple times. Experimental results again 
+indicate that for lazy languages this gives a substantial performance 
+improvement.
+\comment{Maybe add a diagram}
+
+\subsubsection{Issues with mixing LGC and RGC}
 A consequence of  copying everything under a closure  is the fact that

### 2015-06-04 lazy-scheme/src/gc.cpp

@@ -1669,7 +1669,7 @@ void liveness_gc()
 
 
 
-#ifndef __DEBUG__GC
+#ifdef __DEBUG__GC
 	  ofstream pre("GC" + to_string(gccount) + ".txt", ios_base::out);
 	  pre << "Doing liveness based GC #" << gccount << " after " << num_of_allocations << " allocations"<<endl;
 	  ofstream pregc("PreGC" + to_string(gccount) + ".txt", ios_base::out);
@@ -1693,7 +1693,6 @@ void liveness_gc()
     	  cerr << "Looking up " << nodeName << endl;
     	  if (got != statemap.end())
     	  {
-
     		  cons *addr   = followpaths(static_cast<cons*>(vhit->ref), got->second);
     		  vhit->ref = addr;
 
@@ -1707,6 +1706,8 @@ void liveness_gc()
   cerr << "Completed liveness GC" << endl;
   update_heap_ref_stack(pre, 1);
   clear_live_buffer(pre);
+  cerr << "Copied " << numcopied << " cells" << endl;
+  numcopied = 0;
 #ifdef __DEBUG__GC
 	pre.close();
 	ofstream postgc("PostGC" + to_string(gccount) + ".txt", ios_base::out);
@@ -1801,6 +1802,7 @@ cons* followpaths(cons* loc, state_index index)
   case funcApplicationExprClosure:
   case funcArgClosure:
   {
+      //What happens to arg1? Shouldn't it get updated during GC?
 	  string liveness_string = "L/" + *(loccopy->val.closure.prog_pt) + "/" + *(loccopy->val.closure.arg2_name);
 
 	  auto liveness_state = statemap.find(liveness_string);
@@ -1809,6 +1811,15 @@ cons* followpaths(cons* loc, state_index index)
 		  cerr << "Doing GC corresponding to " << liveness_string << endl;
 		  auto new_arg2 = followpaths(loccopy->val.closure.arg2, liveness_state->second);
 		  loccopy->val.closure.arg2 = new_arg2;
+	  }
+
+	  if (loccopy->val.closure.arg1 != NULL)
+	  {
+		  //More arguments exist and have to be copied
+		  auto liveness_state = statemap.find("L/-1/c");
+		  //Using dummy state to always make it live
+		  auto new_arg1 = followpaths(loccopy->val.closure.arg1, liveness_state->second);
+		  loccopy->val.closure.arg1 = new_arg1;
 	  }
   }
   break;

### 2015-06-04 paper/paper.tex
liveness

--- 2015-06-03 paper/paper.tex
@@
 \newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
-\newcommand{\cred}[1]{{\color{red}{(#1)}}}
+\newcommand{\cred}[1]{{\color{red}{#1}}}
 \newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
@@
 \begin{align*}
- \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
+ \cbar\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                        & \{\epsilon\} ~\mbox{otherwise}
 \end{align*}
-
+Note that \cbar\ is idempotent. 
 We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
@@
   \begin{tabular}{cc}
-    \begin{minipage}{.45\textwidth}
+    \begin{minipage}{.40\textwidth}
         \small
@@
         \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
-              [\;ep_1 \mapsto \lbrace \px.\sigma_{\length},
-\xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace,\\
- &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma_{\length},
-                \py.\cbar\sigma_{\length},
-\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},\\
+              [\;ep_1 \mapsto \lbrace \px.\cbar \sigma,
+\xl.\cbar\sigma\rbrace ,~ep_2 \mapsto \lbrace \cred{
+\pv.\sigma}  \rbrace,\\
+ &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma,
+                \py.\cbar\sigma,
+\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma},\\
 &&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; 
-\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
-                \rbrace~\rbrack \\
+\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\cbar\sigma} \cup 
+%%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
+\cred{\cbar\Lf{\length}{1}{\cbar\sigma}}\}\}
+             \rbrack \\
             \cal{P_{\mbox{\tiny \length}}} &=&
@@
                 [\;
-                  ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}},
-\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots\rbrace
-                  ]\\
+                  ep_4 \mapsto \lbrace \pw.\sigma,
+           \pb.\Lf{\length}{\mbox{1}}{\sigma}, \ldots\rbrace
+                   ] \\
                \cal{P_{\mbox{\tiny \main}}} &=&
-                  \lbrack~\pi_6 \mapsto \lbrace ep_4 \rbrace,  \
-                   \pi_7 \mapsto \lbrace ep_4 \rbrace,  \
-                  \pi_8 \mapsto \lbrace ep_4 \rbrace~\rbrack
+                   \lbrack~\pi_9 \mapsto \lbrace ep_4 \rbrace,  \
+                    \pi_{10} \mapsto \lbrace ep_4 \rbrace,  \
+                   \pi_{11} \mapsto \lbrace ep_4 \rbrace,  \pi_{12} \mapsto \lbrace ep_4 \rbrace~\rbrack
 \end{eqnarray*}
@@
 && \Lanv{1}{\xl} = \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
+  \cup \cbar\Lf{\length}{1}{\cbar\sigma}
 \end{eqnarray*}
@@
    && \Lf{\length}{1}{\sigma}
-    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
+    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma} \cup \cbar\Lf{\length}{1}{\cbar\sigma}
 \end{eqnarray*}
@@
 where \Df{f}{i} are sets of strings over the alphabet mentioned above.
-  Substituting the guessed form in
-the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
-$\sigma$,  we get  an equation  for \Df{f}{i}  that is  independent of
-$\sigma$.   Any   solution  for   \Df{f}{i}  yields  a   solution  for
-\Lfonly$_{\mathit  f}$.  Applied  to  \Lfonly$_{\mathit \length}$,  we
-get:
+Substituting   the   guessed   form   in   the   equation   describing
+\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
+for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
+\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
+\Lfonly$_{\mathit \length}$, we get:
   \begin{eqnarray*}
-  &&  \Df{\length}{1}.\sigma = \cbar.\sigma
-    \cup \acdr\Df{\length}{1}(\cbar\sigma), ~\mbox{or}\\
-  &&  \Df{\length}{1} = \lbrace\cbar\rbrace
-    \cup \acdr\Df{\length}{1}\cbar
+&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}.\sigma,~\mbox{and}\\
+&&   \Df{\length}{1} = \cbar \cup \acdr\Df{\length}{1}\cbar 
+       \cup \cbar\Df{\length}{1}.{\cbar}
   \end{eqnarray*}
@@
 Note that this equation can also be viewed as a CFG with \{\acdr,
-\cbar\} as the terminal symbols and \Df{\length}{1} as the sole
+\cbar\} as terminal symbols and \Df{\length}{1} as the sole
 non-terminal.
@@
 
-To avoid analysing the body of a function for each call, we calculated
+To avoid analyzing the body of a function for each call, we calculated
 the liveness  for the arguments and  the variables in  a function with
@@
 calls by a {\em summary demand} and calculate the liveness at each GC
-point inside the  function based on this summary  demand.  Even though
+point \comment{have we defined what a garbage collection point
+  is?} inside the  function based on this summary  demand.  Even though
 we  lose precision  due to  this, our  experiments show  that  this is
@@
 Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
-$\Lanv{i}{x}$ to  the demand $\sigma_f$.  Assuming that  there are $k$
+$\Lanv{i}{x}$ to  the demand $\sigma_f$. Let us denote the
+contribution of a call site $\pi$ in a function $g$ to the overall
+demand on the function $f$ as $\deltacall{f}{\pi}{g}$. Assuming that  there are $k$
 call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
@@
 $\pi_{10}$ and a recursive call at $\pi_6$.
-%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
 So $\sigma_{\length} =
@@
 \sigma_{\length}    &=&
- \sigma_{\!all}  ~\cup~
-\acdr\Lf{\length}{1}{\cbar\sigma_{\mathit{\length}}}
+ \sigma_{\!all}  ~\cup~{\cbar\sigma_{\mathit{\length}}}
 \end{eqnarray*}
-For some of the program points in \length\  and \main,  this gives
-{\color {red} Prasanna, please fill this correctly}. You could use the
-set notation if the map notation spills over:
+As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pb} in terms
+of  $\sigma_{\length}$ are:
 \begin{align*}
-\Lan{1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length},
-\pz.\sigma_{\length},\
-              \py.\cbar\sigma_{\length}, \\
-&~~~~~~~\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
-\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length},
-\pz.\sigma_{\length},\
-                  \py.\cbar\sigma_{\length}, \\
-&\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
-\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
-\Lan{5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
-\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
-\Lan{8} &= \lbrace \pw.\sigma_{\mathit{\!all}},
-\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace
+\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Lf{\length}{1}{\cbar\sigma_{\length}}
+  \cup \cbar\Lf{\length}{1}{\cbar\sigma_{\length}}
+ \\
+\Lanv{9}{\pb} &= \Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}
 \end{align*}
@@
 the summary demand  $\sigma_{\mathit f}$ on $e_f$.
-% \marginpar{Consistency: summary vs cumulative?}
 \item For each function $\mathit{f}$ (including $\mainpgm$ for
-  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages?
-No, we define it only in the implementation section as of now.}
+  $e_\mainpgm$) 
 an equation defining
   liveness at each GC point of $e_{\mathit f}$.
-
+\end{enumerate}
 \subsection{Solving liveness equations---the grammar
@@
 
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
-pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
-process is  continued till  a fixed point  is reached.   When the
-fixed point  is reached, the resulting NFA  contains all possible
-reductions corresponding  to all the  paths in the  original NFA.
-The proofs  of the termination  and correctness of  the algorithm
-are given in an earlier paper~\cite{karkare07liveness}.
+The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
+of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
+is continued till  a fixed point is reached.  When  the fixed point is
+reached,   the  resulting   NFA  contains   all   possible  reductions
+corresponding to all the paths in the original NFA.  The proofs of the
+termination and correctness  of the algorithm are given  in an earlier
+paper~\cite{karkare07liveness}.
 
 %%\input{liveness-dfa.tex}
-\comment{change according to example}
-We  illustrate  the  algorithm  in  Figure~\ref{algo:simplify-NFA}  by
-constructing      the       automaton      for      \var{\Lanv{9}{y}}.
-Figure~\ref{fig:automata-construction}  (c)  shows  the automaton  for
-\var{\Lanv{9}{y}}   constructed   by   composing  the   automata   for
-\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
-\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
-get (d). We  add an $\epsilon$ edge from $q_2$  to $q_3$ bypassing the
-\bcdr\acdr\ pair from $q_2$ to $q_2$  and then to $q_3$. This is shown
-in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
-is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
-$q_2$.   Removing this $\epsilon$  edge gives  the automaton  shown in
-(g).  Restricting  this automaton  to forward edges  only, we  get the
-final   automaton   shown    in   (h).   This   automaton   recognizes
-$\acdr^{*}\!\mid\acdr^{*}\acar  \sigma_{\mathit \!all}$,  showing that
-the entire list \py, including  its elements, is live at $\pi_9$. Also
-note that the  language accepted by the final  automaton satisfies the
-prefix-closed property.
+\comment{change according  to example} We illustrate  the algorithm in
+Figure~\ref{algo:simplify-NFA}  by   constructing  the  automaton  for
+\var{\Lanv{9}{y}}.   Figure~\ref{fig:automata-construction}  (c) shows
+the  automaton  for  \var{\Lanv{9}{y}}  constructed by  composing  the
+automata   for    \var{\Uf{\append}{1}},   \var{\Df{\append}{1}}   and
+$(\epsilon \mid \acdr \mid \acdr\acar\sigma_{\mathit{\!all}})$.  After
+$\epsilon$ removal we get (d). We add an $\epsilon$ edge from $q_2$ to
+$q_3$ bypassing the  \bcdr\acdr\ pair from $q_2$ to  $q_2$ and then to
+$q_3$. This  is shown in (e).   The $\epsilon$ edge is  removed in (f)
+and  a second  $\epsilon$  is  added to  the  automaton bypassing  the
+\bcar\acar\ pair  from $q_1$ to $q_2$.  Removing  this $\epsilon$ edge
+gives  the automaton  shown  in (g).   Restricting  this automaton  to
+forward edges  only, we  get the final  automaton shown in  (h).  This
+automaton  recognizes  $\acdr^{*}\!\mid\acdr^{*}\acar  \sigma_{\mathit
+  \!all}$, showing  that the entire list \py,  including its elements,
+is live at $\pi_9$. Also note  that the language accepted by the final
+automaton satisfies the prefix-closed property.
 
@@
 \end{verbatim}
-In the simple algorithm presented we consult liveness automata for copying both \CONS cells and closures. 
-Copying \CONS cells are simple and just involve consulting the liveness automaton corresponding to them 
-and recursively copying. In case of closures, it is more complicated as they might refer to data/closures 
-which might be created in a different environment(function), all of
-which might not even exist on the program stack. To overcome this we we carry some extra data 
-with each closure, we carry the program point at which it was evaluated and also the names of the free 
-variables contained in the closure. This gives us precise liveness information of the free variables of a 
-closure but involves run-time overhead of updating the program point correctly. To avoid this overhead we 
-go for a safer approximation by storing the creation point of the closure and using it to check liveness. 
-This is safe since the liveness at the creation point of the closure encompasses all possible uses of the 
-closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
-what is found at the creation point. \cred{repeated traversals if sharing}.
+In  the simple algorithm  presented we  consult liveness  automata for
+copying both \CONS cells and closures.  Copying \CONS cells are simple
+and just  involve consulting  the liveness automaton  corresponding to
+them  and  recursively  copying.  In  case of  closures,  it  is  more
+complicated  as  they might  refer  to  data/closures  which might  be
+created in  a different environment(function), all of  which might not
+even exist  on the program  stack. To overcome  this we we  carry some
+extra data with  each closure, we carry the program  point at which it
+was evaluated  and also the names  of the free  variables contained in
+the closure.  This gives us  precise liveness information of  the free
+variables of a closure but  involves run-time overhead of updating the
+program  point correctly. To  avoid this  overhead we  go for  a safer
+approximation by storing  the creation point of the  closure and using
+it to check liveness.  This is safe since the liveness at the creation
+point of the closure encompasses  all possible uses of the closure but
+it might be very imprecise as at run-time the actual liveness required
+might   be   way  less   than   what   is   found  at   the   creation
+point. \cred{repeated traversals if sharing}.
 
@@
 % That's all folks!
-\end{enumerate}
 

### 2015-06-04 paper/paper.tex

@@
 \end{verbatim}
-In  the simple algorithm  presented we  consult liveness  automata for
-copying both \CONS cells and closures.  Copying \CONS cells are simple
-and just  involve consulting  the liveness automaton  corresponding to
-them  and  recursively  copying.  In  case of  closures,  it  is  more
-complicated  as  they might  refer  to  data/closures  which might  be
-created in  a different environment(function), all of  which might not
-even exist  on the program  stack. To overcome  this we we  carry some
-extra data with  each closure, we carry the program  point at which it
-was evaluated  and also the names  of the free  variables contained in
-the closure.  This gives us  precise liveness information of  the free
-variables of a closure but  involves run-time overhead of updating the
-program  point correctly. To  avoid this  overhead we  go for  a safer
-approximation by storing  the creation point of the  closure and using
-it to check liveness.  This is safe since the liveness at the creation
-point of the closure encompasses  all possible uses of the closure but
-it might be very imprecise as at run-time the actual liveness required
-might   be   way  less   than   what   is   found  at   the   creation
-point. \cred{repeated traversals if sharing}.
+In the simple algorithm presented we consult liveness automata for copying both \CONS cells and closures. 
+Copying \CONS cells are simple and just involve consulting the liveness automaton corresponding to them 
+and recursively copying. In case of closures, it is more complicated as they might refer to data/closures 
+which might be created in a different environment(function), all of
+which might not even exist on the program stack. To overcome this we need to carry extra information 
+with each closure, the program point at which it was evaluated and also the names of the free 
+variables contained in the closure. This gives us precise liveness information of the free variables of a 
+closure but involves run-time overhead of updating the program point correctly. To avoid this overhead we 
+go for a safer approximation by storing the creation point of the closure and using it to check liveness. 
+This is safe since the liveness at the creation point of the closure encompasses all possible uses of the 
+closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
+what is found at the creation point.
+
+The advantage of this method is that it collects all dead cells even if they are part of a closure and hence 
+will reduce the memory requirements and the number of garbage collections. The main drawbacks of this method 
+is the fact that the algorithm might do multiple traversals over the same structure. Assume that two references 
+point to the same list, now using reachability if it is garbage collected using one reference there is no need 
+to traverse using the other reference as everything reachable from the root node would have been copied. 
+Unfortunately, this is not the case with liveness based GC as the two references might make different sections 
+of the list as live and copy them. Liveness based algorithm will have to make one traversal per reference. 
+Experimentally we found that this was a major source of overhead in the liveness based algorithm. 
+This can be mitigated by maintaining a list of liveness automata states that a particular heap cell was 
+traversed with and ensuring that we do not repeat traversals with the same state. This requires that each 
+cell maintain a list of state which potentially might contain all the states. Hence this is not a feasible 
+solution.
+\cred{repeated traversals if sharing}.
 

### 2015-06-05 paper/paper.tex
Updated print stack subsection

@@
 \comment{Once experimental results are obtained compare the two algorithms here}
-Since we mix both liveness based GC and reachability the garbage collector has to be more careful 
-and ensure references marked non-live are never copied using reachability. We discuss this issue 
-and its solution in sub-section[]\cred{hanging like a sore thumb}.
+Mixing liveness based GC and reachability has its own set of pros and cons which 
+we discuss further.\cred{hanging like a sore thumb}.
 
@@
 closures is  driven by the print  function\comment{Refer to SPJ's
-book}.  Hence, during  a GC there
-will be heap  references that will be present on  the stack which also
-have  to  copied/collected.   One  solution  is  to  assume  that  all
-references  present  on  the  print  stack are  fully  live  and  copy
-everything pointed  by them  akin to reachability.   This is  safe but
-copies heap  cells which  have already been  printed. Since  the print
-function is  not part of  the executing program its  liveness automata
-are not  considered. Therefore we  simulate the liveness of  the print
-function by using a flag in  the heap cell. The flag ensures that only
-the required parts  of a cons cell are copied.\cred{Very sketchy: Make
-  another attempt to write}
+book}.  
+Printing of atomic values is simple and does not trigger a garbage collection.
+In case of \CONS\ cells, the \CAR\ part is first printed and then the \CDR\ part.
+In a lazy language, both the \CAR\ and the \CDR\ parts could be closures, thus 
+requiring evaluation. A garbage collection might be triggered while printing 
+the \CAR\ part or the \CDR\ part. Therefore, whenever a garbage collection is 
+triggered we have to consider any references that might be on the print stack 
+and copy them. Liveness automata does not exist for the print function as it 
+is not part of the executing program. Since the print function makes everything 
+reachable from a cell live we can copy all cells reachable from the reference.
+
+Consider the case where a \CONS\ cells is being printed and a garbage collection 
+is triggered when printing the \CDR\ part. In this case, since the refernce to the 
+\CONS\ cells is still on the print stack both the \CAR\ and the \CDR\ parts are 
+reachable but we know that the \CAR\ part has already been printed and can be 
+safely garbage collected. A liveness analysis of the print function would have 
+given the same information. In our implementation we mimic the behavior of 
+the liveness analysis by using a flag in every \CONS\ cell which says whether 
+the \CAR\ part has been printed or not. For any \CONS\ cell if the \CAR\ part 
+has not been printed, a reference to the \CAR\ part will also be on the stack but 
+the \CDR\ part has to be explicitly copied.  
+In case the \CAR\ part has been printed, then a reference to the \CDR\ part will 
+be on the stack and since the \CAR\ part is not required anymore we do not 
+copy anything. This optimization subsumes tail call optimization. Again, 
+experimental results show that this optimization minimizes a lot of copying. 
+\cred{Very sketchy: Make another attempt to write}
 

### 2015-06-05 paper/paper.tex
Adding terminology to implementation

@@
 \cred{Introduce terminology}
+We introduce certain terminology that we use in our discussion of the 
+garbage collection algorithms. A reference is a field which holds the 
+address of a heap location. Root set refers to the set of all references 
+that are on the program stack during garbage collection.
+A heap cell can either contain an atomic value, a \CONS\ cell or a closure. A closure 
+has fields to identify the closure type and 2 arguments. In case of function calls 
+with more than 2 arguments we use currying (one of the argument fields will point to another
+ heap location containing the next argument and so on). By reachability we mean all heap 
+locations that are reachable traversing the argument fields of a given reference.
 % Abstract description of the GC algorithm
@@
     IF {celltype is closure && cell is not copied}
-       new_cell      <-  dup_heapcell(cell)
-       new_cell.arg1 <-  copy(cell.arg1)
-       new_cell.arg2 <-  copy(cell.arg2)
+       new_cell  <- copy_and_mark_reachable_cells(cell)
     ENDIF

### 2015-06-05 paper/paper.tex
the statement of undecidability

--- 2015-06-04 paper/paper.tex
@@
 \newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{lemma}[theorem]{Lemma}
 \begin{document}
@@
   \acdr\sigma_2',~\mbox{and}\\
+&\sigma_1\cbar \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
+  \emptyset \\
 &\sigma_1\cbar \emptyset \hookrightarrow \emptyset
 \end{align*}
-
 The reflexive  transitive closure of  $\hookrightarrow$ will
@@
 \label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} = \Df{f}{i}.\sigma
+  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
 \end{eqnarray}
@@
   \begin{eqnarray*}
-&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}.\sigma,~\mbox{and}\\
+&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
 &&   \Df{\length}{1} = \cbar \cup \acdr\Df{\length}{1}\cbar 
-       \cup \cbar\Df{\length}{1}.{\cbar}
+       \cup \cbar\Df{\length}{1}{\cbar}
   \end{eqnarray*}
@@
 calls by a {\em summary demand}  and calculate the liveness at each GC
-point \comment{have we defined what a garbage collection point
-  is?} inside the  function based on this summary  demand.  Even though
+point inside the  function based on this summary  demand.  Even though
 we  lose precision  due to  this, our  experiments show  that  this is
@@
 Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
-$\Lanv{i}{x}$ to  the demand $\sigma_f$. Let us denote the
-contribution of a call site $\pi$ in a function $g$ to the overall
-demand on the function $f$ as $\deltacall{f}{\pi}{g}$. Assuming that  there are $k$
+$\Lanv{i}{x}$ to the demand $\sigma_f$. Let us denote the contribution
+of a call  site $\pi$ in a  function $g$ to the overall  demand on the
+function $f$  as $\deltacall{f}{\pi}{g}$. Assuming that  there are $k$
 call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
 (in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
-$\deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
-The distinguished function  \mainpgm\ is a special case.  We assume it
-is called through a  printing mechanism with demand $\sigma_\mainpgm =
-\{\acar,\acdr\}^\ast$ (denoted $\sigma_{\!all}$).
+$\deltacall{f}{\pi^1}{g^1}          \cup          \cdots          \cup
+\deltacall{f}{\pi^k}{g^k}$.  The distinguished function \mainpgm\ is a
+special case.   We assume  it is called  through a  printing mechanism
+with   demand   $\sigma_\mainpgm   =  \{\acar,\acdr\}^\ast$   (denoted
+$\sigma_{\!all}$).
 
@@
 \begin{align*}
-\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Lf{\length}{1}{\cbar\sigma_{\length}}
-  \cup \cbar\Lf{\length}{1}{\cbar\sigma_{\length}}
+\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Df{\length}{1}\cbar\sigma_{\length}
+  \cup \cbar\Df{\length}{1}{\cbar\sigma_{\length}}
  \\
-\Lanv{9}{\pb} &= \Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}
+\Lanv{9}{\pb} &= \Df{\length}{1}\sigma_{\mathit{\!all}}
 \end{align*}
@@
 \var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{5}{\xl}}   comprises    the   following
+grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
 productions:
  \begin{eqnarray*}
-\var{\Lanv{5}{\xl}}  &\rightarrow&   \var{\acdr \Lf
-  {\length}{1}{\cbar\sigma_{\length}}} \mid \cbar\sigma_{\length} \\
+\var{\Lanv{1}{\xl}}  &\rightarrow&   \cbar\sigma_{\length} \mid \acdr \var{\Df
+  {\length}{1}}{\cbar\sigma_{\length}} \mid
+  \cbar\var{\Df{\length}{1}}{\cbar\sigma_{\length}} \\
+  \var{\Df{\length}{1}} &\rightarrow& \cbar \mid
+  \acdr\var{\Df{\length}{1}}\cbar 
+       \mid \cbar\var{\Df{\length}{1}}{\cbar}\\
+\langle {\sigma_{\length}} \rangle
+&\rightarrow&
+\sigma_{\!all}  \mid \cbar\langle{\sigma_{\length}}\rangle
 \end{eqnarray*}
-\comment{Pick up a richer equation or more than one liveness equations}
-Other equations can be converted  similarly.     The  language
-generated  by  \var{\Lanv{i}{x}},
-denoted $\mathscr{L}(\var{\Lanv{i}{x}})
-$,  is the desired  solution of
-\Lanv{i}{x}.
-However, recall from  our earlier  discussion that
-the  decision problem that  we are  interested in  during garbage
-collection is:
+Other equations can be converted similarly.  The language generated by
+\var{\Lanv{i}{x}},  denoted $\mathscr{L}(\var{\Lanv{i}{x}}) $,  is the
+desired  solution of  \Lanv{i}{x}.  However,  recall from  our earlier
+discussion that the decision problem  that we are interested in during
+garbage collection at a program point $\pi_i$ is:
 \begin{quote}
 Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
-\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let
-$\mathscr{L}(\var{\Lanv{i}{x}})
-\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists
-of forward paths only. Then does $\alpha \in \sigma$?
+\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \cbar).     Let
+$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
+where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
+\sigma$?
 \end{quote}
-We could convert  the  rules  defining  $\hookrightarrow$  into
-productions and  add them to the grammar. However, this results
-in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
-membership  problem   for  such  grammars   is
-undecidable. \comment{Rewrite it to say that this is a CFG with 3
-  unrestricted rules and the membership for even this grammar is
-undecidable}
-\comment{State the undecidability of the decision problem above.}
+
+We now model the above problem  as one of deciding the membership of a
+context-free  grammar  augmented  with  a fixed  set  of  unrestricted
+productions.
+
+\begin{definition}\label{def:specialgrammar} 
+Consider a form of  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
+non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \cbar, \$\}$, $p_1$
+is a set of context-free production that contains the distinguished
+production $S \rightarrow \alpha\$$, $\alpha$ is a string of grammar
+symbols that does not contain $S$, and $p_2$ is the fixed set of
+unrestricted productions $\bcar\acar \rightarrow \epsilon$,
+$\bcdr\acdr \rightarrow \epsilon$, $\cbar\acar \rightarrow \cbar$,
+$\cbar\acdr \rightarrow \cbar$, and $\cbar\$ \rightarrow
+\epsilon$. 
+\end{definition}
+
+From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
+that the results of liveness analysis of any program can be modeled by
+the kind  of grammar described above. The  following proposition shows
+that the converse also holds.
+\begin{proposition}
+Given    a     grammar    $G$    of    the     form    described    in
+Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
+program $p$  with program points  $\pi_i$ and variables $x$  such that
+the liveness  analysis of $p$  is same as  $G$ except for a  change in
+non-terminal names which now have the form \var{\Lanv{i}{x}}.
+\end{proposition}
+We now show that the decision problem required at garbage collection
+time is undecidable.
+
+\begin{lemma}
+Consider a grammar $G$ of the kind  described in
+Definition~\ref{def:specialgrammar} and a forward access path $\alpha$ consisting
+of symbols \acar\ and \acdr\  only. The decision problem $\alpha \in
+\mathscr{L}(G)$ is undecidable.  
+\end{lemma}
+
+\comment{Proof:}
+
+%% We could convert the rules defining $\hookrightarrow$ into productions
+%% and  add  them  to the  grammar.  However,  this  results in  an  {\em
+%%   unrestricted   grammar}~\cite{hopcraft90toc},  and   the  membership
+%% problem for  such grammars is undecidable. \comment{Rewrite  it to say
+%%   that this is a CFG with  3 unrestricted rules and the membership for
+%%   even this grammar  is undecidable} \comment{State the undecidability
+%%   of the decision problem above.}
 
   We circumvent the problem by over-approximating the CFG generated by
-the analysis to  {\em
-  strongly  regular}  CFGs which  have   easy
+  the  analysis  to  {\em  strongly  regular}  CFGs  which  have  easy
   translations to  non-deterministic finite state  automata (NFA). The
-NFAs are then simplified on the
-basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
-membership of forward access paths. The resulting NFAs are   finally
-converted to DFAs for use during garbage collection.
-\comment{Fix all comments and produce first draft}
+  NFAs are then simplified on the basis of the $\hookrightarrow$ rules
+  to  enable  checking of  membership  of  forward  access paths.  The
+  resulting NFAs are finally converted  to DFAs for use during garbage
+  collection.  \comment{Fix all comments and produce first draft}
 %------------------------------------------------------------%
@@
 \end{verbatim}
-In the simple algorithm presented we consult liveness automata for copying both \CONS cells and closures. 
-Copying \CONS cells are simple and just involve consulting the liveness automaton corresponding to them 
-and recursively copying. In case of closures, it is more complicated as they might refer to data/closures 
-which might be created in a different environment(function), all of
-which might not even exist on the program stack. To overcome this we need to carry extra information 
-with each closure, the program point at which it was evaluated and also the names of the free 
-variables contained in the closure. This gives us precise liveness information of the free variables of a 
-closure but involves run-time overhead of updating the program point correctly. To avoid this overhead we 
-go for a safer approximation by storing the creation point of the closure and using it to check liveness. 
-This is safe since the liveness at the creation point of the closure encompasses all possible uses of the 
-closure but it might be very imprecise as at run-time the actual liveness required might be way less than 
-what is found at the creation point.
-
-The advantage of this method is that it collects all dead cells even if they are part of a closure and hence 
-will reduce the memory requirements and the number of garbage collections. The main drawbacks of this method 
-is the fact that the algorithm might do multiple traversals over the same structure. Assume that two references 
-point to the same list, now using reachability if it is garbage collected using one reference there is no need 
-to traverse using the other reference as everything reachable from the root node would have been copied. 
-Unfortunately, this is not the case with liveness based GC as the two references might make different sections 
-of the list as live and copy them. Liveness based algorithm will have to make one traversal per reference. 
-Experimentally we found that this was a major source of overhead in the liveness based algorithm. 
-This can be mitigated by maintaining a list of liveness automata states that a particular heap cell was 
-traversed with and ensuring that we do not repeat traversals with the same state. This requires that each 
-cell maintain a list of state which potentially might contain all the states. Hence this is not a feasible 
-solution.
-\cred{repeated traversals if sharing}.
+In  the simple algorithm  presented we  consult liveness  automata for
+copying both \CONS cells and closures.  Copying \CONS cells are simple
+and just  involve consulting  the liveness automaton  corresponding to
+them  and  recursively  copying.  In  case of  closures,  it  is  more
+complicated  as  they might  refer  to  data/closures  which might  be
+created in  a different environment(function), all of  which might not
+even exist  on the program  stack. To overcome  this we need  to carry
+extra information with each closure, the program point at which it was
+evaluated and  also the names of  the free variables  contained in the
+closure.  This  gives us  precise  liveness  information  of the  free
+variables of a closure but  involves run-time overhead of updating the
+program  point correctly. To  avoid this  overhead we  go for  a safer
+approximation by storing  the creation point of the  closure and using
+it to check liveness.  This is safe since the liveness at the creation
+point of the closure encompasses  all possible uses of the closure but
+it might be very imprecise as at run-time the actual liveness required
+might be way less than what is found at the creation point.
+
+The advantage of  this method is that it collects  all dead cells even
+if  they are  part  of a  closure  and hence  will  reduce the  memory
+requirements and the number of garbage collections. The main drawbacks
+of  this method  is  the fact  that  the algorithm  might do  multiple
+traversals over  the same structure. Assume that  two references point
+to the  same list, now using  reachability if it  is garbage collected
+using  one reference  there is  no need  to traverse  using  the other
+reference as everything  reachable from the root node  would have been
+copied.  Unfortunately, this is not the case with liveness based GC as
+the two references  might make different sections of  the list as live
+and  copy  them.  Liveness  based  algorithm will  have  to  make  one
+traversal  per reference.   Experimentally we  found that  this  was a
+major source of overhead in the liveness based algorithm.  This can be
+mitigated by  maintaining a  list of liveness  automata states  that a
+particular heap  cell was traversed with  and ensuring that  we do not
+repeat traversals  with the same  state. This requires that  each cell
+maintain  a list  of state  which  potentially might  contain all  the
+states.  Hence  this  is  not  a  feasible  solution.   \cred{repeated
+  traversals if sharing}.
 
@@
 
-In this algorithm we consult the liveness automata only if it is a cons cell, for closures we 
-assume everything under it to be live and copy everything that is reachable. Although this might 
-lead to copying more cells it is faster as there are no liveness checks. 
-\comment{Once experimental results are obtained compare the two algorithms here}
-Since we mix both liveness based GC and reachability the garbage collector has to be more careful 
-and ensure references marked non-live are never copied using reachability. We discuss this issue 
-and its solution in sub-section[]\cred{hanging like a sore thumb}.
+In this  algorithm we consult  the liveness automata  only if it  is a
+cons cell, for  closures we assume everything under it  to be live and
+copy everything that is reachable. Although this might lead to copying
+more  cells   it  is   faster  as  there   are  no   liveness  checks.
+\comment{Once  experimental  results  are  obtained  compare  the  two
+  algorithms  here}   Since  we  mix   both  liveness  based   GC  and
+reachability the garbage  collector has to be more  careful and ensure
+references  marked non-live  are never  copied using  reachability. We
+discuss this issue and its solution in sub-section[]\cred{hanging like
+  a sore thumb}.
 
@@
 \label{sec:conclusion}
-We have defined a notion of liveness on structured data;
-this generalizes classical liveness and strong liveness.
-We started with a general fully context-sensitive analysis which
-we proved correct with respect to a minefield semantics
-(this models the effect of garbage collection between every
-evaluation step).
-
-To avoid scalability issues (and to avoid performing part of
-the liveness computation at run time)
-we defined an 0-CFA version of this liveness analysis in which
-demands for function $f$ at all calling contexts are conflated into
-a single demand $\sigma_f$.
-This enabled us to treat the liveness equations symbolically obtaining
-context-free grammars for liveness at each GC point (calls to user
-functions and to $\CONS$).
-These were then converted to DFAs for run-time consultation by the
-garbage collector. Experiments confirm the precision of the analysis.
+We  have  defined  a  notion  of liveness  on  structured  data;  this
+generalizes classical liveness and strong liveness.  We started with a
+general fully context-sensitive analysis  which we proved correct with
+respect to  a minefield semantics  (this models the effect  of garbage
+collection between every evaluation step).
+
+To  avoid scalability  issues (and  to  avoid performing  part of  the
+liveness computation at run time)  we defined an 0-CFA version of this
+liveness analysis  in which  demands for function  $f$ at  all calling
+contexts are conflated into  a single demand $\sigma_f$.  This enabled
+us to treat the liveness equations symbolically obtaining context-free
+grammars for liveness at each GC point (calls to user functions and to
+$\CONS$).  These were then converted to DFAs for run-time consultation
+by  the garbage collector.  Experiments confirm  the precision  of the
+analysis.
 
@@
 significant benefit of LGC is  that programs can run in smaller memory
-when
-compared to RGC. This is potentially useful in situations where memory
-is limited---as with embedded systems. For a majority of programs, the
-garbage collection times were reduced.
+when compared to  RGC. This is potentially useful  in situations where
+memory  is  limited---as with  embedded  systems.  For  a majority  of
+programs, the garbage collection times were reduced.
 

### 2015-06-05 paper/paper.tex

@@
 \comment{Once  experimental  results  are  obtained  compare  the  two
-  algorithms  here}   Since  we  mix   both  liveness  based   GC  and
-reachability the garbage  collector has to be more  careful and ensure
-references  marked non-live  are never  copied using  reachability. We
-discuss this issue and its solution in sub-section[]\cred{hanging like
-  a sore thumb}.
+  algorithms here}  Mixing liveness based GC and  reachability has its
+own set of pros and cons which we discuss further.\cred{hanging like a
+  sore thumb}.
 
@@
 An advantage  of using  reachablity for closures  is that it  opens up
-certain opportunities of optimization. As we saw in section[], multiple 
-traversals over the same structure was a major inefficiency in the full 
-liveness based algorithm and overcoming it required a lot of memory.
-Using reachability for closures allows us to overcome this partially 
-using limited amount of memory. Any heap cell which is part of a closure 
-is copied using reachaibilty and hence by maintaining a single flag in each 
-cell to indicate whether it was copied using reachability prevents 
-traversing the same cell multiple times. Experimental results again 
-indicate that for lazy languages this gives a substantial performance 
-improvement.
-\comment{Maybe add a diagram}
+certain  opportunities  of  optimization.  As  we  saw  in  section[],
+multiple traversals  over the same structure was  a major inefficiency
+in the full liveness based  algorithm and overcoming it required a lot
+of memory.  Using reachability for closures allows us to overcome this
+partially using limited amount of  memory. Any heap cell which is part
+of a closure  is copied using reachaibilty and  hence by maintaining a
+single  flag in  each cell  to indicate  whether it  was  copied using
+reachability    prevents   traversing    the   same    cell   multiple
+times.  Experimental results  again indicate  that for  lazy languages
+this gives a  substantial performance improvement.  \comment{Maybe add
+  a diagram}
 

### 2015-06-05 paper/paper.tex
the statement of undecidability

@@
 &\sigma_1\cbar \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
-  \emptyset \\
-&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
+  \emptyset 
 \end{align*}
@@
 
-%% We could convert the rules defining $\hookrightarrow$ into productions
-%% and  add  them  to the  grammar.  However,  this  results in  an  {\em
-%%   unrestricted   grammar}~\cite{hopcraft90toc},  and   the  membership
-%% problem for  such grammars is undecidable. \comment{Rewrite  it to say
-%%   that this is a CFG with  3 unrestricted rules and the membership for
-%%   even this grammar  is undecidable} \comment{State the undecidability
-%%   of the decision problem above.}
-
-  We circumvent the problem by over-approximating the CFG generated by
-  the  analysis  to  {\em  strongly  regular}  CFGs  which  have  easy
-  translations to  non-deterministic finite state  automata (NFA). The
-  NFAs are then simplified on the basis of the $\hookrightarrow$ rules
-  to  enable  checking of  membership  of  forward  access paths.  The
-  resulting NFAs are finally converted  to DFAs for use during garbage
-  collection.  \comment{Fix all comments and produce first draft}
+We circumvent  the problem by over-approximating the  CFG  by
+non-deterministic finite state automata (NFA). The NFA are then
+simplified using the $\hookrightarrow$ rules. Finally the simplified
+NFAs are converted to DFAs.
+
 %------------------------------------------------------------%
@@
 \label{sec:NFA-approx}
-\comment{Change the language and the example}
-We use the
-algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
-approximate a CFG  to a   {\em strongly regular\/} grammar.  The
-transformation has  the  property that  if $\Lv$  is a
-non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
-Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq
-{\mathscr
-  L}_{G'}(\Lv)$.    This  is   required  for the
-approximation to be safe with respect to liveness.
-
-We show the \var{\Lanv{9}{\py}} grammar after the Mohri-Nederhof
-transformation. We  pick the only production that  is affected by
-the transformation---the production for \Df{\append}{1}.  The
-production for \Uf{\append}{1}, while  recursive, is already in
-strongly  regular  form  and   is  therefore  unaffected  by  the
-transformation.
+ We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
+ approximate   a  CFG   to   a  {\em   strongly  regular\/}   grammar.
+ \comment{Rewrite  this after  constructing the  example: We  show the
+   grammar for ????? after  the Mohri-Nederhof transformation. We pick
+   the only  production that  is affected by  the transformation---the
+   production    for     \Df{\append}{1}.     The    production    for
+   \Uf{\append}{1},  while recursive, is  already in  strongly regular
+   form and is therefore unaffected by the transformation.
  \begin{eqnarray*}
@@
 
-\subsubsection{Removing the $\cbar$ edges}
 Once the NFA's are created, we need to remove the transitions on
@@
 $\sigma$ then the NFA reaches a final state throught this path
-otherwise it does not reach a final state.\comment{Need to rewrite
-this}
+otherwise it does not reach a final state.}
+
 %------------------------------------------------------------%
@@
 The   strongly  regular   CFGs  obtained   after   the  Mohri-Nederhof
-transformation are
-  first converted  into  NFAs.  The        algorithm         described
-      in
-Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
+transformation are first converted into NFAs.  The algorithm described
+in  Figure~\ref{algo:simplify-NFA} converts  an NFA  \ntbar\ to  a NFA
 \nt\    such    that   $\lang{\ntbar}    \stackrel{*}{\hookrightarrow}
-\lang{\nt}$, where {\nt} accepts forward paths only.
-\comment{Rewrite the algorithm to handle \cbar and update the example
-automata}
-Thus  \nt\  can  be  used  to  check
-membership of forward paths.
+\lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
+  the algorithm to handle \cbar  and update the example automata} Thus
+\nt\ can be used to check membership of forward paths.
 
 \begin{figure*}[t]
-\begin{boxedminipage}{\textwidth}
+\begin{boxedminipage}{.5\textwidth}
 \begin{center}
@@
   \\
-%
 
   \raggedright{\bf Steps:}
- %% \begin{algorithmic}
-%%    \COMMENT Fix the algorithm issue
-%%    %% \STATE $i \leftarrow 0$
-%%    %% \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-%%    %% without $\epsilon$-moves \cite{hopcraft90toc}
-
-%%    %% \REPEAT
-%%    %% \STATE $\nt'_{i+1} \leftarrow \nt_i$
-%%    %% \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-%%    %%   edge from $q'$ with label $\bcar$ and
-%%    %%   outgoing edge to $q''$ with label $\acar$}
-%%    %% \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-%%    %% label $\epsilon$.
-%%    %% \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-%%    %% \ENDFOR
-
-%%    %% \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-%%    %%   edge from $q'$ with label $\bcdr$ and outgoing edge to
-%%    %%   $q''$ with label $\acdr$}
-%%    %% \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-%%    %%   label $\epsilon$.
-%%    %% \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-%%    %% \ENDFOR
-
-%%    %% \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-%%    %% $\nt'_{i+1}$ without $\epsilon$-moves
-
-%%    %% \STATE $i \leftarrow i+1$
-%%    %% \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-%%    %% \STATE $\nt \leftarrow \nt_i$
-%% %   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
-%%  \end{algorithmic}
+ \begin{algorithmic}
+   \STATE $i \leftarrow 0$
+   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
+   without $\epsilon$-moves \cite{hopcraft90toc}
+
+   \REPEAT
+   \STATE $\nt'_{i+1} \leftarrow \nt_i$
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcar$ and
+     outgoing edge to $q''$ with label $\acar$}
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+   label $\epsilon$.
+   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
+   \ENDFOR
+
+   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+     edge from $q'$ with label $\bcdr$ and outgoing edge to
+     $q''$ with label $\acdr$}
+   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+     label $\epsilon$.
+   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
+   \ENDFOR
+
+   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
+   $\nt'_{i+1}$ without $\epsilon$-moves
+
+   \STATE $i \leftarrow i+1$
+   \UNTIL ($\nt_{i} = \nt_{i-1}$)
+
+   \STATE $\nt \leftarrow \nt_i$
+  \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
+ \end{algorithmic}
 \end{center}
+\end{boxedminipage}
+\begin{boxedminipage}{.5\textwidth}
+The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
+of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
+is continued till  a fixed point is reached.  When  the fixed point is
+reached,   the  resulting   NFA  contains   all   possible  reductions
+corresponding to all the paths in the original NFA.  The proofs of the
+termination and correctness  of th The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
+of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
+is continued till  a fixed point is reached.  When  the fixed point is
+reached,   the  resulting   NFA  contains   all   possible  reductions
+corresponding to all the paths in the original NFA.  The proofs of the
+termination and correctness  of th The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
+of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
+is continued till  a fixed point is reached.  When  the fixed point is
+reached,   the  resulting   NFA  contains   all   possible  reductions
+corresponding to all the paths in the original NFA.  The proofs of the
+termination and correctness  of th The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
+of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
+is continued till  a fixed point is reached.  When  the fixed point is
+reached,   the  resulting   NFA  contains   all   possible  reductions
+corresponding to all the paths in the original NFA.  The proofs of the
+termination and correctness  of th
+
 \end{boxedminipage}

### 2015-06-06 paper/paper.tex
figures in one col.

@@
 
-\begin{figure*}[t!]
+\begin{figure}[t!]
   %%%%{\color{Myblue}
-\begin{picture}(100,130)(0,-90)
+  \psset{unit=1mm}
+  \begin{pspicture}(80,50)(0,-50)
+    %\psframe(80,50)(0,-50)
     %  \begin{center}
-    \begin{tabular}{cc}
+    \begin{tabular}{@{}c@{}}
       \begin{boxedminipage}{.42\textwidth}
@@
       \end{boxedminipage}
-&
+      \\
+      (a) Example program. \\ \\
       \begin{minipage}{.45\textwidth}
-
-\hspace*{1cm}{\epsfig{file=motiv-example.eps, height=4cm}}
-
+        %\hspace*{1cm}
+                {\epsfig{file=motiv-example.eps, height=4cm}}
       \end{minipage}
-
-\\
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%       &
-%%       \raisebox{-25mm}{\scalebox{.75}{
-%% 	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
-%%       \psset{unit=1mm}
-%%       \psset{linewidth=.3mm}
-%%       \begin{pspicture}(0,-5)(70,60)
-%% 	%\psframe(0,0)(73,60)
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% 	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
-%% 	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
-%% %
-%%      \putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
-%%
-%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
-%%      fillstyle=solid,
-%% 	%      fillcolor=white]{$\times$}}}
-%%
-%%  \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox
-%%{.63}{\nilfigure}}}
-%% %	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
-%% 	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
-%%
-%% \putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
-%% 	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
-%%
-%% \aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
-%%fillstyle=solid,
-%% 		  fillcolor=white]{$\times$}}}
-%% 	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% 	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
-%% 	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
-%% 	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
-%% 	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
-%% 	\ncline[nodesepB=-.5]{*->}{c2}{d}
-%% 	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
-%% 	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
-%%
-%% \aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
-%% fillstyle=solid,
-%% 	      fillcolor=white]{$\times$}}}
-%% 	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
-%%
-%% \nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
-%% 	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
-%%
-%% \aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
-%% fillstyle=solid,
-%% 	      fillcolor=white]{$\times$}}}
-%%
-%% \putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
-%%
-%%\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
-%%
-%%\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% 	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
-%% 	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
-%% 	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
-%% %
-%%\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-%% 	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
-%%
-%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
-%%         %here
-%%
-%%\putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
-%%
-%%\putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{
-%%\scalebox{.63}{\nilfigure}}}
-
-%% %
-%%\aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,
-%%fillstyle=solid,
-%% %	      fillcolor=white]{$\times$}}} %and here
-%% 	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
-%%
-%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h}
-%% %
-%%\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,
-%%fillstyle=solid,
-%% %	      fillcolor=white]{$\times$}}} %and here
-
-%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
-%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
-%% 	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
-%%
-%%\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
-%% 	      fillcolor=white]{$\times$}}}
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% 	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
-%% %
-%%\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
-%% 	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
-%%
-%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
-%% %	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-        %%
-%%       \end{pspicture}}} \\
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-      (a) Example program.&\hspace*{2cm}
-      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
-	(b) Memory graph at $\pi$. Thick edges denote  \\
-        {\white (b) }live links. Traversal stops at edges marked\\
-        {\white (b) }$\times$  during garbage collection.
+      \\ \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+          (b) Memory graph at $\pi$. Thick edges denote live
+          links. Traversal \\{\white (b)} stops at edges
+          marked $\times$ during garbage collection.
       \end{tabular}}
     \end{tabular}
-\end{picture}
-
-\kern -3ex
-
-\caption{Example Program and its Memory Graph.{\color{red}Amey will combine this with the figure for syntax}}\label{fig:mot-example}
-\end{figure*}
+  \end{pspicture}
+ \vspace*{-3ex}
+  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
+\end{figure}
 
@@
 
-\begin{figure*}[]
+\begin{figure}[t]
 \footnotesize
@@
    p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
-    \hspace{7em} \mbox{\em --- program}\\
+    \hspace{5em} \mbox{\em --- program}\\
     \mathit{df} \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
     e) 
-    \hspace{2.2em} \ \ \ \ \mbox{\em --- function definition} \\
+    \hspace{0.2em} \ \ \ \ \mbox{\em --- function definition} \\
 e \in \mathit{Expr} & ::= &
-\left\{\begin{array}{ll@{\hspace{4em}}l}
+\left\{\begin{array}{@{}ll@{\hspace{2em}}l}
        (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
@@
 s \in \mathit{Application} & ::= &
-\left\{\begin{array}{lr@{\hspace{1em}}l}
-       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
-       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\
-       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\
-       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester
+\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
+       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
+       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
+       (\CAR\,\, x) & \\ 
+       (\CDR\,\, x) & \mbox{\em --- selectors} \\
+       (\NULLQ\,\, x) & \\ 
+       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester
 and generic arithmetic} \\
-%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\
-       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)}
+       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
 \\
-%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic}
-\\
-       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)}
-            & \mbox{\em --- function application}
+       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
     \end{array}\right.
@@
 \normalsize
-\end{figure*}
+\end{figure}
 

### 2015-06-06 paper/paper.tex
Updated the example program

--- 2015-06-05 paper/paper.tex
@@
 	  \UNL{1} $(\DEFINE\ (\main)$ 
-          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots
+           \UNL{2} \!\!$\pi_9\!\!:\, \cred{(\LET\  \pa\  \leftarrow (\abigfunction\ \acdr)}$ \IN
+          \UNL{3} \!\!$\pi_{10}\!\!:\, \cred{(\LET\  \pb\  \leftarrow (}$\cred{+}$\cred{\ \pa\ \acdr)}$
 \IN
-	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\
-$\leftarrow$  \ldots \IN
-          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
-          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
-          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\,
+	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ \cred{(\LET\ \pc}\
+$\cred{\leftarrow  (\CONS\ \pb\ ())}$ \IN
+          \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
+          $\cred{(\LET\ \pw}\  $\cred{\leftarrow  (\length\ \pc)}$ \IN
+          \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
 (\SRETURN~\psi_4:\pw)))))$
@@
                 [\;
-                  ep_4 \mapsto \lbrace \pw.\sigma,
-           \pb.\Lf{\length}{\mbox{1}}{\sigma}, \ldots\rbrace
+                  ep_4 \mapsto \lbrace \cred {\pw.\sigma,
+                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},}\\
+&& \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
+                                        \cred{\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},}
+                                        \cred{\pa.\cbar\bcar\Lf{\length}{\mbox{1}}{\sigma}} \rbrace
                                ] \\
                \cal{P_{\mbox{\tiny \main}}} &=&
-                   \lbrack~\pi_9 \mapsto \lbrace ep_4 \rbrace,  \
+                  \lbrack~ \cred{\pi_9 \mapsto \lbrace ep_4 \rbrace,  \
                     \pi_{10} \mapsto \lbrace ep_4 \rbrace,  \
-                   \pi_{11} \mapsto \lbrace ep_4 \rbrace,  \pi_{12} \mapsto \lbrace ep_4 \rbrace~\rbrack
+                   \pi_{11} \mapsto \lbrace ep_4 \rbrace,} \\
+                   && \;\; \cred{\pi_{12} \mapsto \lbrace ep_4 \rbrace, \pi_{13} \mapsto \lbrace ep_4 \rbrace}~\rbrack
 \end{eqnarray*}
@@
 For  the  running  example,   $\length$  has  calls  from  $\main$  at
-$\pi_{10}$ and a recursive call at $\pi_6$.
+$\pi_{12}$ and a recursive call at $\pi_6$.
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{11}}{\main}  \cup 
+     \deltacall{\length}{\pi_{12}}{\main}  \cup 
 \deltacall{\length}{\pi_6}{\length}$.
@@
 \end{eqnarray*}
-As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pb} in terms
+As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pc} in terms
 of  $\sigma_{\length}$ are:
@@
  \\
-\Lanv{9}{\pb} &= \Df{\length}{1}\sigma_{\mathit{\!all}}
+\Lanv{9}{\pc} &= \Df{\length}{1}\sigma_{\mathit{\!all}}
 \end{align*}
@@
 &\rightarrow&
-\sigma_{\!all}  \mid \cbar\langle{\sigma_{\length}}\rangle
+\cred{\epsilon}  \mid \cbar\langle{\sigma_{\length}}\rangle
 \end{eqnarray*}
@@
 \begin{figure*}[t]
-\begin{boxedminipage}{.5\textwidth}
-\begin{center}
-  \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
-  alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
-%
-  \raggedright{\bf Output:} NFA  \nt\ with  underlying
-  alphabet $\lbrace \acar, \acdr \rbrace$ such that
-  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
-  \\
-
-  \raggedright{\bf Steps:}
- \begin{algorithmic}
-   \STATE $i \leftarrow 0$
-   \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-   without $\epsilon$-moves \cite{hopcraft90toc}
-
-   \REPEAT
-   \STATE $\nt'_{i+1} \leftarrow \nt_i$
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcar$ and
-     outgoing edge to $q''$ with label $\acar$}
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-   label $\epsilon$.
-   \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-   \ENDFOR
-
-   \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-     edge from $q'$ with label $\bcdr$ and outgoing edge to
-     $q''$ with label $\acdr$}
-   \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-     label $\epsilon$.
-   \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-   \ENDFOR
-
-   \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-   $\nt'_{i+1}$ without $\epsilon$-moves
-
-   \STATE $i \leftarrow i+1$
-   \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-   \STATE $\nt \leftarrow \nt_i$
-  \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
- \end{algorithmic}
-\end{center}
-\end{boxedminipage}
+%% \begin{boxedminipage}{.5\textwidth}
+%% \begin{center}
+%%   \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
+%%   alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
+%% %
+%%   \raggedright{\bf Output:} NFA  \nt\ with  underlying
+%%   alphabet $\lbrace \acar, \acdr \rbrace$ such that
+%%   $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
+%%   \\
+
+%%   \raggedright{\bf Steps:}
+%%  \begin{algorithmic}
+%%    \STATE $i \leftarrow 0$
+%%    \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
+%%    without $\epsilon$-moves \cite{hopcraft90toc}
+
+%%    \REPEAT
+%%    \STATE $\nt'_{i+1} \leftarrow \nt_i$
+%%    \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+%%      edge from $q'$ with label $\bcar$ and
+%%      outgoing edge to $q''$ with label $\acar$}
+%%    \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+%%    label $\epsilon$.
+%%    \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
+%%    \ENDFOR
+
+%%    \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
+%%      edge from $q'$ with label $\bcdr$ and outgoing edge to
+%%      $q''$ with label $\acdr$}
+%%    \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
+%%      label $\epsilon$.
+%%    \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
+%%    \ENDFOR
+
+%%    \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
+%%    $\nt'_{i+1}$ without $\epsilon$-moves
+
+%%    \STATE $i \leftarrow i+1$
+%%    \UNTIL ($\nt_{i} = \nt_{i-1}$)
+
+%%    \STATE $\nt \leftarrow \nt_i$
+%%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
+%%  \end{algorithmic}
+%% \end{center}
+%% \end{boxedminipage}
 \begin{boxedminipage}{.5\textwidth}

### 2015-06-06 paper/paper.tex
Updated example grammar
liveness

@@
 \end{align*}
-and \cbar\ as:
+and \clazy\ as:
 \begin{align*}
- \cbar\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
+ \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                        & \{\epsilon\} ~\mbox{otherwise}
 \end{align*}
-Note that \cbar\ is idempotent. 
+Note that \clazy\ is idempotent. 
 We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
@@
 &\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          =   x.\cbar\sigma \cup x.0\sigma
+          =   x.\clazy\sigma \cup x.0\sigma
 \end{align*}
@@
         \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
-        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
+        &[\,\mathit{ep} \mapsto  x.\clazy \sigma]
 \end{align*}
@@
 similar to \CAR.
-The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
+The  defintions of \bcar\, \bcdr\ and \clazy\  induce   the  following
 relation $\hookrightarrow$ over sets of access paths:
@@
   \acdr\sigma_2',~\mbox{and}\\
-&\sigma_1\cbar \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
+&\sigma_1\clazy \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
   \emptyset 
@@
         \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
-              [\;ep_1 \mapsto \lbrace \px.\cbar \sigma,
-\xl.\cbar\sigma\rbrace ,~ep_2 \mapsto \lbrace \cred{
+              [\;ep_1 \mapsto \lbrace \px.\clazy \sigma,
+\xl.\clazy\sigma\rbrace ,~ep_2 \mapsto \lbrace \cred{
 \pv.\sigma}  \rbrace,\\
  &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma,
-                \py.\cbar\sigma,
-\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma},\\
+                \py.\clazy\sigma,
+\pu.\Lf{\length}{\mbox{1}}{\clazy\sigma},\\
 &&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; 
-\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\cbar\sigma} \cup 
+\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\clazy\sigma} \cup 
 %%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
-\cred{\cbar\Lf{\length}{1}{\cbar\sigma}}\}\}
+\cred{\clazy\Lf{\length}{1}{\clazy\sigma}}\}\}
              \rbrack \\
@@
                                         \cred{\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},}
-                                        \cred{\pa.\cbar\bcar\Lf{\length}{\mbox{1}}{\sigma}} \rbrace
+                                        \cred{\pa.\clazy\bcar\Lf{\length}{\mbox{1}}{\sigma}} \rbrace
                                ] \\
@@
 \begin{eqnarray*}
-&& \Lanv{1}{\xl} = \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
-  \cup \cbar\Lf{\length}{1}{\cbar\sigma}
+&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
+  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
 \end{eqnarray*}
@@
    && \Lf{\length}{1}{\sigma}
-    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma} \cup \cbar\Lf{\length}{1}{\cbar\sigma}
+    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup \clazy\Lf{\length}{1}{\clazy\sigma}
 \end{eqnarray*}
@@
 each  of the liveness  rules modifies  a demand  by prefixing  it with
-symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
+symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \clazy
 \rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
@@
 &&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
-&&   \Df{\length}{1} = \cbar \cup \acdr\Df{\length}{1}\cbar 
-       \cup \cbar\Df{\length}{1}{\cbar}
+&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy 
+       \cup \clazy\Df{\length}{1}{\clazy}
   \end{eqnarray*}
@@
 Note that this equation can also be viewed as a CFG with \{\acdr,
-\cbar\} as terminal symbols and \Df{\length}{1} as the sole
+\clazy\} as terminal symbols and \Df{\length}{1} as the sole
 non-terminal.
@@
 with   demand   $\sigma_\mainpgm   =  \{\acar,\acdr\}^\ast$ (denoted
-$\sigma_{\!all}$).
+$\sigma_{\!all}$) if \mainpgm\ returns a structure and $\epsilon$ if
+it returns a base value.
 
@@
 \sigma_{\length}    &=&
- \sigma_{\!all}  ~\cup~{\cbar\sigma_{\mathit{\length}}}
+ \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
 \end{eqnarray*}
-As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pc} in terms
+As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
 of  $\sigma_{\length}$ are:
 \begin{align*}
-\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Df{\length}{1}\cbar\sigma_{\length}
-  \cup \cbar\Df{\length}{1}{\cbar\sigma_{\length}}
+\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
+  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
  \\
-\Lanv{9}{\pc} &= \Df{\length}{1}\sigma_{\mathit{\!all}}
+\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\epsilon
 \end{align*}
@@
 context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal
+\bcar, \bcdr, \clazy\rbrace$.  Let  \var{$X$} denote the non-terminal
 for
@@
  \begin{eqnarray*}
-\var{\Lanv{1}{\xl}}  &\rightarrow&   \cbar\sigma_{\length} \mid \acdr \var{\Df
-  {\length}{1}}{\cbar\sigma_{\length}} \mid
-  \cbar\var{\Df{\length}{1}}{\cbar\sigma_{\length}} \\
-  \var{\Df{\length}{1}} &\rightarrow& \cbar \mid
-  \acdr\var{\Df{\length}{1}}\cbar 
-       \mid \cbar\var{\Df{\length}{1}}{\cbar}\\
+\var{\Lanv{1}{\xl}}  &\rightarrow&   \clazy\sigma_{\length} \mid \acdr \var{\Df
+  {\length}{1}}{\clazy\sigma_{\length}} \mid
+  \clazy\var{\Df{\length}{1}}{\clazy\sigma_{\length}} \\
+  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
+  \acdr\var{\Df{\length}{1}}\clazy 
+       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
 \langle {\sigma_{\length}} \rangle
 &\rightarrow&
-\cred{\epsilon}  \mid \cbar\langle{\sigma_{\length}}\rangle
+\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
 \end{eqnarray*}
@@
 Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
-\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \cbar).     Let
+\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \clazy).     Let
 $\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
@@
 Consider a form of  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
-non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \cbar, \$\}$, $p_1$
+non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \clazy, \$\}$, $p_1$
 is a set of context-free production that contains the distinguished
@@
 unrestricted productions $\bcar\acar \rightarrow \epsilon$,
-$\bcdr\acdr \rightarrow \epsilon$, $\cbar\acar \rightarrow \cbar$,
-$\cbar\acdr \rightarrow \cbar$, and $\cbar\$ \rightarrow
+$\bcdr\acdr \rightarrow \epsilon$, $\clazy\acar \rightarrow \clazy$,
+$\clazy\acdr \rightarrow \clazy$, and $\clazy\$ \rightarrow
 \epsilon$. 
@@
 %------------------------------------------------------------%
-\subsubsection{Approximating CFGs using NFAs.}
+\subsubsection{From CFGs to GC-ready DFAs}
 \label{sec:NFA-approx}
  We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
- approximate   a  CFG   to   a  {\em   strongly  regular\/}   grammar.
- \comment{Rewrite  this after  constructing the  example: We  show the
-   grammar for ????? after  the Mohri-Nederhof transformation. We pick
-   the only  production that  is affected by  the transformation---the
-   production    for     \Df{\append}{1}.     The    production    for
-   \Uf{\append}{1},  while recursive, is  already in  strongly regular
-   form and is therefore unaffected by the transformation.
+ approximate a CFG to a {\em strongly regular\/} grammar.  The grammar
+ fragment  that  defines  the  non-terminal  $\var{\Df{\length}{1}}$
+   after the Mohri-Nederhof transformation becomes:
+
  \begin{eqnarray*} 
-  \var{\Df{\append}{1}}
-  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
-  \acdr\var{\Df{\append}{1}} \\
-\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
-\mid \epsilon
+   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
+   \acdr\var{\Df{\length}{1}}
+   \mid \clazy\var{\Df{\length}{1}}\\
+   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \epsilon \\
+   \langle {\sigma_{\length}} \rangle
+   &\rightarrow&
+   \epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
  \end{eqnarray*}
-The  languages
-generated for  \var{\Df{\append}{1}} in  the original grammar  and the
-new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
-\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.
-
-Once the NFA's are created, we need to remove the transitions on
-$\cbar$. This edge is replaced by an $\epsilon$ transition and the
-$\cbar$ edge is deleted. If the original grammar had a non-null
-$\sigma$ then the NFA reaches a final state throught this path
-otherwise it does not reach a final state.}
+
+The grammar after the Mohri-Nederhoff transformation is converted into
+a  set of  NFAs,  one for  each  $\var{\Lanv{i}{x}}$. We  now apply  the
+$\hookrightarrow$ simplification  by repeatedly introducing $\epsilon$
+edges to  bypass a pair  of consecutive edges labelled  \bcar\acar\ or
+\bcdr\acdr  and  deriving  the  $\epsilon$-closure  of  the  resulting
+automaton.   The   process  is  continued   till  a  fixed   point  is
+reached.  Finally we  replace  the $\clazy$  edges  by $\epsilon$  and
+determinize the automaton thus formed.
 
@@
 \lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
-  the algorithm to handle \cbar  and update the example automata} Thus
+  the algorithm to handle \clazy\  and update the example automata} Thus
 \nt\ can be used to check membership of forward paths.
@@
 corresponding to all the paths in the original NFA.  The proofs of the
-termination and correctness  of th The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
-of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
-is continued till  a fixed point is reached.  When  the fixed point is
+termination and correctness of  th The algorithm repeatedly introduces
+$\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
+\bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
+point is reached.  When the  fixed point is reached, the resulting NFA
+contains all possible reductions corresponding to all the paths in the
+original NFA.  The proofs of the termination and correctness of th The
+algorithm repeatedly  introduces $\epsilon$ edges to bypass  a pair of
+consecutive edges labelled \bcar\acar\  or \bcdr\acdr.  The process is
+continued  till a fixed  point is  reached.  When  the fixed  point is
 reached,   the  resulting   NFA  contains   all   possible  reductions
 corresponding to all the paths in the original NFA.  The proofs of the
-termination and correctness  of th The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
-of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
-is continued till  a fixed point is reached.  When  the fixed point is
-reached,   the  resulting   NFA  contains   all   possible  reductions
-corresponding to all the paths in the original NFA.  The proofs of the
-termination and correctness  of th The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
-of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
-is continued till  a fixed point is reached.  When  the fixed point is
-reached,   the  resulting   NFA  contains   all   possible  reductions
-corresponding to all the paths in the original NFA.  The proofs of the
-termination and correctness  of th
+termination and correctness of  th The algorithm repeatedly introduces
+$\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
+\bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
+point is reached.  When the  fixed point is reached, the resulting NFA
+contains all possible reductions corresponding to all the paths in the
+original NFA.  The proofs of the termination and correctness of th
 
@@
 \subsection{A simple LGC Algorithm}
-\cred{Introduce terminology}
+\cred{Introduce representation}
 We introduce certain terminology that we use in our discussion of the 

### 2015-06-06 lazy-scheme/post_process/gen_stats_and_plotall.sh
my changes for running test big and small

@@ -1,8 +1,8 @@
 #!/bin/bash 
 DIR=`dirname $0`
 PROGNAME=`basename $0`
-#AllBMs="   lambda sudoku lcss small gc_bench nperm fibheap treejoin knightstour nqueens "
-AllBMs=" lambda sudoku lcss small treejoin fibheap"
+#AllBMs="   lambda sudoku lcss small test gc_bench nperm fibheap treejoin knightstour nqueens "
+AllBMs=" lambda sudoku lcss small test treejoin fibheap"
 help()
 {
     echo ""
@@ -46,6 +46,7 @@ limit["sudoku"]=1728
 limit["fft"]=3300
 limit["lcss"]=22243
 limit["small"]=17
+limit["test"]=247
 limit["gc_bench"]=131091
 limit["nperm"]=27428
 limit["fibheap"]=38200
@@ -61,6 +62,7 @@ freq["sudoku"]=2050;
 freq["fft"]=2050;
 freq["lcss"]=1000;
 freq["small"]=1;
+freq["test"]=1;
 freq["gc_bench"]=4000;
 freq["nperm"]=145;
 freq["fibheap"]=700;
@@ -75,6 +77,7 @@ xtics["sudoku"]=2050; ytics["sudoku"]=250
 xtics["fft"]=2050; ytics["fft"]=250
 xtics["lcss"]=100000; ytics["lcss"]=9000
 xtics["small"]=1; ytics["small"]=1
+xtics["test"]=25; ytics["test"]=25
 xtics["gc_bench"]=400000; ytics["gc_bench"]=20000
 xtics["nperm"]=145000; ytics["nperm"]=35000
 xtics["fibheap"]=70000; ytics["fibheap"]=30000
@@ -89,8 +92,8 @@ fi
 
 for bm in $BMs
 do
-    for opt in gc-plain gc-live
-    #for opt in gc-plain gc-live gc-freq=${freq[$bm]}
+    #for opt in gc-plain gc-live
+    for opt in gc-plain gc-live gc-freq=${freq[$bm]}
     do 
      	echo "Processing $bm [$opt]"
      	$DIR/gen_gc_stats.sh $OPTARG \

### 2015-06-08 paper/paper.tex
Some problem

--- 2015-06-06 paper/paper.tex
@@
 \usepackage{algorithmic}
-
 \input{fh_defs}
@@
 \begin{document}
+
 % --- Author Metadata here ---
@@
 \title{Liveness-Based Garbage Collection for Lazy Languages}
-
 % You need the command \numberofauthors to handle the 'placement
@@
 % the body of your article BEFORE References or any Appendices.
-\cmt{{
-\numberofauthors{3} %  in this sample file, there are a *total*
+%\cmt{{
+%\numberofauthors{3} %  in this sample file, there are a *total*
 % of EIGHT authors. SIX appear on the 'first-page' (for formatting
@@
 rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
-$\emptyset$, if  $\sigma$ itself $\emptyset$.  To  realize these rules
+$\emptyset$, if  $\sigma$ itself is $\emptyset$.  To  realize these rules
 $\sigma$ needs to be known. This creates difficulties since we want to
@@
  We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
- approximate a CFG to a {\em strongly regular\/} grammar.  The grammar
- fragment  that  defines  the  non-terminal  $\var{\Df{\length}{1}}$
-   after the Mohri-Nederhof transformation becomes:
-
+ approximate a CFG to a {\em strongly regular\/} grammar.  For
+ example, the grammar
+ fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
+ Mohri-Nederhof transformation is:
  \begin{eqnarray*} 
@@
    \mid \clazy\var{\Df{\length}{1}}\\
-   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \epsilon \\
-   \langle {\sigma_{\length}} \rangle
-   &\rightarrow&
-   \epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
+   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
+   \mid \epsilon      
  \end{eqnarray*}
 
-The grammar after the Mohri-Nederhoff transformation is converted into
-a  set of  NFAs,  one for  each  $\var{\Lanv{i}{x}}$. We  now apply  the
-$\hookrightarrow$ simplification  by repeatedly introducing $\epsilon$
-edges to  bypass a pair  of consecutive edges labelled  \bcar\acar\ or
-\bcdr\acdr  and  deriving  the  $\epsilon$-closure  of  the  resulting
-automaton.   The   process  is  continued   till  a  fixed   point  is
-reached.  Finally we  replace  the $\clazy$  edges  by $\epsilon$  and
-determinize the automaton thus formed.
+The strongly regular grammar is converted into a set of
+NFAs,   one  for   each  $\var{\Lanv{i}{x}}$.   The  $\hookrightarrow$
+simplification  is now  done  on the  NFAs  by repeatedly  introducing 
+$\epsilon$  edges  to  bypass  pairs  of  consecutive  edges  labelled
+\bcar\acar\ or \bcdr\acdr\  and constructing the
+$\epsilon$-closure. Since the $\epsilon$-closure step may create more
+adjacent \bcar\acar\ and \bcdr\acdr\  edges, the simplification is  
+ continued  till a  fixed point is  reached. The algorithm  for this
+appears  in~\cite{asati14lgc},   Figure[].  Finally  we   replace  the
+$\clazy$  edges  by  $\epsilon$  and determinize  the  automaton  thus
+formed.
+
+\comment{A paragraph more to explain the illustration on the following lines:
 
 %------------------------------------------------------------%
-\subsubsection{Transforming NFAs to Accept Forward Paths:}
-%\label{sec:NFA-elim}
-The   strongly  regular   CFGs  obtained   after   the  Mohri-Nederhof
-transformation are first converted into NFAs.  The algorithm described
-in  Figure~\ref{algo:simplify-NFA} converts  an NFA  \ntbar\ to  a NFA
-\nt\    such    that   $\lang{\ntbar}    \stackrel{*}{\hookrightarrow}
-\lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
-  the algorithm to handle \clazy\  and update the example automata} Thus
-\nt\ can be used to check membership of forward paths.
-
-\begin{figure*}[t]
+%% \subsubsection{Transforming NFAs to Accept Forward Paths:}
+%% %\label{sec:NFA-elim}
+%% The   strongly  regular   CFGs  obtained   after   the  Mohri-Nederhof
+%% transformation are first converted into NFAs.  The algorithm described
+%% in  Figure~\ref{algo:simplify-NFA} converts  an NFA  \ntbar\ to  a NFA
+%% \nt\    such    that   $\lang{\ntbar}    \stackrel{*}{\hookrightarrow}
+%% \lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
+%%   the algorithm to handle \clazy\  and update the example automata} Thus
+%% \nt\ can be used to check membership of forward paths.
+
+%%\begin{figure*}[t]
 %% \begin{boxedminipage}{.5\textwidth}
@@
 %% \end{boxedminipage}
-\begin{boxedminipage}{.5\textwidth}
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
-of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
-is continued till  a fixed point is reached.  When  the fixed point is
-reached,   the  resulting   NFA  contains   all   possible  reductions
-corresponding to all the paths in the original NFA.  The proofs of the
-termination and correctness of  th The algorithm repeatedly introduces
-$\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
-\bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
-point is reached.  When the  fixed point is reached, the resulting NFA
-contains all possible reductions corresponding to all the paths in the
-original NFA.  The proofs of the termination and correctness of th The
-algorithm repeatedly  introduces $\epsilon$ edges to bypass  a pair of
-consecutive edges labelled \bcar\acar\  or \bcdr\acdr.  The process is
-continued  till a fixed  point is  reached.  When  the fixed  point is
-reached,   the  resulting   NFA  contains   all   possible  reductions
-corresponding to all the paths in the original NFA.  The proofs of the
-termination and correctness of  th The algorithm repeatedly introduces
-$\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
-\bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
-point is reached.  When the  fixed point is reached, the resulting NFA
-contains all possible reductions corresponding to all the paths in the
-original NFA.  The proofs of the termination and correctness of th
-
-\end{boxedminipage}
- \caption{Algorithm for transforming an NFA to  accept forward
-   paths only.}\label{algo:simplify-NFA} \figrule
-\end{figure*}
-
-The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
-of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
-is continued till  a fixed point is reached.  When  the fixed point is
-reached,   the  resulting   NFA  contains   all   possible  reductions
-corresponding to all the paths in the original NFA.  The proofs of the
-termination and correctness  of the algorithm are given  in an earlier
-paper~\cite{karkare07liveness}.
+%% \begin{boxedminipage}{.5\textwidth}
+%% The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
+%% of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
+%% is continued till  a fixed point is reached.  When  the fixed point is
+%% reached,   the  resulting   NFA  contains   all   possible  reductions
+%% corresponding to all the paths in the original NFA.  The proofs of the
+%% termination and correctness of  th The algorithm repeatedly introduces
+%% $\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
+%% \bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
+%% point is reached.  When the  fixed point is reached, the resulting NFA
+%% contains all possible reductions corresponding to all the paths in the
+%% original NFA.  The proofs of the termination and correctness of th The
+%% algorithm repeatedly  introduces $\epsilon$ edges to bypass  a pair of
+%% consecutive edges labelled \bcar\acar\  or \bcdr\acdr.  The process is
+%% continued  till a fixed  point is  reached.  When  the fixed  point is
+%% reached,   the  resulting   NFA  contains   all   possible  reductions
+%% corresponding to all the paths in the original NFA.  The proofs of the
+%% termination and correctness of  th The algorithm repeatedly introduces
+%% $\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
+%% \bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
+%% point is reached.  When the  fixed point is reached, the resulting NFA
+%% contains all possible reductions corresponding to all the paths in the
+%% original NFA.  The proofs of the termination and correctness of th
+
+%% \end{boxedminipage}
+%%  \caption{Algorithm for transforming an NFA to  accept forward
+%%    paths only.}\label{algo:simplify-NFA} \figrule
+%% \end{figure*}
+
+%% The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
+%% of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
+%% is continued till  a fixed point is reached.  When  the fixed point is
+%% reached,   the  resulting   NFA  contains   all   possible  reductions
+%% corresponding to all the paths in the original NFA.  The proofs of the
+%% termination and correctness  of the algorithm are given  in an earlier
+%% paper~\cite{karkare07liveness}.
 
@@
 is live at $\pi_9$. Also note  that the language accepted by the final
-automaton satisfies the prefix-closed property.
-
-\hrule
-\hrule
-
-\section{Implementation}
+automaton satisfies the prefix-closed property.}
+\begin{figure}[t!]
+\hspace*{1cm}{\epsfig{file=example-automata.eps, height=2.75cm}}
+$\caption{(a) The grammar rules for $\Df{\length}{1'}$ converted into
+$  an automaton, and simplified in   (b). (c) and (d) show the automata
+$  for   \Lanv{9}{\pa}.}
+\figrule
+\end{figure}
+
+\section{The Garbage Collection Scheme}
 % GC assumptions

### 2015-06-10 paper/paper.tex
Fixing the section on implementation schemes

@@
 \usepackage{bcprules}
+\usepackage{subfigure}
 %\usepackage{program}
-\usepackage{algorithmic}
+%\usepackage{algorithmic}
+\usepackage[boxed]{algorithm2e}
+
 \input{fh_defs}
@@
 \newcommand{\cred}[1]{{\color{red}{#1}}}
-\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
-    #1 \medskip \hrule \medskip}}}
 \newcommand{\blankout}[1]{}
@@
 \newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
-\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
-\newcommand{\stk}{\mbox{S}}       % stack
-\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
+\def\myvec{\mathaccent"017E } 
+\newcommand{\stk}{\mbox{S}}   
+\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
 \newcommand{\bang}{\mbox{\sc bang}}
@@
 \begin{document}
-
 % --- Author Metadata here ---
 \conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
+%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
+%over-ridden - IF NEED BE.
+%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
+%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
 % --- End of Author Metadata ---
@@
 \title{Liveness-Based Garbage Collection for Lazy Languages}
+
 % You need the command \numberofauthors to handle the 'placement
@@
 % the body of your article BEFORE References or any Appendices.
-%\cmt{{
-%\numberofauthors{3} %  in this sample file, there are a *total*
+\cmt{{
+\numberofauthors{3} %  in this sample file, there are a *total*
 % of EIGHT authors. SIX appear on the 'first-page' (for formatting
@@
 While  the runtime  system  of most  functional  languages includes  a
-garbage collector to efficiently  reclaim memory, empirical studies on
-Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
-Haskell~\cite{rojemo96lag}
+garbage  collector  (GC)  to  efficiently  reclaim  memory,  empirical
+studies on Scheme~\cite{karkare06effectiveness} and, more importantly,
+on Haskell~\cite{rojemo96lag}
 %and Java~\cite{shaham02estimating}
-programs have shown that  garbage collectors leave uncollected a large
+programs have shown that  GCs leave uncollected a large
 number of  memory objects that are  reachable but not  live (here {\em
@@
 %%% MOTIVATING EXAMPLE
-\newcommand{\nilfigure}
-{\scalebox{0.75}{
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,0)(1,1)
-%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
-subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
-\putnode{start}{origin}{0}{0}{}
-\putnode{stop}{origin}{10}{10}{}
-\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
-\end{pspicture}
-}}
+%% \newcommand{\nilfigure}
+%% {\scalebox{0.75}{
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,0)(1,1)
+%% %\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
+%% subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
+%% \putnode{start}{origin}{0}{0}{}
+%% \putnode{stop}{origin}{10}{10}{}
+%% \ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
+%% \end{pspicture}
+%% }}
 
@@
         %\hspace*{1cm}
-                {\epsfig{file=motiv-example.eps, height=4cm}}
+                {\includegraphics[width=1\textwidth]{motiv-example}}
       \end{minipage}
@@
 on the program point where  garbage collection has been triggered, the
-garbage-collector consults  a set of automata  to curtail reachability
+GC consults  a set of automata  to curtail reachability
 during marking.  This results in  an increase in the garbage reclaimed
@@
 cells can be  reclaimed.  Thus if a garbage  collection takes place at
-$\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b), only the
-cells referenced by $\pz$,  $(\CDR~ \pz)$ and (the cells constituting)
-the closure  referenced by $(\CDR~(\CDR~\pz))$  need to be  marked and
-preserved.  In  contrast, a  reachability-based  collector would  have
-preserved all cells.
+$\pi$  with  the  heap  shown  in  Figure~\ref{fig:mot-example}(b),  a
+liveness-based collector (LGC) will  collect only the cells referenced
+by  $\pz$, $(\CDR~  \pz)$  and (the  cells  constituting) the  closure
+referenced by $(\CDR~(\CDR~\pz))$ need to be marked and preserved.  In
+contrast,  a reachability-based collector  (RGC) would  have preserved
+all cells.
 
@@
     \hspace{5em} \mbox{\em --- program}\\
-    \mathit{df} \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
+    \mathit{df} \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
+\,\,x_n)\,\,
     e)
@@
 The  semantics  of   expressions  (and  applications\footnote{In  most
-  contexts, we  shall use  the term 'expression' and the notation $e$  for both  expressions and
+  contexts, we  shall use  the term 'expression' and the notation $e$  for
+both  expressions and
   applications.}) are  given by transitions  of the form  $\rho, \stk,
@@
 updated with the  sum of the arguments and  the next expression to be
-evaluated is picked from the continuation on the stack. If either of the arguments is not in
+evaluated is picked from the continuation on the stack. If either of the
+arguments is not in
 WHNF  ({\sc  prim-1-clo}  and   {\sc  prim-2-clo}),  it  is  sent  for
@@
 \rightsquigarrow
-{\color {red}\rho\oplus\rho'}, (\rho, x, (\CAR~x))\!:\!S, H, s$      &  \sc{car-clo}
+{\color {red}\rho\oplus\rho'}, (\rho, x, (\CAR~x))\!:\!S, H, s$      &
+\sc{car-clo}
 \\
@@
   (\SIF~x~e_1~e_2)   \rightsquigarrow
-{\color {red} \rho \oplus \rho'}, (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$} &
+{\color {red} \rho \oplus \rho'}, (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
+&
 \sc{if-clo} \\
@@
 rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
-$\emptyset$, if  $\sigma$ itself is $\emptyset$.  To  realize these rules
+$\emptyset$, if  $\sigma$ itself is $\emptyset$ and
+$\lbrace\epsilon\rbrace$ otherwise.  To  realize these rules
 $\sigma$ needs to be known. This creates difficulties since we want to
@@
 The  solution  is  to   also  treat  the  operations  mentioned  above
-symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
-\bcdr\ are defined as:
+symbolically.   We  introduce three  new  symbols.  These symbols  are
+defined as a relation  $\hookrightarrow$ between sets of access paths.
+First, \bcar\ and \bcdr\ are defined as:
+
 \begin{align*}
- \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
- \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
+  &\sigma_1\bcar\sigma_2  \hookrightarrow
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
+&\sigma_1\bcdr\sigma_2  \hookrightarrow
+  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
+  \acdr\sigma_2'
 \end{align*}
-and \clazy\ as:
+Thus \bcar, for example, removes the leading \acar\ from $\sigma_2$.
+Similarly, the symbol \clazy\ reduces the string following it to an
+$\epsilon$. \clazy\  is defined by addition to
+$\hookrightarrow$ the rule:
 \begin{align*}
- \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
-                       & \{\epsilon\} ~\mbox{otherwise}
+  &\sigma_1\clazy\sigma_2  \hookrightarrow
+  \sigma_1 \mbox{ if } \sigma_2 \subseteq  \{\acar,\acdr\}^\ast
 \end{align*}
-Note that \clazy\ is idempotent. 
-We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
+%% \begin{align*}
+%%  \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
+%%  \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
+%% \end{align*}
+%% and \clazy\ as:
+%% \begin{align*}
+%%  \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
+%%                        & \{\epsilon\} ~\mbox{otherwise}
+%% \end{align*}
+%% Note that \clazy\ is idempotent.
+We can  now rewrite the \CONS\  and the \CAR\  rules of $\mathit{ref}$
+as:
 \begin{align*}
@@
 The rules for  \CDR, \PRIM\ and \NULLQ\ are also  modified in a manner
-similar to \CAR.
-The  defintions of \bcar\, \bcdr\ and \clazy\  induce   the  following
-relation $\hookrightarrow$ over sets of access paths:
-\begin{align*}
-  &\sigma_1\bcar\sigma_2  \hookrightarrow
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
-&\sigma_1\bcdr\sigma_2  \hookrightarrow
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
-  \acdr\sigma_2',~\mbox{and}\\
-&\sigma_1\clazy \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
-  \emptyset 
-\end{align*}
+similar to \CAR.  %% The definitions  of \bcar\, \bcdr\ and \clazy\ induce
+%% the following relation $\hookrightarrow$ over sets of access paths:
+%% \begin{align*}
+%%   &\sigma_1\bcar\sigma_2  \hookrightarrow
+%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
+%% &\sigma_1\bcdr\sigma_2  \hookrightarrow
+%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
+%%   \acdr\sigma_2',~\mbox{and}\\
+%% &\sigma_1\clazy \sigma_2 \hookrightarrow \sigma_1, \mbox{ if }
+%% \sigma_2 \ne
+%%   \emptyset
+%% \end{align*}
 The reflexive  transitive closure of  $\hookrightarrow$ will
@@
 	  \UNL{1} $(\DEFINE\ (\main)$
-           \UNL{2} \!\!$\pi_9\!\!:\, \cred{(\LET\  \pa\  \leftarrow (\abigfunction\ \acdr)}$ \IN
-          \UNL{3} \!\!$\pi_{10}\!\!:\, \cred{(\LET\  \pb\  \leftarrow (}$\cred{+}$\cred{\ \pa\ \acdr)}$
+           \UNL{2} \!\!$\pi_9\!\!:\, \cred{(\LET\  \pa\  \leftarrow
+(\abigfunction\ \acdr)}$ \IN
+          \UNL{3} \!\!$\pi_{10}\!\!:\, \cred{(\LET\  \pb\  \leftarrow
+(}$\cred{+}$\cred{\ \pa\ \acdr)}$
 \IN
@@
                    \pi_{11} \mapsto \lbrace ep_4 \rbrace,} \\
-                   && \;\; \cred{\pi_{12} \mapsto \lbrace ep_4 \rbrace, \pi_{13} \mapsto \lbrace ep_4 \rbrace}~\rbrack
+                   && \;\; \cred{\pi_{12} \mapsto \lbrace ep_4 \rbrace,
+\pi_{13} \mapsto \lbrace ep_4 \rbrace}~\rbrack
 \end{eqnarray*}
@@
    && \Lf{\length}{1}{\sigma}
-    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup \clazy\Lf{\length}{1}{\clazy\sigma}
+    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
+\clazy\Lf{\length}{1}{\clazy\sigma}
 \end{eqnarray*}
@@
  \begin{eqnarray*}
-\var{\Lanv{1}{\xl}}  &\rightarrow&   \clazy\sigma_{\length} \mid \acdr \var{\Df
+\var{\Lanv{1}{\xl}}  &\rightarrow&   \clazy\sigma_{\length} \mid \acdr
+\var{\Df
   {\length}{1}}{\clazy\sigma_{\length}} \mid
@@
 Consider a grammar $G$ of the kind  described in
-Definition~\ref{def:specialgrammar} and a forward access path $\alpha$ consisting
+Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
+consisting
 of symbols \acar\ and \acdr\  only. The decision problem $\alpha \in
@@
 
-\comment{A paragraph more to explain the illustration on the following lines:
-
-%------------------------------------------------------------%
-%% \subsubsection{Transforming NFAs to Accept Forward Paths:}
-%% %\label{sec:NFA-elim}
-%% The   strongly  regular   CFGs  obtained   after   the  Mohri-Nederhof
-%% transformation are first converted into NFAs.  The algorithm described
-%% in  Figure~\ref{algo:simplify-NFA} converts  an NFA  \ntbar\ to  a NFA
-%% \nt\    such    that   $\lang{\ntbar}    \stackrel{*}{\hookrightarrow}
-%% \lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
-%%   the algorithm to handle \clazy\  and update the example automata} Thus
-%% \nt\ can be used to check membership of forward paths.
-
-%%\begin{figure*}[t]
-%% \begin{boxedminipage}{.5\textwidth}
-%% \begin{center}
-%%   \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
-%%   alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
-%% %
-%%   \raggedright{\bf Output:} NFA  \nt\ with  underlying
-%%   alphabet $\lbrace \acar, \acdr \rbrace$ such that
-%%   $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
-%%   \\
-
-%%   \raggedright{\bf Steps:}
-%%  \begin{algorithmic}
-%%    \STATE $i \leftarrow 0$
-%%    \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
-%%    without $\epsilon$-moves \cite{hopcraft90toc}
-
-%%    \REPEAT
-%%    \STATE $\nt'_{i+1} \leftarrow \nt_i$
-%%    \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-%%      edge from $q'$ with label $\bcar$ and
-%%      outgoing edge to $q''$ with label $\acar$}
-%%    \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-%%    label $\epsilon$.
-%%    \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
-%%    \ENDFOR
-
-%%    \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
-%%      edge from $q'$ with label $\bcdr$ and outgoing edge to
-%%      $q''$ with label $\acdr$}
-%%    \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
-%%      label $\epsilon$.
-%%    \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
-%%    \ENDFOR
-
-%%    \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
-%%    $\nt'_{i+1}$ without $\epsilon$-moves
-
-%%    \STATE $i \leftarrow i+1$
-%%    \UNTIL ($\nt_{i} = \nt_{i-1}$)
-
-%%    \STATE $\nt \leftarrow \nt_i$
-%%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
-%%  \end{algorithmic}
-%% \end{center}
-%% \end{boxedminipage}
-%% \begin{boxedminipage}{.5\textwidth}
-%% The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
-%% of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
-%% is continued till  a fixed point is reached.  When  the fixed point is
-%% reached,   the  resulting   NFA  contains   all   possible  reductions
-%% corresponding to all the paths in the original NFA.  The proofs of the
-%% termination and correctness of  th The algorithm repeatedly introduces
-%% $\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
-%% \bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
-%% point is reached.  When the  fixed point is reached, the resulting NFA
-%% contains all possible reductions corresponding to all the paths in the
-%% original NFA.  The proofs of the termination and correctness of th The
-%% algorithm repeatedly  introduces $\epsilon$ edges to bypass  a pair of
-%% consecutive edges labelled \bcar\acar\  or \bcdr\acdr.  The process is
-%% continued  till a fixed  point is  reached.  When  the fixed  point is
-%% reached,   the  resulting   NFA  contains   all   possible  reductions
-%% corresponding to all the paths in the original NFA.  The proofs of the
-%% termination and correctness of  th The algorithm repeatedly introduces
-%% $\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
-%% \bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
-%% point is reached.  When the  fixed point is reached, the resulting NFA
-%% contains all possible reductions corresponding to all the paths in the
-%% original NFA.  The proofs of the termination and correctness of th
-
-%% \end{boxedminipage}
-%%  \caption{Algorithm for transforming an NFA to  accept forward
-%%    paths only.}\label{algo:simplify-NFA} \figrule
-%% \end{figure*}
-
-%% The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
-%% of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
-%% is continued till  a fixed point is reached.  When  the fixed point is
-%% reached,   the  resulting   NFA  contains   all   possible  reductions
-%% corresponding to all the paths in the original NFA.  The proofs of the
-%% termination and correctness  of the algorithm are given  in an earlier
-%% paper~\cite{karkare07liveness}.
-
-%%\input{liveness-dfa.tex}
-\comment{change according  to example} We illustrate  the algorithm in
-Figure~\ref{algo:simplify-NFA}  by   constructing  the  automaton  for
-\var{\Lanv{9}{y}}.   Figure~\ref{fig:automata-construction}  (c) shows
-the  automaton  for  \var{\Lanv{9}{y}}  constructed by  composing  the
-automata   for    \var{\Uf{\append}{1}},   \var{\Df{\append}{1}}   and
-$(\epsilon \mid \acdr \mid \acdr\acar\sigma_{\mathit{\!all}})$.  After
-$\epsilon$ removal we get (d). We add an $\epsilon$ edge from $q_2$ to
-$q_3$ bypassing the  \bcdr\acdr\ pair from $q_2$ to  $q_2$ and then to
-$q_3$. This  is shown in (e).   The $\epsilon$ edge is  removed in (f)
-and  a second  $\epsilon$  is  added to  the  automaton bypassing  the
-\bcar\acar\ pair  from $q_1$ to $q_2$.  Removing  this $\epsilon$ edge
-gives  the automaton  shown  in (g).   Restricting  this automaton  to
-forward edges  only, we  get the final  automaton shown in  (h).  This
-automaton  recognizes  $\acdr^{*}\!\mid\acdr^{*}\acar  \sigma_{\mathit
-  \!all}$, showing  that the entire list \py,  including its elements,
-is live at $\pi_9$. Also note  that the language accepted by the final
-automaton satisfies the prefix-closed property.}
+Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
+from the  grammar for  \var{\Df{\length}{1'}}.  The  NFA is
+simplified by replacing the edges labeled \clazy\ by $\epsilon$. While
+checking for liveness during GC, a forward access-path traversing such
+an edge  will be accepted  only if it  can reach a final  state.  This
+effectively implements  the $\hookrightarrow$ simplification  rule for
+\clazy for forward paths. The resulting automaton, made deterministic,
+is shown in Figure~\ref{fig:example-automata}(b). This expectedly says
+that  given  a demand  $\epsilon$  on  \length,  the liveness  of  its
+argument             is             $\acdr^{*}$.            Similarly,
+Figure~\ref{fig:example-automata}(c)    shows   the    automaton   for
+\var{\Lanv{9}{\pa}}. The simplified automaton, shown in
+Figure~\ref{fig:example-automata}(d) does not accept any forward
+paths. This reflects the lazy nature of our language. \length\  does not
+evaluate its elements and thus \pa\ is dead at $\pi_9$.
+
 \begin{figure}[t!]
 \hspace*{1cm}{\epsfig{file=example-automata.eps, height=2.75cm}}
-$\caption{(a) The grammar rules for $\Df{\length}{1'}$ converted into
-$  an automaton, and simplified in   (b). (c) and (d) show the automata
-$  for   \Lanv{9}{\pa}.}
+\caption{(a) The grammar rules for $\langle \Df{\length}{1'}\rangle$
+converted into
+  an automaton, and simplified in   (b). (c) and (d) show the automata
+  for   \Lanv{9}{\pa}.}\label{fig:example-automata}
 \figrule
@@
 
-In this  section we describe in  detail the process  of liveness based
-garbage collection.  During  the execution of a program  if it runs of
-out  of memory,  execution is  stopped  and the  garbage collector  is
-invoked. At this point all the references on the stack are checked for
-liveness and only live references are copied and the program continues
-execution.  The main  complexity in our algorithm is  the way closures
-are  handled. The  following sub-sections  show two  different  ways of
-handling closures along with their pros and cons.
+Our experimental setup consists of  an interpreter for our language, a
+liveness  analyzer, and  a single  generation copying  collector.  The
+garbage collector  can be  easily configured to  work on the  basis of
+reachability (RGC  mode) or use  the liveness DFAs (LGC  mode).  While
+exploring the  activation record of a  function, LGC uses  the DFAs at
+the latest program  point traversed in the function's  body.  Thus, if
+LGC is invoked  at the program point $\pi$, it  would use the liveness
+DFAs  at  $\pi$ itself  for  exploring the  root  set  of the  current
+activation record. For  any other activation record in  the stack, say
+for a function ${\mathit f}$  calling ${\mathit g}$ in the call-chain,
+the evaluation  point in ${\mathit f}$  which resulted in  the call to
+${\mathit g}$ would be used.
+
+While exploring the heap, the  GC may encounter closures. In RGC mode,
+the  closure will  be explored  as  usual. However,  \cred{  as
+  mentioned earlier}, our liveness analysis predicts the liveness of
+data in its evaluated form. The absence of liveness information of
+closures presents challenges for LGC.
+
 \subsection{A simple LGC Algorithm}
-\cred{Introduce representation}
-We introduce certain terminology that we use in our discussion of the 
-garbage collection algorithms. A reference is a field which holds the 
-address of a heap location. Root set refers to the set of all references 
-that are on the program stack during garbage collection.
-A heap cell can either contain an atomic value, a \CONS\ cell or a closure. A closure 
-has fields to identify the closure type and 2 arguments. In case of function calls 
-with more than 2 arguments we use currying (one of the argument fields will point to another
- heap location containing the next argument and so on). By reachability we mean all heap 
-locations that are reachable traversing the argument fields of a given reference.
-% Abstract description of the GC algorithm
-\begin{verbatim}
-proceudre lgc()
- FOREACH reference in root set
- DO
-   IF reference is live
-     copy(cell)
-   ENDIF
- DONE
-
-procedure copy (cell)
- IF {cell is live}
-   new_cell <-  dup_heapcell(cell)
-   IF{celltype is cons}
-        new_car        <-  copy(cell.car)
-        new_cell.car   <-  new_car
-
-        new_cdr        <-  copy(cell.cdr)
-        new_cell.cdr   <-  new_cdr
-   ENDIF
- ELSE
-   IF {celltype is closure}
-       new_cell.arg1 <-  copy(cell.arg1)
-       new_cell.arg2 <-  copy(cell.arg2)
-   ENDIF
- ENDIF
- copy_references_on_print_stack()
-\end{verbatim}
+We shall call  a unit allocatable memory as a {\em  cell}.  A cell may
+hold  a  basic value,  or  the constructor  \CONS\,  or  (parts of)  a
+closure.  Apart from information to  identify the cell type, each cell
+has the space  for two fields. The second of the  fields may be empty,
+when, for example, the cell holds a basic value or a closure involving
+a unary operator. \cred{Should we say more about the representation}.
+
+\begin{figure*}[t!]
+\label{fig:algos}
+\hfill
+\subfigure[]{\includegraphics[width=.27\linewidth]{lgc-scheme-1}}
+\hfill
+\subfigure[]{\includegraphics[width=.45\linewidth]{lgc-scheme-2}}
+\hfill
+\caption{(a) LGC with closure collection based on reachability  (b)
+  LGC with liveness-based closure collection.}
+
+\end{figure*}
+
 In  the simple algorithm  presented we  consult liveness  automata for
@@
 book}.
-Printing of atomic values is simple and does not trigger a garbage collection.
-In case of \CONS\ cells, the \CAR\ part is first printed and then the \CDR\ part.
-In a lazy language, both the \CAR\ and the \CDR\ parts could be closures, thus 
+Printing of atomic values is simple and does not trigger a garbage
+collection.
+In case of \CONS\ cells, the \CAR\ part is first printed and then the \CDR\
+part.
+In a lazy language, both the \CAR\ and the \CDR\ parts could be closures,
+thus
 requiring evaluation. A garbage collection might be triggered while printing
@@
 and copy them. Liveness automata does not exist for the print function as it
-is not part of the executing program. Since the print function makes everything 
-reachable from a cell live we can copy all cells reachable from the reference.
-
-Consider the case where a \CONS\ cells is being printed and a garbage collection 
-is triggered when printing the \CDR\ part. In this case, since the refernce to the 
-\CONS\ cells is still on the print stack both the \CAR\ and the \CDR\ parts are 
+is not part of the executing program. Since the print function makes
+everything
+reachable from a cell live we can copy all cells reachable from the
+reference.
+
+Consider the case where a \CONS\ cells is being printed and a garbage
+collection
+is triggered when printing the \CDR\ part. In this case, since the refernce
+to the
+\CONS\ cells is still on the print stack both the \CAR\ and the \CDR\ parts
+are
 reachable but we know that the \CAR\ part has already been printed and can be
-safely garbage collected. A liveness analysis of the print function would have 
+safely garbage collected. A liveness analysis of the print function would
+have
 given the same information. In our implementation we mimic the behavior of
@@
 the \CAR\ part has been printed or not. For any \CONS\ cell if the \CAR\ part
-has not been printed, a reference to the \CAR\ part will also be on the stack but 
+has not been printed, a reference to the \CAR\ part will also be on the stack
+but
 the \CDR\ part has to be explicitly copied.
-In case the \CAR\ part has been printed, then a reference to the \CDR\ part will 
+In case the \CAR\ part has been printed, then a reference to the \CDR\ part
+will
 be on the stack and since the \CAR\ part is not required anymore we do not
@@
 \cred{Very sketchy: Make another attempt to write}
-
-%% \begin{figure}
-%% \begin{algorithm}[H]
-%%  \KwData{this text}
-%%  \KwResult{how to write algorithm with \LaTeX2e }
-%%  initialization\;
-%%  \While{not at end of this document}{
-%%   read current\;
-%%   \eIf{understand}{
-%%    go to next section\;
-%%    current section becomes this one\;
-%%    }{
-%%    go back to the beginning of current section\;
-%%   }
-%%  }
-%%  \caption{How to write algorithms}
-%% \end{algorithm}
-%% \label{algo:LGC-Algo} \figrule
-%% \end{figure}
-%% We   provide   an   intuition    about   the   correctness   of   this
-%% algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts
-%% about the print stack during a
-%% GC,
-
-%% \begin{enumerate}
-%% \item The top of the stack always contains a reference to a closure.
-%% \item All other references on the print stack point to atomic values or
-%% cons cells, it can never contain a
-%% reference to a closure. The components of the cons cell can contain
-%% reference to closures.
-%% \item If a reference on the print stack is a cons cell, either its car
-%% part or its cdr part will also be
-%% present on the stack.
-%% \end{enumerate}
-%% %Needs to be formatted and re-written
-%% Sketch of the proof
-%% Since all closures are copied fully and atomic values are copied, only
-%% cons cells need to be handled.
-%% Consider the following cases,
-%% \begin{enumerate}
-%% \item the car part of the cons  cell is on the print stack. This means
-%%   that the GC happened while the car part was being processed. The cdr
-%%   part is yet to be processed and the reference to the cdr part is not
-%%   on the print stack. Hence it is sufficient to copy only the cdr part
-%%   of the  cons cell. Any  references necessary for processing  the car
-%%   part will be copied due to references on the print stack.
-%% \item the cdr part of the cons  cell is on the print stack. This means
-%%   that GC  happened while the cdr  part was being  processed.  The car
-%%   part has already been processed and printed, hence we do not need to
-%%   copy the car  part. Additionally any references that  are needed for
-%%   printing the cdr  part will also be copied due  to the references on
-%%   the print stack and hence nothing needs to be copied in this case.
-%% \end{enumerate}
 \hrule

### 2015-06-10 paper/paper.tex
Fixing the section on implementation schemes

@@
 
-\subsection{A simple LGC Algorithm}
-We shall call  a unit allocatable memory as a {\em  cell}.  A cell may
-hold  a  basic value,  or  the constructor  \CONS\,  or  (parts of)  a
-closure.  Apart from information to  identify the cell type, each cell
-has the space  for two fields. The second of the  fields may be empty,
-when, for example, the cell holds a basic value or a closure involving
-a unary operator. \cred{Should we say more about the representation}.
+\subsection{A liveness-based garbage collection scheme}
+We shall call  a unit allocatable memory as a {\em  cell}.  A cell can
+hold  a  basic  value   ($\mathit{bas}$),  the  constructor  \CONS\  $(\mathit{cons~
+arg_1~arg_2}$) or   a closure.  The closure, in  turn, can be  one of
+($\mathit{unop~arg_1}$),  ($\mathit{ binop~arg_1~arg_2}$)  and  function  application
+($\mathit{app~arg_1~arg_2})$.  Here each $\mathit{ arg_i}$ is a reference to another heap
+cell or terminated through the function $\mathit{makenull}$.
 
@@
 \hfill
-\subfigure[]{\includegraphics[width=.27\linewidth]{lgc-scheme-1}}
+\subfigure[]{\includegraphics[width=.30\linewidth]{lgc-scheme-1}}
 \hfill
-\subfigure[]{\includegraphics[width=.45\linewidth]{lgc-scheme-2}}
+\subfigure[]{\includegraphics[width=.40\linewidth]{lgc-scheme-2}}
 \hfill
-\caption{(a) LGC with closure collection based on reachability  (b)
-  LGC with liveness-based closure collection.}
+\caption{(a) LGC with closure collection based on liveness.  (b)
+  LGC with reachability-based closure collection.}
 
@@
 
-In  the simple algorithm  presented we  consult liveness  automata for
-copying both \CONS cells and closures.  Copying \CONS cells are simple
-and just  involve consulting  the liveness automaton  corresponding to
-them  and  recursively  copying.  In  case of  closures,  it  is  more
-complicated  as  they might  refer  to  data/closures  which might  be
-created in  a different environment(function), all of  which might not
-even exist  on the program  stack. To overcome  this we need  to carry
-extra information with each closure, the program point at which it was
-evaluated and  also the names of  the free variables  contained in the
-closure.  This  gives us  precise  liveness  information  of the  free
-variables of a closure but  involves run-time overhead of updating the
-program  point correctly. To  avoid this  overhead we  go for  a safer
-approximation by storing  the creation point of the  closure and using
-it to check liveness.  This is safe since the liveness at the creation
-point of the closure encompasses  all possible uses of the closure but
-it might be very imprecise as at run-time the actual liveness required
-might be way less than what is found at the creation point.
-
-The advantage of  this method is that it collects  all dead cells even
-if  they are  part  of a  closure  and hence  will  reduce the  memory
-requirements and the number of garbage collections. The main drawbacks
-of  this method  is  the fact  that  the algorithm  might do  multiple
-traversals over  the same structure. Assume that  two references point
-to the  same list, now using  reachability if it  is garbage collected
-using  one reference  there is  no need  to traverse  using  the other
-reference as everything  reachable from the root node  would have been
-copied.  Unfortunately, this is not the case with liveness based GC as
-the two references  might make different sections of  the list as live
-and  copy  them.  Liveness  based  algorithm will  have  to  make  one
-traversal  per reference.   Experimentally we  found that  this  was a
-major source of overhead in the liveness based algorithm.  This can be
-mitigated by  maintaining a  list of liveness  automata states  that a
-particular heap  cell was traversed with  and ensuring that  we do not
-repeat traversals  with the same  state. This requires that  each cell
-maintain  a list  of state  which  potentially might  contain all  the
-states.  Hence  this  is  not  a  feasible  solution.   \cred{repeated
-  traversals if sharing}.
-
-\subsection{Modified GC Algorithm}
-Since the first algorithm requires a lot of bookkeeping and is not
-precise, we have a second version of the algorithm
-which does not use liveness for copying closures.
-\begin{verbatim}
-procedure copy (cell)
-  IF{celltype is cons}
-    IF cell is live &&
-         cell is not copied using reachability
-       new_cell <-  dup_heapcell(cell)
-       new_car        <-  copy(cell.car)
-       new_cell.car   <-  new_car
-       new_cdr        <-  copy(cell.cdr)
-       new_cell.cdr   <-  new_cdr
-    ELSE
-       cell <- NULL
-    ENDIF
-  ELSE
-    IF {celltype is closure && cell is not copied}
-       new_cell  <- copy_and_mark_reachable_cells(cell)
-    ENDIF
-  ENDIF
- copy_references_on_print_stack()
-\end{verbatim}
-
-In this  algorithm we consult  the liveness automata  only if it  is a
-cons cell, for  closures we assume everything under it  to be live and
-copy everything that is reachable. Although this might lead to copying
-more  cells   it  is   faster  as  there   are  no   liveness  checks.
-\comment{Once  experimental  results  are  obtained  compare  the  two
-  algorithms here}  Mixing liveness based GC and  reachability has its
-own set of pros and cons which we discuss further.\cred{hanging like a
-  sore thumb}.
-
-\subsubsection{Optimization for RGC for closures}
-An advantage  of using  reachablity for closures  is that it  opens up
-certain  opportunities  of  optimization.  As  we  saw  in  section[],
-multiple traversals  over the same structure was  a major inefficiency
-in the full liveness based  algorithm and overcoming it required a lot
-of memory.  Using reachability for closures allows us to overcome this
-partially using limited amount of  memory. Any heap cell which is part
-of a closure  is copied using reachaibilty and  hence by maintaining a
-single  flag in  each cell  to indicate  whether it  was  copied using
-reachability    prevents   traversing    the   same    cell   multiple
-times.  Experimental results  again indicate  that for  lazy languages
-this gives a  substantial performance improvement.  \comment{Maybe add
-  a diagram}
-
-\subsubsection{Issues with mixing LGC and RGC}
-A consequence of  copying everything under a closure  is the fact that
-we can end up trying to copy things which are not live and corrupt the
-heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
-liveness. At  a later stage this  cons cell becomes part  of a closure
-and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
-under the  cons cell  thus copying heap  locations which might  not be
-valid.  To avoid  this behavior we either have to choose  to do a full
-liveness based  GC for closures or  we should ensure  that any non-live
-references are  not carried across  GC's. We take the  latter approach
-and ensure that  non-live references are set to  null. After every GC,
-the live part of the buffer  is scanned and any reference which is not
-live is  set to null.  The correctness of  this follows from  the fact
-that any reference which is decalred  not live in a $GC_k$ will not be
-derefernced during later execution.
-% A reference copied using LGC could become part of a closure in a
-later GC
-% and copy wrong locations. Avoiding it by setting non-live references
-to null after every GC.
-% Proof for the correctness.
+Figure~\ref{fig:algos}  shows   a  liveness-based  garbage  collection
+scheme.   The algorithm  assumes that  checking for  liveness involves
+traversing the  liveness-DFAs stored as  a table and does  not mention
+this explicitly.  Starting  with the root set, each  live reference is
+explored   using  $\mathit{  lgcCopy}$.    Copying  \CONS\   cells  is
+simple---it  just involves  copying  the cell  itself and  recursively
+copying the \CAR\ and  the \CDR\ fields.  Liveness-based collection of
+closures  is more  complicated since  our analysis  only  computes the
+liveness of evaluated objects and not closures.
+
+Let   us  consider   determining  the   liveness  of   a   closure  at
+runtime. Assume  that the closure  $(app~f~y)$ has a  reference from
+the root  set variable $x$. If $x$  is live,  then we  know that the
+cell $(app~f~y)$ has  to be copied. However, to find out  the
+liveness of $y$, could find out the program point which triggered the
+evaluation of $(app~f~y)$ and pick up the liveness of $y$ from this
+program point. However, this is only possible if the activation record
+of the function which created this closure is still on the activation
+stack. This does not always happen since a function could pack a
+closure in a \CONS\ cell and return the \CONS\ cell.
+
+An  alternative  could  be   to  carry  extra  information  with  each
+closure---the {\em free variables in the closure and the program point
+  at  which  it  was  evaluated}.   This  gives  us  precise  liveness
+information  of the  free  variables  of a  closure  but involves  the
+overhead of  updating the program point  information during execution.
+To avoid this  overhead we go for a safe  approximation by storing the
+{\em creation  point of the closure}  and using it  to check liveness.
+This is safe  since the liveness at the creation  point of the closure
+dominates all possible evaluation points  of the closure, but it might
+be imprecise  as the actual  liveness might be  way less than  what is
+found at the creation point.
+
+Note that  when a reachability-based  collector visits a cell  for the
+first time,  it is  marked.  Explorations starting  from the  cell are
+curtailed if the cell is  visited subsequently due to sharing. This is
+because the collector can assume  that the region of the heap starting
+from  the  cell has  already  been  explored.  Since a  liveness-based
+collector may  reach the cell each  time with a  different state, this
+curtailment cannot  be done in  the case of  LGC. This is  an inherent
+drawback  of a  liveness-based collector.   This can  be  mitigated by
+maintaining a list of liveness  automata states that a particular heap
+cell was traversed with and  ensuring that we do not repeat traversals
+if we reach the cell in one of these states.
+
+\subsection{Liveness-based collector with reachability-based closure
+  collection}  Since the first  algorithm has  significant bookkeeping
+overheads,  we have  a  second GC  scheme  that not  use liveness  for
+copying closures.  In this  scheme, the liveness automata is consulted
+only for \CONS\ cells; we assume everything under a closure to be live
+and copy everything  that is reachable.
+
+A consequence of copying everything under a closure is that we can end
+up trying to copy things which are not live and corrupt the heap. This
+can happen if a \CONS\  cell copied using liveness during a collection
+subsequently  becomes   part  of  a  closure.    During  a  subsequent
+collection,  the  collector  may  try  to copy  everything  under  the
+\CONS\ cell thus copying heap  locations which might not be valid.  To
+avoid  this, we  should ensure  that any  non-live references  are not
+carried across  GC's.  We  do this  by scanning the  live part  of the
+buffer after every  GC and setting to null any  reference which is not
+live. This  is correct  because any reference  which is dead  during a
+collection will not be dereferenced any further.
+
+An advantage of  using reachablity for closures it  avoids the problem
+of  multiple  explorations  from  the  same  cell.  This  is  done  by
+maintaining  a flag in  each cell  to indicate  whether it  was copied
+using  reachability.
+
 \subsection{Garbage collection for references on print stack}
-Another  difference that is  present in  GC of  lazy language  is the
-presence  of  a print  stack.  In  lazy  languages the  evaluation  of
-closures is  driven by the print  function\comment{Refer to SPJ's
-book}.
-Printing of atomic values is simple and does not trigger a garbage
-collection.
-In case of \CONS\ cells, the \CAR\ part is first printed and then the \CDR\
-part.
-In a lazy language, both the \CAR\ and the \CDR\ parts could be closures,
-thus
-requiring evaluation. A garbage collection might be triggered while printing
-the \CAR\ part or the \CDR\ part. Therefore, whenever a garbage collection is
-triggered we have to consider any references that might be on the print stack
-and copy them. Liveness automata does not exist for the print function as it
-is not part of the executing program. Since the print function makes
-everything
-reachable from a cell live we can copy all cells reachable from the
-reference.
-
-Consider the case where a \CONS\ cells is being printed and a garbage
-collection
-is triggered when printing the \CDR\ part. In this case, since the refernce
-to the
-\CONS\ cells is still on the print stack both the \CAR\ and the \CDR\ parts
-are
-reachable but we know that the \CAR\ part has already been printed and can be
-safely garbage collected. A liveness analysis of the print function would
-have
-given the same information. In our implementation we mimic the behavior of
-the liveness analysis by using a flag in every \CONS\ cell which says whether
-the \CAR\ part has been printed or not. For any \CONS\ cell if the \CAR\ part
-has not been printed, a reference to the \CAR\ part will also be on the stack
-but
-the \CDR\ part has to be explicitly copied.
-In case the \CAR\ part has been printed, then a reference to the \CDR\ part
-will
-be on the stack and since the \CAR\ part is not required anymore we do not
-copy anything. This optimization subsumes tail call optimization. Again,
-experimental results show that this optimization minimizes a lot of copying.
-\cred{Very sketchy: Make another attempt to write}
-\hrule
-\comment{
-Unlike eager languages where GC happens only at \CONS~ points or at
-function calls, every \LET~ point is a potential GC point. The number
-of
-automata that needs to be stored is reduced by sharing automata along
-all
-the program points along an {\em E-path}.
-
-By the  time the program  starts executing, the liveness  automata are
-created and stored in a file.   This is re-used every time the program
-executes.  An important difference in  GC of lazy language compared to
-an  eager language is  the point  of garbage  collection.  In  case of
-eager  languages only  \CONS~ points  and  function calls  are the  GC
-points. In  a lazy  language every \LET~  statement is a  potential GC
-point. Hence we need to store liveness automata corresponding to every
-\LET~  statement in the  program. This  leads to  an explosion  in the
-number of  automata. From our  observation that along an  {\em E-path}
-all program points have the same liveness, we can reduce the number of
-automata state required  by sharing the same automata  for all program
-points along an {\em E-path}.} In  the example program, we only need to
-create automata  for program  points \ldots, since  they are  the only
-distinct {\em E-path}s in the program.
-
-% Liveness for cons cells & reachability for closures
-
-\comment{Mention the fact that liveness at a program point indicates
-  what references will be used by the program eventually. The exact
-  point where a reference will be used cannot be determined statically
-due to the language being lazy. }
-During GC, if the heap cell being processed is a cons cell, we consult
-the  liveness automaton  corresponding to  it and  copy only  the live
-paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
-approximate the  liveness to be {\{0+1\}\*} and  copy everything under
-the closure.
-% Handling of mixture of LGC & RGC
-
-% Difference between eager and lazy GC
-
-% Doing GC for  references on the print stack - using  a flag to avoid
-% copying non-live  cells Optimizing  GC for closures  - use  a copied
-% using rgc flag for closures to avoid multiple traversals
+In lazy  languages the evaluation of  closures is driven  by the print
+function~\cite{Jones87}.  Printing of  atomic values is
+simple  and  does  not  trigger  a garbage  collection.   In  case  of
+\CONS\ cells, the \CAR\ is first  printed followed by \CDR.  In a lazy
+language, both \CAR\ and \CDR\ could be closures requiring evaluation,
+and this  may trigger  a garbage collection.   When this  happens.  we
+have to consider  any references that might be on  the print stack and
+copy them. We extend the liveness analysis to the print function and
+use its results during garbage collection.
+
+%% Liveness automata  does not exist for  the print function  as it is
+%% not part of the executing  program.  Since the print function makes
+%% everything  reachable  from a  cell  live  we  can copy  all  cells
+%% reachable from the reference.
+
+%% Consider  the case  where a  \CONS\ cells  is being  printed  and a
+%% garbage collection  is triggered when  printing the \CDR\  part. In
+%% this case, since  the refernce to the \CONS\ cells  is still on the
+%% print stack both the \CAR\ and the \CDR\ parts are reachable but we
+%% know that the \CAR\ part has already been printed and can be safely
+%% garbage collected. A liveness  analysis of the print function would
+%% have given  the same information.   In our implementation  we mimic
+%% the  behavior of the  liveness analysis  by using  a flag  in every
+%% \CONS\ cell which  says whether the \CAR\ part  has been printed or
+%% not. For any \CONS\ cell if  the \CAR\ part has not been printed, a
+%% reference  to the  \CAR\ part  will also  be on  the stack  but the
+%% \CDR\ part has to be explicitly copied.  In case the \CAR\ part has
+%% been printed,  then a reference  to the \CDR\  part will be  on the
+%% stack and  since the \CAR\ part  is not required anymore  we do not
+%% copy anything.  This  optimization subsumes tail call optimization.
+%% Again, experimental results show that this optimization minimizes a
+%% lot of copying.  \cred{Very sketchy: Make another attempt to write}
+%% \hrule
+
+%% \comment{ Unlike  eager languages where  GC happens only  at \CONS~
+%%   points or at function calls,  every \LET~ point is a potential GC
+%%   point.  The number of automata that needs to be stored is reduced
+%%   by sharing  automata along all  the program points along  an {\em
+%%   E-path}.
+
+%% By the time the program starts executing, the liveness automata are
+%% created  and stored  in a  file.  This  is re-used  every  time the
+%% program executes.   An important difference in GC  of lazy language
+%% compared to an  eager language is the point  of garbage collection.
+%% In case  of eager languages  only \CONS~ points and  function calls
+%% are the  GC points. In a  lazy language every \LET~  statement is a
+%% potential  GC  point. Hence  we  need  to  store liveness  automata
+%% corresponding to  every \LET~ statement in the  program. This leads
+%% to an  explosion in  the number of  automata. From  our observation
+%% that  along  an {\em  E-path}  all  program  points have  the  same
+%% liveness, we  can reduce the  number of automata state  required by
+%% sharing  the same  automata for  all program  points along  an {\em
+%% E-path}.} In the  example program, we only need  to create automata
+%% for program  points \ldots, since  they are the only  distinct {\em
+%% E-path}s in the program.
 
@@
 \subsection{Results}
-\begin{figure}[t]
-\centering
-\begin{boxedminipage}{.75\textwidth}
-{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\
-\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-  then} {\bf skip}\\
-\noindent \hspace*{.25cm}  {\bf else} {\bf if}
-$\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR =
-\mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\
-\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$
-{\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa},
-\mathsf{dfa}(q,1),x.\CDR)$\\
-\noindent \hspace*{.35cm}{\bf return} $y$
-\end{boxedminipage}
-\caption{Function       for       copying       a       root       set
-  variable.}\label{algo:LGC-main-loop} \figrule
-
-\end{figure}
+%% \begin{figure*}[t]
+%% \centering
+%% \begin{boxedminipage}{.75\textwidth}
+%% {\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\
+%% \noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
+%% \noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
+%%   then} {\bf skip}\\
+%% \noindent \hspace*{.25cm}  {\bf else} {\bf if}
+%% $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR =
+%% \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\
+%% \noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$
+%% {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa},
+%% \mathsf{dfa}(q,1),x.\CDR)$\\
+%% \noindent \hspace*{.35cm}{\bf return} $y$
+%% \end{boxedminipage}
+%% \caption{Function       for       copying       a       root       set
+%%   variable.}\label{algo:LGC-main-loop} \figrule
+
+%% \end{figure*}
 
@@
 cells different from the earlier visit.
-\begin{figure}[t]
+\begin{figure*}[t]
 \footnotesize
@@
 \normalsize
-\end{figure}
-
-\section{Collecting more garbage can never slow things down}
-\label{sec:lgc-always-better}
+\end{figure*}
+
+%% \section{Collecting more garbage can never slow things down}
+%% \label{sec:lgc-always-better}
 
 %% Since garbage collection is  effectively asynchronous to the
-allocator
+%% allocator
 %% thread, one might  worry as to how {\em  robust} our measurements
-are.
+%% are.
 %% For example,  while LGC would,  in general, collect more  garbage
-than
+%% than
 %% RGC in the same heap state,  is it possible that on some programs,
-LGC
+%% LGC
 %% might do  a larger number  of collections? We  prove a result  to
-show
+%% show
 %% that   this  cannot   happen.    This  result   applies  to
-classical
+%% classical
 %% mark-and-sweep and copying garbage  collectors and, we hypothesize,
-to
+%% to
 %% generational collectors too.
@@
 %% For the  same mutator,  a liveness-based collector  can never  do
-more
+%% more
 %% garbage collections than a reachability-based collector.
@@
 %% \begin{proof}Assume, as before,  that time is measured in terms of
-the number of
+%% the number of
 %%   cons cells allocated. Let us run  two copies of the mutator with
-the
+%% the
 %%   two    different   collectors    but    with   synchronous
-memory
+%% memory
 %%   allocations.   The   garbage    collections,   however,   happen
@@
 %%   To prove the lemma, it is enough to show the truth of the
-following
+%% following
 %% statement: After
 %%  every  LGC invocation,  the count of LGC invocations   is  no
-greater
+%% greater
 %% than  RGC invocations.   The  base case  holds  since the  first
 %% invocations of both GCs happen at the same time.  Assume the
-statement
+%% statement
 %% to be true after $n$ invocations of LGC.  Since LGC copies a subset
-of
+%% of
 %% reachable cells, its heap would contain no more cells than RGC heap
-at
+%% at
 %% the end  of the $n$th invocation.   Thus either RGC  is invoked next
-before
+%% before
 %% LGC, or LGC and RGC are both invoked next at the same time.  In
-either
+%% either
 %% case, the statement holds after $n+1$ invocations of LGC.
@@
 
-Generated by bibtex from your ~.bib file.  Run latex,
-then bibtex, then latex twice (to resolve references)
-to create the ~.bbl file.  Insert that ~.bbl file into
-the .tex source file and comment out
-the command \texttt{{\char'134}thebibliography}.
-% This next section command marks the start of
-% Appendix B, and does not continue the present hierarchy
-\section{More Help for the Hardy}
-The sig-alternate.cls file itself is chock-full of succinct
-and helpful comments.  If you consider yourself a moderately
-experienced to expert user of \LaTeX, you may find reading
-it useful but please remember not to change it.
-%\balancecolumns % GM June 2007
-% That's all folks!
-
-% {\color{red}
-%   (Note 3.\ includes $e_\mainpgm$ treated as the body of
-%   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
-% }
-
-%% =======
-
-%------------------------------------------------------------%
-%\input{lazy-liveness}
+%% Generated by bibtex from your ~.bib file.  Run latex,
+%% then bibtex, then latex twice (to resolve references)
+%% to create the ~.bbl file.  Insert that ~.bbl file into
+%% the .tex source file and comment out
+%% the command \texttt{{\char'134}thebibliography}.
+%% % This next section command marks the start of
+%% % Appendix B, and does not continue the present hierarchy
+%% \section{More Help for the Hardy}
+%% The sig-alternate.cls file itself is chock-full of succinct
+%% and helpful comments.  If you consider yourself a moderately
+%% experienced to expert user of \LaTeX, you may find reading
+%% it useful but please remember not to change it.
+%% %\balancecolumns % GM June 2007
+%% % That's all folks!
+
+%% % {\color{red}
+%% %   (Note 3.\ includes $e_\mainpgm$ treated as the body of
+%% %   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
+%% % }
+
+%% %% =======
+
+%% %------------------------------------------------------------%
+%% %\input{lazy-liveness}
 

### 2015-06-12 lazy-scheme/src/Demands.cpp
Partial fixes for liveness

@@ -150,7 +150,8 @@ void make_fa(const std::string & start,
     }
 }
 
-void combineEpsilonEdgesFrom(const std::string & start, automaton * nfa) {
+void combineEpsilonEdgesFrom(const std::string & start, automaton * nfa)
+{
     if(marked_states.find(start) != marked_states.end())    return;
     marked_states.insert(start);
 
@@ -354,12 +355,44 @@ void simplifyWithoutEpsilonEdges(const std::string & start, automaton * nfa) {
 
 void removeXbEdgesFrom(automaton * nfa)
 {
-    for(auto & non_terminal : nfa->second) {
-        if(non_terminal.second.find(TXb) != non_terminal.second.end()) {
-            non_terminal.second.erase(TXb);
-            nfa->first.insert(non_terminal.first);
-        }
+
+	//This is not correct. Need to re-write.
+	std::string prev_nt = E;
+	for(auto & non_terminal : nfa->second)
+    {
+		auto Xb_transition = non_terminal.second.find(TXb);
+//
+//        if(Xb_transition != non_terminal.second.end())
+//        {
+//        	//This should handle cases like 0b2 or 1b2 is encountered.
+//        	//In such cases there can be no path hence do not add an epsilon edge
+//            if (prev_nt == T0b || prev_nt == T1b)
+//            {
+//            	non_terminal.second.erase(TXb);
+//            }
+//            else
+//            {
+//            	//For all other cases add an epsilon edge instead of the the 2 edge
+//
+//            	auto dest_states = Xb_transition->second;
+//            	non_terminal.second.erase(TXb);
+//            	(non_terminal.second[E]).insert(dest_states.begin(), dest_states.end());
+//            	//If the epsilon edge takes it to a final state, mark the current state as final
+//            	nfa->first.insert(non_terminal.first);
+//            }
+//
+//        }
+		//Currently add an epsilon edge unconditionally, and make the current state a final state
+		  if(Xb_transition != non_terminal.second.end())
+		  {
+			  std::cerr << "Replacing Xb by epsilon edge for " << non_terminal.first << std::endl;
+			  auto dest_states = Xb_transition->second;
+			  (non_terminal.second[E]).insert(dest_states.begin(), dest_states.end());
+			  non_terminal.second.erase(TXb);
[…]

### 2015-06-12 lazy-scheme/src/SchemeAST_Defs.cpp

@@ -517,7 +517,7 @@ LetExprNode * LetExprNode::clone() const
 cons* LetExprNode::evaluate()
 {
 	curr_return_addr = getLabel();
-//	cout << "Processing let variable " << this->pID->getIDStr() << " at label" << getLabel() <<  endl;
+	cerr << "Processing let variable " << this->pID->getIDStr() << " at label" << getLabel() <<  endl;
 //	cout << "Pointing to " << static_cast<cons*>(getfree()) << endl;
 #ifdef __MYDEBUG 
 //	cout << "Processing let variable " << this->pID->getIDStr() << " at label " << getLabel() <<  endl;
@@ -704,12 +704,13 @@ cons* UnaryPrimExprNode::evaluateCarExpr()
 {
 
 	cons* heap_cell = update_heap_refs.top();
-
+	cerr << "Evaluating car expression for " << heap_cell << endl;
 	if (heap_cell->inWHNF)
 		return heap_cell;
 	else
 	{
 		update_heap_refs.push(heap_cell->val.closure.arg1);
+		cerr << "Evaluating argument 1 for " << heap_cell->val.closure.arg1 << endl;
 		cons* temp = reduceParamToWHNF(heap_cell->val.closure.arg1);
 
 		assert(temp->inWHNF);
@@ -788,6 +789,7 @@ cons* UnaryPrimExprNode::evaluateCdrExpr()
 // that the expression is representing
 cons* UnaryPrimExprNode::evaluateNullqExpr()
 {
+	cerr << "Evaluating nullq expression"<<endl;
 	cons* heap_cell = update_heap_refs.top();
 	if (heap_cell->inWHNF)
 		return heap_cell;

### 2015-06-12 lazy-scheme/src/Simulator.cpp

@@ -146,7 +146,7 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 	bool state_transition_file =  (stat ((outdir + pgmname + "/fsmdump-" + pgmname + "-state-transition-table").c_str(), &buffer) == 0);
 	filesCached = state_map_file && state_transition_file;
 
-	if (gc_type == gc_live && !filesCached)
+	if (gc_type == gc_live && !filesCached )
 	{
 		
 		//Instead of driver.process returning an integer why can't it return the grammar?

### 2015-06-12 lazy-scheme/src/gc.cpp

@@ -238,7 +238,10 @@ cons* deep_copy(cons* node, int gc_type, ostream& out)
 		if (node != NULL)
 		{
 			if (node->typecell != consExprClosure && node->forward != NULL)
+			{
+				//DBG(cout<<"Node already copied " << endl;);
 				return static_cast<cons*>(node->forward);
+			}
 			else if (node->typecell == consExprClosure)
 			{
 
@@ -254,6 +257,7 @@ cons* deep_copy(cons* node, int gc_type, ostream& out)
 				}
 				else
 				{
+					//DBG(cout<<"Node already copied " << endl;);
 					return static_cast<cons*>(node->forward);
 				}
 			}
@@ -1176,22 +1180,25 @@ void update_heap_ref_stack(ostream& out, int gc_type)
 		temp.push(heap_ref);
 		print_stack.pop();
 	}
+
 	while(!temp.empty())
 	{
 		print_stack.push(temp.top());
 		temp.pop();
 	}
+
+
 	int i = 0;
 	while(!update_heap_refs.empty())
 	{
 		cons* heap_ref = update_heap_refs.top();
-		if (!is_valid_address(heap_ref) && heap_ref->forward == NULL)
+		if (heap_ref->forward == NULL && !is_valid_address(heap_ref))
 		{
 			cout << "ERROR !!!! " << heap_ref << endl;
 			cout << "In heap ref stack copying " << (heap_ref - (cons*)buffer_dead) << " with index " << i << endl;
 			assert(false);
 		}
-
+		cerr << "Updating heap ref " << heap_ref << " to " << heap_ref->forward << endl;
 		heap_ref = static_cast<cons*>(heap_ref->forward);
 		temp.push(heap_ref);
 		update_heap_refs.pop();
@@ -1589,12 +1596,9 @@ void liveness_gc()
 cons* followpaths(cons* loc, state_index index, ostream& out)
 {
 
+	DBG(out<<"Entering followpaths"<<endl);
 	if (NULL == loc)
 		return loc;
-
-	cons* loccopy = copy(loc, out);
-
-    DBG(out << "Copying cell from " << loc << " with type " << loc->typecell << endl);
 
 #ifdef ENABLE_SHARING_STATS
   if (loccopy)
@@ -1602,20 +1606,31 @@ cons* followpaths(cons* loc, state_index index, ostream& out)
 #endif
[…]

### 2015-06-29 paper/paper.tex
Updated Results section first draft

@@
 \section{Results}
-To demonstrate the effectiveness of liveness-based garbage collection,
+\comment{To demonstrate the effectiveness of liveness-based garbage collection,
 we  have  built a  prototype  consisting  of  an interpreter  for  our
@@
 becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.
-
+of garbage collection.}
+
+To test the utility of our method we have implemented a garbage-collector 
+for a simple scheme like first order language. It consists of an interpreter, 
+liveness analyzer and an interchangeable garbage-collector that can use reachability or liveness.
+Our benchmark consists of programs from nofib suite and some external programs manually converted to ANF.
+
+\comment{
 While exploring the activation record of a function, LGC uses the DFAs
@@
 call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.
-
+the DFAs at the point of call to  ${\mathit g}$ would be used.}
+
+The process of liveness-based garbage collection involves going through 
+each activation record on the stack and exploring each variable in the 
+current activation record (root set). The liveness of each variable is 
+determined using the program point and the variable name. The liveness 
+of all variables are stored as DFAs. These DFAs are then encoded as a 
+table. A variable is live only if an entry is found in this table.
+All cells that are live are copied to the live semi-space. In case of 
+\CONS\ cells the \CAR\ and \CDR\ pointers are chased and if they are 
+live the copied \CONS\ cell will get the updated addresses of its 
+\CAR\ and \CDR\ fields.
+
+\comment{
 Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
@@
 value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.
-
+pointer otherwise.}
+
+\subsection{Copying closure using liveness vs Copying closures using reachability}
+To compare the effect of using reachability to copy closures instead of liveness we take a couple of programs from our benchmark suite and execute them with 
+both. As shown in figure-add-figure-here, the number of garbage collections 
+in both cases are same but the time taken varies largely. Both these observations can be explained as follows, since we approximate the liveness of closure to 
+the liveness at its creation point, the liveness we get is usually grossly over approximated.
+Using this liveness may not give too much of a benefit compared to reachability as nearly everything would be live at its point of creation.
+As mentioned in section-ref using liveness for closures may entail multiple traversals over the same structure consuming more time.
+Therefore we believe that using reachability for closures is a sweet spot in terms of performance for liveness based garbage collectors.
+
+\comment{
 The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
@@
 {\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.
+manages to collect a good portion of the data that are not live.}
 
@@
 %% \end{figure}
+The table in Figure~\ref{fig:experimental-results} shows that LGC improves the 
+garbage collection time and efficiency for most of the programs. 
+Even for programs where the time taken is more it is very competitive. 
+As the number of garbage collections tends to be very large our graphs show only 
+a window which is representative of the behavior for that particular benchmark. 
+In all the programs we can see that the curve corresponding to liveness regularly 
+dips below the reachability curve.
+
+The only area of concern is the huge gap between the actual liveness and the 
+liveness percieved by our collector. In case of LGC for eager languages the 
+gap was very narrow and almost touched the actual liveness curve. In case of 
+lazy langauges the inefficiency is due to the non-live cells which become part of 
+a closure and hence get dragged around. Our experiments have shown that even if 
+we use the approximated liveness we do not get too much benefits in terms of non-live
+ cells inside closures.
+\comment{
 The  increased effectiveness  of LGC  over RGC  is also  shown  in the
@@
 cells different from the earlier visit.
+}
 \begin{figure*}[t]

### 2015-06-29 lazy-scheme/src/gc.h

@@ -22,7 +22,7 @@ using namespace Scheme::AST;
 
 #ifndef __DEBUG__GC
 #undef __DEBUG__GC
-#define __DEBUG__GC
+//#define __DEBUG__GC
 #endif
 
 //#define REACHABILITY_BFS

### 2015-07-01 paper/paper.tex
Updated results section

@@
 \section{Results}
-\comment{To demonstrate the effectiveness of liveness-based garbage collection,
-we  have  built a  prototype  consisting  of  an interpreter  for  our
-language,  a  liveness  analyzer  and  a copying  collector  that  can
-optionally use the results of liveness analysis for marking instead of
-reachability. When the collector uses liveness for marking, we call it
-a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-  reachability-based  collector}  (RGC).   The  copying  collector  is
-neither incremental nor generational.  As a consequence, any cell that
-becomes unreachable or  dead is assuredly collected in  the next round
-of garbage collection.}
+%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
+%% we  have  built a  prototype  consisting  of  an interpreter  for  our
+%% language,  a  liveness  analyzer  and  a copying  collector  that  can
+%% optionally use the results of liveness analysis for marking instead of
+%% reachability. When the collector uses liveness for marking, we call it
+%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
+%%   reachability-based  collector}  (RGC).   The  copying  collector  is
+%% neither incremental nor generational.  As a consequence, any cell that
+%% becomes unreachable or  dead is assuredly collected in  the next round
+%% of garbage collection.}
 
 To test the utility of our method we have implemented a garbage-collector 
-for a simple scheme like first order language. It consists of an interpreter, 
-liveness analyzer and an interchangeable garbage-collector that can use reachability or liveness.
+for a simple scheme-like first order language. It consists of an interpreter, 
+liveness analyzer and a garbage-collector that can optionally use use reachability or liveness. 
 Our benchmark consists of programs from nofib suite and some external programs manually converted to ANF.
 
-\comment{
-While exploring the activation record of a function, LGC uses the DFAs
-at  the latest  program  point traversed  in the function's body.
-Thus, if  LGC
-is invoked  at the program point   $\pi$,  it would  use  the
-liveness  DFAs at  $\pi$ itself for exploring the root set of the
-current activation record. For any other activation record in the
-stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
-call-chain,
-the DFAs at the point of call to  ${\mathit g}$ would be used.}
+%% \comment{
+%% While exploring the activation record of a function, LGC uses the DFAs
+%% at  the latest  program  point traversed  in the function's body.
+%% Thus, if  LGC
+%% is invoked  at the program point   $\pi$,  it would  use  the
+%% liveness  DFAs at  $\pi$ itself for exploring the root set of the
+%% current activation record. For any other activation record in the
+%% stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
+%% call-chain,
+%% the DFAs at the point of call to  ${\mathit g}$ would be used.}
 
@@
 
-\comment{
-Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-$\dfa{\pi}{x}$.
-Considering   a   DFA   as   a   table,
-$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-\acar\      or     \acdr.       We      shall     also      write
-$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-indicating  whether there  is a  transition from  $q$ on
-$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-pointer to $x$.
-The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-$x$  in the  other semi-space  and returns  the new  address.  It
-hides  details such as  returning the  forwarding pointer  if the
-value  of $x$  is  already copied,  and  creating the  forwarding
-pointer otherwise.}
+%% \comment{
+%% Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
+%% point  pair ($x$,  $\pi$).  We  assume that  there is  a function
+%% $\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
+%% $\dfa{\pi}{x}$.
+%% Considering   a   DFA   as   a   table,
+%% $\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
+%% state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
+%% \acar\      or     \acdr.       We      shall     also      write
+%% $\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
+%% indicating  whether there  is a  transition from  $q$ on
+%% $\mathit{sym}$.  The LGC action to chase the root variable $x$ at
+%% $\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
+%% empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
+%% $\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
+%% in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
+%% pointer to $x$.
+%% The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
+%% $x$  in the  other semi-space  and returns  the new  address.  It
+%% hides  details such as  returning the  forwarding pointer  if the
+%% value  of $x$  is  already copied,  and  creating the  forwarding
+%% pointer otherwise.}
 
@@
 Using this liveness may not give too much of a benefit compared to reachability as nearly everything would be live at its point of creation. 
-As mentioned in section-ref using liveness for closures may entail multiple traversals over the same structure consuming more time.
+As mentioned in section-ref, using liveness for closures may entail multiple traversals over the same structure consuming more time. 
 Therefore we believe that using reachability for closures is a sweet spot in terms of performance for liveness based garbage collectors. 
-
-\comment{
-The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-in the heap over time for RGC and LGC---here time is measured in terms
-of the number of cons cells allocated. In addition, they also show the
-number of  reachable cells and the  number of cells  that are actually
-live\footnote{This   is  statically   approximated  by   our  liveness
-  analysis}.  Since  the programs have  different memory requirements,
-we have tuned the size of heap for each program to ensure a reasonable
-number of  collections. An invocation  of RGC decreases the  number of
-cells  in heap  till  it touches  the  curve of  reachable cells.   An
-invocation of LGC decreases the number  of heap cells to no lower than
-the curve of live cells.
-
-To construct the  reachable and live curves, we  record for every cell
-its creation time (\CreateTime), its last use time (\UseTime), and the
-earliest time  when the  cell becomes unreachable  and can  be garbage
-collected (\CollTime).  For accurate recording of  \CollTime, we force
-frequent invocations  of a reachability based collector  in a separate
-run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-The benchmark programs are drawn from the {\tt no-fib} suite and other
-sources and have been  manually converted to  ANF.  All  graphs  except
- {\tt  fibheap}  show  fewer  garbage
-collector  invocations for  LGC. The  number  of reachable  cells in
-{\tt
-  fibheap} increases  initially, comes  down before the  heap becomes
-gets
-full and then  continues to remains low.  This results in  a single
-garbage
-collection in both RGC and LGC.  The graphs also show the precision of
-our liveness analysis.  For  all programs  except  {\tt nperms} and
-{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-manages to collect a good portion of the data that are not live.}
+%%
+%% \comment{
+%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
+%% in the heap over time for RGC and LGC---here time is measured in terms
+%% of the number of cons cells allocated. In addition, they also show the
+%% number of  reachable cells and the  number of cells  that are actually
+%% live\footnote{This   is  statically   approximated  by   our  liveness
+%%   analysis}.  Since  the programs have  different memory requirements,
+%% we have tuned the size of heap for each program to ensure a reasonable
+%% number of  collections. An invocation  of RGC decreases the  number of
+%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
+%% invocation of LGC decreases the number  of heap cells to no lower than
+%% the curve of live cells.
+
+%% To construct the  reachable and live curves, we  record for every cell
+%% its creation time (\CreateTime), its last use time (\UseTime), and the
+%% earliest time  when the  cell becomes unreachable  and can  be garbage
+%% collected (\CollTime).  For accurate recording of  \CollTime, we force
+%% frequent invocations  of a reachability based collector  in a separate
+%% run.     A    cell   is    live    at    time   \mbox{\exectime}    if
+%% \mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
+%%   \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
+%%   \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
+
+%% The benchmark programs are drawn from the {\tt no-fib} suite and other
+%% sources and have been  manually converted to  ANF.  All  graphs  except
+%%  {\tt  fibheap}  show  fewer  garbage
+%% collector  invocations for  LGC. The  number  of reachable  cells in
+%% {\tt
+%%   fibheap} increases  initially, comes  down before the  heap becomes
+%% gets
+%% full and then  continues to remains low.  This results in  a single
+%% garbage
+%% collection in both RGC and LGC.  The graphs also show the precision of
+%% our liveness analysis.  For  all programs  except  {\tt nperms} and
+%% {\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
+%% manages to collect a good portion of the data that are not live.}
+%%
 
@@
 Even for programs where the time taken is more it is very competitive. 
-As the number of garbage collections tends to be very large our graphs show only 
+As the number of garbage collections tend to be very large our graphs show only 
 a window which is representative of the behavior for that particular benchmark. 
@@
 
-The only area of concern is the huge gap between the actual liveness and the 
+One area of concern is the huge gap between the actual liveness and the 
 liveness percieved by our collector. In case of LGC for eager languages the 
-gap was very narrow and almost touched the actual liveness curve. In case of 
-lazy langauges the inefficiency is due to the non-live cells which become part of 
-a closure and hence get dragged around. Our experiments have shown that even if 
-we use the approximated liveness we do not get too much benefits in terms of non-live
- cells inside closures.
-\comment{
-The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-provides  statistics  regarding the  analysis  itself.  The number  of
-states and the analysis times  are within tolerable limits. Precision
-of analysis refers to the percentage of dead cells that is collected by
-LGC, averaged over all invocations.
-The second table shows garbage  collection statistics for RGC and LGC.
-LGC  collects larger garbage  per invocation,  drags cells  for lesser
-time and requires a smaller heap size ({\em MinHeap}) for program to
-run
-in comparison with  RGC.
-
-There are a couple of  issues of concern.  The garbage collection time
-is larger in the case of LGC for some programs. The reason is that the
-cost of consulting the liveness DFA may outweigh the combined benefits
-of  fewer   garbage  collections   and  fewer  markings   per  garbage
-collection.   The  other issue  is  illustrated  by  the program  {\tt
-  lambda}    (memory   usage  shown   separately    in
-Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-in Figure~\ref{fig:experimental-results}, the  number of
-touched cells\footnote{These  are the  cells which are  visited during
-  marking phase, often  more than once due to  sharing.} in this
-example is much higher for
-LGC.  This  increase is due to  excessive sharing among  heap nodes in
-this program.  Note  that a node re-visited because  of sharing is not
-explored  any   further  during  a  RGC   collection.   However,  this
-curtailment cannot happen in  LGC because of the possibility that the
-node,  re-visited in a  different liveness  state, may  mark a  set of
-cells different from the earlier visit.
-}
+gap was very narrow and almost touched the actual liveness curve. 
+In case of lazy languages, due to using reachability for copying we end up copying 
+more cells. Any data which is part of closure irrespective of whether it is live 
+or not gets copied. Thus many non-live cells get copied during garbage collection 
+increasing the drag time.
+Our experiments have shown that unless we use the exact execution point for 
+closures bringing down the number of non-live cells copied in closures cannot be reduced.
+Using the approximated liveness (liveness at the creation point) does not give benefits in terms of non-live
+ cells inside closures as everything would be live at the point of creation.
+%% \comment{
+%% The  increased effectiveness  of LGC  over RGC  is also  shown  in the
+%% tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
+%% provides  statistics  regarding the  analysis  itself.  The number  of
+%% states and the analysis times  are within tolerable limits. Precision
+%% of analysis refers to the percentage of dead cells that is collected by
+%% LGC, averaged over all invocations.
+%% The second table shows garbage  collection statistics for RGC and LGC.
+%% LGC  collects larger garbage  per invocation,  drags cells  for lesser
+%% time and requires a smaller heap size ({\em MinHeap}) for program to
+%% run
+%% in comparison with  RGC.
+
+%% There are a couple of  issues of concern.  The garbage collection time
+%% is larger in the case of LGC for some programs. The reason is that the
+%% cost of consulting the liveness DFA may outweigh the combined benefits
+%% of  fewer   garbage  collections   and  fewer  markings   per  garbage
+%% collection.   The  other issue  is  illustrated  by  the program  {\tt
+%%   lambda}    (memory   usage  shown   separately    in
+%% Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
+%% in Figure~\ref{fig:experimental-results}, the  number of
+%% touched cells\footnote{These  are the  cells which are  visited during
+%%   marking phase, often  more than once due to  sharing.} in this
+%% example is much higher for
+%% LGC.  This  increase is due to  excessive sharing among  heap nodes in
+%% this program.  Note  that a node re-visited because  of sharing is not
+%% explored  any   further  during  a  RGC   collection.   However,  this
+%% curtailment cannot happen in  LGC because of the possibility that the
+%% node,  re-visited in a  different liveness  state, may  mark a  set of
+%% cells different from the earlier visit.
+%% }
 \begin{figure*}[t]

### 2015-07-03 paper/paper.tex
new abstract

@@
        \affaddr{Mumbai 400076, India}
-       \email{prasanna@cse.iitb.ac.in}
+       \email{prasannak@cse.iitb.ac.in}
 % 2nd. author
@@
 first-order functional programs.  To do  this, we analyze programs for
-liveness of heap  allocated data, i.e. data that  can possibly be used
-in the future.  The accompanying garbage collector is  modified to use
-the result of the analysis to preserve only {\em live} data---a subset
-of {\em reachable data}---during  garbage collection.  This results in
-an  increase in  the garbage  reclaimed and  consequently  reduces the
-memory footprints of programs.   While this technique has already been
-shown to yield benefits  for eager first-order languages, the presence
-of  closures   at  run-time  poses  additional   challenges  for  lazy
-languages.   Closures require  changes  both in  the earlier  liveness
+{\em liveness} of heap allocated data, i.e.  data that can possibly be
+used in the future.  The accompanying garbage collector is modified to
+use the  result of the analysis  to preserve only {\em  live} data---a
+subset  of {\em  reachable  data}---during  garbage collection.   This
+results  in an  increase  in the  garbage  reclaimed and  consequently
+reduces the memory  footprints of programs.  While  this technique has
+already been shown to yield  benefits for eager first-order languages,
+the presence of  closures at run-time poses  additional challenges for
+lazy languages.  Closures require changes both in the earlier liveness
 analysis for heap allocated memory  and also in the garbage collection
@@
 
-We  formulate  a context-sensitive  liveness  analysis  and using  the
-results annotate each potential allocation point in the program with a
-set     of    deterministic    finite-state     automata(DFA).     The
-garbage-collector  inspects  the DFA  to  curtail reachability  during
-marking.  As a result, fewer  objects are marked (with a possibly more
-expensive  marker)  and  preserved  by the  copy  phase.   Experiments
-confirm  that  liveness-based  collection  results in  a  increase  in
-garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
-collections. On  our benchmark programs, we obtain  a decrease between
-x\% and y\% in the memory size required to run programs, when compared
-to reachability-based garbage collection.
+Our  analysis  is a  context-sensitive  analysis  that annotates  each
+potential  garbage collection  point  in  the program  with  a set  of
+deterministic finite-state  automata (DFA) describing the  liveness at
+that  point.   We  also  present a  practical  implementation  of  our
+analysis in  a garbage collector.   The collector, when  invoked, uses
+the DFA set to preserve live  objects.  As a result, fewer objects are
+marked and preserved when compared  to a reachability based marking.
+Our experiments  confirm that liveness-based collection  increases the
+garbage reclaimed,  consequently decreasing the number  of collections
+on many  practical programs.  On  our benchmark programs, we  obtain a
+decrease of  up to  y\% in  the number  of collections,  and up  to x\%
+decrease in overall garbage collection time.
 \end{abstract}

### 2015-07-03 paper/paper.tex
inew graph

@@
 % --- Author Metadata here ---
-\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
 %\CopyrightYear{2007} % Allows default copyright year (20XX) to be
@@
 
-% A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-\category{D.2.8}{Software  Engineering}{Metrics}[complexity  measures,
-  performance measures]
-
-\terms{Theory}
-
-\keywords{ACM proceedings, \LaTeX, text tagging}
-
-By Friday, February 27 2015, 23:59 (UTC-11), submit a full paper of at
-most  12 pages (6  pages for  an Experience  Report), in  standard ACM
-conference format, including bibliography, figures, and appendices.
+\category{D.3.4}{Programming Languages}{Processors}[Memory Management
+  (Garbage Collection), Optimizations]
+\category{F.3.2}{Logic and Meanings Of Programs}{Semantics of
+  Programming Languages}[Program Analysis]
+
+\terms{Algorithms, Languages, Theory}
+
+\keywords{Heap Analysis, Memory Management, Garbage Collection}
 
@@
 on Haskell~\cite{rojemo96lag}
-%and Java~\cite{shaham02estimating}
 programs have shown that  GCs leave uncollected a large
@@
 \begin{figure}[t!]
-  %%%%{\color{Myblue}
   \psset{unit=1mm}
-  \begin{pspicture}(80,50)(0,-50)
-    %\psframe(80,50)(0,-50)
-    %  \begin{center}
+  \begin{pspicture}(90,48)(0,-48)
+    %\psframe(90,48)(0,-48)
     \begin{tabular}{@{}c@{}}
-      \begin{boxedminipage}{.42\textwidth}
-        {\sf
+      \psframebox{\sf
 	\renewcommand{\arraystretch}{1}{
@@
             $\leftarrow$\ (\CONS\ $5$
-              %          \UNL{1}\hspace*{1.5cm}
             (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
@@
       }}
-      \end{boxedminipage}
       \\ 
       (a) Example program. \\  \\
-      \begin{minipage}{.45\textwidth}
-        %\hspace*{1cm}
-                {\includegraphics[width=1\textwidth]{motiv-example}}
-      \end{minipage}
-      \\ \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
+%      \includegraphics[width=.45\textwidth]{motiv-example}
+      \input{mem-graph} \\ 
+      \renewcommand{\arraystretch}{.9}
+      \begin{tabular}[t]{l}
         (b) Memory graph at $\pi$. Thick edges denote live
@@
         marked $\times$ during garbage collection.
-      \end{tabular}}
+      \end{tabular}
     \end{tabular} 

### 2015-07-03 paper/paper.tex
fixed frames.
fixed minor spacing issues.

@@
     \begin{tabular}{@{}c@{}}
-      \psframebox{\sf
+      %\psframebox
+          {\sf
 	\renewcommand{\arraystretch}{1}{
@@
 \begin{eqnarray*}
-   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
+   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\, e_\mainpgm
     \hspace{5em} \mbox{\em --- program}\\
-    \mathit{df} \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
+    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
 \,\,x_n)\,\,
@@
     \hspace{0.2em} \ \ \ \ \mbox{\em --- function definition} \\
-e \in \mathit{Expr} & ::= &
+e \in \mathit{Expr} & \!\!\!::=\!\!\! &
 \left\{\begin{array}{@{}ll@{\hspace{2em}}l}
@@
     \end{array}\right. \\
-s \in \mathit{Application} & ::= &
+s \in \mathit{App} & \!\!\!::=\!\!\!  &
 \left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
@@
        (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
-\\
        (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}

### 2015-07-03 paper/paper.tex
use xelatex is possible

@@
 \[
-\begin{array}{rlcl@{\hspace{0.5em}}l}
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
-\mbox{-- Environment} \\
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times
-  Data}& \mbox{-- Values}\\
-c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{--
-  Closures}\\
-d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \&
-  Closures} \\
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{--
-Heap}
+\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\
+v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
+c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
+d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
 \end{array}

### 2015-07-04 paper/paper.tex
correctness of liveness analysis

--- 2015-07-03 paper/paper.tex
@@
 % --- Author Metadata here ---
-%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
 %\CopyrightYear{2007} % Allows default copyright year (20XX) to be
@@
        \affaddr{Mumbai 400076, India}
-       \email{prasannak@cse.iitb.ac.in}
+       \email{prasanna@cse.iitb.ac.in}
 % 2nd. author
@@
 first-order functional programs.  To  do this, we analyze programs for
-{\em liveness} of heap allocated data, i.e.  data that can possibly be
-used in the future.  The accompanying garbage collector is modified to
-use the  result of the analysis  to preserve only {\em  live} data---a
-subset  of {\em  reachable  data}---during  garbage collection.   This
-results  in an  increase  in the  garbage  reclaimed and  consequently
-reduces the memory  footprints of programs.  While  this technique has
-already been shown to yield  benefits for eager first-order languages,
-the presence of  closures at run-time poses  additional challenges for
-lazy languages.  Closures require changes both in the earlier liveness
+liveness of heap  allocated data, i.e. data that  can possibly be used
+in the future.  The accompanying garbage collector is  modified to use
+the result of the analysis to preserve only {\em live} data---a subset
+of {\em reachable data}---during  garbage collection.  This results in
+an  increase in  the garbage  reclaimed and  consequently  reduces the
+memory footprints of programs.   While this technique has already been
+shown to yield benefits  for eager first-order languages, the presence
+of  closures   at  run-time  poses  additional   challenges  for  lazy
+languages.   Closures require  changes  both in  the earlier  liveness
 analysis for heap allocated memory  and also in the garbage collection
@@
 
-Our  analysis  is a  context-sensitive  analysis  that annotates  each
-potential  garbage collection  point  in  the program  with  a set  of
-deterministic finite-state  automata (DFA) describing the  liveness at
-that  point.   We  also  present a  practical  implementation  of  our
-analysis in  a garbage collector.   The collector, when  invoked, uses
-the DFA set to preserve live  objects.  As a result, fewer objects are
-marked and preserved when compared  to a reachability based marking.
-Our experiments  confirm that liveness-based collection  increases the
-garbage reclaimed,  consequently decreasing the number  of collections
-on many  practical programs.  On  our benchmark programs, we  obtain a
-decrease of  up to  y\% in  the number  of collections,  and up  to x\%
-decrease in overall garbage collection time.
+We  formulate  a context-sensitive  liveness  analysis  and using  the
+results annotate each potential allocation point in the program with a
+set     of    deterministic    finite-state     automata(DFA).     The
+garbage-collector  inspects  the DFA  to  curtail reachability  during
+marking.  As a result, fewer  objects are marked (with a possibly more
+expensive  marker)  and  preserved  by the  copy  phase.   Experiments
+confirm  that  liveness-based  collection  results in  a  increase  in
+garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
+collections. On  our benchmark programs, we obtain  a decrease between
+x\% and y\% in the memory size required to run programs, when compared
+to reachability-based garbage collection.
 \end{abstract}
 
-\category{D.3.4}{Programming Languages}{Processors}[Memory Management
-  (Garbage Collection), Optimizations]
-\category{F.3.2}{Logic and Meanings Of Programs}{Semantics of
-  Programming Languages}[Program Analysis]
-
-\terms{Algorithms, Languages, Theory}
-
-\keywords{Heap Analysis, Memory Management, Garbage Collection}
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+\category{D.2.8}{Software  Engineering}{Metrics}[complexity  measures,
+  performance measures]
+
+\terms{Theory}
+
+\keywords{ACM proceedings, \LaTeX, text tagging}
+
+By Friday, February 27 2015, 23:59 (UTC-11), submit a full paper of at
+most  12 pages (6  pages for  an Experience  Report), in  standard ACM
+conference format, including bibliography, figures, and appendices.
 
@@
 on Haskell~\cite{rojemo96lag}
+%and Java~\cite{shaham02estimating}
 programs have shown that  GCs leave uncollected a large
@@
 \begin{figure}[t!]
+  %%%%{\color{Myblue}
   \psset{unit=1mm}
-  \begin{pspicture}(90,48)(0,-48)
-    %\psframe(90,48)(0,-48)
+  \begin{pspicture}(80,50)(0,-50)
+    %\psframe(80,50)(0,-50)
+    %  \begin{center}
     \begin{tabular}{@{}c@{}}
-      \psframebox{\sf
+      \begin{boxedminipage}{.42\textwidth}
+        {\sf
 	  \renewcommand{\arraystretch}{1}{
@@
               $\leftarrow$\ (\CONS\ $5$
+              %          \UNL{1}\hspace*{1.5cm}
               (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
@@
         }}
+      \end{boxedminipage}
       \\
       (a) Example program. \\ \\
-%      \includegraphics[width=.45\textwidth]{motiv-example}
-      \input{mem-graph} \\ 
-      \renewcommand{\arraystretch}{.9}
-      \begin{tabular}[t]{l}
+      \begin{minipage}{.45\textwidth}
+        %\hspace*{1cm}
+                {\includegraphics[width=1\textwidth]{motiv-example}}
+      \end{minipage}
+      \\ \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
           (b) Memory graph at $\pi$. Thick edges denote live
@@
           marked $\times$ during garbage collection.
-      \end{tabular}
+      \end{tabular}}
     \end{tabular}
@@
 \rightsquigarrow
-{\color {red}\rho\oplus\rho'}, (\rho, x, (\CAR~x))\!:\!S, H, s$      &
+\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
 \sc{car-clo}
@@
 \rightsquigarrow
-{\color {red}\rho\oplus\rho'}, (\rho, x, (+~x~y))\!:\!S, H, s$      &
+\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
 \sc{prim-1-clo} \\
@@
 \rightsquigarrow
-{\color {red}\rho\oplus\rho'}, (\rho, x, (+~x~y))\!:\!S, H, s$      &
+\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
 \sc{prim-2-clo} \\
@@
 $\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
-    \lfloor\rho\rfloor_{FV(s){\color {red}\setminus\{x\}}}  \oplus [x \mapsto
+    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
   \ell]\rangle], e$} &
@@
   (\SIF~x~e_1~e_2)   \rightsquigarrow
-{\color {red} \rho \oplus \rho'}, (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
+\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
 &
@@
   \rightsquigarrow$
-${\color {red}\rho\oplus\rho'},~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
+$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
 \sc{return-clo} \\
@@
 \end{tabular}
-\caption{A small-step semantics for the language. {\color {red} Can we
-    replace  $\rho   \oplus  \rho'$  by  $\rho'$?  Can   we  drop  the
-    $\setminus\{x\}$         in          the         {\sc         let}
-    rule?}\label{fig:lang-semantics}}
+\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
 \end{center}
@@
   new e-path terminating with $\psi$}\\
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
+\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
 \begin{array}{l l}
@@
 \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
-        \emptyset_{\cal{M}?}  & \mbox{otherwise}
+        \emptyset_{\cal{M}}  & \mbox{otherwise}
                  \end{array} \\
@@
 \mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \lambda\;\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
+        \lambda\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
 \mbox{where}~\mathcal{M} =
-\mathcal{L}(e,\sigma,\Lfonly),~\Lv \in
+\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
 \mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
@@
 
-{\color{red} Proof of correctness of the liveness analysis...}
+\subsection{Correctness of liveness analysis}
+
+We shall now  outline a proof of correctness  of the liveness analysis
+described earlier.   This requires  a significant modification  to the
+technique  used  in  \cite{asati14lgc}.   We shall  first  modify  the
+standard semantics. This semantics, which we shall call {\em minefield
+  semantics}, will  model a liveness based garbage  collection at each
+transition. This is done by starting from the root-set and inserting a
+special value  $\bot$ at each reference  location in the  heap that is
+reachable but not live by  our analysis.  Attempts to dereference such
+locations  during the  transition will  result in  entering  a special
+state denoted  \bang.  We shall then  show that no  program enters the
+\bang\ state.
+
+To set up the minefield semantics, we follow these steps:
+\begin{enumerate}
+\item Given an  demand transformer \Lfonly\  we enrich  the abstract machine
+  state  $\rho, S,  H, e$  to $\rho,  S, H,  e, \sigma,  \Sigma$. Here
+  $\sigma$ is  the demand on  the currently active function,  thus the
+  liveness $\Lv$   at $e$ is $\mathcal{L}(e_f, \sigma, \Lfonly)$, and
+  $\Sigma$ is a stack of  demands---one for each function frame pushed
+  in $S$.   Second, we  augment $Val$ with  a value $\bot$.   To model
+  strong  liveness  $\bot$ may  be  copied  freely,  including into  a
+  \CONS\ cell, but explodes  when used computationally (in a primitive
+  operation other  than a copy). Additionally we  define $GC (\Lv,
+  \Sigma) : (\rho, H, S) \rightarrow (\rho , H , S )$ which determines
+  live-reachability using $\rho$  and the $\rho$'s in $S$  as the root
+  set  and  following  links  in   $H$  only  as  far  as  allowed  by
+  \Lv\  and  $\Sigma$.   Hence  $GC  (\Lv,  \Sigma)$  replaces
+  live-unreachable  values---in  $\rho$,  in  $H$ and  the  $\rho$'s  in
+  $S$---with  $\bot$.  For  example,  if $x.\epsilon  \in \Lv$  then
+  $\rho (x) = \bot$.  Only $GC (\cdots )$ introduces $\bot$. 
+\end{enumerate}
 
@@
 
-\subsection{Modifying the liveness rules.}
+\subsection{Modifying liveness rules}
 
@@
 \section{Results}
-%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
-%% we  have  built a  prototype  consisting  of  an interpreter  for  our
-%% language,  a  liveness  analyzer  and  a copying  collector  that  can
-%% optionally use the results of liveness analysis for marking instead of
-%% reachability. When the collector uses liveness for marking, we call it
-%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-%%   reachability-based  collector}  (RGC).   The  copying  collector  is
-%% neither incremental nor generational.  As a consequence, any cell that
-%% becomes unreachable or  dead is assuredly collected in  the next round
-%% of garbage collection.}
-
-To test the utility of our method we have implemented a garbage-collector 
-for a simple scheme-like first order language. It consists of an interpreter, 
-liveness analyzer and a garbage-collector that can optionally use use reachability or liveness. 
-Our benchmark consists of programs from nofib suite and some external programs manually converted to ANF.
-
-%% \comment{
-%% While exploring the activation record of a function, LGC uses the DFAs
-%% at  the latest  program  point traversed  in the function's body.
-%% Thus, if  LGC
-%% is invoked  at the program point   $\pi$,  it would  use  the
-%% liveness  DFAs at  $\pi$ itself for exploring the root set of the
-%% current activation record. For any other activation record in the
-%% stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
-%% call-chain,
-%% the DFAs at the point of call to  ${\mathit g}$ would be used.}
-
-The process of liveness-based garbage collection involves going through 
-each activation record on the stack and exploring each variable in the 
-current activation record (root set). The liveness of each variable is 
-determined using the program point and the variable name. The liveness 
-of all variables are stored as DFAs. These DFAs are then encoded as a 
-table. A variable is live only if an entry is found in this table.
-All cells that are live are copied to the live semi-space. In case of 
-\CONS\ cells the \CAR\ and \CDR\ pointers are chased and if they are 
-live the copied \CONS\ cell will get the updated addresses of its 
-\CAR\ and \CDR\ fields.
-
-%% \comment{
-%% Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-%% point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-%% $\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-%% $\dfa{\pi}{x}$.
-%% Considering   a   DFA   as   a   table,
-%% $\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-%% state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-%% \acar\      or     \acdr.       We      shall     also      write
-%% $\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-%% indicating  whether there  is a  transition from  $q$ on
-%% $\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-%% $\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-%% empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-%% $\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-%% in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-%% pointer to $x$.
-%% The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-%% $x$  in the  other semi-space  and returns  the new  address.  It
-%% hides  details such as  returning the  forwarding pointer  if the
-%% value  of $x$  is  already copied,  and  creating the  forwarding
-%% pointer otherwise.}
-
-\subsection{Copying closure using liveness vs Copying closures using reachability}
-To compare the effect of using reachability to copy closures instead of liveness we take a couple of programs from our benchmark suite and execute them with 
-both. As shown in figure-add-figure-here, the number of garbage collections 
-in both cases are same but the time taken varies largely. Both these observations can be explained as follows, since we approximate the liveness of closure to 
-the liveness at its creation point, the liveness we get is usually grossly over approximated. 
-Using this liveness may not give too much of a benefit compared to reachability as nearly everything would be live at its point of creation. 
-As mentioned in section-ref, using liveness for closures may entail multiple traversals over the same structure consuming more time. 
-Therefore we believe that using reachability for closures is a sweet spot in terms of performance for liveness based garbage collectors. 
-%%
-%% \comment{
-%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-%% in the heap over time for RGC and LGC---here time is measured in terms
-%% of the number of cons cells allocated. In addition, they also show the
-%% number of  reachable cells and the  number of cells  that are actually
-%% live\footnote{This   is  statically   approximated  by   our  liveness
-%%   analysis}.  Since  the programs have  different memory requirements,
-%% we have tuned the size of heap for each program to ensure a reasonable
-%% number of  collections. An invocation  of RGC decreases the  number of
-%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
-%% invocation of LGC decreases the number  of heap cells to no lower than
-%% the curve of live cells.
-
-%% To construct the  reachable and live curves, we  record for every cell
-%% its creation time (\CreateTime), its last use time (\UseTime), and the
-%% earliest time  when the  cell becomes unreachable  and can  be garbage
-%% collected (\CollTime).  For accurate recording of  \CollTime, we force
-%% frequent invocations  of a reachability based collector  in a separate
-%% run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-%% \mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-%%   \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-%%   \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-%% The benchmark programs are drawn from the {\tt no-fib} suite and other
-%% sources and have been  manually converted to  ANF.  All  graphs  except
-%%  {\tt  fibheap}  show  fewer  garbage
-%% collector  invocations for  LGC. The  number  of reachable  cells in
-%% {\tt
-%%   fibheap} increases  initially, comes  down before the  heap becomes
-%% gets
-%% full and then  continues to remains low.  This results in  a single
-%% garbage
-%% collection in both RGC and LGC.  The graphs also show the precision of
-%% our liveness analysis.  For  all programs  except  {\tt nperms} and
-%% {\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-%% manages to collect a good portion of the data that are not live.}
-%%
+To demonstrate the effectiveness of liveness-based garbage collection,
+we  have  built a  prototype  consisting  of  an interpreter  for  our
+language,  a  liveness  analyzer  and  a copying  collector  that  can
+optionally use the results of liveness analysis for marking instead of
+reachability. When the collector uses liveness for marking, we call it
+a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
+  reachability-based  collector}  (RGC).   The  copying  collector  is
+neither incremental nor generational.  As a consequence, any cell that
+becomes unreachable or  dead is assuredly collected in  the next round
+of garbage collection.
+
+While exploring the activation record of a function, LGC uses the DFAs
+at  the latest  program  point traversed  in the function's body.
+Thus, if  LGC
+is invoked  at the program point   $\pi$,  it would  use  the
+liveness  DFAs at  $\pi$ itself for exploring the root set of the
+current activation record. For any other activation record in the
+stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
+call-chain,
+the DFAs at the point of call to  ${\mathit g}$ would be used.
+
+Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
+point  pair ($x$,  $\pi$).  We  assume that  there is  a function
+$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
+$\dfa{\pi}{x}$.
+Considering   a   DFA   as   a   table,
+$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
+state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
+\acar\      or     \acdr.       We      shall     also      write
+$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
+indicating  whether there  is a  transition from  $q$ on
+$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
+$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
+empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
+$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
+in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
+pointer to $x$.
+The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
+$x$  in the  other semi-space  and returns  the new  address.  It
+hides  details such as  returning the  forwarding pointer  if the
+value  of $x$  is  already copied,  and  creating the  forwarding
+pointer otherwise.
+
+The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
+in the heap over time for RGC and LGC---here time is measured in terms
+of the number of cons cells allocated. In addition, they also show the
+number of  reachable cells and the  number of cells  that are actually
+live\footnote{This   is  statically   approximated  by   our  liveness
+  analysis}.  Since  the programs have  different memory requirements,
+we have tuned the size of heap for each program to ensure a reasonable
+number of  collections. An invocation  of RGC decreases the  number of
+cells  in heap  till  it touches  the  curve of  reachable cells.   An
+invocation of LGC decreases the number  of heap cells to no lower than
+the curve of live cells.
+
+To construct the  reachable and live curves, we  record for every cell
+its creation time (\CreateTime), its last use time (\UseTime), and the
+earliest time  when the  cell becomes unreachable  and can  be garbage
+collected (\CollTime).  For accurate recording of  \CollTime, we force
+frequent invocations  of a reachability based collector  in a separate
+run.     A    cell   is    live    at    time   \mbox{\exectime}    if
+\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
+  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
+  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
+
+The benchmark programs are drawn from the {\tt no-fib} suite and other
+sources and have been  manually converted to  ANF.  All  graphs  except
+ {\tt  fibheap}  show  fewer  garbage
+collector  invocations for  LGC. The  number  of reachable  cells in
+{\tt
+  fibheap} increases  initially, comes  down before the  heap becomes
+gets
+full and then  continues to remains low.  This results in  a single
+garbage
+collection in both RGC and LGC.  The graphs also show the precision of
+our liveness analysis.  For  all programs  except  {\tt nperms} and
+{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
+manages to collect a good portion of the data that are not live.
 
@@
 %% \end{figure}
-The table in Figure~\ref{fig:experimental-results} shows that LGC improves the 
-garbage collection time and efficiency for most of the programs. 
-Even for programs where the time taken is more it is very competitive. 
-As the number of garbage collections tend to be very large our graphs show only 
-a window which is representative of the behavior for that particular benchmark. 
-In all the programs we can see that the curve corresponding to liveness regularly 
-dips below the reachability curve.
-
-One area of concern is the huge gap between the actual liveness and the 
-liveness percieved by our collector. In case of LGC for eager languages the 
-gap was very narrow and almost touched the actual liveness curve. 
-In case of lazy languages, due to using reachability for copying we end up copying 
-more cells. Any data which is part of closure irrespective of whether it is live 
-or not gets copied. Thus many non-live cells get copied during garbage collection 
-increasing the drag time.
-Our experiments have shown that unless we use the exact execution point for 
-closures bringing down the number of non-live cells copied in closures cannot be reduced.
-Using the approximated liveness (liveness at the creation point) does not give benefits in terms of non-live
- cells inside closures as everything would be live at the point of creation.
-%% \comment{
-%% The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-%% tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-%% provides  statistics  regarding the  analysis  itself.  The number  of
-%% states and the analysis times  are within tolerable limits. Precision
-%% of analysis refers to the percentage of dead cells that is collected by
-%% LGC, averaged over all invocations.
-%% The second table shows garbage  collection statistics for RGC and LGC.
-%% LGC  collects larger garbage  per invocation,  drags cells  for lesser
-%% time and requires a smaller heap size ({\em MinHeap}) for program to
-%% run
-%% in comparison with  RGC.
-
-%% There are a couple of  issues of concern.  The garbage collection time
-%% is larger in the case of LGC for some programs. The reason is that the
-%% cost of consulting the liveness DFA may outweigh the combined benefits
-%% of  fewer   garbage  collections   and  fewer  markings   per  garbage
-%% collection.   The  other issue  is  illustrated  by  the program  {\tt
-%%   lambda}    (memory   usage  shown   separately    in
-%% Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-%% in Figure~\ref{fig:experimental-results}, the  number of
-%% touched cells\footnote{These  are the  cells which are  visited during
-%%   marking phase, often  more than once due to  sharing.} in this
-%% example is much higher for
-%% LGC.  This  increase is due to  excessive sharing among  heap nodes in
-%% this program.  Note  that a node re-visited because  of sharing is not
-%% explored  any   further  during  a  RGC   collection.   However,  this
-%% curtailment cannot happen in  LGC because of the possibility that the
-%% node,  re-visited in a  different liveness  state, may  mark a  set of
-%% cells different from the earlier visit.
-%% }
+The  increased effectiveness  of LGC  over RGC  is also  shown  in the
+tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
+provides  statistics  regarding the  analysis  itself.  The number  of
+states and the analysis times  are within tolerable limits. Precision
+of analysis refers to the percentage of dead cells that is collected by
+LGC, averaged over all invocations.
+The second table shows garbage  collection statistics for RGC and LGC.
+LGC  collects larger garbage  per invocation,  drags cells  for lesser
+time and requires a smaller heap size ({\em MinHeap}) for program to
+run
+in comparison with  RGC.
+
+There are a couple of  issues of concern.  The garbage collection time
+is larger in the case of LGC for some programs. The reason is that the
+cost of consulting the liveness DFA may outweigh the combined benefits
+of  fewer   garbage  collections   and  fewer  markings   per  garbage
+collection.   The  other issue  is  illustrated  by  the program  {\tt
+  lambda}    (memory   usage  shown   separately    in
+Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
+in Figure~\ref{fig:experimental-results}, the  number of
+touched cells\footnote{These  are the  cells which are  visited during
+  marking phase, often  more than once due to  sharing.} in this
+example is much higher for
+LGC.  This  increase is due to  excessive sharing among  heap nodes in
+this program.  Note  that a node re-visited because  of sharing is not
+explored  any   further  during  a  RGC   collection.   However,  this
+curtailment cannot happen in  LGC because of the possibility that the
+node,  re-visited in a  different liveness  state, may  mark a  set of
+cells different from the earlier visit.
 \begin{figure*}[t]

### 2015-07-04 paper/paper.tex
undecidability

@@
 first-order functional  programs. To do this, we  analyze programs for
-{\em liveness} of heap allocated data, i.e.  data that can possibly be
-used in the future.  The accompanying garbage collector is modified to
-use the  result of the analysis  to preserve only {\em  live} data---a
-subset  of {\em  reachable  data}---during  garbage collection.   This
+liveness of  heap-allocated data.  The  accompanying garbage collector
+is modified  to use the result  of the analysis to  preserve only live
+data---a subset  of reachable data---during  garbage collection.  This
 results  in an  increase  in the  garbage  reclaimed and  consequently
-reduces the memory  footprints of programs.  While  this technique has
+reduces  the memory footprint  of programs.  While this  technique has
 already been shown to  yield benefits for eager first-order languages,
-the presence of  closures at run-time poses  additional challenges for
-lazy languages.  Closures require changes both in the earlier liveness
-analysis for heap allocated memory  and also in the garbage collection
-scheme required to reclaim memory that is not live.
-
-Our  analysis  is a  context-sensitive  analysis  that annotates  each
-potential  garbage collection  point  in  the program  with  a set  of
-deterministic finite-state  automata (DFA) describing the  liveness at
-that  point.   We  also  present a  practical  implementation  of  our
-analysis in  a garbage collector.   The collector, when  invoked, uses
-the DFA set to preserve live  objects.  As a result, fewer objects are
-marked and preserved when compared  to a reachability based marking.
-Our experiments  confirm that liveness-based collection  increases the
-garbage reclaimed,  consequently decreasing the number  of collections
-on many  practical programs.  On  our benchmark programs, we  obtain a
-decrease of  up to  y\% in  the number  of collections,  and up  to x\%
-decrease in overall garbage collection time.
+the lack of a statically determinable execution order and the presence
+of  closures   at  run-time  poses  additional   challenges  for  lazy
+languages.   This  requires  changes  both  in  the  earlier  liveness
+analysis and also in the garbage collection scheme required to reclaim
+memory that is not live.
+
+The implementation  of our  analysis annotates each  potential garbage
+collection  point   in  the  program  with  a   set  of  deterministic
+finite-state automata (DFA) describing  the liveness at that point. We
+also implement a copying collector  that uses the DFA to preserve live
+objects.   Our  experiments  confirm  that  liveness-based  collection
+always  increases the garbage  reclaimed, consequently  decreasing the
+number of collections.  In addition,  it reduces the execution time of
+several programs.  Experiments also reveal  that the best  results are
+obtained  for   a  mixed  garbage   collection  scheme  that   does  a
+liveness-based    collection   for    fully    evaluated   data    and
+reachability-based   collection  for   closures.   On   our  benchmark
+programs,  we  obtain a  decrease  of  up to  50\%  in  the number  of
+collections, and up to 70\% decrease in overall program execution time
+when compared to reachability based collectors.
 \end{abstract}
@@
 
+\begin{figure*}[t!]
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Premise & Transition & Rule name \\
+\hline
+\hline
+          &\makecell{ $\rho, (\rho', x, e)\!:\!S, H, \kappa$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$ }    &  \sc{const}
+\\
+\hline
+          & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CAR~x)$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$}      &
+\sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x)$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x))\!:\!S, H, s$}      &
+\sc{car-clo}
+\\
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
+\sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
+\sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
+\sc{prim-2-clo} \\
+\hline
+{$\mathit{f}~\mbox{defined as}$
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x})$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$}      &
+\sc{funcall} \\
+\hline
+$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+  \ell]\rangle], e$} &
+\sc{let} \\
+\hline
+$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
+\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\SIF~x~e_1~e_2), ep , \sigma$ \\ $\rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2), ep, \sigma)\!:\!S, H,  s, ep_{if}, \sigma$}
+&
+\sc{if-clo} \\
+\hline
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma)\!:\!S, H, (\SRETURN~x), ep_{return}, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, ep, \sigma$} &
+\sc{return-whnf}\\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\SRETURN~x), ep, \sigma$ \\ $
+  \rightsquigarrow$
+$\rho',~ (\rho, x, (\SRETURN~x), ep, \sigma)\!:\!S, H,  s, ep_{return}, \sigma$} &
+\sc{return-clo} \\
+\hline
+\end{tabular}
+\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
+\end{center}
+\end{figure*}
+
 %% Generated by bibtex from your ~.bib file.  Run latex,

### 2015-07-05 lazy-scheme/src/gc.cpp
Added package makecell

@@ -121,7 +121,18 @@ void gen_graphviz_code(Scheme::AST::cons* conscell, ostream& out);
 std::string getNodeLabel(Scheme::AST::cons* conscell, size_t idx);
 void depthfirstpaths(Scheme::AST::cons* loc, Scheme::AST::state_index index);
 
-
+bool isClosure(cons* cell)
+{
+	switch(cell->typecell)
+	{
+	case nilExprClosure:
+	case constIntExprClosure:
+	case constBoolExprClosure:
+	case consExprClosure:
+	case constStringExprClosure: return false;
+	default : return true;
+	}
+}
 
 
 #define FIND_REACHABLE
@@ -806,22 +817,26 @@ int is_end_of_non_ref_vector()
 }
 
 //!!!!!caution update it to check for already existing variable in same environment
-void make_reference_addr(const char *var, void* addr)
+void make_reference_addr(const char *var, void* addr, bool isParam)
 {
 	int found=0;
 	vector<var_heap>& vh=actRecStack.front().heapRefs;
 	vector<var_heap>::iterator it = vh.begin();
-	while(it != vh.end()) {
+	while(it != vh.end())
+	{
 
 		if (strcmp(it->varname.c_str(),var)==0)
 		{
 			it->ref=addr;
+			it->is_param = isParam;
 			found=1;
 			break;
 		}
 		++it;
 	}
-	if(found==0){
+
+	if(found==0)
+	{
 		var_heap newref;
 		newref.varname=var;
 		newref.ref=addr;
@@ -1565,6 +1580,50 @@ void liveness_gc()
     	  }
     	  else
     	  {
+    		  if (vhit->ref == NULL)
+    			  continue;
+
+    		  cerr << "Processing non-live root variable " << vhit->varname << endl;
+    		  for(auto ref : stackit->heapRefs)
+    		  {
+    			  auto cell = static_cast<cons*>(vhit->ref);
+    			  cerr << "Processing "<<ref.varname << " with type " << cell->typecell << endl;
+    			  if (ref.varname == vhit->varname || !isClosure(cell))
+    				  continue;
+
+    			  auto clo = cell->val.closure;
+    			  if (cell->typecell != funcApplicationExprClosure)
+    			  {
+    				  cerr<<"Checking arg1 " << *(clo.arg1_name) << endl;
+    				  if (clo.arg1 && *(clo.arg1_name) == vhit->varname)
+    				  {
[…]

### 2015-07-05 paper/paper.tex
Added package makecell
correctness of liveness analysis

@@
 \usepackage[boxed]{algorithm2e}
+\usepackage{makecell}
 
@@
 time is undecidable.
+\newcommand{\state}{\ensuremath{\mathsf{S}}}
+\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}
 
@@
 Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
-consisting
-of symbols \acar\ and \acdr\  only. The decision problem $\alpha \in
-\mathscr{L}(G)$ is undecidable.
+consisting  of symbols \acar\  and \acdr\  only. The  decision problem
+$\alpha \in \mathscr{L}(G)$ is undecidable.
 \end{lemma}
 
-\comment{Proof:}
+{\em Proof:}  Given a Turing machine  $TM$ and an input  $w$, we shall
+construct an unrestricted  grammar $G$ of the kind  described above, so
+that $TM$ will halt on $w$ if and only if $\epsilon \in \mathscr{L}(G)$.
+
+We  shall  represent  the  configuration  of  $TM$  at  any  point  as
+$w_L(\state,c)w_R$, where $w_L$ is the string on the left of the head,
+$w_R$ is the string on the right  of the head, $c$ is the symbol under
+the head and \state\ is the current state of the machine. For each
+$(\state,c)$ combination of state and symbol, the grammar
+will contain a nonterminal \nont{}{c}. Each step of the machine will
+correspond to step of derivation based on the grammar.  In addition, we shall maintain
+the following invariant between the configuration of $TM$ and the
+current sentential form:
+
+\begin{quote}
+  If the configuration of the machine is $w_L(\state,c)w_R$, then the
+  sentential form will be $\overline{w}_L\nont{}{c}\,w_R$, where
+  $\overline{w}_L$ is the same as $w_L$ but with each symbol $d$
+  replaced with $\overline{d}$.
+\end{quote}
+
+Assume that the TM starts in a state $S_{init}$ with a tape $cw$ and the
+head is positioned on the symbol $c$. Then the start non-terminal  of the
+grammar $\nont{}{}$ is associated with the production $\nont{}{} \rightarrow \nont{init}{c}w$ and the
+sentential form corresponding to the initial configuration is
+$\nont{init}{c}w$. Further each transition of the turing machine is modeled by
+productions as follows:
+
+\begin{enumerate}
+\item A transition $(S_i, c) \rightarrow (S_j,c',L)$ by two productions
+  $\nont{i}{c} \rightarrow \acar \nont{j}{\acar}c'$ and $\nont{i}{c}
+  \rightarrow \acdr \nont{j}{\acdr}c'$. 
+\item A transition $(S_i, c) \rightarrow (S_j,c',R)$ by two productions
+  $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
+  \rightarrow c \nont{j}{\acdr}\bcdr$. 
+\end{enumerate}
+The idea behind these productions is as follows: Assume that the
+current sentential form is
+\ldots\bcar\bcdr\nont{i}{\acar}\acar\acar\ldots and Starting from $(S_i,
+c)$ if the head writes $c'$ and moves to the left, then the symbol
+$c'$ is exposed to the right of the tape head. However, we do not know
+which symbol that is under the head after the movement. Therefore we
+generate two productions assuming that the symbols are \acar\ and
+\bcar. If the guess is right, then to remove the symbol to the left in
+the sentential form
 
@@
 \section{The Garbage Collection Scheme}
-% GC assumptions
-\cred{Introduction}
 
@@
 \hline
-          &\makecell{ $\rho, (\rho', x, e)\!:\!S, H, \kappa$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$ }    &  \sc{const}
+          &\makecell{ $\rho, (\rho', x, e, \{\psi_{i}\}, \sigma')\!:\!S, H, \kappa, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e, \sigma'$ }   &  \sc{const}
 \\
 \hline
-          & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CAR~x)$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$}      &
+          & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (\CONS~x~y), \sigma$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma'$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma' )\!:\!S, H, (\CAR~x), \sigma$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma'$}      &
 \sc{car-whnf} \\
 \hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x)$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x))\!:\!S, H, s$}      &
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x), \{\psi_{i}\}, \sigma)\!:\!S, H, s,\bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x} $}      &
 \sc{car-clo}
@@
 $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
+ & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (+~x~y), \sigma$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma'$}      &
 \sc{prim-whnf} \\
 \hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x}$}      &
 \sc{prim-1-clo} \\
 \hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, y, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{y} $}      &
 \sc{prim-2-clo} \\
@@
 {$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x})$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$}      &
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x}), \sigma$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}, \sigma$}      &
 \sc{funcall} \\
@@
 \hline
-$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
+$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
 \rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
 \hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\SIF~x~e_1~e_2), ep , \sigma$ \\ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2), ep, \sigma)\!:\!S, H,  s, ep_{if}, \sigma$}
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\pi:\SIF~\psi:x~e_1~e_2), \sigma$ \\ $\rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2),  \pmap{}(\pi), \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$}
 &
@@
 \hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma)\!:\!S, H, (\SRETURN~x), ep_{return}, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, ep, \sigma$} &
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma')\!:\!S, H, (\SRETURN~x), \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, \sigma'$} &
 \sc{return-whnf}\\
 \hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\SRETURN~x), ep, \sigma$ \\ $
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\psi:\SRETURN~x), \sigma$ \\ $
   \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x), ep, \sigma)\!:\!S, H,  s, ep_{return}, \sigma$} &
+$\rho',~ (\rho, x, (\SRETURN~x), \{\psi\}, \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$} &
 \sc{return-clo} \\

### 2015-07-05 paper/paper.tex
correctness of liveness analysis

@@
 \usepackage{amsmath}
+%\usepackage{amsthm}
 \usepackage{pstricks}
@@
 \end{lemma} 
-
-{\em Proof:}  Given a Turing machine  $TM$ and an input  $w$, we shall
-construct an unrestricted  grammar $G$ of the kind  described above, so
-that $TM$ will halt on $w$ if and only if $\epsilon \in \mathscr{L}(G)$.
-
-We  shall  represent  the  configuration  of  $TM$  at  any  point  as
-$w_L(\state,c)w_R$, where $w_L$ is the string on the left of the head,
-$w_R$ is the string on the right  of the head, $c$ is the symbol under
-the head and \state\ is the current state of the machine. For each
-$(\state,c)$ combination of state and symbol, the grammar
-will contain a nonterminal \nont{}{c}. Each step of the machine will
-correspond to step of derivation based on the grammar.  In addition, we shall maintain
-the following invariant between the configuration of $TM$ and the
-current sentential form:
+\begin{proof}
+Given a  Turing machine  and an input  $w\in (1+0)^*$, we  construct a
+grammar $G$  such that  the machine will  halt on  $w$ if and  only if
+$\epsilon \in  \mathscr{L}(G)$. The grammar includes the  fixed set of
+unrestricted productions  in Definition~\ref{def:specialgrammar}.
+
+We  shall   denote  the  configuration   of  the  Turing   machine  as
+$w_l(\state,c)w_r$, where $w_l$ is the string to the left of the head,
+$w_r$ is the string to the right, $c$ is the symbol under the head and
+\state\ is the current state  of the machine.  For each combination of
+state  and   symbol  $(\state,c)$,   the  grammar  will   contain  the
+non-terminal \nont{}{c}. We shall synchronize each move of the machine
+to a  derivation step  using a context  free production,  followed, if
+possible, by  a derivation step  using either $\bcar\acar    \rightarrow
+\epsilon$ or $\bcdr\acdr    \rightarrow
+\epsilon$.
+After each synchronization, we shall establish the following invariant
+relation between the machine configuration and the  sentential form:
 
 \begin{quote}
-  If the configuration of the machine is $w_L(\state,c)w_R$, then the
-  sentential form will be $\overline{w}_L\nont{}{c}\,w_R$, where
-  $\overline{w}_L$ is the same as $w_L$ but with each symbol $d$
-  replaced with $\overline{d}$.
+  If the configuration of the machine is $w_l(\state,c)w_r$, then the
+  sentential form will be $\overline{w}_l\nont{}{c}\,w_r$, where
+  $\overline{w}_l$ is the same as $w_l$ but with each symbol $d$ in $w_l$
+  replaced by $\overline{d}$.
 \end{quote}
 
-Assume that the TM starts in a state $S_{init}$ with a tape $cw$ and the
-head is positioned on the symbol $c$. Then the start non-terminal  of the
-grammar $\nont{}{}$ is associated with the production $\nont{}{} \rightarrow \nont{init}{c}w$ and the
+Assume that  the TM  starts in a  state $S_\mathit{init}$ with  a tape
+$cw$  and the  head   positioned on  the symbol  $c$. Then  the the
 sentential  form   corresponding  to  the   initial  configuration  is
-$\nont{init}{c}w$. Further each transition of the turing machine is modeled by
-productions as follows:
+$\nont{\mathit{init}}{c}w$ (we can assume  that there is a production
+$\nont{}{}  \rightarrow \nont{\mathit{init}}{c}w$, where  \nont{}{} is
+the start symbol of  the grammar). Further correspondences between the
+Turing machine moves and the grammar productions are as follows:
 
 \begin{enumerate}
-\item A transition $(S_i, c) \rightarrow (S_j,c',L)$ by two productions
-  $\nont{i}{c} \rightarrow \acar \nont{j}{\acar}c'$ and $\nont{i}{c}
-  \rightarrow \acdr \nont{j}{\acdr}c'$. 
-\item A transition $(S_i, c) \rightarrow (S_j,c',R)$ by two productions
+\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
+  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
+  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
+\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
+  are two productions
   $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
@@
 \end{enumerate}
-The idea behind these productions is as follows: Assume that the
-current sentential form is
-\ldots\bcar\bcdr\nont{i}{\acar}\acar\acar\ldots and Starting from $(S_i,
-c)$ if the head writes $c'$ and moves to the left, then the symbol
-$c'$ is exposed to the right of the tape head. However, we do not know
-which symbol that is under the head after the movement. Therefore we
-generate two productions assuming that the symbols are \acar\ and
-\bcar. If the guess is right, then to remove the symbol to the left in
-the sentential form
-
-We circumvent  the problem by over-approximating the  CFG  by
-non-deterministic finite state automata (NFA). The NFA are then
+The idea behind  the productions is explained with  an example: Assume
+that        the        current        sentential        form        is
+\bcar\bcdr\nont{i}{\acar}\acar\acar.  Assume that  the  machine has  a
+transition  $(S_i,\acar) \rightarrow (S_j,\acdr,  L)$. Since  the next
+corresponding step in the derivation  has to be done without any prior
+knowledge of whether the symbol to the left of the tape is a \acar\ or
+a  \acdr, two  productions are  provided,  and the  invariant will  be
+maintained  only   if  the  production   $\nont{i}{\acar}  \rightarrow
+\acdr\nont{j}{\acdr}\acdr$  is  chosen  for   the  next  step  in  the
+derivation.         This         gives        the        configuration
+\bcar\bcdr\acdr\nont{i}{\acdr}\acdr\acar\acar.    Simplification  with
+the    production    $\bcdr\acdr    \rightarrow    \epsilon$    yields
+\bcar\nont{i}{\acdr}\acdr\acar\acar, which  exactly corresponds to the
+changed configuration  of the machine.  Notice carefully that  a wrong
+choice breaks the invariant and it cannot be recovered subsequently by
+any choice of productions.
+
+After  the Turing machine  has halted,  there are  further ``cleanup''
+derivations  that derive  $\epsilon$ only  if the  invariant  has been
+maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
+$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
+final state of the Turing machine. We add
+productions $\nont{\mathit{final}}{c} \rightarrow \acar
+\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
+\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
+the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
+\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
+\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
+right of the tape head. This completes the reduction.  
+\end{proof}
+
+We circumvent the problem  of undecidability by over- approximating the
+CFG by non-deterministic finite state automata (NFA). The NFA are then
 simplified using  the $\hookrightarrow$ rules.  Finally the simplified

### 2015-07-06 paper/paper.tex
correctness of liveness analysis
Fixed a minor in liveness GC for closures

@@
 \newcommand{\cred}[1]{{\color{red}{#1}}}
+\newcommand{\cblue}[1]{{\color{blue}{#1}}}
 \newcommand{\blankout}[1]{}
@@
 \newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}
-
+\cblue{
 \begin{lemma}
@@
 Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
-$cw$  and the  head   positioned on  the symbol  $c$. Then  the the
-sentential  form   corresponding  to  the   initial  configuration  is
+$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
+form    corresponding     to    the    initial     configuration    is
 $\nont{\mathit{init}}{c}w$ (we  can assume that there  is a production
@@
 \end{proof}
-
+}%end of cblue
 We circumvent the problem  of undecidability by over- approximating the

### 2015-07-06 lazy-scheme/post_process/gen_stats_and_plotall.sh
some updates to get better graphs.

@@ -76,19 +76,35 @@ freq["lambda"]=100;
 
 declare -A xtics
 declare -A ytics
-xtics["sudoku"]=2050; ytics["sudoku"]=250
-xtics["fft"]=2050; ytics["fft"]=250
-xtics["lcss"]=100000; ytics["lcss"]=9000
-xtics["small"]=1; ytics["small"]=1
-xtics["test"]=25; ytics["test"]=25
-xtics["testsmall"]=25; ytics["testsmall"]=25
-xtics["gc_bench"]=400000; ytics["gc_bench"]=20000
-xtics["nperm"]=145000; ytics["nperm"]=35000
-xtics["fibheap"]=70000; ytics["fibheap"]=30000
-xtics["knightstour"]=850000; ytics["knightstour"]=40000
-xtics["treejoin"]=130000; ytics["treejoin"]=110000
-xtics["nqueens"]=6000000; ytics["nqueens"]=105000
-xtics["lambda"]=90000; ytics["lambda"]=10000
+xtics["small"]=1		; ytics["small"]=1
+xtics["test"]=25		; ytics["test"]=25
+xtics["testsmall"]=25		; ytics["testsmall"]=25
+xtics["fft"]=2050		; ytics["fft"]=250
+xtics["sudoku"]=0.25e5		; ytics["sudoku"]=0.6e3
+xtics["lcss"]=.9e5		; ytics["lcss"]=4e3
+xtics["gc_bench"]=1.9e5   	; ytics["gc_bench"]=70e3
+xtics["nperm"]=8.5e5		; ytics["nperm"]=3.5e3
+xtics["fibheap"]=8e5		; ytics["fibheap"]=10e3
+xtics["knightstour"]=8.5e5	; ytics["knightstour"]=40e3
+xtics["treejoin"]=13e5  	; ytics["treejoin"]=330e3
+xtics["nqueens"]=3.5e5    	; ytics["nqueens"]=4.6e3
+xtics["lambda"]=19e5		; ytics["lambda"]=3e3
+
+declare -A xlimit
+declare -A ylimit
+xlimit["fft"]=2050		; ylimit["fft"]=250
+xlimit["lcss"]=1e5:1.7e5	; ylimit["lcss"]=0:24e3
+xlimit["small"]=1		; ylimit["small"]=1
+xlimit["test"]=25		; ylimit["test"]=25
[…]

### 2015-07-06 paper/paper.tex
Added graphs and updated results table

@@
 
-%% \begin{figure}[p]
-%% \begin{tabular}{@{}c@{}c@{}}
-%% \hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
-%% &
-%% {\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
-%% \\
-%% \hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
-%%  &
-%% {\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
-%% \\
-%% \hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
-%% &
-%% {\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
-%% \\
-%% \hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
-%% &
-%% {\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
-%% \end{tabular}
-%%  \caption{Memory  usage of  programs. The blue and the red curves
-%%   indicate the number of cons cells in the active semi-space for RGC
-%% and
-%%   LGC respectively. The black curve represents the number of reachable
-%%   cells and the grey curve represents the number of cells that are
-%%   actually live (of which liveness analysis does  a static
-%%   approximation). x-axis is the time measured in number of cons-cells
-%%   allocated. y-axis is the number of cons-cells.}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-The table in Figure~\ref{fig:experimental-results} shows that LGC improves the 
-garbage collection time and efficiency for most of the programs. 
-Even for programs where the time taken is more it is very competitive. 
-As the number of garbage collections tend to be very large our graphs show only 
-a window which is representative of the behavior for that particular benchmark. 
-In all the programs we can see that the curve corresponding to liveness regularly 
-dips below the reachability curve.
-
-One area of concern is the huge gap between the actual liveness and the 
-liveness percieved by our collector. In case of LGC for eager languages the 
-gap was very narrow and almost touched the actual liveness curve. 
-In case of lazy languages, due to using reachability for copying we end up copying 
-more cells. Any data which is part of closure irrespective of whether it is live 
-or not gets copied. Thus many non-live cells get copied during garbage collection 
-increasing the drag time.
-Our experiments have shown that unless we use the exact execution point for 
-closures bringing down the number of non-live cells copied in closures cannot be reduced.
-Using the approximated liveness (liveness at the creation point) does not give benefits in terms of non-live
- cells inside closures as everything would be live at the point of creation.
+\begin{figure}[p]
+\begin{tabular}{@{}c@{}c@{}}
+\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=4.10cm}}
+&
+{\epsfig{file=lcss.eps, height=4cm, width=4.10cm}}
+\\
+\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=4.10cm}}
+ &
+{\epsfig{file=nperm.eps, height=4cm, width=4.10cm}}
+\\
+\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=4.10cm}}
+&
+{\epsfig{file=fibheap.eps, height=4cm, width=4.10cm}}%%Add knightstour.eps
+\\
+\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=4.10cm}}
+&
+{\epsfig{file=treejoin.eps, height=4cm, width=4.10cm}}
+\end{tabular}
+ \caption{Memory  usage  of programs.  The  blue  and the  red  curves
+   indicate the number of cons cells  in the active semi-space for LGC
+   and  RGC respectively.  The black  curve represents  the number  of
+   reachable cells and  the grey curve represents the  number of cells
+   that are  actually live (of  which liveness analysis does  a static
+   approximation). x-axis is the time measured in number of cons-cells
+   allocated. y-axis is the number of cons-cells.}
+\label{fig:memory-usage} \figrule
+\end{figure}
+The  table  in  Figure~\ref{fig:experimental-results} shows  that  LGC
+improves the  garbage collection time  and efficiency for most  of the
+programs.  Even for  programs where the time taken is  more it is very
+competitive.  Memory usage  graphs are shown for  select benchmarks in
+Figure~\ref{fig:memory-usage}.  As  the number of  garbage collections
+tend  to  be  very large  our  graphs  show  only  a window  which  is
+representative of the behavior for  that particular benchmark.  In all
+the programs we can see that  the curve corresponding to LGC regularly
+dips below the RGC curve.
+
+One area  of concern is the  huge gap between the  actual liveness and
+the liveness  percieved by  our collector.  In case  of LGC  for eager
+languages  the gap  was  very  narrow and  almost  touched the  actual
+liveness curve.  In case of  lazy languages, due to using reachability
+for copying we  end up copying more  cells. Any data which  is part of
+closure irrespective  of whether it is  live or not gets  copied. Thus
+many non-live  cells get  copied during garbage  collection increasing
+the drag  time.  Our  experiments have  shown that  unless we  use the
+exact  execution  point  for  closures bringing  down  the  number  of
+non-live  cells  copied in  closures  cannot  be reduced.   Using  the
+approximated liveness (liveness  at the creation point)  does not give
+benefits  in terms  of non-live  cells inside  closures as  everything
+would be live at the point of creation.
 %% \comment{
@@
 \hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5
-&.007 & .034 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147
-&588 &.045 & .144 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970
- &4 &.086 & .075 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507
-&171878 &76618 &1.406 & .9  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558
-&78720 &0 &.006 & .014  \\
+    {\sf   sudoku}  &381 &1107 &1268 &281 &169 &159 & 1346  &338 &527 &5
+&.008 & .026 \\
+    {\sf  lcss } & 46522 &51101 &6216 &1363&30&24& 52301  &1701 &5147
+&588 &.007 & .008 \\
+    {\sf   gc\_bench}  &129179 &131067 &1894 &4&34&4& 131071   &6 &16970
+ &4 &.13 & .0002 \\
+     {\sf  nperm}  & 47586  &174478 &201585 &60882&710&386& 202597  &37507
+&171878 &76618 &0.250 & .154  \\
+    {\sf  fibheap} &249502  &251525 &5555 &2997&1002&975& 254520  &13558
+&78720 &0 &1.185 & 5.457  \\
     {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225
 &307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150
-&212653 &1954 &.356 & .217 \\
+    {\sf  treejoin} & 288666  &519943 &297570 &5547&116&81& 525488  &7150
+&212653 &1954 &4.335 & 4.4 \\
     {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579
 &501093 &521826 &39465 &70.314 & 24.811 \\
-    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95
-&.093 &2.49  \\
+    {\sf   lambda}  &205 & 556 &2072&90345 &775 &750&966 & 721  &303 &95
+&.179 &1.01  \\
 %    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????
@@
 \centerline{(b)}
-\vskip -2mm
+\vskip -0.75mm
 \caption{Experimental results comparing RGC and LGC. Table (a) gives

### 2015-07-07 paper/paper.tex
graphs in two col.

@@
 
-\begin{figure}[p]
+\newcommand{\hgt}{3cm}
+\begin{figure*}[t]
 \begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}%, width=4.10cm}}
 &
-{\epsfig{file=lcss.eps, height=4cm, width=4.10cm}}
+{\epsfig{file=lcss.eps, height=\hgt}}%, width=4.10cm}}
 \\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}%
  &
-{\epsfig{file=nperm.eps, height=4cm, width=4.10cm}}
+{\epsfig{file=nperm.eps, height=\hgt}}%
 \\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}%
 &
-{\epsfig{file=fibheap.eps, height=4cm, width=4.10cm}}%%Add knightstour.eps
+{\epsfig{file=fibheap.eps, height=\hgt}}%%%Add knightstour.eps
 \\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}%
 &
-{\epsfig{file=treejoin.eps, height=4cm, width=4.10cm}}
+{\epsfig{file=treejoin.eps, height=\hgt}}%
 \end{tabular}
@@
 \label{fig:memory-usage} \figrule
-\end{figure}
+\end{figure*}
 The  table  in  Figure~\ref{fig:experimental-results} shows  that  LGC

### 2015-07-07 paper/paper.tex
Correctness proof

--- 2015-07-06 paper/paper.tex
@@
 \subsection{Liveness Analysis for lazy languages}
-
+\label{sec:liveness-analysis}
 Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
@@
 We shall now  outline a proof of correctness  of the liveness analysis
-described earlier.   This requires  a significant modification  to the
-technique  used  in  \cite{asati14lgc}.   We shall  first  modify  the
-standard semantics. This semantics, which we shall call {\em minefield
-  semantics}, will  model a liveness based garbage  collection at each
-transition. This is done by starting from the root-set and inserting a
-special value  $\bot$ at each reference  location in the  heap that is
-reachable but not live by  our analysis.  Attempts to dereference such
-locations  during the  transition will  result in  entering  a special
-state denoted  \bang.  We shall then  show that no  program enters the
-\bang\ state.
+presented  in  Section~\ref{sec:liveness-analysis}.   This requires  a
+significant modification  to the technique  used in \cite{asati14lgc}.
+We shall first modify the standard semantics. This semantics, which we
+shall  call {\em  minefield semantics},  will model  a  liveness based
+garbage collection at  each transition. This is done  by starting from
+the root-set  and inserting a  special value $\bot$ at  each reference
+location in the  heap that is reachable but not  live by our analysis.
+Attempts  to dereference  such  locations during  the transition  will
+result in entering a special  state denoted \bang.  We shall then show
+that no program enters the \bang\ state.
 
@@
 \begin{enumerate}
-\item Given an  demand transformer \Lfonly\  we enrich  the abstract machine
-  state  $\rho, S,  H, e$  to $\rho,  S, H,  e, \sigma,  \Sigma$. Here
-  $\sigma$ is  the demand on  the currently active function,  thus the
-  liveness $\Lv$   at $e$ is $\mathcal{L}(e_f, \sigma, \Lfonly)$, and
-  $\Sigma$ is a stack of  demands---one for each function frame pushed
-  in $S$.   Second, we  augment $Val$ with  a value $\bot$.   To model
-  strong  liveness  $\bot$ may  be  copied  freely,  including into  a
-  \CONS\ cell, but explodes  when used computationally (in a primitive
-  operation other  than a copy). Additionally we  define $GC (\Lv,
-  \Sigma) : (\rho, H, S) \rightarrow (\rho , H , S )$ which determines
-  live-reachability using $\rho$  and the $\rho$'s in $S$  as the root
-  set  and  following  links  in   $H$  only  as  far  as  allowed  by
+\item  Given an  demand transformer  \Lfonly\ we  enrich  the abstract
+  machine state $\rho, S, H, e$  to $\rho, S, H, e, \sigma$.  $\sigma$
+  is the  demand on the  currently active function. As  a consequence,
+  the liveness $\Lv$ at $e$ is $\mathcal{L}(e, \sigma, \Lfonly)$. Each
+  entry  in   the  stack  $S$   of  suspended  evaluations,   is  also
+  augmented---it is  now $(\rho, x,  e, eps, \sigma)$. The $\sigma$  in the
+  stack entry  represents the demand  on the  function where  the evaluation
+  \ldots.  And $eps$ is \ldots.
+\item We introduce a function $GC$ that takes the 4-tuple $(\rho, H, S,
+  \sigma)$ and  inserts $\bot$ in all locations in the current
+  environment $\rho$, the heap $H$ and the environments in the stack
+  $S$ that are dead. More formally:
+
+and $\Sigma$ is a  stack of demands---one for each function
+  frame pushed in $S$.  Second,  we augment $Val$ with a value $\bot$.
+  To model strong liveness $\bot$ may be copied freely, including into
+  a  \CONS\  cell,  but  explodes  when  used  computationally  (in  a
+  primitive operation  other than a copy). Additionally  we define $GC
+  (\Lv, \Sigma)  : (\rho,  H, S) \rightarrow  (\rho ,  H , S  )$ which
+  determines live-reachability using $\rho$ and the $\rho$'s in $S$ as
+  the root  set and following links in  $H$ only as far  as allowed by
   \Lv\   and   $\Sigma$.     Hence   $GC   (\Lv,   \Sigma)$   replaces
   live-unreachable  values---in $\rho$,  in  $H$ and  the $\rho$'s  in
-  $S$---with  $\bot$.  For  example,  if $x.\epsilon  \in \Lv$  then
-  $\rho (x) = \bot$.  Only $GC (\cdots )$ introduces $\bot$. 
+  $S$---with $\bot$.  For example,  if $x.\epsilon \in \Lv$ then $\rho
+  (x) = \bot$.  Only $GC (\cdots )$ introduces $\bot$.
 \end{enumerate}
@@
 \end{proposition}
-We now show that the decision problem required at garbage collection
-time is undecidable.
+We now show that the decision  problem that is required to be answered
+at       garbage      collection       time       is      undecidable.
 \newcommand{\state}{\ensuremath{\mathsf{S}}}
 \newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
-\cblue{
 \begin{lemma}
@@
 Assume that  the TM  starts in a  state $S_\mathit{init}$ with  a tape
-$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
-form    corresponding     to    the    initial     configuration    is
+$cw$  and the  head   positioned on  the symbol  $c$. Then  the the
+sentential  form   corresponding  to  the   initial  configuration  is
 $\nont{\mathit{init}}{c}w$ (we can assume  that there is a production
@@
 the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
-\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
+\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow 
 \nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
@@
 \end{proof}
-}%end of cblue
+
 We circumvent the problem  of undecidability by over- approximating the
@@
 
-\begin{figure*}[t!]
+\begin{figure*}[h!]
 \begin{center}
@@
 \end{figure*}
+
+\bigskip
+
+\begin{itemize}
+\item State of the operational semantics will be $\rho, S, H, e,
+  \sigma$.
+\item Stack content will be $\rho, x, e, \sigma, eps$.
+\item Liveness environment calculation:
+  \begin{enumerate}
+  \item Of the   current expression being
+  evaluated:\\  
+  $ \Lv = \bigcup_{ep \in \mathcal{P}(e)} \mathcal{M}_\sigma(ep)$, where $
+  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$
+\item Liveness environment of a stacked computation\\
+  $ \Lv = \bigcup_{ep \in eps} \mathcal{M}_\sigma(ep), \text{where }
+  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$ 
+  \end{enumerate}
+\item $GC$ models a liveness based garbage collection:
+$GC(\rho, S, H, L, SL) = (\rho', S', H')$ 
+  \begin{enumerate}
+  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
+    \notin \Lv$.
+\item For each $x \in domain(\rho)$ where $(\rho,\_,\_,\_,\_,\_)$ is a
+  stack entry  with \Lv\ as the  liveness environment, $\rho'(x)=\bot$
+  iff $x.\epsilon \notin \Lv$.
+\item for each location $\ell$, $H'(\ell) = \bot$, iff  there is no
+  $x$ in either the current environment or one of the stacked
+  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
+  \ell$  and $x.\alpha \notin \Lv$, .
+
+  \end{enumerate}
+\end{itemize}
 

### 2015-07-07 paper/paper.tex
new plots
Correctness proof

@@
 
+\pagebreak
+
+\pagebreak
+\section{Appendix}
+
+We shall now  outline a proof of correctness  of the liveness analysis
+presented  in  Section~\ref{sec:liveness-analysis}.   This requires  a
+significant modification  to the technique  used in \cite{asati14lgc}.
+\cblue{Good points to  make in the main text, right  in the small step
+  semantics  section: Note  that the  state of  the  transition system
+  consists of the current evaluation context, given by $\rho$ and $e$,
+  and  the suspended evaluation  contexts recorded  in the  stack $S$.
+  The heap $H$  is common to all contexts.}
+
+The idea behind the proof is  as follows: We shall modify the standard
+semantics to a semantics  called {\em minefield semantics} that models
+a liveness based garbage collection  at each transition.  The state of
+the  transition system  is augmented  to carry  enough  information to
+calculate the liveness environment  at the current evaluation context,
+and  each   of  the   suspended  evaluation  contexts.    Before  each
+transition,  we start  from the  root-set and  insert a  special value
+$\bot$ at each  location in the heap that contains  a reference and is
+reachable but  not live by  our analysis.  Any attempt  to dereference
+such locations during the transition will result in entering a special
+state denoted \bang.  The main proof is to show that no program enters
+the \bang\ state in the minefield semantics.
+
+To set up the minefield semantics, we follow these steps:
+\begin{enumerate}
+\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
+  $\rho,$ $S,$ $H,$  $e,$ $\sigma,$ $e_f$.  $\sigma$ is  the demand on
+  the expression and $e_f$ is the  body of the function.  The body and
+  the demand are both needed to calculate the liveness environment for
+  expressions.  The evaluation of  an expression may make variables in
+  the scope live because of closures that are created outside it.  The
+  information  in a  suspended  evaluation context  is also  similarly
+  augmented with the demand on it and the function body of which the
+  suspended evaluation is a part.  Thus a stacked entry now takes the
+  form $(\rho, x, e, \sigma, e_f)$. A modification of the small step
+  semantics which carries the extra information is shown in Figure[].
+\item Given the current evaluation context $\rho, S, H, e, \sigma,
+  e_f$ and $LF$, we can construct a liveness environment as follows:
+  \begin{enumerate}
+  \item If $e$ is an expression, then:  
+  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
+    \text{where }
+  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
+\item If $e$ is an application, then:  
+  $$ \Lv = ref(e, \sigma, \Lfonly)$$
+  \end{enumerate}
+We can define in the same way the liveness environment for each of the
+suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
+environments that we shall denote $\mathsf{SL}$.
+\item $GC$ models a liveness based garbage collection:
+$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
+  \begin{enumerate}
+  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
+    \notin \Lv$.
+\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
+  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
+  \in domain(\rho)$,  $\rho'(x)=\bot$
+  iff $x.\epsilon \notin \Lv'$.
+\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
+  $x$ in either the current environment or one of the stacked
+  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
+  \ell$  and $x.\alpha \in \Lv$, .
+  \end{enumerate}
+\end{enumerate}
+
 \begin{figure*}[h!]
@@
 \hline
-          &\makecell{ $\rho, (\rho', x, e, \{\psi_{i}\}, \sigma')\!:\!S, H, \kappa, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e, \sigma'$ }   &  \sc{const}
+          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
+  H, \kappa, \sigma, e_f$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
+    \kappa], e, \sigma', e_f'$ }   &  \sc{const}
 \\
 \hline
-          & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (\CONS~x~y), \sigma$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma'$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma' )\!:\!S, H, (\CAR~x), \sigma$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma'$}      &
+          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
+    e_f'})\!:\!S, H, (\CONS~x~y), \sigma, e_f$ \\ $\rightsquigarrow
+  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma', e_f'$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e,
+  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \sigma, e_f$ \\ $
+  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma', e_f'$}      &
 \sc{car-whnf} \\
 \hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x), \{\psi_{i}\}, \sigma)\!:\!S, H, s,\bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x} $}      &
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+  H, (\CAR~x), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, (\clazy \cup \acar)\sigma , \_ $}      &
 \sc{car-clo}
@@
 $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (+~x~y), \sigma$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma'$}      &
+ & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
+  (+~x~y), \sigma, e_f$ \\  $\rightsquigarrow \rho', S, H[\rho(z)
+    \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma', e_f'$}      &
 \sc{prim-whnf} \\
 \hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x}$}      &
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+  H, (+~x~y), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, x,
+  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \clazy\sigma, \_$}      &
 \sc{prim-1-clo} \\
 \hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, y, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{y} $}      &
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+  S, H, (+~x~y), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, y,
+  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \clazy\sigma, \_ $}      &
 \sc{prim-2-clo} \\
 \hline
-{$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x}), \sigma$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}, \sigma$}      &
+{$\mathit{g}~\mbox{defined as}$
+$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
+  (g~\myvec{x}), \sigma, e_f$ \\ $\rightsquigarrow [\myvec{y} \mapsto
+    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \sigma, e_g$}      &
 \sc{funcall} \\
 \hline
-$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e$} &
+$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
+  s~\IN~e), \sigma, e_f$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+  \ell]\rangle], e, \sigma, e_f$} &
 \sc{let} \\
 \hline
-$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
-\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\pi:\SIF~\psi:x~e_1~e_2), \sigma$ \\ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2),  \pmap{}(\pi), \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$}
+$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \sigma, e_f$ \\ $\rightsquigarrow \rho, S, H,  e_1, \sigma, e_f$} & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \sigma, e_f$ \\   $\rightsquigarrow
+\rho, S, H,  e_2, \sigma, e_f$} & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\pi:\SIF~\psi:x~e_1~e_2), \sigma, e_f$ \\ $\rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2),  \sigma, e_f)\!:\!S, H, s,
+\clazy\sigma, e_f$}
 &
@@
 \hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma')\!:\!S, H, (\SRETURN~x), \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, \sigma'$} &
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
+  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \sigma,
+  e_f$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e,
+  \sigma', e_f'$} &
 \sc{return-whnf}\\
 \hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\psi:\SRETURN~x), \sigma$ \\ $
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\psi:\SRETURN~x), \sigma, e_f$ \\ $
   \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x), \{\psi\}, \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$} &
+$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
+  \sigma, e_f$} &
 \sc{return-clo} \\
@@
 \end{figure*}
-
-\bigskip
-
-\begin{itemize}
-\item State of the operational semantics will be $\rho, S, H, e,
-  \sigma$.
-\item Stack content will be $\rho, x, e, \sigma, eps$.
-\item Liveness environment calculation:
-  \begin{enumerate}
-  \item Of the   current expression being
-  evaluated:\\  
-  $ \Lv = \bigcup_{ep \in \mathcal{P}(e)} \mathcal{M}_\sigma(ep)$, where $
-  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$
-\item Liveness environment of a stacked computation\\
-  $ \Lv = \bigcup_{ep \in eps} \mathcal{M}_\sigma(ep), \text{where }
-  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$ 
-  \end{enumerate}
-\item $GC$ models a liveness based garbage collection:
-$GC(\rho, S, H, L, SL) = (\rho', S', H')$ 
-  \begin{enumerate}
-  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
-    \notin \Lv$.
-\item For each $x \in domain(\rho)$ where $(\rho,\_,\_,\_,\_,\_)$ is a
-  stack entry  with \Lv\ as the  liveness environment, $\rho'(x)=\bot$
-  iff $x.\epsilon \notin \Lv$.
-\item for each location $\ell$, $H'(\ell) = \bot$, iff  there is no
-  $x$ in either the current environment or one of the stacked
-  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
-  \ell$  and $x.\alpha \notin \Lv$, .
-
-  \end{enumerate}
-\end{itemize}
 

### 2015-07-07 paper/paper.tex

@@
 
-\begin{figure}[p]
+\newcommand{\hgt}{3cm}
+\begin{figure*}[t]
+\renewcommand{\arraystretch}{.1}
 \begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}%, width=4.10cm}}
 &
-{\epsfig{file=lcss.eps, height=4cm, width=4.10cm}}
+{\epsfig{file=lcss.eps, height=\hgt}}%, width=4.10cm}}
 \\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}%
  &
-{\epsfig{file=nperm.eps, height=4cm, width=4.10cm}}
+{\epsfig{file=nperm.eps, height=\hgt}}%
 \\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}%
 &
-{\epsfig{file=fibheap.eps, height=4cm, width=4.10cm}}%%Add knightstour.eps
+{\epsfig{file=fibheap.eps, height=\hgt}}%%%Add knightstour.eps
 \\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=4.10cm}}
+\hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}%
 &
-{\epsfig{file=treejoin.eps, height=4cm, width=4.10cm}}
-\end{tabular}
- \caption{Memory  usage  of programs.  The  blue  and the  red  curves
-   indicate the number of cons cells  in the active semi-space for LGC
-   and  RGC respectively.  The black  curve represents  the number  of
-   reachable cells and  the grey curve represents the  number of cells
-   that are  actually live (of  which liveness analysis does  a static
-   approximation). x-axis is the time measured in number of cons-cells
-   allocated. y-axis is the number of cons-cells.}
+{\epsfig{file=treejoin.eps, height=\hgt}}%
+\end{tabular}\vskip -10mm
+ \caption{Memory usage  snapshot of  programs.  The  blue and  the red
+   curves indicate the  number of cons cells in  the active semi-space
+   for  LGC and  RGC  respectively.  The  black  curve represents  the
+   number of reachable cells and  the grey curve represents the number
+   of cells that are actually live  (of which liveness analysis does a
+   static approximation).  x-axis is  the time  measured in  number of
+   cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
+   number of cons-cells (scaled down by factor $10^3$).}
 \label{fig:memory-usage} \figrule
-\end{figure}
+\end{figure*}
 The  table  in  Figure~\ref{fig:experimental-results} shows  that  LGC

### 2015-07-08 lazy-scheme/post_process/gen_gc_stats.sh
automated table generation

@@ -105,8 +105,8 @@ for file in $STATOUTFILES; do
     rm -f ${file}.out
 done
 
-PPSCRIPT=$DIRNAME/process
 PPSCRIPT=$DIRNAME/process_fast
+PPSCRIPT=$DIRNAME/process
 OUTFILE=general_stats.out
 $PPSCRIPT $STATSFILE  > $OUTFILE
 

### 2015-07-08 lazy-scheme/post_process/minMemorySearch.sh

@@ -2,7 +2,7 @@
 TEST=$1
 MIN=${3-0}
 MAX=$2
-FULLTEST=../benchmarks/programs/$TEST/anf-$TEST.scm
+FULLTEST=../benchmarks/programs/$TEST/$TEST.scm
 LOG=run.out.$$
 GCOPT=gc-plain
 
@@ -11,8 +11,8 @@ RANGE=`expr $MAX - $MIN`
 while [ $RANGE -gt 1 ]; do 
     CUR=`expr \( $MIN  + $MAX \) / 2`
     echo "Running with Memory = $CUR"
-    python collector_gc.py $FULLTEST $CUR $GCOPT > $LOG 2>&1
-    if grep -q "Memory Exhausted" $LOG; then
+    ./Simulator $FULLTEST $CUR $GCOPT > $LOG 2>&1
+    if grep -q "No Sufficient Memory" $LOG; then
 	# increase MEMORY
 	MIN=$CUR
     else

### 2015-07-08 lazy-scheme/post_process/touchedAndCopied.sh

@@ -1,4 +1,6 @@
 #!/bin/sh
+DIR=`dirname $0`
+SCRIPT=$DIR/touchedAndCopied.awk
 AllBMs="sudoku lcss gc_bench nperm fibheap  knightstour treejoin nqueens "
 AllBMs="sudoku treejoin lambda fibheap"
 for BM in $AllBMs
@@ -7,6 +9,6 @@ do
     do 
 	FILE=output/${BM}_${OPT}_output/garbage-dump.txt
 	echo "==== $BM ($OPT) ==== "
-	awk -f ./post_process/touchedAndCopied.awk $FILE
+	awk -f $SCRIPT $FILE
     done
 done

### 2015-07-08 lazy-scheme/src/Simulator.cpp
Updated code to print DFA gen time and number of states as part of output

@@ -127,8 +127,9 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 {
 	//Do any per program initialization here.
 	Scheme::SchemeDriver driver;
-	cout << "Parsing " << pgmFilePath << endl;
+//	cout << "Parsing " << pgmFilePath << endl;
 	std::pair<bool, long> parseResult = driver.parse(pgmFilePath.c_str());
+	double dfa_gen_time = 0;
 	if(!parseResult.first)
 	{
 		cout << "Error while parsing " << parseResult.second<<endl;
@@ -148,6 +149,7 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 
 	if (gc_type == gc_live && !filesCached )
 	{
+		clock_t dfa_gen_start = clock();
 		
 		//Instead of driver.process returning an integer why can't it return the grammar?
 		int resint = driver.process();
@@ -168,12 +170,12 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 		write_grammar_to_text_file(&gLivenessData, outdir + pgmname + "/simplified-program-cfg.txt");
 		//Convert CFG to strongly regular grammar
 		regular_demand_grammar *reg = regularize(&gLivenessData);
-		cout << "Converted CFG into strongly regular grammar" << endl;
+//		cout << "Converted CFG into strongly regular grammar" << endl;
 		gLivenessData.clear(); //clear the original grammar
 		gLivenessData.swap(*(reg->first));
 		write_grammar_to_text_file(&gLivenessData, outdir + pgmname + "/program-reg.txt");
 		Scheme::Demands::sanitize(&gLivenessData); //Remove empty productions
-		std::cout << "Sanitized the regular grammar"<<std::endl;
+//		std::cout << "Sanitized the regular grammar"<<std::endl;
 
 
 		automaton *nfa = Scheme::Demands::getNFAsFromRegularGrammar(&gLivenessData, pgmname);
@@ -197,6 +199,8 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 		}
 		numkeys = Scheme::Demands::writeDFAToFile(pgmname, dfa, label_set_map);
 
+		dfa_gen_time = (double(clock() - dfa_gen_start)/CLOCKS_PER_SEC);
+
[…]

### 2015-07-08 lazy-scheme/src/run_in_loop.py

@@ -5,8 +5,9 @@ import time
 import datetime
 from collections import namedtuple
 
-prog_size = {'nqueens':22729, 'nperm':27429, 'fibheap':38001, 'treejoin':1616535, 'sudoku':4075, 'lcss':22243, 'lambda':20466, 'knightstour':677800 ,'gc_bench':204850}
-gc_info = namedtuple("gc_info", "heap_total heap_left heap_used gc_invocations gc_time exec_time")
+#'knightstour':677800
+prog_size = {'nqueens':22729, 'nperm':27429, 'fibheap':38001, 'treejoin':1616535, 'sudoku':4075, 'lcss':22243, 'lambda':20466, 'gc_bench':204850}
+gc_info = namedtuple("gc_info", "heap_total heap_left heap_used gc_invocations gc_time dfa_time num_states exec_time")
 gc_stat = namedtuple("gc_stat", "prog_name reachability liveness")
 gc_stat_list = []
 
@@ -30,14 +31,18 @@ def collect_gc_results(filename, prog_dir_path, gc_type):
     heap_used_cmd = grep_cmd + " \"Heap used=[0-9]*\"   " +  result_file_name  + int_grep
     gc_invocations_cmd = grep_cmd + " \"GC Invocations=[0-9]*\"   " +  result_file_name  + int_grep
     gc_time_cmd = grep_cmd + " \"GC Time=(([0-9][0-9]*\.?[0-9]*)|(\.[0-9]+))([Ee][+-]?[0-9]+)?\"   " +  result_file_name  + float_grep
+    dfa_time_cmd = grep_cmd + " \"DFA Gen Time=(([0-9][0-9]*\.?[0-9]*)|(\.[0-9]+))([Ee][+-]?[0-9]+)?\"   " +  result_file_name  + float_grep
+    dfa_states_cmd = grep_cmd + " \"Num of DFA states=(([0-9][0-9]*\.?[0-9]*)|(\.[0-9]+))([Ee][+-]?[0-9]+)?\"   " +  result_file_name  + int_grep
     prog_execution_time_cmd = grep_cmd + " \"Program Execution Time=[0-9]*(\.[0-9]*)?\"   " +  result_file_name  + float_grep
     heap_total = os.popen(heap_total_cmd).read()
     heap_left = os.popen(heap_left_cmd).read()
     heap_used = os.popen(heap_used_cmd).read()
     gc_invocations = os.popen(gc_invocations_cmd).read()
[…]

### 2015-07-09 paper/paper.tex
automated table generation
Updated table
typos
fitting stuff at right place

@@
 
+In this paper,  we propose the use of liveness  analysis of heap cells
+for garbage collection in a lazy first-order functional language.  The
+central notion in our analysis  is a generalization of liveness called
+{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
+expression.   The analysis  has two  parts.  First,  each  function is
+summarized in a context-sensitive manner as a {\em demand transformer}
+that transforms a {\em symbolic} demand  on its body to demands on its
+arguments.  This summary  is then used to step  through function calls
+during analysis.   Second, the concrete  demand on a function  body is
+obtained through a 0-CFA-like conservative approximation that combines
+the  demands on  all the  calls to  the function.   The result  of the
+analysis  is the annotation  of each  program point  with finite-state
+automata capturing  the liveness of variables at  the point. Depending
+on the program point where  garbage collection has been triggered, the
+GC consults  a set of automata  to curtail reachability
+during marking.  This results in  an increase in the garbage reclaimed
+and consequently in fewer garbage collections.
+
+While this idea has been shown to be effective~\cite{asati14lgc} for a
+first-order {\em  eager} language, a straightforward  extension of the
+earlier  technique  is  not  possible  for  lazy  languages,  where  a
+heap-allocated  object may either  be an  unevaluated expression  or a
+closure. Since  data is  made live by  evaluation of closures,  and in
+lazy languages  the place in  the program where this  evaluation takes
+place cannot  be statically determined,  laziness complicates liveness
+analysis itself.  In  addition, apart from evaluated data,  we want to
+extend  liveness-based collection also  to closures---closures with no
+future use should also  be collected.  This involves tracking closures
+to  their points of  creation at  runtime from  where we  obtain their
+liveness.  This  requires significant  modification to the  earlier GC
+scheme.
+
+{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
+single
+  generation  copying  collector   confirm  the  expected  performance
+  benefits.   Liveness-based  collection  results  in an  increase  in
+  garbage reclaimed.   As a  consequence, there is  a decrease  in the
+  number of  collections and a  (some quantification) decrease  in the
+  memory size required  to run programs. Besides there  is a reduction
+  in the overall execution time for some programs.}
+
+%------------------------------------------------------------%
+\subsection{Motivating Example}
+\label{sec:motiv}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
             (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
-            \UNL{3} (\SIF~(\NULLQ~(\CAR~(\CDR~\pz)))~$0$~$\pi$:\
+            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
             (\length\ \pz))))))
@@
       \renewcommand{\arraystretch}{.9}
-      \begin{tabular}[t]{l}
-        (b) Memory graph at $\pi$. Thick edges denote live
-        links. Traversal \\{\white (b)} stops at edges
-        marked $\times$ during garbage collection.
+      \begin{tabular}[t]{p{0.9\columnwidth}}
+        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
+        denotes a closure. Thick edges denote live links. Traversal
+        stops at edges marked $\times$ during garbage collection.
       \end{tabular}
@@
 \end{figure}
-
-In this paper,  we propose the use of liveness  analysis of heap cells
-for garbage collection in a lazy first-order functional language.  The
-central notion in our analysis  is a generalization of liveness called
-{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
-expression.   The analysis  has two  parts.  First,  each  function is
-summarized in a context-sensitive manner as a {\em demand transformer}
-that transforms a {\em symbolic} demand  on its body to demands on its
-arguments.  This summary  is then used to step  through function calls
-during analysis.   Second, the concrete  demand on a function  body is
-obtained through a 0-CFA-like conservative approximation that combines
-the  demands on  all the  calls to  the function.   The result  of the
-analysis  is the annotation  of each  program point  with finite-state
-automata capturing  the liveness of variables at  the point. Depending
-on the program point where  garbage collection has been triggered, the
-GC consults  a set of automata  to curtail reachability
-during marking.  This results in  an increase in the garbage reclaimed
-and consequently in fewer garbage collections.
-
-While this idea has been shown to be effective~\cite{asati14lgc} for a
-first-order {\em  eager} language, a straightforward  extension of the
-earlier  technique  is  not  possible  for  lazy  languages,  where  a
-heap-allocated  object may either  be an  unevaluated expression  or a
-closure. Since  data is  made live by  evaluation of closures,  and in
-lazy languages  the place in  the program where this  evaluation takes
-place cannot  be statically determined,  laziness complicates liveness
-analysis itself.  In  addition, apart from evaluated data,  we want to
-extend  liveness-based collection also  to closures---closures with no
-future use should also  be collected.  This involves tracking closures
-to  their points of  creation at  runtime from  where we  obtain their
-liveness.  This  requires significant  modification to the  earlier GC
-scheme.
-
-{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
-single
-  generation  copying  collector   confirm  the  expected  performance
-  benefits.   Liveness-based  collection  results  in an  increase  in
-  garbage reclaimed.   As a  consequence, there is  a decrease  in the
-  number of  collections and a  (some quantification) decrease  in the
-  memory size required  to run programs. Besides there  is a reduction
-  in the overall execution time for some programs.}
-
-%------------------------------------------------------------%
-\subsection{Motivating Example}
-\label{sec:motiv}
 
@@
 
+{\color {Myblue}Rewrite at the end: 
 Section~\ref{sec:defs} introduces the syntax and semantics of the
@@
 % and Section~\ref{sec:conclusion} concludes the paper.
-
+}
 \section{The target language---syntax and semantics}
@@
 %%We let $x$,  $y$, $z$ range over variables,  $f$ over user-functions
-and
-%% $p$    over    primitive    functions   ($\CONS$,    $\PRIM$ etc.).
+%% and $p$ over primitive functions ($\CONS$, $\PRIM$ etc.).
 Figure~\ref{fig:lang-syntax} describes the  syntax of our language. It
@@
 
+\include{autotable}
+
 \centerline{(b)}
@@
           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
-  H, \kappa, \sigma, e_f$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
-    \kappa], e, \sigma', e_f'$ }   &  \sc{const}
+  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
+    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
 \\
@@
           & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
-    e_f'})\!:\!S, H, (\CONS~x~y), \sigma, e_f$ \\ $\rightsquigarrow
-  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma', e_f'$}     &  \sc{cons} \\
+    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
+  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
 \hline
 $H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e,
-  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \sigma, e_f$ \\ $
-  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma', e_f'$}      &
+  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
+  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \cred{\sigma', e_f'}$}      &
 \sc{car-whnf} \\
@@
 $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-  H, (\CAR~x), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, (\clazy \cup \acar)\sigma , \_ $}      &
+  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
 \sc{car-clo}
@@
  & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
-  (+~x~y), \sigma, e_f$ \\  $\rightsquigarrow \rho', S, H[\rho(z)
-    \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma', e_f'$}      &
+  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho(z)
+    \mapsto H(\rho(x)) + H(\rho(y))], e, \cred{\sigma', e_f'}$}      &
 \sc{prim-whnf} \\
@@
 $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-  H, (+~x~y), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \clazy\sigma, \_$}      &
+  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
+  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
 \sc{prim-1-clo} \\
@@
 $H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-  S, H, (+~x~y), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, y,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \clazy\sigma, \_ $}      &
+  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
+  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
 \sc{prim-2-clo} \\
@@
 $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
-  (g~\myvec{x}), \sigma, e_f$ \\ $\rightsquigarrow [\myvec{y} \mapsto
-    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \sigma, e_g$}      &
+  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
+    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
 \sc{funcall} \\
@@
 $\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
-  s~\IN~e), \sigma, e_f$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e, \sigma, e_f$} &
+  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+  \ell]\rangle], e, \cred{\sigma, e_f}$} &
 \sc{let} \\
@@
 $H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \sigma, e_f$ \\ $\rightsquigarrow \rho, S, H,  e_1, \sigma, e_f$} & \sc{if-true} \\
+  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
 \hline
 $H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \sigma, e_f$ \\   $\rightsquigarrow
-\rho, S, H,  e_2, \sigma, e_f$} & \sc{if-false} \\
+  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
+\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
 \hline
 $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\pi:\SIF~\psi:x~e_1~e_2), \sigma, e_f$ \\ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2),  \sigma, e_f)\!:\!S, H, s,
-\clazy\sigma, e_f$}
+  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
+\cred{\clazy\sigma, e_f}$}
 &
@@
 {$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
-  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \sigma,
-  e_f$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e,
-  \sigma', e_f'$} &
+  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
+  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e,
+  \cred{\sigma', e_f'}$} &
 \sc{return-whnf}\\
@@
 $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\psi:\SRETURN~x), \sigma, e_f$ \\ $
+  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
   \rightsquigarrow$
 $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
-  \sigma, e_f$} &
+  \cred{\sigma, e_f}$} &
 \sc{return-clo} \\

### 2015-07-09 paper/paper.tex

@@
 
-In this paper,  we propose the use of liveness  analysis of heap cells
-for garbage collection in a lazy first-order functional language.  The
-central notion in our analysis  is a generalization of liveness called
-{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
-expression.   The analysis  has two  parts.  First,  each  function is
-summarized in a context-sensitive manner as a {\em demand transformer}
-that transforms a {\em symbolic} demand  on its body to demands on its
-arguments.  This summary  is then used to step  through function calls
-during analysis.   Second, the concrete  demand on a function  body is
-obtained through a 0-CFA-like conservative approximation that combines
-the  demands on  all the  calls to  the function.   The result  of the
-analysis  is the annotation  of each  program point  with finite-state
-automata capturing  the liveness of variables at  the point. Depending
-on the program point where  garbage collection has been triggered, the
-GC consults  a set of automata  to curtail reachability
-during marking.  This results in  an increase in the garbage reclaimed
-and consequently in fewer garbage collections.
-
-While this idea has been shown to be effective~\cite{asati14lgc} for a
-first-order {\em  eager} language, a straightforward  extension of the
-earlier  technique  is  not  possible  for  lazy  languages,  where  a
-heap-allocated  object may either  be an  unevaluated expression  or a
-closure. Since  data is  made live by  evaluation of closures,  and in
-lazy languages  the place in  the program where this  evaluation takes
-place cannot  be statically determined,  laziness complicates liveness
-analysis itself.  In  addition, apart from evaluated data,  we want to
-extend  liveness-based collection also  to closures---closures with no
-future use should also  be collected.  This involves tracking closures
-to  their points of  creation at  runtime from  where we  obtain their
-liveness.  This  requires significant  modification to the  earlier GC
-scheme.
-
-{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
-single
-  generation  copying  collector   confirm  the  expected  performance
-  benefits.   Liveness-based  collection  results  in an  increase  in
-  garbage reclaimed.   As a  consequence, there is  a decrease  in the
-  number of  collections and a  (some quantification) decrease  in the
-  memory size required  to run programs. Besides there  is a reduction
-  in the overall execution time for some programs.}
-
-%------------------------------------------------------------%
-\subsection{Motivating Example}
-\label{sec:motiv}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
             (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
-            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
+            \UNL{3} (\SIF~(\NULLQ~(\CAR~(\CDR~\pz)))~$0$~$\pi$:\
             (\length\ \pz))))))
@@
       \renewcommand{\arraystretch}{.9}
-      \begin{tabular}[t]{p{0.9\columnwidth}}
-        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
-        denotes a closure. Thick edges denote live links. Traversal
-        stops at edges marked $\times$ during garbage collection.
+      \begin{tabular}[t]{l}
+        (b) Memory graph at $\pi$. Thick edges denote live
+        links. Traversal \\{\white (b)} stops at edges
+        marked $\times$ during garbage collection.
       \end{tabular}
@@
 \end{figure}
+
+In this paper,  we propose the use of liveness  analysis of heap cells
+for garbage collection in a lazy first-order functional language.  The
+central notion in our analysis  is a generalization of liveness called
+{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
+expression.   The analysis  has two  parts.  First,  each  function is
+summarized in a context-sensitive manner as a {\em demand transformer}
+that transforms a {\em symbolic} demand  on its body to demands on its
+arguments.  This summary  is then used to step  through function calls
+during analysis.   Second, the concrete  demand on a function  body is
+obtained through a 0-CFA-like conservative approximation that combines
+the  demands on  all the  calls to  the function.   The result  of the
+analysis  is the annotation  of each  program point  with finite-state
+automata capturing  the liveness of variables at  the point. Depending
+on the program point where  garbage collection has been triggered, the
+GC consults  a set of automata  to curtail reachability
+during marking.  This results in  an increase in the garbage reclaimed
+and consequently in fewer garbage collections.
+
+While this idea has been shown to be effective~\cite{asati14lgc} for a
+first-order {\em  eager} language, a straightforward  extension of the
+earlier  technique  is  not  possible  for  lazy  languages,  where  a
+heap-allocated  object may either  be an  unevaluated expression  or a
+closure. Since  data is  made live by  evaluation of closures,  and in
+lazy languages  the place in  the program where this  evaluation takes
+place cannot  be statically determined,  laziness complicates liveness
+analysis itself.  In  addition, apart from evaluated data,  we want to
+extend  liveness-based collection also  to closures---closures with no
+future use should also  be collected.  This involves tracking closures
+to  their points of  creation at  runtime from  where we  obtain their
+liveness.  This  requires significant  modification to the  earlier GC
+scheme.
+
+{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
+single
+  generation  copying  collector   confirm  the  expected  performance
+  benefits.   Liveness-based  collection  results  in an  increase  in
+  garbage reclaimed.   As a  consequence, there is  a decrease  in the
+  number of  collections and a  (some quantification) decrease  in the
+  memory size required  to run programs. Besides there  is a reduction
+  in the overall execution time for some programs.}
+
+%------------------------------------------------------------%
+\subsection{Motivating Example}
+\label{sec:motiv}
 
@@
 
-{\color {Myblue}Rewrite at the end: 
 Section~\ref{sec:defs} introduces the syntax and semantics of the
@@
 % and Section~\ref{sec:conclusion} concludes the paper.
-}
+
 \section{The target language---syntax and semantics}
@@
 %%We let $x$, $y$, $z$ range over variables, $f$ over user-functions
-%% and $p$ over primitive functions ($\CONS$, $\PRIM$ etc.).
+and
+%% $p$    over    primitive    functions   ($\CONS$,    $\PRIM$ etc.).
 Figure~\ref{fig:lang-syntax} describes the  syntax of our language. It

### 2015-07-09 lazy-scheme/post_process/createTable.sh
stablising towards  data

@@ -21,28 +21,28 @@ cat <<EOF
   &   \multicolumn{2}{c|}{Drag Reduction (\%)}
   &   \multicolumn{2}{c|}{(sec)} & \\\\
 \cline{2-11}
-{Program}&RGC&LGC&RGC&LGC&RGC&LGC&\#Cells&Time&LGC&RGC&Speedup \\\\
+{Program}&RGC&LGC&RGC&LGC&RGC&LGC&\#Cells&Time&RGC&LGC&Speedup \\\\
 \hline
 \hline
 
 EOF
 format="%s & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f"
-
+OUTPUTDIR=output.GOLD
 for BM in $AllBMs
 do
     line=""
-    FILE=output/${BM}_gc-plain_output/garbage-dump.txt
+    FILE=${OUTPUTDIR}/${BM}_gc-plain_output/garbage-dump.txt
     line="$line `awk -f $SCRIPT1 $FILE`"
-    FILE=output/${BM}_gc-live_output/garbage-dump.txt
+    FILE=${OUTPUTDIR}/${BM}_gc-live_output/garbage-dump.txt
     line="$line  `awk -f $SCRIPT1 $FILE`"
 
-    FILE=output/${BM}_gc-plain_output/general_stats.out
+    FILE=${OUTPUTDIR}/${BM}_gc-plain_output/general_stats.out
     tmpData=`cut -d: -f2 $FILE`
     data1=`echo $tmpData | cut -d' ' -f2`
     data2=`echo $tmpData | cut -d' ' -f3`
 
     factor=100000
-    FILE=output/${BM}_gc-live_output/general_stats.out
+    FILE=${OUTPUTDIR}/${BM}_gc-live_output/general_stats.out
     tmpData=`cut -d: -f2 $FILE`
     data3=`echo $tmpData | cut -d' ' -f2`
     data4=`echo $tmpData | cut -d' ' -f3`
@@ -53,17 +53,24 @@ do
 
     #GC time to be filled separately
     
-    FILE=output/${BM}_gc-plain_output/runlog.out
-    r1=TBD1
-    FILE=output/${BM}_gc-live_output/runlog.out
-    r2=TBD2
-    line="$line  $r1  $r2  r1/r2"
+    FILE=../benchmarks/programs/${BM}/results/${BM}-gc-plain
+    r1=`grep "GC Time" $FILE | cut -d= -f3`
[…]

### 2015-07-09 lazy-scheme/post_process/gen_stats_and_plotall.sh

@@ -56,7 +56,7 @@ limit["fft"]=3300
 limit["small"]=17
 limit["test"]=247
 limit["testsmall"]=267
-limit["knightstour"]=677800
+limit["knightstour"]=6778000
 
 declare -A freq
 

### 2015-07-09 lazy-scheme/src/run_in_loop.py

@@ -6,7 +6,7 @@ import datetime
 from collections import namedtuple
 
 #'knightstour':677800
-prog_size = {'nqueens':22729, 'nperm':27429, 'fibheap':38001, 'treejoin':1616535, 'sudoku':4075, 'lcss':22243, 'lambda':20466, 'gc_bench':204850}
+prog_size = {'nqueens':22729, 'nperm':27429, 'fibheap':38001, 'treejoin':1616535, 'sudoku':4075, 'lcss':22243, 'lambda':20466, 'gc_bench':204850, 'knightstour':6778000}
 gc_info = namedtuple("gc_info", "heap_total heap_left heap_used gc_invocations gc_time dfa_time num_states exec_time")
 gc_stat = namedtuple("gc_stat", "prog_name reachability liveness")
 gc_stat_list = []

### 2015-07-09 lazy-scheme/post_process/gen_stats_and_plotall.sh
stablising towards  data

@@ -52,11 +52,12 @@ limit["sudoku"]=4075
 limit["lcss"]=22243
 limit["lambda"]=20466
 limit["gc_bench"]=204850
+limit["knightstour"]=6778000
+## UNUSED BMs
 limit["fft"]=3300
 limit["small"]=17
 limit["test"]=247
 limit["testsmall"]=267
-limit["knightstour"]=6778000
 
 declare -A freq
 

### 2015-07-09 paper/paper.tex
figure fixed

--- 2015-07-09 paper/paper.tex
@@
 \usepackage[T1]{fontenc}
-%\usepackage[pdf]{pstricks}
 \usepackage{epsfig}
@@
 \usepackage{tabularx}
-\usepackage [plain,noend,noline,boxed]{algorithm2e}
 \usepackage{amssymb}
 \usepackage{amsmath}
-%\usepackage{amsthm}
 \usepackage{pstricks}
@@
 \usepackage{subfigure}
-%\usepackage{program}
-%\usepackage{algorithmic}
 \usepackage[boxed]{algorithm2e}
@@
 
+\begin{figure*}[t!]
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Premise & Transition & Rule name \\
+\hline
+\hline
+          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
+  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
+\\
+\hline
+          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
+\rightsquigarrow
+$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
+\sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
+\rightsquigarrow
+\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
+\sc{car-clo}
+\\
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
+$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
+\sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
+\rightsquigarrow
+\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
+\sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
+\rightsquigarrow
+\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
+\sc{prim-2-clo} \\
+\hline
+{$\mathit{f}~\mbox{defined as}$
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
+(f~\myvec{x})  \rightsquigarrow
+[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
+\sc{funcall} \\
+\hline
+$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+  \rightsquigarrow$
+$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
+    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+  \ell]\rangle], e$} &
+\sc{let} \\
+\hline
+$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_1$ & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho, S, H,  e_2$ & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
+  (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
+&
+\sc{if-clo} \\
+\hline
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
+e)\!:\!S, H,
+(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
+\sc{return-whnf}\\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
+  \rightsquigarrow$
+$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
+\sc{return-clo} \\
+\hline
+\end{tabular}
+\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
+\end{center}
+\end{figure*}
+
 \subsection{Semantics}
@@
 has to be resumed after the argument is evaluated.
-
-\begin{figure*}[t!]
-\begin{center}
-\begin{tabular}{|c|c|c|}
-\hline
-Premise & Transition & Rule name \\
-\hline
-\hline
-          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
-  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
-\\
-\hline
-          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
-\rightsquigarrow
-$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
-(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
-\rightsquigarrow
-\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
-\sc{car-clo}
-\\
-\hline
-$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
-$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
-\sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
-\rightsquigarrow
-\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
-\sc{prim-1-clo} \\
-\hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
-\rightsquigarrow
-\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
-\sc{prim-2-clo} \\
-\hline
-{$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
-(f~\myvec{x})  \rightsquigarrow
-[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
-\sc{funcall} \\
-\hline
-$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
-  \rightsquigarrow$
-$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
-    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e$} &
-\sc{let} \\
-\hline
-$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho, S, H,  e_1$ & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho, S, H,  e_2$ & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
-  (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
-&
-\sc{if-clo} \\
-\hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
-e)\!:\!S, H,
-(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
-\sc{return-whnf}\\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
-  \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
-\sc{return-clo} \\
-\hline
-\end{tabular}
-\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
-\end{center}
-\end{figure*}
 
@@
 \label{sec:liveness-analysis}
-Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
-parts. The function $\mathit{ref}$,  takes  an application $s$ and
-a demand $\sigma$, and returns  the incremental liveness generated for the
-free  variables   of  $s$  due  to  the   application.   The  function
-$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
-expressions.
-
- Since  in a  lazy language,  an  expression is  not evaluated  unless
- required, the null demand ($\emptyset$) does not generate liveness in
- any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
- non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed  to the
- liveness $\{x.\epsilon, x.\acdr\sigma\}$.   In an opposite sense, the
- demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
- demand  $\sigma$ on  $z$.   Note  that \CONS\  does  not, by  itself,
- dereference its  arguments. The rules for  (\PRIM~x~y) and (\NULLQ~x)
- are similar. Constants do not generate any liveness.
-
-For  the  case  of a  user-function  call  we  use a  third  parameter
-$\Lfonly$  that  represents the  summaries  of  all  functions in  the
-program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
-specific function $f$), expresses how the demand $\sigma$ on a call to
-$f$  is  transformed  into  the  liveness of  its  parameters  at  the
-beginning of the call.  $\Lfonly$  is determined by the judgement form
-$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
-This  rule  describes the  fixed-point  property  to  be satisfied  by
-$\Lfonly$, namely, the demand transformation assumed for each function
-in  the  program should  be  the  same  as the  demand  transformation
-calculated    from    its    body.      As    we    shall    see    in
-Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar  and the  the  language  generated by  this  grammar is  least
-solution satisfying  the rule. We  prefer the least solution  since it
-ensures the safe collection of the greatest amount of garbage.
-
-We next  describe the function $\mathcal{L}$  that propagates liveness
-across expressions.  Note that,  unlike eager languages, the syntactic
-occurrence  of an expression  in a  program denotes  the point  of
-creation of its closure, not necessarily its evaluation.
-To see the  consequences of this, consider
-Figure~\ref{fig:mot-example2} which shows an analysis  of the body of
-a function $\length$ with a demand $\sigma$.  Clearly, this is also the
-demand on the return value $z$ giving the liveness at the program point
-$\pi_8$ as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated
-backwards as in traditional liveness analysis.
-
-Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
-\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
-generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
-the following observations:
-\begin{enumerate}
-\item While the liveness of \py\  should be killed at $\pi_6$ since it
-  is  beyond the scope  of the  definition of  \py, we  do not  do so.
-  However, this  does not cause any  imprecision in the way  of the GC
-  marking more cells as live,  since, during execution, \py\ is yet to
-  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
-  consulted by the garbage collector at this point.
-\item In an eager language, \py\ would not have been live at $\pi_8$,
-  a program point  beyond its last use.  In  a lazy language, however,
-  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
-  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
-  evaluation  actually takes  place while  evaluating $(\SRETURN~\pz)$.
-  We therefore  record  the liveness of $\py$ at the program
-  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
-  the beginning  of the program $\pi_1$ (since  we do not  kill
-liveness).
-\end{enumerate}
-
-We thus regard  the body of the function as a  tree and consider paths
-from  the  root   of  this  tree  to  program   points  which  trigger
-evaluation\footnote{Program   points  which  trigger   evaluation  are
-  labelled  with   $\psi$  }---\SIF\ conditions  and \SRETURN\
-expressions.   We call such
-paths {\em e-paths}.   For the function \length, there  are three such
-paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
-\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
-\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
 
@@
 
+Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
+parts. The function $\mathit{ref}$,  takes  an application $s$ and
+a demand $\sigma$, and returns  the incremental liveness generated for the
+free  variables   of  $s$  due  to  the   application.   The  function
+$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
+expressions.
+
+ Since  in a  lazy language,  an  expression is  not evaluated  unless
+ required, the null demand ($\emptyset$) does not generate liveness in
+ any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
+ non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed  to the
+ liveness $\{x.\epsilon, x.\acdr\sigma\}$.   In an opposite sense, the
+ demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
+ demand  $\sigma$ on  $z$.   Note  that \CONS\  does  not, by  itself,
+ dereference its  arguments. The rules for  (\PRIM~x~y) and (\NULLQ~x)
+ are similar. Constants do not generate any liveness.
+
+For  the  case  of a  user-function  call  we  use a  third  parameter
+$\Lfonly$  that  represents the  summaries  of  all  functions in  the
+program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
+specific function $f$), expresses how the demand $\sigma$ on a call to
+$f$  is  transformed  into  the  liveness of  its  parameters  at  the
+beginning of the call.  $\Lfonly$  is determined by the judgement form
+$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
+This  rule  describes the  fixed-point  property  to  be satisfied  by
+$\Lfonly$, namely, the demand transformation assumed for each function
+in  the  program should  be  the  same  as the  demand  transformation
+calculated    from    its    body.      As    we    shall    see    in
+Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
+grammar  and the  the  language  generated by  this  grammar is  least
+solution satisfying  the rule. We  prefer the least solution  since it
+ensures the safe collection of the greatest amount of garbage.
+
+We next  describe the function $\mathcal{L}$  that propagates liveness
+across expressions.  Note that,  unlike eager languages, the syntactic
+occurrence  of an expression  in a  program denotes  the point  of
+creation of its closure, not necessarily its evaluation.
+To see the  consequences of this, consider
+Figure~\ref{fig:mot-example2} which shows an analysis  of the body of
+a function $\length$ with a demand $\sigma$.  Clearly, this is also the
+demand on the return value $z$ giving the liveness at the program point
+$\pi_8$ as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated
+backwards as in traditional liveness analysis.
+
+Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
+\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
+generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
+the following observations:
+\begin{enumerate}
+\item While the liveness of \py\  should be killed at $\pi_6$ since it
+  is  beyond the scope  of the  definition of  \py, we  do not  do so.
+  However, this  does not cause any  imprecision in the way  of the GC
+  marking more cells as live,  since, during execution, \py\ is yet to
+  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
+  consulted by the garbage collector at this point.
+\item In an eager language, \py\ would not have been live at $\pi_8$,
+  a program point  beyond its last use.  In  a lazy language, however,
+  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
+  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
+  evaluation  actually takes  place while  evaluating $(\SRETURN~\pz)$.
+  We therefore  record  the liveness of $\py$ at the program
+  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
+  the beginning  of the program $\pi_1$ (since  we do not  kill
+liveness).
+\end{enumerate}
+
+We thus regard  the body of the function as a  tree and consider paths
+from  the  root   of  this  tree  to  program   points  which  trigger
+evaluation\footnote{Program   points  which  trigger   evaluation  are
+  labelled  with   $\psi$  }---\SIF\ conditions  and \SRETURN\
+expressions.   We call such
+paths {\em e-paths}.   For the function \length, there  are three such
+paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
+\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
+\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
+
 Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
@@
 
+\begin{proposition}
+Assume that a liveness computation  based on the original set of rules
+gives the liveness  of the variable $x$ at a  program point $\pi_i$ as
+$\sigma$   (symbolically,   $\Lanv{i}{x}=   \sigma$).   Further,   let
+$\Lanv{i}{x}=  \sigma'$ when the  modified rules  are used  instead of
+\Lfunonly.  Then $\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
+\end{proposition}
+
+For  an explanation  of why  the  proposition holds  for the  modified
+\CONS\  rule,   we  refer   the  reader  to   \cite{asati14lgc}.   The
+proposition also holds for other modified rules for similar reasons.
+
+%Thick red line
+%{\color{red} \medskip \hrule \hrule \hrule  \medskip }
+
+\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
+
+\subsection{Generating Liveness Equations}
+
 \begin{figure*}[t!]
@@
 
-\begin{proposition}
-Assume that a liveness computation  based on the original set of rules
-gives the liveness  of the variable $x$ at a  program point $\pi_i$ as
-$\sigma$   (symbolically,   $\Lanv{i}{x}=   \sigma$).   Further,   let
-$\Lanv{i}{x}=  \sigma'$ when the  modified rules  are used  instead of
-\Lfunonly.  Then $\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
-\end{proposition}
-
-For  an explanation  of why  the  proposition holds  for the  modified
-\CONS\  rule,   we  refer   the  reader  to   \cite{asati14lgc}.   The
-proposition also holds for other modified rules for similar reasons.
-
-%Thick red line
-%{\color{red} \medskip \hrule \hrule \hrule  \medskip }
-
-\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
-
-\subsection{Generating Liveness Equations}
 The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
@@
 \begin{eqnarray*}
-\var{\Lanv{1}{\xl}}  &\rightarrow&   \clazy\sigma_{\length} \mid \acdr
-\var{\Df
-  {\length}{1}}{\clazy\sigma_{\length}} \mid
+  \var{\Lanv{1}{\xl}}  &\rightarrow& 
+  \clazy\sigma_{\length} \mid \acdr \var{\Df {\length}{1}}{\clazy\sigma_{\length}}  \\
+  & & \mid
   \clazy\var{\Df{\length}{1}}{\clazy\sigma_{\length}} \\
@@
 
+\begin{figure}[t!]
+  \begin{tabular}{cc}
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(40,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
+  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+  \ncline{t1}{t2}
+  \putnode{l0}{t1}{7}{2}{\clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-.5]{t1}{t1}
+  \putnode{l1}{t1}{0}{10}{\acdr/\clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-.5]{t2}{t2}
+  \putnode{l2}{t2}{0}{10}{\clazy}
+\end{pspicture}
+&
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(40,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
+  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-.5]{t1}{t1}
+  \putnode{l1}{t1}{0}{10}{\acdr}
+\end{pspicture}
+\\
+(a) & (b) \\
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(40,12) %\psframe(0,-5)(40,12)
+  \putnode{t0}{origin}{5}{0}{\Lanv{9}{\pa}}
+  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
+  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+  \ncline{t1}{t2}
+  \putnode{l0}{t1}{5}{2}{\clazy}
+  \ncline{t2}{t3}
+  \putnode{l1}{t2}{5}{2}{\bcar}
+  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.5]{t3}{t3}
+  \putnode{l2}{t3}{0}{10}{\acdr}
+\end{pspicture}
+&
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(40,12) %\psframe(0,-5)(40,12)
+  \putnode{t0}{origin}{5}{0}{\Lanv{9}{\pa}}
+  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{t3}{t1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+  \ncline{t1}{t3}
+  \putnode{l1}{t1}{5}{2}{\bcar}
+  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.5]{t3}{t3}
+  \putnode{l2}{t3}{0}{10}{\acdr}
+\end{pspicture}
+\\
+(c) & (d)  \end{tabular}
+\\ \\ \\ \\
+\caption{(a) The  grammar rules for $\langle  \Df{\length}{1} \rangle$
+  converted into an automaton, and simplified in (b). (c) and (d) show
+  the   automata  for   \Lanv{9}{\pa}.   }\label{fig:example-automata}
+\figrule
+\end{figure}
+
 Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
@@
 
-\begin{figure}[t!]
-\hspace*{1cm}{\epsfig{file=example-automata.eps, height=2.75cm}}
-\caption{(a) The grammar rules for $\langle \Df{\length}{1'}\rangle$
-converted into
-  an automaton, and simplified in   (b). (c) and (d) show the automata
-  for   \Lanv{9}{\pa}.}\label{fig:example-automata}
-\figrule
-\end{figure}
-
 \section{The Garbage Collection Scheme}
@@
 
-\begin{figure*}[t!]
-\label{fig:algos}
-\hfill
-\subfigure[]{\includegraphics[width=.30\linewidth]{lgc-scheme-1}}
-\hfill
-\subfigure[]{\includegraphics[width=.40\linewidth]{lgc-scheme-2}}
-\hfill
-\caption{(a) LGC with closure collection based on liveness.  (b)
-  LGC with reachability-based closure collection.}
-
-\end{figure*}
-
-Figure~\ref{fig:algos}  shows   a  liveness-based  garbage  collection
+  \SetStartEndCondition{ }{}{}%
+  \SetKwProg{Fn}{procedure}{\string:}{}
+  \SetKwFunction{Range}{range}%%
+  \SetKw{KwTo}{in}\SetKwFor{For}{for}{\string:}{}%
+  \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+  \SetKwFor{While}{while}{:}{fintq}%
+  \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+  \SetKwFunction{Lgc}{$lgc$}%
+  \SetKwFunction{Copy}{$lgcCopy$}%
+  \SetKwFunction{LCopy}{$lgcCopy$}%
+  \SetKwFunction{RCopy}{$rgcCopy$}%
+
+\begin{algorithm}[t]
+  \Fn{\Lgc{}}
+     {
+       \For {each reference $\mathit{ref}$ in root set}
+            {\Copy($\mathit{ref}$)\;}
+            $\mathit{copyReferencesOnPrintStack()}$\;  
+     }
+
+     \Fn{\Copy{$\mathit{ref}$}}
+        {
+          \If{$\mathit{ref}$ is live}
+             {
+               $\mathit{newRef} = $  $\mathit{dupHeapCell(ref})$\;
+               \If{cellType($\mathit{ref}$) is $cons$}
+                  {
+                    {
+                      $\mathit{newCar}  = $  \Copy($\mathit{ref\!.car})$\;
+                      $\mathit{newRef.car}   =  newCar$\;
+                    }
+                    {
+                      $\mathit{newCdr}$ =   \Copy($\mathit{ref.cdr})$\;
+                      $\mathit{newRef.cdr}   = \mathit{newCdr}$\;
+                    }    
+                  }
+               \If {cellType($\mathit{ref}$) is closure}
+                   {
+                     $\mathit{newRef.arg1}$ =  \Copy$\mathit{(cell.arg1)}$\;
+                     $\mathit{newRef.arg2}$ =  \Copy$\mathit{(cell.arg2)}$\;
+                   }
+             }
+        }
+
+        \caption{LGC with closure collection based on liveness.\label{algo:lgc-a}}
+\end{algorithm}
+Algorithm~\ref{algo:lgc-a}  describes  a  liveness-based  garbage  collection
 scheme.   The algorithm  assumes that  checking for  liveness involves
@@
 \subsection{Liveness-based collector with reachability-based closure
-  collection}  Since the first  algorithm has  significant bookkeeping
+  collection}
+\begin{algorithm}[t]
+  \Fn{\Lgc{}}
+     {
+       \For {each reference $\mathit{ref}$ in root set}
+            {copy($\mathit{ref}$)\;}
+            $\mathit{copyReferencesOnPrintStack()}$\;  
+     }
+     \Fn{\LCopy{$\mathit{ref}$}}
+        {
+          \eIf{cellType$(\mathit{ref})$ is $cons$}
+              {
+                \eIf {$\mathit{ref}$ is live and has not been copied using
+                  reachability}
+                     {
+                       $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
+                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
+                       $\mathit{ newRef.car}   = \mathit{ newCar}$\;
+                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.cdr)}$\;
+                       $\mathit{ newRef.cdr}   = \mathit{newCar}$\;
+                     }
+                     { $\mathit{makenull(ref)}$\;}}
+              {
+                \If {cellType$\mathit{(ref)}$ is closure and 
+                  has not been copied}
+                    { 
+                      $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
+                    }
+              }
+        }
+        \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
+\end{algorithm}
+  Since the first  algorithm has  significant bookkeeping
 overheads,  we have  a  second GC  scheme  that not  use liveness  for
@@
 only for \CONS\ cells; we assume everything under a closure to be live
-and copy everything  that is reachable.
+and copy everything  that is reachable. Algorithm~\ref{algo:lgc-b}  describes  this scheme.
 
@@
 &
-{\epsfig{file=fibheap.eps, height=\hgt}}%%%Add knightstour.eps
+{\epsfig{file=sudoku.eps, height=\hgt}}%%%Add knightstour.eps
 \\

### 2015-07-09 paper/paper.tex
figure fixed

@@
 (c) & (d)  \end{tabular}
-\\ \\ \\ \\
 \caption{(a) The  grammar rules for $\langle  \Df{\length}{1} \rangle$
@@
 \footnotesize
-
 \centering
-
-\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-\hline
-{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-  nperm} &  {\sf fibheap} & {\sf knightstour} &
-{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-\hline
-\hline
-{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-\hline
-\end{tabular}}
-
-\centerline{(a)}
-
-\medskip
-
-\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r |
-r | r | r | r | r |}
-\hline
-    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\#
-Touched}
-  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-
-                            &   \multicolumn{2}{c|}{cells per GC}
-                            &   \multicolumn{2}{c|}{cells per GC}
-                            &   \multicolumn{2}{c|}{\#GCs}
-                            &   \multicolumn{2}{c|}{MinHeap}
-                            &   \multicolumn{2}{c|}{Avg. Drag}
-                            &   \multicolumn{2}{c|}{(sec)} \\
-\cline{2-13}
-{Program}    &
-RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC &
-LGC \\
-\hline
-\hline
-    {\sf   sudoku}  &381 &1107 &1268 &281 &169 &159 & 1346  &338 &527 &5
-&.008 & .026 \\
-    {\sf  lcss } & 46522 &51101 &6216 &1363&30&24& 52301  &1701 &5147
-&588 &.007 & .008 \\
-    {\sf   gc\_bench}  &129179 &131067 &1894 &4&34&4& 131071   &6 &16970
- &4 &.13 & .0002 \\
-     {\sf  nperm}  & 47586  &174478 &201585 &60882&710&386& 202597  &37507
-&171878 &76618 &0.250 & .154  \\
-    {\sf  fibheap} &249502  &251525 &5555 &2997&1002&975& 254520  &13558
-&78720 &0 &1.185 & 5.457  \\
-    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225
-&307092 &206729 &82112 &464.902 & 14.124  \\
-    {\sf  treejoin} & 288666  &519943 &297570 &5547&116&81& 525488  &7150
-&212653 &1954 &4.335 & 4.4 \\
-    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579
-&501093 &521826 &39465 &70.314 & 24.811 \\
-    {\sf   lambda}  &205 & 556 &2072&90345 &775 &750&966 & 721  &303 &95
-&.179 &1.01  \\
-%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????
-\\
-%    {\sf  nboyer} & &  & &   & &  & & &  \\
-%    {\sf   circsim?} & &  & &   & & & & &   \\
- \hline
-\end{tabular}}
+%% \scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+%% \hline
+%% {Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
+%%   nperm} &  {\sf fibheap} & {\sf knightstour} &
+%% {\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
+%% \hline
+%% \hline
+%% {Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
+%% {DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
+%% {Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
+%% \hline
+%% \end{tabular}}
+%% \centerline{(a)}
+%% \medskip
+%% \scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r |
+%% r | r | r | r | r |}
+%% \hline
+%%     & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\#
+%% Touched}
+%%   & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
+%% \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
+%%                             &   \multicolumn{2}{c|}{cells per GC}
+%%                             &   \multicolumn{2}{c|}{cells per GC}
+%%                             &   \multicolumn{2}{c|}{\#GCs}
+%%                             &   \multicolumn{2}{c|}{MinHeap}
+%%                             &   \multicolumn{2}{c|}{Avg. Drag}
+%%                             &   \multicolumn{2}{c|}{(sec)} \\
+%% \cline{2-13}
+%% {Program}    &
+%% RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC &
+%% LGC \\
+%% \hline
+%% \hline
+%%     {\sf   sudoku}  &381 &1107 &1268 &281 &169 &159 & 1346  &338 &527 &5
+%% &.008 & .026 \\
+%%     {\sf  lcss } & 46522 &51101 &6216 &1363&30&24& 52301  &1701 &5147
+%% &588 &.007 & .008 \\
+%%     {\sf   gc\_bench}  &129179 &131067 &1894 &4&34&4& 131071   &6 &16970
+%%  &4 &.13 & .0002 \\
+%%      {\sf  nperm}  & 47586  &174478 &201585 &60882&710&386& 202597  &37507
+%% &171878 &76618 &0.250 & .154  \\
+%%     {\sf  fibheap} &249502  &251525 &5555 &2997&1002&975& 254520  &13558
+%% &78720 &0 &1.185 & 5.457  \\
+%%     {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225
+%% &307092 &206729 &82112 &464.902 & 14.124  \\
+%%     {\sf  treejoin} & 288666  &519943 &297570 &5547&116&81& 525488  &7150
+%% &212653 &1954 &4.335 & 4.4 \\
+%%     {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579
+%% &501093 &521826 &39465 &70.314 & 24.811 \\
+%%     {\sf   lambda}  &205 & 556 &2072&90345 &775 &750&966 & 721  &303 &95
+%% &.179 &1.01  \\
+%% %    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????
+%% \\
+%% %    {\sf  nboyer} & &  & &   & &  & & &  \\
+%% %    {\sf   circsim?} & &  & &   & & & & &   \\
+%%  \hline
+%% \end{tabular}}
+%% \centerline{(b)}
 
 \include{autotable}
-
-\centerline{(b)}
 \vskip -0.75mm

### 2015-07-09 paper/paper.tex
figures are almost final
Updated conclusions, Future work sections and fixed some issues discussed with Amey

@@
 
-The strongly regular grammar is converted into a set of
-NFAs,   one  for   each  $\var{\Lanv{i}{x}}$.   The  $\hookrightarrow$
-simplification  is now  done  on the  NFAs  by repeatedly  introducing
-$\epsilon$  edges  to  bypass  pairs  of  consecutive  edges  labelled
-\bcar\acar\ or \bcdr\acdr\  and constructing the
-$\epsilon$-closure. Since the $\epsilon$-closure step may create more
-adjacent \bcar\acar\ and \bcdr\acdr\  edges, the simplification is
- continued  till a  fixed point is  reached. The algorithm  for this
-appears  in~\cite{asati14lgc},   Figure[].  Finally  we   replace  the
-$\clazy$  edges  by  $\epsilon$  and determinize  the  automaton  thus
-formed.
+The strongly regular grammar is converted  into a set of NFAs, one for
+each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
+done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
+pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
+constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
+may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
+simplification  is  continued  till  a fixed  point  is  reached.  The
+algorithm for this appears in~\cite{asati14lgc}, Figure[].  Finally we
+replace the  $\clazy$ edges by  $\epsilon$ and check if  the automaton
+has  valid path  (consisting  of only  \acar\ and  \acdr)  to a  final
+state. If it has a valid path,  we remove the \clazy\ edge and mark the
+source node as  final, else we remove the \clazy\  edge without marking
+the source  node as final.   Once the  \clazy\ edges are  eliminated we
+convert the automaton into a deterministic automaton.
 
@@
 effectively implements  the $\hookrightarrow$ simplification  rule for
-\clazy for forward paths. The resulting automaton, made deterministic,
+\clazy\ for forward paths. The resulting automaton, made deterministic,
 is shown in Figure~\ref{fig:example-automata}(b). This expectedly says
@@
         {
-          \If{$\mathit{ref}$ is live}
+          \If{$\mathit{ref}$ is found live using dfa}
              {
@@
        \For {each reference $\mathit{ref}$ in root set}
-            {copy($\mathit{ref}$)\;}
+            {\LCopy($\mathit{ref}$)\;}
             $\mathit{copyReferencesOnPrintStack()}$\;  
@@
               {
-                \eIf {$\mathit{ref}$ is live and has not been copied using
+                \eIf {$\mathit{ref}$ is found live using dfa and has not been copied using
                   reachability}
@@
                        $\mathit{ newRef.car}   = \mathit{ newCar}$\;
-                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.cdr)}$\;
-                       $\mathit{ newRef.cdr}   = \mathit{newCar}$\;
+                       $\mathit{ newCdr}  =$  \LCopy$\mathit{(ref.cdr)}$\;
+                       $\mathit{ newRef.cdr}   = \mathit{newCdr}$\;
                      }
@@
 
-An advantage of  using reachablity for closures it  avoids the problem
-of  multiple  explorations  from  the  same  cell.  This  is  done  by
-maintaining  a flag in  each cell  to indicate  whether it  was copied
-using  reachability.
+Using  reachablity  for  closures   avoids  the  problem  of  multiple
+explorations from the  same cell.  This is done by  maintaining a flag
+in each cell to indicate whether it was copied using reachability.
 
@@
 \subsection{Copying closure using liveness vs Copying closures using reachability}
-To compare the effect of using reachability to copy closures instead of liveness we take a couple of programs from our benchmark suite and execute them with 
-both. As shown in figure-add-figure-here, the number of garbage collections 
-in both cases are same but the time taken varies largely. Both these observations can be explained as follows, since we approximate the liveness of closure to 
-the liveness at its creation point, the liveness we get is usually grossly over approximated. 
-Using this liveness may not give too much of a benefit compared to reachability as nearly everything would be live at its point of creation. 
-As mentioned in section-ref, using liveness for closures may entail multiple traversals over the same structure consuming more time. 
-Therefore we believe that using reachability for closures is a sweet spot in terms of performance for liveness based garbage collectors. 
+To compare the  effect of using reachability to  copy closures instead
+of liveness we take a couple  of programs from our benchmark suite and
+execute them with both. As shown in \cred{figure-add-figure-here}, the
+number of  garbage collections  in both  cases are  same but  the time
+taken  varies largely.  Both these  observations can  be explained  as
+follows, since we approximate the  liveness of closure to the liveness
+at its  creation point, the  liveness we  get is usually  grossly over
+approximated.  Using this liveness may not  give too much of a benefit
+compared to  reachability as  nearly everything would  be live  at its
+point of  creation.  As mentioned  in section-ref, using  liveness for
+closures  may  entail  multiple  traversals over  the  same  structure
+consuming more time.  Therefore we believe that using reachability for
+closures is  a sweet spot in  terms of performance for  liveness based
+garbage collectors.
 %%
@@
 the drag  time.  Our  experiments have  shown that  unless we  use the
-exact  execution  point  for  closures bringing  down  the  number  of
+exact  execution  point  for  closures, bringing  down  the  number  of
 non-live  cells  copied in  closures  cannot  be reduced.   Using  the
@@
 
-\section{Related Work} Previous attempts to  increase  the space
+\section{Related Work} 
+\cred{Need to rewrite this entire section need to probably mention some of these 
+papers at least i found by googling liveness based GC,\\
+http://www.cs.hmc.edu/~oneill/papers/Simplifiers-MSPC.pdf\\
+http://costa.ls.fi.upm.es/papers/costa/AlbertGGZ12.pdf\\
+http://hirzels.com/martin/papers/ecoop01-gc-liveness.pdf\\
+}
+Previous attempts to  increase  the space
 efficiency
@@
 \label{sec:conclusion}
-We  have  defined  a  notion  of liveness  on  structured  data;  this
-generalizes classical liveness and strong liveness.  We started with a
-general fully context-sensitive analysis  which we proved correct with
-respect to  a minefield semantics  (this models the effect  of garbage
-collection between every evaluation step).
-
-To  avoid scalability  issues (and  to  avoid performing  part of  the
-liveness computation at run time)  we defined an 0-CFA version of this
-liveness analysis  in which  demands for function  $f$ at  all calling
-contexts are conflated into  a single demand $\sigma_f$.  This enabled
-us to treat the liveness equations symbolically obtaining context-free
-grammars for liveness at each GC point (calls to user functions and to
-$\CONS$).  These were then converted to DFAs for run-time consultation
-by  the garbage collector.  Experiments confirm  the precision  of the
-analysis.
-
-To obtain performance figures we compared a reachability-based garbage
-collector with a liveness-based  collector.  This showed a decrease in
-the  number  of  GCs,   more  garbage  collected  per  invocation.   A
-significant benefit of LGC is  that programs can run in smaller memory
-when compared to  RGC. This is potentially useful  in situations where
-memory  is  limited---as with  embedded  systems.  For  a majority  of
-programs, the garbage collection times were reduced.
-
-One issue we  highlighted was that while fewer  nodes were marked (and
-more garbage collected), sometimes  cons cells could be visited and
-traversed  multiple times  with different  sets of  liveness  paths to
-explore.  Further work includes  improvements to the classical copying
-collector to reduce the cost of this.
+%% We  have  defined  a  notion  of liveness  on  structured  data;  this
+%% generalizes classical liveness and strong liveness.  We started with a
+%% general fully context-sensitive analysis  which we proved correct with
+%% respect to  a minefield semantics  (this models the effect  of garbage
+%% collection between every evaluation step).
+
+%% To  avoid scalability  issues (and  to  avoid performing  part of  the
+%% liveness computation at run time)  we defined an 0-CFA version of this
+%% liveness analysis  in which  demands for function  $f$ at  all calling
+%% contexts are conflated into  a single demand $\sigma_f$.  This enabled
+%% us to treat the liveness equations symbolically obtaining context-free
+%% grammars for liveness at each GC point (calls to user functions and to
+%% $\CONS$).  These were then converted to DFAs for run-time consultation
+%% by  the garbage collector.  Experiments confirm  the precision  of the
+%% analysis.
+
+%% To obtain performance figures we compared a reachability-based garbage
+%% collector with a liveness-based  collector.  This showed a decrease in
+%% the  number  of  GCs,   more  garbage  collected  per  invocation.   A
+%% significant benefit of LGC is  that programs can run in smaller memory
+%% when compared to  RGC. This is potentially useful  in situations where
+%% memory  is  limited---as with  embedded  systems.  For  a majority  of
+%% programs, the garbage collection times were reduced.
+
+%% One issue we  highlighted was that while fewer  nodes were marked (and
+%% more garbage collected), sometimes  cons cells could be visited and
+%% traversed  multiple times  with different  sets of  liveness  paths to
+%% explore.  Further work includes  improvements to the classical copying
+%% collector to reduce the cost of this.
+{\color{blue}In our current  work we have extended  the liveness based
+  garbage collection to lazy languages  and show that it has benefits.
+  Our  biggest   contribution  is  our  ability   to  handle  closures
+  (suspended evaluations) during garbage collection.  We show that for
+  a  practical  garbage collector  it  is  not possible  to  implement
+  liveness based collector for  closures efficiently.  By implementing
+  a mixed-mode garbage collection scheme, we were able to collect more
+  cells than reachability based  collectors without paying the penalty
+  of  a full  liveness based  collector.  We  believe this  is a  very
+  practical idea and hence easily implementable.
+
+We  also overcome  the drawback  of  multiple traversals  that a  full
+liveness based  collector has  by using  an extra  bit and  setting it
+whenever  the  cell   is  copied  using  reachability(as   part  of  a
+closure). Our experimental results  have shown that such optimizations
+provide performance benefits with very little loss of precision. Using
+the  liveness  annotations  for  the  print  primitive  also  improves
+performance and subsumes compile  time optimizations such as tail-call
+optimization.
+
+\subsection{Future work}
+Although we  provide an  implementation for  a liveness  based garbage
+collection scheme for lazy language, we  do not provide a formal proof
+of its correctness.  A formal  proof of correctness would describe how
+closures  are  handled  during  garbage  collection  and  liveness  is
+propogated inside closures.
+
+ Another interesting exercise is to narrow
+the gap between the actual liveness  and the perceived liveness of our
+garbage collector. Our  experiments show that a  significant number of
+dead cells get trapped inside  closures. Using strictness analysis and
+eagerly evaluating closures might release  more of these dead cells to
+be  garbage  collected.   Since   our  garbage  collector  can  handle
+closures(suspended evaluations)  another challenge would be  to extend
+it to support higher order programs.
+
+Orthogonally,  we can  improve the  efficiency of  the liveness  based
+garbage  collector so  that  even  when the  decrease  in the  garbage
+collections compared  to reachability is  not much, the  total garbage
+collection is  not bad. Making  the DFA  smaller by using  better data
+structures like Binary Decision Diagrams(BDD), truncating the liveness
+automata  after   certain  depth(k-liveness)  are  some   ideas  worth
+investigating.
+
+ }
 
@@
 %Mention k-liveness, use length example to show 1-liveness is not
-sufficient.
+%sufficient.
 %Mention mixed mode GC,
 % - doing RGC most of the times and switching to LGC in case RGC fails
-to collect any garbage
+%to collect any garbage
 % - doing LGC most of the times but regularly checking the reachability
-of the heap and
+%of the heap and
 % switching to RGC if LGC almost equals RGC

### 2015-07-10 paper/paper.tex
paper review.

@@
 the lack of a statically determinable execution order and the presence
-of  closures   at  run-time  poses  additional   challenges  for  lazy
-languages.   This  requires  changes  both  in  the  earlier  liveness
-analysis and also in the garbage collection scheme required to reclaim
-memory that is not live.
+of closures at run-time pose additional challenges for lazy languages.
+This requires changes  both in the earlier liveness  analysis and also
+in the  garbage collection scheme  required to reclaim memory  that is
+not live.
 
@@
 
-\keywords{Heap Analysis, Memory Management, Garbage Collection}
+\keywords{Heap Analysis, Liveness Analysis, Memory Management, Garbage
+  Collection, Lazy Languages}
 
@@
 studies on Scheme~\cite{karkare06effectiveness} and, more importantly,
-on Haskell~\cite{rojemo96lag}
-programs have shown that  GCs leave uncollected a large
-number of  memory objects that are  reachable but not  live (here {\em
-  live} means the  object can potentially be used by  the program at a
-later stage).  This results in unnecessary memory retention.
+on  Haskell~\cite{rojemo96lag}  programs  have shown  that  GCs  leave
+uncollected a  large number of  memory objects that are  reachable but
+not live (here {\em live} means  the object can potentially be used by
+the program  at a  later stage).  This  results in  unnecessary memory
+retention.
 
@@
 the  demands on  all the  calls to  the function.   The result  of the
-analysis  is the annotation  of each  program point  with finite-state
-automata capturing  the liveness of variables at  the point. Depending
+analysis  is an  annotation of  each program  point with  finite-state
+automata capturing the liveness of  variables at that point. Depending
 on the program point where  garbage collection has been triggered, the
-GC consults  a set of automata  to curtail reachability
-during marking.  This results in  an increase in the garbage reclaimed
-and consequently in fewer garbage collections.
+GC consults a set of  automata to restrict reachability during marking.
+This results in an increase  in the garbage reclaimed and consequently
+in fewer garbage collections.
 
@@
 earlier  technique  is  not  possible  for  lazy  languages,  where  a
-heap-allocated  object may either  be an  unevaluated expression  or a
+heap-allocated   object   may   be  an   unevaluated   expression   or
 closure. Since  data is made  live by  evaluation of closures,  and in
@@
 place cannot  be statically determined, laziness  complicates liveness
-analysis itself.  In  addition, apart from evaluated data,  we want to
+analysis itself.   In addition, apart  from evaluated data,  we extend
 extend liveness-based  collection also to closures---closures  with no
@@
 
-{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
-single
-  generation  copying  collector   confirm  the  expected  performance
-  benefits.   Liveness-based  collection  results  in an  increase  in
-  garbage reclaimed.   As a  consequence, there is  a decrease  in the
-  number of  collections and a  (some quantification) decrease  in the
-  memory size required  to run programs. Besides there  is a reduction
-  in the overall execution time for some programs.}
+{\color {Myblue}Rewrite  after the  experiments: Experiments
+  with  a single  generation  copying collector  confirm the  expected
+  performance  benefits.   Liveness-based  collection  results  in  an
+  increase  in  garbage  reclaimed.   As a  consequence,  there  is  a
+  decrease in  the number of  collections and a  (some quantification)
+  decrease in the memory size  required to run programs. Besides there
+  is a reduction in the overall  execution time for some programs. Our
+  experiments further show that a mixed strategy, where we do liveness
+  based GC  up to the closure  boundary and reachability based  GC for
+  the  arguments (free  variables) of  the closure,  gives us  similar
+  benefits as  a fully liveness based  GC (where the arguments  of the
+  closure are also collected using liveness) at much less overhead (?).
+}
 
@@
 \label{sec:motiv}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%
-%%% MOTIVATING EXAMPLE
-%% \newcommand{\nilfigure}
-%% {\scalebox{0.75}{
-%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-%% \begin{pspicture}(0,0)(1,1)
-%% %\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
-%% subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
-%% \putnode{start}{origin}{0}{0}{}
-%% \putnode{stop}{origin}{10}{10}{}
-%% \ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
-%% \end{pspicture}
-%% }}
-
 \begin{figure}[t!]
@@
 The edges shown by thick arrows are  those which are live at $\pi$.  A
-cell is marked  and preserved during garbage collection  only if it is
-reachable from the  root set through a path of  live edges.  All other
-cells can be  reclaimed.  Thus if a garbage  collection takes place at
-$\pi$  with  the  heap  shown  in  Figure~\ref{fig:mot-example}(b),  a
-liveness-based collector (LGC) will  collect only the cells referenced
-by  $\pz$, $(\CDR~  \pz)$  and (the  cells  constituting) the  closure
-referenced by $(\CDR~(\CDR~\pz))$ need to be marked and preserved.  In
-contrast,  a reachability-based collector  (RGC) would  have preserved
-all cells.
+cell is  preserved during garbage  collection only if it  is reachable
+from the root set  through a path of live edges.   All other cells can
+be reclaimed.  Thus if a garbage  collection takes place at $\pi$ with
+the  heap shown  in Figure~\ref{fig:mot-example}(b),  a liveness-based
+collector  (LGC) will  preserve only  the cells  referenced by  $\pz$,
+$(\CDR~ \pz)$ and  (the cells constituting) the  closure referenced by
+$(\CDR~(\CDR~\pz))$.   In  contrast,  a  reachability-based  collector
+(RGC) would have preserved all cells.
 
@@
 \label{sec:defs}
-%%We let $x$,  $y$, $z$ range over variables,  $f$ over user-functions
-%% and $p$ over primitive functions ($\CONS$, $\PRIM$ etc.).
 Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
@@
 definition  is for  ease  of  exposition---generalization to  multiple
-definitions  does  not add conceptual  difficulties.  We  further
-restrict each variable  in a program to be distinct,  so that no scope
-shadowing   occurs---this  simplifies   proofs   of  soundness.    The
-$\SRETURN$ and \SIF\ expressions trigger evaluation of closures.
+definitions does not add conceptual difficulties.  We further restrict
+each variable in a program to  be distinct, so that no scope shadowing
+occurs---this  simplifies  the  reasoning   about  the  program.   The
+$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
+evaluation of closures.
 
@@
 before expression $e$.
-
-% \paragraph{Inlining.}
-{ \color {Myblue} Relook after proof: In spite of the  ANF restrictions
-it
-  is still possible  to inline non-recursive functions (a  fact we use
-  to  prove the safety  of liveness  analysis).  A  user-function call
-  $(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n)  \,\, \IN\,\, e)$ to a
-  function  defined  (after renaming  its  formals  and  locals to  be
-  disjoint         from         existing         variables)         by
-  $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by a sequence of
-  $\LET$'s of  the form  $z_i\leftarrow (\ID\,\,y_i)$ followed  by the
-  body $e_f$  but with  its $(\SRETURN\,\,w)$ expressions  replaced by
-  $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}
 
@@
 \begin{center}
+\renewcommand{\arraystretch}{1.5}
 \begin{tabular}{|c|c|c|}
@@
 constituents) or  a {\em closure}.   A closure  is a pair  $\langle s,
-\rho\rangle$  in which  $s$  is an  unevaluated  application, and  the
-values of its free variables is given by $\rho$.
+\rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
+maps free variables of $s$ to their respective locations.
 
@@
 reachable from a  variable can be represented by a  set of {\em access
-  paths} i.e.  prefix-closed  strings from $\{\acar,\acdr\}^\ast$.  As
-an  example,  if $x$  is  a  list  with liveness  $\{\epsilon,  \acdr,
-\acdr\acar,  \acdr\acdr, \acdr\acdr\acar\}$, then  future computations
-can only  refer up  to the second  and third  members of $x$.   A {\em
-  liveness  environment} is  a mapping  from variables  to  subsets of
-$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
-writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$ instead of
-$[x     \mapsto\{\epsilon,\acdr,\acdr\acdr\},    y\mapsto\{\epsilon\},
-  z\mapsto\{\}]$.    In  this   notation,  $x   \mapsto  \{\epsilon\}$
-represents access using $x$ itself and $x \mapsto \{\}$ indicates that
-$x$ is dead.  We  associate liveness environments with program points.
-The liveness at $\pi:e$ is the liveness just before executing $e$.
+  paths}     i.e.     prefix-closed     set     of    strings     from
+$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
+$\{\epsilon, \acdr,  \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
+future computations can only refer up  to the second and third members
+of $x$.  A  {\em liveness environment} is a mapping  from variables to
+subsets of $\{\acar,\acdr\}^\ast$,  but often expressed as  a set, for
+example by writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$
+instead       of       $[x       \mapsto\{\epsilon,\acdr,\acdr\acdr\},
+  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $x \mapsto
+\{\epsilon\}$ represents access using $x$  itself and $x \mapsto \{\}$
+indicates that $x$  is dead.  We associate  liveness environments with
+program points.  The  liveness at $\pi:e$ is the  liveness just before
+executing $e$.
 
@@
 of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
-\rbrace$  paths   of  the  result   of  $(\CDR~y)$\footnote{Reproduced
-  from~\cite{asati14lgc}: In  classical strong liveness  analysis, $y$
-  and $z$  are live at the  entry $n: x:=y+z$,  if and only if  $x$ is
-  live  at exit  of $n$.   In our  terminology, the  liveness $\lbrace
-  \epsilon\rbrace $ of $x$ at the  exit from $n$ becomes the demand on
-  $y+z$, and this, in turn generates the liveness $\lbrace y.\epsilon,
-  z.\epsilon \rbrace$ at the entry of $n$.}.
-%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
-%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\
-%% field of the cons cell pointed by
-%% $y$.
-
+\rbrace$  paths  of  the  result of  $(\CDR~y)$.  In  strong  liveness
+analysis       (dual      of       classical      faint       variable
+analysis~\cite{horwitz.faint}),  $y$ and  $z$  are live  at the  entry
+$\pi: x:=y+z$, if  and only if $x$  is live at exit of  $\pi$.  In our
+terminology, the  liveness $\lbrace  \epsilon\rbrace $  of $x$  at the
+exit from $n$ becomes the demand on $y+z$, and this, in turn generates
+the liveness $\lbrace y.\epsilon, z.\epsilon  \rbrace$ at the entry of
+$n$.
+
+{
+\color {Myblue}[I don't think this is valid anymore!!]
 Note that  liveness refers to  access using variables and  fields, and
@@
 locations themselves).
+}
+
+We follow the notations appearing in \cite{asati14lgc}. In particular,
 We use $\sigma$  to range over demands, $\alpha$ to  range over access
@@
 abuse notation to  juxtapose an edge label and a  set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$. A more
-elaborate explanation of these notions appear in \cite{asati14lgc}
+$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
 
@@
 \\[1ex]
-\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{ep}
+\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
 \mapsto
-x.\sigma\rbrack, \mbox{ where $\mathit{ep}$ is a
+x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
   new e-path terminating with $\psi$}\\
@@
         \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
-        \lbrack\lbrace \mathit{ep} \mapsto
+        \lbrack\lbrace \mathit{\epath} \mapsto
 \{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
@@
                  \end{array} \\
-&&\mbox{where $\mathit{ep}$ is a new e-path
+&&\mbox{where $\mathit{\epath}$ is a new e-path
   terminating with $\psi$}\\
 \mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \lambda\mathit{ep}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
+        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
 \mbox{where}~\mathcal{M} =
 \mathcal{L}(e,\sigma,\Lfonly),~\Lv =
-\mathcal{M}(ep),~\mbox{and}~f~\mbox{is the new function}\\
+\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
 && \makebox[0mm]{\hspace*{7cm}
@@
 
-For  the  case  of a  user-function  call  we  use a  third  parameter
-$\Lfonly$  that  represents the  summaries  of  all  functions in  the
-program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
-specific function $f$), expresses how the demand $\sigma$ on a call to
-$f$  is  transformed  into  the  liveness of  its  parameters  at  the
-beginning of the call.  $\Lfonly$  is determined by the judgement form
+For the  case of a  function call we  use a third  parameter $\Lfonly$
+that  represents  the  summaries  of all  functions  in  the  program.
+$\Lfonly_{\mathit  f}$  (the component  of  $\Lfonly$  for a  specific
+function $f$), expresses  how the demand $\sigma$ on a  call to $f$ is
+transformed into  the liveness of  its parameters at the  beginning of
+the   call.    $\Lfonly$  is   determined   by   the  judgement   form
 $\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
@@
 Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar  and the  the  language  generated by  this  grammar is  least
+grammar and  the the language generated  by this grammar is  the least
 solution satisfying  the rule. We  prefer the least solution  since it
@@
   consulted by the garbage collector at this point.
-\item In an eager language, \py\ would not have been live at $\pi_8$,
-  a program point  beyond its last use.  In  a lazy language, however,
-  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
-  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
-  evaluation  actually takes  place while  evaluating $(\SRETURN~\pz)$.
-  We therefore  record  the liveness of $\py$ at the program
-  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
-  the beginning  of the program $\pi_1$ (since  we do not  kill
-liveness).
+\item In an  eager language, the heap location referred  by \py\ would
+  not have been live at $\pi_8$,  a program point beyond its last use.
+  In  a  lazy  language,   however,  the  definition  $\pz  \leftarrow
+  (1+\py)$, does not  result in an evaluation of  $(1+\py)$; instead a
+  closure  is  created.  The  evaluation  actually  takes place  while
+  evaluating $(\SRETURN~\pz)$.   We therefore  record the  liveness of
+  $\py$ at  the program point  $\pi_8$, indeed at every  program point
+  from $\pi_8$ up to the beginning of the program $\pi_1$ (since we do
+  not kill liveness).
 \end{enumerate}
 
-We thus regard  the body of the function as a  tree and consider paths
-from  the  root   of  this  tree  to  program   points  which  trigger
-evaluation\footnote{Program   points  which  trigger   evaluation  are
-  labelled  with   $\psi$  }---\SIF\ conditions  and \SRETURN\
-expressions.   We call such
-paths {\em e-paths}.   For the function \length, there  are three such
-paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
-\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
-\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.
+We thus  regard the  body of  the function as  an expression  tree and
+consider paths from  the root of this tree to  the nodes corresponding
+to the program points  which trigger evaluation---\SIF\ conditions and
+\SRETURN\     expressions.      We     call    such     paths     {\em
+  e-paths}.\footnote{Program  points  which   trigger  evaluation  are
+  labelled with $\psi$. By an abuse of notation, we use the same label
+  $\psi$ to refer  to the e-path terminating at a  point $\psi$.}  For
+example, for the function \length, there  are three such paths $\epath_1 =
+\langle  \pi_1, \pi_2,  \psi_1\rangle$, $\epath_2  = \langle\pi_1,  \pi_2,
+\pi_3, \pi_4, \psi_2\rangle$ and $\epath_3  = \langle \pi_1, \pi_2, \pi_5,
+\pi_6, \pi_7, \pi_8, \psi_3\rangle$.
 
@@
 liveness  environments  of the  individual  e-paths.   In the  example
-program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
-ep_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(ep_1)  \cup {\cal
-  M}(ep_2) \cup  {\cal M}(ep_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
+program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  \epath_1, \epath_2,
+\epath_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(\epath_1)  \cup {\cal
+  M}(\epath_2) \cup  {\cal M}(\epath_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
 \sigma, \Lfonly)$.  In the  sequel we shall only define $\mathcal{L}$.
@@
 The  expression $(\SRETURN~\psi:x)$  with a  demand $\sigma$  forces an
-evaluation  of   $x$  and  hence  a  new   map  $[\mathit{ep}  \mapsto
-  \{x.\sigma\}] $  needs to  be created, where the $ep$ is a fresh
+evaluation  of   $x$  and  hence  a  new   map  $[\mathit{\epath}  \mapsto
+  \{x.\sigma\}] $  needs to  be created, where the $\epath$ is a fresh
 evaluation path terminating in $\psi$.  The  map for  the expression
@@
 $e_2$. In addition, since the condition also triggers an evaluation of
-$x$,  the map  $[\mathit{ep} \mapsto  \{x.\epsilon\}]$ is  created and
+$x$,  the map  $[\mathit{\epath} \mapsto  \{x.\epsilon\}]$ is  created and
 added  to the  union.  Also  notice a  consequences  of laziness---the
@@
         \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
-        &[\,\mathit{ep} \mapsto  x.\clazy \sigma]
+        &[\,\mathit{\epath} \mapsto  x.\clazy \sigma]
 \end{align*}
@@
         \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
-              [\;ep_1 \mapsto \lbrace \px.\clazy \sigma,
-\xl.\clazy\sigma\rbrace ,~ep_2 \mapsto \lbrace \cred{
+              [\;\epath_1 \mapsto \lbrace \px.\clazy \sigma,
+\xl.\clazy\sigma\rbrace ,~\epath_2 \mapsto \lbrace \cred{
 \pv.\sigma}  \rbrace,\\
- &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma,
+ &&          \;\;  \epath_3 \mapsto \lbrace \pz.\sigma,
                 \py.\clazy\sigma,
@@
             \cal{P_{\mbox{\tiny \length}}} &=&
-                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
-                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
+                [\;\pi_1 \mapsto \lbrace \epath_1, \epath_2, \epath_3 \rbrace,  \
+                 \pi_2 \mapsto \lbrace \epath_1, \epath_2, \epath_3
                   \rbrace, \\
-&&\;\;  \pi_3 \mapsto \lbrace ep_2
-                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
-                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
- &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
+&&\;\;  \pi_3 \mapsto \lbrace \epath_2
+                  \rbrace ,  \pi_4 \mapsto \lbrace \epath_2
+                  \rbrace, \pi_5 \mapsto \lbrace  \epath_3 \rbrace, \\
+ &&  \;\;                        \pi_6 \mapsto \lbrace \epath_3 \rbrace,
 \pi_7 \mapsto
-                   \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
+                   \lbrace \epath_3\rbrace, \pi_8 \mapsto \lbrace \epath_3
 \rbrace~\rbrack\\
@@
                 [\;
-                  ep_4 \mapsto \lbrace \cred {\pw.\sigma,
+                  \epath_4 \mapsto \lbrace \cred {\pw.\sigma,
                                        \pc.\Lf{\length}{\mbox{1}}{\sigma},}\\
@@
                \cal{P_{\mbox{\tiny \main}}} &=&
-                  \lbrack~ \cred{\pi_9 \mapsto \lbrace ep_4 \rbrace,  \
-                    \pi_{10} \mapsto \lbrace ep_4 \rbrace,  \
-                   \pi_{11} \mapsto \lbrace ep_4 \rbrace,} \\
-                   && \;\; \cred{\pi_{12} \mapsto \lbrace ep_4 \rbrace,
-\pi_{13} \mapsto \lbrace ep_4 \rbrace}~\rbrack
+                  \lbrack~ \cred{\pi_9 \mapsto \lbrace \epath_4 \rbrace,  \
+                    \pi_{10} \mapsto \lbrace \epath_4 \rbrace,  \
+                   \pi_{11} \mapsto \lbrace \epath_4 \rbrace,} \\
+                   && \;\; \cred{\pi_{12} \mapsto \lbrace \epath_4 \rbrace,
+\pi_{13} \mapsto \lbrace \epath_4 \rbrace}~\rbrack
 \end{eqnarray*}
@@
 \begin{figure*}[t]
-\footnotesize
+%\small
 \centering

### 2015-07-10 paper/paper.tex
minor rewrite.

@@
 consider paths from  the root of this tree to  the nodes corresponding
-to the program points  which trigger evaluation---\SIF\ conditions and
-\SRETURN\     expressions.      We     call    such     paths     {\em
-  e-paths}.\footnote{Program  points  which   trigger  evaluation  are
-  labelled with $\psi$. By an abuse of notation, we use the same label
-  $\psi$ to refer  to the e-path terminating at a  point $\psi$.}  For
-example, for the function \length, there  are three such paths $\epath_1 =
-\langle  \pi_1, \pi_2,  \psi_1\rangle$, $\epath_2  = \langle\pi_1,  \pi_2,
-\pi_3, \pi_4, \psi_2\rangle$ and $\epath_3  = \langle \pi_1, \pi_2, \pi_5,
-\pi_6, \pi_7, \pi_8, \psi_3\rangle$.
+to the program points which trigger evaluation\footnote{Program points
+  which      trigger      evaluation       are      labelled      with
+  $\psi$.}---\SIF\ conditions and \SRETURN\ expressions.  We call such
+paths {\em e-paths}.  For example, for the function \length, there are
+three  such paths  $\epath_1 =  \langle \pi_1,  \pi_2, \psi_1\rangle$,
+$\epath_2  = \langle\pi_1,  \pi_2,  \pi_3,  \pi_4, \psi_2\rangle$  and
+$\epath_3  =  \langle  \pi_1,   \pi_2,  \pi_5,  \pi_6,  \pi_7,  \pi_8,
+\psi_3\rangle$.
 

### 2015-07-10 paper/paper.tex
Review up to Section 4.1 over.

@@
 evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
-  \{x.\sigma\}] $  needs to  be created, where the $\epath$ is a fresh
+  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
 evaluation path  terminating in  $\psi$.  The  map for  the expression
@@
 entire  expression including  the condition  is not  evaluated if  the
-demand on it is $\emptyset$.  This results in the empty map, which, by
-abuse of notation, is also denoted by $\emptyset_{\cal{M}}$.
+demand on it is $\emptyset$.  This  results in the empty map, which is
+denoted by $\emptyset_{\cal{M}}$.
 
@@
    in  the definition  of ${\mathit  f}$ represents  non-deterministic
-   choice.  During   liveness  it  can  be  assumed   to  generate  no
+   choice.  During liveness analysis it  can be assumed to generate no
    liveness.}.  We  pass to  the demand  transformer $\Lfonly_{\mathit
@@
 classically) is due to the functional formulation here.
-}%mycomment
+}
 
@@
 The  solution  is  to  also   treat  the  operations  mentioned  above
-symbolically.   We  introduce three  new  symbols.  These symbols  are
-defined as a relation  $\hookrightarrow$ between sets of access paths.
-First, \bcar\ and \bcdr\ are defined as:
+symbolically.  We  introduce three new symbols:  \bcar, \bcdr, \clazy.
+These symbols are defined as a relation $\hookrightarrow$ between sets
+of access paths.  First, \bcar\ and \bcdr\ are defined as:
 
 \begin{align*}
-  &\sigma_1\bcar\sigma_2  \hookrightarrow
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
-&\sigma_1\bcdr\sigma_2  \hookrightarrow
-  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
-  \acdr\sigma_2'
+  &\bcar\sigma \hookrightarrow \sigma' \mbox{ where } \sigma' = \{\alpha \mid \acar\alpha \in \sigma\}\\
+  &\bcdr\sigma \hookrightarrow \sigma' \mbox{ where } \sigma' = \{\alpha \mid \acdr\alpha \in \sigma\}
 \end{align*}
-Thus \bcar, for example, removes the leading \acar\ from $\sigma_2$.
-Similarly, the symbol \clazy\ reduces the string following it to an
-$\epsilon$. \clazy\  is defined by addition to
-$\hookrightarrow$ the rule:
+Thus \bcar\ selects those entries in $\sigma$ that have leading \acar, and removes the leading \acar\ from the.
+The symbol \clazy\ reduces the set of strings following it to a set containing only $\epsilon$. It filters out, however, the empty set of strings.
 \begin{align*}
-  &\sigma_1\clazy\sigma_2  \hookrightarrow
-  \sigma_1 \mbox{ if } \sigma_2 \subseteq  \{\acar,\acdr\}^\ast
+  \clazy\sigma \hookrightarrow & \left\{ 
+  \begin{array}{ll}
+    \emptyset&\mbox{if}~\sigma = \emptyset\\
+    \{\epsilon\} & \mbox{otherwise}
+  \end{array}\right.
 \end{align*}
-%% \begin{align*}
-%%  \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
-%%  \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
-%% \end{align*}
-%% and \clazy\ as:
-%% \begin{align*}
-%%  \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
-%%                        & \{\epsilon\} ~\mbox{otherwise}
-%% \end{align*}
-%% Note that \clazy\ is idempotent.
 We can  now rewrite the \CONS\  and the \CAR\  rules of $\mathit{ref}$
@@
 %% \end{align*}
-The reflexive  transitive closure of  $\hookrightarrow$ will
-be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
-following proposition relates  the original and the modified
-liveness rules.
+When  there  are  multiple  occurences of  \bcar,  \bcdr  and  \clazy,
+$\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
+transitive   closure  of   $\hookrightarrow$   will   be  denoted   as
+$\stackrel{*}{\hookrightarrow}$.   The  following proposition  relates
+the original and the modified liveness rules.
 

### 2015-07-10 paper/paper.tex
typo fixes.

@@
 
-{\color {Myblue}Rewrite  after the  experiments: Experiments
-  with  a single  generation  copying collector  confirm the  expected
-  performance  benefits.   Liveness-based  collection  results  in  an
-  increase  in  garbage  reclaimed.   As a  consequence,  there  is  a
-  decrease in  the number of  collections and a  (some quantification)
-  decrease in the memory size  required to run programs. Besides there
-  is a reduction in the overall  execution time for some programs. Our
-  experiments further show that a mixed strategy, where we do liveness
-  based GC  up to the closure  boundary and reachability based  GC for
-  the  arguments (free  variables) of  the closure,  gives us  similar
-  benefits as  a fully liveness based  GC (where the arguments  of the
-  closure are also collected using liveness) at much less overhead (?).
-}
+{\color  {Myblue}   Experiments  with  a  single   generation  copying
+  collector confirm the expected performance benefits.  Liveness-based
+  collection  results  in an  increase  in  garbage reclaimed.   As  a
+  consequence, there is a decrease in  the number of collections and a
+  (some quantification)  decrease in the  memory size required  to run
+  programs. Besides there is a reduction in the overall execution time
+  for  some  programs.  Our  experiments  further  show that  a  mixed
+  strategy, where we  do liveness based GC up to  the closure boundary
+  and reachability based GC for  the arguments (free variables) of the
+  closure,  gives us  similar benefits  as a  fully liveness  based GC
+  (where  the  arguments  of  the closure  are  also  collected  using
+  liveness)  at  much  less  overhead (?).  [Overhead:  extra  memory,
+    book-keeping during run-time to update, multiple traversal of same
+    set of cells][Refer to Section~\ref{some-sec-in-impl}] }
 
@@
         denotes a closure. Thick edges denote live links. Traversal
-        stops at edges marked $\times$ during garbage collection.
+        stops at edges marked $\times$ during garbage collection for
+        an ideal collector.
       \end{tabular}
@@
 the heap at  the program point $\pi$, i.e. just  before the evaluation
-of $(\length\ \pz)$.  The heap is represented by a graph in which
-a node  either represents atomic values ($\NIL$,  integers etc.), or
-a \CONS\  cell  containing $\CAR$  and  $\CDR$  fields, or a chunk of
-memory representing an unevaluated
-value  also called a  {\em closure}  (represented by  hatched boxes).
-Edges in the graph are  {\em references} and emanate from variables or
-fields.  The figure shows the lists \px\ and \pz\ partially evaluated.
-The evaluation was triggered by the test (\NULLQ~(\CAR~(\CDR~\pz))) in
-the \SIF\ expression.
-
-The edges shown by thick arrows are  those which are live at $\pi$.  A
-cell is  preserved during garbage  collection only if it  is reachable
-from the root set  through a path of live edges.   All other cells can
-be reclaimed.  Thus if a garbage  collection takes place at $\pi$ with
-the  heap shown  in Figure~\ref{fig:mot-example}(b),  a liveness-based
-collector  (LGC) will  preserve only  the cells  referenced by  $\pz$,
-$(\CDR~ \pz)$ and  (the cells constituting) the  closure referenced by
-$(\CDR~(\CDR~\pz))$.   In  contrast,  a  reachability-based  collector
-(RGC) would have preserved all cells.
+of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
+node either  represents atomic  values ($\NIL$,  integers etc.),  or a
+\CONS\ cell containing $\CAR$ and $\CDR$  fields, or a chunk of memory
+representing  an  unevaluated  value   also  called  a  {\em  closure}
+(represented  by  hatched  boxes).   Edges   in  the  graph  are  {\em
+  references} and emanate from variables  or fields.  The figure shows
+the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
+triggered    by   the    test   (\NULLQ~(\CAR~(\CDR~\pz)))    in   the
+\SIF\ expression.
+
+The edges  shown by thick  arrows are those  which are live  at $\pi$.
+Ideally, a cell should be  preserved during garbage collection only if
+it is reachable from  the root set through a path  of live edges.  All
+other  cells can  be reclaimed.   Thus if  a garbage  collection takes
+place at $\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b),
+an ideal liveness-based  collector (LGC) will preserve  only the cells
+referenced by  $\pz$, $(\CDR~ \pz)$  and (the cells  constituting) the
+closure   referenced   by   $(\CDR~(\CDR~\pz))$.    In   contrast,   a
+reachability-based collector (RGC) would have preserved all cells.
 
@@
 $n$.
-
-{
-\color {Myblue}[I don't think this is valid anymore!!]
-Note that  liveness refers to  access using variables and  fields, and
-not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
-locations themselves).
-}
 
@@
  demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
- demand  $\sigma$ on  $z$.   Note  that \CONS\  does  not, by  itself,
- dereference its  arguments. The rules for  (\PRIM~x~y) and (\NULLQ~x)
- are similar. Constants do not generate any liveness.
+ demand  $\sigma$  on $z$.   Since  \CONS\  does not  dereference  its
+ arguments, there  is no $\epsilon$ demand  on $y$ and $z$.  The rules
+ for (\PRIM~x~y) and (\NULLQ~x) are similar. Constants do not generate
+ any liveness.
 
@@
 $\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
-analysis) can be safely approximated by a {\em procedure summary}. for
+analysis) can be safely approximated  by a {\em procedure summary} for
 each function.   The summary is  in the  form of a  demand transformer

### 2015-07-10 paper/paper.tex
proof marked?
review done up to 4.1 end

@@
 
+{\color{red}
 \subsection{Correctness of liveness analysis}
@@
 
+ }
+
 \section{Computing liveness and its encoding as DFA}\label{sec:computing}
@@
 function.
-
+%%
 To  answer these  questions, it  is convenient  to cast  the equations
 arising  out of  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
-grammar.   But first  we modify  the rules  themselves to  a different
-form.
+grammar.   To do  so, we  need  to modify  the rules  themselves to  a
+different form.
 
@@
 These symbols are defined as a relation $\hookrightarrow$ between sets
-of access paths.  First, \bcar\ and \bcdr\ are defined as:
-
+of access paths:
 \begin{align*}
@@
 \end{align*}
-Thus \bcar\ selects those entries in $\sigma$ that have leading \acar, and removes the leading \acar\ from the.
+Thus \bcar\ selects those entries in $\sigma$ that have leading \acar, and removes the leading \acar\ from them.
 The symbol \clazy\ reduces the set of strings following it to a set containing only $\epsilon$. It filters out, however, the empty set of strings.
@@
 The rules for  \CDR, \PRIM\ and \NULLQ\ are also  modified in a manner
-similar to \CAR.  %% The definitions  of \bcar\, \bcdr\ and \clazy\ induce
-%% the following relation $\hookrightarrow$ over sets of access paths:
-%% \begin{align*}
-%%   &\sigma_1\bcar\sigma_2  \hookrightarrow
-%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
-%% &\sigma_1\bcdr\sigma_2  \hookrightarrow
-%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
-%%   \acdr\sigma_2',~\mbox{and}\\
-%% &\sigma_1\clazy \sigma_2 \hookrightarrow \sigma_1, \mbox{ if }
-%% \sigma_2 \ne
-%%   \emptyset
-%% \end{align*}
-When  there  are  multiple  occurences of  \bcar,  \bcdr  and  \clazy,
+similar to \CAR.
+
+When  there  are  multiple  occurences of  \bcar,  \bcdr\  and  \clazy,
 $\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
@@
 the original and the modified liveness rules.
-
 \begin{proposition}
@@
 
-%Thick red line
-%{\color{red} \medskip \hrule \hrule \hrule  \medskip }
-
 \newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
-
 \subsection{Generating Liveness Equations}

### 2015-07-10 paper/paper.tex
changed CFG nonterminal \sigma.

@@
 	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ \cred{(\LET\ \pc}\
-$\cred{\leftarrow  (\CONS\ \pb\ ())}$ \IN
+$\cred{\leftarrow  (\CONS\ \pb\ \NIL)}$ \IN
           \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
@@
 
-The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
-example.   Given a  function $\mathit{f}$,  we now  describe  how to
-generate    equations    for     the    demand    transformation
-\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
-determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
-consider $\Lanv{1}{x_i}$,  the  liveness of    the
-$i^{\mbox{\footnotesize  th}}$
+Given  a  function  $\mathit{f}$,  we now  describe  how  to  generate
+equations  for the  demand  transformation \Lfonly$_\mathit{f}$.   The
+program in Figure~\ref{fig:mot-example2} serves  as a running example.
+Starting   with    a   symbolic   demand   $\sigma$,    we   determine
+\Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In  particular, we  consider
+$\Lanv{1}{x_i}$,  the liveness  of  the $i^{\mbox{\footnotesize  th}}$
 parameter   $x_i$  at   the  program   point  at   the  beginning   of
 $e_{\mathit{f}}$   (assumed  to   be  $\pi_1$).   By  the   rule  {\sc
-  live-define}, this should be the
-same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
+  live-define},      this     should      be      the     same      as
+$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
 \begin{eqnarray*}
@@
 calculate an over-approximation of the  actual demands made by all the
-calls by a {\em summary demand}  and calculate the liveness at each GC
-point inside the  function based on this summary  demand.  Even though
-we  lose precision  due to  this, our  experiments show  that  this is
-acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
+calls and calculate the liveness at each GC
+point inside the  function based on this approximation.
+The  0-CFA-style  {\em
   summary demand}  is calculated by taking  a union of  the demands at
@@
  \\
-\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\epsilon
+\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
 \end{align*}
@@
 \begin{enumerate}
-\item   For    each   function   $\mathit{f}$,    equations   defining
-\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item  For  each   function  $\mathit{f}$, an  equation  defining
-the summary demand  $\sigma_{\mathit f}$ on $e_f$.
+\item For each function $\mathit{f}$, equations defining \Df{f}{i} for
+  use by \Lfonly$_{\mathit f}$.
+\item For each function $\mathit{f}$, an equation defining the summary
+  demand $\sigma_{\mathit f}$ on $e_f$.
 \item  For  each  function   $\mathit{f}$  (including  $\mainpgm$  for
-  $e_\mainpgm$)
-an equation defining
-  liveness at each GC point of $e_{\mathit f}$.
+  $e_\mainpgm$)  an equation  defining liveness  at each  GC point  of
+  $e_{\mathit f}$.
 \end{enumerate}
@@
   \var{\Lanv{1}{\xl}}  &\rightarrow& 
-  \clazy\sigma_{\length} \mid \acdr \var{\Df {\length}{1}}{\clazy\sigma_{\length}}  \\
+  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
   & & \mid
-  \clazy\var{\Df{\length}{1}}{\clazy\sigma_{\length}} \\
+  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
   \var{\Df{\length}{1}} &\rightarrow& \clazy \mid

### 2015-07-10 paper/paper.tex
forward access path defined.

@@
 Other equations can be converted similarly.  The language generated by
-\var{\Lanv{i}{x}},  denoted $\mathscr{L}(\var{\Lanv{i}{x}}) $,  is the
-desired  solution of  \Lanv{i}{x}.  However,  recall from  our earlier
-discussion that the decision problem  that we are interested in during
+\var{\Lanv{i}{x}},  denoted  $\mathscr{L}(\var{\Lanv{i}{x}})$, is  the
+desired solution  of \Lanv{i}{x}.  However,  note that the  strings in
+the  language are  over alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,
+\clazy \rbrace$, but  we want {\em forward} access paths  only, i. e.,
+access paths over the alphabet $\lbrace \acar, \acdr \rbrace$. In
+other words,  the decision  problem that we  are interested  in during
 garbage collection at a program point $\pi_i$ is:

### 2015-07-10 paper/paper.tex
automata fixed.
fig 6 fixed
CFG->NFA->DFA

@@
 &\rightarrow&
-\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
+\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
+\var{\Lanv{9}{\pa}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
 \end{eqnarray*}
@@
 may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
-simplification  is  continued  till  a fixed  point  is  reached.  The
-algorithm for this appears in~\cite{asati14lgc}, Figure[].  Finally we
-replace the  $\clazy$ edges by  $\epsilon$ and check if  the automaton
-has  valid path  (consisting  of only  \acar\ and  \acdr)  to a  final
-state. If it has a valid path,  we remove the \clazy\ edge and mark the
-source node as  final, else we remove the \clazy\  edge without marking
-the source  node as final.   Once the  \clazy\ edges are  eliminated we
-convert the automaton into a deterministic automaton.
+simplification is continued till a fixed point is reached, after which
+the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
+algorithm, its correctness and termination proofs are given by Karkare
+et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
+has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
+automaton, for every edge labelled \clazy, we check if the source node
+of the edge has  a valid path to a final final state.  If yes, we mark
+the mark  the source node as  final. Finally, we remove  all the edges
+labelled  \clazy\  and  convert  the automaton  into  a  deterministic
+automaton.
 
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(40,12)
+\begin{pspicture}(0,-5)(50,12)
   \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
@@
   \putnode{l0}{t1}{7}{2}{\clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-.5]{t1}{t1}
-  \putnode{l1}{t1}{0}{10}{\acdr/\clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-.5]{t2}{t2}
+  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{t1}{t1}
+  \putnode{l1}{t1}{0}{10}{\acdr, \clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t2}{t2}
   \putnode{l2}{t2}{0}{10}{\clazy}
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(40,12)
+\begin{pspicture}(0,-5)(30,12)
   \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
-  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
   \psset{arrows=->}
   \ncline{t0}{t1}
-  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-.5]{t1}{t1}
+  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1}
   \putnode{l1}{t1}{0}{10}{\acdr}
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(40,12) %\psframe(0,-5)(40,12)
-  \putnode{t0}{origin}{5}{0}{\Lanv{9}{\pa}}
+\begin{pspicture}(0,-5)(50,12) %\psframe(0,-5)(40,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{9}{\pa}}}
   \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
@@
   \putnode{l1}{t2}{5}{2}{\bcar}
-  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.5]{t3}{t3}
+  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
   \putnode{l2}{t3}{0}{10}{\acdr}
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(40,12) %\psframe(0,-5)(40,12)
-  \putnode{t0}{origin}{5}{0}{\Lanv{9}{\pa}}
+\begin{pspicture}(0,-5)(20,12) %\psframe(0,-5)(40,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{9}{\pa}}}
   \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \putnode{t3}{t1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
   \psset{arrows=->}
   \ncline{t0}{t1}
-  \ncline{t1}{t3}
-  \putnode{l1}{t1}{5}{2}{\bcar}
-  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.5]{t3}{t3}
-  \putnode{l2}{t3}{0}{10}{\acdr}
 \end{pspicture}
@@
 (c) & (d)  \end{tabular}
-\caption{(a) The  grammar rules for $\langle  \Df{\length}{1} \rangle$
+\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
   converted into an automaton, and simplified in (b). (c) and (d) show
-  the   automata  for   \Lanv{9}{\pa}.   }\label{fig:example-automata}
+  the   same  for   \var{\Lanv{9}{\pa}}.   }\label{fig:example-automata}
 \figrule

### 2015-07-10 paper/paper.tex
CFG->NFA->DFA fig fixed

@@
   \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{t1}{t1}
-  \putnode{l1}{t1}{0}{10}{\acdr, \clazy}
+  \putnode{l1}{t1}{0}{8}{\acdr, \clazy}
   \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t2}{t2}
-  \putnode{l2}{t2}{0}{10}{\clazy}
+  \putnode{l2}{t2}{0}{8}{\clazy}
 \end{pspicture}
@@
   \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1}
-  \putnode{l1}{t1}{0}{10}{\acdr}
+  \putnode{l1}{t1}{0}{8}{\acdr}
 \end{pspicture}
@@
   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
-  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+%  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
   \psset{arrows=->}
@@
   \putnode{l0}{t1}{5}{2}{\clazy}
-  \ncline{t2}{t3}
+%  \ncline{t2}{t3}
+%  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
+%  \putnode{l2}{tD1}{0}{10}{\acdr}
+ \putnode{tD1}{t2}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{tD2}{tD1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t2}{tD1}
   \putnode{l1}{t2}{5}{2}{\bcar}
-  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
-  \putnode{l2}{t3}{0}{10}{\acdr}
+  \ncline{tD1}{tD2}
+  \putnode{lD0}{tD1}{5}{2}{\clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{tD1}{tD1}
+  \putnode{lD1}{tD1}{0}{8}{\acdr, \clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{tD2}{tD2}
+  \putnode{lD2}{tD2}{0}{8}{\clazy}
 \end{pspicture}
@@
 \caption{(a) The  grammar rules for \var{\Df{\length}{1}}
-  converted into an automaton, and simplified in (b). (c) and (d) show
-  the   same  for   \var{\Lanv{9}{\pa}}.   }\label{fig:example-automata}
+  converted into an automaton, and simplified in (b). The same for \var{\Lanv{9}{\pa}}
+  is shown in (c) and (d)}\label{fig:example-automata}
 \figrule
@@
 Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
-from the  grammar for  \var{\Df{\length}{1'}}.  The  NFA is
+from the  grammar for  \var{\Df{\length}{1}}.  The  NFA is
 simplified by replacing the edges labeled \clazy\ by $\epsilon$. While

### 2015-07-10 paper/paper.tex
CFG->NFA->DFA fig fixed

@@
 \caption{(a) The  grammar rules for \var{\Df{\length}{1}}
-  converted into an automaton, and simplified in (b). The same for \var{\Lanv{9}{\pa}}
-  is shown in (c) and (d)}\label{fig:example-automata}
+  converted into an automaton, and (b) its DFA. The same for \var{\Lanv{9}{\pa}}
+  are shown in (c) and (d)}\label{fig:example-automata}
 \figrule

### 2015-07-10 paper/paper.tex
CFG->NFA->DFA fig fixed
Setion 4 done

@@
 labelled  \clazy\  and  convert  the automaton  into  a  deterministic
-automaton.
+automaton.    This   effectively  implements   the   $\hookrightarrow$
+simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
+paths. While  checking for liveness  during GC, a  forward access-path
+only if it can reach a final state.
 
@@
   converted into an automaton, and (b) its DFA. The same for \var{\Lanv{9}{\pa}}
-  are shown in (c) and (d)}\label{fig:example-automata}
+  are shown in (c) and (d).}\label{fig:example-automata}
 \figrule
@@
 Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
-from the  grammar for  \var{\Df{\length}{1}}.  The  NFA is
-simplified by replacing the edges labeled \clazy\ by $\epsilon$. While
-checking for liveness during GC, a forward access-path traversing such
-an edge  will be accepted  only if it  can reach a final  state.  This
-effectively implements  the $\hookrightarrow$ simplification  rule for
-\clazy\ for forward paths. The resulting automaton, made deterministic,
-is shown in Figure~\ref{fig:example-automata}(b). This expectedly says
-that  given  a demand  $\epsilon$  on  \length,  the liveness  of  its
-argument             is             $\acdr^{*}$.            Similarly,
-Figure~\ref{fig:example-automata}(c)    shows   the    automaton   for
-\var{\Lanv{9}{\pa}}. The simplified automaton, shown in
+from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
+Figure~\ref{fig:example-automata}(b).  This expectedly says that given
+a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
+$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
+Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
+\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
 Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
-paths. This reflects the lazy nature of our language. \length\  does not
-evaluate its elements and thus \pa\ is dead at $\pi_9$.
+paths.   This  reflects  the  lazy  nature  of  our  language.   Since
+\length\ does not  evaluate the member elements of  the argument list,
+the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
+liveness based GC is triggered beyond $\pi_9$.
 

### 2015-07-10 paper/paper.tex
Moved introduction to cell type upwards
Updated closure processing text

@@
 
-While exploring the heap, the  GC may encounter closures. In RGC mode,
-the  closure will  be explored  as  usual. However,  \cred{  as
-  mentioned earlier}, our liveness analysis predicts the liveness of
-data in its evaluated form. The absence of liveness information of
-closures presents challenges for LGC.
-
-\subsection{A liveness-based garbage collection scheme}
 We shall call  a unit allocatable memory as a {\em  cell}.  A cell can
@@
 cell or terminated through the function $\mathit{makenull}$.
+
+While exploring the heap, the GC  may encounter closures. In RGC mode,
+the  closure will  be  explored as  usual and  copied  by passing  the
+references. Since our liveness analysis  predicts the liveness of data
+in its evaluated form only, closures  present a challenge for LGC. LGC
+will need  to map the  liveness of the  free variables of  the closure
+back  to their  evaluation point.  This  situation can  be handled  in
+multiple ways and hence we present 2 algorithms for handling closures.
+
+\subsection{A liveness-based garbage collection scheme}
 

### 2015-07-10 paper/paper.tex
Updated

@@
 While exploring the heap, the GC  may encounter closures. In RGC mode,
-the  closure will  be  explored as  usual and  copied  by passing  the
-references. Since our liveness analysis  predicts the liveness of data
-in its evaluated form only, closures  present a challenge for LGC. LGC
-will need  to map the  liveness of the  free variables of  the closure
-back  to their  evaluation point.  This  situation can  be handled  in
-multiple ways and hence we present 2 algorithms for handling closures.
+the closure  will be explored  as usual  and copied by  traversing the
+references corresponding to the arguments. Since our liveness analysis
+predicts the  liveness of  data in its  evaluated form  only, closures
+present a challenge for LGC. LGC will  need to map the liveness of the
+free variables of the closure to  their evaluation point.  This is not
+straight  forward as  variables may  escape their  creation scope  and
+needs to  be handled in an  approximate way.  We describe this in detail next.
 
@@
 Algorithm~\ref{algo:lgc-a}  describes  a  liveness-based  garbage  collection
-scheme.   The algorithm  assumes that  checking for  liveness involves
-traversing the  liveness-DFAs stored as  a table and does  not mention
-this explicitly.  Starting  with the root set, each  live reference is
+scheme.    Starting  with the root set, each  live reference is
 explored   using  $\mathit{  lgcCopy}$.    Copying  \CONS\   cells  is
 simple---it  just involves  copying  the cell  itself and  recursively
-copying the \CAR\ and  the \CDR\ fields.  Liveness-based collection of
-closures  is more  complicated since  our analysis  only  computes the
-liveness of evaluated objects and not closures.
+copying the \CAR\ and  the \CDR\ fields.
 

### 2015-07-10 paper/paper.tex
Updated

@@
 \end{algorithm}
-Algorithm~\ref{algo:lgc-a}  describes  a  liveness-based  garbage  collection
-scheme.    Starting  with the root set, each  live reference is
+Algorithm~\ref{algo:lgc-a}   describes    a   liveness-based   garbage
+collection scheme.  Starting with the root set, each live reference is
 explored  using   $\mathit{  lgcCopy}$.    Copying  \CONS\   cells  is
 simple---it  just involves  copying  the cell  itself and  recursively
-copying the \CAR\ and  the \CDR\ fields.
+copying the  \CAR\ and the  \CDR\ fields.   On the other  hand copying
+closures  requires  determining  the  liveness  of  its  arguments  at
+run-time.
 
 Let   us  consider   determining  the   liveness  of   a   closure  at
-runtime. Assume  that the closure  $(app~f~y)$ has a  reference from
+runtime. Consider  a closure  $(app~f~y)$ has a  reference from
 the root  set variable $x$. If $x$  is live,  then we  know that the
-cell $(app~f~y)$ has  to be copied. However, to find out  the
-liveness of $y$, could find out the program point which triggered the
+cell $(app~f~y)$ has  to be copied. To find out  the
+liveness of $y$, we need the program point which triggered the
 evaluation of $(app~f~y)$ and pick up the liveness of $y$ from this

### 2015-07-10 paper/paper.tex
Updated

@@
 Let   us  consider   determining  the   liveness  of   a  closure   at
-runtime. Consider  a closure  $(app~f~y)$ has a  reference from
-the root  set variable $x$. If $x$  is live,  then we  know that the
-cell $(app~f~y)$ has  to be copied. To find out  the
-liveness of $y$, we need the program point which triggered the
-evaluation of $(app~f~y)$ and pick up the liveness of $y$ from this
-program point. However, this is only possible if the activation record
-of the function which created this closure is still on the activation
-stack. This does not always happen since a function could pack a
-closure in a \CONS\ cell and return the \CONS\ cell.
-
-An  alternative  could  be   to  carry  extra  information  with  each
-closure---the {\em free variables in the closure and the program point
-  at  which  it  was  evaluated}.   This  gives  us  precise  liveness
-information  of the  free  variables  of a  closure  but involves  the
-overhead of  updating the program point  information during execution.
-To avoid this  overhead we go for a safe  approximation by storing the
-{\em creation  point of the closure}  and using it  to check liveness.
-This is safe  since the liveness at the creation  point of the closure
-dominates all possible evaluation points  of the closure, but it might
-be imprecise  as the actual  liveness might be  way less than  what is
-found at the creation point.
+runtime. Consider a closure $(app~f~y)$  has a reference from the root
+set  variable  $x$.  If $x$  is  live,  then  we  know that  the  cell
+$(app~f~y)$ has  to be copied.  To copy  the heap structure  rooted at
+$y$, we need the liveness of  $y$ at the program point which triggered
+the evaluation of  $(app~f~y)$. However, this is only  possible if the
+activation record of the function  which created this closure is still
+on the activation stack. This does  not always happen since a function
+could pack a closure in a \CONS\ cell and return the \CONS\ cell.
+
+An alternative could be to  carry with each closure, extra information
+about the program point which triggered its evaluation.  This gives us
+precise liveness  information of the  free variables of a  closure but
+involves the overhead of updating the program point information during
+execution.  To avoid  this overhead we go for a  safe approximation by
+storing the {\em creation point of  the closure} and using it to check
+liveness.  This  is safe since the  liveness at the creation  point of
+the closure dominates  all possible evaluation points  of the closure,
+but it  might be imprecise  as the actual  liveness might be  way less
+than what is found at the creation point.
 

### 2015-07-10 paper/paper.tex
Updated

@@
 first time,  it is  marked.  Explorations starting  from the  cell are
-curtailed if the cell is  visited subsequently due to sharing. This is
-because the collector can assume  that the region of the heap starting
-from  the  cell has  already  been  explored.  Since a  liveness-based
-collector may  reach the cell each  time with a  different state, this
-curtailment cannot  be done in  the case of  LGC. This is  an inherent
-drawback  of a  liveness-based collector.   This can  be  mitigated by
-maintaining a list of liveness  automata states that a particular heap
-cell was traversed with and  ensuring that we do not repeat traversals
-if we reach the cell in one of these states.
+curtailed if the cell is visited subsequently due to sharing.  Since a
+liveness-based collector may reach the  same cell multiple times, each
+time with a different automaton state, this curtailment cannot be done
+in the case  of LGC. This is an inherent  drawback of a liveness-based
+collector.  This  can be mitigated  by maintaining a list  of liveness
+automata states  that a  particular heap cell  was traversed  with and
+avoid repeated traversals if we reach the cell in one of these states.
 

### 2015-07-10 paper/paper.tex
Updated

@@
 avoid repeated traversals if we reach the cell in one of these states.
+However this also involves storage and run-time penalty.
+
+To  avoid  performance  penalty,  we experimented  with  a  mixed-mode
+garbage collection scheme  which uses liveness for  evaluated data and
+reachability for closures.
 
@@
 \end{algorithm}
-  Since the first  algorithm has  significant bookkeeping
-overheads,  we have  a  second GC  scheme  that not  use liveness  for
-copying closures.  In this  scheme, the liveness automata is consulted
-only for \CONS\ cells; we assume everything under a closure to be live
-and copy everything  that is reachable. Algorithm~\ref{algo:lgc-b}  describes  this scheme.
-
-A consequence of copying everything under a closure is that we can end
-up trying to copy things which are not live and corrupt the heap. This
-can happen if a \CONS\  cell copied using liveness during a collection
-subsequently  becomes   part  of  a  closure.    During  a  subsequent
-collection,  the  collector  may  try  to copy  everything  under  the
-\CONS\ cell thus copying heap  locations which might not be valid.  To
-avoid  this, we  should ensure  that any  non-live references  are not
-carried across  GC's.  We  do this  by scanning the  live part  of the
-buffer after every  GC and setting to null any  reference which is not
-live. This  is correct  because any reference  which is dead  during a
-collection will not be dereferenced any further.
-
-Using  reachablity  for  closures   avoids  the  problem  of  multiple
-explorations from the  same cell.  This is done by  maintaining a flag
-in each cell to indicate whether it was copied using reachability.
+  Algorithm~\ref{algo:lgc-b}  describes  a garbage  collection  scheme
+  where the liveness automata is  consulted only for \CONS\ cells; and
+  assumes everything under a closure  to be live and copies everything
+  that is reachable.
+
+A consequence of  copying everything under a closure is  that we might
+reach  a non-live  cell that  has been  reclaimed earlier  by liveness
+based GC.   This can  happen if  a \CONS\  cell having  partially live
+sub-structure  is copied  using liveness  during a  collection and  it
+subsequently  becomes part  of a  live closure.   During a  subsequent
+collection, the collector may try to copy everything under the closure
+including the dead  sub-structure of the \CONS\ cell.   To avoid this,
+we ensure  that any non-live  references are not carried  across GC's.
+We do this by scanning the live  part of the buffer after every GC and
+nullifying any  reference which is  not live. This is  correct because
+any reference which  is dead during a collection is  guaranteed not be
+dereferenced any further by the program.
+
+This mixed scheme avoids the problem of multiple explorations from the
+same  cell.  This  is  done by  maintaining  a flag  in  each cell  to
+indicate whether it was copied using reachability.
 

### 2015-07-10 paper/paper.tex
Updated

@@
 
-This mixed scheme avoids the problem of multiple explorations from the
-same  cell.  This  is  done by  maintaining  a flag  in  each cell  to
-indicate whether it was copied using reachability.
+Another advnatage of  this mixed garbage collection scheme  is that it
+avoids the problem of multiple  explorations from the same cell.  This
+is done by maintaining a flag in  each cell to indicate whether it was
+copied using reachability.
 
@@
 copy them. We extend the liveness analysis to the print function and
-use its results during garbage collection.
+use its result during garbage collection.
 

### 2015-07-10 paper/paper.tex
Updated

@@
 
-Another advnatage of  this mixed garbage collection scheme  is that it
+Another advantage of  this mixed garbage collection scheme  is that it
 avoids the problem of multiple  explorations from the same cell.  This
@@
 
-\section{Results}
+\section{Experimental evaluation}
 %% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
@@
 
-To test the utility of our method we have implemented a garbage-collector 
-for a simple scheme-like first order language. It consists of an interpreter, 
-liveness analyzer and a garbage-collector that can optionally use use reachability or liveness. 
-Our benchmark consists of programs from nofib suite and some external programs manually converted to ANF.
+To   test  the   utility  of   our  method   we  have   implemented  a
+garbage-collector for a simple scheme-like first order language having
+non-strict semantics. It consists of an interpreter, liveness analyzer
+and a  garbage-collector that can  optionally use  reachability or
+liveness.  Our  benchmark consists  of programs  from nofib~\cite{nofib}  suite and
+some external programs manually converted to ANF.
 

### 2015-07-10 paper/paper.tex
Updated

@@
 
-\section{Experimental evaluation}
+\section{Experimental Evaluation}
 %% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
@@
 
-\subsection{Copying closure using liveness vs Copying closures using reachability}
+\subsection{copying closure using liveness vs copying closures using reachability}
 To compare the  effect of using reachability to  copy closures instead

### 2015-07-10 paper/paper.tex
Updated

@@
 \subsection{A liveness-based garbage collection scheme}
-
+\label{sec:live-clo}
   \SetStartEndCondition{ }{}{}%
@@
 
-\subsection{copying closure using liveness vs copying closures using reachability}
+\subsection{Copying closure using liveness vs copying closures using reachability}
 To compare the  effect of using reachability to  copy closures instead
@@
 compared to  reachability as  nearly everything would  be live  at its
-point of  creation.  As mentioned  in section-ref, using  liveness for
+point of  creation.  As mentioned  in Section~\ref{sec:live-clo}, using  liveness for
 closures  may  entail  multiple  traversals over  the  same  structure

### 2015-07-10 paper/paper.tex
Sec 5 done

--- 2015-07-10 paper/paper.tex
@@
 
+\begin{table*}[t!]
+\caption{Experimental results comparing RGC and LGC}
+\label{tab:experimental-results}
+
+\centering
+\include{autotable}
+\end{table*}
 \section{Experimental Evaluation}
+
 %% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
@@
 \subsection{Results}
-%% \begin{figure*}[t]
-%% \centering
-%% \begin{boxedminipage}{.75\textwidth}
-%% {\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\
-%% \noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
-%% \noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
-%%   then} {\bf skip}\\
-%% \noindent \hspace*{.25cm}  {\bf else} {\bf if}
-%% $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR =
-%% \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\
-%% \noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$
-%% {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa},
-%% \mathsf{dfa}(q,1),x.\CDR)$\\
-%% \noindent \hspace*{.35cm}{\bf return} $y$
-%% \end{boxedminipage}
-%% \caption{Function       for       copying       a       root       set
-%%   variable.}\label{algo:LGC-main-loop} \figrule
-
-%% \end{figure*}
-
 \newcommand{\hgt}{3cm}
-\begin{figure*}[t]
+\begin{figure*}[t!]
 \renewcommand{\arraystretch}{.1}
@@
 \end{figure*}
-The  table  in  Figure~\ref{fig:experimental-results} shows  that  LGC
+
+Table~\ref{tab:experimental-results} shows  that  LGC
 improves the  garbage collection time  and efficiency for most  of the
@@
 would be live at the point of creation.
-%% \comment{
-%% The  increased effectiveness  of LGC  over RGC  is also  shown  in the
-%% tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
-%% provides  statistics  regarding the  analysis  itself.  The number  of
-%% states and the analysis times  are within tolerable limits. Precision
-%% of analysis refers to the percentage of dead cells that is collected by
-%% LGC, averaged over all invocations.
-%% The second table shows garbage  collection statistics for RGC and LGC.
-%% LGC  collects larger garbage  per invocation,  drags cells  for lesser
-%% time and requires a smaller heap size ({\em MinHeap}) for program to
-%% run
-%% in comparison with  RGC.
-
-%% There are a couple of  issues of concern.  The garbage collection time
-%% is larger in the case of LGC for some programs. The reason is that the
-%% cost of consulting the liveness DFA may outweigh the combined benefits
-%% of  fewer   garbage  collections   and  fewer  markings   per  garbage
-%% collection.   The  other issue  is  illustrated  by  the program  {\tt
-%%   lambda}    (memory   usage  shown   separately    in
-%% Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
-%% in Figure~\ref{fig:experimental-results}, the  number of
-%% touched cells\footnote{These  are the  cells which are  visited during
-%%   marking phase, often  more than once due to  sharing.} in this
-%% example is much higher for
-%% LGC.  This  increase is due to  excessive sharing among  heap nodes in
-%% this program.  Note  that a node re-visited because  of sharing is not
-%% explored  any   further  during  a  RGC   collection.   However,  this
-%% curtailment cannot happen in  LGC because of the possibility that the
-%% node,  re-visited in a  different liveness  state, may  mark a  set of
-%% cells different from the earlier visit.
-%% }
-\begin{figure*}[t]
-%\small
-\centering
-%% \scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-%% \hline
-%% {Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
-%%   nperm} &  {\sf fibheap} & {\sf knightstour} &
-%% {\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
-%% \hline
-%% \hline
-%% {Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
-%% {DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
-%% {Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
-%% \hline
-%% \end{tabular}}
-%% \centerline{(a)}
-%% \medskip
-%% \scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r |
-%% r | r | r | r | r |}
-%% \hline
-%%     & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\#
-%% Touched}
-%%   & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
-%% \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
-%%                             &   \multicolumn{2}{c|}{cells per GC}
-%%                             &   \multicolumn{2}{c|}{cells per GC}
-%%                             &   \multicolumn{2}{c|}{\#GCs}
-%%                             &   \multicolumn{2}{c|}{MinHeap}
-%%                             &   \multicolumn{2}{c|}{Avg. Drag}
-%%                             &   \multicolumn{2}{c|}{(sec)} \\
-%% \cline{2-13}
-%% {Program}    &
-%% RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC &
-%% LGC \\
-%% \hline
-%% \hline
-%%     {\sf   sudoku}  &381 &1107 &1268 &281 &169 &159 & 1346  &338 &527 &5
-%% &.008 & .026 \\
-%%     {\sf  lcss } & 46522 &51101 &6216 &1363&30&24& 52301  &1701 &5147
-%% &588 &.007 & .008 \\
-%%     {\sf   gc\_bench}  &129179 &131067 &1894 &4&34&4& 131071   &6 &16970
-%%  &4 &.13 & .0002 \\
-%%      {\sf  nperm}  & 47586  &174478 &201585 &60882&710&386& 202597  &37507
-%% &171878 &76618 &0.250 & .154  \\
-%%     {\sf  fibheap} &249502  &251525 &5555 &2997&1002&975& 254520  &13558
-%% &78720 &0 &1.185 & 5.457  \\
-%%     {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225
-%% &307092 &206729 &82112 &464.902 & 14.124  \\
-%%     {\sf  treejoin} & 288666  &519943 &297570 &5547&116&81& 525488  &7150
-%% &212653 &1954 &4.335 & 4.4 \\
-%%     {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579
-%% &501093 &521826 &39465 &70.314 & 24.811 \\
-%%     {\sf   lambda}  &205 & 556 &2072&90345 &775 &750&966 & 721  &303 &95
-%% &.179 &1.01  \\
-%% %    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????
-%% \\
-%% %    {\sf  nboyer} & &  & &   & &  & & &  \\
-%% %    {\sf   circsim?} & &  & &   & & & & &   \\
-%%  \hline
-%% \end{tabular}}
-%% \centerline{(b)}
-
-\include{autotable}
-\vskip -0.75mm
-\caption{Experimental results comparing RGC and LGC. Table (a) gives
-  data related to liveness analysis, and (b) gives garbage collection
-  data. The average drag time is measured in allocated cons cells.}
-\label{fig:experimental-results}
-\normalsize
-\end{figure*}
-
-%% \section{Collecting more garbage can never slow things down}
-%% \label{sec:lgc-always-better}
-
-%% Since garbage collection is  effectively asynchronous to the
-%% allocator
-%% thread, one might  worry as to how {\em  robust} our measurements
-%% are.
-%% For example,  while LGC would,  in general, collect more  garbage
-%% than
-%% RGC in the same heap state,  is it possible that on some programs,
-%% LGC
-%% might do  a larger number  of collections? We  prove a result  to
-%% show
-%% that   this  cannot   happen.    This  result   applies  to
-%% classical
-%% mark-and-sweep and copying garbage  collectors and, we hypothesize,
-%% to
-%% generational collectors too.
-
-%% \begin{lemma}
-%% For the  same mutator,  a liveness-based collector  can never  do
-%% more
-%% garbage collections than a reachability-based collector.
-%% \end{lemma}
-
-%% \begin{proof}Assume, as before,  that time is measured in terms of
-%% the number of
-%%   cons cells allocated. Let us run  two copies of the mutator with
-%% the
-%%   two    different   collectors    but    with   synchronous
-%% memory
-%%   allocations.   The   garbage    collections,   however,   happen
-%%   asynchronously.
-
-%%   To prove the lemma, it is enough to show the truth of the
-%% following
-%% statement: After
-%%  every  LGC invocation,  the count of LGC invocations   is  no
-%% greater
-%% than  RGC invocations.   The  base case  holds  since the  first
-%% invocations of both GCs happen at the same time.  Assume the
-%% statement
-%% to be true after $n$ invocations of LGC.  Since LGC copies a subset
-%% of
-%% reachable cells, its heap would contain no more cells than RGC heap
-%% at
-%% the end  of the $n$th invocation.   Thus either RGC  is invoked next
-%% before
-%% LGC, or LGC and RGC are both invoked next at the same time.  In
-%% either
-%% case, the statement holds after $n+1$ invocations of LGC.
-%% \end{proof}
-
-%% \begin{figure}[h]
-%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
-%% \caption{Memory  usage of  programs. The  solid  lines represents
-%%   number of  live cells and the  dashed lines show  the number of
-%%   reachable cells.  The red and blue  lines represent invocations
-%%   of RGC and  LGC---their heights indicate  the amount of memory
-%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
-%%       paper and are being used as placeholder}}}
-%% \label{fig:memory-usage} \figrule
-%% \end{figure}
-%%\clearpage
 

### 2015-07-10 paper/paper.tex
working on Results

@@
 \end{algorithm}
+\begin{table*}[ht!]
+\caption{Experimental results comparing RGC and LGC}
+\label{tab:experimental-results}
+\centering
+\include{autotable}
+\end{table*}
+
   Algorithm~\ref{algo:lgc-b}  describes  a garbage  collection  scheme
@@
 
-\begin{table*}[t!]
-\caption{Experimental results comparing RGC and LGC}
-\label{tab:experimental-results}
-
-\centering
-\include{autotable}
-\end{table*}
 \section{Experimental Evaluation}

### 2015-07-10 paper/paper.tex
new graph

@@
 \newcommand{\hgt}{3cm}
-\begin{figure*}[t!]
+\begin{figure*}[p]
 \renewcommand{\arraystretch}{.1}
 \begin{tabular}{@{}c@{}c@{}}
-\hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}%, width=4.10cm}}
-&
-{\epsfig{file=lcss.eps, height=\hgt}}%, width=4.10cm}}
-\\
-\hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}%
- &
-{\epsfig{file=nperm.eps, height=\hgt}}%
-\\
-\hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}%
-&
-{\epsfig{file=sudoku.eps, height=\hgt}}%%%Add knightstour.eps
-\\
-\hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}%
-&
-{\epsfig{file=treejoin.eps, height=\hgt}}%
+   \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
+   & {\epsfig{file=lambda_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=gc_bench_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
 \end{tabular}\vskip -10mm

### 2015-07-10 paper/paper.tex
removed unwanted stuff

@@
 
-{\color{red}
+\mycomment{\color{red}
 \subsection{Correctness of liveness analysis}
@@
 Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
-and  {\color  {red}  proved  it  correct  with  reference  to  a  {\em
+and  \mycomment{\color  {red}  proved  it  correct  with  reference  to  a  {\em
     minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
@@
 
+\mycomment{
 \pagebreak
-
 \pagebreak
@@
 %% %\input{lazy-liveness}
-
+}
 \end{document}

### 2015-07-10 paper/paper.tex
concl

@@
 
-\section{Conclusions}
+\section{Conclusions and Future Work}
 \label{sec:conclusion}
@@
 
-\subsection{Future work}
+%%\subsection{Future work}
 Although we  provide an  implementation for  a liveness  based garbage

### 2015-07-10 paper/paper.tex
conclusions ...

@@
 %% collector to reduce the cost of this.
-{\color{blue}In our current  work we have extended  the liveness based
-  garbage collection to lazy languages  and show that it has benefits.
-  Our  biggest   contribution  is  our  ability   to  handle  closures
-  (suspended evaluations) during garbage collection.  We show that for
-  a  practical  garbage collector  it  is  not possible  to  implement
-  liveness based collector for  closures efficiently.  By implementing
-  a mixed-mode garbage collection scheme, we were able to collect more
-  cells than reachability based  collectors without paying the penalty
-  of  a full  liveness based  collector.  We  believe this  is a  very
-  practical idea and hence easily implementable.
-
-We  also overcome  the drawback  of  multiple traversals  that a  full
-liveness based  collector has  by using  an extra  bit and  setting it
-whenever  the  cell   is  copied  using  reachability(as   part  of  a
-closure). Our experimental results  have shown that such optimizations
-provide performance benefits with very little loss of precision. Using
-the  liveness  annotations  for  the  print  primitive  also  improves
-performance and subsumes compile  time optimizations such as tail-call
-optimization.
-
-%%\subsection{Future work}
+{\color{blue}We extended the liveness based garbage collection to lazy
+  languages and shown its benefit  for garbage collection in practical
+  programs. We  defined context  senstive liveness analysis  that uses
+  context independent summaries of functions obtained using a symbolic
+  demand. We show  that obtaining precise solution  to these equations
+  is undecidable in  our formulation, and hence  we safely approximate
+  the result using DFAs. These DFAs are consulted by garbage collector
+  to improve to collection.
+
+  Liveness   of   closure   presented  further   challenges   to   our
+  implementation.   We compared  different approximate  strategies for
+  handling   liveness   of   closures,   and  found   that   a   mixed
+  strategy---reachability  based  collection   of  closure  arguments,
+  liveness  based  collection  of   everything  else---works  best  in
+  practice as it  avoid runtime and space  overheads.  With mixed-mode
+  garbage collection scheme,  we were able to collect  more cells than
+  reachability based collectors at a reasonable overhead.
+
   Although we  provide an  implementation for  a liveness  based garbage
@@
 
- Another interesting exercise is to narrow
-the gap between the actual liveness  and the perceived liveness of our
-garbage collector. Our  experiments show that a  significant number of
-dead cells get trapped inside  closures. Using strictness analysis and
-eagerly evaluating closures might release  more of these dead cells to
-be  garbage  collected.   Since   our  garbage  collector  can  handle
-closures(suspended evaluations)  another challenge would be  to extend
-it to support higher order programs.
-
-Orthogonally,  we can  improve the  efficiency of  the liveness  based
-garbage  collector so  that  even  when the  decrease  in the  garbage
-collections compared  to reachability is  not much, the  total garbage
-collection is  not bad. Making  the DFA  smaller by using  better data
-structures like Binary Decision Diagrams(BDD), truncating the liveness
-automata  after   certain  depth(k-liveness)  are  some   ideas  worth
-investigating.
-
- }
+  Another interesting exercise is to narrow the gap between the actual
+  liveness and  the perceived liveness  of our garbage  collector. Our
+  experiments (not  reported here) show  that a significant  number of
+  dead cells  get trapped inside closures.   Using strictness analysis
+  to eagerly evaluate closures might  release more of these dead cells
+  to be  garbage collected.   Since our  garbage collector  can handle
+  closures  (suspended evaluations),  another  major  challenge is  to
+  extend it to support higher order programs.
+
+  Orthogonally,  we plan  to improve  the efficiency  of the  liveness
+  based garbage collector using heuristics  such as limiting the depth
+  of  DFAs,   merging  nearly-equivalent   states  and   using  better
+  representation  and algorithms  for automata  manipulation. We  also
+  need  to   investigate  the  interaction  of   liveness  with  other
+  collection schemes, such as incremental and generational collection.
+  In short,  we need to  investigate the  ways to make  liveness based
+  garbage collection attractive for practical collectors.  }
 

### 2015-07-10 paper/paper.tex
removed extrenous subsection.
Committing changes

--- 2015-07-10 paper/paper.tex
@@
 
-\mycomment{\color{red}
+{\color{red}
 \subsection{Correctness of liveness analysis}
@@
 Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
-and  \mycomment{\color  {red}  proved  it  correct  with  reference  to  a  {\em
+and  {\color  {red}  proved  it  correct  with  reference  to  a  {\em
     minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
@@
 \section{Related Work} 
-\cred{Need to rewrite this entire section need to probably mention some of these 
-papers at least i found by googling liveness based GC,\\
-http://www.cs.hmc.edu/~oneill/papers/Simplifiers-MSPC.pdf\\
-http://costa.ls.fi.upm.es/papers/costa/AlbertGGZ12.pdf\\
-http://hirzels.com/martin/papers/ecoop01-gc-liveness.pdf\\
-}
-Previous attempts to  increase  the space
-efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
+\cred{Need to  rewrite this  entire section  need to  probably mention
+  some of  these papers at  least i  found by googling  liveness based
+  GC,\\ http://www.cs.hmc.edu/~oneill/papers/Simplifiers-MSPC.pdf\\ http://costa.ls.fi.upm.es/papers/costa/AlbertGGZ12.pdf\\ http://hirzels.com/martin/papers/ecoop01-gc-liveness.pdf\\ }
+Previous  attempts  to increase  the  space  efficiency of  functional
+programs  by  additional  reclamation  of memory  fall  in  two  broad
+categories. In the first, the program itself is instrumented to manage
+reclamation and reallocation without the aid of the garbage collector.
+Such      attempts      include:      sharing      analysis      based
 reallocation~\cite{jones89compile},                      deforestation
@@
 methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
+garbage~\cite{inoue88analysis,lee05static}  by  explicitly  nullifying
+pointers  that are  not  live.  However,  the  nullification, done  at
+compile  time,  requires sharing  (alias)  analysis.   Our method,  in
+contrast, does not require sharing  because of the availability of the
+heap itself  at runtime.  To  the best of  our knowledge, this  is the
+first attempt  at liveness-based  marking of  the heap  during garbage
+collection.
 
@@
 
-\mycomment{
 \pagebreak
+
 \pagebreak
@@
 %% %\input{lazy-liveness}
-}
+
 \end{document}

### 2015-07-10 lazy-scheme/post_process/gen_gc_stats.sh
fast cpp processing

@@ -105,8 +105,8 @@ for file in $STATOUTFILES; do
     rm -f ${file}.out
 done
 
-PPSCRIPT=$DIRNAME/process_fast
 PPSCRIPT=$DIRNAME/process
+PPSCRIPT=$DIRNAME/process_fast
 OUTFILE=general_stats.out
 $PPSCRIPT $STATSFILE  > $OUTFILE
 

### 2015-07-10 lazy-scheme/post_process/gen_stats_and_plotall.sh

@@ -116,11 +116,11 @@ do
     for opt in gc-plain gc-live
     # for opt in gc-plain gc-live gc-freq=${freq[$bm]}
     do 
-     	echo "Processing $bm [$opt]"
-     	$DIR/gen_gc_stats.sh $OPTARG \
-     	    -bmdir ../benchmarks/programs/$bm \
-     	    -gcopt $opt -heap ${limit[$bm]}
-    done
+      	echo "Processing $bm [$opt]"
+      	$DIR/gen_gc_stats.sh $OPTARG \
+      	    -bmdir ../benchmarks/programs/$bm \
+      	    -gcopt $opt -heap ${limit[$bm]}
+     done
     #bash $DIR/val.plot $bm ${xtics[$bm]} ${ytics[$bm]}
     #bash $DIR/window.plot $bm ${xlimit[$bm]} ${ylimit[$bm]} 
     gnuplot $DIR/plots/${bm}.gnu

### 2015-07-10 lazy-scheme/post_process/process_fast.cpp

@@ -1,0 +1,218 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <assert.h>
+#include <vector>
+
+using namespace std;
+
+#define MAXLINE 256
+#define MAXCELL 50000000
+#define MAXGC 100000
+
+static const char *FS="|";
+static long c_total  = 0;
+static long c_dragged = 0;
+static long t_dragged = 0;
+static long last_gc_time = 0;
+static long this_gc_time;
+
+static vector<long> gc_reachable;
+static vector<long> gc_used;
+static vector<long> gc_rgcied;
+static vector<long> gc_times;
+static int  gc_count = 0;
+
+void parse(const char* line);
+int main(int argc, char* argv[]) 
+{
+    int len;
+    size_t n = MAXLINE, ncell = 0;
+    char *line = new char[MAXLINE];
+    const char *filename = (argc == 1) ? "GC_STATS.txt" : argv[1];
+    FILE* statsfile = fopen(filename, "r");
+    
+    while ((len = getline(&line, &n, statsfile)) >0) {
+        ncell++;
+        parse(line);
+    }
+    fclose(statsfile);
+    /* Dump the data */
+    printf("# Total Cells               : %ld\n", c_total);
+    printf("# Dragged Cells             : %ld\n", c_dragged);
+    if (c_dragged > 0)
+    {
+        printf("# Total Drag Time           : %ld\n", t_dragged);
+        printf("# Average Drag Time         : %lf\n", t_dragged*1.0/c_dragged);
+    }
+    
+    FILE* file1 = fopen("rch.out", "w");
+    FILE* file2 = fopen("use.out", "w");
+    FILE* file3 = fopen("gbg.out", "w");
+    FILE* file4 = fopen("rgc.out", "w");
+    unsigned int the_time;    
+    for(the_time = 0; the_time <= this_gc_time;  the_time++) {
+        long rch = gc_reachable[the_time];
+        long use = gc_used[the_time];
+        long rgc = gc_rgcied[the_time];
+        fprintf(file2, "%u %ld\n", the_time, use);
+        if ((the_time > 0) && (rch == 0)) /* ignore */
+            continue;
+        
+        fprintf(file1, "%u %ld\n", the_time, rch);
+        fprintf(file3, "%u %ld\n", the_time, rch-use);
+        // if (rgc > 0) : approx use is rgc + use, else it is use.
[…]

### 2015-07-10 paper/paper.tex
Fixed minor issues
Updated related work

--- 2015-07-10 paper/paper.tex
@@
 \section{Related Work} 
-\cred{Need to rewrite this entire section need to probably mention some of these 
-papers at least i found by googling liveness based GC,\\
-http://www.cs.hmc.edu/~oneill/papers/Simplifiers-MSPC.pdf\\
-http://costa.ls.fi.upm.es/papers/costa/AlbertGGZ12.pdf\\
-http://hirzels.com/martin/papers/ecoop01-gc-liveness.pdf\\
-}
-Previous attempts to  increase  the space
-efficiency
-of functional programs by additional  reclamation of memory fall in two
-broad categories. In the first,  the program itself is instrumented to
-manage  reclamation and reallocation  without the  aid of  the garbage
-collector.    Such    attempts   include:   sharing    analysis   based
-reallocation~\cite{jones89compile},                       deforestation
-techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-methods   that  enable   the   garbage  collector   to  collect   more
-garbage~\cite{inoue88analysis,lee05static}   by
-explicitly  nullifying  pointers  that  are not  live.   However,  the
-nullification,  done   at  compile  time,   requires  sharing  (alias)
-analysis.  Our  method, in contrast, does not  require sharing because
-of the availability of the heap  itself at runtime.  To the best of our
-knowledge, this is the first  attempt at liveness-based marking of the
-heap during garbage collection.
+Although augmenting  garbage collection with liveness  information has
+been  studied earlier,  it was  mainly  in the  context of  imperative
+languages~\cite{Albert,  Hirzel}.  Due  to   the  presence  of  global
+variables  and mutation,  practical utility  of simple  liveness based
+techniques are found to be inefficient.
+
+In the space  of functional languages, improving  space efficiency has
+been  mainly  studied  as  a compile  time  activity.  Either  through
+rewriting                methods                such                as
+deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
+sharing  analysis   based  reallocation~\cite{jones89compile},  region
+based  analysis~\cite{tofte98region} or  through insertion  of compile
+time              nullifying              statements              such
+as~\cite{inoue88analysis,lee05static,Hamilton}.    The  compile   time
+marking approaches all rely on an efficient and precise alias analysis
+and in the absence of it cannot provide notable improvement.
+
+The closest  to our approach  is the liveness based  garbage collector
+implemented in~\cite{asati14lgc}. We extend  their work to handle lazy
+evaluation and closures.
+
+%% A similar approach has been suggested in~\cite{ONeill} where they suggest augmenting garbage collection with extra information using {\it simplifiers}.
+
+%% Previous  attempts  to increase  the  space  efficiency of  functional
+%% programs  by  additional  reclamation  of memory  fall  in  two  broad
+%% categories. In the first, the program itself is instrumented to manage
+%% reclamation and reallocation without the aid of the garbage collector.
+%% Such      attempts      include:      sharing      analysis      based
+%% reallocation~\cite{jones89compile},                      deforestation
+%% techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
+%% methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
+%% analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
+%% methods   that  enable   the   garbage  collector   to  collect   more
+%% garbage~\cite{inoue88analysis,lee05static}  by  explicitly  nullifying
+%% pointers  that are  not  live.  However,  the  nullification, done  at
+%% compile  time,  requires sharing  (alias)  analysis.   Our method,  in
+%% contrast, does not require sharing  because of the availability of the
+%% heap itself  at runtime.  To  the best of  our knowledge, this  is the
+%% first attempt  at liveness-based  marking of  the heap  during garbage
+%% collection.
 
@@
   collection schemes, such as incremental and generational collection.
-  In short,  we need to  investigate the  ways to make  liveness based
+  In  summary, we  need to  investigate  ways to  make liveness  based
   garbage collection attractive for practical collectors.  }
@@
 
-\subsection{References}
 \bibliography{fun_hra}{}

### 2015-07-10 paper/paper.tex
Added a new table comparing liveness vs reachability for closures

@@
 \subsection{Copying closure using liveness vs copying closures using reachability}
+
+\begin{table*}[ht!]
+\caption{Comparing LGC with liveness for closures and reachability for closures}
+\label{tab:compare-live-clo-reach-clo}
+\centering
+\begin{tabular}
+{| l | r | r | r | r |}
+\hline
+Prog Name & Heap size & Num GC & Reachablity for closure & Liveness for closure \\ \hline
+\verb@nqueens(8)@ & 10000 & 183 & 0.0087 & 0.2267\\ \hline
+\verb@lambda@&10000&11&0.0010&4.9922\\ \hline
+\verb@lcss@&10000&1&0.00008&34.6684\\ \hline
+\end{tabular}
+\end{table*}
 To compare the  effect of using reachability to  copy closures instead
 of liveness we take a couple  of programs from our benchmark suite and
-execute them with both. As shown in \cred{figure-add-figure-here}, the
-number of  garbage collections  in both  cases are  same but  the time
-taken  varies largely.  Both these  observations can  be explained  as
-follows, since we approximate the  liveness of closure to the liveness
-at its  creation point, the  liveness we  get is usually  grossly over
-approximated.  Using this liveness may not  give too much of a benefit
-compared to  reachability as  nearly everything would  be live  at its
-point of  creation.  As mentioned  in Section~\ref{sec:live-clo}, using  liveness for
+execute       them      with       both.        As      shown       in
+Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
+collections in both cases are same but copying closures using liveness
+takes a lot more time compared to reachability based copying.
+
+The number  of garbage  collections is same  since we  approximate the
+liveness of  a closure  to the  liveness at  its creation  point.  The
+liveness  we get  is usually  grossly over  approximated.  Using  this
+liveness may not  give too much of a benefit  compared to reachability
+as nearly everything would be live at its point of creation.
+
+As  mentioned   in  Section~\ref{sec:live-clo},  using   liveness  for
 closures  may  entail  multiple  traversals over  the  same  structure
-consuming more time.  Therefore we believe that using reachability for
-closures is  a sweet spot in  terms of performance for  liveness based
-garbage collectors.
+consuming more  time.  The  advantage of avoiding  multiple traversals
+can be clearly seen from the lcss example where a single GC takes less
+than a millisecond for reachability  based closure collection where as
+the  liveness  based collection  is  of  the  order of  seconds.  Lazy
+languages tend  to have a  lot of sharing  and hence using  a liveness
+based collection for closures is impractical.
+
+Therefore we believe  that using reachability for closures  is a sweet
+spot   in   terms   of   performance  for   liveness   based   garbage
+collectors.  For the  remainder  of  the section  all  LGC results  we
+discuss uses  a mixed-mode collector--liveness for  evaluated data and
+reachability for closures.
 %%
@@
 sharing  analysis   based  reallocation~\cite{jones89compile},  region
-based  analysis~\cite{tofte98region} or  through insertion  of compile
+based  analysis~\cite{tofte98region}, or  through insertion  of compile
 time              nullifying              statements              such
@@
   is undecidable in  our formulation, and hence  we safely approximate
-  the result using DFAs. These DFAs are consulted by garbage collector
-  to improve to collection.
+  the result using DFA. These DFA are consulted by the garbage collector
+  to improve  collection.
 
@@
   liveness  based  collection  of   everything  else---works  best  in
-  practice as it  avoid runtime and space  overheads.  With mixed-mode
+  practice as it  avoids runtime and space  overheads.  With mixed-mode
   garbage collection scheme,  we were able to collect  more cells than

### 2015-07-10 paper/paper.tex
almost final plots

--- 2015-07-10 paper/paper.tex
@@
 \renewcommand{\arraystretch}{.1}
-\begin{tabular}{@{}c@{}c@{}}
+\begin{tabular}{@{}c@{}@{}c@{}}
    \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
    & {\epsfig{file=lambda_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
 \\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=gc_bench_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
 \\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
 \\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
-\end{tabular}\vskip -10mm
- \caption{Memory usage  snapshot of  programs.  The  blue and  the red
-   curves indicate the  number of cons cells in  the active semi-space
-   for  LGC and  RGC  respectively.  The  black  curve represents  the
-   number of reachable cells and  the grey curve represents the number
-   of cells that are actually live  (of which liveness analysis does a
-   static approximation).  x-axis is  the time  measured in  number of
+\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+\\ (a) & (b)
+\end{tabular}%\vskip -10mm
+ \caption{Memory  usage  for (a)  full  benchmarks  and (b)  for  some
+   interesting part  of the benchmarks.   The blue and the  red curves
+   indicate the number of cons cells  in the active semi-space for RGC
+   and LGC  respectively.  The  black curve  represents the  number of
+   reachable cells and  the grey curve represents the  number of cells
+   that are  actually live (of  which liveness analysis does  a static
+   approximation).   x-axis   is  the  time  measured   in  number  of
    cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
@@
 \section{Related Work} 
-Although augmenting  garbage collection with liveness  information has
-been  studied earlier,  it was  mainly  in the  context of  imperative
-languages~\cite{Albert,  Hirzel}.  Due  to   the  presence  of  global
-variables  and mutation,  practical utility  of simple  liveness based
-techniques are found to be inefficient.
-
-In the space  of functional languages, improving  space efficiency has
-been  mainly  studied  as  a compile  time  activity.  Either  through
-rewriting                methods                such                as
-deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-sharing  analysis   based  reallocation~\cite{jones89compile},  region
-based  analysis~\cite{tofte98region} or  through insertion  of compile
-time              nullifying              statements              such
-as~\cite{inoue88analysis,lee05static,Hamilton}.    The  compile   time
-marking approaches all rely on an efficient and precise alias analysis
-and in the absence of it cannot provide notable improvement.
-
-The closest  to our approach  is the liveness based  garbage collector
-implemented in~\cite{asati14lgc}. We extend  their work to handle lazy
-evaluation and closures.
-
-%% A similar approach has been suggested in~\cite{ONeill} where they suggest augmenting garbage collection with extra information using {\it simplifiers}.
-
-%% Previous  attempts  to increase  the  space  efficiency of  functional
-%% programs  by  additional  reclamation  of memory  fall  in  two  broad
-%% categories. In the first, the program itself is instrumented to manage
-%% reclamation and reallocation without the aid of the garbage collector.
-%% Such      attempts      include:      sharing      analysis      based
-%% reallocation~\cite{jones89compile},                      deforestation
-%% techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-%% methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-%% analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-%% methods   that  enable   the   garbage  collector   to  collect   more
-%% garbage~\cite{inoue88analysis,lee05static}  by  explicitly  nullifying
-%% pointers  that are  not  live.  However,  the  nullification, done  at
-%% compile  time,  requires sharing  (alias)  analysis.   Our method,  in
-%% contrast, does not require sharing  because of the availability of the
-%% heap itself  at runtime.  To  the best of  our knowledge, this  is the
-%% first attempt  at liveness-based  marking of  the heap  during garbage
-%% collection.
+\cred{Need to  rewrite this  entire section  need to  probably mention
+  some of  these papers at  least i  found by googling  liveness based
+  GC,\\ http://www.cs.hmc.edu/~oneill/papers/Simplifiers-MSPC.pdf\\ http://costa.ls.fi.upm.es/papers/costa/AlbertGGZ12.pdf\\ http://hirzels.com/martin/papers/ecoop01-gc-liveness.pdf\\ }
+Previous  attempts  to increase  the  space  efficiency of  functional
+programs  by  additional  reclamation  of memory  fall  in  two  broad
+categories. In the first, the program itself is instrumented to manage
+reclamation and reallocation without the aid of the garbage collector.
+Such      attempts      include:      sharing      analysis      based
+reallocation~\cite{jones89compile},                      deforestation
+techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
+methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
+analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
+methods   that  enable   the   garbage  collector   to  collect   more
+garbage~\cite{inoue88analysis,lee05static}  by  explicitly  nullifying
+pointers  that are  not  live.  However,  the  nullification, done  at
+compile  time,  requires sharing  (alias)  analysis.   Our method,  in
+contrast, does not require sharing  because of the availability of the
+heap itself  at runtime.  To  the best of  our knowledge, this  is the
+first attempt  at liveness-based  marking of  the heap  during garbage
+collection.
 

### 2015-07-10 lazy-scheme/post_process/createTable.sh

@@ -1,9 +1,8 @@
 #!/bin/sh
 DIR=`dirname $0`
 SCRIPT1=$DIR/touchedAndCopied.awk
-
-#AllBMs="sudoku lcss gc_bench nperm fibheap  knightstour treejoin nqueens "
 AllBMs="lambda nperm treejoin lcss sudoku fibheap nqueens gc_bench"
+AllBMs="lambda nperm treejoin lcss sudoku fibheap nqueens knightstour gc_bench "
 #AllBMs="sudoku"
 
 cat <<EOF
@@ -21,14 +20,15 @@ cat <<EOF
   &   \multicolumn{2}{@{}c@{}|}{cells per GC}
   &   \multicolumn{2}{@{}c@{}|}{\#GCs}
   &   \multicolumn{2}{@{}c@{}|}{Drag Reduction (\%)}
-  &   \multicolumn{2}{@{}c@{}|}{(sec)} & \\\\
+  &   \multicolumn{2}{@{}c@{}|}{(sec)} &Speedup \\\\
 \cline{4-13}
-{Program}&States&Time (sec) &RGC&LGC&RGC&LGC&RGC&LGC&\#Cells&Time&RGC&LGC&Speedup \\\\
+{Program}&States&Time (sec) &RGC&LGC&RGC&LGC&RGC&LGC&\#Cells&\#Allocs&RGC&LGC&RGC/LGC \\\\
 \hline
 \hline
 
 EOF
 format="%s & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f & %.2f"
+
 OUTPUTDIR=output.GOLD
 for BM in $AllBMs
 do
@@ -49,8 +49,16 @@ do
     data3=`echo $tmpData | cut -d' ' -f2`
     data4=`echo $tmpData | cut -d' ' -f3`
     
-    d1=`expr $data1 \* 100 / $data3 - 100`
-    d2=`expr $data2 \* 100 / $data4 - 100`
+    d1=`bc -l << EOF                                                  
+scale=2
+$data1 * 100/$data3 - 100
+EOF
+`
+    d2=`bc -l << EOF                                                  
+scale=2
+$data2 * 100/$data4 - 100
+EOF
+`
     line="$line  $d1  $d2"
 
     #GC time to be filled separately

### 2015-07-10 lazy-scheme/post_process/gen_stats_and_plotall.sh

@@ -52,7 +52,7 @@ limit["sudoku"]=4075
 limit["lcss"]=22243
 limit["lambda"]=20466
 limit["gc_bench"]=204850
-limit["knightstour"]=6778000
+limit["knightstour"]=680000
 ## UNUSED BMs
 limit["fft"]=3300
 limit["small"]=17
@@ -113,8 +113,9 @@ fi
 
 for bm in $BMs
 do
-    for opt in gc-plain gc-live
-    # for opt in gc-plain gc-live gc-freq=${freq[$bm]}
+    #for opt in gc-freq=${freq[$bm]}
+    for opt in gc-plain gc-live gc-freq=${freq[$bm]}
+    #for opt in gc-plain gc-live
     do 
       	echo "Processing $bm [$opt]"
       	$DIR/gen_gc_stats.sh $OPTARG \

### 2015-07-10 lazy-scheme/post_process/process_fast.cpp

@@ -152,7 +152,6 @@ void parse(const char* line)
     getValues(line, &time_gc, &address, &create_time, &first_use,
               &last_use, &last_copied);
 
-    //assert(last_use < MAXCELL);
     if (time_gc != this_gc_time) {
         fprintf(stderr, "Error_1 in GC statistics\n\t");
         fprintf(stderr, "%ld != %ld\n", time_gc, this_gc_time);
@@ -171,13 +170,11 @@ void parse(const char* line)
         c_dragged++;
         t_dragged += (time_gc - last_use);
     }
-    assert(gc_count < MAXGC);
  
     int idx;
     long the_time;
     for (idx = 0; idx < gc_count; idx++) {
         the_time = gc_times[idx];
-        //assert(the_time < MAXCELL);
         if (time_gc > this_gc_time) {
             fprintf(stderr, "Error_2 in GC statistics\n\t");
             fprintf(stderr, "%ld > %ld\n", time_gc, this_gc_time);

### 2015-07-10 lazy-scheme/src/run_in_loop.py

@@ -5,8 +5,7 @@ import time
 import datetime
 from collections import namedtuple
 
-#'knightstour':677800
-prog_size = {'nqueens':22729, 'nperm':27429, 'fibheap':38001, 'treejoin':1616535, 'sudoku':4075, 'lcss':22243, 'lambda':20466, 'gc_bench':204850, 'knightstour':6778000}
+prog_size = {'nqueens':22729, 'nperm':27429, 'fibheap':38001, 'treejoin':1616535, 'sudoku':4075, 'lcss':22243, 'lambda':20466, 'gc_bench':204850, 'knightstour':680000}
 gc_info = namedtuple("gc_info", "heap_total heap_left heap_used gc_invocations gc_time dfa_time num_states exec_time")
 gc_stat = namedtuple("gc_stat", "prog_name reachability liveness")
 gc_stat_list = []

### 2015-07-11 paper/paper.tex
Fixed minor naming issues

@@
 determined using the program point and the variable name. The liveness 
-of all variables are stored as DFAs. These DFAs are then encoded as a 
+of all variables are stored as DFA. These DFA are then encoded as a 
 table. A variable is live only if an entry is found in this table.
@@
 \begin{table*}[ht!]
-\caption{Comparing LGC with liveness for closures and reachability for closures}
+\caption{Comparing LGC times for different strategies}
 \label{tab:compare-live-clo-reach-clo}
@@
 \hline
-Prog Name & Heap size & Num GC & Reachablity for closure & Liveness for closure \\ \hline
-\verb@nqueens(8)@ & 10000 & 183 & 0.0087 & 0.2267\\ \hline
-\verb@lambda@&10000&11&0.0010&4.9922\\ \hline
-\verb@lcss@&10000&1&0.00008&34.6684\\ \hline
+Prog Name & Heap size & Num GC & Time taken using reachablity for closure  & Time taken using liveness for closure \\ \hline
+\verb@nqueens(8)@ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
+\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
+\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
 \end{tabular}

### 2015-07-11 paper/paper.tex
abstract

@@
 reachability-based   collection  for   closures.   On   our  benchmark
-programs,  we  obtain a  decrease  of  up to  50\%  in  the number  of
-collections, and up to 70\% decrease in overall program execution time
-when compared to reachability based collectors.
+programs,  we  obtain a  decrease  of  up to  45\%  in  the number  of
+collections, and  a decrease of  up to 88\%  in the number  of dragged
+cells.  Even  though  liveness   based  collector  uses  an  expensive
+traversal scheme,  the execution  time is  competitive (0.2X  - 2.08X)
+when compared to a reachability based collector.
 \end{abstract}
@@
 \end{tabular}%\vskip -10mm
- \caption{Memory  usage  for (a)  full  benchmarks  and (b)  for  some
-   interesting part  of the benchmarks.   The blue and the  red curves
+ \caption{Memory usage.  Column (a) shows complete  usage while column
+   (b) enlarges  a part  of the  usage.  The blue  and the  red curves
    indicate the number of cons cells  in the active semi-space for RGC

### 2015-07-11 paper/paper.tex

@@
 
-{\color {Myblue}Rewrite at the end: 
-Section~\ref{sec:defs} introduces the syntax and semantics of the
-language used to illustrate our
-analysis.
-% Section~\ref{sec:operational} describes the operational semantics of
-% the language.
-Liveness analysis for this language is  described in
-Section~\ref{sec:liveness}. This section also has by  a  sketch  of  a
-correctness proof  relative  to  a  non-standard
-semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
-as   finite-state  automata.    Section~\ref{sec:experiments}  reports
-experimental  results  and Section~\ref{sec:lgc-always-better}  proves
-that a liveness based collector  can never do more garbage collections
-than a reachability based collector.
-% and Section~\ref{sec:conclusion} concludes the paper.
-}
+{\color {Myblue}Section~\ref{sec:defs} introduces
+  the syntax and  describes a small-step operational  semantics of the
+  programming language  considered in  our analysis.  We  describe our
+  liveness     analysis     for      this     lazy     language     in
+  Section~\ref{sec:liveness}.   Section~\ref{sec:computing}  describes
+  the formulation of this liveness  as a grammar and its approximation
+  to DFA. It  also contains a proof showing the  undecidability of the
+  precise  liveness in  our formulation.   Section~\ref{sec:GC-scheme}
+  discusses different  ways of  handling closures  and their  pros and
+  cons.      We     report     our     experimental     results     in
+  Section~\ref{sec:experiments} along  with some of  our observations.
+  Section~\ref{sec:relatedwork}  discusses  previous work  related  to
+  garbage  collection  and liveness  and  Section~\ref{sec:conclusion}
+  discusses possible extensions and concludes the paper.  }
 \section{The target language---syntax and semantics}
@@
 
-\section{Liveness}\label{sec:liveness}
+\section{Liveness}
+\label{sec:liveness}
 
@@
 \section{The Garbage Collection Scheme}
-
+\label{sec:GC-scheme}
 Our experimental setup consists of  an interpreter for our language, a
@@
 straight  forward as  variables may  escape their  creation scope  and
-needs to  be handled in an  approximate way.  We describe this in detail next.
+need to  be handled in an  approximate way.  We describe this in detail next.
 
@@
 \section{Experimental Evaluation}
-
+\label{sec:experiments}
 %% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
@@
 \section{Related Work} 
+\label{sec:relatedwork}
 Although augmenting  garbage collection with liveness  information has

### 2015-07-11 paper/paper.tex
Updated organization of the paper

@@
 reachability-based   collection  for   closures.   On   our  benchmark
-programs,  we  obtain a  decrease  of  up to  45\%  in  the number  of
-collections, and  a decrease of  up to 88\%  in the number  of dragged
-cells.  Even  though  liveness   based  collector  uses  an  expensive
-traversal scheme,  the execution  time is  competitive (0.2X  - 2.08X)
-when compared to a reachability based collector.
+programs,  we  obtain a  decrease  of  up to  50\%  in  the number  of
+collections, and up to 70\% decrease in overall program execution time
+when compared to reachability based collectors.
 \end{abstract}
@@
 \end{tabular}%\vskip -10mm
- \caption{Memory usage.  Column (a) shows complete  usage while column
-   (b) enlarges  a part  of the  usage.  The blue  and the  red curves
+ \caption{Memory  usage  for (a)  full  benchmarks  and (b)  for  some
+   interesting part  of the benchmarks.   The blue and the  red curves
    indicate the number of cons cells  in the active semi-space for RGC

### 2015-07-11 paper/paper.tex
sec 1 done.

--- 2015-07-11 paper/paper.tex
@@
 cells.  Even though liveness based collector uses an expensive
-traversal scheme,  the execution  time is  competitive (0.2X  - 2.08X)
-when compared to a reachability based collector.
+traversal scheme, the execution time is competitive (0.2X - 2.1X) when
+compared to a reachability based collector.
 \end{abstract}
@@
 
-{\color  {Myblue}   Experiments  with  a  single   generation  copying
-  collector confirm the expected performance benefits.  Liveness-based
-  collection  results  in an  increase  in  garbage reclaimed.   As  a
-  consequence, there is a decrease in  the number of collections and a
-  (some quantification)  decrease in the  memory size required  to run
-  programs. Besides there is a reduction in the overall execution time
-  for  some  programs.  Our  experiments  further  show that  a  mixed
+Experiments    with   a    single    generation   copying    collector
+(Section~\ref{sec:experiments})  confirm   the  expected  performance
+  benefits.   Liveness-based  collection  results in  an  increase  in
+  garbage reclaimed.   As a  consequence, there is  a decrease  in the
+  number of  collections (up to 45\%)  and a decrease (up  to 88\%) in
+  the number  of {\em dragged}  cells (cells  that survive a  GC after
+  their  last  use). Besides  there  is  a  reduction in  the  overall
+  execution time for some programs.  Our experiments show that a mixed
   strategy, where we  do liveness based GC up to  the closure boundary
@@
   (where  the  arguments  of  the closure  are  also  collected  using
-  liveness)  at  much  less  overhead (?).  [Overhead:  extra  memory,
-    book-keeping during run-time to update, multiple traversal of same
-    set of cells][Refer to Section~\ref{some-sec-in-impl}] }
+  liveness) at much less overheads (Section~\ref{sec:strategies}).
 
@@
 
-\subsection{Copying closure using liveness vs copying closures using reachability}
+\subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
 

### 2015-07-11 paper/paper.tex
Fixed some references

--- 2015-07-11 paper/paper.tex
@@
 
-{\color {Myblue}Section~\ref{sec:defs} introduces
-  the syntax and  describes a small-step operational  semantics of the
-  programming language  considered in  our analysis.  We  describe our
-  liveness     analysis     for      this     lazy     language     in
-  Section~\ref{sec:liveness}.   Section~\ref{sec:computing}  describes
-  the formulation of this liveness  as a grammar and its approximation
-  to DFA. It  also contains a proof showing the  undecidability of the
-  precise  liveness in  our formulation.   Section~\ref{sec:GC-scheme}
-  discusses different  ways of  handling closures  and their  pros and
-  cons.      We     report     our     experimental     results     in
-  Section~\ref{sec:experiments} along  with some of  our observations.
-  Section~\ref{sec:relatedwork}  discusses  previous work  related  to
-  garbage  collection  and liveness  and  Section~\ref{sec:conclusion}
-  discusses possible extensions and concludes the paper.  }
+{\color  {Myblue}Section~\ref{sec:defs}  introduces   the  syntax  and
+  describes  a small-step  operational  semantics  of the  programming
+  language  considered  in our  analysis.   We  describe our  liveness
+  analysis  for  this  lazy  language  in  Section~\ref{sec:liveness}.
+  Section~\ref{sec:computing}  describes   the  formulation   of  this
+  liveness as a grammar and its approximation to DFA. It also contains
+  a proof  showing the undecidability  of the precise liveness  in our
+  formulation.   Section~\ref{sec:GC-scheme} discusses  different ways
+  of  handling  closures and  their  pros  and  cons.  We  report  our
+  experimental  results  in Section~\ref{sec:experiments}  along  with
+  some observations.  Section~\ref{sec:relatedwork} discusses previous
+  work   related    to   garbage    collection   and    liveness   and
+  Section~\ref{sec:conclusion}   discusses  possible   extensions  and
+  concludes the paper.  }
 \section{The target language---syntax and semantics}
@@
 To compare the  effect of using reachability to  copy closures instead
-of liveness we take a couple  of programs from our benchmark suite and
+of liveness we take some representative programs from our benchmark suite and
 execute       them      with       both.        As      shown       in

### 2015-07-11 paper/paper.tex
Updated organization

@@
 
-{\color  {Myblue}Section~\ref{sec:defs}  introduces   the  syntax  and
-  describes  a small-step  operational  semantics  of the  programming
-  language  considered  in our  analysis.   We  describe our  liveness
+{\color  {Myblue}Section~\ref{sec:defs}  introduces   the  syntax  
+  of the  programming  language  considered  in our  analysis and describes describes  a small-step  operational semnatics for it.   We  describe our  liveness
   analysis  for  this  lazy  language  in  Section~\ref{sec:liveness}.
-  Section~\ref{sec:computing}  describes   the  formulation   of  this
+  Section~\ref{sec:computing}  describes   the  formulation   of  the
   liveness as a grammar and its approximation to DFA. It also contains
   a proof  showing the undecidability  of the precise liveness  in our
-  formulation.   Section~\ref{sec:GC-scheme} discusses  different ways
+  formulation which necessitates the approximation to DFA.   Section~\ref{sec:GC-scheme} discusses  different ways
   of  handling  closures and  their  pros  and  cons.  We  report  our

### 2015-07-11 paper/paper.tex
contribs

--- 2015-07-11 paper/paper.tex
@@
 (Section~\ref{sec:experiments})   confirm  the   expected  performance
-  benefits.   Liveness-based  collection  results in  an  increase  in
-  garbage reclaimed.   As a  consequence, there is  a decrease  in the
-  number of  collections (up to 45\%)  and a decrease (up  to 88\%) in
-  the number  of {\em dragged}  cells (cells  that survive a  GC after
-  their  last  use). Besides  there  is  a  reduction in  the  overall
-  execution time for some programs.  Our experiments show that a mixed
-  strategy, where we  do liveness based GC up to  the closure boundary
-  and reachability based GC for  the arguments (free variables) of the
-  closure,  gives us  similar benefits  as a  fully liveness  based GC
-  (where  the  arguments  of  the closure  are  also  collected  using
-  liveness) at much less overheads (Section~\ref{sec:strategies}).
+benefits.  Liveness-based collection results in an increase in garbage
+reclaimed.  As  a consequence, there  is a  decrease in the  number of
+collections (up to 45\%) and a decrease  (up to 88\%) in the number of
+{\em  dragged}  cells  (cells  that  survive a  GC  after  their  last
+use). Besides, there is a reduction  in the overall execution time for
+some programs.  Our  experiments show that a mixed  strategy, where we
+do liveness based GC up to the closure boundary and reachability based
+GC for the arguments (free variables) of the closure, gives us similar
+benefits as  a fully  liveness based  GC (where  the arguments  of the
+closure  are  also  collected  using liveness)  with  lesser  overhead
+(Section~\ref{sec:strategies}).
 
@@
 
-{\color {Myblue}Rewrite at the end: The specific contributions of the
-paper
-  are: The formulation of a  liveness analysis for a lazy language and
-  the proof of its correctness, the undecidability of the grammar that
-  results from the analysis and  its approximation by DFAs, the design
-  of a  garbage collector that can  use the DFAs to  retain live cells
-  that  also include  closures, and  empirical results  that  show the
-  effectiveness of liveness-based garbage collection.}
+The specific contributions of this work are
+  \begin{itemize}
+  \item  We formulate  a  liveness  analysis for  a  lazy first  order
+    functional language. We show how  to compute function summaries in
+    a  context  independent  manner  to achievs  a  context  sensitive
+    analysis. We  further describe the transformation  of the liveness
+    equations to grammar rules in a  way that the language accepted by
+    the resulting grammar describes live heap locations.\mycomment{the
+      proof of its correctness}
+  \item We prove  the undecidability of the grammar  that results from
+    the  analysis.   To overcome  this,  we  present an  algorithm  to
+    transform grammar  to finite  state automata  (DFA) such  that the
+    language accepted by the  DFAs over-approximates the liveness (and
+    hence a safe approximation).
+  \item We have implemented a garbage  collector that can use the DFAs
+    to  retain live  cells that  also include  closures. We  have done
+    extensive  experiments with  this garbage  collector and  obtained
+    empirical results  that show  the effectiveness  of liveness-based
+    garbage collection.
+  \end{itemize}
 

### 2015-07-11 paper/paper.tex
DFAs->DFA

@@
     transform grammar  to finite  state automata  (DFA) such  that the
-    language accepted by the  DFAs over-approximates the liveness (and
+    language accepted by the  DFA over-approximates the liveness (and
     hence a safe approximation).
-  \item We have implemented a garbage  collector that can use the DFAs
+  \item We have implemented a garbage  collector that can use the DFA
     to  retain live  cells that  also include  closures. We  have done
@@
 simplified using  the $\hookrightarrow$ rules.  Finally the simplified
-NFAs are converted to DFAs.
+NFAs are converted to DFA.
 
 %------------------------------------------------------------%
-\subsubsection{From CFGs to GC-ready DFAs}
+\subsubsection{From CFGs to GC-ready DFA}
 \label{sec:NFA-approx}
@@
 garbage collector  can be  easily configured to  work on the  basis of
-reachability (RGC  mode) or use  the liveness DFAs (LGC  mode).  While
-exploring the  activation record of a  function, LGC uses  the DFAs at
+reachability (RGC  mode) or use  the liveness DFA (LGC  mode).  While
+exploring the  activation record of a  function, LGC uses  the DFA at
 the latest program  point traversed in the function's  body.  Thus, if
 LGC is invoked  at the program point $\pi$, it  would use the liveness
-DFAs  at  $\pi$ itself  for  exploring the  root  set  of the  current
+DFA  at  $\pi$ itself  for  exploring the  root  set  of the  current
 activation record. For  any other activation record in  the stack, say
@@
 %% \comment{
-%% While exploring the activation record of a function, LGC uses the DFAs
+%% While exploring the activation record of a function, LGC uses the DFA
 %% at  the latest  program  point traversed  in the function's body.
@@
 %% is invoked  at the program point   $\pi$,  it would  use  the
-%% liveness  DFAs at  $\pi$ itself for exploring the root set of the
+%% liveness  DFA at  $\pi$ itself for exploring the root set of the
 %% current activation record. For any other activation record in the
@@
 %% call-chain,
-%% the DFAs at the point of call to  ${\mathit g}$ would be used.}
+%% the DFA at the point of call to  ${\mathit g}$ would be used.}
 
@@
 %% grammars for liveness at each GC point (calls to user functions and to
-%% $\CONS$).  These were then converted to DFAs for run-time consultation
+%% $\CONS$).  These were then converted to DFA for run-time consultation
 %% by  the garbage collector.  Experiments confirm  the precision  of the
@@
   based garbage collector using heuristics  such as limiting the depth
-  of  DFAs,   merging  nearly-equivalent   states  and   using  better
+  of  DFA,   merging  nearly-equivalent   states  and   using  better
   representation  and algorithms  for automata  manipulation. We  also

### 2015-07-11 paper/paper.tex
minor-changes

@@
 
-{\color  {Myblue}Section~\ref{sec:defs}  introduces   the  syntax  
-  of the  programming  language  considered  in our  analysis and describes describes  a small-step  operational semnatics for it.   We  describe our  liveness
-  analysis  for  this  lazy  language  in  Section~\ref{sec:liveness}.
-  Section~\ref{sec:computing}  describes   the  formulation   of  the
-  liveness as a grammar and its approximation to DFA. It also contains
-  a proof  showing the undecidability  of the precise liveness  in our
-  formulation which necessitates the approximation to DFA.   Section~\ref{sec:GC-scheme} discusses  different ways
-  of  handling  closures and  their  pros  and  cons.  We  report  our
-  experimental  results  in Section~\ref{sec:experiments}  along  with
-  some observations.  Section~\ref{sec:relatedwork} discusses previous
-  work   related    to   garbage    collection   and    liveness   and
+Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
+language  considered  in our  analysis  and  a small-step  operational
+semnatics  for it.   We present  our liveness  analysis for  this lazy
+language  in Section~\ref{sec:liveness}.   Section~\ref{sec:computing}
+describes  the  formulation of  the  liveness  as  a grammar  and  its
+approximation  to   DFA.  It  also   contains  a  proof   showing  the
+undecidability  of  the  precise  liveness in  our  formulation  which
+necessitates  the approximation  to DFA.   Section~\ref{sec:GC-scheme}
+discusses different ways of handling closures and their pros and cons.
+We  report our  experimental results  in Section~\ref{sec:experiments}
+along with some observations.  Section~\ref{sec:relatedwork} discusses
+previous  work   related  to  garbage  collection   and  liveness  and
 Section~\ref{sec:conclusion}   discusses   possible   extensions   and
-  concludes the paper.  }
+concludes the paper.
 \section{The target language---syntax and semantics}
@@
 \begin{figure}[t]
-\footnotesize
+\renewcommand{\arraystretch}{1.5}
 \begin{eqnarray*}
@@
 
-The closest  to our approach  is the liveness based  garbage collector
-implemented in~\cite{asati14lgc}. We extend  their work to handle lazy
-evaluation and closures.
+Another  interesting  research  in  this   field  is  the  concept  of
+Simplifiers~\cite{ONeill}. They are  described as lightweight daeomons
+which improve the efficiency of the program in general and the garbage
+collector in particular. Most of the simplifications mentioned in this
+work are subsumed by a liveness  based collector. We feel this work is
+orthogonal and can augment our  liveness based collector.  The closest
+to our  approach is the  liveness based garbage  collector implemented
+in~\cite{asati14lgc}. We  extend their work to  handle lazy evaluation
+and closures.
 

### 2015-07-11 paper/paper.tex
Updated related work

--- 2015-07-11 paper/paper.tex
@@
 (Section~\ref{sec:experiments})  confirm   the  expected  performance
-benefits.  Liveness-based collection results in an increase in garbage
-reclaimed.  As  a consequence, there  is a  decrease in the  number of
-collections (up to 45\%) and a decrease  (up to 88\%) in the number of
-{\em  dragged}  cells  (cells  that  survive a  GC  after  their  last
-use). Besides, there is a reduction  in the overall execution time for
-some programs.  Our  experiments show that a mixed  strategy, where we
-do liveness based GC up to the closure boundary and reachability based
-GC for the arguments (free variables) of the closure, gives us similar
-benefits as  a fully  liveness based  GC (where  the arguments  of the
-closure  are  also  collected  using liveness)  with  lesser  overhead
-(Section~\ref{sec:strategies}).
+  benefits.   Liveness-based  collection  results in  an  increase  in
+  garbage reclaimed.   As a  consequence, there is  a decrease  in the
+  number of  collections (up to 45\%)  and a decrease (up  to 88\%) in
+  the number  of {\em dragged}  cells (cells  that survive a  GC after
+  their  last  use). Besides  there  is  a  reduction in  the  overall
+  execution time for some programs.  Our experiments show that a mixed
+  strategy, where we  do liveness based GC up to  the closure boundary
+  and reachability based GC for  the arguments (free variables) of the
+  closure,  gives us  similar benefits  as a  fully liveness  based GC
+  (where  the  arguments  of  the closure  are  also  collected  using
+  liveness) at much less overheads (Section~\ref{sec:strategies}).
 
@@
 
-The specific contributions of this work are
-  \begin{itemize}
-  \item  We formulate  a  liveness  analysis for  a  lazy first  order
-    functional language. We show how  to compute function summaries in
-    a  context  independent  manner  to achievs  a  context  sensitive
-    analysis. We  further describe the transformation  of the liveness
-    equations to grammar rules in a  way that the language accepted by
-    the resulting grammar describes live heap locations.\mycomment{the
-      proof of its correctness}
-  \item We prove  the undecidability of the grammar  that results from
-    the  analysis.   To overcome  this,  we  present an  algorithm  to
-    transform grammar  to finite  state automata  (DFA) such  that the
-    language accepted by the  DFAs over-approximates the liveness (and
-    hence a safe approximation).
-  \item We have implemented a garbage  collector that can use the DFAs
-    to  retain live  cells that  also include  closures. We  have done
-    extensive  experiments with  this garbage  collector and  obtained
-    empirical results  that show  the effectiveness  of liveness-based
-    garbage collection.
-  \end{itemize}
+{\color {Myblue}Rewrite at the end: The specific contributions of the
+paper
+  are: The formulation of a  liveness analysis for a lazy language and
+  the proof of its correctness, the undecidability of the grammar that
+  results from the analysis and  its approximation by DFAs, the design
+  of a  garbage collector that can  use the DFAs to  retain live cells
+  that  also include  closures, and  empirical results  that  show the
+  effectiveness of liveness-based garbage collection.}
 

### 2015-07-11 paper/paper.tex
Fixed typos
spacing..

--- 2015-07-11 paper/paper.tex
@@
   \psset{unit=1mm}
-  \begin{pspicture}(90,48)(0,-48)
+  \begin{pspicture}(90,52)(0,-52)
     %\psframe(90,48)(0,-48)
+    \renewcommand{\arraystretch}{1}
     \begin{tabular}{@{}c@{}}
@@
       }}
-      \\ 
+      \\ \\
       (a) Example program. \\  \\
 %      \includegraphics[width=.45\textwidth]{motiv-example}
-      \input{mem-graph} \\ 
+      \input{mem-graph} \\ \\
       \renewcommand{\arraystretch}{.9}
@@
   \end{pspicture}
-  \vspace*{-3ex}
+%  \vspace*{-3ex}
   \caption{Example Program and its Memory Graph}\label{fig:mot-example}
@@
 \begin{figure}[t]
-\renewcommand{\arraystretch}{1.5}
+\renewcommand{\arraystretch}{1.2}
 \begin{eqnarray*}
@@
     e)
-    \hspace{0.2em} \ \ \ \ \mbox{\em --- function definition} \\
+    \hspace{0.2em} \ \ \ \ \mbox{\em --- function def} \\
 e \in \mathit{Expr} & \!\!\!::=\!\!\! &
-\left\{\begin{array}{@{}ll@{\hspace{2em}}l}
+\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
        (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
@@
        (\NULLQ\,\, x) & \\
-       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester
-and generic arithmetic} \\
-       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
+       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
+%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
        (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
@@
 \[
+\renewcommand{\arraystretch}{1.1}
 \begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
@@
 \begin{figure*}[t]
+\renewcommand{\arraystretch}{1.5}
 \begin{eqnarray*}
@@
 \begin{figure*}[t!]
+\renewcommand{\arraystretch}{1.1}
   \begin{tabular}{cc}
     \begin{minipage}{.40\textwidth}
-        \small
-        \renewcommand{\arraystretch}{1}{
+%        \small
+        \renewcommand{\arraystretch}{1.1}{
           \begin{uprogram}
@@
         \end{uprogram}}
-        \renewcommand{\arraystretch}{1}{
+        \renewcommand{\arraystretch}{1.1}{
 	  \begin{uprogram}
@@
 
-        \small
+      %\small
 \begin{eqnarray*}

### 2015-07-11 paper/paper.tex
removed all red and blue coloring

@@
 analysis itself.   In addition, apart  from evaluated data,  we extend
-extend liveness-based  collection also to closures---closures  with no
-future use should also be  collected.  This involves tracking closures
-to their  points of  creation at  runtime from  where we  obtain their
-liveness.  This  requires significant  modification to the  earlier GC
-scheme.
+liveness-based collection  to closures---closures  with no  future use
+should also  be collected.  This  involves tracking closures  to their
+points of  creation at  runtime from where  we obtain  their liveness.
+This requires significant modification to the earlier GC scheme.
 
@@
         \begin{uprogram}
+          \UNL{1} (\DEFINE\ (\abigfunction~\py)
+          \UNL{2} $(Create\ a\ big\ closure\ and\ return)$ 
+          \end{uprogram}}
+        \renewcommand{\arraystretch}{1.1}{
+          \begin{uprogram}
             \UNL{1} (\DEFINE\ (\length~\xl)
@@
 	  \UNL{1} $(\DEFINE\ (\main)$
-           \UNL{2} \!\!$\pi_9\!\!:\, \cred{(\LET\  \pa\  \leftarrow
-(\abigfunction\ \acdr)}$ \IN
-          \UNL{3} \!\!$\pi_{10}\!\!:\, \cred{(\LET\  \pb\  \leftarrow
-(}$\cred{+}$\cred{\ \pa\ \acdr)}$
+           \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow
+(\abigfunction\ \acdr)$ \IN
+          \UNL{3} \!\!$\pi_{10}\!\!:\, (\LET\  \pb\  \leftarrow ($+$\ \pa\ \acdr)$
 \IN
-	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ \cred{(\LET\ \pc}\
-$\cred{\leftarrow  (\CONS\ \pb\ \NIL)}$ \IN
+	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ (\LET\ \pc\
+$\leftarrow  (\CONS\ \pb\ \NIL)$ \IN
           \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
-          $\cred{(\LET\ \pw}\  $\cred{\leftarrow  (\length\ \pc)}$ \IN
+          $(\LET\ \pw\  $\leftarrow  (\length\ \pc)$ \IN
           \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
@@
               [\;\epath_1 \mapsto \lbrace \px.\clazy \sigma,
-\xl.\clazy\sigma\rbrace ,~\epath_2 \mapsto \lbrace \cred{
-\pv.\sigma}  \rbrace,\\
+\xl.\clazy\sigma\rbrace ,~\epath_2 \mapsto \lbrace 
+\pv.\sigma  \rbrace,\\
  &&          \;\;  \epath_3 \mapsto \lbrace \pz.\sigma,
@@
 %%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
-\cred{\clazy\Lf{\length}{1}{\clazy\sigma}}\}\}
+\clazy\Lf{\length}{1}{\clazy\sigma}\}\}
              \rbrack \\
@@
                 [\;
-                  \epath_4 \mapsto \lbrace \cred {\pw.\sigma,
-                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},}\\
+                  \epath_4 \mapsto \lbrace \pw.\sigma,
+                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},\\
 && \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
-\cred{\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},}
-\cred{\pa.\clazy\bcar\Lf{\length}{\mbox{1}}{\sigma}} \rbrace
+\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},
+\pa.\clazy\bcar\Lf{\length}{\mbox{1}}{\sigma} \rbrace
                                ] \\
                \cal{P_{\mbox{\tiny \main}}} &=&
-                  \lbrack~ \cred{\pi_9 \mapsto \lbrace \epath_4 \rbrace,  \
+                  \lbrack~ \pi_9 \mapsto \lbrace \epath_4 \rbrace,  \
                     \pi_{10} \mapsto \lbrace \epath_4 \rbrace,  \
-                   \pi_{11} \mapsto \lbrace \epath_4 \rbrace,} \\
-                   && \;\; \cred{\pi_{12} \mapsto \lbrace \epath_4 \rbrace,
-\pi_{13} \mapsto \lbrace \epath_4 \rbrace}~\rbrack
+                   \pi_{11} \mapsto \lbrace \epath_4 \rbrace, \\
+                   && \;\; \pi_{12} \mapsto \lbrace \epath_4 \rbrace,
+\pi_{13} \mapsto \lbrace \epath_4 \rbrace~\rbrack
 \end{eqnarray*}
@@
 %% collector to reduce the cost of this.
-{\color{blue}We extended the liveness based garbage collection to lazy
+We extended the liveness based garbage collection to lazy
   languages and shown its benefit  for garbage collection in practical
@@
   In  summary, we  need to  investigate  ways to  make liveness  based
-  garbage collection attractive for practical collectors.  }
+  garbage collection attractive for practical collectors.
 

### 2015-07-11 paper/paper.tex
Fixed issues in op-semantics after discussion with Sanyal Sir
Updated op-semantics and minefield semantics
reorg of fig
appendix is back

@@
 
-Functional programs extensively use dynamically allocated memory.  The
-allocation  is either  explicit (for  example, using  constructors) or
-implicit  (during creation  of  closures).  Programs  written in  lazy
+Functional programs use dynamically allocated memory extensively.  The
+allocation  is either  explicit (while using  constructors, for example) or
+implicit (as while creating  closures).  Programs  written in  lazy
 functional languages put additional  demands on memory as they require
@@
 \rightsquigarrow
-$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
-(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
+$  $\rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
 \sc{car-whnf} \\
@@
  & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
-$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
+$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
 \sc{prim-whnf} \\
@@
 e)\!:\!S, H,
-(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
+(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
 \sc{return-whnf}\\
@@
 %------------------------------------------------------------%
-\subsubsection{From CFGs to GC-ready DFA}
-\label{sec:NFA-approx}
- We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
- approximate a CFG to a {\em strongly regular\/} grammar.  For
- example, the grammar
- fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
- Mohri-Nederhof transformation is:
- \begin{eqnarray*}
-   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
-   \acdr\var{\Df{\length}{1}}
-   \mid \clazy\var{\Df{\length}{1}}\\
-   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
-   \mid \epsilon
- \end{eqnarray*}
-
-The strongly regular grammar is converted  into a set of NFAs, one for
-each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
-done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
-pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
-constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
-may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
-simplification is continued till a fixed point is reached, after which
-the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
-algorithm, its correctness and termination proofs are given by Karkare
-et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
-has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
-automaton, for every edge labelled \clazy, we check if the source node
-of the edge has a valid path to  a final final state.  If yes, we mark
-the mark  the source node as  final. Finally, we remove  all the edges
-labelled  \clazy\  and  convert  the automaton  into  a  deterministic
-automaton.    This   effectively  implements   the   $\hookrightarrow$
-simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
-paths. While  checking for liveness  during GC, a  forward access-path
-only if it can reach a final state.
 
@@
 
+\subsubsection{From CFGs to GC-ready DFA}
+\label{sec:NFA-approx}
+ We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
+ approximate a CFG to a {\em strongly regular\/} grammar.  For
+ example, the grammar
+ fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
+ Mohri-Nederhof transformation is:
+ \begin{eqnarray*}
+   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
+   \acdr\var{\Df{\length}{1}}
+   \mid \clazy\var{\Df{\length}{1}}\\
+   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
+   \mid \epsilon
+ \end{eqnarray*}
+
+The strongly regular grammar is converted  into a set of NFAs, one for
+each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
+done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
+pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
+constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
+may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
+simplification is continued till a fixed point is reached, after which
+the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
+algorithm, its correctness and termination proofs are given by Karkare
+et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
+has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
+automaton, for every edge labelled \clazy, we check if the source node
+of the edge has a valid path to  a final final state.  If yes, we mark
+the mark  the source node as  final. Finally, we remove  all the edges
+labelled  \clazy\  and  convert  the automaton  into  a  deterministic
+automaton.    This   effectively  implements   the   $\hookrightarrow$
+simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
+paths. While  checking for liveness  during GC, a  forward access-path
+only if it can reach a final state.
 Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
@@
 
-\mycomment{
-\pagebreak
 \pagebreak
@@
 \item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
-  $\rho,$ $S,$ $H,$  $e,$ $\sigma,$ $e_f$.  $\sigma$ is  the demand on
-  the expression and $e_f$ is the  body of the function.  The body and
-  the demand are both needed to calculate the liveness environment for
-  expressions.  The evaluation of  an expression may make variables in
-  the scope live because of closures that are created outside it.  The
-  information  in a  suspended  evaluation context  is also  similarly
-  augmented with the demand on it and the function body of which the
-  suspended evaluation is a part.  Thus a stacked entry now takes the
-  form $(\rho, x, e, \sigma, e_f)$. A modification of the small step
-  semantics which carries the extra information is shown in Figure[].
+  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
+  on the expression $e$, and $e_f$  is the body of the function. $e_f$
+  and $\sigma$ are  both needed to calculate  the liveness environment
+  for expressions.  The information  in a suspended evaluation context
+  is also similarly  augmented with the demand on it  and the function
+  body of  which the suspended evaluation  is a part.  Thus  a stacked
+  entry now takes the form $(\rho, x, e, \sigma, e_f)$. A modification
+  of the small  step semantics which carries the  extra information is
+  shown in Figure~\ref{fig:minefield-semantics}.
 \item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
@@
 
+Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
+(\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
+preceded by  a garbage  collection using $GC$.  We consider  a garbage
+collection followed  by a transition as  a step. We show  by induction
+that, starting  from the  initial state, any  transition of  $n$ steps
+occurs without a \bang.
+
+A few observations: First observe that the transitions  which involving dereferencing and can lead to \bang are those which have a $H(\rho(x))$ term in their premise for some variable $x$.
+
 \begin{figure*}[h!]
@@
     e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e,
+  \rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
   \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \cred{\sigma', e_f'}$}      &
+  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
 \sc{car-whnf} \\
@@
  & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
-  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho(z)
-    \mapsto H(\rho(x)) + H(\rho(y))], e, \cred{\sigma', e_f'}$}      &
+  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
+    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
 \sc{prim-whnf} \\
@@
   (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
-  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e,
+  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
   \cred{\sigma', e_f'}$} &
@@
 \end{figure*}
-
-%% Generated by bibtex from your ~.bib file.  Run latex,
-%% then bibtex, then latex twice (to resolve references)
-%% to create the ~.bbl file.  Insert that ~.bbl file into
-%% the .tex source file and comment out
-%% the command \texttt{{\char'134}thebibliography}.
-%% % This next section command marks the start of
-%% % Appendix B, and does not continue the present hierarchy
-%% \section{More Help for the Hardy}
-%% The sig-alternate.cls file itself is chock-full of succinct
-%% and helpful comments.  If you consider yourself a moderately
-%% experienced to expert user of \LaTeX, you may find reading
-%% it useful but please remember not to change it.
-%% %\balancecolumns % GM June 2007
-%% % That's all folks!
-
-%% % {\color{red}
-%% %   (Note 3.\ includes $e_\mainpgm$ treated as the body of
-%% %   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
-%% % }
-
-%% %% =======
-
-%% %------------------------------------------------------------%
-%% %\input{lazy-liveness}
-}
 \end{document}

### 2015-07-11 paper/paper.tex
results...

@@
 
-Table~\ref{tab:experimental-results} shows  that  LGC
-improves the  garbage collection time  and efficiency for most  of the
-programs.  Even for  programs where the time taken is  more it is very
-competitive.  Memory usage  graphs are shown for  select benchmarks in
-Figure~\ref{fig:memory-usage}.  As  the number of  garbage collections
-tend  to  be  very large  our  graphs  show  only  a window  which  is
-representative of the behavior for  that particular benchmark.  In all
-the programs we can see that  the curve corresponding to LGC regularly
-dips below the RGC curve.
+Table~\ref{tab:experimental-results}  compares the  garbage collection
+statisctics for  RGC and LGC  with reachability for  copying closures.
+We report the  number of GC events, average number  of cells reclaimed
+per  GC, average  number  of cells  touched per  GC,  total number  of
+dragged cells  and total time  to perform  all collections.  It  is no
+surprise  that  number  of  cells  reclaimed per  GC  is  higher  (and
+consequently the  number of  GCs is lower)  for LGC  based collection.
+LGC also reduces  the number of dragged  cells significantly. However,
+the cost for LGC is that per  GC execution time is more for LGC, which
+results in higher overall execution  time for garbage collection, even
+with reduced  number of collections.   However, LGC run time  is still
+competitive (slowdown within 5X of RGC  for worst case), and is better
+for   3    benchmarks   (best    case   2X   speedup).     Note   that
+\verb@gc_bench@~\cite{gc_bench}   is  a   synthestic  benchmark   that
+allocates complete binary  trees of various sizes.  These binary trees
+are  never used  by the  program. As  a result,  the benchmark  highly
+favours LGC.   We have included  this benchmark for  completeness only
+and the numbers for it  are not considered representative of practical
+programs.
+
+Memory   usage   graphs   for   select   benchmarks   in   are   shown
+Figure~\ref{fig:memory-usage}, Column  (a).  As the number  of garbage
+collections tend to  be very large, for each benchmark  we also show a
+window which  is representative  of the  behavior for  that particular
+benchmark (Column (b)).  In all the programs we can see that the curve
+corresponding to  LGC (red  line) regularly dips  below the  RGC curve
+(blue line), the dip is specially visible for the benchmarks where LGC
+outperforms                      RGC                     substantially
+(Table~\ref{tab:experimental-results}). The graphs  also include curve
+for reachable cells  (black, obtained approximately by  forcing RGC to
+run at  a very high frequency)  and the live cells  for particular run
+(grey, obtained  by post processing  the heap access  at the end  of a
+program).
 
@@
 the liveness  percieved by our  collector.  In  case of LGC  for eager
-languages  the gap  was  very  narrow and  almost  touched the  actual
-liveness curve.  In case of  lazy languages, due to using reachability
-for copying we  end up copying more  cells. Any data which  is part of
-closure irrespective  of whether it is  live or not gets  copied. Thus
-many non-live  cells get  copied during garbage  collection increasing
-the drag  time.  Our  experiments have  shown that  unless we  use the
-exact  execution  point  for  closures, bringing  down  the  number  of
-non-live  cells  copied in  closures  cannot  be reduced.   Using  the
-approximated liveness (liveness  at the creation point)  does not give
-benefits  in terms  of non-live  cells inside  closures as  everything
-would be live at the point of creation.
+languages~\ref{asati14lgc} the gap was  very narrow and almost touched
+the actual  liveness curve.  In  case of  lazy languages, this  gap is
+largely   due  to   approximate  reachability   based  collection   of
+closures. To implement liveness based  collection of closures, we need
+to  record liveness  data  for  each closure  and  update  it at  each
+evaluation  point.   These  space  and  run  time  overheads  make  it
+infeasible  to  incorporate  liveness  based  GC  for  closures  in  a
+practical garbage collector.
 

### 2015-07-11 paper/paper.tex
Added machine info

--- 2015-07-11 paper/paper.tex
@@
 the lack of a statically determinable execution order and the presence
-of closures at run-time pose additional challenges for lazy languages.
+of closures at runtime pose additional challenges for lazy languages.
 This requires changes  both in the earlier liveness  analysis and also
@@
     functional language. We show how  to compute function summaries in
-    a  context  independent  manner  to achievs  a  context  sensitive
+    a  context  independent  manner  to achieve  a  context  sensitive
     analysis. We  further describe the transformation  of the liveness
@@
 language  considered  in our  analysis  and  a small-step  operational
-semnatics  for it.   We present  our liveness  analysis for  this lazy
+semantics  for it.   We present  our liveness  analysis for  this lazy
 language  in Section~\ref{sec:liveness}.   Section~\ref{sec:computing}
@@
 \rightsquigarrow
-$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
-(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
+$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
+(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
 \sc{car-whnf} \\
@@
 distinguished  variable  that  would  be  updated with  the  value  of
-(\mainpgm).  In  addition, \print\ is a function  modelling a printing
-mechanism---a standard  runtime support assumption  for lazy languages
+(\mainpgm).  In  addition, \print\ is a function  modeling a printing
+mechanism---a standard  run-time support assumption  for lazy languages
 that prints the  value of (\mainpgm).  Notice the  use of the operator
@@
 and  $\heap[\ell :=  d]$
-indicates the  updation of  a heap \heap\  at $\ell$ with  $d$.  $\rho
+indicates the  updating of  a heap \heap\  at $\ell$ with  $d$.  $\rho
 \oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
@@
 to the program points which trigger evaluation\footnote{Program points
-  which      trigger      evaluation       are      labelled      with
+  which      trigger      evaluation       are      labeled      with
   $\psi$.}---\SIF\ conditions and \SRETURN\ expressions.  We call such
@@
 evaluation path  terminating in  $\psi$.  The  map for  the expression
-$(\SIF~\psi:x~e_1~e_2)$ is a pointwise-union of  the maps of $e_1$ and
+$(\SIF~\psi:x~e_1~e_2)$ is a point-wise-union of  the maps of $e_1$ and
 $e_2$. In addition, since the condition also triggers an evaluation of
@@
 
+\subsubsection{From CFGs to GC-ready DFA}
+\label{sec:NFA-approx}
+ We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
+ approximate a CFG to a {\em strongly regular\/} grammar.  For
+ example, the grammar
+ fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
+ Mohri-Nederhof transformation is:
+ \begin{eqnarray*}
+   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
+   \acdr\var{\Df{\length}{1}}
+   \mid \clazy\var{\Df{\length}{1}}\\
+   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
+   \mid \epsilon
+ \end{eqnarray*}
+
+The strongly regular grammar is converted  into a set of NFAs, one for
+each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
+done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
+pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
+constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
+may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
+simplification is continued till a fixed point is reached, after which
+the edges labeled \bcar \ and  \bcdr are deleted.  The details of the
+algorithm, its correctness and termination proofs are given by Karkare
+et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
+has  edges labeled  with  \acar,  \acdr\ and  \clazy\  only. In  this
+automaton, for every edge labeled \clazy, we check if the source node
+of the edge has a valid path to  a final final state.  If yes, we mark
+the mark  the source node as  final. Finally, we remove  all the edges
+labeled  \clazy\  and  convert  the automaton  into  a  deterministic
+automaton.    This   effectively  implements   the   $\hookrightarrow$
+simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
+paths. While  checking for liveness  during GC, a  forward access-path
+only if it can reach a final state.
 Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
@@
 \hline
-Prog Name & Heap size & Num GC & Time taken using reachablity for closure  & Time taken using liveness for closure \\ \hline
+Prog Name & Heap size & Num GC & Time taken using reachability for closure  & Time taken using liveness for closure \\ \hline
 \verb@nqueens(8)@ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
@@
 the programs we can see that  the curve corresponding to LGC regularly
-dips below the RGC curve.
+dips below the RGC curve. All our benchmarks were executed on a desktop with 64 bit Ubuntu 14.04,  8 core
+Intel(R) Core(TM) i7-4770 CPU @ 3.40GHz, L2 cache of 8192 KB and 
+15922 MiB (16 GB) RAM.
 
 One area  of concern is the  huge gap between the  actual liveness and
-the liveness  percieved by  our collector.  In case  of LGC  for eager
+the liveness  perceived by  our collector.  In case  of LGC  for eager
 languages  the gap  was  very  narrow and  almost  touched the  actual
@@
 Another  interesting  research  in  this   field  is  the  concept  of
-Simplifiers~\cite{ONeill}. They are  described as lightweight daeomons
+Simplifiers~\cite{ONeill}. They are  described as lightweight daemons
 which improve the efficiency of the program in general and the garbage
@@
   languages and shown its benefit  for garbage collection in practical
-  programs. We  defined context  senstive liveness analysis  that uses
+  programs. We  defined context  sensitive liveness analysis  that uses
   context independent summaries of functions obtained using a symbolic
@@
     e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e,
+  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
   \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \cred{\sigma', e_f'}$}      &
+  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
 \sc{car-whnf} \\

### 2015-07-11 paper/paper.tex
typo

--- 2015-07-11 paper/paper.tex
@@
 Table~\ref{tab:experimental-results}  compares the  garbage collection
-statisctics for  RGC and LGC  with reachability for  copying closures.
+statistics for RGC and LGC  (with reachability for copying closures).
 We report the  number of GC events, average number  of cells reclaimed
@@
 the liveness  percieved by our  collector.  In  case of LGC  for eager
-languages~\ref{asati14lgc} the gap was  very narrow and almost touched
+languages~\cite{asati14lgc} the gap was  very narrow and almost touched
 the actual  liveness curve.  In  case of  lazy languages, this  gap is

### 2015-07-11 paper/paper.tex

--- 2015-07-11 paper/paper.tex
@@
 \rightsquigarrow
-$  $\rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e$}     &  \sc{cons} \\
+$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
 \hline
@@
 done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
-pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
+pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
 constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
@@
 simplification is continued till a fixed point is reached, after which
-the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
+the edges labeled \bcar \ and  \bcdr are deleted.  The details of the
 algorithm, its correctness and termination proofs are given by Karkare
 et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
-has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
-automaton, for every edge labelled \clazy, we check if the source node
+has  edges labeled  with  \acar,  \acdr\ and  \clazy\  only. In  this
+automaton, for every edge labeled \clazy, we check if the source node
 of the edge has a valid path to  a final final state.  If yes, we mark
 the mark  the source node as  final. Finally, we remove  all the edges
-labelled  \clazy\  and  convert  the automaton  into  a  deterministic
+labeled  \clazy\  and  convert  the automaton  into  a  deterministic
 automaton.    This   effectively  implements   the   $\hookrightarrow$
@@
 and the numbers for it  are not considered representative of practical
-programs.
+programs. All  our benchmarks were executed  on a desktop with  64 bit
+Ubuntu 14.04, 8 core Intel(R) Core(TM) i7-4770 CPU @ 3.40GHz, L2 cache
+of 8192 KB and 15922 MiB (16 GB) RAM.
 
@@
 practical garbage collector.
+>>>>>>> 86509285950f09b0dc95515465513023c4b6fa40
 
@@
     e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-  \rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
+  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
 \hline

### 2015-07-11 paper/paper.tex
Updated machine ininfo and merged changes

@@
 Table~\ref{tab:experimental-results}  compares the  garbage collection
-statisctics for  RGC and LGC  with reachability for  copying closures.
+statisctics for  RGC and LGC.
 We report the  number of GC events, average number  of cells reclaimed
@@
 practical garbage collector.
->>>>>>> 86509285950f09b0dc95515465513023c4b6fa40
 
@@
 \bibliographystyle{abbrv}
+\end{document}
 

### 2015-07-11 paper/paper.tex

@@
 Table~\ref{tab:experimental-results}  compares the  garbage collection
-statisctics for  RGC and LGC.
-We report the  number of GC events, average number  of cells reclaimed
-per  GC, average  number  of cells  touched per  GC,  total number  of
-dragged cells  and total time  to perform  all collections.  It  is no
-surprise  that  number  of  cells  reclaimed per  GC  is  higher  (and
-consequently the  number of  GCs is lower)  for LGC  based collection.
-LGC also reduces  the number of dragged  cells significantly. However,
-the cost for LGC is that per  GC execution time is more for LGC, which
-results in higher overall execution  time for garbage collection, even
-with reduced  number of collections.   However, LGC run time  is still
-competitive (slowdown within 5X of RGC  for worst case), and is better
-for   3    benchmarks   (best    case   2X   speedup).     Note   that
-\verb@gc_bench@~\cite{gc_bench}   is  a   synthestic  benchmark   that
-allocates complete binary trees of  various sizes.  These binary trees
-are  never used  by the  program. As  a result,  the benchmark  highly
-favours LGC.   We have included  this benchmark for  completeness only
-and the numbers for it  are not considered representative of practical
-programs. All  our benchmarks were executed  on a desktop with  64 bit
-Ubuntu 14.04, 8 core Intel(R) Core(TM) i7-4770 CPU @ 3.40GHz, L2 cache
-of 8192 KB and 15922 MiB (16 GB) RAM.
+statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
+average  number of  cells reclaimed  per GC,  average number  of cells
+touched  per GC,  total  number of  dragged cells  and  total time  to
+perform  all collections.   It is  no  surprise that  number of  cells
+reclaimed per  GC is  higher (and  consequently the  number of  GCs is
+lower)  for LGC  based collection.   LGC  also reduces  the number  of
+dragged cells significantly. However, the cost  for LGC is that per GC
+execution  time is  more  for  LGC, which  results  in higher  overall
+execution time  for garbage  collection, even  with reduced  number of
+collections.   However, LGC  run time  is still  competitive (slowdown
+within 5X of RGC for worst case), and is better for 3 benchmarks (best
+case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
+synthestic benchmark  that allocates complete binary  trees of various
+sizes.  These binary trees are never used by the program. As a result,
+the benchmark highly favours LGC.  We have included this benchmark for
+completeness  only  and   the  numbers  for  it   are  not  considered
+representative of practical programs. All our benchmarks were executed
+on  a desktop  with  64 bit  Ubuntu 14.04,  8  core Intel(R)  Core(TM)
+i7-4770 CPU @ 3.40GHz, L2 cache of 8192 KB and 15922 MiB (16 GB) RAM.
 

### 2015-07-11 paper/paper.tex
change s in fig5 and images.
closures.
Updated proof of correctness

@@
 
-A few observations: First observe that the transitions  which involving dereferencing and can lead to \bang are those which have a $H(\rho(x))$ term in their premise for some variable $x$.
+A few observations: First observe that the transitions which involving
+dereferencing and can lead to \bang are those which have a
+$H(...)$ term in the transition rule. 
+Of these most of the terms are of the form $H(\rho(x))$ which involve a root variable. They start from a root variable.
+There is exactly once occurance of $H(l)$ which means a derefencing from a heap location.
+
+In both the cases we show that the dereferences cannot lead to a bang state in a function call less program.
+
+For this we need the following lemma,
+1.Consider a let-defined variable $x \leftarrow...$ in .
+If there is an evaluation path in which the definition is followed by a use i.e an occurrence of (car x) (cdr x) (+ x y) if x & return x, then the demand on $x$ is non-null.
+2.
+Consider a let definition of the form $z->pi1:(cons x y) in ...$, then the demand on $(\cons\ x\ y)$ includes $0\alpha$ for some $\alpha$ if there is an evaluation path in which there is an occurrence of pi2: (car x) pi2 follows pi1.
 

### 2015-07-11 paper/paper.tex
one more ref.

--- 2015-07-11 paper/paper.tex
@@
           \begin{uprogram}
-          \UNL{1} (\DEFINE\ (\abigfunction~\py)
-          \UNL{2} $(Create\ a\ big\ closure\ and\ return)$ 
-          \end{uprogram}}
-        \renewcommand{\arraystretch}{1.1}{
-          \begin{uprogram}
             \UNL{1} (\DEFINE\ (\length~\xl)
@@
 	  \UNL{1} $(\DEFINE\ (\main)$
-           \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow
-(\abigfunction\ \acdr)$ \IN
+           \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$
+           (
+           \scalebox{0.8}{\psframebox[framearc=.5,
+	fillcolor=lightgray,fillstyle=solid,framesep=2pt]{%
+             \begin{tabular}{@{}c@{}}
+               {a BIG closure}
+             \end{tabular}}}
+           ) \IN  
           \UNL{3} \!\!$\pi_{10}\!\!:\, (\LET\  \pb\  \leftarrow ($+$\ \pa\ \acdr)$
@@
 been  studied earlier,  it was  mainly  in the  context of  imperative
-languages~\cite{Albert,  Hirzel}.  Due  to   the  presence  of  global
-variables  and mutation,  practical utility  of simple  liveness based
-techniques are found to be inefficient.
+languages~\cite{Albert, Hirzel,  khedker07heap}.  Due to  the presence
+of global variables and mutation, practical utility of simple liveness
+based techniques are found to be inefficient.
 

### 2015-07-11 paper/paper.tex
tables rewritten
Added proof of correctness

--- 2015-07-11 paper/paper.tex
@@
 \bibliographystyle{abbrv}
-\end{document}
+%\end{document}
 
@@
 
-A few observations: First observe that the transitions  which involving dereferencing and can lead to \bang are those which have a $H(\rho(x))$ term in their premise for some variable $x$.
+We observe that  the transitions involving dereferencing  and can lead
+to \bang are those which have  a $H(...)$ term in the transition rule.
+Of these most  of the terms are of the  form $H(\rho(x))$, involving a
+root variable.  There is exactly  one occurrence of $H(l)$ which means
+dereferencing using a heap location.
+
+In both  the cases  we show  that the dereferencing  cannot lead  to a
+\bang\ state in a function call less program.
+
+For this we need the following lemma,
+\begin{lemma}
+\label{lem:correctness}
+Consider, 
+\begin{enumerate}
+\item
+ A \LET\ definition $\pi_i:(\LET~x \leftarrow~s~\IN~e)$ such
+that there is  an evaluation path in $e$ in  which $\pi_i$ is followed
+by a  use i.e an  occurrence of $\pi_j:(\CAR\ x)$,  $\pi_j:(\CDR\ x)$,
+$\pi_j:(+\ x\ y)$, $\pi_j:(\SIF\ x\ e1\ e2)$ or $\pi_j:(\SRETURN\ x)$
+\item
+ A \LET\ definition $\pi_i(\LET~z  \leftarrow~(\CONS~x~y)~\IN~e$, such that
+ $e$ includes  an evaluation path in  which there is an  occurrence of
+ $\pi_j: (\CAR~z)$ and $\pi_j$ follows $\pi_i$.
+\end{enumerate},
+the demand on $x$ is non-null between $\pi_i$ and $\pi_j$.
+\end{lemma} 
+\begin{proof}
+In the first case it is clear  that at all points $\pi_k$ which occurs
+in between $\pi_i$ and $\pi_j$, $x$  has non-null demand as $x$ is used
+directly in a strict computation.
+
+In the second case, it can be seen that the liveness of $x$ due to the
+\CONS\ is $\bcar\sigma$, where $\sigma$ is the demand on $z$. $\sigma$
+definitely  includes  the  demand  $\acar\alpha$ due  to  its  use  in
+$\pi_j$, therefore the demand on $x$ is at least $\alpha$.
+
+ Since we do not kill the liveness of any variable along an evaluation
+ path, and all program points along  an evaluation path share the same
+ liveness environment, at any point along the evaluation path $x$ will
+ not be set to $\bot$.
+\end{proof}
+
+For a program  executing according to {\em minefield  semantics}, if a
+\bang\ has to occur then some variable which is dereferenced along the
+evaluation path  must be set to  $\bot$, between its creation  and use
+points.  This can  only happen  if the  demand on  the variable  being
+dereferenced  was null.  From  the  lemma~\ref{lem:correctness} it  is
+clear that this cannot happen.
+
+The rest  of the proof depends  on converting a program  with function
+calls to an equivalent program without function calls for which we 
+refer the user to~\cite{asati14lgc}.
 

### 2015-07-11 paper/paper.tex
table split.

--- 2015-07-11 paper/paper.tex
@@
 \end{algorithm}
+
 \begin{table*}[ht!]
-\caption{Experimental results comparing RGC and LGC}
-\label{tab:experimental-results}
+\caption{Comparing LGC times for different strategies}
+\label{tab:compare-live-clo-reach-clo}
 \centering
+\begin{tabular}
+{| l | r | r | r | r |}
+\hline
+Prog Name & Heap size & Num GC & Time taken using reachability for closure  & Time taken using liveness for closure \\ \hline
+\verb@nqueens (reduced) @ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
+\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
+\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
+\end{tabular}
+\end{table*}
+
+\begin{table*}[ht!]
+\caption{Compile time and Garbage Collection statistics}
+\label{tab:exp-results}
+\centering
+ \begin{tabular}{|l||c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c|} \hline
+   \multicolumn{1}{|r||}{Program} & \verb@lambda@ & \verb@nperm@ & \verb@treejoin@ & \verb@lcss@ & \verb@sudoku@ & \verb@fibheap@ & \verb@nqueens@ & \verb@knightstour@ & \verb@gc_bench@ \\ \hline\hline
+   \#CFG Nonterminals            & 1078          & 794          & 709             & 839         & 1758          & 780            & 529            & 659                & 417             \\ \hline
+   \#CFG Rules                   & 1823          & 1038         & 1610            & 1752        & 2509          & 1358           & 748            & 909                & 486             \\  \hline 
+   \#DFA States                  & 3312          & 2055         & 2332            & 2340        & 5672          & 2254           & 1229           & 1742               & 729             \\\hline
+   \#DFA Transitions             & 5792          & 3632         & 4235            & 4261        & 10857         & 4076           & 2078           & 3101               & 1124            \\\hline
+   DFA Generation Time (sec)      & 14.7          & 1.1          & 2690.5          & 13.3        & 798.5         & 49.8           & 0.5            & 5.1                & 0.1             \\\hline
+ \end{tabular}\\
+ (a) Data for Compile-time Liveness Analysis \\[3pt]
 \include{autotable}
@@
 
-\begin{table*}[ht!]
-\caption{Comparing LGC times for different strategies}
-\label{tab:compare-live-clo-reach-clo}
-\centering
-\begin{tabular}
-{| l | r | r | r | r |}
-\hline
-Prog Name & Heap size & Num GC & Time taken using reachability for closure  & Time taken using liveness for closure \\ \hline
-\verb@nqueens(8)@ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
-\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
-\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
-\end{tabular}
-\end{table*}
 To compare the  effect of using reachability to  copy closures instead

### 2015-07-11 paper/paper.tex
stats for other table.

--- 2015-07-11 paper/paper.tex
@@
 \begin{table*}[ht!]
-\caption{Compile time and Garbage Collection statistics}
+\caption{Liveness Analysis and Garbage Collection statistics}
 \label{tab:exp-results}
@@
  \end{tabular}\\
- (a) Data for Compile-time Liveness Analysis \\[3pt]
+ (a) Data for Liveness Analysis \\[3pt]
 \include{autotable}
@@
 
-%% Liveness automata  does not exist for  the print function  as it is
-%% not part of the executing  program.  Since the print function makes
-%% everything  reachable  from a  cell  live  we  can copy  all  cells
-%% reachable from the reference.
-
-%% Consider  the case  where a  \CONS\ cells  is being  printed  and a
-%% garbage collection  is triggered when  printing the \CDR\  part. In
-%% this case, since  the refernce to the \CONS\ cells  is still on the
-%% print stack both the \CAR\ and the \CDR\ parts are reachable but we
-%% know that the \CAR\ part has already been printed and can be safely
-%% garbage collected. A liveness  analysis of the print function would
-%% have given  the same information.   In our implementation  we mimic
-%% the  behavior of the  liveness analysis  by using  a flag  in every
-%% \CONS\ cell which  says whether the \CAR\ part  has been printed or
-%% not. For any \CONS\ cell if  the \CAR\ part has not been printed, a
-%% reference  to the  \CAR\ part  will also  be on  the stack  but the
-%% \CDR\ part has to be explicitly copied.  In case the \CAR\ part has
-%% been printed,  then a reference  to the \CDR\  part will be  on the
-%% stack and  since the \CAR\ part  is not required anymore  we do not
-%% copy anything.  This  optimization subsumes tail call optimization.
-%% Again, experimental results show that this optimization minimizes a
-%% lot of copying.  \cred{Very sketchy: Make another attempt to write}
-%% \hrule
-
-%% \comment{ Unlike  eager languages where  GC happens only  at \CONS~
-%%   points or at function calls,  every \LET~ point is a potential GC
-%%   point.  The number of automata that needs to be stored is reduced
-%%   by sharing  automata along all  the program points along  an {\em
-%%   E-path}.
-
-%% By the time the program starts executing, the liveness automata are
-%% created  and stored  in a  file.  This  is re-used  every  time the
-%% program executes.   An important difference in GC  of lazy language
-%% compared to an  eager language is the point  of garbage collection.
-%% In case  of eager languages  only \CONS~ points and  function calls
-%% are the  GC points. In a  lazy language every \LET~  statement is a
-%% potential  GC  point. Hence  we  need  to  store liveness  automata
-%% corresponding to  every \LET~ statement in the  program. This leads
-%% to an  explosion in  the number of  automata. From  our observation
-%% that  along  an {\em  E-path}  all  program  points have  the  same
-%% liveness, we  can reduce the  number of automata state  required by
-%% sharing  the same  automata for  all program  points along  an {\em
-%% E-path}.} In the  example program, we only need  to create automata
-%% for program  points \ldots, since  they are the only  distinct {\em
-%% E-path}s in the program.
-
 \section{Experimental Evaluation}
 \label{sec:experiments}
-%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
-%% we  have  built a  prototype  consisting  of  an interpreter  for  our
-%% language,  a  liveness  analyzer  and  a copying  collector  that  can
-%% optionally use the results of liveness analysis for marking instead of
-%% reachability. When the collector uses liveness for marking, we call it
-%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
-%%   reachability-based  collector}  (RGC).   The  copying  collector  is
-%% neither incremental nor generational.  As a consequence, any cell that
-%% becomes unreachable or  dead is assuredly collected in  the next round
-%% of garbage collection.}
-
 To   test  the   utility  of   our  method   we  have   implemented  a
@@
 some external programs manually converted to ANF.
-
-%% \comment{
-%% While exploring the activation record of a function, LGC uses the DFA
-%% at  the latest  program  point traversed  in the function's body.
-%% Thus, if  LGC
-%% is invoked  at the program point   $\pi$,  it would  use  the
-%% liveness  DFA at  $\pi$ itself for exploring the root set of the
-%% current activation record. For any other activation record in the
-%% stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
-%% call-chain,
-%% the DFA at the point of call to  ${\mathit g}$ would be used.}
 
@@
 
-%% \comment{
-%% Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
-%% point  pair ($x$,  $\pi$).  We  assume that  there is  a function
-%% $\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
-%% $\dfa{\pi}{x}$.
-%% Considering   a   DFA   as   a   table,
-%% $\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
-%% state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
-%% \acar\      or     \acdr.       We      shall     also      write
-%% $\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
-%% indicating  whether there  is a  transition from  $q$ on
-%% $\mathit{sym}$.  The LGC action to chase the root variable $x$ at
-%% $\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
-%% empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
-%% $\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
-%% in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
-%% pointer to $x$.
-%% The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
-%% $x$  in the  other semi-space  and returns  the new  address.  It
-%% hides  details such as  returning the  forwarding pointer  if the
-%% value  of $x$  is  already copied,  and  creating the  forwarding
-%% pointer otherwise.}
-
 \subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
@@
 To compare the  effect of using reachability to  copy closures instead
-of liveness we take some representative programs from our benchmark suite and
-execute       them      with       both.        As      shown       in
+of liveness  we take some  representative programs from  our benchmark
+suite    and    execute    them    with    both.     As    shown    in
 Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
@@
 
-Table~\ref{tab:experimental-results}  compares the  garbage collection
+The statistics related to compile  time liveness analysis are shown in
+Table~\ref{tab:exp-results}(a). We observe that the analysis requires
+a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
+The bottleneck in our analysis is the NFA to DFA conversion
+(worst-case exponential behaviour). To avoid this compile-time
+overhead, we have added a caching mechanism in our simulator that can
+reuse the DFAs from earlier analysis if the program has not changed..
+
+Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
 statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
@@
 (blue line), the dip is specially visible for the benchmarks where LGC
-outperforms                      RGC                     substantially
-(Table~\ref{tab:experimental-results}). The graphs  also include curve
-for reachable cells  (black, obtained approximately by  forcing RGC to
-run at  a very high frequency)  and the live cells  for particular run
-(grey, obtained  by post processing  the heap access  at the end  of a
-program).
+outperforms  RGC substantially.   The  graphs also  include curve  for
+reachable cells (black,  obtained approximately by forcing  RGC to run
+at a very high frequency) and the live cells for particular run (grey,
+obtained by post processing the heap access at the end of a program).
 

### 2015-07-11 paper/paper.tex
reorg.

@@
 \renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-
-\section{Liveness}
-\label{sec:liveness}
-
-A variable is {\em live} if there  is a possibility of its value being
-used in  future computations and  dead if  it is definitely  not used.
-Heap-allocated data needs a richer model of liveness which talks about
-liveness of  references.  Using  $\acar$, $\acdr$ to  represent access
-using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
-reachable from a  variable can be represented by a  set of {\em access
-  paths}     i.e.     prefix-closed     set     of    strings     from
-$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
-$\{\epsilon, \acdr,  \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
-future computations can only refer up  to the second and third members
-of $x$.  A  {\em liveness environment} is a mapping  from variables to
-subsets of $\{\acar,\acdr\}^\ast$,  but often expressed as  a set, for
-example by writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$
-instead       of       $[x       \mapsto\{\epsilon,\acdr,\acdr\acdr\},
-  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $x \mapsto
-\{\epsilon\}$ represents access using $x$  itself and $x \mapsto \{\}$
-indicates that $x$  is dead.  We associate  liveness environments with
-program points.  The  liveness at $\pi:e$ is the  liveness just before
-executing $e$.
-
-A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
-represents the future use of variables, a demand represents the future
-use of the value of an expression.  The demand on an expression $e$ is
-again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
-which the context of $e$ may explore of $e$'s result.  To see the need
-for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
-(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context of  this
-expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
-  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
-$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes the  demand on
-$(\CDR~y)$.    This,  in   turn,  generates   the  liveness   $\lbrace
-y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
-of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
-\rbrace$  paths  of  the  result of  $(\CDR~y)$.  In  strong  liveness
-analysis       (dual      of       classical      faint       variable
-analysis~\cite{horwitz.faint}),  $y$ and  $z$  are live  at the  entry
-$\pi: x:=y+z$, if  and only if $x$  is live at exit of  $\pi$.  In our
-terminology, the  liveness $\lbrace  \epsilon\rbrace $  of $x$  at the
-exit from $n$ becomes the demand on $y+z$, and this, in turn generates
-the liveness $\lbrace y.\epsilon, z.\epsilon  \rbrace$ at the entry of
-$n$.
-
-We follow the notations appearing in \cite{asati14lgc}. In particular,
-We use $\sigma$  to range over demands, $\alpha$ to  range over access
-paths and  $\Lv$ to  range over  liveness environments.   The notation
-$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
-\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
-abuse notation to  juxtapose an edge label and a  set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
-
-\subsection{Liveness Analysis for lazy languages}
-\label{sec:liveness-analysis}
-
 %==============================================================
@@
 
+\section{Liveness}
+\label{sec:liveness}
+
+A variable is {\em live} if there  is a possibility of its value being
+used in  future computations and  dead if  it is definitely  not used.
+Heap-allocated data needs a richer model of liveness which talks about
+liveness of  references.  Using  $\acar$, $\acdr$ to  represent access
+using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
+reachable from a  variable can be represented by a  set of {\em access
+  paths}     i.e.     prefix-closed     set     of    strings     from
+$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
+$\{\epsilon, \acdr,  \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
+future computations can only refer up  to the second and third members
+of $x$.  A  {\em liveness environment} is a mapping  from variables to
+subsets of $\{\acar,\acdr\}^\ast$,  but often expressed as  a set, for
+example by writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$
+instead       of       $[x       \mapsto\{\epsilon,\acdr,\acdr\acdr\},
+  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $x \mapsto
+\{\epsilon\}$ represents access using $x$  itself and $x \mapsto \{\}$
+indicates that $x$  is dead.  We associate  liveness environments with
+program points.  The  liveness at $\pi:e$ is the  liveness just before
+executing $e$.
+
+A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
+represents the future use of variables, a demand represents the future
+use of the value of an expression.  The demand on an expression $e$ is
+again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
+which the context of $e$ may explore of $e$'s result.  To see the need
+for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
+(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context of  this
+expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
+  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
+$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes the  demand on
+$(\CDR~y)$.    This,  in   turn,  generates   the  liveness   $\lbrace
+y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
+of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
+\rbrace$  paths  of  the  result of  $(\CDR~y)$.  In  strong  liveness
+analysis       (dual      of       classical      faint       variable
+analysis~\cite{horwitz.faint}),  $y$ and  $z$  are live  at the  entry
+$\pi: x:=y+z$, if  and only if $x$  is live at exit of  $\pi$.  In our
+terminology, the  liveness $\lbrace  \epsilon\rbrace $  of $x$  at the
+exit from $n$ becomes the demand on $y+z$, and this, in turn generates
+the liveness $\lbrace y.\epsilon, z.\epsilon  \rbrace$ at the entry of
+$n$.
+
+We follow the notations appearing in \cite{asati14lgc}. In particular,
+We use $\sigma$  to range over demands, $\alpha$ to  range over access
+paths and  $\Lv$ to  range over  liveness environments.   The notation
+$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
+\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
+abuse notation to  juxtapose an edge label and a  set of access paths:
+$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
+
+\subsection{Liveness Analysis for lazy languages}
+\label{sec:liveness-analysis}
+
 Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
@@
 \section{Computing liveness and its encoding as DFA}\label{sec:computing}
-Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
-and  \mycomment{\color  {red}  proved  it  correct  with  reference  to  a  {\em
+Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis%
+  \mycomment{\color  {red}  and proved  it  correct  with  reference  to  a  {\em
     minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
@@
 
-\begin{table*}[ht!]
-\caption{Comparing LGC times for different strategies}
+\begin{table}[ht!]
+\caption{Time taken by LGC for different strategies.  The programs are
+  modified (smaller) instances of original benchmarks.}
 \label{tab:compare-live-clo-reach-clo}
@@
 \begin{tabular}
-{| l | r | r | r | r |}
-\hline
-Prog Name & Heap size & Num GC & Time taken using reachability for closure  & Time taken using liveness for closure \\ \hline
-\verb@nqueens (reduced) @ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
-\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
-\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
+{| l | r | r | r |}
+\hline
+Program        & \#GC & \multicolumn{2}{c|}{Closure collected using} \\ \cline{3-4}
+               &      & Reachability & Liveness                      \\ \hline
+\verb@nqueens@ & 183  & 0.009 sec    & 0.228 sec                     \\\hline
+\verb@lambda@  & 11   & 0.001 sec    & 4.992 sec                     \\ \hline
+\verb@lcss@    & 1    & 0.000 sec    & 34.668 sec                    \\ \hline
 \end{tabular}
-\end{table*}
+\end{table}
+
+Algorithm~\ref{algo:lgc-b} describes a garbage collection scheme where
+the liveness automata is consulted  only for \CONS\ cells; and assumes
+everything under  a closure to be  live and copies everything  that is
+reachable.
+
+A consequence of  copying everything under a closure is  that we might
+reach  a non-live  cell that  has been  reclaimed earlier  by liveness
+based GC.   This can  happen if  a \CONS\  cell having  partially live
+sub-structure  is copied  using liveness  during a  collection and  it
+subsequently  becomes part  of a  live closure.   During a  subsequent
+collection, the collector may try to copy everything under the closure
+including the dead  sub-structure of the \CONS\ cell.   To avoid this,
+we ensure  that any non-live  references are not carried  across GC's.
+We do this by scanning the live  part of the buffer after every GC and
+nullifying any  reference which is  not live. This is  correct because
+any reference which  is dead during a collection is  guaranteed not be
+dereferenced any further by the program.
+
+Another advantage of  this mixed garbage collection scheme  is that it
+avoids the problem of multiple  explorations from the same cell.  This
+is done by maintaining a flag in  each cell to indicate whether it was
+copied using reachability.
+
+\subsection{Garbage collection for references on print stack}
+In lazy  languages the evaluation of  closures is driven  by the print
+function~\cite{Jones87}.  Printing of  atomic values is
+simple  and  does  not  trigger  a garbage  collection.   In  case  of
+\CONS\ cells, the \CAR\ is first  printed followed by \CDR.  In a lazy
+language, both \CAR\ and \CDR\ could be closures requiring evaluation,
+and this  may trigger  a garbage collection.   When this  happens.  we
+have to consider  any references that might be on  the print stack and
+copy them. We extend the liveness analysis to the print function and
+use its result during garbage collection.
+
+\section{Experimental Evaluation}
+\label{sec:experiments}
+To   test  the   utility  of   our  method   we  have   implemented  a
+garbage-collector for a simple scheme-like first order language having
+non-strict semantics. It consists of an interpreter, liveness analyzer
+and  a  garbage-collector  that  can optionally  use  reachability  or
+liveness.  Our benchmark consists  of programs from nofib~\cite{nofib}
+suite and some  external programs manually converted to  ANF.  All our
+benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
+i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
+Ubuntu 14.04.
+
+The process of liveness-based garbage collection involves going through 
+each activation record on the stack and exploring each variable in the 
+current activation record (root set). The liveness of each variable is 
+determined using the program point and the variable name. The liveness 
+of all variables are stored as DFA. These DFA are then encoded as a 
+table. A variable is live only if an entry is found in this table.
+All cells that are live are copied to the live semi-space. In case of 
+\CONS\ cells the \CAR\ and \CDR\ pointers are chased and if they are 
+live the copied \CONS\ cell will get the updated addresses of its 
+\CAR\ and \CDR\ fields.
+
+\subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
+
+To compare the  effect of using reachability to  copy closures instead
+of liveness  we take some  representative programs from  our benchmark
+suite    and    execute    them    with    both.     As    shown    in
+Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
+collections in both cases are same but copying closures using liveness
+takes a lot more time compared to reachability based copying.
+
+The number  of garbage  collections is same  since we  approximate the
+liveness of  a closure  to the  liveness at  its creation  point.  The
+liveness  we get  is usually  grossly over  approximated.  Using  this
+liveness may not  give too much of a benefit  compared to reachability
+as nearly everything would be live at its point of creation.
+
+As  mentioned   in  Section~\ref{sec:live-clo},  using   liveness  for
+closures  may  entail  multiple  traversals over  the  same  structure
+consuming more  time.  The  advantage of avoiding  multiple traversals
+can be clearly seen from the lcss example where a single GC takes less
+than a millisecond for reachability  based closure collection where as
+the  liveness  based collection  is  of  the  order of  seconds.  Lazy
+languages tend  to have a  lot of sharing  and hence using  a liveness
+based collection for closures is impractical.
+
+Therefore we believe  that using reachability for closures  is a sweet
+spot   in   terms   of   performance  for   liveness   based   garbage
+collectors.  For the  remainder  of  the section  all  LGC results  we
+discuss uses  a mixed-mode collector--liveness for  evaluated data and
+reachability for closures.
 
 \begin{table*}[ht!]
-\caption{Liveness Analysis and Garbage Collection statistics}
+\caption{Statistics for liveness analysis and garbage collection}
 \label{tab:exp-results}
@@
 
-  Algorithm~\ref{algo:lgc-b}  describes  a garbage  collection  scheme
-  where the liveness automata is  consulted only for \CONS\ cells; and
-  assumes everything under a closure  to be live and copies everything
-  that is reachable.
-
-A consequence of  copying everything under a closure is  that we might
-reach  a non-live  cell that  has been  reclaimed earlier  by liveness
-based GC.   This can  happen if  a \CONS\  cell having  partially live
-sub-structure  is copied  using liveness  during a  collection and  it
-subsequently  becomes part  of a  live closure.   During a  subsequent
-collection, the collector may try to copy everything under the closure
-including the dead  sub-structure of the \CONS\ cell.   To avoid this,
-we ensure  that any non-live  references are not carried  across GC's.
-We do this by scanning the live  part of the buffer after every GC and
-nullifying any  reference which is  not live. This is  correct because
-any reference which  is dead during a collection is  guaranteed not be
-dereferenced any further by the program.
-
-Another advantage of  this mixed garbage collection scheme  is that it
-avoids the problem of multiple  explorations from the same cell.  This
-is done by maintaining a flag in  each cell to indicate whether it was
-copied using reachability.
-
-\subsection{Garbage collection for references on print stack}
-In lazy  languages the evaluation of  closures is driven  by the print
-function~\cite{Jones87}.  Printing of  atomic values is
-simple  and  does  not  trigger  a garbage  collection.   In  case  of
-\CONS\ cells, the \CAR\ is first  printed followed by \CDR.  In a lazy
-language, both \CAR\ and \CDR\ could be closures requiring evaluation,
-and this  may trigger  a garbage collection.   When this  happens.  we
-have to consider  any references that might be on  the print stack and
-copy them. We extend the liveness analysis to the print function and
-use its result during garbage collection.
-
-\section{Experimental Evaluation}
-\label{sec:experiments}
-To   test  the   utility  of   our  method   we  have   implemented  a
-garbage-collector for a simple scheme-like first order language having
-non-strict semantics. It consists of an interpreter, liveness analyzer
-and a  garbage-collector that can  optionally use  reachability or
-liveness.  Our  benchmark consists  of programs  from nofib~\cite{nofib}  suite and
-some external programs manually converted to ANF.
-
-The process of liveness-based garbage collection involves going through 
-each activation record on the stack and exploring each variable in the 
-current activation record (root set). The liveness of each variable is 
-determined using the program point and the variable name. The liveness 
-of all variables are stored as DFA. These DFA are then encoded as a 
-table. A variable is live only if an entry is found in this table.
-All cells that are live are copied to the live semi-space. In case of 
-\CONS\ cells the \CAR\ and \CDR\ pointers are chased and if they are 
-live the copied \CONS\ cell will get the updated addresses of its 
-\CAR\ and \CDR\ fields.
-
-\subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
-
-To compare the  effect of using reachability to  copy closures instead
-of liveness  we take some  representative programs from  our benchmark
-suite    and    execute    them    with    both.     As    shown    in
-Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
-collections in both cases are same but copying closures using liveness
-takes a lot more time compared to reachability based copying.
-
-The number  of garbage  collections is same  since we  approximate the
-liveness of  a closure  to the  liveness at  its creation  point.  The
-liveness  we get  is usually  grossly over  approximated.  Using  this
-liveness may not  give too much of a benefit  compared to reachability
-as nearly everything would be live at its point of creation.
-
-As  mentioned   in  Section~\ref{sec:live-clo},  using   liveness  for
-closures  may  entail  multiple  traversals over  the  same  structure
-consuming more  time.  The  advantage of avoiding  multiple traversals
-can be clearly seen from the lcss example where a single GC takes less
-than a millisecond for reachability  based closure collection where as
-the  liveness  based collection  is  of  the  order of  seconds.  Lazy
-languages tend  to have a  lot of sharing  and hence using  a liveness
-based collection for closures is impractical.
-
-Therefore we believe  that using reachability for closures  is a sweet
-spot   in   terms   of   performance  for   liveness   based   garbage
-collectors.  For the  remainder  of  the section  all  LGC results  we
-discuss uses  a mixed-mode collector--liveness for  evaluated data and
-reachability for closures.
-%%
-%% \comment{
-%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
-%% in the heap over time for RGC and LGC---here time is measured in terms
-%% of the number of cons cells allocated. In addition, they also show the
-%% number of  reachable cells and the  number of cells  that are actually
-%% live\footnote{This   is  statically   approximated  by   our  liveness
-%%   analysis}.  Since  the programs have  different memory requirements,
-%% we have tuned the size of heap for each program to ensure a reasonable
-%% number of  collections. An invocation  of RGC decreases the  number of
-%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
-%% invocation of LGC decreases the number  of heap cells to no lower than
-%% the curve of live cells.
-
-%% To construct the  reachable and live curves, we  record for every cell
-%% its creation time (\CreateTime), its last use time (\UseTime), and the
-%% earliest time  when the  cell becomes unreachable  and can  be garbage
-%% collected (\CollTime).  For accurate recording of  \CollTime, we force
-%% frequent invocations  of a reachability based collector  in a separate
-%% run.     A    cell   is    live    at    time   \mbox{\exectime}    if
-%% \mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
-%%   \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
-%%   \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
-
-%% The benchmark programs are drawn from the {\tt no-fib} suite and other
-%% sources and have been  manually converted to  ANF.  All  graphs  except
-%%  {\tt  fibheap}  show  fewer  garbage
-%% collector  invocations for  LGC. The  number  of reachable  cells in
-%% {\tt
-%%   fibheap} increases  initially, comes  down before the  heap becomes
-%% gets
-%% full and then  continues to remains low.  This results in  a single
-%% garbage
-%% collection in both RGC and LGC.  The graphs also show the precision of
-%% our liveness analysis.  For  all programs  except  {\tt nperms} and
-%% {\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
-%% manages to collect a good portion of the data that are not live.}
-%%
-
 \subsection{Results}
+
+The statistics related to compile  time liveness analysis are shown in
+Table~\ref{tab:exp-results}(a). We observe that the analysis requires
+a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
+The bottleneck in our analysis is the NFA to DFA conversion
+(worst-case exponential behaviour). To avoid this compile-time
+overhead, we have added a caching mechanism in our simulator that can
+reuse the DFAs from earlier analysis if the program has not changed..
+
+Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
+statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
+average  number of  cells reclaimed  per GC,  average number  of cells
+touched  per GC,  total  number of  dragged cells  and  total time  to
+perform  all collections.   It is  no  surprise that  number of  cells
+reclaimed per  GC is  higher (and  consequently the  number of  GCs is
+lower)  for LGC  based collection.   LGC  also reduces  the number  of
+dragged cells significantly. However, the cost  for LGC is that per GC
+execution  time is  more  for  LGC, which  results  in higher  overall
+execution time  for garbage  collection, even  with reduced  number of
+collections.   However, LGC  run time  is still  competitive (slowdown
+within 5X of RGC for worst case), and is better for 3 benchmarks (best
+case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
+synthestic benchmark  that allocates complete binary  trees of various
+sizes.  These binary trees are never used by the program. As a result,
+the benchmark highly favours LGC.  We have included this benchmark for
+completeness  only  and   the  numbers  for  it   are  not  considered
+representative of practical programs.
+
 \newcommand{\hgt}{3cm}
@@
 \end{figure*}
-
-The statistics related to compile  time liveness analysis are shown in
-Table~\ref{tab:exp-results}(a). We observe that the analysis requires
-a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
-The bottleneck in our analysis is the NFA to DFA conversion
-(worst-case exponential behaviour). To avoid this compile-time
-overhead, we have added a caching mechanism in our simulator that can
-reuse the DFAs from earlier analysis if the program has not changed..
-
-Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
-statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
-average  number of  cells reclaimed  per GC,  average number  of cells
-touched  per GC,  total  number of  dragged cells  and  total time  to
-perform  all collections.   It is  no  surprise that  number of  cells
-reclaimed per  GC is  higher (and  consequently the  number of  GCs is
-lower)  for LGC  based collection.   LGC  also reduces  the number  of
-dragged cells significantly. However, the cost  for LGC is that per GC
-execution  time is  more  for  LGC, which  results  in higher  overall
-execution time  for garbage  collection, even  with reduced  number of
-collections.   However, LGC  run time  is still  competitive (slowdown
-within 5X of RGC for worst case), and is better for 3 benchmarks (best
-case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
-synthestic benchmark  that allocates complete binary  trees of various
-sizes.  These binary trees are never used by the program. As a result,
-the benchmark highly favours LGC.  We have included this benchmark for
-completeness  only  and   the  numbers  for  it   are  not  considered
-representative of practical programs. All our benchmarks were executed
-on  a desktop  with  64 bit  Ubuntu 14.04,  8  core Intel(R)  Core(TM)
-i7-4770 CPU @ 3.40GHz, L2 cache of 8192 KB and 15922 MiB (16 GB) RAM.
 

### 2015-07-11 paper/paper.tex
spell check.

@@
 
-When  there  are  multiple  occurences of  \bcar,  \bcdr\  and  \clazy,
+When  there  are  multiple  occurrences of  \bcar,  \bcdr\  and  \clazy,
 $\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
@@
 simplified using  the $\hookrightarrow$ rules.  Finally the simplified
-NFAs are converted to DFA.
+NFA are converted to DFA.
 
@@
 
-The strongly regular grammar is converted  into a set of NFAs, one for
+The strongly regular grammar is converted  into a set of NFA, one for
 each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
-done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
+done on the NFA by  repeatedly introducing $\epsilon$ edges to bypass
 pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
@@
 overhead, we have added a caching mechanism in our simulator that can
-reuse the DFAs from earlier analysis if the program has not changed..
+reuse the DFA from earlier analysis if the program has not changed..
 
@@
 case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
-synthestic benchmark  that allocates complete binary  trees of various
+synthetic benchmark  that allocates complete binary  trees of various
 sizes.  These binary trees are never used by the program. As a result,
@@
 One area  of concern is the  huge gap between the  actual liveness and
-the liveness  percieved by our  collector.  In  case of LGC  for eager
+the liveness  perceived by our  collector.  In  case of LGC  for eager
 languages~\cite{asati14lgc} the gap was  very narrow and almost touched
@@
   closures  are  handled  during  garbage  collection  and  liveness  is
-  propogated inside closures.
+  propagated inside closures.
 
@@
 
-%Add graphs for memort usage LGC vs RGC
+%Add graphs for memory usage LGC vs RGC
 %Add graphs/table showing the GC times and other statistics

### 2015-07-11 paper/paper.tex

@@
 marking approaches all rely on an efficient and precise alias analysis
-and in the absence of it cannot provide notable improvement.
+and in the  absence of it cannot provide  notable improvement. Another
+important approach  advocated by  Hofmann~\cite{HofmannJ03} is  to use
+linear  typing to  analyze  first-order programs  for  heap usage  and
+annotated  them. The  system  then uses  these  annotations to  re-use
+allocated memory  cells instead  of requesting  for newer  cells. This
+requires the  user to write programs  in a specific way  and hence may
+not be practical
 

### 2015-07-11 paper/paper.tex
Updated related work

@@
 \renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
+
+\section{Liveness}
+\label{sec:liveness}
+
+A variable is {\em live} if there  is a possibility of its value being
+used in  future computations and  dead if  it is definitely  not used.
+Heap-allocated data needs a richer model of liveness which talks about
+liveness of  references.  Using  $\acar$, $\acdr$ to  represent access
+using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
+reachable from a  variable can be represented by a  set of {\em access
+  paths}     i.e.     prefix-closed     set     of    strings     from
+$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
+$\{\epsilon, \acdr,  \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
+future computations can only refer up  to the second and third members
+of $x$.  A  {\em liveness environment} is a mapping  from variables to
+subsets of $\{\acar,\acdr\}^\ast$,  but often expressed as  a set, for
+example by writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$
+instead       of       $[x       \mapsto\{\epsilon,\acdr,\acdr\acdr\},
+  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $x \mapsto
+\{\epsilon\}$ represents access using $x$  itself and $x \mapsto \{\}$
+indicates that $x$  is dead.  We associate  liveness environments with
+program points.  The  liveness at $\pi:e$ is the  liveness just before
+executing $e$.
+
+A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
+represents the future use of variables, a demand represents the future
+use of the value of an expression.  The demand on an expression $e$ is
+again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
+which the context of $e$ may explore of $e$'s result.  To see the need
+for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
+(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context of  this
+expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
+  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
+$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes the  demand on
+$(\CDR~y)$.    This,  in   turn,  generates   the  liveness   $\lbrace
+y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
+of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
+\rbrace$  paths  of  the  result of  $(\CDR~y)$.  In  strong  liveness
+analysis       (dual      of       classical      faint       variable
+analysis~\cite{horwitz.faint}),  $y$ and  $z$  are live  at the  entry
+$\pi: x:=y+z$, if  and only if $x$  is live at exit of  $\pi$.  In our
+terminology, the  liveness $\lbrace  \epsilon\rbrace $  of $x$  at the
+exit from $n$ becomes the demand on $y+z$, and this, in turn generates
+the liveness $\lbrace y.\epsilon, z.\epsilon  \rbrace$ at the entry of
+$n$.
+
+We follow the notations appearing in \cite{asati14lgc}. In particular,
+We use $\sigma$  to range over demands, $\alpha$ to  range over access
+paths and  $\Lv$ to  range over  liveness environments.   The notation
+$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
+\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
+abuse notation to  juxtapose an edge label and a  set of access paths:
+$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
+
+\subsection{Liveness Analysis for lazy languages}
+\label{sec:liveness-analysis}
+
 %==============================================================
@@
 
-\section{Liveness}
-\label{sec:liveness}
-
-A variable is {\em live} if there  is a possibility of its value being
-used in  future computations and  dead if  it is definitely  not used.
-Heap-allocated data needs a richer model of liveness which talks about
-liveness of  references.  Using  $\acar$, $\acdr$ to  represent access
-using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
-reachable from a  variable can be represented by a  set of {\em access
-  paths}     i.e.     prefix-closed     set     of    strings     from
-$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
-$\{\epsilon, \acdr,  \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
-future computations can only refer up  to the second and third members
-of $x$.  A  {\em liveness environment} is a mapping  from variables to
-subsets of $\{\acar,\acdr\}^\ast$,  but often expressed as  a set, for
-example by writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$
-instead       of       $[x       \mapsto\{\epsilon,\acdr,\acdr\acdr\},
-  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $x \mapsto
-\{\epsilon\}$ represents access using $x$  itself and $x \mapsto \{\}$
-indicates that $x$  is dead.  We associate  liveness environments with
-program points.  The  liveness at $\pi:e$ is the  liveness just before
-executing $e$.
-
-A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
-represents the future use of variables, a demand represents the future
-use of the value of an expression.  The demand on an expression $e$ is
-again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
-which the context of $e$ may explore of $e$'s result.  To see the need
-for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
-(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context of  this
-expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
-  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
-$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes the  demand on
-$(\CDR~y)$.    This,  in   turn,  generates   the  liveness   $\lbrace
-y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
-of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
-\rbrace$  paths  of  the  result of  $(\CDR~y)$.  In  strong  liveness
-analysis       (dual      of       classical      faint       variable
-analysis~\cite{horwitz.faint}),  $y$ and  $z$  are live  at the  entry
-$\pi: x:=y+z$, if  and only if $x$  is live at exit of  $\pi$.  In our
-terminology, the  liveness $\lbrace  \epsilon\rbrace $  of $x$  at the
-exit from $n$ becomes the demand on $y+z$, and this, in turn generates
-the liveness $\lbrace y.\epsilon, z.\epsilon  \rbrace$ at the entry of
-$n$.
-
-We follow the notations appearing in \cite{asati14lgc}. In particular,
-We use $\sigma$  to range over demands, $\alpha$ to  range over access
-paths and  $\Lv$ to  range over  liveness environments.   The notation
-$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
-\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
-abuse notation to  juxtapose an edge label and a  set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
-
-\subsection{Liveness Analysis for lazy languages}
-\label{sec:liveness-analysis}
-
 Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
@@
 \section{Computing liveness and its encoding as DFA}\label{sec:computing}
-Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis%
-  \mycomment{\color  {red}  and proved  it  correct  with  reference  to  a  {\em
+Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis
+and  \mycomment{\color  {red}  proved  it  correct  with  reference  to  a  {\em
     minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
@@
 
-When  there  are  multiple  occurrences of  \bcar,  \bcdr\  and  \clazy,
+When  there  are  multiple  occurences of  \bcar,  \bcdr\  and  \clazy,
 $\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
@@
 simplified using  the $\hookrightarrow$ rules.  Finally the simplified
-NFA are converted to DFA.
+NFAs are converted to DFA.
 
@@
 
-The strongly regular grammar is converted  into a set of NFA, one for
+The strongly regular grammar is converted  into a set of NFAs, one for
 each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
-done on the NFA by  repeatedly introducing $\epsilon$ edges to bypass
+done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
 pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
@@
 
-\begin{table}[ht!]
-\caption{Time taken by LGC for different strategies.  The programs are
-  modified (smaller) instances of original benchmarks.}
+\begin{table*}[ht!]
+\caption{Comparing LGC times for different strategies}
 \label{tab:compare-live-clo-reach-clo}
@@
 \begin{tabular}
-{| l | r | r | r |}
-\hline
-Program        & \#GC & \multicolumn{2}{c|}{Closure collected using} \\ \cline{3-4}
-               &      & Reachability & Liveness                      \\ \hline
-\verb@nqueens@ & 183  & 0.009 sec    & 0.228 sec                     \\\hline
-\verb@lambda@  & 11   & 0.001 sec    & 4.992 sec                     \\ \hline
-\verb@lcss@    & 1    & 0.000 sec    & 34.668 sec                    \\ \hline
+{| l | r | r | r | r |}
+\hline
+Prog Name & Heap size & Num GC & Time taken using reachability for closure  & Time taken using liveness for closure \\ \hline
+\verb@nqueens (reduced) @ & 10000 & 183 & 0.0087s & 0.2267s\\ \hline
+\verb@lambda@&10000&11&0.0010s&4.9922s\\ \hline
+\verb@lcss@&10000&1&0.00008s&34.6684s\\ \hline
 \end{tabular}
-\end{table}
-
-Algorithm~\ref{algo:lgc-b} describes a garbage collection scheme where
-the liveness automata is consulted  only for \CONS\ cells; and assumes
-everything under  a closure to be  live and copies everything  that is
-reachable.
+\end{table*}
+
+\begin{table*}[ht!]
+\caption{Compile time and Garbage Collection statistics}
+\label{tab:exp-results}
+\centering
+ \begin{tabular}{|l||c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c|} \hline
+   \multicolumn{1}{|r||}{Program} & \verb@lambda@ & \verb@nperm@ & \verb@treejoin@ & \verb@lcss@ & \verb@sudoku@ & \verb@fibheap@ & \verb@nqueens@ & \verb@knightstour@ & \verb@gc_bench@ \\ \hline\hline
+   \#CFG Nonterminals            & 1078          & 794          & 709             & 839         & 1758          & 780            & 529            & 659                & 417             \\ \hline
+   \#CFG Rules                   & 1823          & 1038         & 1610            & 1752        & 2509          & 1358           & 748            & 909                & 486             \\  \hline 
+   \#DFA States                  & 3312          & 2055         & 2332            & 2340        & 5672          & 2254           & 1229           & 1742               & 729             \\\hline
+   \#DFA Transitions             & 5792          & 3632         & 4235            & 4261        & 10857         & 4076           & 2078           & 3101               & 1124            \\\hline
+   DFA Generation Time (sec)      & 14.7          & 1.1          & 2690.5          & 13.3        & 798.5         & 49.8           & 0.5            & 5.1                & 0.1             \\\hline
+ \end{tabular}\\
+ (a) Data for Compile-time Liveness Analysis \\[3pt]
+\include{autotable}
+\end{table*}
+
+  Algorithm~\ref{algo:lgc-b}  describes  a garbage  collection  scheme
+  where the liveness automata is  consulted only for \CONS\ cells; and
+  assumes everything under a closure  to be live and copies everything
+  that is reachable.
 
@@
 
+%% Liveness automata  does not exist for  the print function  as it is
+%% not part of the executing  program.  Since the print function makes
+%% everything  reachable  from a  cell  live  we  can copy  all  cells
+%% reachable from the reference.
+
+%% Consider  the case  where a  \CONS\ cells  is being  printed  and a
+%% garbage collection  is triggered when  printing the \CDR\  part. In
+%% this case, since  the refernce to the \CONS\ cells  is still on the
+%% print stack both the \CAR\ and the \CDR\ parts are reachable but we
+%% know that the \CAR\ part has already been printed and can be safely
+%% garbage collected. A liveness  analysis of the print function would
+%% have given  the same information.   In our implementation  we mimic
+%% the  behavior of the  liveness analysis  by using  a flag  in every
+%% \CONS\ cell which  says whether the \CAR\ part  has been printed or
+%% not. For any \CONS\ cell if  the \CAR\ part has not been printed, a
+%% reference  to the  \CAR\ part  will also  be on  the stack  but the
+%% \CDR\ part has to be explicitly copied.  In case the \CAR\ part has
+%% been printed,  then a reference  to the \CDR\  part will be  on the
+%% stack and  since the \CAR\ part  is not required anymore  we do not
+%% copy anything.  This  optimization subsumes tail call optimization.
+%% Again, experimental results show that this optimization minimizes a
+%% lot of copying.  \cred{Very sketchy: Make another attempt to write}
+%% \hrule
+
+%% \comment{ Unlike  eager languages where  GC happens only  at \CONS~
+%%   points or at function calls,  every \LET~ point is a potential GC
+%%   point.  The number of automata that needs to be stored is reduced
+%%   by sharing  automata along all  the program points along  an {\em
+%%   E-path}.
+
+%% By the time the program starts executing, the liveness automata are
+%% created  and stored  in a  file.  This  is re-used  every  time the
+%% program executes.   An important difference in GC  of lazy language
+%% compared to an  eager language is the point  of garbage collection.
+%% In case  of eager languages  only \CONS~ points and  function calls
+%% are the  GC points. In a  lazy language every \LET~  statement is a
+%% potential  GC  point. Hence  we  need  to  store liveness  automata
+%% corresponding to  every \LET~ statement in the  program. This leads
+%% to an  explosion in  the number of  automata. From  our observation
+%% that  along  an {\em  E-path}  all  program  points have  the  same
+%% liveness, we  can reduce the  number of automata state  required by
+%% sharing  the same  automata for  all program  points along  an {\em
+%% E-path}.} In the  example program, we only need  to create automata
+%% for program  points \ldots, since  they are the only  distinct {\em
+%% E-path}s in the program.
+
 \section{Experimental Evaluation}
 \label{sec:experiments}
+%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
+%% we  have  built a  prototype  consisting  of  an interpreter  for  our
+%% language,  a  liveness  analyzer  and  a copying  collector  that  can
+%% optionally use the results of liveness analysis for marking instead of
+%% reachability. When the collector uses liveness for marking, we call it
+%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
+%%   reachability-based  collector}  (RGC).   The  copying  collector  is
+%% neither incremental nor generational.  As a consequence, any cell that
+%% becomes unreachable or  dead is assuredly collected in  the next round
+%% of garbage collection.}
+
 To   test  the   utility  of   our  method   we  have   implemented  a
@@
 and a  garbage-collector that can  optionally use  reachability or
-liveness.  Our benchmark consists  of programs from nofib~\cite{nofib}
-suite and some  external programs manually converted to  ANF.  All our
-benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
-i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
-Ubuntu 14.04.
+liveness.  Our  benchmark consists  of programs  from nofib~\cite{nofib}  suite and
+some external programs manually converted to ANF.
+
+%% \comment{
+%% While exploring the activation record of a function, LGC uses the DFA
+%% at  the latest  program  point traversed  in the function's body.
+%% Thus, if  LGC
+%% is invoked  at the program point   $\pi$,  it would  use  the
+%% liveness  DFA at  $\pi$ itself for exploring the root set of the
+%% current activation record. For any other activation record in the
+%% stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
+%% call-chain,
+%% the DFA at the point of call to  ${\mathit g}$ would be used.}
 
@@
 
+%% \comment{
+%% Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
+%% point  pair ($x$,  $\pi$).  We  assume that  there is  a function
+%% $\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
+%% $\dfa{\pi}{x}$.
+%% Considering   a   DFA   as   a   table,
+%% $\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
+%% state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
+%% \acar\      or     \acdr.       We      shall     also      write
+%% $\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
+%% indicating  whether there  is a  transition from  $q$ on
+%% $\mathit{sym}$.  The LGC action to chase the root variable $x$ at
+%% $\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
+%% empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
+%% $\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
+%% in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
+%% pointer to $x$.
+%% The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
+%% $x$  in the  other semi-space  and returns  the new  address.  It
+%% hides  details such as  returning the  forwarding pointer  if the
+%% value  of $x$  is  already copied,  and  creating the  forwarding
+%% pointer otherwise.}
+
 \subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
@@
 To compare the  effect of using reachability to  copy closures instead
-of liveness  we take some  representative programs from  our benchmark
-suite    and    execute    them    with    both.     As    shown    in
+of liveness we take some representative programs from our benchmark suite and
+execute       them      with       both.        As      shown       in
 Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
@@
 reachability for closures.
-
-\begin{table*}[ht!]
-\caption{Statistics for liveness analysis and garbage collection}
-\label{tab:exp-results}
-\centering
- \begin{tabular}{|l||c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c|} \hline
-   \multicolumn{1}{|r||}{Program} & \verb@lambda@ & \verb@nperm@ & \verb@treejoin@ & \verb@lcss@ & \verb@sudoku@ & \verb@fibheap@ & \verb@nqueens@ & \verb@knightstour@ & \verb@gc_bench@ \\ \hline\hline
-   \#CFG Nonterminals            & 1078          & 794          & 709             & 839         & 1758          & 780            & 529            & 659                & 417             \\ \hline
-   \#CFG Rules                   & 1823          & 1038         & 1610            & 1752        & 2509          & 1358           & 748            & 909                & 486             \\  \hline 
-   \#DFA States                  & 3312          & 2055         & 2332            & 2340        & 5672          & 2254           & 1229           & 1742               & 729             \\\hline
-   \#DFA Transitions             & 5792          & 3632         & 4235            & 4261        & 10857         & 4076           & 2078           & 3101               & 1124            \\\hline
-   DFA Generation Time (sec)      & 14.7          & 1.1          & 2690.5          & 13.3        & 798.5         & 49.8           & 0.5            & 5.1                & 0.1             \\\hline
- \end{tabular}\\
- (a) Data for Liveness Analysis \\[3pt]
-\include{autotable}
-\end{table*}
+%%
+%% \comment{
+%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
+%% in the heap over time for RGC and LGC---here time is measured in terms
+%% of the number of cons cells allocated. In addition, they also show the
+%% number of  reachable cells and the  number of cells  that are actually
+%% live\footnote{This   is  statically   approximated  by   our  liveness
+%%   analysis}.  Since  the programs have  different memory requirements,
+%% we have tuned the size of heap for each program to ensure a reasonable
+%% number of  collections. An invocation  of RGC decreases the  number of
+%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
+%% invocation of LGC decreases the number  of heap cells to no lower than
+%% the curve of live cells.
+
+%% To construct the  reachable and live curves, we  record for every cell
+%% its creation time (\CreateTime), its last use time (\UseTime), and the
+%% earliest time  when the  cell becomes unreachable  and can  be garbage
+%% collected (\CollTime).  For accurate recording of  \CollTime, we force
+%% frequent invocations  of a reachability based collector  in a separate
+%% run.     A    cell   is    live    at    time   \mbox{\exectime}    if
+%% \mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
+%%   \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
+%%   \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.
+
+%% The benchmark programs are drawn from the {\tt no-fib} suite and other
+%% sources and have been  manually converted to  ANF.  All  graphs  except
+%%  {\tt  fibheap}  show  fewer  garbage
+%% collector  invocations for  LGC. The  number  of reachable  cells in
+%% {\tt
+%%   fibheap} increases  initially, comes  down before the  heap becomes
+%% gets
+%% full and then  continues to remains low.  This results in  a single
+%% garbage
+%% collection in both RGC and LGC.  The graphs also show the precision of
+%% our liveness analysis.  For  all programs  except  {\tt nperms} and
+%% {\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
+%% manages to collect a good portion of the data that are not live.}
+%%
 
 \subsection{Results}
-
-The statistics related to compile  time liveness analysis are shown in
-Table~\ref{tab:exp-results}(a). We observe that the analysis requires
-a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
-The bottleneck in our analysis is the NFA to DFA conversion
-(worst-case exponential behaviour). To avoid this compile-time
-overhead, we have added a caching mechanism in our simulator that can
-reuse the DFA from earlier analysis if the program has not changed..
-
-Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
-statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
-average  number of  cells reclaimed  per GC,  average number  of cells
-touched  per GC,  total  number of  dragged cells  and  total time  to
-perform  all collections.   It is  no  surprise that  number of  cells
-reclaimed per  GC is  higher (and  consequently the  number of  GCs is
-lower)  for LGC  based collection.   LGC  also reduces  the number  of
-dragged cells significantly. However, the cost  for LGC is that per GC
-execution  time is  more  for  LGC, which  results  in higher  overall
-execution time  for garbage  collection, even  with reduced  number of
-collections.   However, LGC  run time  is still  competitive (slowdown
-within 5X of RGC for worst case), and is better for 3 benchmarks (best
-case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
-synthetic benchmark  that allocates complete binary  trees of various
-sizes.  These binary trees are never used by the program. As a result,
-the benchmark highly favours LGC.  We have included this benchmark for
-completeness  only  and   the  numbers  for  it   are  not  considered
-representative of practical programs.
-
 \newcommand{\hgt}{3cm}
@@
 
+Table~\ref{tab:experimental-results}  compares the  garbage collection
+statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
+average  number of  cells reclaimed  per GC,  average number  of cells
+touched  per GC,  total  number of  dragged cells  and  total time  to
+perform  all collections.   It is  no  surprise that  number of  cells
+reclaimed per  GC is  higher (and  consequently the  number of  GCs is
+lower)  for LGC  based collection.   LGC  also reduces  the number  of
+dragged cells significantly. However, the cost  for LGC is that per GC
+execution  time is  more  for  LGC, which  results  in higher  overall
+execution time  for garbage  collection, even  with reduced  number of
+collections.   However, LGC  run time  is still  competitive (slowdown
+within 5X of RGC for worst case), and is better for 3 benchmarks (best
+case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
+synthestic benchmark  that allocates complete binary  trees of various
+sizes.  These binary trees are never used by the program. As a result,
+the benchmark highly favours LGC.  We have included this benchmark for
+completeness  only  and   the  numbers  for  it   are  not  considered
+representative of practical programs. All our benchmarks were executed
+on  a desktop  with  64 bit  Ubuntu 14.04,  8  core Intel(R)  Core(TM)
+i7-4770 CPU @ 3.40GHz, L2 cache of 8192 KB and 15922 MiB (16 GB) RAM.
+
 Memory   usage   graphs   for   select   benchmarks   in   are   shown
@@
 (blue line), the dip is specially visible for the benchmarks where LGC
-outperforms  RGC substantially.   The  graphs also  include curve  for
-reachable cells (black,  obtained approximately by forcing  RGC to run
-at a very high frequency) and the live cells for particular run (grey,
-obtained by post processing the heap access at the end of a program).
+outperforms                      RGC                     substantially
+(Table~\ref{tab:experimental-results}). The graphs  also include curve
+for reachable cells  (black, obtained approximately by  forcing RGC to
+run at  a very high frequency)  and the live cells  for particular run
+(grey, obtained  by post processing  the heap access  at the end  of a
+program).
 
 One area  of concern is the  huge gap between the  actual liveness and
-the liveness  perceived by our  collector.  In  case of LGC  for eager
+the liveness  percieved by our  collector.  In  case of LGC  for eager
 languages~\cite{asati14lgc} the gap was  very narrow and almost touched
@@
   closures  are  handled  during  garbage  collection  and  liveness  is
-  propagated inside closures.
+  propogated inside closures.
 
@@
 
-%Add graphs for memory usage LGC vs RGC
+%Add graphs for memort usage LGC vs RGC
 %Add graphs/table showing the GC times and other statistics

### 2015-07-11 paper/paper.tex

--- 2015-07-11 paper/paper.tex
@@
 overhead, we have added a caching mechanism in our simulator that can
-reuse the DFA from earlier analysis if the program has not changed..
+reuse the DFAs from earlier analysis if the program has not changed..
 
@@
 case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
-synthetic benchmark  that allocates complete binary  trees of various
+synthestic benchmark  that allocates complete binary  trees of various
 sizes.  These binary trees are never used by the program. As a result,
@@
 One area  of concern is the  huge gap between the  actual liveness and
-the liveness  perceived by our  collector.  In  case of LGC  for eager
+the liveness  percieved by our  collector.  In  case of LGC  for eager
 languages~\cite{asati14lgc} the gap was  very narrow and almost touched

### 2015-07-11 paper/paper.tex
bug fix.
minor

--- 2015-07-11 paper/paper.tex
@@
 the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
-triggered    by   the    test   (\NULLQ~(\CAR~(\CDR~\pz)))    in   the
+triggered    by   the    test   (\NULLQ~(\CAR~\pz))    in   the
 \SIF\ expression.
@@
 
-The specific contributions of this work are
+While an ideal garbage collector is  impractical, in this work we show
+that  static analysis  of  heap data  can help  practical garbage
+collectors in reclaiming more  garbage.  The specific contributions of
+this paper are
 \begin{itemize}
@@
 \bibliographystyle{abbrv}
-%\end{document}
+\end{document}
 

### 2015-07-11 paper/paper.tex
proof.

@@
 \usepackage{makecell}
+\usepackage{balance}
 
@@
 
-\mycomment{\color{red}
 \subsection{Correctness of liveness analysis}
@@
 significant modification  to the technique  used in \cite{asati14lgc}.
-We shall first modify the standard semantics. This semantics, which we
-shall  call {\em  minefield semantics},  will model  a  liveness based
-garbage collection at  each transition. This is done  by starting from
-the root-set  and inserting a  special value $\bot$ at  each reference
-location in the  heap that is reachable but not  live by our analysis.
-Attempts  to dereference  such  locations during  the transition  will
-result in entering a special  state denoted \bang.  We shall then show
-that no program enters the \bang\ state.
+\cblue{Good points to  make in the main text, right  in the small step
+  semantics  section: Note  that the  state of  the  transition system
+  consists of the current evaluation context, given by $\rho$ and $e$,
+  and  the suspended evaluation  contexts recorded  in the  stack $S$.
+  The heap $H$  is common to all contexts.}
+
+The idea behind the proof is  as follows: We shall modify the standard
+semantics to a semantics  called {\em minefield semantics} that models
+a liveness based garbage collection  at each transition.  The state of
+the  transition system  is augmented  to carry  enough  information to
+calculate the liveness environment  at the current evaluation context,
+and  each   of  the   suspended  evaluation  contexts.    Before  each
+transition,  we start  from the  root-set and  insert a  special value
+$\bot$ at each  location in the heap that contains  a reference and is
+reachable but  not live by  our analysis.  Any attempt  to dereference
+such locations during the transition will result in entering a special
+state denoted \bang.  The main proof is to show that no program enters
+the \bang\ state in the minefield semantics.
 
@@
 \begin{enumerate}
-\item  Given an  demand transformer  \Lfonly\ we  enrich  the abstract
-  machine state $\rho, S, H, e$  to $\rho, S, H, e, \sigma$.  $\sigma$
-  is the  demand on the  currently active function. As  a consequence,
-  the liveness $\Lv$ at $e$ is $\mathcal{L}(e, \sigma, \Lfonly)$. Each
-  entry  in   the  stack  $S$   of  suspended  evaluations,   is  also
-  augmented---it is  now $(\rho, x,  e, eps, \sigma)$. The $\sigma$  in the
-  stack entry  represents the demand  on the  function where  the evaluation
-  \ldots.  And $eps$ is \ldots.
-\item We introduce a function $GC$ that takes the 4-tuple $(\rho, H, S,
-  \sigma)$ and  inserts $\bot$ in all locations in the current
-  environment $\rho$, the heap $H$ and the environments in the stack
-  $S$ that are dead. More formally:
-
-and $\Sigma$ is a  stack of demands---one for each function
-  frame pushed in $S$.  Second,  we augment $Val$ with a value $\bot$.
-  To model strong liveness $\bot$ may be copied freely, including into
-  a  \CONS\  cell,  but  explodes  when  used  computationally  (in  a
-  primitive operation  other than a copy). Additionally  we define $GC
-  (\Lv, \Sigma)  : (\rho,  H, S) \rightarrow  (\rho ,  H , S  )$ which
-  determines live-reachability using $\rho$ and the $\rho$'s in $S$ as
-  the root  set and following links in  $H$ only as far  as allowed by
-  \Lv\   and   $\Sigma$.     Hence   $GC   (\Lv,   \Sigma)$   replaces
-  live-unreachable  values---in $\rho$,  in  $H$ and  the $\rho$'s  in
-  $S$---with $\bot$.  For example,  if $x.\epsilon \in \Lv$ then $\rho
-  (x) = \bot$.  Only $GC (\cdots )$ introduces $\bot$.
+\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
+  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
+  on the expression $e$, and $e_f$ is the body of the function. As the
+  section on liveness  analysis shows, The entire  function body $e_f$
+  and $\sigma$  are needed  to calculate  the liveness  environment at
+  different program  points. The  demand sigma  is ``dynamic''  in the
+  sense  that it  arises from  the actual  function calls  made during
+  execution.   The  0-CFA  demand  is an  over-approximation  of  this
+  demand.  The information  in a suspended evaluation  context is also
+  similarly augmented with  the demand on it and the  function body of
+  which the suspended evaluation is a  part.  Thus a stacked entry now
+  takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
+  small step semantics which carries the extra information is shown in
+  Figure~\ref{fig:minefield-semantics}.
+\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
+  e_f$ and $LF$, we can construct a liveness environment as follows:
+  \begin{enumerate}
+  \item If $e$ is an expression, then:  
+  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
+    \text{where }
+  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
+\item If $e$ is an application, then:  
+  $$ \Lv = ref(e, \sigma, \Lfonly)$$
   \end{enumerate}
-
- }
+We can define in the same way the liveness environment for each of the
+suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
+environments that we shall denote $\mathsf{SL}$.
+\item $GC$ models a liveness based garbage collection:
+$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
+  \begin{enumerate}
+  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
+    \notin \Lv$.
+\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
+  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
+  \in domain(\rho)$,  $\rho'(x)=\bot$
+  iff $x.\epsilon \notin \Lv'$.
+\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
+  $x$ in either the current environment or one of the stacked
+  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
+  \ell$  and $x.\alpha \in \Lv$, .
+  \end{enumerate}
+\end{enumerate}
+
+Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
+(\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
+preceded by  a garbage collection  using $GC$.  We consider  a garbage
+collection followed  by a transition as  a step. We show  by induction
+that, for programs  without function calls, starting  from the initial
+state that  has $e_{\main}$  as the first  expression to  be evaluated
+instead  of $\main$,  any transition  of  $n$ steps  occurs without  a
+\bang.
+
+Observe  that  the  transitions  that dereference  and  therefore  can
+potentially lead to \bang\ are those  which have a $H(\ldots)$ term in
+the  transition  rule.   Of  these,  terms  of  the  form  $H(\ldots)$
+dereference a root  variable, and the term $H(l)$  dereferences a heap
+location.  In  both the  cases we show  that the  dereferencing cannot
+lead to a \bang\ state in programs without function calls.
+
+We  need the  following proposition that  connects  liveness analysis  with
+minefield semantics transitions. 
+\begin{proposition}
+\label{prop:correctness}
+\begin{enumerate} 
+\item Consider a \LET\ definition $:\pi: (\LET~x
+  \leftarrow~\ldots~\IN~e$, such that there is a use of $x$
+  in $e$, i.e.  an occurrence of $\LET~y~\leftarrow~s~in
+  \ldots$, where $s$ is $(\CAR\ x)$, $(\CDR\ x)$ or $(+\ x\ y)$, or
+  $(\SIF\ x\ e1\ e2)$ or $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null. 
+\item Assume that a  \LET\ definition $\pi: (\LET~z  \leftarrow~(\CONS~x~y)~\IN~e$ has an occurrence of 
+ $(\CAR~z)$ in $e$. Then the liveness of $x$ at $\pi$ includes $\epsilon$.  Similarly if $e$ contains $(CDR~z)$, then the liveness of$y$ at $\pi$ includes $\epsilon$.  
+\end{enumerate},
+\end{proposition}
+
+In  the second  case for  example,  the liveness  of $x$  due to  the
+\CONS\ is  $\bcar\sigma$, where $\sigma$  is the demand on  $z$. Since
+$\sigma$  includes  the  demand  $\acar\alpha$   due  to  its  use  in
+$(\CAR~z)$, the liveness on $x$  includes $\epsilon$. We thus have the
+following lemma:
+
+\begin{lemma}\label{lemma:no-calls-correctness} A program without function calls cannot enter the \bang\ state in the minefield semantics.
+\end{lemma}
+
+\begin{proof}
+For a program  executing according to {\em minefield  semantics}, if a
+\bang\ has to  occur then some variable which is  dereferenced must be
+set to  $\bot$, between its  creation and  use points.  This  can only
+happen if  the liveness of  the variable being dereferenced  was null.
+From  the lemma~\ref{lem:correctness}  it  is clear  that this  cannot
+happen.
+\end{proof}
+
+For   a    programs   with   functions,   we    use   the   techniques
+in~\cite{asati14lgc}. We  assume as  a contradiction that  the program
+can enter a \bang\ state  and inline the  functions called to  obtain a
+program that is without functions  but has the same execution behavior
+and    demands    as   the    original    program.     We   then    use
+Lemma~\ref{lemma:no-calls-correctness}  to  show  that  the  resulting
+functionless program cannot enter a bang state.
 
@@
 
-%Add graphs for memory usage LGC vs RGC
-%Add graphs/table showing the GC times and other statistics
-
-%Mention lower peak memory usage.
-%Mention faster GC than full LGC and more garbage collected than RGC.
-%Pitch it as sweet spot between full LGC and RGC.
-%Future work
-%Mention full LGC
-%Mention k-liveness, use length example to show 1-liveness is not
-%sufficient.
-%Mention mixed mode GC,
-% - doing RGC most of the times and switching to LGC in case RGC fails
-%to collect any garbage
-% - doing LGC most of the times but regularly checking the reachability
-%of the heap and
-% switching to RGC if LGC almost equals RGC
-
+\balance
 \bibliography{fun_hra}{}
 \bibliographystyle{abbrv}
-\end{document}
-
-\pagebreak
-\section{Appendix}
-
-We shall now  outline a proof of correctness  of the liveness analysis
-presented  in  Section~\ref{sec:liveness-analysis}.   This requires  a
-significant modification  to the technique  used in \cite{asati14lgc}.
-\cblue{Good points to  make in the main text, right  in the small step
-  semantics  section: Note  that the  state of  the  transition system
-  consists of the current evaluation context, given by $\rho$ and $e$,
-  and  the suspended evaluation  contexts recorded  in the  stack $S$.
-  The heap $H$  is common to all contexts.}
-
-The idea behind the proof is  as follows: We shall modify the standard
-semantics to a semantics  called {\em minefield semantics} that models
-a liveness based garbage collection  at each transition.  The state of
-the  transition system  is augmented  to carry  enough  information to
-calculate the liveness environment  at the current evaluation context,
-and  each   of  the   suspended  evaluation  contexts.    Before  each
-transition,  we start  from the  root-set and  insert a  special value
-$\bot$ at each  location in the heap that contains  a reference and is
-reachable but  not live by  our analysis.  Any attempt  to dereference
-such locations during the transition will result in entering a special
-state denoted \bang.  The main proof is to show that no program enters
-the \bang\ state in the minefield semantics.
-
-To set up the minefield semantics, we follow these steps:
-\begin{enumerate}
-\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
-  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
-  on the expression $e$, and $e_f$  is the body of the function. $e_f$
-  and $\sigma$ are  both needed to calculate  the liveness environment
-  for expressions.  The information  in a suspended evaluation context
-  is also similarly  augmented with the demand on it  and the function
-  body of  which the suspended evaluation  is a part.  Thus  a stacked
-  entry now takes the form $(\rho, x, e, \sigma, e_f)$. A modification
-  of the small  step semantics which carries the  extra information is
-  shown in Figure~\ref{fig:minefield-semantics}.
-\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
-  e_f$ and $LF$, we can construct a liveness environment as follows:
-  \begin{enumerate}
-  \item If $e$ is an expression, then:  
-  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
-    \text{where }
-  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
-\item If $e$ is an application, then:  
-  $$ \Lv = ref(e, \sigma, \Lfonly)$$
-  \end{enumerate}
-We can define in the same way the liveness environment for each of the
-suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
-environments that we shall denote $\mathsf{SL}$.
-\item $GC$ models a liveness based garbage collection:
-$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
-  \begin{enumerate}
-  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
-    \notin \Lv$.
-\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
-  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
-  \in domain(\rho)$,  $\rho'(x)=\bot$
-  iff $x.\epsilon \notin \Lv'$.
-\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
-  $x$ in either the current environment or one of the stacked
-  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
-  \ell$  and $x.\alpha \in \Lv$, .
-  \end{enumerate}
-\end{enumerate}
-
-Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
-(\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
-preceded by  a garbage  collection using $GC$.  We consider  a garbage
-collection followed  by a transition as  a step. We show  by induction
-that, starting  from the  initial state, any  transition of  $n$ steps
-occurs without a \bang.
-
-We observe that  the transitions involving dereferencing  and can lead
-to \bang are those which have  a $H(...)$ term in the transition rule.
-Of these most  of the terms are of the  form $H(\rho(x))$, involving a
-root variable.  There is exactly  one occurrence of $H(l)$ which means
-dereferencing using a heap location.
-
-In both  the cases  we show  that the dereferencing  cannot lead  to a
-\bang\ state in a function call less program.
-
-For this we need the following lemma,
-\begin{lemma}
-\label{lem:correctness}
-Consider, 
-\begin{enumerate}
-\item
- A \LET\ definition $\pi_i:(\LET~x \leftarrow~s~\IN~e)$ such
-that there is  an evaluation path in $e$ in  which $\pi_i$ is followed
-by a  use i.e an  occurrence of $\pi_j:(\CAR\ x)$,  $\pi_j:(\CDR\ x)$,
-$\pi_j:(+\ x\ y)$, $\pi_j:(\SIF\ x\ e1\ e2)$ or $\pi_j:(\SRETURN\ x)$
-\item
- A \LET\ definition $\pi_i(\LET~z  \leftarrow~(\CONS~x~y)~\IN~e$, such that
- $e$ includes  an evaluation path in  which there is an  occurrence of
- $\pi_j: (\CAR~z)$ and $\pi_j$ follows $\pi_i$.
-\end{enumerate},
-the demand on $x$ is non-null between $\pi_i$ and $\pi_j$.
-\end{lemma} 
-\begin{proof}
-In the first case it is clear  that at all points $\pi_k$ which occurs
-in between $\pi_i$ and $\pi_j$, $x$  has non-null demand as $x$ is used
-directly in a strict computation.
-
-In the second case, it can be seen that the liveness of $x$ due to the
-\CONS\ is $\bcar\sigma$, where $\sigma$ is the demand on $z$. $\sigma$
-definitely  includes  the  demand  $\acar\alpha$ due  to  its  use  in
-$\pi_j$, therefore the demand on $x$ is at least $\alpha$.
-
- Since we do not kill the liveness of any variable along an evaluation
- path, and all program points along  an evaluation path share the same
- liveness environment, at any point along the evaluation path $x$ will
- not be set to $\bot$.
-\end{proof}
-
-For a program  executing according to {\em minefield  semantics}, if a
-\bang\ has to occur then some variable which is dereferenced along the
-evaluation path  must be set to  $\bot$, between its creation  and use
-points.  This can  only happen  if the  demand on  the variable  being
-dereferenced  was null.  From  the  lemma~\ref{lem:correctness} it  is
-clear that this cannot happen.
-
-The rest  of the proof depends  on converting a program  with function
-calls to an equivalent program without function calls for which we 
-refer the user to~\cite{asati14lgc}.
+
+\appendix
+
+\section{Minefield semantics for the language}
 
@@
 \end{tabular}
-\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
+\caption{Minefield semantics.\label{fig:minefield-semantics}}
 \end{center}
 \end{figure*}
+
 \end{document}

### 2015-07-11 paper/paper.tex
proof again.

@@
       %probably doesn't scale anyway.
-\newcommand{\mycomment}[1]{}
+%\newcommand{\mycomment}[1]{}
 \definecolor{Myblue}{rgb}{.2,0,1}
-\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
-\newcommand{\cred}[1]{{\color{red}{#1}}}
-\newcommand{\cblue}[1]{{\color{blue}{#1}}}
-\newcommand{\blankout}[1]{}
+%\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
+%\newcommand{\cblue}[1]{{\color{blue}{#1}}}
+%\newcommand{\blankout}[1]{}
 
@@
 \item  We  formulate  a  liveness  analysis for  a  lazy  first  order
-  functional language. We show how  to compute function summaries in
-  a  context  independent  manner  to achieve  a  context  sensitive
-  analysis. We  further describe the transformation  of the liveness
-  equations to grammar rules in a  way that the language accepted by
-  the resulting grammar describes live heap locations.\mycomment{the
-    proof of its correctness}
+  functional language  and prove  of its correctness.  We show  how to
+  compute  function  summaries  in  a context  independent  manner  to
+  achieve  a  context  sensitive  analysis. We  further  describe  the
+  transformation of the  liveness equations to grammar rules  in a way
+  that the language  accepted by the resulting  grammar describes live
+  heap locations.
 \item We prove  the undecidability of the grammar  that results from
@@
 The  semantics  of   expressions  (and  applications\footnote{In  most
-  contexts, we  shall use  the term 'expression' and the notation $e$  for
-both  expressions and
-  applications.}) are  given by transitions  of the form  $\rho, \stk,
-\heap, e \rightarrow \rho', \stk', \heap', e'$.  Here \stk\ is a stack
-of  continuation  frames.  Each  continuation  frame  is  of the  form
-$(\rho, x, e)$, signifying that the  location of $x$ has to be updated
-with the  value of the currently  evaluating expression, and $e$  is to be
-evaluated  next  in  the  environment  $\rho$.   The  start  state  is
-$([\;]_\rho,([\;]_\rho,  \ans,   (\print~\ans)):[\;]_{S}  ,  [\;]_{H},
-(\mainpgm)$.  In  this, the initial  environment and the  initial heap
-are  denoted  by  $[\;]_\rho$  and  $[\;]_H$, and  the  initial  stack
-consists  of  a  single  continuation   frame  in  which  \ans\  is  a
-distinguished  variable  that  would  be  updated with  the  value  of
-(\mainpgm).  In  addition, \print\ is a function  modeling a printing
-mechanism---a standard  run-time support assumption  for lazy languages
-that prints the  value of (\mainpgm).  Notice the  use of the operator
-$:$ for adding elements to the top of the stack.
+  contexts, we  shall use the  term 'expression' and the  notation $e$
+  for both expressions and applications.}) are given by transitions of
+the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
+Here \stk\ is a stack of continuation frames.  Each continuation frame
+is of the form $(\rho, x, e)$, signifying that the location of $x$ has
+to be updated  with the value of the  currently evaluating expression,
+and $e$ is to be evaluated  next in the environment $\rho$.  Note that
+the state of the transition  system consists of the current evaluation
+context,  given  by  $\rho$  and $e$,  and  the  suspended  evaluation
+contexts recorded  in the stack  $S$.  The heap  $H$ is common  to all
+contexts.
+
+The     start      state     is      $([\;]_\rho,([\;]_\rho,     \ans,
+(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm)$.  In this, the initial
+environment  and  the initial  heap  are  denoted by  $[\;]_\rho$  and
+$[\;]_H$,  and the  initial stack  consists of  a single  continuation
+frame in which \ans\ is a distinguished variable that would be updated
+with  the value  of (\mainpgm).   In addition,  \print\ is  a function
+modeling a printing mechanism---a standard run-time support assumption
+for lazy  languages that prints  the value of (\mainpgm).   Notice the
+use of the operator $:$ for adding elements to the top of the stack.
 
@@
 $x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
-added  to the  union.  Also  notice a  consequences of  laziness---the
+added  to the  union.   Also notice  a  consequence of  laziness---the
 entire  expression including  the condition  is not  evaluated if  the
@@
 
-\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
-  of  an   expression  $e$.   The  cases  $\SRETURN$   and  $\SIF$  are
-  straightforward, but note the liveness $x.\epsilon$ generated by the
-  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
-  three-address instruction:  the liveness of  $e$ is given  by taking
-  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
-  any incremental  liveness from  $s$.  The main  subtlety is  how the
-  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
-  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.
-
-We make three observations: firstly the rule ({\sc live-define}) has a
-least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
-secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
-inference of mutually recursive  function definitions, and thirdly the
-asymmetry of  demand and liveness (compared to  post- and pre-liveness
-classically) is due to the functional formulation here.
-}
-
 Section~\ref{sec:computing}   shows   how  the   demand   transformers
@@
 We shall now  outline a proof of correctness of  the liveness analysis
-presented  in  Section~\ref{sec:liveness-analysis}.   This requires  a
-significant modification  to the technique  used in \cite{asati14lgc}.
-\cblue{Good points to  make in the main text, right  in the small step
-  semantics  section: Note  that the  state of  the  transition system
-  consists of the current evaluation context, given by $\rho$ and $e$,
-  and  the suspended evaluation  contexts recorded  in the  stack $S$.
-  The heap $H$  is common to all contexts.}
+presented   in    Section~\ref{sec:liveness-analysis}.    This   proof
+technique  is  similar to  the  technique  used in  \cite{asati14lgc}.
 
@@
 \section{Computing liveness and its encoding as DFA}\label{sec:computing}
-Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis%
-  \mycomment{\color  {red}  and proved  it  correct  with  reference  to  a  {\em
-    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
-together described a liveness set for each program point in a function
-in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
-describe how to obtain demand transformers \Lfonly\ from the rule {\sc
-  live-define} and how to compute the specific demand $\sigma$ on each
-function.
+Section~\ref{sec:liveness} gave a  context-sensitive liveness analysis
+and proved it  correct with reference to a  {\em minefield} semantics.
+In  our  analysis, $\mathcal{L}$  and  \pmap{f}  together described  a
+liveness set for each program point in  a function in terms of a given
+$\sigma$  and \Lfonly.   However, we  still  have to  describe how  to
+obtain demand  transformers \Lfonly\  from the rule  {\sc live-define}
+and how to compute the specific demand $\sigma$ on each function.
 %%
@@
 \appendix
-
+\newcommand{\cred}[1]{{\color{red}{#1}}}
 \section{Minefield semantics for the language}

### 2015-07-11 paper/paper.tex
lemma ref.
lemma

@@
 \begin{figure}[t]
-\renewcommand{\arraystretch}{1.2}
+\renewcommand{\arraystretch}{1}
 \begin{eqnarray*}
@@
 \begin{center}
-\renewcommand{\arraystretch}{1.5}
+\renewcommand{\arraystretch}{1}
 \begin{tabular}{|c|c|c|}
@@
 \[
-\renewcommand{\arraystretch}{1.1}
+\renewcommand{\arraystretch}{1}
 \begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
@@
 \begin{figure*}[t]
-\renewcommand{\arraystretch}{1.5}
+\renewcommand{\arraystretch}{1}
 \begin{eqnarray*}
@@
 
+\newcommand{\cred}[1]{{\color{red}{#1}}}
+\section{Minefield semantics for the language}
+\begin{figure*}[ht!]
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Premise & Transition & Rule name \\ 
+\hline
+\hline
+          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
+  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
+    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
+\\
+\hline
+          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
+    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
+  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
+  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
+  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
+\sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
+\sc{car-clo}
+\\
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
+  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
+    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
+\sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
+  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
+\sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
+  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
+\sc{prim-2-clo} \\
+\hline
+{$\mathit{g}~\mbox{defined as}$
+$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
+  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
+    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
+\sc{funcall} \\
+\hline
+$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+  \ell]\rangle], e, \cred{\sigma, e_f}$} &
+\sc{let} \\
+\hline
+$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
+\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
+\cred{\clazy\sigma, e_f}$}
+&
+\sc{if-clo} \\
+\hline
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
+  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
+  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
+  \cred{\sigma', e_f'}$} &
+\sc{return-whnf}\\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
+  \rightsquigarrow$
+$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
+  \cred{\sigma, e_f}$} &
+\sc{return-clo} \\
+\hline
+\end{tabular}
+\caption{Minefield semantics.\label{fig:minefield-semantics}}
+\end{center}
+\end{figure*}
+
 To set up the minefield semantics, we follow these steps:
@@
   on the expression $e$, and $e_f$ is the body of the function. As the
-  section on liveness  analysis shows, The entire  function body $e_f$
+  section on liveness  analysis shows, the entire  function body $e_f$
   and $\sigma$  are needed  to calculate  the liveness  environment at
@@
 \begin{enumerate} 
-\item Consider a \LET\ definition $:\pi: (\LET~x
+\item Consider a \LET\ definition \linebreak $:\pi: (\LET~x
   \leftarrow~\ldots~\IN~e$, such that there is a use of $x$
@@
 happen if  the liveness of  the variable being dereferenced  was null.
-From  the lemma~\ref{lem:correctness}  it  is clear  that this  cannot
+From  the lemma~\ref{lemna:no-calls-correctness}  it  is clear  that this  cannot
 happen.
@@
 \begin{figure*}[t!]
-\renewcommand{\arraystretch}{1.1}
+\renewcommand{\arraystretch}{1}
   \begin{tabular}{cc}
@@
 %        \small
-        \renewcommand{\arraystretch}{1.1}{
+        \renewcommand{\arraystretch}{1}{
           \begin{uprogram}
@@
         \end{uprogram}}
-        \renewcommand{\arraystretch}{1.1}{
+        \renewcommand{\arraystretch}{1}{
 	  \begin{uprogram}
@@
 \bibliographystyle{abbrv}
-
-\appendix
-\newcommand{\cred}[1]{{\color{red}{#1}}}
-\section{Minefield semantics for the language}
-
-\begin{figure*}[h!]
-\begin{center}
-\begin{tabular}{|c|c|c|}
-\hline
-Premise & Transition & Rule name \\ 
-\hline
-\hline
-          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
-  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
-    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
-\\
-\hline
-          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
-    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
-  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
-\sc{car-clo}
-\\
-\hline
-$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
-  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
-    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
-\sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
-\sc{prim-1-clo} \\
-\hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
-\sc{prim-2-clo} \\
-\hline
-{$\mathit{g}~\mbox{defined as}$
-$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
-  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
-    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
-\sc{funcall} \\
-\hline
-$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e, \cred{\sigma, e_f}$} &
-\sc{let} \\
-\hline
-$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
-\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
-\cred{\clazy\sigma, e_f}$}
-&
-\sc{if-clo} \\
-\hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
-  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
-  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
-  \cred{\sigma', e_f'}$} &
-\sc{return-whnf}\\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
-  \cred{\sigma, e_f}$} &
-\sc{return-clo} \\
-\hline
-\end{tabular}
-\caption{Minefield semantics.\label{fig:minefield-semantics}}
-\end{center}
-\end{figure*}
-
 \end{document}

### 2015-07-11 paper/paper.tex
fit to 12 apage.
shading.

@@
 
-\begin{figure}[t]
+\begin{figure}[t]\footnotesize
 \renewcommand{\arraystretch}{1}
@@
 \begin{figure*}[t!]
-\begin{center}
+\begin{center}\footnotesize
 \renewcommand{\arraystretch}{1}
@@
 
-\newcommand{\cred}[1]{{\color{red}{#1}}}
+\newcommand{\cred}[1]{\psframebox[linestyle=none,
+    fillcolor=lightgray,fillstyle=solid,framesep=0.2pt]{#1}}
 \section{Minefield semantics for the language}
 \begin{figure*}[ht!]
-\begin{center}
+\begin{center}\footnotesize
 \begin{tabular}{|c|c|c|}
@@
 happen if  the liveness of  the variable being dereferenced  was null.
-From  the lemma~\ref{lemna:no-calls-correctness}  it  is clear  that this  cannot
+From  the lemma~\ref{lemma:no-calls-correctness}  it  is clear  that this  cannot
 happen.
@@
     \begin{minipage}{.40\textwidth}
-%        \small
+        \small
         \renewcommand{\arraystretch}{1}{
@@
 
-      %\small
+      \small
 \begin{eqnarray*}
@@
  \end{tabular}\\
- (a) Data for Liveness Analysis \\[3pt]
+ (a) Data for Liveness Analysis \\[1pt]
 \include{autotable}
+\vskip -5mm
 \end{table*}

### 2015-07-11 paper/paper.tex
Fixed minor spelling mistakes
changed lemma label
final

@@
 \newcommand{\cred}[1]{\psframebox[linestyle=none,
-    fillcolor=lightgray,fillstyle=solid,framesep=0.2pt]{#1}}
+    fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
 \section{Minefield semantics for the language}
@@
 \end{center}
+\vskip -5mm
 \end{figure*}
@@
 
-For   a    programs   with   functions,   we    use   the   techniques
+For   a    program   with   functions,   we    use   the   techniques
 in~\cite{asati14lgc}. We  assume as  a contradiction that  the program
@@
 \begin{figure}[t!]
+\scalebox{0.8}{
   \begin{tabular}{cc}
@@
 \\
-(c) & (d)  \end{tabular}
+(c) & (d)  \end{tabular}}
 \caption{(a) The  grammar rules for \var{\Df{\length}{1}}
   converted into an automaton, and (b) its DFA. The same for \var{\Lanv{9}{\pa}}
-  are shown in (c) and (d).}\label{fig:example-automata}
+  are in (c) and (d).}\label{fig:example-automata}
 \figrule
@@
 straight  forward as  variables may  escape their  creation scope  and
-need to  be handled in an  approximate way.  We describe this in detail next.
+need to  be handled in an  approximate way.
 
@@
 
-Another  interesting  research  in  this   field  is  the  concept  of
-Simplifiers~\cite{ONeill}. They are  described as lightweight daemons
-which improve the efficiency of the program in general and the garbage
+Simplifiers~\cite{ONeill}, described as lightweight daemons, 
+improve the efficiency of the program in general and the garbage
 collector in particular. Most of the simplifications mentioned in this
@@
 %% \end{figure}
-
+\vskip -5mm
 \section{Conclusions and Future Work}
@@
 
-\balance
+%\balance
 \bibliography{fun_hra}{}

### 2015-07-15 paper/paper.tex
Fixed typos and corrected reference to proposition

--- 2015-07-11 paper/paper.tex
@@
 \item Assume that a  \LET\ definition $\pi: (\LET~z  \leftarrow~(\CONS~x~y)~\IN~e$ has an occurrence of 
- $(\CAR~z)$ in $e$. Then the liveness of $x$ at $\pi$ includes $\epsilon$.  Similarly if $e$ contains $(CDR~z)$, then the liveness of$y$ at $\pi$ includes $\epsilon$.  
-\end{enumerate},
+ $(\CAR~z)$ in $e$. Then the liveness of $x$ at $\pi$ includes $\epsilon$.  Similarly if $e$ contains $(\CDR~z)$, then the liveness of $y$ at $\pi$ includes $\epsilon$.  
+\end{enumerate}
 \end{proposition}
@@
 happen if  the liveness of  the variable being dereferenced  was null.
-From  the lemma~\ref{lemma:no-calls-correctness}  it  is clear  that this  cannot
+From  the Proposition~\ref{prop:correctness}  it  is clear  that this  cannot
 happen.

### 2015-07-15 paper/paper.tex
typos...

--- 2015-07-11 paper/paper.tex
@@
 \begin{enumerate} 
-\item Consider a \LET\ definition \linebreak $:\pi: (\LET~x
-  \leftarrow~\ldots~\IN~e$, such that there is a use of $x$
-  in $e$, i.e.  an occurrence of $\LET~y~\leftarrow~s~in
-  \ldots$, where $s$ is $(\CAR\ x)$, $(\CDR\ x)$ or $(+\ x\ y)$, or
-  $(\SIF\ x\ e1\ e2)$ or $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null. 
-\item Assume that a  \LET\ definition $\pi: (\LET~z  \leftarrow~(\CONS~x~y)~\IN~e$ has an occurrence of 
- $(\CAR~z)$ in $e$. Then the liveness of $x$ at $\pi$ includes $\epsilon$.  Similarly if $e$ contains $(CDR~z)$, then the liveness of$y$ at $\pi$ includes $\epsilon$.  
-\end{enumerate},
+\item   Consider  a   \LET\  definition:   \linebreak  $\pi\!\!\!:   (\LET~x
+  \leftarrow~\ldots~\IN~e$, such  that there is  a use of $x$  in $e$,
+  i.e.  an occurrence of $\LET~y~\leftarrow~s~in \ldots$, where $s$ is
+  $(\CAR\ x)$, $(\CDR\ x)$  or $(+\ x\ y)$, or $(\SIF\  x\ e1\ e2)$ or
+  $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null.
+\item    Assume    that    a   \LET\    definition    $\pi:    (\LET~z
+  \leftarrow~(\CONS~x~y)~\IN~e$  has an  occurrence  of $(\CAR~z)$  in
+  $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
+  Similarly  if $e$  contains $(CDR~z)$,  then the  liveness of$y$  at
+  $\pi$ includes $\epsilon$.
+\end{enumerate}
 \end{proposition}
@@
 
-\begin{lemma}\label{lemma:no-calls-correctness} A program without function calls cannot enter the \bang\ state in the minefield semantics.
+\begin{lemma}\label{lemma:no-calls-correctness} 
+A program without function calls cannot  enter the \bang\ state in the
+minefield semantics.
 \end{lemma}
@@
 happen if  the liveness of  the variable being dereferenced  was null.
-From  the lemma~\ref{lemma:no-calls-correctness}  it  is clear  that this  cannot
-happen.
+From the lemma~\ref{lemma:no-calls-correctness} it  is clear that this
+cannot happen. {\color{Myblue}  Too bad - self  reference!! Should this
+  be Proposition~\ref{prop:correctness}? }
 \end{proof}
@@
 
-\newcommand{\hgt}{3cm}
+\newcommand{\hgt}{2.9cm}
 \begin{figure*}[p]
@@
 \end{tabular}%\vskip -10mm
- \caption{Memory usage.  Column (a) shows complete  usage while column
+ \caption{Memory usage.  Column (a) shows complete  usage, column
    (b) enlarges  a part  of the  usage.  The blue  and the  red curves
@@
    and LGC  respectively.  The  black curve  represents the  number of
-   reachable cells and  the grey curve represents the  number of cells
+   reachable cells and  the lightblue curve represents the  number of cells
    that are  actually live (of  which liveness analysis does  a static
@@
    cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
-   number of cons-cells (scaled down by factor $10^3$).}
+   number of cons-cells (scaled down by  $10^3$).}
 \label{fig:memory-usage} \figrule
@@
 reachable cells (black,  obtained approximately by forcing  RGC to run
-at a very high frequency) and the live cells for particular run (grey,
+at a very high frequency) and the live cells for particular run (lightblue,
 obtained by post processing the heap access at the end of a program).

### 2015-07-15 paper/paper.tex

@@
   $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
-  Similarly  if $e$  contains $(CDR~z)$,  then the  liveness of$y$  at
+  Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of$y$  at
   $\pi$ includes $\epsilon$.
@@
 happen if  the liveness of  the variable being dereferenced  was null.
-From the lemma~\ref{lemma:no-calls-correctness} it  is clear that this
-cannot happen. {\color{Myblue}  Too bad - self  reference!! Should this
-  be Proposition~\ref{prop:correctness}? }
+From  the Proposition~\ref{prop:correctness}  it  is clear  that this  cannot
+happen.
 \end{proof}

### 2015-10-03 paper/paper.tex
Added a space
Fixed typos

@@
 	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
-            (\PRIM\ 1\ (\CAR\  \pl))))
+            (\PRIM\ 1\ (\length (\CDR\  \pl)))))
             \UNL{0}
@@
 \item  We  formulate  a  liveness  analysis for  a  lazy  first  order
-  functional language  and prove  of its correctness.  We show  how to
+  functional language  and prove its correctness.  We show  how to
   compute  function  summaries  in  a context  independent  manner  to
@@
   the  analysis.   To overcome  this,  we  present an  algorithm  to
-  transform grammar  to finite  state automata  (DFA) such  that the
+  transform grammars  to finite  state automata  (DFA) such  that the
   language accepted by the  DFA over-approximates the liveness (and
@@
 \end{tabular}
-\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
+\caption{The small-step semantics for the language. \label{fig:lang-semantics}}
 \end{center}
@@
 The     start      state     is      $([\;]_\rho,([\;]_\rho,     \ans,
-(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm)$.  In this, the initial
+(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm))$.  In this, the initial
 environment  and  the initial  heap  are  denoted by  $[\;]_\rho$  and
@@
 modeling a printing mechanism---a standard run-time support assumption
-for lazy  languages that prints  the value of (\mainpgm).   Notice the
+for lazy  languages---that prints  the value of (\mainpgm).   Notice the
 use of the operator $:$ for adding elements to the top of the stack.
@@
  Since  in a  lazy language,  an  expression is  not evaluated  unless
- required, the null demand ($\emptyset$) does not generate liveness in
+ required. The null demand ($\emptyset$) does not generate liveness in
  any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
@@
 transformed into  the liveness of  its parameters at the  beginning of
-the   call.    $\Lfonly$  is   determined   by   the  judgement   form
+the   call.    $\Lfonly$  is   determined   by   the  judgement 
 $\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
@@
   come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
-  consulted by the garbage collector at this point.
+  considered by the garbage collector at this point.
 \item In an  eager language, the heap location referred  by \py\ would
@@
 analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
-computes a  map that takes a  e-path in ${\mathit f}$  and returns the
+computes a  map that takes an  e-path in ${\mathit f}$  and returns the
 liveness environment  arising out of  all uses of variables  along the
@@
 $x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
-added  to the  union.   Also notice  a  consequence of  laziness---the
+added  to the  union.   Also notice  a  consequence of  laziness: the
 entire  expression including  the condition  is not  evaluated if  the
@@
 \item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
-  e_f$ and $LF$, we can construct a liveness environment as follows:
+  e_f$ and $\mathcal{LF}$, we can construct a liveness environment as follows:
   \begin{enumerate}
@@
 \item $GC$ models a liveness based garbage collection:
-$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
+$GC(\rho,$ $ S,$ $ H,$ $ \mathcal{L},$ $ \mathsf{SL}) = (\rho', S', H')$ 
   \begin{enumerate}
@@
 \item   Consider  a   \LET\  definition:   \linebreak  $\pi\!\!\!:   (\LET~x
-  \leftarrow~\ldots~\IN~e$, such  that there is  a use of $x$  in $e$,
+  \leftarrow~\ldots~\IN~e)$, such  that there is  a use of $x$  in $e$,
   i.e.  an occurrence of $\LET~y~\leftarrow~s~in \ldots$, where $s$ is
@@
 \item    Assume    that    a   \LET\    definition    $\pi:    (\LET~z
-  \leftarrow~(\CONS~x~y)~\IN~e$  has an  occurrence  of $(\CAR~z)$  in
+  \leftarrow~(\CONS~x~y)~\IN~e)$  has an  occurrence  of $(\CAR~z)$  in
   $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
-  Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of$y$  at
+  Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of $y$  at
   $\pi$ includes $\epsilon$.
@@
 For a program  executing according to {\em minefield  semantics}, if a
-\bang\ has to  occur then some variable which is  dereferenced must be
+\bang\ occurs then some variable which is  dereferenced must be
 set to  $\bot$, between its  creation and  use points.  This  can only
@@
 To  answer these  questions, it  is convenient  to cast  the equations
-arising  out of  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
+arising from  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
 grammar.   To do  so, we  need  to modify  the rules  themselves to  a
@@
 $\sigma$ needs to be known. This creates difficulties since we want to
-solve the equations arising out of liveness symbolically.
+solve the equations arising from liveness symbolically.
 
@@
 $\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
-where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
+where $\sigma$ consists  of forward access paths only. Then  does $\alpha \in
 \sigma$?
@@
 \begin{definition}\label{def:specialgrammar}
-Consider a form of  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
+Consider the  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
 non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \clazy, \$\}$, $p_1$
-is a set of context-free production that contains the distinguished
+is a set of context-free productions that contains the distinguished
 production $S \rightarrow \alpha\$$, $\alpha$ is a string of grammar
@@
 We circumvent the problem  of undecidability by over- approximating the
-CFG by non-deterministic finite state automata (NFA). The NFA are then
+CFG by non-deterministic finite state automata (NFA). The NFAs are then
 simplified using  the $\hookrightarrow$ rules.  Finally the simplified
-NFA are converted to DFA.
+NFAs are converted to DFAs.
 
@@
  We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
- approximate a CFG to a {\em strongly regular\/} grammar.  For
+ approximate a CFG by a {\em strongly regular\/} grammar.  For
  example, the grammar
@@
 simplification is continued till a fixed point is reached, after which
-the edges labeled \bcar \ and  \bcdr are deleted.  The details of the
+the edges labeled \bcar \ and  \bcdr\ are deleted.  The details of the
 algorithm, its correctness and termination proofs are given by Karkare
@@
 paths. While  checking for liveness  during GC, a  forward access-path
-only if it can reach a final state.
+is valid only if it can reach a final state.
 Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained

### 2015-10-06 paper/paper.tex
Moved GC to start of function calls. Need to fix errors

@@
 
-The number  of garbage  collections is same  since we  approximate the
+The number  of garbage  collections is the same  since we  approximate the
 liveness of  a closure  to the  liveness at  its creation  point.  The
@@
 
-Memory   usage   graphs   for   select   benchmarks   in   are   shown
+Memory   usage   graphs   for   select   benchmarks    are   shown
 Figure~\ref{fig:memory-usage}, Column  (a).  As the number  of garbage
@@
   languages and shown its benefit  for garbage collection in practical
-  programs. We  defined context  sensitive liveness analysis  that uses
+  programs. We  defined a context  sensitive liveness analysis  that uses
   context independent summaries of functions obtained using a symbolic
-  demand. We show  that obtaining precise solution  to these equations
+  demand. We showed  that obtaining precise solution  to these equations
   is undecidable in  our formulation, and hence  we safely approximate
@@
 
-  Liveness   of   closure   presented  further   challenges   to   our
+  Liveness   of   closures   presented  further   challenges   to   our
   implementation.   We compared  different approximate  strategies for

### 2015-10-06 lazy-scheme/src/SchemeAST.h

@@ -166,6 +166,7 @@ public:
 	enum exprType type;
 	const std::string node_name;
 	std::unordered_set<std::string> label_set;
+	unsigned int heap_cells_required;
 protected:
 	std::string label;
 
@@ -550,6 +551,8 @@ public:
 	std::string getFunctionName() {return pID->getIDStr();}
 	std::vector<std::string> getArgs();
 	ExprNode* getFunctionBody() {return pExpr;}
+
+	unsigned int heap_cells_required;
 
 protected:
 	ExprNode * pExpr;

### 2015-10-06 lazy-scheme/src/SchemeAST_Defs.cpp

@@ -525,9 +525,9 @@ cons* LetExprNode::evaluate()
 //	cout << "Current heap = " << current_heap() << endl;
 
 #endif
-	bool isFunctionCall = getVarExpr()->isFunctionCallExpression();
-    /* -------------------------------------------------------------------------------------------------------------*/
-    /* This is where we do the GC. Can this be made independent of ExprNode class? */
+	/*bool isFunctionCall = getVarExpr()->isFunctionCallExpression();
+     -------------------------------------------------------------------------------------------------------------
+     This is where we do the GC. Can this be made independent of ExprNode class?
     static clock_tick last_gc_clock = 0;
 
 	if (gc_status != gc_disable) 
@@ -583,7 +583,7 @@ cons* LetExprNode::evaluate()
             }
         }
     }
-    /* End of GC related stuff */
+*/    /* End of GC related stuff */
     /* -------------------------------------------------------------------------------------------------------------*/
 
 
@@ -1376,14 +1376,70 @@ std::string FuncExprNode::getFunction()
 
 cons* FuncExprNode::evaluate()
 {
+
 	cons* heap_cell = update_heap_refs.top();
 
 	if (heap_cell->inWHNF)
 		return heap_cell;
 	DefineNode* funcDef = (DefineNode*)pgm->getFunction(this->getFunction());
 
+	if (gc_status != gc_disable)
+	{
+		static clock_tick last_gc_clock = 0;
+		auto num_cells_reqd = funcDef->heap_cells_required;
+		cout << "Processing function " << getFunction() <<endl;
+		cout << "Num heap cells required " << num_cells_reqd<<endl;
+		if (gc_status == gc_freq)
+		{
+			if (GC_STAT_GET_CLOCK() - last_gc_clock > GC_FREQ_THRESHOLD())
+			{
+				reachability_gc();
+
+				GC_STAT_DUMP_GARBAGE_STATS();
+				last_gc_clock = GC_STAT_GET_CLOCK();
+			}
+		}
+		if (current_heap() < num_cells_reqd)
+		{
+			if ((gc_status == gc_plain) || (gc_status == gc_freq)) // try to get more space even if gc-freq
+			{
+				//cout << "Num heap cells required " << num_cells_reqd<<endl;
+				reachability_gc();
+				detail_gc();
+
+				GC_STAT_DUMP_GARBAGE_STATS();
+			}
+			else
+			{
+				assert(gc_status == gc_live);
+
[…]

### 2015-10-06 lazy-scheme/src/SchemeAST_Demands.cpp

@@ -41,6 +41,7 @@ unordered_map<string, expr_demand_grammars*> ReturnExprNode::transformDemand(con
 	label_set.insert(label);
 	pID->label_set.insert(label);
 
+	heap_cells_required = 0;
 
 	return gLivenessMap;
 }
@@ -54,6 +55,8 @@ expr_demand_grammars * IdExprNode::transformDemandRef(const rule & demand)
  	auto retval = new expr_demand_grammars({ new demand_grammar({{ label, demand }}),
         new demand_grammar({{ *pID, demand }})
     });
+
+ 	heap_cells_required = 1;
 	return retval;
 }
 
@@ -62,6 +65,7 @@ expr_demand_grammars * IdExprNode::transformDemandRef(const rule & demand)
 expr_demand_grammars * ConstExprNode::transformDemandRef(const rule & demand)
 {
 
+	heap_cells_required = 1;
 	return new expr_demand_grammars({ new demand_grammar({{ label, demand }}), new demand_grammar({{ }}) }); //TODO: verify this rule
 
 }
@@ -73,6 +77,8 @@ bool UnaryPrimExprNode::isExpressionRecursive(const std::string var) const
 	std::string argString = ((IdExprNode*)pArg)->getIDStr();
 	if (0==argString.compare(var))
 		isRecursive = true;
+
+
 	return isRecursive;
 }
 
@@ -125,6 +131,7 @@ expr_demand_grammars * UnaryPrimExprNode::transformDemandRef(const rule & demand
 
     label_set.insert(pArg->label_set.begin(), pArg->label_set.end());
 
+    heap_cells_required = 1;
     return result;
 }
 
@@ -179,6 +186,7 @@ expr_demand_grammars * BinaryPrimExprNode::transformDemandRef(const rule & deman
     result->first->emplace(label, demand);
     
 
+    heap_cells_required = 1;
     return result;
  
 }
@@ -219,7 +227,11 @@ unordered_map<string, expr_demand_grammars*> IfExprNode::transformDemand(const r
 
     label_set.insert(label);
     label_set.insert(pThen->label_set.begin(), pThen->label_set.end());
-    label_set.insert(pElse->label_set.begin(), pElse->label_set.end());
+    label_set.insert(pElse->label_set.begin(), pElse->label_set.end());\
+
+    heap_cells_required = (pThen->heap_cells_required > pElse->heap_cells_required) ?
+    		                        pThen->heap_cells_required
+								  : pElse->heap_cells_required;
 
[…]

### 2015-10-26 paper/proof.tex
fixing the proof

--- 2015-10-06 paper/paper.tex
@@
 \definecolor{Myblue}{rgb}{.2,0,1}
-%\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
+\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
 %\newcommand{\cblue}[1]{{\color{blue}{#1}}}
@@
 \begin{figure*}[t!]
+\comment{
+
 \begin{center}\footnotesize
-\renewcommand{\arraystretch}{1}
+\renewcommand{\arraystretch}{1.5}
 \begin{tabular}{|c|c|c|}
@@
 \hline 
-          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
-  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
+          & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, \kappa
+  \rightsquigarrow \rho',\, S,\, H[\ell := \kappa],\, e$    &  \sc{const}
 \\
 \hline
-          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
+          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, H,\, (\CONS~x~y)
 \rightsquigarrow
-$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
-(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
+$  $\rho',\, S,\, H[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
+\hline
+
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (\CAR~x)
 \rightsquigarrow
-\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
+\rho',\, (\rho, x, (\CAR~x))\!:\!S, \,H,\, s$      &
 \sc{car-clo}
@@
 \hline
+
+$H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
+(\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho,\rangle), (\CAR~x))\!:\!S,\, H,\, s$      &
+\sc{car-1-clo} \\
+\hline
+
+$H(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\,
+(\CAR~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$      &
+\sc{car-select} \\
+\hline
+
 $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
-$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
-\sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
+ & {$\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, (+~x~y)  \rightsquigarrow$
+$\rho', \,S,\, H[\ell := H(\rho'(x)) + H(\rho'(y))],\, e$}      &
+\sc{prim-add} \\
+\hline
+
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (+~x~y)
 \rightsquigarrow
-\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
+\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, H,\, s$      &
 \sc{prim-1-clo} \\
 \hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, H,\, (+~x~y)
 \rightsquigarrow
-\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
+\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, H,\, s$      &
 \sc{prim-2-clo} \\
@@
 {$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, H,\,
 (f~\myvec{x})  \rightsquigarrow
-[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
+[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, H,\, e_{\mathit{f}}$      &
 \sc{funcall} \\
 \hline
-$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+$\ell$ is a new location& {$\rho,\, S,\, H,\, (\LET~x\leftarrow s~\IN~e)
   \rightsquigarrow$
-$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
+$\rho\oplus[x \mapsto \ell],\, S,\, H[\ell \mapsto \langle s,
     \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e$} &
+  \ell]\rangle],\, e$} &
 \sc{let} \\
 \hline
-$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho, S, H,  e_1$ & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho, S, H,  e_2$ & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
+$H(\rho(x)) \ne 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho,\, S,\, H,\,  e_1$ & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho,\, S, \,H, \, e_2$ & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\,
   (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
+\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, H, \, s$}
 &
@@
 \hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
-e)\!:\!S, H,
-(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
+{$H(\rho(x))~\mbox{is}$ $\mbox{WHNF with value}~v$}& $\rho,\, (\rho', \ell,
+e)\!:\!S,\, H,\,
+(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$ &
 \sc{return-whnf}\\
 \hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\, (\SRETURN~x)
   \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
+$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, H,\,  s$} &
 \sc{return-clo} \\
@@
 \end{center}
+}
 \end{figure*}
@@
 \subsection{Semantics}
+\comment{
 We now give  a small-step semantics for our  language.
@@
 \begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}\\
+d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
 v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
-c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
-d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
+c:   & \mathit{Clo} &=& \mathit{(App \times Env)}& \mbox{-- Closures}\\
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
+\mbox{-- Environment} \\
 \end{array}
@@
 Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
-Since all data objects are boxed, we model an environment as a mapping
-from the set  of variables of the program  $\mathit{Var}$ to locations
-in the  heap.  These  either contain  a value in  WHNF (a  number, the
+These  either contain  a value in  \emph{WHNF} (a  number, the
 empty  list  $\NIL$,  or  a  \CONS\  cell  with  possibly  unevaluated
@@
 \rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
-maps free variables of $s$ to their respective locations.
+maps free variables of $s$ to their respective locations. Since all data objects are boxed, we model an environment as a mapping
+from the set  of variables of the program  $\mathit{Var}$ to locations
+in the  heap.
 
@@
 Here \stk\ is a stack of continuation frames.  Each continuation frame
-is of the form $(\rho, x, e)$, signifying that the location of $x$ has
-to be updated  with the value of the  currently evaluating expression,
-and $e$ is to be evaluated  next in the environment $\rho$.  Note that
-the state of the transition  system consists of the current evaluation
-context,  given  by  $\rho$  and $e$,  and  the  suspended  evaluation
-contexts recorded  in the stack  $S$.  The heap  $H$ is common  to all
-contexts.
-
-The     start      state     is      $([\;]_\rho,([\;]_\rho,     \ans,
-(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm))$.  In this, the initial
-environment  and  the initial  heap  are  denoted by  $[\;]_\rho$  and
-$[\;]_H$,  and the  initial stack  consists of  a single  continuation
-frame in which \ans\ is a distinguished variable that would be updated
-with  the value  of (\mainpgm).   In addition,  \print\ is  a function
-modeling a printing mechanism---a standard run-time support assumption
-for lazy  languages---that prints  the value of (\mainpgm).   Notice the
-use of the operator $:$ for adding elements to the top of the stack.
-
-%% In absence of a static type-checking mechanism, programs with other
-%% than syntactic  errors lead to states  not covered by  any rules in
-%% the small-step semantics.
+is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
+has  to  be  updated  with  the  value  of  the  currently  evaluating
+expression, and $e$ is to be evaluated next in the environment $\rho$.
+Note that the  state of the transition system consists  of the current
+evaluation  context,  given  by  $\rho$ and  $e$,  and  the  suspended
+evaluation contexts (represented as  continuation frames) in the stack
+$S$.  The heap $H$ is common to all contexts.
+
+The  initial  state  of  the  transition  system  described  above  is
+$([\;]_\rho,\,   ([\;]_\rho,   \,   \ans,\,   (\print~\ans)):[\;]_{S},
+[\;]_{H},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_H$  are the
+initial environment and  the initial heap. The  initial stack consists
+of  a single  continuation frame  in  which \ans\  is a  distinguished
+variable  that  would   eventually  be  updated  with   the  value  of
+(\mainpgm).  In  addition, \print\ is  a function modeling  a printing
+mechanism---a   standard   run-time   support  assumption   for   lazy
+languages---that prints  the value of  (\mainpgm).  Notice the  use of
+the operator $:$ for adding elements to the top of the stack.
+
 The   notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
-environment that maps variables $x_i$
-%$x_1,  \ldots, x_n$
-to locations $\ell_i$
-%$\ell_1,  \ldots,  \ell_n$,
-and  $\heap[\ell :=  d]$
-indicates the  updating of  a heap \heap\  at $\ell$ with  $d$.  $\rho
-\oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
-and $\lfloor \rho \rfloor_X$  represents the environment restricted to
-the locations in $X$. Finally $FV(s)$ represents the free variables in
-the application $s$.
+environment  that  maps  variables  $x_i$ to  locations  $\ell_i$  and
+$\heap[\ell := d]$  indicates the updating of a heap  \heap\ at $\ell$
+with  $d$.   $\rho \oplus  \rho'$  represents  the environment  $\rho$
+shadowed  by  $\rho'$  and  $\lfloor \rho  \rfloor_X$  represents  the
+environment  restricted  to  the  variables in  $X$.  Finally  $FV(s)$
+represents the  free variables  in the  application $s$  and $addr(c)$
+gives the address of the closure $c$ in the heap.
 
 The small-step semantics  is shown in Figure~\ref{fig:lang-semantics}.
-As  a sample,  consider  the  rules  for  $(+~x~y)$, an  application
-involving a binary operator strict in both arguments.   When
-both $x$  and $y$ are already  in WHNF ({\sc prim-whnf}),  the heap is
-updated with the  sum of the arguments and  the next expression to be
-evaluated is picked from the continuation on the stack. If either of the
-arguments is not in
-WHNF  ({\sc  prim-1-clo}  and   {\sc  prim-2-clo}),  it  is  sent  for
-evaluation, and a continuation  marks that the evaluation of $(+~x~y)$
-has to be resumed after the argument is evaluated.
-
+As a  sample, consider the  three rules for  $(\CAR~x)$.  If $x$  is a
+closure, it is evaluated to WHNF,  say $(d_1, d_2)$.  This is given by
+the  rule  {\sc  car-clo}.  If  $d_1$  is  not  in  WHNF, it  is  also
+evaluated  ({\sc car-1-clo}). The address ($addr (d_1)$) to be updated with the evaluated
+value is  recorded in  a continuation frame.  This is
+required for lazy evaluation,
+%of an application involving a selector such  as $\CAR$, 
+else, in the presence of sharing,   $d_1$  may  be  evaluated  more  than  once~\cite{Jones87}.    Only after this  is the
+actual selection done ({\sc car-select}).
+
+}
 %==============================================================
@@
 
-The idea behind the proof is  as follows: We shall modify the standard
+The  idea behind  the  proof is  as follows:  We  modify the  standard
 semantics to a semantics called  {\em minefield semantics} that models
-a liveness based garbage collection  at each transition.  The state of
-the  transition system  is augmented  to carry  enough  information to
-calculate the liveness environment  at the current evaluation context,
-and  each   of  the   suspended  evaluation  contexts.    Before  each
+a liveness based  garbage collection at each  transition.  Before each
+transition,  we start  from the  root-set and  insert a  special value
+$\bot$ at each  location in the heap that contains  a reference and is
+reachable but not live. The proof is based on a context-sensitive form
+of    liveness   of    which   our    calculated   liveness    is   an
+over-approximation. The state of the transition system is augmented to
+carry enough information to calculate  the liveness environment at the
+current  evaluation  context  and  each of  the  suspended  evaluation
+contexts on  the stack.   To simulate  garbage collection, before each
 transition,  we start  from the  root-set and  insert a  special value
@@
     fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
+
+\comment{
+\begin{enumerate}
+\item Do you need to carry the function body?
+\item Minefield doesn't insert bottom inside closures. Separate rule to
+be inserted.
+\item Examine Proposition 4.1, See that different cases are disjoint
+  and together cover all possible cases
+\item Suppose the statement that we want to prove is this: Every step
+  (step = GC + $\rightsquigarrow$) goes without a bang. How to divide
+  a (n+1)-step transition into 1 and n. What is the exact invariant
+  required before each step. 
+\end{enumerate}
+
+}
+
 \section{Minefield semantics for the language}
@@
 minefield semantics transitions.
+
 \begin{proposition}
@@
 \end{lemma}
+
 \begin{proof}

### 2015-11-24 paper/paper.tex
Updated paper with latest example

@@
   \begin{tabular}{cc}
-    \begin{minipage}{.40\textwidth}
+    \begin{minipage}{.50\textwidth}
         \small
@@
           \begin{uprogram}
-            \UNL{1} (\DEFINE\ (\length~\xl)
-            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
-(\NULLQ~\xl) \IN
-            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
-$\psi_1\!\!:\,$ \px
-            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
-            $(\LET\ \pv\ $\leftarrow 0$ \IN
-            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
-(\SRETURN~\psi_2:\pv)$
-            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
-$\leftarrow$  (\CDR~\xl)  \IN
-            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
-$\leftarrow$  (\length~\pu)  \IN
-            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
-            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
-            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
-(\SRETURN~\psi_3:\pz)$)))))))
+            \UNL{1} (\DEFINE\ (\fun~\px\ \pn)
+            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \py\ $\leftarrow $\
+(\length~\px) \IN
+            \UNL{3} $\pi_2\!\!:\, $(\LET\ \pl\ $\leftarrow $\
+(\makelist~\pn) \IN
+            \UNL{4} \hspace*{.05cm} $\pi_3\!\!:\,$(\SIF\
+$\psi_1\!\!:\,$ \py
+            \UNL{5} \hspace*{.27cm} $\pi_4\!\!:\, $(\LET\ \xhd\ $\leftarrow$\ (\CAR\ \pl) \IN
+            \UNL{6} \hspace*{.27cm} $\pi_5\!\!:\, $(\LET\ \xtl\ $\leftarrow$\ (\CDR\ \px) \IN
+            \UNL{7} \hspace*{.27cm} $\pi_6\!\!:\, $(\LET\ \rest\ $\leftarrow$\ (\fun~ \xtl\ \xhd) \IN
+            \UNL{8} \hspace*{.27cm} $\pi_7\!\!:\, $(\LET\ \ret\ $\leftarrow$\ (\CONS\ \py\ \rest) \IN
+            \UNL{9} \hspace*{.32cm} $\pi_8\!\!:\,
+(\SRETURN~\psi_2:\ret)))))$
+            \UNL{5} \hspace*{.29cm}    $\pi_9\!\!:\, $(\LET~\ps\
+$\leftarrow$  (\Sum~\pl)  \IN
+            \UNL{6} \hspace*{.34cm}   $\pi_{10}\!\!:\, $(\LET~\pu\
+$\leftarrow$  (+~1~\py)  \IN
+            \UNL{7} \hspace*{.34cm} $\pi_{11}\!\!:\,
+            $(\LET~\pr\ $\leftarrow$ ($/$~\ps~\pu) \IN
+            \UNL{8} \hspace*{.27cm} $\pi_{12}\!\!:\, $(\LET\ \res\ $\leftarrow$\ (\CONS\ \pr\ \NIL) \IN
+            \UNL{9} \hspace*{.34cm} $\pi_{13}\!\!:\,
+(\SRETURN~\psi_3:\res)$)))))))))
         \end{uprogram}}
@@
 	  \UNL{1} $(\DEFINE\ (\main)$
-           \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$
-           (
-           \scalebox{0.8}{\psframebox[framearc=.5,
-	fillcolor=lightgray,fillstyle=solid,framesep=2pt]{%
-             \begin{tabular}{@{}c@{}}
-               {a BIG closure}
-             \end{tabular}}}
-           ) \IN  
-          \UNL{3} \!\!$\pi_{10}\!\!:\, (\LET\  \pb\  \leftarrow ($+$\ \pa\ \acdr)$
+           \UNL{2} \!\!$\pi_{14}\!\!:\, (\LET\  \pa\  \leftarrow$
+           (\CONS\ $1$\ (\CONS\ $2$\ \NIL) \IN  
+          \UNL{3} \!\!$\pi_{15}\!\!:\, (\LET\  \pb\  \leftarrow $5$)$
 \IN
-	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ (\LET\ \pc\
-$\leftarrow  (\CONS\ \pb\ \NIL)$ \IN
-          \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
-          $(\LET\ \pw\  $\leftarrow  (\length\ \pc)$ \IN
-          \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
-(\SRETURN~\psi_4:\pw)))))$
+	  \UNL{4}   \hspace*{.05cm}$\pi_{16}\!\!:\,      $ (\LET\ \pc\
+$\leftarrow  (\fun\ \pa\ \pb)$ \IN
+          \UNL{5}   \hspace*{.15cm}    $\pi_{17}\!\!:\,
+          (\SRETURN~\psi_4:\pc)))))$
 \end{uprogram}}
@@
 
-    \begin{minipage}{.51\textwidth}
+    \begin{minipage}{.48\textwidth}
 
@@
 $\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
+
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-and  the only equation defining \Lfone{\length} is:
+and  the equation defining \Lfone{\fun} is:
   \begin{eqnarray*}
-   && \Lf{\length}{1}{\sigma}
-    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
-\clazy\Lf{\length}{1}{\clazy\sigma}
+   && \Lf{\fun}{1}{\sigma}
+    =  \clazy\sigma \cup \Lf{\length}{1}{\bcar\sigma_{\mathit{\length}}} \cup  \Lf{\length}{1}{\clazy\sigma} 
+       \cup  \acdr\Lf{\fun}{1}{\bcdr\sigma_{\mathit{\fun}}}
 \end{eqnarray*}
@@
 In general, the equations for  \Lfonly\ are recursive since, as in the
-case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
+case  of  \fun,  \Lan{1}\  may   have  been  defined  in  terms  of
 \Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
@@
 \Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
-\Lfonly$_{\mathit \length}$, we get:
+\Lfonly$_{\mathit \fun}$, we get:
   \begin{eqnarray*}
-&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
-&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy
-       \cup \clazy\Df{\length}{1}{\clazy}
+&&  \Lf{\fun }{1}{\sigma} = \Df{\fun}{1}\sigma,~\mbox{and}\\
+&&   \Df{\fun}{1} =  \clazy \cup \Df{\length}{1}\bcar \cup  \Df{\length}{1}\clazy 
+       \cup  \acdr\Df{\fun}{1}\bcdr
   \end{eqnarray*}
 
-Note that this equation can also be viewed as a CFG with \{\acdr,
-\clazy\} as terminal symbols and \Df{\length}{1} as the sole
-non-terminal.
+Note that this equation can also be viewed as a CFG with \{\acdr, \bcar, \bcdr,
+\clazy\} as terminal symbols and \Df{\length}{1} and \Df{\fun}{1} as the 
+non-terminals.
 
@@
 
-For  the  running  example,   $\length$  has  calls  from  $\main$  at
-$\pi_{12}$ and a recursive call at $\pi_6$.
+For  the  running  example,   $\fun$  has  calls  from  $\main$  at
+$\pi_{16}$ and a recursive call at $\pi_6$.
 So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{12}}{\main}  \cup
-\deltacall{\length}{\pi_6}{\length}$.
+     \deltacall{\fun}{\pi_{16}}{\main}  \cup
+\deltacall{\fun}{\pi_6}{\length}$.
 Filling in  the values gives:
 \begin{eqnarray*}
-\sigma_{\length}    &=&
- \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
+\sigma_{\fun}    &=&
+ \{\sigma_{\mathit{all}}\}  ~\cup~{\bcdr\sigma_{\mathit{\fun}}}
 \end{eqnarray*}
-As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
-of  $\sigma_{\length}$ are:
+As examples, the liveness of \Lanv{2}{\pl} and \Lanv{1}{\py} in terms
+of  $\sigma_{\fun}$ are:
 \begin{align*}
-\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
-  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
+\Lanv{2}{\pl} &= \clazy \cup \acar\Lf{\fun}{2}{\bcdr\sigma_{\mathit{\fun}}}
  \\
-\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
+\Lanv{1}{\py} &= \clazy \cup \bcar\sigma_{\mathit{\fun}}
 \end{align*}

### 2015-11-26 paper/paper.tex
Updated liveness eqns
latest on 26th

@@
-\documentclass[9pt]{sigplanconf}
+\documentclass[preprint,9pt]{sigplanconf}
 \usepackage{etex}
@@
 \usepackage{mathrsfs}
+\usepackage{lineno}
 \usepackage{booktabs}
@@
 \definecolor{Myblue}{rgb}{.2,0,1}
-%\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
-%\newcommand{\cblue}[1]{{\color{blue}{#1}}}
+\definecolor{Myred}{rgb}{1,.2,.2}
+
+\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
+\newcommand{\warning}[1]{{\color{Myred}{#1}}}
 %\newcommand{\blankout}[1]{}
 
+\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
 %\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
@@
 \begin{document}
+\linenumbers
 % --- Author Metadata here ---
@@
 % the body of your article BEFORE References or any Appendices.
-\cmt{{
-\numberofauthors{3} %  in this sample file, there are a *total*
+%\cmt{{
+%
+%\numberofauthors{3} %  in this sample file, there are a *total*
 % of EIGHT authors. SIX appear on the 'first-page' (for formatting
 % reasons) and the remaining two appear in the \additionalauthors
-section.
+%section.
 %
-\author{
+%\author{
 % 1st. author
-\alignauthor K. Prasanna Kumar \\
-       \affaddr{IIT Bombay,}\\
-       \affaddr{Mumbai 400076, India}
-       \email{prasannak@cse.iitb.ac.in}
+\authorinfo {Prasanna Kumar. K} 
+       {IIT Bombay, Mumbai 400076, India}
+       {prasannak@cse.iitb.ac.in}
 % 2nd. author
-\alignauthor Amey Karkare\\
-       \affaddr{IIT Kanpur}\\
-       \affaddr{Kanpur 208016, India}
-       \email{karkare@cse.iitk.ac.in}
+\authorinfo {Amitabha Sanyal} 
+       {IIT Bombay, Mumbai 400076, India}
+       {as@cse.iitb.ac.in}
 % 3rd. author
-\alignauthor Amitabha Sanyal \\
-       \affaddr{IIT Bombay}\\
-       \affaddr{Mumbai 400076, India}\\
-       \email{as@cse.iitb.ac.in}
-}
-}}
-\authorinfo{Double Blind Review}{}{}
+\authorinfo {Amey Karkare}
+       {IIT Kanpur, Kanpur 208016, India}
+       {karkare@cse.iitk.ac.in}
+% 3rd. author
+%% \alignauthor Amey Karkare\\
+%%        \affaddr{IIT Kanpur}\\
+%%        \affaddr{Kanpur 208016, India}
+%%        \email{karkare@cse.iitk.ac.in}
+%}
+%}}
+%\authorinfo{Double Blind Review}{}{}
 \maketitle
@@
 We consider  the problem of reducing  the memory required to  run lazy
-first-order functional programs.  To do this, we  analyze programs for
-liveness of  heap-allocated data.  The accompanying  garbage collector
-is modified  to use the result  of the analysis to  preserve only live
-data---a subset  of reachable data---during garbage  collection.  This
-results  in an  increase  in the  garbage  reclaimed and  consequently
-reduces the  memory footprint of  programs.  While this  technique has
-already been shown to yield  benefits for eager first-order languages,
-the lack of a statically determinable execution order and the presence
-of closures at runtime pose additional challenges for lazy languages.
-This requires changes  both in the earlier liveness  analysis and also
-in the  garbage collection scheme  required to reclaim memory  that is
-not live.
-
-The implementation  of our  analysis annotates each  potential garbage
-collection  point  in   the  program  with  a   set  of  deterministic
-finite-state automata (DFA) describing the  liveness at that point. We
-also implement a copying collector that  uses the DFA to preserve live
-objects.   Our  experiments  confirm  that  liveness-based  collection
-always increases  the garbage  reclaimed, consequently  decreasing the
-number of collections.  In addition,  it reduces the execution time of
-several programs.  Experiments  also reveal that the  best results are
-obtained  for   a  mixed  garbage   collection  scheme  that   does  a
-liveness-based    collection   for    fully    evaluated   data    and
-reachability-based   collection  for   closures.   On our benchmark
-programs, we obtain a decrease of up to 45\% in the number of
-collections, and a decrease of up to 88\% in the number of dragged
-cells.  Even though liveness based collector uses an expensive
-traversal scheme, the execution time is competitive (0.2X - 2.1X) when
-compared to a reachability based collector.
+first-order functional programs. We  address the problem by analyzing
+programs  for liveness  of  heap-allocated data.   The  result of  the
+analysis is  used to preserve  only live data---a subset  of reachable
+data---during garbage  collection.  The result  is an increase  in the
+garbage reclaimed and  a  reduction in  the peak memory requirement 
+of programs.   While this  technique has already  been shown  to yield
+benefits for  eager first-order  languages, the  lack of  a statically
+determinable execution  order and  the presence  of closures  pose new
+challenges  for lazy  languages.  These  requires changes  both in  the
+liveness analysis itself and in the design of the garbage collector.
+
+%% We use the  result of the analysis to annotate  each potential garbage
+%% collection  point  in   the  program  with  a   set  of  deterministic
+%% finite-state automata (DFA) describing the liveness at that point.
+We
+implement a copying collector that uses the results of the liveness
+analysis to preserve live
+objects, both evaluated (i.e.  in WHNF) and closures.  Our experiments
+confirm  that  for  programs  running with  a  liveness-based  garbage
+collector,   there  is   a   significant  decrease   in  peak   memory
+requirements.   In  addition,   a  sizable  reduction  in   the  number  of
+collections  ensures that  in spite  of using  a more  complex garbage
+collector,  the execution  time  of  a program  remains comparable  with
+the same program running with a reachabilty-based collector.
+\warning{On our benchmark programs, we obtain a decrease of up to 75\%
+  in the number of collections, and up to 50\% decrease in peak memory
+  requirements.}
 \end{abstract}
@@
 
-Functional programs use dynamically allocated memory extensively.  The
-allocation  is either  explicit (while using  constructors, for example) or
-implicit (as while creating  closures).  Programs  written in  lazy
+Functional programs make extensive use of dynamically allocated
+memory.  The
+allocation  is either  explicit (using  constructors, for example) or
+implicit (creating  closures).  Programs  written in  lazy
 functional languages put additional  demands on memory as they require
@@
 {\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
-expression.   The analysis  has two  parts.  First,  each function  is
-summarized in a context-sensitive manner as a {\em demand transformer}
-that transforms a {\em symbolic} demand  on its body to demands on its
-arguments.  This summary  is then used to step  through function calls
-during analysis.   Second, the concrete  demand on a function  body is
-obtained through a 0-CFA-like conservative approximation that combines
-the  demands on  all the  calls to  the function.   The result  of the
-analysis  is an  annotation of  each program  point with  finite-state
-automata capturing the liveness of  variables at that point. Depending
-on the program point where  garbage collection has been triggered, the
-GC consults a set of  automata to restrict reachability during marking.
-This results in an increase  in the garbage reclaimed and consequently
-in fewer garbage collections.
-
-While this idea has been shown to be effective~\cite{asati14lgc} for a
-first-order {\em  eager} language, a straightforward  extension of the
-earlier  technique  is  not  possible  for  lazy  languages,  where  a
-heap-allocated   object   may   be  an   unevaluated   expression   or
-closure. Since  data is made  live by  evaluation of closures,  and in
-lazy languages  the place in  the program where this  evaluation takes
-place cannot  be statically determined, laziness  complicates liveness
-analysis itself.   In addition, apart  from evaluated data,  we extend
-liveness-based collection  to closures---closures  with no  future use
-should also  be collected.  This  involves tracking closures  to their
-points of  creation at  runtime from where  we obtain  their liveness.
-This requires significant modification to the earlier GC scheme.
-
-Experiments    with   a    single    generation   copying    collector
+expression.   The  analysis has  two  parts.   First, we  calculate  a
+context-sensitive  summary   of  each   function  as  a   {\em  demand
+  transformer} that transforms a {\em  symbolic} demand on its body to
+demands on its  arguments.  This summary is used  to step through
+function  calls during  analysis.  Second,  the concrete  demand on  a
+function   body  is   obtained  through   a  0-CFA-like   conservative
+approximation  that combines  the  demands  on all  the  calls to  the
+function.   The result  of the  analysis is  an annotation  of certain
+program points  with deterministic finite-state  automata (DFA)  capturing the  liveness of
+variables  at  these points.  Depending  on  the program  point  where
+garbage  collection has  been  triggered,  the garbage collector consults  a set  of
+automata to restrict reachability during  marking.  This results in an
+increase in  the garbage reclaimed  and consequently in  fewer garbage
+collections.
+
+While this idea has been shown to be effective for a
+first-order {\em  eager} language~\cite{asati14lgc}, a straightforward  extension of the
+technique   is  not   possible  for  lazy   languages,  where
+heap-allocated objects  may include closures  (runtime representations
+of unevaluated expressions). Since data  is made live by evaluation of
+closures, and  in lazy languages the  place in the program  where this
+evaluation  takes  place  cannot be  statically  determined,  laziness
+complicates  liveness  analysis  itself.    In  addition,  apart  from
+evaluated   data,  we  need to  extend   liveness-based  collection   to
+closures i.e. we reclaim the space for closures  which have no future  use.
+Since a closure can  escape the scope in which it  was created, it has
+to carry the liveness information  of its free variables. Moreover, as
+execution  progresses and  possible  future uses  are eliminated,  the
+liveness information is updated with a more precise version.  All this
+requires significant  modification to  the earlier  garbage collection
+scheme for eager languages.
+
+\warning{Experiments  with  a   single  generation  copying  collector
   (Section~\ref{sec:experiments})  confirm  the  expected  performance
-benefits.  Liveness-based collection results in an increase in garbage
-reclaimed.  As  a consequence, there  is a  decrease in the  number of
-collections (up to 45\%) and a decrease  (up to 88\%) in the number of
-{\em  dragged}  cells  (cells  that  survive a  GC  after  their  last
-use). Besides, there is a reduction  in the overall execution time for
-some programs.  Our  experiments show that a mixed  strategy, where we
-do liveness based GC up to the closure boundary and reachability based
-GC for the arguments (free variables) of the closure, gives us similar
-benefits as  a fully  liveness based  GC (where  the arguments  of the
-closure  are  also  collected  using liveness)  with  lesser  overhead
-(Section~\ref{sec:strategies}).
+  benefits.   Liveness-based  collection  results in  an  increase  in
+  garbage reclaimed.   As a  consequence, there is  a decrease  in the
+  number of  collections (up to 45\%)  and a decrease (up  to 88\%) in
+  the number  of {\em dragged}  cells (cells  that survive a  GC after
+  their  last use).  Besides,  there  is a  reduction  in the  overall
+  execution time for some programs.  Our experiments show that a mixed
+  strategy, where we  do liveness based GC up to  the closure boundary
+  and reachability based GC for  the arguments (free variables) of the
+  closure,  gives us  similar benefits  as a  fully liveness  based GC
+  (where  the  arguments  of  the closure  are  also  collected  using
+  liveness) with lesser overhead (Section~\ref{sec:strategies}).}
 
@@
 	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
-            (\PRIM\ 1\ (\length (\CDR\  \pl)))))
+            (\PRIM\ 1\ (\length\  (\CDR\  \pl)))))
             \UNL{0}
@@
         stops at edges marked $\times$ during garbage collection for
-        an ideal collector.
+        a liveness-based collector.
       \end{tabular}
@@
 node either  represents atomic  values ($\NIL$,  integers etc.),  or a
-\CONS\ cell containing $\CAR$ and $\CDR$  fields, or a chunk of memory
-representing  an  unevaluated  value   also  called  a  {\em  closure}
-(represented  by  hatched  boxes).   Edges   in  the  graph  are  {\em
-  references} and emanate from variables  or fields.  The figure shows
+\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
+(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
+  references} and represent  variables  or fields.  The figure shows
 the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
-triggered    by   the    test   (\NULLQ~(\CAR~\pz))    in   the
-\SIF\ expression.
+triggered by the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.
 
 The edges  shown by thick  arrows are those  which are live  at $\pi$.
-Ideally, a cell should be  preserved during garbage collection only if
-it is reachable from  the root set through a path  of live edges.  All
-other  cells can  be reclaimed.   Thus if  a garbage  collection takes
-place at $\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b),
-an ideal liveness-based  collector (LGC) will preserve  only the cells
-referenced by  $\pz$, $(\CDR~ \pz)$  and (the cells  constituting) the
-closure   referenced   by   $(\CDR~(\CDR~\pz))$.    In   contrast,   a
-reachability-based collector (RGC) would have preserved all cells.
-
-While an ideal garbage collector is  impractical, in this work we show
-that  static analysis  of  heap data  can help  practical garbage
-collectors in reclaiming more  garbage.  The specific contributions of
-this paper are
+%% Only those  cell which have a future use  should be preserved
+%% during garbage collection, all other
+%% cells can be  reclaimed.
+Thus if a garbage collection  takes place at
+$\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b), a
+liveness-based collector (LGC) will preserve only the cells referenced
+by  $\pz$,  and the live cells  constituting the  closure
+referenced by $(\CDR~\pz)$.   In contrast, a reachability-based
+collector (RGC) will preserve all cells.
+
+In  this work  we show  that  static analysis  of heap  data can  help
+garbage collectors in reclaiming more garbage.  The specific
+contributions of this paper are
 \begin{itemize}
 \item  We  formulate  a  liveness  analysis for  a  lazy  first  order
-  functional language  and prove its correctness.  We show  how to
-  compute  function  summaries  in  a context  independent  manner  to
-  achieve  a  context  sensitive  analysis. We  further  describe  the
-  transformation of the  liveness equations to grammar rules  in a way
-  that the language  accepted by the resulting  grammar describes live
-  heap locations.
-\item We prove  the undecidability of the grammar  that results from
-  the  analysis.   To overcome  this,  we  present an  algorithm  to
-  transform grammars  to finite  state automata  (DFA) such  that the
-  language accepted by the  DFA over-approximates the liveness (and
-  hence a safe approximation).
-\item We have implemented a garbage  collector that can use the DFA
-  to  retain live  cells that  also include  closures. We  have done
-  extensive  experiments with  this garbage  collector and  obtained
-  empirical results  that show  the effectiveness  of liveness-based
-  garbage collection.
+  functional language and  prove its correctness. The correctness
+  proof involves setting up a non-standard semantics as a
+  specification of liveness and then proving that the analysis is
+  correct with respect to the specification.   
+\item Our  analysis results  in a set  of context-free  grammars along
+  with  a fixed  set  of non-context-free  productions.  The  decision
+  regarding  whether   to  copy  a  cell   during  garbage  collection
+  translates  to the  membership problem  of such  grammars.  In  this
+  paper, we  have shown the  undecidability of the  membership problem
+  for  the  class  of  grammars  that result  from  our  analysis.  To
+  circumvent the  problem, we  over-approximate the  grammars generated
+  by the analysis with DFAs.
+\item We  have implemented a garbage  collector that uses the  DFAs to
+  retain live  cells.  Our  experiments reveal  interesting space-time
+  trade-offs  in  the  engineering  of  the  collector---for  example,
+  updating liveness  information carried in closures  during execution
+  results  in   more  garbage  being  collected.    Empirical  results
+  demonstrate the effectiveness of liveness-based garbage collection.
 \end{itemize}
+To the  best of our
+  knowledge, this is the first work on liveness-based garbage
+  collection      for      lazy     languages.       All      previous
+  attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
+    asati14lgc, karkare06effectiveness} have targeted either
+  imperative or  eager functional languages.
 
@@
 
-Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
-language  considered  in our  analysis  and  a small-step  operational
-semantics  for it.   We present  our liveness  analysis for  this lazy
-language  in Section~\ref{sec:liveness}.   Section~\ref{sec:computing}
-describes  the  formulation of  the  liveness  as  a grammar  and  its
-approximation  to   DFA.  It  also   contains  a  proof   showing  the
-undecidability  of  the  precise  liveness in  our  formulation  which
-necessitates  the approximation  to DFA.   Section~\ref{sec:GC-scheme}
-discusses different ways of handling closures and their pros and cons.
-We  report our  experimental results  in Section~\ref{sec:experiments}
-along with some observations.  Section~\ref{sec:relatedwork} discusses
-previous  work   related  to  garbage  collection   and  liveness  and
-Section~\ref{sec:conclusion}   discusses   possible   extensions   and
-concludes the paper.
+\warning{Section~\ref{sec:defs}   introduces   the   syntax   of   the
+  programming language  considered and gives a  small-step operational
+  semantics for it.   The liveness analysis for this  language and its
+  soundness   proof   is  presented   in   Section~\ref{sec:liveness}.
+  Section~\ref{sec:computing} describes the formulation of liveness as
+  grammars. We  also give a  proof of undecidability of  such grammars
+  and    show    how   they    can    be    approximated   by    DFAs.
+  Section~\ref{sec:GC-scheme}   discusses  details   of  the   garbage
+  collector,  in  particular the  use  of  liveness DFAs  for  garbage
+  collection.     We    report    our    experimental    results    in
+  Section~\ref{sec:experiments}   along    with   some   observations.
+  Section~\ref{sec:relatedwork}  discusses  previous work  related  to
+  garbage  collection  and liveness  and  Section~\ref{sec:conclusion}
+  discusses possible extensions and concludes the paper.}
 \section{The target language---syntax and semantics}
@@
 expressibility,  this form  has  the benefit  of  making explicit  the
-creation of  closures through  the $\LET$  construct.  $\LET$s  in our
-language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
-\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
-structures in  a pure language. The  restriction of \LET\ to  a single
+creation of  closures through  the $\LET$  construct.  \warning{ For
+  now, we assume  $\LET$s  in our
+language  are non-recursive;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
+\IN\,\, e$, $x$ should not occur in  $s$}. The  restriction of \LET\ to  a single
 definition  is for  ease  of  exposition---generalization to  multiple
@@
 each variable in a program to  be distinct, so that no scope shadowing
-occurs---this  simplifies  the  reasoning   about  the  program.   The
-$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
-evaluation of closures.
+occurs---this  simplifies  reasoning.
 
 The body  of a function ${\mathit  f}$ is denoted  as $e_{\mathit f}$.
-We   shall   extend   this   convention   to   the   main   expression
-$e_\mainpgm$. We assume that each program has a distinguished function
+ We assume that each program has a distinguished function
 \mainpgm\      with      the      definition     $(\DEFINE\      ({\tt
@@
 \begin{eqnarray*}
-   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\, e_\mainpgm
-    \hspace{5em} \mbox{\em --- program}\\
+   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
+   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
     \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
@@
     e)
-    \hspace{0.2em} \ \ \ \ \mbox{\em --- function def} \\
+    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
 e \in \mathit{Expr} & \!\!\!::=\!\!\! &
@@
 \begin{figure*}[t!]
+%\comment{
+
 \begin{center}\footnotesize
-\renewcommand{\arraystretch}{1}
+\renewcommand{\arraystretch}{1.5}
 \begin{tabular}{|c|c|c|}
@@
 \hline 
-          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
-  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
+          & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, \kappa
+  \rightsquigarrow \rho',\, S,\, H[\ell := \kappa],\, e$    &  \sc{const}
 \\
 \hline
-          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
+          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, H,\, (\CONS~x~y)
 \rightsquigarrow
-$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
-(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
+$  $\rho',\, S,\, H[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
+\hline
+
+$H(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\,
+(\CAR~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$      &
+\sc{car-select} \\
+\hline
+
+$H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
+(\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho'\rangle), (\CAR~x))\!:\!S,\, H,\, s$      &
+\sc{car-1-clo} \\
+\hline
+
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (\CAR~x)
 \rightsquigarrow
-\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
+\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,H,\, s$      &
 \sc{car-clo}
@@
 \hline
+
 $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
-$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
-\sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
+ & {$\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, (+~x~y)  \rightsquigarrow$
+$\rho', \,S,\, H[\ell := H(\rho'(x)) + H(\rho'(y))],\, e$}      &
+\sc{prim-add} \\
+\hline
+
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (+~x~y)
 \rightsquigarrow
-\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
+\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, H,\, s$      &
 \sc{prim-1-clo} \\
 \hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, H,\, (+~x~y)
 \rightsquigarrow
-\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
+\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, H,\, s$      &
 \sc{prim-2-clo} \\
@@
 {$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, H,\,
 (f~\myvec{x})  \rightsquigarrow
-[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
+[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, H,\, e_{\mathit{f}}$      &
 \sc{funcall} \\
 \hline
-$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
+$\ell$ is a new location& {$\rho,\, S,\, H,\, (\LET~x\leftarrow s~\IN~e)
   \rightsquigarrow$
-$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
-    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e$} &
+$\rho\oplus[x \mapsto \ell],\, S,\, H[\ell \mapsto \langle s,
+    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
+    \rangle],\, e$} &
 \sc{let} \\
 \hline
-$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho, S, H,  e_1$ & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho, S, H,  e_2$ & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
+$H(\rho(x)) \ne 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho,\, S,\, H,\,  e_1$ & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho,\, S, \,H, \, e_2$ & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\,
   (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
+\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, H, \, s$}
 &
@@
 \hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
-e)\!:\!S, H,
-(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
+{$H(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
+e)\!:\!S,\, H,\,
+(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$ &
 \sc{return-whnf}\\
 \hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
+$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\, (\SRETURN~x)
   \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
+$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, H,\,  s$} &
 \sc{return-clo} \\
@@
 \end{center}
+%}\comment
 \end{figure*}
@@
 \subsection{Semantics}
+%\comment{
 We now give  a small-step semantics for our  language.
@@
 \begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (Data + \{empty\}) & \mbox{-- Heap}\\
+d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
 v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
-c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
-d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
+c:   & \mathit{Clo} &=& \mathit{(App \times Env)}& \mbox{-- Closures}\\
+\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
+\mbox{-- Environment} \\
 \end{array}
@@
 Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
-Since all data objects are boxed, we model an environment as a mapping
-from the set  of variables of the program  $\mathit{Var}$ to locations
-in the  heap.  These  either contain  a value in  WHNF (a  number, the
-empty  list  $\NIL$,  or  a  \CONS\  cell  with  possibly  unevaluated
+A non-empty location either contains  a value in  \emph{WHNF} (a  number, the
+empty  list  $\NIL$ or  a  \CONS\  cell  with  possibly  unevaluated
 constituents) or  a {\em closure}.   A closure  is a pair  $\langle s,
 \rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
-maps free variables of $s$ to their respective locations.
+maps free variables of $s$ to their respective locations. Since all data objects are boxed, we model an environment as a mapping
+from the set  of variables of the program  $\mathit{Var}$ to locations
+in the  heap.
 
@@
 Here \stk\ is a stack of continuation frames.  Each continuation frame
-is of the form $(\rho, x, e)$, signifying that the location of $x$ has
-to be updated  with the value of the  currently evaluating expression,
-and $e$ is to be evaluated  next in the environment $\rho$.  Note that
-the state of the transition  system consists of the current evaluation
-context,  given  by  $\rho$  and $e$,  and  the  suspended  evaluation
-contexts recorded  in the stack  $S$.  The heap  $H$ is common  to all
-contexts.
-
-The     start      state     is      $([\;]_\rho,([\;]_\rho,     \ans,
-(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm))$.  In this, the initial
-environment  and  the initial  heap  are  denoted by  $[\;]_\rho$  and
-$[\;]_H$,  and the  initial stack  consists of  a single  continuation
-frame in which \ans\ is a distinguished variable that would be updated
-with  the value  of (\mainpgm).   In addition,  \print\ is  a function
-modeling a printing mechanism---a standard run-time support assumption
-for lazy  languages---that prints  the value of (\mainpgm).   Notice the
-use of the operator $:$ for adding elements to the top of the stack.
-
-%% In absence of a static type-checking mechanism, programs with other
-%% than syntactic  errors lead to states  not covered by  any rules in
-%% the small-step semantics.
+is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
+has  to  be  updated  with  the  value  of  the  currently  evaluating
+expression, and $e$ is to be evaluated next in the environment $\rho$.
+Note  that  the  state  of  the  transition  system  consists  of  the
+\emph{current evaluation  context}, given by  $\rho$ and $e$,  and the
+\emph{suspended  evaluation  contexts}  (represented  as  continuation
+frames) in the stack $S$.   \warning{The
+3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
+system will be denoted $\mathcal{S}$.}
+
+The  initial  state  of  the  transition  system  described  above  is
+$([\;]_\rho,\,   ([\;]_\rho,   \,   \rho(\ans),\,   (\print~\ans)):[\;]_{S},
+[\;]_{H},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_H$  are the
+empty environment and  the empty heap. The  initial stack consists
+of  a single  continuation frame  in  which \ans\  is a  distinguished
+variable  that  would   eventually  be  updated  with   the  value  of
+(\mainpgm).  In  addition, \print\ is  a function modeling  a printing
+mechanism---a   standard   run-time   support  assumption   for   lazy
+languages---that prints  the value of  (\mainpgm).  The operator $:$
+pushes elements on top of the stack.
+
 The   notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
-environment that maps variables $x_i$
-%$x_1,  \ldots, x_n$
-to locations $\ell_i$
-%$\ell_1,  \ldots,  \ell_n$,
-and  $\heap[\ell :=  d]$
-indicates the  updating of  a heap \heap\  at $\ell$ with  $d$.  $\rho
-\oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
-and $\lfloor \rho \rfloor_X$  represents the environment restricted to
-the locations in $X$. Finally $FV(s)$ represents the free variables in
-the application $s$.
+environment  that  maps  variables  $x_i$ to  locations  $\ell_i$  and
+$\heap[\ell := d]$  indicates the updating of a heap  \heap\ at $\ell$
+with  $d$.   $\rho \oplus  \rho'$  represents  the environment  $\rho$
+shadowed  by  $\rho'$  and  $\lfloor \rho  \rfloor_X$  represents  the
+environment  restricted  to  the  variables in  $X$.  Finally  $FV(s)$
+represents the  free variables  in the  application $s$  and $addr(c)$
+gives the address of the closure $c$ in the heap.
 
 The small-step semantics  is shown in Figure~\ref{fig:lang-semantics}.
-As  a sample,  consider  the  rules  for  $(+~x~y)$, an  application
-involving a binary operator strict in both arguments.   When
-both $x$  and $y$ are already  in WHNF ({\sc prim-whnf}),  the heap is
-updated with the  sum of the arguments and  the next expression to be
-evaluated is picked from the continuation on the stack. If either of the
-arguments is not in
-WHNF  ({\sc  prim-1-clo}  and   {\sc  prim-2-clo}),  it  is  sent  for
-evaluation, and a continuation  marks that the evaluation of $(+~x~y)$
-has to be resumed after the argument is evaluated.
-
+Unlike   an   eager  language,   evaluation   of   a  let   expression
+$(\LET~x\leftarrow  s~\IN~e)$ does  not  result in  the evaluation  of
+$s$. Instead,  as the {\sc let}  rule shows, a closure  is created and
+bound to  $x$. The  program points which  trigger evaluation  of these
+closure   are  the   condition  of   an  $\SIF$   ({\sc  IF-clo})   and
+\SRETURN\ ({\sc  return-clo}).  We  call such  points \emph{evaluation
+  points   $(\epath)$}  and   label  them   with  $\psi$   instead  of
+$\pi$. As an  example of closure  evaluation, we  explain the
+three rules for  $(\CAR~x)$.  If $x$ is a closure,  it is evaluated to
+WHNF, say $(d_1, d_2)$.  This is  given by the rule {\sc car-clo}.  If
+$d_1$ is  not in WHNF,  it is  also evaluated ({\sc  car-1-clo}).  The
+address  to be  updated  with the  evaluated value  is
+recorded  in  a  continuation  frame.    This  is  required  for  lazy
+evaluation, else $d_1$ may  be evaluated more than once
+due to  sharing~\cite{Jones87}.  Only after this  is the actual selection  done ({\sc
+  car-select}).
+
+%}\comment
 %==============================================================
@@
 \begin{figure*}[t]
+\comment{
 \renewcommand{\arraystretch}{1}
@@
 $\NIL$}\\
-\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha
+\mathit{ref\/}(\pi:(\CONS~x~y),\sigma,\Lfonly)
+          &=& \{x_{\pi}.\alpha \mid \acar\alpha \in \sigma\} \cup
+  \{y_{\pi}.\alpha 
 \mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
+\mathit{ref\/}(\pi:(\CAR~x),\sigma,\Lfonly)
           &=&    \begin{array}{l l}
-                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
+                    \{x_{\pi}.\epsilon\} \cup \{x_{\pi}.\acar\alpha \mid \alpha \in
 \sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
@@
                  \end{array} \\
-\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
+\mathit{ref\/}(\pi:(\CDR~x),\sigma,\Lfonly)
           &=&    \begin{array}{l l}
-                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
+                    \{x_{\pi}.\epsilon\} \cup \{x_{\pi}.\acdr\alpha \mid \alpha \in
 \sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
@@
                  \end{array} \\
-\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
+\mathit{ref\/}(\pi:(\PRIM~x~y),\sigma,\Lfonly)
           &=&    \begin{array}{l l}
-                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne
+                    \{x_{\pi}.\epsilon, y_{\pi}.\epsilon\},  & \mbox{if}~\sigma \ne
 \emptyset\\
@@
                  \end{array} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
+\mathit{ref\/}(\pi:(\NULLQ~x),\sigma,\Lfonly)
           &=&    \begin{array}{l l}
+                    \{x_{\pi}.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \emptyset  & \mbox{otherwise}
+                 \end{array} \\
+\mathit{ref\/}(\pi:(f~\myvec{x}),\sigma,\Lfonly)
+%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
+          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
+               \bigcup_{i=1}^n x_{i_{\pi}}.\Lf{f}{i}{\sigma}
+               \end{array}
+%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
+\\[1ex]
+\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& 
+x.\sigma\\
+%% \mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
+%% \mapsto
+%% x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
+%%   new e-path terminating with $\psi$}\\
+\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
+\begin{array}{l l}
+                    \mathcal{L}(e_1,\sigma,\Lfonly) \cup
+        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
 \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
@@
                  \end{array} \\
-\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
-\mapsto
-x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
-  new e-path terminating with $\psi$}\\
-\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
-\begin{array}{l l}
-                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
-        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
-        \lbrack\lbrace \mathit{\epath} \mapsto
-\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
-        \emptyset_{\cal{M}}  & \mbox{otherwise}
-                 \end{array} \\
-&&\mbox{where $\mathit{\epath}$ is a new e-path
-  terminating with $\psi$}\\
-\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
-\mbox{where}~\mathcal{M} =
-\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
-\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
-&& \makebox[0mm]{\hspace*{7cm}
- $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
-           x_2\, \ldots\, x_n)/x]))$,} \\
- \makebox[0mm]{\hspace*{8cm} where
-     $x_1\, \ldots\, x_n$ are the variables in $s$}
+%% \mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+%%         \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
+%% \mbox{where}~\mathcal{M} =
+%% \mathcal{L}(e,\sigma,\Lfonly),~\Lv =
+%% \mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
+%% && \makebox[0mm]{\hspace*{7cm}
+%%  $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
+%%            x_2\, \ldots\, x_n)/x]))$,} \\
+%%  \makebox[0mm]{\hspace*{8cm} where
+%%      $x_1\, \ldots\, x_n$ are the variables in $s$}
+\mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+\mathit{ref\/}(s,\sigma',\Lfonly) \cup \Lv \cup \{x.\sigma'\} 
+~\mbox{where}~ \mathcal{\Lv} = \mathcal{L}(e,\sigma,\Lfonly), ~\mbox{and}~\sigma' = \bigcup_{\pi}  \Lv(x_{\pi})
 \end{eqnarray*}
@@
 \end{minipage}
-  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
+  \caption{Liveness equations and judgement
+    rule}\label{fig:live-judge}
+}%comment
 \end{figure*}
@@
 $\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
-$\{\epsilon, \acdr,  \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
-future computations can only refer up  to the second and third members
-of $x$.  A  {\em liveness environment} is a mapping  from variables to
-subsets of $\{\acar,\acdr\}^\ast$,  but often expressed as  a set, for
-example by writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$
-instead       of       $[x       \mapsto\{\epsilon,\acdr,\acdr\acdr\},
-  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $x \mapsto
-\{\epsilon\}$ represents access using $x$  itself and $x \mapsto \{\}$
-indicates that $x$  is dead.  We associate  liveness environments with
-program points.  The  liveness at $\pi:e$ is the  liveness just before
-executing $e$.
-
-A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
-represents the future use of variables, a demand represents the future
-use of the value of an expression.  The demand on an expression $e$ is
-again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
-which the context of $e$ may explore of $e$'s result.  To see the need
-for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
-(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context of  this
-expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
-  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
-$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes the  demand on
-$(\CDR~y)$.    This,  in   turn,  generates   the  liveness   $\lbrace
-y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
-of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
-\rbrace$  paths  of  the  result of  $(\CDR~y)$.  In  strong  liveness
-analysis       (dual      of       classical      faint       variable
-analysis~\cite{horwitz.faint}),  $y$ and  $z$  are live  at the  entry
-$\pi: x:=y+z$, if  and only if $x$  is live at exit of  $\pi$.  In our
-terminology, the  liveness $\lbrace  \epsilon\rbrace $  of $x$  at the
-exit from $n$ becomes the demand on $y+z$, and this, in turn generates
-the liveness $\lbrace y.\epsilon, z.\epsilon  \rbrace$ at the entry of
-$n$.
-
-We follow the notations appearing in \cite{asati14lgc}. In particular,
-We use $\sigma$  to range over demands, $\alpha$ to  range over access
-paths and  $\Lv$ to  range over  liveness environments.   The notation
-$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
-\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
-abuse notation to  juxtapose an edge label and a  set of access paths:
-$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
+$\{\epsilon, \acar, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$,
+then future  computations can only  refer up  to the second  and third
+members  of $x$.   A  {\em  liveness environment}  is  a mapping  from
+variables to access paths, but often expressed as
+a set,  for example  by writing $\{x.\epsilon,  x.\acdr, x.\acdr\acdr,
+y.\epsilon\}$  instead  of  $[x  \mapsto\{\epsilon,\acdr,\acdr\acdr\},
+  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $y \mapsto
+\{\epsilon\}$ represents access using $y$  itself and $z \mapsto \{\}$
+indicates that $z$ is dead. 
+In lazy languages, liveness environments are associated with regions of
+programs instead of program points.
+
+A notion that generalizes liveness is {\em demand}.  While liveness 
+gives the patterns of future uses of a  variable, demand
+represents the future  use of the value of an  expression.  The demand
+on an expression  $e$ is again a set of  access paths---that subset of
+$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
+result.   To see  the need  for demands,  consider the  let expression
+$\pi:(\LET~x\leftarrow  (\CDR~y)~\IN~ \pi':\SRETURN~x)$.   Assume that
+the context  of this expression  places the demand  $\lbrace \epsilon,
+\acar \rbrace$. Since the value of the expression is
+the value  of $x$, the demand  translates to the liveness  $[x \mapsto
+  \lbrace   \epsilon,  \acar   \rbrace]$  at   $\pi'$.   Due   to  the
+\LET\ definition which binds $x$ to $(\CDR~y)$, the liveness of $x$ at
+$\pi'$ now becomes the demand on $(\CDR~y)$.  This, in turn, generates
+the liveness  $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar  \rbrace$ at
+$\pi$.  These are  the set of $y$-rooted accesses  required to explore
+$\lbrace \epsilon, \acar  \rbrace$ paths of the  result of $(\CDR~y)$.
+%% As an analogy  with classical (strong) liveness analysis,  $y$ and $z$
+%% are live  at the entry $\pi:  x:=y+z$, if and  only if $x$ is  live at
+%% exit   of  $\pi$.    In   our  terminology,   the  liveness   $\lbrace
+%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand on
+%% $y+z$, and this,  in turn generates the  liveness $\lbrace y.\epsilon,
+%% z.\epsilon \rbrace$ at the entry of $\pi$.
+
+We use $\sigma$
+to  range over  demands,  $\alpha$  to range  over
+access  paths and  $\Lv$  to  range over  liveness
+environments.   The   notation  $\sigma_1\sigma_2$
+denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
+\alpha_1     \in     \sigma_1,    \alpha_2     \in
+\sigma_2\rbrace$.  Often  we shall  abuse notation
+to juxtapose  an edge  label and  a set  of access
+paths;   $\acar\sigma$   is    a   shorthand   for
+$\lbrace\acar\rbrace\sigma$.
+
+\input{example-program-and-liveness-maps}
+%%\label{fig:mot-example2}
 
@@
 
+%\comment{ 
+Consider  the program in  Figure~\ref{fig:mot-example2}.  As
+  mentioned earlier,  a lazy  evaluation of  the $\LET$  expression at
+  $\pi_1$ creates a closure for $(length~x)$ instead of evaluating it.
+  Since the  closure may escape the  scope in which it  is created, it
+  carries a copy  of $x$ within itself.   We treat the copy  of $x$ in
+  the closure as being separate from the $x$ introduced by the $\LET$,
+  and call  it a \emph{closure variable}.   For liveness calculations,
+  such variables are distinguished from variables introduced by \LET s
+  and  function arguments  (called \emph{stack  variables} since  they
+  reside in the activation stack). Notationally, we distinguish
+  closure variables from their corresponding stack variables by
+  subscripting them with their program points of
+  creation.\footnote{Mutiple occurrences of the same variable in an
+    application are further distinguished by their positions in the
+    application.}
+
+  Since a  closure is evaluated  only at evaluation points,  a closure
+  variable is attributed  with the same liveness in the  region of the
+  program from the  point of creation to  reachable evaluation points.
+  This is  also true  of stack  variables, because,  as we  shall see,
+  stack variables derive their  liveness from closure variables. Thus,
+  there are  two major differences  in our formulation of  liveness of
+  lazy   languages  with   liveness  of   eager 
+  languages~\cite{asati14lgc}:    (i)  the
+  introduction of  closure variables in the  liveness calculation, and
+  (ii) a  single liveness value  for each variable that  is applicable
+  from its creation point to evaluation points.
+
+  Closure  variables get  their  liveness values  through  a chain  of
+  dependences beginning at  a variable at an evaluation  point.  As an
+  example,  the  variable $z$  returned  at  $\psi_2$ depends  on  $y$
+  through  the  expression $(+~y~1)$.   $y$  in  turn depends  on  the
+  closure  variable  $x_{\pi_1}$   through  $(length~x_{\pi_1})$.   We
+  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!z \leftarrow
+  (+~y~1),\,  y  \leftarrow   (length~x_{\pi_1})\rbrack$.   Indeed,  the
+  closures  in  the  heap  are   a  runtime  representation  of  these
+  dependences.  Since  $z$  is  evaluated   at  $\psi_2$  due  to  the
+  expression $\SRETURN~z$,  the demand made by  the calling context(s)
+  of  $f$ places  a demand  on  $z$ which  will impart  a liveness  to
+  $x_{\pi_1}$.  Other dependence chains which result in a liveness for
+  $x_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!y       \leftarrow
+  (length~x_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!w \leftarrow (/~u~y),\,\,
+  y  \leftarrow  (length~x_{\pi_1})\rbrack$.   The  liveness  analysis
+  described in this section declares the liveness of $x_{\pi_1}$ to be
+  a union of the liveness arising  out of these dependence chains.  To
+  be safe, a  garbage collection during evaluation of  $y$ at $\psi_1$
+  has  to  use  this  liveness  to copy  the  heap  from  $x_{\pi_1}$.
+  However,  notice that  if  a garbage  collection  takes place  while
+  evaluating $z$ at $\psi_2$, it can safely consider only the liveness
+  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!z \leftarrow
+  (y+1),\,   y  \leftarrow   (length~x_{\pi_1})\rbrack$.   The   garbage
+  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
+  generalization  of   this  observation   to  dynamically   select  a
+  evaluation point specific liveness in order to collect more garbage.
+
 Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
-parts. The function $\mathit{ref}$,  takes  an application $s$ and
-a demand $\sigma$, and returns  the incremental liveness generated for the
-free  variables   of  $s$  due  to  the   application.   The  function
-$\mathcal{L}$  uses   $\mathit{ref}$  to  propagate   liveness  across
-expressions.
-
- Since  in a  lazy language,  an  expression is  not evaluated  unless
- required. The null demand ($\emptyset$) does not generate liveness in
- any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
- non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed to  the
+parts. The  function $\mathit{ref}$,  takes an  application $s$  and a
+demand $\sigma$ and returns the incremental liveness generated for the
+free variables of $s$ due to  the application.  This will be consulted
+during garbage collection  while exploring the heap  starting from the
+closure  variables.  The function $\mathcal{L}$  uses  $\mathit{ref}$ to  propagate
+liveness across expressions.
+
+In  a   lazy  language,   an  expression   is  not   evaluated  unless
+required. Therefore  the null  demand ($\emptyset$) does  not generate
+liveness in any of the rules defining $\mathit{ref}$ or $\mathcal{L}$.
+A non-null  demand of  $\sigma$ on (\CDR~$x$),  is transformed  to the
 liveness $\{x.\epsilon,  x.\acdr\sigma\}$.  In an opposite  sense, the
@@
 
-For the  case of a  function call we  use a third  parameter $\Lfonly$
+In case of a  function call, we  use the third  parameter $\Lfonly$
 that  represents  the  summaries  of all  functions  in  the  program.
-$\Lfonly_{\mathit  f}$  (the component  of  $\Lfonly$  for a  specific
-function $f$), expresses  how the demand $\sigma$ on a  call to $f$ is
+$\Lfonly_{\mathit  f}$  (the summary for a  specific
+function $f$) expresses  how the demand $\sigma$ on a  call to $f$ is
 transformed into  the liveness of  its parameters at the  beginning of
-the   call.    $\Lfonly$  is   determined   by   the  judgement 
-$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
-This  rule  describes the  fixed-point  property  to be  satisfied  by
-$\Lfonly$, namely, the demand transformation assumed for each function
-in  the  program should  be  the  same  as the  demand  transformation
-calculated    from    its    body.      As    we    shall    see    in
+the  call.  $\Lfonly$  is determined  by the  judgement $\mathit{Prog}
+\len  \Lfonly$ using  inference rule  ({\sc live-define}).   This rule
+describes  the  fixed-point property  to  be  satisfied by  $\Lfonly$,
+namely, the  demand transformation  assumed for  each function  in the
+program should  be the  same as  the demand  transformation calculated
+from      its      body.       As       we      shall      see      in
 Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
@@
 We next  describe the function $\mathcal{L}$  that propagates liveness
-across expressions.  Note that,  unlike eager languages, the syntactic
-occurrence  of an expression  in a  program denotes  the point  of
-creation of its closure, not necessarily its evaluation.
-To see the  consequences of this, consider
-Figure~\ref{fig:mot-example2} which shows an analysis  of the body of
-a function $\length$ with a demand $\sigma$.  Clearly, this is also the
-demand on the return value $z$ giving the liveness at the program point
-$\pi_8$ as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated
-backwards as in traditional liveness analysis.
-
-Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
-\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
-generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
-the following observations:
+across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
+{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
+value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
+liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
+$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
+$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
+liveness $\{x.\epsilon\}$ is created and  added to the union.  
+%% Also  notice  a  consequence  of laziness:  the  entire  expression
+%% including the  condition is not  evaluated if  the demand on  it is
+%% $\emptyset$.  This results in the empty liveness environment.
+To understand the  liveness rule for $\LET~x \leftarrow~s~\IN~e$, observe that the value of  $\LET$ is the value
+of  its body  $e$.   Thus the  liveness environment  $\Lv$  of $e$  is
+calculated for the given demand $\sigma$. Since the stack variable $x$
+is copied to each of the  closure variable $x_{\pi}$, the liveness of
+$x$ is the union of the
+liveness of the  closure variables.  This liveness,  say $\sigma'$, is
+also the demand on $s$, and the resulting liveness environment $\mathit{ref}(s,
+\sigma', \Lfonly)$  is added  to $\Lv \cup  \{x.\sigma\}$ to  give the
+overall liveness environment for $(\LET~x \leftarrow~s~\IN~e)$.
+
+\warning{Note  carefully that  $\Lv_y$ specifies  the liveness  of the
+  reachable heap starting  from $y$ and extending  till closures.  The
+  liveness  of further  heap  contents starting  from  a closure,  say
+  $(\CAR\, x_{\pi})$, is given  by $\Lv_{x_{\pi}}$.  Therefore, in our
+  implementation,   a  closure   carries   the  liveness   environment
+  consisting of its free variables.}
+
+%} %comment
+
+\subsection{Soundness of analysis}
+
+%% \begin{figure*}[t!]
+%% \begin{center}\footnotesize
+%% \renewcommand{\arraystretch}{1.5}
+
+%% \begin{tabular}{|c|c|c|}
+%% \hline
+%% Premise & Transition & Rule name \\ 
+%% \hline
+
+%% \hline
+%% \makecell[t]{$GC(\rho_1, S, H_1, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
+%%   H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
+%% \sc{car-clo-bang}
+%% \\
+%% \hline
+%% \makecell[t]{$GC(\rho_1, S, H_1, \Lv, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x))
+%% \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
+%%   H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
+%%   (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
+%% \sc{car-clo}\\
+%% \hline
+%% \end{tabular}
+
+%% \medskip
+
+%% \begin{tabular}{|c|c|c|}
+%% \hline
+%% Premise & Transition & Rule name \\ 
+%% \hline
+
+%% \hline
+
+%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
+%%   H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
+%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
+%% \\
+%% \hline
+%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
+%%   \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+%% \hline
+%% $H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
+%%   \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
+%%   \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
+%% \sc{car-whnf} \\
+%% \hline
+%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+%%   H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+%%   (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
+%% \sc{car-clo}
+%% \\
+
+%% \hline
+%% $H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
+%% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
+%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, H,\, s, \, \cred{\sigma}$      &
+%% \sc{car-1-clo} \\
+
+%% \hline
+%% $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H,
+%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, H[\rho'(z)
+%%     := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma'}$}      &
+%% \sc{prim-whnf} \\
+%% \hline
+%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+%%   H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+%% \sc{prim-1-clo} \\
+%% \hline
+%% $H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+%%   S, H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
+%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+%% \sc{prim-2-clo} \\
+%% \hline
+%% {$\mathit{g}~\mbox{defined as}$
+%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
+%%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
+%%     \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma}$}      &
+%% \sc{funcall} \\
+%% \hline
+%% $\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
+%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+%%   \ell]\rangle], e, \cred{\sigma}$} &
+%% \sc{let} \\ 
+%% \hline
+%% $H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma}$} & \sc{if-true} \\
+%% \hline
+%% $H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+%%   \cred{\sigma}$   $\rightsquigarrow
+%% \rho, S, H,  e_2, \cred{\sigma}$} & \sc{if-false} \\
+%% \hline
+%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
+%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, H, s,
+%% \cred{\clazy\sigma}$}
+%% &
+%% \sc{if-clo} \\
+%% \hline
+%% {$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
+%%   (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
+%%   \cred{\sigma'}$} &
+%% \sc{return-whnf}\\
+%% \hline
+%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+%%   (\psi:\SRETURN~x), \cred{\sigma}$  $
+%%   \rightsquigarrow$
+%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, H,  s,
+%%   \cred{\sigma}$} &
+%% \sc{return-clo} \\
+%% \hline
+%% \end{tabular}
+%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
+%% \end{center}
+%% \vskip -5mm
+%% \end{figure*}
+
+We shall now present a proof of soundness of the analysis presented in
+Section~\ref{sec:liveness-analysis}.  \warning{While  the proof  is on
+  the broad lines of  \cite{asati14lgc}, there are several differences
+  due to the  lazy semantics of the language in  this paper.} Here are
+the ideas behind the proof.
 \begin{enumerate}
-\item While the liveness of \py\  should be killed at $\pi_6$ since it
-  is  beyond the scope  of the  definition of  \py, we  do not  do so.
-  However, this  does not cause any  imprecision in the way  of the GC
-  marking more cells as live,  since, during execution, \py\ is yet to
-  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
-  considered by the garbage collector at this point.
-\item In an  eager language, the heap location referred  by \py\ would
-  not have been live at $\pi_8$,  a program point beyond its last use.
-  In  a  lazy  language,   however,  the  definition  $\pz  \leftarrow
-  (1+\py)$, does not  result in an evaluation of  $(1+\py)$; instead a
-  closure  is  created.  The  evaluation  actually  takes place  while
-  evaluating $(\SRETURN~\pz)$.   We therefore  record the  liveness of
-  $\py$ at  the program point  $\pi_8$, indeed at every  program point
-  from $\pi_8$ up to the beginning of the program $\pi_1$ (since we do
-  not kill liveness).
+\item       We        enrich       the        standard       semantics
+  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
+  collection  just  before  each  $\rightsquigarrow$  transition.  The
+  simulated garbage collection starts from  the root-set and inserts a
+  special value  $\bot$ at each location  in the heap that  contains a
+  reference that is reachable but not live. Any attempt to dereference
+  such  locations during  the  transition will  result  in entering  a
+  special state denoted \bang. We call the semantics after
+  augmentation,  \emph{minefield semantics}.
+\item \label{inline} Assuming that a  program enters the \bang\ state,
+  we  show  how to  construct,  through  inline expansion,  a  program
+  without function calls which has the same minefield behavior.
+\item  The final  step   shows  that  no program  without
+  function calls can enter a \bang\ state. As a consequence no program
+  (with or without function calls) can enter the \bang\ state. 
 \end{enumerate}
 
-We thus  regard the  body of  the function as  an expression  tree and
-consider paths from  the root of this tree to  the nodes corresponding
-to the program points which trigger evaluation\footnote{Program points
-  which      trigger      evaluation       are      labeled      with
-  $\psi$.}---\SIF\ conditions and \SRETURN\ expressions.  We call such
-paths {\em e-paths}.  For example, for the function \length, there are
-three  such paths  $\epath_1 =  \langle \pi_1,  \pi_2, \psi_1\rangle$,
-$\epath_2  = \langle\pi_1,  \pi_2,  \pi_3,  \pi_4, \psi_2\rangle$  and
-$\epath_3  =  \langle  \pi_1,   \pi_2,  \pi_5,  \pi_6,  \pi_7,  \pi_8,
-\psi_3\rangle$.
-
-Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
-analysis  computes  two   maps:  $\mathcal{L}(e_f,  \sigma,  \Lfonly)$
-computes a  map that takes an  e-path in ${\mathit f}$  and returns the
-liveness environment  arising out of  all uses of variables  along the
-e-path.  A second map $\cal{P}_{\mathit  f}$ maps any program point in
-$\mathit f$  to the set  of e-paths sharing  the point.  If  a program
-point happens to fall on more  than one such e-path, then the liveness
-environment of  the program  point is the  variable-wise union  of the
-liveness  environments  of the  individual  e-paths.   In the  example
-program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  \epath_1, \epath_2,
-\epath_3\rbrace$, the  liveness at $\pi_1$  is ${\cal M}(\epath_1)  \cup {\cal
-  M}(\epath_2) \cup  {\cal M}(\epath_3)$,  where ${\cal M}  = \mathcal{L}(e_f,
-\sigma, \Lfonly)$.  In the  sequel we shall only define $\mathcal{L}$.
-The formulation  of $\cal{P}$ is easy  and we do not  elaborate it any
-further.
-
-Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
-The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
-evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
-  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
-evaluation path  terminating in  $\psi$.  The  map for  the expression
-$(\SIF~\psi:x~e_1~e_2)$ is a point-wise-union of the maps of $e_1$ and
-$e_2$. In addition, since the condition also triggers an evaluation of
-$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
-added  to the  union.   Also notice  a  consequence of  laziness: the
-entire  expression including  the condition  is not  evaluated if  the
-demand on it is $\emptyset$.  This  results in the empty map, which is
-denoted by $\emptyset_{\cal{M}}$.
-
- The liveness  of {\LET} has  an elegant formulation.   Recollect that
- our language permits lazy {\LET}s whose definitions can be recursive,
- i.e.  in the expression $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in
- $s$.   We  model  the  definition  $x  \leftarrow~s$  as  a  possibly
- recursive function (named $\mathit{f}$ in the rule) which captures an
- arbitrary number of ``unrollings'' of the definition\footnote{The $*$
-   in  the definition  of ${\mathit  f}$ represents  non-deterministic
-   choice.  During liveness analysis it  can be assumed to generate no
-   liveness.}.  We  pass to  the demand  transformer $\Lfonly_{\mathit
-   f}$ of this function the demand  arising out of the liveness of $x$
- at the beginning of $e$. This  gives the liveness of the variables of
- $s$.
-
-Section~\ref{sec:computing}   shows   how  the   demand   transformers
-$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
-analysis) can be safely approximated  by a {\em procedure summary} for
-each function.   The summary is  in the  form of a  demand transformer
-that maps a demand on a call to the function to demands on each of its
-arguments.
-
-\subsection{Correctness of liveness analysis}
-
-We shall now  outline a proof of correctness of  the liveness analysis
-presented   in    Section~\ref{sec:liveness-analysis}.    This   proof
-technique  is  similar to  the  technique  used in  \cite{asati14lgc}.
-
-The idea behind the proof is  as follows: We shall modify the standard
-semantics to a semantics  called {\em minefield semantics} that models
-a liveness based garbage collection  at each transition.  The state of
-the  transition system  is augmented  to carry  enough  information to
-calculate the liveness environment  at the current evaluation context,
-and  each   of  the   suspended  evaluation  contexts.    Before  each
-transition,  we start  from the  root-set and  insert a  special value
-$\bot$ at each  location in the heap that contains  a reference and is
-reachable but  not live by  our analysis.  Any attempt  to dereference
-such locations during the transition will result in entering a special
-state denoted \bang.  The main proof is to show that no program enters
-the \bang\ state in the minefield semantics.
-
-\newcommand{\cred}[1]{\psframebox[linestyle=none,
-    fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
-\section{Minefield semantics for the language}
-\begin{figure*}[ht!]
+\begin{figure*}[t!]
 \begin{center}\footnotesize
-\begin{tabular}{|c|c|c|}
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{|@{}c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|}
 \hline
@@
 \hline
-\hline
-          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
-  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
-    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
+
+\hline
+\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
+  H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
+\sc{car-clo-bang} 
 \\
 \hline
-          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
-    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
-  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
-\sc{car-clo}
-\\
-\hline
-$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
-  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
-    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
-\sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
-\sc{prim-1-clo} \\
-\hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
-\sc{prim-2-clo} \\
-\hline
-{$\mathit{g}~\mbox{defined as}$
-$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
-  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
-    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
-\sc{funcall} \\
-\hline
-$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e, \cred{\sigma, e_f}$} &
-\sc{let} \\
-\hline
-$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
-\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
-\cred{\clazy\sigma, e_f}$}
-&
-\sc{if-clo} \\
-\hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
-  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
-  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
-  \cred{\sigma', e_f'}$} &
-\sc{return-whnf}\\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
-  \cred{\sigma, e_f}$} &
-\sc{return-clo} \\
+\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x))
+\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
+  H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
+\sc{car-clo}\\
 \hline
 \end{tabular}
-\caption{Minefield semantics.\label{fig:minefield-semantics}}
 \end{center}
-\vskip -5mm
 \end{figure*}
@@
 \begin{enumerate}
-\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
-  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
-  on the expression $e$, and $e_f$ is the body of the function. As the
-  section on liveness  analysis shows, the entire  function body $e_f$
-  and $\sigma$  are needed  to calculate  the liveness  environment at
-  different program  points. The  demand sigma  is ``dynamic''  in the
-  sense  that it  arises from  the actual  function calls  made during
-  execution.   The  0-CFA  demand  is an  over-approximation  of  this
-  demand.  The information  in a suspended evaluation  context is also
-  similarly augmented with  the demand on it and the  function body of
-  which the suspended evaluation is a  part.  Thus a stacked entry now
-  takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
-  small step semantics which carries the extra information is shown in
-  Figure~\ref{fig:minefield-semantics}.
-\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
-  e_f$ and $\mathcal{LF}$, we can construct a liveness environment as follows:
+\item   We   enrich    the   abstract   machine   state
+  $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
+We  call such a state  a \emph{minefield state}.
+Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
+  $e$, i.e., it arises
+  from the  actual sequence of function  calls that led
+  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
+  %% during    liveness     analysis    is     a    static
+  %% over-approximation   of   $\sigma$   that   considers
+  %% \emph{all  possible calls}  that  could  lead to  the
+  %% evaluation of  $e$.
+  The  information in  a suspended
+  evaluation  context  on  the   stack  $S$  is  also
+  similarly augmented with its  demand.  Thus a stacked
+  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
+  %% modification  of  the   small  step  semantics  which
+  %% carries the extra information  is shown in the bottom
+  %% table of Figure~\ref{fig:minefield-semantics}.
+\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
+  collection that returns $(\rho', S', \heap')$, where the changes in
+  $\rho, S$ and  $\heap$ are due
+  to certain references being replaced by $\bot$.  This represents the
+  act of  not retaining  (effectively garbage collecting)  the objects
+  pointed to by these references  during an actual garbage collection.
+  This is done in the following way:
   \begin{enumerate}
-  \item If $e$ is an expression, then:  
-  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
-    \text{where }
-  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
-\item If $e$ is an application, then:  
-  $$ \Lv = ref(e, \sigma, \Lfonly)$$
-  \end{enumerate}
-We can define in the same way the liveness environment for each of the
-suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
-environments that we shall denote $\mathsf{SL}$.
-\item $GC$ models a liveness based garbage collection:
-$GC(\rho,$ $ S,$ $ H,$ $ \mathcal{L},$ $ \mathsf{SL}) = (\rho', S', H')$ 
-  \begin{enumerate}
-  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
-    \notin \Lv$.
-\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
-  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
-  \in domain(\rho)$,  $\rho'(x)=\bot$
-  iff $x.\epsilon \notin \Lv'$.
-\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
-  $x$ in either the current environment or one of the stacked
-  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
-  \ell$  and $x.\alpha \in \Lv$, .
+  \item Given the current  evaluation context $\rho, S,
+    \heap,   e,  \sigma$   and   the  demand   transformers
+    $\Lfonly$, we construct a liveness environment
+    \Lv.  %%  If  $e$  is  an expression,  then  $  \Lv  =
+    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
+    %% application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
+    We  similarly  define the  liveness environment
+    for each  of the  suspended evaluation  contexts in
+    $S$, giving  a stack of liveness  environments that
+    we shall denote $\mathsf{SL}$.
+  \item \emph{Garbage collection  of objects pointed by
+    root   set:}  For   each   $x  \in   domain(\rho)$,
+    $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
+    \Lv$.    Similarly,    for   each    stack    entry
+    $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
+    corresponding      liveness     environment      in
+    $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
+    $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
+  \item  \emph{Garbage  collection  of objects  in  the
+    heap:} \warning{ For each location  $\ell$, $\heap'(\ell) = \bot$
+    iff  there   is  no  $x$  in   either  the  current
+    environment  or  one  of the  stacked  environments
+    $\rho$, such  that for some  $\alpha$ corresponding
+    to a  \warning{closure-free path}  in the heap  starting from
+    $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
+    \Lv$}. Here \Lv\ is the liveness of the variables in
+    the current environment or the stacked environment.
   \end{enumerate}
 \end{enumerate}
-
-Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
-(\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
-preceded by  a garbage collection  using $GC$.  We consider  a garbage
-collection followed  by a transition as  a step. We show  by induction
-that, for programs  without function calls, starting  from the initial
-state that  has $e_{\main}$  as the first  expression to  be evaluated
-instead  of $\main$,  any transition  of  $n$ steps  occurs without  a
-\bang.
-
-Observe  that  the  transitions  that dereference  and  therefore  can
-potentially lead to \bang\ are those  which have a $H(\ldots)$ term in
-the  transition  rule.   Of  these,  terms  of  the  form  $H(\ldots)$
-dereference a root  variable, and the term $H(l)$  dereferences a heap
-location.  In  both the  cases we show  that the  dereferencing cannot
-lead to a \bang\ state in programs without function calls.
-
-We  need the  following proposition that  connects  liveness analysis  with
-minefield semantics transitions. 
-\begin{proposition}
-\label{prop:correctness}
-\begin{enumerate} 
-\item   Consider  a   \LET\  definition:   \linebreak  $\pi\!\!\!:   (\LET~x
-  \leftarrow~\ldots~\IN~e)$, such  that there is  a use of $x$  in $e$,
-  i.e.  an occurrence of $\LET~y~\leftarrow~s~in \ldots$, where $s$ is
-  $(\CAR\ x)$, $(\CDR\ x)$  or $(+\ x\ y)$, or $(\SIF\  x\ e1\ e2)$ or
-  $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null.
-\item    Assume    that    a   \LET\    definition    $\pi:    (\LET~z
-  \leftarrow~(\CONS~x~y)~\IN~e)$  has an  occurrence  of $(\CAR~z)$  in
-  $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
-  Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of $y$  at
-  $\pi$ includes $\epsilon$.
-\end{enumerate}
-\end{proposition}
-
-In  the second  case for  example,  the liveness  of $x$  due to  the
-\CONS\ is  $\bcar\sigma$, where $\sigma$  is the demand on  $z$. Since
-$\sigma$  includes  the  demand  $\acar\alpha$   due  to  its  use  in
-$(\CAR~z)$, the liveness on $x$  includes $\epsilon$. We thus have the
-following lemma:
-
-\begin{lemma}\label{lemma:no-calls-correctness} 
-A program without function calls cannot  enter the \bang\ state in the
-minefield semantics.
+Starting from the initial state,
+%%  $([\;]_\rho,([\;]_\rho,
+%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
+every transition  is preceded  by a  garbage collection
+using $GC$.  The details of the transition for the {\sc
+  car-clo}  rule   is  shown   in  the  top   table  of
+Figure~\ref{fig:minefield-semantics}.  If  this  or  an
+earlier  call to  $GC$  results  in $H(\rho(x))$  being
+bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
+enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
+the transition is the same as the earlier {\sc car-clo}
+rule.
+
+Consider a trace  of a minefield execution of a  program $p$, possibly
+ending in  a \bang\ state. We  can construct a call-tree  based on the
+trace in  which each node represents  a function that was  called (but
+did not necessarily  return because of a \bang).  Assume  that each of
+the nodes of  the tree is also annotated with  the program point where
+the corresponding call  was invoked.  This tree can be  used to inline
+function calls in a hierarchical fashion.  The details of the inlining
+can be found in~\cite{asati14lgc}.
+%% The details of inlining a single
+%% function      call     $(\LET\,\,      x     \leftarrow
+%% (f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
+%% Let  $f$ be  defined  (after renaming  its formals  and
+%% locals  to  be  disjoint from  existing  variables)  by
+%% $(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
+%% replaced  by   a  sequence  of  $\LET$s   of  the  form
+%% $z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
+%% but with its  $(\SRETURN\,\,w)$ expressions replaced by
+%% $(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
+%% We prefer to use the  identity function $\ID$ as a form
+%% of no-op rather than  introducing the form $(\LET\,\, x
+%% \leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
+%% $\LET$-definition   is  a   simple  variable. 
+For  a
+call-less program,  the initial state of  the minefield
+semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
+\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{H},
+e_{\mainpgm}$.
+
+\subsection{Soundness Results}
+
+  %% We  consider  a  round of garbage  collection  followed  by  a
+  %% transition as a step.  We show by induction that, for
+  %% programs  without function  calls, %% starting  from the
+  %% %% initial  state  that has $e_{\main}$  as  the  
+  %% %% expression to  be evaluated,
+  %% any
+  %% transition of $n$ steps occurs without a \bang.
+
+We  first  need  an auxiliary  result  about  minefield
+semantics. Consider  a trace of a  minefield execution.
+For every minefield state  $(\rho, S, H, e(s), \sigma)$
+that appears  on the LHS of  a $\rightsquigarrow$ step,
+the  demand   $\sigma$  on   the  expression   $e$  (or
+application $s$) is non-null.  This can be proved by an
+induction  on  the  number  of  steps  leading  to  the
+minefield  state.   The  base step  holds  because  the
+demand $\sigma_{all}$  on $(\mainpgm)$ is  non-null. For the 
+inductive step  we observe   that for
+each  step of  the minefield  semantics, if  the demand
+$\sigma$ on  the LHS of  a minefield step  is non-null,
+the demand on  the RHS is a  transformation of $\sigma$
+(for example $\clazy\sigma$) which is also non-null.
+
+%% Let  us clarify  this  observation: Since  the minefield  semantics
+%% models lazy evaluation,  parts of the program may  not be evaluated
+%% at all.  However,  if an expression (or application)  happens to be
+%% evaluated, it will  appear on the LHS of a  minefield step, and the
+%% context-sensitive demand on it will be non-null.
+
+Note that our proofs will be for a single round of minefield execution
+i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
+programs) to  its WHNF driven  by the printing mechanism.   With minor
+variations,  the   proof  will  also  be   applicable  for  subsequent
+evaluations initiated by $\print$.
+%% We  show next that  the minefield execution of  a call-less
+%% program cannot go \bang.
+
+\begin{lemma}
+\label{lemma:call-less-cannot-go-bang}
+Consider the minefield execution of  a program without function calls.
+Such a program cannot enter the \bang\ state.
 \end{lemma}
-
 \begin{proof}
-For a program  executing according to {\em minefield  semantics}, if a
-\bang\ occurs then some variable which is  dereferenced must be
-set to  $\bot$, between its  creation and  use points.  This  can only
-happen if  the liveness of  the variable being dereferenced  was null.
-From  the Proposition~\ref{prop:correctness}  it  is clear  that this  cannot
-happen.
+Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
+step. Consider  a state  $(\rho, S,  H, e,  \sigma)$ in  the minefield
+execution of a  program.  We show by induction on  the number of steps
+$n$ leading  to this  state that  the next  transition cannot  enter a
+\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\_, [\;]_{S} ,
+            [\;]_{H}, e_{\mainpgm}, \sigma_{all})$.  Since the call to
+            $GC(\ldots)$ in this state does  nothing, we just have to show
+            that  the  $\rightsquigarrow$  transition cannot  enter  a
+            \bang\   state.    Since   our  programs   are   in   ANF,
+            $e_{\mainpgm}$ can  only be a $\LET$  expression.  A
+                     {\sc let} step does not involve dereferencing,
+                     and thus
+                     cannot result in a \bang.
+
+For the inductive step, we  shall show that none of the
+minefield steps that  involves dereferencing results in
+a   \bang.   These  are   the   steps   which  have   a
+$H(\rho(...))$ in the premise. %%  Recall that a minefield
+%% step   consists   of   a   GC  step   followed   by   a
+%% $\rightsquigarrow$. 
+Now a  $\rightsquigarrow$ step can  go
+\bang\ because it dereferences a $\bot$ inserted by the
+immediately  preceding  GC  or  the GC  of  an  earlier
+minefield step.  However the  demand $\sigma'$
+  on basis of which an earlier GC would have inserted a
+  $\bot$,  would  have   included  the  current  demand
+  $\sigma$.  Thus,  if the  earlier GC had  inserted a
+$\bot$, so would the current  GC.  Thus it is enough to
+consider the $\rightsquigarrow$  along with the current
+GC and show that it does not lead to a \bang.
+
+We consider the rules {\sc car-clo} and {\sc car-1-clo} only.  
+The rest  of the rules involve similar reasoning. For the {\sc
+  car-clo}  rule in  the state  $\rho, S,  H, (\CAR~x),
+\sigma$, we  know that $\sigma$ is  non-null. Therefore
+the  liveness  of  $x$  includes  $\epsilon$,  and  the
+dereferencing $H(\rho(x))$ will go without \bang.
+
+For the  {\sc car-1-clo}  rule, observe that  there are
+two dereferences.  First $x$  is dereferenced to  get a
+cons  cell  and then  the  head  of  the cons  cell  is
+dereferenced  to  obtain  a  closure.   If  the  demand
+$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
+of    $x$   will    include    both   $\epsilon$    and
+$\acar\epsilon$,  and  a  GC with  this  liveness  will
+neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
+first   component  of   the   cons   cell.  Thus   both
+dereferences    will    happen   normally    and    the
+$\rightsquigarrow$ step will not enter \bang.
 \end{proof}
 
-For   a    program   with   functions,   we    use   the   techniques
-in~\cite{asati14lgc}. We  assume as  a contradiction that  the program
-can enter a \bang\ state  and inline the  functions called to  obtain a
-program that is without functions  but has the same execution behavior
-and    demands    as   the    original    program.     We   then    use
-Lemma~\ref{lemma:no-calls-correctness}  to  show  that  the  resulting
-functionless program cannot enter a bang state.
+Now we are ready to prove the main soundness result.
+
+\begin{theorem}
+The  minefield  execution of  no  program  can enter  a
+\bang\ state.
+\end{theorem}
+
+\begin{proof}
+Assume to  the contrary that  a program $P$  enters the
+\bang\  state.  \warning{We can  transform  $P$  to a  call-less
+program $P'$ such that  the minefield executions of $P$
+and $P'$  are identical  except for change  of variable
+names. }             However,                 by
+Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
+$P'$,   a   call-less   program,   cannot   enter   the
+\bang\  state.  Therefor  $P$  also  cannot  enter  the
+\bang\ state.
+\end{proof}
+
+%% Section~\ref{sec:computing}   shows   how  the   demand   transformers
+%% $\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
+%% analysis) can be safely approximated  by a {\em procedure summary} for
+%% each function.   The summary is  in the  form of a  demand transformer
+%% that maps a demand on a call to the function to demands on each of its
+%% arguments.
 
 \section{Computing liveness and its encoding as DFA}\label{sec:computing}
-Section~\ref{sec:liveness} gave a  context-sensitive liveness analysis
-and proved it  correct with reference to a  {\em minefield} semantics.
-In  our  analysis, $\mathcal{L}$  and  \pmap{f}  together described  a
-liveness set for each program point in  a function in terms of a given
-$\sigma$  and \Lfonly.   However, we  still  have to  describe how  to
-obtain demand  transformers \Lfonly\  from the rule  {\sc live-define}
-and how to compute the specific demand $\sigma$ on each function.
-%%
-To  answer these  questions, it  is convenient  to cast  the equations
-arising from  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
-grammar.   To do  so, we  need  to modify  the rules  themselves to  a
-different form.
+Section~\ref{sec:liveness}  gave   a  fully context-sensitive
+liveness analysis which 
+%$\mathcal{L}$ and $ref$ together
+described the  liveness set for  each variable  in a function  body in
+terms of  a symbolic demand  $\sigma$ and \Lfonly.  However,  we are yet
+to describe  (i) how to obtain demand  transformers \Lfonly\ from
+the  rule {\sc  live-define}, and  (ii)  how to  compute the  concrete
+demand $\sigma$  on each function.   To do so,  we first need to  modify the
+liveness rules to a different form.
 
@@
 The   $\mathit{ref}$   rule   for   \CONS,   shown   in
-Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
-\acar\ and \acdr\  from the access paths in  $\sigma$.  Similarly, the
-rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
-$\emptyset$, if  $\sigma$ itself is $\emptyset$ and
-$\lbrace\epsilon\rbrace$ otherwise.  To  realize these rules
-$\sigma$ needs to be known. This creates difficulties since we want to
-solve the equations arising from liveness symbolically.
-
-The  solution  is  to  also   treat  the  operations  mentioned  above
-symbolically.  We  introduce three new symbols:  \bcar, \bcdr, \clazy.
-These symbols are defined as a relation $\hookrightarrow$ between sets
-of access paths:
+Figure~\ref{fig:live-judge}, requires us  to remove the
+leading  \acar\ and  \acdr\  from the  access paths  in
+$\sigma$.  Similarly, the rules  for \CAR, \CDR, \PRIM,
+\NULLQ, and \SIF\ require  us to return $\emptyset$, if
+$\sigma$      itself       is      $\emptyset$      and
+$\lbrace\epsilon\rbrace$  otherwise.  To  realize these
+rules  $\sigma$   needs  to  be  known.   This  creates
+difficulties  since  we  want to  solve  the  equations
+arising from liveness symbolically.
+
+The solution is to  also treat the operations mentioned
+above  symbolically.  We  introduce three  new symbols:
+\bcar, \bcdr,  \clazy.  These symbols are  defined as a
+relation  $\hookrightarrow$  between   sets  of  access
+paths:
 \begin{align*}
@@
 \mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
-                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
-        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
-        &[\,\mathit{\epath} \mapsto  x.\clazy \sigma]
+                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\cup
+        \mathcal{L}(e_2,\sigma,\Lfonly)~\cup
+          \{x.\clazy \sigma\}
 \end{align*}
-The rules for  \CDR, \PRIM\ and \NULLQ\ are also  modified in a manner
-similar to \CAR.
-
-When  there  are  multiple  occurrences of  \bcar,  \bcdr\  and  \clazy,
-$\hookrightarrow$  is  applied  from  right to  left.   The  reflexive
-transitive   closure  of   $\hookrightarrow$   will   be  denoted   as
-$\stackrel{*}{\hookrightarrow}$.   The  following proposition  relates
-the original and the modified liveness rules.
+The rules for  \CDR, \PRIM\ and \NULLQ\ are also
+modified similarly.
+
+\warning{When   there  are   multiple   occurrences  of   \bcar,
+\bcdr\  and \clazy,  $\hookrightarrow$ is  applied from
+right  to left.}   The reflexive  transitive closure  of
+$\hookrightarrow$      will      be     denoted      as
+$\stackrel{*}{\hookrightarrow}$.      The     following
+proposition  relates  the  original  and  the  modified
+liveness rules. 
 \begin{proposition}
-Assume that a liveness computation  based on the original set of rules
-gives the liveness  of the variable $x$ at a  program point $\pi_i$ as
-$\sigma$   (symbolically,   $\Lanv{i}{x}=   \sigma$).   Further,   let
-$\Lanv{i}{x}=  \sigma'$ when the  modified rules  are used  instead of
-\Lfunonly.  Then $\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
+Assume  that  a  liveness   computation  based  on  the
+original  set  of  rules  gives  the  liveness  of  the
+variable  $x$   as $\sigma$
+(symbolically,  $\Lanv{x}{}=  \sigma$).  Further,  let
+$\Lanv{x}{}= \sigma'$ when the modified rules are used
+instead      of      \Lfunonly.      Then      $\sigma'
+\stackrel{*}{\hookrightarrow} \sigma$.
 \end{proposition}
 
-For  an explanation  of why  the  proposition holds  for the  modified
-\CONS\  rule,   we  refer   the  reader  to   \cite{asati14lgc}.   The
-proposition also holds for other modified rules for similar reasons.
+An  explanation of  why the  proposition holds  for the
+modified  \CONS\ rule  is  given in  \cite{asati14lgc}.
+The proposition also holds for other modified rules for
+similar reasons.
 
 \newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
-\subsection{Generating Liveness Equations}
-
-\begin{figure*}[t!]
-\renewcommand{\arraystretch}{1}
-  \begin{tabular}{cc}
-    \begin{minipage}{.50\textwidth}
-        \small
-        \renewcommand{\arraystretch}{1}{
-          \begin{uprogram}
-            \UNL{1} (\DEFINE\ (\fun~\px\ \pn)
-            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \py\ $\leftarrow $\
-(\length~\px) \IN
-            \UNL{3} $\pi_2\!\!:\, $(\LET\ \pl\ $\leftarrow $\
-(\makelist~\pn) \IN
-            \UNL{4} \hspace*{.05cm} $\pi_3\!\!:\,$(\SIF\
-$\psi_1\!\!:\,$ \py
-            \UNL{5} \hspace*{.27cm} $\pi_4\!\!:\, $(\LET\ \xhd\ $\leftarrow$\ (\CAR\ \pl) \IN
-            \UNL{6} \hspace*{.27cm} $\pi_5\!\!:\, $(\LET\ \xtl\ $\leftarrow$\ (\CDR\ \px) \IN
-            \UNL{7} \hspace*{.27cm} $\pi_6\!\!:\, $(\LET\ \rest\ $\leftarrow$\ (\fun~ \xtl\ \xhd) \IN
-            \UNL{8} \hspace*{.27cm} $\pi_7\!\!:\, $(\LET\ \ret\ $\leftarrow$\ (\CONS\ \py\ \rest) \IN
-            \UNL{9} \hspace*{.32cm} $\pi_8\!\!:\,
-(\SRETURN~\psi_2:\ret)))))$
-            \UNL{5} \hspace*{.29cm}    $\pi_9\!\!:\, $(\LET~\ps\
-$\leftarrow$  (\Sum~\pl)  \IN
-            \UNL{6} \hspace*{.34cm}   $\pi_{10}\!\!:\, $(\LET~\pu\
-$\leftarrow$  (+~1~\py)  \IN
-            \UNL{7} \hspace*{.34cm} $\pi_{11}\!\!:\,
-            $(\LET~\pr\ $\leftarrow$ ($/$~\ps~\pu) \IN
-            \UNL{8} \hspace*{.27cm} $\pi_{12}\!\!:\, $(\LET\ \res\ $\leftarrow$\ (\CONS\ \pr\ \NIL) \IN
-            \UNL{9} \hspace*{.34cm} $\pi_{13}\!\!:\,
-(\SRETURN~\psi_3:\res)$)))))))))
-        \end{uprogram}}
-        \renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-	  \UNL{1} $(\DEFINE\ (\main)$
-           \UNL{2} \!\!$\pi_{14}\!\!:\, (\LET\  \pa\  \leftarrow$
-           (\CONS\ $1$\ (\CONS\ $2$\ \NIL) \IN  
-          \UNL{3} \!\!$\pi_{15}\!\!:\, (\LET\  \pb\  \leftarrow $5$)$
-\IN
-	  \UNL{4}   \hspace*{.05cm}$\pi_{16}\!\!:\,      $ (\LET\ \pc\
-$\leftarrow  (\fun\ \pa\ \pb)$ \IN
-          \UNL{5}   \hspace*{.15cm}    $\pi_{17}\!\!:\,
-          (\SRETURN~\psi_4:\pc)))))$
-\end{uprogram}}
-        %%}
-    \end{minipage}
-
-    &
-
-    \begin{minipage}{.48\textwidth}
-
-      \small
-\begin{eqnarray*}
-        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
-              [\;\epath_1 \mapsto \lbrace \px.\clazy \sigma,
-\xl.\clazy\sigma\rbrace ,~\epath_2 \mapsto \lbrace 
-\pv.\sigma  \rbrace,\\
- &&          \;\;  \epath_3 \mapsto \lbrace \pz.\sigma,
-                \py.\clazy\sigma,
-\pu.\Lf{\length}{\mbox{1}}{\clazy\sigma},\\
-&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
-\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\clazy\sigma} \cup
-%%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
-\clazy\Lf{\length}{1}{\clazy\sigma}\}\}
-             \rbrack \\
-            \cal{P_{\mbox{\tiny \length}}} &=&
-                [\;\pi_1 \mapsto \lbrace \epath_1, \epath_2, \epath_3 \rbrace,  \
-                 \pi_2 \mapsto \lbrace \epath_1, \epath_2, \epath_3
-                  \rbrace, \\
-&&\;\;  \pi_3 \mapsto \lbrace \epath_2
-                  \rbrace ,  \pi_4 \mapsto \lbrace \epath_2
-                  \rbrace, \pi_5 \mapsto \lbrace  \epath_3 \rbrace, \\
- &&  \;\;                        \pi_6 \mapsto \lbrace \epath_3 \rbrace,
-\pi_7 \mapsto
-                   \lbrace \epath_3\rbrace, \pi_8 \mapsto \lbrace \epath_3
-\rbrace~\rbrack\\
-              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
-                [\;
-                  \epath_4 \mapsto \lbrace \pw.\sigma,
-                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},\\
-&& \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
-\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},
-\pa.\clazy\bcar\Lf{\length}{\mbox{1}}{\sigma} \rbrace
-                               ] \\
-               \cal{P_{\mbox{\tiny \main}}} &=&
-                  \lbrack~ \pi_9 \mapsto \lbrace \epath_4 \rbrace,  \
-                    \pi_{10} \mapsto \lbrace \epath_4 \rbrace,  \
-                   \pi_{11} \mapsto \lbrace \epath_4 \rbrace, \\
-                   && \;\; \pi_{12} \mapsto \lbrace \epath_4 \rbrace,
-\pi_{13} \mapsto \lbrace \epath_4 \rbrace~\rbrack
-\end{eqnarray*}
-    \end{minipage} \\
-(a)&
-(b)
-    %% \end{ tabular}}
-  \end{tabular}
-%\end{picture}
-\kern -3ex
-%\vspace{0.5in}
-\caption{(a) Example program and (b) its liveness
-maps.}\label{fig:mot-example2}
-\end{figure*}
-%================================================================
-
-Given  a  function  $\mathit{f}$,  we now  describe  how  to  generate
-equations  for the  demand  transformation \Lfonly$_\mathit{f}$.   The
-program in Figure~\ref{fig:mot-example2} serves  as a running example.
-Starting   with    a   symbolic   demand   $\sigma$,    we   determine
-\Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In  particular, we  consider
-$\Lanv{1}{x_i}$,  the liveness  of  the $i^{\mbox{\footnotesize  th}}$
-parameter   $x_i$  at   the  program   point  at   the  beginning   of
-$e_{\mathit{f}}$   (assumed  to   be  $\pi_1$).   By  the   rule  {\sc
+\subsection{Generating Equations for \Lfonly$_\mathit{f}$}
+Given a  function $\mathit{f}$, we now  describe how to
+generate  equations   for  the   demand  transformation
+\Lfonly$_\mathit{f}$.        The       program       in
+Figure~\ref{fig:mot-example2}   serves  as   a  running
+example.  Starting with a  symbolic demand $\sigma$, we
+determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
+particular, we  consider $\Lanv{1}{x_i}$,  the liveness
+of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$ at
+the program point at  the beginning of $e_{\mathit{f}}$
+(assumed   to   be   $\pi_1$).   By   the   rule   {\sc
   live-define},   this   should    be   the   same   as
@@
 \end{eqnarray*}
+%% and  the only equation defining \Lfone{\length} is:
+%%   \begin{eqnarray*}
+%%    && \Lf{\length}{1}{\sigma}
+%%     =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
+%% \clazy\Lf{\length}{1}{\clazy\sigma}
+%% \end{eqnarray*}
 and  the equation defining \Lfone{\fun} is:
@@
 
-In general, the equations for  \Lfonly\ are recursive since, as in the
-case  of  \fun,  \Lan{1}\  may   have  been  defined  in  terms  of
-\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
-solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
-each  of the liveness  rules modifies  a demand  by prefixing  it with
-symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \clazy
-\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
-closed form:
+In general, the equations for \Lfonly\ are recursive as
+in the  case of  \Lfonly$_\mathit{f}$. As  mentioned in
+\cite{asati14lgc},   a   closed   form   solution   for
+$\Lfone{\mathit{f}}$ can  be derived by  observing that
+each of  the liveness rules  modifies a demand  only by
+prefixing  it with  symbols  in  the alphabet  $\lbrace
+\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore
+we can  assume that $\Lf{f}{i}{\sigma}$ has  the closed
+form:
 \begin{eqnarray}
@@
 \end{eqnarray}
-where \Df{f}{i} are sets of strings over the alphabet mentioned above.
-Substituting   the   guessed   form   in   the   equation   describing
-\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
-for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
-\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
-\Lfonly$_{\mathit \fun}$, we get:
+where \Df{f}{i}  are sets of strings  over the alphabet
+mentioned above.  Substituting the  guessed form in the
+equation   describing    \Lfonly$_{\mathit   f}$,   and
+factoring  out   $\sigma$,  we  get  an   equation  for
+\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
+solution   for   \Df{f}{i}   yields  a   solution   for
+\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
+  \length}$, we get:
   \begin{eqnarray*}
@@
 
-To avoid analyzing the body of a function for each call, we calculated
-the liveness  for the arguments and  the variables in  a function with
-respect to a symbolic demand  $\sigma$.  To get the actual liveness we
-calculate an over-approximation of the  actual demands made by all the
-calls and calculate the liveness at each GC
-point inside the  function based on this approximation.
-The  0-CFA-style  {\em
-  summary demand}  is calculated by taking  a union of  the demands at
-every call site of a function.
-
-Consider a function $g$ containing a  call to $f$ at a site $\pi$, say
-$\pi\!\!:\!(\LET~x   \leftarrow   (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.
-Let the  demand on $g$  be $\sigma_g$ and,  based on this  demand, the
-liveness of  $x$ at $\pi_i$ be  $\Lanv{i}{x}$.  By the  $\LET$ rule of
-Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
-$\Lanv{i}{x}$ to the demand $\sigma_f$. Let us denote the contribution
-of a call  site $\pi$ in a  function $g$ to the overall  demand on the
-function $f$  as $\deltacall{f}{\pi}{g}$. Assuming that  there are $k$
-call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
-(in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
-$\deltacall{f}{\pi^1}{g^1}          \cup          \cdots          \cup
-\deltacall{f}{\pi^k}{g^k}$.  The distinguished function \mainpgm\ is a
-special case.   We assume  it is called  through a  printing mechanism
-with   demand   $\sigma_\mainpgm   =  \{\acar,\acdr\}^\ast$ (denoted
-$\sigma_{\!all}$) if \mainpgm\ returns a structure and $\epsilon$ if
-it returns a base value.
+To  avoid analyzing  the body  of a  function for  each
+call, we calculated the  liveness for the arguments and
+the variables in a function  with respect to a symbolic
+demand  $\sigma$.   To  get   the  actual  liveness  we
+calculate an  over-approximation of the  actual demands
+made by  all the  calls and  calculate the  liveness at
+each  GC  point  inside  the  function  based  on  this
+approximation.  The 0-CFA-style {\em summary demand} is
+calculated by  taking a union  of the demands  at every
+call site of a function.
+
+%% Consider a function  $g$ containing a call to  $f$ at a
+%% site    $\pi$,   say    $\pi\!\!:\!(\LET~x   \leftarrow
+%% (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.  Let the demand on
+%% $g$  be  $\sigma_g$  and,  based on  this  demand,  the
+%% liveness of  $x$ at  $\pi_i$ be $\Lanv{i}{x}$.   By the
+%% $\LET$ rule of Figure~\ref{fig:live-judge}, the call at
+%% $\pi$   contributes   $\Lanv{i}{x}$   to   the   demand
+%% $\sigma_f$. Let  us denote  the contribution of  a call
+%% site $\pi$ in  a function $g$ to the  overall demand on
+%% the function  $f$ as  $\deltacall{f}{\pi}{g}$. Assuming
+%% that there are $k$ call  sites to function $f$, $\pi^1$
+%% (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
+%% the     over-approximation     of     $\sigma_f$     is
+%% $\deltacall{f}{\pi^1}{g^1}     \cup     \cdots     \cup
+%% \deltacall{f}{\pi^k}{g^k}$.  The distinguished function
+%% \mainpgm\ is  a special case.   We assume it  is called
+%% through    a    printing    mechanism    with    demand
+%% $\sigma_\mainpgm   =   \{\acar,\acdr\}^\ast$   (denoted
+%% $\sigma_{\!all}$) if \mainpgm\  returns a structure and
+%% $\epsilon$ if it returns a base value.
 
@@
 $\pi_{16}$ and a recursive call at $\pi_6$.
-So $\sigma_{\length} =
+So $\sigma_{\fun} =
      \deltacall{\fun}{\pi_{16}}{\main}  \cup
-\deltacall{\fun}{\pi_6}{\length}$.
+\deltacall{\fun}{\pi_6}{\fun}$.
+
+%% For  the  running  example,   $\length$  has  calls  from  $\main$  at
+%% $\pi_{12}$ and a recursive call at $\pi_6$.
+%% So $\sigma_{\length} =
+%%      \deltacall{\length}{\pi_{12}}{\main}  \cup
+%% \deltacall{\length}{\pi_6}{\length}$.
 Filling in  the values gives:
@@
 
-In summary, the equations generated during liveness analysis are:
+In  summary, the  equations  generated during  liveness
+analysis are:
 \begin{enumerate}
-\item For each function $\mathit{f}$, equations defining \Df{f}{i} for
-  use by \Lfonly$_{\mathit f}$.
+\item   For  each   function  $\mathit{f}$,   equations
+  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
 \item For each function $\mathit{f}$, an equation defining the summary
   demand $\sigma_{\mathit f}$ on $e_f$.
-\item  For  each  function   $\mathit{f}$  (including  $\mainpgm$  for
-  $e_\mainpgm$)  an equation  defining liveness  at each  GC point  of
-  $e_{\mathit f}$.
+\item   For  each   function  $\mathit{f}$   (including
+  $\mainpgm$  for  $e_\mainpgm$) an  equation  defining
+  liveness at each GC point of $e_{\mathit f}$.
 \end{enumerate}
 \subsection{Solving liveness equations---the grammar
-interpretation}\label{sec:grammar-formulation}
-The  equations above can now be re-interpreted as a
-context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
-\bcar, \bcdr, \clazy\rbrace$.  Let  \var{$X$} denote the non-terminal
-for
-a variable  $X$ occurring on  the LHS of the  equations generated
-from the analysis.  We can  think of the resulting productions as
-being associated with several  grammars, one for each non-terminal
-\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
-grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
+interpretation}\label{sec:grammar-formulation}      The
+equations  above   can  now  be  re-interpreted   as  a
+context-free    grammar   (CFG)    on   the    alphabet
+$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
+Let \var{$X$}  denote the  non-terminal for  a variable
+$X$  occurring on  the LHS  of the  equations generated
+from  the  analysis.  We  can  think  of the  resulting
+productions as being  associated with several grammars,
+one for each non-terminal \var{\Lanv{i}{x}} regarded as
+a  start  symbol.   As  an  example,  the  grammar  for
+\var{\Lanv{1}{\xl}}     comprises     the     following
 productions:
 \begin{eqnarray*}
-  \var{\Lanv{1}{\xl}}  &\rightarrow& 
-  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
+  \var{\Lanv{3}{\xl}}  &\rightarrow& 
+  \clazy \mid \var{\Df {\Sum}{1}}{\clazy\var{\sigma_{\length}}}  \\
   & & \mid
-  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
+  \clazy\var{\Df{\length}{1}}\clazy \mid \acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
   \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
@@
 &\rightarrow&
-\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
-\var{\Lanv{9}{\pa}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
+\clazy  \mid \bcar\langle{\sigma_{\Sum}}\rangle \\
+ \var{\Df{\fun}{2}} &\rightarrow& \var{\Df{\makelist}{1}}\clazy \mid \var{\Df{\makelist}{1}}\var{\Df{\Sum}{1}}\clazy \\ 
+& & \mid \var{\Df{\makelist}{1}}\acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
+\langle\sigma_{\mathit{\fun}} \rangle &\rightarrow& \sigma_{all} \mid \bcdr\var{\sigma_{\mathit{\fun}}} \\
+\var{\Df{\Sum}{1}} &\rightarrow& \clazy \mid \acar\clazy \mid \acdr\var{\Df{\Sum}{1}}\clazy
+%% \var{\Lanv{3}{\py}} &\rightarrow& \clazy \mid \bcar\var{\sigma_{\mathit{\fun}}}
 \end{eqnarray*}
-Other equations can be converted similarly.  The language generated by
-\var{\Lanv{i}{x}},  denoted  $\mathscr{L}(\var{\Lanv{i}{x}})$, is  the
-desired solution  of \Lanv{i}{x}.  However,  note that the  strings in
-the  language are  over alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,
-\clazy \rbrace$, but  we want {\em forward} access paths  only, i. e.,
-access paths over the alphabet $\lbrace \acar, \acdr \rbrace$. In
-other words,  the decision  problem that we  are interested  in during
-garbage collection at a program point $\pi_i$ is:
+Other  equations  can   be  converted  similarly.   The
+language   generated   by  \var{\Lanv{i}{x}},   denoted
+$\mathscr{L}(\var{\Lanv{i}{x}})$,   is    the   desired
+solution  of  \Lanv{i}{x}.    However,  note  that  the
+strings  in the  language  are  over alphabet  $\lbrace
+\acar, \acdr,\bcar, \bcdr, \clazy \rbrace$, but we want
+     {\em  forward} access  paths only,  i. e.,  access
+     paths  over  the  alphabet $\lbrace  \acar,  \acdr
+     \rbrace$.  In other  words,  the decision  problem
+     that   we  are   interested   in  during   garbage
+     collection at a program point $\pi_i$ is:
 \begin{quote}
-Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
-\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \clazy).     Let
-$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
-where $\sigma$ consists  of forward access paths only. Then  does $\alpha \in
-\sigma$?
+Let $x.\alpha$ be a forward access path consisting only
+of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
+\clazy).       Let      $\mathscr{L}(\var{\Lanv{i}{x}})
+\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
+consists  of  forward  access  paths  only.  Then  does
+$\alpha \in \sigma$?
 \end{quote}
 
-We now model the above problem  as one of deciding the membership of a
-context-free  grammar  augmented  with  a fixed  set  of  unrestricted
-productions.
+We now model  the above problem as one  of deciding the
+membership of  a context-free grammar augmented  with a
+fixed set of unrestricted productions.
 
 \begin{definition}\label{def:specialgrammar}
-Consider the  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
-non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \clazy, \$\}$, $p_1$
-is a set of context-free productions that contains the distinguished
-production $S \rightarrow \alpha\$$, $\alpha$ is a string of grammar
-symbols that does not contain $S$, and $p_2$ is the fixed set of
-unrestricted productions $\bcar\acar \rightarrow \epsilon$,
-$\bcdr\acdr \rightarrow \epsilon$, $\clazy\acar \rightarrow \clazy$,
-$\clazy\acdr \rightarrow \clazy$, and $\clazy\$ \rightarrow
-\epsilon$.
+Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
+$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
+\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
+context-free    productions     that    contains    the
+distinguished  production   $S  \rightarrow  \alpha\$$,
+$\alpha$ is a  string of grammar symbols  that does not
+contain $S$, and $p_2$ is the fixed set of unrestricted
+productions    $\bcar\acar    \rightarrow    \epsilon$,
+$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
+\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
+and $\clazy\$ \rightarrow \epsilon$.
 \end{definition}
@@
 \end{lemma}
+
 \begin{proof}
-Given a  Turing machine  and an input  $w\in (1+0)^*$, we  construct a
-grammar $G$  such that  the machine will  halt on  $w$ if and  only if
-$\epsilon \in  \mathscr{L}(G)$. The grammar includes the  fixed set of
-unrestricted productions  in Definition~\ref{def:specialgrammar}.
-
-We  shall   denote  the  configuration   of  the  Turing   machine  as
-$w_l(\state,c)w_r$, where $w_l$ is the string to the left of the head,
-$w_r$ is the string to the right, $c$ is the symbol under the head and
-\state\ is the current state  of the machine.  For each combination of
-state  and   symbol  $(\state,c)$,   the  grammar  will   contain  the
-non-terminal \nont{}{c}. We shall synchronize each move of the machine
-to a  derivation step  using a context  free production,  followed, if
-possible, by  a derivation step  using either $\bcar\acar    \rightarrow
-\epsilon$ or $\bcdr\acdr    \rightarrow
-\epsilon$.
-After each synchronization, we shall establish the following invariant
-relation between the machine configuration and the  sentential form:
+Given a Turing machine and  an input $w\in (1+0)^*$, we
+construct a grammar $G$ such that the machine will halt
+on    $w$    if    and   only    if    $\epsilon    \in
+\mathscr{L}(G)$. The grammar includes  the fixed set of
+unrestricted               productions               in
+Definition~\ref{def:specialgrammar}.
+
+We shall denote the configuration of the Turing machine
+as $w_l(\state,c)w_r$, where $w_l$ is the string to the
+left of the head, $w_r$ is the string to the right, $c$
+is the symbol under the head and \state\ is the current
+state of  the machine.   For each combination  of state
+and symbol  $(\state,c)$, the grammar will  contain the
+non-terminal \nont{}{c}. We shall synchronize each move
+of the  machine to  a derivation  step using  a context
+free production, followed, if possible, by a derivation
+step using either  $\bcar\acar \rightarrow \epsilon$ or
+$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
+synchronization,  we  shall   establish  the  following
+invariant  relation between  the machine  configuration
+and the sentential form:
 
 \begin{quote}
-  If the configuration of the machine is $w_l(\state,c)w_r$, then the
-  sentential form will be $\overline{w}_l\nont{}{c}\,w_r$, where
-  $\overline{w}_l$ is the same as $w_l$ but with each symbol $d$ in $w_l$
-  replaced by $\overline{d}$.
+  If    the   configuration    of   the    machine   is
+  $w_l(\state,c)w_r$, then the  sentential form will be
+  $\overline{w}_l\nont{}{c}\,w_r$,                where
+  $\overline{w}_l$ is  the same as $w_l$  but with each
+  symbol $d$ in $w_l$ replaced by $\overline{d}$.
 \end{quote}
@@
 \end{enumerate}
-The idea behind  the productions is explained with  an example: Assume
-that        the        current        sentential        form        is
-\bcar\bcdr\nont{i}{\acar}\acar\acar.  Assume that  the  machine has  a
-transition  $(S_i,\acar) \rightarrow (S_j,\acdr,  L)$. Since  the next
-corresponding step in the derivation  has to be done without any prior
-knowledge of whether the symbol to the left of the tape is a \acar\ or
-a  \acdr, two  productions are  provided,  and the  invariant will  be
-maintained  only   if  the  production   $\nont{i}{\acar}  \rightarrow
-\acdr\nont{j}{\acdr}\acdr$  is  chosen  for   the  next  step  in  the
-derivation.         This         gives        the        configuration
-\bcar\bcdr\acdr\nont{i}{\acdr}\acdr\acar\acar.    Simplification  with
-the    production    $\bcdr\acdr    \rightarrow    \epsilon$    yields
-\bcar\nont{i}{\acdr}\acdr\acar\acar, which  exactly corresponds to the
-changed configuration  of the machine.  Notice carefully that  a wrong
-choice breaks the invariant and it cannot be recovered subsequently by
-any choice of productions.
+The idea  behind the  productions is explained  with an
+example:  Assume that  the current  sentential form  is
+\bcar\bcdr\nont{i}{\acar}\acar\acar.   Assume that  the
+machine  has  a   transition  $(S_i,\acar)  \rightarrow
+(S_j,\acdr, L)$.  Since the next corresponding  step in
+the  derivation  has  to  be  done  without  any  prior
+knowledge of whether the symbol to the left of the tape
+is a \acar\  or a \acdr, two  productions are provided,
+and  the  invariant  will  be maintained  only  if  the
+production         $\nont{i}{\acar}         \rightarrow
+\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
+in  the  derivation.    This  gives  the  configuration
+\bcar\bcdr\acdr\nont{i}{\acdr}\acdr\acar\acar.
+Simplification   with    the   production   $\bcdr\acdr
+\rightarrow               \epsilon$              yields
+\bcar\nont{i}{\acdr}\acdr\acar\acar,    which   exactly
+corresponds  to   the  changed  configuration   of  the
+machine.  Notice  carefully that a wrong  choice breaks
+the invariant  and it cannot be  recovered subsequently
+by any choice of productions.
 
@@
 
-We circumvent the problem  of undecidability by over- approximating the
-CFG by non-deterministic finite state automata (NFA). The NFAs are then
-simplified using  the $\hookrightarrow$ rules.  Finally the simplified
-NFAs are converted to DFAs.
+We circumvent  the problem  of undecidability  by over-
+approximating the CFG by non-deterministic finite state
+automata (NFA). The NFAs  are then simplified using the
+$\hookrightarrow$ rules.   Finally the  simplified NFAs
+are converted to DFAs.
 
@@
 \label{sec:GC-scheme}
-Our experimental setup consists of  an interpreter for our language, a
-liveness  analyzer, and  a single  generation copying  collector.  The
-garbage collector  can be  easily configured to  work on the  basis of
-reachability (RGC  mode) or use  the liveness DFA (LGC  mode).  While
-exploring the  activation record of a  function, LGC uses  the DFA at
-the latest program  point traversed in the function's  body.  Thus, if
-LGC is invoked  at the program point $\pi$, it  would use the liveness
-DFA  at  $\pi$ itself  for  exploring the  root  set  of the  current
-activation record. For  any other activation record in  the stack, say
-for a function ${\mathit f}$  calling ${\mathit g}$ in the call-chain,
-the evaluation  point in ${\mathit f}$  which resulted in  the call to
-${\mathit g}$ would be used.
-
-We shall call  a unit allocatable memory as a {\em  cell}.  A cell can
-hold  a  basic  value   ($\mathit{bas}$),  the  constructor  \CONS\  $(\mathit{cons~
-arg_1~arg_2}$) or   a closure.  The closure, in  turn, can be  one of
-($\mathit{unop~arg_1}$),  ($\mathit{ binop~arg_1~arg_2}$)  and  function  application
-($\mathit{app~arg_1~arg_2})$.  Here each $\mathit{ arg_i}$ is a reference to another heap
-cell or terminated through the function $\mathit{makenull}$.
-
-While exploring the heap, the GC  may encounter closures. In RGC mode,
-the closure  will be explored  as usual  and copied by  traversing the
-references corresponding to the arguments. Since our liveness analysis
-predicts the  liveness of  data in its  evaluated form  only, closures
-present a challenge for LGC. LGC will  need to map the liveness of the
-free variables of the closure to  their evaluation point.  This is not
-straight  forward as  variables may  escape their  creation scope  and
-need to  be handled in an  approximate way.
-
-\subsection{A liveness-based garbage collection scheme}
+
+Our experimental setup  consists of an interpreter
+for  our  language,  a liveness  analyzer,  and  a
+single generation copying  collector.  The garbage
+collector can be configured to work on the
+basis  of  reachability  (RGC  mode)  or  use  
+liveness  DFAs  (LGC  mode).
+
+\warning{The reader has to be prepared for several, terms and concepts
+  in   this  paragraph}   In  the   liveness  analysis   described  in
+Section~\ref{sec:liveness-analysis},   the   liveness   of   variables
+appearing in an  application $s$ that lies on  e-paths, say $\epath_1$
+and  $\epath_2$, is  derived from  the dependence  chains along  these
+e-paths.  This  is  a   static  approximation  without  any  knowledge
+regarding  which of  these  e-paths will  be  eventually taken  during
+execution. During  GC, we would  like to  use a more  precise liveness
+using  the knowledge  of the  actual e-paths  taken during  execution.
+Thus  we   have  liveness  automata  separately   for  $\epath_1$  and
+$\epath_2$  as  well  as  a  automaton  corresponding  to  both  paths
+$\epath_1$ and  $\epath_2$.  The closure corresponding  to $s$ carries
+the  liveness  environment for  its  free  variables (as  pointers  to
+automata, one for each  variable).  Initially the liveness environment
+is   based   on   the    dependences   along   both   $\epath_1$   and
+$\epath_2$. However, after evaluating an \SIF\ condition, the liveness
+environments  are  updated  to  one  based  on  either  $\epath_1$  or
+$\epath_2$, so that subsequent GCs  are based on more precise liveness
+information.
+
+Based on  the above considerations,  we restrict the  possible garbage
+collection points to the following:
+\begin{enumerate}
+\item We  maintain static estimates  of the memory required  to create
+  the closures for each function body.  On entering a function, if the
+  available memory is  less than this requirement for  the function, a
+  GC is triggered.
+\item \warning{Since the evaluation of a  \SIF\ condition may trigger a garbage
+  collection, the
+  available memory is checked once again against a revised estimate of
+  the memory  required to execute  the program.  A GC is  triggered if
+  required.} 
+\end{enumerate}
+
+%% While  exploring  the
+%% activation record of a  function, LGC uses the DFA
+%% at  the  latest  program point  traversed  in  the
+%% function's body.   Thus, if LGC is  invoked at the
+%% program point $\pi$, it would use the liveness DFA
+%% at $\pi$ itself for exploring  the root set of the
+%% current   activation   record.   For   any   other
+%% activation record in the stack, say for a function
+%% ${\mathit  f}$   calling  ${\mathit  g}$   in  the
+%% call-chain, the evaluation  point in ${\mathit f}$
+%% which resulted in the  call to ${\mathit g}$ would
+%% be used.
+
+\warning{We shall call a unit  allocatable memory as a {\em
+  cell}.    A  cell   can  hold   a  basic   value
+($\mathsf{bas}$),          the         constructor
+\CONS\   $(\mathsf{cons~   arg_1~arg_2}$)   or   a
+closure.   The closure,  in  turn, can  be one  of
+($\mathsf{unop~arg}$),                ($\mathsf{
+  binop~arg_1~arg_2}$)  and  function  application
+($\mathsf{f~arg})$.  Here each $\mathsf{
+  arg_i}$ is  a reference to another  heap cell.  In
+addition,  the closure  also carries a pointer to
+an automaton (denoted $\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.}
+
+%% While  exploring the  heap, the  GC may  encounter
+%% closures.  In  RGC  mode,   the  closure  will  be
+%% copied as usual  by traversing  the
+%% references corresponding  to the  arguments. Since
+%% our  liveness analysis  predicts  the liveness  of
+%% data in its evaluated  form only, closures present
+%% a  challenge for  LGC. LGC  will need  to map  the
+%% liveness of  the free variables of  the closure to
+%% their  evaluation  point.   This is  not  straight
+%% forward  as variables  may  escape their  creation
+%% scope  and need  to be  handled in  an approximate
+%% way.
+
+%\subsection{A liveness-based garbage collection scheme}
 \label{sec:live-clo}
   \SetStartEndCondition{ }{}{}%
-  \SetKwProg{Fn}{procedure}{\string:}{}
+  \SetKwProg{Pro}{procedure}{\string:}{}
+  \SetKwProg{Fn}{function}{\string:}{}
   \SetKwFunction{Range}{range}%%
@@
   \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-  \SetKwFunction{Lgc}{$lgc$}%
-  \SetKwFunction{Copy}{$lgcCopy$}%
+  \SetKwFunction{Lgc}{$\mathsf{lgc}$}%
+  \SetKwFunction{Copy}{$\mathsf{copy}$}%
   \SetKwFunction{LCopy}{$lgcCopy$}%
   \SetKwFunction{RCopy}{$rgcCopy$}%
-
 \begin{algorithm}[t]
-  \Fn{\Lgc{}}
+  \Pro{\Lgc{}}
      {
-       \For {each reference $\mathit{ref}$ in root set}
-            {\Copy($\mathit{ref}$)\;}
-            $\mathit{copyReferencesOnPrintStack()}$\;  
+       \For {each reference $\mathsf{ref}$ in root set}
+            {$\mathsf{ref}$ = \Copy($\mathsf{ref}, \mathsf{init}(\mathsf{ref.dfa})$)\;}
+            ${\mathsf{copyReferencesOnPrintStack}()}$\;  
      }
-
-     \Fn{\Copy{$\mathit{ref}$}}
+     \Fn{\Copy{$\mathsf{ref}, \mathsf{state}$}}
         {
-          \If{$\mathit{ref}$ is found live using dfa}
+          \eIf {$\mathsf{final}(\mathsf{state}$)}
              {
-               $\mathit{newRef} = $  $\mathit{dupHeapCell(ref})$\;
-               \If{cellType($\mathit{ref}$) is $cons$}
+               $\mathsf{newRef} = \mathsf{dupHeapCell}(\mathsf{ref})$\;
+               \If{$\mathsf{ref\!.cell}$($\mathsf{ref}$)
+                 is a cons cell $\mathsf{(cons~arg_1~arg_2)}$}
                   {
                     {
-                      $\mathit{newCar}  = $  \Copy($\mathit{ref\!.car})$\;
-                      $\mathit{newRef.car}   =  newCar$\;
+                      $\mathsf{newRef}\!.\mathsf{arg_1}  = \Copy(\mathsf{arg_1}, \mathsf{next}(\mathsf{state}, 0))$\;
+%                      $\mathsf{newRef}\!.\mathsf{arg_1}   =  \mathsf{newCar}$\;
                     }
                     {
-                      $\mathit{newCdr}$ =   \Copy($\mathit{ref.cdr})$\;
-                      $\mathit{newRef.cdr}   = \mathit{newCdr}$\;
+                      $\mathsf{newRef}\!.\mathsf{arg_2} =  \Copy(\mathsf{arg_2},
+                      \mathsf{next}(\mathsf{state} , 1))$\;
+%                      $\mathsf{newRef}\!.\mathsf{arg_2}   = \mathsf{newCdr}$\;
                     }    
                   }
-               \If {cellType($\mathit{ref}$) is closure}
+               \If
+                  {$\mathsf{ref\!.cell}$ is a
+                    closure cell, generically  $\mathsf{(binop~arg_1~arg_2)}$}
+              %      
                    { 
-                     $\mathit{newRef.arg1}$ =  \Copy$\mathit{(cell.arg1)}$\;
-                     $\mathit{newRef.arg2}$ =  \Copy$\mathit{(cell.arg2)}$\;
+                     $\mathsf{newRef}\!.\mathsf{arg_1} = \Copy(\mathsf{arg_1}, \mathsf{init}(\mathsf{arg_1.dfa}))$\;
+                     $\mathsf{newRef}\!.\mathsf{arg_2} = \Copy(\mathsf{arg_2}, \mathsf{init}(\mathsf{arg_2.dfa}))$\;
                    }
              }
+             {$\mathsf{newRef = ref}$}
+\KwRet $\mathsf{newRef}$\;
         }
-
-        \caption{LGC with closure collection based on liveness.\label{algo:lgc-a}}
+        \caption{Liveness-based
+          garbage collection.  \label{algo:lgc-a}}
 \end{algorithm}
-Algorithm~\ref{algo:lgc-a}   describes    a   liveness-based   garbage
-collection scheme.  Starting with the root set, each live reference is
-explored  using   $\mathit{  lgcCopy}$.    Copying  \CONS\   cells  is
-simple---it  just involves  copying  the cell  itself and  recursively
-copying the  \CAR\ and the  \CDR\ fields.   On the other  hand copying
-closures  requires  determining  the  liveness  of  its  arguments  at
-run-time.
-
-Let   us  consider   determining  the   liveness  of   a  closure   at
-runtime. Consider a closure $(app~f~y)$  has a reference from the root
-set  variable  $x$.  If $x$  is  live,  then  we  know that  the  cell
-$(app~f~y)$ has  to be copied.  To copy  the heap structure  rooted at
-$y$, we need the liveness of  $y$ at the program point which triggered
-the evaluation of  $(app~f~y)$. However, this is only  possible if the
-activation record of the function  which created this closure is still
-on the activation stack. This does  not always happen since a function
-could pack a closure in a \CONS\ cell and return the \CONS\ cell.
-
-An alternative could be to  carry with each closure, extra information
-about the program point which triggered its evaluation.  This gives us
-precise liveness  information of the  free variables of a  closure but
-involves the overhead of updating the program point information during
-execution.  To avoid  this overhead we go for a  safe approximation by
-storing the {\em creation point of  the closure} and using it to check
-liveness.  This  is safe since the  liveness at the creation  point of
-the closure dominates  all possible evaluation points  of the closure,
-but it  might be imprecise  as the actual  liveness might be  way less
-than what is found at the creation point.
-
-Note that  when a reachability-based  collector visits a cell  for the
-first time,  it is  marked.  Explorations starting  from the  cell are
-curtailed if the cell is visited subsequently due to sharing.  Since a
-liveness-based collector may reach the  same cell multiple times, each
-time with a different automaton state, this curtailment cannot be done
-in the case  of LGC. This is an inherent  drawback of a liveness-based
-collector.  This  can be mitigated  by maintaining a list  of liveness
-automata states  that a  particular heap cell  was traversed  with and
-avoid repeated traversals if we reach the cell in one of these states.
-However this also involves storage and run-time penalty.
-
-To  avoid  performance  penalty,  we experimented  with  a  mixed-mode
-garbage collection scheme  which uses liveness for  evaluated data and
-reachability for closures.
-
-\subsection{Liveness-based collector with reachability-based closure
-  collection}
-\begin{algorithm}[t]
-  \Fn{\Lgc{}}
-     {
-       \For {each reference $\mathit{ref}$ in root set}
-            {\LCopy($\mathit{ref}$)\;}
-            $\mathit{copyReferencesOnPrintStack()}$\;  
-     }
-     \Fn{\LCopy{$\mathit{ref}$}}
-        {
-          \eIf{cellType$(\mathit{ref})$ is $cons$}
-              {
-                \eIf {$\mathit{ref}$ is found live using dfa and has not been copied using
-                  reachability}
-                     {
-                       $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
-                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
-                       $\mathit{ newRef.car}   = \mathit{ newCar}$\;
-                       $\mathit{ newCdr}  =$  \LCopy$\mathit{(ref.cdr)}$\;
-                       $\mathit{ newRef.cdr}   = \mathit{newCdr}$\;
-                     }
-                     { $\mathit{makenull(ref)}$\;}}
-              {
-                \If {cellType$\mathit{(ref)}$ is closure and 
-                  has not been copied}
-                    { 
-                      $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
-                    }
-              }
-        }
-        \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
-\end{algorithm}
-
-\begin{table}[ht!]
-\caption{Time taken by LGC for different strategies.  The programs are
-  modified (smaller) instances of original benchmarks.}
-\label{tab:compare-live-clo-reach-clo}
+Algorithm~\ref{algo:lgc-a}   describes  the    liveness-based   garbage
+collection scheme.  Starting  with the root set, each  cell pointed by
+each live  reference (i.e. whose associated  DFA state is final)  is copied
+using  $\mathsf{copy}$.   Copying  \CONS\ cells  is  simple---it  just
+involves  copying  the  cell  itself  and  conditionally  copying  the
+\CAR\ and  the \CDR\  fields by  referring to the  next states  of the
+automaton.   If the  reference points  to  a closure,  then, as  noted
+earlier, the closure carries pointers to the liveness automata for its
+arguments.   These automata are  used to  recursively initiate  copying of  the
+arguments.     Note     that      the     copying     strategy     for
+$\mathsf{(unop~arg_1~arg_2)}$  or  $\mathsf{(f~arg_1)}$  are  similar  to
+$\mathsf{(binop~arg_1~arg_2)}$ and have not been shown in the algorithm.
+
+%% Let   us  consider   determining  the   liveness  of   a  closure   at
+%% runtime. Consider a closure $(app~f~y)$  has a reference from the root
+%% set  variable  $x$.  If $x$  is  live,  then  we  know that  the  cell
+%% $(app~f~y)$ has  to be copied.  To copy  the heap structure  rooted at
+%% $y$, we need the liveness of  $y$ at the program point which triggered
+%% the evaluation of  $(app~f~y)$. However, this is only  possible if the
+%% activation record of the function  which created this closure is still
+%% on the activation stack. This does  not always happen since a function
+%% could pack a closure in a \CONS\ cell and return the \CONS\ cell.
+
+%% An alternative could be to  carry with each closure, extra information
+%% about the program point which triggered its evaluation.  This gives us
+%% precise liveness  information of the  free variables of a  closure but
+%% involves the overhead of updating the program point information during
+%% execution.  To avoid  this overhead we go for a  safe approximation by
+%% storing the {\em creation point of  the closure} and using it to check
+%% liveness.  This  is safe since the  liveness at the creation  point of
+%% the closure dominates  all possible evaluation points  of the closure,
+%% but it  might be imprecise  as the actual  liveness might be  way less
+%% than what is found at the creation point.
+
+%% Note that  when a reachability-based  collector visits a cell  for the
+%% first time,  it is  marked.  Explorations starting  from the  cell are
+%% curtailed if the cell is visited subsequently due to sharing.  Since a
+%% liveness-based collector may reach the  same cell multiple times, each
+%% time with a different automaton state, this curtailment cannot be done
+%% in the case  of LGC. This is an inherent  drawback of a liveness-based
+%% collector.  This  can be mitigated  by maintaining a list  of liveness
+%% automata states  that a  particular heap cell  was traversed  with and
+%% avoid repeated traversals if we reach the cell in one of these states.
+%% However this also involves storage and run-time penalty.
+
+%% To  avoid  performance  penalty,  we experimented  with  a  mixed-mode
+%% garbage collection scheme  which uses liveness for  evaluated data and
+%% reachability for closures.
+
+%% \subsection{Liveness-based collector with reachability-based closure
+%%   collection}
+%% \begin{algorithm}[t]
+%%   \Fn{\Lgc{}}
+%%      {
+%%        \For {each reference $\mathit{ref}$ in root set}
+%%             {\LCopy($\mathit{ref}$)\;}
+%%             $\mathit{copyReferencesOnPrintStack()}$\;  
+%%      }
+%%      \Fn{\LCopy{$\mathit{ref}$}}
+%%         {
+%%           \eIf{cellType$(\mathit{ref})$ is $cons$}
+%%               {
+%%                 \eIf {$\mathit{ref}$ is found live using dfa and has not been copied using
+%%                   reachability}
+%%                      {
+%%                        $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
+%%                        $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
+%%                        $\mathit{ newRef.car}   = \mathit{ newCar}$\;
+%%                        $\mathit{ newCdr}  =$  \LCopy$\mathit{(ref.cdr)}$\;
+%%                        $\mathit{ newRef.cdr}   = \mathit{newCdr}$\;
+%%                      }
+%%                      { $\mathit{makenull(ref)}$\;}}
+%%               {
+%%                 \If {cellType$\mathit{(ref)}$ is closure and 
+%%                   has not been copied}
+%%                     { 
+%%                       $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
+%%                     }
+%%               }
+%%         }
+%%         \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
+%% \end{algorithm}
+
+%% \begin{table}[ht!]
+%% \caption{Time taken by LGC for different strategies.  The programs are
+%%   modified (smaller) instances of original benchmarks.}
+%% \label{tab:compare-live-clo-reach-clo}
+%% \centering
+%% \begin{tabular}
+%% {| l | r | r | r |}
+%% \hline
+%% Program        & \#GC & \multicolumn{2}{c|}{Closure collected using} \\ \cline{3-4}
+%%                &      & Reachability & Liveness                      \\ \hline
+%% \verb@nqueens@ & 183  & 0.009 sec    & 0.228 sec                     \\\hline
+%% \verb@lambda@  & 11   & 0.001 sec    & 4.992 sec                     \\ \hline
+%% \verb@lcss@    & 1    & 0.000 sec    & 34.668 sec                    \\ \hline
+%% \end{tabular}
+%% \end{table}
+
+%% Algorithm~\ref{algo:lgc-b} describes a garbage collection scheme where
+%% the liveness automata is consulted  only for \CONS\ cells; and assumes
+%% everything under  a closure to be  live and copies everything  that is
+%% reachable.
+
+%% A consequence of  copying everything under a closure is  that we might
+%% reach  a non-live  cell that  has been  reclaimed earlier  by liveness
+%% based GC.   This can  happen if  a \CONS\  cell having  partially live
+%% sub-structure  is copied  using liveness  during a  collection and  it
+%% subsequently  becomes part  of a  live closure.   During a  subsequent
+%% collection, the collector may try to copy everything under the closure
+%% including the dead  sub-structure of the \CONS\ cell.   To avoid this,
+%% we ensure  that any non-live  references are not carried  across GC's.
+%% We do this by scanning the live  part of the buffer after every GC and
+%% nullifying any  reference which is  not live. This is  correct because
+%% any reference which  is dead during a collection is  guaranteed not be
+%% dereferenced any further by the program.
+
+%% Another advantage of  this mixed garbage collection scheme  is that it
+%% avoids the problem of multiple  explorations from the same cell.  This
+%% is done by maintaining a flag in  each cell to indicate whether it was
+%% copied using reachability.
+
+\begin{table*}[t!]
+\caption{Statistics for liveness analysis and garbage collection}
+\label{tab:exp-results}
 \centering
-\begin{tabular}
-{| l | r | r | r |}
-\hline
-Program        & \#GC & \multicolumn{2}{c|}{Closure collected using} \\ \cline{3-4}
-               &      & Reachability & Liveness                      \\ \hline
-\verb@nqueens@ & 183  & 0.009 sec    & 0.228 sec                     \\\hline
-\verb@lambda@  & 11   & 0.001 sec    & 4.992 sec                     \\ \hline
-\verb@lcss@    & 1    & 0.000 sec    & 34.668 sec                    \\ \hline
-\end{tabular}
-\end{table}
-
-Algorithm~\ref{algo:lgc-b} describes a garbage collection scheme where
-the liveness automata is consulted  only for \CONS\ cells; and assumes
-everything under  a closure to be  live and copies everything  that is
-reachable.
-
-A consequence of  copying everything under a closure is  that we might
-reach  a non-live  cell that  has been  reclaimed earlier  by liveness
-based GC.   This can  happen if  a \CONS\  cell having  partially live
-sub-structure  is copied  using liveness  during a  collection and  it
-subsequently  becomes part  of a  live closure.   During a  subsequent
-collection, the collector may try to copy everything under the closure
-including the dead  sub-structure of the \CONS\ cell.   To avoid this,
-we ensure  that any non-live  references are not carried  across GC's.
-We do this by scanning the live  part of the buffer after every GC and
-nullifying any  reference which is  not live. This is  correct because
-any reference which  is dead during a collection is  guaranteed not be
-dereferenced any further by the program.
-
-Another advantage of  this mixed garbage collection scheme  is that it
-avoids the problem of multiple  explorations from the same cell.  This
-is done by maintaining a flag in  each cell to indicate whether it was
-copied using reachability.
-
-\subsection{Garbage collection for references on print stack}
-In lazy  languages the evaluation of  closures is driven  by the print
+\include{data-liveness-analysis}
+\include{autotable}
+\vskip -5mm
+\end{table*}
+
+%\subsection{Garbage collection for references on print stack}
+In lazy  languages the evaluation of the top-level expression is driven  by the print
 function~\cite{Jones87}.  Printing of  atomic values is
@@
 language, both \CAR\ and \CDR\ could be closures requiring evaluation,
-and this  may trigger  a garbage collection.   When this  happens.  we
+and this  may trigger  more garbage collections.   When this  happens.  we
 have to consider  any references that might be on  the print stack and
 copy them. We extend the liveness analysis to the print function and
-use its result during garbage collection.
+use its result during garbage collection. \warning{Is this enough?}
 
@@
 
-\subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
-
-To compare the  effect of using reachability to  copy closures instead
-of liveness  we take some  representative programs from  our benchmark
-suite    and    execute    them    with    both.     As    shown    in
-Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
-collections in both cases are same but copying closures using liveness
-takes a lot more time compared to reachability based copying.
-
-The number  of garbage  collections is the same  since we  approximate the
-liveness of  a closure  to the  liveness at  its creation  point.  The
-liveness  we get  is usually  grossly over  approximated.  Using  this
-liveness may not  give too much of a benefit  compared to reachability
-as nearly everything would be live at its point of creation.
-
-As  mentioned   in  Section~\ref{sec:live-clo},  using   liveness  for
-closures  may  entail  multiple  traversals over  the  same  structure
-consuming more  time.  The  advantage of avoiding  multiple traversals
-can be clearly seen from the lcss example where a single GC takes less
-than a millisecond for reachability  based closure collection where as
-the  liveness  based collection  is  of  the  order of  seconds.  Lazy
-languages tend  to have a  lot of sharing  and hence using  a liveness
-based collection for closures is impractical.
-
-Therefore we believe  that using reachability for closures  is a sweet
-spot   in   terms   of   performance  for   liveness   based   garbage
-collectors.  For the  remainder  of  the section  all  LGC results  we
-discuss uses  a mixed-mode collector--liveness for  evaluated data and
-reachability for closures.
-
-\begin{table*}[ht!]
-\caption{Statistics for liveness analysis and garbage collection}
-\label{tab:exp-results}
-\centering
- \begin{tabular}{|l||c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|@{\ }c|} \hline
-   \multicolumn{1}{|r||}{Program} & \verb@lambda@ & \verb@nperm@ & \verb@treejoin@ & \verb@lcss@ & \verb@sudoku@ & \verb@fibheap@ & \verb@nqueens@ & \verb@knightstour@ & \verb@gc_bench@ \\ \hline\hline
-   \#CFG Nonterminals            & 1078          & 794          & 709             & 839         & 1758          & 780            & 529            & 659                & 417             \\ \hline
-   \#CFG Rules                   & 1823          & 1038         & 1610            & 1752        & 2509          & 1358           & 748            & 909                & 486             \\  \hline 
-   \#DFA States                  & 3312          & 2055         & 2332            & 2340        & 5672          & 2254           & 1229           & 1742               & 729             \\\hline
-   \#DFA Transitions             & 5792          & 3632         & 4235            & 4261        & 10857         & 4076           & 2078           & 3101               & 1124            \\\hline
-   DFA Generation Time (sec)      & 14.7          & 1.1          & 2690.5          & 13.3        & 798.5         & 49.8           & 0.5            & 5.1                & 0.1             \\\hline
- \end{tabular}\\
- (a) Data for Liveness Analysis \\[1pt]
-\include{autotable}
-\vskip -5mm
-\end{table*}
+%% \subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
+
+%% To compare the  effect of using reachability to  copy closures instead
+%% of liveness  we take some  representative programs from  our benchmark
+%% suite    and    execute    them    with    both.     As    shown    in
+%% Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
+%% collections in both cases are same but copying closures using liveness
+%% takes a lot more time compared to reachability based copying.
+
+%% The number  of garbage  collections is the same  since we  approximate the
+%% liveness of  a closure  to the  liveness at  its creation  point.  The
+%% liveness  we get  is usually  grossly over  approximated.  Using  this
+%% liveness may not  give too much of a benefit  compared to reachability
+%% as nearly everything would be live at its point of creation.
+
+%% As  mentioned   in  Section~\ref{sec:live-clo},  using   liveness  for
+%% closures  may  entail  multiple  traversals over  the  same  structure
+%% consuming more  time.  The  advantage of avoiding  multiple traversals
+%% can be clearly seen from the lcss example where a single GC takes less
+%% than a millisecond for reachability  based closure collection where as
+%% the  liveness  based collection  is  of  the  order of  seconds.  Lazy
+%% languages tend  to have a  lot of sharing  and hence using  a liveness
+%% based collection for closures is impractical.
+
+%% Therefore we believe  that using reachability for closures  is a sweet
+%% spot   in   terms   of   performance  for   liveness   based   garbage
+%% collectors.  For the  remainder  of  the section  all  LGC results  we
+%% discuss uses  a mixed-mode collector--liveness for  evaluated data and
+%% reachability for closures.
 
@@
 \newcommand{\hgt}{2.9cm}
-\begin{figure*}[p]
+\begin{figure*}[t]
 \renewcommand{\arraystretch}{.1}
 \begin{tabular}{@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
-   & {\epsfig{file=lambda_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+%%    \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
+%%    & {\epsfig{file=lambda_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+   \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+\\   \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+\\   \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+\\  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
-\\ (a) & (b)
+%\\ (a) & (b)
 \end{tabular}%\vskip -10mm
- \caption{Memory usage.  Column (a) shows complete  usage, column
-   (b) enlarges  a part  of the  usage.  The blue  and the  red curves
-   indicate the number of cons cells  in the active semi-space for RGC
-   and LGC  respectively.  The  black curve  represents the  number of
-   reachable cells and  the lightblue curve represents the  number of cells
-   that are  actually live (of  which liveness analysis does  a static
-   approximation).   x-axis   is  the  time  measured   in  number  of
-   cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
-   number of cons-cells (scaled down by  $10^3$).}
+ \caption{Memory usage.  
+%% Column (a) shows complete  usage, column 
+%%    (b) enlarges  a part  of the  usage.  
+The blue and the red curves indicate the number of cons
+cells  in  the  active   semi-space  for  RGC  and  LGC
+respectively.  The black curve represents the number of
+reachable cells and the  lightblue curve represents the
+number  of  cells  that  are actually  live  (of  which
+liveness analysis does a static approximation).  x-axis
+is the time measured  in number of cons-cells allocated
+(scaled down by factor $10^5$). y-axis is the number of
+cons-cells (scaled down by $10^3$).}
 \label{fig:memory-usage} \figrule
@@
 Memory  usage graphs  for select  benchmarks are  shown
-Figure~\ref{fig:memory-usage}, Column  (a).  As the number  of garbage
-collections tend to  be very large, for each benchmark  we also show a
-window which  is representative  of the  behavior for  that particular
-benchmark (Column (b)).  In all the programs we can see that the curve
-corresponding to  LGC (red  line) regularly dips  below the  RGC curve
-(blue line), the dip is specially visible for the benchmarks where LGC
-outperforms  RGC substantially.   The  graphs also  include curve  for
-reachable cells (black,  obtained approximately by forcing  RGC to run
-at a very high frequency) and the live cells for particular run (lightblue,
-obtained by post processing the heap access at the end of a program).
-
-One area  of concern is the  huge gap between the  actual liveness and
-the liveness  perceived by our  collector.  In  case of LGC  for eager
-languages~\cite{asati14lgc} the gap was  very narrow and almost touched
-the actual  liveness curve.  In  case of  lazy languages, this  gap is
-largely   due  to   approximate  reachability   based  collection   of
-closures. To implement liveness based  collection of closures, we need
-to  record liveness  data  for  each closure  and  update  it at  each
-evaluation  point.   These  space  and  run  time  overheads  make  it
-infeasible  to  incorporate  liveness  based  GC  for  closures  in  a
-practical garbage collector.
+Figure~\ref{fig:memory-usage},  Column   (a).   As  the
+number of  garbage collections  tend to be  very large,
+for  each benchmark  we  also show  a  window which  is
+representative  of  the  behavior for  that  particular
+benchmark (Column (b)).  In all the programs we can see
+that  the   curve  corresponding  to  LGC   (red  line)
+regularly dips below the RGC curve (blue line), the dip
+is  specially  visible  for the  benchmarks  where  LGC
+outperforms RGC substantially.  The graphs also include
+curve    for   reachable    cells   (black,    obtained
+approximately  by forcing  RGC to  run at  a very  high
+frequency)  and  the  live  cells  for  particular  run
+(lightblue, obtained by post processing the heap access
+at the end of a program).
+
+One area of concern is  the huge gap between the actual
+liveness and  the liveness perceived by  our collector.
+In  case of  LGC for  eager languages~\cite{asati14lgc}
+the gap was  very narrow and almost  touched the actual
+liveness curve.  In case of lazy languages, this gap is
+largely   due   to   approximate   reachability   based
+collection  of closures.  To  implement liveness  based
+collection of closures, we need to record liveness data
+for  each  closure and  update  it  at each  evaluation
+point.   These space  and  run time  overheads make  it
+infeasible  to   incorporate  liveness  based   GC  for
+closures in a practical garbage collector.
 
@@
 \label{sec:relatedwork}
-Although augmenting  garbage collection with liveness  information has
-been  studied earlier,  it was  mainly  in the  context of  imperative
-languages~\cite{Albert, Hirzel,  khedker07heap}.  Due to  the presence
-of global variables and mutation, practical utility of simple liveness
-based techniques are found to be inefficient.
-
-In the space  of functional languages, improving  space efficiency has
-been  mainly  studied as  a  compile  time activity.   Either  through
-rewriting                methods                such                as
+Although  augmenting garbage  collection with  liveness
+information has been studied  earlier, it was mainly in
+the   context  of   imperative  languages~\cite{Albert,
+  Hirzel,  khedker07heap}.   Due  to  the  presence  of
+global  variables and  mutation,  practical utility  of
+simple  liveness  based  techniques  are  found  to  be
+inefficient.
+
+In the  space of functional languages,  improving space
+efficiency has  been mainly  studied as a  compile time
+activity.   Either through  rewriting  methods such  as
 deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-sharing  analysis   based  reallocation~\cite{jones89compile},  region
-based analysis~\cite{tofte98region},  or through insertion  of compile
-time              nullifying              statements              such
-as~\cite{inoue88analysis,lee05static,Hamilton}.    The  compile   time
-marking approaches all rely on an efficient and precise alias analysis
-and in the  absence of it cannot provide  notable improvement. Another
-important approach  advocated by  Hofmann~\cite{HofmannJ03} is  to use
-linear  typing to  analyze  first-order programs  for  heap usage  and
-annotated  them. The  system  then uses  these  annotations to  re-use
-allocated memory  cells instead  of requesting  for newer  cells. This
-requires the  user to write programs  in a specific way  and hence may
-not be practical
-
-Simplifiers~\cite{ONeill}, described as lightweight daemons, 
-improve the efficiency of the program in general and the garbage
-collector in particular. Most of the simplifications mentioned in this
-work are subsumed by a liveness  based collector. We feel this work is
-orthogonal and can augment our  liveness based collector.  The closest
-to our  approach is the  liveness based garbage  collector implemented
-in~\cite{asati14lgc}. We  extend their work to  handle lazy evaluation
-and closures.
+sharing                  analysis                 based
+reallocation~\cite{jones89compile},     region    based
+analysis~\cite{tofte98region}, or  through insertion of
+compile     time     nullifying     statements     such
+as~\cite{inoue88analysis,lee05static,Hamilton}.     The
+compile  time   marking  approaches  all  rely   on  an
+efficient and precise alias analysis and in the absence
+of  it  cannot  provide  notable  improvement.  Another
+important         approach         advocated         by
+Hofmann~\cite{HofmannJ03}  is to  use linear  typing to
+analyze  first-order   programs  for  heap   usage  and
+annotated them. The system  then uses these annotations
+to re-use allocated memory  cells instead of requesting
+for  newer  cells.  This  requires the  user  to  write
+programs  in  a  specific  way and  hence  may  not  be
+practical
+
+Simplifiers~\cite{ONeill},  described   as  lightweight
+daemons,  improve  the  efficiency of  the  program  in
+general and  the garbage collector in  particular. Most
+of  the  simplifications  mentioned in  this  work  are
+subsumed by  a liveness  based collector. We  feel this
+work is  orthogonal and can augment  our liveness based
+collector.  The closest to our approach is the liveness
+based        garbage       collector        implemented
+in~\cite{asati14lgc}.  We extend  their work  to handle
+lazy evaluation and closures.
 
@@
 
-%% Previous  attempts  to increase  the  space  efficiency of  functional
-%% programs  by  additional  reclamation  of memory  fall  in  two  broad
-%% categories. In the first, the program itself is instrumented to manage
-%% reclamation and reallocation without the aid of the garbage collector.
-%% Such      attempts      include:      sharing      analysis      based
-%% reallocation~\cite{jones89compile},                      deforestation
+%% Previous attempts  to increase the  space efficiency
+%% of functional programs  by additional reclamation of
+%% memory fall  in two broad categories.  In the first,
+%% the  program   itself  is  instrumented   to  manage
+%% reclamation and reallocation without  the aid of the
+%% garbage collector.   Such attempts  include: sharing
+%% analysis  based  reallocation~\cite{jones89compile},
+%% deforestation
 %% techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-%% methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
-%% analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
-%% methods   that  enable   the   garbage  collector   to  collect   more
-%% garbage~\cite{inoue88analysis,lee05static}  by  explicitly  nullifying
-%% pointers  that are  not  live.  However,  the  nullification, done  at
-%% compile  time,  requires sharing  (alias)  analysis.   Our method,  in
-%% contrast, does not require sharing  because of the availability of the
-%% heap itself  at runtime.  To  the best of  our knowledge, this  is the
-%% first attempt  at liveness-based  marking of  the heap  during garbage
+%% methods based on linear logic~\cite{hofmann00linear}
+%% and region analysis~\cite{tofte98region}.  Closer to
+%% our  approach, there  are  methods  that enable  the
+%% garbage      collector      to     collect      more
+%% garbage~\cite{inoue88analysis,lee05static}        by
+%% explicitly  nullifying pointers  that are  not live.
+%% However,  the nullification,  done at  compile time,
+%% requires sharing  (alias) analysis.  Our  method, in
+%% contrast, does  not require  sharing because  of the
+%% availability of the heap  itself at runtime.  To the
+%% best of our knowledge, this  is the first attempt at
+%% liveness-based  marking of  the heap  during garbage
 %% collection.
@@
 \label{sec:conclusion}
-%% We  have  defined  a  notion  of liveness  on  structured  data;  this
-%% generalizes classical liveness and strong liveness.  We started with a
-%% general fully context-sensitive analysis  which we proved correct with
-%% respect to  a minefield semantics  (this models the effect  of garbage
-%% collection between every evaluation step).
-
-%% To  avoid scalability  issues (and  to  avoid performing  part of  the
-%% liveness computation at run time)  we defined an 0-CFA version of this
-%% liveness analysis  in which  demands for function  $f$ at  all calling
-%% contexts are conflated into  a single demand $\sigma_f$.  This enabled
-%% us to treat the liveness equations symbolically obtaining context-free
-%% grammars for liveness at each GC point (calls to user functions and to
-%% $\CONS$).  These were then converted to DFA for run-time consultation
-%% by  the garbage collector.  Experiments confirm  the precision  of the
-%% analysis.
-
-%% To obtain performance figures we compared a reachability-based garbage
-%% collector with a liveness-based  collector.  This showed a decrease in
-%% the  number  of  GCs,   more  garbage  collected  per  invocation.   A
-%% significant benefit of LGC is  that programs can run in smaller memory
-%% when compared to  RGC. This is potentially useful  in situations where
-%% memory  is  limited---as with  embedded  systems.  For  a majority  of
-%% programs, the garbage collection times were reduced.
-
-%% One issue we  highlighted was that while fewer  nodes were marked (and
-%% more garbage collected), sometimes  cons cells could be visited and
-%% traversed  multiple times  with different  sets of  liveness  paths to
-%% explore.  Further work includes  improvements to the classical copying
-%% collector to reduce the cost of this.
-We extended the liveness based garbage collection to lazy
-  languages and shown its benefit  for garbage collection in practical
-  programs. We  defined a context  sensitive liveness analysis  that uses
-  context independent summaries of functions obtained using a symbolic
-  demand. We showed  that obtaining precise solution  to these equations
-  is undecidable in  our formulation, and hence  we safely approximate
-  the result using DFA. These DFA are consulted by the garbage collector
-  to improve  collection.
-
-  Liveness   of   closures   presented  further   challenges   to   our
-  implementation.   We compared  different approximate  strategies for
-  handling   liveness   of   closures,   and  found   that   a   mixed
-  strategy---reachability  based  collection   of  closure  arguments,
-  liveness  based  collection  of   everything  else---works  best  in
-  practice as it  avoids runtime and space  overheads.  With mixed-mode
-  garbage collection scheme,  we were able to collect  more cells than
-  reachability based collectors at a reasonable overhead.
-
-  Although we  provide an  implementation for  a liveness  based garbage
-  collection scheme for lazy language, we  do not provide a formal proof
-  of its correctness.  A formal  proof of correctness would describe how
-  closures  are  handled  during  garbage  collection  and  liveness  is
+%% We have  defined a notion of  liveness on structured
+%% data; this generalizes classical liveness and strong
+%% liveness.    We  started   with   a  general   fully
+%% context-sensitive analysis  which we  proved correct
+%% with respect  to a minefield semantics  (this models
+%% the  effect  of  garbage  collection  between  every
+%% evaluation step).
+
+%% To avoid scalability issues (and to avoid performing
+%% part  of the  liveness computation  at run  time) we
+%% defined an  0-CFA version of this  liveness analysis
+%% in  which demands  for function  $f$ at  all calling
+%% contexts   are  conflated   into  a   single  demand
+%% $\sigma_f$.  This  enabled us to treat  the liveness
+%% equations   symbolically    obtaining   context-free
+%% grammars  for liveness  at each  GC point  (calls to
+%% user  functions and  to $\CONS$).   These were  then
+%% converted to  DFA for  run-time consultation  by the
+%% garbage   collector.     Experiments   confirm   the
+%% precision of the analysis.
+
+%% To   obtain  performance   figures  we   compared  a
+%% reachability-based   garbage    collector   with   a
+%% liveness-based collector.  This showed a decrease in
+%% the  number  of  GCs,  more  garbage  collected  per
+%% invocation.  A  significant benefit  of LGC  is that
+%% programs can run in  smaller memory when compared to
+%% RGC. This is potentially  useful in situations where
+%% memory is limited---as with embedded systems.  For a
+%% majority of  programs, the garbage  collection times
+%% were reduced.
+
+%% One issue we highlighted  was that while fewer nodes
+%% were marked (and  more garbage collected), sometimes
+%% cons cells  could be visited and  traversed multiple
+%% times  with  different  sets of  liveness  paths  to
+%% explore.  Further work  includes improvements to the
+%% classical copying  collector to  reduce the  cost of
+%% this.
+We extended  the liveness  based garbage  collection to
+lazy  languages  and  shown  its  benefit  for  garbage
+collection in practical programs.  We defined a context
+sensitive   liveness   analysis   that   uses   context
+independent  summaries of  functions  obtained using  a
+symbolic  demand.  We  showed  that  obtaining  precise
+solution  to  these  equations is  undecidable  in  our
+formulation, and hence we safely approximate the result
+using  DFA.  These DFA  are  consulted  by the  garbage
+collector to improve collection.
+
+Liveness  of closures  presented further  challenges to
+our implementation.  We  compared different approximate
+strategies for handling liveness of closures, and found
+that a  mixed strategy---reachability  based collection
+of  closure  arguments,  liveness based  collection  of
+everything else---works  best in practice as  it avoids
+runtime and  space overheads.  With  mixed-mode garbage
+collection scheme,  we were able to  collect more cells
+than  reachability  based  collectors at  a  reasonable
+overhead.
+
+Although we  provide an  implementation for  a liveness
+based garbage  collection scheme for lazy  language, we
+do not  provide a formal  proof of its  correctness.  A
+formal proof of correctness would describe how closures
+are handled  during garbage collection and  liveness is
 propagated inside closures.
 
-  Another interesting exercise is to narrow the gap between the actual
-  liveness and  the perceived liveness  of our garbage  collector. Our
-  experiments (not  reported here) show  that a significant  number of
-  dead cells  get trapped inside closures.   Using strictness analysis
-  to eagerly evaluate closures might  release more of these dead cells
-  to be  garbage collected.   Since our  garbage collector  can handle
-  closures  (suspended evaluations),  another  major  challenge is  to
-  extend it to support higher order programs.
-
-  Orthogonally,  we plan  to improve  the efficiency  of the  liveness
-  based garbage collector using heuristics  such as limiting the depth
-  of  DFA,   merging  nearly-equivalent   states  and   using  better
-  representation  and algorithms  for automata  manipulation. We  also
-  need  to   investigate  the  interaction  of   liveness  with  other
-  collection schemes, such as incremental and generational collection.
-  In  summary, we  need to  investigate  ways to  make liveness  based
-  garbage collection attractive for practical collectors.
-
-%\balance
+  Another  interesting exercise  is to  narrow the  gap
+  between  the   actual  liveness  and   the  perceived
+  liveness  of our  garbage collector.  Our experiments
+  (not reported here) show that a significant number of
+  dead  cells  get   trapped  inside  closures.   Using
+  strictness  analysis  to  eagerly  evaluate  closures
+  might release more of these  dead cells to be garbage
+  collected.   Since our  garbage collector  can handle
+  closures   (suspended  evaluations),   another  major
+  challenge  is to  extend it  to support  higher order
+  programs.
+
+  Orthogonally, we  plan to  improve the  efficiency of
+  the liveness based garbage collector using heuristics
+  such   as  limiting   the  depth   of  DFA,   merging
+  nearly-equivalent    states    and    using    better
+  representation    and    algorithms   for    automata
+  manipulation.  We   also  need  to   investigate  the
+  interaction   of  liveness   with  other   collection
+  schemes,   such  as   incremental  and   generational
+  collection.  In summary, we  need to investigate ways
+  to make liveness  based garbage collection attractive
+  for practical collectors.
+
 \bibliography{fun_hra}{}
 \bibliographystyle{abbrv}
+
+\newpage
+\ 
+\newpage
+
+\section{Appendix: Soundness of liveness analysis}
+
+\begin{figure*}[t!]
+\begin{center}\footnotesize
+\renewcommand{\arraystretch}{1.5}
+
+\begin{tabular}{|c|c|c|}
+\hline
+Premise & Transition & Rule name \\ 
+\hline
+
+\hline
+
+          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
+  H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
+    \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
+\\
+\hline
+          & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
+  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
+  \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
+  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
+\sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+  H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
+\sc{car-clo}
+\\
+
+\hline
+$H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
+(\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, H,\, s, \, \cred{\sigma}$      &
+\sc{car-1-clo} \\
+
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H,
+  (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, H[\rho'(z)
+    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma'}$}      &
+\sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+  H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+  (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+\sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+  S, H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
+  (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+\sc{prim-2-clo} \\
+\hline
+{$\mathit{g}~\mbox{defined as}$
+$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
+  (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
+    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma}$}      &
+\sc{funcall} \\
+\hline
+$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+  \ell]\rangle], e, \cred{\sigma}$} &
+\sc{let} \\ 
+\hline
+$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \cred{\sigma}$  $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma}$} & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \cred{\sigma}$   $\rightsquigarrow
+\rho, S, H,  e_2, \cred{\sigma}$} & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, H, s,
+\cred{\clazy\sigma}$}
+&
+\sc{if-clo} \\
+\hline
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
+  (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
+  \cred{\sigma'}$} &
+\sc{return-whnf}\\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\psi:\SRETURN~x), \cred{\sigma}$  $
+  \rightsquigarrow$
+$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, H,  s,
+  \cred{\sigma}$} &
+\sc{return-clo} \\
+\hline
+\end{tabular}
+\caption{Minefield semantics.\label{fig:minefield-semantics}}
+\end{center}
+\vskip -5mm
+\end{figure*}
+
+We shall now  present a proof of soundness of  the liveness analysis
+presented in Section~\ref{sec:liveness-analysis}.   While the proof is
+on the broad lines of \cite{asati14lgc}, there are several differences
+due to  the lazy  semantics of  the language  being 
+analyzed in this paper. Here is the idea behind the proof.
+\begin{enumerate}
+\item       We        enrich       the        standard       semantics
+  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
+  collection  just  before  each  $\rightsquigarrow$  transition.  The
+  simulated garbage collection starts from  the root-set and inserts a
+  special value  $\bot$ at each location  in the heap that  contains a
+  reference that is reachable but not live. Any attempt to dereference
+  such  locations during  the  transition will  result  in entering  a
+  special state denoted \bang. We call the semantics after
+  augmentation  \emph{minefield semantics}.
+\item \label{inline} Assuming that a  program enters the \bang\ state,
+  we  show  how to  construct,  through  inline expansion,  a  program
+  without function calls which has the same minefield behavior.
+\item  The final  part  of the  proof shows  that  no program  without
+  function calls can enter a \bang\ state. As a consequence no program
+  (with or without function calls) can enter the \bang\ state. 
+\end{enumerate}
+
+\subsection{Minefield semantics}
+
+To set up the minefield semantics, we follow these steps:
+\begin{enumerate}
+\item   We   enrich    the   abstract   machine   state
+  $\rho,\,S,\,H,\,e$   to  $\rho,\,S,\,H,\,e,\,\sigma$,
+  where  $\sigma$  is  the  demand  on  the  expression
+  $e$. We  call such a state  a \emph{minefield state}.
+  The demand $\sigma$ is  ``dynamic'' in that it arises
+  from the  actual sequence of function  calls that led
+  to  the evaluation  of  $e$.  The  0-CFA demand  used
+  during    liveness     analysis    is     a    static
+  over-approximation   of   $\sigma$   that   considers
+  \emph{all  possible calls}  that  could  lead to  the
+  evaluation of  $e$.  The  information in  a suspended
+  evaluation  context  (on  the   stack  $S$)  is  also
+  similarly augmented with its  demand.  Thus a stacked
+  entry now takes  the form $(\rho, x,  e, \sigma)$.  A
+  modification  of  the   small  step  semantics  which
+  carries the extra information  is shown in the bottom
+  table of Figure~\ref{fig:minefield-semantics}.
+\item $GC$  models a liveness based  garbage collection
+  $GC(\rho, S,  H, \sigma)$ returns $(\rho',  S', H')$,
+  where the changes are due to certain references being
+  replaced by  $\bot$.  This represents the  act of not
+  retaining   (effectively   garbage  collecting)   the
+  objects  pointed to  by  these  references during  an
+  actual  garbage  collection.  This  is  done  in  the
+  following way:
+  \begin{enumerate}
+  \item Given the current  evaluation context $\rho, S,
+    H,   e,  \sigma$   and   the  demand   transformers
+    $\Lfonly$, we can  construct a liveness environment
+    \Lv.   If  $e$  is  an expression,  then  $  \Lv  =
+    \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
+    application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
+    We  can similarly  define the  liveness environment
+    for each  of the  suspended evaluation  contexts in
+    $S$, giving  a stack of liveness  environments that
+    we shall denote $\mathsf{SL}$.
+  \item \emph{Garbage collection  of objects pointed by
+    root   set:}  For   each   $x  \in   domain(\rho)$,
+    $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
+    \Lv$.    Similarly,    for   each    stack    entry
+    $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
+    corresponding      liveness     environment      in
+    $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
+    $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
+  \item  \emph{Garbage  collection  of objects  in  the
+    heap:} For each location  $\ell$, $H'(\ell) = \bot$
+    iff  there   is  no  $x$  in   either  the  current
+    environment  or  one  of the  stacked  environments
+    $\rho$, such  that for some  $\alpha$ corresponding
+    to a  closure-free path  in the heap  starting from
+    $x$,  $H[x.\alpha]   =  \ell$  and   $x.\alpha  \in
+    \Lv$. Here \Lv\ is the liveness of the variables in
+    the current environment or the stacked environment.
+
+\warning{We have to tell  the reader about closure free
+  paths  earlier  on.   Also,  has  $H[x.\alpha]$  been
+  defined?}
+  \end{enumerate}
+Starting from the initial state $([\;]_\rho,([\;]_\rho,
+\ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
+every transition  is preceded  by a  garbage collection
+using $GC$.  The details of the transition for the {\sc
+  car-clo}  rule   is  shown   in  the  top   table  of
+Figure~\ref{fig:minefield-semantics}.  If  this  or  an
+earlier  call to  $GC$  results  in $H(\rho(x))$  being
+bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
+enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
+the transition is the same as the earlier {\sc car-clo}
+rule.
+\end{enumerate}
+
+\subsection{Inlining}
+
+Consider a call tree, each  of whose nodes represents a
+function  that  was  called (but  did  not  necessarily
+return because  of a \bang).   Assume that each  of the
+nodes of  the tree is  also annotated with  the program
+point where  the corresponding call was  invoked.  This
+tree  can  be  used  to  inline  function  calls  in  a
+hierarchical fashion.  The details of inlining a single
+function      call     $(\LET\,\,      x     \leftarrow
+(f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
+Let  $f$ be  defined  (after renaming  its formals  and
+locals  to  be  disjoint from  existing  variables)  by
+$(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
+replaced  by   a  sequence  of  $\LET$s   of  the  form
+$z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
+but with its  $(\SRETURN\,\,w)$ expressions replaced by
+$(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
+We prefer to use the  identity function $\ID$ as a form
+of no-op rather than  introducing the form $(\LET\,\, x
+\leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
+$\LET$-definition   is  a   simple  variable.    For  a
+call-less program,  the initial state of  the minefield
+semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
+\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{H},
+(e_{\mainpgm})$.
+
+\subsection{Soundness Results}
+
+  We  consider  a  garbage  collection  followed  by  a
+  transition as a step.  We show by induction that, for
+  programs  without function  calls, starting  from the
+  initial  state  that  has $e_{\main}$  as  the  first
+  expression to  be evaluated  instead of  $\main$, any
+  transition of $n$ steps occurs without a \bang.
+
+We  first  need  an auxiliary  result  about  minefield
+semantics. Consider  a trace of a  minefield execution.
+For every minefield state  $(\rho, S, H, e(s), \sigma)$
+that appears  on the LHS of  a $\rightsquigarrow$ step,
+the  demand   $\sigma$  on   the  expression   $e$  (or
+application $s$) is non-null.  This can be proved by an
+induction  on  the  number  of  steps  leading  to  the
+minefield  state.   The  base step  holds  because  the
+demand $\sigma_{all}$  on $\mainpgm$ is  non-null.  The
+inductive step  follows from  the observation  that for
+each  step of  the minefield  semantics, if  the demand
+$\sigma$ on  the LHS of  a minefield step  is non-null,
+the demand on  the RHS is a  transformation of $\sigma$
+(for example $\clazy\sigma$) which is also non-null.
+
+Let us clarify this observation:  Since the minefield semantics models
+lazy  evaluation,  parts  of  the  program may  not  be  evaluated  at
+all.  However,  if  an  expression  (or  application)  happens  to  be
+evaluated, it  will appear  on the  LHS of a  minefield step,  and the
+context-sensitive demand on it will be non-null.
+
+We  show  next  that   the  minefield  execution  of  a
+call-less program  cannot go \bang. Note  that, for the
+sake of  concreteness, by a single  minefield execution
+we   shall  mean   the  evaluation   of  $(main)$   (or
+$e_{\mainpgm}$,  for call-less  programs)  to its  WHNF
+driven  by   the  printing  mechanism   represented  by
+\print. With  minor variations, the proof  will also be
+applicable   for   subsequent  evaluations   fired   by
+$\print$.
+
+\begin{lemma}
+\label{lemma:call-less-cannot-go-bang}
+Consider the minefield execution of  a program without function calls.
+Such a program cannot enter the \bang\ state.
+\end{lemma}
+\begin{proof}
+Consider a state in the minefield execution of a program $(\rho, S, H,
+e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
+steps  leading to  this state  that  the next  step cannot  lead to  a
+\bang\ state.
+
+When $n$ is 0, the  state is $([\;]_\rho,\_, [\;]_{S} ,
+[\;]_{H},  e_{\mainpgm},  \sigma_{all})$.  Further,  we
+have   to  show   that   a  $GC(..)$   followed  by   a
+$\rightsquigarrow$ step cannot lead  to a \bang\ state.
+Now, because  our programs  are in  ANF, $e_{\mainpgm}$
+can  only  be a  $\LET$.  A  {\sc  let} step  does  not
+dereference anything and thus cannot result in a \bang.
+
+For the inductive step, we  shall show that none of the
+minefield steps that  involves dereferencing results in
+a   \bang.   These  are   the   steps   which  have   a
+$H(\rho(...))$ in the premise.  Recall that a minefield
+step   consists   of   a   GC  step   followed   by   a
+$\rightsquigarrow$.   The   $\rightsquigarrow$  can  go
+\bang\ because it dereferences a $\bot$ inserted by the
+immediately  preceding  GC  or  the GC  of  an  earlier
+minefield step.  \warning{However the  demand $\sigma'$
+  on basis of which an earlier GC would have inserted a
+  $\bot$,  would  have   included  the  current  demand
+  $\sigma$.}  Thus,  if the  earlier GC had  inserted a
+$\bot$, so would the current  GC.  Thus it is enough to
+consider the $\rightsquigarrow$  along with the current
+GC and show that it does not lead to a \bang.
+
+We consider the steps {\sc car-clo} and {\sc car-1-clo}
+only. The rest  of the steps are similar.  For the {\sc
+  car-clo}  step in  the state  $\rho, S,  H, (\CAR~x),
+\sigma$, we  know that $\sigma$ is  non-null. There for
+the  liveness  of  $x$  includes  $\epsilon$,  and  the
+dereferencing $H(\rho(x))$ will go without \bang.
+
+For the  {\sc car-1-clo}  step, observe that  there are
+two dereferences.  First $x$  is dereferenced to  get a
+cons  cell  and then  the  head  of  the cons  cell  is
+dereferenced  to  obtain  a  closure.   If  the  demand
+$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
+of    $x$   will    include    both   $\epsilon$    and
+$\acar\epsilon$,  and  a  GC with  this  liveness  will
+neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
+first   component  of   the   cons   cell.  Thus   both
+dereferences    will    happen   normally    and    the
+$\rightsquigarrow$ step will not enter \bang.
+\end{proof}
+
+Now we are ready to prove the main soundness result.
+
+\begin{theorem}
+The  minefield  execution of  no  program  can enter  a
+\bang\ state.
+\end{theorem}
+
+\begin{proof}
+Assume to  the contrary that  a program $P$  enters the
+\bang\  state.  We can  transform  $P$  to a  call-less
+program $P'$ such that  the minefield executions of $P$
+and $P'$  are identical  except for change  of variable
+names.  \warning{Check  whether  the previous  line  is
+  correct.}                 However,                 by
+Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
+$P'$,   a   call-less   program,   cannot   enter   the
+\bang\  state.  Therefor  $P$  also  cannot  enter  the
+\bang\ state.
+\end{proof}
+
 \end{document}
+
+%%  LocalWords:  dependences subscripting

### 2015-11-26 lazy-scheme/post_process/gen_gc_stats.sh
Fixed a bug in RGC

@@ -79,7 +79,7 @@ mkdir -p $OUTPUT;
 LOG=$OUTPUT/runtime
 echo -n 'Start At : ' > $LOG
 date                 >> $LOG
-EXECNAME=${HOME}/Work/LiveNStrict/lazy-scheme/src/Simulator_Stats
+EXECNAME=${DIRNAME}/../src/Simulator_Stats
 STATSFILE=GC_STATS.txt
 SCMFILE=$TEST/${TESTNAME}.scm
 if [ "$PPONLY" = "false" ]; then

### 2015-11-26 lazy-scheme/post_process/gen_stats_and_plotall.sh

@@ -3,7 +3,7 @@
 DIR=`dirname $0`
 PROGNAME=`basename $0`
 TestBMs="  test testsmall small"
-SmallBMs=" nperm fibheap lambda sudoku lcss nqueens fft huffman "
+SmallBMs=" nperm fibheap lambda sudoku lcss nqueens huffman "
 LargeBMs=" treejoin knightstour "
 AllBMs="$SmallBMs $LargeBMs"
 help()
@@ -46,7 +46,7 @@ declare -A limit
 
 limit["nqueens"]=10101
 limit["nperm"]=27428
-limit["fibheap"]=37041
+limit["fibheap"]=37043
 limit["treejoin"]=1616533
 limit["sudoku"]=4066
 limit["lcss"]=22243

### 2015-11-26 lazy-scheme/post_process/gen_gc_stats.sh
fix path

@@ -8,6 +8,7 @@ CPWD=`pwd`
 echo "Current dir is `pwd`"
 DIRNAME=`echo $DIRNAME | sed -e "s|\.|$CPWD|"`
 echo "Script home is $DIRNAME"
+PROJDIR=`dirname $DIRNAME`
 
 parse()
 {
@@ -79,7 +80,8 @@ mkdir -p $OUTPUT;
 LOG=$OUTPUT/runtime
 echo -n 'Start At : ' > $LOG
 date                 >> $LOG
-EXECNAME=${HOME}/Work/LiveNStrict/lazy-scheme/src/Simulator_Stats
+#EXECNAME=${HOME}/Work/LiveNStrict/lazy-scheme/src/Simulator_Stats
+EXECNAME=${PROJDIR}/src/Simulator_Stats
 STATSFILE=GC_STATS.txt
 SCMFILE=$TEST/${TESTNAME}.scm
 if [ "$PPONLY" = "false" ]; then

### 2015-11-26 paper/proof.tex

@@
 \definecolor{Myblue}{rgb}{.2,0,1}
+\definecolor{Myred}{rgb}{1,.2,.2}
+
 \newcommand{\comment}[1]{{\color{Myblue}{#1}}}
-%\newcommand{\cblue}[1]{{\color{blue}{#1}}}
+\newcommand{\warning}[1]{{\color{Myred}{#1}}}
 %\newcommand{\blankout}[1]{}
 
+\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
 %\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
@@
 \rightsquigarrow
-\rho',\, (\rho, x, (\CAR~x))\!:\!S, \,H,\, s$      &
+\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,H,\, s$      &
 \sc{car-clo}
@@
 \begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}\\
+\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data + \{empty\} & \mbox{-- Heap}\\
 d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
@@
 evaluation contexts (represented as  continuation frames) in the stack
-$S$.  The heap $H$ is common to all contexts.
+$S$.  The heap $H$ is common to all contexts. The 3-tuple $\rho, \stk,
+\heap$ representing 
+the state of the transition system will be denoted $\mathcal{S}$.
 
@@
 %of an application involving a selector such  as $\CAR$, 
-else, in the presence of sharing,   $d_1$  may  be  evaluated  more  than  once~\cite{Jones87}.    Only after this  is the
-actual selection done ({\sc car-select}).
+else, in  the presence of  sharing, $d_1$  may be evaluated  more than
+once~\cite{Jones87}.   Only after  this is  the actual  selection done
+({\sc car-select}).  $\stackrel{+}{\rightsquigarrow}$
+denotes  the  transitive closure  of  $\rightsquigarrow$.
+
+%% We call the sequence of  $\rightsquigarrow$ steps that starts with the
+%% evaluation of  an application  $s$ in a  state $\mathcal{S}$  and ends
+%% with its  value overwriting  the location of  $s$ as  an \emph{closure
+%%   evaluation subtrace (CES)} of $s$ in state $\mathcal{S}$. As an example,
+%% an $CES$ starting  with the {\sc car-clo} step may  have the form shown
+%% below.   In  this   $H(\rho(x))$  is  assumed  to   be  $\langle  s_1,
+%% \rho_1\rangle$ and $d_1$ is $\langle s_2, \rho_2\rangle$. \\
+
+%% 1:$\rho,\, (\_, \ell, \_)\!:\!S,\, H,\, (\CAR~x)
+%% \rightsquigarrow$\\
+%% \hspace*{.4cm}2:$\rho_1,\, (\rho, \rho(x), (\CAR~x))\!:\!(\_, \ell,
+%% \_)\!:\!S, \,H,\, s_1 \stackrel{+}{\rightsquigarrow}$ \\ 
+%% \hspace*{.4cm}3:$\rho,\, (\_, \ell, \_)\!:\!S,\, H'[\rho(x)=(d_1, d_2)],\, (\CAR~x)
+%%  \rightsquigarrow$\\
+%% \hspace*{.4cm}4:$\rho_2,\, (\rho, \rho(x), (\CAR~x))\!:\!(\_, \ell, \_)\!:\!S, \,H',\,
+%% s_2 \stackrel{+}{\rightsquigarrow}$\\
+%% \hspace*{.4cm}5:$\rho,\, (\_, \ell, \_)\!:\!S,\, H''[\rho(x)=(v_1, d_2)],\, (\CAR~x)
+%% \rightsquigarrow$\\ 
+%% \hspace*{.4cm}6:$\rho,\, S,\, H''[\ell=(v_1, d_2)],\, (\CAR~x)$ \\
+
+%% Notice that the $CES$ that begins  with a {\sc car-clo} is composed of
+%% four  more $CES$s  that  start at  lines 2  (evaluation  of $s_1$),  3
+%% (evaluation of $(\CAR~x)$ in a different state, 4 (evaluation of $s_2$
+%% and 5 (evaluation of $(\CAR~x)$ in a yet different state).
 
@@
                  \end{array} \\
-\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
+\mathit{ref\/}(\pi:(\NULLQ~x),\sigma,\Lfonly)
           &=&    \begin{array}{l l}
-                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
+                    \{x_{\pi}.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                     \emptyset  & \mbox{otherwise}
                  \end{array} \\
-\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
+\mathit{ref\/}(\pi:(f~\myvec{x}),\sigma,\Lfonly)
 %          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
           &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
+               \bigcup_{i=1}^n x_{i_{\pi}}.\Lf{f}{i}{\sigma}
                \end{array}
@@
   terminating with $\psi$}\\
-\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
-\mbox{where}~\mathcal{M} =
-\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
-\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
-&& \makebox[0mm]{\hspace*{7cm}
- $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
-           x_2\, \ldots\, x_n)/x]))$,} \\
- \makebox[0mm]{\hspace*{8cm} where
-     $x_1\, \ldots\, x_n$ are the variables in $s$}
+%% \mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+%%         \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
+%% \mbox{where}~\mathcal{M} =
+%% \mathcal{L}(e,\sigma,\Lfonly),~\Lv =
+%% \mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
+%% && \makebox[0mm]{\hspace*{7cm}
+%%  $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
+%%            x_2\, \ldots\, x_n)/x]))$,} \\
+%%  \makebox[0mm]{\hspace*{8cm} where
+%%      $x_1\, \ldots\, x_n$ are the variables in $s$}
+\mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
+        \lambda\mathit{\epath}.\; 
+\mathit{ref\/}(s,\sigma,\Lfonly) \cup \Lv\\
+&&\mbox{where}~ \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly), ~\Lv =
+\mathcal{M}(\epath),~\mbox{and}~\sigma = \bigcup_{{\pi}\in \epath} \Lv(x_{\pi})
 \end{eqnarray*}
@@
 
-We shall now  outline a proof of correctness of  the liveness analysis
-presented   in    Section~\ref{sec:liveness-analysis}.    This   proof
-technique  is  similar to  the  technique  used in  \cite{asati14lgc}.
-
-The  idea behind  the  proof is  as follows:  We  modify the  standard
-semantics to a semantics called  {\em minefield semantics} that models
-a liveness based  garbage collection at each  transition.  Before each
-transition,  we start  from the  root-set and  insert a  special value
-$\bot$ at each  location in the heap that contains  a reference and is
-reachable but not live. The proof is based on a context-sensitive form
-of    liveness   of    which   our    calculated   liveness    is   an
-over-approximation. The state of the transition system is augmented to
-carry enough information to calculate  the liveness environment at the
-current  evaluation  context  and  each of  the  suspended  evaluation
-contexts on  the stack.   To simulate  garbage collection, before each
-transition,  we start  from the  root-set and  insert a  special value
-$\bot$ at each  location in the heap that contains  a reference and is
-reachable but  not live by  our analysis.  Any attempt  to dereference
-such locations during the transition will result in entering a special
-state denoted \bang.  The main proof is to show that no program enters
-the \bang\ state in the minefield semantics.
-
-\newcommand{\cred}[1]{\psframebox[linestyle=none,
-    fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
-
-\comment{
-\begin{enumerate}
-\item Do you need to carry the function body?
-\item Minefield doesn't insert bottom inside closures. Separate rule to
-be inserted.
-\item Examine Proposition 4.1, See that different cases are disjoint
-  and together cover all possible cases
-\item Suppose the statement that we want to prove is this: Every step
-  (step = GC + $\rightsquigarrow$) goes without a bang. How to divide
-  a (n+1)-step transition into 1 and n. What is the exact invariant
-  required before each step. 
-\end{enumerate}
-
-}
-
-\section{Minefield semantics for the language}
-\begin{figure*}[ht!]
-\begin{center}\footnotesize
-\begin{tabular}{|c|c|c|}
-\hline
-Premise & Transition & Rule name \\ 
-\hline
-\hline
-          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
-  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
-    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
-\\
-\hline
-          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
-    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
-  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
-\sc{car-clo}
-\\
-\hline
-$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
-  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
-    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
-\sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
-\sc{prim-1-clo} \\
-\hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
-  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
-\sc{prim-2-clo} \\
-\hline
-{$\mathit{g}~\mbox{defined as}$
-$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
-  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
-    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
-\sc{funcall} \\
-\hline
-$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e, \cred{\sigma, e_f}$} &
-\sc{let} \\
-\hline
-$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
-\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
-\cred{\clazy\sigma, e_f}$}
-&
-\sc{if-clo} \\
-\hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
-  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
-  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
-  \cred{\sigma', e_f'}$} &
-\sc{return-whnf}\\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
-  \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
-  \cred{\sigma, e_f}$} &
-\sc{return-clo} \\
-\hline
-\end{tabular}
-\caption{Minefield semantics.\label{fig:minefield-semantics}}
-\end{center}
-\vskip -5mm
-\end{figure*}
-
-To set up the minefield semantics, we follow these steps:
-\begin{enumerate}
-\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
-  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
-  on the expression $e$, and $e_f$ is the body of the function. As the
-  section on liveness  analysis shows, the entire  function body $e_f$
-  and $\sigma$  are needed  to calculate  the liveness  environment at
-  different program  points. The  demand sigma  is ``dynamic''  in the
-  sense  that it  arises from  the actual  function calls  made during
-  execution.   The  0-CFA  demand  is an  over-approximation  of  this
-  demand.  The information  in a suspended evaluation  context is also
-  similarly augmented with  the demand on it and the  function body of
-  which the suspended evaluation is a  part.  Thus a stacked entry now
-  takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
-  small step semantics which carries the extra information is shown in
-  Figure~\ref{fig:minefield-semantics}.
-\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
-  e_f$ and $\mathcal{LF}$, we can construct a liveness environment as follows:
-  \begin{enumerate}
-  \item If $e$ is an expression, then:  
-  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
-    \text{where }
-  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
-\item If $e$ is an application, then:  
-  $$ \Lv = ref(e, \sigma, \Lfonly)$$
-  \end{enumerate}
-We can define in the same way the liveness environment for each of the
-suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
-environments that we shall denote $\mathsf{SL}$.
-\item $GC$ models a liveness based garbage collection:
-$GC(\rho,$ $ S,$ $ H,$ $ \mathcal{L},$ $ \mathsf{SL}) = (\rho', S', H')$ 
-  \begin{enumerate}
-  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
-    \notin \Lv$.
-\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
-  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
-  \in domain(\rho)$,  $\rho'(x)=\bot$
-  iff $x.\epsilon \notin \Lv'$.
-\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
-  $x$ in either the current environment or one of the stacked
-  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
-  \ell$  and $x.\alpha \in \Lv$, .
-  \end{enumerate}
-\end{enumerate}
-
-Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
-(\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
-preceded by  a garbage collection  using $GC$.  We consider  a garbage
-collection followed  by a transition as  a step. We show  by induction
-that, for programs  without function calls, starting  from the initial
-state that  has $e_{\main}$  as the first  expression to  be evaluated
-instead  of $\main$,  any transition  of  $n$ steps  occurs without  a
-\bang.
-
-Observe  that  the  transitions  that dereference  and  therefore  can
-potentially lead to \bang\ are those  which have a $H(\ldots)$ term in
-the  transition  rule.   Of  these,  terms  of  the  form  $H(\ldots)$
-dereference a root  variable, and the term $H(l)$  dereferences a heap
-location.  In  both the  cases we show  that the  dereferencing cannot
-lead to a \bang\ state in programs without function calls.
-
-We  need the  following proposition that  connects  liveness analysis  with
-minefield semantics transitions.
-
-\begin{proposition}
-\label{prop:correctness}
-\begin{enumerate} 
-\item   Consider  a   \LET\  definition:   \linebreak  $\pi\!\!\!:   (\LET~x
-  \leftarrow~\ldots~\IN~e)$, such  that there is  a use of $x$  in $e$,
-  i.e.  an occurrence of $\LET~y~\leftarrow~s~in \ldots$, where $s$ is
-  $(\CAR\ x)$, $(\CDR\ x)$  or $(+\ x\ y)$, or $(\SIF\  x\ e1\ e2)$ or
-  $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null.
-\item    Assume    that    a   \LET\    definition    $\pi:    (\LET~z
-  \leftarrow~(\CONS~x~y)~\IN~e)$  has an  occurrence  of $(\CAR~z)$  in
-  $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
-  Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of $y$  at
-  $\pi$ includes $\epsilon$.
-\end{enumerate}
-\end{proposition}
-
-In  the second  case for  example,  the liveness  of $x$  due to  the
-\CONS\ is  $\bcar\sigma$, where $\sigma$  is the demand on  $z$. Since
-$\sigma$  includes  the  demand  $\acar\alpha$   due  to  its  use  in
-$(\CAR~z)$, the liveness on $x$  includes $\epsilon$. We thus have the
-following lemma:
-
-\begin{lemma}\label{lemma:no-calls-correctness} 
-A program without function calls cannot  enter the \bang\ state in the
-minefield semantics.
-\end{lemma}
-
-\begin{proof}
-For a program  executing according to {\em minefield  semantics}, if a
-\bang\ occurs then some variable which is  dereferenced must be
-set to  $\bot$, between its  creation and  use points.  This  can only
-happen if  the liveness of  the variable being dereferenced  was null.
-From  the Proposition~\ref{prop:correctness}  it  is clear  that this  cannot
-happen.
-\end{proof}
-
-For   a    program   with   functions,   we    use   the   techniques
-in~\cite{asati14lgc}. We  assume as  a contradiction that  the program
-can enter a \bang\ state  and inline the  functions called to  obtain a
-program that is without functions  but has the same execution behavior
-and    demands    as   the    original    program.     We   then    use
-Lemma~\ref{lemma:no-calls-correctness}  to  show  that  the  resulting
-functionless program cannot enter a bang state.
+\warning{The (hopefully fool proof) full proof is now being developed
+  in the Appendix.}
+
+%% We shall now  outline a proof of correctness of  the liveness analysis
+%% presented   in    Section~\ref{sec:liveness-analysis}.    This   proof
+%% technique  is  similar to  the  technique  used in  \cite{asati14lgc}.
+
+%% The  idea behind  the  proof is  as follows:  We  modify the  standard
+%% semantics to a semantics called  {\em minefield semantics} that models
+%% a liveness based  garbage collection at each  transition.  Before each
+%% transition,  we start  from the  root-set and  insert a  special value
+%% $\bot$ at each  location in the heap that contains  a reference and is
+%% reachable but not live. The proof is based on a context-sensitive form
+%% of    liveness   of    which   our    calculated   liveness    is   an
+%% over-approximation. The state of the transition system is augmented to
+%% carry enough information to calculate  the liveness environment at the
+%% current  evaluation  context  and  each of  the  suspended  evaluation
+%% contexts on  the stack.   To simulate  garbage collection, before each
+%% transition,  we start  from the  root-set and  insert a  special value
+%% $\bot$ at each  location in the heap that contains  a reference and is
+%% reachable but  not live by  our analysis.  Any attempt  to dereference
+%% such locations during the transition will result in entering a special
+%% state denoted \bang.  The main proof is to show that no program enters
+%% the \bang\ state in the minefield semantics.
+
+%% \newcommand{\cred}[1]{\psframebox[linestyle=none,
+%%     fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
+
+%% \comment{
+%% \begin{enumerate}
+%% \item Do you need to carry the function body?
+%% \item Minefield doesn't insert bottom inside closures. Separate rule to
+%% be inserted.
+%% \item Examine Proposition 4.1, See that different cases are disjoint
+%%   and together cover all possible cases
+%% \item Suppose the statement that we want to prove is this: Every step
+%%   (step = GC + $\rightsquigarrow$) goes without a bang. How to divide
+%%   a (n+1)-step transition into 1 and n. What is the exact invariant
+%%   required before each step. 
+%% \end{enumerate}
+
+%% }
+
+%% \section{Minefield semantics for the language}
+%% \begin{figure*}[ht!]
+%% \begin{center}\footnotesize
+%% \begin{tabular}{|c|c|c|}
+%% \hline
+%% Premise & Transition & Rule name \\ 
+%% \hline
+%% \hline
+%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
+%%   H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
+%%     \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
+%% \\
+%% \hline
+%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
+%%     e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
+%%   \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
+%% \hline
+%% $H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
+%%   \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
+%%   \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
+%% \sc{car-whnf} \\
+%% \hline
+%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+%%   H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
+%%   (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
+%% \sc{car-clo}
+%% \\
+%% \hline
+%% $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
+%%   (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
+%%     := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
+%% \sc{prim-whnf} \\
+%% \hline
+%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+%%   H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
+%%   (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
+%% \sc{prim-1-clo} \\
+%% \hline
+%% $H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+%%   S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
+%%   (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
+%% \sc{prim-2-clo} \\
+%% \hline
+%% {$\mathit{g}~\mbox{defined as}$
+%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
+%%   (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
+%%     \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
+%% \sc{funcall} \\
+%% \hline
+%% $\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
+%%   s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+%%   \ell]\rangle], e, \cred{\sigma, e_f}$} &
+%% \sc{let} \\
+%% \hline
+%% $H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+%%   \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
+%% \hline
+%% $H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+%%   \cred{\sigma, e_f}$ \\   $\rightsquigarrow
+%% \rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
+%% \hline
+%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
+%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
+%% \cred{\clazy\sigma, e_f}$}
+%% &
+%% \sc{if-clo} \\
+%% \hline
+%% {$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
+%%   (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
+%%   e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
+%%   \cred{\sigma', e_f'}$} &
+%% \sc{return-whnf}\\
+%% \hline
+%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+%%   (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
+%%   \rightsquigarrow$
+%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
+%%   \cred{\sigma, e_f}$} &
+%% \sc{return-clo} \\
+%% \hline
+%% \end{tabular}
+%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
+%% \end{center}
+%% \vskip -5mm
+%% \end{figure*}
+
+%% To set up the minefield semantics, we follow these steps:
+%% \begin{enumerate}
+%% \item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
+%%   $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
+%%   on the expression $e$, and $e_f$ is the body of the function. As the
+%%   section on liveness  analysis shows, the entire  function body $e_f$
+%%   and $\sigma$  are needed  to calculate  the liveness  environment at
+%%   different program  points. The  demand sigma  is ``dynamic''  in the
+%%   sense  that it  arises from  the actual  function calls  made during
+%%   execution.   The  0-CFA  demand  is an  over-approximation  of  this
+%%   demand.  The information  in a suspended evaluation  context is also
+%%   similarly augmented with  the demand on it and the  function body of
+%%   which the suspended evaluation is a  part.  Thus a stacked entry now
+%%   takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
+%%   small step semantics which carries the extra information is shown in
+%%   Figure~\ref{fig:minefield-semantics}.
+%% \item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
+%%   e_f$ and $\mathcal{LF}$, we can construct a liveness environment as follows:
+%%   \begin{enumerate}
+%%   \item If $e$ is an expression, then:  
+%%   $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
+%%     \text{where }
+%%   \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
+%% \item If $e$ is an application, then:  
+%%   $$ \Lv = ref(e, \sigma, \Lfonly)$$
+%%   \end{enumerate}
+%% We can define in the same way the liveness environment for each of the
+%% suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
+%% environments that we shall denote $\mathsf{SL}$.
+%% \item $GC$ models a liveness based garbage collection:
+%% $GC(\rho,$ $ S,$ $ H,$ $ \mathcal{L},$ $ \mathsf{SL}) = (\rho', S', H')$ 
+%%   \begin{enumerate}
+%%   \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
+%%     \notin \Lv$.
+%% \item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
+%%   the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
+%%   \in domain(\rho)$,  $\rho'(x)=\bot$
+%%   iff $x.\epsilon \notin \Lv'$.
+%% \item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
+%%   $x$ in either the current environment or one of the stacked
+%%   environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
+%%   \ell$  and $x.\alpha \in \Lv$, .
+%%   \end{enumerate}
+%% \end{enumerate}
+
+%% Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
+%% (\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
+%% preceded by  a garbage collection  using $GC$.  We consider  a garbage
+%% collection followed  by a transition as  a step. We show  by induction
+%% that, for programs  without function calls, starting  from the initial
+%% state that  has $e_{\main}$  as the first  expression to  be evaluated
+%% instead  of $\main$,  any transition  of  $n$ steps  occurs without  a
+%% \bang.
+
+%% Observe  that  the  transitions  that dereference  and  therefore  can
+%% potentially lead to \bang\ are those  which have a $H(\ldots)$ term in
+%% the  transition  rule.   Of  these,  terms  of  the  form  $H(\ldots)$
+%% dereference a root  variable, and the term $H(l)$  dereferences a heap
+%% location.  In  both the  cases we show  that the  dereferencing cannot
+%% lead to a \bang\ state in programs without function calls.
+
+%% We  need the  following proposition that  connects  liveness analysis  with
+%% minefield semantics transitions.
+
+%% \begin{proposition}
+%% \label{prop:correctness}
+%% \begin{enumerate} 
+%% \item   Consider  a   \LET\  definition:   \linebreak  $\pi\!\!\!:   (\LET~x
+%%   \leftarrow~\ldots~\IN~e)$, such  that there is  a use of $x$  in $e$,
+%%   i.e.  an occurrence of $\LET~y~\leftarrow~s~in \ldots$, where $s$ is
+%%   $(\CAR\ x)$, $(\CDR\ x)$  or $(+\ x\ y)$, or $(\SIF\  x\ e1\ e2)$ or
+%%   $(\SRETURN\ x)$. Then the demand on $x$ at $\pi$ is non-null.
+%% \item    Assume    that    a   \LET\    definition    $\pi:    (\LET~z
+%%   \leftarrow~(\CONS~x~y)~\IN~e)$  has an  occurrence  of $(\CAR~z)$  in
+%%   $e$.  Then  the  liveness  of  $x$  at  $\pi$  includes  $\epsilon$.
+%%   Similarly  if $e$  contains $(\CDR~z)$,  then the  liveness of $y$  at
+%%   $\pi$ includes $\epsilon$.
+%% \end{enumerate}
+%% \end{proposition}
+
+%% In  the second  case for  example,  the liveness  of $x$  due to  the
+%% \CONS\ is  $\bcar\sigma$, where $\sigma$  is the demand on  $z$. Since
+%% $\sigma$  includes  the  demand  $\acar\alpha$   due  to  its  use  in
+%% $(\CAR~z)$, the liveness on $x$  includes $\epsilon$. We thus have the
+%% following lemma:
+
+%% \comment{
+%% \begin{lemma}\label{lemma:no-calls-correctness} 
+%% A program without function calls cannot  enter the \bang\ state in the
+%% minefield semantics.
+%% \end{lemma}
+
+%% \begin{proof}
+%% For a program  executing according to {\em minefield  semantics}, if a
+%% \bang\ occurs then some variable which is  dereferenced must be
+%% set to  $\bot$, between its  creation and  use points.  This  can only
+%% happen if  the liveness of  the variable being dereferenced  was null.
+%% From  the Proposition~\ref{prop:correctness}  it  is clear  that this  cannot
+%% happen.
+%% \end{proof}
+%% }
+%% For   a    program   with   functions,   we    use   the   techniques
+%% in~\cite{asati14lgc}. We  assume as  a contradiction that  the program
+%% can enter a \bang\ state  and inline the  functions called to  obtain a
+%% program that is without functions  but has the same execution behavior
+%% and    demands    as   the    original    program.     We   then    use
+%% Lemma~\ref{lemma:no-calls-correctness}  to  show  that  the  resulting
+%% functionless program cannot enter a bang state.
 
@@
 \newcommand{\hgt}{2.9cm}
-\begin{figure*}[p]
+\begin{figure*}[t]
 \renewcommand{\arraystretch}{.1}
 \begin{tabular}{@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
-   & {\epsfig{file=lambda_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+%%    \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
+%%    & {\epsfig{file=lambda_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+   \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+\\   \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+\\   \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+\\  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
-\\ (a) & (b)
+%\\ (a) & (b)
 \end{tabular}%\vskip -10mm
- \caption{Memory usage.  Column (a) shows complete  usage, column
-   (b) enlarges  a part  of the  usage.  The blue  and the  red curves
+ \caption{Memory usage.  
+%% Column (a) shows complete  usage, column 
+%%    (b) enlarges  a part  of the  usage.  
+The blue  and the  red curves
    indicate the number of cons cells  in the active semi-space for RGC
@@
 %% collector to reduce the cost of this.
-We extended the liveness based garbage collection to lazy
-  languages and shown its benefit  for garbage collection in practical
-  programs. We  defined a context  sensitive liveness analysis  that uses
-  context independent summaries of functions obtained using a symbolic
+We extended  the liveness based  garbage collection to  lazy languages
+and shown its benefit for garbage collection in practical programs. We
+defined  a  context  sensitive  liveness analysis  that  uses  context
+independent  summaries   of  functions   obtained  using   a  symbolic
 demand. We showed  that obtaining precise solution  to these equations
-  is undecidable in  our formulation, and hence  we safely approximate
-  the result using DFA. These DFA are consulted by the garbage collector
-  to improve  collection.
+is undecidable in our formulation, and hence we safely approximate the
+result using DFA. These DFA are  consulted by the garbage collector to
+improve collection.
 
@@
 strategy---reachability   based  collection   of  closure   arguments,
-  liveness  based  collection  of   everything  else---works  best  in
-  practice as it  avoids runtime and space  overheads.  With mixed-mode
-  garbage collection scheme,  we were able to collect  more cells than
+liveness based collection of  everything else---works best in practice
+as it  avoids runtime  and space  overheads.  With  mixed-mode garbage
+collection  scheme,   we  were  able   to  collect  more   cells  than
 reachability based collectors at a reasonable overhead.
@@
 
-%\balance
 \bibliography{fun_hra}{}
 \bibliographystyle{abbrv}
+
+\newpage
+
+\section{Appendix: Soundness of liveness analysis}
+
+\begin{figure*}[t!]
+\begin{center}\footnotesize
+\renewcommand{\arraystretch}{1.5}
+
+\begin{tabular}{|c|c|c|}
+\hline
+Premise & Transition & Rule name \\ 
+\hline
+
+\hline
+\makecell[t]{$GC(\rho_1, S, H_1, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
+  H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
+\sc{car-clo-bang}
+\\
+\hline
+\makecell[t]{$GC(\rho_1, S, H_1, \Lv, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x))
+\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
+  H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
+\sc{car-clo}\\
+\hline
+\end{tabular}
+
+\medskip
+
+\begin{tabular}{|c|c|c|}
+\hline
+Premise & Transition & Rule name \\ 
+\hline
+
+\hline
+
+          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
+  H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
+    \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
+\\
+\hline
+          & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
+  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+\hline
+$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
+  \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
+  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
+\sc{car-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+  H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
+\sc{car-clo}
+\\
+
+\hline
+$H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
+(\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
+s, \rho,\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, H,\, s, \, \cred{\sigma}$      &
+\sc{car-1-clo} \\
+
+\hline
+$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
+ & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H,
+  (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, H[\rho'(z)
+    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma'}$}      &
+\sc{prim-whnf} \\
+\hline
+$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+  H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+  (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+\sc{prim-1-clo} \\
+\hline
+$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+  S, H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
+  (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+\sc{prim-2-clo} \\
+\hline
+{$\mathit{g}~\mbox{defined as}$
+$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
+  (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
+    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma}$}      &
+\sc{funcall} \\
+\hline
+$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+  \ell]\rangle], e, \cred{\sigma}$} &
+\sc{let} \\ 
+\hline
+$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \cred{\sigma}$  $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma}$} & \sc{if-true} \\
+\hline
+$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+  \cred{\sigma}$   $\rightsquigarrow
+\rho, S, H,  e_2, \cred{\sigma}$} & \sc{if-false} \\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
+\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, H, s,
+\cred{\clazy\sigma}$}
+&
+\sc{if-clo} \\
+\hline
+{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
+  (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
+  \cred{\sigma'}$} &
+\sc{return-whnf}\\
+\hline
+$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+  (\psi:\SRETURN~x), \cred{\sigma}$  $
+  \rightsquigarrow$
+$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, H,  s,
+  \cred{\sigma}$} &
+\sc{return-clo} \\
+\hline
+\end{tabular}
+\caption{Minefield semantics.\label{fig:minefield-semantics}}
+\end{center}
+\vskip -5mm
+\end{figure*}
+
+We shall now  present a proof of soundness of  the liveness analysis
+presented in Section~\ref{sec:liveness-analysis}.   While the proof is
+on the broad lines of \cite{asati14lgc}, there are several differences
+due to  the lazy  semantics of  the language  being 
+analyzed in this paper. Here is the idea behind the proof.
+\begin{enumerate}
+\item       We        enrich       the        standard       semantics
+  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
+  collection  just  before  each  $\rightsquigarrow$  transition.  The
+  simulated garbage collection starts from  the root-set and inserts a
+  special value  $\bot$ at each location  in the heap that  contains a
+  reference that is reachable but not live. Any attempt to dereference
+  such  locations during  the  transition will  result  in entering  a
+  special state denoted \bang. We call the semantics after
+  augmentation  \emph{minefield semantics}.
+\item \label{inline} Assuming that a  program enters the \bang\ state,
+  we  show  how to  construct,  through  inline expansion,  a  program
+  without function calls which has the same minefield behavior.
+\item  The final  part  of the  proof shows  that  no program  without
+  function calls can enter a \bang\ state. As a consequence no program
+  (with or without function calls) can enter the \bang\ state. 
+\end{enumerate}
+
+\subsection{Minefield semantics}
+
+To set up the minefield semantics, we follow these steps:
+\begin{enumerate}
+\item  We  enrich the  abstract  machine  state $\rho,\,S,\,H,\,e$  to
+  $\rho,\,S,\,H,\,e,\,\sigma$,  where $\sigma$  is the  demand on  the
+  expression $e$. We call such  a state a \emph{minefield state}.  The
+  demand $\sigma$  is ``dynamic''  in that it  arises from  the actual
+  sequence of function  calls that led to the evaluation  of $e$.  The
+  0-CFA   demand   used  during   liveness   analysis   is  a   static
+  over-approximation  of $\sigma$  that  considers \emph{all  possible
+    calls} that could lead to  the evaluation of $e$.  The information
+  in  a  suspended evaluation  context  (on  the  stack $S$)  is  also
+  similarly augmented with its demand.  Thus a stacked entry now takes
+  the form $(\rho,  x, e, \sigma)$.  A modification of  the small step
+  semantics which carries the extra information is shown in the bottom
+  table of Figure~\ref{fig:minefield-semantics}.
+\item $GC$ models a liveness  based garbage collection $GC(\rho, S, H,
+  \sigma)$ returns  $(\rho', S',  H')$, where the  changes are  due to
+  certain references  being replaced  by $\bot$.  This  represents the
+  act of  not retaining  (effectively garbage collecting)  the objects
+  pointed   to  by   these   references  during   an  actual   garbage
+  collection. This is done in the following way:
+  \begin{enumerate}
+  \item Given the  current evaluation context $\rho, S,  H, e, \sigma$
+    and the demand transformers $\Lfonly$, we can construct a liveness
+    environment  \Lv.   If  $e$  is   an  expression,  then  $  \Lv  =
+    \mathcal{L}(e, \sigma,  \Lfonly)$, else if $e$  is an application,
+    then $  \Lv = ref(e,  \sigma, \Lfonly)$.  We can  similarly define
+    the  liveness environment  for  each of  the suspended  evaluation
+    contexts in $S$,  giving a stack of liveness  environments that we
+    shall denote $\mathsf{SL}$.
+  \item \emph{Garbage collection of objects pointed by root set:} For each $x \in domain(\rho)$, $\rho'(x)=\bot$ iff $x.\epsilon
+    \notin \Lv$. Similarly, for  each stack entry $(\rho,\_,\_,\_)$ in
+    $S$  with  $\Lv'$ as  the  corresponding  liveness environment  in
+    $\mathsf{SL}$, and for each  $x \in domain(\rho)$, $\rho'(x)=\bot$
+    iff $x.\epsilon \notin \Lv'$. 
+  \item \emph{Garbage collection of objects in the heap:} For each location $\ell$, $H'(\ell) = \bot$ iff  there is no
+  $x$ in either the current environment or one of the stacked
+  environments $\rho$, such that for some $\alpha$ corresponding to a
+  closure-free path in the heap starting from $x$,  $H[x.\alpha] =
+  \ell$  and $x.\alpha \in \Lv$. Here \Lv\ is the liveness of the
+  variables in the  current environment or the stacked environment.
+
+\warning{We have to tell the reader about closure free paths earlier
+  on. Also, has  $H[x.\alpha]$ been defined?}
+  \end{enumerate}
+Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
+(\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
+preceded  by a  garbage collection  using  $GC$.  The  details of  the
+transition for  the {\sc car-clo}  rule is shown  in the top  table of
+Figure~\ref{fig:minefield-semantics}. If  this or  an earlier  call to
+$GC$  results  in  $H(\rho(x))$  being   bound  to  $\bot$,  then  the
+$\rightsquigarrow$    step    enters    the    \bang\    state    {\sc
+  car-clo-bang}. Otherwise the  transition is the same  as the earlier
+{\sc car-clo} rule.
+\end{enumerate}
+
+\subsection{Inlining}
+
+Consider a call  tree, each of whose nodes represents  a function that
+was  called (but  did  not  necessarily return  because  of a  \bang).
+Assume that each of  the nodes of the tree is  also annotated with the
+program point where the corresponding call was invoked.  This tree can
+be  used to  inline function  calls  in a  hierarchical fashion.   The
+details of  inlining a  single function  call $(\LET\,\,  x \leftarrow
+(f\,y_1\,\ldots\,y_n)  \,\, \IN\,\,  e)$ is  as follows.   Let $f$  be
+defined (after  renaming its  formals and locals  to be  disjoint from
+existing variables)  by $(\DEFINE\ (f\  z_1\ \ldots\ z_n)\  e_f)$. The
+call is replaced  by a sequence of $\LET$s of  the form $z_i\leftarrow
+(\ID\,\,y_i)$   followed   by   the    body   $e_f$   but   with   its
+$(\SRETURN\,\,w)$  expressions  replaced  by $(\LET\,\,  x  \leftarrow
+(\ID\,\, w) \,\, \IN\,\, e)$.  We  prefer to use the identity function
+$\ID$ as a form of no-op rather than introducing the form $(\LET\,\, x
+\leftarrow w \,\,  \IN\,\, e)$ where the RHS  of the $\LET$-definition
+is a  simple variable.  For  a call-less program, the initial  state of
+the minefield semantics is assumed to be  $([\;]_\rho,([\;]_\rho, \ans,
+(\print~\ans)):[\;]_{S} , [\;]_{H}, (e_{\mainpgm})$.
+
+\subsection{Soundness Results}
+
+  We  consider a  garbage collection  followed  by a  transition as  a
+  step.  We show  by  induction that,  for  programs without  function
+  calls, starting from  the initial state that has  $e_{\main}$ as the
+  first expression to be evaluated  instead of $\main$, any transition
+  of $n$ steps occurs without a \bang.
+
+We first need an auxiliary  result about minefield semantics. Consider
+a trace of  a minefield execution.  For every  minefield state $(\rho,
+S, H, e(s),  \sigma)$ that appears on the LHS  of a $\rightsquigarrow$
+step, the demand  $\sigma$ on the expression $e$  (or application $s$)
+is non-null.   This can  be proved  by an induction  on the  number of
+steps leading to the minefield state.  The base step holds because the
+demand $\sigma_{all}$  on $\mainpgm$ is non-null.   The inductive step
+follows  from the  observation that  for  each step  of the  minefield
+semantics, if  the demand $\sigma$ on  the LHS of a  minefield step is
+non-null, the demand  on the RHS is a transformation  of $\sigma$ (for
+example $\clazy\sigma$) which is also non-null.
+
+Let us clarify this observation:  Since the minefield semantics models
+lazy  evaluation,  parts  of  the  program may  not  be  evaluated  at
+all.  However,  if  an  expression  (or  application)  happens  to  be
+evaluated, it  will appear  on the  LHS of a  minefield step,  and the
+context-sensitive demand on it will be non-null.
+
+We  show next  that the  minefield  execution of  a call-less  program
+cannot go \bang. Note that, for  the sake of concreteness, by a single
+minefield  execution we  shall  mean the  evaluation  of $(main)$  (or
+$e_{\mainpgm}$,  for call-less  programs) to  its WHNF  driven by  the
+printing mechanism  represented by \print. With  minor variations, the
+proof  will also  be applicable  for subsequent  evaluations fired  by
+$\print$.
+
+\begin{lemma}
+\label{lemma:call-less-cannot-go-bang}
+Consider the minefield execution of  a program without function calls.
+Such a program cannot enter the \bang\ state.
+\end{lemma}
+\begin{proof}
+Consider a state in the minefield execution of a program $(\rho, S, H,
+e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
+steps  leading to  this state  that  the next  step cannot  lead to  a
+\bang\ state.
+
+When $n$ is 0, the state is $([\;]_\rho,\_, [\;]_{S} ,  [\;]_{H},
+e_{\mainpgm}, \sigma_{all})$. Further, we have to show that a $GC(..)$
+followed by a $\rightsquigarrow$ step cannot lead to a \bang\ state.
+Now, because our programs are in  ANF, $e_{\mainpgm}$ can only be a 
+$\LET$. A {\sc let} step  does not  dereference anything
+and thus cannot  result in a \bang.
+
+For the inductive step, we shall show that none of the minefield steps
+that involves  dereferencing results in  a \bang. These are  the steps
+which have a  $H(\rho(...))$ in the premise.  Recall  that a minefield
+step  consists of  a GC  step followed  by a  $\rightsquigarrow$.  The
+$\rightsquigarrow$  can go  \bang\  because it  dereferences a  $\bot$
+inserted  by the  immediately preceding  GC or  the GC  of an  earlier
+minefield  step. \warning{However  the  demand $\sigma'$  on basis  of
+  which  an  earlier GC  would  have  inserted  a $\bot$,  would  have
+  included the current demand $\sigma$.}   Thus, if the earlier GC had
+inserted a  $\bot$, so  would the  current GC.  Thus  it is  enough to
+consider the  $\rightsquigarrow$ along  with the  current GC  and show
+that it does not lead to a \bang.
+
+We consider the steps {\sc car-clo} and {\sc car-1-clo} only. The rest
+of the  steps are  similar. For  the {\sc car-clo}  step in  the state
+$\rho,   S,  H,   (\CAR~x),  \sigma$,   we  know   that  $\sigma$   is
+non-null. There for the liveness of $x$ includes $\epsilon$, and the
+dereferencing $H(\rho(x))$ will go without \bang.
+
+For   the  {\sc   car-1-clo}  step,   observe  that   there  are   two
+dereferences. First  $x$ is dereferenced to  get a cons cell  and then
+the head of the cons cell is dereferenced to obtain a closure.  If the
+demand $\sigma$  on $(\CAR~x)$ is  non-null, then the liveness  of $x$
+will include both  $\epsilon$ and $\acar\epsilon$, and a  GC with this
+liveness will  neither bind $x$ to  a $\bot$, nor insert  $\bot$ at the
+first component of  the cons cell. Thus both  dereferences will happen
+normally and the $\rightsquigarrow$ step will not enter \bang.
+\end{proof}
+
+Now we are ready to prove the main soundness result.
+
+\begin{theorem}
+The minefield execution of no program can enter a \bang\ state. 
+\end{theorem}
+
+\begin{proof}
+Assume to the contrary that a  program $P$ enters the \bang\ state. We
+can transform $P$ to a call-less  program $P'$ such that the minefield
+executions of $P$ and $P'$ are identical except for change of variable
+names. \warning{Check whether the  previous line is correct.} However,
+by  Lemma~\ref{lemma:call-less-cannot-go-bang}  we   know  that   $P'$,  a
+call-less program,  cannot enter the  \bang\ state. Therefor  $P$ also
+cannot enter the \bang\ state.
+\end{proof}
+
 \end{document}

### 2015-11-27 lazy-scheme/post_process/gen_stats_and_plotall.sh
Added example-program.tex file

@@ -114,17 +114,17 @@ fi
 for bm in $BMs
 do
     #for opt in gc-plain gc-live gc-freq=${freq[$bm]}
-    for opt in gc-plain gc-live
-    do 
-      	echo "Processing $bm [$opt]"
-      	$DIR/gen_gc_stats.sh $OPTARG \
-      	    -bmdir ../benchmarks/programs/$bm \
-      	    -gcopt $opt -heap ${limit[$bm]}
-     done
-    #bash $DIR/val.plot $bm ${xtics[$bm]} ${ytics[$bm]}
-    #bash $DIR/window.plot $bm ${xlimit[$bm]} ${ylimit[$bm]} 
-    # gnuplot $DIR/plots/${bm}.gnu
-    # gnuplot $DIR/plots/${bm}_win.gnu
+    # for opt in gc-plain gc-live
+    # do 
+    #   	echo "Processing $bm [$opt]"
+    #   	$DIR/gen_gc_stats.sh $OPTARG \
+    #   	    -bmdir ../benchmarks/programs/$bm \
+    #   	    -gcopt $opt -heap ${limit[$bm]}
+    #  done
+    bash $DIR/val.plot $bm ${xtics[$bm]} ${ytics[$bm]}
+    bash $DIR/window.plot $bm ${xlimit[$bm]} ${ylimit[$bm]} 
+    gnuplot $DIR/plots/${bm}.gnu
+    gnuplot $DIR/plots/${bm}_win.gnu
 
     echo "$bm Done"
 done

### 2015-11-27 paper/paper.tex

@@
 We consider  the problem of reducing  the memory required to  run lazy
-first-order functional programs. We  address the problem by analyzing
+first-order functional programs. Our approach is to analyze
 programs  for liveness  of  heap-allocated data.   The  result of  the
@@
 %% finite-state automata (DFA) describing the liveness at that point.
-We
+To show the effectiveness of our method, we
 implement a copying collector that uses the results of the liveness
@@
 function  calls during  analysis.  Second,  the concrete  demand on  a
-function   body  is   obtained  through   a  0-CFA-like   conservative
-approximation  that combines  the  demands  on all  the  calls to  the
-function.   The result  of the  analysis is  an annotation  of certain
-program points  with deterministic finite-state  automata (DFA)  capturing the  liveness of
-variables  at  these points.  Depending  on  the program  point  where
-garbage  collection has  been  triggered,  the garbage collector consults  a set  of
-automata to restrict reachability during  marking.  This results in an
-increase in  the garbage reclaimed  and consequently in  fewer garbage
-collections.
+function  body is  obtained through  a conservative approximation
+similar to 0-CFA~\cite{Shivers:1988} that combines the  demands on all the calls
+to  the function.   The result  of the  analysis is  an annotation  of
+certain program points with  deterministic finite-state automata (DFA)
+capturing the liveness of variables at these points.  Depending on the
+program point where garbage collection has been triggered, the garbage
+collector consults a  set of automata to  restrict reachability during
+marking.  This  results in  an increase in  the garbage  reclaimed and
+consequently in fewer garbage collections.
 
@@
 evaluation  takes  place  cannot be  statically  determined,  laziness
-complicates  liveness  analysis  itself.    In  addition,  apart  from
+complicates  liveness  analysis  itself.    In  addition, laziness
+also complicates garbage collection, since, 
+ apart  from
 evaluated   data,  we  need to extend   liveness-based  collection   to
-closures i.e. we reclaim the space for closures  which have no future  use.
+closures.
 Since a closure can  escape the scope in which it  was created, it has
@@
 execution  progresses and  possible  future uses  are eliminated,  the
-liveness information is updated with a more precise version.  All this
+liveness information is updated with more precise versions.  All this
 requires significant  modification to  the earlier  garbage collection
@@
 \newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
+
 \subsection{Generating Equations for \Lfonly$_\mathit{f}$}
@@
 \end{eqnarray*}
+
 %% and  the only equation defining \Lfone{\length} is:
@@
 \end{eqnarray}
+
 where \Df{f}{i}  are sets of strings  over the alphabet
@@
 %% \deltacall{\length}{\pi_6}{\length}$.
+
 Filling in  the values gives:

### 2015-11-27 paper/paper.tex
fix labels

@@
   augmentation,  \emph{minefield semantics}.
-\item \label{inline} Assuming that a  program enters the \bang\ state,
+\item \label{inline1} Assuming that a  program enters the \bang\ state,
   we  show  how to  construct,  through  inline expansion,  a  program
@@
   augmentation  \emph{minefield semantics}.
-\item \label{inline} Assuming that a  program enters the \bang\ state,
+\item \label{inline2} Assuming that a  program enters the \bang\ state,
   we  show  how to  construct,  through  inline expansion,  a  program
@@
 \begin{lemma}
-\label{lemma:call-less-cannot-go-bang}
+\label{lemma:call-less-cannot-go-bang-app}
 Consider the minefield execution of  a program without function calls.
@@
   correct.}                 However,                 by
-Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
+Lemma~\ref{lemma:call-less-cannot-go-bang-app} we know that
 $P'$,   a   call-less   program,   cannot   enter   the

### 2015-11-27 paper/paper.tex
fixed minefield semantics
H and l fixed

@@
 \hline 
-          & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, \kappa
-  \rightsquigarrow \rho',\, S,\, H[\ell := \kappa],\, e$    &  \sc{const}
+          & $\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\, \kappa
+  \rightsquigarrow \rho',\, S,\, \heap[\ell := \kappa],\, e$    &  \sc{const}
 \\
 \hline
-          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, H,\, (\CONS~x~y)
+          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, \heap,\, (\CONS~x~y)
 \rightsquigarrow
-$  $\rho',\, S,\, H[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
-\hline
-
-$H(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\,
-(\CAR~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$      &
+$  $\rho',\, S,\, \heap[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
+\hline
+
+$\heap(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\,
+(\CAR~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$      &
 \sc{car-select} \\
@@
 
-$H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
-(\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho'\rangle), (\CAR~x))\!:\!S,\, H,\, s$      &
+$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
+(\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho'\rangle), (\CAR~x))\!:\!S,\, \heap,\, s$      &
 \sc{car-1-clo} \\
@@
 
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (\CAR~x)
+$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (\CAR~x)
 \rightsquigarrow
-\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,H,\, s$      &
+\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,\heap,\, s$      &
 \sc{car-clo}
@@
 
-$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & {$\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, (+~x~y)  \rightsquigarrow$
-$\rho', \,S,\, H[\ell := H(\rho'(x)) + H(\rho'(y))],\, e$}      &
+$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
+ & {$\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\, (+~x~y)  \rightsquigarrow$
+$\rho', \,S,\, \heap[\ell := \heap(\rho'(x)) + \heap(\rho'(y))],\, e$}      &
 \sc{prim-add} \\
@@
 
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (+~x~y)
+$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
 \rightsquigarrow
-\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, H,\, s$      &
+\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
 \sc{prim-1-clo} \\
 \hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, H,\, (+~x~y)
+$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
 \rightsquigarrow
-\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, H,\, s$      &
+\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
 \sc{prim-2-clo} \\
@@
 {$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, H,\,
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
 (f~\myvec{x})  \rightsquigarrow
-[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, H,\, e_{\mathit{f}}$      &
+[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
 \sc{funcall} \\
 \hline
-$\ell$ is a new location& {$\rho,\, S,\, H,\, (\LET~x\leftarrow s~\IN~e)
+$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
   \rightsquigarrow$
-$\rho\oplus[x \mapsto \ell],\, S,\, H[\ell \mapsto \langle s,
+$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell \mapsto \langle s,
     \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
@@
 \hline
-$H(\rho(x)) \ne 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho,\, S,\, H,\,  e_1$ & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho,\, S, \,H, \, e_2$ & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\,
+$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho,\, S,\, \heap,\,  e_1$ & \sc{if-true} \\
+\hline
+$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
+\rho,\, S, \,\heap, \, e_2$ & \sc{if-false} \\
+\hline
+$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
   (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, H, \, s$}
+\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
 &
@@
 \hline
-{$H(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
-e)\!:\!S,\, H,\,
-(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$ &
+{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
+e)\!:\!S,\, \heap,\,
+(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
 \sc{return-whnf}\\
 \hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\, (\SRETURN~x)
+$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
   \rightsquigarrow$
-$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, H,\,  s$} &
+$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
 \sc{return-clo} \\
@@
 $([\;]_\rho,\,   ([\;]_\rho,   \,   \rho(\ans),\,   (\print~\ans)):[\;]_{S},
-[\;]_{H},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_H$  are the
+[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_{\heap}$  are the
 empty environment and  the empty heap. The  initial stack consists
@@
 %% \hline
-%% \makecell[t]{$GC(\rho_1, S, H_1, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
-%%   H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
+%% \makecell[t]{$GC(\rho_1, S, \heap_1, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
+%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
 %% \sc{car-clo-bang}
@@
 %% \hline
-%% \makecell[t]{$GC(\rho_1, S, H_1, \Lv, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x))
+%% \makecell[t]{$GC(\rho_1, S, \heap_1, \Lv, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x))
 %% \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
-%%   H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
-%%   (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
+%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
+%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
 %% \sc{car-clo}\\
@@
 %%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
-%%   H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
+%%   \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(x) :=
 %%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
@@
 %% \hline
-%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
-%%   \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
+%%   \rho', S, \heap[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
 %% \hline
-%% $H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
-%%   \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
-%%   \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
+%% $\heap(\rho(x)) \mbox{ is } (\ell_1, \ell_2)$ & \makecell{$\rho, (\rho', z, e,
+%%   \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
+%%   \rightsquigarrow \rho', S, \heap[\rho'(z) := \heap(\ell_1)], e, \cred{\sigma'}$}      &
 %% \sc{car-whnf} \\
 %% \hline
-%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-%%   H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-%%   (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
+%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
 %% \sc{car-clo}
@@
 %% \hline
-%% $H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
+%% $\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
 %% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
-%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, H,\, s, \, \cred{\sigma}$      &
+%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
 %% \sc{car-1-clo} \\
@@
 %% \hline
-%% $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
-%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H,
-%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, H[\rho'(z)
-%%     := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma'}$}      &
+%% $\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
+%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap,
+%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, \heap[\rho'(z)
+%%     := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
 %% \sc{prim-whnf} \\
 %% \hline
-%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-%%   H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+%%   \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
 %% \sc{prim-1-clo} \\
 %% \hline
-%% $H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-%%   S, H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
-%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+%% $\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+%%   S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
+%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
 %% \sc{prim-2-clo} \\
@@
 %% {$\mathit{g}~\mbox{defined as}$
-%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
+%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, \heap,
 %%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
-%%     \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma}$}      &
+%%     \rho(\myvec{x})], S, \heap, e_{\mathit{g}}, \cred{\sigma}$}      &
 %% \sc{funcall} \\
 %% \hline
-%% $\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
-%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
+%% $\ell$ is a new location& \makecell{$\rho, S, \heap, (\LET~x\leftarrow
+%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
 %%   \ell]\rangle], e, \cred{\sigma}$} &
@@
 %% \hline
-%% $H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma}$} & \sc{if-true} \\
+%% $\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
+%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
 %% \hline
-%% $H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+%% $\heap(\rho(x)) = 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
 %%   \cred{\sigma}$   $\rightsquigarrow
-%% \rho, S, H,  e_2, \cred{\sigma}$} & \sc{if-false} \\
+%% \rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
 %% \hline
-%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
 %%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
-%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, H, s,
+%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
 %% \cred{\clazy\sigma}$}
@@
 %% \hline
-%% {$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
-%%   (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
+%% {$\heap(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
+%%   (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(z) := v], e,
 %%   \cred{\sigma'}$} &
@@
 %% \hline
-%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
 %%   (\psi:\SRETURN~x), \cred{\sigma}$  $
 %%   \rightsquigarrow$
-%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, H,  s,
+%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
 %%   \cred{\sigma}$} &
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
-  H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
 \sc{car-clo-bang} 
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x))
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
 \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
-  H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
 \sc{car-clo}\\
@@
 \item   We   enrich    the   abstract   machine   state
-  $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
+  $\rho,S,\heap,e$   to  $\rho,S,\heap,e,\sigma$. 
 We  call such a state  a \emph{minefield state}.
@@
 %%  $([\;]_\rho,([\;]_\rho,
-%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
+%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
 every transition  is preceded  by a  garbage collection
@@
 Figure~\ref{fig:minefield-semantics}.  If  this  or  an
-earlier  call to  $GC$  results  in $H(\rho(x))$  being
+earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
 bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
@@
 semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
-\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{H},
+\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
 e_{\mainpgm}$.
@@
 semantics. Consider  a trace of a  minefield execution.
-For every minefield state  $(\rho, S, H, e(s), \sigma)$
+For every minefield state  $(\rho, S, \heap, e(s), \sigma)$
 that appears  on the LHS of  a $\rightsquigarrow$ step,
@@
 Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
-step. Consider  a state  $(\rho, S,  H, e,  \sigma)$ in  the minefield
+step. Consider  a state  $(\rho, S,  \heap, e,  \sigma)$ in  the minefield
 execution of a  program.  We show by induction on  the number of steps
@@
 \bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\_, [\;]_{S} ,
-            [\;]_{H}, e_{\mainpgm}, \sigma_{all})$.  Since the call to
+            [\;]_{\heap}, e_{\mainpgm}, \sigma_{all})$.  Since the call to
             $GC(\ldots)$ in this state does  nothing, we just have to show
@@
 a   \bang.   These  are   the   steps   which  have   a
-$H(\rho(...))$ in the premise. %%  Recall that a minefield
+$\heap(\rho(...))$ in the premise. %%  Recall that a minefield
 %% step   consists   of   a   GC  step   followed   by   a
@@
 The rest  of the rules involve similar reasoning. For the {\sc
-  car-clo}  rule in  the state  $\rho, S,  H, (\CAR~x),
+  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
 \sigma$, we  know that $\sigma$ is  non-null. Therefore
 the  liveness  of  $x$  includes  $\epsilon$,  and  the
-dereferencing $H(\rho(x))$ will go without \bang.
+dereferencing $\heap(\rho(x))$ will go without \bang.
 
@@
 \hline
-
-\hline
-
-&\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
-  H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
+\hline
+&\makecell{ $\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S,
+  \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell :=
     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
@@
 \hline
-& \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
-  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
-  \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
-  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-  H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
+& \makecell{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
+  \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+\hline
+$\heap(\rho(x)) \mbox{ is } (v, d)$ & \makecell{$\rho, (\rho', \ell, e,
+  \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
+  \rightsquigarrow \rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
+\sc{car-select} \\
+\hline
+$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
+(\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
+\sc{car-1-clo} \\
+
+\hline
+
+$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+  \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
 \sc{car-clo}
@@
 \hline
-$H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
-(\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
-s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, H,\, s, \, \cred{\sigma}$      &
-\sc{car-1-clo} \\
-
-\hline
-$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
- & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H,
-  (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, H[\rho'(z)
-    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma'}$}      &
-\sc{prim-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-  H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-  (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
+ & \makecell{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
+  (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, \heap[\ell
+    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
+\sc{prim-add} \\
+\hline
+$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
+  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
 \sc{prim-1-clo} \\
 \hline
-$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-  S, H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
-  (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
+$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
+  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
+  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
 \sc{prim-2-clo} \\
 \hline
-{$\mathit{g}~\mbox{defined as}$
-$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
-  (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
-    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma}$}      &
+{$\mathit{f}~\mbox{defined as}$
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, \heap,
+  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
+    \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
 \sc{funcall} \\
 \hline
-$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-  \ell]\rangle], e, \cred{\sigma}$} &
+$\ell$ is a new location& \makecell{$\rho, S, \heap, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
 \sc{let} \\ 
 \hline
-$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma}$  $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma}$} & \sc{if-true} \\
-\hline
-$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
+$\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
+  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
+\hline
+$\heap(\rho(x)) = 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
   \cred{\sigma}$   $\rightsquigarrow
-\rho, S, H,  e_2, \cred{\sigma}$} & \sc{if-false} \\
-\hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
+\rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
+\hline
+$\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, H, s,
+\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
 \cred{\clazy\sigma}$}
@@
 \hline
-{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
-  (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
+{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& \makecell{$\rho,
+  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell := v], e, 
   \cred{\sigma'}$} &
@@
 \hline
-$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
-  (\psi:\SRETURN~x), \cred{\sigma}$  $
+$\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
+  (\SRETURN~x), \cred{\sigma}$  $
   \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, H,  s,
+$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
   \cred{\sigma}$} &
@@
 \item   We   enrich    the   abstract   machine   state
-  $\rho,\,S,\,H,\,e$   to  $\rho,\,S,\,H,\,e,\,\sigma$,
+  $\rho,\,S,\,\heap,\,e$   to  $\rho,\,S,\,\heap,\,e,\,\sigma$,
   where  $\sigma$  is  the  demand  on  the  expression
@@
 \item $GC$  models a liveness based  garbage collection
-  $GC(\rho, S,  H, \sigma)$ returns $(\rho',  S', H')$,
+  $GC(\rho, S,  \heap, \sigma)$ returns $(\rho',  S', \heap')$,
   where the changes are due to certain references being
@@
   \item Given the current  evaluation context $\rho, S,
-    H,   e,  \sigma$   and   the  demand   transformers
+    \heap,   e,  \sigma$   and   the  demand   transformers
     $\Lfonly$, we can  construct a liveness environment
@@
   \item  \emph{Garbage  collection  of objects  in  the
-    heap:} For each location  $\ell$, $H'(\ell) = \bot$
+    heap:} For each location  $\ell$, $\heap'(\ell) = \bot$
     iff  there   is  no  $x$  in   either  the  current
@@
     to a  closure-free path  in the heap  starting from
-    $x$,  $H[x.\alpha]   =  \ell$  and   $x.\alpha  \in
+    $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
     \Lv$. Here \Lv\ is the liveness of the variables in
@@
 \warning{We have to tell  the reader about closure free
-  paths  earlier  on.   Also,  has  $H[x.\alpha]$  been
+  paths  earlier  on.   Also,  has  $\heap[x.\alpha]$  been
   defined?}
@@
 Starting from the initial state $([\;]_\rho,([\;]_\rho,
-\ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
+\ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
 every transition  is preceded  by a  garbage collection
@@
 semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
-\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{H},
+\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
 (e_{\mainpgm})$.
@@
 \begin{proof}
-Consider a state in the minefield execution of a program $(\rho, S, H,
+Consider a state in the minefield execution of a program $(\rho, S, \heap,
 e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
@@
 When $n$ is 0, the  state is $([\;]_\rho,\_, [\;]_{S} ,
-[\;]_{H},  e_{\mainpgm},  \sigma_{all})$.  Further,  we
+[\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.  Further,  we
 have   to  show   that   a  $GC(..)$   followed  by   a
@@
 a   \bang.   These  are   the   steps   which  have   a
-$H(\rho(...))$ in the premise.  Recall that a minefield
+$\heap(\rho(...))$ in the premise.  Recall that a minefield
 step   consists   of   a   GC  step   followed   by   a
@@
 only. The rest  of the steps are similar.  For the {\sc
-  car-clo}  step in  the state  $\rho, S,  H, (\CAR~x),
+  car-clo}  step in  the state  $\rho, S,  \heap, (\CAR~x),
 \sigma$, we  know that $\sigma$ is  non-null. There for
 the  liveness  of  $x$  includes  $\epsilon$,  and  the
-dereferencing $H(\rho(x))$ will go without \bang.
+dereferencing $\heap(\rho(x))$ will go without \bang.
 

### 2015-11-27 lazy-scheme/post_process/minMemorySearch.sh
Working deriv benchmark

@@ -4,7 +4,7 @@ MIN=${3-0}
 MAX=$2
 FULLTEST=../benchmarks/programs/$TEST/$TEST.scm
 LOG=run.out.$$
-GCOPT=gc-plain
+GCOPT=gc-live
 
 
 RANGE=`expr $MAX - $MIN`

### 2015-11-27 paper/paper.tex
latest - 27th

--- 2015-11-27 paper/paper.tex
@@
 \usepackage{booktabs}
+\usepackage{stmaryrd}
 \usepackage{boxedminipage}
@@
 %\newcommand{\blankout}[1]{}
-
+\newcommand{\oxlb}{\ensuremath{[\!|}}
+\newcommand{\oxrb}{\ensuremath{|\!]}}
 \newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
@@
 \def\myvec{\mathaccent"017E } 
-\newcommand{\stk}{\mbox{S}}   
+\newcommand{\stk}{{S}}      
 \newcommand{\ID}{\mbox{$\mathbf{ id}$}}
@@
 
-While this idea has been shown to be effective for a
-first-order {\em  eager} language~\cite{asati14lgc}, a straightforward  extension of the
-technique   is  not   possible  for  lazy   languages,  where
-heap-allocated objects  may include closures  (runtime representations
-of unevaluated expressions). Since data  is made live by evaluation of
+While this idea has been shown  to be effective for a first-order {\em
+  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
+the technique is not possible for lazy languages, where heap-allocated
+objects may  include closures (runtime representations  of unevaluated
+expressions).   Firstly, since  data  is made  live  by evaluation  of
 closures, and  in lazy languages the  place in the program  where this
 evaluation  takes  place  cannot be  statically  determined,  laziness
-complicates  liveness  analysis  itself.    In  addition, laziness
-also complicates garbage collection, since, 
- apart  from
-evaluated   data,  we  need to extend   liveness-based  collection   to
-closures.
-Since a closure can  escape the scope in which it  was created, it has
-to carry the liveness information  of its free variables. Moreover, as
-execution  progresses and  possible  future uses  are eliminated,  the
-liveness information is updated with more precise versions.  All this
-requires significant  modification to  the earlier  garbage collection
-scheme for eager languages.
+complicates  liveness analysis  itself.  Secondly,  for liveness-based
+garbage collection to  be effective, we need  to extend liveness-based
+collection to closures apart from evaluated data.  Since a closure can
+escape the scope in which it was created, it has to carry the liveness
+information of  its free variables. Moreover,  as execution progresses
+and  possible  future uses  are  eliminated,  we update  the  liveness
+information  in a  closure with  a more  precise version.   Because of
+these reasons,  the garbage collector also  becomes significantly more
+complicated than a liveness-based collector for a eager language.
 
@@
   attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
-    asati14lgc, karkare06effectiveness} have targeted either
+    asati14lgc, karkare06effectiveness} have involved either
   imperative or  eager functional languages.
@@
 
-The  semantics  of   expressions  (and  applications\footnote{In  most
-  contexts, we  shall use the  term 'expression' and the  notation $e$
-  for both expressions and applications.}) are given by transitions of
+The  semantics  of   expressions  (and  applications)
+%% \footnote{In  most
+%%   contexts, we  shall use the  term 'expression' and the  notation $e$
+%%   for both expressions and applications.} 
+are given by transitions of
 the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
@@
 \emph{suspended  evaluation  contexts}  (represented  as  continuation
-frames) in the stack $S$.   \warning{The
-3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
-system will be denoted $\mathcal{S}$.}
+frames) in the stack $S$.   
+%% \warning{The
+%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
+%% system will be denoted $\mathcal{S}$.}
 
 The  initial  state  of  the  transition  system  described  above  is
-$([\;]_\rho,\,   ([\;]_\rho,   \,   \rho(\ans),\,   (\print~\ans)):[\;]_{S},
-[\;]_{H},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_H$  are the
-empty environment and  the empty heap. The  initial stack consists
+$([\;]_\rho,\,   (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
+(\print~\ans)):[\;]_{S}, 
+[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$,
+$[\;]_{\mbox \footnotesize {H}}$
+and $[\;]_{S}$ are    are the
+empty environment, heap and stack respectively. The  initial stack consists
 of  a single  continuation frame  in  which \ans\  is a  distinguished
 variable  that  would   eventually  be  updated  with   the  value  of
-(\mainpgm).  In  addition, \print\ is  a function modeling  a printing
+(\mainpgm) and $\rho_\mathit{init}$ maps \ans\ to a location $\ell_{\ans}$.   In  addition, \print\ is  a function modeling  a printing
 mechanism---a   standard   run-time   support  assumption   for   lazy
-languages---that prints  the value of  (\mainpgm).  The operator $:$
+languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The operator $:$
 pushes elements on top of the stack.
@@
 used in  future computations and  dead if  it is definitely  not used.
-Heap-allocated data needs a richer model of liveness which talks about
-liveness of  references.  Using  $\acar$, $\acdr$ to  represent access
-using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
-reachable from a  variable can be represented by a  set of {\em access
-  paths}     i.e.     prefix-closed     set     of    strings     from
-$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
-$\{\epsilon, \acar, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$,
-then future  computations can only  refer up  to the second  and third
-members  of $x$.   A  {\em  liveness environment}  is  a mapping  from
-variables to access paths, but often expressed as
-a set,  for example  by writing $\{x.\epsilon,  x.\acdr, x.\acdr\acdr,
-y.\epsilon\}$  instead  of  $[x  \mapsto\{\epsilon,\acdr,\acdr\acdr\},
-  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $y \mapsto
-\{\epsilon\}$ represents access using $y$  itself and $z \mapsto \{\}$
-indicates that $z$ is dead. 
-In lazy languages, liveness environments are associated with regions of
-programs instead of program points.
+Heap-allocated data  needs a richer model  than classical liveness---a
+model  which  talks  about  liveness of  references.   Using  $\acar$,
+$\acdr$  to  represent access  using  $\CAR$  and $\CDR$  fields,  the
+liveness of the structure reachable from a variable can be represented
+by a set of {\em access paths} i.e.  prefix-closed set of strings from
+$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (typically a
+variable  reference) and  a heap  \heap, semantically  an access  path
+$\alpha$ represents  a reference, denoted $\heap\oxlb\ell,  \alpha\oxrb$, in
+the heap  that is  obtained by  starting with  $\ell$ and  chasing the
+\CAR\ or  \CDR\ fields  as specified  by the  access path.  
+$\heap\oxlb\ell, \alpha\oxrb$  is defined only  if the path followed  in the
+heap  is \emph{closure-free}  (does not  cross closures),  else it  is
+undefined.
+
+Access paths can be used to  represent liveness. As an example, if $x$
+is  a  list  with  liveness  $\{\epsilon,  \acar,  \acdr,  \acdr\acar,
+\acdr\acdr,  \acdr\acdr\acar\}$,  then  future computations  can  only
+refer up  to the  second and  third members of  $x$.  A  {\em liveness
+  environment} is a mapping from  variables to access paths, but often
+expressed as  a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
+x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
+  \mapsto\{\epsilon,\acdr,\acdr\acdr\},          y\mapsto\{\epsilon\},
+  z\mapsto\{\}]$.    In  this   notation,  $y   \mapsto  \{\epsilon\}$
+represents access using $y$ itself and $z \mapsto \{\}$ indicates that
+$z$ is dead.  In lazy  languages, liveness environments are associated
+with regions of programs instead of program points.
 
 A notion  that generalizes liveness  is {\em demand}.   While liveness
-gives the patterns of future uses of a  variable, demand
-represents the future  use of the value of an  expression.  The demand
-on an expression  $e$ is again a set of  access paths---that subset of
+gives the patterns of future uses of a variable, demand represents the
+future use of the value of an expression.  The demand on an expression
+$e$   is   again   a   set    of   access   paths---that   subset   of
 $\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
 result.   To see  the need  for demands,  consider the  let expression
-$\pi:(\LET~x\leftarrow  (\CDR~y)~\IN~ \pi':\SRETURN~x)$.   Assume that
-the context  of this expression  places the demand  $\lbrace \epsilon,
-\acar \rbrace$. Since the value of the expression is
-the value  of $x$, the demand  translates to the liveness  $[x \mapsto
-  \lbrace   \epsilon,  \acar   \rbrace]$  at   $\pi'$.   Due   to  the
-\LET\ definition which binds $x$ to $(\CDR~y)$, the liveness of $x$ at
-$\pi'$ now becomes the demand on $(\CDR~y)$.  This, in turn, generates
-the liveness  $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar  \rbrace$ at
-$\pi$.  These are  the set of $y$-rooted accesses  required to explore
-$\lbrace \epsilon, \acar  \rbrace$ paths of the  result of $(\CDR~y)$.
-%% As an analogy  with classical (strong) liveness analysis,  $y$ and $z$
-%% are live  at the entry $\pi:  x:=y+z$, if and  only if $x$ is  live at
-%% exit   of  $\pi$.    In   our  terminology,   the  liveness   $\lbrace
-%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand on
-%% $y+z$, and this,  in turn generates the  liveness $\lbrace y.\epsilon,
-%% z.\epsilon \rbrace$ at the entry of $\pi$.
-
-We use $\sigma$
-to  range over  demands,  $\alpha$  to range  over
-access  paths and  $\Lv$  to  range over  liveness
-environments.   The   notation  $\sigma_1\sigma_2$
-denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
-\alpha_1     \in     \sigma_1,    \alpha_2     \in
-\sigma_2\rbrace$.  Often  we shall  abuse notation
-to juxtapose  an edge  label and  a set  of access
-paths;   $\acar\sigma$   is    a   shorthand   for
-$\lbrace\acar\rbrace\sigma$.
+$(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
+context of this expression places  the demand $\lbrace \epsilon, \acar
+\rbrace$. Since the  value of the expression is the  value of $x$, the
+demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
+  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
+$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
+This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
+y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
+required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
+result of $(\CDR~y)$.
+%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
+%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
+%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
+%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
+%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
+%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.
+
+We use $\sigma$  to range over demands, $\alpha$ to  range over access
+paths  and $\Lv$  to range  over liveness  environments. $\Lv_y$  will
+denote  the liveness  of an  individual  variable $y$  in $\Lv$.   The
+notation $\sigma_1\sigma_2$ denotes  the set $\lbrace \alpha_1\alpha_2
+\mid \alpha_1  \in \sigma_1, \alpha_2 \in  \sigma_2\rbrace$.  Often we
+shall abuse  notation to juxtapose an  edge label and a  set of access
+paths; $\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
 
@@
 %% $\emptyset$.  This results in the empty liveness environment.
-To understand the  liveness rule for $\LET~x \leftarrow~s~\IN~e$, observe that the value of  $\LET$ is the value
-of  its body  $e$.   Thus the  liveness environment  $\Lv$  of $e$  is
-calculated for the given demand $\sigma$. Since the stack variable $x$
-is copied to each of the  closure variable $x_{\pi}$, the liveness of
-$x$ is the union of the
+To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
+observe that the value  of $\LET$ is the value of  its body $e$.  Thus
+the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
+demand $\sigma$. Since the stack variable $x$ is copied to each of the
+closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
 liveness of the  closure variables.  This liveness,  say $\sigma'$, is
-also the demand on $s$, and the resulting liveness environment $\mathit{ref}(s,
-\sigma', \Lfonly)$  is added  to $\Lv \cup  \{x.\sigma\}$ to  give the
-overall liveness environment for $(\LET~x \leftarrow~s~\IN~e)$.
-
-\warning{Note  carefully that  $\Lv_y$ specifies  the liveness  of the
-  reachable heap starting  from $y$ and extending  till closures.  The
-  liveness  of further  heap  contents starting  from  a closure,  say
-  $(\CAR\, x_{\pi})$, is given  by $\Lv_{x_{\pi}}$.  Therefore, in our
-  implementation,   a  closure   carries   the  liveness   environment
-  consisting of its free variables.}
+also  the  demand  on  $s$, and  the  resulting  liveness  environment
+$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
+\{x.\sigma\}$ to  give the  overall liveness environment  for $(\LET~x
+\leftarrow~s~\IN~e)$.
+
+As noted earlier, $x.\alpha \in  \Lv_y$ specifies the liveness  of the
+reference $\heap\oxlb\rho(x), \alpha\oxrb$  only if $\alpha$ corresponds to  a closure-free path
+in \heap\   starting from $\rho(x)$.  If this  path is intercepted  by a
+closure,  say  $\CAR~{x_{\pi}}$, then  the  liveness  of further  heap
+contents starting from $x_{\pi}$ is given by $\Lv_{x_{\pi}}$.
 
@@
 We shall now present a proof of soundness of the analysis presented in
-Section~\ref{sec:liveness-analysis}.  \warning{While  the proof  is on
-  the broad lines of  \cite{asati14lgc}, there are several differences
-  due to the  lazy semantics of the language in  this paper.} Here are
+Section~\ref{sec:liveness-analysis}.  %% \warning{While  the proof  is on
+  %% the broad lines of  \cite{asati14lgc}, there are several differences
+  %% due to the  lazy semantics of the language in  this paper.}
+Here are
 the ideas behind the proof.
 \begin{enumerate}
-\item       We        enrich       the        standard       semantics
+\item       We   augment     the        standard       semantics
   (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
@@
 \end{center}
+\caption{Minefield semantics for \sc{car-clo-bang} and
+  \sc{car-clo-bang} \label{fig:minefield-semantics-for-some}}
 \end{figure*}
@@
 \begin{enumerate}
-\item   We   enrich    the   abstract   machine   state
+\item   We   extend    the   abstract   machine   state
   $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
@@
   \begin{enumerate} 
-  \item Given the current  evaluation context $\rho, S,
+  \item \label{gc-1}Using the current  evaluation context $\rho, \stk,
     \heap,   e,  \sigma$   and   the  demand   transformers
-    $\Lfonly$, we construct a liveness environment
-    \Lv.  %%  If  $e$  is  an expression,  then  $  \Lv  =
+    $\Lfonly$, we construct a liveness environment \Lv\ for the
+    expression $e$.
+      %%  If  $e$  is  an expression,  then  $  \Lv  =
     %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
     %% application,  then  $  \Lv  =  ref(e,  \sigma,  \Lfonly)$.  
-    We  similarly  define the  liveness environment
-    for each  of the  suspended evaluation  contexts in
-    $S$, giving  a stack of liveness  environments that
-    we shall denote $\mathsf{SL}$.
-  \item \emph{Garbage collection  of objects pointed by
-    root   set:}  For   each   $x  \in   domain(\rho)$,
-    $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
-    \Lv$.    Similarly,    for   each    stack    entry
-    $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
-    corresponding      liveness     environment      in
-    $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
-    $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
-  \item  \emph{Garbage  collection  of objects  in  the
-    heap:} \warning{ For each location  $\ell$, $\heap'(\ell) = \bot$
-    iff  there   is  no  $x$  in   either  the  current
-    environment  or  one  of the  stacked  environments
-    $\rho$, such  that for some  $\alpha$ corresponding
-    to a  \warning{closure-free path}  in the heap  starting from
-    $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
-    \Lv$}. Here \Lv\ is the liveness of the variables in
-    the current environment or the stacked environment.
+ We  similarly  define  the  liveness  environment  for  the
+ expressions $e$ in each of the suspended evaluation contexts $(\rho, \ell,  e, \sigma)$ in $S$, giving
+ a stack of liveness environments that we shall denote $\mathsf{SL}$.
+  \item \emph{Garbage collection of objects pointed by root sets:} For
+    the environment $\rho$ in the current  context and for each $x \in
+    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
+    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
+    in $S$ 
+    with  $\Lv'$   as  the   corresponding  liveness   environment  in
+    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
+    to $\bot$ iff $x.\epsilon \notin \Lv'$.
+  \item \emph{Garbage  collection of  objects in  the heap:}  For each
+    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff there  is no
+    variable  $x$  in either  the  current  environment or  a  stacked
+    environment,  say  $\rho$,  such   that  $x.\alpha$  is  live  and
+    corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
+    domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
+    \oxrb = \ell$  and $x.\alpha \in \Lv$, where \Lv\  is the liveness
+    environment corresponding to $\rho$ mentioned in step~\ref{gc-1}.
   \end{enumerate}
@@
 using $GC$.  The details of the transition for the {\sc
-  car-clo}  rule   is  shown   in  the  top   table  of
-Figure~\ref{fig:minefield-semantics}.  If  this  or  an
+  car-clo}  rule   is  shown   in
+Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
 earlier  call to  $GC$  results  in $H(\rho(x))$  being
 bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
-enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
+enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
 the transition is the same as the earlier {\sc car-clo}
@@
 
-\warning{The reader has to be prepared for several, terms and concepts
-  in   this  paragraph}   In  the   liveness  analysis   described  in
-Section~\ref{sec:liveness-analysis},   the   liveness   of   variables
-appearing in an  application $s$ that lies on  e-paths, say $\epath_1$
-and  $\epath_2$, is  derived from  the dependence  chains along  these
-e-paths.  This  is  a   static  approximation  without  any  knowledge
-regarding  which of  these  e-paths will  be  eventually taken  during
-execution. During  GC, we would  like to  use a more  precise liveness
-using  the knowledge  of the  actual e-paths  taken during  execution.
-Thus  we   have  liveness  automata  separately   for  $\epath_1$  and
-$\epath_2$  as  well  as  a  automaton  corresponding  to  both  paths
-$\epath_1$ and  $\epath_2$.  The closure corresponding  to $s$ carries
-the  liveness  environment for  its  free  variables (as  pointers  to
-automata, one for each  variable).  Initially the liveness environment
-is   based   on   the    dependences   along   both   $\epath_1$   and
-$\epath_2$. However, after evaluating an \SIF\ condition, the liveness
-environments  are  updated  to  one  based  on  either  $\epath_1$  or
-$\epath_2$, so that subsequent GCs  are based on more precise liveness
-information.
+In        the       liveness        analysis       described        in
+Section~\ref{sec:liveness-analysis}, the liveness of closure variables
+in  an application  $s$ is  derived from  dependence chains  along all
+paths from the program point where the closure is created to reachable
+evaluation  points.  Assume,  for   the  sake  of  concreteness,  that
+$\epath_1$  and $\epath_2$  two  such evaluation  points.   This is  a
+static approximation  without any  knowledge regarding which  of these
+paths will be  eventually taken during execution. During  GC, we would
+like to use a more precise  liveness based on the  actual
+paths  taken  during  execution.    Therefore we create  liveness  automata
+separately  for the dependences in   $\epath_1$ and  $\epath_2$  as  well as  a  automaton
+corresponding to  both paths  $\epath_1$ and $\epath_2$.   The closure
+corresponding to  $s$ carries  the liveness  environment for  its free
+variables (as pointers to automata, one for each variable).  Initially
+the  liveness  environment is  based  on  the dependences  along  both
+$\epath_1$    and   $\epath_2$.    However,   after    evaluating   an
+\SIF\ condition, the liveness environments are updated to one based on
+either $\epath_1$ or  $\epath_2$, so that subsequent GCs  are based on
+more precise liveness information.
 
 Based on  the above considerations,  we restrict the  possible garbage
-collection points to the following:
+collection points in a function body to the following:
 \begin{enumerate}
@@
   GC is triggered.
-\item \warning{Since the evaluation of a  \SIF\ condition may trigger a garbage
-  collection, the
-  available memory is checked once again against a revised estimate of
-  the memory  required to execute  the program.  A GC is  triggered if
-  required.} 
+\item Since the evaluation of a  \SIF\ condition may trigger a 
+  collection, after evaluating the condition   the available memory
+  is  checked once  again against  a  revised estimate  of the  memory
+  (based on  value of the condition)  required to execute the  rest of
+  the program.  A GC is triggered if enough memory is unavailable.
 \end{enumerate}
@@
 
-\warning{We shall call a unit  allocatable memory as a {\em
-  cell}.    A  cell   can  hold   a  basic   value
-($\mathsf{bas}$),          the         constructor
-\CONS\   $(\mathsf{cons~   arg_1~arg_2}$)   or   a
-closure.   The closure,  in  turn, can  be one  of
-($\mathsf{unop~arg}$),                ($\mathsf{
-  binop~arg_1~arg_2}$)  and  function  application
-($\mathsf{f~arg})$.  Here each $\mathsf{
-  arg_i}$ is  a reference to another  heap cell.  In
-addition,  the closure  also carries a pointer to
-an automaton (denoted $\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.}
-
-%% While  exploring the  heap, the  GC may  encounter
-%% closures.  In  RGC  mode,   the  closure  will  be
-%% copied as usual  by traversing  the
-%% references corresponding  to the  arguments. Since
-%% our  liveness analysis  predicts  the liveness  of
-%% data in its evaluated  form only, closures present
-%% a  challenge for  LGC. LGC  will need  to map  the
-%% liveness of  the free variables of  the closure to
-%% their  evaluation  point.   This is  not  straight
-%% forward  as variables  may  escape their  creation
-%% scope  and need  to be  handled in  an approximate
-%% way.
+We shall call a  unit allocatable memory as a {\em  cell}.  A cell can
+hold    a    basic    value    ($\mathsf{bas}$),    the    constructor
+\CONS\ $(\mathsf{cons~  arg_1~arg_2}$) or a closure.   The closure, in
+turn,    can    be    one   of    ($\mathsf{unop~arg}$),    ($\mathsf{
+  binop~arg_1~arg_2}$)  and  function application  ($\mathsf{f~arg})$.
+Here each $\mathsf{  arg_i}$ is a reference to another  heap cell.  In
+addition, the closure also carries  a pointer to a DFA (denoted
+$\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.
 
@@
 \end{algorithm}
-Algorithm~\ref{algo:lgc-a}   describes  the    liveness-based   garbage
-collection scheme.  Starting  with the root set, each  cell pointed by
-each live  reference (i.e. whose associated  DFA state is final)  is copied
-using  $\mathsf{copy}$.   Copying  \CONS\ cells  is  simple---it  just
-involves  copying  the  cell  itself  and  conditionally  copying  the
-\CAR\ and  the \CDR\  fields by  referring to the  next states  of the
-automaton.   If the  reference points  to  a closure,  then, as  noted
-earlier, the closure carries pointers to the liveness automata for its
-arguments.   These automata are  used to  recursively initiate  copying of  the
-arguments.     Note     that      the     copying     strategy     for
-$\mathsf{(unop~arg_1~arg_2)}$  or  $\mathsf{(f~arg_1)}$  are  similar  to
-$\mathsf{(binop~arg_1~arg_2)}$ and have not been shown in the algorithm.
-
-%% Let   us  consider   determining  the   liveness  of   a  closure   at
-%% runtime. Consider a closure $(app~f~y)$  has a reference from the root
-%% set  variable  $x$.  If $x$  is  live,  then  we  know that  the  cell
-%% $(app~f~y)$ has  to be copied.  To copy  the heap structure  rooted at
-%% $y$, we need the liveness of  $y$ at the program point which triggered
-%% the evaluation of  $(app~f~y)$. However, this is only  possible if the
-%% activation record of the function  which created this closure is still
-%% on the activation stack. This does  not always happen since a function
-%% could pack a closure in a \CONS\ cell and return the \CONS\ cell.
-
-%% An alternative could be to  carry with each closure, extra information
-%% about the program point which triggered its evaluation.  This gives us
-%% precise liveness  information of the  free variables of a  closure but
-%% involves the overhead of updating the program point information during
-%% execution.  To avoid  this overhead we go for a  safe approximation by
-%% storing the {\em creation point of  the closure} and using it to check
-%% liveness.  This  is safe since the  liveness at the creation  point of
-%% the closure dominates  all possible evaluation points  of the closure,
-%% but it  might be imprecise  as the actual  liveness might be  way less
-%% than what is found at the creation point.
-
-%% Note that  when a reachability-based  collector visits a cell  for the
-%% first time,  it is  marked.  Explorations starting  from the  cell are
-%% curtailed if the cell is visited subsequently due to sharing.  Since a
-%% liveness-based collector may reach the  same cell multiple times, each
-%% time with a different automaton state, this curtailment cannot be done
-%% in the case  of LGC. This is an inherent  drawback of a liveness-based
-%% collector.  This  can be mitigated  by maintaining a list  of liveness
-%% automata states  that a  particular heap cell  was traversed  with and
-%% avoid repeated traversals if we reach the cell in one of these states.
-%% However this also involves storage and run-time penalty.
-
-%% To  avoid  performance  penalty,  we experimented  with  a  mixed-mode
-%% garbage collection scheme  which uses liveness for  evaluated data and
-%% reachability for closures.
-
-%% \subsection{Liveness-based collector with reachability-based closure
-%%   collection}
-%% \begin{algorithm}[t]
-%%   \Fn{\Lgc{}}
-%%      {
-%%        \For {each reference $\mathit{ref}$ in root set}
-%%             {\LCopy($\mathit{ref}$)\;}
-%%             $\mathit{copyReferencesOnPrintStack()}$\;  
-%%      }
-%%      \Fn{\LCopy{$\mathit{ref}$}}
-%%         {
-%%           \eIf{cellType$(\mathit{ref})$ is $cons$}
-%%               {
-%%                 \eIf {$\mathit{ref}$ is found live using dfa and has not been copied using
-%%                   reachability}
-%%                      {
-%%                        $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
-%%                        $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
-%%                        $\mathit{ newRef.car}   = \mathit{ newCar}$\;
-%%                        $\mathit{ newCdr}  =$  \LCopy$\mathit{(ref.cdr)}$\;
-%%                        $\mathit{ newRef.cdr}   = \mathit{newCdr}$\;
-%%                      }
-%%                      { $\mathit{makenull(ref)}$\;}}
-%%               {
-%%                 \If {cellType$\mathit{(ref)}$ is closure and 
-%%                   has not been copied}
-%%                     { 
-%%                       $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
-%%                     }
-%%               }
-%%         }
-%%         \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
-%% \end{algorithm}
-
-%% \begin{table}[ht!]
-%% \caption{Time taken by LGC for different strategies.  The programs are
-%%   modified (smaller) instances of original benchmarks.}
-%% \label{tab:compare-live-clo-reach-clo}
-%% \centering
-%% \begin{tabular}
-%% {| l | r | r | r |}
-%% \hline
-%% Program        & \#GC & \multicolumn{2}{c|}{Closure collected using} \\ \cline{3-4}
-%%                &      & Reachability & Liveness                      \\ \hline
-%% \verb@nqueens@ & 183  & 0.009 sec    & 0.228 sec                     \\\hline
-%% \verb@lambda@  & 11   & 0.001 sec    & 4.992 sec                     \\ \hline
-%% \verb@lcss@    & 1    & 0.000 sec    & 34.668 sec                    \\ \hline
-%% \end{tabular}
-%% \end{table}
-
-%% Algorithm~\ref{algo:lgc-b} describes a garbage collection scheme where
-%% the liveness automata is consulted  only for \CONS\ cells; and assumes
-%% everything under  a closure to be  live and copies everything  that is
-%% reachable.
-
-%% A consequence of  copying everything under a closure is  that we might
-%% reach  a non-live  cell that  has been  reclaimed earlier  by liveness
-%% based GC.   This can  happen if  a \CONS\  cell having  partially live
-%% sub-structure  is copied  using liveness  during a  collection and  it
-%% subsequently  becomes part  of a  live closure.   During a  subsequent
-%% collection, the collector may try to copy everything under the closure
-%% including the dead  sub-structure of the \CONS\ cell.   To avoid this,
-%% we ensure  that any non-live  references are not carried  across GC's.
-%% We do this by scanning the live  part of the buffer after every GC and
-%% nullifying any  reference which is  not live. This is  correct because
-%% any reference which  is dead during a collection is  guaranteed not be
-%% dereferenced any further by the program.
-
-%% Another advantage of  this mixed garbage collection scheme  is that it
-%% avoids the problem of multiple  explorations from the same cell.  This
-%% is done by maintaining a flag in  each cell to indicate whether it was
-%% copied using reachability.
+Algorithm~\ref{algo:lgc-a}   describes   the   garbage
+collection scheme.  Starting with the root set, each cell pointed by a
+live reference  (i.e. whose associated  DFA state is final)  is copied
+using $\mathsf{copy}$.   Copying a \CONS\ cells  just involves copying
+the  cell  itself   and  conditionally  copying  the   \CAR\  and  the
+\CDR\ fields after referring to the  next states of the DFA.  If
+the reference points to a closure, then, as noted earlier, the closure
+carries pointers  to the liveness  DFAs of its  arguments.  These
+are  used to recursively  initiate copying of  the arguments.
+Note that  the copying  strategy for  $\mathsf{(unop~arg_1~arg_2)}$ or
+$\mathsf{(f~arg_1)}$ are similar to $\mathsf{(binop~arg_1~arg_2)}$ and
+have not been shown.
 
@@
 %\subsection{Garbage collection for references on print stack}
-In lazy  languages the evaluation of the top-level expression is driven  by the print
-function~\cite{Jones87}.  Printing of  atomic values is
-simple  and  does  not  trigger  a garbage  collection.   In  case  of
-\CONS\ cells, the \CAR\ is first  printed followed by \CDR.  In a lazy
-language, both \CAR\ and \CDR\ could be closures requiring evaluation,
-and this  may trigger  more garbage collections.   When this  happens.  we
-have to consider  any references that might be on  the print stack and
-copy them. We extend the liveness analysis to the print function and
-use its result during garbage collection. \warning{Is this enough?}
+The evaluation of  the top-level expression in a program  is driven by
+the \print\  function.  Printing of  atomic values is simple  and does
+not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
+\CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
+\CAR\ and \CDR\  could be closures requiring evaluation,  and this may
+trigger  more garbage  collections.  When  this happens.   we have  to
+consider any  references that  might be  on the  print stack  and copy
+them.  We extend liveness-based  garbage  collection  to
+the function \print.
 
@@
 \label{sec:experiments}
-To   test  the   utility  of   our  method   we  have   implemented  a
-garbage-collector for a simple scheme-like first order language having
-non-strict semantics. It consists of an interpreter, liveness analyzer
-and  a  garbage-collector  that  can optionally  use  reachability  or
-liveness.  Our benchmark consists  of programs from nofib~\cite{nofib}
-suite and some  external programs manually converted to  ANF.  All our
+  Our benchmark mostly consists  of programs from nofib~\cite{nofib}
+suite.  All our
 benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
@@
 
-The process of liveness-based garbage collection involves going through 
-each activation record on the stack and exploring each variable in the 
-current activation record (root set). The liveness of each variable is 
-determined using the program point and the variable name. The liveness 
-of all variables are stored as DFA. These DFA are then encoded as a 
-table. A variable is live only if an entry is found in this table.
-All cells that are live are copied to the live semi-space. In case of 
-\CONS\ cells the \CAR\ and \CDR\ pointers are chased and if they are 
-live the copied \CONS\ cell will get the updated addresses of its 
-\CAR\ and \CDR\ fields.
-
-%% \subsection{Copying closure using liveness vs copying closures using reachability}\label{sec:strategies}
-
-%% To compare the  effect of using reachability to  copy closures instead
-%% of liveness  we take some  representative programs from  our benchmark
-%% suite    and    execute    them    with    both.     As    shown    in
-%% Table~\ref{tab:compare-live-clo-reach-clo},  the   number  of  garbage
-%% collections in both cases are same but copying closures using liveness
-%% takes a lot more time compared to reachability based copying.
-
-%% The number  of garbage  collections is the same  since we  approximate the
-%% liveness of  a closure  to the  liveness at  its creation  point.  The
-%% liveness  we get  is usually  grossly over  approximated.  Using  this
-%% liveness may not  give too much of a benefit  compared to reachability
-%% as nearly everything would be live at its point of creation.
-
-%% As  mentioned   in  Section~\ref{sec:live-clo},  using   liveness  for
-%% closures  may  entail  multiple  traversals over  the  same  structure
-%% consuming more  time.  The  advantage of avoiding  multiple traversals
-%% can be clearly seen from the lcss example where a single GC takes less
-%% than a millisecond for reachability  based closure collection where as
-%% the  liveness  based collection  is  of  the  order of  seconds.  Lazy
-%% languages tend  to have a  lot of sharing  and hence using  a liveness
-%% based collection for closures is impractical.
-
-%% Therefore we believe  that using reachability for closures  is a sweet
-%% spot   in   terms   of   performance  for   liveness   based   garbage
-%% collectors.  For the  remainder  of  the section  all  LGC results  we
-%% discuss uses  a mixed-mode collector--liveness for  evaluated data and
-%% reachability for closures.
+%% The  process  of  liveness-based  garbage  collection  involves  going
+%% through  each  activation  record  on the  stack  and  exploring  each
+%% variable in the current activation  record (root set). The liveness of
+%% each variable is  determined using the program point  and the variable
+%% name. The liveness  of all variables are stored as  DFA. These DFA are
+%% then encoded as a table. A variable  is live only if an entry is found
+%% in  this table.   All  cells that  are  live are  copied  to the  live
+%% semi-space. In case  of \CONS\ cells the \CAR\ and  \CDR\ pointers are
+%% chased  and if  they are  live  the copied  \CONS\ cell  will get  the
+%% updated addresses of its \CAR\ and \CDR\ fields.
 

### 2015-11-27 paper/paper.tex

@@
 \begin{enumerate}
+
 \item   We   extend    the   abstract   machine   state
-  $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
+  $\rho,S,\heap,e$   to  $\rho,S,\heap,e,\sigma$. 
 We  call such a state  a \emph{minefield state}.
@@
 Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
-earlier  call to  $GC$  results  in $H(\rho(x))$  being
+earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
 bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
@@
 \hline
-
-\hline
-
-          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
-  H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
+\hline
+&\makecell{ $\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S,
+  \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell :=
     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
@@
 \hline
-          & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
-  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
-\hline
-$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
-  \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
-  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
-\sc{car-whnf} \\
-\hline
-$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-  H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
+& \makecell{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
+  \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+\hline
+$\heap(\rho(x)) \mbox{ is } (v, d)$ & \makecell{$\rho, (\rho', \ell, e,
+  \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
+  \rightsquigarrow \rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
+\sc{car-select} \\
+\hline
+$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
+(\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
+\sc{car-1-clo} \\
+
+\hline
+
+$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+  \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
 \sc{car-clo}

### 2015-11-27 paper/paper.tex
Fig 1 (motiv example) fixed

@@
 \label{sec:motiv}
-\begin{figure}[t!]
-  \psset{unit=1mm}
-  \begin{pspicture}(90,52)(0,-52)
-    %\psframe(90,48)(0,-48)
-    \renewcommand{\arraystretch}{1}
-    \begin{tabular}{@{}c@{}}
-      %\psframebox
-          {\sf
-	\renewcommand{\arraystretch}{1}{
-	  \begin{uprogram}
-            \UFL\
-            \UNL{0} (\DEFINE\ (\length\  \pl)
-	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
-            (\PRIM\ 1\ (\length\  (\CDR\  \pl)))))
-            \UNL{0}
-	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
-	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
-	    \listb
-	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\ \lista) (\append\
-            (\CDR\  \lista)
-            \listb))))
-            \UNL{0}
-            \UNL{0} (\LET\ \px\
-            $\leftarrow$\ (\CONS\ $5$
-            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
-	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
-	    \UNL{2}
-            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
-            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
-            (\length\ \pz))))))
-	  \end{uprogram}
-      }}
-      \\ \\
-      (a) Example program. \\  \\
-%      \includegraphics[width=.45\textwidth]{motiv-example}
-      \input{mem-graph} \\ \\
-      \renewcommand{\arraystretch}{.9}
-      \begin{tabular}[t]{p{0.9\columnwidth}}
-        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
-        denotes a closure. Thick edges denote live links. Traversal
-        stops at edges marked $\times$ during garbage collection for
-        a liveness-based collector.
-      \end{tabular}
-    \end{tabular} 
-  \end{pspicture}
-%  \vspace*{-3ex}
-  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
-\end{figure}
-
+\input{fig_motiv}
 Figure~\ref{fig:mot-example} shows an example program and the state of
@@
 
-\newpage
-\ 
-\newpage
-
+\clearpage
 \section{Appendix: Soundness of liveness analysis}

### 2015-11-27 lazy-scheme/post_process/gen_stats_and_plotall.sh
plots...

@@ -74,6 +74,7 @@ freq["knightstour"]=8500;
 freq["treejoin"]=1300;
 freq["nqueens"]=3000;
 freq["lambda"]=100;
+freq["huffman"]=100;
 
 declare -A xtics
 declare -A ytics
@@ -90,6 +91,7 @@ xtics["knightstour"]=8.5e5	; ytics["knightstour"]=40e3
 xtics["treejoin"]=13e5  	; ytics["treejoin"]=330e3
 xtics["nqueens"]=3.5e5    	; ytics["nqueens"]=4.6e3
 xtics["lambda"]=19e5		; ytics["lambda"]=3e3
+xtics["huffman"]=19e5		; ytics["huffman"]=3e3
 
 declare -A xlimit
 declare -A ylimit
@@ -106,6 +108,7 @@ xlimit["knightstour"]=850000	; ylimit["knightstour"]=40000
 xlimit["treejoin"]=3.9e6:4.5e6	; ylimit["treejoin"]=1320e3:1650e3
 xlimit["nqueens"]=3e5:4e5	; ylimit["nqueens"]=0:23e3
 xlimit["lambda"]=50e5:56e5	; ylimit["lambda"]=:
+xlimit["huffman"]=:             ; ylimit["huffman"]=:
 
 if [ "x$BMs" = "x" ]; then
     BMs=$AllBMs
@@ -121,10 +124,10 @@ do
       	    -bmdir ../benchmarks/programs/$bm \
       	    -gcopt $opt -heap ${limit[$bm]}
      done
-    #bash $DIR/val.plot $bm ${xtics[$bm]} ${ytics[$bm]}
-    #bash $DIR/window.plot $bm ${xlimit[$bm]} ${ylimit[$bm]} 
-    # gnuplot $DIR/plots/${bm}.gnu
-    # gnuplot $DIR/plots/${bm}_win.gnu
+    bash $DIR/val.plot $bm ${xtics[$bm]} ${ytics[$bm]}
+    bash $DIR/window.plot $bm ${xlimit[$bm]} ${ylimit[$bm]} 
+    gnuplot $DIR/plots/${bm}.gnu
+    gnuplot $DIR/plots/${bm}_win.gnu
 
     echo "$bm Done"
 done

### 2015-11-28 paper/paper.tex
latest - 27th

--- 2015-11-27 paper/paper.tex
@@
 
-  %% We  consider  a  round of garbage  collection  followed  by  a
-  %% transition as a step.  We show by induction that, for
-  %% programs  without function  calls, %% starting  from the
-  %% %% initial  state  that has $e_{\main}$  as  the  
-  %% %% expression to  be evaluated,
-  %% any
-  %% transition of $n$ steps occurs without a \bang.
-
-We  first  need  an auxiliary  result  about  minefield
-semantics. Consider  a trace of a  minefield execution.
-For every minefield state  $(\rho, S, \heap, e(s), \sigma)$
-that appears  on the LHS of  a $\rightsquigarrow$ step,
-the  demand   $\sigma$  on   the  expression   $e$  (or
-application $s$) is non-null.  This can be proved by an
-induction  on  the  number  of  steps  leading  to  the
-minefield  state.   The  base step  holds  because  the
-demand $\sigma_{all}$  on $(\mainpgm)$ is  non-null. For the 
-inductive step  we observe   that for
-each  step of  the minefield  semantics, if  the demand
-$\sigma$ on  the LHS of  a minefield step  is non-null,
-the demand on  the RHS is a  transformation of $\sigma$
-(for example $\clazy\sigma$) which is also non-null.
+%% We consider a round of  garbage collection followed by a transition
+%% as  a  step.  We  show  by  induction  that, for  programs  without
+%% function  calls, %%  starting from  the %%  initial state  that has
+%% $e_{\main}$ as the %% expression to be evaluated, any transition of
+%% $n$ steps occurs without a \bang.
+
+We first need an auxiliary  result about minefield semantics. Consider
+a trace of  a minefield execution.  For every  minefield state $(\rho,
+S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
+$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
+application $s$) is  non-null.  This can be proved by  an induction on
+the number  of steps leading  to the  minefield state.  The  base step
+holds   because  the   demand   $\sigma_{all}$   on  $(\mainpgm)$   is
+non-null. For the inductive step we  observe that for each step of the
+minefield semantics, if the demand $\sigma$  on the LHS of a minefield
+step  is non-null,  the  demand  on the  RHS  is  a transformation  of
+$\sigma$ (for example $\clazy\sigma$) which is also non-null.
 
@@
 evaluations initiated by $\print$.
-%% We  show next that  the minefield execution of  a call-less
-%% program cannot go \bang.
+%% We show  next that the  minefield execution of a  call-less program
+%% cannot go \bang.
 
@@
 \bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\_, [\;]_{S} ,
-            [\;]_{\heap}, e_{\mainpgm}, \sigma_{all})$.  Since the call to
-            $GC(\ldots)$ in this state does  nothing, we just have to show
-            that  the  $\rightsquigarrow$  transition cannot  enter  a
-            \bang\   state.    Since   our  programs   are   in   ANF,
-            $e_{\mainpgm}$ can  only be a $\LET$  expression.  A
-                     {\sc let} step does not involve dereferencing,
-                     and thus
+            [\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.   Since  the
+            call to $GC(\ldots)$  in this state does  nothing, we just
+            have to show that the $\rightsquigarrow$ transition cannot
+            enter  a \bang\  state.  Since  our programs  are in  ANF,
+            $e_{\mainpgm}$ can  only be  a $\LET$ expression.   A {\sc
+              let}  step  does  not involve  dereferencing,  and  thus
             cannot result in a \bang.
 
-For the inductive step, we  shall show that none of the
-minefield steps that  involves dereferencing results in
-a   \bang.   These  are   the   steps   which  have   a
-$\heap(\rho(...))$ in the premise. %%  Recall that a minefield
-%% step   consists   of   a   GC  step   followed   by   a
-%% $\rightsquigarrow$. 
+For the inductive step, we shall show that none of the minefield steps
+that involves dereferencing  results in a \bang.  These  are the steps
+which  have a  $\heap(\rho(...))$ in  the  premise. 
 Now a  $\rightsquigarrow$ step can  go
@@
 $\mathscr{L}(\var{\Lanv{i}{x}})$,   is    the   desired
-solution  of  \Lanv{i}{x}.    However,  note  that  the
-strings  in the  language  are  over alphabet  $\lbrace
-\acar, \acdr,\bcar, \bcdr, \clazy \rbrace$, but we want
-     {\em  forward} access  paths only,  i. e.,  access
-     paths  over  the  alphabet $\lbrace  \acar,  \acdr
-     \rbrace$.  In other  words,  the decision  problem
-     that   we  are   interested   in  during   garbage
-     collection at a program point $\pi_i$ is:
+solution  of  \Lanv{i}{x}.
+
+\warning{
+\hrule
+\medskip
+Fixing from here
+\medskip
+\hrule
+}
+
+However, recall that the decision problem that we are interested in
+during garbage collection is the following:
 \begin{quote}
@@
 
-From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
-that the results of liveness analysis of any program can be modeled by
-the kind  of grammar described above. The  following proposition shows
-that the converse also holds.
-\begin{proposition}
-Given    a     grammar    $G$    of    the     form    described    in
-Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
-program $p$  with program points  $\pi_i$ and variables $x$  such that
-the liveness  analysis of $p$  is same as  $G$ except for a  change in
-non-terminal names which now have the form \var{\Lanv{i}{x}}.
-\end{proposition}
-We now show that the decision  problem that is required to be answered
-at       garbage      collection       time       is      undecidable.
+%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
+%% that the results of liveness analysis of any program can be modeled by
+%% the kind  of grammar described above. The  following proposition shows
+%% that the converse also holds.
+%% \begin{proposition}
+%% Given    a     grammar    $G$    of    the     form    described    in
+%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
+%% program $p$  with program points  $\pi_i$ and variables $x$  such that
+%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
+%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
+%% \end{proposition}
+We first show that membership problem of the class of grammars in
+Definition~\ref{def:specialgrammar} is undecidable. However note that
+the grammar may not be necessarily generated from liveness analysis.
 \newcommand{\state}{\ensuremath{\mathsf{S}}}
 \newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
-\begin{lemma}
+\begin{lemma}\label{lemma:grammar-undecidable}
 Consider a grammar    $G$    of   the    kind    described    in
 Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
-consisting  of symbols \acar\  and \acdr\  only. The  decision problem
-$\alpha \in \mathscr{L}(G)$ is undecidable.
+consisting  of symbols \acar\  and \acdr\  only. The membership problem
+$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
 \end{lemma}
@@
 
-We shall denote the configuration of the Turing machine
+We shall represent a Turing machine (TM) configuration 
 as $w_l(\state,c)w_r$, where $w_l$ is the string to the
@@
 is the symbol under the head and \state\ is the current
-state of  the machine.   For each combination  of state
+state of  the TM.   For each combination  of state
 and symbol  $(\state,c)$, the grammar will  contain the
 non-terminal \nont{}{c}. We shall synchronize each move
-of the  machine to  a derivation  step using  a context
+of the TM to  a derivation  step using  a context
 free production, followed, if possible, by a derivation
@@
 synchronization,  we  shall   establish  the  following
-invariant  relation between  the machine  configuration
+invariant  relation between  the TM  configuration
 and the sentential form:
@@
 \begin{quote}
-  If    the   configuration    of   the    machine   is
+  If    the   configuration    of   the    TM   is
   $w_l(\state,c)w_r$, then the  sentential form will be
@@
 form    corresponding     to    the    initial     configuration    is
-$\nont{\mathit{init}}{c}w$ (we  can assume that there  is a production
-$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w$, where  \nont{}{} is
+$\nont{\mathit{init}}{c}w\$$ (we  can assume that there  is a production
+$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w\$$, where  \nont{}{} is
 the start symbol of the  grammar). Further correspondences between the
-Turing machine moves and the grammar productions are as follows:
+TM moves and the grammar productions are as follows:
 
@@
 example:  Assume that  the current  sentential form  is
-\bcar\bcdr\nont{i}{\acar}\acar\acar.   Assume that  the
-machine  has  a   transition  $(S_i,\acar)  \rightarrow
+\bcar\bcdr\nont{i}{\acar}\acar\acar. Also  assume that  the
+TM  has  a   transition  $(S_i,\acar)  \rightarrow
 (S_j,\acdr, L)$.  Since the next corresponding  step in
@@
 corresponds  to   the  changed  configuration   of  the
-machine.  Notice  carefully that a wrong  choice breaks
+TM.  Notice  carefully that a wrong  choice breaks
 the invariant  and it cannot be  recovered subsequently
@@
 
-After  the Turing machine  has halted,  there are  further ``cleanup''
+After  the TM  has halted,  there are  further ``cleanup''
 derivations  that derive  $\epsilon$ only  if the  invariant  has been
@@
 $\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
-final state of the Turing machine. We add
+final state of the TM. We add
 productions $\nont{\mathit{final}}{c} \rightarrow \acar
@@
 \nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
-right of the tape head. This completes the reduction.
+right of the tape head. Finally the production $\clazy\$ \rightarrow
+\epsilon$ is used to derive $\epsilon$.This completes the reduction.
 \end{proof}
+
+We now show that the proof can be replayed for the class of grammars
+generated from liveness analysis of programs.
+
+\begin{lemma}
+Consider a grammar    $G$    of   the    kind    described    in
+Definition~\ref{def:specialgrammar} that is generated by liveness analysis
+of a program and a forward access path $\alpha$. The membership problem
+$\alpha\$ \in \mathscr{L}(G)$ is also undecidable.
+\end{lemma}
+
+Given a Turing machine and an input string, we now give program fragments whose liveness
+analysis generate the same productions as in
+Lemma~\ref{lemma:grammar-undecidable}, but with  extra productions
+that necessarily contain the \clazy\ symbol. For instance,
+corresponding to the production $\nont{i}{c} \rightarrow c
+\nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, we have
+the program (define ($\nont{i}{c}$ a) (let b <- (car a) in (let c <-
+($\nont{j}{\acar}$ b) in (let d <- 5 in (let e <- (cons c d) in (let f
+<- (null? e) in return f)))))). The reader can check that this
+produces the grammar fragment ...
+
+It is obvious that if the TM halts on the input, then the grammar will
+produce the $\epsilon$ string. However, if the TM does not halt on the
+input, then the productions containg
 
@@
 
-\subsubsection{From CFGs to GC-ready DFA}
+\subsubsection{From CFGs to GC-ready DFAs}
 \label{sec:NFA-approx}
@@
 The evaluation of  the top-level expression in a program  is driven by
-the \print\  function.  Printing of  atomic values is simple  and does
-not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
-\CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
-\CAR\ and \CDR\  could be closures requiring evaluation,  and this may
-trigger  more garbage  collections.  When  this happens.   we have  to
-consider any  references that  might be  on the  print stack  and copy
-them.  We extend liveness-based  garbage  collection  to
-the function \print.
+a printing function.  
+%% Printing of  atomic values is simple  and does
+%% not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
+%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
+%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
+%% trigger  more garbage  collections.  When  this happens.   we have  to
+%% consider any  references that  might be  on the  print stack  and copy
+%% them.
+We extend liveness-based  garbage  collection  to
+this function.
 

### 2015-11-28 paper/paper.tex

@@
 %==============================================================
-\begin{figure*}[t]
-\comment{
-\renewcommand{\arraystretch}{1}
-\begin{eqnarray*}
-\mathit{ref\/}(\kappa,\sigma,\Lfonly)
-          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
-$\NIL$}\\
-\mathit{ref\/}(\pi:(\CONS~x~y),\sigma,\Lfonly)
-          &=& \{x_{\pi}.\alpha \mid \acar\alpha \in \sigma\} \cup
-  \{y_{\pi}.\alpha 
-\mid \acdr\alpha \in \sigma\} \\
-\mathit{ref\/}(\pi:(\CAR~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x_{\pi}.\epsilon\} \cup \{x_{\pi}.\acar\alpha \mid \alpha \in
-\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}(\pi:(\CDR~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x_{\pi}.\epsilon\} \cup \{x_{\pi}.\acdr\alpha \mid \alpha \in
-\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}(\pi:(\PRIM~x~y),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x_{\pi}.\epsilon, y_{\pi}.\epsilon\},  & \mbox{if}~\sigma \ne
-\emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}(\pi:(\NULLQ~x),\sigma,\Lfonly)
-          &=&    \begin{array}{l l}
-                    \{x_{\pi}.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
-                    \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-\mathit{ref\/}(\pi:(f~\myvec{x}),\sigma,\Lfonly)
-%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
-          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
-               \bigcup_{i=1}^n x_{i_{\pi}}.\Lf{f}{i}{\sigma}
-               \end{array}
-%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
-\\[1ex]
-\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& 
-x.\sigma\\
-%% \mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
-%% \mapsto
-%% x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
-%%   new e-path terminating with $\psi$}\\
-\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
-\begin{array}{l l}
-                    \mathcal{L}(e_1,\sigma,\Lfonly) \cup
-        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
-\{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
-        \emptyset  & \mbox{otherwise}
-                 \end{array} \\
-%% \mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-%%         \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
-%% \mbox{where}~\mathcal{M} =
-%% \mathcal{L}(e,\sigma,\Lfonly),~\Lv =
-%% \mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
-%% && \makebox[0mm]{\hspace*{7cm}
-%%  $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
-%%            x_2\, \ldots\, x_n)/x]))$,} \\
-%%  \makebox[0mm]{\hspace*{8cm} where
-%%      $x_1\, \ldots\, x_n$ are the variables in $s$}
-\mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
-\mathit{ref\/}(s,\sigma',\Lfonly) \cup \Lv \cup \{x.\sigma'\} 
-~\mbox{where}~ \mathcal{\Lv} = \mathcal{L}(e,\sigma,\Lfonly), ~\mbox{and}~\sigma' = \bigcup_{\pi}  \Lv(x_{\pi})
-\end{eqnarray*}
-\begin{minipage}{0.85\textwidth}
-\infrule[live-define]
-        {\mathcal{L}(e_f,\sigma,\Lfonly) =
-           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
-              \mbox{ for each $f$ and $\sigma$}
-        }
-        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
-\\ \makebox[0mm]{where
-     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
-\ldots \mathit{df_k}$}}
-\end{minipage}
-  \caption{Liveness equations and judgement
-    rule}\label{fig:live-judge}
-}%comment
-\end{figure*}
+\input{fig-liveness}
 

### 2015-11-28 lazy-scheme/post_process/gen_stats_and_plotall.sh
added paraffins and deriv

@@ -1,9 +1,8 @@
-
 #!/bin/bash 
 DIR=`dirname $0`
 PROGNAME=`basename $0`
-TestBMs="  test testsmall small"
-SmallBMs=" nperm fibheap lambda sudoku lcss nqueens huffman "
+TestBMs="  test testsmall small paraffins deriv"
+SmallBMs=" nperm fibheap lambda sudoku lcss nqueens deriv"
 LargeBMs=" treejoin knightstour "
 AllBMs="$SmallBMs $LargeBMs"
 help()
@@ -31,8 +30,8 @@ while [ $# -gt 0 ]; do
     if [ "$ARG" = "-genStats" ]; then
         OPTARG=""
     elif [ "$ARG" = "-bm" ]; then
-	BMs="$BMs $1"
-	shift
+        BMs="$BMs $1"
+        shift
     else
         echo ""
         echo "      ERROR!!! Bag argument or filename" $ARG
@@ -54,6 +53,8 @@ limit["lambda"]=20466
 limit["knightstour"]=677800
 limit["fft"]=172762
 limit["huffman"]=100070
+limit["paraffins"]=425476
+limit["deriv"]=425476
 ## UNUSED BMs
 limit["small"]=17
 limit["test"]=247
@@ -75,40 +76,46 @@ freq["treejoin"]=1300;
 freq["nqueens"]=3000;
 freq["lambda"]=100;
 freq["huffman"]=100;
+freq["deriv"]=100;
+freq["paraffins"]=100;
 
 declare -A xtics
 declare -A ytics
-xtics["small"]=1		; ytics["small"]=1
-xtics["test"]=25		; ytics["test"]=25
-xtics["testsmall"]=25		; ytics["testsmall"]=25
-xtics["fft"]=2050		; ytics["fft"]=250
-xtics["sudoku"]=0.25e5		; ytics["sudoku"]=0.6e3
-xtics["lcss"]=.9e5		; ytics["lcss"]=4e3
-xtics["gc_bench"]=1.9e5   	; ytics["gc_bench"]=70e3
-xtics["nperm"]=8.5e5		; ytics["nperm"]=3.5e3
-xtics["fibheap"]=8e5		; ytics["fibheap"]=10e3
-xtics["knightstour"]=8.5e5	; ytics["knightstour"]=40e3
-xtics["treejoin"]=13e5  	; ytics["treejoin"]=330e3
[…]

### 2015-11-28 lazy-scheme/post_process/gen_stats_and_plotall.sh
fixed gc-freq bug.

@@ -1,8 +1,8 @@
 #!/bin/bash 
 DIR=`dirname $0`
 PROGNAME=`basename $0`
-TestBMs="  test testsmall small paraffins deriv"
-SmallBMs=" nperm fibheap lambda sudoku lcss nqueens deriv"
+TestBMs="  test testsmall small "
+SmallBMs=" nperm fibheap lambda sudoku lcss nqueens paraffins deriv"
 LargeBMs=" treejoin knightstour "
 AllBMs="$SmallBMs $LargeBMs"
 help()
@@ -54,7 +54,7 @@ limit["knightstour"]=677800
 limit["fft"]=172762
 limit["huffman"]=100070
 limit["paraffins"]=425476
-limit["deriv"]=425476
+limit["deriv"]=11124
 ## UNUSED BMs
 limit["small"]=17
 limit["test"]=247
@@ -76,7 +76,7 @@ freq["treejoin"]=1300;
 freq["nqueens"]=3000;
 freq["lambda"]=100;
 freq["huffman"]=100;
-freq["deriv"]=100;
+freq["deriv"]=20;
 freq["paraffins"]=100;
 
 declare -A xtics
@@ -95,7 +95,7 @@ xtics["treejoin"]=13e5          ; ytics["treejoin"]=330e3
 xtics["nqueens"]=3.5e5          ; ytics["nqueens"]=4.6e3
 xtics["lambda"]=19e5            ; ytics["lambda"]=3e3
 xtics["huffman"]=19e5           ; ytics["huffman"]=3e3
-xtics["deriv"]=19e5             ; ytics["deriv"]=3e3
+xtics["deriv"]=100             ; ytics["deriv"]=100
 xtics["paraffins"]=19e5         ; ytics["paraffins"]=3e3
 
 declare -A xlimit
@@ -123,9 +123,9 @@ fi
 
 for bm in $BMs
 do
+    for opt in gc-plain gc-live gc-freq=${freq[$bm]}
+    #for opt in gc-freq=${freq[$bm]}
     #for opt in gc-plain gc-live
-    #for opt in gc-plain gc-live gc-freq=${freq[$bm]}
-    for opt in gc-freq=${freq[$bm]}
     do 
         echo "Processing $bm [$opt]"
         $DIR/gen_gc_stats.sh $OPTARG \

### 2015-11-28 lazy-scheme/src/SchemeAST_Defs.cpp

@@ -588,82 +588,82 @@ cons* IfExprNode::evaluate()
 {
 
 	curr_return_addr = getLabel();
-//	DBG(cout << "At If updating curr_return_address to " << curr_return_addr << endl);
+	//	DBG(cout << "At If updating curr_return_address to " << curr_return_addr << endl);
 
 	if (gc_status == gc_live)
 	{
 		updateEvaluationPoint(this);
 	}
 
-//	for(vector<var_heap>::iterator vhit = actRecStack.begin()->heapRefs.begin(); vhit != actRecStack.begin()->heapRefs.end(); ++vhit)
-//	{
-//
-//		cons* heap_cell = (cons*)vhit->ref;
-////		DBG(cout << "Processing variable " << vhit->varname <<  " is_param " << vhit-> is_param << endl);
-//		if (vhit->is_param)
-//		{
-////			DBG(cout << "Processing a parameter, not updating prog point" << endl);
-//			continue;
-//		}
-//
-//		if (heap_cell && !heap_cell->inWHNF)
-//		{
-//			switch(heap_cell->typecell)
-//			{
-//			case constIntExprClosure:
-//			case constBoolExprClosure:
-//			case constStringExprClosure:
-//			case nilExprClosure:
-//				break;
-//			case unaryprimopExprClosure:
-//			{
-////				cout << "Unaryprimop" << endl;
-//				auto arg_num = heap_cell->val.closure.arg1_name->substr(heap_cell->val.closure.arg1_name->rfind("/")+1);
-////				cout << "Updating argument prg_pt " << *(heap_cell->val.closure.arg1_name)  << " to " << this->getLabel() + "/" + arg_num <<
-////						" for " << vhit->varname << endl;
-//				heap_cell->val.closure.arg1_name = new string(this->getLabel() + "/" + arg_num);
-//				break;
-//			}
-//			case binaryprimopExprClosure:
-//			{
-//				assert(!heap_cell->inWHNF);
-////				cout << "Binaryprimop" << endl;
-//				auto arg_num = heap_cell->val.closure.arg1_name->substr(heap_cell->val.closure.arg1_name->rfind("/")+1);
-//				heap_cell->val.closure.arg1_name = new string(this->getLabel() + "/" + arg_num);
-////				cout << "Updating argument prg_pt to " << this->getLabel() + "/" + arg_num << endl;
-//				arg_num = heap_cell->val.closure.arg2_name->substr(heap_cell->val.closure.arg2_name->rfind("/")+1);
[…]

### 2015-11-28 lazy-scheme/src/SchemeAST_Defs.cpp
added debugging statements

@@ -807,7 +807,8 @@ cons* LetExprNode::evaluate()
 	if (gc_status != gc_disable && (gc_status == gc_plain || gc_status == gc_freq))
     {
 
-		if (gc_status == gc_freq && (GC_STAT_GET_CLOCK() - last_gc_clock > GC_FREQ_THRESHOLD())) {
+		if (gc_status == gc_freq && (GC_STAT_GET_CLOCK() - last_gc_clock > GC_FREQ_THRESHOLD()))
+		{
 			reachability_gc();
 			//return_stack().return_point = curr_let_pgmpt;
 			GC_STAT_DUMP_GARBAGE_STATS();

### 2015-11-28 lazy-scheme/src/Simulator.cpp

@@ -192,7 +192,7 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 
 		automaton *nfa = Scheme::Demands::getNFAsFromRegularGrammar(&gLivenessData, pgmname);
 		Scheme::Demands::printNFAToFile(nfa, outdir + pgmname + "/program-nfa.txt");
-
+//		exit(1);
 		std::unordered_set<std::string> start_states;
 		for (auto nt:gLivenessData)
 			start_states.insert(nt.first);

### 2015-11-28 lazy-scheme/post_process/gen_stats_and_plotall.sh
bm data updates

@@ -53,7 +53,7 @@ limit["lambda"]=20466
 limit["knightstour"]=677800
 limit["fft"]=172762
 limit["huffman"]=100070
-limit["paraffins"]=425476
+limit["paraffins"]=5185
 limit["deriv"]=11124
 ## UNUSED BMs
 limit["small"]=17
@@ -77,7 +77,7 @@ freq["nqueens"]=3000;
 freq["lambda"]=100;
 freq["huffman"]=100;
 freq["deriv"]=20;
-freq["paraffins"]=100;
+freq["paraffins"]=2;
 
 declare -A xtics
 declare -A ytics
@@ -96,7 +96,7 @@ xtics["nqueens"]=3.5e5          ; ytics["nqueens"]=4.6e3
 xtics["lambda"]=19e5            ; ytics["lambda"]=3e3
 xtics["huffman"]=19e5           ; ytics["huffman"]=3e3
 xtics["deriv"]=100             ; ytics["deriv"]=100
-xtics["paraffins"]=19e5         ; ytics["paraffins"]=3e3
+xtics["paraffins"]=100         ; ytics["paraffins"]=100
 
 declare -A xlimit
 declare -A ylimit

### 2015-11-28 lazy-scheme/src/run_in_loop.py

@@ -5,7 +5,7 @@ import time
 import datetime
 from collections import namedtuple
 
-prog_size = {'nqueens':22729, 'nperm':27429, 'fibheap':38001, 'treejoin':1616535, 'sudoku':4075, 'lcss':22243, 'lambda':20466, 'gc_bench':204850, 'knightstour':680000}
+prog_size = {'nqueens':10101, 'nperm':27428, 'fibheap':37043, 'treejoin':1616533, 'sudoku':4066, 'lcss':22243, 'lambda':20466, 'gc_bench':204850, 'knightstour':677800, 'huffman':100070, 'paraffins':5185, 'deriv':11124}
 gc_info = namedtuple("gc_info", "heap_total heap_left heap_used gc_invocations gc_time dfa_time num_states exec_time")
 gc_stat = namedtuple("gc_stat", "prog_name reachability liveness")
 gc_stat_list = []

### 2015-11-28 lazy-scheme/src/run_in_loop.py
python script.

@@ -1,3 +1,4 @@
+#!/usr/bin/python
 import subprocess
 import sys,os
 import os.path

### 2015-11-28 paper/paper.tex
latest - 28th

@@
 
-The  semantics  of   expressions  (and  applications)
+\warning{The  semantics  of   expressions  (and  applications)
 %% \footnote{In  most
@@
 %%   for both expressions and applications.} 
-are given by transitions of
-the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
-Here \stk\ is a stack of continuation frames.  Each continuation frame
-is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
-has  to  be  updated  with  the  value  of  the  currently  evaluating
-expression, and $e$ is to be evaluated next in the environment $\rho$.
-Note  that  the  state  of  the  transition  system  consists  of  the
-\emph{current evaluation  context}, given by  $\rho$ and $e$,  and the
-\emph{suspended  evaluation  contexts}  (represented  as  continuation
-frames) in the stack $S$.   
+are given by transitions of the form $\rho, \stk, \heap, e \rightarrow
+\rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
+frames.  Each  continuation frame  is of the  form $(\rho,  \ell, e)$,
+signifying that the  location $\ell$ has to be updated  with the value
+of the  currently evaluating  expression, and $e$  is to  be evaluated
+next  in the  environment $\rho$.   We shall call the $\rho$ and
+the $e$ on the LHS of a $ \stk,
+\heap,  e$ is the state of  the  transition system, then we shall there is a   \emph{current
+  evaluation  context}, given  by  $\rho$, $\heap$  and  $e$, and  
+\emph{suspended  evaluation  contexts}  (represented  by the
+environment, heap and expression in each   continuation
+frame) in the stack $S$.}
 %% \warning{The
@@
 
-\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
-
-\subsection{Generating Equations for \Lfonly$_\mathit{f}$}
+\subsection{Generating Equations for $\Lfonly_{\mathit f}$}
 Given a  function $\mathit{f}$, we now  describe how to
@@
 $\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
-
-\begin{eqnarray*}
-&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
+$$
+ \Lf{\length}{1}{\sigma} = \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
   \cup \clazy\Lf{\length}{1}{\clazy\sigma}
-\end{eqnarray*}
-
+$$
 %% and  the only equation defining \Lfone{\length} is:
@@
 %% \end{eqnarray*}
-and  the equation defining \Lfone{\fun} is:
-\begin{eqnarray*}
-  && \Lf{\fun}{1}{\sigma}
-  =  \clazy\sigma \cup \Lf{\length}{1}{\bcar\sigma_{\mathit{\length}}} \cup  \Lf{\length}{1}{\clazy\sigma} 
-  \cup  \acdr\Lf{\fun}{1}{\bcdr\sigma_{\mathit{\fun}}}
-\end{eqnarray*}
 
@@
 \end{eqnarray}
-
 where \Df{f}{i}  are sets of strings  over the alphabet
@@
   \begin{eqnarray*}
-&&  \Lf{\fun }{1}{\sigma} = \Df{\fun}{1}\sigma,~\mbox{and}\\
-&&   \Df{\fun}{1} =  \clazy \cup \Df{\length}{1}\bcar \cup  \Df{\length}{1}\clazy 
-       \cup  \acdr\Df{\fun}{1}\bcdr
+&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
+&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy
+       \cup \clazy\Df{\length}{1}{\clazy}
   \end{eqnarray*}
 
-Note that this equation can also be viewed as a CFG with \{\acdr, \bcar, \bcdr,
-\clazy\} as terminal symbols and \Df{\length}{1} and \Df{\fun}{1} as the 
-non-terminals.
+Note that this equation can also be viewed as a CFG with \{\acdr,
+\clazy\} as terminal symbols and \Df{\length}{1} as the sole
+non-terminal.
 
@@
 
-For  the  running  example,   $\fun$  has  calls  from  $\main$  at
-$\pi_{16}$ and a recursive call at $\pi_6$.
-So $\sigma_{\fun} =
-     \deltacall{\fun}{\pi_{16}}{\main}  \cup
-\deltacall{\fun}{\pi_6}{\fun}$.
-
-%% For  the  running  example,   $\length$  has  calls  from  $\main$  at
-%% $\pi_{12}$ and a recursive call at $\pi_6$.
-%% So $\sigma_{\length} =
-%%      \deltacall{\length}{\pi_{12}}{\main}  \cup
-%% \deltacall{\length}{\pi_6}{\length}$.
-
+For  the  running  example,   $\length$  has  calls  from  $\main$  at
+$\pi_{12}$ and a recursive call at $\pi_6$.
+So $\sigma_{\length} =
+     \deltacall{\length}{\pi_{12}}{\main}  \cup
+\deltacall{\length}{\pi_6}{\length}$.
 Filling in  the values gives:
 \begin{eqnarray*}
-\sigma_{\fun}    &=&
- \{\sigma_{\mathit{all}}\}  ~\cup~{\bcdr\sigma_{\mathit{\fun}}}
+\sigma_{\length}    &=&
+ \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
 \end{eqnarray*}
-As examples, the liveness of \Lanv{2}{\pl} and \Lanv{1}{\py} in terms
-of  $\sigma_{\fun}$ are:
+As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
+of  $\sigma_{\length}$ are:
 \begin{align*}
-\Lanv{2}{\pl} &= \clazy \cup \acar\Lf{\fun}{2}{\bcdr\sigma_{\mathit{\fun}}}
+\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
+  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
  \\
-\Lanv{1}{\py} &= \clazy \cup \bcar\sigma_{\mathit{\fun}}
+\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
 \end{align*}
@@
 \begin{eqnarray*}
-  \var{\Lanv{3}{\xl}}  &\rightarrow& 
-  \clazy \mid \var{\Df {\Sum}{1}}{\clazy\var{\sigma_{\length}}}  \\
+  \var{\Lanv{1}{\xl}}  &\rightarrow& 
+  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
   & & \mid
-  \clazy\var{\Df{\length}{1}}\clazy \mid \acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
+  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
   \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
@@
 &\rightarrow&
-\clazy  \mid \bcar\langle{\sigma_{\Sum}}\rangle \\
- \var{\Df{\fun}{2}} &\rightarrow& \var{\Df{\makelist}{1}}\clazy \mid \var{\Df{\makelist}{1}}\var{\Df{\Sum}{1}}\clazy \\ 
-& & \mid \var{\Df{\makelist}{1}}\acar\var{\Df{\fun}{2}}\bcdr\var{\sigma_{\mathit{\fun}}} \\
-\langle\sigma_{\mathit{\fun}} \rangle &\rightarrow& \sigma_{all} \mid \bcdr\var{\sigma_{\mathit{\fun}}} \\
-\var{\Df{\Sum}{1}} &\rightarrow& \clazy \mid \acar\clazy \mid \acdr\var{\Df{\Sum}{1}}\clazy
-%% \var{\Lanv{3}{\py}} &\rightarrow& \clazy \mid \bcar\var{\sigma_{\mathit{\fun}}}
+\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
+\var{\Lanv{9}{\pa}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
 \end{eqnarray*}
@@
 solution  of  \Lanv{i}{x}.    
-
-\warning{
-\hrule
-\medskip
-Fixing from here
-\medskip
-\hrule
-}
-
 However, recall that the decision problem that we are interested in
@@
 Definition~\ref{def:specialgrammar} is undecidable. However note that
-the grammar may not be necessarily generated from liveness analysis.
+the grammars may not be necessarily generated from liveness analysis.
 \newcommand{\state}{\ensuremath{\mathsf{S}}}
@@
 
-After  the TM  has halted,  there are  further ``cleanup''
-derivations  that derive  $\epsilon$ only  if the  invariant  has been
-maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
+After the  TM has  halted, there  are further  ``cleanup'' derivations
+that derive  $\epsilon$ only if  the invariant has been  maintained so
+far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
 $\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
-final state of the TM. We add
-productions $\nont{\mathit{final}}{c} \rightarrow \acar
-\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
-\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
-the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
-\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow 
-\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
-right of the tape head. Finally the production $\clazy\$ \rightarrow
-\epsilon$ is used to derive $\epsilon$.This completes the reduction.
+final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
+\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
+$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
+for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
+and                $\nont{\mathit{final}}{c}               \rightarrow
+\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
+\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
+\acdr\ on the right of the tape head. Finally the production $\clazy\$
+\rightarrow \epsilon$ is used  to derive $\epsilon$.This completes the
+reduction.
 \end{proof}
@@
 
-\begin{lemma}
+\begin{lemma}\ref{lemma:grammar-from-analysis-undecidable}
 Consider    a    grammar    $G$    of   the    kind    described    in
-Definition~\ref{def:specialgrammar} that is generated by liveness analysis
-of a program and a forward access path $\alpha$. The membership problem
-$\alpha\$ \in \mathscr{L}(G)$ is also undecidable.
+Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
+analysis  of a  program,  and  a forward  access  path $\alpha$.   The
+membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
 \end{lemma}
 
-Given a Turing machine and an input string, we now give program fragments whose liveness
-analysis generate the same productions as in
-Lemma~\ref{lemma:grammar-undecidable}, but with  extra productions
-that necessarily contain the \clazy\ symbol. For instance,
-corresponding to the production $\nont{i}{c} \rightarrow c
-\nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, we have
-the program (define ($\nont{i}{c}$ a) (let b <- (car a) in (let c <-
-($\nont{j}{\acar}$ b) in (let d <- 5 in (let e <- (cons c d) in (let f
-<- (null? e) in return f)))))). The reader can check that this
-produces the grammar fragment ...
-
-It is obvious that if the TM halts on the input, then the grammar will
-produce the $\epsilon$ string. However, if the TM does not halt on the
-input, then the productions containg
+\begin{proof}
+Given a Turing machine and an input string, we shall define a function
+for   each  of   the  non-terminals   introduced  in   the  proof   of
+Lemma~\ref{lemma:grammar-undecidable}.
+%% Liveness  analysis  of  $f$  will  result in  not  only  contain  a
+%% production  $P$ itself,  but could  also contain  extra productions
+%% that necessarily contain the \clazy\ symbol.
+For instance, corresponding to the production $\nont{i}{c} \rightarrow
+c \nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, consider
+the program:
+
+If there more than one productions with the same LHS, then the
+corresponding function creates joins them with a $\SIF$.   If there
+are $n$ such functions,  then \mainpgm\  creates  a $n$-way
+branch  and inserts  a  single call  to a  distinct  function in  each
+branch. It can  be verified that the relevant part  of the grammar for
+the production $\nont{i}{c} \rightarrow
+c \nont{j}{\acar}\bcar$ shown earlier is:
+
+Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
+had a single non-terminal on the RHS. There are two characteristics of
+the grammar produced from liveness-analysis  that are are relevant for
+replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
+\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
+grammar  generated from  liveness  analysis of the constructed program
+will  have the  production
+$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
+\nont{\mathit{all}}{}$,  and (b)  other  productions are  of the  form
+$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
+\nont{\mathit{all}}{}$, where $\beta_1'$ and $\beta_2'$ do not contain
+any non-terminals.
+
+It is clear that if the TM given as an instance of the halting problem
+accepts the input string, then the earlier derivation can be replayed,
+replacing   $\nont{\mathit{all}}{}$   in   the  sentential   form   by
+$\epsilon$. However, if the TM does  not accept the input string, then
+any  sentential  form derived  from  the  start  symbol would  have  a
+non-terminal different from $\nont{\mathit{final}}{c}$. Thus the empty
+string would not be derivable from the grammar.
+\end{proof}
 

### 2015-11-28 paper/paper.tex
latest - 28th

@@
 
-\begin{lemma}\ref{lemma:grammar-from-analysis-undecidable}
+\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
 Consider    a    grammar    $G$    of   the    kind    described    in
@@
 
+\input{equivalent-program}
+
 If there more than one productions with the same LHS, then the
@@
 branch  and inserts  a  single call  to a  distinct  function in  each
-branch. It can  be verified that the relevant part  of the grammar for
-the production $\nont{i}{c} \rightarrow
-c \nont{j}{\acar}\bcar$ shown earlier is:
+branch. It can  be verified that the grammar for the function shown
+above, includes the productions:
+\begin{align*}
+\nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{all}{} \\
+\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{all}{} \\
+\nont{all}{} &\rightarrow \epsilon \mid \acar\nont{all}{} \mid
+\acdr\nont{all}{}
+\end{align*}
 

### 2015-11-28 paper/paper.tex
latest - 28th

@@
 
-\warning{The  semantics  of   expressions  (and  applications)
-%% \footnote{In  most
-%%   contexts, we  shall use the  term 'expression' and the  notation $e$
-%%   for both expressions and applications.} 
-are given by transitions of the form $\rho, \stk, \heap, e \rightarrow
+The  semantics  of   expressions  (and  applications\footnote{In  most
+ contexts, we  shall use the  term 'expression' and the  notation $e$
+  for both expressions and applications.}) 
+are given by transitions of the form $\rho, \stk, \heap, e \rightsquigarrow
 \rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
@@
 of the  currently evaluating  expression, and $e$  is to  be evaluated
-next  in the  environment $\rho$.   We shall call the $\rho$ and
-the $e$ on the LHS of a $ \stk,
-\heap,  e$ is the state of  the  transition system, then we shall there is a   \emph{current
-  evaluation  context}, given  by  $\rho$, $\heap$  and  $e$, and  
-\emph{suspended  evaluation  contexts}  (represented  by the
-environment, heap and expression in each   continuation
-frame) in the stack $S$.}
-%% \warning{The
+next  in the  environment $\rho$.  
+%% Given a state $\rho, \stk, \heap,
+%% e$  of  the  transition system, we shall call the pair $(\rho, e)$ as
+%% the    \emph{current
+%%   evaluation  context}, and for a continuation frame $(\rho',  \ell, e')$ in
+%% \stk, we shall call the pair $(\rho', e')$ as a   
+%% \emph{suspended  evaluation  context}.}
+%% %% \warning{The
 %% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
 %% system will be denoted $\mathcal{S}$.}
-
 The  initial  state  of  the  transition  system  described  above  is
@@
 gives the patterns of future uses of a variable, demand represents the
-future use of the value of an expression.  The demand on an expression
+future uses of the value of an expression.  The demand on an expression
 $e$   is   again   a   set    of   access   paths---that   subset   of
@@
 replacing   $\nont{\mathit{all}}{}$   in   the  sentential   form   by
-$\epsilon$. However, if the TM does  not accept the input string, then
-any  sentential  form derived  from  the  start  symbol would  have  a
-non-terminal different from $\nont{\mathit{final}}{c}$. Thus the empty
-string would not be derivable from the grammar.
+$\epsilon$. Conversely,  if the TM  does not accept the  input string,
+then  any sentential  form derived  from the  start symbol  would have
+exactly one non-terminal, and this non-terminal will be different from
+$\nont{\mathit{final}}{c}$.  Thus  the  empty   string  would  not  be
+derivable from the grammar.
 \end{proof}

### 2015-11-28 lazy-scheme/post_process/createTable.sh
min cell reqd in script

@@ -1,24 +1,50 @@
-#!/bin/sh
+#!/bin/bash
 DIR=`dirname $0`
 SCRIPT1=$DIR/touchedAndCopied.awk
 AllBMs="lambda nperm treejoin lcss sudoku fibheap nqueens gc_bench"
-AllBMs="lambda nperm treejoin lcss sudoku fibheap nqueens knightstour gc_bench "
+AllBMs="fibheap sudoku nperm paraffins lcss huffman knightstour nqueens deriv treejoin lambda "
 #AllBMs="sudoku"
-echo "\\\\newcommand{\\\\rlratio}{$\\\\frac{RGC}{LGC}$}"
+echo "\\newcommand{\\rlratio}{$\\frac{RGC}{LGC}$}"
 cat <<EOF
-\begin{tabular}{|@{\ }c@{\ }| @{\ }r@{\ }|@{\ }r@{\ }| @{\ }r@{\ }|@{\ }r@{\ }| r@{\ }|r@{\ }| r@{\ }|r@{\ }|r@{\ }| r@{\ }|r@{\ }|r@{\ }|}
+\begin{tabular}{|@{\ }c@{\ }| @{\ }r@{\ }|@{\ }r@{\ }| @{\ }r@{\ }|@{\ }r@{\ }|r@{\ }| r@{\ }|r@{\ }|r@{\ }| r@{\ }|r@{\ }|r@{\ }|}
 \hline
   &   \multicolumn{2}{@{}c@{}|}{\#Cells collected/GC}
   &   \multicolumn{2}{c@{}|}{\#Cells touched/GC}
   &   \multicolumn{2}{c@{}|}{\#GCs}
-  &   \multicolumn{3}{c@{}|}{\#Drag Cells}
+  &   \multicolumn{2}{c@{}|}{Min. Cells Required}
   &   \multicolumn{2}{c@{}|}{GC time (sec)} &Speedup \\\\
 \cline{2-12}
-{Program}&RGC&LGC&RGC&LGC&RGC&LGC&RGC&LGC&\rlratio&RGC&LGC&\rlratio\\\\
+{Program}&RGC&LGC&RGC&LGC&RGC&LGC&RGC&LGC&RGC&LGC&\rlratio\\\\
 \hline
 \hline
 
 EOF
+declare -A Lmin
+Lmin["fibheap"]=37043
+Lmin["sudoku"]=2960
+Lmin["nperm"]=25343
[…]

### 2015-11-28 paper/paper.tex

@@
   reside in the activation stack). Notationally, we distinguish
-  closure variables from their corresponding stack variables by
-  subscripting them with their program points of
-  creation.\footnote{Mutiple occurrences of the same variable in an
+  a closure variable from its corresponding stack variable by
+  subscripting it with the program point where the closure was created\footnote{Mutiple occurrences of the same variable in an
     application are further distinguished by their positions in the
-    application.}
+    application.}.
 

### 2015-11-28 paper/paper.tex
table and fig.

@@
 \renewcommand{\arraystretch}{.1}
-\begin{tabular}{@{}c@{}@{}c@{}}
-%%    \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
-%%    & {\epsfig{file=lambda_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
-   \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
+\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
+   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-\\   \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=paraffins_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-\\   \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-\\  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
 %\\ (a) & (b)

### 2015-11-29 paper/paper.tex
minefield bang, and \sc command fix
Added graphs for all BMs

@@
           & $\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\, \kappa
-  \rightsquigarrow \rho',\, S,\, \heap[\ell := \kappa],\, e$    &  \sc{const}
+  \rightsquigarrow \rho',\, S,\, \heap[\ell := \kappa],\, e$    &  {\sc const}
 \\
@@
 \rightsquigarrow
-$  $\rho',\, S,\, \heap[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
+$  $\rho',\, S,\, \heap[\ell := (\rho(x), \rho(y))],\, e$}     &  {\sc cons} \\
 \hline
@@
 (\CAR~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$      &
-\sc{car-select} \\
+{\sc car-select} \\
 \hline
@@
 (\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho'\rangle), (\CAR~x))\!:\!S,\, \heap,\, s$      &
-\sc{car-1-clo} \\
+{\sc car-1-clo} \\
 \hline
@@
 \rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,\heap,\, s$      &
-\sc{car-clo}
+{\sc car-clo}
 \\
@@
 $\rho', \,S,\, \heap[\ell := \heap(\rho'(x)) + \heap(\rho'(y))],\, e$}      &
-\sc{prim-add} \\
+{\sc prim-add} \\
 \hline
@@
 \rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
-\sc{prim-1-clo} \\
+{\sc prim-1-clo} \\
 \hline
@@
 \rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
-\sc{prim-2-clo} \\
+{\sc prim-2-clo} \\
 \hline
@@
 [\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
-\sc{funcall} \\
+{\sc funcall} \\
 \hline
@@
     \rangle],\, e$} &
-\sc{let} \\
+{\sc let} \\
 \hline
 $\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho,\, S,\, \heap,\,  e_1$ & \sc{if-true} \\
+\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
 \hline
 $\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho,\, S, \,\heap, \, e_2$ & \sc{if-false} \\
+\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
 \hline
@@
 &
-\sc{if-clo} \\
+{\sc if-clo} \\
 \hline
@@
 (\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
-\sc{return-whnf}\\
+{\sc return-whnf}\\
 \hline
@@
 $\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
-\sc{return-clo} \\
+{\sc return-clo} \\
 \hline
@@
 %%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
-%% \sc{car-clo-bang}
+%% {\sc car-clo-bang}
 %% \\
@@
 %%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
-%% \sc{car-clo}\\
+%% {\sc car-clo}\\
 %% \hline
@@
 %%   \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(x) :=
-%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
+%%     \kappa], e, \cred{\sigma'}$ }   &  {\sc const}
 %% \\
@@
 %%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
-%%   \rho', S, \heap[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+%%   \rho', S, \heap[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  {\sc cons} \\
 %% \hline
@@
 %%   \rightsquigarrow \rho', S, \heap[\rho'(z) := \heap(\ell_1)], e, \cred{\sigma'}$}      &
-%% \sc{car-whnf} \\
+%% {\sc car-whnf} \\
 %% \hline
@@
 %%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
-%% \sc{car-clo}
+%% {\sc car-clo}
 %% \\
@@
 %% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
-%% \sc{car-1-clo} \\
+%% {\sc car-1-clo} \\
 
@@
 %%     := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
-%% \sc{prim-whnf} \\
+%% {\sc prim-whnf} \\
 %% \hline
@@
 %%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-%% \sc{prim-1-clo} \\
+%% {\sc prim-1-clo} \\
 %% \hline
@@
 %%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-%% \sc{prim-2-clo} \\
+%% {\sc prim-2-clo} \\
 %% \hline
@@
 %%     \rho(\myvec{x})], S, \heap, e_{\mathit{g}}, \cred{\sigma}$}      &
-%% \sc{funcall} \\
+%% {\sc funcall} \\
 %% \hline
@@
 %%   \ell]\rangle], e, \cred{\sigma}$} &
-%% \sc{let} \\ 
+%% {\sc let} \\ 
 %% \hline
 %% $\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
-%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
+%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
 %% \hline
@@
 %%   \cred{\sigma}$   $\rightsquigarrow
-%% \rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
+%% \rho, S, \heap,  e_2, \cred{\sigma}$} & {\sc if-false} \\
 %% \hline
@@
 %% &
-%% \sc{if-clo} \\
+%% {\sc if-clo} \\
 %% \hline
@@
 %%   \cred{\sigma'}$} &
-%% \sc{return-whnf}\\
+%% {\sc return-whnf}\\
 %% \hline
@@
 %%   \cred{\sigma}$} &
-%% \sc{return-clo} \\
+%% {\sc return-clo} \\
 %% \hline
@@
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
-\sc{car-clo-bang} 
+{\sc car-clo-bang} 
 \\
@@
 \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \ell,
   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
-\sc{car-clo}\\
+{\sc car-clo}\\
 \hline
@@
 \end{center}
-\caption{Minefield semantics for \sc{car-clo-bang} and
-  \sc{car-clo-bang} \label{fig:minefield-semantics-for-some}}
+\caption{Minefield semantics for {\sc car-clo-bang} and
+  {\sc car-clo} \label{fig:minefield-semantics-for-some}}
 \end{figure*}
@@
 &  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=gc_bench_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
 %\\ (a) & (b)
@@
   \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell :=
-    \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
+    \kappa], e, \cred{\sigma'}$ }   &  {\sc const}
 \\
@@
 & \makecell{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
-  \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
+  \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  {\sc cons} \\
 \hline
@@
   \rightsquigarrow \rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
-\sc{car-select} \\
+{\sc car-select} \\
 \hline
@@
 s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
-\sc{car-1-clo} \\
+{\sc car-1-clo} \\
 
@@
   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
-\sc{car-clo}
+{\sc car-clo}
 \\
@@
     := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
-\sc{prim-add} \\
+{\sc prim-add} \\
 \hline
@@
   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-\sc{prim-1-clo} \\
+{\sc prim-1-clo} \\
 \hline
@@
   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-\sc{prim-2-clo} \\
+{\sc prim-2-clo} \\
 \hline
@@
     \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
-\sc{funcall} \\
+{\sc funcall} \\
 \hline
@@
   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
-\sc{let} \\ 
+{\sc let} \\ 
 \hline
 $\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
+  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
 \hline
@@
   \cred{\sigma}$   $\rightsquigarrow
-\rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
+\rho, S, \heap,  e_2, \cred{\sigma}$} & {\sc if-false} \\
 \hline
@@
   \cred{\sigma'}$} &
-\sc{return-whnf}\\
+{\sc return-whnf}\\
 \hline
@@
   \cred{\sigma}$} &
-\sc{return-clo} \\
+{\sc return-clo} \\
 \hline

### 2015-11-29 lazy-scheme/post_process/createTable.sh
name change for col header

@@ -12,7 +12,7 @@ cat <<EOF
   &   \multicolumn{2}{c@{}|}{\#Cells touched/GC}
   &   \multicolumn{2}{c@{}|}{\#GCs}
   &   \multicolumn{2}{c@{}|}{Min. Cells Required}
-  &   \multicolumn{2}{c@{}|}{GC time (sec)} &Speedup \\\\
+  &   \multicolumn{2}{c@{}|}{GC time (sec)} & Ratio \\\\
 \cline{2-12}
 {Program}&RGC&LGC&RGC&LGC&RGC&LGC&RGC&LGC&RGC&LGC&\rlratio\\\\
 \hline

### 2015-11-29 lazy-scheme/post_process/createTable.sh
added gc_bench.

@@ -2,7 +2,7 @@
 DIR=`dirname $0`
 SCRIPT1=$DIR/touchedAndCopied.awk
 AllBMs="lambda nperm treejoin lcss sudoku fibheap nqueens gc_bench"
-AllBMs="fibheap sudoku nperm paraffins lcss huffman knightstour nqueens deriv treejoin lambda "
+AllBMs="fibheap sudoku nperm paraffins lcss huffman knightstour nqueens deriv treejoin lambda gc_bench "
 #AllBMs="sudoku"
 echo "\\newcommand{\\rlratio}{$\\frac{RGC}{LGC}$}"
 cat <<EOF
@@ -31,6 +31,7 @@ Lmin["nqueens"]=1082
 Lmin["deriv"]=589
 Lmin["treejoin"]=677800
 Lmin["lambda"]=18169
+Lmin["gc_bench"]=72
 
 declare -A Rmin
 Rmin["fibheap"]=37043
@@ -44,6 +45,7 @@ Rmin["nqueens"]=10101
 Rmin["deriv"]=11124
 Rmin["treejoin"]=1616533
 Rmin["lambda"]=20466
+Rmin["gc_bench"]=204813
 
 OUTPUTDIR=output.GOLD
 for BM in $AllBMs

### 2015-11-29 paper/paper.tex
new plots

@@
 &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=paraffins_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
@@
 \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=gc_bench_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}

### 2015-11-29 paper/paper.tex
latest - 29th

--- 2015-11-28 paper/paper.tex
@@
 determinable execution  order and  the presence  of closures  pose new
-challenges  for lazy  languages.  These  requires changes  both in  the
+challenges  for lazy  languages.  These  require changes  both in  the
 liveness analysis itself and in the design of the garbage collector.
@@
 While  the runtime  system  of most  functional  languages includes  a
-garbage  collector  (GC)  to  efficiently  reclaim  memory,  empirical
+garbage  collector    to  efficiently  reclaim  memory,  empirical
 studies on Scheme~\cite{karkare06effectiveness} and, more importantly,
-on  Haskell~\cite{rojemo96lag}  programs  have shown  that  GCs  leave
+on  Haskell~\cite{rojemo96lag}  programs  have shown  that  garbage collectors  leave
 uncollected a  large number of  memory objects that are  reachable but
@@
 {\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
-expression.   The  analysis has  two  parts.   First, we  calculate  a
+expression.   The  analysis  has  two parts.   We  first  calculate  a
 context-sensitive  summary   of  each   function  as  a   {\em  demand
@@
 demands  on its  arguments.   This  summary is  used  to step  through
-function  calls during  analysis.  Second,  the concrete  demand on  a
-function  body is  obtained through  a conservative approximation
-similar to 0-CFA~\cite{Shivers:1988} that combines the  demands on all the calls
-to  the function.   The result  of the  analysis is  an annotation  of
+function  calls  during analysis.   Next,  the  concrete demand  on  a
+function body is obtained through a conservative approximation similar
+to  0-CFA~\cite{Shivers:1988} that  combines  the demands  on all  the
+calls to the function.  The result of the analysis is an annotation of
 certain program points with  deterministic finite-state automata (DFA)
 capturing the liveness of variables at these points.  Depending on the
-program point where garbage collection has been triggered, the garbage
-collector consults a  set of automata to  restrict reachability during
-marking.  This  results in  an increase in  the garbage  reclaimed and
-consequently in fewer garbage collections.
+program  point where  garbage collection  has been  triggered, the
+garbage collector
+consults a  set of automata  to restrict reachability  during marking.
+This results in an increase  in the garbage reclaimed and consequently
+in fewer garbage collections.
 
@@
 evaluation  takes  place  cannot be  statically  determined,  laziness
-complicates  liveness analysis  itself.  Secondly,  for liveness-based
-garbage collection to  be effective, we need  to extend liveness-based
-collection to closures apart from evaluated data.  Since a closure can
-escape the scope in which it was created, it has to carry the liveness
-information of  its free variables. Moreover,  as execution progresses
-and  possible  future uses  are  eliminated,  we update  the  liveness
-information  in a  closure with  a more  precise version.   Because of
-these reasons,  the garbage collector also  becomes significantly more
-complicated than a liveness-based collector for a eager language.
+complicates  liveness analysis  itself.  Moreover,  for liveness-based
+garbage collection to  be effective, we need to extend  it to closures
+apart from  evaluated data.  Since a  closure can escape the  scope in
+which it was created, it has  to carry the liveness information of its
+free variables. Moreover, as  execution progresses and possible future
+uses are eliminated,  we update the liveness information  in a closure
+with a  more precise version.  Because of these reasons,   the garbage
+collector  also   becomes  significantly   more  complicated   than  a
+liveness-based collector for a eager language.
 
@@
   benefits.   Liveness-based  collection  results in  an  increase  in
-  garbage reclaimed.   As a  consequence, there is  a decrease  in the
-  number of  collections (up to 45\%)  and a decrease (up  to 88\%) in
-  the number  of {\em dragged}  cells (cells  that survive a  GC after
-  their  last use).  Besides,  there  is a  reduction  in the  overall
-  execution time for some programs.  Our experiments show that a mixed
-  strategy, where we  do liveness based GC up to  the closure boundary
-  and reachability based GC for  the arguments (free variables) of the
-  closure,  gives us  similar benefits  as a  fully liveness  based GC
-  (where  the  arguments  of  the closure  are  also  collected  using
-  liveness) with lesser overhead (Section~\ref{sec:strategies}).}
+  garbage reclaimed.   As a  consequence, there is  a reduction  in the
+  number of collections up to ???\% and a decrease in the
+  minimum memory requirement up to ???\%. 
+  As an added benefit, there is also a reduction in the overall
+  execution time for some programs.}
 
@@
 (represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
-  references} and represent  variables  or fields.  The figure shows
+  references} and represent values of  variables  or fields.  The figure shows
 the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
@@
 %\comment{  
-Consider  the program in  Figure~\ref{fig:mot-example2}.  As
-  mentioned earlier,  a lazy  evaluation of  the $\LET$  expression at
-  $\pi_1$ creates a closure for $(length~x)$ instead of evaluating it.
-  Since the  closure may escape the  scope in which it  is created, it
-  carries a copy  of $x$ within itself.   We treat the copy  of $x$ in
-  the closure as being separate from the $x$ introduced by the $\LET$,
-  and call  it a \emph{closure variable}.   For liveness calculations,
-  such variables are distinguished from variables introduced by \LET s
-  and  function arguments  (called \emph{stack  variables} since  they
-  reside in the activation stack). Notationally, we distinguish
-  a closure variable from its corresponding stack variable by
-  subscripting it with the program point where the closure was created\footnote{Mutiple occurrences of the same variable in an
-    application are further distinguished by their positions in the
-    application.}.
+Consider the  program in Figure~\ref{fig:mot-example2}.   As mentioned
+earlier, a lazy evaluation of the $\LET$ expression at $\pi_1$ creates
+a closure  for $(\length~\px)$  instead of  evaluating it.   Since the
+closure may escape the scope in which it is created, it carries a copy
+of $\px$ within itself.  We treat the  copy of $\px$ in the closure as
+being separate from the $\px$ introduced  by the $\LET$, and call it a
+\emph{closure  variable}.  For  liveness calculations,  such variables
+are distinguished  from variables  introduced by  \LET s  and function
+arguments  (called \emph{stack  variables}  since they  reside in  the
+activation  stack). Notationally,  we distinguish  a closure  variable
+from  its corresponding  stack variable  by subscripting  it with  the
+program   point  where   the   closure  was   created\footnote{Mutiple
+  occurrences  of the  same  variable in  an  application are  further
+  distinguished by their positions in the application.}.
 
@@
   dependences beginning at  a variable at an evaluation  point.  As an
-  example,  the  variable $z$  returned  at  $\psi_2$ depends  on  $y$
-  through  the  expression $(+~y~1)$.   $y$  in  turn depends  on  the
-  closure  variable  $x_{\pi_1}$   through  $(length~x_{\pi_1})$.   We
-  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!z \leftarrow
-  (+~y~1),\,  y  \leftarrow   (length~x_{\pi_1})\rbrack$.   Indeed,  the
+  example, consider the following program:
+
+\input{example-to-illustrate-liveness}
+
+The  variable \pz\   returned  at  $\psi_2$ depends  on  \py\ 
+  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
+  closure  variable  $\px_{\pi_1}$   through  $(\length~\px_{\pi_1})$.   We
+  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
+  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\pi_1})\rbrack$.   Indeed,  the
   closures  in  the  heap  are   a  runtime  representation  of  these
-  dependences.  Since  $z$  is  evaluated   at  $\psi_2$  due  to  the
-  expression $\SRETURN~z$,  the demand made by  the calling context(s)
-  of  $f$ places  a demand  on  $z$ which  will impart  a liveness  to
-  $x_{\pi_1}$.  Other dependence chains which result in a liveness for
-  $x_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!y       \leftarrow
-  (length~x_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!w \leftarrow (/~u~y),\,\,
-  y  \leftarrow  (length~x_{\pi_1})\rbrack$.   The  liveness  analysis
-  described in this section declares the liveness of $x_{\pi_1}$ to be
+  dependences.  Since  $\pz$  is  evaluated   at  $\psi_2$  due  to  the
+  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
+  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
+  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
+  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
+  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
+  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
+  described in this section declares the liveness of $\px_{\pi_1}$ to be
   a union of the liveness arising  out of these dependence chains.  To
-  be safe, a  garbage collection during evaluation of  $y$ at $\psi_1$
-  has  to  use  this  liveness  to copy  the  heap  from  $x_{\pi_1}$.
+  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
+  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
   However,  notice that  if  a garbage  collection  takes place  while
-  evaluating $z$ at $\psi_2$, it can safely consider only the liveness
-  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!z \leftarrow
-  (y+1),\,   y  \leftarrow   (length~x_{\pi_1})\rbrack$.   The   garbage
+  evaluating $\pz$ at $\psi_2$, it can safely consider only the liveness
+  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
+  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
   collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
@@
 
-%% \begin{figure*}[t!]
-%% \begin{center}\footnotesize
-%% \renewcommand{\arraystretch}{1.5}
-
-%% \begin{tabular}{|c|c|c|}
-%% \hline
-%% Premise & Transition & Rule name \\ 
-%% \hline
-
-%% \hline
-%% \makecell[t]{$GC(\rho_1, S, \heap_1, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
-%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
-%% \sc{car-clo-bang}
-%% \\
-%% \hline
-%% \makecell[t]{$GC(\rho_1, S, \heap_1, \Lv, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x))
-%% \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
-%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
-%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
-%% \sc{car-clo}\\
-%% \hline
-%% \end{tabular}
-
-%% \medskip
-
-%% \begin{tabular}{|c|c|c|}
-%% \hline
-%% Premise & Transition & Rule name \\ 
-%% \hline
-
-%% \hline
-
-%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
-%%   \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(x) :=
-%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
-%% \\
-%% \hline
-%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
-%%   \rho', S, \heap[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
-%% \hline
-%% $\heap(\rho(x)) \mbox{ is } (\ell_1, \ell_2)$ & \makecell{$\rho, (\rho', z, e,
-%%   \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
-%%   \rightsquigarrow \rho', S, \heap[\rho'(z) := \heap(\ell_1)], e, \cred{\sigma'}$}      &
-%% \sc{car-whnf} \\
-%% \hline
-%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
-%% \sc{car-clo}
-%% \\
-
-%% \hline
-%% $\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
-%% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
-%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
-%% \sc{car-1-clo} \\
-
-%% \hline
-%% $\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
-%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap,
-%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, \heap[\rho'(z)
-%%     := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
-%% \sc{prim-whnf} \\
-%% \hline
-%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-%%   \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-%% \sc{prim-1-clo} \\
-%% \hline
-%% $\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-%%   S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
-%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-%% \sc{prim-2-clo} \\
-%% \hline
-%% {$\mathit{g}~\mbox{defined as}$
-%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, \heap,
-%%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
-%%     \rho(\myvec{x})], S, \heap, e_{\mathit{g}}, \cred{\sigma}$}      &
-%% \sc{funcall} \\
-%% \hline
-%% $\ell$ is a new location& \makecell{$\rho, S, \heap, (\LET~x\leftarrow
-%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
-%%   \ell]\rangle], e, \cred{\sigma}$} &
-%% \sc{let} \\ 
-%% \hline
-%% $\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
-%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
-%% \hline
-%% $\heap(\rho(x)) = 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
-%%   \cred{\sigma}$   $\rightsquigarrow
-%% \rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
-%% \hline
-%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
-%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
-%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
-%% \cred{\clazy\sigma}$}
-%% &
-%% \sc{if-clo} \\
-%% \hline
-%% {$\heap(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
-%%   (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(z) := v], e,
-%%   \cred{\sigma'}$} &
-%% \sc{return-whnf}\\
-%% \hline
-%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
-%%   (\psi:\SRETURN~x), \cred{\sigma}$  $
-%%   \rightsquigarrow$
-%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
-%%   \cred{\sigma}$} &
-%% \sc{return-clo} \\
-%% \hline
-%% \end{tabular}
-%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
-%% \end{center}
-%% \vskip -5mm
-%% \end{figure*}
-
 We shall now present a proof of soundness of the analysis presented in
-Section~\ref{sec:liveness-analysis}.  %% \warning{While  the proof  is on
-  %% the broad lines of  \cite{asati14lgc}, there are several differences
-  %% due to the  lazy semantics of the language in  this paper.}
-Here are
-the ideas behind the proof.
+Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
+proof.
 \begin{enumerate}
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,}
+  $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
@@
 \caption{Minefield semantics for \sc{car-clo-bang} and
-  \sc{car-clo-bang} \label{fig:minefield-semantics-for-some}}
+  \sc{car-clo} \label{fig:minefield-semantics-for-some}}
 \end{figure*}
@@
 
-\item   We   extend    the   abstract   machine   state
-  $\rho,S,\heap,e$   to  $\rho,S,\heap,e,\sigma$. 
-We  call such a state  a \emph{minefield state}.
-Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
-  $e$, i.e., it arises
-  from the  actual sequence of function  calls that led
-  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
-  %% during    liveness     analysis    is     a    static
-  %% over-approximation   of   $\sigma$   that   considers
-  %% \emph{all  possible calls}  that  could  lead to  the
-  %% evaluation of  $e$.
-  The  information in  a suspended
-  evaluation  context  on  the   stack  $S$  is  also
-  similarly augmented with its  demand.  Thus a stacked
-  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
-  %% modification  of  the   small  step  semantics  which
-  %% carries the extra information  is shown in the bottom
-  %% table of Figure~\ref{fig:minefield-semantics}.
+\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
+  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
+  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
+  $e$, i.e., it arises from the actual sequence of function calls that
+  led  to the  evaluation  of $e$.
+The demand for the initial rule is  $\sigma_{all}$ and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it the expression after the step.
+The  information in  continuation
+  frames  on the  stack $S$  are also  similarly augmented  with their
+  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
+  \sigma)$.
 \item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
@@
   $\rho, S$ and  $\heap$ are due
-  to certain references being replaced by $\bot$.  This represents the
+  to certain references being replaced by $\bot$.  This simulates the
   act of  not retaining  (effectively garbage collecting)  the objects
@@
   \begin{enumerate} 
-  \item \label{gc-1}Using the current  evaluation context $\rho, \stk,
-    \heap,   e,  \sigma$   and   the  demand   transformers
-    $\Lfonly$, we construct a liveness environment \Lv\ for the
-    expression $e$.
-      %%  If  $e$  is  an expression,  then  $  \Lv  =
-    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
-    %% application,  then  $  \Lv  =  ref(e,  \sigma,  \Lfonly)$.  
- We  similarly  define  the  liveness  environment  for  the
- expressions $e$ in each of the suspended evaluation contexts $(\rho, \ell,  e, \sigma)$ in $S$, giving
- a stack of liveness environments that we shall denote $\mathsf{SL}$.
+  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
+    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
+    liveness environment  \Lv\ for  the expression $e$.   We similarly
+    define the liveness environment for the expressions $e$ in each of
+    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
+    $S$, giving a stack of  liveness environments that we shall denote
+    $\mathsf{SL}$.
   \item \emph{Garbage collection of objects pointed by root sets:} For
-    the environment $\rho$ in the current  context and for each $x \in
+    the environment $\rho$ in the current state and for each $x \in
     domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
@@
   \item \emph{Garbage  collection of  objects in  the heap:}  For each
-    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff there  is no
-    variable  $x$  in either  the  current  environment or  a  stacked
-    environment,  say  $\rho$,  such   that  $x.\alpha$  is  live  and
-    corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
-    domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
-    \oxrb = \ell$  and $x.\alpha \in \Lv$, where \Lv\  is the liveness
-    environment corresponding to $\rho$ mentioned in step~\ref{gc-1}.
+    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
+    environment $\rho$ in either  the  current state or in a
+    continuation frame, for each $x \in domain(\rho)$ and each forward
+    access
+    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\oxlb\rho(x), \alpha
+    \oxrb = \ell$. Here 
+%% $x.\alpha$  is  live  and
+%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
+%%     domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
+%%     \oxrb = \ell$  and $x.\alpha \in \Lv$, where 
+\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
+step~\ref{gc-1}.
   \end{enumerate}
@@
 enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
-the transition is the same as the earlier {\sc car-clo}
+the transition is the similar to the earlier {\sc car-clo}
 rule.
@@
 can be found in~\cite{asati14lgc}.
-%% The details of inlining a single
-%% function      call     $(\LET\,\,      x     \leftarrow
-%% (f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
-%% Let  $f$ be  defined  (after renaming  its formals  and
-%% locals  to  be  disjoint from  existing  variables)  by
-%% $(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
-%% replaced  by   a  sequence  of  $\LET$s   of  the  form
-%% $z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
-%% but with its  $(\SRETURN\,\,w)$ expressions replaced by
-%% $(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
-%% We prefer to use the  identity function $\ID$ as a form
-%% of no-op rather than  introducing the form $(\LET\,\, x
-%% \leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
-%% $\LET$-definition   is  a   simple  variable. 
 For  a
@@
 \bang\ because it dereferences a $\bot$ inserted by the
-immediately  preceding  GC  or  the GC  of  an  earlier
-minefield step.  However the  demand $\sigma'$
-  on basis of which an earlier GC would have inserted a
+immediately  preceding  garbage collection or  the garbage collection of  an  earlier
+minefield step.  \warning{However the  demand $\sigma'$
+  on basis of which an earlier collection would have inserted a
   $\bot$,  would  have   included  the  current  demand
-  $\sigma$.  Thus,  if the  earlier GC had  inserted a
-$\bot$, so would the current  GC.  Thus it is enough to
+  $\sigma$.  Thus,  if the  earlier garbage collection had  inserted a
+$\bot$, so would the current  garbage collection.  Thus it is enough to
 consider the $\rightsquigarrow$  along with the current
-GC and show that it does not lead to a \bang.
+GC and show that it does not lead to a \bang.}
 
@@
 of    $x$   will    include    both   $\epsilon$    and
-$\acar\epsilon$,  and  a  GC with  this  liveness  will
+$\acar\epsilon$,  and  a  garbage collection with  this  liveness  will
 neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
@@
 determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
-particular, we  consider $\Lanv{1}{x_i}$,  the liveness
-of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$ at
-the program point at  the beginning of $e_{\mathit{f}}$
-(assumed   to   be   $\pi_1$).   By   the   rule   {\sc
+particular, we  consider $\Lanv{x_i}{}$,  the liveness
+of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
   live-define},   this   should    be   the   same   as
@@
 $$
- \Lf{\length}{1}{\sigma} = \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
+ \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
   \cup \clazy\Lf{\length}{1}{\clazy\sigma}
 $$
-%% and  the only equation defining \Lfone{\length} is:
-%%   \begin{eqnarray*}
-%%    && \Lf{\length}{1}{\sigma}
-%%     =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
-%% \clazy\Lf{\length}{1}{\clazy\sigma}
-%% \end{eqnarray*}
 
@@
 For  the  running  example,   $\length$  has  calls  from  $\main$  at
-$\pi_{12}$ and a recursive call at $\pi_6$.
-So $\sigma_{\length} =
-     \deltacall{\length}{\pi_{12}}{\main}  \cup
-\deltacall{\length}{\pi_6}{\length}$.
-Filling in  the values gives:
+$\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
+are $\epsilon$ and $\clazy\sigma_{\mathit{\length}}$. Thus:
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
+As examples, the liveness of \Lanv{\pl}{} and \Lanv{\pa}{} in terms
 of  $\sigma_{\length}$ are:
 \begin{align*}
-\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
+\Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
   \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
  \\
-\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
+\Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
 \end{align*}
@@
 \item   For  each   function  $\mathit{f}$   (including
-  $\mainpgm$  for  $e_\mainpgm$) an  equation  defining
-  liveness at each GC point of $e_{\mathit f}$.
+  $\mainpgm$) an  equation  defining
+  liveness for each garbage in  $e_{\mathit f}$.
 \end{enumerate}
@@
 productions as being  associated with several grammars,
-one for each non-terminal \var{\Lanv{i}{x}} regarded as
+one for each non-terminal \var{\Lanv{x}{}} regarded as
 a  start  symbol.   As  an  example,  the  grammar  for
-\var{\Lanv{1}{\xl}}     comprises     the     following
+\var{\Lanv{\xl}{}} and \var{\Lanv{\pa}{}}    comprise     the     following
 productions:
 \begin{eqnarray*}
-  \var{\Lanv{1}{\xl}}  &\rightarrow& 
+  \var{\Lanv{\xl}{}}  &\rightarrow& 
   \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
@@
 \epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
-\var{\Lanv{9}{\pa}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
+\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
 \end{eqnarray*}
 Other  equations  can   be  converted  similarly.   The
-language   generated   by  \var{\Lanv{i}{x}},   denoted
-$\mathscr{L}(\var{\Lanv{i}{x}})$,   is    the   desired
-solution  of  \Lanv{i}{x}.    
+language   generated   by  \var{\Lanv{x}{}},   denoted
+$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
+solution  of  \Lanv{x}{}.    
 However, recall that the decision problem that we are interested in
@@
 of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
-\clazy).       Let      $\mathscr{L}(\var{\Lanv{i}{x}})
+\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
 \stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
@@
 \begin{pspicture}(0,-5)(50,12) %\psframe(0,-5)(40,12)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{9}{\pa}}}
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
   \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
@@
 \begin{pspicture}(0,-5)(20,12) %\psframe(0,-5)(40,12)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{9}{\pa}}}
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
   \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
@@
 \caption{(a) The  grammar rules for \var{\Df{\length}{1}}
-  converted into an automaton, and (b) its DFA. The same for \var{\Lanv{9}{\pa}}
+  converted into an automaton, and (b) its DFA. The same for \var{\Lanv{\pa}{}}
   are in (c) and (d).}\label{fig:example-automata}
@@
 The strongly regular grammar is converted  into a set of NFAs, one for
-each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
+each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
 done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
 pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
-constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
-may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
-simplification is continued till a fixed point is reached, after which
-the edges labeled \bcar \ and  \bcdr\ are deleted.  The details of the
-algorithm, its correctness and termination proofs are given by Karkare
-et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
-has  edges labeled  with  \acar,  \acdr\ and  \clazy\  only. In  this
-automaton, for every edge labeled \clazy, we check if the source node
-of the edge has a valid path to  a final final state.  If yes, we mark
-the mark  the source node as  final. Finally, we remove  all the edges
-labeled  \clazy\  and  convert  the automaton  into  a  deterministic
-automaton.    This   effectively  implements   the   $\hookrightarrow$
-simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
-paths. While  checking for liveness  during GC, a  forward access-path
-is valid only if it can reach a final state. 
-Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
-from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
-Figure~\ref{fig:example-automata}(b).  This expectedly says that given
-a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
-$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
-Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
-\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
+constructing  the $\epsilon$-closure  till a  fixed point  is reached,
+after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
+details of the  algorithm, its correctness and  termination proofs are
+given  in~\cite{karkare07liveness,asati14lgc}.   The
+resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
+\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
+check if the source node of the edge has a path to a final 
+state.  If yes, \warning{we mark the mark the source node as final. Finally, we
+remove all the edges labeled \clazy\  and convert the automaton into a
+deterministic    automaton.}    These steps   effectively   implement    the
+$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
+\clazy\ to  obtain forward paths.  While checking for  liveness during
+garbage  collection, a  forward access-path  is valid  only if  it can
+reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
+NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
+final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
+expectedly  says that for a  demand  \warning{$\epsilon$  on  \length},  the
+liveness of  its argument  is $\acdr^{*}$  (the spine  of the  list is
+traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
+NFA     for     \var{\Lanv{\pa}{}}.      The     DFA     shown     in
 Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
@@
 the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
-liveness based GC is triggered beyond $\pi_9$.
+liveness based garbage collection is triggered beyond $\pi_9$.
 
@@
 static approximation  without any  knowledge regarding which  of these
-paths will be  eventually taken during execution. During  GC, we would
+paths will be  eventually taken during execution. During  garbage collection, we would
 like to use a more precise  liveness based on the  actual
@@
 \SIF\ condition, the liveness environments are updated to one based on
-either $\epath_1$ or  $\epath_2$, so that subsequent GCs  are based on
+either $\epath_1$ or  $\epath_2$, so that subsequent garbage collections  are based on
 more precise liveness information.
@@
   available memory is  less than this requirement for  the function, a
-  GC is triggered.
+  garbage collection is triggered.
 \item Since the evaluation of a  \SIF\ condition may trigger a 
@@
   (based on  value of the condition)  required to execute the  rest of
-  the program.  A GC is triggered if enough memory is unavailable.
+  the program.  A garbage collection is triggered if enough memory is unavailable.
 \end{enumerate}
@@
 Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
-statistics  for RGC  and  LGC.  We  report the  number  of GC  events,
-average  number of  cells reclaimed  per GC,  average number  of cells
-touched  per GC,  total  number of  dragged cells  and  total time  to
+statistics  for RGC  and  LGC.  We  report the  number  of garbage collection  events,
+average  number of  cells reclaimed  per garbage collection,  average number  of cells
+touched  per garbage collection,  total  number of  dragged cells  and  total time  to
 perform  all collections.   It is  no  surprise that  number of  cells
-reclaimed per  GC is  higher (and  consequently the  number of  GCs is
+reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
 lower)  for LGC  based collection.   LGC  also reduces  the number  of
-dragged cells significantly. However, the cost  for LGC is that per GC
+dragged cells significantly. However, the cost  for LGC is that per garbage collection
 execution  time is  more  for  LGC, which  results  in higher  overall
@@
 \renewcommand{\arraystretch}{.1}
-\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
+\begin{tabular}{@{}c@{}@{}c@{}}
+%%    \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
+%%    & {\epsfig{file=lambda_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+   \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=paraffins_win.eps, height=\hgt}}
+\\   \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
+\\   \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
+\\  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
 %\\ (a) & (b)
@@
 point.   These space  and  run time  overheads make  it
-infeasible  to   incorporate  liveness  based   GC  for
-closures in a practical garbage collector.
+infeasible  to   incorporate  liveness  based   garbage collection  for
+closures in a practical garbage collectors.
 
@@
 \label{sec:relatedwork}
-Although  augmenting garbage  collection with  liveness
-information has been studied  earlier, it was mainly in
-the   context  of   imperative  languages~\cite{Albert,
-  Hirzel,  khedker07heap}.   Due  to  the  presence  of
-global  variables and  mutation,  practical utility  of
-simple  liveness  based  techniques  are  found  to  be
-inefficient.
-
-In the  space of functional languages,  improving space
-efficiency has  been mainly  studied as a  compile time
-activity.   Either through  rewriting  methods such  as
+Most previous work on enhancing  garbage collection with liveness were
+in   the  context   of   imperative  languages~\cite{Albert,   Hirzel,
+  khedker07heap}.    Due  to  the  presence  of  global  variables  and
+mutation, these methods are of  limited practical  utility.
+
+There  have been  several efforts  in improving  space utilization  of
+functional programs  by compile-time  analysis.  These  are: rewriting
+methods                             such                            as
 deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-sharing                  analysis                 based
-reallocation~\cite{jones89compile},     region    based
-analysis~\cite{tofte98region}, or  through insertion of
-compile     time     nullifying     statements     such
-as~\cite{inoue88analysis,lee05static,Hamilton}.     The
-compile  time   marking  approaches  all  rely   on  an
-efficient and precise alias analysis and in the absence
-of  it  cannot  provide  notable  improvement.  Another
-important         approach         advocated         by
-Hofmann~\cite{HofmannJ03}  is to  use linear  typing to
-analyze  first-order   programs  for  heap   usage  and
-annotated them. The system  then uses these annotations
-to re-use allocated memory  cells instead of requesting
-for  newer  cells.  This  requires the  user  to  write
-programs  in  a  specific  way and  hence  may  not  be
-practical
-
-Simplifiers~\cite{ONeill},  described   as  lightweight
-daemons,  improve  the  efficiency of  the  program  in
-general and  the garbage collector in  particular. Most
-of  the  simplifications  mentioned in  this  work  are
-subsumed by  a liveness  based collector. We  feel this
-work is  orthogonal and can augment  our liveness based
-collector.  The closest to our approach is the liveness
-based        garbage       collector        implemented
-in~\cite{asati14lgc}.  We extend  their work  to handle
-lazy evaluation and closures.
+sharing  analysis   based  reallocation~\cite{jones89compile},  region
+based   analysis~\cite{tofte98region},   or   through   insertion   of
+compile-time                                                nullifying
+statements~\cite{inoue88analysis,lee05static,Hamilton}.            All
+compile-time marking approaches rely on an efficient and precise alias
+analysis and  cannot provide  significant improvement in  its absence.
+An interesting  approach suggested by Hofmann~\cite{HofmannJ03}  is to
+annotate the heap usage of  first-order programs through linear types.
+The annotations are then used to transform the program to serve memory
+allocation requests by re-using allocated memory cells.  However, this
+requires the user to write programs in a specific way.
+
+Simplifiers~\cite{ONeill}, described  as lightweight  daemons, improve
+the efficiency of the program in general and the garbage collectors in
+particular.  \warning{Most  of the  simplifications mentioned  in this
+  work are subsumed by a liveness  based collector.  We feel this work
+  is orthogonal  and can augment  our liveness based  collector.}  The
+approach that is closest to the  method described in this paper is the
+liveness-based garbage collector implemented in~\cite{asati14lgc}.  We
+have extend this to handle lazy evaluation and closures.
 
@@
 %% reclamation and reallocation without  the aid of the
-%% garbage collector.   Such attempts  include: sharing
+%% GC.   Such attempts  include: sharing
 %% analysis  based  reallocation~\cite{jones89compile},
@@
 %% this.
-We extended  the liveness  based garbage  collection to
-lazy  languages  and  shown  its  benefit  for  garbage
-collection in practical programs.  We defined a context
-sensitive   liveness   analysis   that   uses   context
-independent  summaries of  functions  obtained using  a
-symbolic  demand.  We  showed  that  obtaining  precise
-solution  to  these  equations is  undecidable  in  our
-formulation, and hence we safely approximate the result
-using  DFA.  These DFA  are  consulted  by the  garbage
+We have  extended liveness-based garbage collection  to lazy languages
+and shown its benefit for  standard benchmark programs.  This was done
+by defining a liveness analysis of heap objects that we proved correct
+with  respect   to  a   non-standard  semantics   that  served   as  a
+specification of  liveness.  The result  of the  analysis is a  set of
+grammars, whose membership question was  proved to be undecidable. The
+grammars are  therefore approximated by  DFAs and used by  the garbage
 collector to improve collection.
 
-Liveness  of closures  presented further  challenges to
-our implementation.  We  compared different approximate
-strategies for handling liveness of closures, and found
-that a  mixed strategy---reachability  based collection
-of  closure  arguments,  liveness based  collection  of
-everything else---works  best in practice as  it avoids
-runtime and  space overheads.  With  mixed-mode garbage
-collection scheme,  we were able to  collect more cells
-than  reachability  based  collectors at  a  reasonable
-overhead.
-
-Although we  provide an  implementation for  a liveness
+In addition to collecting evaluated  data, our collector also reclaims
+closures.   Because  of this  we  had  to  make modifications  to  the
+standard run-time representation of a closure to carry liveness of its
+free variables and  periodically and update it  during execution time.
+As  expected, our  garbage collector  could reclaim  more garbage  and
+reduce memory requirements of programs.  An additional benefit was the
+reduction in  execution times for  some programs  in spite of  using a
+more expensive collection scheme.
+
+\warning{Although we  provide an  implementation for  a liveness
 based garbage  collection scheme for lazy  language, we
@@
   to make liveness  based garbage collection attractive
-  for practical collectors.
+  for practical collectors.}
 
@@
   \emph{all  possible calls}  that  could  lead to  the
-  evaluation of  $e$.  The  information in  a suspended
-  evaluation  context  (on  the   stack  $S$)  is  also
-  similarly augmented with its  demand.  Thus a stacked
+  evaluation of  $e$.  The  information in  the continuation frames on
+  the   stack  $S$ are  also
+  similarly augmented with their  demands.  Thus a stacked
   entry now takes  the form $(\rho, x,  e, \sigma)$.  A
@@
   \begin{enumerate}
-  \item Given the current  evaluation context $\rho, S,
+  \item Given the current state $\rho, S,
     \heap,   e,  \sigma$   and   the  demand   transformers
     $\Lfonly$, we can  construct a liveness environment
-    \Lv.   If  $e$  is  an expression,  then  $  \Lv  =
+    \Lv for $e$.   If  $e$  is  an expression,  then  $  \Lv  =
     \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an

### 2015-11-29 paper/paper.tex
latest - 29th

@@
 \subsection{Soundness Results}
-
-%% We consider a round of  garbage collection followed by a transition
-%% as  a  step.  We  show  by  induction  that, for  programs  without
-%% function  calls, %%  starting from  the %%  initial state  that has
-%% $e_{\main}$ as the %% expression to be evaluated, any transition of
-%% $n$ steps occurs without a \bang.
-
 We first need an auxiliary  result about minefield semantics. Consider
@@
 step  is non-null,  the  demand  on the  RHS  is  a transformation  of
-$\sigma$ (for example $\clazy\sigma$) which is also non-null.
-
-%% Let  us clarify  this  observation: Since  the minefield  semantics
-%% models lazy evaluation,  parts of the program may  not be evaluated
-%% at all.  However,  if an expression (or application)  happens to be
-%% evaluated, it will  appear on the LHS of a  minefield step, and the
-%% context-sensitive demand on it will be non-null.
+$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.
 
@@
 $n$ leading  to this  state that  the next  transition cannot  enter a
-\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\_, [\;]_{S} ,
-            [\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.   Since  the
-            call to $GC(\ldots)$  in this state does  nothing, we just
+\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\,    (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
+(\print~\ans)):[\;]_{S}, 
+[\;]_{\heap},\, (\mainpgm))$.   Since  the
+            call to $GC$  in this state does  nothing, we just
             have to show that the $\rightsquigarrow$ transition cannot
@@
 that involves dereferencing  results in a \bang.  These  are the steps
-which  have a  $\heap(\rho(...))$ in  the  premise. 
-Now a  $\rightsquigarrow$ step can  go
-\bang\ because it dereferences a $\bot$ inserted by the
-immediately  preceding  garbage collection or  the garbage collection of  an  earlier
-minefield step.  \warning{However the  demand $\sigma'$
-  on basis of which an earlier collection would have inserted a
-  $\bot$,  would  have   included  the  current  demand
-  $\sigma$.  Thus,  if the  earlier garbage collection had  inserted a
-$\bot$, so would the current  garbage collection.  Thus it is enough to
-consider the $\rightsquigarrow$  along with the current
-GC and show that it does not lead to a \bang.}
+which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
+$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
+inserted  by  the  immediately  preceding garbage  collection  or  the
+garbage collection of an earlier minefield step.  \warning{However the
+  demand $\sigma'$ on basis of  which an earlier collection would have
+  inserted a $\bot$, would have  included the current demand $\sigma$.
+  Thus, if  the earlier garbage  collection had inserted a  $\bot$, so
+  would the current garbage collection.  Thus it is enough to consider
+  the $\rightsquigarrow$  along with the  current GC and show  that it
+  does not lead to a \bang.}
 
@@
 first   component  of   the   cons   cell.  Thus   both
-dereferences    will    happen   normally    and    the
-$\rightsquigarrow$ step will not enter \bang.
+dereferences    can take place without 
+$\rightsquigarrow$ entering the   \bang\ state.
 \end{proof}

### 2015-11-29 lazy-scheme/post_process/liveness_stats.sh
liveness stat gen

@@ -1,0 +1,41 @@
+#!/bin/bash
+DIR=`dirname $0`
+AllBMs="fibheap sudoku nperm paraffins lcss huffman knightstour nqueens deriv treejoin lambda gc_bench "
+OUTDIR=output.DFA
+
+declare -A bmdata
+
+bmdata["HEADER"]="\\multicolumn{1}{|r||}{Program};\#CFG Nonterminals;\#CFG Rules;\#DFA States;\#DFA Transitions;DFA Gen Time (sec)"
+
+echo -n "\begin{tabular}{|l|"
+for bm in $AllBMs
+do 
+  echo -n "|c@{\\ }"
+done
+echo "|}"
+
+for bm in $AllBMs
+do 
+   cfgnt=`wc -l $OUTDIR/$bm/program-cfg.txt | cut -d' ' -f1`
+   cfgrl=`fgrep -o '|' $OUTDIR/$bm/program-cfg.txt | wc -l`
+   cfgrl=`expr $cfgrl + $cfgnt`
+   dfast=`wc -l $OUTDIR/$bm/fsmdump-$bm-state-map | cut -d' ' -f1`
+   dfatr=`wc -l $OUTDIR/$bm/fsmdump-$bm-state-transition-table | cut -d' ' -f1`
+   gentm=`grep "Program.Exec" ../benchmarks/programs/$bm/results/$bm-gc-live | cut -d'=' -f2 | cut -d' ' -f1`
+   gentm=`printf "%.2f\n" $gentm`
+   bmdata[$bm]="\\rot{\\tt $bm};$cfgnt;$cfgrl;$dfast;$dfatr;$gentm" 
+done
+
+for i in 1 2 3 4 5 6 
+do
+  for bm in "HEADER" $AllBMs
+  do
+    if [ $bm = "HEADER" ]; then sep=""; else sep="&"; fi
+    col=`echo ${bmdata[$bm]} | cut -d';' -f$i`
+    echo -n "$sep $col "
+  done
+  echo "\\\\"
+done
+echo "\end{tabular}"
+echo " (a) Data for Liveness Analysis "
+

### 2015-11-29 paper/paper.tex

@@
 \end{center}
-\caption{Minefield semantics for \sc{car-clo-bang} and
-  \sc{car-clo} \label{fig:minefield-semantics-for-some}}
+\caption{Minefield semantics for {\sc car-clo-bang} and
+  {\sc car-clo} \label{fig:minefield-semantics-for-some}}
 \end{figure*}
@@
 the number  of steps leading  to the  minefield state.  The  base step
-holds   because  the   demand   $\sigma_{all}$   on  $(\mainpgm)$   is
+holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
 non-null. For the inductive step we  observe that for each step of the
@@
 \begin{lemma}\label{lemma:grammar-undecidable}
-Consider a grammar    $G$    of   the    kind    described    in
+Given a grammar    $G$    of   the    kind    described    in
 Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
-consisting  of symbols \acar\  and \acdr\  only. The membership problem
+consisting  of symbols \acar\  and \acdr\  only, the membership problem
 $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
@@
 \begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
-Consider    a    grammar    $G$    of   the    kind    described    in
+Given   a    grammar    $G$    of   the    kind    described    in
 Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
-analysis  of a  program,  and  a forward  access  path $\alpha$.   The
+analysis  of a  program  and  a forward  access  path $\alpha$,    the
 membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
@@
 Given a Turing machine and an input string, we shall define a function
-for   each  of   the  non-terminals   introduced  in   the  proof   of
+for   each  non-terminal   introduced  in   the  proof   of
 Lemma~\ref{lemma:grammar-undecidable}.
@@
 %% that necessarily contain the \clazy\ symbol.
-For instance, corresponding to the production $\nont{i}{c} \rightarrow
-c \nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, consider
+For instance, corresponding to the production $\nont{i}{\acar} \rightarrow
+\acar \nont{j}{\acar}\bcar$,  consider
 the program:
@@
 branch. It can  be verified that the grammar for the function shown
-above, includes the productions:
+above, includes the productions---$\nont{\mathit{all}}{}$
+corresponds to the demand $\sigma_{\mathit{all}}$:
 \begin{align*}
-\nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{all}{} \\
-\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{all}{} \\
-\nont{all}{} &\rightarrow \epsilon \mid \acar\nont{all}{} \mid
-\acdr\nont{all}{}
+\nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
+\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{\mathit{all}}{} \\
+\nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
+\acdr\nont{\mathit{all}}{}
 \end{align*}
@@
 $\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
-\nont{\mathit{all}}{}$,  and (b)  other  productions are  of the  form
+\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
 $\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
@@
 accepts the input string, then the earlier derivation can be replayed,
-replacing   $\nont{\mathit{all}}{}$   in   the  sentential   form   by
+every time replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the  sentential   form   by
 $\epsilon$. Conversely,  if the TM  does not accept the  input string,
@@
 In        the       liveness        analysis       described        in
-Section~\ref{sec:liveness-analysis}, the liveness of closure variables
-in  an application  $s$ is  derived from  dependence chains  along all
-paths from the program point where the closure is created to reachable
-evaluation  points.  Assume,  for   the  sake  of  concreteness,  that
-$\epath_1$  and $\epath_2$  two  such evaluation  points.   This is  a
-static approximation  without any  knowledge regarding which  of these
-paths will be  eventually taken during execution. During  garbage collection, we would
-like to use a more precise  liveness based on the  actual
-paths  taken  during  execution.    Therefore we create  liveness  automata
-separately  for the dependences in   $\epath_1$ and  $\epath_2$  as  well as  a  automaton
-corresponding to  both paths  $\epath_1$ and $\epath_2$.   The closure
-corresponding to  $s$ carries  the liveness  environment for  its free
-variables (as pointers to automata, one for each variable).  Initially
-the  liveness  environment is  based  on  the dependences  along  both
-$\epath_1$    and   $\epath_2$.    However,   after    evaluating   an
+Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
+variable is  derived from dependence  chains along all paths  from the
+program point  where the closure  was created to  reachable evaluation
+points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
+$\epath_2$ are two such evaluation points.  During garbage collection,
+we would like to use a more precise liveness, based on the actual paths
+taken  during  execution.   Therefore   we  create  separate  liveness
+automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
+addition to  automata for dependences  along paths to  both $\epath_1$
+and $\epath_2$.  The closure carries  the liveness environment for its
+free  variables (as  pointers  to automata,  one  for each  variable).
+Initially the liveness  environment is based on  the dependences along
+both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
 \SIF\ condition, the liveness environments are updated to one based on
-either $\epath_1$ or  $\epath_2$, so that subsequent garbage collections  are based on
-more precise liveness information.
+either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
+collections are based on more precise liveness information.
 
@@
 \renewcommand{\arraystretch}{.1}
-\begin{tabular}{@{}c@{}@{}c@{}}
-%%    \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
-%%    & {\epsfig{file=lambda_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
-   \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
+\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
+   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
-\\   \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-\\   \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-\\  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
 %\\ (a) & (b)

### 2015-11-29 lazy-scheme/post_process/liveness_stats.sh
minor fixes in script

@@ -12,10 +12,12 @@ for bm in $AllBMs
 do 
   echo -n "|c@{\\ }"
 done
-echo "|}"
+echo "|} \\\\ \hline"
 
 for bm in $AllBMs
 do 
+   bmname=`echo $bm | sed -e 's/_/\\\\_/'`
+   echo $bmname
    cfgnt=`wc -l $OUTDIR/$bm/program-cfg.txt | cut -d' ' -f1`
    cfgrl=`fgrep -o '|' $OUTDIR/$bm/program-cfg.txt | wc -l`
    cfgrl=`expr $cfgrl + $cfgnt`
@@ -23,7 +25,7 @@ do
    dfatr=`wc -l $OUTDIR/$bm/fsmdump-$bm-state-transition-table | cut -d' ' -f1`
    gentm=`grep "Program.Exec" ../benchmarks/programs/$bm/results/$bm-gc-live | cut -d'=' -f2 | cut -d' ' -f1`
    gentm=`printf "%.2f\n" $gentm`
-   bmdata[$bm]="\\rot{\\tt $bm};$cfgnt;$cfgrl;$dfast;$dfatr;$gentm" 
+   bmdata[$bm]="\\rot{\\tt $bmname};$cfgnt;$cfgrl;$dfast;$dfatr;$gentm" 
 done
 
 for i in 1 2 3 4 5 6 
@@ -34,8 +36,8 @@ do
     col=`echo ${bmdata[$bm]} | cut -d';' -f$i`
     echo -n "$sep $col "
   done
-  echo "\\\\"
+  echo "\\\\ \\hline"
 done
-echo "\end{tabular}"
+echo "\end{tabular}\\\\"
 echo " (a) Data for Liveness Analysis "
 

### 2015-11-29 lazy-scheme/post_process/liveness_stats.sh
minor fixes in script

@@ -17,7 +17,6 @@ echo "|} \\\\ \hline"
 for bm in $AllBMs
 do 
    bmname=`echo $bm | sed -e 's/_/\\\\_/'`
-   echo $bmname
    cfgnt=`wc -l $OUTDIR/$bm/program-cfg.txt | cut -d' ' -f1`
    cfgrl=`fgrep -o '|' $OUTDIR/$bm/program-cfg.txt | wc -l`
    cfgrl=`expr $cfgrl + $cfgnt`

### 2015-11-29 lazy-scheme/post_process/liveness_stats.sh
minor fixes in script

@@ -12,7 +12,7 @@ for bm in $AllBMs
 do 
   echo -n "|c@{\\ }"
 done
-echo "|} \\\\ \hline"
+echo "|} \hline"
 
 for bm in $AllBMs
 do 

### 2015-11-29 lazy-scheme/post_process/liveness_stats.sh
minor fixes in script

@@ -22,7 +22,7 @@ do
    cfgrl=`expr $cfgrl + $cfgnt`
    dfast=`wc -l $OUTDIR/$bm/fsmdump-$bm-state-map | cut -d' ' -f1`
    dfatr=`wc -l $OUTDIR/$bm/fsmdump-$bm-state-transition-table | cut -d' ' -f1`
-   gentm=`grep "Program.Exec" ../benchmarks/programs/$bm/results/$bm-gc-live | cut -d'=' -f2 | cut -d' ' -f1`
+   gentm=`grep "DFA.Gen" ../benchmarks/programs/$bm/results/$bm-gc-live | cut -d'=' -f2 | cut -d' ' -f1`
    gentm=`printf "%.2f\n" $gentm`
    bmdata[$bm]="\\rot{\\tt $bmname};$cfgnt;$cfgrl;$dfast;$dfatr;$gentm" 
 done

### 2015-11-29 paper/paper.tex
Updated Results and Future work

@@
 \label{sec:experiments}
-  Our benchmark mostly consists  of programs from nofib~\cite{nofib}
-suite.  All our
+  Our benchmark consists  of programs taken from nofib~\cite{nofib}
+and other similar fucntional programming test suites.  All our
 benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
@@
 
-The statistics related to compile  time liveness analysis are shown in
+The statistics related to liveness analysis and DFA generation are shown in
 Table~\ref{tab:exp-results}(a). We observe that the analysis requires
@@
 The bottleneck in our analysis is the NFA to DFA conversion
-(worst-case exponential behaviour). To avoid this compile-time
-overhead, we have added a caching mechanism in our simulator that can
-reuse the DFA from earlier analysis if the program has not changed..
+(worst-case exponential behaviour). To avoid this 
+overhead, we use a caching mechanism in our simulator that can
+reuse the DFA from earlier analysis if the program has not changed.
 
@@
 average  number of  cells reclaimed  per garbage collection,  average number  of cells
-touched  per garbage collection,  total  number of  dragged cells  and  total time  to
+touched  per garbage collection,  and  total time  to
 perform  all collections.   It is  no  surprise that  number of  cells
 reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
-lower)  for LGC  based collection.   LGC  also reduces  the number  of
-dragged cells significantly. However, the cost  for LGC is that per garbage collection
+lower)  for LGC  based collection.   However, the cost  for LGC is that per garbage collection
 execution  time is  more  for  LGC, which  results  in higher  overall
@@
 collections.   However, LGC  run time  is still  competitive (slowdown
-within 5X of RGC for worst case), and is better for 3 benchmarks (best
+within \warning{5X} of RGC for worst case), and is better for 3 benchmarks (best
 case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
@@
 \newcommand{\hgt}{2.9cm}
+\newcommand{\LGCLine}{red}
+\newcommand{\RGCLine}{blue}
+\newcommand{\ReachLine}{black}
+\newcommand{\UseLine}{lightblue}
+
 \begin{figure*}[t]
@@
 %%    (b) enlarges  a part  of the  usage.  
-The blue and the red curves indicate the number of cons
+The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
 cells  in  the  active   semi-space  for  RGC  and  LGC
-respectively.  The black curve represents the number of
-reachable cells and the  lightblue curve represents the
+respectively.  The \ReachLine\ curve represents the number of
+reachable cells and the  \UseLine\ curve represents the
 number  of  cells  that  are actually  live  (of  which
@@
 Memory  usage graphs  for select  benchmarks are  shown
-Figure~\ref{fig:memory-usage},  Column   (a).   As  the
+Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
 number of  garbage collections  tend to be  very large,
-for  each benchmark  we  also show  a  window which  is
+hence  we  show only  a  slice of the original graph  which  is
 representative  of  the  behavior for  that  particular
 benchmark (Column (b)).  In all the programs we can see
-that  the   curve  corresponding  to  LGC   (red  line)
-regularly dips below the RGC curve (blue line), the dip
+that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
+regularly dips below the RGC curve (\RGCLine\ line), the dip
 is  specially  visible  for the  benchmarks  where  LGC
-outperforms RGC substantially.  The graphs also include
-curve    for   reachable    cells   (black,    obtained
+outperforms RGC substantially.  The graphs also include the
+curve    for   reachable    cells   (\ReachLine,    obtained
 approximately  by forcing  RGC to  run at  a very  high
-frequency)  and  the  live  cells  for  particular  run
-(lightblue, obtained by post processing the heap access
-at the end of a program).
-
-One area of concern is  the huge gap between the actual
-liveness and  the liveness perceived by  our collector.
-In  case of  LGC for  eager languages~\cite{asati14lgc}
-the gap was  very narrow and almost  touched the actual
-liveness curve.  In case of lazy languages, this gap is
-largely   due   to   approximate   reachability   based
-collection  of closures.  To  implement liveness  based
-collection of closures, we need to record liveness data
-for  each  closure and  update  it  at each  evaluation
-point.   These space  and  run time  overheads make  it
-infeasible  to   incorporate  liveness  based   garbage collection  for
-closures in a practical garbage collectors.
+frequency)  and  the  live  cells  for a particular run
+(\UseLine, obtained by post processing the heap access
+at the end of the program).
+
+As demonstrated by the gap between the \RGCLine\ and the \UseLine, a large number of cells 
+which are unused by the program are still copied during RGC. LGC does a much better job of closing this gap 
+but still falls short of the precision achieved by LGC in case of eager languages
+~\cite{asati14lgc}. 
+A case for concern is the huge difference in the number of cells touched by LGC vs RGC for some programs.
+This is due to heap locations being  traversed multiple times during LGC. \warning{We have implemented heuristics to 
+overcome common cases and they show promising results}.
 
@@
 We have  extended liveness-based garbage collection  to lazy languages
-and shown its benefit for  standard benchmark programs.  This was done
+and shown its benefit for a set of benchmark programs.  This was done
 by defining a liveness analysis of heap objects that we proved correct
@@
 \warning{Although we  provide an  implementation for  a liveness
-based garbage  collection scheme for lazy  language, we
+based garbage  collection scheme for a lazy  language, we
 do not  provide a formal  proof of its  correctness.  A
@@
   between  the   actual  liveness  and   the  perceived
-  liveness  of our  garbage collector.  Our experiments
-  (not reported here) show that a significant number of
-  dead  cells  get   trapped  inside  closures.   Using
+  liveness  of our  garbage collector. Using
   strictness  analysis  to  eagerly  evaluate  closures
-  might release more of these  dead cells to be garbage
+  might release more dead cells to be garbage
   collected.  Since our  garbage collector  can handle

### 2015-11-29 lazy-scheme/src/gc.cpp

@@ -2086,6 +2086,7 @@ cons* followpaths(cons* loc, state_index index, ostream& out)
 			 loccopy->val.closure.arg2 = new_arg2;
 
 		 }
+
 	 }
 
 	  if (loccopy->val.closure.arg1 != NULL)
@@ -2791,7 +2792,7 @@ void create_heap_bft(ostream& out)
 	add_elements_to_vector(print_vector, print_stack);
 	for (auto ele : print_vector)
 	{
-		if (heap_map.find(ele) == heap_map.end())
+		if (ele && heap_map.find(ele) == heap_map.end())
 		{
 			assert(is_valid_address(ele));
 			heap_map[ele] = index;

### 2015-11-29 paper/paper.tex
fixed spelling mistakes
fix graphs

@@
 \newcommand{\hgt}{2.9cm}
-\newcommand{\LGCLine}{red}
-\newcommand{\RGCLine}{blue}
+\newcommand{\LGCLine}{blue}
+\newcommand{\RGCLine}{red}
 \newcommand{\ReachLine}{black}
@@
 \begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+   \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
 \\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
 \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
 %\\ (a) & (b)

### 2015-11-29 paper/paper.tex

@@
   Our benchmark consists  of programs taken from nofib~\cite{nofib}
-and other similar fucntional programming test suites.  All our
+and other similar functional programming test suites.  All our
 benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
@@
 \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
 %\\ (a) & (b)

### 2015-11-29 paper/paper.tex
added missing file

--- 2015-11-29 paper/paper.tex
@@
 demands  on its  arguments.   This  summary is  used  to step  through
-function  calls  during analysis.   Next,  the  concrete demand  on  a
+function  calls  during analysis.   The  concrete demand  on  a
 function body is obtained through a conservative approximation similar
@@
 capturing the liveness of variables at these points.  Depending on the
-program  point where  garbage collection  has been  triggered, the
-garbage collector
+where  garbage collection  has been  triggered, the
+collector
 consults a  set of automata  to restrict reachability  during marking.
@@
   references} and represent values of  variables  or fields.  The figure shows
-the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
-triggered by the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.
+the  lists \px\  and  \pz\ partially  evaluated due to  the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.
 
@@
 \begin{itemize}
-\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
-  functional language and  prove its correctness. The correctness
-  proof involves setting up a non-standard semantics as a
-  specification of liveness and then proving that the analysis is
+\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
+  functional language and  prove its correctness. The 
+  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
   correct with respect to the specification.   
-\item Our  analysis results  in a set  of context-free  grammars along
+\item The analysis results  in a set  of context-free  grammars along
   with  a fixed  set  of non-context-free  productions.  The  decision
-  regarding  whether   to  copy  a  cell   during  garbage  collection
-  translates  to the  membership problem  of such  grammars.  In  this
-  paper, we  have shown the  undecidability of the  membership problem
-  for  the  class  of  grammars  that result  from  our  analysis.  To
-  circumvent the  problem, we  over-approximate the  grammars generated
-  by the analysis with DFAs.
+  whether   to  copy  a  cell   during  garbage  collection
+  translates  to a  membership problem  for such  grammars.  In  this
+  paper, we  have shown the  undecidability of this problem and 
+  circumvent the  problem by over-approximating the  grammars
+  with DFAs.
 \item We  have implemented a garbage  collector that uses the  DFAs to
@@
   results  in   more  garbage  being  collected.    Empirical  results
-  demonstrate the effectiveness of liveness-based garbage collection.
+  show the effectiveness of liveness-based garbage collection.
 \end{itemize}
-To the  best of our
-  knowledge, this is the first work on liveness-based garbage
+As far as we know,  this is the first work on liveness-based garbage
   collection      for      lazy     languages.       All      previous
@@
 
-The body  of a function ${\mathit  f}$ is denoted  as $e_{\mathit f}$.
+We denote the body of a function ${\mathit  f}$   as $e_{\mathit f}$.
  We assume that each program has a distinguished function
-\mainpgm\      with      the      definition     $(\DEFINE\      ({\tt
-  \mainpgm})\  e_\mainpgm)$ and  the execution  of the  program starts
+\mainpgm,    defined as    $(\DEFINE\      ({\tt
+  \mainpgm})\  e_\mainpgm)$,  and  the program begins execution
 with the call to \mainpgm.  We write $\pi\!:\!e$ to associate the label
-$\pi$ (not  part of the language  syntax) with the  program point just
+$\pi$ (not  part of the syntax) with the  program point just
 before expression $e$.
@@
 environment  that  maps  variables  $x_i$ to  locations  $\ell_i$  and
-$\heap[\ell := d]$  indicates the updating of a heap  \heap\ at $\ell$
+$\heap[\ell := d]$  indicates updation of  \heap\ at $\ell$
 with  $d$.   $\rho \oplus  \rho'$  represents  the environment  $\rho$
@@
 variable  reference) and  a heap  \heap, semantically  an access  path
-$\alpha$ represents  a reference, denoted $\heap\oxlb\ell,  \alpha\oxrb$, in
+$\alpha$ represents  a reference, denoted $\heap\llbracket\ell,  \alpha\rrbracket$, in
 the heap  that is  obtained by  starting with  $\ell$ and  chasing the
-\CAR\ or  \CDR\ fields  as specified  by the  access path.  
-$\heap\oxlb\ell, \alpha\oxrb$  is defined only  if the path followed  in the
+\CAR\ or  \CDR\ fields in the heap as specified  by the  access path.  
+$\heap\llbracket\ell, \alpha\rrbracket$  is defined only  if the path followed  in the
 heap  is \emph{closure-free}  (does not  cross closures),  else it  is
@@
 
-Access paths can be used to  represent liveness. As an example, if $x$
-is  a  list  with  liveness  $\{\epsilon,  \acar,  \acdr,  \acdr\acar,
-\acdr\acdr,  \acdr\acdr\acar\}$,  then  future computations  can  only
+Access paths are used to  represent liveness. As an example, a list
+$x$ with 
+liveness  $\{\epsilon,  \acar,  \acdr,  \acdr\acar,
+\acdr\acdr,  \acdr\acdr\acar\}$ means that future computations    only
 refer up  to the  second and  third members of  $x$.  A  {\em liveness
@@
 $\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
-result.   To see  the need  for demands,  consider the  let expression
+result.   To see  the need  for demands,  consider the expression
 $(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
@@
   variable is attributed  with the same liveness in the  region of the
-  program from the  point of creation to  reachable evaluation points.
-  This is  also true  of stack  variables, because,  as we  shall see,
-  stack variables derive their  liveness from closure variables. Thus,
-  there are  two major differences  in our formulation of  liveness of
-  lazy   languages  with   liveness  of   eager 
-  languages~\cite{asati14lgc}:    (i)  the
-  introduction of  closure variables in the  liveness calculation, and
-  (ii) a  single liveness value  for each variable that  is applicable
-  from its creation point to evaluation points.
+  program  from the  point of  creation  of the  closure to  reachable
+  evaluation points.  This  is also true of  stack variables, because,
+  as we shall see, stack  variables derive their liveness from closure
+  variables. Thus, there are two  major differences in our formulation
+  of   liveness   of   lazy   languages   with   liveness   of   eager
+  languages~\cite{asati14lgc}:   (i)  the   introduction  of   closure
+  variables in  the liveness calculation,  and (ii) a  single liveness
+  value for each  variable that is applicable from  its creation point
+  to evaluation points.
 
@@
   through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
-  closure  variable  $\px_{\pi_1}$   through  $(\length~\px_{\pi_1})$.   We
+  closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
   denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
-  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\pi_1})\rbrack$.   Indeed,  the
+  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
   closures  in  the  heap  are   a  runtime  representation  of  these
@@
 liveness $\{x.\epsilon\}$ is created and  added to the union.  
-%% Also  notice  a  consequence  of laziness:  the  entire  expression
-%% including the  condition is not  evaluated if  the demand on  it is
-%% $\emptyset$.  This results in the empty liveness environment.
 To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
@@
 liveness of the  closure variables.  This liveness,  say $\sigma'$, is
-also  the  demand  on  $s$, and  the  resulting  liveness  environment
+also  the  demand  on  $s$, thus the  liveness  environment
 $\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
@@
 As noted earlier, $x.\alpha \in  \Lv_y$ specifies the liveness  of the
-reference $\heap\oxlb\rho(x), \alpha\oxrb$  only if $\alpha$ corresponds to  a closure-free path
+reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$ corresponds to  a closure-free path
 in \heap\   starting from $\rho(x)$.  If this  path is intercepted  by a
 closure,  say  $\CAR~{x_{\pi}}$, then  the  liveness  of further  heap
-contents starting from $x_{\pi}$ is given by $\Lv_{x_{\pi}}$.
+contents starting from $\mathit{x_{\pi}}$ is given by $\Lv_\mathit{x_{\pi}}$.
 
@@
     access
-    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\oxlb\rho(x), \alpha
-    \oxrb = \ell$. Here 
+    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
+    \rrbracket = \ell$. Here 
 %% $x.\alpha$  is  live  and
 %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
-%%     domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
-%%     \oxrb = \ell$  and $x.\alpha \in \Lv$, where 
+%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
+%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
 \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
@@
 \label{sec:experiments}
-  Our benchmark consists  of programs taken from nofib~\cite{nofib}
-and other similar fucntional programming test suites.  All our
+  Our benchmark mostly consists  of programs from nofib~\cite{nofib}
+suite.  All our
 benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
@@
 
-The statistics related to liveness analysis and DFA generation are shown in
+The statistics related to compile  time liveness analysis are shown in
 Table~\ref{tab:exp-results}(a). We observe that the analysis requires
@@
 The bottleneck in our analysis is the NFA to DFA conversion
-(worst-case exponential behaviour). To avoid this 
-overhead, we use a caching mechanism in our simulator that can
-reuse the DFA from earlier analysis if the program has not changed.
+(worst-case exponential behaviour). To avoid this compile-time
+overhead, we have added a caching mechanism in our simulator that can
+reuse the DFA from earlier analysis if the program has not changed..
 
@@
 average  number of  cells reclaimed  per garbage collection,  average number  of cells
-touched  per garbage collection,  and  total time  to
+touched  per garbage collection,  total  number of  dragged cells  and  total time  to
 perform  all collections.   It is  no  surprise that  number of  cells
 reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
-lower)  for LGC  based collection.   However, the cost  for LGC is that per garbage collection
+lower)  for LGC  based collection.   LGC  also reduces  the number  of
+dragged cells significantly. However, the cost  for LGC is that per garbage collection
 execution  time is  more  for  LGC, which  results  in higher  overall
@@
 collections.   However, LGC  run time  is still  competitive (slowdown
-within \warning{5X} of RGC for worst case), and is better for 3 benchmarks (best
+within 5X of RGC for worst case), and is better for 3 benchmarks (best
 case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
@@
 \newcommand{\hgt}{2.9cm}
-\newcommand{\LGCLine}{red}
-\newcommand{\RGCLine}{blue}
-\newcommand{\ReachLine}{black}
-\newcommand{\UseLine}{lightblue}
-
 \begin{figure*}[t]
@@
 %%    (b) enlarges  a part  of the  usage.  
-The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
+The blue and the red curves indicate the number of cons
 cells  in  the  active   semi-space  for  RGC  and  LGC
-respectively.  The \ReachLine\ curve represents the number of
-reachable cells and the  \UseLine\ curve represents the
+respectively.  The black curve represents the number of
+reachable cells and the  lightblue curve represents the
 number  of  cells  that  are actually  live  (of  which
@@
 Memory  usage graphs  for select  benchmarks are  shown
-Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
+Figure~\ref{fig:memory-usage},  Column   (a).   As  the
 number of  garbage collections  tend to be  very large,
-hence  we  show only  a  slice of the original graph  which  is
+for  each benchmark  we  also show  a  window which  is
 representative  of  the  behavior for  that  particular
 benchmark (Column (b)).  In all the programs we can see
-that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
-regularly dips below the RGC curve (\RGCLine\ line), the dip
+that  the   curve  corresponding  to  LGC   (red  line)
+regularly dips below the RGC curve (blue line), the dip
 is  specially  visible  for the  benchmarks  where  LGC
-outperforms RGC substantially.  The graphs also include the
-curve    for   reachable    cells   (\ReachLine,    obtained
+outperforms RGC substantially.  The graphs also include
+curve    for   reachable    cells   (black,    obtained
 approximately  by forcing  RGC to  run at  a very  high
-frequency)  and  the  live  cells  for a particular run
-(\UseLine, obtained by post processing the heap access
-at the end of the program).
-
-As demonstrated by the gap between the \RGCLine\ and the \UseLine, a large number of cells 
-which are unused by the program are still copied during RGC. LGC does a much better job of closing this gap 
-but still falls short of the precision achieved by LGC in case of eager languages
-~\cite{asati14lgc}. 
-A case for concern is the huge difference in the number of cells touched by LGC vs RGC for some programs.
-This is due to heap locations being  traversed multiple times during LGC. \warning{We have implemented heuristics to 
-overcome common cases and they show promising results}.
+frequency)  and  the  live  cells  for  particular  run
+(lightblue, obtained by post processing the heap access
+at the end of a program).
+
+One area of concern is  the huge gap between the actual
+liveness and  the liveness perceived by  our collector.
+In  case of  LGC for  eager languages~\cite{asati14lgc}
+the gap was  very narrow and almost  touched the actual
+liveness curve.  In case of lazy languages, this gap is
+largely   due   to   approximate   reachability   based
+collection  of closures.  To  implement liveness  based
+collection of closures, we need to record liveness data
+for  each  closure and  update  it  at each  evaluation
+point.   These space  and  run time  overheads make  it
+infeasible  to   incorporate  liveness  based   garbage collection  for
+closures in a practical garbage collectors.
 
@@
 We have  extended liveness-based garbage collection  to lazy languages
-and shown its benefit for a set of benchmark programs.  This was done
+and shown its benefit for  standard benchmark programs.  This was done
 by defining a liveness analysis of heap objects that we proved correct
@@
 \warning{Although we  provide an  implementation for  a liveness
-based garbage  collection scheme for a lazy  language, we
+based garbage  collection scheme for lazy  language, we
 do not  provide a formal  proof of its  correctness.  A
@@
   between  the   actual  liveness  and   the  perceived
-  liveness  of our  garbage collector. Using
+  liveness  of our  garbage collector.  Our experiments
+  (not reported here) show that a significant number of
+  dead  cells  get   trapped  inside  closures.   Using
   strictness  analysis  to  eagerly  evaluate  closures
-  might release more dead cells to be garbage
+  might release more of these  dead cells to be garbage
   collected.   Since our  garbage collector  can handle
@@
 
-\clearpage
-\section{Appendix: Soundness of liveness analysis}
-
-\begin{figure*}[t!]
+\newpage
+\ 
+\newpage
+
+\section{Appendix: Not to be included in the paper.}
+
+\begin{figure*}[h!]
 \begin{center}\footnotesize
@@
 \hline
-& \makecell{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
+& \makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
+  (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $
   \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  {\sc cons} \\
 \hline
-$\heap(\rho(x)) \mbox{ is } (v, d)$ & \makecell{$\rho, (\rho', \ell, e,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\
+  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
+{\sc car-bang} 
+\\
+\hline
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S,
+    \heap)}$, \\ $\heap(\rho(x)) \mbox{ is } (v, d)$} & \makecell[t]{$\rho, (\rho', \ell, e,
   \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
-  \rightsquigarrow \rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
+  \rightsquigarrow $ \\ $\rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
 {\sc car-select} \\
 \hline
-$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
-(\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
-s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
+    S, \heap)}$, \\ $\heap(\rho(x))
+\mbox{ is } \langle s, \rho'\rangle$ }& \makecell[t]{$\rho, S,
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow $\\ $  \rho', (\rho, \ell,
+  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$ }        &
+{\sc car-clo}\\
+\hline
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
+    S, \heap)}$, \\$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
+(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
 {\sc car-1-clo} \\
-
-\hline
-
-$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-  \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
-{\sc car-clo}
+%% \hline
+
+%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
+%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
+%% {\sc car-clo}
+%% \\
+\hline
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\
+  $\heap(\rho(x)) \mbox{ is } \bot$  or $\heap(\rho(y)) \mbox{ is }
+  \bot$} & $\rho, S,
+  \heap, (+~x~y), \cred{\sigma} \rightsquigarrow \bang$   &
+{\sc prim-1-bang} 
 \\
-
-\hline
-$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
- & \makecell{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
-  (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, \heap[\ell
-    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
+\hline
+
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho,
+    S, \heap)}$, \\ 
+$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$}
+ &
+
+\makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
+  (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $\rho', S, \heap[\ell
+    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}     
+ &
 {\sc prim-add} \\
 \hline
-$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
-  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
+  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho, \rho(x),
   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
@@
 \hline
-$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
-  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
+  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho,\rho(y),
   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
@@
 \hline
-{$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, \heap,
-  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (f~\myvec{x}), \sigma) =
+    (\rho, S, \heap)}$,\\ $\mathit{f}~\mbox{defined as}$
+$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell[t]{$\rho, S, \heap,
+  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow$\\$ [\myvec{y} \mapsto
     \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
@@
 \hline
-$\ell$ is a new location& \makecell{$\rho, S, \heap, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\LET~x\leftarrow
+  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
 {\sc let} \\ 
 \hline
-$\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
+    \sigma) = (\rho, S, \heap)}$, \\
+  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
+  \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
+{\sc if-bang} 
+\\
+\hline
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
+    \sigma) = (\rho, S, \heap)}$, \\$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
 \hline
-$\heap(\rho(x)) = 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
   \cred{\sigma}$   $\rightsquigarrow
@@
 \hline
-$\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
-  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
-\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
+  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow$ \\ $
+\rho', (\rho, \rho(x), (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
 \cred{\clazy\sigma}$}
@@
 \hline
-{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& \makecell{$\rho,
-  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell := v], e, 
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\
+  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
+  \heap, (\SRETURN~x), \cred{\sigma} \rightsquigarrow \bang$   &
+{\sc return-bang} 
+\\
+\hline
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) =
+    (\rho, S, \heap)}$,\\ $\heap(\rho(x))~\mbox{is in WHNF with
+    value}~v$} & \makecell[t]{$\rho,
+  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow$\\$ \rho', S, \heap[\ell := v], e, 
   \cred{\sigma'}$} &
@@
 \hline
-$\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
   (\SRETURN~x), \cred{\sigma}$  $
-  \rightsquigarrow$
-$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
+  \rightsquigarrow$ \\
+$\rho',~ (\rho, \rho(x), (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
   \cred{\sigma}$} &
@@
 \end{tabular}
-\caption{Minefield semantics.\label{fig:minefield-semantics}}
+\caption{Minefield semantics. The differences with the small-step
+  semantics have been highlighted by shading. \label{fig:minefield-semantics}}
 \end{center}
-\vskip -5mm
 \end{figure*}
 
-We shall now  present a proof of soundness of  the liveness analysis
-presented in Section~\ref{sec:liveness-analysis}.   While the proof is
-on the broad lines of \cite{asati14lgc}, there are several differences
-due to  the lazy  semantics of  the language  being 
-analyzed in this paper. Here is the idea behind the proof.
-\begin{enumerate}
-\item       We        enrich       the        standard       semantics
-  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
-  collection  just  before  each  $\rightsquigarrow$  transition.  The
-  simulated garbage collection starts from  the root-set and inserts a
-  special value  $\bot$ at each location  in the heap that  contains a
-  reference that is reachable but not live. Any attempt to dereference
-  such  locations during  the  transition will  result  in entering  a
-  special state denoted \bang. We call the semantics after
-  augmentation  \emph{minefield semantics}.
-\item \label{inline2} Assuming that a  program enters the \bang\ state,
-  we  show  how to  construct,  through  inline expansion,  a  program
-  without function calls which has the same minefield behavior.
-\item  The final  part  of the  proof shows  that  no program  without
-  function calls can enter a \bang\ state. As a consequence no program
-  (with or without function calls) can enter the \bang\ state. 
-\end{enumerate}
-
-\subsection{Minefield semantics}
-
-To set up the minefield semantics, we follow these steps:
-\begin{enumerate}
-\item   We   enrich    the   abstract   machine   state
-  $\rho,\,S,\,\heap,\,e$   to  $\rho,\,S,\,\heap,\,e,\,\sigma$,
-  where  $\sigma$  is  the  demand  on  the  expression
-  $e$. We  call such a state  a \emph{minefield state}.
-  The demand $\sigma$ is  ``dynamic'' in that it arises
-  from the  actual sequence of function  calls that led
-  to  the evaluation  of  $e$.  The  0-CFA demand  used
-  during    liveness     analysis    is     a    static
-  over-approximation   of   $\sigma$   that   considers
-  \emph{all  possible calls}  that  could  lead to  the
-  evaluation of  $e$.  The  information in  the continuation frames on
-  the   stack  $S$ are  also
-  similarly augmented with their  demands.  Thus a stacked
-  entry now takes  the form $(\rho, x,  e, \sigma)$.  A
-  modification  of  the   small  step  semantics  which
-  carries the extra information  is shown in the bottom
-  table of Figure~\ref{fig:minefield-semantics}.
-\item $GC$  models a liveness based  garbage collection
-  $GC(\rho, S,  \heap, \sigma)$ returns $(\rho',  S', \heap')$,
-  where the changes are due to certain references being
-  replaced by  $\bot$.  This represents the  act of not
-  retaining   (effectively   garbage  collecting)   the
-  objects  pointed to  by  these  references during  an
-  actual  garbage  collection.  This  is  done  in  the
-  following way:
-  \begin{enumerate}
-  \item Given the current state $\rho, S,
-    \heap,   e,  \sigma$   and   the  demand   transformers
-    $\Lfonly$, we can  construct a liveness environment
-    \Lv for $e$.   If  $e$  is  an expression,  then  $  \Lv  =
-    \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
-    application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
-    We  can similarly  define the  liveness environment
-    for each  of the  suspended evaluation  contexts in
-    $S$, giving  a stack of liveness  environments that
-    we shall denote $\mathsf{SL}$.
-  \item \emph{Garbage collection  of objects pointed by
-    root   set:}  For   each   $x  \in   domain(\rho)$,
-    $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
-    \Lv$.    Similarly,    for   each    stack    entry
-    $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
-    corresponding      liveness     environment      in
-    $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
-    $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
-  \item  \emph{Garbage  collection  of objects  in  the
-    heap:} For each location  $\ell$, $\heap'(\ell) = \bot$
-    iff  there   is  no  $x$  in   either  the  current
-    environment  or  one  of the  stacked  environments
-    $\rho$, such  that for some  $\alpha$ corresponding
-    to a  closure-free path  in the heap  starting from
-    $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
-    \Lv$. Here \Lv\ is the liveness of the variables in
-    the current environment or the stacked environment.
-
-\warning{We have to tell  the reader about closure free
-  paths  earlier  on.   Also,  has  $\heap[x.\alpha]$  been
-  defined?}
-  \end{enumerate}
-Starting from the initial state $([\;]_\rho,([\;]_\rho,
-\ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
-every transition  is preceded  by a  garbage collection
-using $GC$.  The details of the transition for the {\sc
-  car-clo}  rule   is  shown   in  the  top   table  of
-Figure~\ref{fig:minefield-semantics}.  If  this  or  an
-earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
-bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
-enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
-the transition is the same as the earlier {\sc car-clo}
-rule.
-\end{enumerate}
-
-\subsection{Inlining}
-
-Consider a call tree, each  of whose nodes represents a
-function  that  was  called (but  did  not  necessarily
-return because  of a \bang).   Assume that each  of the
-nodes of  the tree is  also annotated with  the program
-point where  the corresponding call was  invoked.  This
-tree  can  be  used  to  inline  function  calls  in  a
-hierarchical fashion.  The details of inlining a single
-function      call     $(\LET\,\,      x     \leftarrow
-(f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
-Let  $f$ be  defined  (after renaming  its formals  and
-locals  to  be  disjoint from  existing  variables)  by
-$(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
-replaced  by   a  sequence  of  $\LET$s   of  the  form
-$z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
-but with its  $(\SRETURN\,\,w)$ expressions replaced by
-$(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
-We prefer to use the  identity function $\ID$ as a form
-of no-op rather than  introducing the form $(\LET\,\, x
-\leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
-$\LET$-definition   is  a   simple  variable.    For  a
-call-less program,  the initial state of  the minefield
-semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
-\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
-(e_{\mainpgm})$.
-
-\subsection{Soundness Results}
-
-  We  consider  a  garbage  collection  followed  by  a
-  transition as a step.  We show by induction that, for
-  programs  without function  calls, starting  from the
-  initial  state  that  has $e_{\main}$  as  the  first
-  expression to  be evaluated  instead of  $\main$, any
-  transition of $n$ steps occurs without a \bang.
-
-We  first  need  an auxiliary  result  about  minefield
-semantics. Consider  a trace of a  minefield execution.
-For every minefield state  $(\rho, S, \heap, e(s), \sigma)$
-that appears  on the LHS of  a $\rightsquigarrow$ step,
-the  demand   $\sigma$  on   the  expression   $e$  (or
-application $s$) is non-null.  This can be proved by an
-induction  on  the  number  of  steps  leading  to  the
-minefield  state.   The  base step  holds  because  the
-demand $\sigma_{all}$  on $\mainpgm$ is  non-null.  The
-inductive step  follows from  the observation  that for
-each  step of  the minefield  semantics, if  the demand
-$\sigma$ on  the LHS of  a minefield step  is non-null,
-the demand on  the RHS is a  transformation of $\sigma$
-(for example $\clazy\sigma$) which is also non-null.
-
-Let us clarify this observation:  Since the minefield semantics models
-lazy  evaluation,  parts  of  the  program may  not  be  evaluated  at
-all.  However,  if  an  expression  (or  application)  happens  to  be
-evaluated, it  will appear  on the  LHS of a  minefield step,  and the
-context-sensitive demand on it will be non-null.
-
-We  show  next  that   the  minefield  execution  of  a
-call-less program  cannot go \bang. Note  that, for the
-sake of  concreteness, by a single  minefield execution
-we   shall  mean   the  evaluation   of  $(main)$   (or
-$e_{\mainpgm}$,  for call-less  programs)  to its  WHNF
-driven  by   the  printing  mechanism   represented  by
-\print. With  minor variations, the proof  will also be
-applicable   for   subsequent  evaluations   fired   by
-$\print$.
-
-\begin{lemma}
-\label{lemma:call-less-cannot-go-bang-app}
-Consider the minefield execution of  a program without function calls.
-Such a program cannot enter the \bang\ state.
-\end{lemma}
-\begin{proof}
-Consider a state in the minefield execution of a program $(\rho, S, \heap,
-e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
-steps  leading to  this state  that  the next  step cannot  lead to  a
-\bang\ state.
-
-When $n$ is 0, the  state is $([\;]_\rho,\_, [\;]_{S} ,
-[\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.  Further,  we
-have   to  show   that   a  $GC(..)$   followed  by   a
-$\rightsquigarrow$ step cannot lead  to a \bang\ state.
-Now, because  our programs  are in  ANF, $e_{\mainpgm}$
-can  only  be a  $\LET$.  A  {\sc  let} step  does  not
-dereference anything and thus cannot result in a \bang.
-
-For the inductive step, we  shall show that none of the
-minefield steps that  involves dereferencing results in
-a   \bang.   These  are   the   steps   which  have   a
-$\heap(\rho(...))$ in the premise.  Recall that a minefield
-step   consists   of   a   GC  step   followed   by   a
-$\rightsquigarrow$.   The   $\rightsquigarrow$  can  go
-\bang\ because it dereferences a $\bot$ inserted by the
-immediately  preceding  GC  or  the GC  of  an  earlier
-minefield step.  \warning{However the  demand $\sigma'$
-  on basis of which an earlier GC would have inserted a
-  $\bot$,  would  have   included  the  current  demand
-  $\sigma$.}  Thus,  if the  earlier GC had  inserted a
-$\bot$, so would the current  GC.  Thus it is enough to
-consider the $\rightsquigarrow$  along with the current
-GC and show that it does not lead to a \bang.
-
-We consider the steps {\sc car-clo} and {\sc car-1-clo}
-only. The rest  of the steps are similar.  For the {\sc
-  car-clo}  step in  the state  $\rho, S,  \heap, (\CAR~x),
-\sigma$, we  know that $\sigma$ is  non-null. There for
-the  liveness  of  $x$  includes  $\epsilon$,  and  the
-dereferencing $\heap(\rho(x))$ will go without \bang.
-
-For the  {\sc car-1-clo}  step, observe that  there are
-two dereferences.  First $x$  is dereferenced to  get a
-cons  cell  and then  the  head  of  the cons  cell  is
-dereferenced  to  obtain  a  closure.   If  the  demand
-$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
-of    $x$   will    include    both   $\epsilon$    and
-$\acar\epsilon$,  and  a  GC with  this  liveness  will
-neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
-first   component  of   the   cons   cell.  Thus   both
-dereferences    will    happen   normally    and    the
-$\rightsquigarrow$ step will not enter \bang.
-\end{proof}
-
-Now we are ready to prove the main soundness result.
-
-\begin{theorem}
-The  minefield  execution of  no  program  can enter  a
-\bang\ state.
-\end{theorem}
-
-\begin{proof}
-Assume to  the contrary that  a program $P$  enters the
-\bang\  state.  We can  transform  $P$  to a  call-less
-program $P'$ such that  the minefield executions of $P$
-and $P'$  are identical  except for change  of variable
-names.  \warning{Check  whether  the previous  line  is
-  correct.}                 However,                 by
-Lemma~\ref{lemma:call-less-cannot-go-bang-app} we know that
-$P'$,   a   call-less   program,   cannot   enter   the
-\bang\  state.  Therefor  $P$  also  cannot  enter  the
-\bang\ state.
-\end{proof}
+%% We shall now  present a proof of soundness of  the liveness analysis
+%% presented in Section~\ref{sec:liveness-analysis}.   While the proof is
+%% on the broad lines of \cite{asati14lgc}, there are several differences
+%% due to  the lazy  semantics of  the language  being 
+%% analyzed in this paper. Here is the idea behind the proof.
+%% \begin{enumerate}
+%% \item       We        enrich       the        standard       semantics
+%%   (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
+%%   collection  just  before  each  $\rightsquigarrow$  transition.  The
+%%   simulated garbage collection starts from  the root-set and inserts a
+%%   special value  $\bot$ at each location  in the heap that  contains a
+%%   reference that is reachable but not live. Any attempt to dereference
+%%   such  locations during  the  transition will  result  in entering  a
+%%   special state denoted \bang. We call the semantics after
+%%   augmentation  \emph{minefield semantics}.
+%% \item \label{inline2} Assuming that a  program enters the \bang\ state,
+%%   we  show  how to  construct,  through  inline expansion,  a  program
+%%   without function calls which has the same minefield behavior.
+%% \item  The final  part  of the  proof shows  that  no program  without
+%%   function calls can enter a \bang\ state. As a consequence no program
+%%   (with or without function calls) can enter the \bang\ state. 
+%% \end{enumerate}
+
+%% \subsection{Minefield semantics}
+
+%% To set up the minefield semantics, we follow these steps:
+%% \begin{enumerate}
+%% \item   We   enrich    the   abstract   machine   state
+%%   $\rho,\,S,\,\heap,\,e$   to  $\rho,\,S,\,\heap,\,e,\,\sigma$,
+%%   where  $\sigma$  is  the  demand  on  the  expression
+%%   $e$. We  call such a state  a \emph{minefield state}.
+%%   The demand $\sigma$ is  ``dynamic'' in that it arises
+%%   from the  actual sequence of function  calls that led
+%%   to  the evaluation  of  $e$.  The  0-CFA demand  used
+%%   during    liveness     analysis    is     a    static
+%%   over-approximation   of   $\sigma$   that   considers
+%%   \emph{all  possible calls}  that  could  lead to  the
+%%   evaluation of  $e$.  The  information in  the continuation frames on
+%%   the   stack  $S$ are  also
+%%   similarly augmented with their  demands.  Thus a stacked
+%%   entry now takes  the form $(\rho, x,  e, \sigma)$.  A
+%%   modification  of  the   small  step  semantics  which
+%%   carries the extra information  is shown in the bottom
+%%   table of Figure~\ref{fig:minefield-semantics}.
+%% \item $GC$  models a liveness based  garbage collection
+%%   $GC(\rho, S,  \heap, \sigma)$ returns $(\rho',  S', \heap')$,
+%%   where the changes are due to certain references being
+%%   replaced by  $\bot$.  This represents the  act of not
+%%   retaining   (effectively   garbage  collecting)   the
+%%   objects  pointed to  by  these  references during  an
+%%   actual  garbage  collection.  This  is  done  in  the
+%%   following way:
+%%   \begin{enumerate}
+%%   \item Given the current state $\rho, S,
+%%     \heap,   e,  \sigma$   and   the  demand   transformers
+%%     $\Lfonly$, we can  construct a liveness environment
+%%     \Lv for $e$.   If  $e$  is  an expression,  then  $  \Lv  =
+%%     \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
+%%     application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
+%%     We  can similarly  define the  liveness environment
+%%     for each  of the  suspended evaluation  contexts in
+%%     $S$, giving  a stack of liveness  environments that
+%%     we shall denote $\mathsf{SL}$.
+%%   \item \emph{Garbage collection  of objects pointed by
+%%     root   set:}  For   each   $x  \in   domain(\rho)$,
+%%     $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
+%%     \Lv$.    Similarly,    for   each    stack    entry
+%%     $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
+%%     corresponding      liveness     environment      in
+%%     $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
+%%     $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
+%%   \item  \emph{Garbage  collection  of objects  in  the
+%%     heap:} For each location  $\ell$, $\heap'(\ell) = \bot$
+%%     iff  there   is  no  $x$  in   either  the  current
+%%     environment  or  one  of the  stacked  environments
+%%     $\rho$, such  that for some  $\alpha$ corresponding
+%%     to a  closure-free path  in the heap  starting from
+%%     $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
+%%     \Lv$. Here \Lv\ is the liveness of the variables in
+%%     the current environment or the stacked environment.
+
+%% \warning{We have to tell  the reader about closure free
+%%   paths  earlier  on.   Also,  has  $\heap[x.\alpha]$  been
+%%   defined?}
+%%   \end{enumerate}
+%% Starting from the initial state $([\;]_\rho,([\;]_\rho,
+%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
+%% every transition  is preceded  by a  garbage collection
+%% using $GC$.  The details of the transition for the {\sc
+%%   car-clo}  rule   is  shown   in  the  top   table  of
+%% Figure~\ref{fig:minefield-semantics}.  If  this  or  an
+%% earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
+%% bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
+%% enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
+%% the transition is the same as the earlier {\sc car-clo}
+%% rule.
+%% \end{enumerate}
+
+%% \subsection{Inlining}
+
+%% Consider a call tree, each  of whose nodes represents a
+%% function  that  was  called (but  did  not  necessarily
+%% return because  of a \bang).   Assume that each  of the
+%% nodes of  the tree is  also annotated with  the program
+%% point where  the corresponding call was  invoked.  This
+%% tree  can  be  used  to  inline  function  calls  in  a
+%% hierarchical fashion.  The details of inlining a single
+%% function      call     $(\LET\,\,      x     \leftarrow
+%% (f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
+%% Let  $f$ be  defined  (after renaming  its formals  and
+%% locals  to  be  disjoint from  existing  variables)  by
+%% $(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
+%% replaced  by   a  sequence  of  $\LET$s   of  the  form
+%% $z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
+%% but with its  $(\SRETURN\,\,w)$ expressions replaced by
+%% $(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
+%% We prefer to use the  identity function $\ID$ as a form
+%% of no-op rather than  introducing the form $(\LET\,\, x
+%% \leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
+%% $\LET$-definition   is  a   simple  variable.    For  a
+%% call-less program,  the initial state of  the minefield
+%% semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
+%% \ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
+%% (e_{\mainpgm})$.
+
+%% \subsection{Soundness Results}
+
+%%   We  consider  a  garbage  collection  followed  by  a
+%%   transition as a step.  We show by induction that, for
+%%   programs  without function  calls, starting  from the
+%%   initial  state  that  has $e_{\main}$  as  the  first
+%%   expression to  be evaluated  instead of  $\main$, any
+%%   transition of $n$ steps occurs without a \bang.
+
+%% We  first  need  an auxiliary  result  about  minefield
+%% semantics. Consider  a trace of a  minefield execution.
+%% For every minefield state  $(\rho, S, \heap, e(s), \sigma)$
+%% that appears  on the LHS of  a $\rightsquigarrow$ step,
+%% the  demand   $\sigma$  on   the  expression   $e$  (or
+%% application $s$) is non-null.  This can be proved by an
+%% induction  on  the  number  of  steps  leading  to  the
+%% minefield  state.   The  base step  holds  because  the
+%% demand $\sigma_{all}$  on $\mainpgm$ is  non-null.  The
+%% inductive step  follows from  the observation  that for
+%% each  step of  the minefield  semantics, if  the demand
+%% $\sigma$ on  the LHS of  a minefield step  is non-null,
+%% the demand on  the RHS is a  transformation of $\sigma$
+%% (for example $\clazy\sigma$) which is also non-null.
+
+%% Let us clarify this observation:  Since the minefield semantics models
+%% lazy  evaluation,  parts  of  the  program may  not  be  evaluated  at
+%% all.  However,  if  an  expression  (or  application)  happens  to  be
+%% evaluated, it  will appear  on the  LHS of a  minefield step,  and the
+%% context-sensitive demand on it will be non-null.
+
+%% We  show  next  that   the  minefield  execution  of  a
+%% call-less program  cannot go \bang. Note  that, for the
+%% sake of  concreteness, by a single  minefield execution
+%% we   shall  mean   the  evaluation   of  $(main)$   (or
+%% $e_{\mainpgm}$,  for call-less  programs)  to its  WHNF
+%% driven  by   the  printing  mechanism   represented  by
+%% \print. With  minor variations, the proof  will also be
+%% applicable   for   subsequent  evaluations   fired   by
+%% $\print$.
+
+%% \begin{lemma}
+%% \label{lemma:call-less-cannot-go-bang-app}
+%% Consider the minefield execution of  a program without function calls.
+%% Such a program cannot enter the \bang\ state.
+%% \end{lemma}
+%% \begin{proof}
+%% Consider a state in the minefield execution of a program $(\rho, S, \heap,
+%% e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
+%% steps  leading to  this state  that  the next  step cannot  lead to  a
+%% \bang\ state.
+
+%% When $n$ is 0, the  state is $([\;]_\rho,\_, [\;]_{S} ,
+%% [\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.  Further,  we
+%% have   to  show   that   a  $GC(..)$   followed  by   a
+%% $\rightsquigarrow$ step cannot lead  to a \bang\ state.
+%% Now, because  our programs  are in  ANF, $e_{\mainpgm}$
+%% can  only  be a  $\LET$.  A  {\sc  let} step  does  not
+%% dereference anything and thus cannot result in a \bang.
+
+%% For the inductive step, we  shall show that none of the
+%% minefield steps that  involves dereferencing results in
+%% a   \bang.   These  are   the   steps   which  have   a
+%% $\heap(\rho(...))$ in the premise.  Recall that a minefield
+%% step   consists   of   a   GC  step   followed   by   a
+%% $\rightsquigarrow$.   The   $\rightsquigarrow$  can  go
+%% \bang\ because it dereferences a $\bot$ inserted by the
+%% immediately  preceding  GC  or  the GC  of  an  earlier
+%% minefield step.  \warning{However the  demand $\sigma'$
+%%   on basis of which an earlier GC would have inserted a
+%%   $\bot$,  would  have   included  the  current  demand
+%%   $\sigma$.}  Thus,  if the  earlier GC had  inserted a
+%% $\bot$, so would the current  GC.  Thus it is enough to
+%% consider the $\rightsquigarrow$  along with the current
+%% GC and show that it does not lead to a \bang.
+
+%% We consider the steps {\sc car-clo} and {\sc car-1-clo}
+%% only. The rest  of the steps are similar.  For the {\sc
+%%   car-clo}  step in  the state  $\rho, S,  \heap, (\CAR~x),
+%% \sigma$, we  know that $\sigma$ is  non-null. There for
+%% the  liveness  of  $x$  includes  $\epsilon$,  and  the
+%% dereferencing $\heap(\rho(x))$ will go without \bang.
+
+%% For the  {\sc car-1-clo}  step, observe that  there are
+%% two dereferences.  First $x$  is dereferenced to  get a
+%% cons  cell  and then  the  head  of  the cons  cell  is
+%% dereferenced  to  obtain  a  closure.   If  the  demand
+%% $\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
+%% of    $x$   will    include    both   $\epsilon$    and
+%% $\acar\epsilon$,  and  a  GC with  this  liveness  will
+%% neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
+%% first   component  of   the   cons   cell.  Thus   both
+%% dereferences    will    happen   normally    and    the
+%% $\rightsquigarrow$ step will not enter \bang.
+%% \end{proof}
+
+%% Now we are ready to prove the main soundness result.
+
+%% \begin{theorem}
+%% The  minefield  execution of  no  program  can enter  a
+%% \bang\ state.
+%% \end{theorem}
+
+%% \begin{proof}
+%% Assume to  the contrary that  a program $P$  enters the
+%% \bang\  state.  We can  transform  $P$  to a  call-less
+%% program $P'$ such that  the minefield executions of $P$
+%% and $P'$  are identical  except for change  of variable
+%% names.  \warning{Check  whether  the previous  line  is
+%%   correct.}                 However,                 by
+%% Lemma~\ref{lemma:call-less-cannot-go-bang-app} we know that
+%% $P'$,   a   call-less   program,   cannot   enter   the
+%% \bang\  state.  Therefor  $P$  also  cannot  enter  the
+%% \bang\ state.
+%% \end{proof}
 

### 2015-11-29 paper/paper.tex

@@
 \label{sec:experiments}
-  Our benchmark mostly consists  of programs from nofib~\cite{nofib}
-suite.  All our
+  Our benchmark consists  of programs taken from nofib~\cite{nofib}
+and other similar functional programming test suites.  All our
 benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
@@
 \newcommand{\hgt}{2.9cm}
+\newcommand{\LGCLine}{blue}
+\newcommand{\RGCLine}{red}
+\newcommand{\ReachLine}{black}
+\newcommand{\UseLine}{lightblue}
+
 \begin{figure*}[t]
@@
 \begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
+   \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
 \\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
+\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
 \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
 %\\ (a) & (b)

### 2015-11-29 paper/paper.tex
added missing file

@@
 by a set of {\em access paths} i.e.  prefix-closed set of strings from
-$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (typically a
-variable  reference) and  a heap  \heap, semantically  an access  path
+$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (usually a 
+reference corresponding to a variable) and  a heap  \heap, semantically  an access  path
 $\alpha$ represents  a reference, denoted $\heap\llbracket\ell,  \alpha\rrbracket$, in
@@
 in \heap\   starting from $\rho(x)$.  If this  path is intercepted  by a
-closure,  say  $\CAR~{x_{\pi}}$, then  the  liveness  of further  heap
+closure,  say  $(\CAR~{x_{\pi}})$, then  the  liveness  of further  heap
 contents starting from $\mathit{x_{\pi}}$ is given by $\Lv_\mathit{x_{\pi}}$.
@@
 every time replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the  sentential   form   by
-$\epsilon$. Conversely,  if the TM  does not accept the  input string,
+$\epsilon$. However,  if the TM  does not accept the  input string,
 then  any sentential  form derived  from the  start symbol  would have

### 2015-11-29 lazy-scheme/post_process/gen_stats_and_plotall.sh
ktour

@@ -126,15 +126,15 @@ for bm in $BMs
 do
     #for opt in gc-plain gc-live gc-freq=${freq[$bm]}
     #for opt in gc-freq=${freq[$bm]}
-    #for opt in gc-plain gc-live
-    #do 
-    #   echo "Processing $bm [$opt]"
-    #   $DIR/gen_gc_stats.sh $OPTARG \
-    #       -bmdir ../benchmarks/programs/$bm \
-    #       -gcopt $opt -heap ${limit[$bm]}
-    #one
-    #ash $DIR/val.plot $bm ${xtics[$bm]} ${ytics[$bm]}
-    #ash $DIR/window.plot $bm ${xlimit[$bm]} ${ylimit[$bm]} 
+    for opt in gc-plain gc-live
+    do 
+       echo "Processing $bm [$opt]"
+       $DIR/gen_gc_stats.sh $OPTARG \
+           -bmdir ../benchmarks/programs/$bm \
+           -gcopt $opt -heap ${limit[$bm]}
+    done
+    #bash $DIR/val.plot $bm ${xtics[$bm]} ${ytics[$bm]}
+    #bash $DIR/window.plot $bm ${xlimit[$bm]} ${ylimit[$bm]} 
     gnuplot $DIR/plots/${bm}.gnu
     gnuplot $DIR/plots/${bm}_win.gnu
 

### 2015-11-29 lazy-scheme/src/SchemeAST_Defs.cpp
Added optimization for cons revisits

@@ -47,10 +47,10 @@ cons* reduceParamToWHNF(cons* cell)
 {
 	cons* heap_cell = update_heap_refs.top();
 //    DBG(
-//	if (!is_valid_address(heap_cell))
-//		cout << "!!!Address may be invalid for " << heap_cell << endl;
-//	//Removed assertion, to test passing unused parameters. Need to confirm if it works.
-//	//assert(is_valid_address(heap_cell));
+	if (!is_valid_address(heap_cell))
+		cout << "!!!Address may be invalid for " << heap_cell << endl;
+	//Removed assertion, to test passing unused parameters. Need to confirm if it works.
+	assert(is_valid_address(heap_cell));
 //    return heap_cell;
 //	)
 	if (heap_cell->inWHNF)

### 2015-11-29 lazy-scheme/src/gc.cpp

@@ -50,6 +50,7 @@ actRec *stack_start, *stack_top;
 GCStatus gc_status = gc_plain;
 extern int gccount;
 extern unordered_map<string, unsigned int> func_heap_cell_reqd;
+unordered_map<cons* , unordered_set<state_index>> heap_dfa_map;
 
 //ofstream gout("gc_addr.txt", ios::app);
 
@@ -1887,6 +1888,8 @@ void liveness_gc()
 #endif
 	  DBG(pre << "Doing liveness based GC #" << gccount << " after " << num_of_allocations << " allocations"<<endl);
 	  swap_buffer();
+	  DBG(pre << "Number of unique cons cells " << heap_dfa_map.size()<<endl);
+	  heap_dfa_map.clear();
 
   for (deque<actRec>::iterator stackit = actRecStack.begin();stackit != actRecStack.end(); ++stackit)
     {
@@ -1899,6 +1902,7 @@ void liveness_gc()
     	  DBG(pre << "Looking up " << nodeName << endl);
     	  if (got != statemap.end())
     	  {
+
     		  DBG(pre << "Processing var " << vhit->varname << endl);
     		  DBG(pre << "Index is " << ((cons*)vhit->ref - (cons*)buffer_dead) << endl);
     		  cons *addr   = followpaths(static_cast<cons*>(vhit->ref), got->second, pre);
@@ -1962,6 +1966,24 @@ cons* followpaths(cons* loc, state_index index, ostream& out)
   {
 	  DBG(out << "Copying cell from " << loc << " with type " << loc->typecell << endl);
  	  DBG(out << "state index is " << index << endl);
+
+ 	  if (heap_dfa_map.find(loc) != heap_dfa_map.end() &&
+ 			  heap_dfa_map[loc].find(index) != heap_dfa_map[loc].end())
+ 	  {
+ 		  //Already copied using the same liveness state
+ 		  assert(loc->forward);
+ 		  loccopy = (cons*)loc->forward;
+ 		  DBG(out << "REVISITING LOC " << loc << " with index " << index << endl);
+ 		  return loccopy;
+
+ 	  }
+ 	  else
+ 	  {
+ 		  auto &visited_set = heap_dfa_map[loc];
+ 		  visited_set.insert(index);
+ 		  DBG(out << "FIRST visit to cons cell " << loc << " with state " << index << endl);
+ 	  }
+
  	  loccopy = copy(loc, out);
[…]

### 2015-11-29 lazy-scheme/post_process/createTable.sh
new set of touched cells

@@ -4,7 +4,7 @@ SCRIPT1=$DIR/touchedAndCopied.awk
 AllBMs="lambda nperm treejoin lcss sudoku fibheap nqueens gc_bench"
 AllBMs="fibheap sudoku nperm paraffins lcss huffman knightstour nqueens deriv treejoin lambda gc_bench "
 #AllBMs="sudoku"
-echo "\\newcommand{\\rlratio}{$\\frac{RGC}{LGC}$}"
+echo "\\newcommand{\\rlratio}{$\\frac{LGC}{RGC}$}"
 cat <<EOF
 \begin{tabular}{|@{\ }c@{\ }| @{\ }r@{\ }|@{\ }r@{\ }| @{\ }r@{\ }|@{\ }r@{\ }|r@{\ }| r@{\ }|r@{\ }|r@{\ }| r@{\ }|r@{\ }|r@{\ }|}
 \hline
@@ -83,7 +83,7 @@ do
     r2=`grep "GC Time" $FILE | cut -d= -f3`
     ratio=`bc -l << EOF                                                  
 scale=2
-$r1/$r2
+$r2/$r1
 EOF
 `
     if [ "x$ratio" = "x" ]; then

### 2015-11-29 paper/paper.tex
compressing...

@@
 To show the effectiveness of our method, we
-implement a copying collector that uses the results of the liveness
+implemented a copying collector that uses the results of the liveness
 analysis to preserve live 
-objects, both evaluated (i.e.  in WHNF) and closures.  Our experiments
+objects, both evaluated (i.e., in WHNF) and closures.  Our experiments
 confirm  that  for  programs  running with  a  liveness-based  garbage
@@
 memory.  The
-allocation  is either  explicit (using  constructors, for example) or
+allocation  is either  explicit (i. e., using  constructors) or
 implicit (creating  closures).  Programs  written in  lazy
@@
 While  the runtime  system  of most  functional  languages includes  a
-garbage  collector    to  efficiently  reclaim  memory,  empirical
-studies on Scheme~\cite{karkare06effectiveness} and, more importantly,
-on  Haskell~\cite{rojemo96lag}  programs  have shown  that  garbage collectors  leave
-uncollected a  large number of  memory objects that are  reachable but
-not live (here {\em live} means  the object can potentially be used by
-the program  at a  later stage).  This  results in  unnecessary memory
+garbage collector to efficiently  reclaim memory, empirical studies on
+Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
+Haskell~\cite{rojemo96lag} programs have shown that garbage collectors
+leave uncollected a large number  of memory objects that are reachable
+but not live (here {\em live} means the object can potentially be used
+by the program at a  later stage).  This results in unnecessary memory
 retention.
@@
 contributions of this paper are
-\begin{itemize}
+\begin{myitemize}
 \item  We  formulate  a  liveness  analysis for  a  lazy  first-order
@@
   show the effectiveness of liveness-based garbage collection.
-\end{itemize}
+\end{myitemize}
 As far as we know,  this is the first work on liveness-based garbage
@@
 \end{eqnarray*}
+\vspace*{-6pt}
   \caption{The syntax of our language}\label{fig:lang-syntax}
-\figrule
+%\figrule
 \normalsize
+\vspace*{-6pt}
 \end{figure}
@@
 %}\comment
+\vspace*{-6pt}
 \end{figure*}
@@
 
-\newpage
-\ 
-\newpage
-
-\section{Appendix: Not to be included in the paper.}
-
+\clearpage
+\appendix
+%\section{Appendix: Not to be included in the paper.}
 \begin{figure*}[h!]
+\noindent{\Large \bf A. Minefield Semantics}
 \begin{center}\footnotesize
-\renewcommand{\arraystretch}{1.5}
+\renewcommand{\arraystretch}{1.2}
 

### 2015-11-29 lazy-scheme/post_process/createTable.sh
Updated memory requirements for BMs

@@ -29,7 +29,7 @@ Lmin["huffman"]=72
 Lmin["knightstour"]=642303
 Lmin["nqueens"]=1082
 Lmin["deriv"]=589
-Lmin["treejoin"]=677800
+Lmin["treejoin"]=887005
 Lmin["lambda"]=18169
 Lmin["gc_bench"]=72
 

### 2015-11-29 paper/paper.tex
compressing...
more refs

@@
-\documentclass[preprint,9pt]{sigplanconf}
+\documentclass[preprint, 9pt]{sigplanconf}
 \usepackage{etex}
@@
 \usepackage{mathrsfs}
-\usepackage{lineno}
+%%\usepackage{lineno}
 \usepackage{booktabs}
@@
 \begin{document}
-\linenumbers 
+%% \linenumbers 
 % --- Author Metadata here ---
@@
 \label{sec:experiments}
-  Our benchmark consists  of programs taken from nofib~\cite{nofib}
-and other similar functional programming test suites.  All our
-benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
-i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
-Ubuntu 14.04.
+  Our benchmark consists of programs taken from nofib~\cite{nofib} and
+  other  similar  functional  programming  resources~\cite{PLT-Scheme,
+    gc_bench, huffman-sicp}.   All our  benchmarks were executed  on a
+  machine having 8  core Intel(R) Core(TM) i7-4770 3.40GHz  CPU , 8192
+  KB L2 cache, 16 GB RAM running 64 bit Ubuntu 14.04.
 

### 2015-11-30 paper/paper.tex
latest - 30th - d-day

@@
 For instance, corresponding to the production $\nont{i}{\acar} \rightarrow
-\acar \nont{j}{\acar}\bcar$,  consider
-the program:
+\acar \nont{j}{\acar}\bcar$,
+the program is:
 
@@
 
-If there more than one productions with the same LHS, then the
-corresponding function creates joins them with a $\SIF$.   If there
-are $n$ such functions,  then \mainpgm\  creates  a $n$-way
-branch  and inserts  a  single call  to a  distinct  function in  each
-branch. It can  be verified that the grammar for the function shown
-above, includes the productions---$\nont{\mathit{all}}{}$
-corresponds to the demand $\sigma_{\mathit{all}}$:
+The body of the function corresponds  to the RHS of the production. If
+there are  more than one  productions with the same  LHS non-terminal,
+then the function for the non-terminal joins the program fragments for
+the  RHS  with a  $\SIF$.   If  there  are  $n$ such  functions,  then
+\mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
+distinct function in each branch. It  can be verified that the grammar
+for      the     function      shown      above,     includes      the
+productions shown below. $\nont{\mathit{all}}{}$   corresponds   to  the   demand
+$\sigma_{\mathit{all}}$:
 \begin{align*}
@@
 $\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
-\nont{\mathit{all}}{}$, where $\beta_1'$ and $\beta_2'$ do not contain
-any non-terminals.
+\nont{\mathit{all}}{}$.
 
@@
 accepts the input string, then the earlier derivation can be replayed,
-every time replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the  sentential   form   by
-$\epsilon$. However,  if the TM  does not accept the  input string,
-then  any sentential  form derived  from the  start symbol  would have
-exactly one non-terminal, and this non-terminal will be different from
-$\nont{\mathit{final}}{c}$.  Thus  the  empty   string  would  not  be
+every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
+sentential form by $\epsilon$. However, if  the TM does not accept the
+input string, then every sentential form derived from the start symbol
+would    have     at    least    one    non-terminal from the grammar
+of Lemma~\ref{lemma:grammar-undecidable} that is different
+from $\nont{\mathit{final}}{c}$.   Thus  the  empty  string  would  not  be
 derivable from the grammar.
@@
   Our benchmark consists of programs taken from nofib~\cite{nofib} and
-  other  similar  functional  programming  resources~\cite{PLT-Scheme,
+  other repositories for  functional  programs~\cite{PLT-Scheme,
     gc_bench, huffman-sicp}.   All our  benchmarks were executed  on a
@@
 
-The statistics related to liveness analysis and DFA generation are shown in
-Table~\ref{tab:exp-results}(a). We observe that the analysis requires
-a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
-The bottleneck in our analysis is the NFA to DFA conversion
-(worst-case exponential behaviour). To avoid this 
-overhead, we use a caching mechanism in our simulator that can
-reuse the DFA from earlier analysis if the program has not changed.
-
-Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
-statistics  for RGC  and  LGC.  We  report the  number  of garbage collection  events,
-average  number of  cells reclaimed  per garbage collection,  average number  of cells
-touched  per garbage collection,  and  total time  to
-perform  all collections.   It is  no  surprise that  number of  cells
-reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
-lower)  for LGC  based collection.   However, the cost  for LGC is that per garbage collection
-execution  time is  more  for  LGC, which  results  in higher  overall
-execution time  for garbage  collection, even  with reduced  number of
-collections.   However, LGC  run time  is still  competitive (slowdown
-within \warning{5X} of RGC for worst case), and is better for 3 benchmarks (best
-case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
-synthetic benchmark  that allocates complete binary  trees of various
-sizes.  These binary trees are never used by the program. As a result,
-the benchmark highly favours LGC.  We have included this benchmark for
-completeness  only  and   the  numbers  for  it   are  not  considered
-representative of practical programs.
+The statistics  related to  liveness analysis  and DFA  generation are
+shown in Table~\ref{tab:exp-results}(a). We  observe that the analysis
+of  all  programs  except \verb@treejoin@  and  \verb@sudoku@  require
+reasonable time.   The bottleneck in  our analysis  is the NFA  to DFA
+conversion with  worst-case exponential behaviour. However,  since the
+analysis has  to be  done only  once, and its  results can be cached and
+re-used, the time spent in analysis may be considered acceptable.
+
+Table~\ref{tab:exp-results}(b) compares  garbage collection statistics
+for RGC and  LGC.  We report the number of  garbage collection events,
+average  number of  cells  reclaimed per  garbage collection,  average
+number of cells  touched per garbage collection and the  total time to
+perform all collections.   It is no surprise that the  number of cells
+reclaimed per garbage  collection is higher and the  number of garbage
+collections  lower  for  LGC.   The  cost of  LGC  is  higher  garbage
+collection time, which increases the  overall execution time even with
+reduced number of  collections.  However, the execution time  of LGC is
+still comparable for most benchmarks (slowdown within \warning{5X} of
+RGC in the worst case) and better  for 3 benchmarks (2X speedup in the
+best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
+benchmark \warning{in the no-fib suite} that allocates complete binary
+trees  of various  sizes that  are  never used  by the  program. As  a
+result,  the benchmark  highly favours  LGC.  The  benchmark has  been
+included  for  completeness,  and  we  do  not  consider  its  numbers
+representative of real programs.
 
@@
 number of  garbage collections  tend to be  very large,
-hence  we  show only  a  slice of the original graph  which  is
-representative  of  the  behavior for  that  particular
-benchmark (Column (b)).  In all the programs we can see
+hence  we  show only  a  slice of the original graph  with 
+representative    behavior  (Column (b)).  In all the programs we can see
 that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
-regularly dips below the RGC curve (\RGCLine\ line), the dip
-is  specially  visible  for the  benchmarks  where  LGC
-outperforms RGC substantially.  The graphs also include the
-curve    for   reachable    cells   (\ReachLine,    obtained
-approximately  by forcing  RGC to  run at  a very  high
-frequency)  and  the  live  cells  for a particular run
-(\UseLine, obtained by post processing the heap access
-at the end of the program).
+dips   below the RGC curve 
+(\RGCLine\ line) during garbage collection.  The graphs also include the
+curve    for   reachable    cells   (\ReachLine)  and live cells
+(\UseLine). These were  obtained
+ by forcing  RGC to  run at   very  high
+frequency. The curve for live cells were obtained by recording heap
+access times and post processing the data 
+at the end of the program.
 
@@
 \label{sec:relatedwork}
+
 Most previous work on enhancing  garbage collection with liveness were
 in   the  context   of   imperative  languages~\cite{Albert,   Hirzel,
-  khedker07heap}.    Due  to  the  presence  of  global  variables  and
-mutation, these methods are of  limited practical  utility.
+  khedker07heap}.    \warning{Due  to  the  presence  of  global  variables  and
+mutation, these methods are of  limited practical  utility.}
 
@@
 analysis and  cannot provide  significant improvement in  its absence.
-An interesting  approach suggested by Hofmann~\cite{HofmannJ03}  is to
+An interesting  approach suggested in~\cite{HofmannJ03}  is to
 annotate the heap usage of  first-order programs through linear types.
-The annotations are then used to transform the program to serve memory
-allocation requests by re-using allocated memory cells.  However, this
+The annotations are then used to serve memory
+requests through re-allocation.  However, this
 requires the user to write programs in a specific way.
@@
 \vskip -5mm
-\section{Conclusions and Future Work}
+\section{Future Work and Conclusions}
 \label{sec:conclusion}
@@
 %% this.
-We have  extended liveness-based garbage collection  to lazy languages
-and shown its benefit for a set of benchmark programs.  This was done
-by defining a liveness analysis of heap objects that we proved correct
-with  respect   to  a   non-standard  semantics   that  served   as  a
-specification of  liveness.  The result  of the  analysis is a  set of
-grammars, whose membership question was  proved to be undecidable. The
-grammars are  therefore approximated by  DFAs and used by  the garbage
-collector to improve collection.
-
-In addition to collecting evaluated  data, our collector also reclaims
-closures.   Because  of this  we  had  to  make modifications  to  the
-standard run-time representation of a closure to carry liveness of its
-free variables and  periodically and update it  during execution time.
-As  expected, our  garbage collector  could reclaim  more garbage  and
-reduce memory requirements of programs.  An additional benefit was the
-reduction in  execution times for  some programs  in spite of  using a
-more expensive collection scheme.
+We have  extended liveness-based  garbage collection  to lazy
+  languages  and shown  its benefit  for a  set of  standard benchmark
+  programs.  We  defined a liveness analysis  of programs manipulating
+  the  heap and  proved  it  correct with  respect  to a  non-standard
+  semantics that served as a specification of liveness.  The result of
+  the analysis  is a  set of grammars,  whose membership  question was
+  shown to be undecidable. The  grammars are therefore approximated by
+  DFAs and used by the garbage collector to improve collection.
+
+In addition to  evaluated data, our collector  also reclaims closures.
+For this, we  had to modify the standard run-time  representation of a
+closures  to carry  liveness  of its free  variables and  periodically
+update the liveness during execution  time.  As expected, our garbage
+collector  reclaims more  garbage and  reduces memory  requirement of
+programs.  An  additional benefit is  that in  spite of using  a more
+expensive collector, the execution times remains comparable in
+most cases, and even improves for some programs.
 

### 2015-11-30 paper/paper.tex
reviews on 3oth

@@
 collector,  the execution  time  of  a program  remains comparable  with
-the same program running with a reachabilty-based collector.
+the same program running with a reachability-based collector.
 \warning{On our benchmark programs, we obtain a decrease of up to 75\%
@@
 demands  on its  arguments.   This  summary is  used  to step  through
-function  calls  during analysis.   The  concrete demand  on  a
-function body is obtained through a conservative approximation similar
-to  0-CFA~\cite{Shivers:1988} that  combines  the demands  on all  the
-calls to the function.  The result of the analysis is an annotation of
+function calls  during analysis.   The concrete  demand on  a function
+body  is  obtained through  a  conservative  approximation similar  to
+0-CFA~\cite{Shivers:1988} that  combines the demands on  all the calls
+to  the function.   The result  of the  analysis is  an annotation  of
 certain program points with  deterministic finite-state automata (DFA)
-capturing the liveness of variables at these points.  Depending on the
-where  garbage collection  has been  triggered, the
-collector
-consults a  set of automata  to restrict reachability  during marking.
-This results in an increase  in the garbage reclaimed and consequently
-in fewer garbage collections.
+capturing the  liveness of  variables at  these points.   Depending on
+where garbage collection is triggered, the collector consults a set of
+automata to restrict reachability during  marking.  This results in an
+increase in  the garbage reclaimed  and consequently in  fewer garbage
+collections.
 
@@
 which it was created, it has  to carry the liveness information of its
-free variables. Moreover, as  execution progresses and possible future
+free variables. As execution progresses and possible future
 uses are eliminated,  we update the liveness information  in a closure
@@
 
-\warning{Experiments  with  a   single  generation  copying  collector
+Experiments  with  a   single  generation  copying  collector
   (Section~\ref{sec:experiments})  confirm  the  expected  performance
@@
   garbage reclaimed.   As a  consequence, there is  a reduction  in the
-  number of collections up to ???\% and a decrease in the
-  minimum memory requirement up to ???\%. 
+  number of collections from \warning{1.6X to 23X} and a decrease in the
+  minimum memory requirement from  \warning{1X to 1389X}. 
   As an added benefit, there is also a reduction in the overall
-  execution time for some programs.}
+  execution time in 5 out of 12 benchmark programs.
 
@@
   \rightsquigarrow$
-$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell \mapsto \langle s,
+$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell := \langle s,
     \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
@@
 
-\subsection{Semantics}
+\subsection{Semantics}\label{sec:semantics}
 %\comment{
@@
 The  initial  state  of  the  transition  system  described  above  is
-$([\;]_\rho,\,    (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
-(\print~\ans)):[\;]_{S}, 
-[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$,
-$[\;]_{\mbox \footnotesize {H}}$
-and $[\;]_{S}$ are    are the
-empty environment, heap and stack respectively. The  initial stack consists
-of  a single  continuation frame  in  which \ans\  is a  distinguished
-variable  that  would   eventually  be  updated  with   the  value  of
-(\mainpgm) and $\rho_\mathit{init}$ maps \ans\ to a location $\ell_{\ans}$.   In  addition, \print\ is  a function modeling  a printing
+\[([\;]_\rho,\,     (\rho_\mathit{init},     \,    \ell_{\ans}     ,\,
+(\print~\ans)):[\;]_{S},   [\;]_{\heap},\,  (\mainpgm))\]   in  which
+$[\;]_\rho$, $[\;]_{\mbox  \footnotesize {H}}$ and $[\;]_{S}$  are are
+the empty environment, heap and  stack respectively. The initial stack
+consists  of  a  single  continuation   frame  in  which  \ans\  is  a
+distinguished variable that would eventually be updated with the value
+of  (\mainpgm)  and  $\rho_\mathit{init}$  maps \ans\  to  a  location
+$\ell_{\ans}$.  In addition, \print\ is a function modeling a printing
 mechanism---a   standard   run-time   support  assumption   for   lazy
-languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The operator $:$
-pushes elements on top of the stack.
+languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The
+operator~$:$ pushes elements on top of the stack.
 
@@
 
-%}\comment
-%==============================================================
-\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-%==============================================================
-\input{fig-liveness}
-
 \section{Liveness}
@@
 Heap-allocated data  needs a richer model  than classical liveness---a
-model  which  talks  about  liveness of  references.   Using  $\acar$,
-$\acdr$  to  represent access  using  $\CAR$  and $\CDR$  fields,  the
-liveness of the structure reachable from a variable can be represented
-by a set of {\em access paths} i.e.  prefix-closed set of strings from
-$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (usually a 
-reference corresponding to a variable) and  a heap  \heap, semantically  an access  path
-$\alpha$ represents  a reference, denoted $\heap\llbracket\ell,  \alpha\rrbracket$, in
-the heap  that is  obtained by  starting with  $\ell$ and  chasing the
-\CAR\ or  \CDR\ fields in the heap as specified  by the  access path.  
-$\heap\llbracket\ell, \alpha\rrbracket$  is defined only  if the path followed  in the
-heap  is \emph{closure-free}  (does not  cross closures),  else it  is
-undefined.
+model which talks about liveness of references.  An {\em access path}
+is a  prefix-closed set of strings  over $\{\acar,\acdr\}^\ast$, where
+$\acar$,  $\acdr$ represent  access  using $\CAR$  and $\CDR$  fields.
+Given an initial location $\ell$ (usually a reference corresponding to
+a variable)  and a  heap \heap, semantically  an access  path $\alpha$
+represents     a      reference,     denoted     $\heap\llbracket\ell,
+\alpha\rrbracket$,  in the  heap  that is  obtained  by starting  with
+$\ell$ and chasing the \CAR\ or  \CDR\ fields in the heap as specified
+by  the  access   path.   $\heap\llbracket\ell,  \alpha\rrbracket$  is
+defined only if  the path followed in the  heap is \emph{closure-free}
+(does not cross closures), else it is undefined.
 
@@
 
+%==============================================================
+\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
+%==============================================================
+\input{fig-liveness}
+
 \subsection{Liveness Analysis for lazy languages}
@@
 
-%\comment{  
 Consider the  program in Figure~\ref{fig:mot-example2}.   As mentioned
@@
 from  its corresponding  stack variable  by subscripting  it with  the
-program   point  where   the   closure  was   created\footnote{Mutiple
+program   point  where   the   closure  was   created\footnote{Multiple
   occurrences  of the  same  variable in  an  application are  further
@@
 
-The  variable \pz\   returned  at  $\psi_2$ depends  on  \py\ 
+The  variable \pz\   returned  at  $\psi_3$ depends  on  \py\ 
   through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
   closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
-  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
+  denote  this chain  of dependences  as $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
   (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
   closures  in  the  heap  are   a  runtime  representation  of  these
-  dependences.  Since  $\pz$  is  evaluated   at  $\psi_2$  due  to  the
+  dependences.  Since  $\pz$  is  evaluated   at  $\psi_3$  due  to  the
   expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
@@
   $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
-  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
+  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_2\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
   \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
@@
   However,  notice that  if  a garbage  collection  takes place  while
-  evaluating $\pz$ at $\psi_2$, it can safely consider only the liveness
-  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
+  evaluating $\pz$ at $\psi_3$, it can safely consider only the liveness
+  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
   (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
@@
 i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
-programs) to  its WHNF driven  by the printing mechanism.   With minor
-variations,  the   proof  will  also  be   applicable  for  subsequent
-evaluations initiated by $\print$.
+programs)  to   its  WHNF  driven  by   the  printing  mechanism~(Section
+\ref{sec:semantics}).  With  minor variations,  the proof will  also be
+applicable for subsequent evaluations initiated by $\print$.
 %% We show  next that the  minefield execution of a  call-less program
@@
 $P'$,   a   call-less   program,   cannot   enter   the
-\bang\  state.  Therefor  $P$  also  cannot  enter  the
+\bang\  state.  Therefore   $P$  also  cannot  enter  the
 \bang\ state.
@@
 
-In general, the equations for \Lfonly\ are recursive as
-in the  case of  \Lfonly$_\mathit{f}$. As  mentioned in
-\cite{asati14lgc},   a   closed   form   solution   for
-$\Lfone{\mathit{f}}$ can  be derived by  observing that
-each of  the liveness rules  modifies a demand  only by
-prefixing  it with  symbols  in  the alphabet  $\lbrace
-\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore
-we can  assume that $\Lf{f}{i}{\sigma}$ has  the closed
-form:
+In general, the equations for \Lfonly\ are recursive as in the case of
+\Lfonly$_\mathit{f}$. A closed  form solution for $\Lfone{\mathit{f}}$
+can be derived by observing that each of the liveness rules modifies a
+demand  only by  prefixing it  with symbols  in the  alphabet $\lbrace
+\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore we  can assume
+that $\Lf{f}{i}{\sigma}$ has the closed form:
 \begin{eqnarray}
@@
 \rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
-and $\clazy\$ \rightarrow \epsilon$.
+and $\clazy\$ \rightarrow \$ $.
 \end{definition}
@@
 construct a grammar $G$ such that the machine will halt
-on    $w$    if    and   only    if    $\epsilon    \in
+on    $w$    if    and   only    if    $\$    \in
 \mathscr{L}(G)$. The grammar includes  the fixed set of
@@
   $w_l(\state,c)w_r$, then the  sentential form will be
-  $\overline{w}_l\nont{}{c}\,w_r$,                where
+  $\overline{w}_l\nont{}{c}\,w_r\$ $,                where
   $\overline{w}_l$ is  the same as $w_l$  but with each
@@
   are two productions
-  $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
-  \rightarrow c \nont{j}{\acdr}\bcdr$. 
+  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
+  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
 \end{enumerate}
@@
 example:  Assume that  the current  sentential form  is
-\bcar\bcdr\nont{i}{\acar}\acar\acar. Also  assume that  the
+$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
 TM  has  a   transition  $(S_i,\acar)  \rightarrow
@@
 in  the  derivation.    This  gives  the  configuration
-\bcar\bcdr\acdr\nont{i}{\acdr}\acdr\acar\acar.
+$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
 Simplification   with    the   production   $\bcdr\acdr
 \rightarrow               \epsilon$              yields
-\bcar\nont{i}{\acdr}\acdr\acar\acar,    which   exactly
+$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
 corresponds  to   the  changed  configuration   of  the
@@
 \rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
-\acdr\ on the right of the tape head. Finally the production $\clazy\$
-\rightarrow \epsilon$ is used  to derive $\epsilon$.This completes the
-reduction.
+\acdr\ on the right of the tape head.  This completes the
+reduction.\\ \mbox{}
 \end{proof}
@@
 then the function for the non-terminal joins the program fragments for
-the  RHS  with a  $\SIF$.   If  there  are  $n$ such  functions,  then
+the  RHS  with an  $\SIF$.   If  there  are  $n$ such  functions,  then
 \mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
@@
 of Lemma~\ref{lemma:grammar-undecidable} that is different
-from $\nont{\mathit{final}}{c}$.   Thus  the  empty  string  would  not  be
+from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
 derivable from the grammar.
@@
 The evaluation of  the top-level expression in a program  is driven by
-a printing function.  
+a printing function~(Section
+\ref{sec:semantics}).  
 %% Printing of  atomic values is simple  and does
@@
 conversion with  worst-case exponential behaviour. However,  since the
-analysis has  to be  done only  once, and its  results can be cached and
+analysis has  to be  done only  once and its  results can be cached and
 re-used, the time spent in analysis may be considered acceptable.
@@
 approach that is closest to the  method described in this paper is the
-liveness-based garbage collector implemented in~\cite{asati14lgc}.  We
+liveness-based garbage collector implemented in~\cite{karkare07liveness,asati14lgc}.  We
 have extend this to handle lazy evaluation and closures.
@@
 We have  extended liveness-based  garbage collection  to lazy
-  languages  and shown  its benefit  for a  set of  standard benchmark
+  languages  and shown  its benefit  for a  set of  benchmark
   programs.  We  defined a liveness analysis  of programs manipulating
@@
   s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
+    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
 {\sc let} \\ 

### 2015-11-30 paper/paper.tex
fixing reviewr 1 comments

@@
 %------------------------------------------------------------%
-\subsection{Motivating Example}
+\subsection{Motivating example}
 \label{sec:motiv}
@@
 Figure~\ref{fig:mot-example} shows an example program and the state of
-the heap at  the program point $\pi$, i.e. just  before the evaluation
+the heap at  the program point $\pi$ (we write $\pi\!:\!e$ to associate a label
+$\pi$  with the  program point just
+before the expression $e$), i.e. just  before the evaluation
 of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
@@
   \mainpgm})\  e_\mainpgm)$,  and  the program begins execution
-with the call to \mainpgm.  We write $\pi\!:\!e$ to associate the label
-$\pi$ (not  part of the syntax) with the  program point just
-before expression $e$.
+with the call to \mainpgm.  Note that program point labels are not  part of the syntax.
 
@@
 \end{tabular}
-\caption{The small-step semantics for the language. \label{fig:lang-semantics}}
+\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
 \end{center}
@@
 consists  of  a  single  continuation   frame  in  which  \ans\  is  a
-distinguished variable that would eventually be updated with the value
+distinguished variable that will eventually  be updated with the value
 of  (\mainpgm)  and  $\rho_\mathit{init}$  maps \ans\  to  a  location
@@
 
-\input{example-program-and-liveness-maps}
+\input{example-program}
 %%\label{fig:mot-example2}
@@
 
-\subsection{Liveness Analysis for lazy languages}
+\subsection{Liveness analysis for lazy languages}
 \label{sec:liveness-analysis}
 
-Consider the  program in Figure~\ref{fig:mot-example2}.   As mentioned
+Consider the  program in Figure~\ref{fig:mot-example2-a}.   As mentioned
 earlier, a lazy evaluation of the $\LET$ expression at $\pi_1$ creates
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,}
-  $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$, \\
+  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
-{\sc car-clo-bang} 
+{\sc car-bang} 
 \\
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
-\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\ $\heap(\rho(x))
+\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \ell,
-  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
+  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
+\cred{\renewcommand{\arraystretch}{1}\left\{
+\begin{array}{lr}
+  {\emptyset}&\mbox{if } \sigma = \emptyset \\       
+  {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
+\end{array}\right.
+}$         
+&
 {\sc car-clo}\\
@@
 \end{center}
-\caption{Minefield semantics for {\sc car-clo-bang} and
+\caption{Minefield semantics for {\sc car-bang} and
   {\sc car-clo} \label{fig:minefield-semantics-for-some}}
@@
 
-\subsection{Soundness Results}
+\subsection{Soundness results}
 We first need an auxiliary  result about minefield semantics. Consider
@@
 
-\subsection{Generating Equations for $\Lfonly_{\mathit f}$}
+\subsection{Generating equations for $\Lfonly_{\mathit f}$}
 Given a  function $\mathit{f}$, we now  describe how to
@@
 \Lfonly$_\mathit{f}$.        The       program       in
-Figure~\ref{fig:mot-example2}   serves  as   a  running
+Figure~\ref{fig:mot-example2-a}   serves  as   a  running
 example.  Starting with a  symbolic demand $\sigma$, we
@@
 
-For  the  running  example,   $\length$  has  calls  from  $\main$  at
+For  the  running  example (Figure~\ref{fig:mot-example2-a}),  $\length$  has  calls  from  $\main$  at
 $\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
@@
 \end{eqnarray*}
-As examples, the liveness of \Lanv{\pl}{} and \Lanv{\pa}{} in terms
-of  $\sigma_{\length}$ are:
+As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
+stack variables $\pl$ and $\pa$ have the following liveness in terms
+of  $\sigma_{\length}$:
 \begin{align*}
+  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
+  \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
+  \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
   \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
-  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
- \\
+  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
   \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
@@
 
-\section{The Garbage Collection Scheme}
+\section{The garbage collection scheme}
 \label{sec:GC-scheme}
@@
 
-\section{Experimental Evaluation}
+\section{Experimental evaluation}
 \label{sec:experiments}
@@
 
-\section{Related Work} 
+\section{Related work} 
 \label{sec:relatedwork}
@@
 \vskip -5mm
-\section{Future Work and Conclusions}
+\section{Future work and conclusions}
 \label{sec:conclusion}
@@
 \appendix
-%\section{Appendix: Not to be included in the paper.}
+%\section{Appendix: not to be included in the paper.}
 \begin{figure*}[h!]
-\noindent{\Large \bf A. Minefield Semantics}
+\noindent{\Large \bf A. Minefield semantics}
 \begin{center}\footnotesize

### 2015-11-30 paper/paper.tex
fixing reviewr 1 comments

@@
   described in this section declares the liveness of $\px_{\pi_1}$ to be
-  a union of the liveness arising  out of these dependence chains.  To
+  a union of the liveness arising from these dependence chains.  To
   be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
@@
   evaluating $\pz$ at $\psi_3$, it can safely consider only the liveness
-  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
+  arising from the dependence  chain\linebreak $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
   (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
@@
 \begin{quote}
-Let $x.\alpha$ be a forward access path consisting only
-of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
+Let $x.\alpha$ be a forward access path (consisting only
+of edges  \acar\ and \acdr\  but not \bcar,  \bcdr\ or
 \clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
@@
 $\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
-\clazy\ to  obtain forward paths.  While checking for  liveness during
-garbage  collection, a  forward access-path  is valid  only if  it can
+\clazy\ to  obtain forward access paths.  While checking for  liveness during
+garbage  collection, a  forward access path  is valid  only if  it can
 reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the

### 2015-11-30 paper/paper.tex
minor
latest - 30th - d-day

@@
 Functional  programs  make  extensive  use  of  dynamically  allocated
-memory.  The
-allocation  is either  explicit (i. e., using  constructors) or
-implicit (creating  closures).  Programs  written in  lazy
-functional languages put additional  demands on memory as they require
-closures to  be carried  from the  point of creation  to the  point of
-evaluation.
+memory.  The allocation is either explicit (i. e., using constructors)
+or  implicit   (creating  closures).   Programs  in   lazy  functional
+languages put additional  demands on memory, as  they require closures
+to be carried from the point of creation to the point of evaluation.
 
@@
 which it was created, it has  to carry the liveness information of its
-free variables. As execution progresses and possible future
-uses are eliminated,  we update the liveness information  in a closure
-with a  more precise version.  Because of these reasons,   the garbage
-collector  also   becomes  significantly   more  complicated   than  a
-liveness-based collector for a eager language.
+free variables. As  execution progresses and possible  future uses are
+eliminated, we  update the  liveness information in  a closure  with a
+more precise version.  Because of these reasons, the garbage collector
+also  becomes significantly  more  complicated  than a  liveness-based
+collector for a eager language.
 
@@
 (Section~\ref{sec:experiments})   confirm  the   expected  performance
-  benefits.   Liveness-based  collection  results in  an  increase  in
-  garbage reclaimed.   As a  consequence, there is  a reduction  in the
-  number of collections from \warning{1.6X to 23X} and a decrease in the
-  minimum memory requirement from  \warning{1X to 1389X}. 
-  As an added benefit, there is also a reduction in the overall
-  execution time in 5 out of 12 benchmark programs.
+benefits.  Liveness-based collection results in an increase in garbage
+reclaimed.  As  a consequence, there is  a reduction in the  number of
+collections from \warning{1.6X  to 23X} and a decrease  in the minimum
+memory requirement from  \warning{1X to 1389X}.  As  an added benefit,
+there is also a reduction in the overall execution time in 5 out of 12
+benchmark programs.
 
@@
 \label{sec:motiv}
-\input{fig_motiv}
-Figure~\ref{fig:mot-example} shows an example program and the state of
-the heap at  the program point $\pi$ (we write $\pi\!:\!e$ to associate a label
-$\pi$  with the  program point just
-before the expression $e$), i.e. just  before the evaluation
-of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
-node either  represents atomic  values ($\NIL$,  integers etc.),  or a
+\input{fig_motiv}   Figure~\ref{fig:mot-example}   shows  an   example
+program and the state of the heap at the program point $\pi$\footnote{We write
+$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
+before  the  expression  $e$. The label is not part of the syntax.},  i.e. just  before  the  evaluation  of
+$(\length\ \pz)$.  The heap is represented  by a graph in which a node
+either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
 \CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
 (represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
-  references} and represent values of  variables  or fields.  The figure shows
-the  lists \px\  and  \pz\ partially  evaluated due to  the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.
-
+  references} and represent values of variables or fields.  The figure
+shows the  lists \px\  and \pz\  partially evaluated  due to  the
+\SIF\ condition 
+(\NULLQ~(\CAR~\pz)).
 The edges  shown by thick  arrows are those  which are live  at $\pi$.
@@
 %% cells can be  reclaimed.
-Thus if a garbage collection  takes place at
-$\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b), a
-liveness-based collector (LGC) will preserve only the cells referenced
-by  $\pz$,  and the live cells  constituting the  closure
-referenced by $(\CDR~\pz)$.   In contrast, a reachability-based
-collector (RGC) will preserve all cells.
+Thus if a garbage collection takes  place at $\pi$ with the heap shown
+in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
+will preserve only  the cells referenced by $\pz$, and  the live cells
+constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
+reachability-based collector (RGC) will preserve all cells.
 
@@
 attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
-    asati14lgc, karkare06effectiveness} have involved either
-  imperative or  eager functional languages.
+  asati14lgc, karkare06effectiveness} have  involved either imperative
+or eager functional languages.
 
@@
 
-\warning{Section~\ref{sec:defs}   introduces   the   syntax   of   the
-  programming language  considered and gives a  small-step operational
-  semantics for it.   The liveness analysis for this  language and its
-  soundness   proof   is  presented   in   Section~\ref{sec:liveness}.
+Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
+language considered  and gives a small-step  operational semantics for
+it.  The liveness  analysis for this language and  its soundness proof
+is          presented          in          Section~\ref{sec:liveness}.
 Section~\ref{sec:computing} describes  the formulation of  liveness as
-  grammars. We  also give a  proof of undecidability of  such grammars
-  and    show    how   they    can    be    approximated   by    DFAs.
+grammars. We also give a proof  of undecidability of such grammars and
+show     how     they     can     be     approximated     by     DFAs.
 Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
@@
 garbage  collection  and   liveness  and  Section~\ref{sec:conclusion}
-  discusses possible extensions and concludes the paper.}
+discusses possible extensions and concludes the paper.
 \section{The target language---syntax and semantics}
@@
 to        be        in        Administrative        Normal        Form
-(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
+(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
 functions  are  variables.  While  this  restriction  does not  affect
 expressibility,  this form  has  the benefit  of  making explicit  the
-creation of  closures through  the $\LET$  construct.  \warning{ For
-  now, we assume  $\LET$s  in our
-language  are non-recursive;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
-\IN\,\, e$, $x$ should not occur in  $s$}. The  restriction of \LET\ to  a single
-definition  is for  ease  of  exposition---generalization to  multiple
-definitions does not add conceptual difficulties.  We further restrict
-each variable in a program to  be distinct, so that no scope shadowing
-occurs---this  simplifies  reasoning.
+creation of closures through the $\LET$ construct.  \warning{ We
+  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
+  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
+  occur in  $s$}. The restriction of  \LET\ to a single  definition is
+for ease  of exposition---generalization to multiple  definitions does
+not add conceptual difficulties.  We further restrict each variable in
+a program  to be  distinct, so that  no scope  shadowing occurs; this
+simplifies reasoning.
 
@@
   \mainpgm})\  e_\mainpgm)$,  and  the program begins execution
-with the call to \mainpgm.  Note that program point labels are not  part of the syntax.
+with the call to \mainpgm.
 
@@
  contexts, we  shall use the  term 'expression' and the  notation $e$
-  for both expressions and applications.}) 
+to stand   for both expressions and applications.}) 
 are given by transitions of the form $\rho, \stk, \heap, e \rightsquigarrow
-\rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
-frames.  Each  continuation frame  is of the  form $(\rho,  \ell, e)$,
+\rho',  \stk', \heap',  e'$.  Here  $\stk$ is  a stack  of continuation
+frames.  Each  continuation frame  is a triple $(\rho,  \ell, e)$,
 signifying that the  location $\ell$ has to be updated  with the value
-of the  currently evaluating  expression, and $e$  is to  be evaluated
+of the  currently evaluating  expression and $e$  is to  be evaluated
 next  in the  environment $\rho$.  
-%% Given a state $\rho, \stk, \heap,
-%% e$  of  the  transition system, we shall call the pair $(\rho, e)$ as
-%% the    \emph{current
-%%   evaluation  context}, and for a continuation frame $(\rho',  \ell, e')$ in
-%% \stk, we shall call the pair $(\rho', e')$ as a   
-%% \emph{suspended  evaluation  context}.}
-%% %% \warning{The
-%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
-%% system will be denoted $\mathcal{S}$.}
-The  initial  state  of  the  transition  system  described  above  is
+The  initial  state  of  the  transition  system  is:
 \[([\;]_\rho,\,     (\rho_\mathit{init},     \,    \ell_{\ans}     ,\,
@@
 distinguished variable that will eventually  be updated with the value
-of  (\mainpgm)  and  $\rho_\mathit{init}$  maps \ans\  to  a  location
+of  (\mainpgm),  and  $\rho_\mathit{init}$  maps \ans\  to  a  location
 $\ell_{\ans}$.  In addition, \print\ is a function modeling a printing
@@
 bound to  $x$. The  program points which  trigger evaluation  of these
-closure   are  the   condition  of   an  $\SIF$   ({\sc  IF-clo})   and
-\SRETURN\ ({\sc  return-clo}).  We  call such  points \emph{evaluation
+closures   are  an $\SIF$   condition ({\sc  IF-clo})   and
+ a \SRETURN\ ({\sc  return-clo}).  We  call such  points \emph{evaluation
   points   $(\epath)$}  and   label  them   with  $\psi$   instead  of
@@
 address  to be  updated  with the  evaluated value  is
-recorded  in  a  continuation  frame.    This  is  required  for  lazy
-evaluation, else $d_1$ may  be evaluated more than once
+recorded  in  a  continuation  frame.    This  is  required  for the
+evaluation  to be lazy, else $d_1$ may  be evaluated more than once
 due to  sharing~\cite{Jones87}.  Only after this  is the actual selection  done ({\sc
@@
 Heap-allocated data  needs a richer model  than classical liveness---a
-model which talks about liveness of references.  An {\em access path}
+model which talks about liveness of references.
+
+%% The model use the notion of  {\em access paths}, i.e.
+%% a  prefix-closed set of strings  over $\{\acar,\acdr\}^\ast$, in
+%% which 
+%% $\acar$,  $\acdr$ represent  access  using $\CAR$  and $\CDR$  fields.
+
+An {\em access path}
 is a  prefix-closed set of strings  over $\{\acar,\acdr\}^\ast$, where
-$\acar$,  $\acdr$ represent  access  using $\CAR$  and $\CDR$  fields.
+$\acar$,  $\acdr$ represent  access  using $\CAR$  and $\CDR$  fields respectively.
 Given an initial location $\ell$ (usually a reference corresponding to
@@
   z\mapsto\{\}]$.    In  this   notation,  $y   \mapsto  \{\epsilon\}$
-represents access using $y$ itself and $z \mapsto \{\}$ indicates that
+represents access using $y$ itself and $z \mapsto \{\}$ indicates 
 $z$ is dead.  In lazy  languages, liveness environments are associated
@@
 result.   To see  the need  for demands,  consider the expression
-$(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
+$(\LET~x\leftarrow  (\CDR~y)~\IN~   (\SRETURN~x))$.   Assume   that  the
 context of this expression places  the demand $\lbrace \epsilon, \acar
@@
 We use $\sigma$  to range over demands, $\alpha$ to  range over access
-paths  and $\Lv$  to range  over liveness  environments. $\Lv_y$  will
-denote  the liveness  of an  individual  variable $y$  in $\Lv$.   The
-notation $\sigma_1\sigma_2$ denotes  the set $\lbrace \alpha_1\alpha_2
-\mid \alpha_1  \in \sigma_1, \alpha_2 \in  \sigma_2\rbrace$.  Often we
-shall abuse  notation to juxtapose an  edge label and a  set of access
-paths; $\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
+paths and $\Lv$  to range over liveness environments.  The liveness of
+an  individual variable  $y$ in  \Lv\  is $\Lv(y)$,  but more  commonly
+written as  $\Lv_y$.  The notation $\sigma_1\sigma_2$  denotes the set
+$\lbrace  \alpha_1\alpha_2 \mid  \alpha_1 \in  \sigma_1, \alpha_2  \in
+\sigma_2\rbrace$.  Often we shall abuse  notation to juxtapose an edge
+label and  a set  of access  paths; $\acar\sigma$  is a  shorthand for
+$\lbrace\acar\rbrace\sigma$.
 
@@
 %==============================================================
-\input{fig-liveness}
 
@@
 are distinguished  from variables  introduced by  \LET s  and function
-arguments  (called \emph{stack  variables}  since they  reside in  the
-activation  stack). Notationally,  we distinguish  a closure  variable
+arguments  (called \emph{stack  variables},  since they  reside in  the
+activation  stack).   We notationally distinguish  a closure  variable
 from  its corresponding  stack variable  by subscripting  it with  the
-program   point  where   the   closure  was   created\footnote{Multiple
+label of the program   point  where   the   closure  was   created\footnote{Multiple
   occurrences  of the  same  variable in  an  application are  further
@@
   Since a  closure is evaluated  only at evaluation points,  a closure
-  variable is attributed  with the same liveness in the  region of the
-  program  from the  point of  creation  of the  closure to  reachable
-  evaluation points.  This  is also true of  stack variables, because,
-  as we shall see, stack  variables derive their liveness from closure
-  variables. Thus, there are two  major differences in our formulation
-  of   liveness   of   lazy   languages   with   liveness   of   eager
+  variable is attributed  with the same liveness in  the entire region
+  of  the  program from  the  point  of  creation  of the  closure  to
+  reachable evaluation points.  This is  also true of stack variables,
+  because, as we shall see, stack variables derive their liveness from
+  closure variables. Thus, there are two major differences between our
+  formulation of  liveness of  lazy languages  with liveness  of eager
   languages~\cite{asati14lgc}:   (i)  the   introduction  of   closure
-  variables in  the liveness calculation,  and (ii) a  single liveness
+  variables in the  liveness calculations, and (ii)  a single liveness
   value for each  variable that is applicable from  its creation point
   to evaluation points.
+
+\input{fig-liveness}
 
@@
 
-The  variable \pz\   returned  at  $\psi_3$ depends  on  \py\ 
-  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
-  closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
-  denote  this chain  of dependences  as $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
-  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
-  closures  in  the  heap  are   a  runtime  representation  of  these
-  dependences.  Since  $\pz$  is  evaluated   at  $\psi_3$  due  to  the
-  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
-  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
-  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
-  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
-  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_2\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
-  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
+The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
+expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
+$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
+denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
+\leftarrow            (+~\py~1),\,           \py            \leftarrow
+(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
+heap are a  runtime representation of these  dependences.  Since $\pz$
+is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
+demand made by the calling context(s)  of $f$ places a demand on $\pz$
+which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
+chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
+\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
+$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
+\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
 described in this section declares the liveness of $\px_{\pi_1}$ to be
-  a union of the liveness arising from these dependence chains.  To
-  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
-  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
-  However,  notice that  if  a garbage  collection  takes place  while
-  evaluating $\pz$ at $\psi_3$, it can safely consider only the liveness
-  arising from the dependence  chain\linebreak $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
+a union of  the liveness arising from these dependence  chains.  To be
+safe, a garbage collection during  evaluation of $\py$ at $\psi_1$ has
+to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
+notice that if a garbage collection takes place while evaluating $\pz$
+at $\psi_3$, it can safely consider only the liveness arising from the
+dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
 (+~\py~1),\, \py \leftarrow (\length~\x_{\pi_1})\rbrack$.  The garbage
 collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
-  generalization  of   this  observation   to  dynamically   select  a
-  evaluation point specific liveness in order to collect more garbage.
+generalization of this observation  to dynamically select a evaluation
+point specific liveness in order to collect more garbage.
 
@@
 Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar and  the the language generated  by this grammar is  the least
+grammar and  the language generated  by this grammar is  the least
 solution satisfying  the rule. We  prefer the least solution  since it

### 2015-11-30 paper/paper.tex
Fixed minor errors

@@
 the same program running with a reachability-based collector.
-\warning{On our benchmark programs, we obtain a decrease of up to 75\%
-  in the number of collections, and up to 50\% decrease in peak memory
-  requirements.}
 \end{abstract}
@@
   led  to the  evaluation  of $e$.
-The demand for the initial rule is  $\sigma_{all}$ and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it the expression after the step.
+The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.
 The  information in  continuation
@@
     liveness environment  \Lv\ for  the expression $e$.   We similarly
-    define the liveness environment for the expressions $e$ in each of
+    define the liveness environments for the expressions $e$ in each of
     the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
@@
 bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
-enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
-the transition is the similar to the earlier {\sc car-clo}
+enters the  \bang\ state ({\sc  car-bang}). Otherwise
+the transition is similar to the earlier {\sc car-clo}
 rule.

### 2015-12-01 paper/paper.tex
Almost final version

--- 2015-11-30 paper/paper.tex
@@
 reclaimed.  As  a consequence, there is  a reduction in the  number of
-collections from \warning{1.6X  to 23X} and a decrease  in the minimum
-memory requirement from  \warning{1X to 1389X}.  As  an added benefit,
+collections from 1.6X  to 23X and a decrease  in the minimum
+memory requirement from  1X to 1389X.  As  an added benefit,
 there is also a reduction in the overall execution time in 5 out of 12
@@
 expressibility,  this form  has  the benefit  of  making explicit  the
-creation of closures through the $\LET$ construct.  \warning{ We
+creation of closures through the $\LET$ construct.   We
   assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
   expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
-  occur in  $s$}. The restriction of  \LET\ to a single  definition is
+  occur in  $s$. The restriction of  \LET\ to a single  definition is
 for ease  of exposition---generalization to multiple  definitions does
@@
 $\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
-\{x.\sigma\}$ to  give the  overall liveness environment  for $(\LET~x
+\{x.\sigma\}$. Finally, the stack variables corresponding to the free
+variables of $s$ are updated and added 
+  to  give the  overall liveness environment  for $(\LET~x
 \leftarrow~s~\IN~e)$.
@@
 inserted  by  the  immediately  preceding garbage  collection  or  the
-garbage collection of an earlier minefield step.  \warning{However the
+garbage collection of an earlier minefield step.  However the
   demand $\sigma'$ on basis of  which an earlier collection would have
@@
   the $\rightsquigarrow$  along with the  current GC and show  that it
-  does not lead to a \bang.}
+  does not lead to a \bang.
 
@@
 Assume to  the contrary that  a program $P$  enters the
-\bang\  state.  \warning{We can  transform  $P$  to a  call-less
+\bang\  state.  We can  transform  $P$  to a  call-less
 program $P'$ such that  the minefield executions of $P$
 and $P'$  are identical  except for change  of variable
-names. }             However,                 by
+names.             However,                 by
 Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
@@
 
-\warning{When   there  are   multiple   occurrences  of   \bcar,
+When   there  are   multiple   occurrences  of   \bcar,
 \bcdr\  and \clazy,  $\hookrightarrow$ is  applied from
-right  to left.}   The reflexive  transitive closure  of
+right  to left.   The reflexive  transitive closure  of
 $\hookrightarrow$      will      be     denoted      as
@@
 check if the source node of the edge has a path to a final 
-state.  If yes, \warning{we mark the mark the source node as final. Finally, we
+state.  If it does, we mark  the source node as final. Finally, we
 remove all the edges labeled \clazy\  and convert the automaton into a
-deterministic    automaton.}    These steps   effectively   implement    the
+deterministic    automaton.    These steps   effectively   implement    the
 $\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
@@
 final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
-expectedly  says that for a  demand  \warning{$\epsilon$  on  \length},  the
-liveness of  its argument  is $\acdr^{*}$  (the spine  of the  list is
+expectedly  says that for a  demand  $\sigma_{\length}$,  the
+liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
 traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
@@
 reduced number of  collections.  However, the execution time  of LGC is
-still comparable for most benchmarks (slowdown within \warning{5X} of
-RGC in the worst case) and better  for 3 benchmarks (2X speedup in the
+still comparable for most benchmarks (slowdown within 5X of
+RGC in most cases) and better  for 3 benchmarks (2X speedup in the
 best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
-benchmark \warning{in the no-fib suite} that allocates complete binary
+benchmark that allocates complete binary
 trees  of various  sizes that  are  never used  by the  program. As  a
@@
 included  for  completeness,  and  we  do  not  consider  its  numbers
-representative of real programs.
+as being representative of real programs.
 
@@
 \label{sec:relatedwork}
-
-Most previous work on enhancing  garbage collection with liveness were
-in   the  context   of   imperative  languages~\cite{Albert,   Hirzel,
-  khedker07heap}.    \warning{Due  to  the  presence  of  global  variables  and
-mutation, these methods are of  limited practical  utility.}
-
-There  have been  several efforts  in improving  space utilization  of
-functional programs  by compile-time  analysis.  These  are: rewriting
+The important paper
+by~\cite{Hirzel} investigates the impact of liveness on the
+effectiveness of garbage collection. Their observation is that liveness
+can significantly impact garbage collection but only when it is
+interprocedural. As far as memory requirement is concerned, our paper
+demonstrates this observation.
+
+There have been several attempts to use liveness analysis to improve 
+garbage collection for imperative languages.~\cite{khedker07heap}
+presents a liveness analysis and uses the results for inserting 
+nullifying statements in Java programs. ~\cite{ran.shaham-sas03} is
+another work which constructs an automaton modelling heap accesses and
+checks temporal properties like liveness against the model.  
+However both the approaches are restricted to intraprocedural scope.
+
+In the space of functional languages, efforts towards improving
+space efficiency using compile-time  analysis are: rewriting
 methods                             such                            as
@@
 
-Simplifiers~\cite{ONeill}, described  as lightweight  daemons, improve
-the efficiency of the program in general and the garbage collectors in
-particular.  \warning{Most  of the  simplifications mentioned  in this
-  work are subsumed by a liveness  based collector.  We feel this work
-  is orthogonal  and can augment  our liveness based  collector.}  The
+Simplifiers~\cite{ONeill} are abstractly described as lightweight
+daemons that attach themselves to program data and, when activated, improve
+the efficiency of the program. Our liveness based garbage collection
+as an instance of a  simplifier which is tightly coupled with garbage collectors.  The
 approach that is closest to the  method described in this paper is the
-liveness-based garbage collector implemented in~\cite{karkare07liveness,asati14lgc}.  We
-have extend this to handle lazy evaluation and closures.
+liveness-based garbage collector implemented
+in~\cite{karkare07liveness,asati14lgc} and address eager languages. We
+extend this to handle lazy evaluation and closures.
 
@@
 %\section{Appendix: not to be included in the paper.}
+%%%%% prasanna minefield
 \begin{figure*}[h!]
-\noindent{\Large \bf A. Minefield semantics}
+\noindent{\Large \bf A. Minefield semantics(Not included in the paper)}
 \begin{center}\footnotesize
@@
 \end{figure*}
+%%%%%%%%%%%%%%%%%%%%%%
 

### 2015-12-01 paper/paper.tex
Almost final version

--- 2015-11-30 paper/paper.tex
@@
 \item       We   augment     the        standard       semantics
-  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
-  collection  just  before  each  $\rightsquigarrow$  transition.  The
-  simulated garbage collection starts from  the root-set and inserts a
-  special value  $\bot$ at each location  in the heap that  contains a
-  reference that is reachable but not live. Any attempt to dereference
-  such  locations during  the  transition will  result  in entering  a
+  (Figure~\ref{fig:lang-semantics}) to model  a dynamic liveness analysis 
+that records liveness values at each reference created during execution.
+Any attempt to dereference a location with a null-liveness  
+  during  a  transition will  result  in entering  a
   special state denoted \bang. We call the semantics after
@@
 Simplifiers~\cite{ONeill}, described  as lightweight  daemons, improve
-the efficiency of the program in general and the garbage collectors in
+the efficiency of the program in general and garbage collectors in
 particular.  \warning{Most  of the  simplifications mentioned  in this

### 2015-12-01 paper/paper.tex

--- 2015-12-01 paper/paper.tex
@@
 access times and post processing the data 
-at the end of the program.
+at the end of the program. Note that the size of an LGC cell is 1.16
+times the size of a RGC cell.
 
@@
 ~\cite{asati14lgc}. 
-A case for concern is the huge difference in the number of cells touched by LGC vs RGC for some programs.
-This is due to heap locations being  traversed multiple times during LGC. \warning{We have implemented heuristics to 
-overcome common cases and they show promising results}.
+A major source of inefficiency in LGC is multiple traversals of
+already copied heap cells. Since LGC does not mark the heap cells
+after the first visit, the same cells can be repeatedly visited with
+different liveness states. We have mitigated this problem by
+implementing a heuristic which tags \CONS\ cells with the set of
+states already visited.
 
@@
 
-\warning{Although we  provide an  implementation for  a liveness
-based garbage  collection scheme for a lazy  language, we
-do not  provide a formal  proof of its  correctness.  A
-formal proof of correctness would describe how closures
-are handled  during garbage collection and  liveness is
-propagated inside closures.
-
-  Another  interesting exercise  is to  narrow the  gap
-  between  the   actual  liveness  and   the  perceived
-  liveness  of our  garbage collector. Using
-  strictness  analysis  to  eagerly  evaluate  closures
-  might release more dead cells to be garbage
-  collected.  Since our  garbage collector  can handle
-  closures   (suspended  evaluations),   another  major
-  challenge  is to  extend it  to support  higher order
-  programs.
+The graphs show that even in lazy languages there are large amount of
+dead cells that can be collected early. In spite of collecting
+closures there still exists a gap between actual and perceived
+livenesses.
+Language features that could be introduced are higher-order functions
+and recursive lets. We believe that functions can be used to model
+recursive lets and this should be a simple extension to our liveness
+analysis.
+For higher-order we intend to use firstification techniques~\cite{Mitchell:2009}.
 
@@
   nearly-equivalent    states    and    using    better
-  representation    and    algorithms   for    automata
+  representation(for example BDDs~\cite{Bryant86})    and    algorithms   for    automata
   manipulation.  We   also  need  to   investigate  the
@@
   schemes,   such  as   incremental  and   generational
-  collection.  In summary, we  need to investigate ways
+  collection.  It might be interesting to use a mixed mode garbage collection
+  scheme which allows the costs of LGC to be amortized over several runs
+  of RGC. In summary, we  need to investigate ways
   to make liveness  based garbage collection attractive
-  for practical collectors.}
+  for practical collectors.
 

### 2015-12-01 paper/paper.tex
Third submission

@@
 \makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$, \\
-  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
+  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
@@
 \hline
+\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
+    S, \heap)}$, \\$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
+(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
+{\sc car-1-clo} 
+\\
+\hline
 \makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\ $\heap(\rho(x))
 \mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \ell,
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \rho(x),
   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
@@
 Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
-earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
+earlier  call to  $GC$  results  in $\rho(x)$  being
 bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
@@
 
-\subsection{Soundness results}
+\subsection{Soundness result}
 We first need an auxiliary  result about minefield semantics. Consider
@@
 \makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\
-  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
+  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
@@
 \mbox{ is } \langle s, \rho'\rangle$ }& \makecell[t]{$\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow $\\ $  \rho', (\rho, \ell,
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow $\\ $  \rho', (\rho, \rho(x),
   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$ }        &
@@
 \makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\
-  $\heap(\rho(x)) \mbox{ is } \bot$  or $\heap(\rho(y)) \mbox{ is }
+  $\rho(x) \mbox{ is } \bot$  or $\rho(y) \mbox{ is }
   \bot$} & $\rho, S,
   \heap, (+~x~y), \cred{\sigma} \rightsquigarrow \bang$   &
-{\sc prim-1-bang} 
+{\sc prim-bang} 
 \\
@@
     \sigma) = (\rho, S, \heap)}$, \\
-  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
+  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
@@
 \makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\
-  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
+  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\SRETURN~x), \cred{\sigma} \rightsquigarrow \bang$   &

### 2015-12-01 paper/paper.tex
Fourth submission

@@
-\documentclass[preprint, 9pt]{sigplanconf}
+\documentclass[9pt]{sigplanconf}
 \usepackage{etex}
@@
 %% finite-state automata (DFA) describing the liveness at that point.
-To show the effectiveness of our method, we
-implemented a copying collector that uses the results of the liveness
-analysis to preserve live 
-objects, both evaluated (i.e., in WHNF) and closures.  Our experiments
-confirm  that  for  programs  running with  a  liveness-based  garbage
-collector,   there  is   a   significant  decrease   in  peak   memory
+To  show the  effectiveness of  our method,  we implemented  a copying
+collector that uses  the results of the liveness  analysis to preserve
+live  objects,  both evaluated  (i.e.,  in  WHNF) and  closures.   Our
+experiments confirm  that for  programs running with  a liveness-based
+garbage  collector, there  is a  significant decrease  in peak  memory
 requirements.   In addition,  a  sizable reduction  in  the number  of
 collections  ensures that  in spite  of using  a more  complex garbage
-collector,  the execution  time  of  a program  remains comparable  with
-the same program running with a reachability-based collector.
+collector, the execution  times of programs running  with liveness and
+reachability-based collectors remain comparable.
 \end{abstract}
@@
 dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
-(+~\py~1),\, \py \leftarrow (\length~\x_{\pi_1})\rbrack$.  The garbage
+(+~\py~1),\, \py \leftarrow (\length~\px_{\pi_1})\rbrack$.  The garbage
 collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
@@
 \newcommand{\ReachLine}{black}
-\newcommand{\UseLine}{lightblue}
+\newcommand{\UseLine}{light-blue}
 
@@
 
-As demonstrated by the gap between the \RGCLine\ and the \UseLine, a large number of cells 
-which are unused by the program are still copied during RGC. LGC does a much better job of closing this gap 
-but still falls short of the precision achieved by LGC in case of eager languages
-~\cite{asati14lgc}. 
-A major source of inefficiency in LGC is multiple traversals of
-already copied heap cells. Since LGC does not mark the heap cells
-after the first visit, the same cells can be repeatedly visited with
-different liveness states. We have mitigated this problem by
-implementing a heuristic which tags \CONS\ cells with the set of
-states already visited.
+As  demonstrated by  the gap  between the  \RGCLine\ and  the \UseLine\
+lines, a  large number of  cells which are  unused by the  program are
+still copied  during RGC. LGC does  a much better job  of closing this
+gap but still falls short of the  precision achieved by LGC in case of
+eager languages ~\cite{asati14lgc}.  A major source of inefficiency in
+LGC is  multiple traversals  of already copied  heap cells.  Since LGC
+does not mark the heap cells after the first visit, the same cells can
+be  repeatedly  visited  with   different  liveness  states.  We  have
+mitigated  this  problem  by   implementing  a  heuristic  which  tags
+\CONS\ cells with the set of states already visited.
 
@@
 \label{sec:relatedwork}
-The important paper
-by~\cite{Hirzel} investigates the impact of liveness on the
-effectiveness of garbage collection. Their observation is that liveness
-can significantly impact garbage collection but only when it is
+The impact of  liveness on the effectiveness of  garbage collection is
+investigated in~\cite{Hirzel}.   The observation is that  liveness can
+significantly  impact   garbage  collection,  but  only   when  it  is
 interprocedural.  As far as memory requirement is concerned, our paper
@@
 presents a liveness analysis and uses the results for inserting 
-nullifying statements in Java programs. ~\cite{ran.shaham-sas03} is
-another work which constructs an automaton modelling heap accesses and
-checks temporal properties like liveness against the model.  
+nullifying statements in Java programs. In ~\cite{ran.shaham-sas03},
+an automaton modelling heap accesses is created, and
+temporal properties like liveness are checked  against the model.  
 However both the approaches are restricted to intraprocedural scope.
 
-In the space of functional languages, efforts towards improving
-space efficiency using compile-time  analysis are: rewriting
-methods                             such                            as
+In the space of functional  languages, there are: rewriting methods such as
 deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
 sharing  analysis   based  reallocation~\cite{jones89compile},  region
-based   analysis~\cite{tofte98region},   or   through   insertion   of
+based   analysis~\cite{tofte98region},   and    insertion   of
 compile-time                                                nullifying
@@
 analysis and  cannot provide  significant improvement in  its absence.
-An interesting  approach suggested in~\cite{HofmannJ03}  is to
-annotate the heap usage of  first-order programs through linear types.
-The annotations are then used to serve memory
-requests through re-allocation.  However, this
-requires the user to write programs in a specific way.
+An interesting approach suggested  in~\cite{HofmannJ03} is to annotate
+the  heap usage  of first-order  programs through  linear types.   The
+annotations   are  then   used  to   serve  memory   requests  through
+re-allocation.  However, this requires the user to write programs in a
+specific way.
 

### 2015-12-02 lazy-scheme/src/Simulator.cpp
Added code for getting extra memory stats

@@ -12,6 +12,8 @@
 #include "Demands.h"
 #include <ctime>
 #include <sys/stat.h>
+#include<algorithm>
+#include<stdint.h>
 
 using namespace std;
 using namespace Scheme::AST;
@@ -20,9 +22,65 @@ using namespace Scheme::output;
 
 double gctime=0;
 int gccount=0;
+vector<int> heap_map_size;
 extern demand_grammar gLivenessData;
 extern unordered_map<string, expr_demand_grammars*> localLivenessMap;
 extern unordered_map<string, unsigned int> func_heap_cell_reqd;
+
+
+/*
+ * For finding out the ratio of heap cell size
+ */
+struct live_heap_cell
+{
+	cell_type typecell;
+	union
+	{
+		struct
+		{
+			cons *car, *cdr;
+			bool can_delete_car;
+		}cell;
+		int intVal;
+		bool boolval;
+		std::string* stringVal;
+		struct
+		{
+			ExprNode* expr;
+			cons* arg1;
+			cons* arg2;
+			int16_t arg1_name;
+			int16_t arg2_name;
+		}closure;
+	}val;
+	void *forward;
+	bool inWHNF;
+};
+struct reach_heap_cell
+{
+	cell_type typecell;
+	union
+	{
+		struct
+		{
+			cons *car, *cdr;
+		}cell;
+		int intVal;
+		bool boolval;
+		std::string* stringVal;
+		struct
+		{
+			ExprNode* expr;
+			cons* arg1;
+			cons* arg2;
+		}closure;
+	}val;
+	void *forward;
+	bool inWHNF;
+};
+//End of declarations.
+
+
 
 demand_grammar filter_grammar(demand_grammar gLivenessData, vector<string> filter_criteria)
 {
@@ -276,7 +334,12 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize, int numkeys) //Thi
 	cout << "Heap total="<<hsize<<" Heap left="<<current_heap()<<" Heap used="<<(hsize - current_heap())<<endl;
 	detail_gc();
 	if (gc_type == gc_live)
+	{
 		cleanup(numkeys-1);
+		cout << "Maximum extra memory required during LGC " << *std::max_element(heap_map_size.begin(), heap_map_size.end()) << endl;
+		cout << "size of rgc heap cell = " << sizeof(reach_heap_cell) << endl;
[…]

### 2015-12-02 lazy-scheme/src/gc.cpp

@@ -51,6 +51,7 @@ GCStatus gc_status = gc_plain;
 extern int gccount;
 extern unordered_map<string, unsigned int> func_heap_cell_reqd;
 unordered_map<cons* , unordered_set<state_index>> heap_dfa_map;
+extern vector<int> heap_map_size;
 
 //ofstream gout("gc_addr.txt", ios::app);
 
@@ -1861,6 +1862,18 @@ void update_heap_ref_stack(ostream& out, int gc_type)
 
 }
 
+int calculate_heap_map_size()
+{
+	int total_elems = 0;
+	for (auto cell : heap_dfa_map)
+	{
+		total_elems += cell.second.size();
+	}
+	heap_map_size.push_back(total_elems);
+	DBG(cout << "Total number of visits to cons cells " << total_elems << endl);
+	return total_elems;
+}
+
 void liveness_gc()
 {
 	clock_t pstart = clock();
@@ -1888,7 +1901,8 @@ void liveness_gc()
 #endif
 	  DBG(pre << "Doing liveness based GC #" << gccount << " after " << num_of_allocations << " allocations"<<endl);
 	  swap_buffer();
-	  DBG(pre << "Number of unique cons cells " << heap_dfa_map.size()<<endl);
+
+
 	  heap_dfa_map.clear();
 
   for (deque<actRec>::iterator stackit = actRecStack.begin();stackit != actRecStack.end(); ++stackit)
@@ -1942,6 +1956,9 @@ void liveness_gc()
  // cout << "GC Time for " << gccount << " = " << gc_time << endl;
   gctime += gc_time;
   //cerr << "Number of cells copied = " << numcopied << endl;
+  calculate_heap_map_size();
+  DBG(cout << "Number of unique cons cells " << heap_dfa_map.size() << " at GC#"<<heap_map_size.size()<<endl);
+
   return;
 }
 

### 2016-01-28 paper/paper.tex
merging my changes
towards 10 page limit

@@
 
-We consider the rules {\sc car-clo} and {\sc car-1-clo} only.  
+We consider the rules {\sc car-clo} and {\sc car-bang} only.  
 The rest  of the rules involve similar reasoning. For the {\sc
@@
 
-For the  {\sc car-1-clo}  rule, observe that  there are
+For the  {\sc car-bang}  rule, observe that  there are
 two dereferences.  First $x$  is dereferenced to  get a
@@
 
-\subsubsection{From CFGs to GC-ready DFAs}
-\label{sec:NFA-approx}
- We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
- approximate a CFG by a {\em strongly regular\/} grammar.  For
- example, the grammar
- fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
- Mohri-Nederhof transformation is:
- \begin{eqnarray*}
-   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
-   \acdr\var{\Df{\length}{1}}
-   \mid \clazy\var{\Df{\length}{1}}\\
-   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
-   \mid \epsilon
- \end{eqnarray*}
-
-The strongly regular grammar is converted  into a set of NFAs, one for
-each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
-done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
-pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
-constructing  the $\epsilon$-closure  till a  fixed point  is reached,
-after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
-details of the  algorithm, its correctness and  termination proofs are
-given  in~\cite{karkare07liveness,asati14lgc}.   The
-resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
-\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
-check if the source node of the edge has a path to a final 
-state.  If it does, we mark  the source node as final. Finally, we
-remove all the edges labeled \clazy\  and convert the automaton into a
-deterministic    automaton.    These steps   effectively   implement    the
-$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
-\clazy\ to  obtain forward access paths.  While checking for  liveness during
-garbage  collection, a  forward access path  is valid  only if  it can
-reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
-NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
-final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
-expectedly  says that for a  demand  $\sigma_{\length}$,  the
-liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
-traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
-NFA     for     \var{\Lanv{\pa}{}}.      The     DFA     shown     in
-Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
-paths.   This  reflects  the  lazy  nature  of  our  language.   Since
-\length\ does not  evaluate the member elements of  the argument list,
-the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
-liveness based garbage collection is triggered beyond $\pi_9$.
-
-\section{The garbage collection scheme}
-\label{sec:GC-scheme}
-
-Our experimental setup  consists of an interpreter
-for  our  language,  a liveness  analyzer,  and  a
-single generation copying  collector.  The garbage
-collector can be configured to work on the
-basis  of  reachability  (RGC  mode)  or  use  
-liveness  DFAs  (LGC  mode).
-
-In        the       liveness        analysis       described        in
-Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
-variable is  derived from dependence  chains along all paths  from the
-program point  where the closure  was created to  reachable evaluation
-points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
-$\epath_2$ are two such evaluation points.  During garbage collection,
-we would like to use a more precise liveness, based on the actual paths
-taken  during  execution.   Therefore   we  create  separate  liveness
-automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
-addition to  automata for dependences  along paths to  both $\epath_1$
-and $\epath_2$.  The closure carries  the liveness environment for its
-free  variables (as  pointers  to automata,  one  for each  variable).
-Initially the liveness  environment is based on  the dependences along
-both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
-\SIF\ condition, the liveness environments are updated to one based on
-either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
-collections are based on more precise liveness information.
-
-Based on  the above considerations,  we restrict the  possible garbage
-collection points in a function body to the following:
-\begin{enumerate}
-\item We  maintain static estimates  of the memory required  to create
-  the closures for each function body.  On entering a function, if the
-  available memory is  less than this requirement for  the function, a
-  garbage collection is triggered.
-\item Since the evaluation of a  \SIF\ condition may trigger a 
-  collection, after evaluating the condition   the available memory
-  is  checked once  again against  a  revised estimate  of the  memory
-  (based on  value of the condition)  required to execute the  rest of
-  the program.  A garbage collection is triggered if enough memory is unavailable.
-\end{enumerate}
-
-%% While  exploring  the
-%% activation record of a  function, LGC uses the DFA
-%% at  the  latest  program point  traversed  in  the
-%% function's body.   Thus, if LGC is  invoked at the
-%% program point $\pi$, it would use the liveness DFA
-%% at $\pi$ itself for exploring  the root set of the
-%% current   activation   record.   For   any   other
-%% activation record in the stack, say for a function
-%% ${\mathit  f}$   calling  ${\mathit  g}$   in  the
-%% call-chain, the evaluation  point in ${\mathit f}$
-%% which resulted in the  call to ${\mathit g}$ would
-%% be used.
-
-We shall call a  unit allocatable memory as a {\em  cell}.  A cell can
-hold    a    basic    value    ($\mathsf{bas}$),    the    constructor
-\CONS\ $(\mathsf{cons~  arg_1~arg_2}$) or a closure.   The closure, in
-turn,    can    be    one   of    ($\mathsf{unop~arg}$),    ($\mathsf{
-  binop~arg_1~arg_2}$)  and  function application  ($\mathsf{f~arg})$.
-Here each $\mathsf{  arg_i}$ is a reference to another  heap cell.  In
-addition, the closure also carries  a pointer to a DFA (denoted
-$\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.
-
 %\subsection{A liveness-based garbage collection scheme}
@@
   \SetKwFunction{RCopy}{$rgcCopy$}%
-\begin{algorithm}[t]
+\begin{algorithm}[ht!]
   \Pro{\Lgc{}}
@@
 \end{algorithm}
-Algorithm~\ref{algo:lgc-a}   describes   the   garbage
-collection scheme.  Starting with the root set, each cell pointed by a
-live reference  (i.e. whose associated  DFA state is final)  is copied
-using $\mathsf{copy}$.   Copying a \CONS\ cells  just involves copying
-the  cell  itself   and  conditionally  copying  the   \CAR\  and  the
-\CDR\ fields after referring to the  next states of the DFA.  If
-the reference points to a closure, then, as noted earlier, the closure
-carries pointers  to the liveness  DFAs of its  arguments.  These
-are  used to recursively  initiate copying of  the arguments.
-Note that  the copying  strategy for  $\mathsf{(unop~arg_1~arg_2)}$ or
-$\mathsf{(f~arg_1)}$ are similar to $\mathsf{(binop~arg_1~arg_2)}$ and
-have not been shown.
-
-\begin{table*}[t!]
-\caption{Statistics for liveness analysis and garbage collection}
-\label{tab:exp-results}
-\centering
-\include{data-liveness-analysis}
-\include{autotable}
-\vskip -5mm
-\end{table*}
-
-%\subsection{Garbage collection for references on print stack}
-The evaluation of  the top-level expression in a program  is driven by
-a printing function~(Section
-\ref{sec:semantics}).  
-%% Printing of  atomic values is simple  and does
-%% not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
-%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
-%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
-%% trigger  more garbage  collections.  When  this happens.   we have  to
-%% consider any  references that  might be  on the  print stack  and copy
-%% them.
-We extend liveness-based  garbage  collection  to
-this function.
-
-\section{Experimental evaluation}
-\label{sec:experiments}
-  Our benchmark consists of programs taken from nofib~\cite{nofib} and
-  other repositories for  functional  programs~\cite{PLT-Scheme,
-    gc_bench, huffman-sicp}.   All our  benchmarks were executed  on a
-  machine having 8  core Intel(R) Core(TM) i7-4770 3.40GHz  CPU , 8192
-  KB L2 cache, 16 GB RAM running 64 bit Ubuntu 14.04.
-
-%% The  process  of  liveness-based  garbage  collection  involves  going
-%% through  each  activation  record  on the  stack  and  exploring  each
-%% variable in the current activation  record (root set). The liveness of
-%% each variable is  determined using the program point  and the variable
-%% name. The liveness  of all variables are stored as  DFA. These DFA are
-%% then encoded as a table. A variable  is live only if an entry is found
-%% in  this table.   All  cells that  are  live are  copied  to the  live
-%% semi-space. In case  of \CONS\ cells the \CAR\ and  \CDR\ pointers are
-%% chased  and if  they are  live  the copied  \CONS\ cell  will get  the
-%% updated addresses of its \CAR\ and \CDR\ fields.
-
-\subsection{Results}
-
-The statistics  related to  liveness analysis  and DFA  generation are
-shown in Table~\ref{tab:exp-results}(a). We  observe that the analysis
-of  all  programs  except \verb@treejoin@  and  \verb@sudoku@  require
-reasonable time.   The bottleneck in  our analysis  is the NFA  to DFA
-conversion with  worst-case exponential behaviour. However,  since the
-analysis has  to be  done only  once and its  results can be cached and
-re-used, the time spent in analysis may be considered acceptable.
-
-Table~\ref{tab:exp-results}(b) compares  garbage collection statistics
-for RGC and  LGC.  We report the number of  garbage collection events,
-average  number of  cells  reclaimed per  garbage collection,  average
-number of cells  touched per garbage collection and the  total time to
-perform all collections.   It is no surprise that the  number of cells
-reclaimed per garbage  collection is higher and the  number of garbage
-collections  lower  for  LGC.   The  cost of  LGC  is  higher  garbage
-collection time, which increases the  overall execution time even with
-reduced number of  collections.  However, the execution time  of LGC is
-still comparable for most benchmarks (slowdown within 5X of
-RGC in most cases) and better  for 3 benchmarks (2X speedup in the
-best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
-benchmark that allocates complete binary
-trees  of various  sizes that  are  never used  by the  program. As  a
-result,  the benchmark  highly favours  LGC.  The  benchmark has  been
-included  for  completeness,  and  we  do  not  consider  its  numbers
-as being representative of real programs.
+
+\subsubsection{From CFGs to GC-ready DFAs}
+\label{sec:NFA-approx}
+ We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
+ approximate a CFG by a {\em strongly regular\/} grammar.  For
+ example, the grammar
+ fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
+ Mohri-Nederhof transformation is:
+ \begin{eqnarray*}
+   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
+   \acdr\var{\Df{\length}{1}}
+   \mid \clazy\var{\Df{\length}{1}}\\
+   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
+   \mid \epsilon
+ \end{eqnarray*}
+
+The strongly regular grammar is converted  into a set of NFAs, one for
+each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
+done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
+pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
+constructing  the $\epsilon$-closure  till a  fixed point  is reached,
+after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
+details of the  algorithm, its correctness and  termination proofs are
+given  in~\cite{karkare07liveness,asati14lgc}.   The
+resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
+\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
+check if the source node of the edge has a path to a final 
+state.  If it does, we mark  the source node as final. Finally, we
+remove all the edges labeled \clazy\  and convert the automaton into a
+deterministic    automaton.    These steps   effectively   implement    the
+$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
+\clazy\ to  obtain forward access paths.  While checking for  liveness during
+garbage  collection, a  forward access path  is valid  only if  it can
+reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
+NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
+final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
+expectedly  says that for a  demand  $\sigma_{\length}$,  the
+liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
+traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
+NFA     for     \var{\Lanv{\pa}{}}.      The     DFA     shown     in
+Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
+paths.   This  reflects  the  lazy  nature  of  our  language.   Since
+\length\ does not  evaluate the member elements of  the argument list,
+the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
+liveness based garbage collection is triggered beyond $\pi_9$.
 
@@
 
-\begin{figure*}[t]
+\begin{figure*}[t!]
 \renewcommand{\arraystretch}{.1}
@@
 
+\section{The garbage collection scheme}
+\label{sec:GC-scheme}
+
+Our experimental setup  consists of an interpreter
+for  our  language,  a liveness  analyzer,  and  a
+single generation copying  collector.  The garbage
+collector can be configured to work on the
+basis  of  reachability  (RGC  mode)  or  use  
+liveness  DFAs  (LGC  mode).
+
+In        the       liveness        analysis       described        in
+Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
+variable is  derived from dependence  chains along all paths  from the
+program point  where the closure  was created to  reachable evaluation
+points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
+$\epath_2$ are two such evaluation points.  During garbage collection,
+we would like to use a more precise liveness, based on the actual paths
+taken  during  execution.   Therefore   we  create  separate  liveness
+automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
+addition to  automata for dependences  along paths to  both $\epath_1$
+and $\epath_2$.  The closure carries  the liveness environment for its
+free  variables (as  pointers  to automata,  one  for each  variable).
+Initially the liveness  environment is based on  the dependences along
+both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
+\SIF\ condition, the liveness environments are updated to one based on
+either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
+collections are based on more precise liveness information.
+
+Based on  the above considerations,  we restrict the  possible garbage
+collection points in a function body to the following:
+\begin{enumerate}
+\item We  maintain static estimates  of the memory required  to create
+  the closures for each function body.  On entering a function, if the
+  available memory is  less than this requirement for  the function, a
+  garbage collection is triggered.
+\item Since the evaluation of a  \SIF\ condition may trigger a 
+  collection, after evaluating the condition   the available memory
+  is  checked once  again against  a  revised estimate  of the  memory
+  (based on  value of the condition)  required to execute the  rest of
+  the program.  A garbage collection is triggered if enough memory is unavailable.
+\end{enumerate}
+
+%% While  exploring  the
+%% activation record of a  function, LGC uses the DFA
+%% at  the  latest  program point  traversed  in  the
+%% function's body.   Thus, if LGC is  invoked at the
+%% program point $\pi$, it would use the liveness DFA
+%% at $\pi$ itself for exploring  the root set of the
+%% current   activation   record.   For   any   other
+%% activation record in the stack, say for a function
+%% ${\mathit  f}$   calling  ${\mathit  g}$   in  the
+%% call-chain, the evaluation  point in ${\mathit f}$
+%% which resulted in the  call to ${\mathit g}$ would
+%% be used.
+
+We shall call a  unit allocatable memory as a {\em  cell}.  A cell can
+hold    a    basic    value    ($\mathsf{bas}$),    the    constructor
+\CONS\ $(\mathsf{cons~  arg_1~arg_2}$) or a closure.   The closure, in
+turn,    can    be    one   of    ($\mathsf{unop~arg}$),    ($\mathsf{
+  binop~arg_1~arg_2}$)  and  function application  ($\mathsf{f~arg})$.
+Here each $\mathsf{  arg_i}$ is a reference to another  heap cell.  In
+addition, the closure also carries  a pointer to a DFA (denoted
+$\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.
+Algorithm~\ref{algo:lgc-a}   describes   the   garbage
+collection scheme.  Starting with the root set, each cell pointed by a
+live reference  (i.e. whose associated  DFA state is final)  is copied
+using $\mathsf{copy}$.   Copying a \CONS\ cells  just involves copying
+the  cell  itself   and  conditionally  copying  the   \CAR\  and  the
+\CDR\ fields after referring to the  next states of the DFA.  If
+the reference points to a closure, then, as noted earlier, the closure
+carries pointers  to the liveness  DFAs of its  arguments.  These
+are  used to recursively  initiate copying of  the arguments.
+Note that  the copying  strategy for  $\mathsf{(unop~arg_1~arg_2)}$ or
+$\mathsf{(f~arg_1)}$ are similar to $\mathsf{(binop~arg_1~arg_2)}$ and
+have not been shown.
+
+\begin{table*}[t!]
+\caption{Statistics for liveness analysis and garbage collection}
+\label{tab:exp-results}
+\centering
+\include{data-liveness-analysis}
+\include{autotable}
+\vskip -5mm
+\end{table*}
+
+%\subsection{Garbage collection for references on print stack}
+The evaluation of  the top-level expression in a program  is driven by
+a printing function~(Section
+\ref{sec:semantics}).  
+%% Printing of  atomic values is simple  and does
+%% not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
+%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
+%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
+%% trigger  more garbage  collections.  When  this happens.   we have  to
+%% consider any  references that  might be  on the  print stack  and copy
+%% them.
+We extend liveness-based  garbage  collection  to
+this function.
+\section{Experimental evaluation}
+\label{sec:experiments}
+  Our benchmark consists of programs taken from nofib~\cite{nofib} and
+  other repositories for  functional  programs~\cite{PLT-Scheme,
+    gc_bench, huffman-sicp}.   All our  benchmarks were executed  on a
+  machine having 8  core Intel(R) Core(TM) i7-4770 3.40GHz  CPU , 8192
+  KB L2 cache, 16 GB RAM running 64 bit Ubuntu 14.04.
+
+%% The  process  of  liveness-based  garbage  collection  involves  going
+%% through  each  activation  record  on the  stack  and  exploring  each
+%% variable in the current activation  record (root set). The liveness of
+%% each variable is  determined using the program point  and the variable
+%% name. The liveness  of all variables are stored as  DFA. These DFA are
+%% then encoded as a table. A variable  is live only if an entry is found
+%% in  this table.   All  cells that  are  live are  copied  to the  live
+%% semi-space. In case  of \CONS\ cells the \CAR\ and  \CDR\ pointers are
+%% chased  and if  they are  live  the copied  \CONS\ cell  will get  the
+%% updated addresses of its \CAR\ and \CDR\ fields.
+
+\subsection{Results}
+
+The statistics  related to  liveness analysis  and DFA  generation are
+shown in Table~\ref{tab:exp-results}(a). We  observe that the analysis
+of  all  programs  except \verb@treejoin@  and  \verb@sudoku@  require
+reasonable time.   The bottleneck in  our analysis  is the NFA  to DFA
+conversion with  worst-case exponential behaviour. However,  since the
+analysis has  to be  done only  once and its  results can be cached and
+re-used, the time spent in analysis may be considered acceptable.
+
+Table~\ref{tab:exp-results}(b) compares  garbage collection statistics
+for RGC and  LGC.  We report the number of  garbage collection events,
+average  number of  cells  reclaimed per  garbage collection,  average
+number of cells  touched per garbage collection and the  total time to
+perform all collections.   It is no surprise that the  number of cells
+reclaimed per garbage  collection is higher and the  number of garbage
+collections  lower  for  LGC.   The  cost of  LGC  is  higher  garbage
+collection time, which increases the  overall execution time even with
+reduced number of  collections.  However, the execution time  of LGC is
+still comparable for most benchmarks (slowdown within 5X of
+RGC in most cases) and better  for 3 benchmarks (2X speedup in the
+best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
+benchmark that allocates complete binary
+trees  of various  sizes that  are  never used  by the  program. As  a
+result,  the benchmark  highly favours  LGC.  The  benchmark has  been
+included  for  completeness,  and  we  do  not  consider  its  numbers
+as being representative of real programs.
+
 Memory  usage graphs  for select  benchmarks are  shown
@@
 
+\clearpage
+\balance
 \bibliography{fun_hra}{}
 \bibliographystyle{abbrv}
-
 \clearpage
+
 \appendix
@@
 %%%%% prasanna minefield
-\begin{figure*}[h!]
+\begin{figure*}[b!]
 \noindent{\Large \bf A. Minefield semantics(Not included in the paper)}
@@
 \end{tabular}
-\caption{Minefield semantics. The differences with the small-step
+
+\caption {Minefield semantics. The differences with the small-step
   semantics have been highlighted by shading. \label{fig:minefield-semantics}}
 \end{center}
+
 \end{figure*}
-%%%%%%%%%%%%%%%%%%%%%%
-
-%% We shall now  present a proof of soundness of  the liveness analysis
-%% presented in Section~\ref{sec:liveness-analysis}.   While the proof is
-%% on the broad lines of \cite{asati14lgc}, there are several differences
-%% due to  the lazy  semantics of  the language  being 
-%% analyzed in this paper. Here is the idea behind the proof.
-%% \begin{enumerate}
-%% \item       We        enrich       the        standard       semantics
-%%   (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
-%%   collection  just  before  each  $\rightsquigarrow$  transition.  The
-%%   simulated garbage collection starts from  the root-set and inserts a
-%%   special value  $\bot$ at each location  in the heap that  contains a
-%%   reference that is reachable but not live. Any attempt to dereference
-%%   such  locations during  the  transition will  result  in entering  a
-%%   special state denoted \bang. We call the semantics after
-%%   augmentation  \emph{minefield semantics}.
-%% \item \label{inline2} Assuming that a  program enters the \bang\ state,
-%%   we  show  how to  construct,  through  inline expansion,  a  program
-%%   without function calls which has the same minefield behavior.
-%% \item  The final  part  of the  proof shows  that  no program  without
-%%   function calls can enter a \bang\ state. As a consequence no program
-%%   (with or without function calls) can enter the \bang\ state. 
-%% \end{enumerate}
-
-%% \subsection{Minefield semantics}
-
-%% To set up the minefield semantics, we follow these steps:
-%% \begin{enumerate}
-%% \item   We   enrich    the   abstract   machine   state
-%%   $\rho,\,S,\,\heap,\,e$   to  $\rho,\,S,\,\heap,\,e,\,\sigma$,
-%%   where  $\sigma$  is  the  demand  on  the  expression
-%%   $e$. We  call such a state  a \emph{minefield state}.
-%%   The demand $\sigma$ is  ``dynamic'' in that it arises
-%%   from the  actual sequence of function  calls that led
-%%   to  the evaluation  of  $e$.  The  0-CFA demand  used
-%%   during    liveness     analysis    is     a    static
-%%   over-approximation   of   $\sigma$   that   considers
-%%   \emph{all  possible calls}  that  could  lead to  the
-%%   evaluation of  $e$.  The  information in  the continuation frames on
-%%   the   stack  $S$ are  also
-%%   similarly augmented with their  demands.  Thus a stacked
-%%   entry now takes  the form $(\rho, x,  e, \sigma)$.  A
-%%   modification  of  the   small  step  semantics  which
-%%   carries the extra information  is shown in the bottom
-%%   table of Figure~\ref{fig:minefield-semantics}.
-%% \item $GC$  models a liveness based  garbage collection
-%%   $GC(\rho, S,  \heap, \sigma)$ returns $(\rho',  S', \heap')$,
-%%   where the changes are due to certain references being
-%%   replaced by  $\bot$.  This represents the  act of not
-%%   retaining   (effectively   garbage  collecting)   the
-%%   objects  pointed to  by  these  references during  an
-%%   actual  garbage  collection.  This  is  done  in  the
-%%   following way:
-%%   \begin{enumerate}
-%%   \item Given the current state $\rho, S,
-%%     \heap,   e,  \sigma$   and   the  demand   transformers
-%%     $\Lfonly$, we can  construct a liveness environment
-%%     \Lv for $e$.   If  $e$  is  an expression,  then  $  \Lv  =
-%%     \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
-%%     application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
-%%     We  can similarly  define the  liveness environment
-%%     for each  of the  suspended evaluation  contexts in
-%%     $S$, giving  a stack of liveness  environments that
-%%     we shall denote $\mathsf{SL}$.
-%%   \item \emph{Garbage collection  of objects pointed by
-%%     root   set:}  For   each   $x  \in   domain(\rho)$,
-%%     $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
-%%     \Lv$.    Similarly,    for   each    stack    entry
-%%     $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
-%%     corresponding      liveness     environment      in
-%%     $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
-%%     $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
-%%   \item  \emph{Garbage  collection  of objects  in  the
-%%     heap:} For each location  $\ell$, $\heap'(\ell) = \bot$
-%%     iff  there   is  no  $x$  in   either  the  current
-%%     environment  or  one  of the  stacked  environments
-%%     $\rho$, such  that for some  $\alpha$ corresponding
-%%     to a  closure-free path  in the heap  starting from
-%%     $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
-%%     \Lv$. Here \Lv\ is the liveness of the variables in
-%%     the current environment or the stacked environment.
-
-%% \warning{We have to tell  the reader about closure free
-%%   paths  earlier  on.   Also,  has  $\heap[x.\alpha]$  been
-%%   defined?}
-%%   \end{enumerate}
-%% Starting from the initial state $([\;]_\rho,([\;]_\rho,
-%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
-%% every transition  is preceded  by a  garbage collection
-%% using $GC$.  The details of the transition for the {\sc
-%%   car-clo}  rule   is  shown   in  the  top   table  of
-%% Figure~\ref{fig:minefield-semantics}.  If  this  or  an
-%% earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
-%% bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
-%% enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
-%% the transition is the same as the earlier {\sc car-clo}
-%% rule.
-%% \end{enumerate}
-
-%% \subsection{Inlining}
-
-%% Consider a call tree, each  of whose nodes represents a
-%% function  that  was  called (but  did  not  necessarily
-%% return because  of a \bang).   Assume that each  of the
-%% nodes of  the tree is  also annotated with  the program
-%% point where  the corresponding call was  invoked.  This
-%% tree  can  be  used  to  inline  function  calls  in  a
-%% hierarchical fashion.  The details of inlining a single
-%% function      call     $(\LET\,\,      x     \leftarrow
-%% (f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
-%% Let  $f$ be  defined  (after renaming  its formals  and
-%% locals  to  be  disjoint from  existing  variables)  by
-%% $(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
-%% replaced  by   a  sequence  of  $\LET$s   of  the  form
-%% $z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
-%% but with its  $(\SRETURN\,\,w)$ expressions replaced by
-%% $(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
-%% We prefer to use the  identity function $\ID$ as a form
-%% of no-op rather than  introducing the form $(\LET\,\, x
-%% \leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
-%% $\LET$-definition   is  a   simple  variable.    For  a
-%% call-less program,  the initial state of  the minefield
-%% semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
-%% \ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
-%% (e_{\mainpgm})$.
-
-%% \subsection{Soundness Results}
-
-%%   We  consider  a  garbage  collection  followed  by  a
-%%   transition as a step.  We show by induction that, for
-%%   programs  without function  calls, starting  from the
-%%   initial  state  that  has $e_{\main}$  as  the  first
-%%   expression to  be evaluated  instead of  $\main$, any
-%%   transition of $n$ steps occurs without a \bang.
-
-%% We  first  need  an auxiliary  result  about  minefield
-%% semantics. Consider  a trace of a  minefield execution.
-%% For every minefield state  $(\rho, S, \heap, e(s), \sigma)$
-%% that appears  on the LHS of  a $\rightsquigarrow$ step,
-%% the  demand   $\sigma$  on   the  expression   $e$  (or
-%% application $s$) is non-null.  This can be proved by an
-%% induction  on  the  number  of  steps  leading  to  the
-%% minefield  state.   The  base step  holds  because  the
-%% demand $\sigma_{all}$  on $\mainpgm$ is  non-null.  The
-%% inductive step  follows from  the observation  that for
-%% each  step of  the minefield  semantics, if  the demand
-%% $\sigma$ on  the LHS of  a minefield step  is non-null,
-%% the demand on  the RHS is a  transformation of $\sigma$
-%% (for example $\clazy\sigma$) which is also non-null.
-
-%% Let us clarify this observation:  Since the minefield semantics models
-%% lazy  evaluation,  parts  of  the  program may  not  be  evaluated  at
-%% all.  However,  if  an  expression  (or  application)  happens  to  be
-%% evaluated, it  will appear  on the  LHS of a  minefield step,  and the
-%% context-sensitive demand on it will be non-null.
-
-%% We  show  next  that   the  minefield  execution  of  a
-%% call-less program  cannot go \bang. Note  that, for the
-%% sake of  concreteness, by a single  minefield execution
-%% we   shall  mean   the  evaluation   of  $(main)$   (or
-%% $e_{\mainpgm}$,  for call-less  programs)  to its  WHNF
-%% driven  by   the  printing  mechanism   represented  by
-%% \print. With  minor variations, the proof  will also be
-%% applicable   for   subsequent  evaluations   fired   by
-%% $\print$.
-
-%% \begin{lemma}
-%% \label{lemma:call-less-cannot-go-bang-app}
-%% Consider the minefield execution of  a program without function calls.
-%% Such a program cannot enter the \bang\ state.
-%% \end{lemma}
-%% \begin{proof}
-%% Consider a state in the minefield execution of a program $(\rho, S, \heap,
-%% e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
-%% steps  leading to  this state  that  the next  step cannot  lead to  a
-%% \bang\ state.
-
-%% When $n$ is 0, the  state is $([\;]_\rho,\_, [\;]_{S} ,
-%% [\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.  Further,  we
-%% have   to  show   that   a  $GC(..)$   followed  by   a
-%% $\rightsquigarrow$ step cannot lead  to a \bang\ state.
-%% Now, because  our programs  are in  ANF, $e_{\mainpgm}$
-%% can  only  be a  $\LET$.  A  {\sc  let} step  does  not
-%% dereference anything and thus cannot result in a \bang.
-
-%% For the inductive step, we  shall show that none of the
-%% minefield steps that  involves dereferencing results in
-%% a   \bang.   These  are   the   steps   which  have   a
-%% $\heap(\rho(...))$ in the premise.  Recall that a minefield
-%% step   consists   of   a   GC  step   followed   by   a
-%% $\rightsquigarrow$.   The   $\rightsquigarrow$  can  go
-%% \bang\ because it dereferences a $\bot$ inserted by the
-%% immediately  preceding  GC  or  the GC  of  an  earlier
-%% minefield step.  \warning{However the  demand $\sigma'$
-%%   on basis of which an earlier GC would have inserted a
-%%   $\bot$,  would  have   included  the  current  demand
-%%   $\sigma$.}  Thus,  if the  earlier GC had  inserted a
-%% $\bot$, so would the current  GC.  Thus it is enough to
-%% consider the $\rightsquigarrow$  along with the current
-%% GC and show that it does not lead to a \bang.
-
-%% We consider the steps {\sc car-clo} and {\sc car-1-clo}
-%% only. The rest  of the steps are similar.  For the {\sc
-%%   car-clo}  step in  the state  $\rho, S,  \heap, (\CAR~x),
-%% \sigma$, we  know that $\sigma$ is  non-null. There for
-%% the  liveness  of  $x$  includes  $\epsilon$,  and  the
-%% dereferencing $\heap(\rho(x))$ will go without \bang.
-
-%% For the  {\sc car-1-clo}  step, observe that  there are
-%% two dereferences.  First $x$  is dereferenced to  get a
-%% cons  cell  and then  the  head  of  the cons  cell  is
-%% dereferenced  to  obtain  a  closure.   If  the  demand
-%% $\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
-%% of    $x$   will    include    both   $\epsilon$    and
-%% $\acar\epsilon$,  and  a  GC with  this  liveness  will
-%% neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
-%% first   component  of   the   cons   cell.  Thus   both
-%% dereferences    will    happen   normally    and    the
-%% $\rightsquigarrow$ step will not enter \bang.
-%% \end{proof}
-
-%% Now we are ready to prove the main soundness result.
-
-%% \begin{theorem}
-%% The  minefield  execution of  no  program  can enter  a
-%% \bang\ state.
-%% \end{theorem}
-
-%% \begin{proof}
-%% Assume to  the contrary that  a program $P$  enters the
-%% \bang\  state.  We can  transform  $P$  to a  call-less
-%% program $P'$ such that  the minefield executions of $P$
-%% and $P'$  are identical  except for change  of variable
-%% names.  \warning{Check  whether  the previous  line  is
-%%   correct.}                 However,                 by
-%% Lemma~\ref{lemma:call-less-cannot-go-bang-app} we know that
-%% $P'$,   a   call-less   program,   cannot   enter   the
-%% \bang\  state.  Therefor  $P$  also  cannot  enter  the
-%% \bang\ state.
-%% \end{proof}
 

### 2016-01-28 paper/paper.tex
fitting exercise

@@
       %probably doesn't scale anyway.
-%\newcommand{\mycomment}[1]{} 
+\newcommand{\mycomment}[1]{} 
 \definecolor{Myblue}{rgb}{.2,0,1}
@@
 
-\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
 \newcommand{\warning}[1]{{\color{Myred}{#1}}}
@@
 % 1st. author
+\cmt{{
 \authorinfo {Prasanna Kumar. K} 
@@
        {karkare@cse.iitk.ac.in}
-% 3rd. author
-%% \alignauthor Amey Karkare\\
-%%        \affaddr{IIT Kanpur}\\
-%%        \affaddr{Kanpur 208016, India}
-%%        \email{karkare@cse.iitk.ac.in}
-%}
-%}}
-%\authorinfo{Double Blind Review}{}{}
+}}
+\authorinfo{Double Blind Review}{}{}
 \maketitle
@@
  We assume that each program has a distinguished function
-\mainpgm,    defined as    $(\DEFINE\      ({\tt
+\mainpgm,    defined as\linebreak    $(\DEFINE\      ({\tt
   \mainpgm})\  e_\mainpgm)$,  and  the program begins execution
@@
 \begin{figure}[t]\footnotesize
-\renewcommand{\arraystretch}{1}
+\renewcommand{\arraystretch}{0.9}
 \begin{eqnarray*}
@@
 \begin{center}\footnotesize
-\renewcommand{\arraystretch}{1.5}
+\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{|c|c|c|}
@@
 \begin{center}\footnotesize
-\renewcommand{\arraystretch}{1.5}
+\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{|@{}c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|}
@@
 \include{data-liveness-analysis}
+\vskip -5mm
 \include{autotable}
@@
 other repositories for functional programs~\cite{PLT-Scheme,
-    gc_bench, huffman-sicp}.   All our  benchmarks were executed  on a
-  machine having 8  core Intel(R) Core(TM) i7-4770 3.40GHz  CPU , 8192
-  KB L2 cache, 16 GB RAM running 64 bit Ubuntu 14.04.
+  gc_bench, huffman-sicp}.  We ran the experiments on 8 core Intel\textsuperscript{\textregistered} Core\textsuperscript{TM}
+i7-4770 3.40GHz CPU having 8192KB L2 cache, 16GB RAM, running 64 bit
+Ubuntu 14.04.
 
@@
 \renewcommand{\arraystretch}{1.2}
-
 \begin{tabular}{|c|c|c|}

### 2016-01-28 paper/paper.tex
more savings;

@@
 %}\comment
-\vspace*{-6pt}
+\vspace*{-16pt}
 \end{figure*}
@@
 
+\begin{figure}[t]
+  \input{example-to-illustrate-liveness}  
+  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
+\end{figure}
 \input{fig-liveness}
@@
 dependences beginning at  a variable at an evaluation  point.  As an
-  example, consider the following program:
-
-\input{example-to-illustrate-liveness}
-
+example, consider the program in Figure~\ref{fig:lv-closure}.
 The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
@@
 proof.
-\begin{enumerate}
+\begin{myenumerate}
 \item       We   augment     the        standard       semantics
@@
   (with or without function calls) can enter the \bang\ state. 
-\end{enumerate}
+\end{myenumerate}
 
@@
 To set up the minefield semantics, we follow these steps:
-\begin{enumerate}
+\begin{myenumerate}
 
@@
   \sigma)$.
-\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
+\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
   collection that returns $(\rho', S', \heap')$, where the changes in
@@
   This is done in the following way:
-  \begin{enumerate} 
+  \begin{myenumerate} 
   \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
@@
 step~\ref{gc-1}.
-  \end{enumerate}
-\end{enumerate}
+  \end{myenumerate}
+\end{myenumerate}
 Starting from the initial state,
@@
 
-In  summary, the  equations  generated during  liveness
+In  summary, the  equations  generated during the
 analysis are:
-\begin{enumerate}
+\begin{myenumerate}
 \item   For  each   function  $\mathit{f}$,   equations
@@
   liveness for each garbage in  $e_{\mathit f}$.
-\end{enumerate}
+\end{myenumerate}
+
 \subsection{Solving liveness equations---the grammar
-interpretation}\label{sec:grammar-formulation}      The
-equations  above   can  now  be  re-interpreted   as  a
-context-free    grammar   (CFG)    on   the    alphabet
-$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
-Let \var{$X$}  denote the  non-terminal for  a variable
-$X$  occurring on  the LHS  of the  equations generated
-from  the  analysis.  We  can  think  of the  resulting
-productions as being  associated with several grammars,
-one for each non-terminal \var{\Lanv{x}{}} regarded as
-a  start  symbol.   As  an  example,  the  grammar  for
-\var{\Lanv{\xl}{}} and \var{\Lanv{\pa}{}}    comprise     the     following
-productions:
+interpretation}\label{sec:grammar-formulation}
+
+The  equations  above can  now  be  re-interpreted as  a  context-free
+grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
+\clazy\rbrace$.  Let \var{$X$} denote  the non-terminal for a variable
+$X$ occurring on the LHS of the equations generated from the analysis.
+We can  think of  the resulting productions  as being  associated with
+several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
+as a start symbol.  As  an example, the grammar for \var{\Lanv{\xl}{}}
+and \var{\Lanv{\pa}{}} comprise the following productions:
 \begin{eqnarray*}
   \var{\Lanv{\xl}{}}  &\rightarrow& 
-  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
-  & & \mid
+  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
   \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
@@
 However, recall that the decision problem that we are interested in
-during garbage collection is the following: 
-\begin{quote}
+during garbage collection is: 
+%%\begin{quote} 
+{\em
 Let $x.\alpha$ be a forward access path (consisting only
@@
 $\alpha \in \sigma$?
-\end{quote}
-
-We now model  the above problem as one  of deciding the
-membership of  a context-free grammar augmented  with a
+}
+%%\end{quote}
+We model  this problem as one  of deciding the
+membership of  a CFG augmented  with a
 fixed set of unrestricted productions.
@@
 
-\begin{enumerate}
+\begin{myenumerate}
 \item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
@@
   \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
-\end{enumerate}
+\end{myenumerate}
 The idea  behind the  productions is explained  with an
@@
 \acdr\ on the right of the tape head.  This completes the
-reduction.\\ \mbox{}
+reduction.\!
 \end{proof}
@@
 \end{proof}
-
-We circumvent  the problem  of undecidability  by over-
-approximating the CFG by non-deterministic finite state
-automata (NFA). The NFAs  are then simplified using the
-$\hookrightarrow$ rules.   Finally the  simplified NFAs
-are converted to DFAs.
 
@@
   \SetKwFunction{RCopy}{$rgcCopy$}%
-\begin{algorithm}[ht!]
+
+\begin{algorithm}[t!]
   \Pro{\Lgc{}}
@@
         }
-        \caption{Liveness-based
-          garbage collection.  \label{algo:lgc-a}}
+        \caption{Liveness-based garbage collection.  \label{algo:lgc-a}}
 \end{algorithm}
 
-\subsubsection{From CFGs to GC-ready DFAs}
-\label{sec:NFA-approx}
- We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
- approximate a CFG by a {\em strongly regular\/} grammar.  For
- example, the grammar
- fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
- Mohri-Nederhof transformation is:
+%%\subsubsection{From CFGs to GC-ready DFAs}
+%%\label{sec:NFA-approx}
+
+We circumvent the problem of undecidability by over approximating the
+CFG by non-deterministic finite state automata (NFA) using
+%% The NFAs  are then simplified using the
+%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
+%%are converted to DFAs.
+%%
+%% We use  the algorithm by 
+Mohri  and Nederhof~\cite{mohri00regular} method.
+%to approximate a CFG by a {\em strongly regular\/} grammar.  
+For example, the grammar fragment for the non-terminal
+$\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
  \begin{eqnarray*}
@@
 constructing  the $\epsilon$-closure  till a  fixed point  is reached,
-after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
-details of the  algorithm, its correctness and  termination proofs are
-given  in~\cite{karkare07liveness,asati14lgc}.   The
+after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  
+\cmt{ %% REMOVED FOR BLIND REVIEW
+  The details of the  algorithm, its correctness and  termination proofs are
+given  in~\cite{karkare07liveness,asati14lgc}. }
+The
 resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
@@
 traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
-NFA     for     \var{\Lanv{\pa}{}}.      The     DFA     shown     in
+NFA     for     \var{\Lanv{\pa}{}}.      The     DFA in
 Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
-paths.   This  reflects  the  lazy  nature  of  our  language.   Since
-\length\ does not  evaluate the member elements of  the argument list,
+paths, reflecting  the  lazy  nature  of  our  language.   Since
+\length\ does not  evaluate the elements of  the argument list,
 the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
-liveness based garbage collection is triggered beyond $\pi_9$.
+liveness-based GC triggers beyond $\pi_9$.
 
@@
 
+\vspace*{-2.5mm}
 \section{The garbage collection scheme}
@@
 collection points in a function body to the following:
-\begin{enumerate}
+\begin{myenumerate}
 \item We  maintain static estimates  of the memory required  to create
@@
   the program.  A garbage collection is triggered if enough memory is unavailable.
-\end{enumerate}
+\end{myenumerate}
 
@@
 
+\vspace*{-2mm}
 \section{Related work} 
@@
 daemons that attach themselves to program data and, when activated, improve
-the efficiency of the program. Our liveness based garbage collection
+the efficiency of the program. Our liveness-based garbage collection
 as an instance of a  simplifier which is tightly coupled with garbage collectors.  The
@@
   Orthogonally, we  plan to  improve the  efficiency of
-  the liveness based garbage collector using heuristics
+  the liveness-based garbage collector using heuristics
   such   as  limiting   the  depth   of  DFA,   merging
@@
   of RGC. In summary, we  need to investigate ways
-  to make liveness  based garbage collection attractive
+  to make liveness-based garbage collection attractive
   for practical collectors.

### 2016-01-31 paper/paper.tex
Added changes to the contributions

@@
       %probably doesn't scale anyway.
-\newcommand{\mycomment}[1]{} 
+
 \definecolor{Myblue}{rgb}{.2,0,1}
 \definecolor{Myred}{rgb}{1,.2,.2}
+\newcommand{\mycomment}[1]{{\color{Myblue}{#1}} }
+\newcommand{\added}[1]{{\color{Myblue}{#1}}}
+\newcommand{\todelete}[1]{{\color{Myred}{#1}}}
 
@@
 
-While this idea has been shown  to be effective for a first-order {\em
+\warning{While this idea has been shown  to be effective for a first-order {\em
   eager}  language~\cite{asati14lgc}, a  straightforward extension  of
@@
 objects may  include closures (runtime representations  of unevaluated
-expressions).   Firstly, since  data  is made  live  by evaluation  of
+expressions).}   Firstly, since  data  is made  live  by evaluation  of
 closures, and  in lazy languages the  place in the program  where this
@@
 contributions of this paper are
+\added{
+\begin{myitemize}
+\item  To the best of our knowledge this is the first effort to define a
+liveness based garbage collection for a lazy language. We provide a neat way 
+of handling liveness of functional values or unevaluated thunks (closures).
+\item 
+We  formulate  a  liveness  analysis for  a  lazy  first-order
+  functional language and  prove its correctness. The 
+  proof involves specifying liveness through a non-standard semantics  and then proving the analysis   correct with respect to the specification.   
+\item 
+In this paper, we formally prove that the membership question of liveness grammars is undecidable. We use standard  over-approximating techniques to circumvent this problem.
+\item 
+We  have implemented a garbage  collector that uses the  DFAs to
+  retain live  cells.
+Our  experiments reveal  interesting space-time
+  trade-offs  in  the  engineering  of  the  collector---for  example,
+  updating liveness  information carried in closures  during execution
+  results  in   more  garbage  being  collected.    Empirical  results
+  show the effectiveness of liveness-based garbage collection.
+\end{myitemize}
+}
+\todelete{
 \begin{myitemize}
@@
 \end{myitemize}
+}
 As far  as we know, this  is the first work  on liveness-based garbage
@@
 that records liveness values at each reference created during execution.
-Any attempt to dereference a location with a null-liveness  
+Any attempt to dereference a reference \mycomment{that has been declared not live by our analysis} %%with a null-liveness   
  during  a  transition  will  result  in entering  a
@@
 \caption{Minefield semantics for {\sc car-bang} and
-  {\sc car-clo} \label{fig:minefield-semantics-for-some}}
+  {\sc car-clo} \warning{Add a note mentioning that we handle the issue of requiring a specific demand by introducing a new symbol}\label{fig:minefield-semantics-for-some}}
 \end{figure*}

### 2016-02-02 paper/paper.tex
Updated introduction and contributions

@@
 
-\warning{While this idea has been shown  to be effective for a first-order {\em
+\todelete{While this idea has been shown  to be effective for a first-order {\em
   eager}  language~\cite{asati14lgc}, a  straightforward extension  of
@@
 objects may  include closures (runtime representations  of unevaluated
-expressions).}   Firstly, since  data  is made  live  by evaluation  of
+expressions).}
+We encounter significant challenges in defining a liveness based garbage 
+collector for lazy languages. 
+\added{
+Firstly, during garbage collection the heap contains both fully 
+evaluated data and unevaluated thunks/closures. 
+For liveness-based
+garbage collection to  be effective, we need to extend  it to closures
+apart from  evaluated data. Since closures can escape the scope in which 
+they were created, they need to carry liveness information of its free 
+variables. 
+Moreover, since  data  is made  live  by evaluation  of
+closures, and  in lazy languages the  place in the program  where this
+evaluation  takes  place  cannot be  statically  determined,  laziness
+complicates  liveness analysis  itself.  
+Also, as  execution progresses and possible  future uses are
+eliminated, we  update the  liveness information in  a closure  with a
+more precise version.  Because of these reasons, the garbage collector
+also  becomes significantly  more  complicated  than a  liveness-based
+collector for a eager language.
+}
+
+\todelete{
+Firstly, since  data  is made  live  by evaluation  of
 closures, and  in lazy languages the  place in the program  where this
@@
 collector for a eager language.
-
+}
 Experiments    with   a    single    generation   copying    collector
@@
 \begin{myitemize}
-\item  To the best of our knowledge this is the first effort to define a
-liveness based garbage collection for a lazy language. We provide a neat way 
-of handling liveness of functional values or unevaluated thunks (closures).
+\item  We propose a liveness based garbage collection scheme for a lazy first order 
+functional language and prove its correctness. 
+To the best of our knowledge this is the first effort to define a
+liveness based garbage collection for a lazy language. 
+Our main contribution is the way we handle liveness of functional values i.e closures which 
+distinguishes our work from earlier work in this area.
+
 \item
-We  formulate  a  liveness  analysis for  a  lazy  first-order
+  We  formulate  a  liveness  analysis for  the  lazy  first-order
   functional language and  prove its correctness. The 
-  proof involves specifying liveness through a non-standard semantics  and then proving the analysis   correct with respect to the specification.   
+  proof involves specifying liveness through a non-standard semantics  and then proving the analysis  
+  correct with respect to the specification.
+
 \item 
-In this paper, we formally prove that the membership question of liveness grammars is undecidable. We use standard  over-approximating techniques to circumvent this problem.
+  The analysis results  in a set  of context-free  grammars along
+  with  a fixed  set  of non-context-free  productions.  The  decision
+  whether   to  copy  a  cell   during  garbage  collection
+  translates  to a  membership problem  for such  grammars.  In  this
+  paper, we  have shown the  undecidability of this problem. We 
+  circumvent the  problem by over-approximating the  grammars
+  with DFAs.
 \item 
-We  have implemented a garbage  collector that uses the  DFAs to
+We  have implemented a garbage  collector that uses the  result of liveness analysis to
   retain live  cells.
@@
 \end{myitemize}
-}
+
 As far  as we know, this  is the first work  on liveness-based garbage
@@
 or eager functional languages.
-
+}
 \subsection{Organization of the paper}

### 2016-02-03 paper/paper.tex
rewrite

@@
 
-\todelete{While this idea has been shown  to be effective for a first-order {\em
-  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
-the technique is not possible for lazy languages, where heap-allocated
-objects may  include closures (runtime representations  of unevaluated
-expressions).}
-We encounter significant challenges in defining a liveness based garbage 
-collector for lazy languages. 
-\added{
-Firstly, during garbage collection the heap contains both fully 
-evaluated data and unevaluated thunks/closures. 
-For liveness-based
-garbage collection to  be effective, we need to extend  it to closures
-apart from  evaluated data. Since closures can escape the scope in which 
-they were created, they need to carry liveness information of its free 
-variables. 
-Moreover, since  data  is made  live  by evaluation  of
-closures, and  in lazy languages the  place in the program  where this
-evaluation  takes  place  cannot be  statically  determined,  laziness
-complicates  liveness analysis  itself.  
-Also, as  execution progresses and possible  future uses are
-eliminated, we  update the  liveness information in  a closure  with a
-more precise version.  Because of these reasons, the garbage collector
-also  becomes significantly  more  complicated  than a  liveness-based
-collector for a eager language.
-}
-
-\todelete{
-Firstly, since  data  is made  live  by evaluation  of
-closures, and  in lazy languages the  place in the program  where this
-evaluation  takes  place  cannot be  statically  determined,  laziness
-complicates  liveness analysis  itself.  Moreover,  for liveness-based
-garbage collection to  be effective, we need to extend  it to closures
-apart from  evaluated data.  Since a  closure can escape the  scope in
-which it was created, it has  to carry the liveness information of its
-free variables. As  execution progresses and possible  future uses are
-eliminated, we  update the  liveness information in  a closure  with a
-more precise version.  Because of these reasons, the garbage collector
-also  becomes significantly  more  complicated  than a  liveness-based
-collector for a eager language.
-}
+\todelete{While  this  idea has  been  shown  to  be effective  for  a
+  first-order     {\em     eager}    language~\cite{asati14lgc},     a
+  straightforward extension of the technique  is not possible for lazy
+  languages,  where   heap-allocated  objects  may   include  closures
+  (runtime   representations    of   unevaluated    expressions).    }
+\added{Designing a liveness based garbage collector for lazy languages
+  poses significant challenges.}  Firstly, since  data is made live by
+evaluation of closures, and in lazy languages the place in the program
+where  this evaluation  takes place  cannot be  statically determined,
+laziness   complicates  liveness   analysis  itself.    Moreover,  for
+liveness-based garbage collection  to be effective, we  need to extend
+it to closures apart from evaluated  data.  Since a closure can escape
+the  scope in  which it  was  created, it  has to  carry the  liveness
+information  of  its  free  variables.  As  execution  progresses  and
+possible  future   uses  are   eliminated,  we  update   the  liveness
+information  in a  closure with  a more  precise version.   Because of
+these reasons,  the garbage collector also  becomes significantly more
+complicated than a liveness-based collector for a eager language.
+
 Experiments    with   a    single    generation   copying    collector

### 2016-02-03 paper/paper.tex
rewrite

@@
 \begin{myitemize}
-\item  We propose a liveness based garbage collection scheme for a lazy first order 
-functional language and prove its correctness. 
-To the best of our knowledge this is the first effort to define a
-liveness based garbage collection for a lazy language. 
-Our main contribution is the way we handle liveness of functional values i.e closures which 
-distinguishes our work from earlier work in this area.
-
-\item
-  We  formulate  a  liveness  analysis for  the  lazy  first-order
-  functional language and  prove its correctness. The 
-  proof involves specifying liveness through a non-standard semantics  and then proving the analysis  
-  correct with respect to the specification.
-
-\item 
-  The analysis results  in a set  of context-free  grammars along
+\item We propose a liveness based garbage collection scheme for a lazy
+  first order functional  language and prove its  correctness.  To the
+  best  of our  knowledge, this  is the  first work  on liveness-based
+  garbage    collection   for    lazy    languages.    All    previous
+  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
+    asati14lgc,    karkare06effectiveness}   have    involved   either
+  imperative  or  eager  functional  languages.   The  novely  of  our
+  approach is in  handling of  liveness of  functional values  (i.e.,
+  closures), an issue that arises due to laziness.
+
+\item  We  formulate a  liveness  analysis  for the  lazy  first-order
+  functional language  and prove  its correctness. The  proof involves
+  specifying  liveness   for  the  language  through   a  non-standard
+  semantics and then proving the  analysis correct with respect to the
+  specification.
+
+\item The  analysis results  in a set  of context-free  grammars along
   with  a fixed  set  of non-context-free  productions.  The  decision
-  whether   to  copy  a  cell   during  garbage  collection
-  translates  to a  membership problem  for such  grammars.  In  this
-  paper, we  have shown the  undecidability of this problem. We 
-  circumvent the  problem by over-approximating the  grammars
-  with DFAs.
-\item 
-We  have implemented a garbage  collector that uses the  result of liveness analysis to
-  retain live  cells.
-Our  experiments reveal  interesting space-time
-  trade-offs  in  the  engineering  of  the  collector---for  example,
-  updating liveness  information carried in closures  during execution
-  results  in   more  garbage  being  collected.    Empirical  results
-  show the effectiveness of liveness-based garbage collection.
+  whether to  copy a  cell during garbage  collection translates  to a
+  membership problem for such grammars.   In this paper, we have shown
+  the undecidability  of this  problem. We  circumvent the  problem by
+  over-approximating the grammars with DFAs.
+
+\item We have implemented a garbage  collector that uses the result of
+  liveness  analysis to  retain  live cells.   Our experiments  reveal
+  interesting  space-time   trade-offs  in  the  engineering   of  the
+  collector---for  example, updating  liveness information  carried in
+  closures during  execution results in more  garbage being collected.
+  Empirical results  show the effectiveness of  liveness-based garbage
+  collection.
 \end{myitemize}

### 2016-02-03 paper/paper.tex
compressed enum and items

@@
 
+% to save space in lists
+\usepackage{enumitem}
+\setlist{nolistsep}
+
 \input{fh_defs}
@@
 \newcommand{\added}[1]{{\color{Myblue}{#1}}}
-\newcommand{\todelete}[1]{{\color{Myred}{#1}}}
+\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
 
@@
 \added{
-\begin{myitemize}
+\begin{itemize}
 \item We propose a liveness based garbage collection scheme for a lazy
@@
   collection.
-\end{myitemize}
+\end{itemize}
 }
 \todelete{
-\begin{myitemize}
+\begin{itemize}
 \item  We  formulate  a  liveness  analysis for  a  lazy  first-order
@@
   show the effectiveness of liveness-based garbage collection.
-\end{myitemize}
+\end{itemize}
 
@@
 As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
-reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$ corresponds to  a closure-free path
-in \heap\   starting from $\rho(x)$.  If this  path is intercepted  by a
-closure,  say  $(\CAR~{x_{\pi}})$, then  the  liveness  of further  heap
-contents starting from $\mathit{x_{\pi}}$ is given by $\Lv_\mathit{x_{\pi}}$.
+reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$
+corresponds to a closure-free path  in \heap\ starting from $\rho(x)$.
+If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
+the liveness of further heap contents starting from $\mathit{x_{\pi}}$
+is given by $\Lv_\mathit{x_{\pi}}$.
 
@@
 proof.
-\begin{myenumerate}
+\begin{enumerate}
 \item       We   augment     the        standard       semantics
@@
   (with or without function calls) can enter the \bang\ state. 
-\end{myenumerate}
+\end{enumerate}
 
@@
 
-To set up the minefield semantics, we follow these steps:
-\begin{myenumerate}
+\noindent To set up the minefield semantics, we follow these steps:
+\begin{enumerate}
 
@@
   This is done in the following way:
-  \begin{myenumerate} 
+  \begin{enumerate} 
   \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
@@
 step~\ref{gc-1}.
-  \end{myenumerate}
-\end{myenumerate}
+  \end{enumerate}
+\end{enumerate}
 Starting from the initial state,
@@
 analysis are:
-\begin{myenumerate}
+\begin{enumerate}
 \item   For  each   function  $\mathit{f}$,   equations
@@
   liveness for each garbage in  $e_{\mathit f}$.
-\end{myenumerate}
+\end{enumerate}
 
@@
 
-\begin{myenumerate}
+\begin{enumerate}
 \item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
@@
   \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
-\end{myenumerate}
+\end{enumerate}
 The idea  behind the  productions is explained  with an
@@
 collection points in a function body to the following:
-\begin{myenumerate}
+\begin{enumerate}
 \item We  maintain static estimates  of the memory required  to create
@@
   the program.  A garbage collection is triggered if enough memory is unavailable.
-\end{myenumerate}
+\end{enumerate}
 

### 2016-02-03 paper/paper.tex
garbage collection --> GC

@@
 In this paper,  we propose the use of liveness  analysis of heap cells
-for garbage collection in a lazy first-order functional language.  The
+for garbage collection (GC) in a lazy first-order functional language.  The
 central notion in our analysis  is a generalization of liveness called
@@
 capturing the  liveness of  variables at  these points.   Depending on
-where garbage collection is triggered, the collector consults a set of
+where GC is triggered, the collector consults a set of
 automata to restrict reachability during  marking.  This results in an
-increase in  the garbage reclaimed  and consequently in  fewer garbage
+increase in  the garbage reclaimed  and consequently in  fewer 
 collections.
@@
 laziness   complicates  liveness   analysis  itself.    Moreover,  for
-liveness-based garbage collection  to be effective, we  need to extend
+liveness-based GC  to be effective, we  need to extend
 it to closures apart from evaluated  data.  Since a closure can escape
@@
 %% cells can be  reclaimed.
-Thus if a garbage collection takes  place at $\pi$ with the heap shown
+Thus if a GC takes  place at $\pi$ with the heap shown
 in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
@@
 \begin{itemize}
-\item We propose a liveness based garbage collection scheme for a lazy
+\item We propose a liveness based GC scheme for a lazy
   first order functional  language and prove its  correctness.  To the
   best  of our  knowledge, this  is the  first work  on liveness-based
-  garbage    collection   for    lazy    languages.    All    previous
+  GC   for    lazy    languages.    All    previous
   attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
@@
   with  a fixed  set  of non-context-free  productions.  The  decision
-  whether to  copy a  cell during garbage  collection translates  to a
+  whether to  copy a  cell during GC translates  to a
   membership problem for such grammars.   In this paper, we have shown
@@
   closures during  execution results in more  garbage being collected.
-  Empirical results  show the effectiveness of  liveness-based garbage
-  collection.
+  Empirical results  show the effectiveness of  liveness-based GC.
 \end{itemize}
@@
   with  a fixed  set  of non-context-free  productions.  The  decision
-  whether   to  copy  a  cell   during  garbage  collection
-  translates  to a  membership problem  for such  grammars.  In  this
+  whether   to  copy  a  cell   during  GC  translates  to a  membership problem  for such  grammars.  In  this
   paper, we  have shown the  undecidability of this problem and 
@@
   results  in   more  garbage  being  collected.    Empirical  results
-  show the effectiveness of liveness-based garbage collection.
+  show the effectiveness of liveness-based GC.
 \end{itemize}
 
-As far  as we know, this  is the first work  on liveness-based garbage
-collection      for      lazy       languages.       All      previous
+As far  as we know, this  is the first work  on liveness-based GC      for      lazy       languages.       All      previous
 attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
@@
 Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
-collector,  in  particular  the  use  of  liveness  DFAs  for  garbage
-collection.     We     report    our    experimental     results    in
+collector,  in  particular  the  use  of  liveness  DFAs  for  GC.     We     report    our    experimental     results    in
 Section~\ref{sec:experiments}    along    with   some    observations.
 Section~\ref{sec:relatedwork}  discusses  previous   work  related  to
-garbage  collection  and   liveness  and  Section~\ref{sec:conclusion}
+GC  and   liveness  and  Section~\ref{sec:conclusion}
 discusses possible extensions and concludes the paper.
@@
 a union of  the liveness arising from these dependence  chains.  To be
-safe, a garbage collection during  evaluation of $\py$ at $\psi_1$ has
+safe, a GC during  evaluation of $\py$ at $\psi_1$ has
 to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
-notice that if a garbage collection takes place while evaluating $\pz$
+notice that if a GC takes place while evaluating $\pz$
 at $\psi_3$, it can safely consider only the liveness arising from the
@@
 free variables of $s$ due to  the application.  This will be consulted
-during garbage collection  while exploring the heap  starting from the
+during GC  while exploring the heap  starting from the
 closure  variables.  The function $\mathcal{L}$  uses  $\mathit{ref}$ to  propagate
@@
 If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
-the liveness of further heap contents starting from $\mathit{x_{\pi}}$
+the liveness of heap contents starting from $\mathit{x_{\pi}}$
 is given by $\Lv_\mathit{x_{\pi}}$.
@@
   act of  not retaining  (effectively garbage collecting)  the objects
-  pointed to by these references  during an actual garbage collection.
+  pointed to by these references  during an actual GC.
   This is done in the following way:
@@
 %% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
-every transition  is preceded  by a  garbage collection
+every transition  is preceded  by a  GC
 using $GC$.  The details of the transition for the {\sc
@@
 inserted  by  the  immediately  preceding garbage  collection  or  the
-garbage collection of an earlier minefield step.  However the
+GC of an earlier minefield step.  However the
   demand $\sigma'$ on basis of  which an earlier collection would have
@@
   Thus, if  the earlier garbage  collection had inserted a  $\bot$, so
-  would the current garbage collection.  Thus it is enough to consider
+  would the current GC.  Thus it is enough to consider
   the $\rightsquigarrow$  along with the  current GC and show  that it
@@
 of    $x$   will    include    both   $\epsilon$    and
-$\acar\epsilon$,  and  a  garbage collection with  this  liveness  will
+$\acar\epsilon$,  and  a  GC with  this  liveness  will
 neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
@@
 \begin{align*}
-  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
+  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} 
+\end{align*}
+\begin{align*}
   \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
@@
 However, recall that the decision problem that we are interested in
-during garbage collection is: 
+during GC is: 
 %%\begin{quote} 
@@
 
-%\subsection{A liveness-based garbage collection scheme}
+%\subsection{A liveness-based GC scheme}
 \label{sec:live-clo}
@@
 %to approximate a CFG by a {\em strongly regular\/} grammar.  
-For example, the grammar fragment for the non-terminal
+For example, the grammar fragment for %the non-terminal
 $\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
@@
 \vspace*{-2.5mm}
-\section{The garbage collection scheme}
+\section{The GC scheme}
 \label{sec:GC-scheme}
@@
 points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
-$\epath_2$ are two such evaluation points.  During garbage collection,
+$\epath_2$ are two such evaluation points.  During GC,
 we would like to use a more precise liveness, based on the actual paths
@@
   available memory is  less than this requirement for  the function, a
-  garbage collection is triggered.
+  GC is triggered.
 \item Since the evaluation of a  \SIF\ condition may trigger a 
@@
   (based on  value of the condition)  required to execute the  rest of
-  the program.  A garbage collection is triggered if enough memory is unavailable.
+  the program.  A GC is triggered if enough memory is unavailable.
 \end{enumerate}
@@
 %% Printing of  atomic values is simple  and does
-%% not  trigger a  garbage  collection.   In case  of  \CONS\ cells,  the
+%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
 %% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
 %% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
-%% trigger  more garbage  collections.  When  this happens.   we have  to
+%% trigger  more GCs.  When  this happens.   we have  to
 %% consider any  references that  might be  on the  print stack  and copy
@@
 
-Table~\ref{tab:exp-results}(b) compares  garbage collection statistics
-for RGC and  LGC.  We report the number of  garbage collection events,
-average  number of  cells  reclaimed per  garbage collection,  average
-number of cells  touched per garbage collection and the  total time to
+Table~\ref{tab:exp-results}(b) compares  GC statistics
+for RGC and  LGC.  We report the number of  GC events,
+average  number of  cells  reclaimed per  GC,  average
+number of cells  touched per GC and the  total time to
 perform all collections.   It is no surprise that the  number of cells
@@
 Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
-number of  garbage collections  tend to be  very large,
+number of  GCs  tend to be  very large,
 hence  we  show only  a  slice of the original graph  with 
@@
 dips   below the RGC curve 
-(\RGCLine\ line) during garbage collection.  The graphs also include the
+(\RGCLine\ line) during GC.  The graphs also include the
 curve    for   reachable    cells   (\ReachLine)  and live cells
@@
 \label{sec:relatedwork}
-The impact of  liveness on the effectiveness of  garbage collection is
+The impact of  liveness on the effectiveness of  GC is
 investigated in~\cite{Hirzel}.   The observation is that  liveness can
@@
 There have been several attempts to use liveness analysis to improve 
-garbage collection for imperative languages.~\cite{khedker07heap}
+GC for imperative languages.~\cite{khedker07heap}
 presents a liveness analysis and uses the results for inserting 
@@
 daemons that attach themselves to program data and, when activated, improve
-the efficiency of the program. Our liveness-based garbage collection
-as an instance of a  simplifier which is tightly coupled with garbage collectors.  The
+the efficiency of the program. Our liveness-based GC
+is an instance of a  simplifier which is tightly coupled with garbage collectors.  The
 approach that is closest to the  method described in this paper is the
@@
 
-%% A similar approach has been suggested in~\cite{ONeill} where they suggest augmenting garbage collection with extra information using {\it simplifiers}.
+%% A similar approach has been suggested in~\cite{ONeill} where they suggest augmenting GC with extra information using {\it simplifiers}.
 
@@
 %% this.
-We have  extended liveness-based  garbage collection  to lazy
+We have  extended liveness-based  GC  to lazy
   languages  and shown  its benefit  for a  set of  benchmark
@@
   schemes,   such  as   incremental  and   generational
-  collection.  It might be interesting to use a mixed mode garbage collection
+  collection.  It might be interesting to use a mixed mode GC
   scheme which allows the costs of LGC to be amortized over several runs
   of RGC. In summary, we  need to investigate ways
-  to make liveness-based garbage collection attractive
+  to make liveness-based GC attractive
   for practical collectors.
@@
 \begin{figure*}[b!]
-\noindent{\Large \bf A. Minefield semantics(Not included in the paper)}
+\noindent{\Large \bf A. Minefield semantics (Not to be included in the paper) }
 \begin{center}\footnotesize

### 2016-02-03 paper/paper.tex
Changed order of statements in introduction

--- 2016-02-03 paper/paper.tex
@@
 
-\todelete{While  this  idea has  been  shown  to  be effective  for  a
-  first-order     {\em     eager}    language~\cite{asati14lgc},     a
-  straightforward extension of the technique  is not possible for lazy
-  languages,  where   heap-allocated  objects  may   include  closures
-  (runtime   representations    of   unevaluated    expressions).    }
-\added{Designing a liveness based garbage collector for lazy languages
-  poses significant challenges.}  Firstly, since  data is made live by
-evaluation of closures, and in lazy languages the place in the program
-where  this evaluation  takes place  cannot be  statically determined,
-laziness   complicates  liveness   analysis  itself.    Moreover,  for
-liveness-based garbage collection  to be effective, we  need to extend
-it to closures apart from evaluated  data.  Since a closure can escape
-the  scope in  which it  was  created, it  has to  carry the  liveness
-information  of  its  free  variables.  As  execution  progresses  and
-possible  future   uses  are   eliminated,  we  update   the  liveness
-information  in a  closure with  a more  precise version.   Because of
-these reasons,  the garbage collector also  becomes significantly more
-complicated than a liveness-based collector for a eager language.
-
+\todelete{While this idea has been shown  to be effective for a first-order {\em
+  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
+the technique is not possible for lazy languages, where heap-allocated
+objects may  include closures (runtime representations  of unevaluated
+expressions).}
+We encounter significant challenges in defining a liveness based garbage 
+collector for lazy languages. 
+\added{
+Firstly, during garbage collection the heap contains both fully 
+evaluated data and unevaluated thunks/closures. 
+For liveness-based
+garbage collection to  be effective, we need to extend  it to closures
+apart from  evaluated data. 
+Moreover, since  data  is made  live  by evaluation  of
+closures, and  in lazy languages the  place in the program  where this
+evaluation  takes  place  cannot be  statically  determined,  laziness
+complicates  liveness analysis  itself.
+Since closures can escape the scope in which 
+they were created, they need to carry liveness information of its free 
+variables. 
+As  execution progresses and possible  future uses are
+eliminated, we  update the  liveness information in  a closure  with a
+more precise version.  Because of these reasons, the garbage collector
+also  becomes significantly  more  complicated  than a  liveness-based
+collector for a eager language.
+}
+
+\todelete{
+Firstly, since  data  is made  live  by evaluation  of
+closures, and  in lazy languages the  place in the program  where this
+evaluation  takes  place  cannot be  statically  determined,  laziness
+complicates  liveness analysis  itself.  Moreover,  for liveness-based
+garbage collection to  be effective, we need to extend  it to closures
+apart from  evaluated data.  Since a  closure can escape the  scope in
+which it was created, it has  to carry the liveness information of its
+free variables. As  execution progresses and possible  future uses are
+eliminated, we  update the  liveness information in  a closure  with a
+more precise version.  Because of these reasons, the garbage collector
+also  becomes significantly  more  complicated  than a  liveness-based
+collector for a eager language.
+}
 Experiments    with   a    single    generation   copying    collector
@@
 \begin{myitemize}
-\item We propose a liveness based garbage collection scheme for a lazy
-  first order functional  language and prove its  correctness.  To the
-  best  of our  knowledge, this  is the  first work  on liveness-based
-  garbage    collection   for    lazy    languages.    All    previous
-  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
-    asati14lgc,    karkare06effectiveness}   have    involved   either
-  imperative  or  eager  functional  languages.   The  novely  of  our
-  approach is in  handling of  liveness of  functional values  (i.e.,
-  closures), an issue that arises due to laziness.
-
-\item  We  formulate a  liveness  analysis  for the  lazy  first-order
-  functional language  and prove  its correctness. The  proof involves
-  specifying  liveness   for  the  language  through   a  non-standard
-  semantics and then proving the  analysis correct with respect to the
-  specification.
-
-\item The  analysis results  in a set  of context-free  grammars along
+\item  We propose a liveness based garbage collection scheme for a lazy first order 
+functional language and prove its correctness. 
+To the best of our knowledge this is the first effort to define a
+liveness based garbage collection for a lazy language. 
+Our main contribution is the way we handle liveness of functional values i.e closures which 
+distinguishes our work from earlier work in this area.
+
+\item
+  We  formulate  a  liveness  analysis for  the  lazy  first-order
+  functional language and  prove its correctness. The 
+  proof involves specifying liveness through a non-standard semantics  and then proving the analysis  
+  correct with respect to the specification.
+
+\item 
+  The analysis results  in a set  of context-free  grammars along
   with  a fixed  set  of non-context-free  productions.  The  decision
-  whether to  copy a  cell during garbage  collection translates  to a
-  membership problem for such grammars.   In this paper, we have shown
-  the undecidability  of this  problem. We  circumvent the  problem by
-  over-approximating the grammars with DFAs.
-
-\item We have implemented a garbage  collector that uses the result of
-  liveness  analysis to  retain  live cells.   Our experiments  reveal
-  interesting  space-time   trade-offs  in  the  engineering   of  the
-  collector---for  example, updating  liveness information  carried in
-  closures during  execution results in more  garbage being collected.
-  Empirical results  show the effectiveness of  liveness-based garbage
-  collection.
+  whether   to  copy  a  cell   during  garbage  collection
+  translates  to a  membership problem  for such  grammars.  In  this
+  paper, we  have shown the  undecidability of this problem. We 
+  circumvent the  problem by over-approximating the  grammars
+  with DFAs.
+\item 
+We  have implemented a garbage  collector that uses the  result of liveness analysis to
+  retain live  cells.
+Our  experiments reveal  interesting space-time
+  trade-offs  in  the  engineering  of  the  collector---for  example,
+  updating liveness  information carried in closures  during execution
+  results  in   more  garbage  being  collected.    Empirical  results
+  show the effectiveness of liveness-based garbage collection.
 \end{myitemize}

### 2016-02-03 paper/paper.tex
spell check

--- 2016-02-03 paper/paper.tex
@@
     asati14lgc,    karkare06effectiveness}   have    involved   either
-  imperative  or  eager  functional  languages.   The  novely  of  our
+  imperative  or  eager  functional  languages.   The  novelty  of  our
   approach is in  handling of  liveness of  functional values  (i.e.,
@@
 
-In the space of functional  languages, there are: rewriting methods such as
+In the  space of  functional languages,  there are:  rewriting methods
+such                                                                as
 deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
 sharing  analysis   based  reallocation~\cite{jones89compile},  region
-based   analysis~\cite{tofte98region},   and    insertion   of
-compile-time                                                nullifying
-statements~\cite{inoue88analysis,lee05static,Hamilton}.            All
-compile-time marking approaches rely on an efficient and precise alias
-analysis and  cannot provide  significant improvement in  its absence.
-An interesting approach suggested  in~\cite{HofmannJ03} is to annotate
-the  heap usage  of first-order  programs through  linear types.   The
-annotations   are  then   used  to   serve  memory   requests  through
+based  analysis~\cite{tofte98region},  and insertion  of  compile-time
+nullifying     statements~\cite{inoue88analysis,lee05static,Hamilton}.
+All compile-time marking  approaches rely on an  efficient and precise
+alias  analysis  and cannot  provide  significant  improvement in  its
+absence.  An interesting approach suggested in~\cite{HofmannJ03} is to
+annotate the heap usage of  first-order programs through linear types.
+The  annotations  are  then  used to  serve  memory  requests  through
 re-allocation.  However, this requires the user to write programs in a
@@
 Simplifiers~\cite{ONeill}  are  abstractly  described  as  lightweight
-daemons that attach themselves to program data and, when activated, improve
-the efficiency of the program. Our liveness-based GC
-is an instance of a  simplifier which is tightly coupled with garbage collectors.  The
-approach that is closest to the  method described in this paper is the
-liveness-based garbage collector implemented
-in~\cite{karkare07liveness,asati14lgc} and address eager languages. We
-extend this to handle lazy evaluation and closures.
-
-%% A similar approach has been suggested in~\cite{ONeill} where they suggest augmenting GC with extra information using {\it simplifiers}.
+daemons that  attach themselves to  program data and,  when activated,
+improve the  efficiency of the  program. Our liveness-based GC  can be
+seen as  an instance  of a  simplifier which  is tightly  coupled with
+garbage  collectors.   The approach  that  is  closest to  the  method
+described  in  this  paper  is the  liveness-based  garbage  collector
+implemented  in~\cite{karkare07liveness,asati14lgc} and  address eager
+languages. We extend this to handle lazy evaluation and closures.
+
+%% A similar  approach has been suggested  in~\cite{ONeill} where they
+%% suggest   augmenting  GC   with   extra   information  using   {\it
+%% simplifiers}.
 

### 2016-02-04 paper/paper.tex
Added reference to safe for space paper
references and space saving

@@
 The impact of  liveness on the effectiveness of  GC is
-investigated in~\cite{Hirzel}.   The observation is that  liveness can
+investigated in~\cite{Hirzel}.   They observe that  liveness can
 significantly  impact   garbage  collection,  but  only   when  it  is
@@
 presents a liveness analysis and uses the results for inserting 
-nullifying statements in Java programs. In ~\cite{ran.shaham-sas03},
-an automaton modelling heap accesses is created, and
-temporal properties like liveness are checked  against the model.  
-However both the approaches are restricted to intraprocedural scope.
+nullifying statements in Java programs. In~\cite{ran.shaham-sas03},
+temporal properties like liveness are checked  against an automaton modelling heap accesses.
+Both these approaches are intraprocedural in scope.
 
@@
 specific way. 
+%
+Safe-for-space~\cite{appel.cps}              approaches~\cite{Clinger,
+  shao00efficient}  reduce the  amount of  heap used  by a  program by
+allocating   closures    in   registers   and   through    tail   call
+optimizations. However, these approaches take care of only part of the
+problem addressed  by our  analysis as the  program can  still contain
+unused objects and closures that are reachable.
+%
 
@@
 %% \end{figure}
+\vspace*{-2mm}
 \vskip -5mm

### 2016-02-04 paper/paper.tex
Cleaned up paper for first submission

@@
 \newcommand{\mycomment}[1]{{\color{Myblue}{#1}} }
-\newcommand{\added}[1]{{\color{Myblue}{#1}}}
+%\newcommand{\added}[1]{{\color{Myblue}{#1}}}
+\newcommand{\added}[1]{#1}
 \newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
 
-\newcommand{\warning}[1]{{\color{Myred}{#1}}}
+%\newcommand{\warning}[1]{{\color{Myred}{#1}}}
+\newcommand{\warning}[1]{}
 %\newcommand{\blankout}[1]{}
@@
 that records liveness values at each reference created during execution.
-Any attempt to dereference a reference \mycomment{that has been declared not live by our analysis} %%with a null-liveness   
+Any attempt to dereference a reference \added{that has been declared not live by our analysis} %%with a null-liveness   
  during  a  transition  will  result  in entering  a

### 2016-02-04 paper/paper.tex
Renamed sections

@@
 
-\section{Computing liveness and its encoding as DFA}\label{sec:computing}
+%\section{Computing liveness and its encoding as DFA}\label{sec:computing}
+\section{Generating and storing liveness information}\label{sec:computing}
 Section~\ref{sec:liveness}  gave   a  fully context-sensitive
@@
 
-\subsection{Modifying liveness rules}
+%\subsection{Modifying liveness rules}
+\subsection{Tweaking the rules}
 
@@
 
-\subsection{Generating equations for $\Lfonly_{\mathit f}$}
+%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
+\subsection{Computing function summaries $\Lfonly_{\mathit f}$}
 Given a  function $\mathit{f}$, we now  describe how to
@@
 
-\subsection{Generating liveness equations \Lv\  for function bodies}
+%\subsection{Generating liveness equations \Lv\  for function bodies}
+\subsection{Handling user-defined functions}
 \label{sec:bodylivenessbodies}
@@
 
-\subsection{Solving liveness equations---the grammar
-interpretation}\label{sec:grammar-formulation}
+%\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
+\subsection{From liveness sets to context-free grammars}\label{sec:grammar-formulation}
 

### 2016-02-06 paper/paper.tex
Updated soundness proof

@@
 
+{\color{Myblue}
+As we shall see in section??, the liveness of the closure variable $x_{\pi}$
+is recorded along with the closure so that the GC can refer to it
+during garbage collection. 
+}
+
 %} %comment
@@
 \begin{enumerate}
-\item       We   augment     the        standard       semantics
-  (Figure~\ref{fig:lang-semantics}) to model  a dynamic liveness analysis 
-that records liveness values at each reference created during execution.
-Any attempt to dereference a reference \added{that has been declared not live by our analysis} %%with a null-liveness   
- during  a  transition  will  result  in entering  a
-  special state denoted \bang. We call the semantics after
-  augmentation,  \emph{minefield semantics}.
+\item       We       augment        the       standard       semantics in
+  Figure~\ref{fig:lang-semantics} to model a GC before the execution
+  of each \LET\ expression. {\color{Myblue}Note  that memory is allocated only during
+  execution of \LET\  expressions}. During GC, we  track each reference
+  in the root set and heap that is not declared live by our analysis.
+Any  attempt to dereference such references results
+  in the transition system entering  a special state denoted \bang.
+We
+  call the semantics after augmentation, \emph{minefield semantics}.
+
+%% During GC, we bind  a special value
+%%   $\bot$ to each  reference in the root  set and heap  that is not
+%%   live by  our analysis.  This  models the de-allocation  of locations
+%%   without any  live references.  Any  attempt to dereference  a $\bot$
+%%   during execution results
+%% %\added{that  has   been   declared  not   live  by   our
+%% %    analysis} %%with  a null-liveness during a  transition will result
+%%   in the transition system entering  a special state denoted \bang.  We
+%%   call the semantics after augmentation, \emph{minefield semantics}.
 \item \label{inline1} Assuming that a program enters the \bang\ state,
   we  show  how to  construct,  through  inline expansion,  a  program
-  without function calls which has the same minefield behavior.
-\item  The final  step   shows  that  no program  without
-  function calls can enter a \bang\ state. As a consequence no program
-  (with or without function calls) can enter the \bang\ state. 
+  without function  calls which has  the same minefield  behavior. The
+  final step shows that no program  without function calls can enter the
+  \bang\ state. As a consequence  no program (with or without function
+  calls) can enter the \bang\ state.
 \end{enumerate}
@@
   state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
-  $e$, i.e., it arises from the actual sequence of function calls that
-  led  to the  evaluation  of $e$.
-The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.
-The  information in  continuation
-  frames  on the  stack $S$  are also  similarly augmented  with their
-  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
-  \sigma)$.
+  $e$, that arises from the actual sequence of function calls that led
+  to  the  evaluation of  $e$.\footnote{The  static  liveness that  is
+    consulted during actual GC  is computed from an over-approximation
+    of this demand.  Thus the soundness  result on the modeled GC will
+    also apply for the actual GC.} The demand for the initial state is
+  $\sigma_{\mathit  {all}}$,  and each  $\rightsquigarrow$  transition
+  transforms   the  demand   according  to   the  liveness   rules  of
+  Section~\ref{sec:liveness-analysis}.      The     information     in
+  continuation frames  on the stack  $S$ are also  similarly augmented
+  with their demands.  Thus a stacked entry now takes the form $(\rho,
+  \ell, e, \sigma)$.
+
+\item The closure created by a let expression is now a 3-tuple
+  $\langle s,  \rho, \sigma \rangle$, where $\sigma$ represents the
+  demand on the closure.
+
 \item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
@@
   $\rho, S$ and  $\heap$ are due
-  to certain references being replaced by $\bot$.  This simulates the
-  act of  not retaining  (effectively garbage collecting)  the objects
+  to non-live references being replaced by $\bot$.  This simulates the
+  act of  not retaining  (effectively garbage collecting)  the cells
   pointed to by these references  during an actual GC.
@@
   \begin{enumerate} 
+ \item Consider the following environments, 1) environment in the
+   current state, 2) environments in each of the stacked items and 3)
+   the environment in each of the closures in the heap. For each of
+   these environments we can calculate a corresponding liveness environment
+   using the  $e\ and\ \sigma$. 
+\item \emph{GC}
+   for each $x \in
+    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
+    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
+    in $S$ 
+    with  $\Lv'$   as  the   corresponding  liveness   environment  in
+    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
+    to $\bot$ iff $x.\epsilon \notin \Lv'$.
+
   \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
@@
     $\mathsf{SL}$.
-  \item \emph{Garbage collection of objects pointed by root sets:} For
+  \item \emph{Garbage collection of objects pointed by root sets:}
+    Consider each of the following environments: (i) The 
+    the environment $\rho$ in the current state (ii) and for each $x \in
+    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
+    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
+    in $S$ 
+    with  $\Lv'$   as  the   corresponding  liveness   environment  in
+    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
+    to $\bot$ iff $x.\epsilon \notin \Lv'$.
+
+Consider each of the following environments: (i) The 
     the environment $\rho$ in the current state and for each $x \in
@@
     to $\bot$ iff $x.\epsilon \notin \Lv'$.
-  \item \emph{Garbage  collection of  objects in  the heap:}  For each
+
+  \item \emph{Garbage  collection of  objects in  the heap:}
+
+    For each
     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each

### 2016-02-06 paper/paper.tex
Changed as per discussion on 5th Feb

--- 2016-02-04 paper/paper.tex
@@
 
-\todelete{While  this  idea has  been  shown  to  be effective  for  a
-  first-order     {\em     eager}    language~\cite{asati14lgc},     a
-  straightforward extension of the technique  is not possible for lazy
-  languages,  where   heap-allocated  objects  may   include  closures
-  (runtime   representations    of   unevaluated    expressions).    }
-\added{Designing a liveness based garbage collector for lazy languages
-  poses significant challenges.}  Firstly, since  data is made live by
-evaluation of closures, and in lazy languages the place in the program
-where  this evaluation  takes place  cannot be  statically determined,
-laziness   complicates  liveness   analysis  itself.    Moreover,  for
-liveness-based GC  to be effective, we  need to extend
-it to closures apart from evaluated  data.  Since a closure can escape
-the  scope in  which it  was  created, it  has to  carry the  liveness
-information  of  its  free  variables.  As  execution  progresses  and
-possible  future   uses  are   eliminated,  we  update   the  liveness
-information  in a  closure with  a more  precise version.   Because of
-these reasons,  the garbage collector also  becomes significantly more
-complicated than a liveness-based collector for a eager language.
+\added{While  the idea  of  using static  analysis  to improve  memory
+  utilization  has  been  shown  to   be  effective  for  {\em  eager}
+  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
+    lee05static}, a straightforward extension  of the technique is not
+  possible  for  lazy  languages,  where  heap-allocated  objects  may
+  include   closures    (runtime   representations    of   unevaluated
+  expressions).   }   \todelete{Designing  a  liveness  based  garbage
+  collector   for  lazy   languages  poses   significant  challenges.}
+Firstly, since  data is made  live by  evaluation of closures,  and in
+lazy languages  the place in  the program where this  evaluation takes
+place cannot  be statically determined, laziness  complicates liveness
+analysis itself.  Moreover, for liveness-based  GC to be effective, we
+need to  extend it  to closures  apart from  evaluated data.   Since a
+closure can escape the scope in which  it was created, it has to carry
+the  liveness  information  of   its  free  variables.   As  execution
+progresses  and possible  future uses  are eliminated,  we update  the
+liveness  information  in  a  closure with  a  more  precise  version.
+Because  of   these  reasons,  the  garbage   collector  also  becomes
+significantly more  complicated than a liveness-based  collector for a
+eager language.
 
@@
 \begin{itemize}
-\item We propose a liveness based GC scheme for a lazy
-  first order functional  language and prove its  correctness.  To the
-  best  of our  knowledge, this  is the  first work  on liveness-based
-  GC   for    lazy    languages.    All    previous
-  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
-    asati14lgc,    karkare06effectiveness}   have    involved   either
-  imperative  or  eager  functional  languages.   The  novelty  of  our
-  approach is in  handling of  liveness of  functional values  (i.e.,
-  closures), an issue that arises due to laziness.
+\item We  propose a liveness  based GC scheme  for a lazy  first order
+  functional language and  prove its correctness.  To the  best of our
+  knowledge,  this is  the first  work on  liveness-based GC  for lazy
+  languages.       All      previous      attempts~\cite{shaham01heap,
+    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
+    karkare06effectiveness} have  involved either imperative  or eager
+  functional languages.  The novelty of our approach is in handling of
+  liveness  of closures,  an  issue  that arises  due  to laziness  at
+  runtime.
 
@@
   with  a fixed  set  of non-context-free  productions.  The  decision
-  whether to  copy a  cell during GC translates  to a
-  membership problem for such grammars.   In this paper, we have shown
-  the undecidability  of this  problem. We  circumvent the  problem by
-  over-approximating the grammars with DFAs.
+  whether to copy a cell during  GC translates to a membership problem
+  for  such  grammars.  Earlier  assumed  the  undecidability of  this
+  membership question  and used an over-approximation  to overcome it.
+  In this  paper, we provide a  formal proof of the  undecidability of
+  this problem.
 
@@
 
-%\section{Computing liveness and its encoding as DFA}\label{sec:computing}
-\section{Generating and storing liveness information}\label{sec:computing}
+\section{Computing liveness and its encoding as DFA}\label{sec:computing}
+%\section{Generating and storing liveness information}\label{sec:computing}
 Section~\ref{sec:liveness}  gave   a  fully context-sensitive
@@
 %\subsection{Modifying liveness rules}
-\subsection{Tweaking the rules}
+\paragraph{Symbolic representation of operations}
 
@@
 %\subsection{Generating equations for $\Lfonly_{\mathit f}$}
-\subsection{Computing function summaries $\Lfonly_{\mathit f}$}
+\paragraph{Computing function summaries $\Lfonly_{\mathit f}$}
 Given a  function $\mathit{f}$, we now  describe how to
@@
 %\subsection{Generating liveness equations \Lv\  for function bodies}
-\subsection{Handling user-defined functions}
+\paragraph{Handling user-defined functions}
 \label{sec:bodylivenessbodies}
@@
 Our benchmark  consists of programs taken  from nofib~\cite{nofib} and
-other repositories for functional programs~\cite{PLT-Scheme,
-  gc_bench, huffman-sicp}.  We ran the experiments on 8 core Intel\textsuperscript{\textregistered} Core\textsuperscript{TM}
+other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
+  huffman-sicp}.     We    ran    the   experiments    on    8    core
+Intel\textsuperscript{\textregistered}        Core\textsuperscript{TM}
 i7-4770 3.40GHz CPU  having 8192KB L2 cache, 16GB RAM,  running 64 bit
@@
 There have been  several attempts to use liveness  analysis to improve
-GC for imperative languages.~\cite{khedker07heap}
-presents a liveness analysis and uses the results for inserting 
-nullifying statements in Java programs. In~\cite{ran.shaham-sas03},
-temporal properties like liveness are checked  against an automaton modelling heap accesses.
+GC for imperative  languages.~\cite{khedker07heap} presents a liveness
+analysis and uses  the results for inserting  nullifying statements in
+Java  programs. In~\cite{ran.shaham-sas03},  temporal properties  like
+liveness  are checked  against an  automaton modelling  heap accesses.
 Both these approaches are intraprocedural in scope.
@@
 based analysis~\cite{tofte98region}, and insertion of compile-time
-nullifying     statements~\cite{inoue88analysis,lee05static,Hamilton}.
-All compile-time marking  approaches rely on an  efficient and precise
-alias  analysis  and cannot  provide  significant  improvement in  its
-absence.  An interesting approach suggested in~\cite{HofmannJ03} is to
-annotate the heap usage of  first-order programs through linear types.
-The  annotations  are  then  used to  serve  memory  requests  through
+nullifying statements~\cite{inoue88analysis}.  All compile-time
+marking approaches rely on an efficient and precise alias analysis and
+cannot provide significant improvement in its absence. The only work
+in the space of lazy languages seems to be~\cite{lee05static,Hamilton}
+which touches upon only basic techniques of compile-time garbage
+marking, explicit deallocation and desctructive allocation. An
+interesting approach suggested in~\cite{HofmannJ03} is to annotate the
+heap usage of first-order programs through linear types.  The
+annotations are then used to serve memory requests through
 re-allocation.  However, this requires the user to write programs in a
@@
 closures there still exists a gap between actual and perceived
-livenesses.
-Language features that could be introduced are higher-order functions
-and recursive lets. We believe that functions can be used to model
-recursive lets and this should be a simple extension to our liveness
-analysis.
-For higher-order we intend to use firstification techniques~\cite{Mitchell:2009}.
-
-  Orthogonally, we  plan to  improve the  efficiency of
-  the liveness-based garbage collector using heuristics
-  such   as  limiting   the  depth   of  DFA,   merging
-  nearly-equivalent    states    and    using    better
-  representation(for example BDDs~\cite{Bryant86})    and    algorithms   for    automata
-  manipulation.  We   also  need  to   investigate  the
-  interaction   of  liveness   with  other   collection
-  schemes,   such  as   incremental  and   generational
-  collection.  It might be interesting to use a mixed mode GC
-  scheme which allows the costs of LGC to be amortized over several runs
-  of RGC. In summary, we  need to investigate ways
-  to make liveness-based GC attractive
-  for practical collectors.
+livenesses.  Language features that could be introduced are
+higher-order functions and recursive lets. We believe that functions
+can be used to model recursive lets and this should be a simple
+extension to our liveness analysis.  For higher-order we intend to use
+firstification techniques~\cite{Mitchell:2009}.
+
+  Orthogonally,   we   plan  to   improve   the   efficiency  of   the
+  liveness-based garbage  collector using heuristics such  as limiting
+  the depth of DFA, merging  nearly-equivalent states and using better
+  representation(for example BDDs~\cite{Bryant86})  and algorithms for
+  automata manipulation.  We also  need to investigate the interaction
+  of liveness with  other collection schemes, such  as incremental and
+  generational collection.   It might  be interesting  to use  a mixed
+  mode GC  scheme which allows the  costs of LGC to  be amortized over
+  several runs of RGC. In summary, we need to investigate ways to make
+  liveness-based GC attractive for practical collectors.
 

### 2016-02-07 paper/paper.tex
Modified minefield semantics

@@
 
-{\color{Myblue}
-As we shall see in section??, the liveness of the closure variable $x_{\pi}$
-is recorded along with the closure so that the GC can refer to it
-during garbage collection. 
-}
+ As we  shall see in section??,  the liveness of the  closure variable
+ $x_{\pi}$ is recorded along with the closure so that the GC can refer
+ to it during garbage collection.
 
@@
   Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
-  of each \LET\ expression. {\color{Myblue}Note  that memory is allocated only during
-  execution of \LET\  expressions}. During GC, we  track each reference
-  in the root set and heap that is not declared live by our analysis.
-Any  attempt to dereference such references results
-  in the transition system entering  a special state denoted \bang.
-We
-  call the semantics after augmentation, \emph{minefield semantics}.
+  of  each  \LET\  expression.   {\color{Myblue}Note  that  memory  is
+    allocated only during execution  of \LET\ expressions}. During GC,
+  we  track each  reference  in the  root  set and  heap  that is  not
+  declared  live by  our analysis.   Any attempt  to dereference  such
+  references results in the transition system entering a special state
+  denoted   \bang.   We   call  the   semantics  after   augmentation,
+  \emph{minefield semantics}.
 
@@
 \hline 
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$, \\
-  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
+\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
-    S, \heap)}$, \\$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
+\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
 (\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\ $\heap(\rho(x))
+\makecell[t]{ $\heap(\rho(x))
 \mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
@@
   \begin{enumerate}
+
 \item  Consider the following environments, 1) environment in the
@@
    using the  $e\ and\ \sigma$. 
-\item \emph{GC}
-   for each $x \in
-    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
-    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
-    in $S$ 
-    with  $\Lv'$   as  the   corresponding  liveness   environment  in
-    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
-    to $\bot$ iff $x.\epsilon \notin \Lv'$.
+\item \emph{GC}  for each  $x \in domain(\rho)$,  we set  $\rho(x)$ to
+  $\bot$ iff $x.\epsilon \notin \Lv$.  Similarly, for each stack entry
+  $(\rho,\ell,e,\sigma)$  in  $S$  with $\Lv'$  as  the  corresponding
+  liveness environment in  $\mathsf{SL}$, and finally for  each $x \in
+  domain(\rho)$  for some  closure in  the heap,  we set  $\rho(x)$ to
+  $\bot$ iff $x.\epsilon \notin \Lv'$ corresponding to the closure.
 
@@
 based analysis~\cite{tofte98region}, and insertion of compile-time
-nullifying statements~\cite{inoue88analysis}.  All compile-time
+nullifying statements~\cite{inoue88analysis, lee05static}.  All compile-time
 marking approaches rely on an efficient and precise alias analysis and
 cannot provide significant improvement in its absence. The only work
-in the space of lazy languages seems to be~\cite{lee05static,Hamilton}
+in the space of lazy languages seems to be~\cite{Hamilton}
 which touches upon only basic techniques of compile-time garbage
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\
-  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
+\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S,
-    \heap)}$, \\ $\heap(\rho(x)) \mbox{ is } (v, d)$} & \makecell[t]{$\rho, (\rho', \ell, e,
+\makecell[t]{$\heap(\rho(x)) \mbox{ is } (v, d)$} & \makecell[t]{$\rho, (\rho', \ell, e,
   \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
-    S, \heap)}$, \\ $\heap(\rho(x))
+\makecell[t]{ $\heap(\rho(x))
 \mbox{ is } \langle s, \rho'\rangle$ }& \makecell[t]{$\rho, S,
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\
-  $\rho(x) \mbox{ is } \bot$  or $\rho(y) \mbox{ is }
+\makecell[t]{$\rho(x) \mbox{ is } \bot$  or $\rho(y) \mbox{ is }
   \bot$} & $\rho, S,
@@
 
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho,
-    S, \heap)}$, \\ 
-$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$}
+\makecell[t]{$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$}
  &
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
+\makecell[t]{$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
   \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho, \rho(x),
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
+\makecell[t]{$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
   S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho,\rho(y),
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (f~\myvec{x}), \sigma) =
-    (\rho, S, \heap)}$,\\ $\mathit{f}~\mbox{defined as}$
+\makecell[t]{$\mathit{f}~\mbox{defined as}$
 $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell[t]{$\rho, S, \heap,
@@
   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
-    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
+    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
+    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
 {\sc let} \\ 
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
-    \sigma) = (\rho, S, \heap)}$, \\
-  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
+\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
   \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
-    \sigma) = (\rho, S, \heap)}$, \\$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
+\makecell[t]{$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
+\makecell[t]{$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
   \cred{\sigma}$   $\rightsquigarrow
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
+\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow$ \\ $
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\
+\makecell[t]{
   $\rho(x) \mbox{ is } \bot$} & $\rho, S,
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) =
-    (\rho, S, \heap)}$,\\ $\heap(\rho(x))~\mbox{is in WHNF with
+\makecell[t]{$\heap(\rho(x))~\mbox{is in WHNF with
     value}~v$} & \makecell[t]{$\rho,
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
+\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
   (\SRETURN~x), \cred{\sigma}$  $

### 2016-02-08 paper/paper.tex

@@
 \definecolor{Myred}{rgb}{1,.2,.2}
-\newcommand{\mycomment}[1]{{\color{Myblue}{#1}} }
-%\newcommand{\added}[1]{{\color{Myblue}{#1}}}
-\newcommand{\added}[1]{#1}
-\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
+\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
+\newcommand{\added}[1]{{\color{Myblue}{#1}}}
+%\newcommand{\added}[1]{#1}
+\newcommand{\todelete}[1]{{\color{Myred}{#1}}}
 
@@
 In this paper,  we propose the use of liveness  analysis of heap cells
-for garbage collection (GC) in a lazy first-order functional language.  The
-central notion in our analysis  is a generalization of liveness called
-{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
+for garbage collection (GC) in a lazy first-order functional language.
+The central  notion in  our analysis is  a generalization  of liveness
+called {\em  demand}---the pattern of future  uses of the value  of an
 expression.   The  analysis  has  two parts.   We  first  calculate  a
@@
 capturing the  liveness of  variables at  these points.   Depending on
-where GC is triggered, the collector consults a set of
-automata to restrict reachability during  marking.  This results in an
-increase in  the garbage reclaimed  and consequently in  fewer 
-collections.
+where GC  is triggered, the  collector consults  a set of  automata to
+restrict reachability during marking.  This  results in an increase in
+the garbage reclaimed and consequently in fewer collections.
 
@@
 \leftarrow~s~\IN~e)$.
-
+{\color{Myblue}
 As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
@@
  to it during garbage collection.
-
+}
 %} %comment
@@
   the environment  in each of  the closures in  the heap. For  each of
-   these environments we can calculate a corresponding liveness environment
-   using the  $e\ and\ \sigma$. 
-\item \emph{GC}  for each  $x \in domain(\rho)$,  we set  $\rho(x)$ to
-  $\bot$ iff $x.\epsilon \notin \Lv$.  Similarly, for each stack entry
-  $(\rho,\ell,e,\sigma)$  in  $S$  with $\Lv'$  as  the  corresponding
-  liveness environment in  $\mathsf{SL}$, and finally for  each $x \in
-  domain(\rho)$  for some  closure in  the heap,  we set  $\rho(x)$ to
-  $\bot$ iff $x.\epsilon \notin \Lv'$ corresponding to the closure.
+  these  environments  we  can   calculate  a  corresponding  liveness
+  environment using the $e\ and\ \sigma$.
+
+  \item {\color{Myblue}\emph{GC} for  each $x \in domain(\rho)$, we  set $\rho(x)$ to
+    $\bot$  iff $x.\epsilon  \notin \Lv$.   Similarly, for  each stack
+    entry   $(\rho,\ell,e,\sigma)$  in   $S$   with   $\Lv'$  as   the
+    corresponding liveness  environment in $\mathsf{SL}$,  and finally
+    for  each  $x  \in  domain(\rho)$ for  some  closure  $\langle  s,
+    \lfloor\rho\rfloor_{FV(s)}, \sigma\rangle$ in the heap, for all $x
+    \in  FV(s)$ we  set $x$  to  $\bot$ iff  $x.\epsilon \notin  \Lv'$.}
 
@@
     liveness environment  \Lv\ for  the expression $e$.   We similarly
-    define the liveness environments for the expressions $e$ in each of
-    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
-    $S$, giving a stack of  liveness environments that we shall denote
+    define the liveness  environments for the expressions  $e$ in each
+    of  the continuation  frames  $(\rho, \ell,  e,  \sigma)$ in  $S$,
+    giving  a stack  of  liveness environments  that  we shall  denote
     $\mathsf{SL}$.
   \item  \emph{Garbage collection  of objects  pointed by  root sets:}
-    Consider each of the following environments: (i) The 
-    the environment $\rho$ in the current state (ii) and for each $x \in
+    Consider  each   of  the  following  environments:   (i)  The  the
+    environment $\rho$ in  the current state (ii) and for  each $x \in
     domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
-    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
-    in $S$ 
-    with  $\Lv'$   as  the   corresponding  liveness   environment  in
+    \Lv$.  Similarly,  for each stack entry  $(\rho,\ell,e,\sigma)$ in
+    $S$  with  $\Lv'$ as  the  corresponding  liveness environment  in
     $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
@@
 
-Consider each of the following environments: (i) The 
-    the environment $\rho$ in the current state and for each $x \in
-    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
-    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
-    in $S$ 
-    with  $\Lv'$   as  the   corresponding  liveness   environment  in
-    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
-    to $\bot$ iff $x.\epsilon \notin \Lv'$.
+Consider each of  the following environments: (i)  The the environment
+$\rho$ in the current state and  for each $x \in domain(\rho)$, we set
+$\rho(x)$ to $\bot$ iff $x.\epsilon  \notin \Lv$.  Similarly, for each
+stack  entry   $(\rho,\ell,e,\sigma)$  in  $S$  with   $\Lv'$  as  the
+corresponding liveness  environment in $\mathsf{SL}$, and  for each $x
+\in domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
+\Lv'$.
 
@@
 $n$ leading  to this  state that  the next  transition cannot  enter a
-\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\,    (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
-(\print~\ans)):[\;]_{S}, 
-[\;]_{\heap},\, (\mainpgm))$.   Since  the
-            call to $GC$  in this state does  nothing, we just
-            have to show that the $\rightsquigarrow$ transition cannot
-            enter  a \bang\  state.  Since  our programs  are in  ANF,
-            $e_{\mainpgm}$ can  only be  a $\LET$ expression.   A {\sc
-              let}  step  does  not involve  dereferencing,  and  thus
-            cannot result in a \bang.
+\bang\  state.    When  $n$   is  0,   the  state   is  $([\;]_\rho,\,
+(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
+        [\;]_{\heap},\, (\mainpgm))$.  Since the  call to $GC$ in this
+        state  does   nothing,  we  ju1st   have  to  show   that  the
+        $\rightsquigarrow$  transition cannot  enter  a \bang\  state.
+        Since our  programs are in  ANF, $e_{\mainpgm}$ can only  be a
+        $\LET$  expression.   A  {\sc   let}  step  does  not  involve
+        dereferencing, and thus cannot result in a \bang.
 
@@
 $\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
-inserted  by  the  immediately  preceding garbage  collection  or  the
-GC of an earlier minefield step.  However the
-  demand $\sigma'$ on basis of  which an earlier collection would have
-  inserted a $\bot$, would have  included the current demand $\sigma$.
-  Thus, if  the earlier garbage  collection had inserted a  $\bot$, so
-  would the current GC.  Thus it is enough to consider
-  the $\rightsquigarrow$  along with the  current GC and show  that it
-  does not lead to a \bang.
+inserted by the immediately preceding  garbage collection or the GC of
+an earlier minefield  step.  However the demand $\sigma'$  on basis of
+which an earlier  collection would have inserted a  $\bot$, would have
+included the  current demand $\sigma$.   Thus, if the  earlier garbage
+collection had inserted a $\bot$, so would the current GC.  Thus it is
+enough to  consider the $\rightsquigarrow$  along with the  current GC
+and show that it does not lead to a \bang.
 

### 2016-02-13 paper/paper.tex
proof writing

@@
 %\newcommand{\added}[1]{#1}
-\newcommand{\todelete}[1]{{\color{Myred}{#1}}}
+\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
 
@@
 \leftarrow~s~\IN~e)$.
-{\color{Myblue}
-As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
-reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$
-corresponds to a closure-free path  in \heap\ starting from $\rho(x)$.
-If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
-the liveness of heap contents starting from $\mathit{x_{\pi}}$
-is given by $\Lv_\mathit{x_{\pi}}$.
-
- As we  shall see in section??,  the liveness of the  closure variable
- $x_{\pi}$ is recorded along with the closure so that the GC can refer
- to it during garbage collection.
-}
+
+{\color{Myblue}  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
+  liveness of the reference $\heap\llbracket\rho(x), \alpha\rrbracket$
+  only   if   $\alpha$  corresponds   to   a   closure-free  path   in
+  \heap\ starting  from $\rho(x)$.  If  this path is intercepted  by a
+  closure,  say  $(\CAR~{y_{\pi}})$, then  the  liveness  of the  path
+  starting from $\mathit{y_{\pi}}$ is given by $\Lv_\mathit{y_{\pi}}$.
+  As we shall see in  Section~\ref{sec:GC-scheme}, the liveness of the
+  closure variable  $y_{\pi}$ is recorded  along with the  closure for
+  $s$ so that the GC can refer to it during garbage collection.  }
 %} %comment
@@
 We shall now present a proof of soundness of the analysis presented in
-Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
-proof.
+Section~\ref{sec:liveness-analysis}. {\color{Myblue} It is easy to see that the analysis
+correctly identifies the liveness of stack variables. A stack variable
+is live
+between its introduction through a \LET\  and its last use to
+create a closure variable. This is correctly captured by the
+\LET\ rule in Figure~\ref{fig:live-judge}.}
+
+Proving soundness  for closure  variables is more complex. Here  are the
+ideas behind the proof. 
 \begin{enumerate}
@@
   Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
-  of  each  \LET\  expression.   {\color{Myblue}Note  that  memory  is
-    allocated only during execution  of \LET\ expressions}. During GC,
-  we  track each  reference  in the  root  set and  heap  that is  not
-  declared  live by  our analysis.   Any attempt  to dereference  such
-  references results in the transition system entering a special state
-  denoted   \bang.   We   call  the   semantics  after   augmentation,
-  \emph{minefield semantics}.
+  of  each \LET\  expression.  {\color{Myblue}Note  that, unlike  eager
+    languages,   memory  is   allocated  only   during  execution   of
+    \LET\ expressions}. During GC, we track each reference in the root
+  set and heap that is declared  dead by our analysis.  Any attempt to
+  dereference  such  references  results   in  the  transition  system
+  entering a special state denoted \bang.  We call the semantics after
+  augmentation, \emph{minefield semantics}.
 
@@
 \item \label{inline1} Assuming that a program enters the \bang\ state,
-  we  show  how to  construct,  through  inline expansion,  a  program
+  we  construct,  through  inline expansion,  a  program
   without function  calls which has  the same minefield  behavior. The
@@
 \hline
+\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
+  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
+    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
+    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
+{\sc let} \\ 
+\hline
 \end{tabular}
 \end{center}
-\caption{Minefield semantics for {\sc car-bang} and
-  {\sc car-clo} \warning{Add a note mentioning that we handle the issue of requiring a specific demand by introducing a new symbol}\label{fig:minefield-semantics-for-some}}
+\caption{Minefield  semantics for  {\sc  car-bang}  and {\sc  car-clo}
+  \warning{Add a note mentioning that we handle the issue of requiring
+    a      specific     demand      by      introducing     a      new
+    symbol}\label{fig:minefield-semantics-for-some}}
 \end{figure*}
@@
   to non-live references being replaced by $\bot$.  This simulates the
-  act of  not retaining  (effectively garbage collecting)  the cells
-  pointed to by these references  during an actual GC.
-  This is done in the following way:
+  act of  garbage collecting  the cells
+  pointed to by these references  during an actual garbage collection.
+  To do so, $GC$ needs to consider the following environments:  (1) the  environment in  the
+  current state, (2) the  environment in each of the stacked  continuations and (3)
+  the environment  in each of  the closures in  the heap.
+
  \begin{enumerate} 
-
- \item  Consider  the following  environments,  1)  environment in  the
-  current state, 2)  environments in each of the stacked  items and 3)
-  the environment  in each of  the closures in  the heap. For  each of
-  these  environments  we  can   calculate  a  corresponding  liveness
-  environment using the $e\ and\ \sigma$.
-
-  \item {\color{Myblue}\emph{GC} for  each $x \in domain(\rho)$, we  set $\rho(x)$ to
-    $\bot$  iff $x.\epsilon  \notin \Lv$.   Similarly, for  each stack
-    entry   $(\rho,\ell,e,\sigma)$  in   $S$   with   $\Lv'$  as   the
-    corresponding liveness  environment in $\mathsf{SL}$,  and finally
-    for  each  $x  \in  domain(\rho)$ for  some  closure  $\langle  s,
-    \lfloor\rho\rfloor_{FV(s)}, \sigma\rangle$ in the heap, for all $x
-    \in  FV(s)$ we  set $x$  to  $\bot$ iff  $x.\epsilon \notin  \Lv'$.}
-
-  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
-    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
-    liveness environment  \Lv\ for  the expression $e$.   We similarly
-    define the liveness  environments for the expressions  $e$ in each
-    of  the continuation  frames  $(\rho, \ell,  e,  \sigma)$ in  $S$,
-    giving  a stack  of  liveness environments  that  we shall  denote
-    $\mathsf{SL}$.
-  \item  \emph{Garbage collection  of objects  pointed by  root sets:}
-    Consider  each   of  the  following  environments:   (i)  The  the
-    environment $\rho$ in  the current state (ii) and for  each $x \in
-    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
-    \Lv$.  Similarly,  for each stack entry  $(\rho,\ell,e,\sigma)$ in
-    $S$  with  $\Lv'$ as  the  corresponding  liveness environment  in
-    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
-    to $\bot$ iff $x.\epsilon \notin \Lv'$.
-
-Consider each of  the following environments: (i)  The the environment
-$\rho$ in the current state and  for each $x \in domain(\rho)$, we set
-$\rho(x)$ to $\bot$ iff $x.\epsilon  \notin \Lv$.  Similarly, for each
-stack  entry   $(\rho,\ell,e,\sigma)$  in  $S$  with   $\Lv'$  as  the
-corresponding liveness  environment in $\mathsf{SL}$, and  for each $x
-\in domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
-\Lv'$.
-
-  \item \emph{Garbage  collection of  objects in  the heap:}
-
+{\color{Myblue}
+
+ \item \label{env}  For each of  these environments, $GC$  calculates a
+   liveness  environment $\Lv$  using  the  corresponding  $e$  (or  $s$)  and
+   $\sigma$.
+ \item 
    For each
-    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
-    environment $\rho$ in either  the  current state or in a
-    continuation frame, for each $x \in domain(\rho)$ and each forward
+    location $\ell$,  $GC$  sets $\heap(\ell)$  to $\bot$ iff for each
+    environment $\rho$ above, for each $x \in domain(\rho)$, and each forward
     access
-    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
-    \rrbracket = \ell$. Here 
+    path $\alpha$, it is \emph{not} the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
+    \rrbracket = \ell$. 
 %% $x.\alpha$  is  live  and
@@
 %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
-\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
-step~\ref{gc-1}.
+}
+
+  %% \item {\color{Myblue}\emph{GC} for  each $x \in domain(\rho)$, we  set $\rho(x)$ to
+  %%   $\bot$  iff $x.\epsilon  \notin \Lv$.   Similarly, for  each stack
+  %%   entry   $(\rho,\ell,e,\sigma)$  in   $S$   with   $\Lv'$  as   the
+  %%   corresponding liveness  environment in $\mathsf{SL}$,  and finally
+  %%   for  each  $x  \in  domain(\rho)$ for  some  closure  $\langle  s,
+  %%   \lfloor\rho\rfloor_{FV(s)}, \sigma\rangle$ in the heap, for all $x
+  %%   \in  FV(s)$ we  set $x$  to  $\bot$ iff  $x.\epsilon \notin  \Lv'$.}
+
+  %% \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
+  %%   \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
+  %%   liveness environment  \Lv\ for  the expression $e$.   We similarly
+  %%   define the liveness  environments for the expressions  $e$ in each
+  %%   of  the continuation  frames  $(\rho, \ell,  e,  \sigma)$ in  $S$,
+  %%   giving  a stack  of  liveness environments  that  we shall  denote
+  %%   $\mathsf{SL}$.
+  %% \item  \emph{Garbage collection  of objects  pointed by  root sets:}
+  %%   Consider  each   of  the  following  environments:   (i)  The  the
+  %%   environment $\rho$ in  the current state (ii) and for  each $x \in
+  %%   domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
+  %%   \Lv$.  Similarly,  for each stack entry  $(\rho,\ell,e,\sigma)$ in
+  %%   $S$  with  $\Lv'$ as  the  corresponding  liveness environment  in
+  %%   $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
+  %%   to $\bot$ iff $x.\epsilon \notin \Lv'$.
+
+%% Consider each of  the following environments: (i)  The the environment
+%% $\rho$ in the current state and  for each $x \in domain(\rho)$, we set
+%% $\rho(x)$ to $\bot$ iff $x.\epsilon  \notin \Lv$.  Similarly, for each
+%% stack  entry   $(\rho,\ell,e,\sigma)$  in  $S$  with   $\Lv'$  as  the
+%% corresponding liveness  environment in $\mathsf{SL}$, and  for each $x
+%% \in domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
+%% \Lv'$.
+
+%%   \item \emph{Garbage  collection of  objects in  the heap:}
+
+%%     For each
+%%     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
+%%     environment $\rho$ in either  the  current state or in a
+%%     continuation frame, for each $x \in domain(\rho)$ and each forward
+%%     access
+%%     path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
+%%     \rrbracket = \ell$. Here 
+%% %% $x.\alpha$  is  live  and
+%% %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
+%% %%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
+%% %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
+%% \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
+%% step~\ref{gc-1}.
   \end{enumerate}
@@
 \hline
-\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\LET~x\leftarrow
+\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
   s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow

### 2016-02-13 paper/paper.tex
proof writing done, now for space saving

@@
 
-{\color{Myblue}  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
+\added{  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
   liveness of the reference $\heap\llbracket\rho(x), \alpha\rrbracket$
@@
 We shall now present a proof of soundness of the analysis presented in
-Section~\ref{sec:liveness-analysis}. {\color{Myblue} It is easy to see that the analysis
+Section~\ref{sec:liveness-analysis}. \added{It is easy to see that the analysis
 correctly identifies the liveness of stack variables. A stack variable
@@
 \LET\ rule in Figure~\ref{fig:live-judge}.}
-
 Proving soundness  for closure  variables is more complex. Here  are the
@@
   Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
-  of  each \LET\  expression.  {\color{Myblue}Note  that, unlike  eager
+  of  each \LET\  expression.  \added{Note  that, unlike  eager
     languages,   memory  is   allocated  only   during  execution   of
@@
 \end{center}
-\caption{Minefield  semantics for  {\sc  car-bang}  and {\sc  car-clo}
+\caption{Minefield  semantics for  \CAR\   and \LET.
   \warning{Add a note mentioning that we handle the issue of requiring
@@
 \item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
-  collection that returns $(\rho', S', \heap')$, where the changes in
+  collection that returns $(\rho', S', \heap')$. The changes in
   $\rho, S$ and  $\heap$ are due
@@
   pointed to by these references  during an actual garbage collection.
-  To do so, $GC$ needs to consider the following environments:  (1) the  environment in  the
+  To do so, $GC(\ldots)$ needs to consider the following environments:  (1) the  environment in  the
   current state, (2) the  environment in each of the stacked  continuations and (3)
@@
  \begin{enumerate} 
-{\color{Myblue}
-
- \item \label{env}  For each of  these environments, $GC$  calculates a
+\added{
+
+ \item \label{env}  For each of  these environments, $GC(\ldots)$  calculates a
    liveness  environment $\Lv$  using  the  corresponding  $e$  (or  $s$)  and
@@
    For each
-    location $\ell$,  $GC$  sets $\heap(\ell)$  to $\bot$ iff for each
+    location $\ell$,  $GC(\ldots)$  sets $\heap(\ell)$  to $\bot$ iff for each
     environment $\rho$ above, for each $x \in domain(\rho)$, and each forward
@@
 \end{enumerate}
-Starting from the initial state,
+Figure ~\ref{fig:minefield-semantics-for-some} shows some of the
+minefield rules. As mentioned earlier, the transition for a \LET\ 
 %%  $([\;]_\rho,([\;]_\rho,
 %% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
-every transition  is preceded  by a  GC
-using $GC$.  The details of the transition for the {\sc
-  car-clo}  rule   is  shown   in
-Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
-earlier  call to  $GC$  results  in $\rho(x)$  being
+is preceded  by $GC(\ldots)$.  The details of the transition for the {\sc
+  car-clo}  rule   is  also shown.  If    an
+earlier  call to  $GC(\ldots)$  results  in $\rho(x)$  being
 bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
@@
 call-less program,  the initial state of  the minefield
-semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
+semantics  is  assumed  to  be  $([\;]_\rho,(\rho_\mathit{init},
 \ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
-e_{\mainpgm}$.
+e_{\mainpgm})$.
 
@@
 Note that our proofs will be for a single round of minefield execution
-i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
-programs)  to   its  WHNF  driven  by   the  printing  mechanism~(Section
+i.e.  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
 \ref{sec:semantics}).  With  minor variations,  the proof will  also be
@@
 \begin{proof}
-Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
-step. Consider a state $(\rho, S,  \heap, e, \sigma)$ in the minefield
-execution of a  program.  We show by induction on  the number of steps
-$n$ leading  to this  state that  the next  transition cannot  enter a
-\bang\  state.    When  $n$   is  0,   the  state   is  $([\;]_\rho,\,
+Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
+minefield execution of a program.  We  show by induction on the number $n$
+of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
+enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
 (\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
-        [\;]_{\heap},\, (\mainpgm))$.  Since the  call to $GC$ in this
-        state  does   nothing,  we  ju1st   have  to  show   that  the
-        $\rightsquigarrow$  transition cannot  enter  a \bang\  state.
-        Since our  programs are in  ANF, $e_{\mainpgm}$ can only  be a
-        $\LET$  expression.   A  {\sc   let}  step  does  not  involve
-        dereferencing, and thus cannot result in a \bang.
+[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
+does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
+transition cannot  enter a  \bang\ state.  Since  our programs  are in
+ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
+does not involve dereferencing, and thus cannot result in a \bang.
 
@@
 $\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
-inserted by the immediately preceding  garbage collection or the GC of
-an earlier minefield  step.  However the demand $\sigma'$  on basis of
-which an earlier  collection would have inserted a  $\bot$, would have
-included the  current demand $\sigma$.   Thus, if the  earlier garbage
-collection had inserted a $\bot$, so would the current GC.  Thus it is
-enough to  consider the $\rightsquigarrow$  along with the  current GC
-and show that it does not lead to a \bang.
-
-We consider the rules {\sc car-clo} and {\sc car-bang} only.  
+inserted by an earlier $GC(\ldots)$.   However the demand $\sigma'$  on basis of
+which the $GC(\ldots)$ would have inserted a  $\bot$ would have
+included the  current demand $\sigma$.     Thus it is
+enough to  show that the $\rightsquigarrow$ step would not lead to a
+\bang, even if $GC(\ldots)$ had been done with the current demand
+$\sigma$.
+
+We consider the rules for \CAR\  only.  
 The rest  of the rules involve similar reasoning. For the {\sc
@@
 
-For the  {\sc car-bang}  rule, observe that  there are
+For the  {\sc car-1-clo}  rule, observe that  there are
 two dereferences.  First $x$  is dereferenced to  get a
@@
 %\subsection{Modifying liveness rules}
-\paragraph{Symbolic representation of operations}
-
+\paragraph{Symbolic representation of operations:}
 The   $\mathit{ref}$   rule   for   \CONS,   shown   in
@@
 %\subsection{Generating equations for $\Lfonly_{\mathit f}$}
-\paragraph{Computing function summaries $\Lfonly_{\mathit f}$}
+\paragraph{Computing function summaries $\Lfonly_{\mathit f}$:}
 Given a  function $\mathit{f}$, we now  describe how to
@@
 %\subsection{Generating liveness equations \Lv\  for function bodies}
-\paragraph{Handling user-defined functions}
+\paragraph{Handling user-defined functions:}
 \label{sec:bodylivenessbodies}
-
 To  avoid analyzing  the body  of a  function for  each
@@
 %\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
-\subsection{From liveness sets to context-free grammars}\label{sec:grammar-formulation}
-
+\paragraph{From liveness sets to context-free grammars:}\label{sec:grammar-formulation}      
 The  equations  above can  now  be  re-interpreted as  a  context-free

### 2016-02-13 paper/paper.tex
proof writing done, now for space saving

@@
 
-\begin{figure*}[t!]
+\begin{figure}[t!]
 \begin{center}\footnotesize
@@
 \hline
-
 \hline 
@@
 (\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
-s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,$\\ $\quad \heap,\, s, \, \cred{\sigma}$ }     &
 {\sc car-1-clo} 
@@
 \hline
-\makecell[t]{ $\heap(\rho(x))
-\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \rho(x),
-  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
-\cred{\renewcommand{\arraystretch}{1}\left\{
+\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
+\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
+  \rightsquigarrow$ \\
+  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S,$\\ $\quad \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
     \begin{array}{lr}
@@
     \end{array}\right.
-}$         
+  }$   }      
 &
@@
 \makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
-  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
+    s~\IN~e), \sigma)}$\\ $\cred{ = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
-    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
+    \mapsto \ell], S,$\\$ \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
     where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
@@
     symbol}\label{fig:minefield-semantics-for-some}}
-\end{figure*}
+\end{figure}
 
@@
 \begin{align*}
-  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} 
-\end{align*}
-\begin{align*}
+  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
+%\end{align*}
+%\begin{align*}
   \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
@@
 \begin{proof}
-Given a Turing machine and an input string, we shall define a function
-for   each  non-terminal   introduced  in   the  proof   of
-Lemma~\ref{lemma:grammar-undecidable}.
-%% Liveness  analysis  of  $f$  will  result in  not  only  contain  a
-%% production  $P$ itself,  but could  also contain  extra productions
-%% that necessarily contain the \clazy\ symbol.
-For instance, corresponding to the production $\nont{i}{\acar} \rightarrow
-\acar \nont{j}{\acar}\bcar$,
-the program is:
-
-\input{equivalent-program}
-
-The body of the function corresponds  to the RHS of the production. If
-there are  more than one  productions with the same  LHS non-terminal,
-then the function for the non-terminal joins the program fragments for
-the  RHS  with an  $\SIF$.   If  there  are  $n$ such  functions,  then
+Given a Turing machine and an input string, the  proof   of
+Lemma~\ref{lemma:grammar-undecidable} generates a grammar. 
+We shall define a function
+for   each  non-terminal $\nont{i}{c}$   introduced  in   this grammar
+such that the
+liveness  analysis  of  the function will  result
+in a set of productions that includes the productions for
+$\nont{i}{c}$ generated for the proof.
+As an example, it  can be verified that the grammar
+for the function shown in Figure~\ref{fig:turing-pgm}(b)    includes      the
+productions shown in Figure~\ref{fig:turing-pgm}(a). Here $\nont{\mathit{all}}{}$   corresponds   to  the   demand
+$\sigma_{\mathit{all}}$.
+
+The body of the function corresponds  to the RHS of the productions
+for $\nont{i}{c}$. Productions with the same  LHS non-terminal, but
+with differing non-terminal on RHS can be generated by 
+joining program fragments with $\SIF$.   If  there  are  $n$ such  functions,  then
 \mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
-distinct function in each branch. It  can be verified that the grammar
-for      the     function      shown      above,     includes      the
-productions shown below. $\nont{\mathit{all}}{}$   corresponds   to  the   demand
-$\sigma_{\mathit{all}}$:
+distinct function in each branch.
+
+\begin{figure}[t]
+  \begin{minipage}{.45\columnwidth}
     \begin{align*}
       \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
-\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{\mathit{all}}{} \\
+      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
       \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
@@
     \end{align*}
+  \end{minipage}
+  \begin{minipage}{.45\columnwidth}
+    \input{equivalent-program}
+  \end{minipage}\\
+  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
+  \caption{A possible grammar generated by the proof of
+    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
+  grammar.}\label{fig:turing-pgm}
+\vspace*{-5mm}
+\end{figure}
 
@@
 %------------------------------------------------------------%
-
 \begin{figure}[t!]
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(50,12)
+\begin{pspicture}(0,-3)(50,12)
   \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(30,12)
+\begin{pspicture}(0,-3)(30,12)
   \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(50,12) %\psframe(0,-5)(40,12)
+\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
@@
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(20,12) %\psframe(0,-5)(40,12)
+\begin{pspicture}(0,-5)(20,8) %\psframe(0,-5)(40,8)
   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
@@
   are in (c) and (d).}\label{fig:example-automata}
+\vspace*{-2mm}
 \figrule
@@
 
-%\subsection{A liveness-based GC scheme}
-\label{sec:live-clo}
+%\subsection{A liveness-based GC scheme} \label{sec:live-clo}
+
+%%\subsubsection{From CFGs to GC-ready DFAs}
+%%\label{sec:NFA-approx}
+
+We circumvent the problem of undecidability by over approximating the
+CFG by non-deterministic finite state automata (NFA) using
+%% The NFAs  are then simplified using the
+%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
+%%are converted to DFAs.
+%%
+%% We use  the algorithm by 
+Mohri  and Nederhof~\cite{mohri00regular} method.
+%to approximate a CFG by a {\em strongly regular\/} grammar.  
+For example, the grammar fragment for %the non-terminal
+$\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
+ \begin{eqnarray*}
+   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
+   \acdr\var{\Df{\length}{1}}
+   \mid \clazy\var{\Df{\length}{1}}\\
+   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
+   \mid \epsilon
+ \end{eqnarray*}
+
+The strongly regular grammar is converted  into a set of NFAs, one for
+each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
+done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
+pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
+constructing  the $\epsilon$-closure  till a  fixed point  is reached,
+after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  
+\cmt{ %% REMOVED FOR BLIND REVIEW
+  The details of the  algorithm, its correctness and  termination proofs are
+given  in~\cite{karkare07liveness,asati14lgc}. }
+The
+resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
+\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
+check if the source node of the edge has a path to a final 
+state.  If it does, we mark  the source node as final. Finally, we
+remove all the edges labeled \clazy\  and convert the automaton into a
+deterministic    automaton.    These steps   effectively   implement    the
+$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
+\clazy\ to  obtain forward access paths.  While checking for  liveness during
+garbage  collection, a  forward access path  is valid  only if  it can
+reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
+NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
+final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
+expectedly  says that for a  demand  $\sigma_{\length}$,  the
+liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
+traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
+NFA     for     \var{\Lanv{\pa}{}}.      The     DFA in
+Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
+paths, reflecting  the  lazy  nature  of  our  language.   Since
+\length\ does not  evaluate the elements of  the argument list,
+the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
+liveness-based GC triggers beyond $\pi_9$.
+
   \SetStartEndCondition{ }{}{}%
@@
 
-%%\subsubsection{From CFGs to GC-ready DFAs}
-%%\label{sec:NFA-approx}
-
-We circumvent the problem of undecidability by over approximating the
-CFG by non-deterministic finite state automata (NFA) using
-%% The NFAs  are then simplified using the
-%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
-%%are converted to DFAs.
-%%
-%% We use  the algorithm by 
-Mohri  and Nederhof~\cite{mohri00regular} method.
-%to approximate a CFG by a {\em strongly regular\/} grammar.  
-For example, the grammar fragment for %the non-terminal
-$\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
- \begin{eqnarray*}
-   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
-   \acdr\var{\Df{\length}{1}}
-   \mid \clazy\var{\Df{\length}{1}}\\
-   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
-   \mid \epsilon
- \end{eqnarray*}
-
-The strongly regular grammar is converted  into a set of NFAs, one for
-each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
-done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
-pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
-constructing  the $\epsilon$-closure  till a  fixed point  is reached,
-after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  
-\cmt{ %% REMOVED FOR BLIND REVIEW
-  The details of the  algorithm, its correctness and  termination proofs are
-given  in~\cite{karkare07liveness,asati14lgc}. }
-The
-resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
-\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
-check if the source node of the edge has a path to a final 
-state.  If it does, we mark  the source node as final. Finally, we
-remove all the edges labeled \clazy\  and convert the automaton into a
-deterministic    automaton.    These steps   effectively   implement    the
-$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
-\clazy\ to  obtain forward access paths.  While checking for  liveness during
-garbage  collection, a  forward access path  is valid  only if  it can
-reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
-NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
-final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
-expectedly  says that for a  demand  $\sigma_{\length}$,  the
-liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
-traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
-NFA     for     \var{\Lanv{\pa}{}}.      The     DFA in
-Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
-paths, reflecting  the  lazy  nature  of  our  language.   Since
-\length\ does not  evaluate the elements of  the argument list,
-the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
-liveness-based GC triggers beyond $\pi_9$.
-
 \newcommand{\hgt}{2.9cm}

### 2016-02-13 paper/paper.tex
nearly there -- 12 pages

@@
-\documentclass[9pt]{sigplanconf}
+\documentclass[9pt,preprint]{sigplanconf}
 \usepackage{etex}
@@
 garbage  collectors   in  reclaiming   more  garbage.    The  specific
-contributions of this paper are
+contributions of this paper are:
 \added{
@@
 future uses of the value of an expression.  The demand on an expression
-$e$   is   again   a   set    of   access   paths---that   subset   of
+$e$   is   also   a   set    of   access   paths---the   subset   of
 $\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
@@
 This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
-y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
+y.\acdr\acar  \rbrace$.   These are  the $y$-rooted  accesses
 required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
@@
 
-\begin{figure}[t!]
+\begin{figure*}[t!]
 \begin{center}\footnotesize
 \renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|@{}c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|}
+\begin{tabular}{|@{}c@{\ }|@{\ }c@{}|@{\ }c@{\ }|}
 \hline
@@
 \makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
-(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
-s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,$\\ $\quad \heap,\, s, \, \cred{\sigma}$ }     &
+(\CAR~x), \cred{\sigma} \rightsquigarrow$ $ \rho', \,(\rho, addr(\langle
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S, \heap,\, s, \, \cred{\sigma}$ }     &
 {\sc car-1-clo} 
@@
 \makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
-  \rightsquigarrow$ \\
-  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S,$\\ $\quad \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
+  \rightsquigarrow$ 
+  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
     \begin{array}{lr}
@@
 \makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
-    s~\IN~e), \sigma)}$\\ $\cred{ = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
-    \mapsto \ell], S,$\\$ \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
+    s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x
+    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
     where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
@@
     symbol}\label{fig:minefield-semantics-for-some}}
-\end{figure}
+\vspace*{-2mm}
+\end{figure*}
 
@@
   grammar.}\label{fig:turing-pgm}
-\vspace*{-5mm}
+\vspace*{-6mm}
 \end{figure}
@@
 %\\ (a) & (b)
-\end{tabular}%\vskip -10mm
+\end{tabular}
+\vspace*{-8mm}
  \caption{Memory usage.  
-%% Column (a) shows complete  usage, column 
-%%    (b) enlarges  a part  of the  usage.  
 The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
@@
 \label{fig:memory-usage} \figrule
+\vspace*{-1mm}
 \end{figure*}
 
-\vspace*{-2.5mm}
+\vspace*{-6.5mm}
 \section{The GC scheme}

### 2016-02-13 paper/paper.tex
in 12 pages...
fixed for compliance with ISMM
spell checks

@@
-\documentclass[9pt,preprint]{sigplanconf}
+\documentclass[9pt,preprint,letter,nonatbib]{sigplanconf}
 \usepackage{etex}
@@
 \definecolor{Myred}{rgb}{1,.2,.2}
+
 \newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
-\newcommand{\added}[1]{{\color{Myblue}{#1}}}
-%\newcommand{\added}[1]{#1}
+%\newcommand{\added}[1]{{\color{Myblue}{#1}}}
+\newcommand{\added}[1]{#1}
 \newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
-
 %\newcommand{\warning}[1]{{\color{Myred}{#1}}}
@@
 %\newcommand{\blankout}[1]{}
+
 \newcommand{\oxlb}{\ensuremath{[\!|}}
@@
 \begin{document}
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
 %% \linenumbers 
@@
 \include{autotable}
-\vskip -5mm
+\vskip -7mm
 \end{table*}
@@
 Java  programs. In~\cite{ran.shaham-sas03},  temporal properties  like
-liveness  are checked  against an  automaton modelling  heap accesses.
+liveness  are checked  against an  automaton modeling  heap accesses.
 Both these approaches are intraprocedural in scope.
@@
 which touches upon only basic techniques of compile-time garbage
-marking, explicit deallocation and desctructive allocation. An
+marking, explicit deallocation and destructive allocation. An
 interesting approach suggested in~\cite{HofmannJ03} is to annotate the

### 2016-02-15 paper/paper.tex
minor cleanups
minor rewrite

@@
-\documentclass[9pt,preprint,letter,nonatbib]{sigplanconf}
+\documentclass[9pt,preprint,nonatbib]{sigplanconf}
 \usepackage{etex}
@@
 
-\section{Computing liveness and its encoding as DFA}\label{sec:computing}
+\section{Towards a  computable form of liveness}\label{sec:computing}
 %\section{Generating and storing liveness information}\label{sec:computing}
-Section~\ref{sec:liveness}  gave   a  fully context-sensitive
-liveness analysis which 
-%$\mathcal{L}$ and $ref$ together
-described the  liveness set for  each variable  in a function  body in
-terms of  a symbolic demand  $\sigma$ and \Lfonly.  However,  we are yet
-to describe  (i) how to obtain demand  transformers \Lfonly\ from
-the  rule {\sc  live-define}, and  (ii)  how to  compute the  concrete
-demand $\sigma$  on each function.   To do so,  we first need to  modify the
-liveness rules to a different form.
+The analysis in Section~\ref{sec:liveness} is fully context-sensitive, 
+describing the  liveness sets  in a function body in
+terms of a symbolic demand $\sigma$  and \Lfonly. However, we are yet to
+describe (i) how to obtain demand transformers \Lfonly\ from the rule
+{\sc live-define} , and (ii) how to compute the concrete demand $\sigma$ on
+each function. To do so, we first need to modify the liveness rules
+to a slightly different form.
 
@@
 
-\vspace*{-6.5mm}
+%\vspace*{-6.5mm}
 \section{The GC scheme}
@@
 \begin{figure*}[b!]
-\noindent{\Large \bf A. Minefield semantics (Not to be included in the paper) }
+\noindent{\Large \bf A. Complete minefield semantics}
 \begin{center}\footnotesize

### 2016-03-25 paper/paper.tex
typos fixed, layout fixed, minor mistakes fixed AFTER ISMM REVIEWS

@@
 $\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
-before  the  expression  $e$. The label is not part of the syntax.},  i.e. just  before  the  evaluation  of
+before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
 $(\length\ \pz)$.  The heap is represented  by a graph in which a node
@@
   whether to copy a cell during  GC translates to a membership problem
-  for  such  grammars.  Earlier  assumed  the  undecidability of  this
-  membership question  and used an over-approximation  to overcome it.
-  In this  paper, we provide a  formal proof of the  undecidability of
-  this problem.
+  for such  grammars.  Earlier research assumed  the undecidability of
+  this membership question and  used an over-approximation to overcome
+  it.  In this paper, we provide  a formal proof of the undecidability
+  of this problem.
 
@@
 
-%% The model use the notion of  {\em access paths}, i.e.
+%% The model use the notion of  {\em access paths}, i.e.,
 %% a  prefix-closed set of strings  over $\{\acar,\acdr\}^\ast$, in
@@
 
-\input{example-program}
-%%\label{fig:mot-example2}
-
 %==============================================================
@@
 
+\begin{figure}[t]
+  \input{example-to-illustrate-liveness}  
+  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
+\end{figure}
+\input{fig-liveness}
+
 \subsection{Liveness analysis for lazy languages}
@@
 
-Consider the  program in Figure~\ref{fig:mot-example2-a}.   As mentioned
+Consider the  program in Figure~\ref{fig:lv-closure}.   As mentioned
 earlier, a lazy evaluation of the $\LET$ expression at $\pi_1$ creates
@@
 
-\begin{figure}[t]
-  \input{example-to-illustrate-liveness}  
-  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
-\end{figure}
-\input{fig-liveness}
-
 Closure  variables  get  their  liveness values  through  a  chain  of
 dependences beginning  at a  variable at an  evaluation point.   As an
-example, consider the program in Figure~\ref{fig:lv-closure}.
+example,  consider again  the program  in Figure~\ref{fig:lv-closure}.
 The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
@@
 \leftarrow            (+~\py~1),\,           \py            \leftarrow
-(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
-heap are a  runtime representation of these  dependences.  Since $\pz$
-is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
-demand made by the calling context(s)  of $f$ places a demand on $\pz$
-which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
-chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
-\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
-$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
-\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
-described in this section declares the liveness of $\px_{\pi_1}$ to be
-a union of  the liveness arising from these dependence  chains.  To be
-safe, a GC during  evaluation of $\py$ at $\psi_1$ has
-to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
-notice that if a GC takes place while evaluating $\pz$
-at $\psi_3$, it can safely consider only the liveness arising from the
-dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
-(+~\py~1),\, \py \leftarrow (\length~\px_{\pi_1})\rbrack$.  The garbage
-collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
-generalization of this observation  to dynamically select a evaluation
-point specific liveness in order to collect more garbage.
+(\length~\px_{\mathit{\pi}_1})\rbrack$.    Indeed,   the   chains   of
+closures  in   the  heap  are   a  runtime  representation   of  these
+dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
+expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
+of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
+$\px_{\pi_1}$.  Other dependence chains which result in a liveness for
+$\px_{\pi_1}$    are    $\lbrack    \psi_1\!\!\!:\!\!\py    \leftarrow
+(\length~\px_{\pi_1})\rbrack$   and  $\lbrack   \psi_2\!\!\!:\!\!\!\pw
+\leftarrow          (/~\pu~\py),\,\,           \py          \leftarrow
+(\length~\px_{\pi_1})\rbrack$.   The  liveness analysis  described  in
+this section declares  the liveness of $\px_{\pi_1}$ to be  a union of
+the liveness arising  from these dependence chains.  To be  safe, a GC
+during evaluation  of $\py$ at  $\psi_1$ has  to use this  liveness to
+copy the heap starting from  $\px_{\pi_1}$.  However, notice that if a
+GC  takes place  while evaluating  $\pz$  at $\psi_3$,  it can  safely
+consider only the liveness arising from the dependence chain\linebreak
+$\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow (+~\py~1),\, \py \leftarrow
+(\length~\px_{\pi_1})\rbrack$.    The    garbage   collection   scheme
+described in Section~\ref{sec:GC-scheme} uses a generalization of this
+observation  to  dynamically  select   an  evaluation  point  specific
+liveness in order to collect more garbage.
 
@@
 Note that our proofs will be for a single round of minefield execution
-i.e.  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
+i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
 \ref{sec:semantics}).  With  minor variations,  the proof will  also be
@@
 similar reasons.
+
+\input{example-program}
 
@@
 \end{tabular}
-\vspace*{-8mm}
+\vspace*{-2mm}
  \caption{Memory usage.  
@@
 
-We shall call a  unit allocatable memory as a {\em  cell}.  A cell can
+We shall call a  unit of allocatable memory as a {\em  cell}.  A cell can
 hold    a    basic    value    ($\mathsf{bas}$),    the    constructor
@@
 collection scheme.  Starting with the root set, each cell pointed by a
-live reference  (i.e. whose associated  DFA state is final)  is copied
+live reference  (i.e., whose associated  DFA state is final)  is copied
 using $\mathsf{copy}$.   Copying a \CONS\ cells  just involves copying
@@
 \label{tab:exp-results}
+\vskip -2mm
 \centering
@@
 
-Memory  usage graphs  for select  benchmarks are  shown
-Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
-number of  GCs  tend to be  very large,
-hence  we  show only  a  slice of the original graph  with 
-representative    behavior  (Column (b)).  In all the programs we can see
-that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
-dips   below the RGC curve 
-(\RGCLine\ line) during GC.  The graphs also include the
-curve    for   reachable    cells   (\ReachLine)  and live cells
-(\UseLine). These were  obtained
- by forcing  RGC to  run at   very  high
-frequency. The curve for live cells were obtained by recording heap
-access times and post processing the data 
-at the end of the program. Note that the size of an LGC cell is 1.16
-times the size of a RGC cell.
+Memory    usage    graphs    for     the    benchmarks    are    shown
+Figure~\ref{fig:memory-usage}.  In  all the  programs we can  see that
+the curve  corresponding to  LGC (\LGCLine\ line)  dips below  the RGC
+curve (\RGCLine\ line)  during GC.  The graphs also  include the curve
+for reachable cells (\ReachLine) and live cells (\UseLine). These were
+obtained by forcing  RGC to run at very high  frequency. The curve for
+live  cells were  obtained by  recording  heap access  times and  post
+processing the data at  the end of the program. Note  that the size of
+an LGC cell is 1.16 times the size of a RGC cell.
 

### 2016-04-12 paper/paper.tex
space savings

@@
-\documentclass[9pt,preprint,nonatbib]{sigplanconf}
+\documentclass[9pt,draft,nonatbib]{sigplanconf}
 \usepackage{etex}
@@
 \label{sec:motiv}
-\input{fig_motiv}   Figure~\ref{fig:mot-example}   shows  an   example
+\input{fig_motiv}   
+%%
+Figure~\ref{fig:mot-example}   shows  an   example
 program and the state of the heap at the program point $\pi$\footnote{We write
@@
 membership of  a CFG augmented  with a
-fixed set of unrestricted productions.
-
-\begin{definition}\label{def:specialgrammar}
-Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
-$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
-\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
-context-free    productions     that    contains    the
-distinguished  production   $S  \rightarrow  \alpha\$$,
-$\alpha$ is a  string of grammar symbols  that does not
-contain $S$, and $p_2$ is the fixed set of unrestricted
-productions    $\bcar\acar    \rightarrow    \epsilon$,
-$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
-\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
-and $\clazy\$ \rightarrow \$ $.
-\end{definition}
-
-%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
-%% that the results of liveness analysis of any program can be modeled by
-%% the kind  of grammar described above. The  following proposition shows
-%% that the converse also holds.
-%% \begin{proposition}
-%% Given    a     grammar    $G$    of    the     form    described    in
-%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
-%% program $p$  with program points  $\pi_i$ and variables $x$  such that
-%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
-%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
-%% \end{proposition}
-We first show that membership problem of the class of grammars in
-Definition~\ref{def:specialgrammar} is undecidable. However note that
-the grammars may not be necessarily generated from liveness analysis.
-\newcommand{\state}{\ensuremath{\mathsf{S}}}
-\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
-\begin{lemma}\label{lemma:grammar-undecidable}
-Given a grammar    $G$    of   the    kind    described    in
-Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
-consisting  of symbols \acar\  and \acdr\  only, the membership problem
-$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
-\end{lemma}
-
-\begin{proof}
-Given a Turing machine and  an input $w\in (1+0)^*$, we
-construct a grammar $G$ such that the machine will halt
-on    $w$    if    and   only    if    $\$    \in
-\mathscr{L}(G)$. The grammar includes  the fixed set of
-unrestricted               productions               in
-Definition~\ref{def:specialgrammar}.
-
-We shall represent a Turing machine (TM) configuration 
-as $w_l(\state,c)w_r$, where $w_l$ is the string to the
-left of the head, $w_r$ is the string to the right, $c$
-is the symbol under the head and \state\ is the current
-state of  the TM.   For each combination  of state
-and symbol  $(\state,c)$, the grammar will  contain the
-non-terminal \nont{}{c}. We shall synchronize each move
-of the TM to  a derivation  step using  a context
-free production, followed, if possible, by a derivation
-step using either  $\bcar\acar \rightarrow \epsilon$ or
-$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
-synchronization,  we  shall   establish  the  following
-invariant  relation between  the TM  configuration
-and the sentential form:
-
-\begin{quote}
-  If    the   configuration    of   the    TM   is
-  $w_l(\state,c)w_r$, then the  sentential form will be
-  $\overline{w}_l\nont{}{c}\,w_r\$ $,                where
-  $\overline{w}_l$ is  the same as $w_l$  but with each
-  symbol $d$ in $w_l$ replaced by $\overline{d}$.
-\end{quote}
-
-Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
-$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
-form    corresponding     to    the    initial     configuration    is
-$\nont{\mathit{init}}{c}w\$$ (we  can assume that there  is a production
-$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w\$$, where  \nont{}{} is
-the start symbol of the  grammar). Further correspondences between the
-TM moves and the grammar productions are as follows:
-
-\begin{enumerate}
-\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
-  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
-  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
-\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
-  are two productions
-  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
-  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
-\end{enumerate}
-The idea  behind the  productions is explained  with an
-example:  Assume that  the current  sentential form  is
-$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
-TM  has  a   transition  $(S_i,\acar)  \rightarrow
-(S_j,\acdr, L)$.  Since the next corresponding  step in
-the  derivation  has  to  be  done  without  any  prior
-knowledge of whether the symbol to the left of the tape
-is a \acar\  or a \acdr, two  productions are provided,
-and  the  invariant  will  be maintained  only  if  the
-production         $\nont{i}{\acar}         \rightarrow
-\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
-in  the  derivation.    This  gives  the  configuration
-$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
-Simplification   with    the   production   $\bcdr\acdr
-\rightarrow               \epsilon$              yields
-$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
-corresponds  to   the  changed  configuration   of  the
-TM.  Notice  carefully that a wrong  choice breaks
-the invariant  and it cannot be  recovered subsequently
-by any choice of productions.
-
-After the  TM has  halted, there  are further  ``cleanup'' derivations
-that derive  $\epsilon$ only if  the invariant has been  maintained so
-far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
-$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
-final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
-\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
-$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
-for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
-and                $\nont{\mathit{final}}{c}               \rightarrow
-\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
-\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
-\acdr\ on the right of the tape head.  This completes the
-reduction.\!
-\end{proof}
-
-We now show that the proof can be replayed for the class of grammars
-generated from liveness analysis of programs.
-
-\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
-Given   a    grammar    $G$    of   the    kind    described    in
-Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
-analysis  of a  program  and  a forward  access  path $\alpha$,    the
-membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
-\end{lemma}
-
-\begin{proof}
-Given a Turing machine and an input string, the  proof   of
-Lemma~\ref{lemma:grammar-undecidable} generates a grammar. 
-We shall define a function
-for   each  non-terminal $\nont{i}{c}$   introduced  in   this grammar
-such that the
-liveness  analysis  of  the function will  result
-in a set of productions that includes the productions for
-$\nont{i}{c}$ generated for the proof.
-As an example, it  can be verified that the grammar
-for the function shown in Figure~\ref{fig:turing-pgm}(b)    includes      the
-productions shown in Figure~\ref{fig:turing-pgm}(a). Here $\nont{\mathit{all}}{}$   corresponds   to  the   demand
-$\sigma_{\mathit{all}}$.
-
-The body of the function corresponds  to the RHS of the productions
-for $\nont{i}{c}$. Productions with the same  LHS non-terminal, but
-with differing non-terminal on RHS can be generated by 
-joining program fragments with $\SIF$.   If  there  are  $n$ such  functions,  then
-\mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
-distinct function in each branch.
-
-\begin{figure}[t]
-  \begin{minipage}{.45\columnwidth}
-    \begin{align*}
-      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
-      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
-      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
-      \acdr\nont{\mathit{all}}{}
-    \end{align*}
-  \end{minipage}
-  \begin{minipage}{.45\columnwidth}
-    \input{equivalent-program}
-  \end{minipage}\\
-  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
-  \caption{A possible grammar generated by the proof of
-    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
-  grammar.}\label{fig:turing-pgm}
-\vspace*{-6mm}
-\end{figure}
-
-Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
-had a single non-terminal on the RHS. There are two characteristics of
-the grammar produced from liveness-analysis  that are are relevant for
-replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
-\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
-grammar  generated from  liveness  analysis of the constructed program
-will  have the  production
-$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
-\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
-$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
-\nont{\mathit{all}}{}$.
-
-It is clear that if the TM given as an instance of the halting problem
-accepts the input string, then the earlier derivation can be replayed,
-every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
-sentential form by $\epsilon$. However, if  the TM does not accept the
-input string, then every sentential form derived from the start symbol
-would    have     at    least    one    non-terminal from the grammar
-of Lemma~\ref{lemma:grammar-undecidable} that is different
-from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
-derivable from the grammar.
-\end{proof}
-
+fixed set of unrestricted productions. As shown in Appendix~\ref{sec:undec}, the  membership problem of this class of grammars is undecidable.
 %------------------------------------------------------------%
@@
 
+\appendix
+\section{Proof of Undecidability}
+\label{sec:undec}
+We show that  membership problem  of the  class of  grammars defined below 
+%%Definition~\ref{def:specialgrammar}
+is undecidable.   We then  show that  the grammars  in this  class are
+indeed generated from liveness analysis.
+\begin{definition}\label{def:specialgrammar}
+Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
+$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
+\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
+context-free    productions     that    contains    the
+distinguished  production   $S  \rightarrow  \alpha\$$,
+$\alpha$ is a  string of grammar symbols  that does not
+contain $S$, and $p_2$ is the fixed set of unrestricted
+productions    $\bcar\acar    \rightarrow    \epsilon$,
+$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
+\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
+and $\clazy\$ \rightarrow \$ $.
+\end{definition}
+
+%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
+%% that the results of liveness analysis of any program can be modeled by
+%% the kind  of grammar described above. The  following proposition shows
+%% that the converse also holds.
+%% \begin{proposition}
+%% Given    a     grammar    $G$    of    the     form    described    in
+%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
+%% program $p$  with program points  $\pi_i$ and variables $x$  such that
+%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
+%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
+%% \end{proposition}
+
+\newcommand{\state}{\ensuremath{\mathsf{S}}}
+\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}
+\begin{lemma}\label{lemma:grammar-undecidable}
+Given a grammar    $G$    of   the    kind    described    in
+Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
+consisting  of symbols \acar\  and \acdr\  only, the membership problem
+$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
+\end{lemma}
+
+\begin{proof}
+Given a Turing machine and  an input $w\in (1+0)^*$, we
+construct a grammar $G$ such that the machine will halt
+on    $w$    if    and   only    if    $\$    \in
+\mathscr{L}(G)$. The grammar includes  the fixed set of
+unrestricted               productions               in
+Definition~\ref{def:specialgrammar}.
+
+We shall represent a Turing machine (TM) configuration 
+as $w_l(\state,c)w_r$, where $w_l$ is the string to the
+left of the head, $w_r$ is the string to the right, $c$
+is the symbol under the head and \state\ is the current
+state of  the TM.   For each combination  of state
+and symbol  $(\state,c)$, the grammar will  contain the
+non-terminal \nont{}{c}. We shall synchronize each move
+of the TM to  a derivation  step using  a context
+free production, followed, if possible, by a derivation
+step using either  $\bcar\acar \rightarrow \epsilon$ or
+$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
+synchronization,  we  shall   establish  the  following
+invariant  relation between  the TM  configuration
+and the sentential form:
+
+\begin{quote}
+  If    the   configuration    of   the    TM   is
+  $w_l(\state,c)w_r$, then the  sentential form will be
+  $\overline{w}_l\nont{}{c}\,w_r\$ $,                where
+  $\overline{w}_l$ is  the same as $w_l$  but with each
+  symbol $d$ in $w_l$ replaced by $\overline{d}$.
+\end{quote}
+
+Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
+$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
+form    corresponding     to    the    initial     configuration    is
+$\nont{\mathit{init}}{c}w\$$ (we  can assume that there  is a production
+$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w\$$, where  \nont{}{} is
+the start symbol of the  grammar). Further correspondences between the
+TM moves and the grammar productions are as follows:
+
+\begin{enumerate}
+\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
+  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
+  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
+\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
+  are two productions
+  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
+  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
+\end{enumerate}
+The idea  behind the  productions is explained  with an
+example:  Assume that  the current  sentential form  is
+$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
+TM  has  a   transition  $(S_i,\acar)  \rightarrow
+(S_j,\acdr, L)$.  Since the next corresponding  step in
+the  derivation  has  to  be  done  without  any  prior
+knowledge of whether the symbol to the left of the tape
+is a \acar\  or a \acdr, two  productions are provided,
+and  the  invariant  will  be maintained  only  if  the
+production         $\nont{i}{\acar}         \rightarrow
+\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
+in  the  derivation.    This  gives  the  configuration
+$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
+Simplification   with    the   production   $\bcdr\acdr
+\rightarrow               \epsilon$              yields
+$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
+corresponds  to   the  changed  configuration   of  the
+TM.  Notice  carefully that a wrong  choice breaks
+the invariant  and it cannot be  recovered subsequently
+by any choice of productions.
+
+After the  TM has  halted, there  are further  ``cleanup'' derivations
+that derive  $\epsilon$ only if  the invariant has been  maintained so
+far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
+$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
+final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
+\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
+$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
+for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
+and                $\nont{\mathit{final}}{c}               \rightarrow
+\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
+\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
+\acdr\ on the right of the tape head.  This completes the
+reduction.\!
+\end{proof}
+
+We now show that the proof can be replayed for the class of grammars
+generated from liveness analysis of programs.
+
+\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
+Given   a    grammar    $G$    of   the    kind    described    in
+Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
+analysis  of a  program  and  a forward  access  path $\alpha$,    the
+membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
+\end{lemma}
+
+\begin{proof}
+Given a Turing machine and an input string, the  proof   of
+Lemma~\ref{lemma:grammar-undecidable} generates a grammar. 
+We shall define a function
+for   each  non-terminal $\nont{i}{c}$   introduced  in   this grammar
+such that the
+liveness  analysis  of  the function will  result
+in a set of productions that includes the productions for
+$\nont{i}{c}$ generated for the proof.
+As an example, it  can be verified that the grammar
+for the function shown in Figure~\ref{fig:turing-pgm}(b)    includes      the
+productions shown in Figure~\ref{fig:turing-pgm}(a). Here $\nont{\mathit{all}}{}$   corresponds   to  the   demand
+$\sigma_{\mathit{all}}$.
+
+The body of the function corresponds  to the RHS of the productions
+for $\nont{i}{c}$. Productions with the same  LHS non-terminal, but
+with differing non-terminal on RHS can be generated by 
+joining program fragments with $\SIF$.   If  there  are  $n$ such  functions,  then
+\mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
+distinct function in each branch.
+
+\begin{figure}[t]
+  \begin{minipage}{.45\columnwidth}
+    \begin{align*}
+      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
+      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
+      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
+      \acdr\nont{\mathit{all}}{}
+    \end{align*}
+  \end{minipage}
+  \begin{minipage}{.45\columnwidth}
+    \input{equivalent-program}
+  \end{minipage}\\
+  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
+  \caption{A possible grammar generated by the proof of
+    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
+  grammar.}\label{fig:turing-pgm}
+\vspace*{-6mm}
+\end{figure}
+
+Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
+had a single non-terminal on the RHS. There are two characteristics of
+the grammar produced from liveness-analysis  that are are relevant for
+replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
+\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
+grammar  generated from  liveness  analysis of the constructed program
+will  have the  production
+$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
+\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
+$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
+\nont{\mathit{all}}{}$.
+
+It is clear that if the TM given as an instance of the halting problem
+accepts the input string, then the earlier derivation can be replayed,
+every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
+sentential form by $\epsilon$. However, if  the TM does not accept the
+input string, then every sentential form derived from the start symbol
+would    have     at    least    one    non-terminal from the grammar
+of Lemma~\ref{lemma:grammar-undecidable} that is different
+from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
+derivable from the grammar.
+\end{proof}
+
 \clearpage

### 2016-04-13 paper/cfg-to-nfa.tex
created file cfg-to-nfa.tex and added wrapfig for fig4

--- 2016-03-25 paper/paper.tex
@@
-\documentclass[9pt,preprint,nonatbib]{sigplanconf}
-\usepackage{etex}
-\usepackage{float}
-\usepackage{mathrsfs}
-%%\usepackage{lineno}
-\usepackage{booktabs}
-\usepackage{stmaryrd}
-\usepackage{boxedminipage}
-\usepackage[T1]{fontenc}
-\usepackage{epsfig}
-\usepackage{multirow}
-\usepackage{url}
-\usepackage[normalem]{ulem}
-\usepackage{schemepgm}
-\usepackage{graphicx}
-\usepackage{graphpap}
-\usepackage{tabularx}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{pstricks}
-\usepackage{pst-text}
-\usepackage{pst-node}
-\usepackage{pst-tree}
-\usepackage{pst-rel-points}
-\usepackage{bcprules}
-\usepackage{subfigure}
-\usepackage[boxed]{algorithm2e}
-\usepackage{makecell}
-\usepackage{balance}
-
-% to save space in lists
-\usepackage{enumitem}
-\setlist{nolistsep}
-
-\input{fh_defs}
-\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
-      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
-      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
-      #1 \vfill}}
-      %Poor man's typography
-\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
-      %It would be better to specify these in font-relative measures, but it
-      %probably doesn't scale anyway.
-
-\definecolor{Myblue}{rgb}{.2,0,1}
-\definecolor{Myred}{rgb}{1,.2,.2}
-
-\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
-%\newcommand{\added}[1]{{\color{Myblue}{#1}}}
-\newcommand{\added}[1]{#1}
-\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
-%\newcommand{\warning}[1]{{\color{Myred}{#1}}}
-\newcommand{\warning}[1]{}
-%\newcommand{\blankout}[1]{}
-
-\newcommand{\oxlb}{\ensuremath{[\!|}}
-\newcommand{\oxrb}{\ensuremath{|\!]}}
-\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
-%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
-\newcommand{\scmin} {\mbox{\sf\em in}}
-\newcommand{\scmnull}{\mbox{\sf\em null?}}
-\newcommand{\scmpair}{\mbox{\sf\em pair?}}
-\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
-\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
-\def\myvec{\mathaccent"017E } 
-\newcommand{\stk}{{S}}      
-\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
-\newcommand{\bang}{\mbox{\sc bang}}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{lemma}[theorem]{Lemma}
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-%% \linenumbers 
-% --- Author Metadata here ---
-%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
-%over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
-%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE. 
-% --- End of Author Metadata ---
-
-\title{Liveness-Based Garbage Collection for Lazy Languages}
-
-% You need the command \numberofauthors to handle the 'placement
-% and alignment' of the authors beneath the title.
-%
-% For aesthetic reasons, we recommend 'three authors at a time'
-% i.e. three 'name/affiliation blocks' be placed beneath the title.
-%
-% NOTE: You are NOT restricted in how many 'rows' of
-% "name/affiliations" may appear. We just ask that you restrict
-% the number of 'columns' to three.
-%
-% Because of the available 'opening page real-estate'
-% we ask you to refrain from putting more than six authors
-% (two rows with three columns) beneath the article title.
-% More than six makes the first-page appear very cluttered indeed.
-%
-% Use the \alignauthor commands to handle the names
-% and affiliations for an 'aesthetic maximum' of six authors.
-% Add names, affiliations, addresses for
-% the seventh etc. author(s) as the argument for the
-% \additionalauthors command.
-% These 'additional authors' will be output/set for you
-% without further effort on your part as the last section in
-% the body of your article BEFORE References or any Appendices.
-%\cmt{{
-%
-%\numberofauthors{3} %  in this sample file, there are a *total*
-% of EIGHT authors. SIX appear on the 'first-page' (for formatting
-% reasons) and the remaining two appear in the \additionalauthors
-%section.
-%
-%\author{
-% 1st. author
-\cmt{{
-\authorinfo {Prasanna Kumar. K} 
-       {IIT Bombay, Mumbai 400076, India}
-       {prasannak@cse.iitb.ac.in}
-% 2nd. author
-\authorinfo {Amitabha Sanyal} 
-       {IIT Bombay, Mumbai 400076, India}
-       {as@cse.iitb.ac.in}
-% 3rd. author
-\authorinfo {Amey Karkare}
-       {IIT Kanpur, Kanpur 208016, India}
-       {karkare@cse.iitk.ac.in}
-}}
-\authorinfo{Double Blind Review}{}{}
-\maketitle
-
-\begin{abstract} 
-We consider  the problem of reducing  the memory required to  run lazy
-first-order functional programs. Our approach is to analyze
-programs  for liveness  of  heap-allocated data.   The  result of  the
-analysis is  used to preserve  only live data---a subset  of reachable
-data---during garbage  collection.  The result  is an increase  in the
-garbage reclaimed and  a  reduction in  the peak memory requirement 
-of programs.   While this  technique has already  been shown  to yield
-benefits for  eager first-order  languages, the  lack of  a statically
-determinable execution  order and  the presence  of closures  pose new
-challenges  for lazy  languages.  These  require changes  both in  the
-liveness analysis itself and in the design of the garbage collector.
-
-%% We use the  result of the analysis to annotate  each potential garbage
-%% collection  point  in   the  program  with  a   set  of  deterministic
-%% finite-state automata (DFA) describing the liveness at that point.
-To  show the  effectiveness of  our method,  we implemented  a copying
-collector that uses  the results of the liveness  analysis to preserve
-live  objects,  both evaluated  (i.e.,  in  WHNF) and  closures.   Our
-experiments confirm  that for  programs running with  a liveness-based
-garbage  collector, there  is a  significant decrease  in peak  memory
-requirements.   In addition,  a  sizable reduction  in  the number  of
-collections  ensures that  in spite  of using  a more  complex garbage
-collector, the execution  times of programs running  with liveness and
-reachability-based collectors remain comparable.
-\end{abstract}
-
-\category{D.3.4}{Programming Languages}{Processors}[Memory Management
-  (Garbage Collection), Optimizations]
-\category{F.3.2}{Logic and Meanings Of Programs}{Semantics of
-  Programming Languages}[Program Analysis]
-
-\terms{Algorithms, Languages, Theory}
-
-\keywords{Heap Analysis, Liveness Analysis, Memory Management, Garbage
-  Collection, Lazy Languages}
-
-\section{Introduction}
-\label{sec:intro}
-
-Functional  programs  make  extensive  use  of  dynamically  allocated
-memory.  The allocation is either explicit (i. e., using constructors)
-or  implicit   (creating  closures).   Programs  in   lazy  functional
-languages put additional  demands on memory, as  they require closures
-to be carried from the point of creation to the point of evaluation.
-
-While  the runtime  system  of most  functional  languages includes  a
-garbage collector to efficiently  reclaim memory, empirical studies on
-Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
-Haskell~\cite{rojemo96lag} programs have shown that garbage collectors
-leave uncollected a large number  of memory objects that are reachable
-but not live (here {\em live} means the object can potentially be used
-by the program at a  later stage).  This results in unnecessary memory
-retention.
-
-In this paper,  we propose the use of liveness  analysis of heap cells
-for garbage collection (GC) in a lazy first-order functional language.
-The central  notion in  our analysis is  a generalization  of liveness
-called {\em  demand}---the pattern of future  uses of the value  of an
-expression.   The  analysis  has  two parts.   We  first  calculate  a
-context-sensitive  summary   of  each   function  as  a   {\em  demand
-  transformer} that transforms a {\em  symbolic} demand on its body to
-demands  on its  arguments.   This  summary is  used  to step  through
-function calls  during analysis.   The concrete  demand on  a function
-body  is  obtained through  a  conservative  approximation similar  to
-0-CFA~\cite{Shivers:1988} that  combines the demands on  all the calls
-to  the function.   The result  of the  analysis is  an annotation  of
-certain program points with  deterministic finite-state automata (DFA)
-capturing the  liveness of  variables at  these points.   Depending on
-where GC  is triggered, the  collector consults  a set of  automata to
-restrict reachability during marking.  This  results in an increase in
-the garbage reclaimed and consequently in fewer collections.
-
-\added{While  the idea  of  using static  analysis  to improve  memory
-  utilization  has  been  shown  to   be  effective  for  {\em  eager}
-  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
-    lee05static}, a straightforward extension  of the technique is not
-  possible  for  lazy  languages,  where  heap-allocated  objects  may
-  include   closures    (runtime   representations    of   unevaluated
-  expressions).   }   \todelete{Designing  a  liveness  based  garbage
-  collector   for  lazy   languages  poses   significant  challenges.}
-Firstly, since  data is made  live by  evaluation of closures,  and in
-lazy languages  the place in  the program where this  evaluation takes
-place cannot  be statically determined, laziness  complicates liveness
-analysis itself.  Moreover, for liveness-based  GC to be effective, we
-need to  extend it  to closures  apart from  evaluated data.   Since a
-closure can escape the scope in which  it was created, it has to carry
-the  liveness  information  of   its  free  variables.   As  execution
-progresses  and possible  future uses  are eliminated,  we update  the
-liveness  information  in  a  closure with  a  more  precise  version.
-Because  of   these  reasons,  the  garbage   collector  also  becomes
-significantly more  complicated than a liveness-based  collector for a
-eager language.
-
-Experiments    with   a    single    generation   copying    collector
-(Section~\ref{sec:experiments})   confirm  the   expected  performance
-benefits.  Liveness-based collection results in an increase in garbage
-reclaimed.  As  a consequence, there is  a reduction in the  number of
-collections from 1.6X  to 23X and a decrease  in the minimum
-memory requirement from  1X to 1389X.  As  an added benefit,
-there is also a reduction in the overall execution time in 5 out of 12
-benchmark programs.
-
-%------------------------------------------------------------%
-\subsection{Motivating example}
-\label{sec:motiv}
-\input{fig_motiv}   Figure~\ref{fig:mot-example}   shows  an   example
-program and the state of the heap at the program point $\pi$\footnote{We write
-$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
-before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
-$(\length\ \pz)$.  The heap is represented  by a graph in which a node
-either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
-\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
-(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
-  references} and represent values of variables or fields.  The figure
-shows the  lists \px\  and \pz\  partially evaluated  due to  the
-\SIF\ condition 
-(\NULLQ~(\CAR~\pz)).
-The edges  shown by thick  arrows are those  which are live  at $\pi$.
-%% Only those  cell which have a future use  should be preserved
-%% during garbage collection, all other
-%% cells can be  reclaimed.
-Thus if a GC takes  place at $\pi$ with the heap shown
-in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
-will preserve only  the cells referenced by $\pz$, and  the live cells
-constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
-reachability-based collector (RGC) will preserve all cells.
-
-In  this work  we show  that  static analysis  of heap  data can  help
-garbage  collectors   in  reclaiming   more  garbage.    The  specific
-contributions of this paper are:
-\added{
-\begin{itemize}
-\item We  propose a liveness  based GC scheme  for a lazy  first order
-  functional language and  prove its correctness.  To the  best of our
-  knowledge,  this is  the first  work on  liveness-based GC  for lazy
-  languages.       All      previous      attempts~\cite{shaham01heap,
-    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
-    karkare06effectiveness} have  involved either imperative  or eager
-  functional languages.  The novelty of our approach is in handling of
-  liveness  of closures,  an  issue  that arises  due  to laziness  at
-  runtime.
-
-\item  We  formulate a  liveness  analysis  for the  lazy  first-order
-  functional language  and prove  its correctness. The  proof involves
-  specifying  liveness   for  the  language  through   a  non-standard
-  semantics and then proving the  analysis correct with respect to the
-  specification.
-
-\item The  analysis results  in a set  of context-free  grammars along
-  with  a fixed  set  of non-context-free  productions.  The  decision
-  whether to copy a cell during  GC translates to a membership problem
-  for such  grammars.  Earlier research assumed  the undecidability of
-  this membership question and  used an over-approximation to overcome
-  it.  In this paper, we provide  a formal proof of the undecidability
-  of this problem.
-
-\item We have implemented a garbage collector that uses the result of
-  liveness  analysis to  retain  live cells.   Our experiments  reveal
-  interesting  space-time   trade-offs  in  the  engineering   of  the
-  collector---for  example, updating  liveness information  carried in
-  closures during  execution results in more  garbage being collected.
-  Empirical results  show the effectiveness of  liveness-based GC.
-\end{itemize}
-}
-\todelete{
-\begin{itemize}
-\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
-  functional language and  prove its correctness. The 
-  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
-  correct with respect to the specification.   
-\item The analysis results  in a set  of context-free  grammars along
-  with  a fixed  set  of non-context-free  productions.  The  decision
-  whether   to  copy  a  cell   during  GC  translates  to a  membership problem  for such  grammars.  In  this
-  paper, we  have shown the  undecidability of this problem and 
-  circumvent the  problem by over-approximating the  grammars
-  with DFAs.
-\item We  have implemented a garbage  collector that uses the  DFAs to
-  retain live  cells.  Our  experiments reveal  interesting space-time
-  trade-offs  in  the  engineering  of  the  collector---for  example,
-  updating liveness  information carried in closures  during execution
-  results  in   more  garbage  being  collected.    Empirical  results
-  show the effectiveness of liveness-based GC.
-\end{itemize}
-
-As far  as we know, this  is the first work  on liveness-based GC      for      lazy       languages.       All      previous
-attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
-  asati14lgc, karkare06effectiveness} have  involved either imperative
-or eager functional languages.
-}
-\subsection{Organization of the paper}
-
-Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
-language considered  and gives a small-step  operational semantics for
-it.  The liveness  analysis for this language and  its soundness proof
-is          presented          in          Section~\ref{sec:liveness}.
-Section~\ref{sec:computing} describes  the formulation of  liveness as
-grammars. We also give a proof  of undecidability of such grammars and
-show     how     they     can     be     approximated     by     DFAs.
-Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
-collector,  in  particular  the  use  of  liveness  DFAs  for  GC.     We     report    our    experimental     results    in
-Section~\ref{sec:experiments}    along    with   some    observations.
-Section~\ref{sec:relatedwork}  discusses  previous   work  related  to
-GC  and   liveness  and  Section~\ref{sec:conclusion}
-discusses possible extensions and concludes the paper.
-\section{The target language---syntax and semantics}
-\label{sec:defs}
-Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
-is a first order language with lazy semantics. Programs are restricted
-to        be        in        Administrative        Normal        Form
-(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
-functions  are  variables.  While  this  restriction  does not  affect
-expressibility,  this form  has  the benefit  of  making explicit  the
-creation of closures through the $\LET$ construct.   We
-  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
-  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
-  occur in  $s$. The restriction of  \LET\ to a single  definition is
-for ease  of exposition---generalization to multiple  definitions does
-not add conceptual difficulties.  We further restrict each variable in
-a program  to be  distinct, so that  no scope  shadowing occurs; this
-simplifies reasoning.
-
-We denote the body of a function ${\mathit  f}$   as $e_{\mathit f}$.
- We assume that each program has a distinguished function
-\mainpgm,    defined as\linebreak    $(\DEFINE\      ({\tt
-  \mainpgm})\  e_\mainpgm)$,  and  the program begins execution
-with the call to \mainpgm.
-
-\begin{figure}[t]\footnotesize
-\renewcommand{\arraystretch}{0.9}
-\begin{eqnarray*}
-   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
-   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
-    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
-\,\,x_n)\,\,
-    e)
-    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
-e \in \mathit{Expr} & \!\!\!::=\!\!\! &
-\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
-       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
-       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
-binding} \\
-       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
-    \end{array}\right. \\
-s \in \mathit{App} & \!\!\!::=\!\!\!  &
-\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
-       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
-       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
-       (\CAR\,\, x) & \\
-       (\CDR\,\, x) & \mbox{\em --- selectors} \\
-       (\NULLQ\,\, x) & \\
-       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
-%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
-       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
-    \end{array}\right.
-\end{eqnarray*}
-\vspace*{-6pt}
-  \caption{The syntax of our language}\label{fig:lang-syntax}
-%\figrule
-\normalsize
-\vspace*{-6pt}
-\end{figure}
-
-\begin{figure*}[t!]
-%\comment{
-
-\begin{center}\footnotesize
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|c|c|c|}
-\hline
-Premise & Transition & Rule name \\
-\hline
-\hline 
-          & $\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\, \kappa
-  \rightsquigarrow \rho',\, S,\, \heap[\ell := \kappa],\, e$    &  {\sc const}
-\\
-\hline
-          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, \heap,\, (\CONS~x~y)
-\rightsquigarrow
-$  $\rho',\, S,\, \heap[\ell := (\rho(x), \rho(y))],\, e$}     &  {\sc cons} \\
-\hline
-
-$\heap(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\,
-(\CAR~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$      &
-{\sc car-select} \\
-\hline
-
-$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
-(\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho'\rangle), (\CAR~x))\!:\!S,\, \heap,\, s$      &
-{\sc car-1-clo} \\
-\hline
-
-$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (\CAR~x)
-\rightsquigarrow
-\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,\heap,\, s$      &
-{\sc car-clo}
-\\
-\hline
-
-$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
- & {$\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\, (+~x~y)  \rightsquigarrow$
-$\rho', \,S,\, \heap[\ell := \heap(\rho'(x)) + \heap(\rho'(y))],\, e$}      &
-{\sc prim-add} \\
-\hline
-
-$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
-\rightsquigarrow
-\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
-{\sc prim-1-clo} \\
-\hline
-$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
-\rightsquigarrow
-\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
-{\sc prim-2-clo} \\
-\hline
-{$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
-(f~\myvec{x})  \rightsquigarrow
-[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
-{\sc funcall} \\
-\hline
-$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
-  \rightsquigarrow$
-$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell := \langle s,
-    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
-    \rangle],\, e$} &
-{\sc let} \\
-\hline
-$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
-\hline
-$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
-\hline
-$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
-  (\SIF~x~e_1~e_2)   \rightsquigarrow
-\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
-&
-{\sc if-clo} \\
-\hline
-{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
-e)\!:\!S,\, \heap,\,
-(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
-{\sc return-whnf}\\
-\hline
-$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
-  \rightsquigarrow$
-$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
-{\sc return-clo} \\
-\hline
-\end{tabular}
-\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
-\end{center}
-%}\comment
-\vspace*{-16pt}
-\end{figure*}
-
-\subsection{Semantics}\label{sec:semantics}
-%\comment{
-We now give  a small-step semantics for our  language.
-We first specify the domains used by the semantics:
-\[
-\renewcommand{\arraystretch}{1}
-\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
-\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow (Data + \{empty\}) & \mbox{-- Heap}\\
-d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
-v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
-c:   & \mathit{Clo} &=& \mathit{(App \times Env)}& \mbox{-- Closures}\\
-\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
-\mbox{-- Environment} \\
-\end{array}
-\]
-
-Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
-A non-empty location either contains  a value in  \emph{WHNF} (a  number, the
-empty  list  $\NIL$ or  a  \CONS\  cell  with  possibly  unevaluated
-constituents) or  a {\em closure}.   A closure  is a pair  $\langle s,
-\rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
-maps free variables of $s$ to their respective locations. Since all data objects are boxed, we model an environment as a mapping
-from the set  of variables of the program  $\mathit{Var}$ to locations
-in the  heap.
-
-The  semantics  of   expressions  (and  applications\footnote{In  most
- contexts, we  shall use the  term 'expression' and the  notation $e$
-to stand   for both expressions and applications.}) 
-are given by transitions of the form $\rho, \stk, \heap, e \rightsquigarrow
-\rho',  \stk', \heap',  e'$.  Here  $\stk$ is  a stack  of continuation
-frames.  Each  continuation frame  is a triple $(\rho,  \ell, e)$,
-signifying that the  location $\ell$ has to be updated  with the value
-of the  currently evaluating  expression and $e$  is to  be evaluated
-next  in the  environment $\rho$.  
-The  initial  state  of  the  transition  system  is:
-\[([\;]_\rho,\,     (\rho_\mathit{init},     \,    \ell_{\ans}     ,\,
-(\print~\ans)):[\;]_{S},   [\;]_{\heap},\,   (\mainpgm))\]  in   which
-$[\;]_\rho$, $[\;]_{\mbox  \footnotesize {H}}$ and $[\;]_{S}$  are are
-the empty environment, heap and  stack respectively. The initial stack
-consists  of  a  single  continuation   frame  in  which  \ans\  is  a
-distinguished variable that will eventually  be updated with the value
-of  (\mainpgm),  and  $\rho_\mathit{init}$  maps \ans\  to  a  location
-$\ell_{\ans}$.  In addition, \print\ is a function modeling a printing
-mechanism---a   standard   run-time   support  assumption   for   lazy
-languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The
-operator~$:$ pushes elements on top of the stack.
-
-The   notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
-environment  that  maps  variables  $x_i$ to  locations  $\ell_i$  and
-$\heap[\ell := d]$  indicates updation of  \heap\ at $\ell$
-with  $d$.   $\rho \oplus  \rho'$  represents  the environment  $\rho$
-shadowed  by  $\rho'$  and  $\lfloor \rho  \rfloor_X$  represents  the
-environment  restricted  to  the  variables in  $X$.  Finally  $FV(s)$
-represents the  free variables  in the  application $s$  and $addr(c)$
-gives the address of the closure $c$ in the heap.
-
-The small-step semantics  is shown in Figure~\ref{fig:lang-semantics}.
-Unlike   an   eager  language,   evaluation   of   a  let   expression
-$(\LET~x\leftarrow  s~\IN~e)$ does  not  result in  the evaluation  of
-$s$. Instead,  as the {\sc let}  rule shows, a closure  is created and
-bound to  $x$. The  program points which  trigger evaluation  of these
-closures   are  an $\SIF$   condition ({\sc  IF-clo})   and
- a \SRETURN\ ({\sc  return-clo}).  We  call such  points \emph{evaluation
-  points   $(\epath)$}  and   label  them   with  $\psi$   instead  of
-$\pi$. As an  example of closure  evaluation, we  explain the
-three rules for  $(\CAR~x)$.  If $x$ is a closure,  it is evaluated to
-WHNF, say $(d_1, d_2)$.  This is  given by the rule {\sc car-clo}.  If
-$d_1$ is  not in WHNF,  it is  also evaluated ({\sc  car-1-clo}).  The
-address  to be  updated  with the  evaluated value  is
-recorded  in  a  continuation  frame.    This  is  required  for the
-evaluation  to be lazy, else $d_1$ may  be evaluated more than once
-due to  sharing~\cite{Jones87}.  Only after this  is the actual selection  done ({\sc
-  car-select}).
-
-\section{Liveness}
-\label{sec:liveness}
-
-A variable is {\em live} if there  is a possibility of its value being
-used in  future computations and  dead if  it is definitely  not used.
-Heap-allocated data  needs a richer model  than classical liveness---a
-model which talks about liveness of references.
-
-%% The model use the notion of  {\em access paths}, i.e.,
-%% a  prefix-closed set of strings  over $\{\acar,\acdr\}^\ast$, in
-%% which 
-%% $\acar$,  $\acdr$ represent  access  using $\CAR$  and $\CDR$  fields.
-
-An {\em access path}
-is a  prefix-closed set of strings  over $\{\acar,\acdr\}^\ast$, where
-$\acar$,  $\acdr$ represent  access  using $\CAR$  and $\CDR$  fields respectively.
-Given an initial location $\ell$ (usually a reference corresponding to
-a variable)  and a  heap \heap, semantically  an access  path $\alpha$
-represents     a      reference,     denoted     $\heap\llbracket\ell,
-\alpha\rrbracket$,  in the  heap  that is  obtained  by starting  with
-$\ell$ and chasing the \CAR\ or  \CDR\ fields in the heap as specified
-by  the  access   path.   $\heap\llbracket\ell,  \alpha\rrbracket$  is
-defined only if  the path followed in the  heap is \emph{closure-free}
-(does not cross closures), else it is undefined.
-
-Access paths are used to  represent liveness. As an example, a list
-$x$ with 
-liveness  $\{\epsilon,  \acar,  \acdr,  \acdr\acar,
-\acdr\acdr,  \acdr\acdr\acar\}$ means that future computations    only
-refer up  to the  second and  third members of  $x$.  A  {\em liveness
-  environment} is a mapping from  variables to access paths, but often
-expressed as  a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
-x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
-  \mapsto\{\epsilon,\acdr,\acdr\acdr\},          y\mapsto\{\epsilon\},
-  z\mapsto\{\}]$.    In  this   notation,  $y   \mapsto  \{\epsilon\}$
-represents access using $y$ itself and $z \mapsto \{\}$ indicates 
-$z$ is dead.  In lazy  languages, liveness environments are associated
-with regions of programs instead of program points.
-
-A notion  that generalizes liveness  is {\em demand}.   While liveness
-gives the patterns of future uses of a variable, demand represents the
-future uses of the value of an expression.  The demand on an expression
-$e$   is   also   a   set    of   access   paths---the   subset   of
-$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
-result.   To see  the need  for demands,  consider the expression
-$(\LET~x\leftarrow  (\CDR~y)~\IN~   (\SRETURN~x))$.   Assume   that  the
-context of this expression places  the demand $\lbrace \epsilon, \acar
-\rbrace$. Since the  value of the expression is the  value of $x$, the
-demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
-  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
-$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
-This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
-y.\acdr\acar  \rbrace$.   These are  the $y$-rooted  accesses
-required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
-result of $(\CDR~y)$.
-%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
-%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
-%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
-%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
-%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
-%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.
-
-We use $\sigma$  to range over demands, $\alpha$ to  range over access
-paths and $\Lv$  to range over liveness environments.  The liveness of
-an  individual variable  $y$ in  \Lv\  is $\Lv(y)$,  but more  commonly
-written as  $\Lv_y$.  The notation $\sigma_1\sigma_2$  denotes the set
-$\lbrace  \alpha_1\alpha_2 \mid  \alpha_1 \in  \sigma_1, \alpha_2  \in
-\sigma_2\rbrace$.  Often we shall abuse  notation to juxtapose an edge
-label and  a set  of access  paths; $\acar\sigma$  is a  shorthand for
-$\lbrace\acar\rbrace\sigma$.
-
-%==============================================================
-\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
-%==============================================================
-
-\begin{figure}[t]
-  \input{example-to-illustrate-liveness}  
-  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
-\end{figure}
-\input{fig-liveness}
-
-\subsection{Liveness analysis for lazy languages}
-\label{sec:liveness-analysis}
-
-Consider the  program in Figure~\ref{fig:lv-closure}.   As mentioned
-earlier, a lazy evaluation of the $\LET$ expression at $\pi_1$ creates
-a closure  for $(\length~\px)$  instead of  evaluating it.   Since the
-closure may escape the scope in which it is created, it carries a copy
-of $\px$ within itself.  We treat the  copy of $\px$ in the closure as
-being separate from the $\px$ introduced  by the $\LET$, and call it a
-\emph{closure  variable}.  For  liveness calculations,  such variables
-are distinguished  from variables  introduced by  \LET s  and function
-arguments  (called \emph{stack  variables},  since they  reside in  the
-activation  stack).   We notationally distinguish  a closure  variable
-from  its corresponding  stack variable  by subscripting  it with  the
-label of the program   point  where   the   closure  was   created\footnote{Multiple
-  occurrences  of the  same  variable in  an  application are  further
-  distinguished by their positions in the application.}.
-
-  Since a  closure is evaluated  only at evaluation points,  a closure
-  variable is attributed  with the same liveness in  the entire region
-  of  the  program from  the  point  of  creation  of the  closure  to
-  reachable evaluation points.  This is  also true of stack variables,
-  because, as we shall see, stack variables derive their liveness from
-  closure variables. Thus, there are two major differences between our
-  formulation of  liveness of  lazy languages  with liveness  of eager
-  languages~\cite{asati14lgc}:   (i)  the   introduction  of   closure
-  variables in the  liveness calculations, and (ii)  a single liveness
-  value for each  variable that is applicable from  its creation point
-  to evaluation points.
-
-Closure  variables  get  their  liveness values  through  a  chain  of
-dependences beginning  at a  variable at an  evaluation point.   As an
-example,  consider again  the program  in Figure~\ref{fig:lv-closure}.
-The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
-expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
-$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
-denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
-\leftarrow            (+~\py~1),\,           \py            \leftarrow
-(\length~\px_{\mathit{\pi}_1})\rbrack$.    Indeed,   the   chains   of
-closures  in   the  heap  are   a  runtime  representation   of  these
-dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
-expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
-of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
-$\px_{\pi_1}$.  Other dependence chains which result in a liveness for
-$\px_{\pi_1}$    are    $\lbrack    \psi_1\!\!\!:\!\!\py    \leftarrow
-(\length~\px_{\pi_1})\rbrack$   and  $\lbrack   \psi_2\!\!\!:\!\!\!\pw
-\leftarrow          (/~\pu~\py),\,\,           \py          \leftarrow
-(\length~\px_{\pi_1})\rbrack$.   The  liveness analysis  described  in
-this section declares  the liveness of $\px_{\pi_1}$ to be  a union of
-the liveness arising  from these dependence chains.  To be  safe, a GC
-during evaluation  of $\py$ at  $\psi_1$ has  to use this  liveness to
-copy the heap starting from  $\px_{\pi_1}$.  However, notice that if a
-GC  takes place  while evaluating  $\pz$  at $\psi_3$,  it can  safely
-consider only the liveness arising from the dependence chain\linebreak
-$\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow (+~\py~1),\, \py \leftarrow
-(\length~\px_{\pi_1})\rbrack$.    The    garbage   collection   scheme
-described in Section~\ref{sec:GC-scheme} uses a generalization of this
-observation  to  dynamically  select   an  evaluation  point  specific
-liveness in order to collect more garbage.
-
-Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
-parts. The  function $\mathit{ref}$,  takes an  application $s$  and a
-demand $\sigma$ and returns the incremental liveness generated for the
-free variables of $s$ due to  the application.  This will be consulted
-during GC  while exploring the heap  starting from the
-closure  variables.  The function $\mathcal{L}$  uses  $\mathit{ref}$ to  propagate
-liveness across expressions.
-
-In  a   lazy  language,   an  expression   is  not   evaluated  unless
-required. Therefore  the null  demand ($\emptyset$) does  not generate
-liveness in any of the rules defining $\mathit{ref}$ or $\mathcal{L}$.
-A non-null  demand of  $\sigma$ on (\CDR~$x$),  is transformed  to the
-liveness $\{x.\epsilon,  x.\acdr\sigma\}$.  In an opposite  sense, the
-demand  of  $\acdr\sigma$ on  (\CONS~$y$~$z$)  is  transformed to  the
-demand  $\sigma$  on  $z$.   Since \CONS\  does  not  dereference  its
-arguments, there  is no $\epsilon$ demand  on $y$ and $z$.   The rules
-for (\PRIM~x~y) and (\NULLQ~x) are  similar. Constants do not generate
-any liveness.
-
-In case of a  function call, we  use the third  parameter $\Lfonly$
-that  represents  the  summaries  of all  functions  in  the  program.
-$\Lfonly_{\mathit  f}$  (the summary for a  specific
-function $f$) expresses  how the demand $\sigma$ on a  call to $f$ is
-transformed into  the liveness of  its parameters at the  beginning of
-the  call.  $\Lfonly$  is determined  by the  judgement $\mathit{Prog}
-\len  \Lfonly$ using  inference rule  ({\sc live-define}).   This rule
-describes  the  fixed-point property  to  be  satisfied by  $\Lfonly$,
-namely, the  demand transformation  assumed for  each function  in the
-program should  be the  same as  the demand  transformation calculated
-from      its      body.       As       we      shall      see      in
-Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
-grammar and  the language generated  by this grammar is  the least
-solution satisfying  the rule. We  prefer the least solution  since it
-ensures the safe collection of the greatest amount of garbage.
-
-We next  describe the function $\mathcal{L}$  that propagates liveness
-across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
-{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
-value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
-liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
-$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
-$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
-liveness $\{x.\epsilon\}$ is created and  added to the union.  
-To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
-observe that the value  of $\LET$ is the value of  its body $e$.  Thus
-the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
-demand $\sigma$. Since the stack variable $x$ is copied to each of the
-closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
-liveness of the  closure variables.  This liveness,  say $\sigma'$, is
-also  the  demand  on  $s$, thus the  liveness  environment
-$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
-\{x.\sigma\}$. Finally, the stack variables corresponding to the free
-variables of $s$ are updated and added 
-  to  give the  overall liveness environment  for $(\LET~x
-\leftarrow~s~\IN~e)$.
-
-\added{  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
-  liveness of the reference $\heap\llbracket\rho(x), \alpha\rrbracket$
-  only   if   $\alpha$  corresponds   to   a   closure-free  path   in
-  \heap\ starting  from $\rho(x)$.  If  this path is intercepted  by a
-  closure,  say  $(\CAR~{y_{\pi}})$, then  the  liveness  of the  path
-  starting from $\mathit{y_{\pi}}$ is given by $\Lv_\mathit{y_{\pi}}$.
-  As we shall see in  Section~\ref{sec:GC-scheme}, the liveness of the
-  closure variable  $y_{\pi}$ is recorded  along with the  closure for
-  $s$ so that the GC can refer to it during garbage collection.  }
-%} %comment
-
-\subsection{Soundness of analysis}
-
-We shall now present a proof of soundness of the analysis presented in
-Section~\ref{sec:liveness-analysis}. \added{It is easy to see that the analysis
-correctly identifies the liveness of stack variables. A stack variable
-is live
-between its introduction through a \LET\  and its last use to
-create a closure variable. This is correctly captured by the
-\LET\ rule in Figure~\ref{fig:live-judge}.}
-Proving soundness  for closure  variables is more complex. Here  are the
-ideas behind the proof. 
-\begin{enumerate}
-\item      We     augment      the      standard     semantics      in
-  Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
-  of  each \LET\  expression.  \added{Note  that, unlike  eager
-    languages,   memory  is   allocated  only   during  execution   of
-    \LET\ expressions}. During GC, we track each reference in the root
-  set and heap that is declared  dead by our analysis.  Any attempt to
-  dereference  such  references  results   in  the  transition  system
-  entering a special state denoted \bang.  We call the semantics after
-  augmentation, \emph{minefield semantics}.
-
-%% During GC, we bind  a special value
-%%   $\bot$ to each  reference in the root  set and heap  that is not
-%%   live by  our analysis.  This  models the de-allocation  of locations
-%%   without any  live references.  Any  attempt to dereference  a $\bot$
-%%   during execution results
-%% %\added{that  has   been   declared  not   live  by   our
-%% %    analysis} %%with  a null-liveness during a  transition will result
-%%   in the transition system entering  a special state denoted \bang.  We
-%%   call the semantics after augmentation, \emph{minefield semantics}.
-\item \label{inline1} Assuming that a program enters the \bang\ state,
-  we  construct,  through  inline expansion,  a  program
-  without function  calls which has  the same minefield  behavior. The
-  final step shows that no program  without function calls can enter the
-  \bang\ state. As a consequence  no program (with or without function
-  calls) can enter the \bang\ state.
-\end{enumerate}
-
-\begin{figure*}[t!]
-\begin{center}\footnotesize
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|@{}c@{\ }|@{\ }c@{}|@{\ }c@{\ }|}
-\hline
-Premise & Transition & Rule name \\ 
-\hline
-\hline 
-\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
-{\sc car-bang} 
-\\
-\hline
-\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
-(\CAR~x), \cred{\sigma} \rightsquigarrow$ $ \rho', \,(\rho, addr(\langle
-s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S, \heap,\, s, \, \cred{\sigma}$ }     &
-{\sc car-1-clo} 
-\\
-\hline
-\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
-\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
-  \rightsquigarrow$ 
-  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
-    \begin{array}{lr}
-      {\emptyset}&\mbox{if } \sigma = \emptyset \\       
-      {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
-    \end{array}\right.
-  }$   }      
-&
-{\sc car-clo}\\
-\hline
-\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
-    s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x
-    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
-    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
-{\sc let} \\ 
-\hline
-\end{tabular}
-\end{center}
-\caption{Minefield  semantics for  \CAR\   and \LET.
-  \warning{Add a note mentioning that we handle the issue of requiring
-    a      specific     demand      by      introducing     a      new
-    symbol}\label{fig:minefield-semantics-for-some}}
-\vspace*{-2mm}
-\end{figure*}
-
-\noindent To set up the minefield semantics, we follow these steps:
-\begin{enumerate}
-
-\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
-  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
-  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
-  $e$, that arises from the actual sequence of function calls that led
-  to  the  evaluation of  $e$.\footnote{The  static  liveness that  is
-    consulted during actual GC  is computed from an over-approximation
-    of this demand.  Thus the soundness  result on the modeled GC will
-    also apply for the actual GC.} The demand for the initial state is
-  $\sigma_{\mathit  {all}}$,  and each  $\rightsquigarrow$  transition
-  transforms   the  demand   according  to   the  liveness   rules  of
-  Section~\ref{sec:liveness-analysis}.      The     information     in
-  continuation frames  on the stack  $S$ are also  similarly augmented
-  with their demands.  Thus a stacked entry now takes the form $(\rho,
-  \ell, e, \sigma)$.
-
-\item The closure created by a let expression is now a 3-tuple
-  $\langle s,  \rho, \sigma \rangle$, where $\sigma$ represents the
-  demand on the closure.
-
-\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
-  collection that returns $(\rho', S', \heap')$. The changes in
-  $\rho, S$ and  $\heap$ are due
-  to non-live references being replaced by $\bot$.  This simulates the
-  act of  garbage collecting  the cells
-  pointed to by these references  during an actual garbage collection.
-  To do so, $GC(\ldots)$ needs to consider the following environments:  (1) the  environment in  the
-  current state, (2) the  environment in each of the stacked  continuations and (3)
-  the environment  in each of  the closures in  the heap.
-
- \begin{enumerate} 
-\added{
-
- \item \label{env}  For each of  these environments, $GC(\ldots)$  calculates a
-   liveness  environment $\Lv$  using  the  corresponding  $e$  (or  $s$)  and
-   $\sigma$.
- \item 
-   For each
-    location $\ell$,  $GC(\ldots)$  sets $\heap(\ell)$  to $\bot$ iff for each
-    environment $\rho$ above, for each $x \in domain(\rho)$, and each forward
-    access
-    path $\alpha$, it is \emph{not} the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
-    \rrbracket = \ell$. 
-%% $x.\alpha$  is  live  and
-%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
-%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
-%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
-}
-
-  %% \item {\color{Myblue}\emph{GC} for  each $x \in domain(\rho)$, we  set $\rho(x)$ to
-  %%   $\bot$  iff $x.\epsilon  \notin \Lv$.   Similarly, for  each stack
-  %%   entry   $(\rho,\ell,e,\sigma)$  in   $S$   with   $\Lv'$  as   the
-  %%   corresponding liveness  environment in $\mathsf{SL}$,  and finally
-  %%   for  each  $x  \in  domain(\rho)$ for  some  closure  $\langle  s,
-  %%   \lfloor\rho\rfloor_{FV(s)}, \sigma\rangle$ in the heap, for all $x
-  %%   \in  FV(s)$ we  set $x$  to  $\bot$ iff  $x.\epsilon \notin  \Lv'$.}
-
-  %% \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
-  %%   \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
-  %%   liveness environment  \Lv\ for  the expression $e$.   We similarly
-  %%   define the liveness  environments for the expressions  $e$ in each
-  %%   of  the continuation  frames  $(\rho, \ell,  e,  \sigma)$ in  $S$,
-  %%   giving  a stack  of  liveness environments  that  we shall  denote
-  %%   $\mathsf{SL}$.
-  %% \item  \emph{Garbage collection  of objects  pointed by  root sets:}
-  %%   Consider  each   of  the  following  environments:   (i)  The  the
-  %%   environment $\rho$ in  the current state (ii) and for  each $x \in
-  %%   domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
-  %%   \Lv$.  Similarly,  for each stack entry  $(\rho,\ell,e,\sigma)$ in
-  %%   $S$  with  $\Lv'$ as  the  corresponding  liveness environment  in
-  %%   $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
-  %%   to $\bot$ iff $x.\epsilon \notin \Lv'$.
-
-%% Consider each of  the following environments: (i)  The the environment
-%% $\rho$ in the current state and  for each $x \in domain(\rho)$, we set
-%% $\rho(x)$ to $\bot$ iff $x.\epsilon  \notin \Lv$.  Similarly, for each
-%% stack  entry   $(\rho,\ell,e,\sigma)$  in  $S$  with   $\Lv'$  as  the
-%% corresponding liveness  environment in $\mathsf{SL}$, and  for each $x
-%% \in domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
-%% \Lv'$.
-
-%%   \item \emph{Garbage  collection of  objects in  the heap:}
-
-%%     For each
-%%     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
-%%     environment $\rho$ in either  the  current state or in a
-%%     continuation frame, for each $x \in domain(\rho)$ and each forward
-%%     access
-%%     path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
-%%     \rrbracket = \ell$. Here 
-%% %% $x.\alpha$  is  live  and
-%% %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
-%% %%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
-%% %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
-%% \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
-%% step~\ref{gc-1}.
-  \end{enumerate}
-\end{enumerate}
-Figure ~\ref{fig:minefield-semantics-for-some} shows some of the
-minefield rules. As mentioned earlier, the transition for a \LET\ 
-%%  $([\;]_\rho,([\;]_\rho,
-%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
-is preceded  by $GC(\ldots)$.  The details of the transition for the {\sc
-  car-clo}  rule   is  also shown.  If    an
-earlier  call to  $GC(\ldots)$  results  in $\rho(x)$  being
-bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
-enters the  \bang\ state ({\sc  car-bang}). Otherwise
-the transition is similar to the earlier {\sc car-clo}
-rule.
-
-Consider a trace  of a minefield execution of a  program $p$, possibly
-ending in  a \bang\ state. We  can construct a call-tree  based on the
-trace in  which each node represents  a function that was  called (but
-did not necessarily  return because of a \bang).  Assume  that each of
-the nodes of  the tree is also annotated with  the program point where
-the corresponding call  was invoked.  This tree can be  used to inline
-function calls in a hierarchical fashion.  The details of the inlining
-can be found in~\cite{asati14lgc}.
-For  a
-call-less program,  the initial state of  the minefield
-semantics  is  assumed  to  be  $([\;]_\rho,(\rho_\mathit{init},
-\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
-e_{\mainpgm})$.
-
-\subsection{Soundness result}
-We first need an auxiliary  result about minefield semantics. Consider
-a trace of  a minefield execution.  For every  minefield state $(\rho,
-S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
-$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
-application $s$) is  non-null.  This can be proved by  an induction on
-the number  of steps leading  to the  minefield state.  The  base step
-holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
-non-null. For the inductive step we  observe that for each step of the
-minefield semantics, if the demand $\sigma$  on the LHS of a minefield
-step  is non-null,  the  demand  on the  RHS  is  a transformation  of
-$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.
-
-Note that our proofs will be for a single round of minefield execution
-i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
-\ref{sec:semantics}).  With  minor variations,  the proof will  also be
-applicable for subsequent evaluations initiated by $\print$.
-%% We show  next that the  minefield execution of a  call-less program
-%% cannot go \bang.
-
-\begin{lemma}
-\label{lemma:call-less-cannot-go-bang}
-Consider the minefield execution of  a program without function calls.
-Such a program cannot enter the \bang\ state.
-\end{lemma}
-\begin{proof}
-Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
-minefield execution of a program.  We  show by induction on the number $n$
-of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
-enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
-(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
-[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
-does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
-transition cannot  enter a  \bang\ state.  Since  our programs  are in
-ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
-does not involve dereferencing, and thus cannot result in a \bang.
-
-For the inductive step, we shall show that none of the minefield steps
-that involves dereferencing  results in a \bang.  These  are the steps
-which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
-$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
-inserted by an earlier $GC(\ldots)$.   However the demand $\sigma'$  on basis of
-which the $GC(\ldots)$ would have inserted a  $\bot$ would have
-included the  current demand $\sigma$.     Thus it is
-enough to  show that the $\rightsquigarrow$ step would not lead to a
-\bang, even if $GC(\ldots)$ had been done with the current demand
-$\sigma$.
-
-We consider the rules for \CAR\  only.  
-The rest  of the rules involve similar reasoning. For the {\sc
-  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
-\sigma$, we  know that $\sigma$ is  non-null. Therefore
-the  liveness  of  $x$  includes  $\epsilon$,  and  the
-dereferencing $\heap(\rho(x))$ will go without \bang.
-
-For the  {\sc car-1-clo}  rule, observe that  there are
-two dereferences.  First $x$  is dereferenced to  get a
-cons  cell  and then  the  head  of  the cons  cell  is
-dereferenced  to  obtain  a  closure.   If  the  demand
-$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
-of    $x$   will    include    both   $\epsilon$    and
-$\acar\epsilon$,  and  a  GC with  this  liveness  will
-neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
-first   component  of   the   cons   cell.  Thus   both
-dereferences    can take place without 
-$\rightsquigarrow$ entering the   \bang\ state.
-\end{proof}
-
-Now we are ready to prove the main soundness result.
-
-\begin{theorem}
-The  minefield  execution of  no  program  can enter  a
-\bang\ state.
-\end{theorem}
-
-\begin{proof}
-Assume to  the contrary that  a program $P$  enters the
-\bang\  state.  We can  transform  $P$  to a  call-less
-program $P'$ such that  the minefield executions of $P$
-and $P'$  are identical  except for change  of variable
-names.             However,                 by
-Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
-$P'$,   a   call-less   program,   cannot   enter   the
-\bang\  state.  Therefore   $P$  also  cannot  enter  the
-\bang\ state.
-\end{proof}
-
-%% Section~\ref{sec:computing}   shows   how  the   demand   transformers
-%% $\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
-%% analysis) can be safely approximated  by a {\em procedure summary} for
-%% each function.   The summary is  in the  form of a  demand transformer
-%% that maps a demand on a call to the function to demands on each of its
-%% arguments.
-
-\section{Towards a  computable form of liveness}\label{sec:computing}
-%\section{Generating and storing liveness information}\label{sec:computing}
-The analysis in Section~\ref{sec:liveness} is fully context-sensitive, 
-describing the  liveness sets  in a function body in
-terms of a symbolic demand $\sigma$  and \Lfonly. However, we are yet to
-describe (i) how to obtain demand transformers \Lfonly\ from the rule
-{\sc live-define} , and (ii) how to compute the concrete demand $\sigma$ on
-each function. To do so, we first need to modify the liveness rules
-to a slightly different form.
-
-%\subsection{Modifying liveness rules}
-\paragraph{Symbolic representation of operations:}
-The   $\mathit{ref}$   rule   for   \CONS,   shown   in
-Figure~\ref{fig:live-judge}, requires us  to remove the
-leading  \acar\ and  \acdr\  from the  access paths  in
-$\sigma$.  Similarly, the rules  for \CAR, \CDR, \PRIM,
-\NULLQ, and \SIF\ require  us to return $\emptyset$, if
-$\sigma$      itself       is      $\emptyset$      and
-$\lbrace\epsilon\rbrace$  otherwise.  To  realize these
-rules  $\sigma$   needs  to  be  known.   This  creates
-difficulties  since  we  want to  solve  the  equations
-arising from liveness symbolically.
-
-The solution is to  also treat the operations mentioned
-above  symbolically.  We  introduce three  new symbols:
-\bcar, \bcdr,  \clazy.  These symbols are  defined as a
-relation  $\hookrightarrow$  between   sets  of  access
-paths:
-\begin{align*}
-  &\bcar\sigma \hookrightarrow \sigma' \mbox{ where } \sigma' = \{\alpha \mid \acar\alpha \in \sigma\}\\
-  &\bcdr\sigma \hookrightarrow \sigma' \mbox{ where } \sigma' = \{\alpha \mid \acdr\alpha \in \sigma\}
-\end{align*}
-Thus \bcar\ selects those entries in $\sigma$ that have leading \acar, and removes the leading \acar\ from them.
-The symbol \clazy\ reduces the set of strings following it to a set containing only $\epsilon$. It filters out, however, the empty set of strings.
-\begin{align*}
-  \clazy\sigma \hookrightarrow & \left\{ 
-  \begin{array}{ll}
-    \emptyset&\mbox{if}~\sigma = \emptyset\\
-    \{\epsilon\} & \mbox{otherwise}
-  \end{array}\right.
-\end{align*}
-We can  now rewrite the \CONS\  and the \CAR\  rules of $\mathit{ref}$
-as:
-\begin{align*}
-&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
-= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
-\mbox{and} \\
-&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
-          =   x.\clazy\sigma \cup x.0\sigma
-\end{align*}
-and the \Lfunonly\ rule
-for \SIF\ as:
-\begin{align*}
-\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
-                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\cup
-        \mathcal{L}(e_2,\sigma,\Lfonly)~\cup
-          \{x.\clazy \sigma\}
-\end{align*}
-The rules for  \CDR, \PRIM\ and \NULLQ\ are also
-modified similarly.
-
-When   there  are   multiple   occurrences  of   \bcar,
-\bcdr\  and \clazy,  $\hookrightarrow$ is  applied from
-right  to left.   The reflexive  transitive closure  of
-$\hookrightarrow$      will      be     denoted      as
-$\stackrel{*}{\hookrightarrow}$.      The     following
-proposition  relates  the  original  and  the  modified
-liveness rules. 
-\begin{proposition}
-Assume  that  a  liveness   computation  based  on  the
-original  set  of  rules  gives  the  liveness  of  the
-variable  $x$   as $\sigma$
-(symbolically,  $\Lanv{x}{}=  \sigma$).  Further,  let
-$\Lanv{x}{}= \sigma'$ when the modified rules are used
-instead      of      \Lfunonly.      Then      $\sigma'
-\stackrel{*}{\hookrightarrow} \sigma$.
-\end{proposition}
-
-An  explanation of  why the  proposition holds  for the
-modified  \CONS\ rule  is  given in  \cite{asati14lgc}.
-The proposition also holds for other modified rules for
-similar reasons.
-
-\input{example-program}
-
-%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
-\paragraph{Computing function summaries $\Lfonly_{\mathit f}$:}
-Given a  function $\mathit{f}$, we now  describe how to
-generate  equations   for  the   demand  transformation
-\Lfonly$_\mathit{f}$.        The       program       in
-Figure~\ref{fig:mot-example2-a}   serves  as   a  running
-example.  Starting with a  symbolic demand $\sigma$, we
-determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
-particular, we  consider $\Lanv{x_i}{}$,  the liveness
-of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
-  live-define},   this   should    be   the   same   as
-$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
-$$
- \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
-  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
-$$
-
-In general, the equations for \Lfonly\ are recursive as in the case of
-\Lfonly$_\mathit{f}$. A closed  form solution for $\Lfone{\mathit{f}}$
-can be derived by observing that each of the liveness rules modifies a
-demand  only by  prefixing it  with symbols  in the  alphabet $\lbrace
-\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore we  can assume
-that $\Lf{f}{i}{\sigma}$ has the closed form:
-\begin{eqnarray}
-\label{eq:LF:DI}
-  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
-\end{eqnarray}
-where \Df{f}{i}  are sets of strings  over the alphabet
-mentioned above.  Substituting the  guessed form in the
-equation   describing    \Lfonly$_{\mathit   f}$,   and
-factoring  out   $\sigma$,  we  get  an   equation  for
-\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
-solution   for   \Df{f}{i}   yields  a   solution   for
-\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
-  \length}$, we get:
-  \begin{eqnarray*}
-&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
-&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy
-       \cup \clazy\Df{\length}{1}{\clazy}
-  \end{eqnarray*}
-
-Note that this equation can also be viewed as a CFG with \{\acdr,
-\clazy\} as terminal symbols and \Df{\length}{1} as the sole
-non-terminal.
-
-%\subsection{Generating liveness equations \Lv\  for function bodies}
-\paragraph{Handling user-defined functions:}
-\label{sec:bodylivenessbodies}
-To  avoid analyzing  the body  of a  function for  each
-call, we calculated the  liveness for the arguments and
-the variables in a function  with respect to a symbolic
-demand  $\sigma$.   To  get   the  actual  liveness  we
-calculate an  over-approximation of the  actual demands
-made by  all the  calls and  calculate the  liveness at
-each  GC  point  inside  the  function  based  on  this
-approximation.  The 0-CFA-style {\em summary demand} is
-calculated by  taking a union  of the demands  at every
-call site of a function.
-
-%% Consider a function  $g$ containing a call to  $f$ at a
-%% site    $\pi$,   say    $\pi\!\!:\!(\LET~x   \leftarrow
-%% (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.  Let the demand on
-%% $g$  be  $\sigma_g$  and,  based on  this  demand,  the
-%% liveness of  $x$ at  $\pi_i$ be $\Lanv{i}{x}$.   By the
-%% $\LET$ rule of Figure~\ref{fig:live-judge}, the call at
-%% $\pi$   contributes   $\Lanv{i}{x}$   to   the   demand
-%% $\sigma_f$. Let  us denote  the contribution of  a call
-%% site $\pi$ in  a function $g$ to the  overall demand on
-%% the function  $f$ as  $\deltacall{f}{\pi}{g}$. Assuming
-%% that there are $k$ call  sites to function $f$, $\pi^1$
-%% (in function $g^1$) \ldots $\pi^k$ (in function $g^k$),
-%% the     over-approximation     of     $\sigma_f$     is
-%% $\deltacall{f}{\pi^1}{g^1}     \cup     \cdots     \cup
-%% \deltacall{f}{\pi^k}{g^k}$.  The distinguished function
-%% \mainpgm\ is  a special case.   We assume it  is called
-%% through    a    printing    mechanism    with    demand
-%% $\sigma_\mainpgm   =   \{\acar,\acdr\}^\ast$   (denoted
-%% $\sigma_{\!all}$) if \mainpgm\  returns a structure and
-%% $\epsilon$ if it returns a base value.
-
-For  the  running  example (Figure~\ref{fig:mot-example2-a}),  $\length$  has  calls  from  $\main$  at
-$\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
-are $\epsilon$ and $\clazy\sigma_{\mathit{\length}}$. Thus:
-\begin{eqnarray*}
-\sigma_{\length}    &=&
- \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
-\end{eqnarray*}
-As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
-stack variables $\pl$ and $\pa$ have the following liveness in terms
-of  $\sigma_{\length}$:
-\begin{align*}
-  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
-%\end{align*}
-%\begin{align*}
-  \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
-  \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
-  \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
-  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
-  \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
-\end{align*}
-
-In  summary, the  equations  generated during the
-analysis are:
-\begin{enumerate}
-\item   For  each   function  $\mathit{f}$,   equations
-  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item For each function $\mathit{f}$, an equation defining the summary
-  demand $\sigma_{\mathit f}$ on $e_f$.
-\item   For  each   function  $\mathit{f}$   (including
-  $\mainpgm$) an  equation  defining
-  liveness for each garbage in  $e_{\mathit f}$.
-\end{enumerate}
-
-%\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
-\paragraph{From liveness sets to context-free grammars:}\label{sec:grammar-formulation}      
-The  equations  above can  now  be  re-interpreted as  a  context-free
-grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
-\clazy\rbrace$.  Let \var{$X$} denote  the non-terminal for a variable
-$X$ occurring on the LHS of the equations generated from the analysis.
-We can  think of  the resulting productions  as being  associated with
-several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
-as a start symbol.  As  an example, the grammar for \var{\Lanv{\xl}{}}
-and \var{\Lanv{\pa}{}} comprise the following productions:
-\begin{eqnarray*}
-  \var{\Lanv{\xl}{}}  &\rightarrow& 
-  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
-  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
-  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
-  \acdr\var{\Df{\length}{1}}\clazy
-       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
-\langle {\sigma_{\length}} \rangle
-&\rightarrow&
-\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
-\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
-\end{eqnarray*}
-Other  equations  can   be  converted  similarly.   The
-language   generated   by  \var{\Lanv{x}{}},   denoted
-$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
-solution  of  \Lanv{x}{}.    
-However, recall that the decision problem that we are interested in
-during GC is: 
-%%\begin{quote} 
-{\em
-Let $x.\alpha$ be a forward access path (consisting only
-of edges  \acar\ and \acdr\  but not \bcar,  \bcdr\ or
-\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
-\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
-consists  of  forward  access  paths  only.  Then  does
-$\alpha \in \sigma$?
-}
-%%\end{quote}
-We model  this problem as one  of deciding the
-membership of  a CFG augmented  with a
-fixed set of unrestricted productions.
-
-\begin{definition}\label{def:specialgrammar}
-Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
-$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
-\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
-context-free    productions     that    contains    the
-distinguished  production   $S  \rightarrow  \alpha\$$,
-$\alpha$ is a  string of grammar symbols  that does not
-contain $S$, and $p_2$ is the fixed set of unrestricted
-productions    $\bcar\acar    \rightarrow    \epsilon$,
-$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
-\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
-and $\clazy\$ \rightarrow \$ $.
-\end{definition}
-
-%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
-%% that the results of liveness analysis of any program can be modeled by
-%% the kind  of grammar described above. The  following proposition shows
-%% that the converse also holds.
-%% \begin{proposition}
-%% Given    a     grammar    $G$    of    the     form    described    in
-%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
-%% program $p$  with program points  $\pi_i$ and variables $x$  such that
-%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
-%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
-%% \end{proposition}
-We first show that membership problem of the class of grammars in
-Definition~\ref{def:specialgrammar} is undecidable. However note that
-the grammars may not be necessarily generated from liveness analysis.
-\newcommand{\state}{\ensuremath{\mathsf{S}}}
-\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
-\begin{lemma}\label{lemma:grammar-undecidable}
-Given a grammar    $G$    of   the    kind    described    in
-Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
-consisting  of symbols \acar\  and \acdr\  only, the membership problem
-$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
-\end{lemma}
-
-\begin{proof}
-Given a Turing machine and  an input $w\in (1+0)^*$, we
-construct a grammar $G$ such that the machine will halt
-on    $w$    if    and   only    if    $\$    \in
-\mathscr{L}(G)$. The grammar includes  the fixed set of
-unrestricted               productions               in
-Definition~\ref{def:specialgrammar}.
-
-We shall represent a Turing machine (TM) configuration 
-as $w_l(\state,c)w_r$, where $w_l$ is the string to the
-left of the head, $w_r$ is the string to the right, $c$
-is the symbol under the head and \state\ is the current
-state of  the TM.   For each combination  of state
-and symbol  $(\state,c)$, the grammar will  contain the
-non-terminal \nont{}{c}. We shall synchronize each move
-of the TM to  a derivation  step using  a context
-free production, followed, if possible, by a derivation
-step using either  $\bcar\acar \rightarrow \epsilon$ or
-$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
-synchronization,  we  shall   establish  the  following
-invariant  relation between  the TM  configuration
-and the sentential form:
-
-\begin{quote}
-  If    the   configuration    of   the    TM   is
-  $w_l(\state,c)w_r$, then the  sentential form will be
-  $\overline{w}_l\nont{}{c}\,w_r\$ $,                where
-  $\overline{w}_l$ is  the same as $w_l$  but with each
-  symbol $d$ in $w_l$ replaced by $\overline{d}$.
-\end{quote}
-
-Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
-$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
-form    corresponding     to    the    initial     configuration    is
-$\nont{\mathit{init}}{c}w\$$ (we  can assume that there  is a production
-$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w\$$, where  \nont{}{} is
-the start symbol of the  grammar). Further correspondences between the
-TM moves and the grammar productions are as follows:
-
-\begin{enumerate}
-\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
-  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
-  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
-\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
-  are two productions
-  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
-  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
-\end{enumerate}
-The idea  behind the  productions is explained  with an
-example:  Assume that  the current  sentential form  is
-$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
-TM  has  a   transition  $(S_i,\acar)  \rightarrow
-(S_j,\acdr, L)$.  Since the next corresponding  step in
-the  derivation  has  to  be  done  without  any  prior
-knowledge of whether the symbol to the left of the tape
-is a \acar\  or a \acdr, two  productions are provided,
-and  the  invariant  will  be maintained  only  if  the
-production         $\nont{i}{\acar}         \rightarrow
-\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
-in  the  derivation.    This  gives  the  configuration
-$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
-Simplification   with    the   production   $\bcdr\acdr
-\rightarrow               \epsilon$              yields
-$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
-corresponds  to   the  changed  configuration   of  the
-TM.  Notice  carefully that a wrong  choice breaks
-the invariant  and it cannot be  recovered subsequently
-by any choice of productions.
-
-After the  TM has  halted, there  are further  ``cleanup'' derivations
-that derive  $\epsilon$ only if  the invariant has been  maintained so
-far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
-$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
-final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
-\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
-$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
-for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
-and                $\nont{\mathit{final}}{c}               \rightarrow
-\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
-\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
-\acdr\ on the right of the tape head.  This completes the
-reduction.\!
-\end{proof}
-
-We now show that the proof can be replayed for the class of grammars
-generated from liveness analysis of programs.
-
-\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
-Given   a    grammar    $G$    of   the    kind    described    in
-Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
-analysis  of a  program  and  a forward  access  path $\alpha$,    the
-membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
-\end{lemma}
-
-\begin{proof}
-Given a Turing machine and an input string, the  proof   of
-Lemma~\ref{lemma:grammar-undecidable} generates a grammar. 
-We shall define a function
-for   each  non-terminal $\nont{i}{c}$   introduced  in   this grammar
-such that the
-liveness  analysis  of  the function will  result
-in a set of productions that includes the productions for
-$\nont{i}{c}$ generated for the proof.
-As an example, it  can be verified that the grammar
-for the function shown in Figure~\ref{fig:turing-pgm}(b)    includes      the
-productions shown in Figure~\ref{fig:turing-pgm}(a). Here $\nont{\mathit{all}}{}$   corresponds   to  the   demand
-$\sigma_{\mathit{all}}$.
-
-The body of the function corresponds  to the RHS of the productions
-for $\nont{i}{c}$. Productions with the same  LHS non-terminal, but
-with differing non-terminal on RHS can be generated by 
-joining program fragments with $\SIF$.   If  there  are  $n$ such  functions,  then
-\mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
-distinct function in each branch.
-
-\begin{figure}[t]
-  \begin{minipage}{.45\columnwidth}
-    \begin{align*}
-      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
-      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
-      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
-      \acdr\nont{\mathit{all}}{}
-    \end{align*}
-  \end{minipage}
-  \begin{minipage}{.45\columnwidth}
-    \input{equivalent-program}
-  \end{minipage}\\
-  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
-  \caption{A possible grammar generated by the proof of
-    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
-  grammar.}\label{fig:turing-pgm}
-\vspace*{-6mm}
-\end{figure}
-
-Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
-had a single non-terminal on the RHS. There are two characteristics of
-the grammar produced from liveness-analysis  that are are relevant for
-replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
-\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
-grammar  generated from  liveness  analysis of the constructed program
-will  have the  production
-$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
-\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
-$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
-\nont{\mathit{all}}{}$.
-
-It is clear that if the TM given as an instance of the halting problem
-accepts the input string, then the earlier derivation can be replayed,
-every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
-sentential form by $\epsilon$. However, if  the TM does not accept the
-input string, then every sentential form derived from the start symbol
-would    have     at    least    one    non-terminal from the grammar
-of Lemma~\ref{lemma:grammar-undecidable} that is different
-from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
-derivable from the grammar.
-\end{proof}
-
-%------------------------------------------------------------%
-\begin{figure}[t!]
-\scalebox{0.8}{
-  \begin{tabular}{cc}
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-3)(50,12)
-  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
-  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-  \ncline{t1}{t2}
-  \putnode{l0}{t1}{7}{2}{\clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{t1}{t1}
-  \putnode{l1}{t1}{0}{8}{\acdr, \clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t2}{t2}
-  \putnode{l2}{t2}{0}{8}{\clazy}
-\end{pspicture}
-&
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-3)(30,12)
-  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
-  \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1}
-  \putnode{l1}{t1}{0}{8}{\acdr}
-\end{pspicture}
-\\
-(a) & (b) \\
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
-  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
-%  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-  \ncline{t1}{t2}
-  \putnode{l0}{t1}{5}{2}{\clazy}
-%  \ncline{t2}{t3}
-%  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
-%  \putnode{l2}{tD1}{0}{10}{\acdr}
- \putnode{tD1}{t2}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \putnode{tD2}{tD1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t2}{tD1}
-  \putnode{l1}{t2}{5}{2}{\bcar}
-  \ncline{tD1}{tD2}
-  \putnode{lD0}{tD1}{5}{2}{\clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{tD1}{tD1}
-  \putnode{lD1}{tD1}{0}{8}{\acdr, \clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{tD2}{tD2}
-  \putnode{lD2}{tD2}{0}{8}{\clazy}
-\end{pspicture}
-&
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(20,8) %\psframe(0,-5)(40,8)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
-  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-\end{pspicture}
-\\ 
-(c) & (d)  \end{tabular}}
-\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
-  converted into an automaton, and (b) its DFA. The same for \var{\Lanv{\pa}{}}
-  are in (c) and (d).}\label{fig:example-automata}
-\vspace*{-2mm}
-\figrule
-\end{figure}
-
-%\subsection{A liveness-based GC scheme} \label{sec:live-clo}
-
-%%\subsubsection{From CFGs to GC-ready DFAs}
-%%\label{sec:NFA-approx}
-
 We circumvent the problem of undecidability by over approximating the
@@
 liveness-based GC triggers beyond $\pi_9$.
-
-  \SetStartEndCondition{ }{}{}%
-  \SetKwProg{Pro}{procedure}{\string:}{}
-  \SetKwProg{Fn}{function}{\string:}{}
-  \SetKwFunction{Range}{range}%%
-  \SetKw{KwTo}{in}\SetKwFor{For}{for}{\string:}{}%
-  \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-  \SetKwFor{While}{while}{:}{fintq}%
-  \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-  \SetKwFunction{Lgc}{$\mathsf{lgc}$}%
-  \SetKwFunction{Copy}{$\mathsf{copy}$}%
-  \SetKwFunction{LCopy}{$lgcCopy$}%
-  \SetKwFunction{RCopy}{$rgcCopy$}%
-
-\begin{algorithm}[t!]
-  \Pro{\Lgc{}}
-     {
-       \For {each reference $\mathsf{ref}$ in root set}
-            {$\mathsf{ref}$ = \Copy($\mathsf{ref}, \mathsf{init}(\mathsf{ref.dfa})$)\;}
-            ${\mathsf{copyReferencesOnPrintStack}()}$\;  
-     }
-     \Fn{\Copy{$\mathsf{ref}, \mathsf{state}$}}
-        {
-          \eIf {$\mathsf{final}(\mathsf{state}$)}
-             {
-               $\mathsf{newRef} = \mathsf{dupHeapCell}(\mathsf{ref})$\;
-               \If{$\mathsf{ref\!.cell}$($\mathsf{ref}$)
-                 is a cons cell $\mathsf{(cons~arg_1~arg_2)}$}
-                  {
-                    {
-                      $\mathsf{newRef}\!.\mathsf{arg_1}  = \Copy(\mathsf{arg_1}, \mathsf{next}(\mathsf{state}, 0))$\;
-%                      $\mathsf{newRef}\!.\mathsf{arg_1}   =  \mathsf{newCar}$\;
-                    }
-                    {
-                      $\mathsf{newRef}\!.\mathsf{arg_2} =  \Copy(\mathsf{arg_2},
-                      \mathsf{next}(\mathsf{state} , 1))$\;
-%                      $\mathsf{newRef}\!.\mathsf{arg_2}   = \mathsf{newCdr}$\;
-                    }    
-                  }
-               \If
-                  {$\mathsf{ref\!.cell}$ is a
-                    closure cell, generically  $\mathsf{(binop~arg_1~arg_2)}$}
-              %      
-                   { 
-                     $\mathsf{newRef}\!.\mathsf{arg_1} = \Copy(\mathsf{arg_1}, \mathsf{init}(\mathsf{arg_1.dfa}))$\;
-                     $\mathsf{newRef}\!.\mathsf{arg_2} = \Copy(\mathsf{arg_2}, \mathsf{init}(\mathsf{arg_2.dfa}))$\;
-                   }
-             }
-             {$\mathsf{newRef = ref}$}
-\KwRet $\mathsf{newRef}$\;
-        }
-        \caption{Liveness-based garbage collection.  \label{algo:lgc-a}}
-\end{algorithm}
-
-\newcommand{\hgt}{2.9cm}
-\newcommand{\LGCLine}{blue}
-\newcommand{\RGCLine}{red}
-\newcommand{\ReachLine}{black}
-\newcommand{\UseLine}{light-blue}
-
-\begin{figure*}[t!]
-\renewcommand{\arraystretch}{.1}
-\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
-%\\ (a) & (b)
-\end{tabular}
-\vspace*{-2mm}
- \caption{Memory usage.  
-The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
-cells  in  the  active   semi-space  for  RGC  and  LGC
-respectively.  The \ReachLine\ curve represents the number of
-reachable cells and the  \UseLine\ curve represents the
-number  of  cells  that  are actually  live  (of  which
-liveness analysis does a static approximation).  x-axis
-is the time measured  in number of cons-cells allocated
-(scaled down by factor $10^5$). y-axis is the number of
-cons-cells (scaled down by $10^3$).}
-\label{fig:memory-usage} \figrule
-\vspace*{-1mm}
-\end{figure*}
-
-%\vspace*{-6.5mm}
-\section{The GC scheme}
-\label{sec:GC-scheme}
-
-Our experimental setup  consists of an interpreter
-for  our  language,  a liveness  analyzer,  and  a
-single generation copying  collector.  The garbage
-collector can be configured to work on the
-basis  of  reachability  (RGC  mode)  or  use  
-liveness  DFAs  (LGC  mode).
-
-In        the       liveness        analysis       described        in
-Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
-variable is  derived from dependence  chains along all paths  from the
-program point  where the closure  was created to  reachable evaluation
-points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
-$\epath_2$ are two such evaluation points.  During GC,
-we would like to use a more precise liveness, based on the actual paths
-taken  during  execution.   Therefore   we  create  separate  liveness
-automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
-addition to  automata for dependences  along paths to  both $\epath_1$
-and $\epath_2$.  The closure carries  the liveness environment for its
-free  variables (as  pointers  to automata,  one  for each  variable).
-Initially the liveness  environment is based on  the dependences along
-both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
-\SIF\ condition, the liveness environments are updated to one based on
-either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
-collections are based on more precise liveness information.
-
-Based on  the above considerations,  we restrict the  possible garbage
-collection points in a function body to the following:
-\begin{enumerate}
-\item We  maintain static estimates  of the memory required  to create
-  the closures for each function body.  On entering a function, if the
-  available memory is  less than this requirement for  the function, a
-  GC is triggered.
-\item Since the evaluation of a  \SIF\ condition may trigger a 
-  collection, after evaluating the condition   the available memory
-  is  checked once  again against  a  revised estimate  of the  memory
-  (based on  value of the condition)  required to execute the  rest of
-  the program.  A GC is triggered if enough memory is unavailable.
-\end{enumerate}
-
-%% While  exploring  the
-%% activation record of a  function, LGC uses the DFA
-%% at  the  latest  program point  traversed  in  the
-%% function's body.   Thus, if LGC is  invoked at the
-%% program point $\pi$, it would use the liveness DFA
-%% at $\pi$ itself for exploring  the root set of the
-%% current   activation   record.   For   any   other
-%% activation record in the stack, say for a function
-%% ${\mathit  f}$   calling  ${\mathit  g}$   in  the
-%% call-chain, the evaluation  point in ${\mathit f}$
-%% which resulted in the  call to ${\mathit g}$ would
-%% be used.
-
-We shall call a  unit of allocatable memory as a {\em  cell}.  A cell can
-hold    a    basic    value    ($\mathsf{bas}$),    the    constructor
-\CONS\ $(\mathsf{cons~  arg_1~arg_2}$) or a closure.   The closure, in
-turn,    can    be    one   of    ($\mathsf{unop~arg}$),    ($\mathsf{
-  binop~arg_1~arg_2}$)  and  function application  ($\mathsf{f~arg})$.
-Here each $\mathsf{  arg_i}$ is a reference to another  heap cell.  In
-addition, the closure also carries  a pointer to a DFA (denoted
-$\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.
-Algorithm~\ref{algo:lgc-a}   describes   the   garbage
-collection scheme.  Starting with the root set, each cell pointed by a
-live reference  (i.e., whose associated  DFA state is final)  is copied
-using $\mathsf{copy}$.   Copying a \CONS\ cells  just involves copying
-the  cell  itself   and  conditionally  copying  the   \CAR\  and  the
-\CDR\ fields after referring to the  next states of the DFA.  If
-the reference points to a closure, then, as noted earlier, the closure
-carries pointers  to the liveness  DFAs of its  arguments.  These
-are  used to recursively  initiate copying of  the arguments.
-Note that  the copying  strategy for  $\mathsf{(unop~arg_1~arg_2)}$ or
-$\mathsf{(f~arg_1)}$ are similar to $\mathsf{(binop~arg_1~arg_2)}$ and
-have not been shown.
-
-\begin{table*}[t!]
-\caption{Statistics for liveness analysis and garbage collection}
-\label{tab:exp-results}
-\vskip -2mm
-\centering
-\include{data-liveness-analysis}
-\vskip -5mm
-\include{autotable}
-\vskip -7mm
-\end{table*}
-
-%\subsection{Garbage collection for references on print stack}
-The evaluation of  the top-level expression in a program  is driven by
-a printing function~(Section
-\ref{sec:semantics}).  
-%% Printing of  atomic values is simple  and does
-%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
-%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
-%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
-%% trigger  more GCs.  When  this happens.   we have  to
-%% consider any  references that  might be  on the  print stack  and copy
-%% them.
-We extend liveness-based  garbage  collection  to
-this function.
-\section{Experimental evaluation}
-\label{sec:experiments}
-Our benchmark  consists of programs taken  from nofib~\cite{nofib} and
-other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
-  huffman-sicp}.     We    ran    the   experiments    on    8    core
-Intel\textsuperscript{\textregistered}        Core\textsuperscript{TM}
-i7-4770 3.40GHz CPU  having 8192KB L2 cache, 16GB RAM,  running 64 bit
-Ubuntu 14.04.
-
-%% The  process  of  liveness-based  garbage  collection  involves  going
-%% through  each  activation  record  on the  stack  and  exploring  each
-%% variable in the current activation  record (root set). The liveness of
-%% each variable is  determined using the program point  and the variable
-%% name. The liveness  of all variables are stored as  DFA. These DFA are
-%% then encoded as a table. A variable  is live only if an entry is found
-%% in  this table.   All  cells that  are  live are  copied  to the  live
-%% semi-space. In case  of \CONS\ cells the \CAR\ and  \CDR\ pointers are
-%% chased  and if  they are  live  the copied  \CONS\ cell  will get  the
-%% updated addresses of its \CAR\ and \CDR\ fields.
-
-\subsection{Results}
-
-The statistics  related to  liveness analysis  and DFA  generation are
-shown in Table~\ref{tab:exp-results}(a). We  observe that the analysis
-of  all  programs  except \verb@treejoin@  and  \verb@sudoku@  require
-reasonable time.   The bottleneck in  our analysis  is the NFA  to DFA
-conversion with  worst-case exponential behaviour. However,  since the
-analysis has  to be  done only  once and its  results can be cached and
-re-used, the time spent in analysis may be considered acceptable.
-
-Table~\ref{tab:exp-results}(b) compares  GC statistics
-for RGC and  LGC.  We report the number of  GC events,
-average  number of  cells  reclaimed per  GC,  average
-number of cells  touched per GC and the  total time to
-perform all collections.   It is no surprise that the  number of cells
-reclaimed per garbage  collection is higher and the  number of garbage
-collections  lower  for  LGC.   The  cost of  LGC  is  higher  garbage
-collection time, which increases the  overall execution time even with
-reduced number of  collections.  However, the execution time  of LGC is
-still comparable for most benchmarks (slowdown within 5X of
-RGC in most cases) and better  for 3 benchmarks (2X speedup in the
-best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
-benchmark that allocates complete binary
-trees  of various  sizes that  are  never used  by the  program. As  a
-result,  the benchmark  highly favours  LGC.  The  benchmark has  been
-included  for  completeness,  and  we  do  not  consider  its  numbers
-as being representative of real programs.
-
-Memory    usage    graphs    for     the    benchmarks    are    shown
-Figure~\ref{fig:memory-usage}.  In  all the  programs we can  see that
-the curve  corresponding to  LGC (\LGCLine\ line)  dips below  the RGC
-curve (\RGCLine\ line)  during GC.  The graphs also  include the curve
-for reachable cells (\ReachLine) and live cells (\UseLine). These were
-obtained by forcing  RGC to run at very high  frequency. The curve for
-live  cells were  obtained by  recording  heap access  times and  post
-processing the data at  the end of the program. Note  that the size of
-an LGC cell is 1.16 times the size of a RGC cell.
-
-As  demonstrated by  the gap  between the  \RGCLine\ and  the \UseLine\
-lines, a  large number of  cells which are  unused by the  program are
-still copied  during RGC. LGC does  a much better job  of closing this
-gap but still falls short of the  precision achieved by LGC in case of
-eager languages ~\cite{asati14lgc}.  A major source of inefficiency in
-LGC is  multiple traversals  of already copied  heap cells.  Since LGC
-does not mark the heap cells after the first visit, the same cells can
-be  repeatedly  visited  with   different  liveness  states.  We  have
-mitigated  this  problem  by   implementing  a  heuristic  which  tags
-\CONS\ cells with the set of states already visited.
-
-\vspace*{-2mm}
-\section{Related work} 
-\label{sec:relatedwork}
-The impact of  liveness on the effectiveness of  GC is
-investigated in~\cite{Hirzel}.   They observe that  liveness can
-significantly  impact   garbage  collection,  but  only   when  it  is
-interprocedural.  As far as memory requirement is concerned, our paper
-demonstrates this observation.
-
-There have been  several attempts to use liveness  analysis to improve
-GC for imperative  languages.~\cite{khedker07heap} presents a liveness
-analysis and uses  the results for inserting  nullifying statements in
-Java  programs. In~\cite{ran.shaham-sas03},  temporal properties  like
-liveness  are checked  against an  automaton modeling  heap accesses.
-Both these approaches are intraprocedural in scope.
-
-In the space of functional languages, there are: rewriting methods
-such as
-deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-sharing analysis based reallocation~\cite{jones89compile}, region
-based analysis~\cite{tofte98region}, and insertion of compile-time
-nullifying statements~\cite{inoue88analysis, lee05static}.  All compile-time
-marking approaches rely on an efficient and precise alias analysis and
-cannot provide significant improvement in its absence. The only work
-in the space of lazy languages seems to be~\cite{Hamilton}
-which touches upon only basic techniques of compile-time garbage
-marking, explicit deallocation and destructive allocation. An
-interesting approach suggested in~\cite{HofmannJ03} is to annotate the
-heap usage of first-order programs through linear types.  The
-annotations are then used to serve memory requests through
-re-allocation.  However, this requires the user to write programs in a
-specific way.
-%
-Safe-for-space~\cite{appel.cps}              approaches~\cite{Clinger,
-  shao00efficient}  reduce the  amount of  heap used  by a  program by
-allocating   closures    in   registers   and   through    tail   call
-optimizations. However, these approaches take care of only part of the
-problem addressed  by our  analysis as the  program can  still contain
-unused objects and closures that are reachable.
-%
-
-Simplifiers~\cite{ONeill}  are  abstractly  described  as  lightweight
-daemons that  attach themselves to  program data and,  when activated,
-improve the  efficiency of the  program. Our liveness-based GC  can be
-seen as  an instance  of a  simplifier which  is tightly  coupled with
-garbage  collectors.   The approach  that  is  closest to  the  method
-described  in  this  paper  is the  liveness-based  garbage  collector
-implemented  in~\cite{karkare07liveness,asati14lgc} and  address eager
-languages. We extend this to handle lazy evaluation and closures.
-
-%% A similar  approach has been suggested  in~\cite{ONeill} where they
-%% suggest   augmenting  GC   with   extra   information  using   {\it
-%% simplifiers}.
-
-%% Previous attempts  to increase the  space efficiency
-%% of functional programs  by additional reclamation of
-%% memory fall  in two broad categories.  In the first,
-%% the  program   itself  is  instrumented   to  manage
-%% reclamation and reallocation without  the aid of the
-%% GC.   Such attempts  include: sharing
-%% analysis  based  reallocation~\cite{jones89compile},
-%% deforestation
-%% techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
-%% methods based on linear logic~\cite{hofmann00linear}
-%% and region analysis~\cite{tofte98region}.  Closer to
-%% our  approach, there  are  methods  that enable  the
-%% garbage      collector      to     collect      more
-%% garbage~\cite{inoue88analysis,lee05static}        by
-%% explicitly  nullifying pointers  that are  not live.
-%% However,  the nullification,  done at  compile time,
-%% requires sharing  (alias) analysis.  Our  method, in
-%% contrast, does  not require  sharing because  of the
-%% availability of the heap  itself at runtime.  To the
-%% best of our knowledge, this  is the first attempt at
-%% liveness-based  marking of  the heap  during garbage
-%% collection.
-
-%% \begin{figure}[t]
-%% \centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
-%%  \caption{Memory  usage and garbage collection pattern of  {\tt
-%% lambda}.}
-%% \label{fig:memory-usage-lambda} \figrule
-%% \end{figure}
-\vspace*{-2mm}
-\vskip -5mm
-\section{Future work and conclusions}
-\label{sec:conclusion}
-%% We have  defined a notion of  liveness on structured
-%% data; this generalizes classical liveness and strong
-%% liveness.    We  started   with   a  general   fully
-%% context-sensitive analysis  which we  proved correct
-%% with respect  to a minefield semantics  (this models
-%% the  effect  of  garbage  collection  between  every
-%% evaluation step).
-
-%% To avoid scalability issues (and to avoid performing
-%% part  of the  liveness computation  at run  time) we
-%% defined an  0-CFA version of this  liveness analysis
-%% in  which demands  for function  $f$ at  all calling
-%% contexts   are  conflated   into  a   single  demand
-%% $\sigma_f$.  This  enabled us to treat  the liveness
-%% equations   symbolically    obtaining   context-free
-%% grammars  for liveness  at each  GC point  (calls to
-%% user  functions and  to $\CONS$).   These were  then
-%% converted to  DFA for  run-time consultation  by the
-%% garbage   collector.     Experiments   confirm   the
-%% precision of the analysis.
-
-%% To   obtain  performance   figures  we   compared  a
-%% reachability-based   garbage    collector   with   a
-%% liveness-based collector.  This showed a decrease in
-%% the  number  of  GCs,  more  garbage  collected  per
-%% invocation.  A  significant benefit  of LGC  is that
-%% programs can run in  smaller memory when compared to
-%% RGC. This is potentially  useful in situations where
-%% memory is limited---as with embedded systems.  For a
-%% majority of  programs, the garbage  collection times
-%% were reduced.
-
-%% One issue we highlighted  was that while fewer nodes
-%% were marked (and  more garbage collected), sometimes
-%% cons cells  could be visited and  traversed multiple
-%% times  with  different  sets of  liveness  paths  to
-%% explore.  Further work  includes improvements to the
-%% classical copying  collector to  reduce the  cost of
-%% this.
-We have  extended liveness-based  GC  to lazy
-  languages  and shown  its benefit  for a  set of  benchmark
-  programs.  We  defined a liveness analysis  of programs manipulating
-  the  heap and  proved  it  correct with  respect  to a  non-standard
-  semantics that served as a specification of liveness.  The result of
-  the analysis  is a  set of grammars,  whose membership  question was
-  shown to be undecidable. The  grammars are therefore approximated by
-  DFAs and used by the garbage collector to improve collection.
-
-In addition to  evaluated data, our collector  also reclaims closures.
-For this, we  had to modify the standard run-time  representation of a
-closures  to carry  liveness  of its free  variables and  periodically
-update the liveness during execution  time.  As expected, our garbage
-collector  reclaims more  garbage and  reduces memory  requirement of
-programs.  An  additional benefit is  that in  spite of using  a more
-expensive collector, the execution times remains comparable in
-most cases, and even improves for some programs.
-
-The graphs show that even in lazy languages there are large amount of
-dead cells that can be collected early. In spite of collecting
-closures there still exists a gap between actual and perceived
-livenesses.  Language features that could be introduced are
-higher-order functions and recursive lets. We believe that functions
-can be used to model recursive lets and this should be a simple
-extension to our liveness analysis.  For higher-order we intend to use
-firstification techniques~\cite{Mitchell:2009}.
-
-  Orthogonally,   we   plan  to   improve   the   efficiency  of   the
-  liveness-based garbage  collector using heuristics such  as limiting
-  the depth of DFA, merging  nearly-equivalent states and using better
-  representation(for example BDDs~\cite{Bryant86})  and algorithms for
-  automata manipulation.  We also  need to investigate the interaction
-  of liveness with  other collection schemes, such  as incremental and
-  generational collection.   It might  be interesting  to use  a mixed
-  mode GC  scheme which allows the  costs of LGC to  be amortized over
-  several runs of RGC. In summary, we need to investigate ways to make
-  liveness-based GC attractive for practical collectors.
-
-\clearpage
-\balance
-\bibliography{fun_hra}{}
-\bibliographystyle{abbrv}
-\clearpage
-
-\appendix
-%\section{Appendix: not to be included in the paper.}
-%%%%% prasanna minefield
-\begin{figure*}[b!]
-\noindent{\Large \bf A. Complete minefield semantics}
-\begin{center}\footnotesize
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|c|c|c|}
-\hline
-Premise & Transition & Rule name \\ 
-\hline
-\hline
-&\makecell{ $\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S,
-  \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell :=
-    \kappa], e, \cred{\sigma'}$ }   &  {\sc const}
-\\
-\hline
-& \makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
-  (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $
-  \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  {\sc cons} \\
-\hline
-\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
-{\sc car-bang} 
-\\
-\hline
-\makecell[t]{$\heap(\rho(x)) \mbox{ is } (v, d)$} & \makecell[t]{$\rho, (\rho', \ell, e,
-  \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
-  \rightsquigarrow $ \\ $\rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
-{\sc car-select} \\
-\hline
-\makecell[t]{ $\heap(\rho(x))
-\mbox{ is } \langle s, \rho'\rangle$ }& \makecell[t]{$\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow $\\ $  \rho', (\rho, \rho(x),
-  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$ }        &
-{\sc car-clo}\\
-\hline
-\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
-(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
-s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
-{\sc car-1-clo} \\
-%% \hline
-
-%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
-%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
-%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
-%% {\sc car-clo}
-%% \\
-\hline
-\makecell[t]{$\rho(x) \mbox{ is } \bot$  or $\rho(y) \mbox{ is }
-  \bot$} & $\rho, S,
-  \heap, (+~x~y), \cred{\sigma} \rightsquigarrow \bang$   &
-{\sc prim-bang} 
-\\
-\hline
-
-\makecell[t]{$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$}
- &
-
-\makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
-  (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $\rho', S, \heap[\ell
-    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}     
- &
-{\sc prim-add} \\
-\hline
-\makecell[t]{$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
-  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho, \rho(x),
-  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-{\sc prim-1-clo} \\
-\hline
-\makecell[t]{$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
-  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho,\rho(y),
-  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
-{\sc prim-2-clo} \\
-\hline
-\makecell[t]{$\mathit{f}~\mbox{defined as}$
-$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell[t]{$\rho, S, \heap,
-  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow$\\$ [\myvec{y} \mapsto
-    \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
-{\sc funcall} \\
-\hline
-\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
-  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
-    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
-    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
-{\sc let} \\ 
-\hline
-\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
-  \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
-{\sc if-bang} 
-\\
-\hline
-\makecell[t]{$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
-\hline
-\makecell[t]{$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
-  \cred{\sigma}$   $\rightsquigarrow
-\rho, S, \heap,  e_2, \cred{\sigma}$} & {\sc if-false} \\
-\hline
-\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
-  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow$ \\ $
-\rho', (\rho, \rho(x), (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
-\cred{\clazy\sigma}$}
-&
-{\sc if-clo} \\
-\hline
-\makecell[t]{
-  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
-  \heap, (\SRETURN~x), \cred{\sigma} \rightsquigarrow \bang$   &
-{\sc return-bang} 
-\\
-\hline
-\makecell[t]{$\heap(\rho(x))~\mbox{is in WHNF with
-    value}~v$} & \makecell[t]{$\rho,
-  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow$\\$ \rho', S, \heap[\ell := v], e, 
-  \cred{\sigma'}$} &
-{\sc return-whnf}\\
-\hline
-\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
-  (\SRETURN~x), \cred{\sigma}$  $
-  \rightsquigarrow$ \\
-$\rho',~ (\rho, \rho(x), (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
-  \cred{\sigma}$} &
-{\sc return-clo} \\
-\hline
-\end{tabular}
-
-\caption {Minefield semantics. The differences with the small-step
-  semantics have been highlighted by shading. \label{fig:minefield-semantics}}
-\end{center}
-
-\end{figure*}
-
-\end{document}
-
-%%  LocalWords:  dependences subscripting

### 2016-04-13 paper/paper.tex
created file cfg-to-nfa.tex and added wrapfig for fig4

@@
 \usepackage{balance}
+\usepackage{wrapfig}
 
@@
 
-\begin{figure}[t]
+%\begin{figure}[t]
   \input{example-to-illustrate-liveness}  
-  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
-\end{figure}
+%  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
+%\end{figure}
 \input{fig-liveness}
@@
 
-We circumvent the problem of undecidability by over approximating the
-CFG by non-deterministic finite state automata (NFA) using
-%% The NFAs  are then simplified using the
-%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
-%%are converted to DFAs.
-%%
-%% We use  the algorithm by 
-Mohri  and Nederhof~\cite{mohri00regular} method.
-%to approximate a CFG by a {\em strongly regular\/} grammar.  
-For example, the grammar fragment for %the non-terminal
-$\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
- \begin{eqnarray*}
-   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
-   \acdr\var{\Df{\length}{1}}
-   \mid \clazy\var{\Df{\length}{1}}\\
-   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
-   \mid \epsilon
- \end{eqnarray*}
-
-The strongly regular grammar is converted  into a set of NFAs, one for
-each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
-done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
-pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
-constructing  the $\epsilon$-closure  till a  fixed point  is reached,
-after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  
-\cmt{ %% REMOVED FOR BLIND REVIEW
-  The details of the  algorithm, its correctness and  termination proofs are
-given  in~\cite{karkare07liveness,asati14lgc}. }
-The
-resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
-\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
-check if the source node of the edge has a path to a final 
-state.  If it does, we mark  the source node as final. Finally, we
-remove all the edges labeled \clazy\  and convert the automaton into a
-deterministic    automaton.    These steps   effectively   implement    the
-$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
-\clazy\ to  obtain forward access paths.  While checking for  liveness during
-garbage  collection, a  forward access path  is valid  only if  it can
-reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
-NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
-final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
-expectedly  says that for a  demand  $\sigma_{\length}$,  the
-liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
-traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
-NFA     for     \var{\Lanv{\pa}{}}.      The     DFA in
-Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
-paths, reflecting  the  lazy  nature  of  our  language.   Since
-\length\ does not  evaluate the elements of  the argument list,
-the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
-liveness-based GC triggers beyond $\pi_9$.
+\input{cfg-to-nfa}
 

### 2016-04-13 paper/cfg-to-nfa.tex
Updated cfg to nfa conversion with example

@@
 %% We use  the algorithm by 
-Mohri  and Nederhof~\cite{mohri00regular} method.
+Mohri  and Nederhof~\cite{mohri00regular} method. This method transforms 
+a given CFG into a strongly regular grammar. A strongly regular grammar 
+is a grammar in which mutually recursive rules are all either right-linear
+or left-linear grammars.
+
+In our example, the rule  $\var{\Df{\length}{1}}$ has two context free
+productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
+$\clazy\var{\Df{\length}{1}}{\clazy}$.   We  add another  non-terminal
+$\var{\Df{\length}{1'}}$ to the grammar and we split each context free
+production at the point where it refers to $\var{\Df{\length}{1}}$ and
+the  remaining  symbols are  then  added  as  productions of  the  new
+variable   $\var{\Df{\length}{1'}}$.   For   the  regular   production
+$\clazy$,  we  add $\var{\Df{\length}{1'}}$  at  the  end and  add  an
+$\epsilon$ transition to $\var{\Df{\length}{1'}}$.
 %to approximate a CFG by a {\em strongly regular\/} grammar.  
-For example, the grammar fragment for %the non-terminal
+The resulting grammar fragment for %the non-terminal
 $\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
@@
 after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  
+ \begin{wrapfigure}{r}{0.25\textwidth}
+%\begin{figure}[t!]
+\scalebox{0.5}{
+  \begin{tabular}{cc}
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-3)(50,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
+  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+  \ncline{t1}{t2}
+  \putnode{l0}{t1}{7}{2}{\clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{t1}{t1}
+  \putnode{l1}{t1}{0}{8}{\acdr, \clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t2}{t2}
+  \putnode{l2}{t2}{0}{8}{\clazy}
+\end{pspicture}
+&
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-3)(30,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
+  \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1}
+  \putnode{l1}{t1}{0}{8}{\acdr}
+\end{pspicture}
+\\
+(a) & (b) \\
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
+  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
+%  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+  \ncline{t1}{t2}
+  \putnode{l0}{t1}{5}{2}{\clazy}
+%  \ncline{t2}{t3}
+%  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
+%  \putnode{l2}{tD1}{0}{10}{\acdr}
+ \putnode{tD1}{t2}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \putnode{tD2}{tD1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->}
+  \ncline{t2}{tD1}
+  \putnode{l1}{t2}{5}{2}{\bcar}
+  \ncline{tD1}{tD2}
+  \putnode{lD0}{tD1}{5}{2}{\clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{tD1}{tD1}
+  \putnode{lD1}{tD1}{0}{8}{\acdr, \clazy}
+  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{tD2}{tD2}
+  \putnode{lD2}{tD2}{0}{8}{\clazy}
+\end{pspicture}
+&
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(20,8) %\psframe(0,-5)(40,8)
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
+  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+  \psset{arrows=->}
+  \ncline{t0}{t1}
+\end{pspicture}
+\\ 
+(c) & (d)  \end{tabular}}
+\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
+  converted into an automaton, and (b) its DFA. The same for \var{\Lanv{\pa}{}}
+  are in (c) and (d).}\label{fig:example-automata}
+\vspace*{-2mm}
+\figrule
+%\end{figure}
+\end{wrapfigure}
 \cmt{ %% REMOVED FOR BLIND REVIEW
@@
 garbage  collection, a  forward access path  is valid  only if  it can
-reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
+reach a  final state. 
+Figure~\ref{fig:example-automata}(a)  shows the
 NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
@@
 liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
-traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
+traversed).
+Similarly, Figure~\ref{fig:example-automata}(c) shows the
 NFA     for     \var{\Lanv{\pa}{}}.      The     DFA in

### 2016-04-13 paper/paper.tex
Updated cfg to nfa conversion with example

@@
 %------------------------------------------------------------%
-\begin{figure}[t!]
-\scalebox{0.8}{
-  \begin{tabular}{cc}
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-3)(50,12)
-  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
-  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-  \ncline{t1}{t2}
-  \putnode{l0}{t1}{7}{2}{\clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{t1}{t1}
-  \putnode{l1}{t1}{0}{8}{\acdr, \clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t2}{t2}
-  \putnode{l2}{t2}{0}{8}{\clazy}
-\end{pspicture}
-&
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-3)(30,12)
-  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
-  \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1}
-  \putnode{l1}{t1}{0}{8}{\acdr}
-\end{pspicture}
-\\
-(a) & (b) \\
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
-  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
-%  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-  \ncline{t1}{t2}
-  \putnode{l0}{t1}{5}{2}{\clazy}
-%  \ncline{t2}{t3}
-%  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
-%  \putnode{l2}{tD1}{0}{10}{\acdr}
- \putnode{tD1}{t2}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \putnode{tD2}{tD1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
-  \psset{arrows=->}
-  \ncline{t2}{tD1}
-  \putnode{l1}{t2}{5}{2}{\bcar}
-  \ncline{tD1}{tD2}
-  \putnode{lD0}{tD1}{5}{2}{\clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{tD1}{tD1}
-  \putnode{lD1}{tD1}{0}{8}{\acdr, \clazy}
-  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{tD2}{tD2}
-  \putnode{lD2}{tD2}{0}{8}{\clazy}
-\end{pspicture}
-&
-\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(20,8) %\psframe(0,-5)(40,8)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
-  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
-  \psset{arrows=->}
-  \ncline{t0}{t1}
-\end{pspicture}
-\\ 
-(c) & (d)  \end{tabular}}
-\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
-  converted into an automaton, and (b) its DFA. The same for \var{\Lanv{\pa}{}}
-  are in (c) and (d).}\label{fig:example-automata}
-\vspace*{-2mm}
-\figrule
-\end{figure}
 

### 2016-04-13 paper/paper.tex
Moved soundness proof to appendix

@@
 \ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
-e_{\mainpgm})$.
-
-\subsection{Soundness result}
-We first need an auxiliary  result about minefield semantics. Consider
-a trace of  a minefield execution.  For every  minefield state $(\rho,
-S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
-$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
-application $s$) is  non-null.  This can be proved by  an induction on
-the number  of steps leading  to the  minefield state.  The  base step
-holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
-non-null. For the inductive step we  observe that for each step of the
-minefield semantics, if the demand $\sigma$  on the LHS of a minefield
-step  is non-null,  the  demand  on the  RHS  is  a transformation  of
-$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.
-
-Note that our proofs will be for a single round of minefield execution
-i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
-\ref{sec:semantics}).  With  minor variations,  the proof will  also be
-applicable for subsequent evaluations initiated by $\print$.
-%% We show  next that the  minefield execution of a  call-less program
-%% cannot go \bang.
-
-\begin{lemma}
-\label{lemma:call-less-cannot-go-bang}
-Consider the minefield execution of  a program without function calls.
-Such a program cannot enter the \bang\ state.
-\end{lemma}
-\begin{proof}
-Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
-minefield execution of a program.  We  show by induction on the number $n$
-of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
-enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
-(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
-[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
-does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
-transition cannot  enter a  \bang\ state.  Since  our programs  are in
-ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
-does not involve dereferencing, and thus cannot result in a \bang.
-
-For the inductive step, we shall show that none of the minefield steps
-that involves dereferencing  results in a \bang.  These  are the steps
-which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
-$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
-inserted by an earlier $GC(\ldots)$.   However the demand $\sigma'$  on basis of
-which the $GC(\ldots)$ would have inserted a  $\bot$ would have
-included the  current demand $\sigma$.     Thus it is
-enough to  show that the $\rightsquigarrow$ step would not lead to a
-\bang, even if $GC(\ldots)$ had been done with the current demand
-$\sigma$.
-
-We consider the rules for \CAR\  only.  
-The rest  of the rules involve similar reasoning. For the {\sc
-  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
-\sigma$, we  know that $\sigma$ is  non-null. Therefore
-the  liveness  of  $x$  includes  $\epsilon$,  and  the
-dereferencing $\heap(\rho(x))$ will go without \bang.
-
-For the  {\sc car-1-clo}  rule, observe that  there are
-two dereferences.  First $x$  is dereferenced to  get a
-cons  cell  and then  the  head  of  the cons  cell  is
-dereferenced  to  obtain  a  closure.   If  the  demand
-$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
-of    $x$   will    include    both   $\epsilon$    and
-$\acar\epsilon$,  and  a  GC with  this  liveness  will
-neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
-first   component  of   the   cons   cell.  Thus   both
-dereferences    can take place without 
-$\rightsquigarrow$ entering the   \bang\ state.
-\end{proof}
-
-Now we are ready to prove the main soundness result.
-
-\begin{theorem}
-The  minefield  execution of  no  program  can enter  a
-\bang\ state.
-\end{theorem}
-
-\begin{proof}
-Assume to  the contrary that  a program $P$  enters the
-\bang\  state.  We can  transform  $P$  to a  call-less
-program $P'$ such that  the minefield executions of $P$
-and $P'$  are identical  except for change  of variable
-names.             However,                 by
-Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
-$P'$,   a   call-less   program,   cannot   enter   the
-\bang\  state.  Therefore   $P$  also  cannot  enter  the
-\bang\ state.
-\end{proof}
+e_{\mainpgm})$. The proof that such a program cannot go \bang\ is provided
+in Appendix~\ref{sec:sound}.
 
@@
 \end{eqnarray*}
-As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
-stack variables $\pl$ and $\pa$ have the following liveness in terms
-of  $\sigma_{\length}$:
-\begin{align*}
-  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
-%\end{align*}
-%\begin{align*}
-  \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
-  \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
-  \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
-  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
-  \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
-\end{align*}
-
-In  summary, the  equations  generated during the
-analysis are:
-\begin{enumerate}
-\item   For  each   function  $\mathit{f}$,   equations
-  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-\item For each function $\mathit{f}$, an equation defining the summary
-  demand $\sigma_{\mathit f}$ on $e_f$.
-\item   For  each   function  $\mathit{f}$   (including
-  $\mainpgm$) an  equation  defining
-  liveness for each garbage in  $e_{\mathit f}$.
-\end{enumerate}
+
+%% COMMENTED OUT BY PRASANNA
+%% As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
+%% stack variables $\pl$ and $\pa$ have the following liveness in terms
+%% of  $\sigma_{\length}$:
+%% \begin{align*}
+%%   \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
+%% %\end{align*}
+%% %\begin{align*}
+%%   \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
+%%   \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
+%%   \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
+%%   \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
+%%   \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
+%% \end{align*}
+
+%% In  summary, the  equations  generated during the
+%% analysis are:
+%% \begin{enumerate}
+%% \item   For  each   function  $\mathit{f}$,   equations
+%%   defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
+%% \item For each function $\mathit{f}$, an equation defining the summary
+%%   demand $\sigma_{\mathit f}$ on $e_f$.
+%% \item   For  each   function  $\mathit{f}$   (including
+%%   $\mainpgm$) an  equation  defining
+%%   liveness for each variable in  $e_{\mathit f}$.
+%% \end{enumerate}
 
@@
 \begin{eqnarray*}
-  \var{\Lanv{\xl}{}}  &\rightarrow& 
-  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
-  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
+  %% \var{\Lanv{\xl}{}}  &\rightarrow& 
+  %% \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
+  %% \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
   \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
@@
 &\rightarrow&
-\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
-\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
+\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle 
+%% \\
+%% \var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
 \end{eqnarray*}
@@
 \appendix
+\section{Soundness result for liveness analysis under minefield semantics}
+\label{sec:sound}
+We first need an auxiliary  result about minefield semantics. Consider
+a trace of  a minefield execution.  For every  minefield state $(\rho,
+S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
+$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
+application $s$) is  non-null.  This can be proved by  an induction on
+the number  of steps leading  to the  minefield state.  The  base step
+holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
+non-null. For the inductive step we  observe that for each step of the
+minefield semantics, if the demand $\sigma$  on the LHS of a minefield
+step  is non-null,  the  demand  on the  RHS  is  a transformation  of
+$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.
+
+Note that our proofs will be for a single round of minefield execution
+i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
+\ref{sec:semantics}).  With  minor variations,  the proof will  also be
+applicable for subsequent evaluations initiated by $\print$.
+%% We show  next that the  minefield execution of a  call-less program
+%% cannot go \bang.
+
+\begin{lemma}
+\label{lemma:call-less-cannot-go-bang}
+Consider the minefield execution of  a program without function calls.
+Such a program cannot enter the \bang\ state.
+\end{lemma}
+\begin{proof}
+Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
+minefield execution of a program.  We  show by induction on the number $n$
+of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
+enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
+(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
+[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
+does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
+transition cannot  enter a  \bang\ state.  Since  our programs  are in
+ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
+does not involve dereferencing, and thus cannot result in a \bang.
+
+For the inductive step, we shall show that none of the minefield steps
+that involves dereferencing  results in a \bang.  These  are the steps
+which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
+$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
+inserted by an earlier $GC(\ldots)$.   However the demand $\sigma'$  on basis of
+which the $GC(\ldots)$ would have inserted a  $\bot$ would have
+included the  current demand $\sigma$.     Thus it is
+enough to  show that the $\rightsquigarrow$ step would not lead to a
+\bang, even if $GC(\ldots)$ had been done with the current demand
+$\sigma$.
+
+We consider the rules for \CAR\  only.  
+The rest  of the rules involve similar reasoning. For the {\sc
+  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
+\sigma$, we  know that $\sigma$ is  non-null. Therefore
+the  liveness  of  $x$  includes  $\epsilon$,  and  the
+dereferencing $\heap(\rho(x))$ will go without \bang.
+
+For the  {\sc car-1-clo}  rule, observe that  there are
+two dereferences.  First $x$  is dereferenced to  get a
+cons  cell  and then  the  head  of  the cons  cell  is
+dereferenced  to  obtain  a  closure.   If  the  demand
+$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
+of    $x$   will    include    both   $\epsilon$    and
+$\acar\epsilon$,  and  a  GC with  this  liveness  will
+neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
+first   component  of   the   cons   cell.  Thus   both
+dereferences    can take place without 
+$\rightsquigarrow$ entering the   \bang\ state.
+\end{proof}
+
+Now we are ready to prove the main soundness result.
+
+\begin{theorem}
+The  minefield  execution of  no  program  can enter  a
+\bang\ state.
+\end{theorem}
+
+\begin{proof}
+Assume to  the contrary that  a program $P$  enters the
+\bang\  state.  We can  transform  $P$  to a  call-less
+program $P'$ such that  the minefield executions of $P$
+and $P'$  are identical  except for change  of variable
+names.             However,                 by
+Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
+$P'$,   a   call-less   program,   cannot   enter   the
+\bang\  state.  Therefore   $P$  also  cannot  enter  the
+\bang\ state.
+\end{proof}
+
 \section{Proof of Undecidability}

### 2016-04-13 paper/cfg-to-nfa.tex
Split example program into 2 parts and updated mohri-nederhof

--- 2016-04-13 paper/cfg-to-nfa.tex
@@
 %% We use  the algorithm by 
-Mohri  and Nederhof~\cite{mohri00regular} method.
-%to approximate a CFG by a {\em strongly regular\/} grammar.  
-For example, the grammar fragment for %the non-terminal
-$\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
- \begin{eqnarray*}
-   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
-   \acdr\var{\Df{\length}{1}}
-   \mid \clazy\var{\Df{\length}{1}}\\
-   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
-   \mid \epsilon
- \end{eqnarray*}
+Mohri   and   Nederhof~\cite{mohri00regular}   method.   This   method
+transforms a  given CFG  into a strongly  regular grammar.
+
+In our example, the rule  $\var{\Df{\length}{1}}$ has two context free
+productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
+$\clazy\var{\Df{\length}{1}}{\clazy}$.  We transform this grammar into
+a strongly regular grammar in two steps,
+
+\begin{enumerate}
+\item
+Add a  new non-terminal  $\var{\Df{\length}{1'}}$ to the  grammar with
+the rule $\var{\Df{\length}{1'}} \rightarrow \epsilon$.
+\item
+Consider each production of $\var{\Df{\length}{1}}$,
+\begin{enumerate}
+\item
+ replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$,     by
+ $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1'}}$
+\item
+ replace
+ $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1}}\clazy$,
+ by  $\var{\Df{\length}{1}}  \rightarrow  \clazy\var{\Df{\length}{1}}$
+ and $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
+\item
+ replace
+ $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}\clazy$,
+ by $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}$ and
+ $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
+\end{enumerate}
+\end{enumerate}
+
+The    resulting    grammar    fragment    for    %the    non-terminal
+$\var{\Df{\length}{1}}$   after   the  Mohri-Nederhof   transformation
+is:               \begin{eqnarray*}              \var{\Df{\length}{1}}
+&\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \acdr\var{\Df{\length}{1}} \mid \clazy\var{\Df{\length}{1}}\\ \var{\Df{\length}{1'}}
+&\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \epsilon \end{eqnarray*}
+Notice that the language generated  by the strongly regular grammar is
+a superset of the original CFG.
 
@@
 constructing  the $\epsilon$-closure  till a  fixed point  is reached,
-after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  
-\cmt{ %% REMOVED FOR BLIND REVIEW
-  The details of the  algorithm, its correctness and  termination proofs are
-given  in~\cite{karkare07liveness,asati14lgc}. }
-The
-resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
+after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
+details of the  algorithm, its correctness and  termination proofs are
+given in~\cite{karkare07liveness,asati14lgc}.
+
+The resulting automaton has edges labeled with \acar, \acdr\ and
 \clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
-check if the source node of the edge has a path to a final 
-state.  If it does, we mark  the source node as final. Finally, we
-remove all the edges labeled \clazy\  and convert the automaton into a
-deterministic    automaton.    These steps   effectively   implement    the
-$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
+check if the source node of the edge  has a path to a final state.  If
+it does, we mark the source node  as final. Finally, we remove all the
+edges labeled \clazy\  and convert the automaton  into a deterministic
+automaton.   These steps  effectively implement  the $\hookrightarrow$
+simplification rules for \bcar, \bcdr, and
 \clazy\ to  obtain forward access paths.  While checking for  liveness during
 garbage collection,  a forward  access path  is valid  only if  it can
-reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
-NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
-final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
-expectedly  says that for a  demand  $\sigma_{\length}$,  the
-liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
-traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
-NFA     for     \var{\Lanv{\pa}{}}.      The     DFA in
-Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
-paths, reflecting  the  lazy  nature  of  our  language.   Since
+reach a final state.
+
+ \begin{wrapfigure}{r}{0.25\textwidth}
+%\begin{figure}[t!]
+\scalebox{0.5}{
+  \begin{tabular}{cc}
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-3)(50,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}} \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \
+  }}} \hspace{5mm} \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
+  }}} \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{7}{2}{\clazy} \nccurve[angleA=45,
+  angleB=135,                                                 ncurv=4,
+  nodesep=-1]{t1}{t1} \putnode{l1}{t1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
+  angleB=135,                                                 ncurv=3,
+  nodesep=-1]{t2}{t2} \putnode{l2}{t2}{0}{8}{\clazy}
+\end{pspicture}
+&
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-3)(30,12)
+  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}} \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
+  }}}     \psset{arrows=->}    \ncline{t0}{t1}     \nccurve[angleA=45,
+  angleB=135,                                                 ncurv=3,
+  nodesep=-1]{t1}{t1} \putnode{l1}{t1}{0}{8}{\acdr}
+\end{pspicture}
+\\
+(a) & (b) \end{tabular}}
+\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
+  converted      into      an      automaton     and      (b)      its
+  DFA.}\label{fig:example-automata}
+\vspace*{-2mm}
+\figrule
+%\end{figure}
+\end{wrapfigure}
+Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
+from the grammar for \var{\Df{\length}{1}}, and the final DFA is shown
+in  Figure~\ref{fig:example-automata}(b).  This  expectedly says  that
+for  a  demand  $\sigma_{\length}$,   the  liveness  of  the  argument
+of \length\ is $\acdr^{*}$ (the spine of the list is traversed).
+
+\input{main-pgm1}
+For the main  program~\ref{fig:main-pgm1}, the automaton corresponding
+to  the  variable   \var{\Lanv{\pa}{}}  and  its  DFA   are  shown  in
+Figure~\ref{fig:main1}.  The DFA in Figure~\ref{fig:main1}(b) does not
+accept any forward paths, reflecting  the lazy nature of our language.
+Since
 \length\ does not  evaluate the elements of  the argument list,
@@
 liveness-based GC triggers beyond $\pi_9$.
+
+\begin{wrapfigure}{r}{.25\textwidth}
+\scalebox{0.5}{
+\begin{tabular}{cc}
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+  }}} \hspace{5mm} \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \
+  }}}
+%  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+  \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
+%  \ncline{t2}{t3}
+%  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
+%  \putnode{l2}{tD1}{0}{10}{\acdr}
+ \putnode{tD1}{t2}{12}{0}{\pscirclebox{\mbox{\       \       \       \
+  }}} \hspace{5mm} \putnode{tD2}{tD1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
+  }}} \psset{arrows=->} \ncline{t2}{tD1} \putnode{l1}{t2}{5}{2}{\bcar} \ncline{tD1}{tD2} \putnode{lD0}{tD1}{5}{2}{\clazy} \nccurve[angleA=45,
+  angleB=135,                                                 ncurv=4,
+  nodesep=-1]{tD1}{tD1} \putnode{lD1}{tD1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
+  angleB=135,                                                 ncurv=3,
+  nodesep=-1]{tD2}{tD2} \putnode{lD2}{tD2}{0}{8}{\clazy}
+\end{pspicture}
+&
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(20,8) %\psframe(0,-5)(40,8)
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+  }}} \hspace{5mm} \psset{arrows=->} \ncline{t0}{t1}
+\end{pspicture}
+\\ 
+(a) & (b)
+\end{tabular}}
+ \caption{(a) grammar  rules for \var{\Lanv{\pa}{}} converted  into an
+  automation and (b) its DFA.}\label{fig:main1}
+\end{wrapfigure}

### 2016-04-13 paper/paper.tex
Split example program into 2 parts and updated mohri-nederhof

@@
 
-For  the  running  example (Figure~\ref{fig:mot-example2-a}),  $\length$  has  calls  from  $\main$  at
-$\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
-are $\epsilon$ and $\clazy\sigma_{\mathit{\length}}$. Thus:
+%% For  the  running  example (Figure~\ref{fig:mot-example2-a}), $\length$  has  calls  from  $\main$  at
+%% $\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
+For   the   running  example   (Figure~\ref{fig:mot-example2-a}),   if
+$\length$  has calls  from  $\main$  with a  demand  $\epsilon$ and  a
+recursive      call      at       $\pi_6$      with      a      demand
+$\clazy\sigma_{\mathit{\length}}$. Thus:
 \begin{eqnarray*}
-\sigma_{\length}    &=&
- \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
+\sigma_{\length}                    &=&                   \{\epsilon\}
+~\cup~{\clazy\sigma_{\mathit{\length}}}
 \end{eqnarray*}
@@
 several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
-as a start symbol.  As  an example, the grammar for \var{\Lanv{\xl}{}}
-and \var{\Lanv{\pa}{}} comprise the following productions:
+as a start  symbol. The grammar for  \var{\Df{\length}{1}} and $\langle
+{\sigma_{\length}} \rangle$ comprise of the following productions:
+%%  As  an example, the grammar for \var{\Lanv{\xl}{}}
+%% and \var{\Lanv{\pa}{}} comprise the following productions:
 \begin{eqnarray*}
@@
 %%\end{quote}
-We model  this problem as one  of deciding the
-membership of  a CFG augmented  with a
-fixed set of unrestricted productions. As shown in Appendix~\ref{sec:undec}, the  membership problem of this class of grammars is undecidable.
+We model this problem as one of deciding the membership of a CFG
+augmented with a fixed set of unrestricted productions. As shown in
+Appendix~\ref{sec:undec}, the membership problem of this class of
+grammars is undecidable.
 %------------------------------------------------------------%

### 2016-04-14 paper/cfg-to-nfa.tex
Updated NFA example automata

@@
 \begin{wrapfigure}{r}{0.25\textwidth}
-%\begin{figure}[t!]
 \scalebox{0.5}{
@@
 \\
-(a) & (b) \end{tabular}}
+(a) & (b)
+ \end{tabular}}
 \caption{(a) The  grammar rules for \var{\Df{\length}{1}}
@@
   DFA.}\label{fig:example-automata}
-\vspace*{-2mm}
 \figrule
-%\end{figure}
+\vspace*{-5mm}
 \end{wrapfigure}
+
 Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
@@
 \scalebox{0.5}{
-\begin{tabular}{cc}
+\begin{tabular}{c}
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
@@
   }}}
-%  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
   \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
-%  \ncline{t2}{t3}
-%  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
-%  \putnode{l2}{tD1}{0}{10}{\acdr}
- \putnode{tD1}{t2}{12}{0}{\pscirclebox{\mbox{\       \       \       \
-  }}} \hspace{5mm} \putnode{tD2}{tD1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
-  }}} \psset{arrows=->} \ncline{t2}{tD1} \putnode{l1}{t2}{5}{2}{\bcar} \ncline{tD1}{tD2} \putnode{lD0}{tD1}{5}{2}{\clazy} \nccurve[angleA=45,
+  \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\       \       \       \
+  }}} 
+  \psset{arrows=->} \ncline{t2}{tD1} \putnode{l1}{t2}{5}{2}{\bcar}
+  \nccurve[angleA=45,
   angleB=135,                                                 ncurv=4,
-  nodesep=-1]{tD1}{tD1} \putnode{lD1}{tD1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
-  angleB=135,                                                 ncurv=3,
-  nodesep=-1]{tD2}{tD2} \putnode{lD2}{tD2}{0}{8}{\clazy}
+  nodesep=-1]{tD1}{tD1} \putnode{lD1}{tD1}{0}{10}{\acdr}
 \end{pspicture} 
-&
+\\
+(a)
+\\
 \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(20,8) %\psframe(0,-5)(40,8)
+\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
+  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+  }}} 
+  \hspace{5mm} 
+  \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \  }}}
+  \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} 
+\end{pspicture}
+\\
+(b)
+\\
+\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
@@
 \\ 
-(a) & (b)
+(c)
 \end{tabular}}
  \caption{(a) grammar  rules for \var{\Lanv{\pa}{}} converted  into an
-  automation and (b) its DFA.}\label{fig:main1}
+  automaton and  its DFAs after simplification.}\label{fig:main1}
 \end{wrapfigure}
+
+If the expression at $\pi_{12}$ in the main program in Figure~\ref{fig:main-pgm1} was 
+$\CAR~\pc$, then the liveness of $\pc$ would be $(\clazy\sigma_{all}, \acar\sigma_{all})$.

### 2016-04-14 paper/cfg-to-nfa.tex
Updated NFA and DFAs

@@
 Mohri   and   Nederhof~\cite{mohri00regular}   method.   This   method
-transforms a  given CFG  into a strongly  regular grammar.
-
-In our example, the rule  $\var{\Df{\length}{1}}$ has two context free
-productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
-$\clazy\var{\Df{\length}{1}}{\clazy}$.  We transform this grammar into
-a strongly regular grammar in two steps,
+transforms a CFG $G$ into a strongly  regular grammar $R$ such that $\lang{G} \subseteq \lang{R}$. 
+%%
+If a CFG consists of a set of mutually recursive non-terminals such that the rules involved are not all left regular or not all right regular, then the method breaks the rules into right regular rules by introducing fresh non-terminals.
+For our example, the rule  $\var{\Df{\length}{1}}$ has two non regular productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
+$\clazy\var{\Df{\length}{1}}{\clazy}$.  The steps for transforming these productions into
+right regular productions are:
 
@@
 \item
-Consider each production of $\var{\Df{\length}{1}}$,
-\begin{enumerate}
-\item
- replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$,     by
+ Replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$,     by
  $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1'}}$
 \item
- replace
+ Replace
  $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1}}\clazy$,
@@
 \item
- replace
+  Replace
  $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}\clazy$,
@@
 \end{enumerate}
-\end{enumerate}
 
-The    resulting    grammar    fragment    for    %the    non-terminal
-$\var{\Df{\length}{1}}$   after   the  Mohri-Nederhof   transformation
-is:               \begin{eqnarray*}              \var{\Df{\length}{1}}
-&\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \acdr\var{\Df{\length}{1}} \mid \clazy\var{\Df{\length}{1}}\\ \var{\Df{\length}{1'}}
-&\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \epsilon \end{eqnarray*}
-Notice that the language generated  by the strongly regular grammar is
-a superset of the original CFG.
+Mohri and Nederhof~\cite{mohri00regular} provide the generalization of
+these  steps.  The  rules for  
+$\var{\Df{\length}{1}}$ after the transformation are:
+\begin{eqnarray*}              
+  \var{\Df{\length}{1}}   &\rightarrow&   \clazy\var{\Df{\length}{1'}}
+  \mid                 \acdr\var{\Df{\length}{1}}                 \mid
+  \clazy\var{\Df{\length}{1}}\\  \var{\Df{\length}{1'}}  &\rightarrow&
+  \clazy\var{\Df{\length}{1'}} \mid \epsilon
+\end{eqnarray*}
 
 The strongly regular grammar is converted  into a set of NFAs, one for
-each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
+each non-terminal.  The  $\hookrightarrow$ simplification is now
 done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
@@
 
-\begin{wrapfigure}{r}{0.25\textwidth}
-\scalebox{0.5}{
-          \begin{tabular}{cc}
-          \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+\newcommand{\nfaD}{%
+  \scalebox{.5}{\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
     \begin{pspicture}(0,-3)(50,12)
-            \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}} \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \
-  }}} \hspace{5mm} \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
-  }}} \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{7}{2}{\clazy} \nccurve[angleA=45,
-  angleB=135,                                                 ncurv=4,
-  nodesep=-1]{t1}{t1} \putnode{l1}{t1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
-  angleB=135,                                                 ncurv=3,
-  nodesep=-1]{t2}{t2} \putnode{l2}{t2}{0}{8}{\clazy}
+      \putnode{t0}{origin}{5}{0}{} 
+      \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \      }}}
+      \hspace{5mm} 
+      \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}} 
+      \psset{arrows=->} \ncline{t0}{t1} 
+      \ncline{t1}{t2} 
+      \putnode{l0}{t1}{7}{2}{\clazy} \nccurve[angleA=45, angleB=135, 
+        ncurv=4, nodesep=-1]{t1}{t1} 
+      \putnode{l1}{t1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
+        angleB=135, ncurv=3, nodesep=-1]{t2}{t2} 
+      \putnode{l2}{t2}{0}{8}{\clazy}
     \end{pspicture}
-&
+}}
+\newcommand{\dfaD}{%
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
@@
     \end{pspicture}
-\\
-(a) & (b)
- \end{tabular}}
-\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
-  converted      into      an      automaton     and      (b)      its
-  DFA.}\label{fig:example-automata}
-\figrule
-\vspace*{-5mm}
-\end{wrapfigure}
-
-Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
-from the grammar for \var{\Df{\length}{1}}, and the final DFA is shown
-in  Figure~\ref{fig:example-automata}(b).  This  expectedly says  that
+}
+The  NFA for \var{\Df{\length}{1}} is 
+\nfaD\ and the final DFA is \dfaD.
+This  expectedly says  that
 for  a  demand  $\sigma_{\length}$,   the  liveness  of  the  argument

### 2016-04-14 paper/cfg-to-nfa.tex
inlined nfa and dfa

@@
 \newcommand{\nfaD}{%
-  \scalebox{.5}{\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(0,-3)(50,12)
-      \putnode{t0}{origin}{5}{0}{} 
-      \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \      }}}
+  \scalebox{.7}{
+    \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+    \begin{pspicture}(2,-2)(30,10)
+      \putnode{t0}{origin}{2}{-1}{} 
+      \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{\ \ \ \      }}}
       \hspace{5mm} 
@@
 \newcommand{\dfaD}{%
+  \scalebox{.7}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(0,-3)(30,12)
-      \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}} \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
-      }}}     \psset{arrows=->}    \ncline{t0}{t1}     \nccurve[angleA=45,
-        angleB=135,                                                 ncurv=3,
-        nodesep=-1]{t1}{t1} \putnode{l1}{t1}{0}{8}{\acdr}
+    \begin{pspicture}(2,-2)(14,8)
+      \putnode{t0}{origin}{2}{0}{}
+      \putnode{t1}{t0}{8}{0}{\pscirclebox[doubleline=true]{
+          \mbox{\ \ \ \ }}}     
+      \psset{arrows=->}    
+      \ncline{t0}{t1}     
+      \nccurve[angleA=45, angleB=135, ncurv=2, nodesep=-1]{t1}{t1} 
+      \putnode{l1}{t1}{0}{8}{\acdr}
     \end{pspicture}
-}
+}}
 The  NFA for \var{\Df{\length}{1}} is 

### 2016-04-14 paper/cfg-to-nfa.tex
Inlined all DFAs

@@
 CFG by non-deterministic finite state automata (NFA) using
-%% The NFAs  are then simplified using the
-%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
-%%are converted to DFAs.
-%%
-%% We use  the algorithm by 
 Mohri   and   Nederhof~\cite{mohri00regular}   method.   This   method
@@
 
-\input{main-pgm1}
-For the main  program~\ref{fig:main-pgm1}, the automaton corresponding
-to  the  variable   \var{\Lanv{\pa}{}}  and  its  DFA   are  shown  in
-Figure~\ref{fig:main1}.  The DFA in Figure~\ref{fig:main1}(b) does not
-accept any forward paths, reflecting  the lazy nature of our language.
-Since
-\length\ does not  evaluate the elements of  the argument list,
-the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
-liveness-based GC triggers beyond $\pi_9$.
-
-\begin{wrapfigure}{r}{.25\textwidth}
-\scalebox{0.5}{
-\begin{tabular}{c}
+\newcommand{\dfaE}{%
+  \scalebox{.7}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+    \begin{pspicture}(7,-2)(40,10)
+     \putnode{t0}{origin}{5}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
       }}} \hspace{5mm} \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \
@@
               \end{pspicture} 
-\\
-(a)
-\\
+              }
+              }
+
+\newcommand{\dfaF}{%
+  \scalebox{.7}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+    \begin{pspicture}(3,-2)(27,10)
+     \putnode{t0}{origin}{5}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
       }}} 
@@
         \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \  }}}
-  \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} 
+         \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
          \end{pspicture}
-\\
-(b)
-\\
+         }
+         }
+
+\newcommand{\dfaG}{%
+  \scalebox{.7}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-\begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-  \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
-  }}} \hspace{5mm} \psset{arrows=->} \ncline{t0}{t1}
+    \begin{pspicture}(3,-2)(16,10)
+    \putnode{t0}{origin}{5}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+      }}} \hspace{1mm} \psset{arrows=->} \ncline{t0}{t1}
       \end{pspicture}
-\\ 
-(c)
-\end{tabular}}
- \caption{(a) grammar  rules for \var{\Lanv{\pa}{}} converted  into an
-  automaton and  its DFAs after simplification.}\label{fig:main1}
-\end{wrapfigure}
+      }
+      }
 
-If the expression at $\pi_{12}$ in the main program in Figure~\ref{fig:main-pgm1} was 
-$\CAR~\pc$, then the liveness of $\pc$ would be $(\clazy\sigma_{all}, \acar\sigma_{all})$.
+\input{main-pgm1}
+For the main  program in Figure~\ref{fig:main-pgm1}, the liveness corresponding
+to  the  variable   \var{\Lanv{\pa}{}} at $\pi_{10}$ is given by 
+$\var{\Lanv{\pa}{}} \rightarrow \clazy \bcar \var{\Df{\length}{1}}$.
+The  DFA   corresponding to this grammar is 
+\dfaE. 
+As there is no $\acar$ symbol to cancel the $\bcar$ symbol in the automaton no 
+$\epsilon$ edges are added and the $\bcar$ edge is deleted to get 
+\dfaF.
+We finally do the $\clazy$ simplification and notice that there is no path to a final state. 
+We delete the $\clazy$ edge without marking any state as final giving the final automata for 
+$\var{\Lanv{\pa}{}}$ 
+\dfaG.
+The final automaton does not
+accept any forward paths, reflecting  the lazy nature of our language.
+Since
+\length\ does not  evaluate the elements of  the argument list,
+the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
+liveness-based GC triggers beyond $\pi_9$.
+
+%% \begin{wrapfigure}{r}{.25\textwidth}
+%% \scalebox{0.5}{
+%% \begin{tabular}{c}
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
+%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+%%   }}} \hspace{5mm} \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \
+%%   }}}
+%%   \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
+%%   \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\       \       \       \
+%%   }}} 
+%%   \psset{arrows=->} \ncline{t2}{tD1} \putnode{l1}{t2}{5}{2}{\bcar}
+%%   \nccurve[angleA=45,
+%%   angleB=135,                                                 ncurv=4,
+%%   nodesep=-1]{tD1}{tD1} \putnode{lD1}{tD1}{0}{10}{\acdr}
+%% \end{pspicture} 
+%% \\
+%% (a)
+%% \\
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
+%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+%%   }}} 
+%%   \hspace{5mm} 
+%%   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \  }}}
+%%   \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
+%% \end{pspicture}
+%% \\
+%% (b)
+%% \\
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
+%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+%%   }}} \hspace{5mm} \psset{arrows=->} \ncline{t0}{t1}
+%% \end{pspicture}
+%% \\ 
+%% (c)
+%% \end{tabular}}
+%%  \caption{(a) grammar  rules for \var{\Lanv{\pa}{}} converted  into an
+%%   automaton and  its DFAs after simplification.}\label{fig:main1}
+%% \end{wrapfigure}

### 2016-04-14 paper/paper.tex
Updated NFA and DFAs
Inlined all DFAs

@@
 \usepackage{balance}
+\usepackage[numbers,sort&compress]{natbib}
 \usepackage{wrapfig}
@@
 each  GC  point  inside  the  function  based  on  this
-approximation.  The 0-CFA-style {\em summary demand} is
+approximation.  The {\em summary demand} is
 calculated by  taking a union  of the demands  at every
-call site of a function.
+call site of a function analogous to 0-CFA~\cite{Shivers:1988}.
 
@@
 %% \end{enumerate}
+We compute the liveness of all variables and arguments of $\length$ in terms of $\sigma_{\length}$.
 
@@
 during GC is:
-%%\begin{quote} 
+
 {\em
@@
 }
-%%\end{quote}
+
 We model this problem as one of deciding the membership of a CFG
@@
 %\subsection{A liveness-based GC scheme} \label{sec:live-clo}
-
 %%\subsubsection{From CFGs to GC-ready DFAs}
@@
 be  repeatedly  visited  with   different  liveness  states.  We  have
-mitigated  this  problem  by   implementing  a  heuristic  which  tags
-\CONS\ cells with the set of states already visited.
+mitigated  this  problem  by  adding checks to avoid multiple traversals 
+of closure and function arguments.
 

### 2016-04-15 paper/cfg-to-nfa.tex
updated experimental setup

@@
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(7,-2)(40,10)
-     \putnode{t0}{origin}{5}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+    \begin{pspicture}(0,-2)(35,8)
+     \putnode{t0}{origin}{0}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
       }}} \hspace{5mm} \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \
@@
              \nccurve[angleA=45,
-              angleB=135,                                                 ncurv=4,
+              angleB=135,                                                 ncurv=2,
               nodesep=-1]{tD1}{tD1} \putnode{lD1}{tD1}{0}{10}{\acdr}
@@
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(3,-2)(27,10)
-     \putnode{t0}{origin}{5}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+    \begin{pspicture}(3,-2)(24,3)
+     \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
       }}} 
@@
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(3,-2)(16,10)
-    \putnode{t0}{origin}{5}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
+    \begin{pspicture}(3,-2)(10,3)
+    \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{7}{0}{\pscirclebox{\mbox{\ \ \ \
       }}} \hspace{1mm} \psset{arrows=->} \ncline{t0}{t1}
@@
 We finally do the $\clazy$ simplification and notice that there is no path to a final state. 
-We delete the $\clazy$ edge without marking any state as final giving the final automata for 
-$\var{\Lanv{\pa}{}}$ 
+Thus, we delete the $\clazy$ edge without marking any state as final giving the final automata for 
+$\var{\Lanv{\pa}{}}$ as 
 \dfaG.

### 2016-04-15 paper/paper.tex
updated experimental setup

@@
 
-Our experimental setup  consists of an interpreter
-for  our  language,  a liveness  analyzer,  and  a
-single generation copying  collector.  The garbage
-collector can be configured to work on the
-basis  of  reachability  (RGC  mode)  or  use  
-liveness  DFAs  (LGC  mode).
-
 In        the       liveness        analysis       described        in
@@
 \label{sec:experiments}
+Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
+interpreter  for our  language, (b)  a  liveness analyzer,  and (c)  a
+single generation  copying garbage  collector.  The  garbage collector
+can be configured  to work on the basis of  reachability (RGC mode) or
+use liveness DFAs (LGC mode).
+%
 Our benchmark  consists of programs taken  from nofib~\cite{nofib} and

### 2016-04-15 paper/paper.tex
reduced number of graphs & added extra columns to autotable

@@
 
-\newcommand{\hgt}{2.9cm}
+\newcommand{\hgt}{4.3cm}
 \newcommand{\LGCLine}{blue}
@@
 
+%% \begin{figure*}[t!]
+%% \renewcommand{\arraystretch}{.1}
+%% \begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
+%%    \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
+%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
+%% &  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
+%% %\\ (a) & (b)
+%% \end{tabular}
+%% \vspace*{-2mm}
+%%  \caption{Memory usage.  
+%% The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
+%% cells  in  the  active   semi-space  for  RGC  and  LGC
+%% respectively.  The \ReachLine\ curve represents the number of
+%% reachable cells and the  \UseLine\ curve represents the
+%% number  of  cells  that  are actually  live  (of  which
+%% liveness analysis does a static approximation).  x-axis
+%% is the time measured  in number of cons-cells allocated
+%% (scaled down by factor $10^5$). y-axis is the number of
+%% cons-cells (scaled down by $10^3$).}
+%% \label{fig:memory-usage} \figrule
+%% \vspace*{-1mm}
+%% \end{figure*}
+
 \begin{figure*}[t!]
-\renewcommand{\arraystretch}{.1}
+\renewcommand{\arraystretch}{.5}
 \begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+   \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+\\
+   \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
 &  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
-\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
-%\\ (a) & (b)
 \end{tabular}

### 2016-04-16 paper/cfg-to-nfa.tex
Minor changes
more bibs fixed

@@
 \item
-Add a  new non-terminal  $\var{\Df{\length}{1'}}$ to the  grammar with
-the rule $\var{\Df{\length}{1'}} \rightarrow \epsilon$.
+Add a  new non-terminal  $\varnew{\Df{\length}{1}}$ to the  grammar with
+the rule $\varnew{\Df{\length}{1}} \rightarrow \epsilon$.
 \item
  Replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$,     by
- $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1'}}$
+ $\var{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
 \item
@@
  by  $\var{\Df{\length}{1}}  \rightarrow  \clazy\var{\Df{\length}{1}}$
- and $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
+ and $\varnew{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
 \item
@@
  by $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}$ and
- $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
+ $\varnew{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
 \end{enumerate}
@@
 \begin{eqnarray*}              
-  \var{\Df{\length}{1}}   &\rightarrow&   \clazy\var{\Df{\length}{1'}}
+  \var{\Df{\length}{1}}   &\rightarrow&   \clazy\varnew{\Df{\length}{1}}
   \mid                 \acdr\var{\Df{\length}{1}}                 \mid
-  \clazy\var{\Df{\length}{1}}\\  \var{\Df{\length}{1'}}  &\rightarrow&
-  \clazy\var{\Df{\length}{1'}} \mid \epsilon
+  \clazy\var{\Df{\length}{1}}\\  \varnew{\Df{\length}{1}}  &\rightarrow&
+  \clazy\varnew{\Df{\length}{1}} \mid \epsilon
 \end{eqnarray*}
@@
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(2,-2)(30,10)
+    \begin{pspicture}(3,-2)(28,10)
       \putnode{t0}{origin}{2}{-1}{} 
@@
       \ncline{t1}{t2} 
-      \putnode{l0}{t1}{7}{2}{\clazy} \nccurve[angleA=45, angleB=135, 
-        ncurv=4, nodesep=-1]{t1}{t1} 
-      \putnode{l1}{t1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
-        angleB=135, ncurv=3, nodesep=-1]{t2}{t2} 
-      \putnode{l2}{t2}{0}{8}{\clazy}
+      \putnode{l0}{t1}{7}{2}{\clazy} 
+      \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1} 
+      \putnode{l1}{t1}{0}{7}{\acdr, \clazy} 
+      \nccurve[angleA=45,
+        angleB=135, ncurv=2, nodesep=-1]{t2}{t2} 
+      \putnode{l2}{t2}{0}{7}{\clazy}
     \end{pspicture}
@@
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(2,-2)(14,8)
+    \begin{pspicture}(2,-2)(13,8)
       \putnode{t0}{origin}{2}{0}{}
-      \putnode{t1}{t0}{8}{0}{\pscirclebox[doubleline=true]{
-          \mbox{\ \ \ \ }}}     
+      \putnode{t1}{t0}{8}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}     
       \psset{arrows=->}    
@@
       \nccurve[angleA=45, angleB=135, ncurv=2, nodesep=-1]{t1}{t1} 
-      \putnode{l1}{t1}{0}{8}{\acdr}
+      \putnode{l1}{t1}{0}{7}{\acdr}
     \end{pspicture}
@@
     \begin{pspicture}(0,-2)(35,8)
-     \putnode{t0}{origin}{0}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
-      }}} \hspace{5mm} \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \
+     \putnode{t0}{origin}{0}{0}{} 
+     \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{\ \ \ \
       }}} 
-       \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
-         \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\       \       \       \
-           }}} 
-            \psset{arrows=->} \ncline{t2}{tD1} \putnode{l1}{t2}{5}{2}{\bcar}
-             \nccurve[angleA=45,
-              angleB=135,                                                 ncurv=2,
-              nodesep=-1]{tD1}{tD1} \putnode{lD1}{tD1}{0}{10}{\acdr}
+     \hspace{5mm} 
+   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \ }}}
+   \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} 
+   \putnode{l0}{t1}{5}{2}{\clazy}
+   \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}} 
+   \psset{arrows=->} \ncline{t2}{tD1} 
+   \putnode{l1}{t2}{5}{2}{\bcar}
+   \nccurve[angleA=45,angleB=135,ncurv=2,nodesep=-1]{tD1}{tD1} 
+   \putnode{lD1}{tD1}{0}{7}{\acdr}
 \end{pspicture} 
-              }
-              }
+  }}
 
@@
 \dfaE. 
-As there is no $\acar$ symbol to cancel the $\bcar$ symbol in the automaton no 
+As there is no $\acar$ symbol to cancel the $\bcar$ symbol in the automaton, no 
 $\epsilon$ edges are added and the $\bcar$ edge is deleted to get 

### 2016-04-16 paper/paper.tex
no more hodong :)
updating bib entries
Updated table with new values
more bibs fixed

@@
-\documentclass[9pt,draft,nonatbib]{sigplanconf}
+\documentclass[9pt,nonatbib]{sigplanconf}
 \usepackage{etex}
@@
 \usepackage{wrapfig}
+\usepackage{spreadtab}
 
@@
 % 1st. author
-\cmt{{
 \authorinfo {Prasanna Kumar. K} 
@@
        {karkare@cse.iitk.ac.in}
-}}
-\authorinfo{Double Blind Review}{}{}
 \maketitle
@@
 grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
-\clazy\rbrace$.  Let \var{$X$} denote  the non-terminal for a variable
+\clazy\rbrace$.  Let \var{X} denote  the non-terminal for a variable
 $X$ occurring on the LHS of the equations generated from the analysis.
@@
 several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
-as a start  symbol. The grammar for  \var{\Df{\length}{1}} and $\langle
-{\sigma_{\length}} \rangle$ comprise of the following productions:
+as a start  symbol. The grammar for  \var{\Df{\length}{1}} and \var{\sigma_{\length}} comprise of the following productions:
 %%  As  an example, the grammar for \var{\Lanv{\xl}{}}
@@
        \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
-\langle {\sigma_{\length}} \rangle
+\var{\sigma_{\length}}
 &\rightarrow&
-\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle 
+\epsilon  \mid \clazy\var{\sigma_{\length}}
 %% \\

### 2016-04-16 paper/cfg-to-nfa.tex
some rewriting

@@
 CFG by non-deterministic finite state automata (NFA) using
-Mohri   and   Nederhof~\cite{mohri00regular}   method.   This   method
+Mohri   and   Nederhof method~\cite{mohri00regular}.   This   method
 transforms a CFG $G$ into a strongly  regular grammar $R$ such that $\lang{G} \subseteq \lang{R}$. 
@@
 \item
- Replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$,     by
+ Replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$     by
  $\var{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
@@
  Replace
- $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1}}\clazy$,
+ $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1}}\clazy$
  by  $\var{\Df{\length}{1}}  \rightarrow  \clazy\var{\Df{\length}{1}}$
- and $\varnew{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
+ and \linebreak $\varnew{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
 \item
   Replace
- $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}\clazy$,
+ $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}\clazy$
  by $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}$ and
- $\varnew{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
+ \linebreak $\varnew{\Df{\length}{1}} \rightarrow \clazy\varnew{\Df{\length}{1}}$
 \end{enumerate}

### 2016-04-16 paper/paper.tex
some rewriting
Updated introduction to refer to new values

--- 2016-04-15 paper/paper.tex
@@
 collections from 1.6X  to 23X and a decrease  in the minimum
-memory requirement from  1X to 1389X.  As  an added benefit,
+memory requirement from  1X to 1198X.  As  an added benefit,
 there is also a reduction in the overall execution time in 5 out of 12

### 2016-04-16 paper/cfg-to-nfa.tex
fixed fig 7&8

@@
 
-\input{main-pgm1}
-For the main  program in Figure~\ref{fig:main-pgm1}, the liveness corresponding
+%\input{main-pgm1}
+For the main  program in Figure~\ref{fig:mot-example2-a}, the liveness corresponding
 to  the  variable   \var{\Lanv{\pa}{}} at $\pi_{10}$ is given by 

### 2016-04-16 paper/paper.tex
added reference to earlier work

@@
 significantly more  complicated than a liveness-based  collector for a
-eager language.
+eager language. An earlier  effort in efficiently garbage collecting
+closures~\cite{Thomas19951}  used  a   similar  approach  where  every
+closure  would maintain  a  sequence of  executable  code called  {\em
+  preserver}  which  identified  exactly which  closure  variables  to
+copy. The main focus of the  above work was also in garbage collecting
+closures without using  an explicit stack. Our work  extends this work
+and tags  liveness values to evaluated  data, as well as  modfiyng the
+liveness associated with closures dynamically  during runtime to make
+it more precise.
 

### 2016-04-16 paper/paper.tex
Updated...

@@
 copy. The main focus of the  above work was also in garbage collecting
-closures without using  an explicit stack. Our work  extends this work
-and tags  liveness values to evaluated  data, as well as  modfiyng the
-liveness associated with closures dynamically  during runtime to make
-it more precise.
+closures without using  an explicit stack. However, it not clear whether
+their method could handle partially live data structures.
 
@@
 \begin{itemize}
-\item We  propose a liveness  based GC scheme  for a lazy  first order
+\item {\color{Myblue}We  propose a liveness  based GC scheme  for a lazy  first order
   functional language and  prove its correctness.  To the  best of our
@@
   liveness  of closures,  an  issue  that arises  due  to laziness  at
-  runtime.
+  runtime.}
 
@@
 solution  of  \Lanv{x}{}.    
-However, recall that the decision problem that we are interested in
+Recall that the decision problem that we are interested in
 during GC is:
@@
 
-\clearpage
+%%\clearpage
 \balance
@@
 \bibliographystyle{abbrv}
+\end{document}
+
 \clearpage

### 2016-04-17 paper/paper.tex
Referred to Thomas's paper and marked all our changes in blue

@@
 significantly more  complicated than a liveness-based  collector for a
-eager language. An earlier  effort in efficiently garbage collecting
-closures~\cite{Thomas19951}  used  a   similar  approach  where  every
-closure  would maintain  a  sequence of  executable  code called  {\em
-  preserver}  which  identified  exactly which  closure  variables  to
-copy. The main focus of the  above work was also in garbage collecting
-closures without using  an explicit stack. However, it not clear whether
-their method could handle partially live data structures.
+eager language.
 
@@
 \begin{itemize}
-\item {\color{Myblue}We  propose a liveness  based GC scheme  for a lazy  first order
-  functional language and  prove its correctness.  To the  best of our
-  knowledge,  this is  the first  work on  liveness-based GC  for lazy
-  languages.       All      previous      attempts~\cite{shaham01heap,
-    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
-    karkare06effectiveness} have  involved either imperative  or eager
-  functional languages.  The novelty of our approach is in handling of
-  liveness  of closures,  an  issue  that arises  due  to laziness  at
-  runtime.}
+\item {\color{Myblue}We propose a liveness based GC scheme for a lazy
+  first order functional language and prove its correctness.
+  Thomas~\cite{Thomas19951} describes a copying garbage collector for
+  Three Instruction Machine (TIM) that only traces live variables in a
+  closure's environment, i.e., the variables that are useful for
+  evaluating the closure.  Their notion of liveness of variables is
+  close to the classical notion of liveness, and does not take into
+  account liveness of parts of structures (say, only the spine of a
+  list).  To the best of our knowledge, it is the only earlier work on
+  GC that takes liveness for closures into account.  All previous
+  attempts~\cite{shaham01heap, ran.shaham-sas03, shaham02estimating,
+    asati14lgc, karkare06effectiveness} have involved either
+  imperative or eager functional languages.  The novelty of our
+  approach is in handling of liveness of closures, an issue that
+  arises due to laziness at runtime.}
 
@@
 non-terminal.
-
+{\color{Myblue}
 %\subsection{Generating liveness equations \Lv\  for function bodies}
@@
 \input{cfg-to-nfa}
-
+}
   \SetStartEndCondition{ }{}{}%

### 2016-04-17 paper/paper.tex
First revision with Sanyal sir

@@
 \begin{itemize}
-\item {\color{Myblue}We propose a liveness based GC scheme for a lazy
-  first order functional language and prove its correctness.
+\item {\color{Myblue}We  propose an interprocedural liveness  based GC
+  scheme  for a  lazy first  order functional  language and  prove its
+  correctness.  To the  best of our knowledge, this is  the first work
+  that uses  the results  of an  interprocedural liveness  analysis to
+  garbage    collect    both     evaluated    data    and    closures.
   Thomas~\cite{Thomas19951} describes a  copying garbage collector for
-  Three Instruction Machine (TIM) that only traces live variables in a
-  closure's environment, i.e., the variables that are useful for
-  evaluating the closure.  Their notion of liveness of variables is
-  close to the classical notion of liveness, and does not take into
-  account liveness of parts of structures (say, only the spine of a
-  list).  To the best of our knowledge, it is the only earlier work on
-  GC that takes liveness for closures into account.  All previous
+  the  Three  Instruction  Machine  (TIM)  that  only  preserves  live
+  closures in a function's environment (also called a frame). However,
+  in the absence of  details, it is not clear whether  a) the scope of
+  the method is interprocedural, and b) it handles algebraic datatypes
+  like lists (the original design of TIM did not).  All other previous
   attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
-    asati14lgc, karkare06effectiveness} have involved either
-  imperative or eager functional languages.  The novelty of our
-  approach is in handling of liveness of closures, an issue that
-  arises due to laziness at runtime.}
+    asati14lgc, karkare06effectiveness} involved  either imperative or
+  eager functional languages.}
+%% The novelty of our approach is in handling of
+%%   liveness  of closures,  an  issue  that arises  due  to laziness  at
+%%   runtime.}
 
@@
   whether to copy a cell during  GC translates to a membership problem
-  for such  grammars.  Earlier research assumed  the undecidability of
-  this membership question and  used an over-approximation to overcome
-  it.  In this paper, we provide  a formal proof of the undecidability
-  of this problem.
+  for such grammars.  Earlier, this decision problem was assumed to be
+  undecidable, and the context-free grammar was over approximated by a
+  regular  grammar.   In  this  paper,  we  provide  a  proof  of  the
+  undecidability of this problem.
 
@@
 similar reasons.
+
+We  now return  back  to the  problem of  computing  liveness using
+\Lfonly\  and a  concrete  demand $\sigma$.  Although  the process  is
+explained through an example, it can be generalized easily.
 

### 2016-04-18 paper/paper.tex
Added reference to TIM

@@
   Thomas~\cite{Thomas19951} describes a  copying garbage collector for
-  the  Three  Instruction  Machine  (TIM)  that  only  preserves  live
-  closures in a function's environment (also called a frame). However,
-  in the absence of  details, it is not clear whether  a) the scope of
-  the method is interprocedural, and b) it handles algebraic datatypes
-  like lists (the original design of TIM did not).  All other previous
-  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
-    asati14lgc, karkare06effectiveness} involved  either imperative or
-  eager functional languages.}
+  the Three  Instruction Machine (TIM)~\cite{Fairbairn1987}  that only
+  preserves live closures  in a function's environment  (also called a
+  frame). However, in the absence of  details, it is not clear whether
+  a) the  scope of the  method is  interprocedural, and b)  it handles
+  algebraic datatypes like lists (the original design of TIM did not).
+  All  other  previous attempts~\cite{shaham01heap,  ran.shaham-sas03,
+    shaham02estimating,  asati14lgc, karkare06effectiveness}  involved
+  either imperative or eager functional languages.}
 %% The novelty of our approach is in handling of

### 2016-04-18 paper/paper.tex
space savings
Updated Table to display total execution time and added dummy reference to Technical Report

@@
 \usepackage{enumitem}
-\setlist{nolistsep}
+\setlist{nolistsep,leftmargin=*}
 
@@
 
-%\begin{figure}[t]
-  \input{example-to-illustrate-liveness}  
-%  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
-%\end{figure}
-\input{fig-liveness}
-
 \subsection{Liveness analysis for lazy languages}
 \label{sec:liveness-analysis}
+\input{example-to-illustrate-liveness}
 
@@
 from its corresponding stack variable by subscripting it with the
-label of the program   point  where   the   closure  was   created\footnote{Multiple
-  occurrences  of the  same  variable in  an  application are  further
-  distinguished by their positions in the application.}.
+label of the program point where the closure was
+created\footnote{Multiple occurrences of the same variable in an
+  application are further distinguished by their positions in the
+  application.}.
 
@@
 
+\input{fig-liveness}
 Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
@@
 
+\begin{figure*}[t!]
+\begin{center}\footnotesize
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{|@{}c@{\ }|@{\ }c@{}|@{\ }c@{\ }|}
+\hline
+Premise & Transition & Rule name \\ 
+\hline
+\hline 
+\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
+  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
+{\sc car-bang} 
+\\
+\hline
+\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
+(\CAR~x), \cred{\sigma} \rightsquigarrow$ $ \rho', \,(\rho, addr(\langle
+s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S, \heap,\, s, \, \cred{\sigma}$ }     &
+{\sc car-1-clo} 
+\\
+\hline
+\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
+\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
+  \rightsquigarrow$ 
+  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
+    \begin{array}{lr}
+      {\emptyset}&\mbox{if } \sigma = \emptyset \\       
+      {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
+    \end{array}\right.
+  }$   }      
+&
+{\sc car-clo}\\
+\hline
+\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
+    s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
+  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x
+    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
+    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
+{\sc let} \\ 
+\hline
+\end{tabular}
+\end{center}
+\caption{Minefield  semantics for  \CAR\   and \LET. Complete minefield semantics is provided in~\cite{TR16-1}.
+  \warning{Add a note mentioning that we handle the issue of requiring
+    a      specific     demand      by      introducing     a      new
+    symbol}\label{fig:minefield-semantics-for-some}}
+\vspace*{-2mm}
+\end{figure*}
+
 \subsection{Soundness of analysis}
@@
 \end{enumerate}
-
-\begin{figure*}[t!]
-\begin{center}\footnotesize
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|@{}c@{\ }|@{\ }c@{}|@{\ }c@{\ }|}
-\hline
-Premise & Transition & Rule name \\ 
-\hline
-\hline 
-\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
-  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
-{\sc car-bang} 
-\\
-\hline
-\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
-(\CAR~x), \cred{\sigma} \rightsquigarrow$ $ \rho', \,(\rho, addr(\langle
-s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S, \heap,\, s, \, \cred{\sigma}$ }     &
-{\sc car-1-clo} 
-\\
-\hline
-\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
-\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
-  \rightsquigarrow$ 
-  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
-    \begin{array}{lr}
-      {\emptyset}&\mbox{if } \sigma = \emptyset \\       
-      {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
-    \end{array}\right.
-  }$   }      
-&
-{\sc car-clo}\\
-\hline
-\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
-    s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
-  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x
-    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
-    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
-{\sc let} \\ 
-\hline
-\end{tabular}
-\end{center}
-\caption{Minefield  semantics for  \CAR\   and \LET.
-  \warning{Add a note mentioning that we handle the issue of requiring
-    a      specific     demand      by      introducing     a      new
-    symbol}\label{fig:minefield-semantics-for-some}}
-\vspace*{-2mm}
-\end{figure*}
 
@@
 }
-  \SetStartEndCondition{ }{}{}%
-  \SetKwProg{Pro}{procedure}{\string:}{}
-  \SetKwProg{Fn}{function}{\string:}{}
-  \SetKwFunction{Range}{range}%%
-  \SetKw{KwTo}{in}\SetKwFor{For}{for}{\string:}{}%
-  \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-  \SetKwFor{While}{while}{:}{fintq}%
-  \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-  \SetKwFunction{Lgc}{$\mathsf{lgc}$}%
-  \SetKwFunction{Copy}{$\mathsf{copy}$}%
-  \SetKwFunction{LCopy}{$lgcCopy$}%
-  \SetKwFunction{RCopy}{$rgcCopy$}%
-
-\begin{algorithm}[t!]
-  \Pro{\Lgc{}}
-     {
-       \For {each reference $\mathsf{ref}$ in root set}
-            {$\mathsf{ref}$ = \Copy($\mathsf{ref}, \mathsf{init}(\mathsf{ref.dfa})$)\;}
-            ${\mathsf{copyReferencesOnPrintStack}()}$\;  
-     }
-     \Fn{\Copy{$\mathsf{ref}, \mathsf{state}$}}
-        {
-          \eIf {$\mathsf{final}(\mathsf{state}$)}
-             {
-               $\mathsf{newRef} = \mathsf{dupHeapCell}(\mathsf{ref})$\;
-               \If{$\mathsf{ref\!.cell}$($\mathsf{ref}$)
-                 is a cons cell $\mathsf{(cons~arg_1~arg_2)}$}
-                  {
-                    {
-                      $\mathsf{newRef}\!.\mathsf{arg_1}  = \Copy(\mathsf{arg_1}, \mathsf{next}(\mathsf{state}, 0))$\;
-%                      $\mathsf{newRef}\!.\mathsf{arg_1}   =  \mathsf{newCar}$\;
-                    }
-                    {
-                      $\mathsf{newRef}\!.\mathsf{arg_2} =  \Copy(\mathsf{arg_2},
-                      \mathsf{next}(\mathsf{state} , 1))$\;
-%                      $\mathsf{newRef}\!.\mathsf{arg_2}   = \mathsf{newCdr}$\;
-                    }    
-                  }
-               \If
-                  {$\mathsf{ref\!.cell}$ is a
-                    closure cell, generically  $\mathsf{(binop~arg_1~arg_2)}$}
-              %      
-                   { 
-                     $\mathsf{newRef}\!.\mathsf{arg_1} = \Copy(\mathsf{arg_1}, \mathsf{init}(\mathsf{arg_1.dfa}))$\;
-                     $\mathsf{newRef}\!.\mathsf{arg_2} = \Copy(\mathsf{arg_2}, \mathsf{init}(\mathsf{arg_2.dfa}))$\;
-                   }
-             }
-             {$\mathsf{newRef = ref}$}
-\KwRet $\mathsf{newRef}$\;
-        }
-        \caption{Liveness-based garbage collection.  \label{algo:lgc-a}}
-\end{algorithm}
-
-\newcommand{\hgt}{4.3cm}
-\newcommand{\LGCLine}{blue}
-\newcommand{\RGCLine}{red}
-\newcommand{\ReachLine}{black}
-\newcommand{\UseLine}{light-blue}
 
@@
 
-\begin{figure*}[t!]
-\renewcommand{\arraystretch}{.5}
-\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
-\\
-   \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
-\end{tabular}
-\vspace*{-2mm}
- \caption{Memory usage.  
-The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
-cells  in  the  active   semi-space  for  RGC  and  LGC
-respectively.  The \ReachLine\ curve represents the number of
-reachable cells and the  \UseLine\ curve represents the
-number  of  cells  that  are actually  live  (of  which
-liveness analysis does a static approximation).  x-axis
-is the time measured  in number of cons-cells allocated
-(scaled down by factor $10^5$). y-axis is the number of
-cons-cells (scaled down by $10^3$).}
-\label{fig:memory-usage} \figrule
-\vspace*{-1mm}
-\end{figure*}
-
 %\vspace*{-6.5mm}
@@
 $\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.
+  \SetStartEndCondition{ }{}{}%
+  \SetKwProg{Pro}{procedure}{\string:}{}
+  \SetKwProg{Fn}{function}{\string:}{}
+  \SetKwFunction{Range}{range}%%
+  \SetKw{KwTo}{in}\SetKwFor{For}{for}{\string:}{}%
+  \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+  \SetKwFor{While}{while}{:}{fintq}%
+  \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+  \SetKwFunction{Lgc}{$\mathsf{lgc}$}%
+  \SetKwFunction{Copy}{$\mathsf{copy}$}%
+  \SetKwFunction{LCopy}{$lgcCopy$}%
+  \SetKwFunction{RCopy}{$rgcCopy$}%
+
+\begin{algorithm}[ht!]
+  \Pro{\Lgc{}}
+     {
+       \For {each reference $\mathsf{ref}$ in root set}
+            {$\mathsf{ref}$ = \Copy($\mathsf{ref}, \mathsf{init}(\mathsf{ref.dfa})$)\;}
+            ${\mathsf{copyReferencesOnPrintStack}()}$\;  
+     }
+     \Fn{\Copy{$\mathsf{ref}, \mathsf{state}$}}
+        {
+          \If {$\neg\mathsf{final}(\mathsf{state}$)} {\KwRet $\mathsf{ref}$\;}
+          $\mathsf{newRef} = \mathsf{dupHeapCell}(\mathsf{ref})$\;
+          \If{$\mathsf{ref\!.cell}$($\mathsf{ref}$)
+            is a cons cell $\mathsf{(cons~arg_1~arg_2)}$}
+             {
+               {
+                 $\mathsf{newRef}\!.\mathsf{arg_1}  = \Copy(\mathsf{arg_1}, \mathsf{next}(\mathsf{state}, 0))$\;
+                 %                      $\mathsf{newRef}\!.\mathsf{arg_1}   =  \mathsf{newCar}$\;
+               }
+               {
+                 $\mathsf{newRef}\!.\mathsf{arg_2} =  \Copy(\mathsf{arg_2},
+                 \mathsf{next}(\mathsf{state} , 1))$\;
+                 %                      $\mathsf{newRef}\!.\mathsf{arg_2}   = \mathsf{newCdr}$\;
+               }    
+             }
+             \If {$\mathsf{ref\!.cell}$ is a closure cell, generically  $\mathsf{(binop~arg_1~arg_2)}$}
+             { 
+               $\mathsf{newRef}\!.\mathsf{arg_1} = \Copy(\mathsf{arg_1}, \mathsf{init}(\mathsf{arg_1.dfa}))$\;
+               $\mathsf{newRef}\!.\mathsf{arg_2} = \Copy(\mathsf{arg_2}, \mathsf{init}(\mathsf{arg_2.dfa}))$\;
+             }
+             \KwRet $\mathsf{newRef}$\;
+        }
+        \caption{Liveness-based garbage collection.  \label{algo:lgc-a}}
+\end{algorithm}
+%\setlength{\textfloatsep}{0pt}% Remove \textfloatsep
+
 Algorithm~\ref{algo:lgc-a}   describes   the   garbage
@@
 
+\begin{figure*}[t!]
+\renewcommand{\arraystretch}{.5}
+\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
+   \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
+\\
+   \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
+&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
+\end{tabular}
+\vspace*{-2mm}
+ \caption{Memory usage.  
+The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
+cells  in  the  active   semi-space  for  RGC  and  LGC
+respectively.  The \ReachLine\ curve represents the number of
+reachable cells and the  \UseLine\ curve represents the
+number  of  cells  that  are actually  live  (of  which
+liveness analysis does a static approximation).  x-axis
+is the time measured  in number of cons-cells allocated
+(scaled down by factor $10^5$). y-axis is the number of
+cons-cells (scaled down by $10^3$). Graphs for other benchmarks are provided in~\cite{TR16-1}}
+\label{fig:memory-usage} \figrule
+\vspace*{-1mm}
+\end{figure*}
+
 Memory    usage    graphs    for     the    benchmarks    are    shown
@@
 We consider the rules for \CAR\  only.  
-The rest  of the rules involve similar reasoning. For the {\sc
+The other rules involve similar reasoning. For the {\sc
   car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
@@
   grammar.}\label{fig:turing-pgm}
-\vspace*{-6mm}
+\vspace*{-2mm}
 \end{figure}

### 2016-04-19 paper-for-archive/paper.tex
Added Acknowledgement and arxiv version

--- 2016-04-17 paper/paper.tex
@@
-\documentclass[9pt,nonatbib]{sigplanconf}
-\usepackage{etex}
+\documentclass[9pt,preprint,nonatbib]{sigplanconf}
+\usepackage{balance}
 \usepackage{float}
@@
 \usepackage{stmaryrd}
+\usepackage{etex}
 \usepackage{boxedminipage}
@@
 \usepackage{amsmath}
-\usepackage{pstricks}
+%\usepackage[pdf]{pstricks}
 \usepackage{pst-text}
@@
 \usepackage{pst-tree}
-\usepackage{pst-rel-points}
+%\usepackage{pst-rel-points}
 \usepackage{bcprules}
@@
 \usepackage{makecell}
-\usepackage{balance}
-\usepackage[numbers,sort&compress]{natbib}
-\usepackage{wrapfig}
-\usepackage{spreadtab}
 
@@
 
-\definecolor{Myblue}{rgb}{.2,0,1}
-\definecolor{Myred}{rgb}{1,.2,.2}
+%\definecolor{Myblue}{rgb}{.2,0,1}
+%\definecolor{Myred}{rgb}{1,.2,.2}
 
@@
 \newcommand{\oxrb}{\ensuremath{|\!]}}
-\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
+%\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{\ensuremath{#1}}}
+\newcommand{\cred}[1]{\colorbox{lightgray}{\ensuremath{#1}}}
 %\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
@@
        {karkare@cse.iitk.ac.in}
+
 \maketitle
@@
 collections from 1.6X  to 23X and a decrease  in the minimum
-memory requirement from  1X to 1198X.  As  an added benefit,
+memory requirement from  1X to 1389X.  As  an added benefit,
 there is also a reduction in the overall execution time in 5 out of 12
@@
 \label{sec:motiv}
-\input{fig_motiv}   
-%%
+
+\begin{figure}[t!]
+  \includegraphics[width=\columnwidth]{fig_motiv.pdf}
+  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
+\end{figure}
+
   Figure~\ref{fig:mot-example}   shows  an   example
@@
 \begin{itemize}
-\item {\color{Myblue}We  propose an interprocedural liveness  based GC
-  scheme  for a  lazy first  order functional  language and  prove its
-  correctness.  To the  best of our knowledge, this is  the first work
-  that uses  the results  of an  interprocedural liveness  analysis to
-  garbage    collect    both     evaluated    data    and    closures.
-  Thomas~\cite{Thomas19951} describes a  copying garbage collector for
-  the  Three  Instruction  Machine  (TIM)  that  only  preserves  live
-  closures in a function's environment (also called a frame). However,
-  in the absence of  details, it is not clear whether  a) the scope of
-  the method is interprocedural, and b) it handles algebraic datatypes
-  like lists (the original design of TIM did not).  All other previous
-  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
-    asati14lgc, karkare06effectiveness} involved  either imperative or
-  eager functional languages.}
-%% The novelty of our approach is in handling of
-%%   liveness  of closures,  an  issue  that arises  due  to laziness  at
-%%   runtime.}
+\item We  propose a liveness  based GC scheme  for a lazy  first order
+  functional language and  prove its correctness.  To the  best of our
+  knowledge,  this is  the first  work on  liveness-based GC  for lazy
+  languages.       All      previous      attempts~\cite{shaham01heap,
+    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
+    karkare06effectiveness} have  involved either imperative  or eager
+  functional languages.  The novelty of our approach is in handling of
+  liveness  of closures,  an  issue  that arises  due  to laziness  at
+  runtime.
 
@@
   whether to copy a cell during  GC translates to a membership problem
-  for such grammars.  Earlier, this decision problem was assumed to be
-  undecidable, and the context-free grammar was over approximated by a
-  regular  grammar.   In  this  paper,  we  provide  a  proof  of  the
-  undecidability of this problem.
+  for such  grammars.  Earlier research assumed  the undecidability of
+  this membership question and  used an over-approximation to overcome
+  it.  In this paper, we provide  a formal proof of the undecidability
+  of this problem.
 
@@
 \end{eqnarray*}
-\vspace*{-6pt}
+%\vspace*{-6pt}
   \caption{The syntax of our language}\label{fig:lang-syntax}
@@
 \normalsize
-\vspace*{-6pt}
+%\vspace*{-6pt}
 \end{figure}
@@
 %}\comment
-\vspace*{-16pt}
+%\vspace*{-16pt}
 \end{figure*}
@@
 
-%\begin{figure}[t]
+\begin{figure}[t]
   \input{example-to-illustrate-liveness}  
-%  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
-%\end{figure}
+  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
+\end{figure}
 \input{fig-liveness}
@@
 \renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|@{}c@{\ }|@{\ }c@{}|@{\ }c@{\ }|}
+\begin{tabular}{|@{}c@{}|@{}c@{}|@{}c@{}|}
 \hline
@@
   $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
-    \begin{array}{lr}
+    \begin{array}{@{}lr@{}}
       {\emptyset}&\mbox{if } \sigma = \emptyset \\       
@@
     symbol}\label{fig:minefield-semantics-for-some}}
-\vspace*{-2mm}
+%\vspace*{-2mm}
 \end{figure*}
-
 \noindent To set up the minefield semantics, we follow these steps:
@@
 the transition is similar to the earlier {\sc car-clo}
-rule.
+rule. The remaining rules for minefield semantics are given in Appendix~\ref{sec:minefield}.
 
@@
 \ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
-e_{\mainpgm})$. The proof that such a program cannot go \bang\ is provided
-in Appendix~\ref{sec:sound}.
+e_{\mainpgm})$.
+
+\subsection{Soundness result}
+We first need an auxiliary  result about minefield semantics. Consider
+a trace of  a minefield execution.  For every  minefield state $(\rho,
+S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
+$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
+application $s$) is  non-null.  This can be proved by  an induction on
+the number  of steps leading  to the  minefield state.  The  base step
+holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
+non-null. For the inductive step we  observe that for each step of the
+minefield semantics, if the demand $\sigma$  on the LHS of a minefield
+step  is non-null,  the  demand  on the  RHS  is  a transformation  of
+$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.
+
+Note that our proofs will be for a single round of minefield execution
+i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
+\ref{sec:semantics}).  With  minor variations,  the proof will  also be
+applicable for subsequent evaluations initiated by $\print$.
+%% We show  next that the  minefield execution of a  call-less program
+%% cannot go \bang.
+
+\begin{lemma}
+\label{lemma:call-less-cannot-go-bang}
+Consider the minefield execution of  a program without function calls.
+Such a program cannot enter the \bang\ state.
+\end{lemma}
+\begin{proof}
+Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
+minefield execution of a program.  We  show by induction on the number $n$
+of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
+enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
+(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
+[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
+does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
+transition cannot  enter a  \bang\ state.  Since  our programs  are in
+ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
+does not involve dereferencing, and thus cannot result in a \bang.
+
+For the inductive step, we shall show that none of the minefield steps
+that involves dereferencing  results in a \bang.  These  are the steps
+which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
+$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
+inserted by an earlier $GC(\ldots)$.   However the demand $\sigma'$  on basis of
+which the $GC(\ldots)$ would have inserted a  $\bot$ would have
+included the  current demand $\sigma$.     Thus it is
+enough to  show that the $\rightsquigarrow$ step would not lead to a
+\bang, even if $GC(\ldots)$ had been done with the current demand
+$\sigma$.
+
+We consider the rules for \CAR\  only.  
+The rest  of the rules involve similar reasoning. For the {\sc
+  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
+\sigma$, we  know that $\sigma$ is  non-null. Therefore
+the  liveness  of  $x$  includes  $\epsilon$,  and  the
+dereferencing $\heap(\rho(x))$ will go without \bang.
+
+For the  {\sc car-1-clo}  rule, observe that  there are
+two dereferences.  First $x$  is dereferenced to  get a
+cons  cell  and then  the  head  of  the cons  cell  is
+dereferenced  to  obtain  a  closure.   If  the  demand
+$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
+of    $x$   will    include    both   $\epsilon$    and
+$\acar\epsilon$,  and  a  GC with  this  liveness  will
+neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
+first   component  of   the   cons   cell.  Thus   both
+dereferences    can take place without 
+$\rightsquigarrow$ entering the   \bang\ state.
+\end{proof}
+
+Now we are ready to prove the main soundness result.
+
+\begin{theorem}
+The  minefield  execution of  no  program  can enter  a
+\bang\ state.
+\end{theorem}
+
+\begin{proof}
+Assume to  the contrary that  a program $P$  enters the
+\bang\  state.  We can  transform  $P$  to a  call-less
+program $P'$ such that  the minefield executions of $P$
+and $P'$  are identical  except for change  of variable
+names.             However,                 by
+Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
+$P'$,   a   call-less   program,   cannot   enter   the
+\bang\  state.  Therefore   $P$  also  cannot  enter  the
+\bang\ state.
+\end{proof}
 
@@
 
-We  now return  back  to the  problem of  computing  liveness using
-\Lfonly\  and a  concrete  demand $\sigma$.  Although  the process  is
-explained through an example, it can be generalized easily.
-
 \input{example-program}
@@
 non-terminal.
-{\color{Myblue}
+
 %\subsection{Generating liveness equations \Lv\  for function bodies}
@@
 each  GC  point  inside  the  function  based  on  this
-approximation.  The {\em summary demand} is
+approximation.  The 0-CFA-style {\em summary demand} is
 calculated by  taking a union  of the demands  at every
-call site of a function analogous to 0-CFA~\cite{Shivers:1988}.
+call site of a function.
 
@@
 
-%% For  the  running  example (Figure~\ref{fig:mot-example2-a}), $\length$  has  calls  from  $\main$  at
-%% $\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
-For   the   running  example   (Figure~\ref{fig:mot-example2-a}),   if
-$\length$  has calls  from  $\main$  with a  demand  $\epsilon$ and  a
-recursive      call      at       $\pi_6$      with      a      demand
-$\clazy\sigma_{\mathit{\length}}$. Thus:
+For  the  running  example (Figure~\ref{fig:mot-example2-a}),  $\length$  has  calls  from  $\main$  at
+$\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
+are $\epsilon$ and $\clazy\sigma_{\mathit{\length}}$. Thus:
 \begin{eqnarray*}
-\sigma_{\length}                    &=&                   \{\epsilon\}
-~\cup~{\clazy\sigma_{\mathit{\length}}}
+\sigma_{\length}    &=&
+ \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
 \end{eqnarray*}
-
-%% COMMENTED OUT BY PRASANNA
-%% As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
-%% stack variables $\pl$ and $\pa$ have the following liveness in terms
-%% of  $\sigma_{\length}$:
-%% \begin{align*}
-%%   \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
-%% %\end{align*}
-%% %\begin{align*}
-%%   \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
-%%   \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
-%%   \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
-%%   \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
-%%   \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
-%% \end{align*}
-
-%% In  summary, the  equations  generated during the
-%% analysis are:
-%% \begin{enumerate}
-%% \item   For  each   function  $\mathit{f}$,   equations
-%%   defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
-%% \item For each function $\mathit{f}$, an equation defining the summary
-%%   demand $\sigma_{\mathit f}$ on $e_f$.
-%% \item   For  each   function  $\mathit{f}$   (including
-%%   $\mainpgm$) an  equation  defining
-%%   liveness for each variable in  $e_{\mathit f}$.
-%% \end{enumerate}
-We compute the liveness of all variables and arguments of $\length$ in terms of $\sigma_{\length}$.
+As examples, the closure variables $\xl_{\mbox{$\pi$}_1}$ and $\xl_{\mbox{$\pi$}_5}$, and the
+stack variables $\pl$ and $\pa$ have the following liveness in terms
+of  $\sigma_{\length}$:
+\begin{align*}
+  \Lv_{\xl_{\mbox{$\pi$}_1}} &= \clazy\sigma_{\mathit{\length}} \\
+%\end{align*}
+%\begin{align*}
+  \Lv_{\xl_{\mbox{$\pi$}_5}} &=  (\acdr\Df{\length}{1}{\clazy} \cup
+  \clazy\Df{\length}{1}{\clazy}) \sigma_{\mathit{\length}} \\
+  \Lanv{\pl}{} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
+  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}} \\
+  \Lanv{\pa}{} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
+\end{align*}
+
+In  summary, the  equations  generated during the
+analysis are:
+\begin{enumerate}
+\item   For  each   function  $\mathit{f}$,   equations
+  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
+\item For each function $\mathit{f}$, an equation defining the summary
+  demand $\sigma_{\mathit f}$ on $e_f$.
+\item   For  each   function  $\mathit{f}$   (including
+  $\mainpgm$) an  equation  defining
+  liveness for each garbage in  $e_{\mathit f}$.
+\end{enumerate}
 
@@
 grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
-\clazy\rbrace$.  Let \var{X} denote  the non-terminal for a variable
+\clazy\rbrace$.  Let \var{$X$} denote  the non-terminal for a variable
 $X$ occurring on the LHS of the equations generated from the analysis.
@@
 several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
-as a start  symbol. The grammar for  \var{\Df{\length}{1}} and \var{\sigma_{\length}} comprise of the following productions:
-%%  As  an example, the grammar for \var{\Lanv{\xl}{}}
-%% and \var{\Lanv{\pa}{}} comprise the following productions:
+as a start symbol.  As  an example, the grammar for \var{\Lanv{\xl}{}}
+and \var{\Lanv{\pa}{}} comprise the following productions:
 \begin{eqnarray*}
-  %% \var{\Lanv{\xl}{}}  &\rightarrow& 
-  %% \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
-  %% \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
+  \var{\Lanv{\xl}{}}  &\rightarrow& 
+  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
+  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
   \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
@@
        \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
-\var{\sigma_{\length}}
+\langle {\sigma_{\length}} \rangle
 &\rightarrow&
-\epsilon  \mid \clazy\var{\sigma_{\length}}
-%% \\
-%% \var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
+\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
+\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
 \end{eqnarray*}
@@
 solution  of  \Lanv{x}{}.    
-Recall that the decision problem that we are interested in
+However, recall that the decision problem that we are interested in
 during GC is: 
-
+%%\begin{quote} 
 {\em
@@
 }
-
-We model this problem as one of deciding the membership of a CFG
-augmented with a fixed set of unrestricted productions. As shown in
-Appendix~\ref{sec:undec}, the membership problem of this class of
-grammars is undecidable.
+%%\end{quote}
+We model  this problem as one  of deciding the
+membership of  a CFG augmented  with a
+fixed set of unrestricted productions.
+
+\begin{definition}\label{def:specialgrammar}
+Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
+$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
+\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
+context-free    productions     that    contains    the
+distinguished  production   $S  \rightarrow  \alpha\$$,
+$\alpha$ is a  string of grammar symbols  that does not
+contain $S$, and $p_2$ is the fixed set of unrestricted
+productions    $\bcar\acar    \rightarrow    \epsilon$,
+$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
+\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
+and $\clazy\$ \rightarrow \$ $.
+\end{definition}
+
+%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
+%% that the results of liveness analysis of any program can be modeled by
+%% the kind  of grammar described above. The  following proposition shows
+%% that the converse also holds.
+%% \begin{proposition}
+%% Given    a     grammar    $G$    of    the     form    described    in
+%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
+%% program $p$  with program points  $\pi_i$ and variables $x$  such that
+%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
+%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
+%% \end{proposition}
+We first show that membership problem of the class of grammars in
+Definition~\ref{def:specialgrammar} is undecidable. However note that
+the grammars may not be necessarily generated from liveness analysis.
+\newcommand{\state}{\ensuremath{\mathsf{S}}}
+\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
+\begin{lemma}\label{lemma:grammar-undecidable}
+Given a grammar    $G$    of   the    kind    described    in
+Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
+consisting  of symbols \acar\  and \acdr\  only, the membership problem
+$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
+\end{lemma}
+
+\begin{proof}
+Given a Turing machine and  an input $w\in (1+0)^*$, we
+construct a grammar $G$ such that the machine will halt
+on    $w$    if    and   only    if    $\$    \in
+\mathscr{L}(G)$. The grammar includes  the fixed set of
+unrestricted               productions               in
+Definition~\ref{def:specialgrammar}.
+
+We shall represent a Turing machine (TM) configuration 
+as $w_l(\state,c)w_r$, where $w_l$ is the string to the
+left of the head, $w_r$ is the string to the right, $c$
+is the symbol under the head and \state\ is the current
+state of  the TM.   For each combination  of state
+and symbol  $(\state,c)$, the grammar will  contain the
+non-terminal \nont{}{c}. We shall synchronize each move
+of the TM to  a derivation  step using  a context
+free production, followed, if possible, by a derivation
+step using either  $\bcar\acar \rightarrow \epsilon$ or
+$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
+synchronization,  we  shall   establish  the  following
+invariant  relation between  the TM  configuration
+and the sentential form:
+
+\begin{quote}
+  If    the   configuration    of   the    TM   is
+  $w_l(\state,c)w_r$, then the  sentential form will be
+  $\overline{w}_l\nont{}{c}\,w_r\$ $,                where
+  $\overline{w}_l$ is  the same as $w_l$  but with each
+  symbol $d$ in $w_l$ replaced by $\overline{d}$.
+\end{quote}
+
+Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
+$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
+form    corresponding     to    the    initial     configuration    is
+$\nont{\mathit{init}}{c}w\$$ (we  can assume that there  is a production
+$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w\$$, where  \nont{}{} is
+the start symbol of the  grammar). Further correspondences between the
+TM moves and the grammar productions are as follows:
+
+\begin{enumerate}
+\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
+  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
+  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
+\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
+  are two productions
+  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
+  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
+\end{enumerate}
+The idea  behind the  productions is explained  with an
+example:  Assume that  the current  sentential form  is
+$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
+TM  has  a   transition  $(S_i,\acar)  \rightarrow
+(S_j,\acdr, L)$.  Since the next corresponding  step in
+the  derivation  has  to  be  done  without  any  prior
+knowledge of whether the symbol to the left of the tape
+is a \acar\  or a \acdr, two  productions are provided,
+and  the  invariant  will  be maintained  only  if  the
+production         $\nont{i}{\acar}         \rightarrow
+\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
+in  the  derivation.    This  gives  the  configuration
+$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
+Simplification   with    the   production   $\bcdr\acdr
+\rightarrow               \epsilon$              yields
+$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
+corresponds  to   the  changed  configuration   of  the
+TM.  Notice  carefully that a wrong  choice breaks
+the invariant  and it cannot be  recovered subsequently
+by any choice of productions.
+
+After the  TM has  halted, there  are further  ``cleanup'' derivations
+that derive  $\epsilon$ only if  the invariant has been  maintained so
+far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
+$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
+final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
+\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
+$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
+for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
+and                $\nont{\mathit{final}}{c}               \rightarrow
+\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
+\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
+\acdr\ on the right of the tape head.  This completes the
+reduction.\!
+\end{proof}
+
+We now show that the proof can be replayed for the class of grammars
+generated from liveness analysis of programs.
+
+\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
+Given   a    grammar    $G$    of   the    kind    described    in
+Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
+analysis  of a  program  and  a forward  access  path $\alpha$,    the
+membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
+\end{lemma}
+
+\begin{proof}
+Given a Turing machine and an input string, the  proof   of
+Lemma~\ref{lemma:grammar-undecidable} generates a grammar. 
+We shall define a function
+for   each  non-terminal $\nont{i}{c}$   introduced  in   this grammar
+such that the
+liveness  analysis  of  the function will  result
+in a set of productions that includes the productions for
+$\nont{i}{c}$ generated for the proof.
+As an example, it  can be verified that the grammar
+for the function shown in Figure~\ref{fig:turing-pgm}(b)    includes      the
+productions shown in Figure~\ref{fig:turing-pgm}(a). Here $\nont{\mathit{all}}{}$   corresponds   to  the   demand
+$\sigma_{\mathit{all}}$.
+
+The body of the function corresponds  to the RHS of the productions
+for $\nont{i}{c}$. Productions with the same  LHS non-terminal, but
+with differing non-terminal on RHS can be generated by 
+joining program fragments with $\SIF$.   If  there  are  $n$ such  functions,  then
+\mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
+distinct function in each branch.
+
+\begin{figure}[t]
+  \begin{minipage}{.45\columnwidth}
+    \begin{align*}
+      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
+      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
+      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
+      \acdr\nont{\mathit{all}}{}
+    \end{align*}
+  \end{minipage}
+  \begin{minipage}{.45\columnwidth}
+    \input{equivalent-program}
+  \end{minipage}\\
+  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
+  \caption{A possible grammar generated by the proof of
+    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
+  grammar.}\label{fig:turing-pgm}
+%\vspace*{-6mm}
+\end{figure}
+
+Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
+had a single non-terminal on the RHS. There are two characteristics of
+the grammar produced from liveness-analysis  that are are relevant for
+replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
+\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
+grammar  generated from  liveness  analysis of the constructed program
+will  have the  production
+$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
+\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
+$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
+\nont{\mathit{all}}{}$.
+
+It is clear that if the TM given as an instance of the halting problem
+accepts the input string, then the earlier derivation can be replayed,
+every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
+sentential form by $\epsilon$. However, if  the TM does not accept the
+input string, then every sentential form derived from the start symbol
+would    have     at    least    one    non-terminal from the grammar
+of Lemma~\ref{lemma:grammar-undecidable} that is different
+from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
+derivable from the grammar.
+\end{proof}
+
 %------------------------------------------------------------%
+\begin{figure}[t!]
+%% \scalebox{0.8}{
+%% \begin{pspicture}(0,-3)(50,12)
+%%   \begin{tabular}{cc}
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,-3)(50,12)
+%%   \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
+%%   \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+%%   \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+%%   \psset{arrows=->}
+%%   \ncline{t0}{t1}
+%%   \ncline{t1}{t2}
+%%   \putnode{l0}{t1}{7}{2}{\clazy}
+%%   \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{t1}{t1}
+%%   \putnode{l1}{t1}{0}{8}{\acdr, \clazy}
+%%   \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t2}{t2}
+%%   \putnode{l2}{t2}{0}{8}{\clazy}
+%% \end{pspicture}
+%% &
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,-3)(30,12)
+%%   \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
+%%   \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+%%   \psset{arrows=->}
+%%   \ncline{t0}{t1}
+%%   \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1}
+%%   \putnode{l1}{t1}{0}{8}{\acdr}
+%% \end{pspicture}
+%% \\
+%% (a) & (b) \\
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
+%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
+%%   \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+%%   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
+%% %  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+%%   \psset{arrows=->}
+%%   \ncline{t0}{t1}
+%%   \ncline{t1}{t2}
+%%   \putnode{l0}{t1}{5}{2}{\clazy}
+%% %  \ncline{t2}{t3}
+%% %  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
+%% %  \putnode{l2}{tD1}{0}{10}{\acdr}
+%%  \putnode{tD1}{t2}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+%%   \putnode{tD2}{tD1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
+%%   \psset{arrows=->}
+%%   \ncline{t2}{tD1}
+%%   \putnode{l1}{t2}{5}{2}{\bcar}
+%%   \ncline{tD1}{tD2}
+%%   \putnode{lD0}{tD1}{5}{2}{\clazy}
+%%   \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{tD1}{tD1}
+%%   \putnode{lD1}{tD1}{0}{8}{\acdr, \clazy}
+%%   \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{tD2}{tD2}
+%%   \putnode{lD2}{tD2}{0}{8}{\clazy}
+%% \end{pspicture}
+%% &
+%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+%% \begin{pspicture}(0,-5)(20,8) %\psframe(0,-5)(40,8)
+%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}}
+%%   \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
+%%   \psset{arrows=->}
+%%   \ncline{t0}{t1}
+%% \end{pspicture}
+%% \\ 
+%% (c) & (d)  \end{tabular}
+%% \end{pspicture}}
+\includegraphics[width=\columnwidth]{grammar-example.pdf}
+\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
+  converted into an automaton, and (b) its DFA. The same for \var{\Lanv{\pa}{}}
+  are in (c) and (d).}\label{fig:example-automata}
+%\vspace*{-2mm}
+\figrule
+\end{figure}
 
 %\subsection{A liveness-based GC scheme} \label{sec:live-clo}
+
 %%\subsubsection{From CFGs to GC-ready DFAs}
@@
 
-\input{cfg-to-nfa}
-}
+We circumvent the problem of undecidability by over approximating the
+CFG by non-deterministic finite state automata (NFA) using
+%% The NFAs  are then simplified using the
+%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
+%%are converted to DFAs.
+%%
+%% We use  the algorithm by 
+Mohri  and Nederhof~\cite{mohri00regular} method.
+%to approximate a CFG by a {\em strongly regular\/} grammar.  
+For example, the grammar fragment for %the non-terminal
+$\var{\Df{\length}{1}}$ after the Mohri-Nederhof transformation is:
+ \begin{eqnarray*}
+   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
+   \acdr\var{\Df{\length}{1}}
+   \mid \clazy\var{\Df{\length}{1}}\\
+   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
+   \mid \epsilon
+ \end{eqnarray*}
+
+The strongly regular grammar is converted  into a set of NFAs, one for
+each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
+done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
+pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
+constructing  the $\epsilon$-closure  till a  fixed point  is reached,
+after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  
+\cmt{ %% REMOVED FOR BLIND REVIEW
+  The details of the  algorithm, its correctness and  termination proofs are
+given  in~\cite{karkare07liveness,asati14lgc}. }
+The
+resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
+\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
+check if the source node of the edge has a path to a final 
+state.  If it does, we mark  the source node as final. Finally, we
+remove all the edges labeled \clazy\  and convert the automaton into a
+deterministic    automaton.    These steps   effectively   implement    the
+$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
+\clazy\ to  obtain forward access paths.  While checking for  liveness during
+garbage  collection, a  forward access path  is valid  only if  it can
+reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
+NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
+final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
+expectedly  says that for a  demand  $\sigma_{\length}$,  the
+liveness of  the argument of \length\  is $\acdr^{*}$  (the spine  of the  list is
+traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
+NFA     for     \var{\Lanv{\pa}{}}.      The     DFA in
+Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
+paths, reflecting  the  lazy  nature  of  our  language.   Since
+\length\ does not  evaluate the elements of  the argument list,
+the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
+liveness-based GC triggers beyond $\pi_9$.
+
   \SetStartEndCondition{ }{}{}%
@@
 
-\newcommand{\hgt}{4.3cm}
+\newcommand{\wdh}{.33\textwidth}
 \newcommand{\LGCLine}{blue}
@@
 
-%% \begin{figure*}[t!]
-%% \renewcommand{\arraystretch}{.1}
-%% \begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-%%    \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
-%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
-%% &  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
-%% %\\ (a) & (b)
-%% \end{tabular}
-%% \vspace*{-2mm}
-%%  \caption{Memory usage.  
-%% The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
-%% cells  in  the  active   semi-space  for  RGC  and  LGC
-%% respectively.  The \ReachLine\ curve represents the number of
-%% reachable cells and the  \UseLine\ curve represents the
-%% number  of  cells  that  are actually  live  (of  which
-%% liveness analysis does a static approximation).  x-axis
-%% is the time measured  in number of cons-cells allocated
-%% (scaled down by factor $10^5$). y-axis is the number of
-%% cons-cells (scaled down by $10^3$).}
-%% \label{fig:memory-usage} \figrule
-%% \vspace*{-1mm}
-%% \end{figure*}
-
 \begin{figure*}[t!]
-\renewcommand{\arraystretch}{.5}
-\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
-   \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
-\\
-   \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
-&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
-\end{tabular}
-\vspace*{-2mm}
+  \includegraphics[width=\wdh]{fibheap.pdf}
+  \includegraphics[width=\wdh]{sudoku.pdf}
+  \includegraphics[width=\wdh]{nperm.pdf}\\
+  \includegraphics[width=\wdh]{paraffins.pdf}
+  \includegraphics[width=\wdh]{lcss.pdf}
+  \includegraphics[width=\wdh]{huffman.pdf}\\
+  \includegraphics[width=\wdh]{knightstour.pdf}
+  \includegraphics[width=\wdh]{nqueens.pdf}
+  \includegraphics[width=\wdh]{deriv.pdf}\\
+  \includegraphics[width=\wdh]{treejoin.pdf}
+  \includegraphics[width=\wdh]{lambda.pdf}
+  \includegraphics[width=\wdh]{gc_bench.pdf}
   \caption{Memory usage.  
@@
 \label{fig:memory-usage} \figrule
-\vspace*{-1mm}
+%\vspace*{-1mm}
 \end{figure*}
@@
 \label{sec:GC-scheme}
+
+Our experimental setup  consists of an interpreter
+for  our  language,  a liveness  analyzer,  and  a
+single generation copying  collector.  The garbage
+collector can be configured to work on the
+basis  of  reachability  (RGC  mode)  or  use  
+liveness  DFAs  (LGC  mode).
 
@@
 \label{tab:exp-results}
-\vskip -2mm
+%\vskip -2mm
 \centering
 \include{data-liveness-analysis}
-\vskip -5mm
+%\vskip -5mm
 \include{autotable}
-\vskip -7mm
+%\vskip -7mm
 \end{table*}
@@
 \label{sec:experiments}
-Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
-interpreter  for our  language, (b)  a  liveness analyzer,  and (c)  a
-single generation  copying garbage  collector.  The  garbage collector
-can be configured  to work on the basis of  reachability (RGC mode) or
-use liveness DFAs (LGC mode).
-%
 Our benchmark  consists of programs taken  from nofib~\cite{nofib} and
@@
 be  repeatedly  visited  with   different  liveness  states.  We  have
-mitigated  this  problem  by  adding checks to avoid multiple traversals 
-of closure and function arguments.
-
-\vspace*{-2mm}
+mitigated  this  problem  by   implementing  a  heuristic  which  tags
+\CONS\ cells with the set of states already visited.
+
+%\vspace*{-2mm}
 \section{Related work} 
@@
 %% \begin{figure}[t]
-%% \centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
+%% \centerline{\epsfig{file=lambda.pdf, height=4cm, width=7cm}}
 %%  \caption{Memory  usage and garbage collection pattern of  {\tt
@@
 %% \end{figure}
-\vspace*{-2mm}
-\vskip -5mm
+%\vspace*{-2mm}
+%\vskip -5mm
 \section{Future work and conclusions}
@@
 
-\appendix
-\section{Soundness result for liveness analysis under minefield semantics}
-\label{sec:sound}
-We first need an auxiliary  result about minefield semantics. Consider
-a trace of  a minefield execution.  For every  minefield state $(\rho,
-S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
-$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
-application $s$) is  non-null.  This can be proved by  an induction on
-the number  of steps leading  to the  minefield state.  The  base step
-holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
-non-null. For the inductive step we  observe that for each step of the
-minefield semantics, if the demand $\sigma$  on the LHS of a minefield
-step  is non-null,  the  demand  on the  RHS  is  a transformation  of
-$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.
-
-Note that our proofs will be for a single round of minefield execution
-i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
-\ref{sec:semantics}).  With  minor variations,  the proof will  also be
-applicable for subsequent evaluations initiated by $\print$.
-%% We show  next that the  minefield execution of a  call-less program
-%% cannot go \bang.
-
-\begin{lemma}
-\label{lemma:call-less-cannot-go-bang}
-Consider the minefield execution of  a program without function calls.
-Such a program cannot enter the \bang\ state.
-\end{lemma}
-\begin{proof}
-Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
-minefield execution of a program.  We  show by induction on the number $n$
-of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
-enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
-(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
-[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
-does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
-transition cannot  enter a  \bang\ state.  Since  our programs  are in
-ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
-does not involve dereferencing, and thus cannot result in a \bang.
-
-For the inductive step, we shall show that none of the minefield steps
-that involves dereferencing  results in a \bang.  These  are the steps
-which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
-$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
-inserted by an earlier $GC(\ldots)$.   However the demand $\sigma'$  on basis of
-which the $GC(\ldots)$ would have inserted a  $\bot$ would have
-included the  current demand $\sigma$.     Thus it is
-enough to  show that the $\rightsquigarrow$ step would not lead to a
-\bang, even if $GC(\ldots)$ had been done with the current demand
-$\sigma$.
-
-We consider the rules for \CAR\  only.  
-The rest  of the rules involve similar reasoning. For the {\sc
-  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
-\sigma$, we  know that $\sigma$ is  non-null. Therefore
-the  liveness  of  $x$  includes  $\epsilon$,  and  the
-dereferencing $\heap(\rho(x))$ will go without \bang.
-
-For the  {\sc car-1-clo}  rule, observe that  there are
-two dereferences.  First $x$  is dereferenced to  get a
-cons  cell  and then  the  head  of  the cons  cell  is
-dereferenced  to  obtain  a  closure.   If  the  demand
-$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
-of    $x$   will    include    both   $\epsilon$    and
-$\acar\epsilon$,  and  a  GC with  this  liveness  will
-neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
-first   component  of   the   cons   cell.  Thus   both
-dereferences    can take place without 
-$\rightsquigarrow$ entering the   \bang\ state.
-\end{proof}
-
-Now we are ready to prove the main soundness result.
-
-\begin{theorem}
-The  minefield  execution of  no  program  can enter  a
-\bang\ state.
-\end{theorem}
-
-\begin{proof}
-Assume to  the contrary that  a program $P$  enters the
-\bang\  state.  We can  transform  $P$  to a  call-less
-program $P'$ such that  the minefield executions of $P$
-and $P'$  are identical  except for change  of variable
-names.             However,                 by
-Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
-$P'$,   a   call-less   program,   cannot   enter   the
-\bang\  state.  Therefore   $P$  also  cannot  enter  the
-\bang\ state.
-\end{proof}
-
-\section{Proof of Undecidability}
-\label{sec:undec}
-We show that  membership problem  of the  class of  grammars defined below 
-%%Definition~\ref{def:specialgrammar}
-is undecidable.   We then  show that  the grammars  in this  class are
-indeed generated from liveness analysis.
-\begin{definition}\label{def:specialgrammar}
-Consider the  grammar $(N,T,  p_1\cup p_2,S)$  in which
-$N$ is  a set  of non-terminals,  $T =  \{\acar, \acdr,
-\bcar,  \bcdr,  \clazy,  \$\}$,   $p_1$  is  a  set  of
-context-free    productions     that    contains    the
-distinguished  production   $S  \rightarrow  \alpha\$$,
-$\alpha$ is a  string of grammar symbols  that does not
-contain $S$, and $p_2$ is the fixed set of unrestricted
-productions    $\bcar\acar    \rightarrow    \epsilon$,
-$\bcdr\acdr    \rightarrow   \epsilon$,    $\clazy\acar
-\rightarrow \clazy$,  $\clazy\acdr \rightarrow \clazy$,
-and $\clazy\$ \rightarrow \$ $.
-\end{definition}
-
-%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
-%% that the results of liveness analysis of any program can be modeled by
-%% the kind  of grammar described above. The  following proposition shows
-%% that the converse also holds.
-%% \begin{proposition}
-%% Given    a     grammar    $G$    of    the     form    described    in
-%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
-%% program $p$  with program points  $\pi_i$ and variables $x$  such that
-%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
-%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
-%% \end{proposition}
-
-\newcommand{\state}{\ensuremath{\mathsf{S}}}
-\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}
-\begin{lemma}\label{lemma:grammar-undecidable}
-Given a grammar    $G$    of   the    kind    described    in
-Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
-consisting  of symbols \acar\  and \acdr\  only, the membership problem
-$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
-\end{lemma}
-
-\begin{proof}
-Given a Turing machine and  an input $w\in (1+0)^*$, we
-construct a grammar $G$ such that the machine will halt
-on    $w$    if    and   only    if    $\$    \in
-\mathscr{L}(G)$. The grammar includes  the fixed set of
-unrestricted               productions               in
-Definition~\ref{def:specialgrammar}.
-
-We shall represent a Turing machine (TM) configuration 
-as $w_l(\state,c)w_r$, where $w_l$ is the string to the
-left of the head, $w_r$ is the string to the right, $c$
-is the symbol under the head and \state\ is the current
-state of  the TM.   For each combination  of state
-and symbol  $(\state,c)$, the grammar will  contain the
-non-terminal \nont{}{c}. We shall synchronize each move
-of the TM to  a derivation  step using  a context
-free production, followed, if possible, by a derivation
-step using either  $\bcar\acar \rightarrow \epsilon$ or
-$\bcdr\acdr   \rightarrow    \epsilon$.    After   each
-synchronization,  we  shall   establish  the  following
-invariant  relation between  the TM  configuration
-and the sentential form:
-
-\begin{quote}
-  If    the   configuration    of   the    TM   is
-  $w_l(\state,c)w_r$, then the  sentential form will be
-  $\overline{w}_l\nont{}{c}\,w_r\$ $,                where
-  $\overline{w}_l$ is  the same as $w_l$  but with each
-  symbol $d$ in $w_l$ replaced by $\overline{d}$.
-\end{quote}
-
-Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
-$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
-form    corresponding     to    the    initial     configuration    is
-$\nont{\mathit{init}}{c}w\$$ (we  can assume that there  is a production
-$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w\$$, where  \nont{}{} is
-the start symbol of the  grammar). Further correspondences between the
-TM moves and the grammar productions are as follows:
-
-\begin{enumerate}
-\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
-  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
-  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
-\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
-  are two productions
-  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
-  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
-\end{enumerate}
-The idea  behind the  productions is explained  with an
-example:  Assume that  the current  sentential form  is
-$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
-TM  has  a   transition  $(S_i,\acar)  \rightarrow
-(S_j,\acdr, L)$.  Since the next corresponding  step in
-the  derivation  has  to  be  done  without  any  prior
-knowledge of whether the symbol to the left of the tape
-is a \acar\  or a \acdr, two  productions are provided,
-and  the  invariant  will  be maintained  only  if  the
-production         $\nont{i}{\acar}         \rightarrow
-\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
-in  the  derivation.    This  gives  the  configuration
-$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
-Simplification   with    the   production   $\bcdr\acdr
-\rightarrow               \epsilon$              yields
-$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
-corresponds  to   the  changed  configuration   of  the
-TM.  Notice  carefully that a wrong  choice breaks
-the invariant  and it cannot be  recovered subsequently
-by any choice of productions.
-
-After the  TM has  halted, there  are further  ``cleanup'' derivations
-that derive  $\epsilon$ only if  the invariant has been  maintained so
-far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
-$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
-final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
-\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
-$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
-for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
-and                $\nont{\mathit{final}}{c}               \rightarrow
-\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
-\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
-\acdr\ on the right of the tape head.  This completes the
-reduction.\!
-\end{proof}
-
-We now show that the proof can be replayed for the class of grammars
-generated from liveness analysis of programs.
-
-\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
-Given   a    grammar    $G$    of   the    kind    described    in
-Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
-analysis  of a  program  and  a forward  access  path $\alpha$,    the
-membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
-\end{lemma}
-
-\begin{proof}
-Given a Turing machine and an input string, the  proof   of
-Lemma~\ref{lemma:grammar-undecidable} generates a grammar. 
-We shall define a function
-for   each  non-terminal $\nont{i}{c}$   introduced  in   this grammar
-such that the
-liveness  analysis  of  the function will  result
-in a set of productions that includes the productions for
-$\nont{i}{c}$ generated for the proof.
-As an example, it  can be verified that the grammar
-for the function shown in Figure~\ref{fig:turing-pgm}(b)    includes      the
-productions shown in Figure~\ref{fig:turing-pgm}(a). Here $\nont{\mathit{all}}{}$   corresponds   to  the   demand
-$\sigma_{\mathit{all}}$.
-
-The body of the function corresponds  to the RHS of the productions
-for $\nont{i}{c}$. Productions with the same  LHS non-terminal, but
-with differing non-terminal on RHS can be generated by 
-joining program fragments with $\SIF$.   If  there  are  $n$ such  functions,  then
-\mainpgm\ creates  a $n$-way  branch and  inserts a  single call  to a
-distinct function in each branch.
-
-\begin{figure}[t]
-  \begin{minipage}{.45\columnwidth}
-    \begin{align*}
-      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
-      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
-      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
-      \acdr\nont{\mathit{all}}{}
-    \end{align*}
-  \end{minipage}
-  \begin{minipage}{.45\columnwidth}
-    \input{equivalent-program}
-  \end{minipage}\\
-  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
-  \caption{A possible grammar generated by the proof of
-    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
-  grammar.}\label{fig:turing-pgm}
-\vspace*{-6mm}
-\end{figure}
-
-Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
-had a single non-terminal on the RHS. There are two characteristics of
-the grammar produced from liveness-analysis  that are are relevant for
-replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
-\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
-grammar  generated from  liveness  analysis of the constructed program
-will  have the  production
-$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
-\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
-$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
-\nont{\mathit{all}}{}$.
-
-It is clear that if the TM given as an instance of the halting problem
-accepts the input string, then the earlier derivation can be replayed,
-every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
-sentential form by $\epsilon$. However, if  the TM does not accept the
-input string, then every sentential form derived from the start symbol
-would    have     at    least    one    non-terminal from the grammar
-of Lemma~\ref{lemma:grammar-undecidable} that is different
-from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
-derivable from the grammar.
-\end{proof}
-
-%%\clearpage
+%\clearpage
 \balance
@@
 \bibliographystyle{abbrv}
-\end{document}
-
-\clearpage
-
+%\clearpage
 \appendix
-%\section{Appendix: not to be included in the paper.}
-%%%%% prasanna minefield
 \begin{figure*}[b!]
-\noindent{\Large \bf A. Complete minefield semantics}
+\section{Complete Minefield Semantics}\label{sec:minefield}
+
 \begin{center}\footnotesize

### 2016-04-20 paper/paper.tex
fit in 10+bib pages

@@
 memory requirement from  1X to 1198X.  As  an added benefit,
-there is also a reduction in the overall execution time in 5 out of 12
+there is also a reduction in the overall execution time in 4 out of 12
 benchmark programs.
@@
 \vskip -5mm
+{\color{Myblue}
 \include{autotable}
+}
 \vskip -7mm
@@
 \label{sec:experiments}
+{\color{Myblue}
 Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
@@
 can be configured  to work on the basis of  reachability (RGC mode) or
-use liveness DFAs (LGC mode).
+use liveness DFAs (LGC mode).}
 %
@@
 RGC in most cases) and better  for 3 benchmarks (2X speedup in the
-best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
-benchmark that allocates complete binary
-trees  of various  sizes that  are  never used  by the  program. As  a
-result,  the benchmark  highly favours  LGC.  The  benchmark has  been
-included  for  completeness,  and  we  do  not  consider  its  numbers
-as being representative of real programs.
+best case).  
+%% Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
+%% benchmark that allocates complete binary
+%% trees  of various  sizes that  are  never used  by the  program. As  a
+%% result,  the benchmark  highly favours  LGC.  The  benchmark has  been
+%% included  for  completeness,  and  we  do  not  consider  its  numbers
+%% as being representative of real programs.
 
@@
 
-Note that our proofs will be for a single round of minefield execution
-i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
-\ref{sec:semantics}).  With  minor variations,  the proof will  also be
-applicable for subsequent evaluations initiated by $\print$.
+Our proofs will be for a single round of minefield execution i.e., the
+evaluation  of  $(\mainpgm)$  to  its  WHNF  driven  by  the  printing
+mechanism~(Section \ref{sec:semantics}).   With minor  variations, the
+proof will also be applicable  for subsequent evaluations initiated by
+$\print$.
 %% We show  next that the  minefield execution of a  call-less program
 %% cannot go \bang.
-
+%
 \begin{lemma}
@@
 
+\section*{Acknowledgement}
+We would like to thank the anonymous referees for their valuable suggestions. We thank 
+Evgeny Kapun for his suggestions regarding the undecidability proof.
+
 %%\clearpage

### 2016-04-22 paper/paper.tex
toappear
removed old colouring

@@
 
-\definecolor{Myblue}{rgb}{.2,0,1}
-\definecolor{Myred}{rgb}{1,.2,.2}
-
-\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
+%\definecolor{Myblue}{rgb}{.2,0,1}
+%\definecolor{Myred}{rgb}{1,.2,.2}
+
+%\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
 %\newcommand{\added}[1]{{\color{Myblue}{#1}}}
-\newcommand{\added}[1]{#1}
-\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
+%\newcommand{\added}[1]{#1}
+%\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
 %\newcommand{\warning}[1]{{\color{Myred}{#1}}}
-\newcommand{\warning}[1]{}
+%\newcommand{\warning}[1]{}
 %\newcommand{\blankout}[1]{}
@@
 \begin{document}
-
+\toappear{}
 \special{papersize=8.5in,11in}
@@
 \setlength{\pdfpagewidth}{\paperwidth}
-
-%% \linenumbers 
-% --- Author Metadata here ---
-%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
-%over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
-%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE. 
-% --- End of Author Metadata ---
 
@@
 
-\added{While  the idea  of  using static  analysis  to improve  memory
+While  the idea  of  using static  analysis  to improve  memory
   utilization  has  been  shown  to   be  effective  for  {\em  eager}
@@
   include   closures    (runtime   representations    of   unevaluated
-  expressions).   }   \todelete{Designing  a  liveness  based  garbage
-  collector   for  lazy   languages  poses   significant  challenges.}
+  expressions).
+%%\todelete{Designing  a  liveness  based  garbage
+%%  collector   for  lazy   languages  poses   significant  challenges.}
 Firstly, since  data is made  live by  evaluation of closures,  and in
@@
 contributions of this paper are:
-\added{
+
 \begin{itemize}
-\item {\color{Myblue}We  propose an interprocedural liveness  based GC
+\item We  propose an interprocedural liveness  based GC
   scheme  for a  lazy first  order functional  language and  prove its
@@
     shaham02estimating,  asati14lgc, karkare06effectiveness}  involved
-  either imperative or eager functional languages.}
+  either imperative or eager functional languages.
 %% The novelty of our approach is in handling of
@@
 \end{itemize}
-}
-\todelete{
-\begin{itemize}
-\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
-  functional language and  prove its correctness. The 
-  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
-  correct with respect to the specification.   
-\item The analysis results  in a set  of context-free  grammars along
-  with  a fixed  set  of non-context-free  productions.  The  decision
-  whether   to  copy  a  cell   during  GC  translates  to a  membership problem  for such  grammars.  In  this
-  paper, we  have shown the  undecidability of this problem and 
-  circumvent the  problem by over-approximating the  grammars
-  with DFAs.
-\item We  have implemented a garbage  collector that uses the  DFAs to
-  retain live  cells.  Our  experiments reveal  interesting space-time
-  trade-offs  in  the  engineering  of  the  collector---for  example,
-  updating liveness  information carried in closures  during execution
-  results  in   more  garbage  being  collected.    Empirical  results
-  show the effectiveness of liveness-based GC.
-\end{itemize}
-
-As far  as we know, this  is the first work  on liveness-based GC      for      lazy       languages.       All      previous
-attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
-  asati14lgc, karkare06effectiveness} have  involved either imperative
-or eager functional languages.
-}
+
 \subsection{Organization of the paper}
@@
 
-\added{  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
+ As noted  earlier, $x.\alpha  \in \Lv$  specifies the
   liveness of the reference $\heap\llbracket\rho(x), \alpha\rrbracket$
@@
   closure variable  $y_{\pi}$ is recorded  along with the  closure for
-  $s$ so that the GC can refer to it during garbage collection.  }
-%} %comment
+  $s$ so that the GC can refer to it during garbage collection.
 
@@
 \end{center}
-\caption{Minefield  semantics for  \CAR\   and \LET. Complete minefield semantics is provided in~\cite{TR16-1}.
-  \warning{Add a note mentioning that we handle the issue of requiring
-    a      specific     demand      by      introducing     a      new
-    symbol}\label{fig:minefield-semantics-for-some}}
+\caption{Minefield  semantics for  \CAR\   and \LET. Complete minefield semantics is provided in~\cite{TR16-1}.}
 \vspace*{-2mm}
@@
 We shall now present a proof of soundness of the analysis presented in
-Section~\ref{sec:liveness-analysis}. \added{It is easy to see that the analysis
+Section~\ref{sec:liveness-analysis}. It is easy to see that the analysis
 correctly identifies the liveness of stack variables. A stack variable
@@
 create a closure variable. This is correctly captured by the
-\LET\ rule in Figure~\ref{fig:live-judge}.}
+\LET\ rule in Figure~\ref{fig:live-judge}.
 Proving soundness  for closure  variables is more complex. Here  are the
@@
   Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
-  of  each \LET\  expression.  \added{Note  that, unlike  eager
+  of  each \LET\  expression. Note  that, unlike  eager
     languages,   memory  is   allocated  only   during  execution   of
-    \LET\ expressions}. During GC, we track each reference in the root
+    \LET\ expressions. During GC, we track each reference in the root
   set and heap that is declared  dead by our analysis.  Any attempt to
@@
  \begin{enumerate} 
-\added{
-
  \item \label{env}  For each of  these environments, $GC(\ldots)$  calculates a
@@
     \rrbracket = \ell$.
-%% $x.\alpha$  is  live  and
-%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
-%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
-%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
-}
 
@@
 non-terminal.
-{\color{Myblue}
 %\subsection{Generating liveness equations \Lv\  for function bodies}
@@
 \input{cfg-to-nfa}
-}
 
@@
 \vskip -5mm
-{\color{Myblue}
 \include{autotable}
-}
 \vskip -7mm
@@
 \label{sec:experiments}
-{\color{Myblue}
 Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
@@
 can be configured  to work on the basis of  reachability (RGC mode) or
-use liveness DFAs (LGC mode).}
+use liveness DFAs (LGC mode).
 %

### 2016-04-22 paper/cfg-to-nfa.tex
all changes from Eric's review

@@
-We circumvent the problem of  undecidability by over approximating the
-CFG by non-deterministic finite state automata (NFA) using
-Mohri   and   Nederhof method~\cite{mohri00regular}.   This   method
+\added{We circumvent the problem of  undecidability by over approximating the
+CFG by non-deterministic finite state automata (NFA) using the method of 
+Mohri   and   Nederhof~\cite{mohri00regular}}.   This   method
 transforms a CFG $G$ into a strongly  regular grammar $R$ such that $\lang{G} \subseteq \lang{R}$. 
@@
 pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
-constructing  the $\epsilon$-closure  till a  fixed point  is reached,
+\added{constructing  the $\epsilon$-closure  until a  fixed point  is reached},
 after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
@@
 
-The resulting automaton has edges labeled with \acar, \acdr\ and
-\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
-check if the source node of the edge  has a path to a final state.  If
-it does, we mark the source node  as final. Finally, we remove all the
-edges labeled \clazy\  and convert the automaton  into a deterministic
-automaton.   These steps  effectively implement  the $\hookrightarrow$
-simplification rules for \bcar, \bcdr, and
-\clazy\ to  obtain forward access paths.  While checking for  liveness during
-garbage collection,  a forward  access path  is valid  only if  it can
-reach a final state.
-
 \newcommand{\nfaD}{%
-  \scalebox{.7}{
+  \scalebox{.70}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(3,-2)(28,10)
-      \putnode{t0}{origin}{2}{-1}{} 
-      \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
+    \begin{pspicture}(3,-1)(28,9)%\psgrid
+      \putnode{t0}{origin}{2}{1}{} 
+      \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{$q_0$}}}
       \hspace{5mm} 
-      \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}} 
+      \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{$q_1$}}} 
       \psset{arrows=->} \ncline{t0}{t1} 
@@
       \putnode{l0}{t1}{7}{2}{\clazy} 
-      \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1} 
+      \nccurve[angleA=45, angleB=135, ncurv=2, nodesep=-1]{t1}{t1} 
+      \putnode{l1}{t1}{0}{6}{\acdr, \clazy} 
+      \nccurve[angleA=45,
+        angleB=135, ncurv=1.5, nodesep=-1]{t2}{t2} 
+      \putnode{l2}{t2}{0}{7}{\clazy}
+    \end{pspicture}
+}}
+\newcommand{\nfaDF}{%
+  \scalebox{.70}{
+    \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
+    \begin{pspicture}(3,-1)(28,9)%\psgrid
+      \putnode{t0}{origin}{2}{1}{} 
+      \putnode{t1}{t0}{8}{0}{\pscirclebox[doubleline=true]{\mbox{$q_0$}}}
+      \hspace{5mm} 
+      \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{$q_1$}}} 
+      \psset{arrows=->} \ncline{t0}{t1} 
+      \ncline{t1}{t2} 
+      \putnode{l0}{t1}{7}{2}{\clazy} 
+      \nccurve[angleA=45, angleB=135, ncurv=1.5, nodesep=-1]{t1}{t1} 
       \putnode{l1}{t1}{0}{7}{\acdr, \clazy} 
-      \nccurve[angleA=45,
-        angleB=135, ncurv=2, nodesep=-1]{t2}{t2} 
+      \nccurve[angleA=45, angleB=135, ncurv=1.5, nodesep=-1]{t2}{t2} 
       \putnode{l2}{t2}{0}{7}{\clazy}
@@
 \newcommand{\dfaD}{%
-  \scalebox{.7}{
+  \scalebox{.70}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-    \begin{pspicture}(2,-2)(13,8)
-      \putnode{t0}{origin}{2}{0}{}
-      \putnode{t1}{t0}{8}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}     
+    \begin{pspicture}(2,-1)(13,9)
+      \putnode{t0}{origin}{2}{1}{}
+      \putnode{t1}{t0}{8}{0}{\pscirclebox[doubleline=true]{\mbox{$q_0$}}}     
       \psset{arrows=->}    
       \ncline{t0}{t1}     
-      \nccurve[angleA=45, angleB=135, ncurv=2, nodesep=-1]{t1}{t1} 
+      \nccurve[angleA=45, angleB=135, ncurv=1.5, nodesep=-1]{t1}{t1} 
       \putnode{l1}{t1}{0}{7}{\acdr}
@@
 }}
-The  NFA for \var{\Df{\length}{1}} is 
-\nfaD\ and the final DFA is \dfaD.
+
+\added{The resulting automaton has edges labeled with \acar, \acdr\ and
+\clazy\ only.  For our example, the  NFA for \var{\Df{\length}{1}} is 
+\nfaD.
+In this  automaton, for every  edge labeled  \clazy, we
+check if the source node of the edge  has a path to a final state.  If
+it does, we mark the source node  as final. For \var{\Df{\length}{1}}, we get \nfaDF. Finally, we remove all the
+edges labeled \clazy\  and convert the automaton  into a deterministic
+automaton.   These steps  effectively implement  the $\hookrightarrow$
+simplification rules for \bcar, \bcdr, and
+\clazy\ to  obtain forward access paths.   The final DFA obtained for \var{\Df{\length}{1}} is \dfaD.}
+While checking for  liveness during
+garbage collection,  a forward  access path  is valid  only if  it can
+reach a final state.
 This  expectedly says  that
@@
 \newcommand{\dfaE}{%
-  \scalebox{.7}{
+  \scalebox{.70}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
@@
      \putnode{t0}{origin}{0}{0}{} 
-     \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{\ \ \ \
-      }}} 
+     \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{$q_0$}}} 
      \hspace{5mm} 
-   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \ }}}
+   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{$q_1$}}}
    \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} 
    \putnode{l0}{t1}{5}{2}{\clazy}
-   \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}} 
+   \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{$q_2$}}} 
    \psset{arrows=->} \ncline{t2}{tD1} 
@@
 \newcommand{\dfaF}{%
-  \scalebox{.7}{
+  \scalebox{.70}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
     \begin{pspicture}(3,-2)(24,3)
-     \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
-      }}} 
+     \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{$q_0$}}} 
        \hspace{5mm} 
-        \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \  }}}
+        \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{$q_1$}}}
          \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
@@
 \newcommand{\dfaG}{%
-  \scalebox{.7}{
+  \scalebox{.70}{
     \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
     \begin{pspicture}(3,-2)(10,3)
-    \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{7}{0}{\pscirclebox{\mbox{\ \ \ \
-      }}} \hspace{1mm} \psset{arrows=->} \ncline{t0}{t1}
+    \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{7}{0}{\pscirclebox{\mbox{$q_0$}}} \hspace{1mm} \psset{arrows=->} \ncline{t0}{t1}
       \end{pspicture}
@@
 liveness-based GC triggers beyond $\pi_9$.
-
-%% \begin{wrapfigure}{r}{.25\textwidth}
-%% \scalebox{0.5}{
-%% \begin{tabular}{c}
-%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-%% \begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
-%%   }}} \hspace{5mm} \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \
-%%   }}}
-%%   \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
-%%   \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\       \       \       \
-%%   }}} 
-%%   \psset{arrows=->} \ncline{t2}{tD1} \putnode{l1}{t2}{5}{2}{\bcar}
-%%   \nccurve[angleA=45,
-%%   angleB=135,                                                 ncurv=4,
-%%   nodesep=-1]{tD1}{tD1} \putnode{lD1}{tD1}{0}{10}{\acdr}
-%% \end{pspicture} 
-%% \\
-%% (a)
-%% \\
-%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-%% \begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
-%%   }}} 
-%%   \hspace{5mm} 
-%%   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\  \ \ \  }}}
-%%   \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
-%% \end{pspicture}
-%% \\
-%% (b)
-%% \\
-%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
-%% \begin{pspicture}(0,-5)(50,8) %\psframe(0,-5)(40,8)
-%%   \putnode{t0}{origin}{5}{0}{\var{\Lanv{\pa}{}}} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \
-%%   }}} \hspace{5mm} \psset{arrows=->} \ncline{t0}{t1}
-%% \end{pspicture}
-%% \\ 
-%% (c)
-%% \end{tabular}}
-%%  \caption{(a) grammar  rules for \var{\Lanv{\pa}{}} converted  into an
-%%   automaton and  its DFAs after simplification.}\label{fig:main1}
-%% \end{wrapfigure}

### 2016-04-22 paper/paper.tex
all changes from Eric's review

@@
 
-%\definecolor{Myblue}{rgb}{.2,0,1}
+\definecolor{Myblue}{rgb}{.2,0,1}
 %\definecolor{Myred}{rgb}{1,.2,.2}
@@
 %\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
-%\newcommand{\added}[1]{{\color{Myblue}{#1}}}
+\newcommand{\added}[1]{{\color{Myblue}{#1}}}
 %\newcommand{\added}[1]{#1}
@@
 \end{center}
-\caption{Minefield  semantics for  \CAR\   and \LET. Complete minefield semantics is provided in~\cite{TR16-1}.}
+\caption{Minefield semantics for \CAR\ and \LET. Complete minefield
+  semantics is provided
+  in~\cite{TR16-1}.\label{fig:minefield-semantics-for-some}}
 \vspace*{-2mm}
@@
 %% $\pi_{12}$ and a recursive call at $\pi_6$. The demands on these calls
-For   the   running  example   (Figure~\ref{fig:mot-example2-a}),   if
+\added{For   the   running  example   (Figure~\ref{fig:mot-example2-a}),
 $\length$  has calls  from  $\main$  with a  demand  $\epsilon$ and  a
 recursive      call      at       $\pi_6$      with      a      demand
-$\clazy\sigma_{\mathit{\length}}$. Thus:
+$\clazy\sigma_{\mathit{\length}}$.} Thus:
 \begin{eqnarray*}
@@
 several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
-as a start  symbol. The grammar for  \var{\Df{\length}{1}} and \var{\sigma_{\length}} comprise of the following productions:
+as a start  symbol. \added{The grammars for  \var{\Df{\length}{1}} and \var{\sigma_{\length}} consist of the following productions}:
 %%  As  an example, the grammar for \var{\Lanv{\xl}{}}
@@
 \stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
-consists  of  forward  access  paths  only.  Then  does
-$\alpha \in \sigma$?
+consists  of  forward  access  paths  only.  \added{Then  is
+$\alpha \in \sigma$?}
 }

### 2016-04-22 paper/cfg-to-nfa.tex
minor rewrites

@@
 
-\added{The resulting automaton has edges labeled with \acar, \acdr\ and
-\clazy\ only.  For our example, the  NFA for \var{\Df{\length}{1}} is 
-\nfaD.
-In this  automaton, for every  edge labeled  \clazy, we
-check if the source node of the edge  has a path to a final state.  If
-it does, we mark the source node  as final. For \var{\Df{\length}{1}}, we get \nfaDF. Finally, we remove all the
-edges labeled \clazy\  and convert the automaton  into a deterministic
-automaton.   These steps  effectively implement  the $\hookrightarrow$
-simplification rules for \bcar, \bcdr, and
-\clazy\ to  obtain forward access paths.   The final DFA obtained for \var{\Df{\length}{1}} is \dfaD.}
-While checking for  liveness during
-garbage collection,  a forward  access path  is valid  only if  it can
-reach a final state.
-This  expectedly says  that
-for  a  demand  $\sigma_{\length}$,   the  liveness  of  the  argument
-of \length\ is $\acdr^{*}$ (the spine of the list is traversed).
+\added{The  resulting   automaton  has   edges  labeled   with  \acar,
+  \acdr\   and  \clazy\   only.   For   our  example,   the  NFA   for
+  \var{\Df{\length}{1}} is  \nfaD.  In this automaton,  for every edge
+  labeled \clazy, we check  if the source node of the  edge has a path
+  to a final state.  If it does, we mark the source node as final. For
+  \var{\Df{\length}{1}}, we get \nfaDF, in which $q_0$ becomes a final
+  state. Finally, we remove all  the edges labeled \clazy\ and convert
+  the  automaton   into  a   deterministic  automaton.    These  steps
+  effectively implement the $\hookrightarrow$ simplification rules for
+  \bcar, \bcdr, and \clazy\ to obtain forward access paths.  The final
+  DFA obtained  for \var{\Df{\length}{1}}  is \dfaD.}   While checking
+for liveness during garbage collection, a forward access path is valid
+only if it can  reach a final state.  This expectedly  says that for a
+demand $\sigma_{\length}$, the liveness of the argument of \length\ is
+$\acdr^{*}$ (the spine of the list is traversed).
 
@@
      \putnode{t0}{origin}{0}{0}{} 
-     \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{$q_0$}}} 
+     \putnode{t1}{t0}{8}{0}{\pscirclebox{\mbox{$q_3$}}} 
      \hspace{5mm} 
-   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{$q_1$}}}
+   \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{$q_4$}}}
    \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} 
    \putnode{l0}{t1}{5}{2}{\clazy}
-   \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{$q_2$}}} 
+   \putnode{tD1}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{$q_5$}}} 
    \psset{arrows=->} \ncline{t2}{tD1} 
    \putnode{l1}{t2}{5}{2}{\bcar}
-   \nccurve[angleA=45,angleB=135,ncurv=2,nodesep=-1]{tD1}{tD1} 
+   \nccurve[angleA=45,angleB=135,ncurv=1.5,nodesep=-1]{tD1}{tD1} 
    \putnode{lD1}{tD1}{0}{7}{\acdr}
@@
     \begin{pspicture}(3,-2)(24,3)
-     \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{$q_0$}}} 
+     \putnode{t0}{origin}{2}{0}{}
+     \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{$q_3$}}} 
      \hspace{5mm} 
-        \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{$q_1$}}}
-         \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{5}{2}{\clazy}
+     \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{$q_4$}}}
+     \psset{arrows=->} \ncline{t0}{t1} 
+     \ncline{t1}{t2} 
+     \putnode{l0}{t1}{5}{2}{\clazy}
     \end{pspicture}
@@
     \begin{pspicture}(3,-2)(10,3)
-    \putnode{t0}{origin}{2}{0}{} \putnode{t1}{t0}{7}{0}{\pscirclebox{\mbox{$q_0$}}} \hspace{1mm} \psset{arrows=->} \ncline{t0}{t1}
+      \putnode{t0}{origin}{2}{0}{}
+      \putnode{t1}{t0}{7}{0}{\pscirclebox{\mbox{$q_3$}}} 
+      \hspace{1mm} \psset{arrows=->} 
+      \ncline{t0}{t1}
     \end{pspicture}

### 2016-04-26 paper/paper.tex
Fixed comments except WHNF

@@
 garbage collector to efficiently  reclaim memory, empirical studies on
-Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
-Haskell~\cite{rojemo96lag} programs have shown that garbage collectors
-leave uncollected a large number  of memory objects that are reachable
-but not live (here {\em live} means the object can potentially be used
-by the program at a  later stage).  This results in unnecessary memory
-retention.
+Scheme~\cite{karkare06effectiveness}  and,  Haskell~\cite{rojemo96lag}
+programs have shown that garbage  collectors leave uncollected a large
+number of  memory objects that are  reachable but not live  (here {\em
+  live} means the  object can potentially be used by  the program at a
+later stage).  This results in unnecessary memory retention.
 
@@
 liveness  information  in  a  closure with  a  more  precise  version.
-Because  of   these  reasons,  the  garbage   collector  also  becomes
-significantly more  complicated than a liveness-based  collector for a
+For these  reasons,  the  garbage   collector  also  becomes
+significantly more  complicated than a liveness-based  collector for an
 eager language.
@@
 $(\length\ \pz)$.  The heap is represented  by a graph in which a node
-either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
+either  represents  atomic  values   ($\NIL$,  integers,  etc.),  or  a
 \CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
@@
 describing the  liveness sets  in a function body in
-terms of a symbolic demand $\sigma$  and \Lfonly. However, we are yet to
+terms of a symbolic demand $\sigma$  and \Lfonly. However, we have yet to
 describe (i) how to obtain demand transformers \Lfonly\ from the rule
@@
 In addition to  evaluated data, our collector  also reclaims closures.
-For this, we  had to modify the standard run-time  representation of a
+For this, we  had to modify the standard run-time  representation for
 closures  to carry  liveness  of its free  variables and  periodically
@@
   the depth of DFA, merging  nearly-equivalent states and using better
-  representation(for example BDDs~\cite{Bryant86})  and algorithms for
+  representation (for example BDDs~\cite{Bryant86})  and algorithms for
   automata manipulation.  We also  need to investigate the interaction

### 2016-04-26 paper/paper.tex
grammary check.

@@
 %% cells can be  reclaimed.
-Thus if a GC takes  place at $\pi$ with the heap shown
+Thus, if a GC takes  place at $\pi$ with the heap shown
 in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
@@
 
-In  this work  we show  that  static analysis  of heap  data can  help
+In  this work,  we show  that  static analysis  of heap  data can  help
 garbage  collectors   in  reclaiming   more  garbage.    The  specific
@@
 \item We  propose an interprocedural liveness  based GC
-  scheme  for a  lazy first  order functional  language and  prove its
+  scheme  for a  lazy first-order functional  language and  prove its
   correctness.  To the  best of our knowledge, this is  the first work
@@
   for such grammars.  Earlier, this decision problem was assumed to be
-  undecidable, and the context-free grammar was over approximated by a
+  undecidable, and the context-free grammar was over-approximated by a
   regular  grammar.   In  this  paper,  we  provide  a  proof  of  the
@@
 Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
-collector,  in  particular  the  use  of  liveness  DFAs  for  GC.     We     report    our    experimental     results    in
+collector,  in  particular,  the  use  of  liveness  DFAs  for  GC.     We     report    our    experimental     results    in
 Section~\ref{sec:experiments}    along    with   some    observations.
@@
 Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
-is a first order language with lazy semantics. Programs are restricted
+is a first-order language with lazy semantics. Programs are restricted
 to        be        in        Administrative        Normal        Form
@@
 $\ell_{\ans}$.  In addition, \print\ is a function modeling a printing
-mechanism---a   standard   run-time   support  assumption   for   lazy
+mechanism---a   standard   runtime   support  assumption   for   lazy
 languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The
@@
 environment  that  maps  variables  $x_i$ to  locations  $\ell_i$  and
-$\heap[\ell := d]$  indicates updation of  \heap\ at $\ell$
+$\heap[\ell := d]$  indicates an update of  \heap\ at $\ell$
 with  $d$.   $\rho \oplus  \rho'$  represents  the environment  $\rho$
@@
 $s$. Instead,  as the {\sc let}  rule shows, a closure  is created and
-bound to  $x$. The  program points which  trigger evaluation  of these
+bound to  $x$. The  program points which  trigger the evaluation  of these
 closures   are  an $\SIF$   condition ({\sc  IF-clo})   and
@@
 refer up  to the  second and  third members of  $x$.  A  {\em liveness
-  environment} is a mapping from  variables to access paths, but often
+  environment} is a mapping from  variables to access paths but often
 expressed as  a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
@@
 dependences beginning  at a  variable at an  evaluation point.   As an
-example,  consider again  the program  in Figure~\ref{fig:lv-closure}.
+example,  consider the program  in Figure~\ref{fig:lv-closure} again.
 The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
@@
 (\length~\px_{\mathit{\pi}_1})\rbrack$.    Indeed,   the   chains   of
-closures  in   the  heap  are   a  runtime  representation   of  these
+closures  in   the  heap  are    runtime  representations   of  these
 dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
@@
 In  a   lazy  language,   an  expression   is  not   evaluated  unless
-required. Therefore  the null  demand ($\emptyset$) does  not generate
+required. Therefore,  the null  demand ($\emptyset$) does  not generate
 liveness in any of the rules defining $\mathit{ref}$ or $\mathcal{L}$.
@@
 
-We shall now present a proof of soundness of the analysis presented in
+We shall now present a proof of the soundness of the analysis presented in
 Section~\ref{sec:liveness-analysis}. It is easy to see that the analysis
@@
   final step shows that no program  without function calls can enter the
-  \bang\ state. As a consequence  no program (with or without function
+  \bang\ state. As a consequence,  no program (with or without function
   calls) can enter the \bang\ state.
@@
     consulted during actual GC  is computed from an over-approximation
-    of this demand.  Thus the soundness  result on the modeled GC will
+    of this demand.  Thus, the soundness  result on the modeled GC will
     also apply for the actual GC.} The demand for the initial state is
@@
   continuation frames  on the stack  $S$ are also  similarly augmented
-  with their demands.  Thus a stacked entry now takes the form $(\rho,
+  with their demands.  Thus, a stacked entry now takes the form $(\rho,
   \ell, e, \sigma)$.
@@
 bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
-enters the  \bang\ state ({\sc  car-bang}). Otherwise
+enters the  \bang\ state ({\sc  car-bang}). Otherwise,
 the transition is similar to the earlier {\sc car-clo}
@@
 
-The solution is to  also treat the operations mentioned
+The solution is to  treat the operations mentioned
 above  symbolically.  We  introduce three  new symbols:
@@
 \end{align*}
-Thus \bcar\ selects those entries in $\sigma$ that have leading \acar, and removes the leading \acar\ from them.
+Thus, \bcar\ selects those entries in $\sigma$ that have leading \acar, and removes the leading \acar\ from them.
 The symbol \clazy\ reduces the set of strings following it to a set containing only $\epsilon$. It filters out, however, the empty set of strings.
@@
 demand  only by  prefixing it  with symbols  in the  alphabet $\lbrace
-\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore we  can assume
+\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore, we  can assume
 that $\Lf{f}{i}{\sigma}$ has the closed form:
@@
 the variables in a function  with respect to a symbolic
-demand  $\sigma$.   To  get   the  actual  liveness  we
+demand  $\sigma$.   To  get   the  actual  liveness,  we
 calculate an  over-approximation of the  actual demands
@@
 we would like to use a more precise liveness, based on the actual paths
-taken  during  execution.   Therefore   we  create  separate  liveness
+taken  during  execution.   Therefore,   we  create  separate  liveness
 automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
@@
 %% To avoid scalability issues (and to avoid performing
-%% part  of the  liveness computation  at run  time) we
+%% part  of the  liveness computation  at runtime) we
 %% defined an  0-CFA version of this  liveness analysis
@@
 %% user  functions and  to $\CONS$).   These were  then
-%% converted to  DFA for  run-time consultation  by the
+%% converted to  DFA for  runtime consultation  by the
 %% garbage   collector.     Experiments   confirm   the
@@
 In addition to  evaluated data, our collector  also reclaims closures.
-For this, we  had to modify the standard run-time  representation for
+For this, we  had to modify the standard runtime  representation for
 closures  to carry  liveness  of its free  variables and  periodically
@@
   car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
-\sigma$, we  know that $\sigma$ is  non-null. Therefore
-the  liveness  of  $x$  includes  $\epsilon$,  and  the
+\sigma$, we  know that $\sigma$ is  non-null. Therefore,
+the  liveness  of  $x$  includes  $\epsilon$  and  the
 dereferencing $\heap(\rho(x))$ will go without \bang.
@@
 two dereferences.  First $x$  is dereferenced to  get a
-cons  cell  and then  the  head  of  the cons  cell  is
+cons  cell,  and then  the  head  of  the cons  cell  is
 dereferenced  to  obtain  a  closure.   If  the  demand
@@
 neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
-first   component  of   the   cons   cell.  Thus   both
+first   component  of   the   cons   cell.  Thus,   both
 dereferences    can take place without 
@@
 names.             However,                 by
-Lemma~\ref{lemma:call-less-cannot-go-bang} we know that
+Lemma~\ref{lemma:call-less-cannot-go-bang}, we know that
 $P'$,   a   call-less   program,   cannot   enter   the
-\bang\  state.  Therefore   $P$  also  cannot  enter  the
+\bang\  state.  Therefore,   $P$  also  cannot  enter  the
 \bang\ state.
@@
 non-terminal \nont{}{c}. We shall synchronize each move
-of the TM to  a derivation  step using  a context
-free production, followed, if possible, by a derivation
+of the TM to  a derivation  step using  a context-free production, followed, if possible, by a derivation
 step using either  $\bcar\acar \rightarrow \epsilon$ or
@@
 Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
-had a single non-terminal on the RHS. There are two characteristics of
-the grammar produced from liveness-analysis  that are are relevant for
+had a single non-terminal on the RHS. There are two characteristics, of
+the grammar produced from liveness-analysis,  that are relevant for
 replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
@@
 of Lemma~\ref{lemma:grammar-undecidable} that is different
-from $\nont{\mathit{final}}{c}$.   Thus  $\$$  would  not  be
+from $\nont{\mathit{final}}{c}$.   Thus,  $\$$  would  not  be
 derivable from the grammar.

### 2016-04-26 paper/paper.tex
Fixed WHNF reference issue

@@
 collector that uses  the results of the liveness  analysis to preserve
-live  objects,  both evaluated  (i.e.,  in  WHNF) and  closures.   Our
+live  objects,  both evaluated  and  closures.   Our
 experiments confirm  that for  programs running with  a liveness-based
@@
 garbage collector to efficiently  reclaim memory, empirical studies on
-Scheme~\cite{karkare06effectiveness}  and,  Haskell~\cite{rojemo96lag}
+Scheme~\cite{karkare06effectiveness}  and  Haskell~\cite{rojemo96lag}
 programs have shown that garbage  collectors leave uncollected a large
@@
 
-Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
-A non-empty location either contains  a value in  \emph{WHNF} (a  number, the
-empty  list  $\NIL$ or  a  \CONS\  cell  with  possibly  unevaluated
-constituents) or  a {\em closure}.   A closure  is a pair  $\langle s,
+Here $\mathit{Loc}$  is a countable set  of locations in the  heap.  A
+non-empty location either contains a {\em closure}, or a value in Weak
+Head Normal Form (WHNF)\cite{Jones87}. In our case, a WHNF value
+can be a number, the empty list  $\NIL$ or a \CONS\ cell with possibly
+unevaluated   constituents.   A   closure  is   a  pair   $\langle  s,
 \rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
-maps free variables of $s$ to their respective locations. Since all data objects are boxed, we model an environment as a mapping
-from the set  of variables of the program  $\mathit{Var}$ to locations
-in the  heap.
+maps free  variables of $s$  to their respective locations.  Since all
+data objects are boxed, we model  an environment as a mapping from the
+set of  variables of  the program $\mathit{Var}$  to locations  in the
+heap.
 

### 2016-04-26 paper/paper.tex
minor updates after discussion with Amey

@@
 Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
-collector,  in  particular,  the  use  of  liveness  DFAs  for  GC.     We     report    our    experimental     results    in
-Section~\ref{sec:experiments}    along    with   some    observations.
-Section~\ref{sec:relatedwork}  discusses  previous   work  related  to
-GC  and   liveness  and  Section~\ref{sec:conclusion}
+collector, in particular, the use of  liveness DFAs for GC.  We report
+our experimental  results in Section~\ref{sec:experiments}  along with
+some  observations.  Section~\ref{sec:relatedwork}  discusses previous
+work  related  to  GC and  liveness  and  Section~\ref{sec:conclusion}
 discusses possible extensions and concludes the paper.
@@
        (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
-       (\CAR\,\, x) & \\
-       (\CDR\,\, x) & \mbox{\em --- selectors} \\
-       (\NULLQ\,\, x) & \\
-       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
+       (\CAR\,\, x) &  \mbox{\em --- selects $1^{st}$ part of \CONS} \\
+       (\CDR\,\, x) & \mbox{\em --- selects $2^{nd}$ part of \CONS} \\
+       (\NULLQ\,\, x) & \mbox{\em --- returns 0 if x is not $\NIL$}\\
+       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  generic arithmetic} \\
 %%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
@@
 dependences beginning  at a  variable at an  evaluation point.   As an
-example,  consider the program  in Figure~\ref{fig:lv-closure} again.
-The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
-expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
-$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
+example,  in  Figure~\ref{fig:lv-closure},   a  dependence  chain  for
+$\px_{\mathit{\pi}_1}$  begins   with  the   variable  $\pz$   at  the
+evaluation  point $\psi_3$.   The variable  \pz\ returned  at $\psi_3$
+depends on  \py\ through  the expression  $(+~\py~1)$.  $\py$  in turn
+depends  on   the  closure  variable   $\px_{\mathit{\pi}_1}$  through
+$(\length~\px_{\mathit{\pi}_1})$.  We
 denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
@@
 \begin{enumerate}
-\item We  maintain static estimates  of the memory required  to create
+\item We  statically over-approximate the memory required  to create
   the closures for each function body.  On entering a function, if the

### 2016-04-27 paper-for-archive/paper.tex
Updated arXiv version with the changes

@@
 \usepackage{makecell}
-
+\usepackage[numbers,sort&compress]{natbib}
 % to save space in lists
@@
 % For aesthetic reasons, we recommend 'three authors at a time'
-% i.e. three 'name/affiliation blocks' be placed beneath the title.
+% i.e., three 'name/affiliation blocks' be placed beneath the title.
 %
@@
 garbage collector to efficiently  reclaim memory, empirical studies on
-Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
+Scheme~\cite{karkare06effectiveness}   and  on
 Haskell~\cite{rojemo96lag} programs have shown that garbage collectors
@@
 liveness  information  in  a  closure with  a  more  precise  version.
-Because  of   these  reasons,  the  garbage   collector  also  becomes
+For  these  reasons,  the  garbage   collector  also  becomes
 significantly more  complicated than a liveness-based  collector for a
@@
 \begin{itemize}
-\item We  propose a liveness  based GC scheme  for a lazy  first order
-  functional language and  prove its correctness.  To the  best of our
-  knowledge,  this is  the first  work on  liveness-based GC  for lazy
-  languages.       All      previous      attempts~\cite{shaham01heap,
-    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
-    karkare06effectiveness} have  involved either imperative  or eager
-  functional languages.  The novelty of our approach is in handling of
-  liveness  of closures,  an  issue  that arises  due  to laziness  at
-  runtime.
+\item We  propose an interprocedural liveness  based GC
+  scheme  for a  lazy first-order functional  language and  prove its
+  correctness.  To the  best of our knowledge, this is  the first work
+  that uses  the results  of an  interprocedural liveness  analysis to
+  garbage    collect    both     evaluated    data    and    closures.
+  Thomas~\cite{Thomas19951} describes a  copying garbage collector for
+  the Three  Instruction Machine (TIM)~\cite{Fairbairn1987}  that only
+  preserves live closures  in a function's environment  (also called a
+  frame). However, in the absence of  details, it is not clear whether
+  a) the  scope of the  method is  interprocedural, and b)  it handles
+  algebraic datatypes like lists (the original design of TIM did not).
+  All  other  previous attempts~\cite{shaham01heap,  ran.shaham-sas03,
+    shaham02estimating,  asati14lgc, karkare06effectiveness}  involved
+  either imperative or eager functional languages.
 
@@
        (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
-       (\CAR\,\, x) & \\
-       (\CDR\,\, x) & \mbox{\em --- selectors} \\
-       (\NULLQ\,\, x) & \\
-       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
+       (\CAR\,\, x) &  \mbox{\em --- selects $1^{st}$ part of \CONS} \\
+       (\CDR\,\, x) &  \mbox{\em --- selects $2^{nd}$ part of \CONS} \\
+       (\NULLQ\,\, x) & \mbox{\em --- returns 0 if x is not $\NIL$} \\
+       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em generic arithmetic} \\
 %%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
@@
 
-Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
-A non-empty location either contains  a value in  \emph{WHNF} (a  number, the
-empty  list  $\NIL$ or  a  \CONS\  cell  with  possibly  unevaluated
-constituents) or  a {\em closure}.   A closure  is a pair  $\langle s,
+Here $\mathit{Loc}$  is a countable set  of locations in the  heap.  A
+non-empty location either contains a {\em closure}, or a value in Weak
+Head Normal Form  (WHNF)\cite{Jones87}. In our case, a  WHNF value can
+be a  number, the  empty list  $\NIL$ or a  \CONS\ cell  with possibly
+unevaluated   constituents.   A   closure  is   a  pair   $\langle  s,
 \rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
-maps free variables of $s$ to their respective locations. Since all data objects are boxed, we model an environment as a mapping
-from the set  of variables of the program  $\mathit{Var}$ to locations
-in the  heap.
+maps free  variables of $s$  to their respective locations.  Since all
+data objects are boxed, we model  an environment as a mapping from the
+set of  variables of  the program $\mathit{Var}$  to locations  in the
+heap.
 
@@
 dependences beginning  at a  variable at an  evaluation point.   As an
-example,  consider again  the program  in Figure~\ref{fig:lv-closure}.
-The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
-expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
-$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
-denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
-\leftarrow            (+~\py~1),\,           \py            \leftarrow
-(\length~\px_{\mathit{\pi}_1})\rbrack$.    Indeed,   the   chains   of
-closures  in   the  heap  are   a  runtime  representation   of  these
+example,  in  Figure~\ref{fig:lv-closure},   a  dependence  chain  for
+$\px_{\mathit{\pi}_1}$  begins   with  the   variable  $\pz$   at  the
+evaluation  point $\psi_3$.  The  variable \pz\  returned at  $\psi_3$
+depends on  \py\ through  the expression  $(+~\py~1)$.  $\py$  in turn
+depends  on   the  closure  variable   $\px_{\mathit{\pi}_1}$  through
+$(\length~\px_{\mathit{\pi}_1})$.  We denote this chain of dependences
+as   $\lbrack  \psi_3\!\!\!:\!\!\!\pz   \leftarrow  (+~\py~1),\,   \py
+\leftarrow (\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed, the chains
+of  closures  in  the  heap  are a  runtime  representation  of  these
 dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
@@
 \label{sec:GC-scheme}
-
-Our experimental setup  consists of an interpreter
-for  our  language,  a liveness  analyzer,  and  a
-single generation copying  collector.  The garbage
-collector can be configured to work on the
-basis  of  reachability  (RGC  mode)  or  use  
-liveness  DFAs  (LGC  mode).
-
 In        the       liveness        analysis       described        in
@@
 \begin{enumerate}
-\item We  maintain static estimates  of the memory required  to create
+\item We  statically over-approximate the memory required  to create
   the closures for each function body. On entering a function, if the
@@
 \label{sec:experiments}
+Our experimental setup  consists of an interpreter
+for  our  language,  a liveness  analyzer,  and  a
+single generation copying  collector.  The garbage
+collector can be configured to work on the
+basis  of  reachability  (RGC  mode)  or  use  
+liveness  DFAs  (LGC  mode).
+
 Our benchmark  consists of programs taken  from nofib~\cite{nofib} and
@@
 In addition to  evaluated data, our collector  also reclaims closures.
-For this, we  had to modify the standard run-time  representation of a
+For this, we  had to modify the standard run-time  representation of
 closures  to carry  liveness  of its free  variables and  periodically

### 2016-04-27 paper-for-archive/paper.tex
Split graphs into subfigures and fixed minor grammar errors

@@
 \usepackage{schemepgm}
-\usepackage{graphicx}
+\usepackage{graphicx, subcaption}
 \usepackage{graphpap}
@@
 \usepackage{bcprules}
-\usepackage{subfigure}
+%\usepackage{subfigure}
 \usepackage[boxed]{algorithm2e}
@@
 For  these  reasons,  the  garbage   collector  also  becomes
-significantly more  complicated than a liveness-based  collector for a
+significantly more  complicated than a liveness-based  collector for an
 eager language.
@@
 
-\newcommand{\wdh}{.33\textwidth}
+\newcommand{\wdh}{.75\textwidth}
 \newcommand{\LGCLine}{blue}
@@
 \begin{figure*}[t!]
-  \includegraphics[width=\wdh]{fibheap.pdf}
-  \includegraphics[width=\wdh]{sudoku.pdf}
-  \includegraphics[width=\wdh]{nperm.pdf}\\
-  \includegraphics[width=\wdh]{paraffins.pdf}
-  \includegraphics[width=\wdh]{lcss.pdf}
-  \includegraphics[width=\wdh]{huffman.pdf}\\
-  \includegraphics[width=\wdh]{knightstour.pdf}
-  \includegraphics[width=\wdh]{nqueens.pdf}
-  \includegraphics[width=\wdh]{deriv.pdf}\\
-  \includegraphics[width=\wdh]{treejoin.pdf}
-  \includegraphics[width=\wdh]{lambda.pdf}
-  \includegraphics[width=\wdh]{gc_bench.pdf}
  \caption{Memory usage.  
@@
 cons-cells (scaled down by $10^3$).}
-\label{fig:memory-usage} \figrule
-%\vspace*{-1mm}
+\label{fig:memory-usage}
+ \begin{subfigure}{\textwidth}
+   \includegraphics[width=\wdh]{fibheap.pdf}
+ \end{subfigure}
+ \begin{subfigure}{\textwidth}
+   \includegraphics[width=\wdh]{sudoku.pdf}
+ \end{subfigure}
+ \begin{subfigure}{\textwidth}
+   \includegraphics[width=\wdh]{nperm.pdf}
+ \end{subfigure}
 \end{figure*}
 
-%\vspace*{-6.5mm}
+\begin{figure*}[t!]
+  \ContinuedFloat
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{lcss.pdf}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{huffman.pdf}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{knightstour.pdf}
+  \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[t!]
+  \ContinuedFloat
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{deriv.pdf}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{treejoin.pdf}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{lambda.pdf}
+  \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[t!]
+  \ContinuedFloat
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{gc_bench.pdf}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\wdh]{nqueens.pdf}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+   \includegraphics[width=\wdh]{paraffins.pdf}
+  \end{subfigure} 
+ \figrule
+\end{figure*}
+
 \section{The GC scheme}
@@
 
-Memory    usage    graphs    for     the    benchmarks    are    shown
+Memory   usage    graphs   for    the   benchmarks   are    shown   in
 Figure~\ref{fig:memory-usage}.  In  all the  programs we can  see that
@@
 processing the data at  the end of the program. Note  that the size of
-an LGC cell is 1.16 times the size of a RGC cell.
+an LGC  cell is 1.16 times  the size of  a RGC cell as  it potentially
+might have  to store references  to liveness DFA of  closure arguments
+(if the cell is a closure).
 
@@
 be  repeatedly  visited  with   different  liveness  states.  We  have
-mitigated  this  problem  by   implementing  a  heuristic  which  tags
-\CONS\ cells with the set of states already visited.
+mitigated  this  problem  by   implementing  a  heuristic  which  avoids 
+revisiting closures and function arguments more than once. 
+%% Since the 
+%% liveness automata for a closure argument encompasses all possible uses 
+%% of the closure, multiple traversals for closures can be safely avoided.
 

### 2016-04-28 paper/paper.tex
Fixed as per discussion with Prof.AS

@@
 number of  memory objects that are  reachable but not live  (here {\em
-  live} means the  object can potentially be used by  the program at a
+  live} means the  object may potentially be used by  the program at a
 later stage).  This results in unnecessary memory retention.
@@
        (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
-       (\CAR\,\, x) &  \mbox{\em --- selects $1^{st}$ part of \CONS} \\
-       (\CDR\,\, x) & \mbox{\em --- selects $2^{nd}$ part of \CONS} \\
-       (\NULLQ\,\, x) & \mbox{\em --- returns 0 if x is not $\NIL$}\\
+       (\CAR\,\, x) &  \mbox{\em --- selects first part of \CONS} \\
+       (\CDR\,\, x) & \mbox{\em --- selects second part of \CONS} \\
+       (\NULLQ\,\, x) & \mbox{\em --- returns true if x is $\NIL$}\\
        (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  generic arithmetic} \\
@@
 non-empty location either contains a {\em closure}, or a value in Weak
-Head Normal Form (WHNF)\cite{Jones87}. In our case, a WHNF value
-can be a number, the empty list  $\NIL$ or a \CONS\ cell with possibly
-unevaluated   constituents.   A   closure  is   a  pair   $\langle  s,
-\rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
+Head Normal Form (WHNF)\cite{Jones87}. For our implementation, a value
+in WHNF is either a number, or  the empty list $\NIL$ or a \CONS\ cell
+with possibly unevaluated constituents.  A  closure is a pair $\langle
+s, \rho\rangle$ in which $s$ is an unevaluated application, and $\rho$
 maps free variables  of $s$ to their respective  locations.  Since all
@@
 
-As  demonstrated by  the gap  between the  \RGCLine\ and  the \UseLine\
-lines, a  large number of  cells which are  unused by the  program are
-still copied  during RGC. LGC does  a much better job  of closing this
-gap but still falls short of the  precision achieved by LGC in case of
-eager languages ~\cite{asati14lgc}.  A major source of inefficiency in
-LGC is  multiple traversals  of already copied  heap cells.  Since LGC
-does not mark the heap cells after the first visit, the same cells can
-be  repeatedly  visited  with   different  liveness  states.  We  have
-mitigated  this  problem  by  adding checks to avoid multiple traversals 
-of closure and function arguments.
+As   demonstrated  by   the  gap   between  the   \RGCLine\  and   the
+\UseLine\  lines, a  large number  of cells  which are  unused by  the
+program are  still copied during  RGC. LGC does  a much better  job of
+closing this  gap but still falls  short of the precision  achieved by
+LGC in case of eager  languages ~\cite{asati14lgc}.  A major source of
+inefficiency  in LGC  is multiple  traversals of  already copied  heap
+cells.  Since LGC does not mark  the heap cells after the first visit,
+the  same cells  can  be repeatedly  visited  with different  liveness
+states.  This can  be  seen in  the cells touched  per  GC for  $huffman$
+benchmark in Table~\ref{fig:memory-usage}, even though it uses only 72
+cells, number of cells touched is  88.  We have mitigated this problem
+by adding checks to avoid  multiple traversals of closure and function
+arguments.
 

### 2016-04-28 paper/paper.tex
space saved

@@
 the  same cells  can  be repeatedly  visited  with different  liveness
-states.  This can  be  seen in  the cells touched  per  GC for  $huffman$
-benchmark in Table~\ref{fig:memory-usage}, even though it uses only 72
-cells, number of cells touched is  88.  We have mitigated this problem
-by adding checks to avoid  multiple traversals of closure and function
-arguments.
+states.   This  can   be  seen  in  the  cells  touched   per  GC  for
+\verb@huffman@  benchmark in  Table~\ref{fig:memory-usage}.  Even  though the  benchmark  uses  72 cells  maximum,  the
+average  number  of cells  touched per GC is 88. During  a  GC,  for every  traversal  the
+arguments of a  closure are garbage collected using  the same automata
+(stored in  the closure itself). We mark the arguments {\em
+  visited}  during  the  first  traversal  and  avoid subsequent
+traversals.
 
@@
 (\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
-[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
-does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
-transition cannot  enter a  \bang\ state.  Since  our programs  are in
+[\;]_{\heap},\, e_{\mainpgm})$.  A call  to $GC(\ldots)$ in  this state
+does  nothing. Also,  the  $\rightsquigarrow$
+transition cannot  enter a  \bang\ state.  This is because  our programs  are in
 ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
@@
 $\rightsquigarrow$ entering the   \bang\ state.
-\end{proof}
+\end{proof} \vspace*{-1mm}
 
@@
 \bang\ state.
-\end{proof}
+\end{proof} \vspace*{-1mm}
 
@@
 reduction.\!
-\end{proof}
+\end{proof} \vspace*{-2mm}
 
@@
   grammar.}\label{fig:turing-pgm}
-\vspace*{-2mm}
+\vspace*{-4mm}
 \end{figure}
@@
 derivable from the grammar.
-\end{proof}
+\end{proof} \vspace*{-2mm}
 
@@
 
-%%\clearpage
+\clearpage
 \balance

### 2016-04-28 paper/paper.tex
replaced While with Whereas as suggested

@@
 garbage reclaimed and  a  reduction in  the peak memory requirement 
-of programs.   While this  technique has already  been shown  to yield
+of programs.   Whereas this  technique has already  been shown  to yield
 benefits for  eager first-order  languages, the  lack of  a statically
@@
 
-While  the runtime  system  of most  functional  languages includes  a
-garbage collector to efficiently  reclaim memory, empirical studies on
+Whereas  the runtime  system  of most  functional  languages includes  a
+garbage collector to reclaim memory efficiently, empirical studies on
 Scheme~\cite{karkare06effectiveness}  and  Haskell~\cite{rojemo96lag}
@@
 
-While  the idea  of  using static  analysis  to improve  memory
+Whereas  the idea  of  using static  analysis  to improve  memory
   utilization  has  been  shown  to   be  effective  for  {\em  eager}
@@
 (ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
-functions  are  variables.  While  this  restriction  does not  affect
+functions  are  variables.  Whereas  this  restriction  does not  affect
 expressibility,  this form  has  the benefit  of  making explicit  the
@@
 
-A notion  that generalizes liveness  is {\em demand}.   While liveness
+A notion  that generalizes liveness  is {\em demand}.   Whereas liveness
 gives the patterns of future uses of a variable, demand represents the

### 2016-05-02 paper-for-archive/data-liveness-analysis.tex

\renewcommand{\arraystretch}{.9}
\begin{tabular}{|l||c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|c@{\ }|} \hline
 \multicolumn{1}{|r||}{Program} & \rot{\tt fibheap} & \rot{\tt sudoku} & \rot{\tt nperm} & \rot{\tt paraffins} & \rot{\tt lcss} & \rot{\tt huffman} & \rotmore{\tt knightstour} & \rot{\tt nqueens} & \rot{\tt deriv} & \rot{\tt treejoin} & \rot{\tt lambda} & \rot{\tt gc\_bench} \\ \hline
 \#CFG Nonterminals & 621 & 1422 & 662 & 1174 & 642 & 499 & 660 & 404 & 328 & 615 & 669 & 390 \\ \hline
 \#CFG Rules & 1176 & 2009 & 866 & 1773 & 1206 & 818 & 883 & 643 & 468 & 1328 & 1088 & 450 \\ \hline
 \#DFA States & 1761 & 4283 & 1546 & 3346 & 1666 & 1414 & 1519 & 889 & 809 & 1803 & 1703 & 571 \\ \hline
 \#DFA Transitions & 2829 & 7690 & 2522 & 6086 & 2726 & 2528 & 2420 & 1170 & 1435 & 2797 & 2580 & 788 \\ \hline
 DFA Gen Time (sec) & 37.28 & 655.41 & 0.94 & 13.22 & 8.66 & 4.00 & 10.97 & 0.36 & 0.61 & 903.14 & 11.01 & 0.10 \\ \hline
\end{tabular}\\
 (a) Data for Liveness Analysis

### 2016-05-02 paper/paper.tex
saved space
fixed plots in arxiv paper deleted non-essential files in paper.

@@
 %\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
-\newcommand{\added}[1]{{\color{Myblue}{#1}}}
-%\newcommand{\added}[1]{#1}
+%\newcommand{\added}[1]{{\color{Myblue}{#1}}}
+\newcommand{\added}[1]{#1}
 %\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}
@@
 future uses of the value of an expression.  The demand on an expression
-$e$   is   also   a   set    of   access   paths---the   subset   of
-$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
+$e$   is   also   a   set    of   access   paths
+%---the   subset   of $\{\acar,\acdr\}^\ast$ 
+which the  context of $e$ may  explore of $e$'s
 result.   To see  the need  for demands,  consider the expression
@@
 
-\section*{Acknowledgement}
+\section*{Acknowledgements}
 We would like to thank the anonymous referees for their valuable suggestions. We thank 

### 2016-05-06 paper/paper.tex
updated name
Fixed name issue
Corrected capitalization of headings

@@
 % 1st. author
-\authorinfo {Prasanna Kumar. K} 
+\authorinfo {Prasanna Kumar. K.} 
        {IIT Bombay, Mumbai 400076, India}
@@
 %------------------------------------------------------------%
-\subsection{Motivating example}
+\subsection{Motivating Example}
 \label{sec:motiv}
@@
 
-\subsection{Organization of the paper}
+\subsection{Organization of the Paper}
 
@@
 discusses possible extensions and concludes the paper.
-\section{The target language---syntax and semantics}
+\section{The Target Language---Syntax and Semantics}
 \label{sec:defs}
@@
 
-\subsection{Liveness analysis for lazy languages}
+\subsection{Liveness analysis for Lazy Languages}
 \label{sec:liveness-analysis}
@@
 
-\subsection{Soundness of analysis}
+\subsection{Soundness of Analysis}
 
@@
 
-\section{Towards a  computable form of liveness}\label{sec:computing}
+\section{Towards a  Computable Form of Liveness}\label{sec:computing}
 %\section{Generating and storing liveness information}\label{sec:computing}
@@
 %\subsection{Modifying liveness rules}
-\paragraph{Symbolic representation of operations:}
+\paragraph{Symbolic Representation of Operations:}
 The   $\mathit{ref}$   rule   for   \CONS,   shown   in
@@
 %\subsection{Generating equations for $\Lfonly_{\mathit f}$}
-\paragraph{Computing function summaries $\Lfonly_{\mathit f}$:}
+\paragraph{Computing Function Summaries $\Lfonly_{\mathit f}$:}
 Given a  function $\mathit{f}$, we now  describe how to
@@
 %\subsection{Generating liveness equations \Lv\  for function bodies}
-\paragraph{Handling user-defined functions:}
+\paragraph{Handling User-defined Functions:}
 \label{sec:bodylivenessbodies}
@@
 %\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
-\paragraph{From liveness sets to context-free grammars:}\label{sec:grammar-formulation}      
+\paragraph{From Liveness Sets to Context-Free Grammars:}\label{sec:grammar-formulation}      
 The  equations  above can  now  be  re-interpreted as  a  context-free
@@
 %\vspace*{-6.5mm}
-\section{The GC scheme}
+\section{The GC Scheme}
 \label{sec:GC-scheme}
@@
 this function.
-\section{Experimental evaluation}
+\section{Experimental Evaluation}
 \label{sec:experiments}
@@
 \vspace*{-2mm}
-\section{Related work} 
+\section{Related Work} 
 \label{sec:relatedwork}
@@
 \vskip -5mm
-\section{Future work and conclusions}
+\section{Future Work and Conclusions}
 \label{sec:conclusion}
@@
 \appendix
-\section{Soundness result for liveness analysis under minefield semantics}
+\section{Soundness Result for Liveness Analysis under Minefield Semantics}
 \label{sec:sound}

### 2016-05-06 paper/cfg-to-nfa.tex

@@
 If a CFG consists of a set of mutually recursive non-terminals such that the rules involved are not all left regular or not all right regular, then the method breaks the rules into right regular rules by introducing fresh non-terminals.
-For our example, the rule  $\var{\Df{\length}{1}}$ has two non regular productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
+For our example, the rule  $\var{\Df{\length}{1}}$ has two non-regular productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
 $\clazy\var{\Df{\length}{1}}{\clazy}$.  The steps for transforming these productions into

### 2020-01-15 eager-lgc/src/DemandStructure.h
updates to sync with asoffice machine

@@ -54,6 +54,8 @@ extern LivenessState PHI;
 extern LivenessState EPSILON;
 extern LivenessState ZERO;
 extern LivenessState ONE;
+extern LivenessState ZEROALL;
+extern LivenessState ONEALL;
 extern LivenessState ONE_STAR;
 extern LivenessState ALL;
 
@@ -64,7 +66,7 @@ bool operator==(const LivenessState l1, const LivenessState l2);
 class LivenessStateHasher
 {
 public:
-	std::size_t operator()(const LivenessState& obj) const { return std::hash<std::string>()(obj.idString); }
+	std::size_t operator()(const LivenessState& obj) const { return std::hash<int>()(obj.latticeNo); }
 };
 
 class LivenessTable;
@@ -72,6 +74,7 @@ bool operator==(const LivenessTable&, const LivenessTable&);
 bool operator!=(const LivenessTable&, const LivenessTable&);
 //Defines a type for liveness table of any variable
 typedef std::map< std::string, LivenessTable> LivenessInformation;
+typedef std::unordered_map<std::string, LivenessTable> LivenessInformation2;
 bool operator==(const LivenessInformation&, const LivenessInformation&);
 LivenessInformation mapLiveness(const LivenessTable&, const LivenessInformation&);
 class LivenessTable
@@ -100,18 +103,22 @@ public:
 
 
 typedef std::map< std::string, LivenessInformation> ProgramLiveness;
+typedef std::unordered_map<std::string, std::vector<std::string> > ArgumentsOrder;
+typedef std::unordered_map<std::string, LivenessInformation2> ProgramLiveness2;
 
 //Printing Demands
 std::ostream& operator<<(std::ostream& out, const LivenessState& l);
 std::ostream& operator<<(std::ostream& out, const LivenessTable& t);
 std::ostream& operator<<(std::ostream& out, const LivenessInformation& t);
 std::ostream& operator<<(std::ostream& out, const ProgramLiveness& t);
-
+bool printLivenessToFile(const ProgramLiveness&, const std::string&);
 
 void doUnion(LivenessInformation &, const LivenessInformation &);
 }
 }
 extern Scheme::Demands::ProgramLiveness progLiveness;
+extern Scheme::Demands::ProgramLiveness functionCallDemands;
+extern Scheme::Demands::ProgramLiveness2 progLiveness2;
 
 
 #endif

### 2020-01-15 eager-lgc/src/DemandStructure_diff.cpp

@@ -1,0 +1,439 @@
+#include <iostream>
+#include <fstream>
+#include <cassert>
+#include <iterator>
+#include <unordered_set>
+#include <unordered_map>
+#include <map>
+#include <string>
+#include <locale>
+#include <boost/graph/adjacency_list.hpp>
+#include <boost/graph/strong_components.hpp>
+#include <boost/algorithm/string.hpp>
+#include<boost/tokenizer.hpp>
+
+//#include <boost/filesystem.hpp>
+
+#include "DemandStructure.h"
+
+using namespace Scheme::Demands;
+
+// LivenessState Constructors
+Scheme::Demands::LivenessState::LivenessState(const std::string& s, const int& n): idString(s), latticeNo(n) {}
+Scheme::Demands::LivenessState::LivenessState(): latticeNo(0) {idString = "";}
+
+// Possible Liveness
+LivenessState Scheme::Demands::PHI = LivenessState("phi", 0);
+LivenessState Scheme::Demands::EPSILON = LivenessState("eps", 1);
+LivenessState Scheme::Demands::ZERO = LivenessState("zero", 2);
+LivenessState Scheme::Demands::ONE = LivenessState("one", 3);
+LivenessState Scheme::Demands::ZEROALL = LivenessState("zeroall", 4);
+LivenessState Scheme::Demands::ONEALL = LivenessState("oneall", 5);
+LivenessState Scheme::Demands::ALL = LivenessState("all", 6);
+
+bool Scheme::Demands::operator<(const LivenessState& s1, const LivenessState& s2)
+{
+  return s1.latticeNo < s2.latticeNo;
+}
+
+std::unordered_map<std::string, std::unordered_map<std::string, LivenessState> > getUnionRules()
+{
+    std::unordered_map<std::string, std::unordered_map<std::string, LivenessState> > m {{"phi", {{"phi", PHI}, 
+                                                                                                  {"eps", EPSILON},
+                                                                                                  {"zero", ZERO},
+                                                                                                  {"one", ONE},
+                                                                                                  {"zeroall", ZEROALL},
+                                                                                                  {"oneall", ONEALL},
+                                                                                                  {"all", ALL}}},
+                                                                                         {"eps", {{"phi", EPSILON}, 
+                                                                                                  {"eps", EPSILON},
+                                                                                                  {"zero", ZERO},
+                                                                                                  {"one", ONE},
+                                                                                                  {"zeroall", ZEROALL},
+                                                                                                  {"oneall", ONEALL},
[…]

### 2020-01-15 eager-lgc/src/DemandStructure_simple.cpp

@@ -35,44 +35,43 @@ bool Scheme::Demands::operator<(const LivenessState& s1, const LivenessState& s2
   return s1.latticeNo < s2.latticeNo;
 }
 
-//Can parse from text file
 std::unordered_map<std::string, std::unordered_map<std::string, LivenessState> > getUnionRules()
 {
     std::unordered_map<std::string, std::unordered_map<std::string, LivenessState> > m { {"phi", { {"phi", PHI}, 
                                                                                                                       {"eps", EPSILON},
-                                                                                                                      {"zero", ZERO},
-                                                                                                                      {"one", ONE},
-                                                                                                                      {"one*", ONE_STAR},
+                                                                                                                      // {"0", ZERO},
+                                                                                                                      // {"1", ONE},
+                                                                                                                      // {"1*", ONE_STAR},
                                                                                                                       {"all", ALL} }},
                                                                                                              {"eps", { {"phi", EPSILON}, 
                                                                                                                       {"eps", EPSILON},
-                                                                                                                      {"zero", ZERO},
-                                                                                                                      {"one", ONE},
-                                                                                                                      {"one*", ONE_STAR},
+                                                                                                                      // {"0", ZERO},
+                                                                                                                      // {"1", ONE},
+                                                                                                                      // {"1*", ONE_STAR},
                                                                                                                       {"all", ALL} }},
-                                                                                                             {"zero", { {"phi", ZERO}, 
-                                                                                                                      {"eps", ZERO},
-                                                                                                                      {"zero", ZERO},
-                                                                                                                      {"one", ALL},
-                                                                                                                      {"one*", ALL},
-                                                                                                                      {"all", ALL} }},
-                                                                                                             {"one", { {"phi", ONE}, 
-                                                                                                                      {"eps", ONE},
-                                                                                                                      {"zero", ALL},
-                                                                                                                      {"one", ONE},
-                                                                                                                      {"one*", ONE_STAR},
-                                                                                                                      {"all", ALL} }},
-                                                                                                             {"one*", { {"phi", ONE_STAR}, 
-                                                                                                                      {"eps", ONE_STAR},
-                                                                                                                      {"zero", ALL},
[…]

### 2020-01-15 eager-lgc/src/Output.cpp

@@ -7,94 +7,3 @@ std::ostream & Scheme::output::indentTill(std::ostream & stream, unsigned depth)
     while(depth--)  stream.put(' ');
     return stream;
 }
-
-void Scheme::output::dumpGrammar(std::ostream & out,
-                                 Scheme::Demands::regular_demand_grammar * rgram) {
-    out << std::left;
-    auto gram = rgram->first;
-    for(auto ipart = 0; ipart < rgram->second->size(); ++ipart) {
-        auto type = "NONE";
-        if(rgram->second->at(ipart).first == Scheme::Demands::LEFT)         type = "LEFT";
-        else if(rgram->second->at(ipart).first == Scheme::Demands::RIGHT)   type = "RIGHT";
-        else if(rgram->second->at(ipart).first == Scheme::Demands::CYCLE)   type = "CYCLE";
-
-        out << "Partition #" << ipart
-            << " | SIZE = " << rgram->second->at(ipart).second.size()
-            << " | TYPE = " << type << "\n";
-        for(auto & non_terminal : rgram->second->at(ipart).second) {
-            out << std::setw(16) << non_terminal << arrow;
-
-            if(!gram->at(non_terminal).size()) {
-                out << empty << "\n";
-                continue;
-            }
-
-            auto path_iter = gram->at(non_terminal).begin();
-
-            if(path_iter->size()) {
-                auto token_iter = path_iter->begin();
-                out << *token_iter;
-                while(++token_iter != path_iter->end())
-                    out << cdot << *token_iter;
-            } else
-                out << empty;
-
-            while(++path_iter != gram->at(non_terminal).end()) {
-                out << vpipe;
-                if(path_iter->size()) {
-                    auto token_iter = path_iter->begin();
-                    out << *token_iter;
-                    while(++token_iter != path_iter->end())
-                        out << cdot << *token_iter;
-                } else
-                    out << empty;
-            }
-
-            out << "\n";
-        }
-        out << "\n";
-    }
-}
-
-void Scheme::output::dumpGrammar(std::ostream & out, Scheme::Demands::demand_grammar * gram) {
-    out << std::left;
-    for(auto & non_terminal : *gram) {
-        out << std::setw(16) << non_terminal.first << arrow;
-
-        if(!non_terminal.second.size()) {
-            out << empty << std::endl;
-            continue;
-        }
-
-        auto path_iter = non_terminal.second.begin();
-
-        if(path_iter->size()) {
-            auto token_iter = path_iter->begin();
[…]

### 2020-01-15 eager-lgc/src/Output.h

@@ -8,16 +8,12 @@
 #include <iomanip>
 #include <iostream>
 
-#include "Demands.h"
-
-
 namespace Scheme {
 
     namespace output {
 
         struct options {
-            std::string cfg, approx_cfg, approx_nfa, reduced_nfa,
-                json_ast, plain_ast, scheme_ast,
+            std::string json_ast, plain_ast, scheme_ast,
                 json_slice, plain_slice, scheme_slice;
             int ast_anf, slice_anf, ast_label, slice_label, dump_all;
         };
@@ -34,9 +30,6 @@ namespace Scheme {
         enum output_t { SCHEME = 0, PLAIN = 1, JSON = 2 };
 
         std::ostream & indentTill(std::ostream &, unsigned = 0);
-        void dumpGrammar(std::ostream &, Scheme::Demands::demand_grammar *);
-        void dumpGrammar(std::ostream &, Scheme::Demands::expr_demand_grammars *);
-        void dumpGrammar(std::ostream &, Scheme::Demands::regular_demand_grammar *);
 
     }
 

### 2020-01-15 eager-lgc/src/SchemeAST.h

@@ -1,21 +1,19 @@
 #ifndef __SCHEME_AST_H__
 #define __SCHEME_AST_H__ 1
 
-//#include "Output.h"
+#include "Output.h"
 #include <vector>
-#include<map>
-#include<iostream>
+#include <map>
+#include <iostream>
 #include <boost/variant.hpp>
-#include "Demands.h"
+#include "DemandStructure.h"
 #include <functional>
 #include <cstring>
-#include<utility>
+#include <utility>
 using namespace std;
 namespace Scheme {
 
 namespace AST {
-
-
 
 struct StrCompare : public std::binary_function<const char*, const char*, bool> {
 public:
@@ -23,12 +21,10 @@ public:
     { return std::strcmp(str1, str2) < 0; }
 };
 class ExprNode;
-typedef std::list<ExprNode*> LivenessInfo;
-typedef std::pair<std::string, LivenessInfo > Liveness;
 
-// Added by Saksham
+
 typedef unordered_set<string> EdgeSet;
-// Added by Saksham
+extern unordered_map<string, EdgeSet> revCallGraph;
 
 enum exprType
 {
@@ -110,17 +106,17 @@ public:
 	virtual Node * clone() const = 0;
 	virtual Node * getANF() const = 0;
 
+	// Return type by reference??
+	virtual Scheme::Demands::LivenessInformation
+	transformDemand() const = 0;
 
-	virtual Scheme::Demands::expr_demand_grammars *
-	transformDemand(const Scheme::Demands::rule&) const = 0;
-
-	// virtual std::ostream & print(std::ostream &, unsigned = 0, bool = true, bool = false,
-	// 		Scheme::output::output_t format = Scheme::output::PLAIN) const = 0;
+	virtual std::ostream & print(std::ostream &, unsigned = 0, bool = true, bool = false,
+			Scheme::output::output_t format = Scheme::output::PLAIN) const = 0;
 	enum exprType type;
 	const std::string node_name;
 protected:
 	std::string label;
-
+	//LivenessInformation livenessInformation;
 
 	Node(const std::string);
 
@@ -139,56 +135,31 @@ public:
 	virtual ExprNode * clone() const = 0;
 	virtual ExprNode * getANF() const = 0;
 	virtual resultValue evaluate() = 0;
-	virtual LivenessInfo analyse(Liveness)
-	{
-		std::cout << "Method should be overriden in the derived class." <<std::endl;
-		exit(-1);
-	}
-
[…]

### 2020-01-15 eager-lgc/src/SchemeAST_ANFs.cpp

@@ -1,5 +1,5 @@
 #include "Utils.hpp"
-#include "SchemeAST_new.h"
+#include "SchemeAST.h"
 
 #include <sstream>
 #include <cassert>

### 2020-01-15 eager-lgc/src/SchemeAST_Defs.cpp

@@ -28,6 +28,7 @@ ProgramNode* pgm;
 unsigned long lbl_count = 0;
 map< string, vector<ExprNode*>> func_call_points;
 
+
 // Added by Saksham
 // REVERSED CALL GRAPH FUNCTION
 
@@ -40,19 +41,16 @@ unordered_map<string, EdgeSet> ProgramNode::makeRevCallGraph()
 		string funcName = i->getFuncName();
 		result[funcName] = {};
 	}
-
 	for (auto i: *pListDefines)
 	{
 		string funcName = i->getFuncName();
 
 		for (string j : i->makeCallGraph())
 		{
-			cout<<"Adding "<<funcName<<" to "<<j<<'\n';
 			result[j].insert(funcName);
 		}
 
 	}
-
 	return result;
 }
 
@@ -81,6 +79,7 @@ EdgeSet IfExprNode::dependentFunctions(EdgeSet& s)
 	pCond->dependentFunctions(s);
 	pThen->dependentFunctions(s);
 	pElse->dependentFunctions(s);
+	return s;
 }
 
 //For every other type of ExprNode
@@ -91,6 +90,7 @@ EdgeSet ExprNode::dependentFunctions(EdgeSet& s)
 
 // Added by Saksham
 
+//OLD CODE BEGINS HERE
 void print_arg_type(resType r)
 {
 		switch(r)
@@ -402,11 +402,11 @@ bool BoolConstExprNode::getVal()
 
 // /*  Definitions for IF node */
 
-// IfExprNode::~IfExprNode() {
-// 	delete pCond;
-// 	delete pThen;
-// 	delete pElse;
-// }
+IfExprNode::~IfExprNode() {
+	delete pCond;
+	delete pThen;
+	delete pElse;
+}
 
 IfExprNode::IfExprNode(ExprNode * cond, ExprNode * then_expr, ExprNode * else_expr)
 : ExprNode("IF"), pCond(cond), pThen(then_expr), pElse(else_expr) {}
@@ -502,6 +502,10 @@ LetExprNode * LetExprNode::clone() const {
 
 resultValue LetExprNode::evaluate()
 {
+	// if (pArg->getIDStr() == "_var181")
+	// {
+	// 	cout<<pExpr
+	// }
 	resultValue var_res = this->getVarExpr()->evaluate();
 
 	//cout << "Completed processing let expression " << endl;
@@ -637,7 +641,7 @@ resultValue UnaryPrimExprNode::evaluateCdrExpr()
 	break;
[…]

### 2020-01-15 eager-lgc/src/SchemeAST_Defs_new.cpp

@@ -29,7 +29,6 @@ unsigned long lbl_count = 0;
 map< string, vector<ExprNode*>> func_call_points;
 
 
-// Added by Saksham
 // REVERSED CALL GRAPH FUNCTION
 
 unordered_map<string, EdgeSet> ProgramNode::makeRevCallGraph()
@@ -45,14 +44,15 @@ unordered_map<string, EdgeSet> ProgramNode::makeRevCallGraph()
 	for (auto i: *pListDefines)
 	{
 		string funcName = i->getFuncName();
-
+		cout << "Processing Function " << funcName << '\n';
 		for (string j : i->makeCallGraph())
 		{
 			result[j].insert(funcName);
+			cout << "Found " << j << '\n';
 		}
 
 	}
-
+	cout << "DONE!\n";
 	return result;
 }
 
@@ -65,7 +65,12 @@ EdgeSet DefineNode::makeCallGraph()
 
 EdgeSet FuncExprNode::dependentFunctions(EdgeSet& s)
 {
-	s.insert(getFunction());
+	auto name = getFunction();
+	if (name == "check") {
+		cout << "got it back!" << '\n';
+	}
+	cout << name << "\n";
+	s.insert(name);
 	return s;
 }
 
@@ -961,6 +966,7 @@ FuncExprNode * FuncExprNode::clone() const {
 
 std::string FuncExprNode::getFunction()
 {
+	cout << "Processing function : " << this->pID->getName() << '\n';
 	return this->pID->getName();
 }
 

### 2020-01-15 eager-lgc/src/SchemeAST_Demands.cpp

@@ -1,34 +1,60 @@
 #include "SchemeAST.h"
+#include "DemandStructure.h"
+#include <iostream>
+#include <typeinfo>
+#include <set>
 
 using namespace Scheme::AST;
 using namespace Scheme::Demands;
 
+ProgramLiveness progLiveness;
+ProgramLiveness functionCallDemands;
+ArgumentsOrder argumentsOrder;
+ProgramLiveness2 progLiveness2;
+unordered_map<string, EdgeSet> Scheme::AST::revCallGraph;
 
-demand_grammar function_call_demands;
-demand_grammar gLivenessData; //Global map which stores the liveness at cons points and the next statement after a function call only.
-
+void Scheme::AST::printRevCallGraph()
+{
+    cout<<"\n\nPrinting the reverse call graph:-\n\n";
+    for (auto i : Scheme::AST::revCallGraph)
+    {
+        cout<<i.first<<": ";
+        for (auto j : i.second)
+        {
+            cout<<j<<", ";
+        }
+        cout<<'\n';
+    }
+    cout<<"\n\n";
+}
 
 /* D( (return x), s, T ) = x.s  */
-expr_demand_grammars * ReturnExprNode::transformDemand(const rule & demand) const
+LivenessInformation ReturnExprNode::transformDemand() const
 {
-	return pID->transformDemand(demand);
+    LivenessInformation returnValue;
+    LivenessTable l= LivenessTable(pID->getIDStr(), true);
+    returnValue[pID->getIDStr()] = l;
+    return returnValue;
 }
 
 
 /* ref( (id x), s, T ) = x.s  */
-expr_demand_grammars * IdExprNode::transformDemand(const rule & demand) const
+// Function calls?
+LivenessInformation IdExprNode::transformDemand() const
 {
 
-	return new expr_demand_grammars({ new demand_grammar({{ label, demand }}),
-        new demand_grammar({{ *pID, demand }})
-    });
+    LivenessInformation returnValue;
+    LivenessTable l= LivenessTable(getIDStr(), true);
+    returnValue[getIDStr()] = l;
+    return returnValue;
 }
 
 
 /* ref( k, s, T ) = {k.s}  */
-expr_demand_grammars * ConstExprNode::transformDemand(const rule & demand) const
+LivenessInformation ConstExprNode::transformDemand() const
 {
[…]

### 2020-01-15 eager-lgc/src/SchemeAST_Print.cpp

@@ -1,4 +1,4 @@
-#include "SchemeAST_new.h"
+#include "SchemeAST.h"
 
 #include <cassert>
 #include <algorithm>
@@ -386,7 +386,3 @@ std::ostream & ProgramNode::print(std::ostream & stream, unsigned depth, bool ne
     }
     return stream;
 }
-
-
-
-

### 2020-01-15 eager-lgc/src/SchemeAST_new.h

@@ -22,9 +22,7 @@ public:
     { return std::strcmp(str1, str2) < 0; }
 };
 class ExprNode;
-// Added by Saksham
 typedef unordered_set<string> EdgeSet;
-// Added by Saksham
 
 enum exprType
 {

### 2020-01-15 eager-lgc/src/SchemeDriver.cpp

@@ -17,27 +17,7 @@ Scheme::SchemeDriver::SchemeDriver()
 
 Scheme::SchemeDriver::~SchemeDriver()
 {
-	// if (program)
-	// 	delete(program);
-	// if (anf_program)
-	// 	delete(anf_program);
-	// if (program_grammars)
-	// {
-	// 	if (program_grammars->second)
-	// 		delete(program_grammars->second);
-	// 	if (program_grammars->first)
-	// 		delete(program_grammars->first);
- //    delete(program_grammars);
-	// }
-	// if (combined_grammar)
-	// 	delete(combined_grammar);
-	// if (approx_grammar)
-	// 	delete(approx_grammar);
 
-	// if(scanner)
-	// 	delete(scanner);
-	// if (parser)
-	// 	delete(parser);
 }
 
 std::pair<bool, long> Scheme::SchemeDriver::parse(const char * infilename)
@@ -80,22 +60,19 @@ std::pair<bool, long> Scheme::SchemeDriver::parse(const char * infilename)
     anf_program = program->getANF();
     anf_program->doLabel(true);
 
-    // Added by Saksham
+    // auto revCallGraph = anf_program->makeRevCallGraph();
 
-    auto revCallGraph = anf_program->makeRevCallGraph();
-
-    cout<<"\n\nPrinting the reverse call graph:-\n\n";
-    for (auto i : revCallGraph)
-    {
-        cout<<i.first<<": ";
-        for (auto j : i.second)
-        {
-            cout<<j<<", ";
-        }
-        cout<<'\n';
-    }
-    cout<<"\n\n";
-    // Added by Saksham
+    // cout<<"\n\nPrinting the reverse call graph:-\n\n";
+    // for (auto i : revCallGraph)
+    // {
+    //     cout<<i.first<<": ";
+    //     for (auto j : i.second)
+    //     {
+    //         cout<<j<<", ";
+    //     }
+    //     cout<<'\n';
+    // }
+    // cout<<"\n\n";
 
     string inputfilename(infilename);
     string inputfilepath = inputfilename.substr(0, inputfilename.find_last_of("/") + 1);
@@ -103,7 +80,7 @@ std::pair<bool, long> Scheme::SchemeDriver::parse(const char * infilename)
     inputfilename = inputfilepath + "anf_" + inputfilename;
     cout << "The anf file is being written to  " << inputfilename << endl;
     ofstream anf_file(inputfilename);
[…]

### 2020-01-15 eager-lgc/src/Simulator.cpp

@@ -22,7 +22,6 @@ using namespace Scheme::output;
 
 double gctime=0;
 int gcinvoke=0;
-extern demand_grammar gLivenessData;
 
 
 
@@ -66,44 +65,8 @@ Simulator::Simulator(int gctype)
 	//Do all global initialization here
 }
 
-void print_path (path prod_path, ostream& out)
-{
-
-	std::stringstream ss;
-	for_each(prod_path.begin(), prod_path.end(), [&ss] (const std::string& s) { ss << "." << s; });
-	if (ss.str().size() > 1)
-	{
-		std::string a = ss.str().substr(1); //Remove the first "." character
-		out << a;
-	}
-}
-
-void write_grammar_to_text_file(demand_grammar *g, string filename)
-{
-	std::ofstream gram_file(filename);
-	//Convert the set into a vector and sort it and print it to a file
-	for(auto l:*g)
-	{
-		vector<list<string> > vec(l.second.begin(), l.second.end());
-		sort(vec.begin(), vec.end(), StringListComparer);
-		int i = 0;
-		gram_file << l.first << "->";
-		for(auto p:vec)
-		{
-			print_path(p, gram_file);
-			++i;
-			if (i < vec.size())
-				gram_file << "|";
-		}
-		gram_file << endl;
-	}
-	gram_file.close();
-	return;
-}
-
 Simulator& Simulator::run(std::string pgmFilePath, int hsize) //This method or the constructor should take other parameters like gctype, heap size etc...
 {
-	//Do any per program initialization here.
 	Scheme::SchemeDriver driver;
 	cout << "Parsing " << pgmFilePath << endl;
 	std::pair<bool, long> parseResult = driver.parse(pgmFilePath.c_str());
@@ -115,108 +78,57 @@ Simulator& Simulator::run(std::string pgmFilePath, int hsize) //This method or t
 	pgm = driver.get_anf_prog();
 	std::string pgmname = pgmFilePath.substr(pgmFilePath.find_last_of("/") + 1,
 			                                 pgmFilePath.find_first_of(".", pgmFilePath.find_last_of("/")) - pgmFilePath.find_last_of("/") - 1);
-///////////////////////////////////////////////////////////////////////
-	//bool filesCached = true;
-	struct stat buffer;
[…]

### 2020-01-15 eager-lgc/src/gc.h

@@ -5,6 +5,8 @@
 #include<unordered_map>
 #include<set>
 #include<utility>
+#include "DemandStructure.h"
+
 using namespace std;
 
 
@@ -22,7 +24,7 @@ using namespace std;
  * to disable stat generation.
  * Author : Amey Karkare
  * */
-#define GC_ENABLE_STATS
+ #define GC_ENABLE_STATS
 #if 0
 #define DBG(stmt) stmt
 #else
@@ -32,12 +34,12 @@ using namespace std;
 enum GCStatus {gc_disable, gc_live, gc_freq, gc_plain};
 
 typedef int state_index;
+typedef int liv_index;
 typedef set<state_index> stateset;
 typedef stateset::iterator setiter;
 typedef unordered_map<string, state_index>::const_iterator stateMapIter;
 typedef unordered_map<string, state_index> stateMap;
 typedef unordered_map<std::string, std::string> call_next_data;
-
 
 
 typedef struct Var_Heap
@@ -62,6 +64,7 @@ typedef struct activationRecord
 	string funcname;
 	vector<var_heap> heapRefs;
 	vector<var_stack> nonRefs;
+  Scheme::Demands::LivenessState funcLiveness;
 }actRec;
 
 typedef unsigned long clock_tick;
@@ -160,6 +163,7 @@ typedef struct cell_cons
 //GC methods
  void reachability_gc();
  void initialize (string program_name, state_index, GCStatus);
+ void initialize2 (GCStatus gc_type);
  void cleanup(state_index);
  void liveness_gc();
  void dump_heap(string label);

### 2020-01-15 eager-lgc/src/gc_diff.cpp

@@ -17,7 +17,7 @@
 using namespace std;
 
 int numcopied = 0; //TODO:Delete after debugging
-int numLiveness = 6; //TODO:Delete after debugging
+int numLiveness = 7; //TODO:Delete after debugging
 
 
 extern GCStatus gc_status;
@@ -987,66 +987,6 @@ int copy_given_scan_children(void* node, int docar, int docdr)
 }
 
 typedef void* (*gcFunc)(void* );
-void* gcPHI(void* cellptr)
-{
-	assert(lt_scan_freept() != 1);
-	return cellptr;
-}
-void* gcEPS(void* cellptr)
-{
-	void* addr = copy(cellptr);
-	// copy_given_scan_children(addr, 1, 1);
-	while(lt_scan_freept()==1)
-		update_scan();
-	assert(lt_scan_freept() != 1);
-	return addr;
-}
-void* gcZERO(void* cellptr)
-{
-	void* addr = copy(cellptr);
-	copy_given_scan_children(addr, 1 , 0);
-	if(lt_scan_freept()==1)
-		update_scan();
-	if(lt_scan_freept()==1)
-		update_scan();
-	assert(lt_scan_freept() != 1);
-	return addr;
-}
-
-void* gcONE(void* cellptr)
-{
-	void* addr = copy(cellptr);
-	copy_given_scan_children(addr, 0, 1);
-	if(lt_scan_freept()==1)
-		update_scan();
-	if(lt_scan_freept()==1)
-		update_scan();
-	assert(lt_scan_freept() != 1);
-	return addr;
-}
-
-void* gcONES(void* cellptr)
-{
-	void* addr = copy(cellptr);
-	while(lt_scan_freept()==1) //will automatically break upon hitting a cell which has been copied under demand of (0+1)*
-	{
-		copy_given_scan_children(getscan(), 0, 1);
-		update_scan();
-	}
-	return addr;
-}
-
-void* gcALL(void* cellptr)
-{
-	void* addr = copy(cellptr);
-		while(lt_scan_freept()==1)
-		{
-			// cout<<"doing loop"<<endl;
-			copy_given_scan_children(getscan(), 1, 1);
-			update_scan();
-		}
-		return addr;
-}
 
 void print_cons(cons* a)
 {
@@ -1113,29 +1053,133 @@ void liveness_gc()
 
 			// auto curstate = progLiveness2["L/" + stackit->return_point][vhit->varname][Scheme::Demands::ALL];
 			// cout<<"curstate"<<curstate<<endl;
[…]

### 2020-01-15 eager-lgc/src/lexer.cpp

@@ -1,0 +1,2140 @@
+#line 2 "lexer.cpp"
+
+#line 4 "lexer.cpp"
+
+#define  YY_INT_ALIGNED short int
+
+/* A lexical scanner generated by flex */
+
+/* %not-for-header */
+/* %if-c-only */
+/* %if-not-reentrant */
+/* %endif */
+/* %endif */
+/* %ok-for-header */
+
+#define FLEX_SCANNER
+#define YY_FLEX_MAJOR_VERSION 2
+#define YY_FLEX_MINOR_VERSION 6
+#define YY_FLEX_SUBMINOR_VERSION 4
+#if YY_FLEX_SUBMINOR_VERSION > 0
+#define FLEX_BETA
+#endif
+
+/* %if-c++-only */
+    /* The c++ scanner is a mess. The FlexLexer.h header file relies on the
+     * following macro. This is required in order to pass the c++-multiple-scanners
+     * test in the regression suite. We get reports that it breaks inheritance.
+     * We will address this in a future release of flex, or omit the C++ scanner
+     * altogether.
+     */
+    #define yyFlexLexer yyFlexLexer
+/* %endif */
+
+/* %if-c-only */
+/* %endif */
+
+/* %if-c-only */
+/* %endif */
+
+/* First, we deal with  platform-specific or compiler-specific issues. */
+
+/* begin standard C headers. */
+/* %if-c-only */
+/* %endif */
+
+/* %if-tables-serialization */
+/* %endif */
+/* end standard C headers. */
+
+/* %if-c-or-c++ */
+/* flex integer type definitions */
+
+#ifndef FLEXINT_H
+#define FLEXINT_H
+
+/* C99 systems have <inttypes.h>. Non-C99 systems may or may not. */
+
+#if defined (__STDC_VERSION__) && __STDC_VERSION__ >= 199901L
+
+/* C99 says to define __STDC_LIMIT_MACROS before including stdint.h,
+ * if you want the limit (max/min) macros for int types. 
+ */
+#ifndef __STDC_LIMIT_MACROS
+#define __STDC_LIMIT_MACROS 1
+#endif
+
+#include <inttypes.h>
+typedef int8_t flex_int8_t;
+typedef uint8_t flex_uint8_t;
+typedef int16_t flex_int16_t;
+typedef uint16_t flex_uint16_t;
+typedef int32_t flex_int32_t;
+typedef uint32_t flex_uint32_t;
+#else
+typedef signed char flex_int8_t;
[…]

### 2020-01-15 eager-lgc/src/minMemory.py

@@ -1,0 +1,42 @@
+import sys
+import os
+import re
+import subprocess
+import json
+
+if len(sys.argv) != 3:
+	print("Usage: python3 minMemorySearch.py <path-to-exec> <path-to-bench>")
+	exit(-1)
+
+if not os.path.isfile(sys.argv[1]):
+	print("Path to executables invalid")
+	exit(-1)
+
+if not os.path.isfile(sys.argv[2]):
+	print("Path to benchmark directory invalid")
+	exit(-1)
+
+sim_lattice = sys.argv[1]
+benchmark = sys.argv[2]
+filename = os.path.basename(benchmark)
+for gcopt in ['gc-plain', 'gc-live']:
+	maxMem = 10000
+	minMem = 1
+	cmdlist = [sim_lattice, benchmark, str(maxMem), gcopt]
+	while re.search("Terminating", subprocess.run(cmdlist, stderr=subprocess.PIPE, stdout=subprocess.PIPE).stderr.decode('utf-8')):
+		minMem = maxMem
+		maxMem = maxMem * 2
+		cmdlist[2] = str(maxMem)
+		print(filename + ": Trying with " + str(maxMem))
+
+	while maxMem != minMem:
+		mid = (minMem + maxMem) // 2
+		cmdlist[2] = str(mid)
+		notWorking = re.search("Terminating", subprocess.run(cmdlist, stderr=subprocess.PIPE, stdout=subprocess.PIPE).stderr.decode('utf-8'))
+
+		if notWorking:
+			minMem = mid + 1
+		else:
+			maxMem = mid
+
+	print(filename, gcopt, str(minMem))

### 2020-01-15 eager-lgc/src/run_in_loop.py

@@ -5,7 +5,7 @@ import time
 import datetime
 from collections import namedtuple
 
-prog_size = {'testgc':143370}
+prog_size = {"append": 5, "gc_bench": 131071, "takl": 3, "testgc": 143355, "testsmall": 12, "paraffins": 2079, "lcss": 10801, "nperm": 93264, "sudoku": 1704, "test": 5, "fibheap": 254520, "knightstour": 423318, "deriv": 313, "fft": 768, "lambda": 7014, "nqueens": 17437}
 
 gc_info = namedtuple("gc_info", "heap_total heap_left heap_used gc_invocations gc_time exec_time")
 gc_stat = namedtuple("gc_stat", "prog_name reachability liveness")
@@ -16,7 +16,7 @@ path = os.getcwd()
 collector_path = os.path.join(path, "Simulator ")
 
 def run_program(filename, prog_dir_path, gc_type):
-    inputfile =  prog_dir_path + '/' + filename + '.scm'
+    inputfile =  prog_dir_path + '/' + filename + ''
     cmd = collector_path + " "  + inputfile + " " +  str(prog_size[filename]) + "   " + gc_type + " > " + prog_dir_path +'/results/' + filename + '-' + gc_type
     subprocess.call(cmd,shell=True)
 

### 2020-01-15 eager-lgc/src/scheme.tab.h

@@ -1,0 +1,520 @@
+// A Bison parser, made by GNU Bison 3.0.4.
+
+// Skeleton interface for Bison LALR(1) parsers in C++
+
+// Copyright (C) 2002-2015 Free Software Foundation, Inc.
+
+// This program is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+
+// This program is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+
+// You should have received a copy of the GNU General Public License
+// along with this program.  If not, see <http://www.gnu.org/licenses/>.
+
+// As a special exception, you may create a larger work that contains
+// part or all of the Bison parser skeleton and distribute that work
+// under terms of your choice, so long as that work isn't itself a
+// parser generator using the skeleton or a modified version thereof
+// as a parser skeleton.  Alternatively, if you modify or redistribute
+// the parser skeleton itself, you may (at your option) remove this
+// special exception, which will cause the skeleton and the resulting
+// Bison output files to be licensed under the GNU General Public
+// License without this special exception.
+
+// This special exception was added by the Free Software Foundation in
+// version 2.2 of Bison.
+
+/**
+ ** \file scheme.tab.h
+ ** Define the Scheme::parser class.
+ */
+
+// C++ LALR(1) parser skeleton written by Akim Demaille.
+
+#ifndef YY_YY_SCHEME_TAB_H_INCLUDED
+# define YY_YY_SCHEME_TAB_H_INCLUDED
+// //                    "%code requires" blocks.
+#line 9 "scheme.y" // lalr1.cc:377
+
+    #include "SchemeAST.h"
+
+    // Forward declare the Scanner and Driver classes, as the Parser needs them
+    // But they themselves cannot be declared without the Parser.
+    namespace Scheme {
+        class SchemeDriver;
+        class SchemeScanner;
+    }
+
[…]
